### 2013-12-18 image_analysis_paper/plos_draft.tex
Adding files

% Template for PLoS
% Version 1.0 January 2009
%
% To compile to pdf, run:
% latex plos.template
% bibtex plos.template
% latex plos.template
% latex plos.template
% dvipdf plos.template

\documentclass[10pt]{article}

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color}

% Use doublespacing - comment out for single spacing
%\usepackage{setspace}
%\doublespacing

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm
\textheight 21cm

% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

% Use the PLoS provided bibtex style
\bibliographystyle{plos2009}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

\pagestyle{myheadings}
%% ** EDIT HERE **

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW
\usepackage{subcaption}
\usepackage{float}
\usepackage{pgffor}
%% END MACROS SECTION

\begin{document}

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Image Analysis for {\em Drosophila} Embryogenesis}
}
% Insert Author names, affiliations and corresponding author email.
\\
Carmeline J. Dsilva$^{1}$,
Bomyi Lim$^{1}$,
Stanislav Y. Shvartsman$^{1,2}$
Ioannis G. Kevrekidis$^{1,3,\ast}$
\\
\bf{1} Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, USA
\\
\bf{2} Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, NJ, USA
\\
\bf{3} Program in Applied and Computational Mathematics, Princeton University, Princeton, NJ, USA
\\
$\ast$ E-mail: yannis@princeton.edu
\end{flushleft}

% Please keep the abstract between 250 and 300 words
\section*{Abstract}

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step.
% Author Summary not valid for PLoS ONE submissions.
\section*{Author Summary}

\section*{Introduction}

\begin{figure}[H]
\begin{center}
\begin{subfigure}{0.55\textwidth}
\includegraphics[width=0.8\textwidth]{drosophila_schematic}\\
\includegraphics[width=0.3\textwidth]{drosophila_membrane}
\includegraphics[width=0.3\textwidth]{drosophila_dpERK}
\includegraphics[width=0.3\textwidth]{drosophila_DI}
\caption{}
\end{subfigure}
\begin{subfigure}{0.35\textwidth}
\centering
\includegraphics[width=0.6\textwidth]{circle_profile}\\
\includegraphics[width=\textwidth, trim=0mm 300mm 0mm 200mm, clip]{line_profile}
\caption{}
\end{subfigure}\\
\begin{subfigure}{0.45\textwidth}
\includegraphics[width=\textwidth]{data_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.45\textwidth}
\includegraphics[width=\textwidth]{data_ordered_membrane}
\caption{}
\end{subfigure}
\end{center}
\caption{{\bf Data collection.} (a) Schematic of {\em Drosophila} embryo (longitudinal and cross-section), and cross-sectional fluorescent images of an embryo stained for the membrane proteins (left), dpERK (center), and dorsal protein (right).
(b) Concentration profile of dpERK extracted from a fluorescent image. The circular profile is ``unwrapped'' at the dorsal midline (determined from the expression of the dorsal protein) to obtain a concentration profile on a line.
(c) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time.
(d) Concentration profiles of dpERK for many embryos from (c), now ordered by the thickness of the membrane. The membrane thickness is known to be one-to-one with time.}
\label{fig:background}
\end{figure}

% Results and Discussion can be combined.
\section*{Results}

\subsection*{dpERK concentration profiles ordered using Principal Component Analysis}
\begin{figure}[H]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_PCA}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{PCA_time_corr}
\caption{}
\end{subfigure}
\caption{{\bf Ordering dpERK concentration profiles using PCA.} (a) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time.
(b) Concentration profiles of dpERK from (a), now ordered by the first PCA projection coefficient.
(c) Correlation between the first PCA projection coefficient and the membrane thickness.}
\label{fig:PCA_ordering}
\end{figure}

\subsection*{dpERK concentration profiles ordered using Diffusion Maps}
\begin{figure}[H]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_DMAPS}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{DMAPS_time_corr}
\caption{}
\end{subfigure}
\caption{{\bf Ordering dpERK concentration profiles using DMAPS.} (a) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time.
(b) Concentration profiles of dpERK from (a), now ordered by the first non-trivial DMAPS embedding coordinate.
(c) Correlation between the first (non-trivial) DMAPS embedding coordinate and the membrane thickness.}
\label{fig:DMAPS_ordering}
\end{figure}

\subsection*{dpERK concentration profiles aligned using angular synchronization}

\begin{figure}[H]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_unaligned_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_aligned_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_angsynch}
\caption{}
\end{subfigure}
\caption{{\bf Ordering spatially unaligned dpERK concentration profiles.} (a) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time. The circluar profiles have not been ``opened'' at the same point; rather, each profile was opened at a random point around the embryo. We are allowed to shift the (linear) profiles left and right.
(b) Concentration profiles of dpERK aligned using angular synchronization.
(c) Concentration profiles of dpERK aligned using angular synchronization and ordered by the first (non-trivial) DMAPS embedding coordinate.}
\label{fig:angsynch_ordering}
\end{figure}

\subsection*{dpERK concentration profiles aligned and ordered using vector diffusion maps}

\begin{figure}[H]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_unaligned_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_vdm}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{VDM_time_corr}
\caption{}
\end{subfigure}
\caption{{\bf Ordering spatially unaligned dpERK concentration profiles using vector diffusion maps.} (a) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time. The circluar profiles have not been ``opened'' at the same point; rather, each profile was opened at a random point around the embryo. We are allowed to shift the (linear) profiles left and right.
(b) Concentration profiles of dpERK aligned and ordered using vector diffusion maps.
(c) Correlation between the VDM embedding coordinate and the membrane thickness.}
\label{fig:angsynch_ordering}
\end{figure}

\subsection*{Two-dimensional dpERK images aligned and ordered using vector diffusion maps}

\begin{figure}[H]
\centering
\begin{subfigure}{0.45\textwidth}
\includegraphics[width=\textwidth]{vdm_2d_time_corr}
\caption{}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\foreach \n in {1,9,17,25,30,34,38,43,49}{
\includegraphics[width=0.3\textwidth]{dpERK_vdm_\n}
\hfill}
\caption{}
\end{subfigure}
\caption{{\bf Ordering two-dimensional fluorescent images of dpERK.}
(a) Correlation between the VDM embedding coordinate and the membrane thickness. 
(b) Some representative fluorescent images, aligned  and ordered using the VDM embedding coordinate. }
\label{fig:scattrans_dpERK_ordering}
\end{figure}

\subsection*{Two-dimensional dpERK images ordered using diffusion maps and the scattering transform}

\begin{figure}[H]
\centering
\begin{subfigure}{0.45\textwidth}
\includegraphics[width=\textwidth]{DMAPS_scat_time_corr}
\caption{}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\foreach \n in {1,9,17,25,30,34,38,43,49}{
\includegraphics[width=0.3\textwidth]{dpERK_scat_\n}
\hfill}
\caption{}
\end{subfigure}
\caption{{\bf Ordering two-dimensional fluorescent images of dpERK.}
(a) Correlation between the first (non-trivial) DMAPS embedding coordinate and the membrane thickness. The DMAPS embedding was computed using the translation- and rotation-invariant scattering transform coefficients as ``features'' of the images.
(b) Some representative fluorescent images, ordered using the first (non-trivial) DMAPS embedding coordinate computed from the scattering transform coefficients. }
\label{fig:scattrans_dpERK_ordering}
\end{figure}

\subsection*{Two-dimensional membrane images ordered using diffusion maps and the scattering transform}

\begin{figure}[H]
\centering
\begin{subfigure}{0.45\textwidth}
\includegraphics[width=\textwidth]{DMAPS_membrane_scat_time_corr}
\caption{}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\foreach \n in {1,9,17,25,30,34,38,43,49}{
\includegraphics[width=0.3\textwidth]{membrane_scat_\n}
\hfill}
\caption{}
\end{subfigure}
\caption{{\bf Ordering two-dimensional fluorescent images of membrane proteins.}
(a) Correlation between the seventh (non-trivial) DMAPS embedding coordinate and the membrane thickness. DMAPS was computed using the translation- and rotation-invariant scattering transform coefficients as ``features'' of the images. The first six non-trivial DMAPS coordinates were found to capture other features of the images, such as  differences in fluorescence intensity within the images.
(b) Some representative fluorescent images, ordered using the seventh (non-trivial) DMAPS embedding coordinate computed from the scattering transform coefficients. }
\label{fig:scattrans_membrane_ordering}
\end{figure}

\section*{Discussion}

% You may title this section "Methods" or "Models".
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis"
\section*{Materials and Methods}

\subsection*{Experimental Setup}

\subsection*{Principal Component Analysis}

\subsection*{Diffusion Maps}

\subsection*{Angular Synchronization in $SO(2)$}

\subsection*{Scattering Transform}

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}

%\section*{References}
% The bibtex filename
\bibliography{template}

\section*{Figure Legends}
%\begin{figure}[!ht]
%\begin{center}
%%\includegraphics[width=4in]{figure_name.2.eps}
%\end{center}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure 2  caption.  Caption
%should be left justified, as specified by the options to the caption
%package.
%}
%\label{Figure_label}
%\end{figure}

\section*{Tables}
%\begin{table}[!ht]
%\caption{
%\bf{Table title}}
%\begin{tabular}{|c|c|c|}
%table information
%\end{tabular}
%\begin{flushleft}Table caption
%\end{flushleft}
%\label{tab:label}
% \end{table}

\end{document}

### 2013-12-18 image_analysis_paper/make_1d_alignment_figures.m

@@ -1,0 +1,142 @@
+addpath ../membrane_pictures/synchron_SO2
+
+dpERK_unaligned = dpERK;
+
+%% scramble data alignments
+rng(12345);
+
+rand_offsets = zeros(m,1);
+for i=1:m
+    rand_offsets(i) = randi(n);
+    dpERK_unaligned(i,:) = circshift(dpERK_unaligned(i,:),[0 rand_offsets(i)]);
+end
+
+%% plot scrambled data
+figure;
+imagesc(dpERK_unaligned)
+xlabel('position (unaligned)')
+ylabel('data point (unordered)')
+set(gca, 'xtick', [])
+set(gca, 'ytick', [])
+if print_figures
+    print('data_unaligned_unordered',fmt, res)
+end
+
+%% calculate pairwise alignments
+[H, W, theta] = align_data(dpERK_unaligned);
+
+%% angular synchronization
+R_opt = ang_synch(H, 2);
+
+dpERK_aligned = zeros(size(dpERK_unaligned));
+R_est = R_opt(1:2, 1:2);
+theta_est = atan2(R_est(2,1),R_est(1,1));
+shift_est0 = round(theta_est/(2*pi)*n) + rand_offsets(1);
+for i=1:m
+    R_est = R_opt(2*i-1:2*i, 1:2);
+    theta_est = atan2(R_est(2,1),R_est(1,1));
+    shift_est = round(theta_est/(2*pi)*n) - shift_est0;
+    dpERK_aligned(i,:) = circshift(dpERK_unaligned(i,:),[0 shift_est]);
+end
+
+%% plot alignment with angular synchronization
+figure;
+imagesc(dpERK_aligned)
+xlabel('position (aligned using angular synchronization)')
+ylabel('data point (unordered)')
+set(gca, 'xtick', [])
+set(gca, 'ytick', [])
+if print_figures
+    print('data_aligned_unordered',fmt, res)
+end
+
+%% dmaps following angular synchronization
+W = squareform(pdist(dpERK_aligned)).^2;
+eps_dmaps = median(W(:));
+[V_dmaps, D_dmaps] = dmaps(W, eps_dmaps, 10);
+
+if corr(V_dmaps(:,2), L(:,1)) < 0
+    V_dmaps(:,2) = -V_dmaps(:,2);
+end
+
+%% plot dmaps ordering
+[~, I] = sort(V_dmaps(:,2));
+figure;
+imagesc(dpERK_aligned(I,:))
[…]

### 2013-12-18 image_analysis_paper/make_2d_membrane_scattering_figures.m

@@ -1,0 +1,109 @@
+
+% pictures are in image channel 2
+image_channel = 2;
+% number of pixels (subsample images)
+npixels = 100;
+
+%set image plotting parameters
+subplot_dim1 = ceil(sqrt(m));
+subplot_dim2 = ceil(m / subplot_dim1);
+
+% store images in 3d array
+image_set = zeros(npixels, npixels, m);
+
+% load in images
+figure;
+for i=1:m
+    % read image
+    im1 = imread(sprintf('%s/emb%02d.tif', dpERK_membrane_dir, i));
+
+    % resize image
+    im1 = imresize(im1, [npixels npixels]);
+
+    % extract relevent color from image
+    im1 = im1(:,:,image_channel);
+
+    % plot image
+    subplot(subplot_dim1,subplot_dim2,i)
+    imshow(im1)
+
+    %store image
+    im1 = double(im1);
+    image_set(:,:,i) = im1;
+end
+
+% set scattering transform parameters
+filt_opt = struct();
+filt_rot_opt = struct();
+% oversampling is set to infty
+% so that scattering of original and rotated
+% image will be sampled at exactly the same points
+scat_opt = struct();
+scat_opt.oversampling = 10;
+
+% compute scattering transform of first image
+x = image_set(:,:,1);
+% define wavelet transforms
+Wop = wavelet_factory_3d(size(x), filt_opt, filt_rot_opt, scat_opt);
+Sx = scat(x, Wop);
+Sx_mat = format_scat(Sx);
+
+% store scattering invariants (one row per image)
+sx_all = zeros(m, size(Sx_mat, 1));
+
+% compute scattering invariants for each image
+for i=1:m
+    x = image_set(:,:,i);
+    
+    Sx = scat(x, Wop);
+
+    sx_all(i,:) = mean(mean(format_scat(Sx),2),3)';
+end
+
+% dmaps
+
+W = squareform(pdist(sx_all)).^2;
+eps = median(W(:));
+[V, D] = dmaps(W, eps, 20);
+
+for i=3:20;
+    figure;
+    plot(V(:,2),V(:,i),'.')
+end
+
+for i=2:20
+    corr(V(:,i),L(:,1))
+end
+
+idx = 8;
+
+if corr(V(:,idx), L(:,1)) < 0
+    V(:,idx) = -V(:,idx);
+end
+
+figure;
+plot(L(:,1), V(:,idx),'.')
+xlabel('membrane thickness')
[…]

### 2013-12-18 image_analysis_paper/make_figures.m

@@ -1,0 +1,140 @@
+clear all
+close all
+
+%% set parameters for saving figures
+set(0,'DefaultLineMarkerSize',14)
+set(0,'DefaultAxesFontSize',20)
+
+res = '-r300';
+fmt = '-djpeg';
+print_figures = true;
+
+dpERK_data = '../membrane_lengths/oct16.mat';
+dpERK_image_dir = '../membrane_pictures/membrane2/dpERK_staining';
+dpERK_membrane_dir = '../membrane_pictures/membrane2';
+
+%% load data
+load(dpERK_data);
+dpERK = dpERK_raw;
+
+[m, n] = size(dpERK);
+
+%% plot images
+
+image_idx = 44;
+
+im1 = imread(sprintf('%s/emb%02d.tif',dpERK_image_dir, image_idx));
+
+im1_dpERK = im1;
+im1_dpERK(:,:,1) = 0;
+im1_dpERK(:,:,3) = 0;
+
+im1_DI = im1;
+im1_DI(:,:,1) = im1_DI(:,:,3);
+im1_DI(:,:,2) = 0;
+
+im1_membrane = imread(sprintf('%s/emb%02d.tif',dpERK_membrane_dir, image_idx));
+
+figure;
+imshow(im1_dpERK)
+if print_figures
+    print('drosophila_dpERK', fmt, res)
+end
+
+figure;
+imshow(im1_DI)
+if print_figures
+    print('drosophila_DI', fmt, res)
+end
+
+figure;
+imshow(im1_membrane)
+if print_figures
+    print('drosophila_membrane', fmt, res)
+end
+
+%% plot circle and line profiles
+
+r = 1;
+theta = linspace(0, 2*pi, n+1);
+theta = theta(1:end-1);
+theta = theta - pi/4;
+
+x = r * cos(theta);
+y = r * sin(theta);
+
+figure;
+scatter(x, y, 2000, dpERK(image_idx,:), '.')
+hold on
+delta=0.1;
+plot([x(1)-delta x(1)+delta], [y(1)+delta y(1)-delta], 'linewidth',6, 'color','k')
+text(x(1)+delta+0.05, y(1)-delta-0.05, 'Open circle')
+set(gca,'xtick',[])
+set(gca,'ytick',[])
+axis equal
+set(gca, 'visible', 'off')
+if print_figures
+    print('circle_profile', fmt, res)
+end
+
+figure;
[…]

### 2013-12-18 group_meeting_5Feb2013/presentation.tex
Added files

\documentclass{beamer}
\usetheme{Frankfurt}
\usepackage{tikz}

\AtBeginSection[]
{
  \begin{frame}
    \frametitle{Table of Contents}
    \tableofcontents[currentsection]
  \end{frame}
}

\title{Dimensionality Reduction for \\Developmental Biology}
%\subtitle[]{Estimation of numerical errors}
\author[C. Dsilva]{Carmeline Dsilva}
\institute[PU]{
  Department of Chemical and Biological Engineering\\
  Princeton University
  Princeton, NJ 08544\\[1ex]
  \texttt{cdsilva@princeton.edu}
}
\date[February 2013]{5 February 2013}

\begin{document}

\begin{frame}[plain]
  \titlepage
\end{frame}

\section{Motivation}

\begin{frame}{Motivation}
    \begin{itemize}
        \item Time series data is prevalent in many applications
        \item It can be difficult to determine the times in an experiment
        \item We would like an ``automated'' way to order data in time
    \end{itemize}

    \centering
    \includegraphics[width=0.4\textwidth]{data1_scrambled.jpg}
    \hspace{0.1in}
    \includegraphics[width=0.4\textwidth]{data1_unscrambled.jpg}

    Above are data from one of Bomyi's experiments.
    Each row is the concentration profile of a morphogen in a fly.
    On the left, the data are scrambled (since each fly is at a different time).
    On the right, the data are ordered using principal component analysis (PCA) to get an idea of the temporal evolution of the morphogen.

    \begin{tikzpicture}[overlay]
    \draw [->] (-0.5,5)--(0.5,5);
    \end{tikzpicture}

\end{frame}

\begin{frame}{Example}

    \begin{columns}[T]
        \begin{column}{0.3 \textwidth}
            \includegraphics[width=\textwidth]{example1_all.jpg}

            {\Tiny This data is clearly ordered in time \par}
        \end{column}

        \begin{column}{0.3 \textwidth}
            \includegraphics[width=\textwidth]{example1_coarse.jpg}

            {\Tiny We can look at a ``coarsened'' representation of the data with only three spatial components \par}
        \end{column}

        \begin{column}{0.3 \textwidth}
            \includegraphics[width=\textwidth]{example1_embedding.jpg}

            {\Tiny If we plot the three-dimensional data and color them in time, we see they fall on a line and are correlated along this line in time \par}
        \end{column}
    \end{columns}
    \vspace{0.1 in}

    \centering
   If we embedded the original data, we would see the same linear trend, but the data would be embedded in a very high-dimensional space and we could not visualize it.

   We would like an automatic way to find this line in higher dimensions, so we can then order our data appropriately.

\end{frame}

\section{Principal Component Analysis}

\begin{frame}{Principal Component Analysis (PCA)}
  \begin{itemize}
  \item The goal of PCA \footnote{\tiny Shlens, J. (2005) A Tutorial on Principal Component Analysis.} is to find the direction(s) of maximum variance in a data set
  \item In our case, we are assuming that there is one direction with significantly more variance than the others
  \item We assume that if we project our data along this line, we will get an embedding that is correlated with time
  \end{itemize}
\end{frame}

\begin{frame}{PCA: Covariance Matrix}
  \begin{itemize}
  \item We have $n$ {\bf data points}, each in $d$ dimensions, denoted $x_1, x_2, \dots, x_n \in \mathbb{R}^d$, where $x_{i,j}$ denotes the $j^{th}$ component of data point $i$
  \item We first define the {\bf mean-centered} data points $\hat{x}_i$ as $$\hat{x}_{i,j} = x_{i,j} - \frac{1}{n} \sum_{i=1}^{n} x_{i,j}$$
  \item Let $X \in \mathbb{R}^{n \times d}$ denote the data matrix, where the $i^{th}$ row of $X$ contains $\hat{x}_i$
  \item The {\bf covariance matrix} $R \in \mathbb{R}^{d \times d}$ can be estimated as $$R = \frac{1}{n-1} X^T X$$
  \end{itemize}
\end{frame}

\begin{frame}{PCA: Eigendecomposition}
\begin{columns}
\begin{column}{0.7\textwidth}
  \begin{itemize}
    \item We then compute the {\bf eigenvectors} $v_1, v_2, \dots, v_d$ and {\bf eigenvalues} $\lambda_1, \lambda_2, \dots, \lambda_d$ of $R$ (all eigenvectors are normalized to 1)
    \item By construction, $R$ is symmetric and positive semi-definite and is therefore guaranteed to have real non-negative eigenvalues and orthogonal real eigenvectors
    \item We {\bf order} the eigenvector/eigenvalue pairs such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_d$
    \item $v_1, v_2, \dots, v_d$ are called the {\bf principal components}, and $\lambda_j$ measures the {\bf variance} captured by principal component $j$
  \end{itemize}
  \end{column}

  \begin{column}{0.3\textwidth}
        \includegraphics[width=\textwidth]{PCA_spectrum1.jpg}\\
        {\tiny The PCA eigenvalue spectrum; each eigenvalue measures the energy captured by the corresponding eigenvector \par}

        \vspace{0.3 in}

        \includegraphics[width=\textwidth]{PCA_mode1.jpg}\\
        {\tiny The first principal component \par}

  \end{column}
  \end{columns}
\end{frame}

\begin{frame}{PCA: Ordering and Projections}
    \begin{itemize}
        \item We can calculate the projection of data point $i$ onto principal component $j$ as $a_{i,j} = \langle \hat{x}_i, v_j \rangle = \sum_{k=1}^{d} \hat{x}_{i,k} v_{j,k}$
        \item $a_{i,j}$ can be viewed as a measure of how much principal component $j$ is represented in data point $i$
        \item We can order the data by $a_{i,1}$
        \item Each principal component defines an important ``direction'' in our data; we can compare data sets by comparing their principal components
    \end{itemize}

    \centering
    \includegraphics[height=0.3\textheight]{data1_scrambled.jpg}
    \includegraphics[height=0.3\textheight]{coeff_scrambled.jpg}\\
    {\Tiny The scrambled data (left) and the projection coefficient for each data point onto the first principal component (right).\\
    We will sort the data by the values of the projection coefficients. \par}

\end{frame}

\begin{frame}{Results}
  \centering
    Below are three replicate data sets.\\
    You can see the scrambled data (left), unscrambled data using PCA (middle), and first PCA mode (right)

    \includegraphics[width=0.3\textheight]{data1_scrambled.jpg}
    \includegraphics[width=0.3\textheight]{data1_unscrambled.jpg}
    \includegraphics[width=0.3\textheight]{PCA_mode1.jpg}\\
    \includegraphics[width=0.3\textheight]{data2_scrambled.jpg}
    \includegraphics[width=0.3\textheight]{data2_unscrambled.jpg}
    \includegraphics[width=0.3\textheight]{PCA_mode2.jpg}\\
    \includegraphics[width=0.3\textheight]{data3_scrambled.jpg}
    \includegraphics[width=0.3\textheight]{data3_unscrambled.jpg}
    \includegraphics[width=0.3\textheight]{PCA_mode3.jpg}

\end{frame}

\section{Diffusion Maps}

\begin{frame}{Some Problems with PCA}
    \centering
    \includegraphics[width=0.5\textwidth]{data1_unscrambled.jpg}
    \includegraphics[width=0.5\textwidth]{data_unscrambled_dmaps.jpg}\\
    The ordering of the large central peaks is comparable.\\
    However, the outer peaks are grouped together in the right picture (we could view this as ``better'').

\end{frame}

\begin{frame}{Nonlinear Orderings}
    \centering
    \includegraphics[width=0.75\textwidth]{coeff_12.jpg}

    The first two projection coefficients for the data (each point corresponds to a data point/concentration profile). \\
    We only use the first coefficient in the PCA ordering.

\end{frame}

\begin{frame}{Nonlinear Orderings}
    \centering
    \includegraphics[width=0.75\textwidth]{coeff_12_colored.jpg}

    The ordering of the data produced with this coloring would perhaps be better. \\
    PCA cannot detect this ordering because it is nonlinear.

\end{frame}

\begin{frame}{Diffusion Maps (DMAPS)}
\centering
Unlike PCA, diffusion maps \footnote{ \tiny Coifman {\em et al}, PNAS, 2005.} is a {\em nonlinear} dimensionality reduction technique

    \includegraphics[width=0.5\textwidth]{example1_embedding.jpg}
    \includegraphics[width=0.5\textwidth]{dmaps_spiral.jpg}

On the left you see data that has been ordered and colored using PCA.
On the right you see data that has been ordered and colored using DMAPS.
PCA would not be able to order the data on the right ``correctly'', since the curve is one-dimensional, but non-linear.

\end{frame}

\begin{frame}{DMAPS: Compute Distances}
\begin{itemize}
    \item Again we have data $x_1, x_2, \dots, x_n \in \mathbb{R}^d$

    \item We need a {\bf distance metric} $d(x_i, x_j)$ for our data; it is typical to use the Euclidean distance, although other metrics can be used
    \item We first construct the matrix $W \in \mathbb{R}^{n \times n}$, with
    $$W_{ij} = \exp \left( -\frac{d^2 (x_i, x_j)}{\epsilon} \right) $$
    where $\epsilon$ is a characteristic distance

    \item We define the diagonal matrix $D$ with $D_{ii} = \sum_{j=1}^{n} W_{ij}$
\end{itemize}
\end{frame}

\begin{frame}{DMAPS: Eigendecomposition}
\begin{itemize}
    \item We compute the {\bf eigenvectors} $v_1, v_2, \dots, v_n$ and {\bf eigenvalues} $\lambda_1, \lambda_2, \dots, \lambda_n$ of $D^{-1} W$.

    \item We {\bf order} the eigenvector/eigenvalue pairs such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_n|$

    \item Since $D^{-1}W \sim D^{-1/2} \left( D^{-1/2} W D^{-1/2} \right) D^{1/2}$, and $D^{-1/2} W D^{-1/2}$ is symmetric, $D^{-1}W$ is guaranteed to have real eigenvalues and eigenvectors.

    \item $D^{-1}W$ is row-stochastic, and so $\lambda_1 = 1$ and $v_1 =  \vec{1}$

    \item $|\lambda_i|$ provides a measure of the ``importance'' of eigenvector $i$.

    \item Based on the eigenvalues $\lambda_i$, we choose to use $m < n$ eigenvectors and define a {\bf diffusion map} $x_i \mapsto \left( v_2(i), v_3(i), \dots, v_m(i) \right)$

    \item In the time-series case, we assume there is only one important mode; our map is $x_i \mapsto v_2(i)$, and we order our data by $v_2(i)$

\end{itemize}
\end{frame}

\begin{frame}{Results}
You can see the scrambled data (left), unscrambled data using PCA (middle), and unscrambled data using DMAPS (right)
    \centering
    \includegraphics[width=0.3\textheight]{data1_scrambled.jpg}
    \includegraphics[width=0.3\textheight]{data1_unscrambled.jpg}
    \includegraphics[width=0.3\textheight]{data1_unscrambled_dmaps.jpg}\\
    \includegraphics[width=0.3\textheight]{data2_scrambled.jpg}
    \includegraphics[width=0.3\textheight]{data2_unscrambled.jpg}
    \includegraphics[width=0.3\textheight]{data2_unscrambled_dmaps.jpg}\\
    \includegraphics[width=0.3\textheight]{data3_scrambled.jpg}
    \includegraphics[width=0.3\textheight]{data3_unscrambled.jpg}
    \includegraphics[width=0.3\textheight]{data3_unscrambled_dmaps.jpg}

%With the first data set, there is a significantly better grouping of the secondary peaks when using DMAPS.
%In the second and third data sets, there is no significant advantage of using DMAPS versus PCA.
\end{frame}

\section{Theory of Diffusion Maps}

\begin{frame}{The Idea Behind DMAPS}
\begin{itemize}
\item We have data that lies on a curve, or more generally, a manifold
\item We assume that the manifold is smooth enough so that locally, it is linear and we trust the Euclidean distances
\item We want a method to compute ``good'' distances between all pairs of data points
\end{itemize}
\end{frame}

\begin{frame}{A Random Walk on the Data}
\begin{itemize}
\item Given data points $x_i$ and $x_j$, what would be a good distance measure between them?
\item We can imagine
\begin{itemize}
\item starting a swarm of random walkers at $x_i$
\item starting a swarm of random walkers at $x_j$
\item letting the random walkers ``hop'' from data point to data point, but they can only hop to ``nearby'' data points
\item after some time, we compare where the random walkers starting from $x_i$ are, and where the random walkers starting from $x_j$ are
\item if $x_i$ and $x_j$ are ``close'', then the two groups of random walkers should be in similar places.
\end{itemize}
\item We call this distance the {\bf diffusion distance}
\item We can mathematically formalize such a thought experiment
\end{itemize}
\end{frame}

\begin{frame}{Random Walker: The Math}
\begin{itemize}
\item We first construct a {\bf graph with weighted edges} on which our random walkers can ``hop''
\begin{itemize}
\item Each node of our graph is a data point
\item Each pair of nodes is connected by an edge with weight $W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon} \right)$
\item High-weight edges have a high probability of being traversed by a random walker
\item There is a large probability to hop to points $< \epsilon$ away and a small probability to hop to points $> \epsilon$ away
\end{itemize}
\item We then place these edge weights into a matrix $W \in \mathbb{R}^{n \times n}$
\item We define the {\bf probability transition matrix} $A$ for a random walker on our graph by normalizing the rows of $W$ (i.e. $A = D^{-1}W$)
\item If $\pi$ is an initial distribution of random walkers on the data, then $\pi A$ is the distribution of random walkers after one timestep, and $\pi A^t$ is the distribution of random walkers after $t$ timesteps
\end{itemize}
\end{frame}

\begin{frame}{Random Walker: Calculating the Diffusion Distance}
\begin{itemize}
\item let $\pi_i$ be the vector with $\pi_i(i) = 1$ and $\pi_i(k \ne i) = 0$
\item So $\pi_i A^t$ is the {\bf probability distribution of random walkers} at time $t$ if we start a swarm of random walkers at $x_i$
\item However, $\pi_i A^t$ is just the $i^{th}$ row of $A^t$, denoted $A_{i,\cdot}^t$
\item To compare random walkers starting at $x_i$ and $x_j$, we need to {\bf compare rows} $i$ and $j$ of $A^t$
\item One can show that the diffusion distance is given by
$$ \| A_{i,\cdot}^t - A_{j,\cdot}^t \|^2 =\sum_{k=1}^{n} \left( \frac{A_{i,k} - A_{j,k}}{D_{kk}} \right)^2 =  \sum_{k=1}^{n} \lambda_k^{2t} \left( v_k(i) - v_k(j) \right)^2 $$
(we weight the norm by the degrees because higher degree nodes should ``accumulate'' more probability)
\item Therefore, embedding the data points using the eigenvectors $v_2, v_3, \dots, v_n$ {\bf preserves the diffusion distance }
\item The eigenvalues measure how much each eigenvector contributes to the diffusion distance
\end{itemize}
\end{frame}

\section{Conclusions}
\begin{frame}{Other applications of PCA and DMAPS}
\begin{itemize}
\item Although we have only been discussing one-dimensional data, both PCA and DMAPS are more generally applicable to cases when there are several important dimensions
\begin{itemize}
\item PCA can find lines, planes, and hyperplanes
\item DMAPS can find curves and general manifolds
\end{itemize}
\item Both methods use eigenvalues as a measure of importance for each component, and so the intrinsic dimensionality of the data set is selected by looking for a ``spectral gap'', i.e. an $m$ such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_m \gg \lambda_{m+1} \ge \dots \ge \lambda_n$
\end{itemize}
\end{frame}

\begin{frame}{Compare/Contrast/Summarize}
\begin{itemize}
\item PCA and DMAPS can find reduced representations of data
\item With time series data, PCA/DMAPS can find the one ``important'' dimension that is correlated with time
\item Both methods require a distance measure for the data
\item Both methods are based on eigendecomposition of a matrix
\begin{itemize}
\item PCA uses a $d \times d$ matrix, where $d$ is the dimension of the ambient space
\item DMAPS uses an $n \times n$ matrix, where $n$ is the number of data points; therefore, DMAPS becomes computationally prohibitive for large data sets
\end{itemize}
\item PCA provides principal components, which are the important direction(s) in the data; DMAPS provides no analog to principal components, only an analog to the projection coefficients $a_{i,j}$
\item DMAPS is nonlinear, and so can often uncover/capture more complex patterns in data than PCA can
\end{itemize}
\end{frame}

\end{document}

### 2013-12-18 group_meeting_5Feb2013/dpERK_example.m

@@ -1,0 +1,150 @@
+clear all
+close all
+
+%% set parameters for saving figures
+set(0,'DefaultLineMarkerSize',14)
+set(0,'DefaultAxesFontSize',20)
+
+res = '-r300';
+fmt = '-djpeg';
+
+%% load data, create arrays to store data
+ndata = 3;
+data = cell(1,ndata);
+
+load ../dpERK_oct16.mat
+data{1,1} = dpERK;
+
+load ../dpERK_jan10.mat
+data{1,2} = dpERK_0110;
+
+load ../dpERK_jan17.mat
+data{1,3} = dpERK_0117;
+
+c = 'brg';
+
+first_mode = zeros(size(data{1,1},2),ndata);
+
+%% PCA for each data set
+for i=1:ndata
+    dpERK = data{1,i};
+    
+    [m, n] = size(dpERK);
+    
+    % plot data
+    figure;
+    imagesc(dpERK)
+    ylabel('data point (scrambled)')
+    xlabel('position')
+    print(strcat(['data' num2str(i) '_scrambled']),fmt,res)
+    
+    % PCA
+    
+    [V_PCA, D_PCA] = PCA(dpERK, 5);
+    
+    % mean - center
+    mean_data = mean(dpERK, 1);
+    data_PCA = dpERK - ones(m, 1) * mean_data;
+    
+    if sum(V_PCA(:,1)) < 0
+        V_PCA(:,1) = -V_PCA(:,1);
+    end
+    first_mode(:,i) = V_PCA(:,1);
+    coeff = data_PCA * V_PCA(:,1);
+    [~, I] = sort(coeff);
+    
+    % plot PCA results
+    
+    % eigenvalue spectrum
+    figure;
+    plot(diag(D_PCA),'.')
+    xlabel('j')
+    ylabel('\lambda_j')
+    xlim([0 5])
+    print(strcat(['PCA_spectrum' num2str(i)]),fmt,res)
+    
+    % data ordered by PCA
+    figure;
+    imagesc(dpERK(I,:))
+    ylabel('time')
+    xlabel('position')
+    print(strcat(['data' num2str(i) '_unscrambled']),fmt,res)
+    
+    % first PCA mode
+    figure;
+    plot(V_PCA(:,1),'linewidth',4)
+    xlabel('position')
[…]

### 2013-12-18 group_meeting_5Feb2013/dpERK_spaced_example.m

@@ -1,0 +1,84 @@
+clear all
+close all
+
+%% set parameters for saving figures
+set(0,'DefaultLineMarkerSize',14)
+set(0,'DefaultAxesFontSize',20)
+
+res = '-r300';
+fmt = '-djpeg';
+
+%% load data
+load ../dpERK_oct16.mat
+
+%% PCA for each data set
+
+[m, n] = size(dpERK);
+
+dpERK2 = 1+max(max(dpERK)) * ones(2*size(dpERK,1),size(dpERK,2));
+dpERK2(2:2:end,:) = dpERK;
+
+jet2 = jet;
+jet2(end,:) = [1,1,1];
+
+% plot data
+figure;
+imagesc(dpERK2)
+ylabel('data point (scrambled)')
+xlabel('position')
+colormap(jet2)
+print('data_spaced_scrambled',fmt,res)
+
+% PCA
+
+[V_PCA, D_PCA] = PCA(dpERK, 5);
+
+% mean - center
+mean_data = mean(dpERK, 1);
+data_PCA = dpERK - ones(m, 1) * mean_data;
+
+if sum(V_PCA(:,1)) < 0
+    V_PCA(:,1) = -V_PCA(:,1);
+end
+coeff = data_PCA * V_PCA(:,1);
+[~, I] = sort(coeff);
+
+% data ordered by PCA
+dpERK2(2:2:end,:) = dpERK(I,:);
+figure;
+imagesc(dpERK2)
+ylabel('time')
+xlabel('position')
+colormap(jet2)
+print('data_spaced_unscrambled',fmt,res)
+
+coeff = data_PCA * V_PCA;
+
+% plot first two PCA coefficients
+figure;
+scatter(coeff(:,1),coeff(:,2),200,'b','.')
+xlabel('a_{i,1}')
+ylabel('a_{i,2}')
+axis equal
+print('coeff_12_new', fmt, res)
+
+% dmaps
+W = squareform(pdist(dpERK));
+[V_dmaps, D_dmaps] = dmaps(W, 100, 10);
+[~,I] = sort(V_dmaps(:,2));
+
+%p = polyfit(coeff(:,1),coeff(:,2),6);
+
+% plot first two PCA coefficients colored by dmaps
+figure;
+scatter(coeff(:,1),coeff(:,2),200,V_dmaps(:,2),'.')
+hold on
[…]

### 2013-12-22 image_analysis_paper/plos_draft.tex
Added some details about PCA

@@
 (c) Correlation between the VDM embedding coordinate and the membrane thickness.}
-\label{fig:angsynch_ordering}
+\label{fig:vdm_ordering}
 \end{figure}
@@
 (b) Some representative fluorescent images, aligned  and ordered using the VDM embedding coordinate. }
-\label{fig:scattrans_dpERK_ordering}
+\label{fig:vdm_image_ordering}
 \end{figure}
@@
 \subsection*{Principal Component Analysis}
+Principal component analysis, or PCA, has been the workhorse dimensionality reduction technique for the past century.
+%
+PCA finds the optimal planes or hyperplanes on which to project the data in order to maximize the variance of the projected data.
+
+We assume we have a data set consisting of $n$ $d$-dimensional observations, $X \in \mathbb{R}^{n \times d}$, where $X_{ij}$ denotes the $j^{th}$ dimension of the $i^{th}$ observation. 
+%
+PCA first begins by mean-centering the data set to produce the matrix $\hat{X}$, where
+\begin{equation}
+\hat{X}_{ij} = X_{ij} - \frac{1}{n} \sum_{k=1}^n X_{kj}.
+\end{equation}
+%
+The empirical covariance of the data set, $C \in mathbb{R}^{d \times d}$ can be computed as
+\begin{equation}
+C = X^T X.
+\end{equation}
+%
+The principal components of the data set $X$ are then given by the eigenvectors $\psi_1, \dots, \psi_d$ of $C$.
+%
+Furthermore, the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_d$ measure the variance of the data capatured by the corresponding eigenvector. 
+%
+Therefore, we choose to order the eigenvectors such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_d$.
+%
+Then, the first few eigenvectors capture a significant portion of the variance in our data set, and projection onto these first few principal components will retain many features of the data set while reducing the dimenstionality.
+%
+In our setup, we assume that the data is inherently one-dimensional, and that projection onto $\psi_1$ will uncover the correct time ordering of the data.
 
 \subsection*{Diffusion Maps}
+Unlike PCA, diffusion maps (DMAPS) is a nonlinear dimensionality reduction technique. 
+%
+DMAPS
 

### 2013-12-22 image_analysis_paper/plos_draft.tex
Added more disucssion to paper

@@
 %
-DMAPS
+DMAPS aims to uncover the parameterization of data sampled from a nonlinear manifold.
+%
+The {\em essential} requirement for DMAPS is an appropriate distance metric $d(x_i, x_j)$ for comparing data points. 
+%
+This can be the standard Euclidean distance, or a more complex metric for other data sets.
 
-\subsection*{Angular Synchronization in $SO(2)$}
+We then calculate the matrix $W \in \mathbb{R}^{n \times n}$, where 
+\begin{equation}
+W_{ij} = \exp \left( -\frac{d^2(x_i, x)j)}{\epsilon} \right)
+\end{equation}
+and $\epsilon$ is a characteristic distance between data points.
+%
+In practice, we choose $\epsilon$ to be the median of the pairwise distances between data points.
+%
+We then compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{n} W_{ij}$, and the matrix $A = D^{-1} W$. 
+%
+We then compute the eigenvectors $\phi_1, \phi_2, \dots, \phi_n$ and
+
+\subsection*{Angular Synchronization}
+
+\subsection*{Vector Diffusion Maps}
 

### 2013-12-25 image_analysis_paper/plos_draft.tex
Updated tex
added angular synchronzation and some vdm writeup

@@
 %
-We then compute the eigenvectors $\phi_1, \phi_2, \dots, \phi_n$ and
+We then compute the eigenvectors $\phi_1, \phi_2, \dots, \phi_n$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_n$ and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_n|$.
+%
+Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthoganol eigenvectors. 
+%
+The eigenvectors $\phi_1, \phi_2, \dots, \phi_n$ give the embedding coordinates, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point. 
+%
+Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector.
+%
+The next few eigenvectors give the ``meaningful'' embedding coordinates for the data.
 
@@
 
+Angular synchronization aims to align data points under some synchronization group, assuming we have estimates of the pairwise alignments to align pairs of data points.
+
+Assume we have data points $x_1, \dots, x_n$ that we wish to align under some symmetry group (typically $SO(2)$ or $SO(3)$, the space of two-dimensional or three-dimensional rotations, respectively).
+%
+Let $R_{ij}$ denote the rotation that aligns $x_i$ to $x_j$. 
+%
+We seek to find the rotations $R_1, R_2, \dots, R_n$ such that $R_i^T R_j$ is most consistent (in some sense) with $R_{ij}$, for every pair $i, j$. 
+%
+We therefore define the objective function 
+\begin{equation} \label{eqn:angsynch_obj}
+\min_{R_1, \dots, R_n \in SO(2)} \sum_{i=1}^{n} \sum_{j=1}^{n} R_i^T R_{ij} R_j.
+\end{equation}
+%
+The solution to \eqref{eqn:angsynch_obj} is not easily computed.
+%
+However, if we relax the problem and allow $R_1, \dots, R_n \in \mathbb{R}^2$, then the solution is given by the top eigenvectors of the matrix $H$, where DEFINE H.
+
+We then project the eigenvector solution back onto the space of $SO(2)$ by $\hat{R}_i = U V^T$, where $U$ and $V$ are the left and right singular vectors of ADDMORE DEF.
+
 \subsection*{Vector Diffusion Maps}
+
+Angular synchronization assumes that the data points are all replicate measurements of the same underlying data point, simply corrupted with rotations and some noise.
+%
+However, in our systems, there is an underlying symmerty group {\em as well as} a dynamical process at work, so the data points are not identical modulo symmetries and noise.
+
+Vector diffusion maps couples the symmetry-factoring of angular synchronization with the uncovering of dynamics of diffusion maps. 
+%
 

### 2013-12-26 image_analysis_paper/plos_draft.tex
Added discussion to the paper

@@
 
+High-throughput techniques have allowed for a dramatic increase in the amount of data that can be collected in a single biological experiment.
+%
+However, in systems where one is interested in studying the dynamics of a single organism, high-throughput techniques often
+
 % Please keep the Author Summary between 150 and 200 words
@@
 \subsection*{dpERK concentration profiles ordered using Principal Component Analysis}
+
+We first used Principal Components Analysis (PCA) to automatically order the one-dimensional concentration profiles in time. 
+%
+We computed the first principal component, $\psi_1$ from the data set, and ordered the data by the projection coefficients onto $\psi_1$. 
+%
+The results are shown in Figure \ref{fig:PCA_ordering}.
+%
+There is a strong correlation between the first projection coefficient, $\langle x_i, \psi_1 \rangle$, and the membrane thickness.
+%
+Therefore, we can conclude that BLAH BLAH.
+
 \begin{figure}[H]
@@
 \caption{}
+\label{subfig:PCA_ordering_image}
 \end{subfigure}
@@
 
+We would like to note that, because the embryos can be oriented with either the anterior or posterior end facing up in the microfluidic device, the directionality of the concentration profiles (clockwise versus counterclockwise around the ring, or right to left versus left to right for the linear profiles) is not meaningful.
+%
+Therefore, we could also do the PCA analysis with an expanded data set, where we include each concentration profile as well as its mirror image.
+%
+We performed this analysis, but it did not change the results significantly, since most data points are nearly symmetric with respect to left-right reflections.
+
 \subsection*{dpERK concentration profiles ordered using Diffusion Maps}
+
+Although PCA was rather successful in ordering the concentration profiles, there are some errors or inconsistencies in the orderings.
+%
+Perhaps the most obvious error is the scrambling of the secondary *ADD NAME* peaks located at the far left and right edges of the concentration profiles.
+%
+It is known that these peaks arise later in the development of {\em Drosophila}, and so any profiles which express these secondary peaks should come later in the ordering.
+%
+However, one can see in Figure \ref{subfig:PCA_ordering_image} that these secondary peaks are scrambled when the data is ordered using PCA.
+%
+We can gain insight into this ``scrambling'' by looking at the projection of the data onto the first two principal components, shown in Figure \ref{...}.
+%
+The data falls onto a one-dimensional {\em nonlinear} curve in this two-dimensional space, implying that nonlinear dimensionality reduction techniques may be more approproate, as it will order the data along the curve, and not simply along the first principal axis.
+
+We choose to use diffusion maps (DMAPS) as our nonlinear dimensionality reduction technique.
+%
+We compute the diffusion maps embedding for our data set, and order the data using the first (nontrivial) diffusion maps coordinate $\phi_2$. 
+%
+The results are shown in Figure \ref{fig:DMAPS_ordering}.
+%
+The secondary peaks are now nicely clustered at the end of the developmental trajectory.
+
 \begin{figure}[H]
@@
 \subsection*{dpERK concentration profiles aligned using angular synchronization}
+
+As we have shown, we can order the concentration profiles using PCA or DMAPS.
+%
+However, ordering the profiles first requires that we factor out the relevant symmetries in the problem.
+%
+For the linear concentration profiles that we extract from the fluorescent images, there is a shift invarance that we must consider.
+%
+In the previous analyses, we have assumed that the profiles have been prealigned using the location of the dorsal peak.
+%
+However, without some {\em a priori} knowledge of the system, we would first need to align the concentration profiles before doing our analysis.
+
+The most common and perhaps intuitive technique for alignment are template-based methods \cite{...}, where one chooses are representative data point as a ``template'', and then shifts or rotates each data point to be optimally aligned with this template.
+%
+Although these methods are simple, they require knowledge of an appropriate template function and can suffer when the data set is noisy.
+
+Angular synchronization \cite{...} is a relatively new technique that is more robust than template-based methods in the presence of noise.
+%
+We used angular synchronization to factor out the shifts in the concentration profiles measured from the fluorescent images.
+%
+Following the synchronization procedure (which yields the optimal shift for each data point so that all points are optimally aligned), we then used DMAPS to order the aligned concentration profiles.
+%
+The results are shown in Figure \ref{fig:angsynch_ordering}.
+%
+One can see that the results are visually consistent with the DMAPS results shown in Figure \ref{fig:DMAPS_ordering}.
 
@@
 \subsection*{dpERK concentration profiles aligned and ordered using vector diffusion maps}
+
+We then applied vector diffusion maps (VDM) to the unaligned data set.
+%
+VDM is similar to angular synchronization, but it is intended for data sets that contain both underlying symmetries and dynamics.
+%
+The results of aligning {\em and} ordering the data using VDM are shown in Figure \ref{fig:vdm_ordering}.
+%
+The results are similar to those from angular syncrhonization (Figure \ref{fig:angsynch_ordering}), but we now obtain the alignments and ordering in one computation.
 
@@
 
-Vector diffusion maps couples the symmetry-factoring of angular synchronization with the uncovering of dynamics of diffusion maps. 
+Vector diffusion maps (VDM) couples the symmetry-factoring of angular synchronization with the uncovering of dynamics of diffusion maps. 
 %
+We construct the matrix $A$ as in diffusion maps, and the matrix $H$ as in angular synchronization.
+%
+We then construct the matrix $S$, where the $(i,j)$ block of $S$ is the $A_{ij} H_{ij}$ where $A_{ij}$ is the $(i,j)$ entry of $A$, and $H_{ij}$ is the $(i,j)$ block of $H$. 
+%
+ADD INFO ABOUT SEPARATING COORDINATES AND ROTATIONS
 
 \subsection*{Scattering Transform}
+Angular syncrhonization and VDM both align symmetry-invariant data points {\em before} computing the diffusion maps embedding.
+%
+An alternative is to construct features that are {\em invariant} to the underlying symmetry group, and then perform DMAPS or other analysis on these features.
+
+We choose to use the scattering transform to construct these features.
+%
+The scattering transform is a specific formulation of a wavelet transform.
+%
+It is constructed to be invariant to rotations and translations.
+%
 

### 2014-01-06 image_analysis_paper/plos_draft.tex
Added more to draft

--- 2013-12-22 image_analysis_paper/plos_draft.tex
@@
 %\doublespacing
+
+\DeclareMathOperator*{\argmin}{arg\,min}
 
@@
 \begin{center}
+\begin{subfigure}{\textwidth}
+\centering
+\includegraphics[width=0.65\textwidth]{drosophila_imaging_setup}
+\caption{}
+\label{subfig:imaging_device}
+\end{subfigure}
 \begin{subfigure}{0.55\textwidth}
@@
 \end{subfigure}\\
-\begin{subfigure}{0.45\textwidth}
+\begin{subfigure}{0.4\textwidth}
 \includegraphics[width=\textwidth]{data_unordered}
@@
 \end{subfigure}
-\begin{subfigure}{0.45\textwidth}
+\begin{subfigure}{0.4\textwidth}
 \includegraphics[width=\textwidth]{data_ordered_membrane}
@@
 
-\subsection*{dpERK concentration profiles ordered using Principal Component Analysis}
+\subsection*{dpERK concentration profiles ordered using principal component analysis}
+
+We first used Principal Components Analysis (PCA) to automatically order the one-dimensional concentration profiles in time. 
+%
+We computed the first principal component, $\psi_1$, from the data set, and ordered the data by the projection coefficients onto $\psi_1$. 
+%
+The results are shown in Figure \ref{fig:PCA_ordering}.
+%
+There is a strong correlation between the first projection coefficient, $\langle x_i, \psi_1 \rangle$, and the membrane thickness.
+
 \begin{figure}[H]
@@
 \caption{}
+\label{subfig:unordered_profiles}
 \end{subfigure}
@@
 
-\subsection*{dpERK concentration profiles ordered using Diffusion Maps}
+We would like to note that, because the embryos can be oriented with either the anterior or posterior end facing up in the microfluidic device (see Figure \ref{subfig:imaging_device}), the directionality of the concentration profiles (clockwise versus counterclockwise around the ring, or right-to-left versus left-to-right for the linear profiles) is not meaningful.
+%
+Therefore, we could also do PCA with an expanded data set where we also include the mirror image of each concentration profile.
+%
+Symmetries in PCA have been studied extensively in previous work \cite{holmes1998turbulence}; it was shown that when mirror images are also included in the data set, the first PCA mode captures the symmetric portion of the variability, and the second mode captures the anti-symmetric portion.
+%
+We found that including the mirror images of each data point did not change our results significantly, since the first principal component (not shown) is nearly symmetric with respect to left-right reflections.
+
+\subsection*{dpERK concentration profiles ordered using diffusion maps}
+
+Although PCA was rather successful in ordering the concentration profiles, there are some inconsistencies in the orderings.
+%
+Perhaps the most obvious feature to note is the scrambling of the secondary *ADD NAME* peaks located at the far left and right edges of the concentration profiles.
+%
+It is known that these peaks arise later in the development of {\em Drosophila};
+however, one can see in Figure \ref{subfig:PCA_ordering_image} that these secondary peaks are scrambled when the data is ordered using PCA.
+%
+We can gain insight into this ``scrambling'' by looking at the projection of the data onto the first two principal components, shown in Figure \ref{subfig:PCA_12}.
+%
+The data falls onto a one-dimensional {\em nonlinear} curve in this two-dimensional space, implying that nonlinear dimensionality reduction techniques may be more approproate, as they can order the data along the curve, and not simply along the first principal axis.
+
+We chose to use diffusion maps (DMAPS) as our nonlinear dimensionality reduction technique.
+%
+We computed the diffusion maps embedding for our data set, using the Euclidean distance between concentration profiles as our distance metric.
+%
+We then ordered the data using the first (nontrivial) diffusion maps coordinate $\phi_2$. 
+%
+The results are shown in Figure \ref{fig:DMAPS_ordering}.
+%
+The secondary peaks are now nicely clustered towards the end of the developmental trajectory.
+
 \begin{figure}[H]
 \begin{subfigure}{0.3\textwidth}
-\includegraphics[width=\textwidth]{data_unordered}
+\includegraphics[width=\textwidth]{coeff_12_new}
 \caption{}
+\label{subfig:PCA_12}
 \end{subfigure}
@@
 \end{subfigure}
-\caption{{\bf Ordering dpERK concentration profiles using DMAPS.} (a) Concentration profiles of dpERK for many embryos.
-Each row represents a different embryo fixed at a slightly different developmental time.
+\caption{{\bf Ordering dpERK concentration profiles using DMAPS.} (a) Projection of data shown in Figure \ref{subfig:unordered_profiles} onto the first two principal components; note that the data falls on an (effectively) one-dimensional, nonlinear curve.
 (b) Concentration profiles of dpERK from (a), now ordered by the first non-trivial DMAPS embedding coordinate.
@@
 \subsection*{dpERK concentration profiles aligned using angular synchronization}
+
+As we have shown, we can order the concentration profiles using PCA or DMAPS.
+%
+However, ordering the profiles first requires that we factor out the relevant symmetries in the problem.
+%
+For the linear concentration profiles that we extract from the fluorescent images, there is a shift invarance that we must consider (because the profiles in fact lie on a ring).
+%
+In the previous analyses, we have assumed that the profiles have been prealigned using the location of the dorsal peak.
+%
+However, without some {\em a priori} knowledge of the system (i.e. knowing that dorsal is expressed at one position), we would need to align the concentration profiles before doing our analysis.
+
+The most common and perhaps intuitive techniques for alignment are template-based methods \cite{ahuja2007template}, where one chooses are representative data point as a ``template'', and then shifts or rotates each data point to be optimally aligned with this template.
+%
+Although these methods are simple, they require knowledge of an appropriate template function and can suffer when the data set is noisy.
+
+Angular synchronization \cite{singer2011angular} is a relatively new technique that is more robust than template-based methods in the presence of noise.
+%
+We used angular synchronization to factor out the shifts in the concentration profiles measured from the fluorescent images.
+%
+After aligning the data points using angular synchronization, we used DMAPS to order the aligned concentration profiles.
+%
+The results are shown in Figure \ref{fig:angsynch_ordering}.
+%
+One can see that the results are visually consistent with the results shown in Figure \ref{fig:DMAPS_ordering}.
 
@@
 
+Instead of using angular synchronization and DMAPS, we can do the alignment and dimensionality reduction/ordering in one computation using vector diffusion maps (VDM).
+%
+VDM is similar to angular synchronization, but it is intended for data sets that contain both underlying symmetries and dynamics.
+%
+The results of aligning {\em and} ordering the data using VDM are shown in Figure \ref{fig:vdm_ordering}; the results are similar to those from angular syncrhonization (Figure \ref{fig:angsynch_ordering}).
+
 \begin{figure}[H]
@@
 
+Thus far, we have focused on the one-dimensional linear concentration profiles that are extracted from the fluorescent images.
+%
+However, the techniques we have outlined can also be applied to the two-dimensional fluorescent images. 
+%
+We use VDM to align and order the fluorescent images stained for dpERK. 
+%
+We now must factor out translations and rotations in the images. 
+%
+To do this, we project the images onto a sphere;
+rotations and translations of a two-dimensional image then correspond to (three-dimensioanl) rotations of the sphere.
+%
+We then use VDM to align and order the (projected) images, with the underlying symmetry group $SO(3)$. 
+%
+The results are shown in Figure \ref{fig:vdm_image_ordering};
+the orderings are of comparable quality to those obtained with the one-dimensional profiles, and the dpERK peaks are nicely aligned.
+
 \begin{figure}[H]
@@
 \subsection*{Two-dimensional dpERK images ordered using diffusion maps and the scattering transform}
+
+An alternative to aligning the data (as in angular synchronization of VDM) is to construct {\em features} that are invariant to the underlying symmetry group, and then compute distances between data points using these features for use in a DMAPS calculation.
+%
+In our setup, we need features for two-dimensional images that are invariant to translations and rotations of the images.
+%
+We chose to use the scattering transform \cite{mallat2012group} to construct our invariant features.
+%
+Not only is this transform invariant to translations and rotations, but it is built on the wavelet transform (which is know to perform well for images), and is also contractive (so that noise and small deformations are not amplified by the transform).
+
+We compute the scattering transform coefficients for each image in our data set.
+%
+We then perform a DMAPS calculation, using the Euclidean distance between the vectors of scattering transform coefficients as our distance metric.
+%
+The results are shown in Figure \ref{fig:scattrans_dpERK_ordering}.
+%
+The ordering is again good; however, using this method, we do not obtain optimal alignments for the images (we only obtain an ordering).
 
@@
 
+We hope that we have illustrated the use of dimensionality reduction techniques in application with underlying symmetries.
+%
+The work here is intended to provide an overview of the different techniques available, as well as the benefits and shortcomings of each method.
+
+\begin{table}
+	\begin{tabular}{| p{0.3\textwidth} | c | c | p{0.3\textwidth} |}
+		\hline 
+		Method & Spearman's $\rho$ & Execution Time & Notes About Method \\ 
+		\hline 
+		PCA on one-dimensional profiles & 0.9244 & & \\
+		\hline 
+		DMAPS on one-dimensional profiles & 0.9209 & & \\
+		\hline 
+		Angular synchornization + DMAPS on one-dimensional profiles & 0.9164 & & \\
+		\hline 
+		VDM on one-dimensional profiles & 0.9232 & & \\	
+		\hline 
+		VDM on two-dimensional images & 0.9051  & & \\
+		\hline 	
+		Scattering transform + DMAPS on two-dimensional profiles & 0.9011  & & \\
+		\hline
+	\end{tabular}
+\end{table}
 % You may title this section "Methods" or "Models".
@@
 \subsection*{Principal Component Analysis}
-Principal component analysis, or PCA, has been the workhorse dimensionality reduction technique for the past century.
+Principal component analysis (PCA) has been the workhorse dimensionality reduction technique for the past century.
 %
-PCA finds the optimal planes or hyperplanes on which to project the data in order to maximize the variance of the projected data.
+PCA finds the optimal planes or hyperplanes on which to project the data in order to maximize the variance of the projected data \cite{shlens2005tutorial}.
 
-We assume we have a data set consisting of $n$ $d$-dimensional observations, $X \in \mathbb{R}^{n \times d}$, where $X_{ij}$ denotes the $j^{th}$ dimension of the $i^{th}$ observation. 
+Let $X \in \mathbb{R}^{n \times d}$ denote our data set consisting of $n$ points in $d$ dimensions, where $X_{ij}$ denotes the $j^{th}$ dimension of the $i^{th}$ observation. 
 %
-PCA first begins by mean-centering the data set to produce the matrix $\hat{X}$, where
+We first mean-center the data set to produce the matrix $\hat{X}$, where
 \begin{equation}
@@
 %
-The empirical covariance of the data set, $C \in mathbb{R}^{d \times d}$ can be computed as
+The empirical covariance of the data set, $C \in \mathbb{R}^{d \times d}$ is then given by
 \begin{equation}
@@
 %
-The principal components of the data set $X$ are then given by the eigenvectors $\psi_1, \dots, \psi_d$ of $C$.
+We compute the eigenvectors $\psi_1, \dots, \psi_d$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_d$ of $C$, and order them such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_d$.
 %
-Furthermore, the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_d$ measure the variance of the data capatured by the corresponding eigenvector. 
+The principal components of the data set $X$ are given by $\psi_1, \dots, \psi_d$, and the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_d$ measure the variability captured by the corresponding eigenvector.
 %
-Therefore, we choose to order the eigenvectors such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_d$.
-%
-Then, the first few eigenvectors capture a significant portion of the variance in our data set, and projection onto these first few principal components will retain many features of the data set while reducing the dimenstionality.
+If there is a {\em spectral gap} such that the first few eigenvectors capture a significant portion of the variance in our data set, projection onto these first few principal components will retain most features of the data set while reducing the dimensionality.
 %
@@
 %
-DMAPS aims to uncover the parameterization of data sampled from a nonlinear manifold.
+DMAPS aims to uncover the parameterization of data sampled from a nonlinear manifold \cite{coifman2005geometric}.
 %
@@
 %
-This can be the standard Euclidean distance, or a more complex metric for other data sets.
+This can as simple as the standard Euclidean distance, or a more complex metric (such as a distance between features of the data points) for other data sets.
 
-We then calculate the matrix $W \in \mathbb{R}^{n \times n}$, where 
+Given data points $x_1, \dots, x_n \in \mathbb{R}^d$, we fist calculate the matrix $W \in \mathbb{R}^{n \times n}$, where 
 \begin{equation}
@@
 %
-In practice, we choose $\epsilon$ to be the median of the pairwise distances between data points.
+$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph}); in practice, we choose $\epsilon$ to be the median of the pairwise distances between data points.
 %
@@
 %
-We then compute the eigenvectors $\phi_1, \phi_2, \dots, \phi_n$ and
+We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_n$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_n$ and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_n|$.
+%
+Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors. 
+%
+The eigenvectors $\phi_1, \phi_2, \dots, \phi_n$ give the embedding coordinates, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point. 
+%
+Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector; the next few eigenvectors give the ``meaningful'' embedding coordinates for the data.
 
@@
 
+When the data set is invariant under an underlying symmetry group (such as translations and/or rotations), we must first factor out the underlying symmetries before doing further analysis.
+%
+Angular synchronization calculates the optimal alignments for a data set, assuming we have estimates of the optimal alignments between pairs of data points \cite{singer2011angular}.
+
+Let $x_1, \dots, x_n$ denote the data points that we wish to align under some symmetry group.
+%
+We will consider the case where the symmetry group is the group of two-dimensional rotations, $SO(2)$; the same construction and ideas hold for other symmetry groups, such as $SO(3)$. 
+%
+Let $R_{ij} \in SO(2)$ denote the rotation that aligns $x_j$ to $x_i$, so that
+\begin{equation}
+R_{ij} = \argmin_{R \in SO(2)} \|Rx_j - x_i \|^2.
+\end{equation}
+%
+We seek to find the rotations $R_1, R_2, \dots, R_n \in SO(2)$ such that $R_i R_j^T \approx R_{ij}$, for every pair $i, j$. (CHECK THIS)
+%
+We would also like to exploit {\em higher-order} consistency relations.
+%
+For example, for noiseless measurements, $R_{ij} R_{jk} R_{ki} = I$, where $I$ is the identity matrix; therefore, we consider measurements which (almost) satisfy such conditions ``good'' measurements, and those measurements which do not satisfy such conditions as most likely inaccurate.
+
+We define the objective function 
+\begin{equation} \label{eqn:angsynch_obj}
+\max_{R_1, \dots, R_n \in SO(2)} \sum_{i=1}^{n} \sum_{j=1}^{n} R_i^T R_{ij} R_j.
+\end{equation}
+%
+In general, the solution to \eqref{eqn:angsynch_obj} is not easily computed.
+
+Instead, we relax the problem and allow $R_1, \dots, R_n \in \mathbb{R}^2$.
+%
+We then define the matrix $H$, which is an $n \times n$ block matrix with $2 \times 2$ blocks (so $H \in \mathbb{R}^{2n \times 2n}$).
+%
+We then compute the top ``block'' eigenvector of $H$, which we denote $\hat{R} \in \mathbb{R}^{2n \times 2}$. 
+%
+The solution to the relaxed problem is given by $\hat{R}$, so that the $i^{th}$ rotation is (approximately) $\hat{R}(i) \in \mathbb{R}^{2 \times 2}$.
+%
+However, because we relaxed the problem to allow our solutions to lie in $\mathbb{R}^{2 \times 2}$, we must project our approximate solution back to $SO(2)$.
+%
+The optimal rotation is therefore given by $R_i = U_i V_i^T$, where $U_i$ and $V_i$ are the left and right singular vectors of $\hat{R}_i$, respectively.
+
 \subsection*{Vector Diffusion Maps}
 
+Angular synchronization assumes that each data point is a replicate measurements of the same underlying configuration, simply corrupted with rotations and some noise.
+%
+However, in system with an underlying symmetry group {\em as well as} a dynamical process, the data points are not identical modulo symmetries and noise, and we would like to place more emphasis on aligning data points which are dynamically close.
+
+Vector diffusion maps (VDM) couples the symmetry-factoring of angular synchronization with the uncovering of dynamics of diffusion maps \cite{singer2012vector}. 
+%
+We construct the matrix $A$ as in diffusion maps, and the matrix $H$ as in angular synchronization.
+%
+We then construct the block matrix $S$, where the blocks of $S$ are defined by
+\begin{equation}
+S_{ij} = \left\{ \begin{array}{l l} 
+A_{ij} H_{ij} & i \ne j \\
+0_{2 \times 2} & i = j
+\end{array}
+\right.
+\end{equation}
+%
+where $A_{ij}$ is the $(i,j)$ entry of $A$, and $H_{ij}$ is the $(i,j)$ block of $H$.
+
+We then compute the eigenvectors $\phi_1, \dots, \phi_{2n}$ and eigenvalues $\lambda_1, \dots, \lambda_{2n}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{2n}|$.
+%
+As in angular synchronization, the first ``block'' eigenvector, i.e. the first two eigenvectors concatenated into a $2 n \times 2$ matrix, gives the optimal rotations. 
+%
+However, the eigenvectors now also give us embedding coordinates.
+%
+Let $\phi_j(i)$ denote the $i^{th}$ block of $\phi_j$. 
+%
+The embedding coordinates are then given by $\langle \phi_j(i), \phi_k(i) \rangle$, where $1 \le j, k, \le 2 n$.
+
 \subsection*{Scattering Transform}
+
+The scattering transform consists of a wavelet transform of the signal, followed by  the appropriate averaging operator.
+%
+This transform is applied to the signal itself, as well as to the residual, in order to preserve as much information about the signal itself, while still keeping the transform coefficients translation- and rotation-invariant.
+% 
+Furthermore, the scattering transform can be shown to be a contractive operator; therefore, noise and small deformations in the signal are not amplified in the transformation.
+
+\subsection*{Spearman's $\rho$}
+
+Spearman's rank correlation coefficient, or Spearman's $\rho$, measures the monotonicity of a function \cite{...}. 
+%
+It therefore serves as an appropriate metric for the ``goodness'' of our orderings.
+%
+Given data points $X_i$ and $Y_i$, we compute the ranks of each data point within the data set, denoted $x_i$ and $y_i$.
+%
+Spearman's $\rho$ is then defined as the correlation between the ranks
+\begin{equation}
+\rho = \frac{\sum_i (x_i - \overline{x})(y_i - \overline{y})}{\sqrt{\sum_i (x_i - \overline{x})^2 \sum_i (y_i - \overline{y})^2}}
+\end{equation}
 
@@
 % The bibtex filename
-\bibliography{template}
+\bibliography{../../references/references}
 

### 2014-01-07 image_analysis_paper/plos_draft.tex
Added timing to draft

@@
 		\hline 
-		PCA on one-dimensional profiles & 0.9244 & & \\
+		PCA on one-dimensional profiles & 0.9244 & < 1 second & \\
 		\hline 
-		DMAPS on one-dimensional profiles & 0.9209 & & \\
+		DMAPS on one-dimensional profiles & 0.9209 & < 1 second & \\
 		\hline 
-		Angular synchornization + DMAPS on one-dimensional profiles & 0.9164 & & \\
+		Angular synchornization + DMAPS on one-dimensional profiles & 0.9164 & < 1 second & \\
 		\hline 
-		VDM on one-dimensional profiles & 0.9232 & & \\	
+		VDM on one-dimensional profiles & 0.9232 & < 1 second & \\	
 		\hline 
-		VDM on two-dimensional images & 0.9051  & & \\
+		VDM on two-dimensional images & 0.9051  & 325 seconds & \\
 		\hline 	
-		Scattering transform + DMAPS on two-dimensional profiles & 0.9011  & & \\
+		Scattering transform + DMAPS on two-dimensional profiles & 0.9011  & 86 seconds & \\
 		\hline

### 2014-01-07 image_analysis_paper/plos_draft.tex
Edited draft

--- 2014-01-06 image_analysis_paper/plos_draft.tex
@@
 		\hline 
-		PCA on one-dimensional profiles & 0.9244 & & \\
+		PCA on one-dimensional profiles & 0.9244 & $< 1$ second & \\
 		\hline 
-		DMAPS on one-dimensional profiles & 0.9209 & & \\
+		DMAPS on one-dimensional profiles & 0.9209 & $< 1$ second & \\
 		\hline 
-		Angular synchornization + DMAPS on one-dimensional profiles & 0.9164 & & \\
+		Angular synchronization + DMAPS on one-dimensional profiles & 0.9164 & $< 1$ second & \\
 		\hline 
-		VDM on one-dimensional profiles & 0.9232 & & \\	
+		VDM on one-dimensional profiles & 0.9232 & $< 1$ second & \\	
 		\hline 
-		VDM on two-dimensional images & 0.9051  & & \\
+		VDM on two-dimensional images & 0.9051  & $325$ seconds & $\mathcal{O}(n^2)$ to compute the pairwise alignments\\
 		\hline 	
-		Scattering transform + DMAPS on two-dimensional profiles & 0.9011  & & \\
+		Scattering transform + DMAPS on two-dimensional profiles & 0.9011  & $86$ seconds &  $\mathcal{O}(n)$ to compute the scattering transform for each data point \\
 		\hline

### 2014-01-08 image_analysis_paper/references_for_plos.tex
Added bacground references readme file for Yannis

\documentclass[12pt]{article}

\title{References for PloS paper}

\usepackage{biblatex}

\bibliography{../../references/references}

\begin{document}

\section{Algorithms}

\fullcite{flusser2000independence}
\begin{itemize}
\item Construct moments (polynomials) of the pixel intensities that are invariant to rotations
\end{itemize}
%
\fullcite{besl1992method}
\begin{itemize}
	\item Iterative closest point (ICP) algorithm
	\item Aligns/registers point clouds
\end{itemize}
%
\fullcite{lowe1999object}
\begin{itemize}
	\item Scale-invartiant feature transform (SIFT)
	\item Computes feature vectors for image; features are scale, translation, and rotation invariant
\end{itemize}
%
\fullcite{sadler1992shift}
\begin{itemize}
	\item Calculate bispectrum for image
	\item Bispectrum is rotation-invariant
	\item Do not need to align the images
\end{itemize}
%
\fullcite{hilai1994recognition}
\begin{itemize}
	\item Construct rotation-invariant basis for images
\end{itemize}
%
\fullcite{ojala2002multiresolution}
\begin{itemize}
	\item Decribe local patches of image (texture) by binary code
	\item Define function on binary code that is rotation-invariant
	\item Classify using binary code
\end{itemize}
%
\fullcite{zhao2013fourier}
\begin{itemize}
	\item Construct PCA basis that is rotation-invariant
\end{itemize}

\section{Applications}

\fullcite{hajnal2010medical}
\begin{itemize}
	\item Medical image registration (MRI, CT, etc.)
\end{itemize}
%
\fullcite{greenspan1994rotation}
\begin{itemize}
\item Texture recognition/classification
\item use Fourier basis; fit texture response (in Fourier basis) to rotations 
\end{itemize}
%
\fullcite{rowley1998rotation}
\begin{itemize}
	\item face recognition
	\item train neural net to estimate face orientation, then orient faces
\end{itemize}
%
\fullcite{simard1992efficient}
\begin{itemize}
	\item recognize digits
	\item only consider small rotations; fit plane to span of rotations of each digit, then see which plane new digit is closest to
\end{itemize}

\end{document}

### 2014-01-08 image_analysis_paper/plos_draft.tex
Added some to writeup

@@
 
+The issue of rotation invariance arises in many applications, most commonly image registration and image alignment.
+%
+In settings such as face recognition, medical imaging, and computer vision, objects from different frames or timepoints must be aligned or ``registered'' before further analysis can be done.
+%
+For example, often medical images (such as MRI scans) are collected over multiple days, and doctors would like to study development of a condition (such as the growth of a tumor).
+%
+However, on each day, the patient and/or sensors may be in slightly different locations; therefore, the images must be aligned before the doctor can look at how the tumor is growing and evolving.
+
+Several different techniques have been developed to solve the image registration problem. 
+%
+Most of these techniques are ``template-based'' techniques, where one selects a prototypical image or ``template'', and then aligns the remaining images to this template \cite{...}.
+%
+Such techniques can suffer if a good template is not know {\em a priori}; when the images are noisy, selecting one image from the data set at random can yield a noisy template and result in many mis-alignments.
+
+In addition, most existing techniques rely on feature extraction algorithms, and these features are often problem-specific.
+%
+For example, (ADD INFO ABOUT FACES).
+
+We are interested in analyzing data collected from developing {\em Drosophila} embryos during nuclear cycle 14 (NC14) (CHECK).
+%
+(ADD MORE ABOUT WHY THIS IS IMPORTANT).
+
+Currently, data is collected using a high-throughput microfluidic device \cite{chung2010microfluidic}; we can therefore image hundreds of embryos in one experiment.
+%
+Each embryo is stained for the membrane proteins, dpERK, and DI. 
+%
+We image the cross-section of each embryo.
+%
+We then use image analysis software to extract the circular dpERK concentration profile from the fluorescent image.
+%
+We use the expression of DI to find the dorsal midline of the embryo, and ``unzip'' the circular profile at the dorsal midline to obtain a one-dimensional, linear dpERK concentration profile.
+%
+We then measure the membrane thickness of them embryo; the thickness of the membrane is known to be monotonic in time during NC14 \cite{...}. 
+%
+We then order the concentration profiles for each embryo by the thickness of the membrane to obtain a ``reconstruction'' of the dynamics of dpERK.
+
 \begin{figure}[H]
@@
 \end{figure}
+
+Here,  we will discuss alternative methodologies to reconstruct the dpERK dynamics from fluorescent images. 
+%
+We will use various data mining techniques to {\em automatically} order the dpERK concentration profiles in time.
+%
+We will not only focus on the linear concentration profiles extracted from the images, but also work directly with the two-dimensional dpERK fluorescent images. 
+%
+Finally, we will look at ordering the two-dimensional membrane images in time; these are much more feature-rich than the dpERK images, and prove to be more challenging.
 
@@
 		\hline
+		Scattering transform + DMAPS on two-dimensional membrane images & 0.8403  & $300$ seconds &  $\mathcal{O}(n)$ to compute the scattering transform for each data point \\
+		\hline
 	\end{tabular}
@@
 
+\subsection*{Edge detection}
+For the membrane images, we are only interested in the location of the edges of the membrane (not other features, such as the intensity of the fluorescent stain).
+%
+We therefore preprocess the
+
 \subsection*{Spearman's $\rho$}

### 2014-01-10 image_analysis_paper/plos_draft.tex
added more to draft

@@
 
-The issue of rotation invariance arises in many applications, most commonly image registration and image alignment.
+The issue of rotation invariance arises in many applications, such as medical imaging \cite{hajnal2010medical}, face recognition \cite{rowley1998rotation}, digit recognition \cite{simard1992efficient}, and texture classification \cite{greenspan1994rotation}.
 %
-In settings such as face recognition, medical imaging, and computer vision, objects from different frames or timepoints must be aligned or ``registered'' before further analysis can be done.
+In these settings, objects from different frames or timepoints must be aligned or ``registered'' before further analysis can be done.
 %
-For example, often medical images (such as MRI scans) are collected over multiple days, and doctors would like to study development of a condition (such as the growth of a tumor).
+%For example, often medical images (such as MRI scans) are collected over multiple days, and doctors would like to study development of a condition (such as the growth of a tumor).
 %
-However, on each day, the patient and/or sensors may be in slightly different locations; therefore, the images must be aligned before the doctor can look at how the tumor is growing and evolving.
-
+%However, on each day, the patient and/or sensors may be in slightly different locations; therefore, the images must be aligned before the doctor can look at how the tumor is growing and evolving.
+%
 Several different techniques have been developed to solve the image registration problem. 
 %
-Most of these techniques are ``template-based'' techniques, where one selects a prototypical image or ``template'', and then aligns the remaining images to this template \cite{...}.
+``Template-based'' techniques, where one selects a prototypical image or ``template'', and then aligns the remaining images to this template \cite{ahuja2007template}.
 %
 Such techniques can suffer if a good template is not know {\em a priori}; when the images are noisy, selecting one image from the data set at random can yield a noisy template and result in many mis-alignments.
-
-In addition, most existing techniques rely on feature extraction algorithms, and these features are often problem-specific.
 %
-For example, (ADD INFO ABOUT FACES).
+Other techniques focus on constructing ``features'' that are rotation-invariant \cite{flusser2000independence, lowe1999object, sadler1992shift, ojala2002multiresolution}, and then using these features (rather than the images themselves) in further analyses and computations.
+%
+However, such features are often problem-dependent, and developing rotation-invariant features that are stable to noise and deformations is still an active area of research.
+%
+One can also construct algorithms that are invariant to rotations of the images \cite{hilai1004recognition, zhao2013fourier}.
+%
+However,
 
@@
 %
-We then use image analysis software to extract the circular dpERK concentration profile from the fluorescent image.
+We then use image analysis software (ADD HOW THIS IS DONE IN THE APPENDIX) to extract the circular dpERK concentration profile from the fluorescent image.
 %
@@
 %
-We then measure the membrane thickness of them embryo; the thickness of the membrane is known to be monotonic in time during NC14 \cite{...}. 
+We measure the membrane thickness of them embryo; the thickness of the membrane is known to be monotonic in time during NC14 \cite{...}. 
 %
@@
 %
-For the linear concentration profiles that we extract from the fluorescent images, there is a shift invarance that we must consider (because the profiles in fact lie on a ring).
+The developmental dynamics are invariant to the oritentation of the embryo in the microfluidic device; therefore, to obtain meaningful results from our analysis, we must first align the embryo snapshots.
 %
-In the previous analyses, we have assumed that the profiles have been prealigned using the location of the dorsal peak.
+Rotation of the embryo cross-sections correspond to translation (with periodic boundary conditions) of the linear profiles we have been considering.
 %
-However, without some {\em a priori} knowledge of the system (i.e. knowing that dorsal is expressed at one position), we would need to align the concentration profiles before doing our analysis.
+In the previous analyses, we have assumed that the profiles have been prealigned using the location of the dorsal protein peak.
+%
+However, without some {\em a priori} knowledge of the system (i.e.
+knowing that dorsal is expressed at one position along the perimeter of the embryo), we would need to align the concentration profiles before doing our analysis.
 
-The most common and perhaps intuitive techniques for alignment are template-based methods \cite{ahuja2007template}, where one chooses are representative data point as a ``template'', and then shifts or rotates each data point to be optimally aligned with this template.
+The most common and perhaps intuitive techniques for alignment are template-based methods \cite{ahuja2007template}, where one chooses are representative data point or shape as a ``template'', and then shifts or rotates each data point to be optimally aligned with this template.
 %
 Although these methods are simple, they require knowledge of an appropriate template function and can suffer when the data set is noisy.
-
-Angular synchronization \cite{singer2011angular} is a relatively new technique that is more robust than template-based methods in the presence of noise.
+%
+Angular synchronization \cite{singer2011angular} is a relatively new technique that is often more robust than template-based methods in the presence of noise.
 %
@@
 
+\subsection*{Image Preprocessing}
+The dpERK images we collect from the confocal microscope are subsampled to $100 \times 100$ pixels; the membrane images are subsampled to $200 \times 200$ pixels (due to the finer features in the membrane pictures).
+
+Because we are only interested in the membrane thickness, we run a simple edge detection algorithm on the membrane images.
+%
+We use a Laplacian of Gaussians filter \cite{...}, as implemented in the Matlab Image processing toolbox \cite{...}.
+
 \subsection*{Principal Component Analysis}

### 2014-01-13 image_analysis_paper/plos_draft.tex
Added more to report

@@
 %
-``Template-based'' techniques, where one selects a prototypical image or ``template'', and then aligns the remaining images to this template \cite{ahuja2007template}.
+``Template-based'' techniques are relatively common, where one selects a prototypical image or ``template'', and then aligns the remaining images to this template \cite{ahuja2007template}.
 %
@@
 %
-Other techniques focus on constructing ``features'' that are rotation-invariant \cite{flusser2000independence, lowe1999object, sadler1992shift, ojala2002multiresolution}, and then using these features (rather than the images themselves) in further analyses and computations.
+Other techniques focus on constructing ``features'' that are rotation-invariant \cite{flusser2000independence, lowe1999object, sadler1992shift, ojala2002multiresolution}, so that one can use these features (rather than the images themselves) in further analyses and computations.
 %
-However, such features are often problem-dependent, and developing rotation-invariant features that are stable to noise and deformations is still an active area of research.
+However, constructing rotation-invariant features that still retain the important information in the images is difficult and often problem-dependent, and developing rotation-invariant features that are stable to noise and deformations is still an active area of research \cite{mallat2012group, sifre2013rotation}.
 %
-One can also construct algorithms that are invariant to rotations of the images \cite{hilai1004recognition, zhao2013fourier}.
+One can also construct algorithms that are invariant to rotations of the images \cite{hilai1994recognition, zhao2013fourier}.
 %
-However,
+ADD WHY THIS IS BAD.
 
@@
 \centering
-\begin{subfigure}{0.45\textwidth}
-\includegraphics[width=\textwidth]{vdm_2d_time_corr}
+\begin{subfigure}{\textwidth}
+\centering
+\includegraphics[width=0.7\textwidth]{vdm_2d_time_corr}
 \caption{}
 \end{subfigure}
-\begin{subfigure}{0.5\textwidth}
+\begin{subfigure}{\textwidth}
 \foreach \n in {1,9,17,25,30,34,38,43,49}{
-\includegraphics[width=0.3\textwidth]{dpERK_vdm_\n}
+\includegraphics[width=0.07\textwidth, trim=6in 0in 6in 0in, clip]{dpERK_vdm_\n}
 \hfill}
@@
 \centering
-\begin{subfigure}{0.45\textwidth}
-\includegraphics[width=\textwidth]{DMAPS_scat_time_corr}
+\begin{subfigure}{\textwidth}
+\includegraphics[width=0.7\textwidth]{DMAPS_scat_time_corr}
 \caption{}
 \end{subfigure}
-\begin{subfigure}{0.5\textwidth}
+\begin{subfigure}{\textwidth}
 \foreach \n in {1,9,17,25,30,34,38,43,49}{
-\includegraphics[width=0.3\textwidth]{dpERK_scat_\n}
+\includegraphics[width=0.07\textwidth, trim=6in 0in 6in 0in, clip]{dpERK_scat_\n}
 \hfill}
@@
 
+$\rho = 0.8403$
+
+$\rho = 0.3525$ for VDM + membrane pics
+
 \begin{figure}[H]
 \centering
-\begin{subfigure}{0.45\textwidth}
-\includegraphics[width=\textwidth]{DMAPS_membrane_scat_time_corr}
+\begin{subfigure}{\textwidth}
+\centering
+\includegraphics[width=0.7\textwidth]{DMAPS_membrane_scat_time_corr}
 \caption{}
 \end{subfigure}
-\begin{subfigure}{0.5\textwidth}
+\begin{subfigure}{\textwidth}
 \foreach \n in {1,9,17,25,30,34,38,43,49}{
-\includegraphics[width=0.3\textwidth]{membrane_scat_\n}
+\includegraphics[width=0.07\textwidth, trim=6in 0in 6in 0in, clip]{membrane_scat_\n}
+\hfill}
+\newline
+\foreach \n in {1,9,17,25,30,34,38,43,49}{
+\includegraphics[width=0.07\textwidth, trim=6in 0in 6in 0in, clip]{membrane_scat_raw_\n}
 \hfill}
@@
 \subsection*{Image Preprocessing}
-The dpERK images we collect from the confocal microscope are subsampled to $100 \times 100$ pixels; the membrane images are subsampled to $200 \times 200$ pixels (due to the finer features in the membrane pictures).
+The dpERK images we collect from the confocal microscope are subsampled to $100 \times 100$ pixels; the membrane images are subsampled to $200 \times 200$ pixels (due to the finer structure in the membrane pictures).
 
@@
 Furthermore, the scattering transform can be shown to be a contractive operator; therefore, noise and small deformations in the signal are not amplified in the transformation.
-
-\subsection*{Edge detection}
-For the membrane images, we are only interested in the location of the edges of the membrane (not other features, such as the intensity of the fluorescent stain).
 %
-We therefore preprocess the
+We utilized the freely available ScatNet software (\texttt{http://www.di.ens.fr/data/software/scatnet/}) to compute the rotation-invariant scattering coefficients.
 
@@
 \end{equation}
+where $\overline{x}$ and $\overline{y}$ are the averages of $x_i$ and $y_i$, respectively.
 

### 2014-01-13 image_analysis_paper/plos_draft.tex
Added more to draft

@@
 \usepackage{pgffor}
+\usepackage{hyperref}
+
 %% END MACROS SECTION
@@
 {\Large
-\textbf{Image Analysis for {\em Drosophila} Embryogenesis}
+\textbf{Rotation-Invariant Image Analysis for Reconstructing Dynamics in {\em Drosophila} Embryogenesis}
 }
@@
 
-The issue of rotation invariance arises in many applications, such as medical imaging \cite{hajnal2010medical}, face recognition \cite{rowley1998rotation}, digit recognition \cite{simard1992efficient}, and texture classification \cite{greenspan1994rotation}.
+The issue of rotation invariance arises in many applications, including medical imaging \cite{hajnal2010medical}, face recognition \cite{rowley1998rotation}, digit recognition \cite{simard1992efficient}, and texture classification \cite{greenspan1994rotation}.
 %
-In these settings, objects from different frames or timepoints must be aligned or ``registered'' before further analysis can be done.
+In these settings, objects from different frames or timepoints must be aligned or ``registered'' prior to further analysis.
 %
@@
 %
-``Template-based'' techniques are relatively common, where one selects a prototypical image or ``template'', and then aligns the remaining images to this template \cite{ahuja2007template}.
+``Template-based'' techniques, where one selects a prototypical image or ``template'' and then aligns the remaining images to this template, are relatively common \cite{ahuja2007template}.
 %
-Such techniques can suffer if a good template is not know {\em a priori}; when the images are noisy, selecting one image from the data set at random can yield a noisy template and result in many mis-alignments.
+Such techniques can suffer if a good template is not know {\em a priori}; if the images are noisy, selecting one image from the data set at random can yield a noisy template and result in many mis-alignments.
 %
@@
 %
-However, constructing rotation-invariant features that still retain the important information in the images is difficult and often problem-dependent, and developing rotation-invariant features that are stable to noise and deformations is still an active area of research \cite{mallat2012group, sifre2013rotation}.
+However, developing rotation-invariant features that still retain the important information in the images is difficult and often problem-dependent, and constructing features that are stable to noise and deformations is still an active area of research \cite{mallat2012group, sifre2013rotation}.
 %
-One can also construct algorithms that are invariant to rotations of the images \cite{hilai1994recognition, zhao2013fourier}.
+One can also construct algorithms that consider all possible image rotations in the analysis \cite{hilai1994recognition, zhao2013fourier}.
 %
-ADD WHY THIS IS BAD.
+%TODO: why do these methods fail?
 
-We are interested in analyzing data collected from developing {\em Drosophila} embryos during nuclear cycle 14 (NC14) (CHECK).
+We are interested in analyzing images of developing {\em Drosophila} embryos during nuclear cycle 14 (NC14).
+%TODO: check that NC14 is correct
 %
-(ADD MORE ABOUT WHY THIS IS IMPORTANT).
-
-Currently, data is collected using a high-throughput microfluidic device \cite{chung2010microfluidic}; we can therefore image hundreds of embryos in one experiment.
+More specifically, we would like to order the images {\em in time} so that we can reconstruct the developmental dynamics from snapshots. 
 %
-Each embryo is stained for the membrane proteins, dpERK, and DI. 
+% TODO: add motivation about why we are interested in this
 %
-We image the cross-section of each embryo.
+Currently, data is collected using a high-throughput microfluidic device \cite{chung2010microfluidic}, so we can image hundreds of embryos in one experiment.
 %
-We then use image analysis software (ADD HOW THIS IS DONE IN THE APPENDIX) to extract the circular dpERK concentration profile from the fluorescent image.
+Each embryo is stained for the membrane proteins, dpERK, and DI, and we collect fluorescent images of the cross-section of each embryo using a confocal microscope.
+%
+We then use image analysis software to extract the dpERK concentration profile around the circumfrence of the embryo from the fluorescent image.
 %
@@
 %
-We measure the membrane thickness of them embryo; the thickness of the membrane is known to be monotonic in time during NC14 \cite{...}. 
+We also measure the membrane thickness of the embryo; the thickness of the membrane is known to be monotonic in time during NC14 \cite{lim2013kinetics, lecuit2002slam}. 
 %
-We then order the concentration profiles for each embryo by the thickness of the membrane to obtain a ``reconstruction'' of the dynamics of dpERK.
+We then order the dpERK concentration profiles for each embryo by the thickness of the membrane to obtain a ``trajectory'' of the dpERK dynamics.
 
-\begin{figure}[H]
+\begin{figure}[!ht]
 \begin{center}
@@
 \includegraphics[width=0.8\textwidth]{drosophila_schematic}\\
-\includegraphics[width=0.3\textwidth]{drosophila_membrane}
-\includegraphics[width=0.3\textwidth]{drosophila_dpERK}
-\includegraphics[width=0.3\textwidth]{drosophila_DI}
+\includegraphics[width=0.25\textwidth]{drosophila_membrane}
+\hfill
+\includegraphics[width=0.25\textwidth]{drosophila_dpERK}
+\hfill
+\includegraphics[width=0.25\textwidth]{drosophila_DI}
 \caption{}
@@
 \centering
-\includegraphics[width=0.6\textwidth]{circle_profile}\\
+\includegraphics[width=0.7\textwidth]{circle_profile}\\
 \includegraphics[width=\textwidth, trim=0mm 300mm 0mm 200mm, clip]{line_profile}
@@
 %
-We will use various data mining techniques to {\em automatically} order the dpERK concentration profiles in time.
+We will use various dimensionality techniques to {\em automatically} order the dpERK concentration profiles in time.
+%
+We assume that the data is approximately one-dimensional, and that this one dimension parameterizes time, so that reducing the data to one-dimension and ordering it along this dimension will (approximately) order it in time.
 %
@@
 
-We first used Principal Components Analysis (PCA) to automatically order the one-dimensional concentration profiles in time. 
+We first used principal components analysis (PCA) to automatically order the one-dimensional concentration profiles in time. 
 %
@@
 %
-There is a strong correlation between the first projection coefficient, $\langle x_i, \psi_1 \rangle$, and the membrane thickness.
+There is a strong correlation between the first projection coefficient, $\langle x_i, \psi_1 \rangle$, and the membrane thickness;
+the Spearman correlation between the first projection coefficient and the membrane thickness is 0.9244.
 
-\begin{figure}[H]
+\begin{figure}[!ht]
 \begin{subfigure}{0.3\textwidth}
@@
 
-We would like to note that, because the embryos can be oriented with either the anterior or posterior end facing up in the microfluidic device (see Figure \ref{subfig:imaging_device}), the directionality of the concentration profiles (clockwise versus counterclockwise around the ring, or right-to-left versus left-to-right for the linear profiles) is not meaningful.
+We would like to note that, because the embryos can be oriented with either the anterior or posterior facing up in the microfluidic device (see Figure \ref{subfig:imaging_device}), the directionality of the concentration profiles (clockwise versus counterclockwise around the ring, or right-to-left versus left-to-right for the linear profiles) is not meaningful.
 %
-Therefore, we could also do PCA with an expanded data set where we also include the mirror image of each concentration profile.
+Therefore, we could do PCA with an expanded data set where we also include the mirror image of each concentration profile.
 %
@@
 %
-Perhaps the most obvious feature to note is the scrambling of the secondary *ADD NAME* peaks located at the far left and right edges of the concentration profiles.
+Perhaps the most obvious feature to note is the scrambling of the secondary peaks located at the far left and right edges of the concentration profiles.
+%TODO: add name of secondary peaks
 %
-It is known that these peaks arise later in the development of {\em Drosophila};
+It is known that these peaks arise later in the development of {\em Drosophila} \cite{...};
 however, one can see in Figure \ref{subfig:PCA_ordering_image} that these secondary peaks are scrambled when the data is ordered using PCA.
@@
 %
-The data falls onto a one-dimensional {\em nonlinear} curve in this two-dimensional space, implying that nonlinear dimensionality reduction techniques may be more approproate, as they can order the data along the curve, and not simply along the first principal axis.
+The data falls onto a one-dimensional {\em nonlinear} curve in this two-dimensional space, implying that nonlinear dimensionality reduction techniques may be more appropriate.
 
@@
 %
-The results are shown in Figure \ref{fig:DMAPS_ordering}.
+The results are shown in Figure \ref{fig:DMAPS_ordering}; the Spearman correlation between $\phi_2$ and the membrane thickness is 0.9209.
 %
@@
 
-\begin{figure}[H]
+\begin{figure}[!ht]
 \begin{subfigure}{0.3\textwidth}
@@
 
-As we have shown, we can order the concentration profiles using PCA or DMAPS.
+We have demonstrated that we can order the concentration profiles using PCA or DMAPS.
 %
@@
 %
-The developmental dynamics are invariant to the oritentation of the embryo in the microfluidic device; therefore, to obtain meaningful results from our analysis, we must first align the embryo snapshots.
+The developmental dynamics are invariant to the orientation of the embryo in the microfluidic device; therefore, to obtain meaningful results from our analysis, we must first align the embryo snapshots.
 %
-Rotation of the embryo cross-sections correspond to translation (with periodic boundary conditions) of the linear profiles we have been considering.
+Rotations of the embryo cross-sections correspond to translations (with periodic boundary conditions) of the linear profiles we have been studying.
 %
@@
 %
-However, without some {\em a priori} knowledge of the system (i.e.
-knowing that dorsal is expressed at one position along the perimeter of the embryo), we would need to align the concentration profiles before doing our analysis.
+However, without some {\em a priori} knowledge of the system (e.g.
+without knowing that dorsal is expressed at one position along the perimeter of the embryo), we would need to align the concentration profiles before doing our analysis.
+%TODO: add when this might be relevant, e.g. when there is no Dorsal
 
@@
 %
-Although these methods are simple, they require knowledge of an appropriate template function and can suffer when the data set is noisy.
+Although these methods are simple, they require knowledge of an appropriate template function and can suffer when the data set is noisy \cite{sonday2013noisy}.
 %
@@
 %
-One can see that the results are visually consistent with the results shown in Figure \ref{fig:DMAPS_ordering}.
+One can see that the results are visually consistent with the results shown in Figure \ref{fig:DMAPS_ordering}; the Spearman correlation between $\phi_2$ and the membrane length is 0.9164.
 
-\begin{figure}[H]
+\begin{figure}[!ht]
 \begin{subfigure}{0.3\textwidth}
@@
 
-Instead of using angular synchronization and DMAPS, we can do the alignment and dimensionality reduction/ordering in one computation using vector diffusion maps (VDM).
+Instead of using angular synchronization followed by DMAPS, we can do the alignment and ordering in one computation using vector diffusion maps (VDM).
 %
-VDM is similar to angular synchronization, but it is intended for data sets that contain both underlying symmetries and dynamics.
+VDM is similar to angular synchronization, but it operates on data sets that contain both underlying symmetries and dynamics.
 %
-The results of aligning {\em and} ordering the data using VDM are shown in Figure \ref{fig:vdm_ordering}; the results are similar to those from angular syncrhonization (Figure \ref{fig:angsynch_ordering}).
+The results of aligning {\em and} ordering the data using VDM are shown in Figure \ref{fig:vdm_ordering}; the results are similar to those from angular syncrhonization (Figure \ref{fig:angsynch_ordering}), and the Spearman correlation between $\langle \phi_3, \phi_1 \rangle$ and the membrane length is 0.9123.
 
-\begin{figure}[H]
+\begin{figure}[!ht]
 \begin{subfigure}{0.3\textwidth}
@@
 %
-We use VDM to align and order the fluorescent images stained for dpERK. 
+We used VDM to align and order the dpERK fluorescent images. 
 %
@@
 %
-To do this, we project the images onto a sphere;
-rotations and translations of a two-dimensional image then correspond to (three-dimensioanl) rotations of the sphere.
+We first projected the images onto a sphere;
+rotations and translations of a two-dimensional image then correspond to (three-dimensional) rotations of the sphere.
 %
-We then use VDM to align and order the (projected) images, with the underlying symmetry group $SO(3)$. 
+We then used VDM to align and order the (projected) images, with the underlying symmetry group $SO(3)$. 
 %
@@
 the orderings are of comparable quality to those obtained with the one-dimensional profiles, and the dpERK peaks are nicely aligned.
+%
+The Spearman correlation between $\langle \phi_4, \phi_1 \rangle$ and the membrane length is 0.9051.
 
-\begin{figure}[H]
+\def\imageindices{1,19,25,30,35,40,45,49}
+%\def\imageindices{1,9,17,25,30,34,38,43,49}
+
+\begin{figure}[!ht]
 \centering
@@
 \begin{subfigure}{\textwidth}
-\foreach \n in {1,9,17,25,30,34,38,43,49}{
-\includegraphics[width=0.07\textwidth, trim=6in 0in 6in 0in, clip]{dpERK_vdm_\n}
+\foreach \n in \imageindices{
+\includegraphics[width=0.1\textwidth]{dpERK_vdm_\n}
 \hfill}
@@
 (a) Correlation between the VDM embedding coordinate and the membrane thickness. 
-(b) Some representative fluorescent images, aligned  and ordered using the VDM embedding coordinate. }
+(b) Some representative fluorescent images, aligned  and ordered using the VDM embedding coordinate. The corresponding data points are circled in (a). }
 \label{fig:vdm_image_ordering}
@@
 
-An alternative to aligning the data (as in angular synchronization of VDM) is to construct {\em features} that are invariant to the underlying symmetry group, and then compute distances between data points using these features for use in a DMAPS calculation.
+An alternative to aligning the data (as in angular synchronization or VDM) is to construct {\em features} that are invariant to the underlying symmetry group, and then compute distances between features of the data for use in a DMAPS calculation.
 %
-In our setup, we need features for two-dimensional images that are invariant to translations and rotations of the images.
+In our setup, we require features of two-dimensional images that are invariant to translations and rotations of the images.
 %
-We chose to use the scattering transform \cite{mallat2012group} to construct our invariant features.
+We chose to use the scattering transform \cite{mallat2012group} to construct our invariant features; the scattering transform utilizes a series of wavelet transforms, which are known to perform well for images \cite{akansu2010emerging}.
 %
-Not only is this transform invariant to translations and rotations, but it is built on the wavelet transform (which is know to perform well for images), and is also contractive (so that noise and small deformations are not amplified by the transform).
+Not only is this transform invariant to translations and rotations, and is also contractive (so that noise and small deformations are not amplified by the transform) \cite{bruna2011classification}.
 
@@
 %
-We then perform a DMAPS calculation, using the Euclidean distance between the vectors of scattering transform coefficients as our distance metric.
+We then perform a DMAPS calculation, using the Euclidean distance between vectors of scattering transform coefficients as our distance metric.
 %
@@
 %
-The ordering is again good; however, using this method, we do not obtain optimal alignments for the images (we only obtain an ordering).
+The ordering is again good, with a Spearman correlation between $\phi_2$ and the membrane length of 0.9011.
+%
+However, we would like to note that using this method, we do not obtain optimal alignments for the images (we only obtain an ordering).
 
-\begin{figure}[H]
+\begin{figure}[!ht]
 \centering
 \begin{subfigure}{\textwidth}
+\centering
 \includegraphics[width=0.7\textwidth]{DMAPS_scat_time_corr}
@@
 \begin{subfigure}{\textwidth}
-\foreach \n in {1,9,17,25,30,34,38,43,49}{
-\includegraphics[width=0.07\textwidth, trim=6in 0in 6in 0in, clip]{dpERK_scat_\n}
+\foreach \n in \imageindices{
+\includegraphics[width=0.1\textwidth]{dpERK_scat_\n}
 \hfill}
@@
 The DMAPS embedding was computed using the translation- and rotation-invariant scattering transform coefficients as ``features'' of the images.
-(b) Some representative fluorescent images, ordered using the first (non-trivial) DMAPS embedding coordinate computed from the scattering transform coefficients. }
+(b) Some representative fluorescent images, ordered using the first (non-trivial) DMAPS embedding coordinate computed from the scattering transform coefficients.
+The corresponding data points are circled in (a).}
 \label{fig:scattrans_dpERK_ordering}
@@
 
-$\rho = 0.8403$
+Thus far, we have focused on ordering the dpERK fluorescent images. 
+%
+However, we used the membrane thickness in Figure~\ref{fig:background} to order the dpERK concentration profiles. 
+%
+Therefore, we could also use the techniques discussed above to order the two-dimensional {\em membrane} images.
+%
+Because we are only interested in the membrane thickness, and not other features of the images (e.g. fluorescent intensity), we preprocess the images using an edge detection filter.
+%
+Some of these images are shown in Figure~\ref{fig:scattrans_membrane_ordering}; the green fluorescent images are those collected from the confocal microscope, and the grayscale images are the result of applying the edge-detection algorithm.
 
-$\rho = 0.3525$ for VDM + membrane pics
+Because these images are much more complex than the dpERK fluorescent images, we expect any alignment results to be poor.
+%
+Using VDM to align and order the images yielded a Spearman correlation coefficient of 0.3525.
+%
+However, the scattering transform is robust to small deformations, and so we expect using DMAPS with the scattering transform coefficients to give better results.
+%
+We computed the scattering transform coeffficients of the edge-detected images, and used DMAPS to order the images in time. 
+%
+The results are shown in Figure~\ref{fig:scattrans_membrane_ordering}.
+%
+The correlation between $\phi_2$ and the membrane thickness is not as strong as the correlations for the dpERK images; the Spearman correlation is 0.8403.
+%
+However, these results are still promising considering the relative simplicity of the methods. 
+%
+%$\rho = 0.8403$, 300 seconds for scat + DMAPS
+%$\rho = 0.3525$ for VDM + membrane pics, 1100 seconds
 
-\begin{figure}[H]
+\begin{figure}[!ht]
 \centering
@@
 \begin{subfigure}{\textwidth}
-\foreach \n in {1,9,17,25,30,34,38,43,49}{
-\includegraphics[width=0.07\textwidth, trim=6in 0in 6in 0in, clip]{membrane_scat_\n}
+\foreach \n in \imageindices{
+\includegraphics[width=0.1\textwidth]{membrane_scat_\n}
 \hfill}
 \newline
-\foreach \n in {1,9,17,25,30,34,38,43,49}{
-\includegraphics[width=0.07\textwidth, trim=6in 0in 6in 0in, clip]{membrane_scat_raw_\n}
+\foreach \n in \imageindices{
+\includegraphics[width=0.1\textwidth]{membrane_scat_raw_\n}
 \hfill}
@@
 The first six non-trivial DMAPS coordinates were found to capture other features of the images, such as  differences in fluorescence intensity within the images.
-(b) Some representative fluorescent images, ordered using the seventh (non-trivial) DMAPS embedding coordinate computed from the scattering transform coefficients. }
+(b) Some representative edge-detected images (top) and fluorescent images (bottom), ordered using the first (non-trivial) DMAPS embedding coordinate computed from the scattering transform coefficients.
+The corresponding data points are circled in (a).}
 \label{fig:scattrans_membrane_ordering}
@@
 
-We hope that we have illustrated the use of dimensionality reduction techniques in application with underlying symmetries.
+We hope that we have demonstrated the utility of dimensionality reduction techniques in dynamical systems with underlying symmetries.
 %
 The work here is intended to provide an overview of the different techniques available, as well as the benefits and shortcomings of each method. 
+%
+The techniques here were presented from least complex (PCA) to most complex (scattering transforms).
+%
+More complex techniques allowed us to deal with more complex data, and allowed us to circumvent much of the preprocessing described in Figure~\ref{fig:background}.
 
-\begin{table}
-	\begin{tabular}{| p{0.3\textwidth} | c | c | p{0.3\textwidth} |}
-		\hline 
-		Method & Spearman's $\rho$ & Execution Time & Notes About Method \\ 
-		\hline 
-		PCA on one-dimensional profiles & 0.9244 & $< 1$ second & \\
-		\hline 
-		DMAPS on one-dimensional profiles & 0.9209 & $< 1$ second & \\
-		\hline 
-		Angular synchronization + DMAPS on one-dimensional profiles & 0.9164 & $< 1$ second & \\
-		\hline 
-		VDM on one-dimensional profiles & 0.9232 & $< 1$ second & \\	
-		\hline 
-		VDM on two-dimensional images & 0.9051  & $325$ seconds & $\mathcal{O}(n^2)$ to compute the pairwise alignments\\
-		\hline 	
-		Scattering transform + DMAPS on two-dimensional profiles & 0.9011  & $86$ seconds &  $\mathcal{O}(n)$ to compute the scattering transform for each data point \\
-		\hline
-		Scattering transform + DMAPS on two-dimensional membrane images & 0.8403  & $300$ seconds &  $\mathcal{O}(n)$ to compute the scattering transform for each data point \\
-		\hline
-	\end{tabular}
-\end{table}
+The more complex techniques that operate on two-dimensional images do come at a higher computational cost. 
+%
+The computational time for analyzing the one-dimensional profiles using any of the techniques described (PCA, DMAPS, angular synchronization, VDM) is less than one second. 
+%
+In contrast, the analysis of two-dimensional images using either VDM or the scattering transform with DMAPS is $\mathcal{O}(100)$ seconds.
+%
+The expensive portion of the VDM calculation is computing the pairwise alignments of the images; the expensive portion of the scattering transform with DMAPS is computing the scattering transform coefficients.
+%
+However, computing the pairwise alignments is $\mathcal{O}(n^2)$, where $n$ is the number of data points, while computing the scattering transform coefficients is $\mathcal{O}(n)$.
+%
+We therefore expect for VDM to be computationally intractable for large data sets.
+%
+The added computational cost of VDM is not without benefits, however; VDM produces optimal alignments for each image (in addition to an ordering), while the scattering transform gives us no alignments.
+
+We would like to note that our system was such that the data was inherently one-dimensional, and that this one dimension was monotonic in time.
+%
+However, this monotonicity was a special feature of our system and is not guaranteed in general.
+
+%\begin{table}
+%	\begin{tabular}{| p{0.3\textwidth} | c | c | p{0.3\textwidth} |}
+%		\hline 
+%		Method & Spearman's $\rho$ & Execution Time & Notes About Method \\ 
+%		\hline 
+%		PCA on one-dimensional profiles & 0.9244 & $< 1$ second & \\
+%		\hline 
+%		DMAPS on one-dimensional profiles & 0.9209 & $< 1$ second & \\
+%		\hline 
+%		Angular synchronization + DMAPS on one-dimensional profiles & 0.9164 & $< 1$ second & \\
+%		\hline 
+%		VDM on one-dimensional profiles & 0.9232 & $< 1$ second & \\	
+%		\hline 
+%		VDM on two-dimensional images & 0.9051  & $325$ seconds & $\mathcal{O}(n^2)$ to compute the pairwise alignments\\
+%		\hline 	
+%		Scattering transform + DMAPS on two-dimensional profiles & 0.9011  & $86$ seconds &  $\mathcal{O}(n)$ to compute the scattering transform for each data point \\
+%		\hline
+%		Scattering transform + DMAPS on two-dimensional membrane images & 0.8403  & $300$ seconds &  $\mathcal{O}(n)$ to compute the scattering transform for each data point \\
+%		\hline
+%	\end{tabular}
+%\end{table}
 % You may title this section "Methods" or "Models".
@@
 
-\subsection*{Experimental Setup}
+\subsection*{Experimental setup}
 
-\subsection*{Image Preprocessing}
-The dpERK images we collect from the confocal microscope are subsampled to $100 \times 100$ pixels; the membrane images are subsampled to $200 \times 200$ pixels (due to the finer structure in the membrane pictures).
+\subsection*{Image preprocessing}
+The dpERK images we collected from the confocal microscope were subsampled to $100 \times 100$ pixels; 
+the membrane images were subsampled to $200 \times 200$ pixels (due to the finer structure in the membrane pictures). 
+%
+Because we are only interested in the membrane thickness, we ran a simple edge detection algorithm on the membrane images.
+%
+We used a Laplacian of Gaussians filter \cite{canny1986computational}, as implemented in the Matlab Image processing toolbox via the \texttt{edge} function (\url{http://www.mathworks.com/help/images/ref/edge.html}).
+%
+To compute the pairwise alignments between images for use in angular synchronization and VDM, we searched over 20 rotations (evenly spaced from $0$ to $2 \pi$) and 10 translations spanning $\pm 10 \%$ of the image pixels.
+%
+We used the Euclidean distance between images to measure the quality of the alignment.
+%
+These translations and rotations were then converted to three-dimensional rotations on the sphere, with the image spanning a $\pi/4 \times \pi/4$ portion of the sphere, so that the angle of rotation of the image corresponded to the Euler angle $\alpha$, $pixels_x/pixels_{max} * \pi/4$ corresponded to the Euler angle $\beta$, and  $pixels_y/pixels_{max} * \pi/4$ corresponded to the Euler angle $\gamma$.
 
-Because we are only interested in the membrane thickness, we run a simple edge detection algorithm on the membrane images.
+\subsection*{Spearman's rank correlation coefficient}
+
+Spearman's rank correlation coefficient, or Spearman's $\rho$, measures the monotonicity of a function \cite{myers2010research}. 
 %
-We use a Laplacian of Gaussians filter \cite{...}, as implemented in the Matlab Image processing toolbox \cite{...}.
+It therefore serves as an appropriate metric for the ``goodness'' of our orderings.
+%
+Given data points $X_i$ and $Y_i$, we compute the ranks of each data point within the data set, denoted $x_i$ and $y_i$.
+%
+Spearman's $\rho$ is then defined as the correlation between the ranks
+\begin{equation}
+\rho = \frac{\sum_i (x_i - \overline{x})(y_i - \overline{y})}{\sqrt{\sum_i (x_i - \overline{x})^2 \sum_i (y_i - \overline{y})^2}}
+\end{equation}
+where $\overline{x}$ and $\overline{y}$ are the averages of $x_i$ and $y_i$, respectively.
 
-\subsection*{Principal Component Analysis}
+\subsection*{Principal component analysis}
 Principal component analysis (PCA) has been the workhorse dimensionality reduction technique for the past century.
@@
 
-\subsection*{Diffusion Maps}
+\subsection*{Diffusion maps}
 Unlike PCA, diffusion maps (DMAPS) is a nonlinear dimensionality reduction technique. 
@@
 
-\subsection*{Angular Synchronization}
+\subsection*{Angular synchronization}
 
@@
 %
-We will consider the case where the symmetry group is the group of two-dimensional rotations, $SO(2)$; the same construction and ideas hold for other symmetry groups, such as $SO(3)$. 
+We will consider the case where the symmetry group is the group of $d$-dimensional rotations, $SO(d)$. 
 %
-Let $R_{ij} \in SO(2)$ denote the rotation that aligns $x_j$ to $x_i$, so that
+Let $R_{ij} \in SO(d)$ denote the rotation that aligns $x_j$ to $x_i$, so that
 \begin{equation}
-R_{ij} = \argmin_{R \in SO(2)} \|Rx_j - x_i \|^2.
+R_{ij} = \argmin_{R \in SO(d)} \|Rx_j - x_i \|^2.
 \end{equation}
 %
-We seek to find the rotations $R_1, R_2, \dots, R_n \in SO(2)$ such that $R_i R_j^T \approx R_{ij}$, for every pair $i, j$. (CHECK THIS)
+We seek to find the rotations $R_1, R_2, \dots, R_n \in SO(d)$ such that $R_i R_j^T \approx R_{ij}$, for every pair $i, j$. 
+%TODO: Check that this order is correct
 %
-We would also like to exploit {\em higher-order} consistency relations.
+We would also like to exploit {\em higher-order} consistency relations;
+for example, the relationship $R_{ij} R_{jk} R_{ki} = I$, where $I$ is the identity matrix, should hold.
 %
-For example, for noiseless measurements, $R_{ij} R_{jk} R_{ki} = I$, where $I$ is the identity matrix; therefore, we consider measurements which (almost) satisfy such conditions ``good'' measurements, and those measurements which do not satisfy such conditions as most likely inaccurate.
+Therefore, we consider measurements which (almost) satisfy such conditions ``good'' measurements, and those measurements which do not satisfy such conditions as most likely inaccurate.
 
@@
 \begin{equation} \label{eqn:angsynch_obj}
-\max_{R_1, \dots, R_n \in SO(2)} \sum_{i=1}^{n} \sum_{j=1}^{n} R_i^T R_{ij} R_j.
+\max_{R_1, \dots, R_n \in SO(d)} \sum_{i=1}^{n} \sum_{j=1}^{n} R_i^T R_{ij} R_j.
 \end{equation}
 %
+%TODO: check if a norm is needed
+This objective function is large if $R_i$ adn $R_j$ are consistent with $R_{ij}$, and will be random with mean 0 if $R_i$ and $R_j$ are inconsistent with $R_{ij}$.
+%
+Therefore, we are seeking the most consistent set of global alignments, given the computed pairwise alignments.
+
 In general, the solution to \eqref{eqn:angsynch_obj} is not easily computed.
+%
+Instead, we relax the problem and allow $R_1, \dots, R_n \in \mathbb{R}^{d \times d}$.
+%
+We then define the matrix $H$, which is an $n \times n$ block matrix with $d \times d$ blocks (so $H \in \mathbb{R}^{nd \times nd}$).
+%
+We then compute the top ``block'' eigenvector of $H$, which we denote $\hat{R} \in \mathbb{R}^{nd \times d}$. 
+%
+The solution to the relaxed problem is given by $\hat{R}$, so that the $i^{th}$ rotation is (approximately) $\hat{R}(i) \in \mathbb{R}^{d \times d}$.
+%
+This formulation also accounts for higher-order relations.
+%
+For example, if we want to optimize over all pairs of pairwise rotations, we would solve $\max_{R_1, \dots, R_n \in SO(d)} \sum_{i=1}^{n} \sum_{j=1}^{n} R_i^T \left( \sum_k R_{ik} R_{kj} \right) R_j$.
+%
+However, the solution to this problem is also given by the top eigenvector of $H$.
 
-Instead, we relax the problem and allow $R_1, \dots, R_n \in \mathbb{R}^2$.
-%
-We then define the matrix $H$, which is an $n \times n$ block matrix with $2 \times 2$ blocks (so $H \in \mathbb{R}^{2n \times 2n}$).
-%
-We then compute the top ``block'' eigenvector of $H$, which we denote $\hat{R} \in \mathbb{R}^{2n \times 2}$. 
-%
-The solution to the relaxed problem is given by $\hat{R}$, so that the $i^{th}$ rotation is (approximately) $\hat{R}(i) \in \mathbb{R}^{2 \times 2}$.
-%
-However, because we relaxed the problem to allow our solutions to lie in $\mathbb{R}^{2 \times 2}$, we must project our approximate solution back to $SO(2)$.
+Because we relaxed the problem to allow our solutions to lie in $\mathbb{R}^{d \times d}$, we must project our approximate solution back to $SO(d)$.
 %
@@
 
-\subsection*{Vector Diffusion Maps}
+\subsection*{Vector diffusion maps}
 
@@
 A_{ij} H_{ij} & i \ne j \\
-0_{2 \times 2} & i = j
+0_{d \times d} & i = j
 \end{array}
@@
 
-We then compute the eigenvectors $\phi_1, \dots, \phi_{2n}$ and eigenvalues $\lambda_1, \dots, \lambda_{2n}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{2n}|$.
+We then compute the eigenvectors $\phi_1, \dots, \phi_{nd}$ and eigenvalues $\lambda_1, \dots, \lambda_{nd}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{nd}|$.
 %
-As in angular synchronization, the first ``block'' eigenvector, i.e. the first two eigenvectors concatenated into a $2 n \times 2$ matrix, gives the optimal rotations. 
+As in angular synchronization, the first ``block'' eigenvector, i.e. the first two eigenvectors concatenated into a $nd \times d$ matrix, gives the optimal rotations. 
 %
@@
 %
-The embedding coordinates are then given by $\langle \phi_j(i), \phi_k(i) \rangle$, where $1 \le j, k, \le 2 n$.
+The embedding coordinates are then given by $\langle \phi_j(i), \phi_k(i) \rangle$, where $1 \le j, k, \le nd$.
 
-\subsection*{Scattering Transform}
+\subsection*{Scattering transform}
 
@@
 %
-We utilized the freely available ScatNet software (\texttt{http://www.di.ens.fr/data/software/scatnet/}) to compute the rotation-invariant scattering coefficients.
-
-\subsection*{Spearman's $\rho$}
-
-Spearman's rank correlation coefficient, or Spearman's $\rho$, measures the monotonicity of a function \cite{...}. 
-%
-It therefore serves as an appropriate metric for the ``goodness'' of our orderings.
-%
-Given data points $X_i$ and $Y_i$, we compute the ranks of each data point within the data set, denoted $x_i$ and $y_i$.
-%
-Spearman's $\rho$ is then defined as the correlation between the ranks
-\begin{equation}
-\rho = \frac{\sum_i (x_i - \overline{x})(y_i - \overline{y})}{\sqrt{\sum_i (x_i - \overline{x})^2 \sum_i (y_i - \overline{y})^2}}
-\end{equation}
-where $\overline{x}$ and $\overline{y}$ are the averages of $x_i$ and $y_i$, respectively.
+We utilized the freely available ScatNet software (\url{http://www.di.ens.fr/data/software/scatnet/}) to compute the rotation-invariant scattering coefficients.
 
@@
 
-\section*{Figure Legends}
+%\section*{Figure Legends}
 %\begin{figure}[!ht]
@@
 
-\section*{Tables}
+%\section*{Tables}
 %\begin{table}[!ht]

### 2014-01-14 image_analysis_paper/plos_draft.tex
Added more edits to paper

@@
 
-The issue of rotation invariance arises in many applications, including medical imaging \cite{hajnal2010medical}, face recognition \cite{rowley1998rotation}, digit recognition \cite{simard1992efficient}, and texture classification \cite{greenspan1994rotation}.
+The issue of rotation invariance arises in many applications, including face recognition \cite{rowley1998rotation}, medical imaging \cite{hajnal2010medical}, digit recognition \cite{simard1992efficient}, and texture classification \cite{greenspan1994rotation}.
 %
-In these settings, objects from different frames or timepoints must be aligned or ``registered'' prior to further analysis.
+In these settings, images from different timepoints, users, or contexts must be aligned or ``registered'' prior to further analysis.
 %
@@
 %
-``Template-based'' techniques, where one selects a prototypical image or ``template'' and then aligns the remaining images to this template, are relatively common \cite{ahuja2007template}.
+Template-based techniques, where one selects a prototypical image or ``template'' and then aligns the remaining images to this template, are relatively common \cite{ahuja2007template}.
 %
@@
 %
-%TODO: why do these methods fail?
+While these techniques are often robust to additive Gaussian noise, they are often sensitive to deformations and other types of noise typically encountered in images.
 
@@
 %
-More specifically, we would like to order the images {\em in time} so that we can reconstruct the developmental dynamics from snapshots. 
+More specifically, we are interested in the activation dynamics of dpERK during NC14.
+%
+We collect fluorescent images of {\em Drosophila} embryos that have been stained for dpERK. 
+%
+Because we must fix the embryos before staining for dpERK, we cannot obtain a continuous time trajectory of dpERK in one embryo; instead, we obtain a set of snapshots, one per embryo, of embryos stained for dpERK.
+% 
+We would like to order the resulting images {\em in time} so that we can reconstruct the developmental dynamics from snapshots. 
 %
@@
 %
+%TODO: add why we cannot always do the procedure outlined above
 We will use various dimensionality techniques to {\em automatically} order the dpERK concentration profiles in time.
 %
-We assume that the data is approximately one-dimensional, and that this one dimension parameterizes time, so that reducing the data to one-dimension and ordering it along this dimension will (approximately) order it in time.
+We assume that the data is approximately one-dimensional, and that this one dimension parameterizes time, so that reducing the data to one dimension and ordering it along this dimension will (approximately) order it in time.
 %
@@
 
-We first used principal components analysis (PCA) to automatically order the one-dimensional concentration profiles in time. 
+We first used principal components analysis (PCA) to order the one-dimensional concentration profiles in time. 
 %
-We computed the first principal component, $\psi_1$, from the data set, and ordered the data by the projection coefficients onto $\psi_1$.
+We computed the first principal component, $\psi_1$, from the data set of concentration profiles, and ordered the data by the projection coefficients onto $\psi_1$.
 %
@@
 %
-We found that including the mirror images of each data point did not change our results significantly, since the first principal component (not shown) is nearly symmetric with respect to left-right reflections.
+We found that including the mirror images of each data point did not change our results significantly, since the first principal component computed using the original data set (not shown) is nearly symmetric with respect to left-right reflections.
 
@@
 %
-The developmental dynamics are invariant to the orientation of the embryo in the microfluidic device; therefore, to obtain meaningful results from our analysis, we must first align the embryo snapshots.
+The developmental dynamics are invariant to the orientation of the embryo in the microfluidic device; therefore, to obtain meaningful results from our analysis, we must first align the concentration profiles.
 %
@@
 %
-We used angular synchronization to factor out the shifts in the concentration profiles measured from the fluorescent images.
+We used angular synchronization to factor out the shifts and align the concentration profiles measured from the fluorescent images.
 %
-After aligning the data points using angular synchronization, we used DMAPS to order the aligned concentration profiles.
+Because the concentration profiles lie on a circle (we unwrap them onto a line for visualization purposes), factoring out shifts of the linear profiles (with periodic boundary conditions) corresponds to factoring out rotations of a circle with the underlying symmetry group $SO(2)$.
+%
+After alignment, we used DMAPS to order the aligned concentration profiles.
 %
@@
 
-Instead of using angular synchronization followed by DMAPS, we can do the alignment and ordering in one computation using vector diffusion maps (VDM).
+Instead of using angular synchronization followed by DMAPS, one can do the alignment and ordering in one computation using vector diffusion maps (VDM).
 %
@@
 %
-We then used VDM to align and order the (projected) images, with the underlying symmetry group $SO(3)$. 
+We used VDM to align and order the (projected) images, with the underlying symmetry group $SO(3)$. 
 %
@@
 %
-Not only is this transform invariant to translations and rotations, and is also contractive (so that noise and small deformations are not amplified by the transform) \cite{bruna2011classification}.
+Not only is this transform invariant to translations and rotations, but is also a contractive operator \cite{bruna2011classification}, so that noise and small deformations are not amplified by the transform.
 
@@
 
-We hope that we have demonstrated the utility of dimensionality reduction techniques in dynamical systems with underlying symmetries.
+We hope we have demonstrated the utility of dimensionality reduction techniques in systems with underlying symmetries and dynamics.
 %
@@
 %
-The techniques here were presented from least complex (PCA) to most complex (scattering transforms).
+The techniques were presented from least complex (PCA) to most complex (scattering transforms).
 %
@@
 %
-The expensive portion of the VDM calculation is computing the pairwise alignments of the images; the expensive portion of the scattering transform with DMAPS is computing the scattering transform coefficients.
+The expensive portion of the VDM calculation is computing the pairwise alignments of the images, and the expensive portion of the scattering transform with DMAPS is computing the scattering transform coefficients.
 %
@@
 %
-The added computational cost of VDM is not without benefits, however; VDM produces optimal alignments for each image (in addition to an ordering), while the scattering transform gives us no alignments.
+The added computational cost of VDM is not without benefits; VDM produces optimal alignments for each image (in addition to an ordering), while the scattering transform gives us no alignments.
 
@@
 However, this monotonicity was a special feature of our system and is not guaranteed in general.
+%TODO: add more conclusions
 
@@
 %
-Because we are only interested in the membrane thickness, we ran a simple edge detection algorithm on the membrane images.
+Because we were only interested in the membrane thickness, we ran a simple edge detection algorithm on the membrane images.
 %
@@
 %
-These translations and rotations were then converted to three-dimensional rotations on the sphere, with the image spanning a $\pi/4 \times \pi/4$ portion of the sphere, so that the angle of rotation of the image corresponded to the Euler angle $\alpha$, $pixels_x/pixels_{max} * \pi/4$ corresponded to the Euler angle $\beta$, and  $pixels_y/pixels_{max} * \pi/4$ corresponded to the Euler angle $\gamma$.
+These translations and rotations were then converted to three-dimensional rotations on the sphere, with the image spanning a $\pi/4 \times \pi/4$ portion of the sphere.%, so that the angle of rotation of the image corresponded to the Euler angle $\alpha$, $pixels_x/pixels_{max} * \pi/4$ corresponded to the Euler angle $\beta$, and  $pixels_y/pixels_{max} * \pi/4$ corresponded to the Euler angle $\gamma$.
 
@@
 %
-It therefore serves as an appropriate metric for the ``goodness'' of our orderings.
+It therefore serves as an appropriate metric for the quality of our orderings.
 %
-Given data points $X_i$ and $Y_i$, we compute the ranks of each data point within the data set, denoted $x_i$ and $y_i$.
+Given two data sets $X$ and $Y$, with data points $X_i$ and $Y_i$, we compute the ranks of each data point within the data set, denoted $x_i$ and $y_i$.
 %
-Spearman's $\rho$ is then defined as the correlation between the ranks
+Spearman's rank correlation between $X$ and $Y$ is then defined as the correlation between the ranks
 \begin{equation}
@@
 
-Let $X \in \mathbb{R}^{n \times d}$ denote our data set consisting of $n$ points in $d$ dimensions, where $X_{ij}$ denotes the $j^{th}$ dimension of the $i^{th}$ observation. 
+Let $X \in \mathbb{R}^{n \times d}$ denote our data set, consisting of $n$ points in $d$ dimensions, where $X_{ij}$ denotes the $j^{th}$ dimension of the $i^{th}$ observation. 
 %
@@
 %
-DMAPS aims to uncover the parameterization of data sampled from a nonlinear manifold \cite{coifman2005geometric}.
+DMAPS aims to uncover a parameterization of high-dimensional data sampled from a low-dimensional nonlinear manifold \cite{coifman2005geometric}.
 %
@@
 
-When the data set is invariant under an underlying symmetry group (such as translations and/or rotations), we must first factor out the underlying symmetries before doing further analysis.
+When the data set is invariant under an underlying symmetry group (such as rotations), we must first factor out the underlying symmetries before doing further analysis.
 %

### 2014-01-14 image_analysis_paper/plos_draft.tex
Updated draft

@@
 %
-More specifically, we are interested in the activation dynamics of dpERK during NC14.
+More specifically, we are interested in the activation dynamics of the ERK pathway during NC14, which we monitor via the production of doubly phosphorylated ERK (dpERK).
 %
@@
 %
-We then use image analysis software to extract the dpERK concentration profile around the circumfrence of the embryo from the fluorescent image.
+We then use image analysis software to extract the dpERK concentration profile around the circumference of the embryo from the fluorescent image.
 %
@@
 \end{center}
-\caption{{\bf Data collection.} (a) Schematic of {\em Drosophila} embryo (longitudinal and cross-section), and cross-sectional fluorescent images of an embryo stained for the membrane proteins (left), dpERK (center), and dorsal protein (right).
-(b) Concentration profile of dpERK extracted from a fluorescent image.
+\caption{{\bf Data collection.} (a) Schematic of microfluidic device used for imaging.
+Embryos are flown through the channels of the device, and one embryo is caught in each ``trap''.
+The device can then be placed in the confocal microscope for imaging.
+(b) Schematic of {\em Drosophila} embryo (longitudinal and cross-section), and cross-sectional fluorescent images of an embryo stained for the membrane proteins (left), dpERK (center), and dorsal protein (right).
+(c) Concentration profile of dpERK extracted from a fluorescent image.
 The circular profile is ``unwrapped'' at the dorsal midline (determined from the expression of the dorsal protein) to obtain a concentration profile on a line.
-(c) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time.
-(d) Concentration profiles of dpERK for many embryos from (c), now ordered by the thickness of the membrane. The membrane thickness is known to be one-to-one with time.}
+(d) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time.
+(e) Concentration profiles of dpERK for many embryos from (d), now ordered by the thickness of the membrane. The membrane thickness is known to be one-to-one with time.}
 \label{fig:background}
@@
 \caption{{\bf Ordering dpERK concentration profiles using DMAPS.} (a) Projection of data shown in Figure \ref{subfig:unordered_profiles} onto the first two principal components; note that the data falls on an (effectively) one-dimensional, nonlinear curve.
-(b) Concentration profiles of dpERK from (a), now ordered by the first non-trivial DMAPS embedding coordinate.
+(b) Concentration profiles of dpERK from \ref{subfig:unordered_profiles}, now ordered by the first non-trivial DMAPS embedding coordinate.
 (c) Correlation between the first (non-trivial) DMAPS embedding coordinate and the membrane thickness.}
@@
 Each row represents a different embryo fixed at a slightly different developmental time.
-The circluar profiles have not been ``opened'' at the same point; rather, each profile was opened at a random point around the embryo.
-We are allowed to shift the (linear) profiles left and right.
+The circular profiles have not been ``opened'' at the same point; rather, each profile was opened at a random point around the embryo.
+We are allowed to shift the (linear) profiles left and right (this corresponds to rotating the circular profile).
 (b) Concentration profiles of dpERK aligned using angular synchronization.
@@
 Each row represents a different embryo fixed at a slightly different developmental time.
-The circluar profiles have not been ``opened'' at the same point; rather, each profile was opened at a random point around the embryo.
+The circular profiles have not been ``opened'' at the same point; rather, each profile was opened at a random point around the embryo.
 We are allowed to shift the (linear) profiles left and right.
@@
 \caption{{\bf Ordering two-dimensional fluorescent images of dpERK.}
-(a) Correlation between the VDM embedding coordinate and the membrane thickness. 
+(a) Correlation between the VDM embedding coordinate calculated from two-dimensional fluorescent images and the membrane thickness. 
 (b) Some representative fluorescent images, aligned  and ordered using the VDM embedding coordinate. The corresponding data points are circled in (a). }
@@
 %
-We computed the scattering transform coeffficients of the edge-detected images, and used DMAPS to order the images in time. 
+We computed the scattering transform coefficients of the edge-detected images, and used DMAPS to order the images in time. 
 %
@@
 \caption{{\bf Ordering two-dimensional fluorescent images of membrane proteins.}
-(a) Correlation between the seventh (non-trivial) DMAPS embedding coordinate and the membrane thickness.
-DMAPS was computed using the translation- and rotation-invariant scattering transform coefficients as ``features'' of the images.
-The first six non-trivial DMAPS coordinates were found to capture other features of the images, such as  differences in fluorescence intensity within the images.
+(a) Correlation between the first (non-trivial) DMAPS embedding coordinate and the membrane thickness.
+DMAPS was computed using the translation- and rotation-invariant scattering transform coefficients of the edge-detected images as ``features'' of the images.
 (b) Some representative edge-detected images (top) and fluorescent images (bottom), ordered using the first (non-trivial) DMAPS embedding coordinate computed from the scattering transform coefficients.
@@
 
-Let $X \in \mathbb{R}^{n \times d}$ denote our data set, consisting of $n$ points in $d$ dimensions, where $X_{ij}$ denotes the $j^{th}$ dimension of the $i^{th}$ observation. 
+Let $X \in \mathbb{R}^{n \times p}$ denote our data set, consisting of $n$ points in $p$ dimensions, where $X_{ij}$ denotes the $j^{th}$ dimension of the $i^{th}$ observation. 
 %
@@
 %
-The empirical covariance of the data set, $C \in \mathbb{R}^{d \times d}$ is then given by
+The empirical covariance of the data set, $C \in \mathbb{R}^{p \times p}$ is then given by
 \begin{equation}
-C = X^T X.
+C = \frac{1}{n-1} X^T X.
 \end{equation}
 %
-We compute the eigenvectors $\psi_1, \dots, \psi_d$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_d$ of $C$, and order them such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_d$.
+We compute the eigenvectors $\psi_1, \dots, \psi_p$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_p$ of $C$, and order them such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_p$.
 %
-The principal components of the data set $X$ are given by $\psi_1, \dots, \psi_d$, and the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_d$ measure the variability captured by the corresponding eigenvector.
+The principal components of the data set $X$ are given by $\psi_1, \dots, \psi_p$, and the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_p$ measure the variability captured by projecting the data onto the corresponding eigenvector.
 %
-If there is a {\em spectral gap} such that the first few eigenvectors capture a significant portion of the variance in our data set, projection onto these first few principal components will retain most features of the data set while reducing the dimensionality.
+If there is a {\em spectral gap} such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_d \gg \lambda_{d+1} \ge \dots \ge \lambda_p$, then projecting the data onto the first $d$ eigenvectors will retain most of the variance of the data set, while reducing the dimensionality from $p$ to $d$.
 %
-In our setup, we assume that the data is inherently one-dimensional, and that projection onto $\psi_1$ will uncover the correct time ordering of the data.
+In our setup, we assume that the data is inherently one-dimensional ($d=1$), and that projection onto $\psi_1$ will uncover the correct time ordering of the data.
 
@@
 
-Given data points $x_1, \dots, x_n \in \mathbb{R}^d$, we fist calculate the matrix $W \in \mathbb{R}^{n \times n}$, where 
+Given data points $x_1, \dots, x_n \in \mathbb{R}^p$, we fist calculate the matrix $W \in \mathbb{R}^{n \times n}$, where 
 \begin{equation}
-W_{ij} = \exp \left( -\frac{d^2(x_i, x)j)}{\epsilon} \right)
+W_{ij} = \exp \left( -\frac{d^2(x_i, x)j)}{\epsilon^2} \right)
 \end{equation}
@@
 %
-Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector; the next few eigenvectors give the ``meaningful'' embedding coordinates for the data.
+Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector.
+%
+The next few eigenvectors give the ``meaningful'' embedding coordinates for the data.
 
@@
 
-When the data set is invariant under an underlying symmetry group (such as rotations), we must first factor out the underlying symmetries before doing further analysis.
+When a data set is invariant under an underlying symmetry group (such as rotations), we must factor out the underlying symmetries before doing further analysis.
 %
-Angular synchronization calculates the optimal alignments for a data set, assuming we have estimates of the optimal alignments between pairs of data points \cite{singer2011angular}.
+Angular synchronization calculates the optimal alignments for each point in a data set, given estimates of the optimal alignments between pairs of data points \cite{singer2011angular}.
 
@@
 We would also like to exploit {\em higher-order} consistency relations;
-for example, the relationship $R_{ij} R_{jk} R_{ki} = I$, where $I$ is the identity matrix, should hold.
+for example, the relationship $R_{ij} R_{jk} \approx R_{ik}$ should hold if the rotation estimates are accurate.
 %
@@
 
-We define the objective function 
+We would like to solve the optimization problem 
 \begin{equation} \label{eqn:angsynch_obj}
-\max_{R_1, \dots, R_n \in SO(d)} \sum_{i=1}^{n} \sum_{j=1}^{n} R_i^T R_{ij} R_j.
+\max_{R_1, \dots, R_n \in SO(d)} \left\| \sum_{i=1}^{n} \sum_{j=1}^{n} R_i^T R_{ij} R_j \right\|^2.
 \end{equation}
@@
 %TODO: check if a norm is needed
-This objective function is large if $R_i$ adn $R_j$ are consistent with $R_{ij}$, and will be random with mean 0 if $R_i$ and $R_j$ are inconsistent with $R_{ij}$.
+Each term in the objective function will be large large if $R_i$ and $R_j$ are consistent with $R_{ij}$, and will be random with mean 0 if $R_i$ and $R_j$ are inconsistent with $R_{ij}$.
 %
-Therefore, we are seeking the most consistent set of global alignments, given the computed pairwise alignments.
+Therefore, solving \eqref{eqn:angsynch_obj} gives us the set of global alignments which are most consistent with the computed pairwise alignments.
 
@@
 %
-We then define the matrix $H$, which is an $n \times n$ block matrix with $d \times d$ blocks (so $H \in \mathbb{R}^{nd \times nd}$).
+We define the matrix $H$, which is an $n \times n$ block matrix with $d \times d$ blocks (so $H \in \mathbb{R}^{nd \times nd}$), with $H_{ij} = R_{ij}$.
 %
-We then compute the top ``block'' eigenvector of $H$, which we denote $\hat{R} \in \mathbb{R}^{nd \times d}$. 
+We can then write the relaxed problem as 
+\begin{equation} \label{eqn:angsynch_relax}
+\max_{R\in \mathbb{nd \times d}} \| R^T H R \|^2.
+\end{equation}
 %
-The solution to the relaxed problem is given by $\hat{R}$, so that the $i^{th}$ rotation is (approximately) $\hat{R}(i) \in \mathbb{R}^{d \times d}$.
+The solution to \eqref{eqn:angsynch_relax} is given by the top ``block'' eigenvector of $H$, which we denote $\hat{R} \in \mathbb{R}^{nd \times d}$. 
 %
-This formulation also accounts for higher-order relations.
+The $i^{th}$ rotation is (approximately) $\hat{R}(i) \in \mathbb{R}^{d \times d}$.
 %
-For example, if we want to optimize over all pairs of pairwise rotations, we would solve $\max_{R_1, \dots, R_n \in SO(d)} \sum_{i=1}^{n} \sum_{j=1}^{n} R_i^T \left( \sum_k R_{ik} R_{kj} \right) R_j$.
+We would like to note that this formulation also accounts for higher-order relations.
+%
+For example, if we want to optimize over all pairs of pairwise rotations, we would solve $\max_{R_1, \dots, R_n \in SO(d)} \sum_{i=1}^{n} \sum_{j=1}^{n} \left\| R_i^T \left( \sum_k R_{ik} R_{kj} \right) R_j \right\| = \| R^T H^2 R \|^2$.
 %
@@
 
-Angular synchronization assumes that each data point is a replicate measurements of the same underlying configuration, simply corrupted with rotations and some noise.
+Angular synchronization assumes that each data point is a replicate measurements of the same underlying configuration, simply corrupted with rotations and some noise, and therefore considers all pairwise measurements equally important.
 %
@@
 
-Vector diffusion maps (VDM) couples the symmetry-factoring of angular synchronization with the uncovering of dynamics of diffusion maps \cite{singer2012vector}. 
+Vector diffusion maps (VDM) couples the symmetry-reduction of angular synchronization with the automatic parameterization of diffusion maps \cite{singer2012vector}. 
 %
@@
 
-We then compute the eigenvectors $\phi_1, \dots, \phi_{nd}$ and eigenvalues $\lambda_1, \dots, \lambda_{nd}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{nd}|$.
+We compute the eigenvectors $\phi_1, \dots, \phi_{nd}$ and eigenvalues $\lambda_1, \dots, \lambda_{nd}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{nd}|$.
 %
-As in angular synchronization, the first ``block'' eigenvector, i.e. the first two eigenvectors concatenated into a $nd \times d$ matrix, gives the optimal rotations. 
+As in angular synchronization, the first ``block'' eigenvector, i.e. the first $d$ eigenvectors concatenated into a $nd \times d$ matrix, gives the (relaxed) optimal rotations. 
 %
@@
 %
-Let $\phi_j(i)$ denote the $i^{th}$ block of $\phi_j$. 
+Let $\phi_j(i)$ denote the $i^{th}$ block of $\phi_j$ (so that $\phi_j(i) \in \mathbb{R}^d$). 
 %
-The embedding coordinates are then given by $\langle \phi_j(i), \phi_k(i) \rangle$, where $1 \le j, k, \le nd$.
+The embedding coordinates of $x_i$ are then given by $\langle \phi_j(i), \phi_k(i) \rangle$, where $1 \le j, k, \le nd$.
 
@@
 % 
-Furthermore, the scattering transform can be shown to be a contractive operator; therefore, noise and small deformations in the signal are not amplified in the transformation.
+Furthermore, the scattering transform can be shown to be a contractive operator, so that noise and small deformations in the signal are not amplified in the transformation.
 %
-We utilized the freely available ScatNet software (\url{http://www.di.ens.fr/data/software/scatnet/}) to compute the rotation-invariant scattering coefficients.
+We utilized the freely available software ScatNet (\url{http://www.di.ens.fr/data/software/scatnet/}) to compute the rotation-invariant scattering coefficients.
 

### 2014-01-28 Amit_notes/Procrustes.tex
Added notes from Amit

\documentclass[12pt]{article}

\begin{document}

Assume we have a signal that consists of $n$ features, each which lie in $d$-dimensional space (e.g. if we have a 10-atom molecule in three-dimensional space, then $n=10$ and $d=3$).

Denote the features (atoms) as $\x_1, \dots, x_n \in \mathbb{R}^d$.

Assume that we observe $m$ rotated noisy copies of the features
$$y_i^{(k)} = O_k x_i + \xi_^{(k)}, k=1, \dots, m, i=1, \dots, n$$

We would like to construct the original features $x_1, \dots, x_n$ from the rotated noisy copies $y_i^{(k)}$

Note that, up to noise
$$ \langle y_i^{(k)}, y_j^{(k)} \rangle  \approx \langle x_i, x_j \rangle $$
because the rotations preserve inner products.

Therefore,  we can approximate $\langle x_i, x_j \rangle$ as 
$$ \langle x_i, x_j \rangle  \approx \frac{1}{m} \sum_{k=1}^m \langle y_i^{(k)}, y_j^{(k)} \rangle $$

We can therefore estimate the matrix $X$, where $X_{ij} = \langle x_i, x_j \rangle$.

However, this is precisely the covariance matrix of the data $x_1, \dots, x_n$.

We can therefore recover the original signal using PCA/eigendecomposition.

\end{document}

### 2014-01-29 Amit_notes/multireference_alignment.tex
added multireference alignment notes

\documentclass[12pt]{article}

\begin{document}

We have images $I_1, \dots, I_n$

Want to find rotations $R_1, \dots, R_n \in SO(2)$ such that the images are optimally aligned

Define a cost function $c(R_i I_i, R_j I_j)$.
%
Then we can write the objective function 
$$\min_{R_1, \dots R_n \in SO(2)} \sum_{i \ne j} c(R_i I_i, R_j I_j)$$

Since the cost is invariant to global rotations, we can rewrite it as
$$\min_{R_1, \dots R_n \in SO(2)} \sum_{i \ne j} c(I_i, R_i^T R_j I_j)$$

We can discretize the space of rotations into $L$ intervals, and write each rotation as a number $l \in \mathbb{Z}_L$. 
%
We also define a cost function between a {\em specific} pair of images $I_i$ and $I_j$ as $c_{ij}$. 
%
This cost function is only a function of the difference in rotations, $l_i - l_j$. 
%
Therefore, we can rewrite the optimization problem as
$$\min_{l_1, \dots, l_n \in \mathbb{Z}_L} \sum_{i \ne j} c_{ij} (l_i - l_j)$$

We now note that each rotation $l_i-l_j \in \mathbb{Z}_L$ maps to a {\em cyclic} permutation matrix $\pi_{ij} \in \mathbb{R}^{L \times L}$, because each rotation can be thought of as a cyclic permutation of the $L$ ``ticks'' on the unit circle.
%
However, the set of cyclic permutation matrices is non-convex (e.g. because the average of two permutations matrices is not a permutation matrix). 
%
The convex hull of permutation matrices is the set of all doubly stochastic matrices:
$$\pi_{ij}(k,l) \ge 0  \forall k, l $$
$$\pi_{ij} \mathbb{1} = \mathbb{1}$$
$$\mathbb{1}^T \pi_{ij} = \mathbb{1}^T$$

We also require the cyclic condition to hold:
$$\pi_{ij}(k,l) = \pi_{ij}(k+1, l+1)$$

We would also like {\em consistency} relations to hold, e.g. $\pi_{ij} \pi_{jk} = \pi_{ik}$.
%
This is trivially satisfied if $\exists \pi_1, \dots, \pi_n$ such that $\pi_{ij} = \pi_i^T \pi_j \forall i, j$. 
%
Therefore, we want
$$\Pi = \left[ \begin{array}{c} \pi_1^T \\ \dots \\ \pi_n^T \end{array} \right] [\pi_1 \dots \pi_n]$$
where $\Pi \in \mathbb{R}^{nL \times nL}$ is the Gram matrix, where the $i,j$ block of $\Pi$ is $\pi_{ij} \in \mathbb{R}^{L \times L}$.

This is true if $rank(\Pi) = L$ and $\Pi$ is positive semidefinite. 
%
We relax the constrant that $rank(\Pi) = L$, and only require that $\Pi$ is psd.

We also require that
$$\pi_{ii} = I){L \times L}$$

We then compute the cost function $c_{ij}$ for every possible $l_i - l_j \in \mathbb{Z}_L$, and put these values into the matrix $C_{ij} \in \mathbb{R}^{L \times L}$.
%
The objective function corresponding to a particular permutation matrix can then be written as 
$$\sum_{i \ne j} Tr(C_{ij} \pi_{ij})$$

Or, if we write it in full matrix notation
$$\min_{\Pi} Tr(C^T \Pi)$$

\end{document}

### 2014-01-29 Amit_notes/Procrustes.tex
Updated notes from Amit

@@
 \documentclass[12pt]{article}
+
+\usepackage{amsmath}
+\usepackage{amssymb}
+
+\title{Solving the Procrustes Problem with PCA}
 
@@
 
-Assume we have a signal that consists of $n$ features, each which lie in $d$-dimensional space (e.g. if we have a 10-atom molecule in three-dimensional space, then $n=10$ and $d=3$).
+\maketitle
 
-Denote the features (atoms) as $\x_1, \dots, x_n \in \mathbb{R}^d$.
+Assume we have a signal that consists of $n$ features, each of which lie in $d$-dimensional space (e.g. if we have a 10-atom molecule in three-dimensional space, then $n=10$ and $d=3$).
+%
+Denote the features as $x_1, \dots, x_n \in \mathbb{R}^d$.
 
-Assume that we observe $m$ rotated noisy copies of the features
-$$y_i^{(k)} = O_k x_i + \xi_^{(k)}, k=1, \dots, m, i=1, \dots, n$$
+We observe $m$ rotated noisy copies of the features
+\begin{equation}
+y_i^{(k)} = O_k x_i + \xi_i^{(k)}, k=1, \dots, m, i=1, \dots, n,
+\end{equation}
+where $O_k$ is an orthoganol transform, and $\xi_i^{(k)}$ is Gaussian white noise. 
+%
+We would like to estimate the original signal $x_1, \dots, x_n$ from the rotated noisy copies $y_i^{(k)}$.
 
-We would like to construct the original features $x_1, \dots, x_n$ from the rotated noisy copies $y_i^{(k)}$
+Note that, because orthoganol transforms preserve inner products,
+\begin{equation}
+\langle y_i^{(k)}, y_j^{(k)} \rangle  \approx \langle x_i, x_j \rangle 
+\end{equation}
+(up to noise).
 
-Note that, up to noise
-$$ \langle y_i^{(k)}, y_j^{(k)} \rangle  \approx \langle x_i, x_j \rangle $$
-because the rotations preserve inner products.
+Therefore,  we can estimate $\langle x_i, x_j \rangle$ as 
+\begin{equation} \label{eq:ip_est}
+\langle x_i, x_j \rangle  \approx \frac{1}{m} \sum_{k=1}^m \langle y_i^{(k)}, y_j^{(k)} \rangle 
+\end{equation}
 
-Therefore,  we can approximate $\langle x_i, x_j \rangle$ as 
-$$ \langle x_i, x_j \rangle  \approx \frac{1}{m} \sum_{k=1}^m \langle y_i^{(k)}, y_j^{(k)} \rangle $$
-
-We can therefore estimate the matrix $X$, where $X_{ij} = \langle x_i, x_j \rangle$.
-
-However, this is precisely the covariance matrix of the data $x_1, \dots, x_n$.
-
-We can therefore recover the original signal using PCA/eigendecomposition.
+We then construct the matrix $C$, where $C_{ij} = \langle x_i, x_j \rangle$, using the estimates in \eqref{eq:ip_est}.
+%
+However, $C$ is precisely the covariance matrix of the data $x_1, \dots, x_n$. 
+%
+We can therefore recover $x_1, \dots, x_n$ using eigendecomposition:
+\begin{equation}
+\begin{bmatrix}
+x_1^T \\
+\vdots \\
+x_n^T 
+\end{bmatrix} = 
+\begin{bmatrix}
+| & | &  & | \\
+\phi_1 & \phi_2 & \cdots & \phi_d \\
+| & | &  & | \\
+\end{bmatrix}
+\begin{bmatrix}
+\sqrt{\lambda_1} & 0 & 0 & \cdots & 0 \\
+0 & \sqrt{\lambda_2} & 0 & \cdots & 0 \\
+\vdots & \vdots & \vdots & \ddots & \vdots \\
+0 & 0 & 0 & \cdots & \sqrt{\lambda_d} 
+\end{bmatrix}
+\end{equation}
+where $\lambda_1, \dots, \lambda_n$ and $\phi_1, \dots, \phi_n$ are the eigenvalues and eigenvectors, respectively, of $C$, and they are ordered such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_n$.
 

### 2014-01-29 Amit_notes/multireference_alignment.tex
Updated notes from Amit

@@
 \documentclass[12pt]{article}
+\usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{dsfont}
+
+\title{Multireference Alignment Notes}
 
@@
 
-We have images $I_1, \dots, I_n$
+\maketitle
 
-Want to find rotations $R_1, \dots, R_n \in SO(2)$ such that the images are optimally aligned
+We are given images $I_1, \dots, I_n$.
 
-Define a cost function $c(R_i I_i, R_j I_j)$.
+We want to find rotations $R_1, \dots, R_n \in SO(2)$ such that the images are optimally aligned.
+
+We define a cost function $c(R_i I_i, R_j I_j)$.
 %
-Then we can write the objective function 
-$$\min_{R_1, \dots R_n \in SO(2)} \sum_{i \ne j} c(R_i I_i, R_j I_j)$$
+We would like to solve the optimization problem
+\begin{equation} \label{eq:opt1}
+\min_{R_1, \dots R_n \in SO(2)} \sum_{i \ne j} c(R_i I_i, R_j I_j)
+\end{equation}
 
-Since the cost is invariant to global rotations, we can rewrite it as
-$$\min_{R_1, \dots R_n \in SO(2)} \sum_{i \ne j} c(I_i, R_i^T R_j I_j)$$
+Since the cost is invariant to global rotations, we can rewrite \eqref{eq:opt1} as
+\begin{equation}
+\min_{R_1, \dots R_n \in SO(2)} \sum_{i \ne j} c(I_i, R_i^T R_j I_j)
+\end{equation}
 
-We can discretize the space of rotations into $L$ intervals, and write each rotation as a number $l \in \mathbb{Z}_L$. 
+We discretize the space of rotations into $L$ intervals, and write each rotation as a number $l_i \in \mathbb{Z}_L, i=1, \dots, n$. 
 %
-We also define a cost function between a {\em specific} pair of images $I_i$ and $I_j$ as $c_{ij}$. 
+We also define a cost function between a {\em specific} pair of images $I_i$ and $I_j$ as $c_{ij}(l_i - l_j)$; 
+this cost function is only a function of the difference in rotations, $l_i - l_j$. 
 %
-This cost function is only a function of the difference in rotations, $l_i - l_j$. 
-%
-Therefore, we can rewrite the optimization problem as
+we rewrite the optimization problem as
 $$\min_{l_1, \dots, l_n \in \mathbb{Z}_L} \sum_{i \ne j} c_{ij} (l_i - l_j)$$
@@
 %
+We will optimize over the space of cyclic permutation matrices, rather than the space of rotation matrices.
+
 However, the set of cyclic permutation matrices is non-convex (e.g. because the average of two permutations matrices is not a permutation matrix). 
 %
-The convex hull of permutation matrices is the set of all doubly stochastic matrices:
-$$\pi_{ij}(k,l) \ge 0  \forall k, l $$
-$$\pi_{ij} \mathbb{1} = \mathbb{1}$$
-$$\mathbb{1}^T \pi_{ij} = \mathbb{1}^T$$
+The convex hull of permutation matrices can be shown to be the set of all doubly stochastic matrices, which satisfy the following constraints
+\begin{equation} \label{eq:doublestoch}
+\begin{aligned}
+\pi_{ij}(k,l) &  \ge 0 \forall k, l \\
+\pi_{ij} \mathds{1} & = \mathds{1}\\
+\mathds{1}^T \pi_{ij} & = \mathds{1}^T 
+\end{aligned}
+\end{equation}
 
 We also require the cyclic condition to hold:
-$$\pi_{ij}(k,l) = \pi_{ij}(k+1, l+1)$$
+\begin{equation} \label{eq:cyclic_const}
+\pi_{ij}(k,l) = \pi_{ij}(k+1, l+1)
+\end{equation}
 
@@
 Therefore, we want
-$$\Pi = \left[ \begin{array}{c} \pi_1^T \\ \dots \\ \pi_n^T \end{array} \right] [\pi_1 \dots \pi_n]$$
+\begin{equation} \label{eq:Gram_const}
+\Pi = 
+\begin{bmatrix}
+\pi_{11} & \pi_{12} & \cdots & \pi_{1n} \\
+\pi_{21} & \pi_{22} & \cdots  & \pi_{2n} \\
+\vdots & \vdots & \ddots & \vdots \\
+\pi_{n1} & \pi_{n2} & \cdots  & \pi_{nn} 
+\end{bmatrix} = 
+\begin{bmatrix}
+\pi_1^T \\
+\vdots \\
+\pi_n^T 
+\end{bmatrix}
+\begin{bmatrix}
+\pi_1^T \cdots \pi_n^T 
+\end{bmatrix}
+\end{equation}
 where $\Pi \in \mathbb{R}^{nL \times nL}$ is the Gram matrix, where the $i,j$ block of $\Pi$ is $\pi_{ij} \in \mathbb{R}^{L \times L}$.
 
-This is true if $rank(\Pi) = L$ and $\Pi$ is positive semidefinite. 
+\eqref{eq:Gram_const} holds if $rank(\Pi) = L$ and $\Pi$ is positive semidefinite. 
 %
-We relax the constrant that $rank(\Pi) = L$, and only require that $\Pi$ is psd.
+We relax the constrant that $rank(\Pi) = L$, and only require  
+\begin{equation} \label{eq:psd_const}
+\Pi \succeq 0
+\end{equation}
 
 We also require that
-$$\pi_{ii} = I){L \times L}$$
+\begin{equation} \label{eq:identity_const}
+\pi_{ii} = I_{L \times L}
+\end{equation}
 
-We then compute the cost function $c_{ij}$ for every possible $l_i - l_j \in \mathbb{Z}_L$, and put these values into the matrix $C_{ij} \in \mathbb{R}^{L \times L}$.
-%
-The objective function corresponding to a particular permutation matrix can then be written as 
-$$\sum_{i \ne j} Tr(C_{ij} \pi_{ij})$$
+We then construct the matrices $C_{ij}  \in \mathbb{R}^{L \times L}$ for $i, j= 1, \dots, n$, where
+\begin{equation}
+C_{ij}(l_1, l_2) = c_{ij}(l_1 - l_2)
+\end{equation}
 
-Or, if we write it in full matrix notation
-$$\min_{\Pi} Tr(C^T \Pi)$$
+The objective function can then be written as
+\begin{equation} \label{eq:one_cost}
+\min_{\pi_{ij}} \sum_{i \ne j} Tr(C_{ij} \pi_{ij})
+\end{equation}
+
+We then define the matrix $C \in \mathbb{R}^{nL \times nL}$, where
+$$C = \begin{bmatrix}
+ C_{11} & C_{12} & \cdots & C_{1n} \\
+ C_{21} & C_{22} & \cdots & C_{2n} \\
+\vdots & \vdots & \ddots & \vdots \\ 
+C_{n1} & C_{n2} & \cdots & C_{nn} \\
+\end{bmatrix} $$
+
+We can then write \eqref{eq:one_cost} as 
+\begin{equation} \label{eq:total_cost}
+\min_{\Pi} Tr(C^T \Pi)
+\end{equation}
+
+\eqref{eq:total_cost}, togther with constraints \eqref{eq:doublestoch}, \eqref{eq:cyclic_const}, \eqref{eq:psd_const}, and \eqref{eq:identity_const}, define the semidefinite program that we would like to solve.
 

### 2014-02-13 image_analysis_paper/background_reading/article_notes.tex

\documentclass[12pt]{article}

\usepackage{biblatex}

\bibliography{references}

\begin{document}

\fullcite{wei2006conditional}

\fullcite{wei2006quantile}

\fullcite{li2009stochastic}

\fullcite{jordan2013behavioral}

\fullcite{tucker2010pseudotemporal}
\begin{itemize}
	\item Uses cross-sectional data of eye disease progression (e.g. glaucoma).
	\item Build time series from cross-sectional data (``pseudo time series'')
	\item Subsample data; from each subsample, build a ``pseudo time series'' by finding the shortest path through the points
	\item Use these shortest paths as time series to train a time series model (i.e. a HMM model).
\end{itemize}

\fullcite{kafri2013dynamics}
\begin{itemize}
	\item Take snapshot of fixed cell population
	\item Fit 1D curve to trajectory (they assert that the two ``important'' parameters are levels of DNA and the level of ``anaphase-promoting complex'')
	\item Since the cells follow a closed cycle, given the closed curve, they can use mass balances to determine the relative rates/fluxes through the different stages
\end{itemize}

\fullcite{wyart2010evaluating}

\fullcite{chen2012nonlinear}

\fullcite{das2013dynamic}

\fullcite{schindler2007inferring}

\fullcite{bigot2012geometric}

\fullcite{escudero2011epithelial}

\fullcite{machta2013parameter}

\fullcite{peng2013detecting}

\fullcite{gerber2010visual}

\fullcite{qiu2011discovering}

\fullcite{gupta2008extracting}

\fullcite{lau2009modeling}

\fullcite{toga2003temporal}

\fullcite{li2012modelling}

\fullcite{denkinger2013set}

\fullcite{xu2012comprehensive}

\fullcite{sisan2010event}

\fullcite{golparvar2011monitoring}

\fullcite{golani1999phenotyping}
\begin{itemize}
	\item Want to construct video of construction progress from snapshots
	\item Building Information Models (BIM) 
\end{itemize}

\fullcite{anavy2014blind}

\end{document}

### 2014-02-13 image_analysis_paper/paper_draft.tex

\documentclass[12pt]{article}

\begin{document}

\section{Introduction}

% only images

\section{Techniques}

% show schematic with images (starfish or zebrafish w/ stripes)

\subsection{Alignment Using Angular Synchronization}

\subsection{Ordering Using Diffusion Maps}

\section{Example: {\em Drosophila}}

\subsection{dpERK}

\subsection{Dorsal}

\section{Discussion}
% compare to TSP algorithms

\end{document}

### 2014-02-14 image_analysis_paper/paper_draft.tex
Upadted draft

@@
 \documentclass[12pt]{article}
+
+\usepackage{amsmath}
+\usepackage{amssymb}
+\DeclareMathOperator*{\argmin}{arg\,min}
 
@@
 
+When a data set is invariant under an underlying symmetry group (such as translations and rotations), we must factor out the underlying symmetries and align the data before doing further analysis.
+%
+In our example, we will translate and rotate images until they are optimally aligned, and then operate on this aligned set. 
+%
+There are many ways to align images. 
+%
+``Template--based'' alignment \cite{...} is one technique, where one selects a specific image or {\em template} as
+
+Angular synchronization calculates the optimal alignments for each point in a data set, given estimates of the optimal alignments between pairs of data points \cite{singer2011angular}.
+
+An important part of angular synchronization is that the symmetry group be {\em compact}. 
+%
+This is a rather technical point, but all we wish to state is that the group of two--dimensional translations and rotations is {\em not} compact.
+%
+We can instead map this group to the group of
+
+Let $x_1, \dots, x_n$ denote the data points that we wish to align under some symmetry group.
+%
+We will consider the case where the symmetry group is the group of $d$-dimensional rotations, $SO(d)$. 
+%
+Let $R_{ij} \in SO(d)$ denote the rotation that aligns $x_j$ to $x_i$, so that
+\begin{equation}
+R_{ij} = \argmin_{R \in SO(d)} \|Rx_j - x_i \|^2.
+\end{equation}
+%
+We seek to find the rotations $R_1, R_2, \dots, R_n \in SO(d)$ such that $R_i R_j^T \approx R_{ij}$, for every pair $i, j$. 
+%TODO: Check that this order is correct
+%
+We would also like to exploit {\em higher-order} consistency relations;
+for example, the relationship $R_{ij} R_{jk} \approx R_{ik}$ should hold if the rotation estimates are accurate.
+%
+Therefore, we consider measurements which (almost) satisfy such conditions ``good'' measurements, and those measurements which do not satisfy such conditions as most likely inaccurate.
+
+We would like to solve the optimization problem 
+\begin{equation} \label{eqn:angsynch_obj}
+\max_{R_1, \dots, R_n \in SO(d)} \left\| \sum_{i=1}^{n} \sum_{j=1}^{n} R_i^T R_{ij} R_j \right\|^2.
+\end{equation}
+%
+%TODO: check if a norm is needed
+Each term in the objective function will be large large if $R_i$ and $R_j$ are consistent with $R_{ij}$, and will be random with mean 0 if $R_i$ and $R_j$ are inconsistent with $R_{ij}$.
+%
+Therefore, solving \eqref{eqn:angsynch_obj} gives us the set of global alignments which are most consistent with the computed pairwise alignments.
+
+In general, the solution to \eqref{eqn:angsynch_obj} is not easily computed.
+%
+Instead, we relax the problem and allow $R_1, \dots, R_n \in \mathbb{R}^{d \times d}$.
+%
+We define the matrix $H$, which is an $n \times n$ block matrix with $d \times d$ blocks (so $H \in \mathbb{R}^{nd \times nd}$), with $H_{ij} = R_{ij}$.
+%
+We can then write the relaxed problem as 
+\begin{equation} \label{eqn:angsynch_relax}
+\max_{R\in \mathbb{nd \times d}} \| R^T H R \|^2.
+\end{equation}
+%
+The solution to \eqref{eqn:angsynch_relax} is given by the top ``block'' eigenvector of $H$, which we denote $\hat{R} \in \mathbb{R}^{nd \times d}$. 
+%
+The $i^{th}$ rotation is (approximately) $\hat{R}(i) \in \mathbb{R}^{d \times d}$.
+%
+We would like to note that this formulation also accounts for higher-order relations.
+%
+For example, if we want to optimize over all pairs of pairwise rotations, we would solve $\max_{R_1, \dots, R_n \in SO(d)} \sum_{i=1}^{n} \sum_{j=1}^{n} \left\| R_i^T \left( \sum_k R_{ik} R_{kj} \right) R_j \right\| = \| R^T H^2 R \|^2$.
+%
+However, the solution to this problem is also given by the top eigenvector of $H$.
+
+Because we relaxed the problem to allow our solutions to lie in $\mathbb{R}^{d \times d}$, we must project our approximate solution back to $SO(d)$.
+%
+The optimal rotation is therefore given by $R_i = U_i V_i^T$, where $U_i$ and $V_i$ are the left and right singular vectors of $\hat{R}_i$, respectively.
+
 \subsection{Ordering Using Diffusion Maps}
+
+Unlike PCA, diffusion maps (DMAPS) is a nonlinear dimensionality reduction technique. 
+%
+DMAPS aims to uncover a parameterization of high-dimensional data sampled from a low-dimensional nonlinear manifold \cite{coifman2005geometric}.
+%
+The {\em essential} requirement for DMAPS is an appropriate distance metric $d(x_i, x_j)$ for comparing data points. 
+%
+This can as simple as the standard Euclidean distance, or a more complex metric (such as a distance between features of the data points) for other data sets.
+
+Given data points $x_1, \dots, x_n \in \mathbb{R}^p$, we fist calculate the matrix $W \in \mathbb{R}^{n \times n}$, where 
+\begin{equation}
+W_{ij} = \exp \left( -\frac{d^2(x_i, x)j)}{\epsilon^2} \right)
+\end{equation}
+and $\epsilon$ is a characteristic distance between data points.
+%
+$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph}); in practice, we choose $\epsilon$ to be the median of the pairwise distances between data points.
+%
+We then compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{n} W_{ij}$, and the matrix $A = D^{-1} W$. 
+%
+We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_n$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_n$ and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_n|$.
+%
+Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors. 
+%
+The eigenvectors $\phi_1, \phi_2, \dots, \phi_n$ give the embedding coordinates, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point. 
+%
+Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector.
+%
+The next few eigenvectors give the ``meaningful'' embedding coordinates for the data.
+
+\subsection{Aligning and Ordering Using Vector diffusion maps}
+
+Angular synchronization assumes that each data point is a replicate measurements of the same underlying configuration, simply corrupted with rotations and some noise, and therefore considers all pairwise measurements equally important.
+%
+However, in system with an underlying symmetry group {\em as well as} a dynamical process, the data points are not identical modulo symmetries and noise, and we would like to place more emphasis on aligning data points which are dynamically close.
+
+Vector diffusion maps (VDM) couples the symmetry-reduction of angular synchronization with the automatic parameterization of diffusion maps \cite{singer2012vector}. 
+%
+We construct the matrix $A$ as in diffusion maps, and the matrix $H$ as in angular synchronization.
+%
+We then construct the block matrix $S$, where the blocks of $S$ are defined by
+\begin{equation}
+S_{ij} = \left\{ \begin{array}{l l} 
+A_{ij} H_{ij} & i \ne j \\
+0_{d \times d} & i = j
+\end{array}
+\right.
+\end{equation}
+%
+where $A_{ij}$ is the $(i,j)$ entry of $A$, and $H_{ij}$ is the $(i,j)$ block of $H$.
+
+We compute the eigenvectors $\phi_1, \dots, \phi_{nd}$ and eigenvalues $\lambda_1, \dots, \lambda_{nd}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{nd}|$.
+%
+As in angular synchronization, the first ``block'' eigenvector, i.e. the first $d$ eigenvectors concatenated into a $nd \times d$ matrix, gives the (relaxed) optimal rotations. 
+%
+However, the eigenvectors now also give us embedding coordinates.
+%
+Let $\phi_j(i)$ denote the $i^{th}$ block of $\phi_j$ (so that $\phi_j(i) \in \mathbb{R}^d$). 
+%
+The embedding coordinates of $x_i$ are then given by $\langle \phi_j(i), \phi_k(i) \rangle$, where $1 \le j, k, \le nd$.
 

### 2014-02-14 image_analysis_paper/background_reading/article_notes.tex
added edits to paper, backgrond reading; added eve data

@@
 
-\fullcite{wei2006conditional}
+\begin{enumerate}
 
-\fullcite{wei2006quantile}
+\item \fullcite{wei2006conditional}
 
-\fullcite{li2009stochastic}
+\item \fullcite{wei2006quantile}
 
-\fullcite{jordan2013behavioral}
+\item \fullcite{li2009stochastic}
 
-\fullcite{tucker2010pseudotemporal}
+\item \fullcite{jordan2013behavioral}
+
+\item \fullcite{tucker2010pseudotemporal}
 \begin{itemize}
@@
 
-\fullcite{kafri2013dynamics}
+\item \fullcite{kafri2013dynamics}
 \begin{itemize}
 	\item Take snapshot of fixed cell population
-	\item Fit 1D curve to trajectory (they assert that the two ``important'' parameters are levels of DNA and the level of ``anaphase-promoting complex'')
+	\item Fit 1D curve to trajectory (they assert that the two ``important'' parameters are levels of DNA and the level of ``anaphase-promoting complex'', so they project onto these two components and then fit a 1D closed curve to the data)
 	\item Since the cells follow a closed cycle, given the closed curve, they can use mass balances to determine the relative rates/fluxes through the different stages
@@
 
-\fullcite{wyart2010evaluating}
+\item \fullcite{wyart2010evaluating}
 
-\fullcite{chen2012nonlinear}
+\begin{itemize}
+	\item Use RNA FISH technique to count levels of individual mRNAs in cells; requires cells to be fixed
+	\item Propose models for dynamics of mRNA throughout cell cycle
+	\item Show how they can use data from population of fixed cells to fit model (periodic model for level of mRNA production)
+\end{itemize}
 
-\fullcite{das2013dynamic}
+\item \fullcite{chen2012nonlinear}
 
-\fullcite{schindler2007inferring}
+\item \fullcite{das2013dynamic}
 
-\fullcite{bigot2012geometric}
+\item \fullcite{schindler2007inferring}
 
-\fullcite{escudero2011epithelial}
+\item \fullcite{bigot2012geometric}
 
-\fullcite{machta2013parameter}
+\item \fullcite{escudero2011epithelial}
 
-\fullcite{peng2013detecting}
+\item \fullcite{machta2013parameter}
 
-\fullcite{gerber2010visual}
+\item \fullcite{peng2013detecting}
 
-\fullcite{qiu2011discovering}
+\item \fullcite{gerber2010visual}
 
-\fullcite{gupta2008extracting}
+\item \fullcite{qiu2011discovering}
 
-\fullcite{lau2009modeling}
+\item \fullcite{gupta2008extracting}
 
-\fullcite{toga2003temporal}
+\item \fullcite{lau2009modeling}
 
-\fullcite{li2012modelling}
+\item \fullcite{toga2003temporal}
 
-\fullcite{denkinger2013set}
+\item \fullcite{li2012modelling}
 
-\fullcite{xu2012comprehensive}
+\item \fullcite{denkinger2013set}
 
-\fullcite{sisan2010event}
+\item \fullcite{xu2012comprehensive}
 
-\fullcite{golparvar2011monitoring}
+\item \fullcite{sisan2010event}
 
-\fullcite{golani1999phenotyping}
+\item \fullcite{golparvar2011monitoring}
+
+\item \fullcite{golani1999phenotyping}
 \begin{itemize}
@@
 
-\fullcite{anavy2014blind}
+\item \fullcite{anavy2014blind}
+
+\end{enumerate}
 

### 2014-02-14 image_analysis_paper/paper_draft.tex
added edits to paper, backgrond reading; added eve data

@@
 
-When a data set is invariant under an underlying symmetry group (such as translations and rotations), we must factor out the underlying symmetries and align the data before doing further analysis.
+Often when we collect images in experiments, each image is taken in a different orientation. 
 %
-In our example, we will translate and rotate images until they are optimally aligned, and then operate on this aligned set. 
+Therefore, the first task in image analysis is to align the set of images. 
 %
+We will consider settings when alignment is done with respect to two--dimensional translations and rotations of the images, as each ``object of interest'' (i.e.
+each embryo) is in a different position and rotated at a different angle within the image frame.
+
 There are many ways to align images. 
 %
-``Template--based'' alignment \cite{...} is one technique, where one selects a specific image or {\em template} as
+``Template--based'' alignment \cite{...} is one technique, where one selects a specific image or {\em template}, and then tries to optimally align each image to that template.
+%
+However, this method can result in many misalignemnts for noisy images, and can be problematic if a ``good'' template image is not known {\em a priori}. 
+%
+Instead, we will use a technique called angular synchronization\cite{singer2011angular}.
+%
+Angular synchronization uses each image as a template for every other image to compute the pairwise alignments for every possible pair of images.
+%
+From this large collection of pairwise alignments, it then tries to find optimal alignment for each image, such that these individual alignments are most consistent (in some metric) with the pairwise alignment.
+%
+Because we consider {\em all} pairwise alignments, angular synchronization is often much more robust to noise.
 
-Angular synchronization calculates the optimal alignments for each point in a data set, given estimates of the optimal alignments between pairs of data points \cite{singer2011angular}.
+An important part of angular synchronization is that the symmetry group $G$ be {\em compact} and have a real (or perhaps complex) representation.
+%
+Compactness guarantees that the minimizer for each pairwise alignment is contained in the group (Extreme Value Theorem). 
+%
+Having a real representation allows us to write combinations of transforms as group operations, and will allow us to therefore write our optimization problem in matrix form. 
+%
+This is a rather technical point;
+we are only discussing this because the group of two--dimensional translations and rotations, $ISO(2)$, is {\em not} compact.
+%
+The angular synchronization problem is therefore not well--defined for this symmetry group.
+%
+However, we can ``compactify'' this group by mapping $ISO(2)$ into the group $SO(3)$, the group of three--dimensional rotations.
+%
+We do this by projecting the image onto a portion of the surface of a (three--dimensional) sphere.
+%
+Now, rotations in two--dimensions correspond to rotations around one principal axis in three--dimensions, and translations in two--dimensions correspond (approximately) to rotations around the other two principal axes in three--dimensions.
 
-An important part of angular synchronization is that the symmetry group be {\em compact}. 
+Let $x_1, \dots, x_m \in \mathbb{R}^{n \times n}$ denote the images that we wish to align with respect to translations and rotations.
 %
-This is a rather technical point, but all we wish to state is that the group of two--dimensional translations and rotations is {\em not} compact.
+The first step in angular synchronization is to compute the alignments between each pair of data points. 
 %
-We can instead map this group to the group of
+For each image pair $x_i$ and $x_j$, we want to find the optimal translation + rotation that aligns $x_i$ to $x_j$.
+%
+We discritize the search space of rotations into 20 possible rotations ($0, \pi/10, \pi/5, \dots, 19\pi/10, 2\pi$), and we allow translations of $-20, -16, -12, \dots, 12, 16, 20$ pixels in both the $x$ and $y$ directions.
+%
+We then exhaustively search over this space for the best rotation + translation that aligns $x_i$ to $x_j$. 
+\begin{equation}
+(\theta_{ij}, dx_{ij}, dy_{ij}) = \argmin_{\theta, dx, dy} \|f(x_j, \theta, dx, dy) - x_i \|^2.
+\end{equation}
+where $f$ is a function that first rotates the image $x_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction.
+%
+The norm is the Frobenius norm of the resulting matrix. 
+%
+This means we check each of the 2420 possible combinations; although this is somewhat time intensive (TODO: add timings), it is not prohibative and can be trivially parallelized for larger data sets if necessary.
+%
+Let $\theta$ denote the optimal angle of rotation, and let $dx$ and $dy$ denote the optimal number of pixels to translate in the $x$ and $y$ directions, respectively.
+%
+Then, in the space of three--dimensional rotations, the Euler angles $\alpha$, $\beta$, and $\gamma$ correspond to
+\begin{eqnarray}
+	\alpha_{ij} &=& \theta_{ij} \\
+	\beta_{ij} &=& \frac{dx_{ij}}{npixels} \times \eta_{proj} \\
+	\gamma_{ij} &=& \frac{dy_{ij}}{npixels} \times \eta_{proj} \\
+\end{eqnarray}
+where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project (we take $\eta_{proj} =  \pi/8$).
+%
+From here, we can write rotations around the three principal axes in terms of the three Euler angles
+\begin{eqnarray}
+	R^x_{ij} &=& \begin{bmatrix}
+	1 & 0 & 0 \\
+    0 & \cos(\alpha) & -\sin(\alpha) \\
+    0 & \sin(\alpha) & \cos(\alpha)
+	\end{bmatrix} \\
+	R^y_{ij} &=& \begin{bmatrix}
+	\cos(\beta) & 0 & \sin(\beta) \\
+    0 & 1 & 0 \\
+    -\sin(\beta) & 0 & \cos(\beta)
+    \end{bmatrix} \\
+	R^z_{ij} &=& \begin{bmatrix} 
+	\cos(\gamma) & -\sin(\gamma) & 0 \\
+    \sin(\gamma) & \cos(\gamma) & 0 \\
+    0 & 0 & 1 
+    \end{bmatrix}
+\end{eqnarray}
+%
+We then write the total rotation $R_{ij} \in SO(3)$ as 
+\begin{equation} \label{eq:total_rot}
+	R_{ij}	 = R^z_{ij} \times R^y_{ij} \times R^x_{ij}
+\end{equation}
+This corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (because the rotation matrices operate on the left, the order of operations/transformations is read from right to left in \eqref{eq:total_rot}.
 
-Let $x_1, \dots, x_n$ denote the data points that we wish to align under some symmetry group.
-%
-We will consider the case where the symmetry group is the group of $d$-dimensional rotations, $SO(d)$. 
-%
-Let $R_{ij} \in SO(d)$ denote the rotation that aligns $x_j$ to $x_i$, so that
+Let $R_{ij} \in SO(3)$ denote the rotation that aligns $x_j$ to $x_i$, so that
 \begin{equation}
-R_{ij} = \argmin_{R \in SO(d)} \|Rx_j - x_i \|^2.
+R_{ij} = \argmin_{R \in SO(3)} \|Rx_j - x_i \|^2.
 \end{equation}
 %
-We seek to find the rotations $R_1, R_2, \dots, R_n \in SO(d)$ such that $R_i R_j^T \approx R_{ij}$, for every pair $i, j$. 
+We want to find the rotations $R_1, R_2, \dots, R_n \in SO(d)$ such that $R_i R_j^T \approx R_{ij}$, for every pair $i, j$. 
 %TODO: Check that this order is correct
@@
 \begin{equation} \label{eqn:angsynch_obj}
-\max_{R_1, \dots, R_n \in SO(d)} \left\| \sum_{i=1}^{n} \sum_{j=1}^{n} R_i^T R_{ij} R_j \right\|^2.
+\max_{R_1, \dots, R_m \in SO(3)} \left\| \sum_{i=1}^{m} \sum_{j=1}^{m} R_i^T R_{ij} R_j \right\|^2.
 \end{equation}
@@
 %
-Instead, we relax the problem and allow $R_1, \dots, R_n \in \mathbb{R}^{d \times d}$.
+Instead, we relax the problem and allow $R_1, \dots, R_m \in \mathbb{R}^{3 \times 3}$.
 %
-We define the matrix $H$, which is an $n \times n$ block matrix with $d \times d$ blocks (so $H \in \mathbb{R}^{nd \times nd}$), with $H_{ij} = R_{ij}$.
+We define the matrix $H$, which is an $m \times m$ block matrix with $3 \times 3$ blocks (so $H \in \mathbb{R}^{3m \times 3m}$), with $H_{ij} = R_{ij}$.
 %
@@
 \begin{equation} \label{eqn:angsynch_relax}
-\max_{R\in \mathbb{nd \times d}} \| R^T H R \|^2.
+\max_{R\in \mathbb{3m \times 3}} \| R^T H R \|^2.
 \end{equation}
 %
-The solution to \eqref{eqn:angsynch_relax} is given by the top ``block'' eigenvector of $H$, which we denote $\hat{R} \in \mathbb{R}^{nd \times d}$. 
+The solution to \eqref{eqn:angsynch_relax} is given by the top ``block'' eigenvector of $H$, which we denote $\hat{R} \in \mathbb{R}^{3m \times 3}$. 
 %
-The $i^{th}$ rotation is (approximately) $\hat{R}(i) \in \mathbb{R}^{d \times d}$.
+The $i^{th}$ rotation is (approximately) $\hat{R}(i) \in \mathbb{R}^{3 \times 3}$.
 %
@@
 
-Because we relaxed the problem to allow our solutions to lie in $\mathbb{R}^{d \times d}$, we must project our approximate solution back to $SO(d)$.
+Because we relaxed the problem to allow our solutions to lie in $\mathbb{R}^{3 \times 3}$, we must project our approximate solution back to $SO(3)$.
 %
 The optimal rotation is therefore given by $R_i = U_i V_i^T$, where $U_i$ and $V_i$ are the left and right singular vectors of $\hat{R}_i$, respectively. 
+%
+Because the eigenvectors of $H$ are determined up to a sign, we adjust the signs of the eigenvectors so that $det(R_i) = +1$.
 

### 2014-02-14 eve data/rotate_image.m

@@ -1,0 +1,23 @@
+function im1 = rotate_image(R, im2, angle_proj)
+
+npixels = size(im2, 1);
+
+[theta, dxpixels, dypixels] = find_specs(R, npixels, angle_proj);
+
+im2 = imrotate(im2, theta, 'crop');
+im2 = circshift(im2, [dxpixels dypixels]);
+
+im1 = im2;
+
+function [theta, dx, dy] = find_specs(R, npixels, angle_proj)
+
+alpha = atan2(R(3,2), R(3,3));
+gamma = atan2(R(2,1), R(1,1));
+beta = -asin(R(3,1));
+%beta = atan2(-R(3,1), R(2,1)/sin(gamma));
+
+theta = alpha * 180 / pi;
+dx = round(beta * npixels / angle_proj);
+dy = round(gamma * npixels / angle_proj);
+
+

### 2014-02-14 eve data/test1.m

@@ -1,0 +1,112 @@
+clear all
+close all
+
+%% set parameters for saving figures
+set(0,'DefaultLineMarkerSize',14)
+set(0,'DefaultAxesFontSize',20)
+
+res = '-r300';
+fmt = '-djpeg';
+print_figures = false;
+
+eve_image_dir = 'Eve_13_0821';
+eve_membrane_times = 'Eve_13_0821/times_08_21_13.mat';
+
+%%
+load(eve_membrane_times);
+
+%% load images
+
+npixels = 100;
+m = 39;
+
+%set image plotting parameters
+subplot_dim1 = ceil(sqrt(m));
+subplot_dim2 = ceil(m / subplot_dim1);
+
+image_set = zeros(npixels, npixels, m);
+
+image_channel = 2;
+
+%figure;
+for i=1:m
+    % read image
+    im1 = imread(sprintf('%s/emb%02d.tif', eve_image_dir, i));
+    
+    % resize image
+    im1 = imresize(im1, [NaN npixels]);
+    
+    % extract relevent color from image
+    im1 = im1(:,:,image_channel);
+    
+    npad = [npixels npixels]-size(im1);
+    im1 = padarray(im1, floor(npad/2),'pre');
+    im1 = padarray(im1, ceil(npad/2),'post');
+    
+    im1 = imrotate(im1, rand*360, 'crop');
+    
+    subplot(subplot_dim1, subplot_dim2, i)
+    imshow(im1);
+    
+    %store image
+    im1 = double(im1);
+    image_set(:, :, i) = im1;
+    
+end
+
+%% raw dpERK images-- synchronization
+
+angle_proj = pi/8;
+shift_max = 20;
+shift_step = 4;
+dim = 3;
+
+tic
+[R, W] = compute_pairwise_alignments(image_set, angle_proj, shift_max, shift_step);
+
+%%
+
+eps = median(W(:));
+neigs = 5;
+
+%R_opt = ang_synch(R, dim);
+[R_opt, embed_coord, embed_idx, D] = vdm(R, W, eps, neigs);
+
+toc
+
+%%
+image_set_aligned = zeros(size(image_set));
+for i=1:m
+    im1 = rotate_image(R_opt(dim*(i-1)+1:dim*i,:), uint8(image_set(:,:,i)), angle_proj);
[…]

### 2014-02-14 eve data/vdm.m

@@ -1,0 +1,55 @@
+function [R_opt, embed_coord, embed_idx, D] = vdm(R, W, eps, neigs)
+
+dim = size(R,1) / size(W,1);
+
+[n, ~] = size(W);
+
+if mod(dim, 1) ~= 0
+    disp('sizes of R and W are not compatible')
+    return
+end
+
+R2 = zeros(size(R));
+W2 = exp(-W/eps);
+W2 = diag(1./sum(W2)) * W2;
+
+for i=1:n
+    for j=1:n
+        R2(dim*(i-1)+1:dim*i,dim*(j-1)+1:dim*j) = W2(i,j) * R(dim*(i-1)+1:dim*i,dim*(j-1)+1:dim*j);
+    end
+end
+
+[V, D] = eigs(R2, neigs);
+for i=1:neigs
+    V(:,i) = V(:,i) / norm(V(:,i));
+end
+if det(V(1:dim,1:dim)*D(1:dim,1:dim)*V(1:dim,1:dim)') < 0
+    V(:,dim) = -V(:,dim);
+end
+
+R_opt = V(:,1:dim);
+for i=1:n
+    [u, s, v] = svd(R_opt(dim*(i-1)+1:dim*i,:));
+    R_opt(dim*(i-1)+1:dim*i,:) = u * v';
+end
+R_opt = R_opt * R_opt(1:dim,:)';
+
+% 
+% R_opt = V(:,1:dim)*D(1:dim,1:dim)*V(1:dim,1:dim)';
+% 
+% for i=1:n
+%     R_opt(dim*(i-1)+1:dim*i,:) = R_opt(dim*(i-1)+1:dim*i,:) /(det(R_opt(dim*(i-1)+1:dim*i,:))^(1/dim));
+% end
+
+embed_coord = zeros(n, neigs*(neigs-1)/2+neigs);
+embed_idx = zeros(2, neigs*(neigs-1)/2+neigs);
+curr_idx = 1;
+for i=1:neigs
+    for j=1:i
+        embed_coord(:, curr_idx) = sum(reshape(V(:,i),dim, []).*reshape(V(:,j),dim, []))';
+        embed_idx(1, curr_idx) = i;
+        embed_idx(2, curr_idx) = j;
+        curr_idx = curr_idx + 1;
+    end
+end
+        

### 2014-02-17 image_analysis_paper/background_reading/article_notes.tex
Added more notes for background reading

@@
 
-\usepackage{biblatex}
+\usepackage[backend=bibtex]{biblatex}
 
@@
 \item \fullcite{jordan2013behavioral}
+\begin{itemize}
+	\item Look at phenotypic diversity of the swimming of the ciliate {\em Tetrahymena}
+	\item Record the swimming trajectories of many samples over their lifetimes
+	\item Define similarity metric between trajectories (compare distributions of speeds and angular velocities)
+	\item Use multidimensional scaling (MDS) to embed the trajectories in a lower-dimensional space
+	\item Find that there are two ``important'' dimensions that describe all trajectories
+	\item Can then look at whether these (low-dimensional) phenotypes are preserved through generations
+\end{itemize}
 
@@
 \item \fullcite{schindler2007inferring}
-
+\begin{itemize}
+	\item Want to order images that span many years
+	\item Solve a constraint satisfaction problem (CSP)
+	\item Find features in images, and then try to match up the features between the different images
+	\item Use structure from motion (SFM) algorithms to construct 3D points and camera angles from 2D images
+	\item Given the sets of features in each image, they then try to order the images in a way that is ``most consistent'' with how buildings are constructed in the real world (i.e.
+buildings are built, persist for some time, and then can be torn down)-- this is what they call the CSP
+\end{itemize}
 \item \fullcite{bigot2012geometric}
@@
 \item \fullcite{sisan2010event}
+\begin{itemize}
+	\item Looks at time correlations in data to determine temporal ordering
+	\item Goal is to determine in what order cellular events happen (i.e. does rxn A lead to rxn B, or does B lead to A)
+	\item They look at the time-correlation between events.
+	\item Asymmetry in time-correlation tells you in what order events happen (depending on model)
+	\item Looking at time correlation over many samples integrates/averages out stocahsticity in system
+\end{itemize}
 

### 2014-02-17 image_analysis_paper/paper_draft.tex
updated paper draft

@@
 %
-Instead, we will use a technique called angular synchronization\cite{singer2011angular}.
+Instead, we will use a technique called angular synchronization\cite{singer2011angular} to align our sets of images.
 %
-Angular synchronization uses each image as a template for every other image to compute the pairwise alignments for every possible pair of images.
+Angular synchronization uses each image as a template for every other image and computes the pairwise alignments for every possible pair of images.
 %
@@
 %
-However, we can ``compactify'' this group by mapping $ISO(2)$ into the group $SO(3)$, the group of three--dimensional rotations.
+However, we can ``compactify'' this group by mapping $ISO(2)$ to the group $SO(3)$, the group of three--dimensional rotations.
 %
@@
 %
-Now, rotations in two--dimensions correspond to rotations around one principal axis in three--dimensions, and translations in two--dimensions correspond (approximately) to rotations around the other two principal axes in three--dimensions.
+Rotations in two--dimensions now correspond to rotations around one principal axis in three--dimensions, and translations in two--dimensions correspond (approximately) to rotations around the other two principal axes in three--dimensions.
 

### 2014-02-17 image_analysis_paper/paper_draft.tex
added more stuff for paper (new figures)

@@
 \DeclareMathOperator*{\argmin}{arg\,min}
+
+\usepackage{graphicx}
+\graphicspath{ {paper_figures/} }
+
+\usepackage{forloop}
 
@@
 %
-``Template--based'' alignment \cite{...} is one technique, where one selects a specific image or {\em template}, and then tries to optimally align each image to that template.
+``Template--based'' alignment \cite{ahuja2007template} is one technique, where one selects a specific image or {\em template}, and then tries to optimally align each image to that template.
 %
@@
 %
-Angular synchronization uses each image as a template for every other image and computes the pairwise alignments for every possible pair of images.
+Angular synchronization first computes the pairwise alignments for every possible pair of images, effectively using each image as a template for every other image.
 %
@@
 
-An important part of angular synchronization is that the symmetry group $G$ be {\em compact} and have a real (or perhaps complex) representation.
+An important part of angular synchronization is that the symmetry group $G$ be {\em compact} and have a real (or perhaps complex) representation \cite{singer2011angular}.
 %
-Compactness guarantees that the minimizer for each pairwise alignment is contained in the group (Extreme Value Theorem). 
-%
-Having a real representation allows us to write combinations of transforms as group operations, and will allow us to therefore write our optimization problem in matrix form. 
+Compactness guarantees (by the extreme value theorem) that the minimizer for each pairwise alignment is contained in the group,
+and having a real representation allows us to write combinations of transforms as group operations, which will allow us to therefore write our optimization problem in matrix form. 
 %
@@
 Rotations in two--dimensions now correspond to rotations around one principal axis in three--dimensions, and translations in two--dimensions correspond (approximately) to rotations around the other two principal axes in three--dimensions.
+%
+Furthermore, successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
 
-Let $x_1, \dots, x_m \in \mathbb{R}^{n \times n}$ denote the images that we wish to align with respect to translations and rotations.
+Let $x_1, \dots, x_m \in \mathbb{R}^{n \times n}$ denote the (grayscale) images that we wish to align with respect to translations and rotations.
 %
-The first step in angular synchronization is to compute the alignments between each pair of data points. 
+The first step in angular synchronization is to compute the translations and rotations to optimally align each pair of data points. 
 %
-For each image pair $x_i$ and $x_j$, we want to find the optimal translation + rotation that aligns $x_i$ to $x_j$.
+For each image pair $x_i$ and $x_j$, we would like to solve
+\begin{equation}\label{eq:opt_pairwise}
+(\theta_{ij}, dx_{ij}, dy_{ij}) = \argmin_{\theta \in [0, 2\pi), dx \in [-\Delta, \Delta], dy \in [-\Delta, \Delta]} \|f(x_j, \theta, dx, dy) - x_i \|_F^2.
+\end{equation}
+where $\| \cdot \|_F$ denotes the Frobenius norm, and $f$ is a function that first rotates the image $x_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction.
 %
-We discritize the search space of rotations into 20 possible rotations ($0, \pi/10, \pi/5, \dots, 19\pi/10, 2\pi$), and we allow translations of $-20, -16, -12, \dots, 12, 16, 20$ pixels in both the $x$ and $y$ directions.
+$\Delta$ is the maximum number of pixels by which we are allowed to translate the image; we take $\Delta=20$, which corresponds to a 20\% shift in the image, $\Delta$ can be taken larger or smaller depending on the nature of the images.
 %
-We then exhaustively search over this space for the best rotation + translation that aligns $x_i$ to $x_j$. 
-\begin{equation}
-(\theta_{ij}, dx_{ij}, dy_{ij}) = \argmin_{\theta, dx, dy} \|f(x_j, \theta, dx, dy) - x_i \|^2.
+The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
+%
+Therefore, instead of using an optimization procedure to compute the solution to \eqref{eq:opt_pairwise}, we choose to discretize the search space and exhaustively search to estimate the solution to \eqref{eq:opt_pairwise}.
+%
+We discritize the search space of rotations into 20 possible rotations ($0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10$), and 11 possible translations in both the $x$ and $y$ directions ($-20, -16, -12, \dots, 12, 16, 20$ pixels).
+%
+We then check all possible combinations for the best rotation and translations that align $x_i$ to $x_j$. 
+%
+Although this can be somewhat time intensive, it is not prohibitive for the size of data sets which we are considering, and can be trivially parallelized for larger data sets if necessary.
+%
+The solution to this search will not be the exact solution to \eqref{eq:opt_pairwise}, but it will (most likely) be a close approximation.
+%
+Since our techniques are robust to noise, close approximations will be sufficient to obtain accurate results.
+
+We now compute the three--dimensional rotation, $R_{ij} \in SO(3)$, from the two--dimensional parameters $\theta_{ij}$, $dx_{ij}$, and $dy_{ij}$.
+%
+In the space of three--dimensional rotations, the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$ correspond to
+\begin{equation} \label{eq:angle_relations}
+\begin{aligned}
+	\alpha_{ij} &= \theta_{ij} \\
+	\beta_{ij} &= \frac{dx_{ij}}{n} \times \eta_{proj} \\
+	\gamma_{ij} &= \frac{dy_{ij}}{n} \times \eta_{proj} \\
+\end{aligned}
 \end{equation}
-where $f$ is a function that first rotates the image $x_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction.
-%
-The norm is the Frobenius norm of the resulting matrix. 
-%
-This means we check each of the 2420 possible combinations; although this is somewhat time intensive (TODO: add timings), it is not prohibative and can be trivially parallelized for larger data sets if necessary.
-%
-Let $\theta$ denote the optimal angle of rotation, and let $dx$ and $dy$ denote the optimal number of pixels to translate in the $x$ and $y$ directions, respectively.
-%
-Then, in the space of three--dimensional rotations, the Euler angles $\alpha$, $\beta$, and $\gamma$ correspond to
-\begin{eqnarray}
-	\alpha_{ij} &=& \theta_{ij} \\
-	\beta_{ij} &=& \frac{dx_{ij}}{npixels} \times \eta_{proj} \\
-	\gamma_{ij} &=& \frac{dy_{ij}}{npixels} \times \eta_{proj} \\
-\end{eqnarray}
 where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project (we take $\eta_{proj} =  \pi/8$).
@@
 From here, we can write rotations around the three principal axes in terms of the three Euler angles
-\begin{eqnarray}
-	R^x_{ij} &=& \begin{bmatrix}
+\begin{equation}
+\begin{aligned}
+	R^x_{ij} &= \begin{bmatrix}
 	1 & 0 & 0 \\
@@
 	\end{bmatrix} \\
-	R^y_{ij} &=& \begin{bmatrix}
+	R^y_{ij} &= \begin{bmatrix}
 	\cos(\beta) & 0 & \sin(\beta) \\
@@
     \end{bmatrix} \\
-	R^z_{ij} &=& \begin{bmatrix} 
+	R^z_{ij} &= \begin{bmatrix} 
 	\cos(\gamma) & -\sin(\gamma) & 0 \\
@@
     \end{bmatrix}
-\end{eqnarray}
+\end{aligned}
+\end{equation}
+where $R^x_{ij}$, $R^y_{ij}$, and $R^z_{ij}$ correspond to rotations around the $x$, $y$, and $z$ axes, respectively.
 %
@@
 \end{equation}
-This corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (because the rotation matrices operate on the left, the order of operations/transformations is read from right to left in \eqref{eq:total_rot}.
+%
+This corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).
 
-Let $R_{ij} \in SO(3)$ denote the rotation that aligns $x_j$ to $x_i$, so that
-\begin{equation}
-R_{ij} = \argmin_{R \in SO(3)} \|Rx_j - x_i \|^2.
-\end{equation}
-%
-We want to find the rotations $R_1, R_2, \dots, R_n \in SO(d)$ such that $R_i R_j^T \approx R_{ij}$, for every pair $i, j$. 
+Given the pairwise rotations that optimally align the imges, $R_{ij}$, we would like to find the rotations $R_1, R_2, \dots, R_n \in SO(3)$ such that $R_i R_j^T \approx R_{ij}$, for every pair $i, j$.
 %TODO: Check that this order is correct
@@
 
-We would like to solve the optimization problem 
+We formulate our problem as an optimization problem:
 \begin{equation} \label{eqn:angsynch_obj}
@@
 %
-We define the matrix $H$, which is an $m \times m$ block matrix with $3 \times 3$ blocks (so $H \in \mathbb{R}^{3m \times 3m}$), with $H_{ij} = R_{ij}$.
+We define the matrix $H \in \mathbb{R}^{3m \times 3m}$, which is an $m \times m$ block matrix with $3 \times 3$ blocks, with $H_{ij} = R_{ij}$.
 %
@@
 \begin{equation} \label{eqn:angsynch_relax}
-\max_{R\in \mathbb{3m \times 3}} \| R^T H R \|^2.
+\max_{R\in \mathbb{R}^{3m \times 3}} \| R^T H R \|^2.
 \end{equation}
@@
 %
-The $i^{th}$ rotation is (approximately) $\hat{R}(i) \in \mathbb{R}^{3 \times 3}$.
+The $i^{th}$ rotation is (approximately) the $i^{th}$ block of $\hat{R}$, $\hat{R}(i) \in \mathbb{R}^{3 \times 3}$.
 %
@@
 %
-For example, if we want to optimize over all pairs of pairwise rotations, we would solve $\max_{R_1, \dots, R_n \in SO(d)} \sum_{i=1}^{n} \sum_{j=1}^{n} \left\| R_i^T \left( \sum_k R_{ik} R_{kj} \right) R_j \right\| = \| R^T H^2 R \|^2$.
+For example, if we want to optimize over all pairs of pairwise rotations, we would solve $\max_{R_1, \dots, R_m \in SO(3)} \sum_{i=1}^{m} \sum_{j=1}^{m} \left\| R_i^T \left( \sum_k R_{ik} R_{kj} \right) R_j \right\| = \| R^T H^2 R \|^2$.
 %
@@
 %
-The optimal rotation is therefore given by $R_i = U_i V_i^T$, where $U_i$ and $V_i$ are the left and right singular vectors of $\hat{R}_i$, respectively. 
+The optimal rotation is therefore given by $R_i = U_i V_i^T$, where $U_i$ and $V_i$ are the left and right singular vectors of $\hat{R}_i$, respectively \cite{...}. 
 %
-Because the eigenvectors of $H$ are determined up to a sign, we adjust the signs of the eigenvectors so that $det(R_i) = +1$.
+Because the eigenvectors of $H$ are determined up to a sign, we adjust the signs of the eigenvectors so that $det(R_i) = +1$ to ensure proper rotations.
+%
+
+From the matrices $R_1, \dots, R_m$, we now wish to find the corresponding translations and rotations of the images.
+%
+Because we are permitted to adjust our rotations by a global rotations, we multiply all rotations by $R_1^T$ to ensure that all the images are (approximately) in the same region of the sphere where we began.
+%
+We then compute the Euler angles $\alpha$, $\beta$, and $\gamma$ from the rotation matrix $R$ using the following relationships
+\begin{equation}
+\begin{aligned}
+R_{1,1} & = \cos(\beta)\cos(\gamma) \\
+R_{2,1} & = \cos(\beta)\sin(\gamma) \\
+R_{3,1} & = -\sin(\beta) \\
+R_{3,2} & = \sin(\alpha)\cos(\beta) \\
+R_{3,3} & = \cos(\alpha)\cos(\beta) 
+\end{aligned}
+\end{equation}
+%
+We can then invert \eqref{eq:angle_relations} to compute the optimal translation and rotation for each image.
 
@@
 
-Unlike PCA, diffusion maps (DMAPS) is a nonlinear dimensionality reduction technique. 
+Given our set of aligned images, we wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
 %
-DMAPS aims to uncover a parameterization of high-dimensional data sampled from a low-dimensional nonlinear manifold \cite{coifman2005geometric}.
+We will use diffusion maps (DMAPS) \cite{coifman2005geometric} to temporally order our data.
+%
+DMAPS aims to uncover a parameterization of high-dimensional data sampled from a low-dimensional nonlinear manifold.
 %
@@
 This can as simple as the standard Euclidean distance, or a more complex metric (such as a distance between features of the data points) for other data sets.
+%
+We will use the Frobenius norm of the difference of our aligned pictures as our distance metric, $d(x_i, x_j) = \| x_i - x_j \|_F$.
 
-Given data points $x_1, \dots, x_n \in \mathbb{R}^p$, we fist calculate the matrix $W \in \mathbb{R}^{n \times n}$, where 
+Given {\em aligned} images $x_1, \dots, x_m \in \mathbb{R}^{n \times n}$, we fist calculate the matrix $W \in \mathbb{R}^{m \times m}$, where 
 \begin{equation}
-W_{ij} = \exp \left( -\frac{d^2(x_i, x)j)}{\epsilon^2} \right)
+W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right)
 \end{equation}
@@
 %
-We then compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{n} W_{ij}$, and the matrix $A = D^{-1} W$. 
+We then compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{m} W_{ij}$, and the matrix $A = D^{-1} W$. 
 %
-We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_n$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_n$ and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_n|$.
+We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_m$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_m$ and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$.
 %
@@
 %
-The eigenvectors $\phi_1, \phi_2, \dots, \phi_n$ give the embedding coordinates, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point. 
+The eigenvectors $\phi_1, \phi_2, \dots, \phi_m$ give the embedding coordinates, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point. 
 %
@@
 
-\subsection{Aligning and Ordering Using Vector diffusion maps}
+In our setup, we assume that our data are inherently one--dimensional, and that this one dimension is correlated with time.
+%
+Therefore, ordering our data by $\phi_2(j)$ will allow us to automatically order our data in time.
 
-Angular synchronization assumes that each data point is a replicate measurements of the same underlying configuration, simply corrupted with rotations and some noise, and therefore considers all pairwise measurements equally important.
+\subsection{Vector Diffusion Maps}
+
+Many times, we would to both align and order our data.
 %
-However, in system with an underlying symmetry group {\em as well as} a dynamical process, the data points are not identical modulo symmetries and noise, and we would like to place more emphasis on aligning data points which are dynamically close.
+Our proposed alignment methodology, angular syncrhonization, utilizes the eigendecompostion of the matrix $R \in \mathbb{R}^{3m \times 3m}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
+%
+We can combine the two steps into one eigencomputation that allows us to {|em simultaneously} recover the optimal alignments and the temporal ordering.
+%
+This technique is called vector diffusion maps \cite{singer2012vector}.
+%
+It is significantly more robust to mistakes in the alignment.
 
-Vector diffusion maps (VDM) couples the symmetry-reduction of angular synchronization with the automatic parameterization of diffusion maps \cite{singer2012vector}. 
-%
-We construct the matrix $A$ as in diffusion maps, and the matrix $H$ as in angular synchronization.
-%
-We then construct the block matrix $S$, where the blocks of $S$ are defined by
-\begin{equation}
-S_{ij} = \left\{ \begin{array}{l l} 
-A_{ij} H_{ij} & i \ne j \\
-0_{d \times d} & i = j
-\end{array}
-\right.
-\end{equation}
-%
-where $A_{ij}$ is the $(i,j)$ entry of $A$, and $H_{ij}$ is the $(i,j)$ block of $H$.
-
-We compute the eigenvectors $\phi_1, \dots, \phi_{nd}$ and eigenvalues $\lambda_1, \dots, \lambda_{nd}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{nd}|$.
-%
-As in angular synchronization, the first ``block'' eigenvector, i.e. the first $d$ eigenvectors concatenated into a $nd \times d$ matrix, gives the (relaxed) optimal rotations. 
-%
-However, the eigenvectors now also give us embedding coordinates.
-%
-Let $\phi_j(i)$ denote the $i^{th}$ block of $\phi_j$ (so that $\phi_j(i) \in \mathbb{R}^d$). 
-%
-The embedding coordinates of $x_i$ are then given by $\langle \phi_j(i), \phi_k(i) \rangle$, where $1 \le j, k, \le nd$.
+For vector diffusion maps, we construct the matrix $S \in \mathbb{R}^{3m \times 3m}$
 
@@
 
+\def \nimages {81}
+\def \imagewidth {0.22in}
+\begin{figure}
+\begin{minipage}{0.5\textwidth}
+\newcounter{n}
+\forloop{n}{1}{\value{n} < \nimages}{
+	\includegraphics[width=\imagewidth]{dpERK_\arabic{n}}
+}
+\hspace{0pt}
+\end{minipage}
+\includegraphics[width=0.45\textwidth]{dpERK_rank_corr}
+\end{figure}
+
 \subsection{Dorsal}
+
+\begin{figure}
+\begin{minipage}{0.5\textwidth}
+\forloop{n}{1}{\value{n} < \nimages}{
+	\includegraphics[width=\imagewidth]{DI_\arabic{n}}
+	\hfill
+}
+\end{minipage}
+\includegraphics[width=0.45\textwidth]{DI_rank_corr}
+\end{figure}
 
@@
 
+\bibliographystyle{plain}
+\bibliography{../../references/references}
+
 \end{document}

### 2014-02-17 image_analysis_paper/make_figures_large_dataset.m

@@ -11,7 +11,7 @@ print_figures = false;
 
 dpERK_data = '../membrane_pictures/large_dataset/time.mat';
 dpERK_image_dir = '../membrane_pictures/large_dataset';
-%dpERK_membrane_dir = '../membrane_pictures/membrane2';
+im_save_dir = 'paper_figures';
 
 %% load membrane lengths
 
@@ -20,7 +20,6 @@ mem_lengths = length;
 clear length;
 
 ind = setdiff(1:90, [3, 20, 25, 32, 53, 59, 61, 66, 82]);
-%ind = randsample(ind, 25);
 
 m = length(ind);
 
@@ -36,7 +35,7 @@ subplot_dim2 = ceil(m / subplot_dim1);
 
 image_set = zeros(npixels, npixels, m);
 
-image_channel = 2;
+image_channel = 3;
 
 %figure;
 for i=1:m
@@ -58,25 +57,6 @@ for i=1:m
     
 end
 
-%%
-% m = 81;
-% [~, ind] = sort(mem_lengths);
-% ind = ind(end-m+1:end);
-% ind = ind(randperm(m));
-% 
-% mem_lengths = mem_lengths(ind);
-% image_set = image_set(:,:,ind);
-% 
-% %set image plotting parameters
-% subplot_dim1 = ceil(sqrt(m));
-% subplot_dim2 = ceil(m / subplot_dim1);
-
-% figure;
-% for i=1:m
-%     subplot(subplot_dim1, subplot_dim2, i)
-%     imshow(uint8(image_set(:,:,i)));
-% end
-
 %% raw dpERK images-- synchronization
 
 angle_proj = pi/8;
@@ -87,7 +67,7 @@ dim = 3;
 tic
 [R, W] = compute_pairwise_alignments(image_set, angle_proj, shift_max, shift_step);
 
-%%
+%% VDM
 
 eps = median(W(:));
 neigs = 5;
@@ -106,28 +86,42 @@ for i=1:m
     imshow(im1);
 end
 
-%[~, idx] = max(abs(corr(embed_coord, mem_lengths)));
-idx = 7;
-
-if corr(mem_lengths, embed_coord(:, idx)) < 0
-    embed_coord(:, idx) = -embed_coord(:, idx);
+image_set_aligned = zeros(size(image_set));
+for i=1:m
+    im1 = rotate_image(R_opt(dim*(i-1)+1:dim*i,:), uint8(image_set(:,:,i)), angle_proj);
[…]

### 2014-02-17 image_analysis_paper/background_reading/article_notes.tex
added more notes for literature

@@
 \item \fullcite{wei2006conditional}
+\begin{itemize}
+		\item Want to develop a more informative growth chart metric
+		\item Instead of looking at how atypical a current measurement is for an individual, they look at how atypical a time series is for an individual 
+		\item This means that a person will be flagged if, for example, their weight changes dramatically in a short period of time, even if both measurements are within the ``normal'' range
+		\item Fit model to data for ``normal'' time--series behavior
+\end{itemize}
 
 \item \fullcite{wei2006quantile}
+\begin{itemize}
+	\item Develop nonparameteric models (using spline regression) for growth charts
+	\item Typical growth charts rely on normality assumptions (data is often transformed to fulfill such assumptions)
+	\item Using nonparametric models can allow them to fit a much broader range of behaviors (i.e. bimodal distributions)
+\end{itemize}
 

### 2014-02-18 image_analysis_paper/paper_draft.tex
UPdated paper draft

@@
 \graphicspath{ {paper_figures/} }
+\usepackage{subcaption}
 
@@
 % only images
+We have cross-sectional data.
+When there is just a few snapshots and differences between them are obvious, such as the overall size or some other morphological feature, ordering is easy.
+We can point to "fish" as our example.
+
+But there is more and more studies with lots of snapshots, lots of noise, and subtle differences between snapshots.
+Our example is the dataset from the embryo.
+We discuss how to align and temporally order cross sectional data in this case.
+
+Maybe figure 1 is fish ordered (B) and unordered (A).
+
+Figure 2 can be the embryo dataset - unordered and unoriented (A) and hand-ordered (by membranes) and hand-oriented (by Dorsal) (B).
+
+This way Figure 2 presents the data and makes it a reference for all future alignments.
 
@@
 
-For vector diffusion maps, we construct the matrix $S \in \mathbb{R}^{3m \times 3m}$
+For vector diffusion maps, we construct the matrix $S \in \mathbb{R}^{3m \times 3m}$, with
+\begin{equation}
+	S_{ij} = A_{ij} R_{ij}
+\end{equation}
+%
+We then compute the eigenvalues and eigenvectors of $S$.
+%
+Again, the top (block) eigenvector of $S$ gives us approxiamtions to the optimal rotations for each image.
+%
+However, we now also obtain embedding coordinates for our images.
+%
+The embedding coordinates are given by 
+\begin{equation}
+\psi_{kl} (i) = \langle \phi_k(i), \phi_l(i) \rangle
+\end{equation}
+where $\phi_j$ are the eigenvectors of $S$.
 
@@
 
-\def \nimages {81}
-\def \imagewidth {0.22in}
+Show how correlation varies with number of images
+
 \begin{figure}
-\begin{minipage}{0.5\textwidth}
-\newcounter{n}
-\forloop{n}{1}{\value{n} < \nimages}{
-	\includegraphics[width=\imagewidth]{dpERK_\arabic{n}}
-}
-\hspace{0pt}
-\end{minipage}
-\includegraphics[width=0.45\textwidth]{dpERK_rank_corr}
+\begin{subfigure}[c]{0.45\textwidth}
+\includegraphics[width=\textwidth]{dpERK_array}
+\caption{}
+\end{subfigure}
+\begin{subfigure}[c]{0.5\textwidth}
+\includegraphics[width=\textwidth]{dpERK_rank_corr}
+\caption{}
+\end{subfigure}
 \end{figure}
+
+ADD TEXT HERE
 
@@
 \begin{figure}
-\begin{minipage}{0.5\textwidth}
-\forloop{n}{1}{\value{n} < \nimages}{
-	\includegraphics[width=\imagewidth]{DI_\arabic{n}}
-	\hfill
-}
-\end{minipage}
-\includegraphics[width=0.45\textwidth]{DI_rank_corr}
+\begin{subfigure}[c]{0.45\textwidth}
+\includegraphics[width=\textwidth]{DI_array}
+\caption{}
+\end{subfigure}
+\begin{subfigure}[c]{0.5\textwidth}
+\includegraphics[width=\textwidth]{DI_rank_corr}
+\caption{}
+\end{subfigure}
 \end{figure}
@@
 % compare to TSP algorithms
+Discussion of what we did and how it relates to other efforts (TSP, etc) and how it can be extended.
 

### 2014-02-18 image_analysis_paper/vdm.m

@@ -20,6 +20,9 @@ for i=1:n
 end
 
 [V, D] = eigs(R2, neigs);
+[~, ind] = sort(abs(diag(D)), 'descend');
+V = V(:, ind);
+D = D(ind, ind);
 for i=1:neigs
     V(:,i) = V(:,i) / norm(V(:,i));
 end

### 2014-02-19 image_analysis_paper/paper_draft.tex
added paper updates (bootsrtap, etc.)

@@
-\documentclass[12pt]{article}
+\documentclass[twocolumn, 11pt]{article}
 
@@
 
+Let $I_1, \dots, I_m \in \mathbb{R}^{n \times n}$ denote the (grayscale) images that we wish to align with respect to translations and rotations.
+%
+The first step in angular synchronization is to compute the translations and rotations to optimally align each pair of data points. 
+%
+For each image pair $I_i$ and $I_j$, we would like to solve
+\begin{equation}\label{eq:opt_pairwise}
+(\theta_{ij}, dx_{ij}, dy_{ij}) = \argmin_{
+\begin{matrix}
+\theta \in [0, 2\pi) \\ 
+dx \in [-\Delta, \Delta]\\ 
+dy \in [-\Delta, \Delta]
+\end{matrix}
+} \|f(I_j, \theta, dx, dy) - I_i \|_F^2.
+\end{equation}
+where $\| \cdot \|_F$ denotes the Frobenius norm, and $f(I_j, \theta, dx, dy)$ is a function that first rotates the image $I_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction.
+%
+$\Delta$ is the maximum number of pixels by which we are allowed to translate the image (we take $\Delta=20$, which corresponds to a 20\% shift in the image). 
+%
+The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
+%
+Therefore, instead of using an optimization procedure to compute the solution to \eqref{eq:opt_pairwise}, we choose to discretize the search space and exhaustively search to estimate the solution to \eqref{eq:opt_pairwise}.
+%
+We discritize the search space of rotations into 20 possible rotations ($0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10$), and 11 possible translations in both the $x$ and $y$ directions ($-20, -16, -12, \dots, 12, 16, 20$ pixels).
+%
+We then check all possible combinations for the best rotation and translations that align $I_i$ to $I_j$. 
+%
+Although this can be somewhat time intensive, it is not prohibitive for the size of data sets which we are considering, and can be trivially parallelized for larger data sets if necessary.
+%
+The solution to this search will not be the exact solution to \eqref{eq:opt_pairwise}, but it will (most likely) be a close approximation.
+%
+Since our techniques are robust to noise, close approximations will be sufficient to obtain accurate results.
+
 An important part of angular synchronization is that the symmetry group $G$ be {\em compact} and have a real (or perhaps complex) representation \cite{singer2011angular}.
@@
 Compactness guarantees (by the extreme value theorem) that the minimizer for each pairwise alignment is contained in the group,
-and having a real representation allows us to write combinations of transforms as group operations, which will allow us to therefore write our optimization problem in matrix form. 
+and having a real representation will allow us to write our problem in matrix form. 
 %
@@
 
-Let $x_1, \dots, x_m \in \mathbb{R}^{n \times n}$ denote the (grayscale) images that we wish to align with respect to translations and rotations.
-%
-The first step in angular synchronization is to compute the translations and rotations to optimally align each pair of data points. 
-%
-For each image pair $x_i$ and $x_j$, we would like to solve
-\begin{equation}\label{eq:opt_pairwise}
-(\theta_{ij}, dx_{ij}, dy_{ij}) = \argmin_{\theta \in [0, 2\pi), dx \in [-\Delta, \Delta], dy \in [-\Delta, \Delta]} \|f(x_j, \theta, dx, dy) - x_i \|_F^2.
-\end{equation}
-where $\| \cdot \|_F$ denotes the Frobenius norm, and $f$ is a function that first rotates the image $x_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction.
-%
-$\Delta$ is the maximum number of pixels by which we are allowed to translate the image; we take $\Delta=20$, which corresponds to a 20\% shift in the image, $\Delta$ can be taken larger or smaller depending on the nature of the images.
-%
-The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
-%
-Therefore, instead of using an optimization procedure to compute the solution to \eqref{eq:opt_pairwise}, we choose to discretize the search space and exhaustively search to estimate the solution to \eqref{eq:opt_pairwise}.
-%
-We discritize the search space of rotations into 20 possible rotations ($0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10$), and 11 possible translations in both the $x$ and $y$ directions ($-20, -16, -12, \dots, 12, 16, 20$ pixels).
-%
-We then check all possible combinations for the best rotation and translations that align $x_i$ to $x_j$. 
-%
-Although this can be somewhat time intensive, it is not prohibitive for the size of data sets which we are considering, and can be trivially parallelized for larger data sets if necessary.
-%
-The solution to this search will not be the exact solution to \eqref{eq:opt_pairwise}, but it will (most likely) be a close approximation.
-%
-Since our techniques are robust to noise, close approximations will be sufficient to obtain accurate results.
-
-We now compute the three--dimensional rotation, $R_{ij} \in SO(3)$, from the two--dimensional parameters $\theta_{ij}$, $dx_{ij}$, and $dy_{ij}$.
+We now compute the three--dimensional rotation, $R_{ij} \in SO(3)$, from the two--dimensional translation and rotation parameters $\theta_{ij}$, $dx_{ij}$, and $dy_{ij}$.
 %
@@
 %
-This corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).
+\eqref{eq:total_rot} corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).
 
@@
 %
-The solution to \eqref{eqn:angsynch_relax} is given by the top ``block'' eigenvector of $H$, which we denote $\hat{R} \in \mathbb{R}^{3m \times 3}$. 
+The solution to \eqref{eqn:angsynch_relax} is given by the top ``block'' eigenvector of $H$ (i.e.
+the three eigenvectors corresponding to the three largest eigenvalues, concatenated into a $3m \times 3$ matrix), which we denote $\hat{R} \in \mathbb{R}^{3m \times 3}$.
 %
@@
 %
-For example, if we want to optimize over all pairs of pairwise rotations, we would solve $\max_{R_1, \dots, R_m \in SO(3)} \sum_{i=1}^{m} \sum_{j=1}^{m} \left\| R_i^T \left( \sum_k R_{ik} R_{kj} \right) R_j \right\| = \| R^T H^2 R \|^2$.
+For example, if we want to optimize over all pairs of pairwise rotations, we would solve 
+\begin{equation}
+\max_{R_1, \dots, R_m \in SO(3)} \sum_{i=1}^{m} \sum_{j=1}^{m} \left\| R_i^T \left( \sum_k R_{ik} R_{kj} \right) R_j \right\|,
+\end{equation} 
 %
-However, the solution to this problem is also given by the top eigenvector of $H$.
+which relaxes to the problem $ \max_{R \in \mathbb{R}^{3m \times 3}} \| R^T H^2 R \|^2$.
+%
+However, the solution to this problem is also given by the top block eigenvector of $H$.
 
@@
 %
-The optimal rotation is therefore given by $R_i = U_i V_i^T$, where $U_i$ and $V_i$ are the left and right singular vectors of $\hat{R}_i$, respectively \cite{...}. 
+The optimal rotations are given by $R_i = U_i V_i^T$, where $U_i$ and $V_i$ are the left and right singular vectors of $\hat{R}_i$, respectively \cite{...}. 
 %
@@
 %
-Because we are permitted to adjust our rotations by a global rotations, we multiply all rotations by $R_1^T$ to ensure that all the images are (approximately) in the same region of the sphere where we began.
+Because we are permitted to adjust our rotations by a global rotations, we first multiply all rotations by $R_1^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
 %
@@
 %
-We can then invert \eqref{eq:angle_relations} to compute the optimal translation and rotation for each image.
+We can then invert \eqref{eq:angle_relations} to compute the optimal translation and rotation for each image from the Euler angles.
 
@@
 
-Given our set of aligned images, we wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
+Given our set of aligned images, we now wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
 %
@@
 %
-The {\em essential} requirement for DMAPS is an appropriate distance metric $d(x_i, x_j)$ for comparing data points. 
+The {\em essential} requirement for DMAPS is an appropriate distance metric $d(I_i, I_j)$ for comparing data points. 
 %
 This can as simple as the standard Euclidean distance, or a more complex metric (such as a distance between features of the data points) for other data sets.
-%
-We will use the Frobenius norm of the difference of our aligned pictures as our distance metric, $d(x_i, x_j) = \| x_i - x_j \|_F$.
 
-Given {\em aligned} images $x_1, \dots, x_m \in \mathbb{R}^{n \times n}$, we fist calculate the matrix $W \in \mathbb{R}^{m \times m}$, where 
-\begin{equation}
-W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right)
+Given {\em aligned} images $I_1, \dots, I_m \in \mathbb{R}^{n \times n}$, we fist calculate the matrix $W \in \mathbb{R}^{m \times m}$, where 
+\begin{equation} \label{eq:dmaps_W}
+W_{ij} = \exp \left( -\frac{d^2(I_i, I_j)}{\epsilon^2} \right)
 \end{equation}
@@
 %
-Our proposed alignment methodology, angular syncrhonization, utilizes the eigendecompostion of the matrix $R \in \mathbb{R}^{3m \times 3m}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
+Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{3m \times 3m}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
 %
-We can combine the two steps into one eigencomputation that allows us to {|em simultaneously} recover the optimal alignments and the temporal ordering.
+We can combine the two steps into one eigencomputation that allows us to {\em simultaneously} recover the optimal alignments and the temporal ordering.
 %
@@
 %
-It is significantly more robust to mistakes in the alignment.
+It not only reduces the number of required eigencomputations, but also is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
 
@@
 \begin{equation}
-	S_{ij} = A_{ij} R_{ij}
+	S_{ij} = A_{ij} H_{ij}
 \end{equation}
 %
-We then compute the eigenvalues and eigenvectors of $S$.
+where $A$ is constructed from the matrix $W$ \eqref{eq:dmaps_W}, and the distance metric $d(I_i, I_j)$ is the Frobenius norm between the images {\em after} optimal pairwise alignment ($d(I_i, I_j) = \| f(I_j, \theta_{ij}, dx_{ij}, dy_{ij}) - I_i \|_F$).
+
+We then compute the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_{3m}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{3m}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{3m}|$.
 %
-Again, the top (block) eigenvector of $S$ gives us approxiamtions to the optimal rotations for each image.
+Again, the top (block) eigenvector of $S$ gives us approximations to the optimal rotations for each image.
 %
@@
 \begin{equation}
-\psi_{kl} (i) = \langle \phi_k(i), \phi_l(i) \rangle
+\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle
 \end{equation}
-where $\phi_j$ are the eigenvectors of $S$.
+where $\phi_k(i) \in \mathbb{R}^3$ denotes the $i^{th}$ block of $\phi_k$.
+%
+If we assume that our data are one--dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, then we can order our data by $\psi_{1,4}$.
+%
+If no such claims can be made, then one often runs a second step of diffusion maps, using $\psi_{k,l}$ as the data points, in order to uncover the true manifold modulo symmetries.
 
@@
 
-Show how correlation varies with number of images
-
-\begin{figure}
+\begin{figure*}[h]
 \begin{subfigure}[c]{0.45\textwidth}
@@
 \end{subfigure}
-\end{figure}
+\caption{(a) Ordering of images fluorescently stained for dpERK. (b) Correlation between ranks from ordering using VDM, and ordering using the membrane length.}
+\label{fig:dpERK}
+\end{figure*}
 
-ADD TEXT HERE
+\begin{figure*}[h]
+\centering
+\includegraphics[width=0.5\textwidth]{bootstrap_rank_corr}
+\caption{Rank correlation as a function of the number of data points.}
+\label{fig:bootstrap}
+\end{figure*}
 
@@
 
-\begin{figure}
+\begin{figure*}[h]
 \begin{subfigure}[c]{0.45\textwidth}
@@
 \end{subfigure}
-\end{figure}
+\caption{(a) Ordering of images fluorescently stained for Dorsal. (b) Correlation between ranks from ordering using VDM, and ordering using the membrane length.}
+\label{fig:Dl}
+\end{figure*}
 

### 2014-02-19 image_analysis_paper/background_reading/article_notes.tex
added updates to background reading notes

@@
 \item \fullcite{li2009stochastic}
+\begin{itemize}
+	\item Want to look at mechanism for {\em eve} stripe formation
+	\item Propose a large gene/protein interaction network
+	\item Use maximum likelihood to prune the network and find the most probable network that would give rise to the observed spatiotemporal dynamics
+\end{itemize}
 
@@
 \item \fullcite{peng2013detecting}
+\begin{itemize}
+	\item Wants to find which genes/SNPs are responsible for variation in facial features
+	\item Has DNA data and 3D face scans for many samples
+	\item Looks at different facial models (all are based on landmarks)
+	\item Tries to find which variations in the facial models correlate with which SNPs
+\end{itemize}
 
 \item \fullcite{gerber2010visual}
+\begin{itemize}
+	\item Want to find a ``good'' way to visualize high--dimensional functions
+	\item First find local maxima and minima in data set
+	\item Use these maxima and minima to segment the domain of the function into ``crystals''
+	\item Approximate each crystal by a line/curve
+	\item Embed the maxima and minima in 2D using PCA; draw lines connecting relevant maxima and minima
+\end{itemize}
 
 \item \fullcite{qiu2011discovering}
+\begin{itemize}
+	\item Want to order microarray data
+	\item Look at specific (high--variance) genes (~3000 genes)
+	\item Cluster genes into ``modules'' using k--means, so that similairly expressed genes are grouped together
+	\item Within each module, construct a minimium spanning tree of the data (distance between nodes is Euclidean distance)
+	\item Find set of MSTs that are most similar/consistent
+	\item Assume these are the ``right'' MST and that they uncover the progression of the data
+\end{itemize}
 
 \item \fullcite{gupta2008extracting}
+\begin{itemize}
+	\item Want to temporally order gene expression data from cancer patients
+	\item Use TSP to order gene expression data (can show that TSP path will recover the correct ordering under some probability model)
+	\item Use a probabilistic gene/patient model to model the variation from patient to patient (model underlying dynamics with splines, and then assume rest in noise from patient-to-patient variability)
+\end{itemize}
 
 \item \fullcite{lau2009modeling}
+\begin{itemize}
+	\item Want to model trajectory/motion data where each sample is a trajectory
+	\item Assume that the trajectories come from a second-order Markov model (i.e. $t+2$ depends on $t$ and $t+1$)
+	\item Fit a Bayesian network that models the trajectory evolution using training data
+	\item Can then produce new sample trajectories from this Bayesian network
+\end{itemize}
 
 \item \fullcite{toga2003temporal}
+\begin{itemize}
+	\item Want to model spatiotemporal changes in the 3D structure of the brain (so that one can detect disease, etc.)
+	\item Model growth of different portions of the brain using deformation mapping (look at deformation of different regions of a {\em single} person's brain from one scan to the next)
+	\item Can fit cortical models for each person in a population; can then look at variations across these models
+\end{itemize}
 
 \item \fullcite{li2012modelling}
+\begin{itemize}
+	\item Want to use cross--sectional data to model disease progression
+	\item Data are things that are known to be important/telling for that particlular disease (i.e.
+breast cancer data contains descriptors of cell morphology, Parkinson's disease data contains descriptors of speech patterns)
+	\item Create ``pseudo time series'' by subsampling data, and from each subsample, finding shortest path that goes from a healthy sample to a diseased sample
+	\item From these pseudo time series, can fit a hidden Markov model to find different states of disease progression and likely transitions from one state to another
+\end{itemize}
 
 \item \fullcite{denkinger2013set}
+\begin{itemize}
+	\item Want to develop a standardized data set that can be used to investigate temporal and causal neural processing
+	\item Developed a set of 222 images which contain different causal and temporal relationships
+\end{itemize}
 
 \item \fullcite{xu2012comprehensive}
-
+\begin{itemize}
+	\item Want to quantify shape of liver lesions
+	\item Take set of 14 common shape descriptors
+	\item Use regression to find optimal combination of shape descriptors that can predict the similairity using a ``similairity reference standard''
+	\item FINISH
+\end{itemize}
 \item \fullcite{sisan2010event}
@@
 
+\item \fullcite{basu2014detecting}
+
 \end{enumerate}

### 2014-02-20 image_analysis_paper/background_reading/article_notes.tex
Updated writing things

@@
 \usepackage[backend=bibtex]{biblatex}
+\usepackage[left=1in, right=1in, top=1in, bottom=1in]{geometry}
 
-\bibliography{references}
+\bibliography{references,../../../references/references}
 
 \begin{document}
+
+\section{References from Stas}
 
@@
 
+\section{Algorithms}
+
+\begin{enumerate}
+
+\item \fullcite{flusser2000independence}
+\begin{itemize}
+\item Construct moments (polynomials) of the pixel intensities that are invariant to rotations
+\end{itemize}
+
+\item \fullcite{besl1992method}
+\begin{itemize}
+	\item Iterative closest point (ICP) algorithm
+	\item Aligns/registers point clouds
+\end{itemize}
+
+\item \fullcite{lowe1999object}
+\begin{itemize}
+	\item Scale-invartiant feature transform (SIFT)
+	\item Computes feature vectors for image; features are scale, translation, and rotation invariant
+\end{itemize}
+
+\item \fullcite{sadler1992shift}
+\begin{itemize}
+	\item Calculate bispectrum for image
+	\item Bispectrum is rotation-invariant
+	\item Do not need to align the images
+\end{itemize}
+
+\item \fullcite{hilai1994recognition}
+\begin{itemize}
+	\item Construct rotation-invariant basis for images
+\end{itemize}
+
+\item \fullcite{ojala2002multiresolution}
+\begin{itemize}
+	\item Decribe local patches of image (texture) by binary code
+	\item Define function on binary code that is rotation-invariant
+	\item Classify using binary code
+\end{itemize}
+
+\item \fullcite{zhao2013fourier}
+\begin{itemize}
+	\item Construct PCA basis that is rotation-invariant
+\end{itemize}
+
+\end{enumerate}
+
+\section{Applications}
+
+\begin{enumerate}
+
+\item \fullcite{hajnal2010medical}
+\begin{itemize}
+	\item Medical image registration (MRI, CT, etc.)
+\end{itemize}
+
+\item \fullcite{greenspan1994rotation}
+\begin{itemize}
+\item Texture recognition/classification
+\item use Fourier basis; fit texture response (in Fourier basis) to rotations 
+\end{itemize}
+
+\item \fullcite{rowley1998rotation}
+\begin{itemize}
+	\item face recognition
+	\item train neural net to estimate face orientation, then orient faces
+\end{itemize}
+
+\item \fullcite{simard1992efficient}
+\begin{itemize}
+	\item recognize digits
+	\item only consider small rotations; fit plane to span of rotations of each digit, then see which plane new digit is closest to
+\end{itemize}
+
+\end{enumerate}
+
 \end{document}

### 2014-02-22 image_analysis_paper/math_description.tex
added math description

--- 2014-02-19 image_analysis_paper/paper_draft.tex
@@
-\documentclass[twocolumn, 11pt]{article}
+\documentclass[11pt]{article}
 
@@
 
-\usepackage{graphicx}
-\graphicspath{ {paper_figures/} }
-\usepackage{subcaption}
-
-\usepackage{forloop}
-
 \begin{document}
 
-\section{Introduction}
-
-% only images
-We have cross-sectional data.
-When there is just a few snapshots and differences between them are obvious, such as the overall size or some other morphological feature, ordering is easy.
-We can point to "fish" as our example.
-
-But there is more and more studies with lots of snapshots, lots of noise, and subtle differences between snapshots.
-Our example is the dataset from the embryo.
-We discuss how to align and temporally order cross sectional data in this case.
-
-Maybe figure 1 is fish ordered (B) and unordered (A).
-
-Figure 2 can be the embryo dataset - unordered and unoriented (A) and hand-ordered (by membranes) and hand-oriented (by Dorsal) (B).
-
-This way Figure 2 presents the data and makes it a reference for all future alignments.
-
 \section{Techniques}
-
-% show schematic with images (starfish or zebrafish w/ stripes)
 
@@
 %
-$\Delta$ is the maximum number of pixels by which we are allowed to translate the image (we take $\Delta=20$, which corresponds to a 20\% shift in the image). 
-%
-The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
-%
-Therefore, instead of using an optimization procedure to compute the solution to \eqref{eq:opt_pairwise}, we choose to discretize the search space and exhaustively search to estimate the solution to \eqref{eq:opt_pairwise}.
-%
-We discritize the search space of rotations into 20 possible rotations ($0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10$), and 11 possible translations in both the $x$ and $y$ directions ($-20, -16, -12, \dots, 12, 16, 20$ pixels).
-%
-We then check all possible combinations for the best rotation and translations that align $I_i$ to $I_j$. 
-%
-Although this can be somewhat time intensive, it is not prohibitive for the size of data sets which we are considering, and can be trivially parallelized for larger data sets if necessary.
-%
-The solution to this search will not be the exact solution to \eqref{eq:opt_pairwise}, but it will (most likely) be a close approximation.
-%
-Since our techniques are robust to noise, close approximations will be sufficient to obtain accurate results.
+$\Delta$ is the maximum number of pixels by which we are allowed to translate the image.
 
@@
 We now compute the three--dimensional rotation, $R_{ij} \in SO(3)$, from the two--dimensional translation and rotation parameters $\theta_{ij}$, $dx_{ij}$, and $dy_{ij}$.
-%
-In the space of three--dimensional rotations, the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$ correspond to
-\begin{equation} \label{eq:angle_relations}
-\begin{aligned}
-	\alpha_{ij} &= \theta_{ij} \\
-	\beta_{ij} &= \frac{dx_{ij}}{n} \times \eta_{proj} \\
-	\gamma_{ij} &= \frac{dy_{ij}}{n} \times \eta_{proj} \\
-\end{aligned}
-\end{equation}
-where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project (we take $\eta_{proj} =  \pi/8$).
-%
-From here, we can write rotations around the three principal axes in terms of the three Euler angles
-\begin{equation}
-\begin{aligned}
-	R^x_{ij} &= \begin{bmatrix}
-	1 & 0 & 0 \\
-    0 & \cos(\alpha) & -\sin(\alpha) \\
-    0 & \sin(\alpha) & \cos(\alpha)
-	\end{bmatrix} \\
-	R^y_{ij} &= \begin{bmatrix}
-	\cos(\beta) & 0 & \sin(\beta) \\
-    0 & 1 & 0 \\
-    -\sin(\beta) & 0 & \cos(\beta)
-    \end{bmatrix} \\
-	R^z_{ij} &= \begin{bmatrix} 
-	\cos(\gamma) & -\sin(\gamma) & 0 \\
-    \sin(\gamma) & \cos(\gamma) & 0 \\
-    0 & 0 & 1 
-    \end{bmatrix}
-\end{aligned}
-\end{equation}
-where $R^x_{ij}$, $R^y_{ij}$, and $R^z_{ij}$ correspond to rotations around the $x$, $y$, and $z$ axes, respectively.
-%
-We then write the total rotation $R_{ij} \in SO(3)$ as 
-\begin{equation} \label{eq:total_rot}
-	R_{ij}	 = R^z_{ij} \times R^y_{ij} \times R^x_{ij}
-\end{equation}
-%
-\eqref{eq:total_rot} corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).
 
@@
 %
-The optimal rotations are given by $R_i = U_i V_i^T$, where $U_i$ and $V_i$ are the left and right singular vectors of $\hat{R}_i$, respectively \cite{...}. 
-%
-Because the eigenvectors of $H$ are determined up to a sign, we adjust the signs of the eigenvectors so that $det(R_i) = +1$ to ensure proper rotations.
-%
-
-From the matrices $R_1, \dots, R_m$, we now wish to find the corresponding translations and rotations of the images.
-%
-Because we are permitted to adjust our rotations by a global rotations, we first multiply all rotations by $R_1^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
-%
-We then compute the Euler angles $\alpha$, $\beta$, and $\gamma$ from the rotation matrix $R$ using the following relationships
-\begin{equation}
-\begin{aligned}
-R_{1,1} & = \cos(\beta)\cos(\gamma) \\
-R_{2,1} & = \cos(\beta)\sin(\gamma) \\
-R_{3,1} & = -\sin(\beta) \\
-R_{3,2} & = \sin(\alpha)\cos(\beta) \\
-R_{3,3} & = \cos(\alpha)\cos(\beta) 
-\end{aligned}
-\end{equation}
-%
-We can then invert \eqref{eq:angle_relations} to compute the optimal translation and rotation for each image from the Euler angles.
 
@@
 
-\section{Example: {\em Drosophila}}
+\section{Appendix}
 
-\subsection{dpERK}
+\subsection{Angular Synchronization Implementation}
+(we take $\Delta=20$, which corresponds to a 20\% shift in the image). 
+%
+The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
+%
+Therefore, instead of using an optimization procedure to compute the solution to \eqref{eq:opt_pairwise}, we choose to discretize the search space and exhaustively search to estimate the solution to \eqref{eq:opt_pairwise}.
+%
+We discritize the search space of rotations into 20 possible rotations ($0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10$), and 11 possible translations in both the $x$ and $y$ directions ($-20, -16, -12, \dots, 12, 16, 20$ pixels).
+%
+We then check all possible combinations for the best rotation and translations that align $I_i$ to $I_j$. 
+%
+Although this can be somewhat time intensive, it is not prohibitive for the size of data sets which we are considering, and can be trivially parallelized for larger data sets if necessary.
+%
+The solution to this search will not be the exact solution to \eqref{eq:opt_pairwise}, but it will (most likely) be a close approximation.
+%
+Since our techniques are robust to noise, close approximations will be sufficient to obtain accurate results.
 
-\begin{figure*}[h]
-\begin{subfigure}[c]{0.45\textwidth}
-\includegraphics[width=\textwidth]{dpERK_array}
-\caption{}
-\end{subfigure}
-\begin{subfigure}[c]{0.5\textwidth}
-\includegraphics[width=\textwidth]{dpERK_rank_corr}
-\caption{}
-\end{subfigure}
-\caption{(a) Ordering of images fluorescently stained for dpERK. (b) Correlation between ranks from ordering using VDM, and ordering using the membrane length.}
-\label{fig:dpERK}
-\end{figure*}
+%
+In the space of three--dimensional rotations, the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$ correspond to
+\begin{equation} \label{eq:angle_relations}
+\begin{aligned}
+	\alpha_{ij} &= \theta_{ij} \\
+	\beta_{ij} &= \frac{dx_{ij}}{n} \times \eta_{proj} \\
+	\gamma_{ij} &= \frac{dy_{ij}}{n} \times \eta_{proj} \\
+\end{aligned}
+\end{equation}
+where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project (we take $\eta_{proj} =  \pi/8$).
+%
+From here, we can write rotations around the three principal axes in terms of the three Euler angles
+\begin{equation}
+\begin{aligned}
+	R^x_{ij} &= \begin{bmatrix}
+	1 & 0 & 0 \\
+    0 & \cos(\alpha) & -\sin(\alpha) \\
+    0 & \sin(\alpha) & \cos(\alpha)
+	\end{bmatrix} \\
+	R^y_{ij} &= \begin{bmatrix}
+	\cos(\beta) & 0 & \sin(\beta) \\
+    0 & 1 & 0 \\
+    -\sin(\beta) & 0 & \cos(\beta)
+    \end{bmatrix} \\
+	R^z_{ij} &= \begin{bmatrix} 
+	\cos(\gamma) & -\sin(\gamma) & 0 \\
+    \sin(\gamma) & \cos(\gamma) & 0 \\
+    0 & 0 & 1 
+    \end{bmatrix}
+\end{aligned}
+\end{equation}
+where $R^x_{ij}$, $R^y_{ij}$, and $R^z_{ij}$ correspond to rotations around the $x$, $y$, and $z$ axes, respectively.
+%
+We then write the total rotation $R_{ij} \in SO(3)$ as 
+\begin{equation} \label{eq:total_rot}
+	R_{ij}	 = R^z_{ij} \times R^y_{ij} \times R^x_{ij}
+\end{equation}
+%
+\eqref{eq:total_rot} corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).
 
-\begin{figure*}[h]
-\centering
-\includegraphics[width=0.5\textwidth]{bootstrap_rank_corr}
-\caption{Rank correlation as a function of the number of data points.}
-\label{fig:bootstrap}
-\end{figure*}
+The optimal rotations are given by $R_i = U_i V_i^T$, where $U_i$ and $V_i$ are the left and right singular vectors of $\hat{R}_i$, respectively \cite{...}. 
+%
+Because the eigenvectors of $H$ are determined up to a sign, we adjust the signs of the eigenvectors so that $det(R_i) = +1$ to ensure proper rotations.
+%
 
-\subsection{Dorsal}
-
-\begin{figure*}[h]
-\begin{subfigure}[c]{0.45\textwidth}
-\includegraphics[width=\textwidth]{DI_array}
-\caption{}
-\end{subfigure}
-\begin{subfigure}[c]{0.5\textwidth}
-\includegraphics[width=\textwidth]{DI_rank_corr}
-\caption{}
-\end{subfigure}
-\caption{(a) Ordering of images fluorescently stained for Dorsal. (b) Correlation between ranks from ordering using VDM, and ordering using the membrane length.}
-\label{fig:Dl}
-\end{figure*}
-
-\section{Discussion}
-% compare to TSP algorithms
-Discussion of what we did and how it relates to other efforts (TSP, etc) and how it can be extended.
+From the matrices $R_1, \dots, R_m$, we now wish to find the corresponding translations and rotations of the images.
+%
+Because we are permitted to adjust our rotations by a global rotations, we first multiply all rotations by $R_1^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
+%
+We then compute the Euler angles $\alpha$, $\beta$, and $\gamma$ from the rotation matrix $R$ using the following relationships
+\begin{equation}
+\begin{aligned}
+R_{1,1} & = \cos(\beta)\cos(\gamma) \\
+R_{2,1} & = \cos(\beta)\sin(\gamma) \\
+R_{3,1} & = -\sin(\beta) \\
+R_{3,2} & = \sin(\alpha)\cos(\beta) \\
+R_{3,3} & = \cos(\alpha)\cos(\beta) 
+\end{aligned}
+\end{equation}
+%
+We can then invert \eqref{eq:angle_relations} to compute the optimal translation and rotation for each image from the Euler angles.
 

### 2014-02-23 image_analysis_paper/math_description.tex
added more desriptions to math

@@
 
-Often when we collect images in experiments, each image is taken in a different orientation. 
+Often when one collects images in experiments, each image is taken in a different orientation. 
 %
@@
 %
-Instead, we will use a technique called angular synchronization\cite{singer2011angular} to align our sets of images.
+Instead, we will use a technique called angular synchronization\cite{singer2011angular} to align sets of images.
 %
@@
 
-Let $I_1, \dots, I_m \in \mathbb{R}^{n \times n}$ denote the (grayscale) images that we wish to align with respect to translations and rotations.
+We will first discuss angular synchronization with {\em only} rotations. 
 %
+Let $I_1, \dots, I_m \in \mathbb{R}^{n \times n}$ denote the (grayscale) images that we wish to align with respect to rotations.
+%
+We assume that each image $I_i$ is a noisy rotated copy of the underlying image $I_{true}$ (which we are {\em not} given), such that 
+\begin{equation}
+I_i = R_i I_{true} + \xi_i,
+\end{equation}
+where $R_i \in \mathbb{R}^{2 \times 2}$ is an unknown rotation matrix and $\xi_i$ is random (often Gaussian) noise \footnote{The rotation matrices $R_i$ do not operate directly on the $n \times n$ image matrix $I_{true}$; however, we feel that this abuse of notation makes our description of the problem more intuitive.}.
+%
+Our goal is to obtain estimates of $R_1, \dots, R_m$ so that we can rotate the images $I_1, \dots, I_m$ to be aligned.
+%
+Up to noise, 
+\begin{equation} \label{eq:pairwise_rot}
+I_i \approx R_i R_j^T I_j;
+\end{equation}
+ note that \eqref{eq:pairwise_rot} does not require knowledge of the true underlying image $I_{true}$.
+%
+We can obtain {\em estimates} of $R_i R_j^T$ by computing the rotations to optimally align $I_j$ to $I_i$. 
+%
+Let $R_{ij}$ denote the rotation that optimally aligns $I_j$ to $I_i$.
+%
+We then construct the matrix $H \in \mathbb{R}^{2m \times 2m}$, where $H$ is an $m \times m$ matrix of $2 \times 2$ blocks with
+\begin{equation} \label{eq:H_to_R}
+H_{ij} = R_{ij}
+\end{equation}
+%
+We note that, under our assumptions, $H_{ij} \approx R_i R_j^T$, and so
+\begin{equation} \label{eq:H_low_rank}
+	H \approx 
+	\begin{bmatrix}
+	R_1 \\
+	R_2 \\
+	\vdots \\
+	R_m
+	\end{bmatrix}
+	\begin{bmatrix}
+	R_1^T R_2^T \dots R_m^T
+	\end{bmatrix}
+\end{equation}
+%
+We can (approximately) recover $R_1, R_2, \dots, R_m$ by computing the top block eigenvector of $H$ (i.e the eigenvectors corresponding to the two largest eigenvalues, stacked into a $2m \times 2$ matrix), and register the images by rotating image $i$ by $R_i^T$.
+%
+Furthermore, this eigendecomposition also considers {\em higher--order} consistency information. 
+%
+For example, given our pairwise estimates $R_{ij}$, we know that relationships of the form
+\begin{equation} \label{eq:triplet_consistency}
+R_{ik} R_{kj} \approx R_i R_k^T R_k R_j^T = R_i R_j^T
+\end{equation}
+should also hold.
+%
+However, 
+\begin{equation}
+(H^2)_{ij} = \sum_k R_{ik} R_{kj}
+\end{equation}
+and so {\em all} infomation of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$.
+%
+Because $H$ and $H^2$ have the same eigenvectors, our problem formulation accounts for not only pairwise alignment information, but also these higher--order considerations.
+
+We now note that, for \eqref{eq:H_low_rank} to be satisfied, we require that the group operation matrices be orthogonal, i.e. $R_i^{-1} = R_i^T$. 
+%
+The typical matrix representation for $ISO(2)$, the group of two--dimensional translations and rotations (using the matrix representation for the affine group \cite{...}), does not satisfy this property.
+%
+However, we can (approximately) represent rotations and translations in two--dimensions using three--dimensional rotation matrices.
+%
+We do this by projecting the image onto a portion of the surface of a (three--dimensional) sphere.
+%
+Rotations in two--dimensions now correspond to rotations around one principal axis in three--dimensions, and translations in two--dimensions correspond (approximately) to rotations around the other two principal axes in three--dimensions \footnote{Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$.
+However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.}.
+%
+These rotation matrices are (by definition) orthoganol, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
+
 The first step in angular synchronization is to compute the translations and rotations to optimally align each pair of data points. 
 %
-For each image pair $I_i$ and $I_j$, we would like to solve
+For each image pair $I_i$ and $I_j$, we solve
 \begin{equation}\label{eq:opt_pairwise}
@@
 \end{equation}
-where $\| \cdot \|_F$ denotes the Frobenius norm, and $f(I_j, \theta, dx, dy)$ is a function that first rotates the image $I_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction.
+where $\| \cdot \|_F$ denotes the Frobenius norm, $\Delta$ is the maximum number of pixels by which we are allowed to translate the image, and $f(I_j, \theta, dx, dy)$ is a function that first rotates the image $I_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction.
 %
-$\Delta$ is the maximum number of pixels by which we are allowed to translate the image.
-
-An important part of angular synchronization is that the symmetry group $G$ be {\em compact} and have a real (or perhaps complex) representation \cite{singer2011angular}.
+We then compute the three--dimensional rotation, $R_{ij} \in SO(3)$, from the two--dimensional translation and rotation parameters $\theta_{ij}$, $dx_{ij}$, and $dy_{ij}$ (see Appendix for more details).
 %
-Compactness guarantees (by the extreme value theorem) that the minimizer for each pairwise alignment is contained in the group,
-and having a real representation will allow us to write our problem in matrix form. 
+We then build the matrix $H \in \mathbb{R}^{3m \times 3m}$ as described in \eqref{eq:H_to_R}, and compute the top block eigenvector of $H$, $\hat{R} \in \mathbb{R}^{3m \times 3}.$ 
 %
-This is a rather technical point;
-we are only discussing this because the group of two--dimensional translations and rotations, $ISO(2)$, is {\em not} compact.
+The $3 \times 3$ blocks of $\hat{R}$ contain approximations of $R_1, \dots, R_m$.  
 %
-The angular synchronization problem is therefore not well--defined for this symmetry group.
+Because the blocks are not guaranteed to lie in $SO(3)$, we then project each block back to $SO(3)$ (see Appendix).
 %
-However, we can ``compactify'' this group by mapping $ISO(2)$ to the group $SO(3)$, the group of three--dimensional rotations.
-%
-We do this by projecting the image onto a portion of the surface of a (three--dimensional) sphere.
-%
-Rotations in two--dimensions now correspond to rotations around one principal axis in three--dimensions, and translations in two--dimensions correspond (approximately) to rotations around the other two principal axes in three--dimensions.
-%
-Furthermore, successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
-
-We now compute the three--dimensional rotation, $R_{ij} \in SO(3)$, from the two--dimensional translation and rotation parameters $\theta_{ij}$, $dx_{ij}$, and $dy_{ij}$.
-
-Given the pairwise rotations that optimally align the imges, $R_{ij}$, we would like to find the rotations $R_1, R_2, \dots, R_n \in SO(3)$ such that $R_i R_j^T \approx R_{ij}$, for every pair $i, j$.
-%TODO: Check that this order is correct
-%
-We would also like to exploit {\em higher-order} consistency relations;
-for example, the relationship $R_{ij} R_{jk} \approx R_{ik}$ should hold if the rotation estimates are accurate.
-%
-Therefore, we consider measurements which (almost) satisfy such conditions ``good'' measurements, and those measurements which do not satisfy such conditions as most likely inaccurate.
-
-We formulate our problem as an optimization problem:
-\begin{equation} \label{eqn:angsynch_obj}
-\max_{R_1, \dots, R_m \in SO(3)} \left\| \sum_{i=1}^{m} \sum_{j=1}^{m} R_i^T R_{ij} R_j \right\|^2.
-\end{equation}
-%
-%TODO: check if a norm is needed
-Each term in the objective function will be large large if $R_i$ and $R_j$ are consistent with $R_{ij}$, and will be random with mean 0 if $R_i$ and $R_j$ are inconsistent with $R_{ij}$.
-%
-Therefore, solving \eqref{eqn:angsynch_obj} gives us the set of global alignments which are most consistent with the computed pairwise alignments.
-
-In general, the solution to \eqref{eqn:angsynch_obj} is not easily computed.
-%
-Instead, we relax the problem and allow $R_1, \dots, R_m \in \mathbb{R}^{3 \times 3}$.
-%
-We define the matrix $H \in \mathbb{R}^{3m \times 3m}$, which is an $m \times m$ block matrix with $3 \times 3$ blocks, with $H_{ij} = R_{ij}$.
-%
-We can then write the relaxed problem as 
-\begin{equation} \label{eqn:angsynch_relax}
-\max_{R\in \mathbb{R}^{3m \times 3}} \| R^T H R \|^2.
-\end{equation}
-%
-The solution to \eqref{eqn:angsynch_relax} is given by the top ``block'' eigenvector of $H$ (i.e.
-the three eigenvectors corresponding to the three largest eigenvalues, concatenated into a $3m \times 3$ matrix), which we denote $\hat{R} \in \mathbb{R}^{3m \times 3}$.
-%
-The $i^{th}$ rotation is (approximately) the $i^{th}$ block of $\hat{R}$, $\hat{R}(i) \in \mathbb{R}^{3 \times 3}$.
-%
-We would like to note that this formulation also accounts for higher-order relations.
-%
-For example, if we want to optimize over all pairs of pairwise rotations, we would solve 
-\begin{equation}
-\max_{R_1, \dots, R_m \in SO(3)} \sum_{i=1}^{m} \sum_{j=1}^{m} \left\| R_i^T \left( \sum_k R_{ik} R_{kj} \right) R_j \right\|,
-\end{equation} 
-%
-which relaxes to the problem $ \max_{R \in \mathbb{R}^{3m \times 3}} \| R^T H^2 R \|^2$.
-%
-However, the solution to this problem is also given by the top block eigenvector of $H$.
-
-Because we relaxed the problem to allow our solutions to lie in $\mathbb{R}^{3 \times 3}$, we must project our approximate solution back to $SO(3)$.
-%
+To register the images, we compute the translations and rotation corresponding to $R_i^T$.
 
@@
 
-Given {\em aligned} images $I_1, \dots, I_m \in \mathbb{R}^{n \times n}$, we fist calculate the matrix $W \in \mathbb{R}^{m \times m}$, where 
+In our framework, we assume that our data lie on a one--dimensional curve or manifold, and that this one dimension is correlated with time. 
+%
+We will uncover a parameterization of this curve by only considering {\em local} neighbors of each data point.
+%
+To illustrate how local neighbors can allow us to uncover a global parameterization, assume we have data $x_1, x_2, \dots, x_m \in \mathbb{R}$ such that $x_1 < x_2 < \dots < x_m$.
+%
+We consider $x_i$ ``close to'' $x_{i+1}$ and $x_{i-1}$, and $x_i$ ``far away from'' all other data points. 
+%
+We then construct the weight matrix, $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if $x_i$ is close to $x_j$,
+\begin{equation}
+W = 
+\begin{bmatrix}
+	1 & 1/2 & 0 & 0 & \dots & 0 \\
+	1/2 & 1 & 1/2 & 0 & \dots & 0 \\
+	\ddots & \ddots & \ddots & \ddots & \ddots & \ddots \\
+	0 & 0 & 0 & \dots & 1 & 1/2 \\
+	0 & 0 & 0 & \dots & 1/2 & 1 
+\end{bmatrix}.
+\end{equation}
+%
+If we normalize this matrix such that $\sum_j W_{ij} = 1$, then one can show that the first eigenvector is the all--ones vector, $\phi_1 = [1 1 \dots 1]^T$, and the second eigenvector, $\phi_2 = [cos(\pi/m) \cos(2 \pi/ m) \dots \cos((m-1) \pi / m) \cos(\pi)]^T$.
+%
+Therefore, the second eigenvector, $\phi_2$ is {\em one--to--one} with the parameterization of the data $x_1, x_2, \dots, x_m$. 
+%
+Therefore, sorting our data by the value of the entries in $\phi_2$ will allow us to order our data.
+
+For the above analysis, the key step was having a notion of ``closeness'' between data points so that we can construct the matrix $W$.
+%
+For general data in high--dimensional space, we typically use a Gaussian kernel,
 \begin{equation} \label{eq:dmaps_W}
-W_{ij} = \exp \left( -\frac{d^2(I_i, I_j)}{\epsilon^2} \right)
+W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right)
 \end{equation}
@@
 %
+Therefore, points closer than $\epsilon$ apart are considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
+%
 $\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph}); in practice, we choose $\epsilon$ to be the median of the pairwise distances between data points.
+
+We then compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{m} W_{ij}$, and the matrix $A$, where
+\begin{equation} \label{eq:dmaps_A}
+A = D^{-1} W.
+\end{equation} 
 %
-We then compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{m} W_{ij}$, and the matrix $A = D^{-1} W$. 
-%
-We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_m$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_m$ and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$.
-%
-Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors. 
-%
-The eigenvectors $\phi_1, \phi_2, \dots, \phi_m$ give the embedding coordinates, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point. 
+We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_m$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_m$ and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$ \footnote{Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors.}.
 %
@@
 %
-The next few eigenvectors give the ``meaningful'' embedding coordinates for the data.
-
+In general, the next few eigenvectors $\phi_2, \dots, \phi_m$ give ``meaningful'' embedding coordinates for the data, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point.
+%
 In our setup, we assume that our data are inherently one--dimensional, and that this one dimension is correlated with time.
@@
 %
-where $A$ is constructed from the matrix $W$ \eqref{eq:dmaps_W}, and the distance metric $d(I_i, I_j)$ is the Frobenius norm between the images {\em after} optimal pairwise alignment ($d(I_i, I_j) = \| f(I_j, \theta_{ij}, dx_{ij}, dy_{ij}) - I_i \|_F$).
+where $A$ is defined in \eqref{eq:dmaps_A}, the distance metric $d(I_i, I_j)$ is the Frobenius norm between the images {\em after} optimal pairwise alignment ($d(I_i, I_j) = \| f(I_j, \theta_{ij}, dx_{ij}, dy_{ij}) - I_i \|_F$), and $H$ is defined in \eqref{eq:H_to_R}.
 
@@
 %
-The embedding coordinates are given by 
+In general, the embedding coordinates are given by 
 \begin{equation}
@@
 %
-If we assume that our data are one--dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, then we can order our data by $\psi_{1,4}$.
+If we assume that our data are one--dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, one can show that the ``correct'' embedding coordinate (i.e.
+the analog of $\phi_2$ from the diffusion maps case) is $\psi_{1,4}$.
 %
@@
 \subsection{Angular Synchronization Implementation}
-(we take $\Delta=20$, which corresponds to a 20\% shift in the image). 
+
+\subsubsection{Computing Optimal Pairwise Alignments}
+
+In \eqref{eq:opt_pairwise}, we take $\Delta=20$, which corresponds to a 20\% shift in the image. 
 %
@@
 %
-We discritize the search space of rotations into 20 possible rotations ($0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10$), and 11 possible translations in both the $x$ and $y$ directions ($-20, -16, -12, \dots, 12, 16, 20$ pixels).
+We discritize the search space of rotations into 20 possible rotations ($d\theta  \in \{0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10 \}$), and 11 possible translations in both the $x$ and $y$ directions ($dx, dy \in \{-20, -16, -12, \dots, 12, 16, 20 \}$).
 %
-We then check all possible combinations for the best rotation and translations that align $I_i$ to $I_j$. 
+We then check all possible combinations for the best rotation and translations that align $I_j$ to $I_i$. 
 %
@@
 
-%
+\subsubsection{Converting from $ISO(2)$ to $SO(3)$}
+
 In the space of three--dimensional rotations, the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$ correspond to
@@
 \end{equation}
-where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project (we take $\eta_{proj} =  \pi/8$).
+where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project the image.
+%
+We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere in $\mathbb{R}^3$.
 %
@@
 
-The optimal rotations are given by $R_i = U_i V_i^T$, where $U_i$ and $V_i$ are the left and right singular vectors of $\hat{R}_i$, respectively \cite{...}. 
+\subsubsection{Computing Optimal Translations and Rotations from the Eigenvectors of $H$}
+
+From the top block eigenvector $\hat{R} \in \mathbb{R}^{3m \times 3}$, we would like to compute the optimal translation and rotation for each image. 
+%
+The optimal three--dimensional rotations are given by $R_i = U_i V_i^T$, where $U_i$ and $V_i$ are the left and right singular vectors of $\hat{R}_i$, respectively \cite{...}. 
 %
 Because the eigenvectors of $H$ are determined up to a sign, we adjust the signs of the eigenvectors so that $det(R_i) = +1$ to ensure proper rotations.
-%
 

### 2014-02-27 image_analysis_paper/background_reading/article_notes.tex
Updated paper again.

@@
 
-\section{References from Stas}
+\section{Biology Applications}
 
@@
 
-\item \fullcite{wei2006conditional}
+%\item \fullcite{golani1999phenotyping}
+
+\item \fullcite{anavy2014blind}
 \begin{itemize}
-		\item Want to develop a more informative growth chart metric
-		\item Instead of looking at how atypical a current measurement is for an individual, they look at how atypical a time series is for an individual 
-		\item This means that a person will be flagged if, for example, their weight changes dramatically in a short period of time, even if both measurements are within the ``normal'' range
-		\item Fit model to data for ``normal'' time--series behavior
+	\item Want to temporally order RNA-Seq data
+	\item First do PCA on RNA-Seq data to denoise/reduce dimension
+	\item Then solve a TSP in this reduced space; assume that the path found by the TSP is the ``correct'' developmental path
+	\item Assume later stages in development are characterized by more variability/higher entropy to determine where the path starts (in time) and where it ends
 \end{itemize}
 
-\item \fullcite{wei2006quantile}
+\item \fullcite{basu2014detecting}
 \begin{itemize}
-	\item Develop nonparameteric models (using spline regression) for growth charts
-	\item Typical growth charts rely on normality assumptions (data is often transformed to fulfill such assumptions)
-	\item Using nonparametric models can allow them to fit a much broader range of behaviors (i.e. bimodal distributions)
+	\item Want to detect changes in cell morphology
+	\item First use image segmentation to extract individual cells from image
+	\item Center images, and align main axis of cell to be vertical
+	\item Consider pictures as mass distributions (more color = more mass); describe distance between cells to be the amount of effort one must exert (mass * distance) to make one picture look like the other
+	\item Then use PCA to visualize this data
+\end{itemize}
+
+\item \fullcite{escudero2011epithelial}
+\begin{itemize}
+	\item Want to be able to characterize differences in cellular organization between organisms
+	\item Describe cells using features of cell morphology (i.e. average size of cells) as well as features of cell organization (i.e. how many neighbors does each cell have?)
+	\item Think of cell organization as a network (each cell is a node, connected by links to its neighbors), and then use network descriptors (degree, average degree of neighbors, clustering coefficient) to describe cell organization
+	\item Then use PCA and discriminant analysis (like regression, but for categorize variables) to show they can separate mutants and WT cells using these network + morphology features
+\end{itemize}
+
+\item \fullcite{qiu2011discovering}
+\begin{itemize}
+	\item Want to order microarray data
+	\item Look at specific (high--variance) genes (~3000 genes)
+	\item Cluster genes into ``modules'' using k--means, so that similairly expressed genes are grouped together
+	\item Within each module, construct a minimium spanning tree of the data (distance between nodes is Euclidean distance)
+	\item Find set of MSTs that are most similar/consistent
+	\item Assume these are the ``right'' MST and that they uncover the progression of the data
+\end{itemize}
+
+\item \fullcite{gupta2008extracting}
+\begin{itemize}
+	\item Want to temporally order gene expression data from cancer patients
+	\item Use TSP to order gene expression data (can show that TSP path will recover the correct ordering under some probability model)
+	\item Use a probabilistic gene/patient model to model the variation from patient to patient (model underlying dynamics with splines, and then assume rest in noise from patient-to-patient variability)
+\end{itemize}
+
+\item \fullcite{kafri2013dynamics}
+\begin{itemize}
+	\item Take snapshot of fixed cell population
+	\item Fit 1D curve to trajectory (they assert that the two ``important'' parameters are levels of DNA and the level of ``anaphase-promoting complex'', so they project onto these two components and then fit a 1D closed curve to the data)
+	\item Since the cells follow a closed cycle, given the closed curve, they can use mass balances to determine the relative rates/fluxes through the different stages
+\end{itemize}
+
+\item \fullcite{wyart2010evaluating}
+
+\begin{itemize}
+	\item Use RNA FISH technique to count levels of individual mRNAs in cells; requires cells to be fixed
+	\item Propose models for dynamics of mRNA throughout cell cycle
+	\item Show how they can use data from population of fixed cells to fit model (periodic model for level of mRNA production)
+\end{itemize}
+
+\item \fullcite{sisan2010event}
+\begin{itemize}
+	\item Looks at time correlations in data to determine temporal ordering
+	\item Goal is to determine in what order cellular events happen (i.e. does rxn A lead to rxn B, or does B lead to A)
+	\item They look at the time-correlation between events.
+	\item Asymmetry in time-correlation tells you in what order events happen (depending on model)
+	\item Looking at time correlation over many samples integrates/averages out stocahsticity in system
+\end{itemize}
+
+\item \fullcite{peng2013detecting}
+\begin{itemize}
+	\item Wants to find which genes/SNPs are responsible for variation in facial features
+	\item Has DNA data and 3D face scans for many samples
+	\item Looks at different facial models (all are based on landmarks)
+	\item Tries to find which variations in the facial models correlate with which SNPs
 \end{itemize}
@@
 
+\end{enumerate}
+
+\section{Medical Statistics Applications}
+
+\begin{enumerate}
+
+\item \fullcite{li2012modelling}
+\begin{itemize}
+	\item Want to use cross--sectional data to model disease progression
+	\item Data are things that are known to be important/telling for that particlular disease (i.e.
+breast cancer data contains descriptors of cell morphology, Parkinson's disease data contains descriptors of speech patterns)
+	\item Create ``pseudo time series'' by subsampling data, and from each subsample, finding shortest path that goes from a healthy sample to a diseased sample
+	\item From these pseudo time series, can fit a hidden Markov model to find different states of disease progression and likely transitions from one state to another
+\end{itemize}
+
 \item \fullcite{tucker2010pseudotemporal}
@@
 
-\item \fullcite{kafri2013dynamics}
+\item \fullcite{das2013dynamic}
 \begin{itemize}
-	\item Take snapshot of fixed cell population
-	\item Fit 1D curve to trajectory (they assert that the two ``important'' parameters are levels of DNA and the level of ``anaphase-promoting complex'', so they project onto these two components and then fit a 1D closed curve to the data)
-	\item Since the cells follow a closed cycle, given the closed curve, they can use mass balances to determine the relative rates/fluxes through the different stages
+	\item Use spline regression to model disease progression from irregular longitudinal data
 \end{itemize}
 
-\item \fullcite{wyart2010evaluating}
-
+\item \fullcite{xu2012comprehensive}
 \begin{itemize}
-	\item Use RNA FISH technique to count levels of individual mRNAs in cells; requires cells to be fixed
-	\item Propose models for dynamics of mRNA throughout cell cycle
-	\item Show how they can use data from population of fixed cells to fit model (periodic model for level of mRNA production)
+	\item Want to quantify shape of liver lesions
+	\item Take set of 14 common shape descriptors
+	\item Use regression to find optimal combination of shape descriptors that can predict the similairity using a ``similairity reference standard''
 \end{itemize}
 
-\item \fullcite{chen2012nonlinear}
+\item \fullcite{wei2006conditional}
+\begin{itemize}
+		\item Want to develop a more informative growth chart metric
+		\item Instead of looking at how atypical a current measurement is for an individual, they look at how atypical a time series is for an individual 
+		\item This means that a person will be flagged if, for example, their weight changes dramatically in a short period of time, even if both measurements are within the ``normal'' range
+		\item Fit model to data for ``normal'' time--series behavior
+\end{itemize}
 
-\item \fullcite{das2013dynamic}
+\item \fullcite{wei2006quantile}
+\begin{itemize}
+	\item Develop nonparameteric models (using spline regression) for growth charts
+	\item Typical growth charts rely on normality assumptions (data is often transformed to fulfill such assumptions)
+	\item Using nonparametric models can allow them to fit a much broader range of behaviors (i.e. bimodal distributions)
+\end{itemize}
+
+\item \fullcite{toga2003temporal}
+\begin{itemize}
+	\item Want to model spatiotemporal changes in the 3D structure of the brain (so that one can detect disease, etc.)
+	\item Model growth of different portions of the brain using deformation mapping (look at deformation of different regions of a {\em single} person's brain from one scan to the next)
+	\item Can fit cortical models for each person in a population; can then look at variations across these models
+\end{itemize}
+
+\item \fullcite{denkinger2013set}
+\begin{itemize}
+	\item Want to develop a standardized data set that can be used to investigate temporal and causal neural processing
+	\item Developed a set of 222 images which contain different causal and temporal relationships
+\end{itemize}
+
+\end{enumerate}
+
+\section{Image Applications}
+
+\begin{enumerate}
+
+\item \fullcite{lau2009modeling}
+\begin{itemize}
+	\item Want to model trajectory/motion data where each sample is a trajectory
+	\item Assume that the trajectories come from a second-order Markov model (i.e. $t+2$ depends on $t$ and $t+1$)
+	\item Fit a Bayesian network that models the trajectory evolution using training data
+	\item Can then produce new sample trajectories from this Bayesian network
+\end{itemize}
 
@@
 \end{itemize}
+
 \item \fullcite{bigot2012geometric}
+\begin{itemize}
+	\item Want to do PCA on 2D images (e.g. handwritten digits)
+	\item Typically, Euclidean distance is not a good metric for images
+	\item They define a {\em deformation operator}, and compute the deformation operator needed to take each image to a specified template
+	\item They then do PCA on these deformation operators to see what are ``typical'' deformations in handwritten digits
+\end{itemize}
 
-\item \fullcite{escudero2011epithelial}
+\item \fullcite{golparvar2011monitoring}
+\begin{itemize}
+	\item Want to construct video of construction progress from snapshots
+	\item Building Information Models (BIM) 
+\end{itemize}
+
+\end{enumerate}
+
+\section{Math/Algorithms (Temporal)}
+
+\begin{enumerate}
 
 \item \fullcite{machta2013parameter}
-
-\item \fullcite{peng2013detecting}
 \begin{itemize}
-	\item Wants to find which genes/SNPs are responsible for variation in facial features
-	\item Has DNA data and 3D face scans for many samples
-	\item Looks at different facial models (all are based on landmarks)
-	\item Tries to find which variations in the facial models correlate with which SNPs
+	\item Look at Fisher Information Matrix (measures sensitvity of model to changes in parameters)
+	\item Eigenvalues of FIM tell you how many ``important'' or stiff parameters there are in the system
+	\item Demonstrate how we can use this to see how many parameters are necessary to describe diffusion and an Ising model
 \end{itemize}
@@
 
-\item \fullcite{qiu2011discovering}
+\item \fullcite{chen2012nonlinear}
 \begin{itemize}
-	\item Want to order microarray data
-	\item Look at specific (high--variance) genes (~3000 genes)
-	\item Cluster genes into ``modules'' using k--means, so that similairly expressed genes are grouped together
-	\item Within each module, construct a minimium spanning tree of the data (distance between nodes is Euclidean distance)
-	\item Find set of MSTs that are most similar/consistent
-	\item Assume these are the ``right'' MST and that they uncover the progression of the data
+	\item Want to look at reduced representations for data, where each data point is a function (e.g. a growth trajectory or curve)
 \end{itemize}
-
-\item \fullcite{gupta2008extracting}
-\begin{itemize}
-	\item Want to temporally order gene expression data from cancer patients
-	\item Use TSP to order gene expression data (can show that TSP path will recover the correct ordering under some probability model)
-	\item Use a probabilistic gene/patient model to model the variation from patient to patient (model underlying dynamics with splines, and then assume rest in noise from patient-to-patient variability)
-\end{itemize}
-
-\item \fullcite{lau2009modeling}
-\begin{itemize}
-	\item Want to model trajectory/motion data where each sample is a trajectory
-	\item Assume that the trajectories come from a second-order Markov model (i.e. $t+2$ depends on $t$ and $t+1$)
-	\item Fit a Bayesian network that models the trajectory evolution using training data
-	\item Can then produce new sample trajectories from this Bayesian network
-\end{itemize}
-
-\item \fullcite{toga2003temporal}
-\begin{itemize}
-	\item Want to model spatiotemporal changes in the 3D structure of the brain (so that one can detect disease, etc.)
-	\item Model growth of different portions of the brain using deformation mapping (look at deformation of different regions of a {\em single} person's brain from one scan to the next)
-	\item Can fit cortical models for each person in a population; can then look at variations across these models
-\end{itemize}
-
-\item \fullcite{li2012modelling}
-\begin{itemize}
-	\item Want to use cross--sectional data to model disease progression
-	\item Data are things that are known to be important/telling for that particlular disease (i.e.
-breast cancer data contains descriptors of cell morphology, Parkinson's disease data contains descriptors of speech patterns)
-	\item Create ``pseudo time series'' by subsampling data, and from each subsample, finding shortest path that goes from a healthy sample to a diseased sample
-	\item From these pseudo time series, can fit a hidden Markov model to find different states of disease progression and likely transitions from one state to another
-\end{itemize}
-
-\item \fullcite{denkinger2013set}
-\begin{itemize}
-	\item Want to develop a standardized data set that can be used to investigate temporal and causal neural processing
-	\item Developed a set of 222 images which contain different causal and temporal relationships
-\end{itemize}
-
-\item \fullcite{xu2012comprehensive}
-\begin{itemize}
-	\item Want to quantify shape of liver lesions
-	\item Take set of 14 common shape descriptors
-	\item Use regression to find optimal combination of shape descriptors that can predict the similairity using a ``similairity reference standard''
-	\item FINISH
-\end{itemize}
-\item \fullcite{sisan2010event}
-\begin{itemize}
-	\item Looks at time correlations in data to determine temporal ordering
-	\item Goal is to determine in what order cellular events happen (i.e. does rxn A lead to rxn B, or does B lead to A)
-	\item They look at the time-correlation between events.
-	\item Asymmetry in time-correlation tells you in what order events happen (depending on model)
-	\item Looking at time correlation over many samples integrates/averages out stocahsticity in system
-\end{itemize}
-
-\item \fullcite{golparvar2011monitoring}
-
-\item \fullcite{golani1999phenotyping}
-\begin{itemize}
-	\item Want to construct video of construction progress from snapshots
-	\item Building Information Models (BIM) 
-\end{itemize}
-
-\item \fullcite{anavy2014blind}
-
-\item \fullcite{basu2014detecting}
 
@@
 
-\section{Algorithms}
+\section{Math/Algorithms (Symmetries)}
 
@@
 
-\section{Applications}
+\section{Applications (Symmetries)}
 

### 2014-02-27 image_analysis_paper/math_description.tex
Updated paper again.

@@
 
+\usepackage{graphicx}
+\usepackage{subcaption}
+\graphicspath{ {paper_figures/} }
+
 \begin{document}
+
+\begin{figure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{fig2a}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{fig2b}
+\caption{}
+\end{subfigure}
+\caption{}
+\end{figure}
 
@@
 
-\subsection{Alignment Using Angular Synchronization}
+\subsection{Registration Using Angular Synchronization} \label{subsec:angsynch}
 
@@
 %
-Therefore, the first task in image analysis is to align the set of images. 
+Therefore, the first task in image analysis is to align or {\em register} the set of images. 
 %
-We will consider settings when alignment is done with respect to two--dimensional translations and rotations of the images, as each ``object of interest'' (i.e.
-each embryo) is in a different position and rotated at a different angle within the image frame.
-
 There are many ways to align images. 
@@
 %
-Let $I_1, \dots, I_m \in \mathbb{R}^{n \times n}$ denote the (grayscale) images that we wish to align with respect to rotations.
+Let $x_1, \dots, x_m \in \mathbb{R}^x$ denote the (grayscale) signal that we wish to align with respect to rotations.
 %
-We assume that each image $I_i$ is a noisy rotated copy of the underlying image $I_{true}$ (which we are {\em not} given), such that 
+We assume that each signal $x_i$ is a noisy rotated copy of the underlying signal $x_{true}$ (which we are {\em not} given), such that 
 \begin{equation}
-I_i = R_i I_{true} + \xi_i,
+x_i = f(x_{true}, \theta_i) + \xi_i
 \end{equation}
-where $R_i \in \mathbb{R}^{2 \times 2}$ is an unknown rotation matrix and $\xi_i$ is random (often Gaussian) noise \footnote{The rotation matrices $R_i$ do not operate directly on the $n \times n$ image matrix $I_{true}$; however, we feel that this abuse of notation makes our description of the problem more intuitive.}.
+where $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ degrees \footnote{add about how we rotate, because it's soooooo complicated}. 
 %
-Our goal is to obtain estimates of $R_1, \dots, R_m$ so that we can rotate the images $I_1, \dots, I_m$ to be aligned.
+Our goal is to obtain estimates of $\theta_1, \dots, \theta_m$ so that we can align $x_1, \dots, x_m$. 
 %
@@
 \begin{equation} \label{eq:pairwise_rot}
-I_i \approx R_i R_j^T I_j;
+x_i \approx f(x_j, \theta_i - \theta_j) ;
 \end{equation}
- note that \eqref{eq:pairwise_rot} does not require knowledge of the true underlying image $I_{true}$.
+ note that \eqref{eq:pairwise_rot} does not require knowledge of $x_{true}$.
 %
-We can obtain {\em estimates} of $R_i R_j^T$ by computing the rotations to optimally align $I_j$ to $I_i$. 
+We can obtain {\em estimates} of $\theta_i - \theta_j$ by computing the rotations to optimally align $x_j$ to $x_i$. 
 %
-Let $R_{ij}$ denote the rotation that optimally aligns $I_j$ to $I_i$.
+Let $\theta_{ij}$ denote the rotation that optimally aligns $x_j$ to $x_i$, i.e.,
 %
-We then construct the matrix $H \in \mathbb{R}^{2m \times 2m}$, where $H$ is an $m \times m$ matrix of $2 \times 2$ blocks with
-\begin{equation} \label{eq:H_to_R}
-H_{ij} = R_{ij}
+\begin{equation} \label{eq:opt_angle}
+\theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2
+\end{equation}
+
+Rather than work with the angles directly, it will be more convenient to work with the rotation matrices 
+\begin{equation} \label{eq:R_theta}
+R(\theta_{ij}) = \begin{bmatrix}
+\cos(\theta_{ij}) & \sin(\theta_{ij}) \\
+-\sin(\theta_{ij}) & \cos(\theta_{ij})
+\end{bmatrix}
 \end{equation}
 %
-We note that, under our assumptions, $H_{ij} \approx R_i R_j^T$, and so
+Successive rotations correspond to multiplication of the rotation matrices, i.e., 
+$R(\alpha_1 + \alpha_2) = R(\alpha_1) R(\alpha_2)$,
+and, due to the orthogonality of rotation matrices, $R(-\alpha) = R(\alpha)^T$.
+
+Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $d=2$).
+%
+ We then construct the matrix $H \in \mathbb{R}^{md \times md}$, where $H$ is an $m \times m$ matrix of $d \times d$ blocks with
+\begin{equation} \label{eq:H_to_R}
+H_{ij} = R(\theta_{ij})
+\end{equation}
+%
+
+We note that, under our assumptions, 
+\begin{equation} 
+H_{ij} = R(\theta_{ij}) \approx R(\theta_i - \theta_j) = R(\theta_i) R(-\theta_j) = R(\theta_i) R(\theta_j)^T,
+\end{equation}
+ and so
 \begin{equation} \label{eq:H_low_rank}
@@
 	\begin{bmatrix}
-	R_1 \\
-	R_2 \\
+	R(\theta_1) \\
+	R(\theta_2) \\
 	\vdots \\
-	R_m
+	R(\theta_m)
 	\end{bmatrix}
 	\begin{bmatrix}
-	R_1^T R_2^T \dots R_m^T
+	R(\theta_1)^T R(\theta_2)^T \dots R(\theta_m)^T
 	\end{bmatrix}
@@
 %
-We can (approximately) recover $R_1, R_2, \dots, R_m$ by computing the top block eigenvector of $H$ (i.e the eigenvectors corresponding to the two largest eigenvalues, stacked into a $2m \times 2$ matrix), and register the images by rotating image $i$ by $R_i^T$.
+We can (approximately) recover $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$ by computing the top block eigenvector of $H$ (i.e the eigenvectors corresponding to the $d$ largest eigenvalues, stacked into a $md \times 2$ matrix).
 %
+We can estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and then register the signals by rotating image $i$ by $-\theta_i$.
+
 Furthermore, this eigendecomposition also considers {\em higher--order} consistency information. 
@@
 \begin{equation} \label{eq:triplet_consistency}
-R_{ik} R_{kj} \approx R_i R_k^T R_k R_j^T = R_i R_j^T
+R(\theta_{ik}) R(\theta_{kj}) \approx R(\theta_i) R(\theta_k)^T R(\theta_k) R(\theta_j)^T = R(\theta_i) R(\theta_j)^T
 \end{equation}
@@
 \begin{equation}
-(H^2)_{ij} = \sum_k R_{ik} R_{kj}
+(H^2)_{ij} = \sum_k R(\theta_{ik}) R(\theta_{kj})
 \end{equation}
@@
 
-We now note that, for \eqref{eq:H_low_rank} to be satisfied, we require that the group operation matrices be orthogonal, i.e. $R_i^{-1} = R_i^T$. 
+To illustrate how angular synchronization works, we have converted the two--dimensional images into concentration profiles on a ring (we will address the full two--dimensional images in a subsequent section).
 %
-The typical matrix representation for $ISO(2)$, the group of two--dimensional translations and rotations (using the matrix representation for the affine group \cite{...}), does not satisfy this property.
+Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one--dimensional concentration profile. 
 %
-However, we can (approximately) represent rotations and translations in two--dimensions using three--dimensional rotation matrices.
+The images are not aligned, and so rotation of this image corresponds to shifting (with periodic boundary conditions) the one--dimensional concentration profile. 
 %
-We do this by projecting the image onto a portion of the surface of a (three--dimensional) sphere.
+Figure \ref{subfig:1d_unaligned_unordered} shows the concentration profiles for all the images in our data set. 
 %
-Rotations in two--dimensions now correspond to rotations around one principal axis in three--dimensions, and translations in two--dimensions correspond (approximately) to rotations around the other two principal axes in three--dimensions \footnote{Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$.
-However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.}.
+The concentration profiles have been stacked in an array, such that each row corresponds to one image.
 %
-These rotation matrices are (by definition) orthoganol, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
+Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles, now aligned using angular synchronization. 
+%
+The peaks are now (mostly) aligned.
 
-The first step in angular synchronization is to compute the translations and rotations to optimally align each pair of data points. 
-%
-For each image pair $I_i$ and $I_j$, we solve
-\begin{equation}\label{eq:opt_pairwise}
-(\theta_{ij}, dx_{ij}, dy_{ij}) = \argmin_{
-\begin{matrix}
-\theta \in [0, 2\pi) \\ 
-dx \in [-\Delta, \Delta]\\ 
-dy \in [-\Delta, \Delta]
-\end{matrix}
-} \|f(I_j, \theta, dx, dy) - I_i \|_F^2.
-\end{equation}
-where $\| \cdot \|_F$ denotes the Frobenius norm, $\Delta$ is the maximum number of pixels by which we are allowed to translate the image, and $f(I_j, \theta, dx, dy)$ is a function that first rotates the image $I_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction.
-%
-We then compute the three--dimensional rotation, $R_{ij} \in SO(3)$, from the two--dimensional translation and rotation parameters $\theta_{ij}$, $dx_{ij}$, and $dy_{ij}$ (see Appendix for more details).
-%
-We then build the matrix $H \in \mathbb{R}^{3m \times 3m}$ as described in \eqref{eq:H_to_R}, and compute the top block eigenvector of $H$, $\hat{R} \in \mathbb{R}^{3m \times 3}.$ 
-%
-The $3 \times 3$ blocks of $\hat{R}$ contain approximations of $R_1, \dots, R_m$.  
-%
-Because the blocks are not guaranteed to lie in $SO(3)$, we then project each block back to $SO(3)$ (see Appendix).
-%
-To register the images, we compute the translations and rotation corresponding to $R_i^T$.
+\begin{figure}
+\begin{subfigure}{0.17\textwidth}
+\includegraphics[width=\textwidth]{illustrate_1d}
+\caption{}
+\label{subfig:1d_image_example}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{unregistered_unordered_1d}
+\caption{}
+\label{subfig:1d_unaligned_unordered}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{registered_unordered_1d}
+\caption{}
+\label{subfig:1d_aligned_unordered}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{registered_ordered_1d}
+\caption{}
+\label{subfig:1d_aligned_ordered}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{registered_ordered_vdm_1d}
+\caption{}
+\label{subfig:1d_aligned_ordered_vdm}
+\end{subfigure}
+\caption{}
+\label{fig:1d_demo}
+\end{figure}
 
@@
 
-Given our set of aligned images, we now wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
+Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we now wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
 %
@@
 %
-The {\em essential} requirement for DMAPS is an appropriate distance metric $d(I_i, I_j)$ for comparing data points. 
+The {\em essential} requirement for DMAPS is an appropriate distance metric $d(x_i, x_j)$ for comparing data points. 
 %
@@
 
+We demonstrate how we can do this with our one--dimensional concentration profiles. 
+%
+We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our distances in \eqref{eq:dmaps_W}.
+%
+Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non--trivial) dmaps component, $\phi_2$.
+
 \subsection{Vector Diffusion Maps}
@@
 %
-Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{3m \times 3m}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
+Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
 %
@@
 %
-This technique is called vector diffusion maps \cite{singer2012vector}.
+This technique is called vector diffusion maps (VDM) \cite{singer2012vector}.
 %
@@
 
-For vector diffusion maps, we construct the matrix $S \in \mathbb{R}^{3m \times 3m}$, with
+For vector diffusion maps, we construct the matrix $S \in \mathbb{R}^{md \times md}$, with
 \begin{equation}
@@
 %
-where $A$ is defined in \eqref{eq:dmaps_A}, the distance metric $d(I_i, I_j)$ is the Frobenius norm between the images {\em after} optimal pairwise alignment ($d(I_i, I_j) = \| f(I_j, \theta_{ij}, dx_{ij}, dy_{ij}) - I_i \|_F$), and $H$ is defined in \eqref{eq:H_to_R}.
+where $A$ is defined in \eqref{eq:dmaps_A} and $H$ is defined in \eqref{eq:H_to_R}.
+%
+The distance $d(x_i, x_j)$ is the distance between samples {\em after} alignment (i.e. it is the minimium obtained in \eqref{eq:opt_pairwise}).
 
-We then compute the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_{3m}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{3m}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{3m}|$.
+We then compute the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_{md}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{md}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$.
 %
@@
 \end{equation}
-where $\phi_k(i) \in \mathbb{R}^3$ denotes the $i^{th}$ block of $\phi_k$.
+where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$.
 %
 If we assume that our data are one--dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, one can show that the ``correct'' embedding coordinate (i.e.
-the analog of $\phi_2$ from the diffusion maps case) is $\psi_{1,4}$.
+the analog of $\phi_2$ from the diffusion maps case) is $\psi_{1,d+1}$.
 %
@@
 
-\section{Appendix}
+Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of aligning and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps. 
+%
+The data has been ordered by $\psi_{1, 3}$.
 
-\subsection{Angular Synchronization Implementation}
+\subsection{Registering Images with Respect to Translations and Rotations}
+
+We now note that, for \eqref{eq:H_low_rank} to be satisfied, we require that the group operation matrices be orthogonal, i.e. $R(\alpha)^{-1} = R(\alpha)^T$. 
+%
+The typical matrix representation for $ISO(2)$, the group of two--dimensional translations and rotations (using the matrix representation for the affine group \cite{...}), does not satisfy this property.
+%
+However, we can (approximately) represent rotations and translations in two--dimensions using three--dimensional rotation matrices.
+%
+We do this by projecting the image onto a portion of the surface of a (three--dimensional) sphere.
+%
+Rotations in two--dimensions now correspond to rotations around one principal axis in three--dimensions, and translations in two--dimensions correspond (approximately) to rotations around the other two principal axes in three--dimensions \footnote{Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$.
+However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.}.
+%
+These rotation matrices are (by definition) orthoganol, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
+%
+See the Appendix for more details
+
+\section{Results}
+
+\begin{figure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{fig4a}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{dpERK_rank_corr}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{fig4c}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{Dl_rank_corr}
+\caption{}
+\end{subfigure}
+\end{figure}
+
+\appendix
+
+\section{Computing Optimal Alignments for One--Dimensional Signals}
+
+To compute the optimal angles in \eqref{eq:opt_angle}, we search over all possible shifts of the signals.
 
 \subsubsection{Computing Optimal Pairwise Alignments}
+The first step in angular synchronization is to compute the translations and rotations to optimally align each pair of data points. 
+%
+For each image pair $I_i$ and $I_j$, we solve
+\begin{equation}\label{eq:opt_pairwise}
+(\theta_{ij}, dx_{ij}, dy_{ij}) = \argmin_{
+\begin{matrix}
+\theta \in [0, 2\pi) \\ 
+dx \in [-\Delta, \Delta]\\ 
+dy \in [-\Delta, \Delta]
+\end{matrix}
+} \|f(I_j, \theta, dx, dy) - I_i \|_F^2.
+\end{equation}
+where $\| \cdot \|_F$ denotes the Frobenius norm, $\Delta$ is the maximum number of pixels by which we are allowed to translate the image, and $f(I_j, \theta, dx, dy)$ is a function that first rotates the image $I_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction.
+%
 
@@
 
+ADD HOW WE ROTATE AND TRANSLATE, FILL IN MISSING PIXELS, ETC.
+
 \subsubsection{Computing Optimal Translations and Rotations from the Eigenvectors of $H$}
+
+%
+We then compute the three--dimensional rotation, $R_{ij} \in SO(3)$, from the two--dimensional translation and rotation parameters $\theta_{ij}$, $dx_{ij}$, and $dy_{ij}$ (see Appendix for more details).
+%
+We then build the matrix $H \in \mathbb{R}^{3m \times 3m}$ as described in \eqref{eq:H_to_R}, and compute the top block eigenvector of $H$, $\hat{R} \in \mathbb{R}^{3m \times 3}.$ 
+%
+The $3 \times 3$ blocks of $\hat{R}$ contain approximations of $R_1, \dots, R_m$.  
+%
+Because the blocks are not guaranteed to lie in $SO(3)$, we then project each block back to $SO(3)$ (see Appendix).
+%
+To register the images, we compute the translations and rotation corresponding to $R_i^T$.
 

### 2014-02-27 image_analysis_paper/ang_synch.m

@@ -8,17 +8,24 @@ end
 
 
 [V, D] = eigs(R, dim);
-if det(V(1:dim,:)*D*V(1:dim,:)') < 0
-    V(:,end) = -V(:,end);
+[~, ind] = sort(abs(diag(D)), 'descend');
+V = V(:, ind);
+D = D(ind, ind);
+for i=1:dim
+    V(:,i) = V(:,i) / norm(V(:,i));
+end
+if det(V(1:dim,1:dim)*D(1:dim,1:dim)*V(1:dim,1:dim)') < 0
+    V(:,dim) = -V(:,dim);
 end
 
 R_opt = V;
 for i=1:m
     [u, s, v] = svd(R_opt(dim*(i-1)+1:dim*i,:));
-    R_opt(dim*(i-1)+1:dim*i,:) = u * v';
+    R_opt(dim*(i-1)+1:dim*i,:) = v * u';
 end
-
-R_opt = R_opt * R_opt(1:dim,:)';
+for i=m:-1:1
+    R_opt(dim*(i-1)+1:dim*i,:) = R_opt(1:dim,:)' * R_opt(dim*(i-1)+1:dim*i,:);
+end
 
 
 % R_opt = V*D*V(1:dim,:)';

### 2014-02-27 image_analysis_paper/compute_pairwise_alignments.m

@@ -32,9 +32,9 @@ for i=1:m
                     d2 = sum(sum((images(:,:,j) - double(imagei_tmp2)).^2));
                     if d2 < W(i, j)
                         W(i,j) = d2;
-                        thetas(i, j) = theta_vec(k);
-                        dx(i, j) = shifts(k1);
-                        dy(i, j) = shifts(k2);
+                        thetas(j, i) = theta_vec(k);
+                        dx(j, i) = shifts(k1);
+                        dy(j, i) = shifts(k2);
                     end
                 end
             end
@@ -45,8 +45,8 @@ end
 for i=1:m
     for j=1:i-1
         W(j,i) = W(i,j);        
-        R(dim*(i-1)+1:dim*i, dim*(j-1)+1:dim*j) = calc_rot_matrix(thetas(i,j), dx(i,j), dy(i,j), npixels, angle_proj);
-        R(dim*(j-1)+1:dim*j, dim*(i-1)+1:dim*i) = R(dim*(i-1)+1:dim*i, dim*(j-1)+1:dim*j)';
+        R(dim*(j-1)+1:dim*j, dim*(i-1)+1:dim*i) = calc_rot_matrix(thetas(j, i), dx(j, i), dy(j, i), npixels, angle_proj);
+        R(dim*(i-1)+1:dim*i, dim*(j-1)+1:dim*j) = R(dim*(j-1)+1:dim*j, dim*(i-1)+1:dim*i)';
     end
 end
 

### 2014-02-27 image_analysis_paper/make_1d_alignment_figures.m

@@ -1,146 +1,0 @@
-addpath ../membrane_pictures/synchron_SO2
-
-%% plot scrambled data
-figure;
-imagesc(dpERK_unaligned)
-xlabel('position (unaligned)')
-ylabel('data point (unordered)')
-set(gca, 'xtick', [])
-set(gca, 'ytick', [])
-if print_figures
-    print('data_unaligned_unordered',fmt, res)
-end
-
-%% calculate pairwise alignments
-tic
-[H, W, theta] = align_data(dpERK_unaligned);
-
-%% angular synchronization
-R_opt = ang_synch(H, 2);
-
-runtime_angsynch = toc
-dpERK_aligned = zeros(size(dpERK_unaligned));
-R_est = R_opt(1:2, 1:2);
-theta_est = atan2(R_est(2,1),R_est(1,1));
-shift_est0 = round(theta_est/(2*pi)*n) + rand_offsets(1);
-for i=1:m
-    R_est = R_opt(2*i-1:2*i, 1:2);
-    theta_est = atan2(R_est(2,1),R_est(1,1));
-    shift_est = round(theta_est/(2*pi)*n) - shift_est0;
-    dpERK_aligned(i,:) = circshift(dpERK_unaligned(i,:),[0 shift_est]);
-end
-
-%% plot alignment with angular synchronization
-figure;
-imagesc(dpERK_aligned)
-xlabel('position (aligned using angular synchronization)')
-ylabel('data point (unordered)')
-set(gca, 'xtick', [])
-set(gca, 'ytick', [])
-if print_figures
-    print('data_aligned_unordered',fmt, res)
-end
-
-%% dmaps following angular synchronization
-W = squareform(pdist(dpERK_aligned)).^2;
-eps_dmaps = median(W(:));
-[V_dmaps, D_dmaps] = dmaps(W, eps_dmaps, 10);
-
-if corr(V_dmaps(:,2), L(:,1)) < 0
-    V_dmaps(:,2) = -V_dmaps(:,2);
-end
-
-%% plot dmaps ordering
-[~, I] = sort(V_dmaps(:,2));
-figure;
-imagesc(dpERK_aligned(I,:))
-ylabel('data point (ordered using DMAPS)')
-xlabel('position (aligned using angular synchronization)')
-set(gca, 'xtick', [])
-set(gca, 'ytick', [])
-if print_figures
-    print('data_ordered_angsynch',fmt, res)
[…]

### 2014-02-27 image_analysis_paper/make_2d_alignment_figures.m

@@ -1,212 +1,0 @@
-addpath('../membrane_pictures/synchron_so3_nosphharm')
-
-%% parameters
-
-% amont of border or buffer to add around the images
-buffer_size = 20;
-
-% maximum shift in pixels
-shift_max = 20;
-
-%size of "portion" of sphere on which to project
-angle_proj = pi/4;
-
-% total number of pixels
-npixels2 = npixels+2*buffer_size;
-
-% dimension of rotations
-dim = 3;
-
-% store images in 3d array
-image_set_buffered = zeros(npixels2, npixels2, m);
-
-%% load in images
-
-for i=1:m
-    %store image
-    im1 = image_set(:,:,i);
-    image_set_buffered(buffer_size+1:buffer_size+npixels,buffer_size+1:buffer_size+npixels,i) = im1;
-end
-
-%% compute pairwise alignments
-tic
-xmax = 0.1;
-ymax = 0.1;
-[R, W, angles] = align_data_nosph(image_set_buffered, xmax, ymax, shift_max);
-
-        
-%% angular synchronization
-% dim = 3;
-% 
-% R_opt = ang_synch(R, dim);
-% 
-% image_set_aligned = zeros(size(image_set));
-% 
-% figure;
-% for i=1:m
-%     
-%     image_tmp = image_set_buffered(:,:,i);
-%     
-%     R_all = R_opt(1:3, 1:3)'*R_opt(3*i-2:3*i,:);
-%     
-%     [alpha, beta, gamma] = calc_angles(R_all);    
-%     image_tmp = shift_image(image_tmp, alpha, beta, gamma, angle_proj);
-%     image_set_aligned(:,:,i) = image_tmp(buffer_size+1:buffer_size+npixels,buffer_size+1:buffer_size+npixels);
-%     
-%     subplot(subplot_dim1, subplot_dim2, i);
-%     imshow(uint8(image_set_aligned(:,:,i)),'InitialMagnification', 'fit')
-% 
-% end
-% 
-% W_new = zeros(m);
-% 
-% for i=1:m
-%     for j=1:i-1
-%         W_new(i,j) = sum(sum((image_set_aligned(:,:,i)-image_set_aligned(:,:,j)).^2));
-%         W_new(j,i) = W_new(i,j);
-%     end
-% end
-% 
-% eps = median(W_new(:));
-% [V, D] = dmaps(W_new, eps, 10);
[…]

### 2014-02-27 image_analysis_paper/make_2d_scattering_figures.m

@@ -1,76 +1,0 @@
-addpath 'C:\Users\cdsilva\Documents\MATLAB\scatnet-0.2';
-%addpath '../../../MATLAB/scatnet-0.2';
-addpath_scatnet
-
-%% compute scattering coefficients
-% set scattering transform parameters
-filt_opt = struct();
-filt_rot_opt = struct();
-% oversampling is set to infty
-% so that scattering of original and rotated
-% image will be sampled at exactly the same points
-scat_opt = struct();
-scat_opt.oversampling = 10;
-
-% compute scattering transform of first image
-x = image_set(:,:,1);
-% define wavelet transforms
-Wop = wavelet_factory_3d(size(x), filt_opt, filt_rot_opt, scat_opt);
-Sx = scat(x, Wop);
-Sx_mat = format_scat(Sx);
-
-% store scattering invariants (one row per image)
-sx_all = zeros(m, size(Sx_mat, 1));
-
-tic
-% compute scattering invariants for each image
-for i=1:m
-    x = image_set(:,:,i);
-    
-    Sx = scat(x, Wop);
-
-    sx_all(i,:) = mean(mean(format_scat(Sx),2),3)';
-end
-
-%% dmaps
-W = squareform(pdist(sx_all)).^2;
-eps = median(W(:));
-[V, D] = dmaps(W, eps, 10);
-toc
-
-if corr(V(:,2), L(:,1)) < 0
-    V(:,2) = -V(:,2);
-end
-
-[~, I] = sort(V(:,2));
-
-figure;
-plot(L(:,1), V(:,2),'.')
-hold on
-plot(L(I(im_save_idx),1), V(I(im_save_idx), 2), 'o')
-xlabel('membrane thickness')
-ylabel('$\phi_2$','interpreter','latex')
-if print_figures
-    print('DMAPS_scat_time_corr',fmt,res)
-end
-fprintf('Scattering transform Spearman coeff: %2.4f \n', corr(L(:,1),V(:,2), 'type','spearman'));
-
-if print_figures
-    figure;
-    for i=im_save_idx
-        %subplot(subplot_dim1,subplot_dim2,i)
-        set(gcf, 'paperposition',[0 0 8 8])
-        imshow(uint8(image_set(:,:,I(i))), 'InitialMagnification', 'fit')
-        set(gca,'position',[0 0 1 1],'units','normalized')
-        % make green colormap
-        cm_green = gray;
-        cm_green(:,1) = 0;
[…]

### 2014-02-27 image_analysis_paper/make_figures_large_dataset.m

@@ -112,6 +112,7 @@ for j=1:length(image_channel)
     ylabel('rank from vdm')
     
     image_set_aligned2 = uint8(image_set_aligned(:,:,I));
-    save(mat_file_names{j}, 'image_set_aligned2', 'ranks_from_membranes', 'ranks_from_vdm');
-    save(mat_file_names_all{j})
+    pause
+    %save(mat_file_names{j}, 'image_set_aligned2', 'ranks_from_membranes', 'ranks_from_vdm');
+    %save(mat_file_names_all{j})
 end

### 2014-02-27 image_analysis_paper/test_alignment.m

@@ -1,88 +1,0 @@
-clear all
-close all
-
-dpERK_data = '../membrane_pictures/large_dataset/time.mat';
-dpERK_image_dir = '../membrane_pictures/large_dataset';
-
-%% load membrane lengths
-
-load(dpERK_data);
-mem_lengths = length;
-clear length;
-
-m   = 16;
-
-%% load images
-
-npixels = 100;
-
-%set image plotting parameters
-subplot_dim1 = ceil(sqrt(m));
-subplot_dim2 = ceil(m / subplot_dim1);
-
-image_set = zeros(npixels, npixels, m);
-
-image_channel = 2;
-
-%%
-[~, ind] = max(mem_lengths);
-
-im_template = imread(sprintf('%s/lat%02d.tif', dpERK_image_dir,ind));
-%resize image
-im_template = imresize(im_template, [npixels npixels]);
-%extract relevent color from image
-im_template = im_template(:,:,image_channel);
-
-theta_vec = linspace(0, 360, 21);
-theta_vec = theta_vec(1:end-1);
-
-thetas = zeros(m,1);
-xdisp = zeros(m,1);
-ydisp = zeros(m,1);
-
-figure;
-for i=1:m
-    thetas(i) = theta_vec(randi(20,1));
-    xdisp(i) = randi([-10 10], 1);
-    ydisp(i) = randi([-10 10], 1);
-    
-    im1 = imrotate(im_template, thetas(i), 'crop');
-    im1 = circshift(im1, [xdisp(i) ydisp(i)]);
-    
-    subplot(subplot_dim1, subplot_dim2, i)
-    imshow(im1);
-    
-    %store image
-    im1 = double(im1);
-    image_set(:, :, i) = im1;
-    
-end
-
-%% raw dpERK images-- synchronization
-
-angle_proj = pi/8;
-shift_max = 20;
-shift_step = 4;
-dim = 3;
-
-tic
-[R, W] = compute_pairwise_alignments(image_set, angle_proj, shift_max, shift_step);
-
-%%
-
-eps = median(W(:));
-neigs = 5;
-
-%R_opt = ang_synch(R, dim);
-[R_opt, embed_coord, embed_idx, D] = vdm(R, W, eps, neigs);
-
-toc
-
-
-figure;
-for i=1:m
-    im1 = rotate_image(R_opt(dim*(i-1)+1:dim*i,:), uint8(image_set(:,:,i)), angle_proj);
[…]

### 2014-02-27 image_analysis_paper/vdm.m

@@ -33,9 +33,11 @@ end
 R_opt = V(:,1:dim);
 for i=1:n
     [u, s, v] = svd(R_opt(dim*(i-1)+1:dim*i,:));
-    R_opt(dim*(i-1)+1:dim*i,:) = u * v';
+    R_opt(dim*(i-1)+1:dim*i,:) = v * u';
 end
-R_opt = R_opt * R_opt(1:dim,:)';
+for i=n:-1:1
+    R_opt(dim*(i-1)+1:dim*i,:) = R_opt(1:dim,:)' * R_opt(dim*(i-1)+1:dim*i,:);
+end
 
 % 
 % R_opt = V(:,1:dim)*D(1:dim,1:dim)*V(1:dim,1:dim)';

### 2014-02-27 image_analysis_paper/math_description.tex
Added more to description

@@
 To compute the optimal angles in \eqref{eq:opt_angle}, we search over all possible shifts of the signals. 
+%
+The signal is discritized into $100$ points.
 
@@
 %
-
+The image rotation is done with the \texttt{imrotate} function in Matlab, using nearest neighbor interpolation.
+%
+The missing pixels at the corners of the rotated image are taken to be black.
+%
+The translations are done by shifting the pixels, and taking this missing pixels to be black. 
+%
 In \eqref{eq:opt_pairwise}, we take $\Delta=20$, which corresponds to a 20\% shift in the image. 

### 2014-02-27 image_analysis_paper/math_description.tex
Added more to description

@@
 
-\subsubsection{Computing Optimal Pairwise Alignments}
+\section{Computing Optimal Pairwise Alignments for Two--Dimensional Images}
 The first step in angular synchronization is to compute the translations and rotations to optimally align each pair of data points. 
@@
 \end{matrix}
-} \|f(I_j, \theta, dx, dy) - I_i \|_F^2.
+} \|g(I_j, \theta, dx, dy) - I_i \|_F^2.
 \end{equation}
-where $\| \cdot \|_F$ denotes the Frobenius norm, $\Delta$ is the maximum number of pixels by which we are allowed to translate the image, and $f(I_j, \theta, dx, dy)$ is a function that first rotates the image $I_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction.
+where $\| \cdot \|_F$ denotes the Frobenius norm, $\Delta$ is the maximum number of pixels by which we are allowed to translate the image, and $g(I_j, \theta, dx, dy)$ is a function that first rotates the image $I_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction.
 %
@@
 %
-The translations are done by shifting the pixels, and taking this missing pixels to be black. 
+The translations are done by shifting the pixels, and taking the missing pixels at the edges of the image to be black; we only consider shifts that correspond to an integer number of pixels, to remove the need for interpolation when doing the translations.
 %
@@
 
-\subsubsection{Converting from $ISO(2)$ to $SO(3)$}
+\section{Converting from $ISO(2)$ to $SO(3)$}
 
@@
 
-\subsubsection{Computing Optimal Translations and Rotations from the Eigenvectors of $H$}
+\section{Computing Optimal Translations and Rotations from the Eigenvectors of $H$}
 

### 2014-03-03 image_analysis_paper/math_description.tex
Added figures with sphere for paper

@@
 \caption{} 
+\label{fig:fluorescent_images}
 \end{figure}
@@
 
-We will first discuss angular synchronization with {\em only} rotations. 
+We will first discuss angular synchronization {\em only} considering rotations. 
 %
-Let $x_1, \dots, x_m \in \mathbb{R}^x$ denote the (grayscale) signal that we wish to align with respect to rotations.
+Let $x_1, \dots, x_m \in \mathbb{R}^x$ denote the signals that we wish to align with respect to rotations.
 %
@@
 \end{equation}
-where $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ degrees \footnote{add about how we rotate, because it's soooooo complicated}. 
+where $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ degrees, and $\xi_i$ is a noise term (often Gaussian noise). 
 %
@@
 %
-We can obtain {\em estimates} of $\theta_i - \theta_j$ by computing the rotations to optimally align $x_j$ to $x_i$. 
-%
-Let $\theta_{ij}$ denote the rotation that optimally aligns $x_j$ to $x_i$, i.e.,
+We can obtain {\em estimates} of $\theta_i - \theta_j$ by computing the rotations to optimally align $x_j$ to $x_i$, 
+i.e., $\theta_{ij} \approx \theta_i - \theta_j$, where
 %
 \begin{equation} \label{eq:opt_angle}
-\theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2
+\theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2.
 \end{equation}
 
-Rather than work with the angles directly, it will be more convenient to work with the rotation matrices 
+Rather than work with the angles $\theta_{ij}$ directly, it will be more convenient to work with the rotation matrices 
 \begin{equation} \label{eq:R_theta}
@@
 
-Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $d=2$).
+Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$; we describe this for general $d$ because we will later consider three--dimensional rotations).
 %
@@
 
-We note that, under our assumptions, 
+We note that, under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, 
 \begin{equation} 
@@
 %
-We can (approximately) recover $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$ by computing the top block eigenvector of $H$ (i.e the eigenvectors corresponding to the $d$ largest eigenvalues, stacked into a $md \times 2$ matrix).
+We can (approximately) recover $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$ by computing the top block eigenvector of $H$, which we will denote $\hat{R}$.
+%
+Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$, ordered such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_{md}$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$.
+%
+Then
+\begin{equation}
+\hat{R} = 
+\begin{bmatrix}
+\hat{R}_1 \\
+\hat{R}_2 \\
+\vdots \\
+\hat{R}_m
+\end{bmatrix} =
+\begin{bmatrix}
+| & | & & | \\
+\phi_1 & \phi_2 & \dots & \phi_d \\
+| & | & & | 
+\end{bmatrix}
+\end{equation}
+where $\hat{R}_i \in \mathbb{R}^{d \times d}$ is the estimate for $R(\theta_i)$. 
+%
+Becaus $\hat{R}_i$ is not guaranteed to be a rotation matrix, we project $\hat{R}_i$ onto the closest rotation matrix. 
+%
+Let $U_i$ and $V_i$ denote the left and right singular vectors, respectively, of $\hat{R}_i$.
+%
+Then
+\begin{equation}
+R_{i, est} = U_i V_i^T
+\end{equation} 
+where $R_{i, est}$ is our estimate of $R(\theta_i)$. 
+%
+Note that $R_{i, est}$ is not guaranteed to be a proper rotation; because we are permitted to scale the eigenvectors of a matrix, we adjust the sign of $\rho_d$ so that (most of) the rotations are proper.
 %
@@
 %
-The {\em essential} requirement for DMAPS is an appropriate distance metric $d(x_i, x_j)$ for comparing data points. 
-%
-This can as simple as the standard Euclidean distance, or a more complex metric (such as a distance between features of the data points) for other data sets.
-
 In our framework, we assume that our data lie on a one--dimensional curve or manifold, and that this one dimension is correlated with time. 
@@
 We will uncover a parameterization of this curve by only considering {\em local} neighbors of each data point.
-%
+
 To illustrate how local neighbors can allow us to uncover a global parameterization, assume we have data $x_1, x_2, \dots, x_m \in \mathbb{R}$ such that $x_1 < x_2 < \dots < x_m$.
@@
 %
-We then construct the weight matrix, $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if $x_i$ is close to $x_j$,
+We then construct the (symmetric) weight matrix, $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if $x_i$ is close to $x_j$,
 \begin{equation}
@@
 %
-If we normalize this matrix such that $\sum_j W_{ij} = 1$, then one can show that the first eigenvector is the all--ones vector, $\phi_1 = [1 1 \dots 1]^T$, and the second eigenvector, $\phi_2 = [cos(\pi/m) \cos(2 \pi/ m) \dots \cos((m-1) \pi / m) \cos(\pi)]^T$.
+If we normalize this matrix such that $\sum_j W_{ij} = 1$, then one can show that the first eigenvector of this normalized matrix is the all--ones vector, $\phi_1 = [1 1 \dots 1]^T$, and the second eigenvector is a cosine mode, $\phi_2 = [cos(\pi/m) \cos(2 \pi/ m) \dots \cos((m-1) \pi / m) \cos(\pi)]^T$.
 %
-Therefore, the second eigenvector, $\phi_2$ is {\em one--to--one} with the parameterization of the data $x_1, x_2, \dots, x_m$. 
+Therefore, the second eigenvector, $\phi_2$ is {\em one--to--one} with the parameterization of the data, i.e. $\phi_2(1) < \phi_2(2) < \dots < \phi_2(m)$, just as $x_1 < x_2 < \dots < x_m$. 
 %
@@
 %
-For general data in high--dimensional space, we typically use a Gaussian kernel,
+For general data in high--dimensional space, we typically use a Gaussian kernel to construct $W$,
 \begin{equation} \label{eq:dmaps_W}
@@
 %
-The distance $d(x_i, x_j)$ is the distance between samples {\em after} alignment (i.e. it is the minimium obtained in \eqref{eq:opt_pairwise}).
+The distance $d(x_i, x_j)$ used to compute $W_{ij}$ is the distance between samples {\em after} pairwise alignment (i.e. it is the minimium obtained in \eqref{eq:opt_pairwise}).
 
@@
 %
-Again, the top (block) eigenvector of $S$ gives us approximations to the optimal rotations for each image.
+Again, the top (block) eigenvector of $S$ contains approximations of the optimal rotations for each image.
 %
@@
 %
-If we assume that our data are one--dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, one can show that the ``correct'' embedding coordinate (i.e.
-the analog of $\phi_2$ from the diffusion maps case) is $\psi_{1,d+1}$.
+If we assume that our data are one--dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, one can show that a parameterization of our data (i.e.
+the analog of $\phi_2$ from the diffusion maps case) is given by $\psi_{1,d+1}$.
 %
-If no such claims can be made, then one often runs a second step of diffusion maps, using $\psi_{k,l}$ as the data points, in order to uncover the true manifold modulo symmetries.
+If no such claims can be made, then one often runs a second step of diffusion maps, using $\psi_{k,l}$ as the coordinates, in order to uncover the true manifold modulo symmetries.
 
@@
 
-\subsection{Registering Images with Respect to Translations and Rotations}
+\subsection{Registering Images with Respect to Translations and Rotations} \label{subsec:trans_rot_register}
 
-We now note that, for \eqref{eq:H_low_rank} to be satisfied, we require that the group operation matrices be orthogonal, i.e. $R(\alpha)^{-1} = R(\alpha)^T$. 
+We are not only interested in registering the one--dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two--dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\em and} translations.
 %
-The typical matrix representation for $ISO(2)$, the group of two--dimensional translations and rotations (using the matrix representation for the affine group \cite{...}), does not satisfy this property.
+However, for \eqref{eq:H_low_rank} to be satisfied, we require that the matrices used to represent the symmetry group be orthogonal, i.e. $R(\alpha)^{-1} = R(\alpha)^T$. 
+%
+The typical matrix representation for the group of two--dimensional translations and rotations (using the matrix representation for the affine group \cite{...}) does not satisfy this property. 
 %
@@
 %
-We do this by projecting the image onto a portion of the surface of a (three--dimensional) sphere.
+We do this by projecting the image onto a portion of the surface of a (three--dimensional) sphere (see Appendix).
 %
@@
 %
-See the Appendix for more details
+See the Appendix for more details.
 
@@
 
-To compute the optimal angles in \eqref{eq:opt_angle}, we search over all possible shifts of the signals. 
+Each signal shown in Figure \ref{fig:1d_demo} is discretized into 100 points. 
 %
-The signal is discritized into $100$ points.
+To compute the optimal angles in \eqref{eq:opt_angle}, we exhaustively search over all 100 possible shifts/rotations of the signals.
 
 \section{Computing Optimal Pairwise Alignments for Two--Dimensional Images}
-The first step in angular synchronization is to compute the translations and rotations to optimally align each pair of data points. 
+
+The first step in angular synchronization or vector diffusion maps is to compute the translations and rotations to optimally align each pair of data points. 
 %
@@
 %
-The image rotation is done with the \texttt{imrotate} function in Matlab, using nearest neighbor interpolation.
+The image rotation is done with the \texttt{imrotate} function in Matlab, using nearest neighbor interpolation to interpolate the intensities of the pixels (since there is no direct mapping of the pixels of the unrotated image to the pixels of the rotated image).
 %
-The missing pixels at the corners of the rotated image are taken to be black.
+Also, the missing pixels in the rotated image due to corner effects are taken to be black.
 %
-The translations are done by shifting the pixels, and taking the missing pixels at the edges of the image to be black; we only consider shifts that correspond to an integer number of pixels, to remove the need for interpolation when doing the translations.
+The translations are done by shifting the pixels with periodic boundary conditions.
 %
-In \eqref{eq:opt_pairwise}, we take $\Delta=20$, which corresponds to a 20\% shift in the image. 
+In \eqref{eq:opt_pairwise}, $\Delta$ is chosen such that only edge pixels with little to no signal are shifted, and the main image is not split or separated when translating (we take $\Delta=20$, which corresponds to a 20\% shift in the image).
 %
+We only consider shifts that correspond to an integer number of pixels, to remove the need for interpolation when doing the translations.
+
 The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
@@
 
+\begin{figure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{sphere_1}
+\includegraphics[width=\textwidth]{sphere2_1}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{sphere_2}
+\includegraphics[width=\textwidth]{sphere2_2}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{sphere_3}
+\includegraphics[width=\textwidth]{sphere2_3}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{sphere_4}
+\includegraphics[width=\textwidth]{sphere2_4}
+\caption{}
+\end{subfigure}
+\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions.
+(a) The original image.
+(b) Rotation around the x--axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image.
+(c) Rotation around the y--axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation.
+(d) Rotation around the z--axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.
+The top row shows the three--dimensional spheres, and the bottom row shows the (two--dimensional) surface of the sphere in which we are interested.}
+\end{figure}
+
+As discussed in Section \ref{subsec:trans_rot_register}, the underlying symmetry group needs to have a real and orthoganol matrix representation. 
+%
+We therefore convert $ISO(2)$, the group of two--dimensional translations and rotations, to $SO(3)$, the group of three--dimensional rotations. 
+%
 In the space of three--dimensional rotations, the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$ correspond to
@@
 
-ADD HOW WE ROTATE AND TRANSLATE, FILL IN MISSING PIXELS, ETC.
-
 \section{Computing Optimal Translations and Rotations from the Eigenvectors of $H$}
 
-%
 We then compute the three--dimensional rotation, $R_{ij} \in SO(3)$, from the two--dimensional translation and rotation parameters $\theta_{ij}$, $dx_{ij}$, and $dy_{ij}$ (see Appendix for more details).
 %
-We then build the matrix $H \in \mathbb{R}^{3m \times 3m}$ as described in \eqref{eq:H_to_R}, and compute the top block eigenvector of $H$, $\hat{R} \in \mathbb{R}^{3m \times 3}.$ 
+We then build the matrix $H$ and compute the estimates $R_{i, est}$ as described in Section \ref{subsec:angsynch}.
 %
-The $3 \times 3$ blocks of $\hat{R}$ contain approximations of $R_1, \dots, R_m$.  
-%
-Because the blocks are not guaranteed to lie in $SO(3)$, we then project each block back to $SO(3)$ (see Appendix).
-%
-To register the images, we compute the translations and rotation corresponding to $R_i^T$.
-
-From the top block eigenvector $\hat{R} \in \mathbb{R}^{3m \times 3}$, we would like to compute the optimal translation and rotation for each image. 
-%
-The optimal three--dimensional rotations are given by $R_i = U_i V_i^T$, where $U_i$ and $V_i$ are the left and right singular vectors of $\hat{R}_i$, respectively \cite{...}. 
-%
-Because the eigenvectors of $H$ are determined up to a sign, we adjust the signs of the eigenvectors so that $det(R_i) = +1$ to ensure proper rotations.
-
-From the matrices $R_1, \dots, R_m$, we now wish to find the corresponding translations and rotations of the images.
+From the matrices $R_{1, opt}, \dots, R_{m, opt}$, we now wish to find the corresponding translations and rotations of the images.
 %
@@
 We can then invert \eqref{eq:angle_relations} to compute the optimal translation and rotation for each image from the Euler angles.
+%
+We round the optimal translations to the nearest integer.
 

### 2014-03-03 image_analysis_paper/make_sphere.m

@@ -1,0 +1,76 @@
+clear all
+close all
+
+res = '-r300';
+fmt = '-djpeg';
+
+im_save_dir = 'paper_figures';
+
+n = 100;
+
+[X, Y, Z] = sphere(n);
+
+C = zeros(size(X));
+C(round(0.4*n):round(0.6*n), round(0.2*n):round(0.3*n)) = 1;
+
+v0 = [0 0];
+v1 = [-1.5 1.5];
+
+dir = [0 0 0;
+    0 1 0;
+    0 0 1;
+    1 0 0];
+
+for i=1:4
+    figure;
+    set(gcf, 'papersize', [1 1])
+    h = surf(X, Y, Z, C, 'edgecolor','none');
+    set(gca,'position',[0 0 1 1],'units','normalized')
+    camlight(45, -20);
+    lighting phong
+    material dull
+    
+    axis off
+    hold on
+    plot3(v1, v0, v0, 'k')
+    plot3(v0, v1, v0, 'k')
+    plot3(v0, v0, v1, 'k')
+    
+    if i > 1
+        rotate(h,dir(i, :),25)
+    end
+    
+    print(sprintf('%s/sphere_%d', im_save_dir, i), fmt, res)
+end
+
+figure;
+for i=1:4
+    figure;
+    set(gcf, 'papersize', [1 1])
+    h = surf(X, Y, Z, C, 'edgecolor','none');
+    set(gca,'position',[0 0 1 1],'units','normalized')
+    
+    camlight(45, -20);
+    lighting phong
+    material dull
+    
+    axis off
+    hold on
+    plot3(v1, v0, v0, 'k')
+    plot3(v0, v1, v0, 'k')
+    plot3(v0, v0, v1, 'k')
+    if i > 1
+        rotate(h,dir(i, :),25)
+    else
+        annotation('doublearrow', [0.4 0.6], [0.55 0.55],'color','w')
+        annotation('doublearrow', [0.55 0.55], [0.4 0.6],'color','w')
+        text(0, -1, 0.2, '\eta_{proj}','color','w')
+        text(0.17, -1, 0, '\eta_{proj}','color','w')
+        
+    end
+    view(0, 0)
+    
+    print(sprintf('%s/sphere2_%d', im_save_dir, i), fmt, res)
+    
+end
+

### 2014-03-03 image_analysis_paper/math_description.tex
Updated math for paper

@@
 %
-Let $x_1, \dots, x_m \in \mathbb{R}^x$ denote the signals that we wish to align with respect to rotations.
+Let $x_1, \dots, x_m \in \mathbb{R}^n$ denote the signals that we wish to align with respect to rotations.
 %
@@
 \end{equation}
-where $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ degrees, and $\xi_i$ is a noise term (often Gaussian noise). 
+where the function $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ degrees, and $\xi_i$ is a noise term (often Gaussian noise). 
 %
@@
 %
-We can obtain {\em estimates} of $\theta_i - \theta_j$ by computing the rotations to optimally align $x_j$ to $x_i$, 
+We can obtain an {\em estimate} of $\theta_i - \theta_j$ by computing the rotation that optimally aligns $x_j$ to $x_i$, 
 i.e., $\theta_{ij} \approx \theta_i - \theta_j$, where
@@
 %
-Successive rotations correspond to multiplication of the rotation matrices, i.e., 
+Successive rotations now correspond to multiplication of rotation matrices, i.e., 
 $R(\alpha_1 + \alpha_2) = R(\alpha_1) R(\alpha_2)$,
@@
 
-Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$; we describe this for general $d$ because we will later consider three--dimensional rotations).
+Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$; we will write our procedure for general $d$ because we will later consider three--dimensional rotations).
 %
-We then construct the matrix $H \in \mathbb{R}^{md \times md}$, where $H$ is an $m \times m$ matrix of $d \times d$ blocks with
+We construct the matrix $H \in \mathbb{R}^{md \times md}$, where $H$ is an $m \times m$ matrix of $d \times d$ blocks with
 \begin{equation} \label{eq:H_to_R}
@@
 	R(\theta_1)^T R(\theta_2)^T \dots R(\theta_m)^T
-	\end{bmatrix}
+	\end{bmatrix}.
 \end{equation}
 %
-We can (approximately) recover $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$ by computing the top block eigenvector of $H$, which we will denote $\hat{R}$.
+From \eqref{eq:H_low_rank}, we can see that we can (approximately) recover $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$ by computing the top block eigenvector of $H$, which we will denote $\hat{R}$.
 %
-Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$, ordered such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_{md}$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$.
+Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$.
 %
-Then
+Then the top block eigenvector of $H$, $\hat{R}$, can be written as
 \begin{equation}
@@
 | & | & & | 
-\end{bmatrix}
+\end{bmatrix},
 \end{equation}
-where $\hat{R}_i \in \mathbb{R}^{d \times d}$ is the estimate for $R(\theta_i)$. 
+and $\hat{R}_i \in \mathbb{R}^{d \times d}$ is the estimate for $R(\theta_i)$. 
 %
-Becaus $\hat{R}_i$ is not guaranteed to be a rotation matrix, we project $\hat{R}_i$ onto the closest rotation matrix. 
+Because $\hat{R}_i$ is not guaranteed to be a rotation matrix, we must project $\hat{R}_i$ onto the closest rotation matrix \cite{...}. 
 %
@@
 \end{equation} 
-where $R_{i, est}$ is our estimate of $R(\theta_i)$. 
+where $R_{i, est}$ is an orthonormal matrix and is our estimate of $R(\theta_i)$. 
 %
-Note that $R_{i, est}$ is not guaranteed to be a proper rotation; because we are permitted to scale the eigenvectors of a matrix, we adjust the sign of $\rho_d$ so that (most of) the rotations are proper.
+Note that $R_{i, est}$ is not guaranteed to be a proper rotation (i.e.
+$det(R_{i, est} = \pm 1)$; because we are permitted to scale the eigenvectors of a matrix, we adjust the sign of $\\phi_d$ so that (most of) the rotations are proper.
 %
-We can estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and then register the signals by rotating image $i$ by $-\theta_i$.
+We can now estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and then register the signals by rotating signal $i$ by $-\theta_i$.
 
-Furthermore, this eigendecomposition also considers {\em higher--order} consistency information. 
+Furthermore, this formulation also considers {\em higher--order} consistency information. 
 %
@@
 %
-However, 
+Note that
 \begin{equation}
-(H^2)_{ij} = \sum_k R(\theta_{ik}) R(\theta_{kj})
+(H^2)_{ij} = \sum_k R(\theta_{ik}) R(\theta_{kj});
 \end{equation}
-and so {\em all} infomation of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$.
+therefore, {\em all} infomation of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$.
 %
@@
 
-To illustrate how angular synchronization works, we have converted the two--dimensional images into concentration profiles on a ring (we will address the full two--dimensional images in a subsequent section).
+To illustrate how angular synchronization works, we have converted the two--dimensional images shown in Figure \ref{fig:fluorescent_images} into concentration profiles on a ring (we will address the full two--dimensional images in a subsequent section).
 %
@@
 %
-In our framework, we assume that our data lie on a one--dimensional curve or manifold, and that this one dimension is correlated with time. 
+In our analysis, we will assume that our data lie on a one--dimensional curve or manifold, and that this one dimension is correlated with time. 
 %
@@
 %
-We then construct the (symmetric) weight matrix, $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if $x_i$ is close to $x_j$,
+We then construct the (symmetric) weight matrix, $W \in \mathbb{R}^{m \times m}$,
 \begin{equation}
@@
 %
-If we normalize this matrix such that $\sum_j W_{ij} = 1$, then one can show that the first eigenvector of this normalized matrix is the all--ones vector, $\phi_1 = [1 1 \dots 1]^T$, and the second eigenvector is a cosine mode, $\phi_2 = [cos(\pi/m) \cos(2 \pi/ m) \dots \cos((m-1) \pi / m) \cos(\pi)]^T$.
+$W_{ij}$ is large if $x_i$ is close to $x_j$,
 %
-Therefore, the second eigenvector, $\phi_2$ is {\em one--to--one} with the parameterization of the data, i.e. $\phi_2(1) < \phi_2(2) < \dots < \phi_2(m)$, just as $x_1 < x_2 < \dots < x_m$. 
+If we normalize this matrix such that $\sum_j W_{ij} = 1$, then one can show that the first eigenvector of this normalized matrix is the all--ones vector, $\phi_1 = [1 1 \dots 1]^T$, and the second eigenvector is a cosine mode, $\phi_2 = [cos(\pi/m) \cos(2 \pi/ m) \cdots \cos((m-1) \pi / m) \cos(\pi)]^T$.
+%
+Therefore, the second eigenvector, $\phi_2$ is {\em one--to--one} with the parameterization of the data, i.e. $\phi_2(1) > \phi_2(2) > \dots > \phi_2(m)$, just as $x_1 < x_2 < \dots < x_m$. 
 %
@@
 
-We demonstrate how we can do this with our one--dimensional concentration profiles. 
+We demonstrate how we can apply these techniques to our (aligned)one--dimensional concentration profiles. 
 %
@@
 %
-Again, the top (block) eigenvector of $S$ contains approximations of the optimal rotations for each image.
+Again, the top (block) eigenvector of $S$ contains approximations of the optimal rotations for each image, and the rotations can be computed as described in Section \ref{subsec:angsynch}.
 %
-However, we now also obtain embedding coordinates for our images.
+However, the eigenvectors now also contain information about the embedding coordinates for our images.
 %
@@
 %
-We do this by projecting the image onto a portion of the surface of a (three--dimensional) sphere (see Appendix).
+We do this by projecting the image onto a portion of the surface of a (three--dimensional) sphere (see Figure \ref{fig:SO3_picture}).
 %
@@
 %
-The image rotation is done with the \texttt{imrotate} function in Matlab, using nearest neighbor interpolation to interpolate the intensities of the pixels (since there is no direct mapping of the pixels of the unrotated image to the pixels of the rotated image).
+The image rotation is done with the \texttt{imrotate} function in Matlab, using nearest neighbor interpolation to estimate the pixel intensities, since there is no direct mapping of the pixels of the unrotated image to the pixels of the rotated image.
 %
-Also, the missing pixels in the rotated image due to corner effects are taken to be black.
+The missing pixels in the rotated image due to corner effects are taken to be black.
 %
@@
 %
-Therefore, instead of using an optimization procedure to compute the solution to \eqref{eq:opt_pairwise}, we choose to discretize the search space and exhaustively search to estimate the solution to \eqref{eq:opt_pairwise}.
+Therefore, instead of using an optimization procedure to compute the solution to \eqref{eq:opt_pairwise}, we choose to discretize the search space and exhaustively search to estimate the solution.
 %
@@
 The top row shows the three--dimensional spheres, and the bottom row shows the (two--dimensional) surface of the sphere in which we are interested.}
+\label{fig:SO3_picture}
 \end{figure}
@@
 \begin{aligned}
-	R^x_{ij} &= \begin{bmatrix}
+	R^x(\alpha) &= \begin{bmatrix}
 	1 & 0 & 0 \\
@@
 	\end{bmatrix} \\
-	R^y_{ij} &= \begin{bmatrix}
+	R^y(\beta) &= \begin{bmatrix}
 	\cos(\beta) & 0 & \sin(\beta) \\
@@
     \end{bmatrix} \\
-	R^z_{ij} &= \begin{bmatrix} 
+	R^z(\gamma) &= \begin{bmatrix} 
 	\cos(\gamma) & -\sin(\gamma) & 0 \\
@@
 \end{equation}
-where $R^x_{ij}$, $R^y_{ij}$, and $R^z_{ij}$ correspond to rotations around the $x$, $y$, and $z$ axes, respectively.
+where $R^x(\alpha)$, $R^y(\beta)$, and $R^z(\gamma)$ correspond to rotations around the $x$, $y$, and $z$ axes, respectively.
 %
@@
 \begin{equation} \label{eq:total_rot}
-	R_{ij}	 = R^z_{ij} \times R^y_{ij} \times R^x_{ij}
+	R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})	 = R^z(\gamma_{ij})  R^y(\beta_{ij})  R^x(\alpha_{ij})
 \end{equation}
@@
 
-We then compute the three--dimensional rotation, $R_{ij} \in SO(3)$, from the two--dimensional translation and rotation parameters $\theta_{ij}$, $dx_{ij}$, and $dy_{ij}$ (see Appendix for more details).
+From the three--dimensional rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$,
+we build the matrix $H$ and compute the estimates $R_{i, est}$ as described in Section \ref{subsec:angsynch}.
 %
-We then build the matrix $H$ and compute the estimates $R_{i, est}$ as described in Section \ref{subsec:angsynch}.
+From the matrices $R_{1, est}, \dots, R_{m, est}$, we now wish to find the corresponding translations and rotations of the images.
 %
-From the matrices $R_{1, opt}, \dots, R_{m, opt}$, we now wish to find the corresponding translations and rotations of the images.
-%
-Because we are permitted to adjust our rotations by a global rotations, we first multiply all rotations by $R_1^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
+Because we are permitted to adjust our rotations by a global rotations, we first multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
 %
@@
 %
-We round the optimal translations to the nearest integer.
+We round the optimal translations to the nearest integer to remove the need for interpolation when translating the images.
 

### 2014-03-03 image_analysis_paper/make_sphere.m

@@ -4,6 +4,10 @@ close all
 res = '-r300';
 fmt = '-djpeg';
 
+%set(0,'DefaultLineMarkerSize',8)
+%set(0,'DefaultLineLineWidth',2)
+%set(0,'DefaultAxesFontSize',20)
+
 im_save_dir = 'paper_figures';
 
 n = 100;
@@ -20,10 +24,13 @@ dir = [0 0 0;
     0 1 0;
     0 0 1;
     1 0 0];
+theta = 0:0.1:pi;
+r = 0.2;
+r1 = 1.25;
 
 for i=1:4
     figure;
-    set(gcf, 'papersize', [1 1])
+    set(gcf, 'papersize', [0.1 0.1])
     h = surf(X, Y, Z, C, 'edgecolor','none');
     set(gca,'position',[0 0 1 1],'units','normalized')
     camlight(45, -20);
@@ -32,16 +39,34 @@ for i=1:4
     
     axis off
     hold on
-    plot3(v1, v0, v0, 'k')
-    plot3(v0, v1, v0, 'k')
-    plot3(v0, v0, v1, 'k')
+    plot3(v1, v0, v0, 'k', 'linewidth', 2)
+    plot3(v0, v1, v0, 'k', 'linewidth', 2)
+    plot3(v0, v0, v1, 'k', 'linewidth', 2)
     
     if i > 1
         rotate(h,dir(i, :),25)
     end
+    if i == 2
+        plot3(r*cos(theta), -r1*ones(size(theta)),r*sin(theta), 'linewidth', 2)
+        plot3(r*cos(theta(end)), -r1,r*sin(theta(end)), '>', 'linewidth', 2, 'markerfacecolor','b')
+        text(-r-0.25, -r1, r+0.25,  '\alpha', 'color','b', 'fontsize', 24)
+        
+    end
+    
+    if i == 3
+        plot3(r*cos(theta), r*sin(theta), -r1*ones(size(theta)), 'linewidth', 2)
+        plot3(r*cos(theta(end)), r*sin(theta(end)), -r1, '<', 'linewidth', 2, 'markerfacecolor','b')
+        text(0, r+0.05, -r1-0.5, '\beta', 'color','b', 'fontsize', 24)
+    end
+    if i == 4
+        plot3(-r1*ones(size(theta)), r*sin(theta), r*cos(theta), 'linewidth', 2)
+        plot3(-r1, r*sin(theta(end)), r*cos(theta(end)), '<', 'linewidth', 2, 'markerfacecolor','b')
[…]

### 2014-03-15 image_analysis_paper/make_figures_large_dataset_color.m
made paper figures have no margins

@@ -33,6 +33,7 @@ npixels = 100;
 %set image plotting parameters
 subplot_dim1 = ceil(sqrt(m));
 subplot_dim2 = ceil(m / subplot_dim1);
+[Y, X] = meshgrid((subplot_dim1:-1:1)/subplot_dim1, (1:subplot_dim2)/subplot_dim2);
 
 nchannels = 3;
 image_set = zeros(npixels, npixels, nchannels, m, 'uint8');
@@ -77,13 +78,10 @@ figure;
 set(gcf, 'paperposition',[0 0 8 8])
 for i=1:m
     im1 = image_set(:,:,:,i);
-    subplot(subplot_dim1, subplot_dim2, i);
+    %subplot(subplot_dim1, subplot_dim2, i);
+    subplot('position', [X(i)-1/subplot_dim1 Y(i)-1/subplot_dim2 1/subplot_dim1 1/subplot_dim2])
     imshow(im1);
     
-    pos = get(gca, 'position');
-    pos(3) = 0.0884;
-    pos(4) = 0.0884;
-    set(gca, 'position', pos)
 end
 print(sprintf('%s/unregistered_unordered_2d', im_save_dir), fmt, res);
 
@@ -97,13 +95,11 @@ figure;
 set(gcf, 'paperposition',[0 0 8 8])
 for i=1:m
     im1 = image_set_aligned(:,:,:,i);
-    subplot(subplot_dim1, subplot_dim2, i);
+    
+    %subplot(subplot_dim1, subplot_dim2, i);
+    subplot('position', [X(i)-1/subplot_dim1 Y(i)-1/subplot_dim2 1/subplot_dim1 1/subplot_dim2])
     imshow(im1);
     
-    pos = get(gca, 'position');
-    pos(3) = 0.0884;
-    pos(4) = 0.0884;
-    set(gca, 'position', pos)
 end
 print(sprintf('%s/registered_unordered_2d', im_save_dir), fmt, res);
 
@@ -133,13 +129,10 @@ end
 figure;
 set(gcf, 'paperposition', [0 0 8 8])
 for i=1:m
-    subplot(subplot_dim1, subplot_dim2, i);
-    imshow(image_set_aligned(:,:,:,I(i)));
+    %subplot(subplot_dim1, subplot_dim2, i);
+    subplot('position', [X(i)-1/subplot_dim1 Y(i)-1/subplot_dim2 1/subplot_dim1 1/subplot_dim2])
[…]

### 2014-03-24 image_analysis_paper/pnas_draft.tex
draft of pnas paper

--- 2014-03-03 image_analysis_paper/math_description.tex
@@
-\documentclass[11pt]{article}
+%% PNAStmpl.tex
+%% Template file to use for PNAS articles prepared in LaTeX
+%% Version: Apr 14, 2008
 
-\usepackage{amsmath}
-\usepackage{amssymb}
-\DeclareMathOperator*{\argmin}{arg\,min}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% BASIC CLASS FILE 
+%% PNAStwo for two column articles is called by default.
+%% Uncomment PNASone for single column articles. One column class
+%% and style files are available upon request from pnas@nas.edu.
+%% (uncomment means get rid of the '%' in front of the command)
 
-\usepackage{graphicx}
+%\documentclass{pnasone}
+\documentclass{pnastwo}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Changing position of text on physical page:
+%% Since not all printers position
+%% the printed page in the same place on the physical page,
+%% you can change the position yourself here, if you need to:
+
+% \advance\voffset -.5in % Minus dimension will raise the printed page on the 
+                         %  physical page; positive dimension will lower it.
+
+%% You may set the dimension to the size that you need.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% OPTIONAL GRAPHICS STYLE FILE
+
+%% Requires graphics style file (graphicx.sty), used for inserting
+%% .eps files into LaTeX articles.
+%% Note that inclusion of .eps files is for your reference only;
+%% when submitting to PNAS please submit figures separately.
+
+%% Type into the square brackets the name of the driver program 
+%% that you are using. If you don't know, try dvips, which is the
+%% most common PC driver, or textures for the Mac. These are the options:
+
+% [dvips], [xdvi], [dvipdf], [dvipdfm], [dvipdfmx], [pdftex], [dvipsone],
+% [dviwindo], [emtex], [dviwin], [pctexps], [pctexwin], [pctexhp], [pctex32],
+% [truetex], [tcidvi], [vtex], [oztex], [textures], [xetex]
+
+%\usepackage[dvips]{graphicx}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% OPTIONAL POSTSCRIPT FONT FILES
+
+%% PostScript font files: You may need to edit the PNASoneF.sty
+%% or PNAStwoF.sty file to make the font names match those on your system. 
+%% Alternatively, you can leave the font style file commands commented out
+%% and typeset your article using the default Computer Modern 
+%% fonts (recommended). If accepted, your article will be typeset
+%% at PNAS using PostScript fonts.
+
+% Choose PNASoneF for one column; PNAStwoF for two column:
+%\usepackage{PNASoneF}
+%\usepackage{PNAStwoF}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% ADDITIONAL OPTIONAL STYLE FILES
+
+%% The AMS math files are commonly used to gain access to useful features
+%% like extended math fonts and math commands.
+
+\usepackage{amssymb,amsfonts,amsmath}
+
 \usepackage{subcaption}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% OPTIONAL MACRO FILES
+%% Insert self-defined macros here.
+%% \newcommand definitions are recommended; \def definitions are supported
+
+%\newcommand{\mfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2}}
+%\def\s{\sigma}
+
+\DeclareMathOperator*{\argmin}{arg\,min}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Don't type in anything in the following section:
+%%%%%%%%%%%%
+%% For PNAS Only:
+\contributor{Submitted to Proceedings
+of the National Academy of Sciences of the United States of America}
+\url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
+\copyrightyear{2008}
+\issuedate{Issue Date}
+\volume{Volume}
+\issuenumber{Issue Number}
+%%%%%%%%%%%%
+
 \begin{document}
 
-\begin{figure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{fig2a}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{fig2b}
-\caption{}
-\end{subfigure}
-\caption{} 
-\label{fig:fluorescent_images}
-\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Techniques}
+%% For titles, only capitalize the first letter
+%% \title{Almost sharp fronts for the surface quasi-geostrophic equation}
 
-\subsection{Registration Using Angular Synchronization} \label{subsec:angsynch}
+\title{Temporal ordering and registration of imaging data in cross--sectional studies of development}
+
+%% Enter authors via the \author command.  
+%% Use \affil to define affiliations.
+%% (Leave no spaces between author name and \affil command)
+
+%% Note that the \thanks{} command has been disabled in favor of
+%% a generic, reserved space for PNAS publication footnotes.
+
+%% \author{<author name>
+%% \affil{<number>}{<Institution>}} One number for each institution.
+%% The same number should be used for authors that
+%% are affiliated with the same institution, after the first time
+%% only the number is needed, ie, \affil{number}{text}, \affil{number}{}
+%% Then, before last author ...
+%% \and
+%% \author{<author name>
+%% \affil{<number>}{}}
+
+%% For example, assuming Garcia and Sonnery are both affiliated with
+%% Universidad de Murcia:
+%% \author{Roberta Graff\affil{1}{University of Cambridge, Cambridge,
+%% United Kingdom},
+%% Javier de Ruiz Garcia\affil{2}{Universidad de Murcia, Bioquimica y Biologia
+%% Molecular, Murcia, Spain}, \and Franklin Sonnery\affil{2}{}}
+
+\author{Carmeline J. Dsilva\affil{1}{Departmnent of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08544}, 
+Bomyi Lim\affil{1}{}, Stanislav Y.
+Shvartsman\affil{1}{} \affil{2}{Lewis--Seigler Instittue for Integrative Genomics, Princeton University, Princeton, NJ 08544}, \and Ioannis G.
+Kevrekidis\affil{1}{} \affil{3}{Program in Applied and Computational Mathematics, Princeton University, Princeton, NJ 08544}}
+
+\contributor{Submitted to Proceedings of the National Academy of Sciences
+of the United States of America}
+
+%% The \maketitle command is necessary to build the title page.
+\maketitle
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{article}
+
+\begin{abstract} 
+In studies of development, researchers are often presented with cross--sectional data, where each data point is a sample from a population fixed at a slightly different developmental time. 
+%
+The goal is then to temporally order the data to reconstruct the developmental dynamics. 
+%
+If each data point is a two-dimensional image, the images must first be aligned or registered before they can be temporally ordered. 
+%
+When such data sets are large, and/or if the developmental changes are subtle, these tasks can be difficult to do by hand. 
+%
+We present an automatic way to register and temporally order cross--sectional data sets of images.
+We use relatively new mathematical techniques (angular synchronization for image registration, and diffusion maps for temporal ordering) which are applicable to a wide variety of data sets and require little {\em a priori} knowledge of the image features or the developmental dynamics.
+%
+We demonstrate the utility of our methods using a collection of images from a study of {\em Drosophila} embryogenesis.
+\end{abstract}
+
+%% When adding keywords, separate each term with a straight line: |
+\keywords{temporal ordering | image registration}
+
+%% Optional for entering abbreviations, separate the abbreviation from
+%% its definition with a comma, separate each pair with a semicolon:
+%% for example:
+%% \abbreviations{SAM, self-assembled monolayer; OTS,
+%% octadecyltrichlorosilane}
+
+% \abbreviations{}
+
+%% The first letter of the article should be drop cap: \dropcap{}
+%\dropcap{I}n this article we study the evolution of ''almost-sharp'' fronts
+
+%% Enter the text of your article beginning here and ending before
+%% \begin{acknowledgements}
+%% Section head commands for your reference:
+%% \section{}
+%% \subsection{}
+%% \subsubsection{}
+
+\begin{itemize}
+\item Want to study development
+\item Often difficult/impossible to obtain longitudinal data
+\item Instead, we are presented with cross--sectional data, and we need to order it to look at dynamics
+\item This can be done by hand for small data sets (Figure \ref{fig:fish})
+\item large data sets-- need automation (Figure \ref{fig:fluorescent_images})
+\item we will present mathematical algorithms to register images and temporally order them
+\item Use {\em Drosophila} as a case study
+\end{itemize}
+
+\section{Results and Discussion}
 
@@
 %
-There are many ways to align images. 
+There are many methods for image registration. 
 %
-``Template--based'' alignment \cite{ahuja2007template} is one technique, where one selects a specific image or {\em template}, and then tries to optimally align each image to that template.
+``Template--based'' registration \cite{ahuja2007template} is a common technique, where one selects a specific image or {\em template}, and then tries to optimally register each image to that template.
 %
-However, this method can result in many misalignemnts for noisy images, and can be problematic if a ``good'' template image is not known {\em a priori}. 
+However, this method can result in many errors for noisy images, and can be problematic if a ``good'' template image is not known {\em a priori}. 
 %
-Instead, we will use a technique called angular synchronization\cite{singer2011angular} to align sets of images.
+Instead, we will use a technique called angular synchronization\cite{singer2011angular} to register sets of images.
 %
@@
 %
-From this large collection of pairwise alignments, it then tries to find optimal alignment for each image, such that these individual alignments are most consistent (in some metric) with the pairwise alignment.
+From this large collection of pairwise alignments, it then tries to find optimal alignment for each image, such that these individual alignments are most consistent (in some metric) with the pairwise alignments.
 %
@@
 %
+We have converted the two--dimensional images shown in Figure \ref{fig:fluorescent_images} into concentration profiles on a ring (we will address the full two--dimensional images in a subsequent section).
+%
+Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one--dimensional concentration profile; the profile is discretized into 100 points. 
+%
+Rotation of this image corresponds to shifting (with periodic boundary conditions) the one--dimensional concentration profile. 
+%
+Figure \ref{subfig:1d_unaligned_unordered} shows the concentration profiles for all the images in our data set. 
+%
+The concentration profiles have been stacked in an array, such that each row corresponds to one image.
+%
+Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles, now aligned using angular synchronization;
+the peaks are aligned, even though we input no information about the existence or location of these peaks into our alogrithm.
+%
+We would like to note that, because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered {\em up to} a global rotation, and we are permitted rotate every point by a constant rotation.
+
+Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we now wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
+%
+We will use diffusion maps (DMAPS) \cite{coifman2005geometric} to temporally order our data.
+%
+DMAPS aims to uncover a parameterization of high-dimensional data sampled from a low-dimensional nonlinear manifold.
+%
+In our analysis, we will assume that our data lie on a one--dimensional curve/ manifold, and that this one dimension is correlated with time. 
+%
+Therefore, sorting the data by the value of DMAPS coordinate will order the data along this one--dimensional curve, and will therefore temporally order the data.
+
+We demonstrate how we can apply these techniques to our (aligned) one--dimensional concentration profiles. 
+%
+We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances for DMAPS (in \eqref{eq:dmaps_W}).
+%
+Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non--trivial) DMAPS component, $\phi_2$. 
+%
+As expected, the intensity of the dpERK peaks grows as a function of time. 
+%
+Although DMAPS can recover a good ordering for the data, it cannot tell us about the absolute time of the data; we therefore have no information about the speed of this trajectory.
+%
+Furthermore, the direction of the trajectory is also unknown; determining which ``end'' of the DMAPS trajectory corresponds to the young embryos, and which ``end'' corresponds to the old embryos is a task that needs to be done by hand.
+
+In many cases, we are interested in both registering and ordering our data.
+%
+%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
+%
+We can combine angular synchronization (for registration) and diffusion maps (for temporal ordering) into one eigencomputation that allows us to {\em simultaneously} recover the optimal alignments and the temporal ordering.
+%
+This technique is called vector diffusion maps (VDM) \cite{singer2012vector}.
+%
+It not only reduces the number of required eigencomputations, but also is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
+%
+Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of aligning and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps. 
+%
+The data has been ordered by $\psi_{1, 3}$, the first (non--trivial) VDM coordinate.
+
+The same methodologies (with small modifications) can be applied to two--dimensional images.
+%
+Instead of having $100$--dimensional concentration vectors, we now work with images containing $10000$ pixels.
+%
+The two--dimensional images are registered with respect to all possible translations and rotations. 
+%
+We used vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_scrambled}.
+%
+The results are shown in Figure~\ref{subfig:images_vdm}.
+%
+The peaks are now aligned, and the intensity of dpERK grows as a function of time, as expected.
+
+In these {\em Drosophila} experiments, we have the luxury of an independent time marker from membranes.
+%
+We can therefore evaluate the quality of our orderings obtained from VDM by comparing with the orderings obtained from the membrane markers.
+%
+The rank of each data point, computed using VDM, versus the rank of each data point computed from the membrane thickness, is shown in Figure~\ref{subfig:rank_corr}.
+%
+The rank correlation is 0.8665, indicating that our ordering from VDM is consistent with the ordering obtained from the membrane thickness.
+%
+There is more scatter for the younger embryos; this is to be expected, as the dpERK signal is very low in this range.
+
+Although our methods are fairly robust, they do require a sufficient amount of data to ``see'' a smooth developmental trajectory.
+%
+We evaluated how much data is required by subsampling our data set (with replacement), and for each subsample, computing the temporal ordering using VDM and calculating the rank correlation between this ordering and the ordering using the membrane markers.
+%
+The results are shown in Figure~\ref{subfig:bootstrap}.
+%
+For small data sets, the average rank correlation is very low, indicating that the orderings from VDM are inaccurate. 
+%
+However, for moderately sized data sets (approximately 30 data points), the 
+rank correlation begins to plateau and VDM can accurately order the data.
+
+\section{Conclusions}
+
+\begin{itemize}
+
+\item Techniques for registration and ordering
+\item Very general-- requires little knowledge of the types of images
+\item Ordering-- different than other studies (TSP, MST, etc.)
+\item assumes one--dimensionality of data, and one dimension is correlated with time
+\item gives no sense of absolute time (speed) or space (where the ``zero point'' is)
+\end{itemize}
+
+%% == end of paper:
+
+%% Optional Materials and Methods Section
+%% The Materials and Methods section header will be added automatically.
+
+%% Enter any subheads and the Materials and Methods text below.
+
+\begin{materials}
+
+\section{Angular synchronization}
+
 Let $x_1, \dots, x_m \in \mathbb{R}^n$ denote the signals that we wish to align with respect to rotations.
@@
 \end{equation}
+%
+For the one--dimensional profiles shown in Figure~\ref{fig:1d_demo}, to compute the optimal angles in \eqref{eq:opt_angle}, we exhaustively search over all 100 possible shifts/rotations of the signals.
 
@@
 
-To illustrate how angular synchronization works, we have converted the two--dimensional images shown in Figure \ref{fig:fluorescent_images} into concentration profiles on a ring (we will address the full two--dimensional images in a subsequent section).
+\section{Diffusion Maps}
+
+We will uncover a parameterization of this curve by only considering {\em local} neighbors of each data point.
 %
-Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one--dimensional concentration profile. 
-%
-The images are not aligned, and so rotation of this image corresponds to shifting (with periodic boundary conditions) the one--dimensional concentration profile. 
-%
-Figure \ref{subfig:1d_unaligned_unordered} shows the concentration profiles for all the images in our data set. 
-%
-The concentration profiles have been stacked in an array, such that each row corresponds to one image.
-%
-Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles, now aligned using angular synchronization. 
-%
-The peaks are now (mostly) aligned.
-
-\begin{figure}
-\begin{subfigure}{0.17\textwidth}
-\includegraphics[width=\textwidth]{illustrate_1d}
-\caption{}
-\label{subfig:1d_image_example}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{unregistered_unordered_1d}
-\caption{}
-\label{subfig:1d_unaligned_unordered}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{registered_unordered_1d}
-\caption{}
-\label{subfig:1d_aligned_unordered}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{registered_ordered_1d}
-\caption{}
-\label{subfig:1d_aligned_ordered}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{registered_ordered_vdm_1d}
-\caption{}
-\label{subfig:1d_aligned_ordered_vdm}
-\end{subfigure}
-\caption{}
-\label{fig:1d_demo}
-\end{figure}
-
-\subsection{Ordering Using Diffusion Maps}
-
-Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we now wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
-%
-We will use diffusion maps (DMAPS) \cite{coifman2005geometric} to temporally order our data.
-%
-DMAPS aims to uncover a parameterization of high-dimensional data sampled from a low-dimensional nonlinear manifold.
-%
-In our analysis, we will assume that our data lie on a one--dimensional curve or manifold, and that this one dimension is correlated with time. 
-%
-We will uncover a parameterization of this curve by only considering {\em local} neighbors of each data point.
-
 To illustrate how local neighbors can allow us to uncover a global parameterization, assume we have data $x_1, x_2, \dots, x_m \in \mathbb{R}$ such that $x_1 < x_2 < \dots < x_m$.
@@
 
-We demonstrate how we can apply these techniques to our (aligned)one--dimensional concentration profiles. 
-%
-We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our distances in \eqref{eq:dmaps_W}.
-%
-Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non--trivial) dmaps component, $\phi_2$.
-
-\subsection{Vector Diffusion Maps}
-
-Many times, we would to both align and order our data.
-%
-Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
-%
-We can combine the two steps into one eigencomputation that allows us to {\em simultaneously} recover the optimal alignments and the temporal ordering.
-%
-This technique is called vector diffusion maps (VDM) \cite{singer2012vector}.
-%
-It not only reduces the number of required eigencomputations, but also is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
+\section{Vector Diffusion Maps}
 
@@
 
-Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of aligning and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps. 
-%
-The data has been ordered by $\psi_{1, 3}$.
-
-\subsection{Registering Images with Respect to Translations and Rotations} \label{subsec:trans_rot_register}
+\section{Registering Images with Respect to Translations and Rotations} \label{subsec:trans_rot_register}
 
@@
 These rotation matrices are (by definition) orthoganol, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
-%
-See the Appendix for more details.
-
-\section{Results}
-
-\begin{figure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{fig4a}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{dpERK_rank_corr}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{fig4c}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{Dl_rank_corr}
-\caption{}
-\end{subfigure}
-\end{figure}
-
-\appendix
-
-\section{Computing Optimal Alignments for One--Dimensional Signals}
-
-Each signal shown in Figure \ref{fig:1d_demo} is discretized into 100 points. 
-%
-To compute the optimal angles in \eqref{eq:opt_angle}, we exhaustively search over all 100 possible shifts/rotations of the signals.
-
-\section{Computing Optimal Pairwise Alignments for Two--Dimensional Images}
 
@@
 Since our techniques are robust to noise, close approximations will be sufficient to obtain accurate results.
-
-\section{Converting from $ISO(2)$ to $SO(3)$}
-
-\begin{figure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{sphere_1}
-\includegraphics[width=\textwidth]{sphere2_1}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{sphere_2}
-\includegraphics[width=\textwidth]{sphere2_2}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{sphere_3}
-\includegraphics[width=\textwidth]{sphere2_3}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{sphere_4}
-\includegraphics[width=\textwidth]{sphere2_4}
-\caption{}
-\end{subfigure}
-\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions.
-(a) The original image.
-(b) Rotation around the x--axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image.
-(c) Rotation around the y--axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation.
-(d) Rotation around the z--axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.
-The top row shows the three--dimensional spheres, and the bottom row shows the (two--dimensional) surface of the sphere in which we are interested.}
-\label{fig:SO3_picture}
-\end{figure}
 
@@
 
-\section{Computing Optimal Translations and Rotations from the Eigenvectors of $H$}
-
 From the three--dimensional rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$,
@@
 We round the optimal translations to the nearest integer to remove the need for interpolation when translating the images.  
+\end{materials}
 
-\bibliographystyle{plain}
+%% Optional Appendix or Appendices
+%% \appendix Appendix text...
+%% or, for appendix with title, use square brackets:
+%% \appendix[Appendix Title]
+
+\begin{acknowledgments}
+-- text of acknowledgments here, including grant info --
+\end{acknowledgments}
+
+%% PNAS does not support submission of supporting .tex files such as BibTeX.
+%% Instead all references must be included in the article .tex document. 
+%% If you currently use BibTeX, your bibliography is formed because the 
+%% command \verb+\bibliography{}+ brings the <filename>.bbl file into your
+%% .tex document. To conform to PNAS requirements, copy the reference listings
+%% from your .bbl file and add them to the article .tex file, using the
+%% bibliography environment described above.
+
+%%  Contact pnas@nas.edu if you need assistance with your
+%%  bibliography.
+
+% Sample bibliography item in PNAS format:
+%% \bibitem{in-text reference} comma-separated author names up to 5,
+%% for more than 5 authors use first author last name et al. (year published)
+%% article title  {\it Journal Name} volume #: start page-end page.
+%% ie,
+% \bibitem{Neuhaus} Neuhaus J-M, Sitcher L, Meins F, Jr, Boller T (1991) 
+% A short C-terminal sequence is necessary and sufficient for the
+% targeting of chitinases to the plant vacuole. 
+% {\it Proc Natl Acad Sci USA} 88:10362-10366.
+
+%% Enter the largest bibliography number in the facing curly brackets
+%% following \begin{thebibliography}
+
+\bibliographystyle{pnas}
 \bibliography{../../references/references}
 
+%\begin{thebibliography}{10}
+
+%\bibitem{ahuja2007template}
+%Sunil Ahuja, Ioannis~G Kevrekidis, and Clarence~W Rowley.
+%\newblock Template-based stabilization of relative equilibria in systems with
+%  continuous symmetry.
+%\newblock {\em Journal of Nonlinear Science}, 17(2):109--143, 2007.
+
+%\bibitem{coifman2005geometric}
+%Ronald~R Coifman, Stephane Lafon, Ann~B Lee, Mauro Maggioni, Boaz Nadler,
+%  Frederick Warner, and Steven~W Zucker.
+%\newblock Geometric diffusions as a tool for harmonic analysis and structure
+%  definition of data: Diffusion maps.
+%\newblock {\em Proceedings of the National Academy of Sciences},
+%  102(21):7426--7431, 2005.
+%
+%\bibitem{coifman2008graph}
+%Ronald~R Coifman, Yoel Shkolnisky, Fred~J Sigworth, and Amit Singer.
+%\newblock Graph laplacian tomography from unknown random projections.
+%\newblock {\em IEEE Transactions on Image Processing}, 17(10):1891--1899, 2008.
+%
+%\bibitem{singer2011angular}
+%A~Singer.
+%\newblock Angular synchronization by eigenvectors and semidefinite programming.
+%\newblock {\em Applied and computational harmonic analysis}, 30(1):20--36,
+%  2011.
+%
+%\bibitem{singer2012vector}
+%Amit Singer and H-T Wu.
+%\newblock Vector diffusion maps and the connection laplacian.
+%\newblock {\em Communications on Pure and Applied Mathematics},
+%  65(8):1067--1144, 2012.
+
+%\end{thebibliography}
+
+\end{article}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%% Adding Figure and Table References
+%% Be sure to add figures and tables after \end{article}
+%% and before \end{document}
+
+%% For figures, put the caption below the illustration.
+%%
+%% \begin{figure}
+%% \caption{Almost Sharp Front}\label{afoto}
+%% \end{figure}
+
+%% For Tables, put caption above table
+%%
+%% Table caption should start with a capital letter, continue with lower case
+%% and not have a period at the end
+%% Using @{\vrule height ?? depth ?? width0pt} in the tabular preamble will
+%% keep that much space between every line in the table.
+
+%% \begin{table}
+%% \caption{Repeat length of longer allele by age of onset class}
+%% \begin{tabular}{@{\vrule height 10.5pt depth4pt  width0pt}lrcccc}
+%% table text
+%% \end{tabular}
+%% \end{table}
+
+%% For two column figures and tables, use the following:
+
+%% \begin{figure*}
+%% \caption{Almost Sharp Front}\label{afoto}
+%% \end{figure*}
+
+%% \begin{table*}
+%% \caption{Repeat length of longer allele by age of onset class}
+%% \begin{tabular}{ccc}
+%% table text
+%% \end{tabular}
+%% \end{table*}
+
+\begin{figure}
+\begin{subfigure}{0.25\textwidth}
+\includegraphics[width=\textwidth]{fish1}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.25\textwidth}
+\includegraphics[width=\textwidth]{fish2}
+\caption{}
+\end{subfigure}
+\caption{Schematic illustration of reconstructing dynamics from cross--sectional data.
+(a) Fish, each in a different orientation and a different stage of development.
+(b) Fish, now registered and temporally ordered.
+This is easy to do by hand because the data set is small and the developmental changes are easy to recognize.}
+\label{fig:fish}
+\end{figure}
+
+\begin{figure}
+\begin{subfigure}{0.25\textwidth}
+\includegraphics[width=\textwidth]{fig2a}
+\caption{}
+\label{subfig:images_scrambled}
+\end{subfigure}
+\begin{subfigure}{0.25\textwidth}
+\includegraphics[width=\textwidth]{fig2b}
+\caption{}
+\end{subfigure}
+\caption{{\em Drosophila} data.
+(a) Dorsal--ventral images of {\em Drosphila} embryos in nuclear cycle 14, each stained for Dorsal (green) and dpERK (red).
+(b) Images from (a), now registered using the location of the Dorsal peak, and temporally ordered using the membrane marker.}
+\label{fig:fluorescent_images}
+\end{figure}
+
+\begin{figure*}
+\begin{subfigure}{0.17\textwidth}
+\includegraphics[width=\textwidth]{illustrate_1d}
+\caption{}
+\label{subfig:1d_image_example}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{unregistered_unordered_1d}
+\caption{}
+\label{subfig:1d_unaligned_unordered}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{registered_unordered_1d}
+\caption{}
+\label{subfig:1d_aligned_unordered}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{registered_ordered_1d}
+\caption{}
+\label{subfig:1d_aligned_ordered}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{registered_ordered_vdm_1d}
+\caption{}
+\label{subfig:1d_aligned_ordered_vdm}
+\end{subfigure}
+\caption{(a) One--dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
+(b) The concentration profiles corresponding to the images in Figure \ref{fig:fluorescent_images}.
+Each row in the array corresponds to one image.
+(c) The concentration profiles in (b), now registered using angular synchronization.
+(d) The concentration profiles in (c), now temporally ordered using diffusion maps.
+(e) The concentration profiles in (b), now simultaneously registered and temporally ordered using vector diffusion maps.
+}
+\label{fig:1d_demo}
+\end{figure*}
+
+\begin{figure*}
+\begin{minipage}{0.6\textwidth}
+\begin{subfigure}{\textwidth}
+\includegraphics[width=\textwidth]{registered_ordered_vdm_2d}
+\caption{}
+\label{subfig:images_vdm}
+\end{subfigure}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}
+\begin{subfigure}{\textwidth}
+\includegraphics[width=\textwidth]{rank_corr_vdm}
+\caption{}
+\label{subfig:rank_corr}
+\end{subfigure}
+\begin{subfigure}{\textwidth}
+\includegraphics[width=\textwidth]{bootstrap_rank_corr_color}
+\caption{}
+\label{subfig:bootstrap}
+\end{subfigure}
+\end{minipage}
+\caption{(a) The images in Figure~\ref{fig:fluorescent_images}, aligned and temporally ordered using vector diffusion maps.
+(b) The rank correlation between the ordering of the images in (a) using vector diffusion maps, and the ordering obtained using the membrane markers.
+(c) The rank correlation coefficient as a function of the number of data points.
+}
+\end{figure*}
+
+%\begin{figure*}
+%\begin{subfigure}{0.2\textwidth}
+%\includegraphics[width=\textwidth]{sphere_1}
+%\includegraphics[width=\textwidth]{sphere2_1}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{0.2\textwidth}
+%\includegraphics[width=\textwidth]{sphere_2}
+%\includegraphics[width=\textwidth]{sphere2_2}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{0.2\textwidth}
+%\includegraphics[width=\textwidth]{sphere_3}
+%\includegraphics[width=\textwidth]{sphere2_3}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{0.2\textwidth}
+%\includegraphics[width=\textwidth]{sphere_4}
+%\includegraphics[width=\textwidth]{sphere2_4}
+%\caption{}
+%\end{subfigure}
+%\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions.
+(a) The original image.
+(b) Rotation around the x--axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image.
+(c) Rotation around the y--axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation.
+(d) Rotation around the z--axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.
+The top row shows the three--dimensional spheres, and the bottom row shows the (two--dimensional) surface of the sphere in which we are interested.}
+%\label{fig:SO3_picture}
+%\end{figure*}
+
 \end{document}

### 2014-03-24 image_analysis_paper/make_figures_large_dataset_color.m

@@ -177,6 +177,8 @@ ranks_from_membranes = compute_ranks(mem_lengths);
 ranks_from_vdm = compute_ranks(embed_coord(:,idx));
 
 figure;
+set(gcf, 'papersize', [11/4 8.5/4])
+set(gcf, 'paperposition', [0 0 11/4 8.5/4])
 plot(ranks_from_membranes, ranks_from_vdm, '.')
 xlabel('rank from membrane lengths')
 ylabel('rank from vdm')

### 2014-03-25 image_analysis_paper/pnas_draft.tex
updated paper

@@
 
-\begin{itemize}
-\item Want to study development
-\item Often difficult/impossible to obtain longitudinal data
-\item Instead, we are presented with cross--sectional data, and we need to order it to look at dynamics
-\item This can be done by hand for small data sets (Figure \ref{fig:fish})
-\item large data sets-- need automation (Figure \ref{fig:fluorescent_images})
-\item we will present mathematical algorithms to register images and temporally order them
-\item Use {\em Drosophila} as a case study
-\end{itemize}
+\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional.  
+%
+In longitudinal studies, a chemical or morphological process is monitored over time within the same embryo.  
+%
+In a cross-sectional study, one embryo provides just a single snapshot of a population-averaged developmental trajectory which must be reconstructed from multiple snapshots provided by different embryos.
+%
+Both experimental designs have their advantages and limitations and are extensively used in studies of embryonic development across species. 
+%
+Our work focuses on cross-sectional studies, which have a time-honored history and still present the only option for most of the organisms.  
+%
+Most commonly, a group of developing embryos is fixed (arresting their development), stained with chemicals that visualize a handful of cellular processes, and imaged by fluorescence microscopy. 
+%
+Recent advances in large scale physical manipulation of embryos produce rapidly increasing volumes of imaging data from such experiments, in which every embryo is observed at a different orientation and at a different point in its developmental trajectory.
+%
+In order to mine such datasets for underlying developmental dynamics, multiple images must be registered (aligned) and ordered in time. 
+%
+Here present an algorithmic framework that greatly accelerates both of these tasks.
+
+Temporal ordering and registration of images is straightforward, when the number of images is small and differences between them are apparent. 
+%
+As an example, Figure~\ref{fig:fish} shows a schematic of fish development, combining the processes of growth and patterning.  
+%
+In this case, temporal ordering can be accomplished by arranging the fish by their size, which correlates with the progress of development, and image alignment is based on obvious morphological landmarks, such positions of head and fins.
+%
+Dealing with real data is far more nontrivial. 
+%
+To illustrate most of the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of pattern formation in the early {\em Drosophila} embryo, one of the leading systems for studies of developmental dynamics.
+%
+This dataset comprises 81 fluorescence microscopy images of optical cross-sections of different embryos fixed during the 3$^{rd}$ hour of their development.  
+%
+During this time interval, most of cell nuclei in the embryo are arranged under the common plasma membrane and embryo undergoes the process of cellularization, whereby membranes that enclose individual nuclei into different cells grow in length (Figure~\ref{subfig:images_scrambled}).
+
+Each of the images reveals a uniform arrangement of nuclei and nonuniform distribution of two proteins (green and red) involved in patterning of the dorsoventral (back-to-belly) patterning of the embryo.
+%
+Each image comes from a different embryo in captured at a different orientation and at a different and unknown point in time (Figure~\ref{subfig:images_scrambled}). 
+%
+The task is to align these images in a globally consistent way and order them in a way that is most representative of the underlying developmental trajectory. 
+%
+For this particular dataset, both of these tasks are simplified by prior knowledge. 
+%
+For instance, it is known that the peak in the distribution of one of the proteins (visualized by the green channel) specifies the ventralmost position in the embryo. 
+%
+Based on this, images translated to make the centers of the embryos coincide and rotated to place the peak of the green channel at the same angular position. 
+%
+At the same time, temporal ordering can be done based on the monotonic progress of cellularization. 
+%
+Figure~\ref{subfig:images_hand_ordered} shows the collection of registered and temporally ordered images.
+
+Note that both of these tasks have been accomplished based on prior knowledge about the underlying developmental processes: We knew that the peak of the green signal can be used to align the images and that the progress of cellularization can order them in time.
+%
+Is this information essential? 
+%
+Below we show that this is not the case and that both the temporal ordering and registration of images can be fully automated, without prior information about the underlying dynamics. 
+%
+Our algorithmic approach is based on diffusion maps, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets. 
+%
+The paper is organized as follows. 
+%
+In the next two sections, we use a simpler version of the data (one-dimensional spatial profiles) to formulate both tasks mathematically.  
+%
+Then, we show that both of the tasks can be effectively accomplished in single step. 
+%
+Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align the images and order them correctly in time. 
+%
+We conclude by discussion of the general requirements that must be satisfied for our approach to work and compare it to other ideas in cross-sectional studies of biological dynamics.
 
 \section{Results and Discussion}
+
+\subsection{Angular synchronization for image registration}
 
@@
 %
+This requires translating and rotating each image, so that the set of images are globally aligned.
+
 There are many methods for image registration. 
 %
-``Template--based'' registration \cite{ahuja2007template} is a common technique, where one selects a specific image or {\em template}, and then tries to optimally register each image to that template.
+``Template-based'' registration \cite{ahuja2007template} is a common technique, where one selects a specific image or {\em template}, and then translates and rotates each image in the data set until it is optimally aligned with this template (optimal for us means minimizing the norm between the difference in pixel intensities).
 %
-However, this method can result in many errors for noisy images, and can be problematic if a ``good'' template image is not known {\em a priori}. 
+However, this method can result in many errors if the images are noisy, and can be problematic if a ``good'' template image is not known {\em a priori}. 
 %
-Instead, we will use a technique called angular synchronization\cite{singer2011angular} to register sets of images.
+Instead, we will use a relatively new technique called angular synchronization\cite{singer2011angular} to register sets of images.
 %
-Angular synchronization first computes the pairwise alignments for every possible pair of images, effectively using each image as a template for every other image.
+Angular synchronization first computes the pairwise alignments (translation + rotation) for every possible pair of images, effectively using each image as a template for every other image.
 %
-From this large collection of pairwise alignments, it then tries to find optimal alignment for each image, such that these individual alignments are most consistent (in some metric) with the pairwise alignments.
+From this large collection of pairwise alignments, it then tries to find the optimal translation and rotation for each image, such that these individual alignments are most consistent (in some metric) with the pairwise alignments.
 %
-Because we consider {\em all} pairwise alignments, angular synchronization is often much more robust to noise.
+The algorithm solves this optimization problem by relaxing it to an eigenproblem; the top eigenvector(s) of a matrix (which contains information about the pairwise alignments) give the translation and rotation for each image.
+%
+Because the method considers {\em all} pairwise alignments, angular synchronization is often much more robust to noise.
 
-We will first discuss angular synchronization {\em only} considering rotations. 
-%
-We have converted the two--dimensional images shown in Figure \ref{fig:fluorescent_images} into concentration profiles on a ring (we will address the full two--dimensional images in a subsequent section).
-%
-Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one--dimensional concentration profile; the profile is discretized into 100 points. 
-%
-Rotation of this image corresponds to shifting (with periodic boundary conditions) the one--dimensional concentration profile. 
-%
-Figure \ref{subfig:1d_unaligned_unordered} shows the concentration profiles for all the images in our data set. 
-%
-The concentration profiles have been stacked in an array, such that each row corresponds to one image.
-%
-Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles, now aligned using angular synchronization;
-the peaks are aligned, even though we input no information about the existence or location of these peaks into our alogrithm.
-%
-We would like to note that, because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered {\em up to} a global rotation, and we are permitted rotate every point by a constant rotation.
+\subsection{Diffusion maps for temporal ordering}
 
-Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we now wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
-%
 We will use diffusion maps (DMAPS) \cite{coifman2005geometric} to temporally order our data.
 %
-DMAPS aims to uncover a parameterization of high-dimensional data sampled from a low-dimensional nonlinear manifold.
+We assume that our data are high-dimensional (for example, we will work with images of 10000 pixels), but that they lie on a one-dimensional, perhaps nonlinear curve.
 %
-In our analysis, we will assume that our data lie on a one--dimensional curve/ manifold, and that this one dimension is correlated with time. 
+We further assume that curve can be parameterized by time.
 %
-Therefore, sorting the data by the value of DMAPS coordinate will order the data along this one--dimensional curve, and will therefore temporally order the data.
+Therefore, our goal is to {\em automatically} uncover a parameterization of this curve, so that we can sort/order our data along it, effectively ordering our data in time.
 
-We demonstrate how we can apply these techniques to our (aligned) one--dimensional concentration profiles. 
+To ``unravel'' this curve, we assume that nearby points (in the high-dimensional space) are also close on the curve.
 %
-We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances for DMAPS (in \eqref{eq:dmaps_W}).
+We first construct the matrix $W$ (see Equation \eqref{eq:dmaps_W}); $W_{ij}$ is large if $i$ and $j$ are ``close''.
 %
-Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non--trivial) DMAPS component, $\phi_2$. 
+We then want to find a coordinate $y$ such that $y(i)$ and $y(j)$ if $W_{ij}$ is large (i.e., $y$ preserves local information, and so points that are close in the original space are also close in the coordinate $y$).
 %
-As expected, the intensity of the dpERK peaks grows as a function of time. 
+We therefore want to solve the following optimization problem
+\begin{equation} \label{eq:dmaps_opt_problem}
+\argmin_{y} \sum_{ij} W_{ij} (y(i) - y(j))
+\end{equation}
 %
-Although DMAPS can recover a good ordering for the data, it cannot tell us about the absolute time of the data; we therefore have no information about the speed of this trajectory.
+The solution \ref{eq:dmaps_opt_problem} is $\phi_2$, the second eigenvector of $A$, where $A = D^{-1} W$, and $D$ is a diagonal matrix with $D_{ii} = \sum_j W_{ij}$ \cite{...}.
 %
-Furthermore, the direction of the trajectory is also unknown; determining which ``end'' of the DMAPS trajectory corresponds to the young embryos, and which ``end'' corresponds to the old embryos is a task that needs to be done by hand.
+Therefore, $\phi_2$ provides a one-dimensional parameterization of the data, and ordering the data points by the value of $\phi_2$ will order them in time.
+
+\subsection{Vector diffusion maps for registration and ordering}
 
@@
 It not only reduces the number of required eigencomputations, but also is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
+
+\subsection{Analysis on one-dimensional concentration profiles}
+
+We will first demonstrate our techniques considering {\em only} considering rotations. 
 %
+We have converted the two-dimensional images shown in Figure \ref{fig:fluorescent_images} into concentration profiles on a ring (we will address the full two-dimensional images in a subsequent section).
+%
+Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one-dimensional concentration profile; the profile is discretized into 100 points. 
+%
+Rotation of this image corresponds to shifting (with periodic boundary conditions) the one--dimensional concentration profile. 
+%
+Figure \ref{subfig:1d_unaligned_unordered} shows the concentration profiles for all the images in our data set. 
+%
+The concentration profiles have been stacked in an array, such that each row corresponds to one image.
+
+Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles, now aligned using angular synchronization;
+the peaks are aligned, even though we input no information about the existence or location of these peaks into our algorithm.
+%
+We would like to note that, because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered {\em up to} a global rotation, and we are permitted rotate every point by a constant rotation.
+
+Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we now wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
+%
+We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances for DMAPS (in \eqref{eq:dmaps_W}).
+%
+Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non-trivial) DMAPS component, $\phi_2$. 
+%
+As expected, the intensity of the dpERK peaks grows as a function of time. 
+%
+Although DMAPS can recover a good ordering for the data, it cannot tell us about the absolute time of the data; we therefore have no information about the speed of this trajectory.
+%
+Furthermore, the direction of the trajectory is also unknown; determining which ``end'' of the DMAPS trajectory corresponds to the young embryos, and which ``end'' corresponds to the old embryos is a task that needs to be done by hand.
+
 Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of aligning and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps. 
 %
-The data has been ordered by $\psi_{1, 3}$, the first (non--trivial) VDM coordinate.
+The data has been ordered by $\psi_{1, 3}$, the first (non-trivial) VDM coordinate.
 
-The same methodologies (with small modifications) can be applied to two--dimensional images.
+\subsection{Analysis of two-dimensional images}
+
+The same methodologies (with small modifications) can be applied to two-dimensional images.
 %
-Instead of having $100$--dimensional concentration vectors, we now work with images containing $10000$ pixels.
+Instead of having $100$-dimensional concentration vectors, we now work with images containing $10000$ pixels.
 %
-The two--dimensional images are registered with respect to all possible translations and rotations. 
+The two-dimensional images are registered with respect to all possible translations and rotations; 
+we convert two-dimensional translations and rotations to three-dimensional rotations so that we can apply angular synchronization or vector diffusion maps. 
 %
@@
 \item Techniques for registration and ordering
-\item Very general-- requires little knowledge of the types of images
-\item Ordering-- different than other studies (TSP, MST, etc.)
-\item assumes one--dimensionality of data, and one dimension is correlated with time
+\item Very general- requires little knowledge of the types of images
+\item Ordering- different than other studies (TSP, MST, etc.)
+\item assumes one-dimensionality of data, and one dimension is correlated with time
 \item gives no sense of absolute time (speed) or space (where the ``zero point'' is)
@@
 %
-For the one--dimensional profiles shown in Figure~\ref{fig:1d_demo}, to compute the optimal angles in \eqref{eq:opt_angle}, we exhaustively search over all 100 possible shifts/rotations of the signals.
+For the one-dimensional profiles shown in Figure~\ref{fig:1d_demo}, to compute the optimal angles in \eqref{eq:opt_angle}, we exhaustively search over all 100 possible shifts/rotations of the signals.
 
@@
 
-Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$; we will write our procedure for general $d$ because we will later consider three--dimensional rotations).
+Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$; we will write our procedure for general $d$ because we will later consider three-dimensional rotations).
 %
@@
 Then
-\begin{equation}
+\begin{equation} \label{eq:R_est}
 R_{i, est} = U_i V_i^T
@@
 Note that $R_{i, est}$ is not guaranteed to be a proper rotation (i.e.
-$det(R_{i, est} = \pm 1)$; because we are permitted to scale the eigenvectors of a matrix, we adjust the sign of $\\phi_d$ so that (most of) the rotations are proper.
+$det(R_{i, est} = \pm 1)$; because we are permitted to scale the eigenvectors of a matrix, we adjust the sign of $\phi_d$ so that (most of) the rotations are proper.
 %
@@
 
-Furthermore, this formulation also considers {\em higher--order} consistency information. 
-%
-For example, given our pairwise estimates $R_{ij}$, we know that relationships of the form
-\begin{equation} \label{eq:triplet_consistency}
-R(\theta_{ik}) R(\theta_{kj}) \approx R(\theta_i) R(\theta_k)^T R(\theta_k) R(\theta_j)^T = R(\theta_i) R(\theta_j)^T
-\end{equation}
-should also hold.
-%
-Note that
-\begin{equation}
-(H^2)_{ij} = \sum_k R(\theta_{ik}) R(\theta_{kj});
-\end{equation}
-therefore, {\em all} infomation of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$.
-%
-Because $H$ and $H^2$ have the same eigenvectors, our problem formulation accounts for not only pairwise alignment information, but also these higher--order considerations.
+%Furthermore, this formulation also considers {\em higher-order} consistency information. 
+%%
+%For example, given our pairwise estimates $R_{ij}$, we know that relationships of the form
+%\begin{equation} \label{eq:triplet_consistency}
+%R(\theta_{ik}) R(\theta_{kj}) \approx R(\theta_i) R(\theta_k)^T R(\theta_k) R(\theta_j)^T = R(\theta_i) R(\theta_j)^T
+%\end{equation}
+%should also hold.
+%%
+%Note that
+%\begin{equation}
+%(H^2)_{ij} = \sum_k R(\theta_{ik}) R(\theta_{kj});
+%\end{equation}
+%therefore, {\em all} infomation of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$.
+%%
+%Because $H$ and $H^2$ have the same eigenvectors, our problem formulation accounts for not only pairwise alignment information, but also these higher-order considerations.
 
@@
 
-We will uncover a parameterization of this curve by only considering {\em local} neighbors of each data point.
+%We will uncover a parameterization of this curve by only considering {\em local} neighbors of each data point.
+%%
+%To illustrate how local neighbors can allow us to uncover a global parameterization, assume we have data $x_1, x_2, \dots, x_m \in \mathbb{R}$ such that $x_1 < x_2 < \dots < x_m$.
+%%
+%We consider $x_i$ ``close to'' $x_{i+1}$ and $x_{i-1}$, and $x_i$ ``far away from'' all other data points. 
+%%
+%We then construct the (symmetric) weight matrix, $W \in \mathbb{R}^{m \times m}$,
+%\begin{equation}
+%W = 
+%\begin{bmatrix}
+%	1 & 1/2 & 0 & 0 & \dots & 0 \\
+%	1/2 & 1 & 1/2 & 0 & \dots & 0 \\
+%	\ddots & \ddots & \ddots & \ddots & \ddots & \ddots \\
+%	0 & 0 & 0 & \dots & 1 & 1/2 \\
+%	0 & 0 & 0 & \dots & 1/2 & 1 
+%\end{bmatrix}.
+%\end{equation}
+%%
+%$W_{ij}$ is large if $x_i$ is close to $x_j$,
+%%
+%If we normalize this matrix such that $\sum_j W_{ij} = 1$, then one can show that the first eigenvector of this normalized matrix is the all-ones vector, $\phi_1 = [1 1 \dots 1]^T$, and the second eigenvector is a cosine mode, $\phi_2 = [cos(\pi/m) \cos(2 \pi/ m) \cdots \cos((m-1) \pi / m) \cos(\pi)]^T$.
+%%
+%Therefore, the second eigenvector, $\phi_2$ is {\em one-to-one} with the parameterization of the data, i.e. $\phi_2(1) > \phi_2(2) > \dots > \phi_2(m)$, just as $x_1 < x_2 < \dots < x_m$. 
+%%
+%Therefore, sorting our data by the value of the entries in $\phi_2$ will allow us to order our data.
+
+For diffusion maps, the first step is to construct a matrix $W \in \mathbb{R}^{m \times m}$ that contains information about which pairs of data points are ``close.''
+%For the above analysis, the key step was having a notion of ``closeness'' between data points so that we can construct the matrix $W$.
 %
-To illustrate how local neighbors can allow us to uncover a global parameterization, assume we have data $x_1, x_2, \dots, x_m \in \mathbb{R}$ such that $x_1 < x_2 < \dots < x_m$.
-%
-We consider $x_i$ ``close to'' $x_{i+1}$ and $x_{i-1}$, and $x_i$ ``far away from'' all other data points. 
-%
-We then construct the (symmetric) weight matrix, $W \in \mathbb{R}^{m \times m}$,
-\begin{equation}
-W = 
-\begin{bmatrix}
-	1 & 1/2 & 0 & 0 & \dots & 0 \\
-	1/2 & 1 & 1/2 & 0 & \dots & 0 \\
-	\ddots & \ddots & \ddots & \ddots & \ddots & \ddots \\
-	0 & 0 & 0 & \dots & 1 & 1/2 \\
-	0 & 0 & 0 & \dots & 1/2 & 1 
-\end{bmatrix}.
-\end{equation}
-%
-$W_{ij}$ is large if $x_i$ is close to $x_j$,
-%
-If we normalize this matrix such that $\sum_j W_{ij} = 1$, then one can show that the first eigenvector of this normalized matrix is the all--ones vector, $\phi_1 = [1 1 \dots 1]^T$, and the second eigenvector is a cosine mode, $\phi_2 = [cos(\pi/m) \cos(2 \pi/ m) \cdots \cos((m-1) \pi / m) \cos(\pi)]^T$.
-%
-Therefore, the second eigenvector, $\phi_2$ is {\em one--to--one} with the parameterization of the data, i.e. $\phi_2(1) > \phi_2(2) > \dots > \phi_2(m)$, just as $x_1 < x_2 < \dots < x_m$. 
-%
-Therefore, sorting our data by the value of the entries in $\phi_2$ will allow us to order our data.
-
-For the above analysis, the key step was having a notion of ``closeness'' between data points so that we can construct the matrix $W$.
-%
-For general data in high--dimensional space, we typically use a Gaussian kernel to construct $W$,
+For general data in high-dimensional space, we typically use a Gaussian kernel to construct $W$,
 \begin{equation} \label{eq:dmaps_W}
@@
 %
-In our setup, we assume that our data are inherently one--dimensional, and that this one dimension is correlated with time.
+In our setup, we assume that our data are inherently one-dimensional, and that this one dimension is correlated with time.
 %
@@
 %
-Again, the top (block) eigenvector of $S$ contains approximations of the optimal rotations for each image, and the rotations can be computed as described in Section \ref{subsec:angsynch}.
+Again, the top (block) eigenvector of $S$ contains approximations of the optimal rotations for each image, and the optimal rotations can be computed as described for angular synchronization.
 %
@@
 %
-If we assume that our data are one--dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, one can show that a parameterization of our data (i.e.
+If we assume that our data are one-dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, one can show that a parameterization of our data (i.e.
 the analog of $\phi_2$ from the diffusion maps case) is given by $\psi_{1,d+1}$.
@@
 
-We are not only interested in registering the one--dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two--dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\em and} translations.
+Registering images with respect to rotations {\em and translations} requires some additional care.
 %
-However, for \eqref{eq:H_low_rank} to be satisfied, we require that the matrices used to represent the symmetry group be orthogonal, i.e. $R(\alpha)^{-1} = R(\alpha)^T$. 
+%We are not only interested in registering the one-dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two-dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\em and} translations.
 %
-The typical matrix representation for the group of two--dimensional translations and rotations (using the matrix representation for the affine group \cite{...}) does not satisfy this property. 
+For \eqref{eq:H_low_rank} to be satisfied, we require that the matrices used to represent the symmetry group be orthogonal, i.e. $R(\alpha)^{-1} = R(\alpha)^T$. 
 %
-However, we can (approximately) represent rotations and translations in two--dimensions using three--dimensional rotation matrices.
+However, the typical matrix representation for the group of two-dimensional translations and rotations (using the matrix representation for the affine group \cite{...}) does not satisfy this property.
 %
-We do this by projecting the image onto a portion of the surface of a (three--dimensional) sphere (see Figure \ref{fig:SO3_picture}).
+Instead, we can (approximately) represent rotations and translations in two-dimensions using three-dimensional rotation matrices.
 %
-Rotations in two--dimensions now correspond to rotations around one principal axis in three--dimensions, and translations in two--dimensions correspond (approximately) to rotations around the other two principal axes in three--dimensions \footnote{Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$.
-However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.}.
+We do this by projecting the image onto a portion of the surface of a (three-dimensional) sphere (see Figure \ref{fig:SO3_picture}).
 %
-These rotation matrices are (by definition) orthoganol, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
+Rotations in two-dimensions now correspond to rotations around one principal axis in three-dimensions, and translations in two-dimensions correspond (approximately) to rotations around the other two principal axes in three-dimensions.
+(Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$.
+However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.)
+%
+These rotation matrices are (by definition) orthogonal, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
 
@@
 
-As discussed in Section \ref{subsec:trans_rot_register}, the underlying symmetry group needs to have a real and orthoganol matrix representation. 
+%As discussed above, for angular synchronization, the underlying symmetry group needs to have a real and orthogonal matrix representation. 
 %
-We therefore convert $ISO(2)$, the group of two--dimensional translations and rotations, to $SO(3)$, the group of three--dimensional rotations. 
+%We therefore convert $ISO(2)$, the group of two-dimensional translations and rotations, to $SO(3)$, the group of three-dimensional rotations. 
 %
-In the space of three--dimensional rotations, the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$ correspond to
+We then convert the two-dimensional translation and rotation to a three dimensional rotation. 
+%
+In the space of three-dimensional rotations, the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$ correspond to
 \begin{equation} \label{eq:angle_relations}
@@
 
-From the three--dimensional rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$,
-we build the matrix $H$ and compute the estimates $R_{i, est}$ as described in Section \ref{subsec:angsynch}.
+From the three-dimensional rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$,
+we build the matrix $H$ in \eqref{eq:H_to_R} and compute the estimates $R_{i, est}$ in \eqref{eq:R_est}.
 %
@@
 %
-We can then invert \eqref{eq:angle_relations} to compute the optimal translation and rotation for each image from the Euler angles.
+We then invert \eqref{eq:angle_relations} to compute the optimal translation and rotation for each image from the Euler angles.
 %
@@
 \begin{acknowledgments}
--- text of acknowledgments here, including grant info --
+- text of acknowledgments here, including grant info -
 \end{acknowledgments}
@@
 %  continuous symmetry.
-%\newblock {\em Journal of Nonlinear Science}, 17(2):109--143, 2007.
+%\newblock {\em Journal of Nonlinear Science}, 17(2):109-143, 2007.
 
@@
 %\newblock {\em Proceedings of the National Academy of Sciences},
-%  102(21):7426--7431, 2005.
+%  102(21):7426-7431, 2005.
 %
@@
 %\newblock Graph laplacian tomography from unknown random projections.
-%\newblock {\em IEEE Transactions on Image Processing}, 17(10):1891--1899, 2008.
+%\newblock {\em IEEE Transactions on Image Processing}, 17(10):1891-1899, 2008.
 %
@@
 %\newblock Angular synchronization by eigenvectors and semidefinite programming.
-%\newblock {\em Applied and computational harmonic analysis}, 30(1):20--36,
+%\newblock {\em Applied and computational harmonic analysis}, 30(1):20-36,
 %  2011.
@@
 %\newblock {\em Communications on Pure and Applied Mathematics},
-%  65(8):1067--1144, 2012.
+%  65(8):1067-1144, 2012.
 
@@
 \end{subfigure}
-\caption{Schematic illustration of reconstructing dynamics from cross--sectional data.
+\caption{Schematic illustration of reconstructing dynamics from cross-sectional data.
 (a) Fish, each in a different orientation and a different stage of development.
@@
 \caption{}
+\label{subfig:images_hand_ordered}
 \end{subfigure}
 \caption{{\em Drosophila} data.
-(a) Dorsal--ventral images of {\em Drosphila} embryos in nuclear cycle 14, each stained for Dorsal (green) and dpERK (red).
+(a) Dorsal-ventral images of {\em Drosphila} embryos in nuclear cycle 14, each stained for Dorsal (green) and dpERK (red).
 (b) Images from (a), now registered using the location of the Dorsal peak, and temporally ordered using the membrane marker.}
@@
 \end{subfigure}
-\caption{(a) One--dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
+\caption{(a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
 (b) The concentration profiles corresponding to the images in Figure \ref{fig:fluorescent_images}.
@@
 (a) The original image.
-(b) Rotation around the x--axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image.
-(c) Rotation around the y--axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation.
-(d) Rotation around the z--axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.
-The top row shows the three--dimensional spheres, and the bottom row shows the (two--dimensional) surface of the sphere in which we are interested.}
+(b) Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image.
+(c) Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation.
+(d) Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.
+The top row shows the three-dimensional spheres, and the bottom row shows the (two-dimensional) surface of the sphere in which we are interested.}
 %\label{fig:SO3_picture}

### 2014-03-26 image_analysis_paper/pnas_draft.tex
updated paper

@@
 
+We will first demonstrate our techniques considering {\em only} considering rotations. 
+%
+We have converted the two-dimensional images shown in Figure \ref{subfig:images_scrambled} into concentration profiles on a ring (we will address the full two-dimensional images in a subsequent section).
+%
+Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one-dimensional concentration profile; the profile is discretized into 100 points. 
+%
+Rotation of this image corresponds to shifting (with periodic boundary conditions) the one--dimensional concentration profile. 
+%
+Figure \ref{subfig:1d_unaligned_unordered} shows the concentration profiles for all the images in our data set. 
+%
+The concentration profiles have been stacked in an array, such that each row corresponds to one image.
+
 \subsection{Angular synchronization for image registration}
@@
 
+Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization;
+the peaks are aligned, even though we input no information about the existence or location of these peaks into our algorithm.
+%
+We would like to note that, because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered {\em up to} a global rotation, and we are permitted rotate every point by a constant rotation.
+
 \subsection{Diffusion maps for temporal ordering}
@@
 %
-We assume that our data are high-dimensional (for example, we will work with images of 10000 pixels), but that they lie on a one-dimensional, perhaps nonlinear curve.
+Although our data are high-dimensional (for example, we will work with 2-color images of 10000 pixels, and so we treat these images as 20000-dimensional vectors, where each dimension describes the intensity of one color in one pixel), we will assume that they lie on a one-dimensional, perhaps nonlinear curve.
 %
@@
 
-To ``unravel'' this curve, we assume that nearby points (in the high-dimensional space) are also close on the curve.
+To ``unravel'' this curve, we assume that data points which are close in the high-dimensional space should also be close on the curve.
 %
-We first construct the matrix $W$ (see Equation \eqref{eq:dmaps_W}); $W_{ij}$ is large if $i$ and $j$ are ``close''.
+We want to find a parameterization of the points, such that the points which are close in the original close are {\em as close as possible} in the parameterization.
 %
-We then want to find a coordinate $y$ such that $y(i)$ and $y(j)$ if $W_{ij}$ is large (i.e., $y$ preserves local information, and so points that are close in the original space are also close in the coordinate $y$).
+This problem can be formulated as an optimization problem \cite{...}, and the solution is given by the top (non-trivial) eigenvector of a matrix which contains information about pairwise distances between the data points.
+
+As an illustration of how local neighbors are important for parameterizing a curve, Figure~\ref{subfig:dmaps_edges} shows a data set which lies on a one-dimensional, nonlinear curve in two dimensions.
 %
-We therefore want to solve the following optimization problem
-\begin{equation} \label{eq:dmaps_opt_problem}
-\argmin_{y} \sum_{ij} W_{ij} (y(i) - y(j))
-\end{equation}
+The edges between the pairs of data points are colored according to the distance between the data points, so that close data points have a dark edge between them.
 %
-The solution \ref{eq:dmaps_opt_problem} is $\phi_2$, the second eigenvector of $A$, where $A = D^{-1} W$, and $D$ is a diagonal matrix with $D_{ii} = \sum_j W_{ij}$ \cite{...}.
+Visually, these dark edges are meaningful and informative of the structure of the data; they clearly delineate the one-dimensional curve in the data that our eye can extract.
 %
-Therefore, $\phi_2$ provides a one-dimensional parameterization of the data, and ordering the data points by the value of $\phi_2$ will order them in time.
+Furthermore, the light edges are not informative and do not give any meaningful structure to the curve. 
+%
+The DMAPS parameterization of these are data are shown in Figure~\ref{subfig:dmaps_color}.
+%
+The parameterization is what we would expect visually.
+%
+Although this data is useful for illustrative purposes, it is
+
+Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we now wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
+%
+We use the total Euclidean distance (both the red and the green channels) between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances for DMAPS (in \eqref{eq:dmaps_W}).
+%
+Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non-trivial) DMAPS component, $\phi_2$. 
+%
+As expected, the intensity of the dpERK peaks grows as a function of time. 
+%
+Although DMAPS can recover a good ordering for the data, it cannot tell us about the absolute time of the data; we therefore have no information about the speed of this trajectory.
+%
+Furthermore, the direction of the trajectory is also unknown; determining which ``end'' of the DMAPS trajectory corresponds to the young embryos, and which ``end'' corresponds to the old embryos is a task that needs to be done by hand.
 
@@
 It not only reduces the number of required eigencomputations, but also is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
-
-\subsection{Analysis on one-dimensional concentration profiles}
-
-We will first demonstrate our techniques considering {\em only} considering rotations. 
-%
-We have converted the two-dimensional images shown in Figure \ref{fig:fluorescent_images} into concentration profiles on a ring (we will address the full two-dimensional images in a subsequent section).
-%
-Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one-dimensional concentration profile; the profile is discretized into 100 points. 
-%
-Rotation of this image corresponds to shifting (with periodic boundary conditions) the one--dimensional concentration profile. 
-%
-Figure \ref{subfig:1d_unaligned_unordered} shows the concentration profiles for all the images in our data set. 
-%
-The concentration profiles have been stacked in an array, such that each row corresponds to one image.
-
-Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles, now aligned using angular synchronization;
-the peaks are aligned, even though we input no information about the existence or location of these peaks into our algorithm.
-%
-We would like to note that, because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered {\em up to} a global rotation, and we are permitted rotate every point by a constant rotation.
-
-Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we now wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
-%
-We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances for DMAPS (in \eqref{eq:dmaps_W}).
-%
-Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non-trivial) DMAPS component, $\phi_2$. 
-%
-As expected, the intensity of the dpERK peaks grows as a function of time. 
-%
-Although DMAPS can recover a good ordering for the data, it cannot tell us about the absolute time of the data; we therefore have no information about the speed of this trajectory.
-%
-Furthermore, the direction of the trajectory is also unknown; determining which ``end'' of the DMAPS trajectory corresponds to the young embryos, and which ``end'' corresponds to the old embryos is a task that needs to be done by hand.
 
@@
 
-\begin{itemize}
+We have presented mathematical techniques for registration and temporal ordering of images arising in studies of developmental biology.
+%
+Although we illustrated our methods using a specific data set arising in a study of {\em Drosophila} embryogenesis, the techniques are very general and require little knowledge of the image features or the developmental dynamics.
 
-\item Techniques for registration and ordering
-\item Very general- requires little knowledge of the types of images
-\item Ordering- different than other studies (TSP, MST, etc.)
-\item assumes one-dimensionality of data, and one dimension is correlated with time
-\item gives no sense of absolute time (speed) or space (where the ``zero point'' is)
-\end{itemize}
+We would like to note that there have been multiple other works with temporally ordering biological data (RNA data, cell morphology data, disease progression data, etc.). 
+%
+The majority of these works temporally order the data either by solving a travelling salesman probelem, or by constructing a minimum spanning tree on the data.
+%
+We, instead, use nonlinear dimensionality reduction techniques, to construct a one-dimensional embedding of the data.
+%
+Furthermore, we work with images that must be first registered before they can be temporally ordered.
+
+Although our methods are applicable to a wide variety of data sets, there are several questions for which our techniques can provide no answers.
+%
+Because our registration algorithm only considers pairwise information, we cannot recover the true ``zero'' point of the images. 
+%
+Similairly, we cannot know which end of our trajectory is the beginning and which is the end.
+%
+Both of these tasks bust be done by hand as a post-processing step.
+%
+Furthermore, our methods can only provide a temporal ordering for the data, and cannot provide any information as to the speed or absolute time of the data samples.
+%
+Regardless, we are confident that our methods hold great promise in areas where cross-sectional imaging data is readily available and relatively inexpensive to obtain. 
+%
+There are many applications for which develpomental dynamics are of interest, and yet there is no time marker to use to reconstruct the dynamics from such cross-sectional data.
 
@@
 
-We then compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{m} W_{ij}$, and the matrix $A$, where
+We then want to find a coordinate $y$ such that $y(i)$ and $y(j)$ if $W_{ij}$ is large (i.e., $y$ preserves local information, and so points that are close in the original space are also close in the coordinate $y$).
+%
+We therefore want to solve the following optimization problem
+\begin{equation} \label{eq:dmaps_opt_problem}
+\argmin_{y} \sum_{ij} W_{ij} (y(i) - y(j))
+\end{equation}
+%
+%The solution \ref{eq:dmaps_opt_problem} is $\phi_2$, the second eigenvector of $A$, where $A = D^{-1} W$, and $D$ is a diagonal matrix with $D_{ii} = \sum_j W_{ij}$ \cite{...}.
+%%
+%Therefore, $\phi_2$ provides a one-dimensional parameterization of the data, and ordering the data points by the value of $\phi_2$ will order them in time.
+
+To solve \eqref{eq:dmaps_opt_problem}, we compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{m} W_{ij}$, and the matrix $A$, where
 \begin{equation} \label{eq:dmaps_A}
@@
 %
-We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_m$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_m$ and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$ \footnote{Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors.}.
+We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_m$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_m$ and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$.
+%
+Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors. 
 %
@@
 %
-In general, the next few eigenvectors $\phi_2, \dots, \phi_m$ give ``meaningful'' embedding coordinates for the data, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point.
+%In general, the next few eigenvectors $\phi_2, \dots, \phi_m$ give ``meaningful'' embedding coordinates for the data, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point.
 %
-In our setup, we assume that our data are inherently one-dimensional, and that this one dimension is correlated with time.
+The next eigenvector, $\phi_2$, can be shown to be the solution to \eqref{eq:dmaps_opt_problem}, so that $\phi_2(j)$ gives the coordinate for data point $x_j$.
+%
+In our setup, we assume that this one dimension, parameterized by $\phi_2$, is correlated with time.
 %
@@
 
-%\begin{thebibliography}{10}
-
-%\bibitem{ahuja2007template}
-%Sunil Ahuja, Ioannis~G Kevrekidis, and Clarence~W Rowley.
-%\newblock Template-based stabilization of relative equilibria in systems with
-%  continuous symmetry.
-%\newblock {\em Journal of Nonlinear Science}, 17(2):109-143, 2007.
-
-%\bibitem{coifman2005geometric}
-%Ronald~R Coifman, Stephane Lafon, Ann~B Lee, Mauro Maggioni, Boaz Nadler,
-%  Frederick Warner, and Steven~W Zucker.
-%\newblock Geometric diffusions as a tool for harmonic analysis and structure
-%  definition of data: Diffusion maps.
-%\newblock {\em Proceedings of the National Academy of Sciences},
-%  102(21):7426-7431, 2005.
-%
-%\bibitem{coifman2008graph}
-%Ronald~R Coifman, Yoel Shkolnisky, Fred~J Sigworth, and Amit Singer.
-%\newblock Graph laplacian tomography from unknown random projections.
-%\newblock {\em IEEE Transactions on Image Processing}, 17(10):1891-1899, 2008.
-%
-%\bibitem{singer2011angular}
-%A~Singer.
-%\newblock Angular synchronization by eigenvectors and semidefinite programming.
-%\newblock {\em Applied and computational harmonic analysis}, 30(1):20-36,
-%  2011.
-%
-%\bibitem{singer2012vector}
-%Amit Singer and H-T Wu.
-%\newblock Vector diffusion maps and the connection laplacian.
-%\newblock {\em Communications on Pure and Applied Mathematics},
-%  65(8):1067-1144, 2012.
+%\begin{thebibliography}{}
 
@@
 \label{fig:fluorescent_images}
+\end{figure}
+
+\begin{figure}
+\begin{subfigure}{0.25\textwidth}
+\includegraphics[width=\textwidth]{dmaps_schematic_edges}
+\caption{}
+\label{subfig:dmaps_edges}
+\end{subfigure}
+\begin{subfigure}{0.25\textwidth}
+\includegraphics[width=\textwidth]{dmaps_schematic_color}
+\caption{}
+\label{subfig:dmaps_color}
+\end{subfigure}
+\caption{Diffusion maps illustration.
+(a) Data points in two dimensions which lie on a one-dimensional nonlinear curve.
+Each pair of data points is connected by an edge, and the edge intensity is related to the distance between the points (close data points are connected by dark edges).
+(b) The data in (a), colored by the first (non-trivial) DMAPS eigenvector, $\phi_2$.
+The color parameterizes the one-dimensional curve.
+}
+\label{fig:dmaps_schematic}
 \end{figure}

### 2014-03-26 image_analysis_paper/dmaps_schematic.m

@@ -1,11 +1,16 @@
 clear all
 close all
 
-n = 20;
+res = '-r300';
+fmt = '-djpeg';
+
+im_save_dir = 'paper_figures';
+%%
+n = 15;
 theta = linspace(0, 3*pi/2, n);
 
-x = cos(theta);
-y = sin(theta);
+x = 0.9*cos(theta);
+y = 0.9*sin(theta);
 
 %x = -1:0.1:1;
 %y = x.^3;
@@ -20,21 +25,28 @@ eps = median(W(:))/2;
 A = exp(-W/eps);
 %A = diag(1./sum(A))*A;
 
-figure;
-plot(x, y, '.', 'markersize', 10)
 
+%%
 
 figure;
 for i=1:n
     for j=1:n-1
         idx = ceil(size(cm, 1) / max(A(:)) * A(i,j));
-        plot([x(i) x(j)], [y(i) y(j)], 'color', cm(idx, :), 'linewidth', 1);
+        plot([x(i) x(j)], [y(i) y(j)], 'color', cm(idx, :), 'linewidth', 2);
         hold on
     end
 end     
-plot(x, y, '.', 'markersize', 20)
+plot(x, y, '.', 'markersize', 50)
+set(gca, 'xtick', [])
+set(gca, 'ytick', [])
+%axis off
+print(sprintf('%s/dmaps_schematic_edges', im_save_dir), fmt, res);
+
 
 [V, D] = dmaps(W, eps, 2);
 figure;
-scatter(x, y, 500, V(:,2), '.')
+scatter(x, y, 2000, V(:,2), '.')
+set(gca, 'xtick', [])
+set(gca, 'ytick', [])
+print(sprintf('%s/dmaps_schematic_color', im_save_dir), fmt, res);
 

### 2014-03-27 image_analysis_paper/pnas_draft.tex
updated paper

@@
 %
-Our algorithmic approach is based on diffusion maps, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets. 
+Our temporal ordering analysis is based on diffusion maps, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets.
+%
+Our image registration is done via angular synchronization, a relatively new technique for registering images which has been applied to reconstructing the three-dimensional structure of molecules from cryo-electron microscopy \cite{singer2011three}; here, we will use the algorithm to register {\em Drosophila} images.
 %
@@
 
-We will first demonstrate our techniques considering {\em only} considering rotations. 
+We will first demonstrate our techniques using one-dimensional concentrations profiles.
 %
-We have converted the two-dimensional images shown in Figure \ref{subfig:images_scrambled} into concentration profiles on a ring (we will address the full two-dimensional images in a subsequent section).
+We have converted the two-dimensional images shown in Figure \ref{subfig:images_scrambled} into profiles defined on a ring, corresponding to the concentrations around the periphery of the embryo (we will address the full two-dimensional images in a subsequent section).
 %
-Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one-dimensional concentration profile; the profile is discretized into 100 points. 
+Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one-dimensional profile; the profile is discretized into 100 points, and so our signals will be 200-dimensional vectors (100 dimensions correspond to the red channel at the 100 discretization points, and 100 dimensions correspond to the green channel at the 100 discretization points).
 %
-Rotation of this image corresponds to shifting (with periodic boundary conditions) the one--dimensional concentration profile. 
+Rotation of this image corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown in the bottom of Figure \ref{subfig:1d_image_example}. 
 %
@@
 The concentration profiles have been stacked in an array, such that each row corresponds to one image.
+%
+The task is now to shift each profile left and right to optimally register/align the signals, and then temporally order the rows in the array to reconstruct the developmental trajectory.
 
@@
 %
-This requires translating and rotating each image, so that the set of images are globally aligned.
+This requires rotating and/or translating each image, so that the set of images are globally aligned.
 
@@
 %
-Angular synchronization first computes the pairwise alignments (translation + rotation) for every possible pair of images, effectively using each image as a template for every other image.
+Angular synchronization first computes the pairwise alignments (rotation for our one-dimensional profiles in Figure \ref{subfig:1d_unaligned_unordered}, or translation + rotation for the two-dimensional images in Figure \ref{subfig:images_scrambled}) for every possible pair of images, effectively using each image as a template for every other image.
 %
-From this large collection of pairwise alignments, it then tries to find the optimal translation and rotation for each image, such that these individual alignments are most consistent (in some metric) with the pairwise alignments.
+From this large collection of pairwise alignments, it then tries to find the optimal alignment for each image, such that these individual alignments are most consistent (in some metric) with the pairwise alignments (i.e., we would like the difference between the optimal rotations for two different concentration profiles to be as close as possible to the pairwise rotation needed to align the two profiles).
 %
-The algorithm solves this optimization problem by relaxing it to an eigenproblem; the top eigenvector(s) of a matrix (which contains information about the pairwise alignments) give the translation and rotation for each image.
+The algorithm solves this optimization problem by relaxing it to an eigenproblem; the top eigenvector(s) of a matrix (which contains information about the pairwise alignments) give the alignment for each image.
 %
@@
 %
-We would like to note that, because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered {\em up to} a global rotation, and we are permitted rotate every point by a constant rotation.
+We would like to note that, because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered {\em up to} a global rotation; we are permitted rotate every point by a constant rotation.
 
@@
 %
-Although our data are high-dimensional (for example, we will work with 2-color images of 10000 pixels, and so we treat these images as 20000-dimensional vectors, where each dimension describes the intensity of one color in one pixel), we will assume that they lie on a one-dimensional, perhaps nonlinear curve.
+Although our data are high-dimensional (for example, we will work with 2-color images of 10000 pixels, and so our data are 20000-dimensional vectors, where each dimension describes the intensity of one color in one pixel), we will assume that they lie on a one-dimensional, but perhaps nonlinear, curve.
 %
-We further assume that curve can be parameterized by time.
+We further assume that this curve is monotonic in time.
 %
 Therefore, our goal is to {\em automatically} uncover a parameterization of this curve, so that we can sort/order our data along it, effectively ordering our data in time.
-
-To ``unravel'' this curve, we assume that data points which are close in the high-dimensional space should also be close on the curve.
 %
-We want to find a parameterization of the points, such that the points which are close in the original close are {\em as close as possible} in the parameterization.
+To parameterize this one-dimensional curve which lies in high-dimensional space, we assume that data points which are close in the high-dimensional space should also be close on the curve, 
+e.g, images which look similar will be close in the high-dimensional pixel space (they will have similar pixel intensities), and we would like these similar images to be close to each other in our trajectory.
 %
-This problem can be formulated as an optimization problem \cite{...}, and the solution is given by the top (non-trivial) eigenvector of a matrix which contains information about pairwise distances between the data points.
+%We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\em as close as possible} in the parameterization.
 
@@
 Furthermore, the light edges are not informative and do not give any meaningful structure to the curve.
+
+The problem of finding a parameterization of the data which forces nearby points together can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a matrix which contains information about pairwise distances between the data points.
+%
+This procedure is called diffusion maps \cite{coifman2005geometric}; the formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).
 %
@@
 %
-The parameterization is what we would expect visually.
+The parameterization agrees with what our eye can detect visually.
 %
-Although this data is useful for illustrative purposes, it is
+However, we will work with very high-dimensional data, rather than these two-dimensional data points; we therefore have no easy way to plot and visualize the data, and rather need DMAPS to automatically uncover such structure in the data.
 
-Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we now wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
+Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we would like to use DMAPS to order them in time.
 %
@@
 %
-It not only reduces the number of required eigencomputations, but also is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
-
-Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of aligning and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps. 
+It both reduces the number of required eigencomputations, and is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
+%
+Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using VDM. 
 %
@@
 %
-Instead of having $100$-dimensional concentration vectors, we now work with images containing $10000$ pixels.
+Instead of having $200$-dimensional concentration vectors, we now work with $20000$-dimensional images ($100x100$ pixel images containing two color channels).
 %
-The two-dimensional images are registered with respect to all possible translations and rotations; 
-we convert two-dimensional translations and rotations to three-dimensional rotations so that we can apply angular synchronization or vector diffusion maps. 
+The two-dimensional images are registered with respect to translations and rotations 
+(we convert the two-dimensional translations and rotations to three-dimensional rotations so that we can apply angular synchronization or vector diffusion maps). 
 %
@@
 
-In these {\em Drosophila} experiments, we have the luxury of an independent time marker from membranes.
+In these {\em Drosophila} experiments, we have the luxury of an independent measure of time from cellularization.
 %
-We can therefore evaluate the quality of our orderings obtained from VDM by comparing with the orderings obtained from the membrane markers.
+We can therefore evaluate the quality of our orderings obtained from VDM by comparing with the orderings obtained from cellularization.
 %
-The rank of each data point, computed using VDM, versus the rank of each data point computed from the membrane thickness, is shown in Figure~\ref{subfig:rank_corr}.
+The rank of each data point, computed using VDM, versus the rank of each data point computed from cellularization, is shown in Figure~\ref{subfig:rank_corr}.
 %
-The rank correlation is 0.8665, indicating that our ordering from VDM is consistent with the ordering obtained from the membrane thickness.
+The rank correlation is 0.8665, indicating that our ordering from VDM is consistent with the ordering obtained from cellularization.
 %
-There is more scatter for the younger embryos; this is to be expected, as the dpERK signal is very low in this range.
+There is more scatter for the younger embryos; this is to be expected, as the dpERK signal, which is the main feature that changes with time, is very low in this range.
 
-Although our methods are fairly robust, they do require a sufficient amount of data to ``see'' a smooth developmental trajectory.
+Although our methods are fairly robust, they do require a sufficient amount of data so that DMAPS can ``see'' a smooth trajectory.
 %
-We evaluated how much data is required by subsampling our data set (with replacement), and for each subsample, computing the temporal ordering using VDM and calculating the rank correlation between this ordering and the ordering using the membrane markers.
+We evaluated how much data is required by subsampling our data set (with replacement), and for each subsample, computing the temporal ordering using VDM and calculating the rank correlation between this ordering and the ordering from cellularization.
 %
-The results are shown in Figure~\ref{subfig:bootstrap}.
+The results are shown in Figure~\ref{subfig:bootstrap} (each data point is averaged over 50 random subsamples).
 %
@@
 %
-Although we illustrated our methods using a specific data set arising in a study of {\em Drosophila} embryogenesis, the techniques are very general and require little knowledge of the image features or the developmental dynamics.
+Although we illustrated our methods using a specific data set from a study of {\em Drosophila} embryogenesis, the techniques are very general and require little knowledge of the image features or the developmental dynamics.
 
-We would like to note that there have been multiple other works with temporally ordering biological data (RNA data, cell morphology data, disease progression data, etc.). 
+We would like to note that there have been multiple other works with temporally ordering biological data (RNA-seq, cell morphology, disease progression, etc.). 
 %
-The majority of these works temporally order the data either by solving a travelling salesman probelem, or by constructing a minimum spanning tree on the data.
+The majority of these works temporally order the data either by solving a traveling salesman problem, or by constructing a minimum spanning tree on the data.
 %
@@
 %
-There are many applications for which develpomental dynamics are of interest, and yet there is no time marker to use to reconstruct the dynamics from such cross-sectional data.
+They provide a simple and cost-effective alternative to live imaging or ordering images by hand in order to gain insight into the developmental dynamics.
 
@@
 
-\section{Angular synchronization}
+\section{Angular synchronization \cite{singer2011angular}}
 
@@
 
-\section{Diffusion Maps}
+\section{Diffusion Maps \cite{coifman2005geometric}}
 
@@
 %
-We therefore want to solve the following optimization problem
+We therefore want to solve the following optimization problem \cite{Belkin2003}
 \begin{equation} \label{eq:dmaps_opt_problem}
@@
 
-\section{Vector Diffusion Maps}
+\section{Vector Diffusion Maps \cite{singer2012vector}}
 
@@
 %
-If we assume that our data are one-dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, one can show that a parameterization of our data (i.e.
-the analog of $\phi_2$ from the diffusion maps case) is given by $\psi_{1,d+1}$.
+If we assume that our data are one-dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, one can show that a parameterization of our data (i.e., the analog of $\phi_2$ from the diffusion maps case) is given by $\psi_{1,d+1}$.
 %
@@
 %
-We do this by projecting the image onto a portion of the surface of a (three-dimensional) sphere (see Figure \ref{fig:SO3_picture}).
+We do this by projecting the image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular}. %(see Figure \ref{fig:SO3_picture}).
 %
@@
 We round the optimal translations to the nearest integer to remove the need for interpolation when translating the images.
+
+\section{Rank correlation}
+
+We use Spearman's rank correlation coefficient as a metric to evaluate the quality of our orderings.
+%
+Given data $x_1, \dots, x_m$ and $y_1, \dots, y_m$ for which we would like to compute the rank correlation coefficient, we first compute the ranks $X_i$ and $Y_i$ of each data point $x_i$ and $y_i$ in their respective data sets.
+%
+The rank correlation coefficient, $\rho$, is then given by
+\begin{equation}
+\rho = \frac{\sum_{i=1}^m (X_i - \overline{X})(Y_i - \overline{Y})}{\sqrt{\sum_{i=1}^m (X_i - \overline{X})^2 \sum_{i=1}^m (Y_i - \overline{Y})^2}}
+\end{equation}
+where $\overline{X}$ and $\overline{Y}$ are the means of $X_1, \dots, X_m$ and $Y_1, \dots, Y_m$, respectively.
+
 \end{materials}
@@
 (b) Fish, now registered and temporally ordered.
-This is easy to do by hand because the data set is small and the developmental changes are easy to recognize.}
+For this example, the registration and ordering is easy to do by hand because the data set is small and the developmental changes are easy to recognize.}
 \label{fig:fish}
@@
 \caption{{\em Drosophila} data.
-(a) Dorsal-ventral images of {\em Drosphila} embryos in nuclear cycle 14, each stained for Dorsal (green) and dpERK (red).
-(b) Images from (a), now registered using the location of the Dorsal peak, and temporally ordered using the membrane marker.}
+(a) Dorsal-ventral images of {\em Drosphila} embryos, each fluorescentally stained for Dorsal (green) and dpERK (red).
+(b) Images from (a), now registered using the location of the Dorsal peak, and temporally ordered using cellularization.}
 \label{fig:fluorescent_images}
@@
 \caption{Diffusion maps illustration.
-(a) Data points in two dimensions which lie on a one-dimensional nonlinear curve.
+(a) Data points (in blue) which lie on a one-dimensional nonlinear curve in two dimensions.
 Each pair of data points is connected by an edge, and the edge intensity is related to the distance between the points (close data points are connected by dark edges).
@@
 \end{subfigure}
-\caption{(a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
+\caption{Illustration of methodologies using one-dimensional profiles.
+(a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
+Rotation of the ring corresponds to shifting the line left and right (with periodic boundary conditions).
 (b) The concentration profiles corresponding to the images in Figure \ref{fig:fluorescent_images}.
@@
 (d) The concentration profiles in (c), now temporally ordered using diffusion maps.
+The profiles are sorted according to the value of the first non-trivial DMAPS component, $\phi_2$.
 (e) The concentration profiles in (b), now simultaneously registered and temporally ordered using vector diffusion maps.
-}
+The profiles are sorted according to the value of the first non-trivial VDM component, $\psi_{1, 3}$.}
 \label{fig:1d_demo}
@@
 \begin{subfigure}{\textwidth}
-\includegraphics[width=\textwidth]{registered_ordered_vdm_2d}
+%\includegraphics[width=\textwidth]{registered_ordered_vdm_2d}
+\includegraphics[width=\textwidth]{test}
 \caption{}
@@
 \caption{(a) The images in Figure~\ref{fig:fluorescent_images}, aligned and temporally ordered using vector diffusion maps.
-(b) The rank correlation between the ordering of the images in (a) using vector diffusion maps, and the ordering obtained using the membrane markers.
-(c) The rank correlation coefficient as a function of the number of data points.
+The images are sorted according to the value of the first non-trivial VDM component, $\psi_{1, 4}$.
+(b) The rank correlation between the ordering of the images in (a) using vector diffusion maps, and the ordering of in the images in Figure~\ref{subfig:images_hand_ordered}, obtained using cellularization.
+(c) The rank correlation coefficient between the VDM ordering and the cellularization ordering as a function of the number of data points.
 }

### 2014-03-27 image_analysis_paper/make_figures_large_dataset_color.m

@@ -180,7 +180,7 @@ figure;
 set(gcf, 'papersize', [11/4 8.5/4])
 set(gcf, 'paperposition', [0 0 11/4 8.5/4])
 plot(ranks_from_membranes, ranks_from_vdm, '.')
-xlabel('rank from membrane lengths')
+xlabel('rank from cellularization')
 ylabel('rank from vdm')
 print(sprintf('%s/rank_corr_vdm', im_save_dir), fmt, res);
 

### 2014-03-27 image_analysis_paper/pnas_draft.tex
updated paper figures; added toy example

@@
 \begin{figure}
-\begin{subfigure}{0.25\textwidth}
-\includegraphics[width=\textwidth]{fig2a}
-\caption{}
-\label{subfig:images_scrambled}
-\end{subfigure}
-\begin{subfigure}{0.25\textwidth}
-\includegraphics[width=\textwidth]{fig2b}
-\caption{}
-\label{subfig:images_hand_ordered}
-\end{subfigure}
-\caption{{\em Drosophila} data.
-(a) Dorsal-ventral images of {\em Drosphila} embryos, each fluorescentally stained for Dorsal (green) and dpERK (red).
-(b) Images from (a), now registered using the location of the Dorsal peak, and temporally ordered using cellularization.}
+\includegraphics[width=0.5\textwidth]{unregistered_unordered_2d}
+\caption{}
 \label{fig:fluorescent_images}
@@
 \begin{subfigure}{0.17\textwidth}
-\includegraphics[width=\textwidth]{illustrate_1d}
+\includegraphics[width=\textwidth]{illustrate_toy}
 \caption{}
@@
 \begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{unregistered_unordered_1d}
+\includegraphics[width=\textwidth]{unregistered_unordered_toy}
 \caption{}
@@
 \begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{registered_unordered_1d}
+\includegraphics[width=\textwidth]{registered_unordered_toy}
 \caption{}
@@
 \begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{registered_ordered_1d}
+\includegraphics[width=\textwidth]{registered_ordered_toy}
 \caption{}
@@
 \begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{registered_ordered_vdm_1d}
+\includegraphics[width=\textwidth]{registered_ordered_vdm_toy}
 \caption{}
@@
 \begin{subfigure}{\textwidth}
-%\includegraphics[width=\textwidth]{registered_ordered_vdm_2d}
-\includegraphics[width=\textwidth]{test}
+\includegraphics[width=\textwidth]{registered_ordered_vdm_2d}
 \caption{}
@@
 \end{minipage}
-\begin{minipage}{0.4\textwidth}
+\begin{minipage}{0.3\textwidth}
 \begin{subfigure}{\textwidth}
-\includegraphics[width=\textwidth]{rank_corr_vdm}
+\includegraphics[width=\textwidth]{registered_ordered_vdm_1d}
 \caption{}
@@
 \begin{subfigure}{\textwidth}
-\includegraphics[width=\textwidth]{bootstrap_rank_corr_color}
-\caption{}
-\label{subfig:bootstrap}
+\includegraphics[width=\textwidth]{registered_ordered_hand}
+\caption{}
+%\label{subfig:bootstrap}
 \end{subfigure}
 \end{minipage}
-\caption{(a) The images in Figure~\ref{fig:fluorescent_images}, aligned and temporally ordered using vector diffusion maps.
+%\caption{(a) The images in Figure~\ref{fig:fluorescent_images}, aligned and temporally ordered using vector diffusion maps.
 The images are sorted according to the value of the first non-trivial VDM component, $\psi_{1, 4}$.

### 2014-03-27 image_analysis_paper/make_figures_large_dataset_1d_color.m

@@ -59,6 +59,20 @@ for i=2:m
 end
 
 %%
+
+[~, I] = sort(mem_lengths);
+
+figure;
+set(gcf, 'paperposition',[0 0 8 8])
+imshow(uint8(dpERK(I, :, :)))
+xlabel('position (registered)', 'fontsize', fontsize)
+ylabel('data sample (ordered)', 'fontsize', fontsize)
+set(gca, 'xtick', [])
+set(gca, 'ytick', [])
+set(gca,'position',[0.1 0.1 0.9 0.9],'units','normalized')
+print(sprintf('%s/registered_ordered_hand', im_save_dir), fmt, res);
+return
+
 figure;
 set(gcf, 'paperposition',[0 0 8 8])
 imshow(uint8(dpERK_unaligned))

### 2014-03-27 image_analysis_paper/pnas_draft.tex
upadted paper

@@
 \dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional.
-%
-In longitudinal studies, a chemical or morphological process is monitored over time within the same embryo.  
-%
-In a cross-sectional study, one embryo provides just a single snapshot of a population-averaged developmental trajectory which must be reconstructed from multiple snapshots provided by different embryos.
-%
-Both experimental designs have their advantages and limitations and are extensively used in studies of embryonic development across species. 
-%
-Our work focuses on cross-sectional studies, which have a time-honored history and still present the only option for most of the organisms.  
-%
-Most commonly, a group of developing embryos is fixed (arresting their development), stained with chemicals that visualize a handful of cellular processes, and imaged by fluorescence microscopy. 
-%
-Recent advances in large scale physical manipulation of embryos produce rapidly increasing volumes of imaging data from such experiments, in which every embryo is observed at a different orientation and at a different point in its developmental trajectory.
-%
-In order to mine such datasets for underlying developmental dynamics, multiple images must be registered (aligned) and ordered in time. 
-%
-Here present an algorithmic framework that greatly accelerates both of these tasks.
-
-Temporal ordering and registration of images is straightforward, when the number of images is small and differences between them are apparent. 
-%
+In longitudinal studies, developmental progress is monitored over time within the same embryo.
+In a cross-sectional study, one embryo contributes only one snapshot of a chemical or morphological process along developmental trajectory which must be reconstructed from multiple snapshots of different embryos.
+Both of these experimental designs have their advantages and limitations and are extensively used by developmental biologists.
+Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most of the organisms.
+In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that visualize a handful of cellular processes.
+Fixed embryos are then imaged using any given number of microscopic techniques.
+Recent advances in large scale physical manipulation of embryos and their imaging produce rapidly increasing volumes of data from such experiments, in which every embryo is observed at a different orientation and at a different time point.
+Importantly, the ``age'' of any given embryo arrested in its development is not known to high accuracy.
+In general, it is known that a collection of embryos belongs to a certain time window.
+In order to mine such datasets for underlying developmental dynamics, snapshots of different embryos must be aligned and ordered in time.
+We present a general framework that greatly accelerates both of these tasks.
+
+Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are apparent.
 As an example, Figure~\ref{fig:fish} shows a schematic of fish development, combining the processes of growth and patterning.
-%
-In this case, temporal ordering can be accomplished by arranging the fish by their size, which correlates with the progress of development, and image alignment is based on obvious morphological landmarks, such positions of head and fins.
-%
+In this case, temporal ordering can be accomplished by arranging the fish by size, which correlates with the progress of development, and image alignment is based on obvious morphological landmarks, such as positions of head and fins.
 Dealing with real data is far more nontrivial.
-%
-To illustrate most of the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of pattern formation in the early {\em Drosophila} embryo, one of the leading systems for studies of developmental dynamics.
-%
-This dataset comprises 81 fluorescence microscopy images of optical cross-sections of different embryos fixed during the 3$^{rd}$ hour of their development.  
-%
-During this time interval, most of cell nuclei in the embryo are arranged under the common plasma membrane and embryo undergoes the process of cellularization, whereby membranes that enclose individual nuclei into different cells grow in length (Figure~\ref{subfig:images_scrambled}).
-
-Each of the images reveals a uniform arrangement of nuclei and nonuniform distribution of two proteins (green and red) involved in patterning of the dorsoventral (back-to-belly) patterning of the embryo.
-%
-Each image comes from a different embryo in captured at a different orientation and at a different and unknown point in time (Figure~\ref{subfig:images_scrambled}). 
-%
-The task is to align these images in a globally consistent way and order them in a way that is most representative of the underlying developmental trajectory. 
-%
-For this particular dataset, both of these tasks are simplified by prior knowledge. 
-%
-For instance, it is known that the peak in the distribution of one of the proteins (visualized by the green channel) specifies the ventralmost position in the embryo. 
-%
-Based on this, images translated to make the centers of the embryos coincide and rotated to place the peak of the green channel at the same angular position. 
-%
-At the same time, temporal ordering can be done based on the monotonic progress of cellularization. 
-%
-Figure~\ref{subfig:images_hand_ordered} shows the collection of registered and temporally ordered images.
-
-Note that both of these tasks have been accomplished based on prior knowledge about the underlying developmental processes: We knew that the peak of the green signal can be used to align the images and that the progress of cellularization can order them in time.
-%
-Is this information essential? 
-%
-Below we show that this is not the case and that both the temporal ordering and registration of images can be fully automated, without prior information about the underlying dynamics. 
-%
-Our temporal ordering analysis is based on diffusion maps, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets.
-%
-Our image registration is done via angular synchronization, a relatively new technique for registering images which has been applied to reconstructing the three-dimensional structure of molecules from cryo-electron microscopy \cite{singer2011three}; here, we will use the algorithm to register {\em Drosophila} images.
+To illustrate most of the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of pattern formation in the early fruit fly embryo, one of the leading systems for studies of developmental dynamics.
+This dataset comprises 81 fluorescence microscopy images of different embryos fixed during the 3$^{rd}$ hour of their development.
+
+The fruit fly embryo is shaped like a rice grain, approximately ½ mm long.
+By end of the 2nd hour of development, sequential divisions of a fertilized egg generated a system where $\sim 6,000$ nuclei are uniformly arranged in under the common plasma membrane.
+During the 3$^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclosing nuclei into different cells grow in length (Figure 2a).
+Concurrently with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs.
+Images in Figure 2b show optical sections through the short axis of the embryos.
+The embryos were fixed and stained with chemicals that visualize nuclei as well as the spatial distributions of two different proteins.
+Each of the images reveals a uniform arrangement of nuclei and nonuniform distribution of two proteins (green and red) involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.
+
+Each image in Figure 2b comes from a different embryo, captured at a different orientation and at a different and unknown point in time (Figure 2a).
+The goal is to align these images in a globally consistent way and order them in a way that is most representative of the underlying developmental trajectory.
+Below we show that both of these tasks can be fully automated.
+%
+We will use angular synchronization \cite{singer2011angular}, to align the images in a consistent way, and use diffusion maps \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
+%
+We will also show how both steps can be combined using methodology called vector diffusion maps \cite{singer2012vector}.
 %
 The paper is organized as follows.
-%
-In the next two sections, we use a simpler version of the data (one-dimensional spatial profiles) to formulate both tasks mathematically.  
-%
-Then, we show that both of the tasks can be effectively accomplished in single step. 
-%
-Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align the images and order them correctly in time. 
-%
-We conclude by discussion of the general requirements that must be satisfied for our approach to work and compare it to other ideas in cross-sectional studies of biological dynamics.
+First, we use a synthetic dataset, where the correct orientation and ordering of data points is known, to formalize the tasks of image registration and temporal ordering.
+Then, we show that both of these tasks can be effectively combined in a single step.
+Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align them and order them correctly in time.
+We conclude by discussing the general requirements that must be satisfied for our approach to work and compare it to other approaches and ideas in cross-sectional studies of biological dynamics.
 
@@
 
-We will first demonstrate our techniques using one-dimensional concentrations profiles.
-%
-We have converted the two-dimensional images shown in Figure \ref{subfig:images_scrambled} into profiles defined on a ring, corresponding to the concentrations around the periphery of the embryo (we will address the full two-dimensional images in a subsequent section).
-%
-Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one-dimensional profile; the profile is discretized into 100 points, and so our signals will be 200-dimensional vectors (100 dimensions correspond to the red channel at the 100 discretization points, and 100 dimensions correspond to the green channel at the 100 discretization points).
-%
-Rotation of this image corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown in the bottom of Figure \ref{subfig:1d_image_example}. 
-%
-Figure \ref{subfig:1d_unaligned_unordered} shows the concentration profiles for all the images in our data set. 
+We will first demonstrate our techniques using a synthetic data set with relatively simple dynamics.
+%
+We consider a concentration profile defined on a ring, shown in Figure~\ref{subfig:1d_image_example}.
+%
+The concentration profile is a Gaussian bump, which increases in intensity as a function of time.
+%
+Figure \ref{subfig:1d_image_example} shows a typical concentration profile; the profile is discretized into 100 points, and so our signals will be 100-dimensional vectors. 
+%
+We assume that the ring is in an arbitrary orientation, and we are allowed to rotate the ring.
+%
+Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown in the bottom of Figure \ref{subfig:1d_image_example}. 
+%
+Figure \ref{subfig:1d_unaligned_unordered} shows all the concentration profiles in our data set. 
 %
@@
 %
-Angular synchronization first computes the pairwise alignments (rotation for our one-dimensional profiles in Figure \ref{subfig:1d_unaligned_unordered}, or translation + rotation for the two-dimensional images in Figure \ref{subfig:images_scrambled}) for every possible pair of images, effectively using each image as a template for every other image.
+Angular synchronization first computes the pairwise alignments (rotation for our one-dimensional profiles in Figure \ref{subfig:1d_unaligned_unordered}, or translation + rotation for the two-dimensional images in Figure \ref{fig:fluorescent_images}) for every possible pair of images, effectively using each image as a template for every other image.
 %
@@
 %
-We use the total Euclidean distance (both the red and the green channels) between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances for DMAPS (in \eqref{eq:dmaps_W}).
+We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances for DMAPS (in \eqref{eq:dmaps_W}).
 %
@@
 %
-Instead of having $200$-dimensional concentration vectors, we now work with $20000$-dimensional images ($100x100$ pixel images containing two color channels).
+Instead of having $100$-dimensional concentration vectors, we now work with $20000$-dimensional images ($100x100$ pixel images containing two color channels).
 %
@@
 %
-We used vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_scrambled}.
+We used vector diffusion maps to align and order the images shown in Figure~\ref{fig:fluorescent_images}
 %
@@
 
-In these {\em Drosophila} experiments, we have the luxury of an independent measure of time from cellularization.
-%
-We can therefore evaluate the quality of our orderings obtained from VDM by comparing with the orderings obtained from cellularization.
-%
-The rank of each data point, computed using VDM, versus the rank of each data point computed from cellularization, is shown in Figure~\ref{subfig:rank_corr}.
-%
-The rank correlation is 0.8665, indicating that our ordering from VDM is consistent with the ordering obtained from cellularization.
-%
-There is more scatter for the younger embryos; this is to be expected, as the dpERK signal, which is the main feature that changes with time, is very low in this range.
-
-Although our methods are fairly robust, they do require a sufficient amount of data so that DMAPS can ``see'' a smooth trajectory.
-%
-We evaluated how much data is required by subsampling our data set (with replacement), and for each subsample, computing the temporal ordering using VDM and calculating the rank correlation between this ordering and the ordering from cellularization.
-%
-The results are shown in Figure~\ref{subfig:bootstrap} (each data point is averaged over 50 random subsamples).
-%
-For small data sets, the average rank correlation is very low, indicating that the orderings from VDM are inaccurate. 
-%
-However, for moderately sized data sets (approximately 30 data points), the 
-rank correlation begins to plateau and VDM can accurately order the data.
+In these {\em Drosophila} experiments, both registration and temporal ordering can be done by hand using prior knowledge about the system. 
+%
+For instance, it is known that the peak in the distribution of one of the proteins (visualized by the green channel) specifies the ventralmost position in the embryo. 
+%
+Based on this, images translated to make the centers of the embryos coincide and rotated to place the peak of the green channel at the same angular position. 
+%
+At the same time, temporal ordering can be done based on the monotonic progress of cellularization. 
+%
+We can therefore evaluate the quality of our orderings obtained from VDM by comparing hand registered and ordering images.
+%
+These results are shown in Figure~\ref{subfig:1dprofiles_hand}; the results are visually consistent with the results in Figure~\ref{subfig:1dprofiles_vdm} obtained using VDM. 
+%
+Furthermore, we can quantify the accuracy of our orderings by computing the rank correlation coefficient between the ordering obtained form VDM, and the orderings obtained from cellularization.
+%
+The rank correlation coefficient is 0.8665, indicating that our ordering from VDM is consistent with the ordering obtained from cellularization.
+
+%Although our methods are fairly robust, they do require a sufficient amount of data so that DMAPS can ``see'' a smooth trajectory.
+%%
+%We evaluated how much data is required by subsampling our data set (with replacement), and for each subsample, computing the temporal ordering using VDM and calculating the rank correlation between this ordering and the ordering from cellularization.
+%%
+%The results are shown in Figure~\ref{subfig:bootstrap} (each data point is averaged over 50 random subsamples).
+%%
+%For small data sets, the average rank correlation is very low, indicating that the orderings from VDM are inaccurate. 
+%%
+%However, for moderately sized data sets (approximately 30 data points), the 
+%rank correlation begins to plateau and VDM can accurately order the data.
 
@@
 %
-Although we illustrated our methods using a specific data set from a study of {\em Drosophila} embryogenesis, the techniques are very general and require little knowledge of the image features or the developmental dynamics.
-
-We would like to note that there have been multiple other works with temporally ordering biological data (RNA-seq, cell morphology, disease progression, etc.). 
-%
-The majority of these works temporally order the data either by solving a traveling salesman problem, or by constructing a minimum spanning tree on the data.
-%
-We, instead, use nonlinear dimensionality reduction techniques, to construct a one-dimensional embedding of the data.
-%
-Furthermore, we work with images that must be first registered before they can be temporally ordered.
-
-Although our methods are applicable to a wide variety of data sets, there are several questions for which our techniques can provide no answers.
-%
-Because our registration algorithm only considers pairwise information, we cannot recover the true ``zero'' point of the images. 
-%
-Similairly, we cannot know which end of our trajectory is the beginning and which is the end.
-%
-Both of these tasks bust be done by hand as a post-processing step.
-%
-Furthermore, our methods can only provide a temporal ordering for the data, and cannot provide any information as to the speed or absolute time of the data samples.
-%
-Regardless, we are confident that our methods hold great promise in areas where cross-sectional imaging data is readily available and relatively inexpensive to obtain. 
-%
-They provide a simple and cost-effective alternative to live imaging or ordering images by hand in order to gain insight into the developmental dynamics.
+Although we illustrated our methods using a specific data set from a study of {\em Drosophila} embryogenesis, our framework is very general and requires little knowledge of the image features or the developmental dynamics.
+
+Temporal ordering of cross-sectional biological data has been studied in a variety of other contexts.
+%
+Cross-sectional RNA sequencing data has been temporally ordered \cite{anavy2014blind}, where each data point is the transcriptome for a single sample.
+%
+Gene expression data from multiple patients has been studied \cite{gupta2008extracting} \cite{qiu2011discovering},
+%
+where each data point is a vector containing the expression levels of the various genes.
+%
+Snapshots of cells throughout the cell cycle can be described by a vector of features (e.g., amount of DNA) which describe the cell's state  \cite{kafri2013dynamics}.
+%
+These problems are no different than our task of ordering images, each of which can be thought of as a large vector of pixel intensities. 
+%
+The majority of these works temporally order the data either by solving a traveling salesman problem (TSP) on the data, or by constructing a minimum spanning tree (MST) of the data, 
+in the hopes that these structures (the path of the TSP, or the MST) characterize the majority of the structure within the data and provide an accurate picture of the dynamical progression.
+%
+We, instead, use diffusion maps, to construct a one-dimensional embedding of the data.
+%
+In general, diffusion maps have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{tenenbaum2000global} \cite{balasubramanian2002isomap}.
+%
+Furthermore, although we use only the first DMAPS component to order the data (we assume the data are one-dimensional), DMAPS is not limited to one-dimensional data and can be used to analyze data containing more complex structure.
+%
+For example, in data which contains several branches or several classes, the first few DMAPS components will delineate the branches or separate the classes.
+
+The task of image registration is widely studied, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
+%
+The main point to note here is that, in most of these applications, registration first requires defining a set of {\em features} or landmark points, and then locating these features within each image.
+%
+For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then compare images by comparing these features.
+%
+Temporal ordering of images has been studied, for example, in ordering images taken of a construction site over time \cite{schindler2007inferring}, 
+but these methods are also based on first extracting features from the images. 
+%
+In our more general setting of applications in developmental biology, ``good'' features may not be known {\em a priori}, and so we prefer a methodology which does not require this preprocessing step.
+%
+The registration algorithm we use, angular synchronization/vector diffusion maps, has been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}.
+%
+However, in this application, each image is a picture of (essentially) the {\em same} molecule; variation in the data comes from differences in viewing angle (rotations of the molecule) and from instrument/measurement noise.
+%
+In contrast, our data contains variation from the different embryo orientations, as well as the temporal variation of the underling developmental dynamics.
+
+In all of these data-driven methods, the essential question is ``what data is necessary?'': we would like to know how much and what type of data is needed to accurately reconstruct the dynamics. 
+%
+For the methods we present here, we require a sufficient amount of data for the one-dimensional curve to be ``seen'' by DMAPS (i.e., in Figure ~\ref{fig:dmaps_schematic}, there are a sufficient number of data points so that our eye can see the curve; if there were only 4 data points, it would most likely be difficult for our eye to detect the one-dimensional structure).
+%
+We also require the data we collect to contain enough information/be ``rich enough'' so that the alignments and dynamics are apparent.
+%
+For example, in the data presented in Figure~\ref{fig:fluorescent_images}, the green Dorsal peaks delineate the position of the ventral axis in the embryo, and the intensity of the red dpERK peaks grow over time.
+%
+Therefore, these images contain information about rotational symmetries of the data as well as the dynamics of the data.
+%
+If we removed the green Dorsal peaks, and only used the red dpERK peaks in our analysis, the registration would not be as accurate.
+%
+Furthermore, if we only used the green Dorsal peaks, then there would be no dynamic information (the shape or intensity of Dorsal does not change as a function of time in the developmental stage we are considering), and so the temporal ordering would be inaccurate.
+%
+However, we are confident that for many applications, these requirements - a sufficient number of data points, and data which contain enough information - are easily met. 
+%
+In these cases, our methods allow for the rapid, automated analysis of complex imaging data sets to help uncover complex dynamics and structure.
 
@@
 From here, we can write rotations around the three principal axes in terms of the three Euler angles
+\small
 \begin{equation}
@@
 \end{equation}
+\normalsize
 where $R^x(\alpha)$, $R^y(\beta)$, and $R^z(\gamma)$ correspond to rotations around the $x$, $y$, and $z$ axes, respectively.
@@
 \bibliographystyle{pnas}
-\bibliography{../../references/references}
+\bibliography{background_reading/references,../../references/references}
 
@@
 \includegraphics[width=0.5\textwidth]{unregistered_unordered_2d}
-\caption{}
+\caption{Fluorescent images of {\em Drosophila} embryos. Each embryo has been stained from Dorsal (green) and dpERK (red).}
 \label{fig:fluorescent_images}
 \end{figure}
+
+\begin{figure*}
+\begin{subfigure}{0.17\textwidth}
+\includegraphics[width=\textwidth]{illustrate_toy}
+\caption{}
+\label{subfig:1d_image_example}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{unregistered_unordered_toy}
+\caption{}
+\label{subfig:1d_unaligned_unordered}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{registered_unordered_toy}
+\caption{}
+\label{subfig:1d_aligned_unordered}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{registered_ordered_toy}
+\caption{}
+\label{subfig:1d_aligned_ordered}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{registered_ordered_vdm_toy}
+\caption{}
+\label{subfig:1d_aligned_ordered_vdm}
+\end{subfigure}
+\caption{Illustration of methodologies using one-dimensional profiles.
+(a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
+Rotation of the ring corresponds to shifting the line left and right (with periodic boundary conditions).
+(b) Many concentration profiles, each of the form described in (a).
+Each row in the array corresponds to one image.
+The overall dynamics are a Gaussian bump growing in time, but this is not readily apparent from the data because the profiles are unregistered and unordered.
+(c) The concentration profiles in (b), now registered using angular synchronization.
+(d) The concentration profiles in (c), now temporally ordered using diffusion maps.
+The profiles are sorted according to the value of the first non-trivial DMAPS component, $\phi_2$.
+(e) The concentration profiles in (b), now simultaneously registered and temporally ordered using vector diffusion maps.
+The profiles are sorted according to the value of the first non-trivial VDM component, $\psi_{1, 3}$.}
+\label{fig:1d_demo}
+\end{figure*}
 
@@
 \begin{figure*}
-\begin{subfigure}{0.17\textwidth}
-\includegraphics[width=\textwidth]{illustrate_toy}
-\caption{}
-\label{subfig:1d_image_example}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{unregistered_unordered_toy}
-\caption{}
-\label{subfig:1d_unaligned_unordered}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{registered_unordered_toy}
-\caption{}
-\label{subfig:1d_aligned_unordered}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{registered_ordered_toy}
-\caption{}
-\label{subfig:1d_aligned_ordered}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{registered_ordered_vdm_toy}
-\caption{}
-\label{subfig:1d_aligned_ordered_vdm}
-\end{subfigure}
-\caption{Illustration of methodologies using one-dimensional profiles.
-(a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
-Rotation of the ring corresponds to shifting the line left and right (with periodic boundary conditions).
-(b) The concentration profiles corresponding to the images in Figure \ref{fig:fluorescent_images}.
-Each row in the array corresponds to one image.
-(c) The concentration profiles in (b), now registered using angular synchronization.
-(d) The concentration profiles in (c), now temporally ordered using diffusion maps.
-The profiles are sorted according to the value of the first non-trivial DMAPS component, $\phi_2$.
-(e) The concentration profiles in (b), now simultaneously registered and temporally ordered using vector diffusion maps.
-The profiles are sorted according to the value of the first non-trivial VDM component, $\psi_{1, 3}$.}
-\label{fig:1d_demo}
-\end{figure*}
-
-\begin{figure*}
 \begin{minipage}{0.6\textwidth}
@@
 \begin{subfigure}{\textwidth}
-\includegraphics[width=\textwidth]{registered_ordered_vdm_1d}
-\caption{}
-\label{subfig:rank_corr}
+\includegraphics[width=\textwidth]{registered_ordered_1dfrom2d}
+\caption{}
+\label{subfig:1dprofiles_vdm}
 \end{subfigure}
@@
 \caption{}
-%\label{subfig:bootstrap}
+\label{subfig:1dprofiles_hand}
 \end{subfigure}
 \end{minipage}
-%\caption{(a) The images in Figure~\ref{fig:fluorescent_images}, aligned and temporally ordered using vector diffusion maps.
+\caption{(a) The images in Figure~\ref{fig:fluorescent_images}, aligned and temporally ordered using vector diffusion maps.
 The images are sorted according to the value of the first non-trivial VDM component, $\psi_{1, 4}$.
-(b) The rank correlation between the ordering of the images in (a) using vector diffusion maps, and the ordering of in the images in Figure~\ref{subfig:images_hand_ordered}, obtained using cellularization.
-(c) The rank correlation coefficient between the VDM ordering and the cellularization ordering as a function of the number of data points.
+(b) The one-dimensional concentration profiles corresponding to the images in (a).
+Each row of the array corresponds to one image.
+(c) The one-dimensional concentration profiles, extracted, registered, and ordered by hand.
+The profiles are extracted using the nuclei positions in the image, registered using the location of the Dorsal peak, and ordered using cellularization.
 }

### 2014-03-27 image_analysis_paper/make_figures_large_dataset_color.m

@@ -38,6 +38,7 @@ subplot_dim2 = ceil(m / subplot_dim1);
 nchannels = 3;
 image_set = zeros(npixels, npixels, nchannels, m, 'uint8');
 
+nuclei = zeros(npixels, npixels, m, 'uint8');
 
 %% load images
 
@@ -47,6 +48,8 @@ for i=1:m
     
     % resize image
     im1 = imresize(im1, [npixels npixels]);
+    
+    nuclei(:, :, i) = im1(:, :, 1);
     
     im1(:, :, 1) = im1(:, :, 2);
     im1(:, :, 2) = im1(:, :, 3);
@@ -78,6 +81,10 @@ figure;
 set(gcf, 'paperposition',[0 0 8 8])
 for i=1:m
     im1 = image_set(:,:,:,i);
+    
+    for j=1:3
+        im1(:,:,j) = im1(:,:,j) + nuclei(:,:,i);
+    end
     %subplot(subplot_dim1, subplot_dim2, i);
     subplot('position', [X(i)-1/subplot_dim1 Y(i)-1/subplot_dim2 1/subplot_dim1-0.01 1/subplot_dim2-0.01])
     imshow(im1);
@@ -87,7 +94,11 @@ print(sprintf('%s/unregistered_unordered_2d', im_save_dir), fmt, res);
 
 image_set_aligned = zeros(size(image_set), 'uint8');
 for i=1:m
-    im1 = rotate_image(R_opt(dim*(i-1)+1:dim*i,:), image_set(:,:,:,i), angle_proj);
+    im_tmp = image_set(:,:,:,i);
+    for j=1:3
+        im_tmp(:,:,j) = im_tmp(:,:,j) + nuclei(:,:,i);
+    end
+    im1 = rotate_image(R_opt(dim*(i-1)+1:dim*i,:), im_tmp, angle_proj);
     image_set_aligned(:,:,:,i) = im1;
 end
 
@@ -153,8 +164,18 @@ print(sprintf('%s/rank_corr_dmaps', im_save_dir), fmt, res);
 
 eps = median(W(:));
 
-
 [R_opt, embed_coord, embed_idx, D] = vdm(R, W, eps, neigs);
+
+image_set_aligned = zeros(size(image_set), 'uint8');
+for i=1:m
+    im_tmp = image_set(:,:,:,i);
+    for j=1:3
+        im_tmp(:,:,j) = im_tmp(:,:,j) + nuclei(:,:,i);
+    end
[…]

### 2014-03-28 image_analysis_paper/make_figures_toy_1d.m
upadted paper; made figures to send to bomyi

@@ -2,8 +2,9 @@ clear all
 close all
 
 res = '-r300';
-fmt = '-djpeg';
-fontsize = 30;
+%fmt = '-djpeg';
+fmt = '-depsc';
+fontsize = 10;
 
 im_save_dir = 'paper_figures';
 
@@ -45,20 +46,19 @@ for i=2:m
     dpERK_unaligned(i,:) = circshift(dpERK(i,:),[0 rand_offsets(i)]);
 end
 
-% dpERK_unaligned = [dpERK_unaligned; fliplr(dpERK_unaligned)];
-% mem_lengths = [mem_lengths; mem_lengths];
-% ranks_from_membranes = compute_ranks(mem_lengths);
-
 %%
 figure;
-set(gcf, 'paperposition',[0 0 8 8])
+set(gcf, 'paperunits', 'centimeters')
+set(gcf, 'papersize', [4 4])
+set(gcf, 'paperposition',[0 0 4 4])
 image(dpERK_unaligned)
 colormap(cm_red)
-xlabel('position (unregistered)', 'fontsize', fontsize)
-ylabel('data sample (unordered)', 'fontsize', fontsize)
+xlabel('position', 'fontsize', fontsize)
+ylabel('sample', 'fontsize', fontsize)
 set(gca, 'xtick', [])
 set(gca, 'ytick', [])
-print(sprintf('%s/unregistered_unordered_toy', im_save_dir), fmt, res);
+%print(sprintf('%s/unregistered_unordered_toy', im_save_dir), fmt, res);
+saveas(gcf, sprintf('%s/fig3c', im_save_dir), 'epsc');
 
 %% 
 
@@ -82,7 +82,9 @@ for i=1:n
 end
 
 figure;
-set(gcf, 'paperposition',[0 0 9 10])
+set(gcf, 'paperunits', 'centimeters')
+set(gcf, 'papersize', [4 4])
+set(gcf, 'paperposition',[0 0 4 4])
 subplot(10, 1, 1:9)
 image(image_test)
 axis off
@@ -93,8 +95,20 @@ subplot(10, 1, 10)
 image(dpERK_unaligned(idx,:));
 axis off
 colormap(cm_red)
+%print(sprintf('%s/illustrate_toy', im_save_dir), fmt, res);
+saveas(gcf, sprintf('%s/fig3a', im_save_dir), 'epsc');
 
-print(sprintf('%s/illustrate_toy', im_save_dir), fmt, res);
+
+figure;
[…]

### 2014-03-31 image_analysis_paper/pnas_draft.tex
updated paper

@@
 \begin{abstract} 
-In studies of development, researchers are often presented with cross--sectional data, where each data point is a sample from a population fixed at a slightly different developmental time. 
+In studies of development, researchers are often presented with cross-sectional data, where each data point is a sample from a population fixed at a slightly different developmental time. 
 %
@@
 \dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional. 
+%
 In longitudinal studies, developmental progress is monitored over time within the same embryo. 
+%
 In a cross-sectional study, one embryo contributes only one snapshot of a chemical or morphological process along developmental trajectory which must be reconstructed from multiple snapshots of different embryos.
+%
 Both of these experimental designs have their advantages and limitations and are extensively used by developmental biologists. 
-Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most of the organisms.
+%
+Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms. 
+%
 In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that visualize a handful of cellular processes. 
+%
 Fixed embryos are then imaged using any given number of microscopic techniques. 
+%
 Recent advances in large scale physical manipulation of embryos and their imaging produce rapidly increasing volumes of data from such experiments, in which every embryo is observed at a different orientation and at a different time point.
+%
 Importantly, the ``age'' of any given embryo arrested in its development is not known to high accuracy. 
+%
 In general, it is known that a collection of embryos belongs to a certain time window.
-In order to mine such datasets for underlying developmental dynamics, snapshots of different embryos must be aligned and ordered in time.
+%
+In order to mine such datasets for underlying developmental dynamics, snapshots of different embryos must be spatially aligned and ordered in time. 
+%
 We present a general framework that greatly accelerates both of these tasks.
@@
 Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are apparent. 
+%
 As an example, Figure~\ref{fig:fish} shows a schematic of fish development, combining the processes of growth and patterning.  
+%
 In this case, temporal ordering can be accomplished by arranging the fish by size, which correlates with the progress of development, and image alignment is based on obvious morphological landmarks, such as positions of head and fins.
+%
 Dealing with real data is far more nontrivial. 
+%
 To illustrate most of the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of pattern formation in the early fruit fly embryo, one of the leading systems for studies of developmental dynamics.
-This dataset comprises 81 fluorescence microscopy images of different embryos fixed during the 3$^{rd}$ hour of their development.
-
-The fruit fly embryo is shaped like a rice grain, approximately ½ mm long.
-By end of the 2nd hour of development, sequential divisions of a fertilized egg generated a system where $\sim 6,000$ nuclei are uniformly arranged in under the common plasma membrane.
-During the 3$^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclosing nuclei into different cells grow in length (Figure 2a).
+%
+This dataset comprises $81$ fluorescence microscopy images of different embryos fixed during the $3^{rd}$ hour of their development.
+
+The fruit fly embryo is shaped like a rice grain, approximately $1/2$~mm long. 
+%
+By end of the $2^{nd}$ hour of development, sequential divisions of a fertilized egg generated a system where $\sim 6,000$ nuclei are uniformly arranged in under the common plasma membrane. 
+%
+During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclosing nuclei into different cells grow in length (Figure~\ref{fig:membrane_compare}a). 
+%
 Concurrently with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs. 
-Images in Figure 2b show optical sections through the short axis of the embryos.
+%
+Images in Figure~\ref{fig:images_unordered} show optical sections through the short axis of the embryos. 
+%
 The embryos were fixed and stained with chemicals that visualize nuclei as well as the spatial distributions of two different proteins. 
+%
 Each of the images reveals a uniform arrangement of nuclei and nonuniform distribution of two proteins (green and red) involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.
 
-Each image in Figure 2b comes from a different embryo, captured at a different orientation and at a different and unknown point in time (Figure 2a).
+Each image in Figure~\ref{fig:images_unordered} comes from a different embryo, captured at a different orientation and at a different and unknown point in time. 
+%
 The goal is to align these images in a globally consistent way and order them in a way that is most representative of the underlying developmental trajectory. 
+%
 Below we show that both of these tasks can be fully automated.
 %
-We will use angular synchronization \cite{singer2011angular}, to align the images in a consistent way, and use diffusion maps \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
+We will use angular synchronization \cite{singer2011angular}, a registration algorithm that is robust to noise, to align the images in a consistent way.
+%
+We will use diffusion maps \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
 %
@@
 The paper is organized as follows. 
+%
 First, we use a synthetic dataset, where the correct orientation and ordering of data points is known, to formalize the tasks of image registration and temporal ordering. 
+%
 Then, we show that both of these tasks can be effectively combined in a single step. 
+%
 Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align them and order them correctly in time. 
+%
 We conclude by discussing the general requirements that must be satisfied for our approach to work and compare it to other approaches and ideas in cross-sectional studies of biological dynamics.
@@
 
-We will first demonstrate our techniques using a synthetic data set with relatively simple dynamics.
-%
-We consider a concentration profile defined on a ring, shown in Figure~\ref{subfig:1d_image_example}.
-%
-The concentration profile is a Gaussian bump, which increases in intensity as a function of time.
-%
-Figure \ref{subfig:1d_image_example} shows a typical concentration profile; the profile is discretized into 100 points, and so our signals will be 100-dimensional vectors. 
-%
-We assume that the ring is in an arbitrary orientation, and we are allowed to rotate the ring.
-%
-Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown in the bottom of Figure \ref{subfig:1d_image_example}. 
-%
-Figure \ref{subfig:1d_unaligned_unordered} shows all the concentration profiles in our data set. 
-%
-The concentration profiles have been stacked in an array, such that each row corresponds to one image.
-%
-The task is now to shift each profile left and right to optimally register/align the signals, and then temporally order the rows in the array to reconstruct the developmental trajectory.
+Our goal is to temporally order images, such as the images shown in Figure~\ref{fig:images_unordered}.
+%
+Our ordering algorithm will require distances between data points. 
+%
+However, because each image is in a different orientation, we need to first align or {\em register} the images.
+%
+Once all the images have been aligned, we can compute meaningful distances and temporally order the profiles to reconstruct the developmental trajectory.
+
+We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set with relatively simple dynamics.
+%
+We construct concentration profiles defined on a ring, and we assume that each ring is in an arbitrairy orientation; an example is shown in Figure~\ref{fig:1d_demo}a.
+%
+Each concentration profile is a Gaussian (shown in Figure~\ref{fig:1d_demo}b), and the Gaussians increase in intensity as  a function of time.
+%
+We discritize the profiles into 100 points, so our data will be 100-dimensional vectors. 
+%
+Rotation of the ring in Figure~\ref{fig:1d_demo}a corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown below the ring. 
+%
+Figure \ref{fig:1d_demo}c shows all the concentration profiles in our data set. 
+%
+The concentration profiles have been stacked in an array, such that each row corresponds to one data point.
+%
+Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian growing in time) are not apparant.
 
@@
 
-Often when one collects images in experiments, each image is taken in a different orientation. 
-%
-Therefore, the first task in image analysis is to align or {\em register} the set of images. 
-%
-This requires rotating and/or translating each image, so that the set of images are globally aligned.
-
-There are many methods for image registration. 
-%
-``Template-based'' registration \cite{ahuja2007template} is a common technique, where one selects a specific image or {\em template}, and then translates and rotates each image in the data set until it is optimally aligned with this template (optimal for us means minimizing the norm between the difference in pixel intensities).
-%
-However, this method can result in many errors if the images are noisy, and can be problematic if a ``good'' template image is not known {\em a priori}. 
-%
-Instead, we will use a relatively new technique called angular synchronization\cite{singer2011angular} to register sets of images.
-%
-Angular synchronization first computes the pairwise alignments (rotation for our one-dimensional profiles in Figure \ref{subfig:1d_unaligned_unordered}, or translation + rotation for the two-dimensional images in Figure \ref{fig:fluorescent_images}) for every possible pair of images, effectively using each image as a template for every other image.
-%
-From this large collection of pairwise alignments, it then tries to find the optimal alignment for each image, such that these individual alignments are most consistent (in some metric) with the pairwise alignments (i.e., we would like the difference between the optimal rotations for two different concentration profiles to be as close as possible to the pairwise rotation needed to align the two profiles).
-%
-The algorithm solves this optimization problem by relaxing it to an eigenproblem; the top eigenvector(s) of a matrix (which contains information about the pairwise alignments) give the alignment for each image.
-%
-Because the method considers {\em all} pairwise alignments, angular synchronization is often much more robust to noise.
-
-Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization;
-the peaks are aligned, even though we input no information about the existence or location of these peaks into our algorithm.
-%
-We would like to note that, because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered {\em up to} a global rotation; we are permitted rotate every point by a constant rotation.
+Our first step is to register the set of images or profiles.
+%
+%``Template-based'' registration \cite{ahuja2007template} is a common technique, where one selects a specific image or {\em template}, and then translates and rotates each image in the data set until it is optimally aligned with this template (optimal for us means minimizing the norm between the difference in pixel intensities).
+%
+%However, this method can result in many errors if the images are noisy, and can be problematic if a ``good'' template image is not known {\em a priori}. 
+%
+We will use a relatively new technique called angular synchronization\cite{singer2011angular}, which is robust to noise and requires no {\em a priori} knowledge of image features.
+%
+Angular synchronization first computes the pairwise alignments %(rotations for our one-dimensional profiles in Figure \ref{fig:1d_demo}c, or translations and rotations for the two-dimensional images in Figure \ref{fig:images_unordered})
+for every possible pair of data points.%, effectively using each image as a template for every other image.
+%
+An illustration is shown in Figure~\ref{fig:schematics}c;
+each data point is a vector, and $\theta_{ij}$ denotes the angle of rotation needed to align vectors $i$ and $j$ (these are the pairwise alignments).
+%
+From this large collection of pairwise alignments, angular synchronization then tries to find the alignment for each data point, such that these individual alignments are most consistent (in some metric) with the pairwise alignments.
+% (i.e., we would like the difference between the optimal rotations for two different concentration profiles to be as close as possible to the pairwise rotation needed to align the two profiles).
+%
+Figure~\ref{fig:schematics}d shows each vector from Figure~\ref{fig:schematics}c, now rotated so that the entire set is aligned.
+%-- these are the angles we would like to recover from the pairwise information. 
+%
+Note that the differences between the rotation angles are consistent with the pairwise angles needed to rotate one vector to another, i.e. $\theta_i - \theta_j \approx \theta_{ij}$.
+%
+We formulate this task as an optimization problem, and then relax it to an eigenproblem.
+%
+The top eigenvector(s) of a matrix (which contains information about the pairwise alignments) give the alignment for each data point.
+%
+%Because the method considers {\em all} pairwise alignments, angular synchronization is often robust to noise.
+
+Figure \ref{fig:1d_demo}d shows the concentration profiles from Figure~\ref{fig:1d_demo}c, now aligned using angular synchronization.
+%
+The peaks are now aligned, even though we input no information about the existence or location of these peaks into our algorithm.
+%
+We would like to note that, because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered up to a global rotation.
 
@@
 
+After registering the concentration profiles, we want to order them in time to reconstuct the developmental dynamics. 
+%
 We will use diffusion maps (DMAPS) \cite{coifman2005geometric} to temporally order our data.
 %
-Although our data are high-dimensional (for example, we will work with 2-color images of 10000 pixels, and so our data are 20000-dimensional vectors, where each dimension describes the intensity of one color in one pixel), we will assume that they lie on a one-dimensional, but perhaps nonlinear, curve.
-%
-We further assume that this curve is monotonic in time.
-%
-Therefore, our goal is to {\em automatically} uncover a parameterization of this curve, so that we can sort/order our data along it, effectively ordering our data in time.
-%
-To parameterize this one-dimensional curve which lies in high-dimensional space, we assume that data points which are close in the high-dimensional space should also be close on the curve, 
-e.g, images which look similar will be close in the high-dimensional pixel space (they will have similar pixel intensities), and we would like these similar images to be close to each other in our trajectory.
+In general, diffusion maps is a dimensionality reduction technique, which aims to uncover low-dimensional, nonlinear structure in a high-dimensional data set. 
+%
+Figure~\ref{fig:schematics}a shows an example of a two-dimensional data set (the ``high-dimensional data'') which lies on a one-dimensional nonlinear curve (the low-dimensional structure).
+%
+Figure~\ref{fig:schematics}b shows the same data, now colored by the first diffusion maps coordinate.
+%
+Clearly, the color parameterizes the one-dimensional curve that our eye can visually detect.
+%
+We assume that our imaging data also falls on a one-dimensional, nonlinear curve, and that this curve is parameterized by time (e.g., in Figure~\ref{fig:schematics}b, we assume that the data points are ordered from youngest to oldest along the curve).
+%
+Therefore, ordering our data along this curve will allow us to order our data in time.
+%
+Because our data will be very high dimensional, we cannot do this by eye, and instead require an algorithm to automatically extract this one-dimensional structure.
+
+To uncover a parameterization of this curve, we assume that data points which are close in the high-dimensional space should also be close on the curve (for example, we would like images which look similar to be close to each other in our developmental trajectory).
 %
 %We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\em as close as possible} in the parameterization.
-
-As an illustration of how local neighbors are important for parameterizing a curve, Figure~\ref{subfig:dmaps_edges} shows a data set which lies on a one-dimensional, nonlinear curve in two dimensions.
-%
-The edges between the pairs of data points are colored according to the distance between the data points, so that close data points have a dark edge between them.
-%
-Visually, these dark edges are meaningful and informative of the structure of the data; they clearly delineate the one-dimensional curve in the data that our eye can extract.
-%
-Furthermore, the light edges are not informative and do not give any meaningful structure to the curve.
-
-The problem of finding a parameterization of the data which forces nearby points together can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a matrix which contains information about pairwise distances between the data points.
-%
-This procedure is called diffusion maps \cite{coifman2005geometric}; the formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).
-%
-The DMAPS parameterization of these are data are shown in Figure~\ref{subfig:dmaps_color}.
-%
-The parameterization agrees with what our eye can detect visually.
-%
-However, we will work with very high-dimensional data, rather than these two-dimensional data points; we therefore have no easy way to plot and visualize the data, and rather need DMAPS to automatically uncover such structure in the data.
-
-Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we would like to use DMAPS to order them in time.
-%
-We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances for DMAPS (in \eqref{eq:dmaps_W}).
-%
-Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non-trivial) DMAPS component, $\phi_2$. 
+%
+As an illustration of how local neighbors are important for parameterizing a curve, the edges between the pairs of data points in Figure~\ref{fig:schematics}a are colored according to the distance between the data points, so that close data points have a dark edge between them.
+%
+Visually, the dark edges are meaningful and delineate the one-dimensional curve.
+%
+Furthermore, the light edges are not informative and do not deptict any meaningful structure. 
+%
+We therefore want points which are close, i.e., connected by dark edges, to be close in our parameterization.
+%
+This problem can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a matrix which contains information about pairwise distances between the data points.
+%
+In general, this procedure is called diffusion maps \cite{coifman2005geometric}.%; the formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).
+
+Given our set of registered signals, shown in Figure \ref{fig:1d_demo}d, we can use diffusion maps to order them in time.
+%
+We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{fig:1d_demo}d) as our pairwise distances between data points.
+%
+Figure \ref{fig:1d_demo}e shows the data from Figure \ref{fig:1d_demo}d, now sorted by the value of the first (non-trivial) eigenvector, $\phi_2$. 
 %
@@
 %
-Although DMAPS can recover a good ordering for the data, it cannot tell us about the absolute time of the data; we therefore have no information about the speed of this trajectory.
-%
-Furthermore, the direction of the trajectory is also unknown; determining which ``end'' of the DMAPS trajectory corresponds to the young embryos, and which ``end'' corresponds to the old embryos is a task that needs to be done by hand.
+Although diffusion maps can order the data, it cannot tell us about the absolute time; we have no information about the speed of this trajectory.
+%
+Furthermore, the direction of the trajectory is also unknown; determining which end of the trajectory corresponds to the beginning of development is a task that needs to be done by hand.
 
@@
 
-In many cases, we are interested in both registering and ordering our data.
+We are often interested in both registering and ordering our data.
 %
@@
 %
-This technique is called vector diffusion maps (VDM) \cite{singer2012vector}.
+This technique is called vector diffusion maps \cite{singer2012vector}.
 %
@@
 %
-Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using VDM. 
-%
-The data has been ordered by $\psi_{1, 3}$, the first (non-trivial) VDM coordinate.
+Figure \ref{fig:1d_demo}f shows the result of simultaneously registering and ordering the data in Figure \ref{fig:1d_demo}c using vector diffusion maps. 
+%
+The data has been ordered by $\psi_{1, 3}$, the first (non-trivial) vector diffusion maps coordinate.
 
@@
 %
-Instead of having $100$-dimensional concentration vectors, we now work with $20000$-dimensional images ($100x100$ pixel images containing two color channels).
-%
-The two-dimensional images are registered with respect to translations and rotations 
+Instead of $100$-dimensional concentration vectors, we now work with $20000$-dimensional images ($100 \times 100$ pixel images with two color channels).
+%
+We register the images with respect to translations and rotations 
 (we convert the two-dimensional translations and rotations to three-dimensional rotations so that we can apply angular synchronization or vector diffusion maps). 
 %
-We used vector diffusion maps to align and order the images shown in Figure~\ref{fig:fluorescent_images}
-%
-The results are shown in Figure~\ref{subfig:images_vdm}.
-%
-The peaks are now aligned, and the intensity of dpERK grows as a function of time, as expected.
+We used vector diffusion maps to align and order the images shown in Figure~\ref{fig:images_unordered}.
+%
+The results are shown in Figure~\ref{fig:images_ordered}.
+%
+Both the red and green peaks are now aligned, and the intensity of the red peaks grows as a function of time.
 
@@
 %
-For instance, it is known that the peak in the distribution of one of the proteins (visualized by the green channel) specifies the ventralmost position in the embryo. 
-%
-Based on this, images translated to make the centers of the embryos coincide and rotated to place the peak of the green channel at the same angular position. 
-%
-At the same time, temporal ordering can be done based on the monotonic progress of cellularization. 
-%
-We can therefore evaluate the quality of our orderings obtained from VDM by comparing hand registered and ordering images.
-%
-These results are shown in Figure~\ref{subfig:1dprofiles_hand}; the results are visually consistent with the results in Figure~\ref{subfig:1dprofiles_vdm} obtained using VDM. 
-%
-Furthermore, we can quantify the accuracy of our orderings by computing the rank correlation coefficient between the ordering obtained form VDM, and the orderings obtained from cellularization.
-%
-The rank correlation coefficient is 0.8665, indicating that our ordering from VDM is consistent with the ordering obtained from cellularization.
+It is known that the peak in the distribution of Dorsal (visualized by the green channel) specifies the ventralmost position in the embryo, and that the nuclei (shown in gray) delineate the circumfrence of the embryo.
+%
+Based on this, images can be translated to make the centers of the embryos coincide and rotated to place the peak of the green channel at the same angular position. 
+%
+At the same time, temporal ordering can be done based on the monotonic progress of cellularization (Figure~\ref{fig:membrane_compare}a). 
+%
+We can evaluate the quality our automatic orderings by comparing with the orderings obtained from cellularization.
+%
+The rank correlation coefficient between the two orderings is 0.9149, indicating that our automatic ordering is consistent with the ordering obtained from cellularization.
+%
+We can also visually compare the two orderings by comparing the arrays of one-dimensional profiles (similar to the format in Figure~\ref{fig:1d_demo});
+these results are shown in Figure~\ref{fig:membrane_compare}.
+%
+The general trends are consistent:
+the green peaks are aligned, and the red peaks grow as a function of time. 
+%
+However, Figure~\ref{fig:membrane_compare}c, which was registered and ordered using our mathematical algorithms, shows a better clustering of the outer peaks in the top corners of the image.
+%
+%These peaks are more scrambled in Figure~\ref{fig:membrane_compare}b, which was analyzed using cellularization.
+%
+These peaks are known to arise later in development, and so it is reassuring that they are at the end of the trajectory.
+%
+Furthermore, the membranes stop growing towards the end of this trajectory, making the ordering of the late signals in Figure~\ref{fig:membrane_compare}b less accurate.
 
@@
 Although we illustrated our methods using a specific data set from a study of {\em Drosophila} embryogenesis, our framework is very general and requires little knowledge of the image features or the developmental dynamics.
+%
+To our knowledge, temporal ordering of {\em Drosophila} images has not been automated to the extent we present here.
+%
+Recently, machine learning algorithms have been used to sort {\em Drosophila} images into different developmental classes \cite{yuan2014automated};
+however, this has the disadvantages that the user must define the approrpiate classes, and that hand-labeled training data is required.
 
@@
 %
-The majority of these works temporally order the data either by solving a traveling salesman problem (TSP) on the data, or by constructing a minimum spanning tree (MST) of the data, 
-in the hopes that these structures (the path of the TSP, or the MST) characterize the majority of the structure within the data and provide an accurate picture of the dynamical progression.
+The majority of these works temporally order the data either by solving a traveling salesman problem on the data, or by constructing a minimum spanning tree of the data, 
+in the hopes that these structures (the path of the travelling salesman, or the minimum spanning tree) characterize the majority of the structure within the data and provide an accurate picture of the dynamical progression.
 %
@@
 %
-In general, diffusion maps have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{tenenbaum2000global} \cite{balasubramanian2002isomap}.
-%
-Furthermore, although we use only the first DMAPS component to order the data (we assume the data are one-dimensional), DMAPS is not limited to one-dimensional data and can be used to analyze data containing more complex structure.
-%
-For example, in data which contains several branches or several classes, the first few DMAPS components will delineate the branches or separate the classes.
+Diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003} \cite{tenenbaum2000global} \cite{Donoho2003} \cite{Roweis2000}.
+%
+In general, diffusion maps have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{tenenbaum2000global} \cite{balasubramanian2002isomap}; we therefore expect them to perform better for noisy data than the algorithms used in previous work.
+%
+Furthermore, although we use only the first diffusion maps coordinate to order the data (we assume the data are one-dimensional), diffusion mapss are not limited to one-dimensional data and can be used to analyze data containing more complex structure.
+%
+For example, in data which contains several branches or several classes, the first few coordinates will delineate the branches or separate the classes.
 
@@
 %
-In our more general setting of applications in developmental biology, ``good'' features may not be known {\em a priori}, and so we prefer a methodology which does not require this preprocessing step.
+In our more general setting of applications in developmental biology, appropriate features may not be known {\em a priori}.
 %
@@
 
-In all of these data-driven methods, the essential question is ``what data is necessary?'': we would like to know how much and what type of data is needed to accurately reconstruct the dynamics. 
-%
-For the methods we present here, we require a sufficient amount of data for the one-dimensional curve to be ``seen'' by DMAPS (i.e., in Figure ~\ref{fig:dmaps_schematic}, there are a sufficient number of data points so that our eye can see the curve; if there were only 4 data points, it would most likely be difficult for our eye to detect the one-dimensional structure).
-%
-We also require the data we collect to contain enough information/be ``rich enough'' so that the alignments and dynamics are apparent.
-%
-For example, in the data presented in Figure~\ref{fig:fluorescent_images}, the green Dorsal peaks delineate the position of the ventral axis in the embryo, and the intensity of the red dpERK peaks grow over time.
+In all of these data-driven methods, the essential question is how much and what type of data is needed to accurately reconstruct the dynamics. 
+%
+For the methods we present here, we require a sufficient amount of data for the one-dimensional curve to be ``seen'' by the algorithms (i.e., in Figure ~\ref{fig:schematics}a, there are a sufficient number of data points so that our eye can see the curve; if there were only 4 data points, it would most likely be difficult for our eye to detect the one-dimensional structure).
+%
+We also require the data we collect to contain enough information so that the alignments and dynamics are apparent.
+%
+For example, in the data presented in Figure~\ref{fig:images_unordered}, the green peaks delineate the position of the ventral axis in the embryo, and the intensity of the red peaks grow over time.
 %
@@
 %
-If we removed the green Dorsal peaks, and only used the red dpERK peaks in our analysis, the registration would not be as accurate.
-%
-Furthermore, if we only used the green Dorsal peaks, then there would be no dynamic information (the shape or intensity of Dorsal does not change as a function of time in the developmental stage we are considering), and so the temporal ordering would be inaccurate.
+If we removed the green signal, and only used the red signal in our analysis, the registration would not be as accurate.
+%
+Furthermore, if we only used the green signal, then there would be no dynamic information (the shape or intensity of Dorsal does not change as a function of time in the developmental stage we are considering), and the temporal ordering would be inaccurate.
 %
@@
 %
-In these cases, our methods allow for the rapid, automated analysis of complex imaging data sets to help uncover complex dynamics and structure.
+In these cases, our methods allow for the rapid, automated analysis of imaging data sets to help uncover complex dynamics and structure.
 
@@
 
-\section{Angular synchronization \cite{singer2011angular}}
+\section{Angular synchronization for rotations\cite{singer2011angular}}
 
@@
 %
-Our goal is to obtain estimates of $\theta_1, \dots, \theta_m$ so that we can align $x_1, \dots, x_m$. 
+Our goal is to obtain estimates of $\theta_1, \dots, \theta_m$.
 %
@@
 %
-For the one-dimensional profiles shown in Figure~\ref{fig:1d_demo}, to compute the optimal angles in \eqref{eq:opt_angle}, we exhaustively search over all 100 possible shifts/rotations of the signals.
+For the one-dimensional profiles shown in Figure~\ref{fig:1d_demo}, to compute the optimal angles in \eqref{eq:opt_angle}, we exhaustively search over all 100 possible rotations of the signals.
 
@@
 \begin{equation} \label{eq:H_to_R}
-H_{ij} = R(\theta_{ij})
+H_{ij} = R(\theta_{ij}).
 \end{equation}
@@
 %
-From \eqref{eq:H_low_rank}, we can see that we can (approximately) recover $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$ by computing the top block eigenvector of $H$, which we will denote $\hat{R}$.
+From \eqref{eq:H_low_rank}, we can see that the top block eigenvector of $H$, which we will denote $\hat{R}$, contains estimates of $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$.
 %
@@
 %
-Then the top block eigenvector of $H$, $\hat{R}$, can be written as
+Then,
 \begin{equation}
@@
 %
-Because $\hat{R}_i$ is not guaranteed to be a rotation matrix, we must project $\hat{R}_i$ onto the closest rotation matrix \cite{...}. 
-%
-Let $U_i$ and $V_i$ denote the left and right singular vectors, respectively, of $\hat{R}_i$.
-%
-Then
+However, $\hat{R}_i$ is not guaranteed to be a rotation matrix; the closet rotation matrix, $R_{i, est}$, is given by
 \begin{equation} \label{eq:R_est}
-R_{i, est} = U_i V_i^T
-\end{equation} 
-where $R_{i, est}$ is an orthonormal matrix and is our estimate of $R(\theta_i)$. 
-%
-Note that $R_{i, est}$ is not guaranteed to be a proper rotation (i.e.
-$det(R_{i, est} = \pm 1)$; because we are permitted to scale the eigenvectors of a matrix, we adjust the sign of $\phi_d$ so that (most of) the rotations are proper.
-%
-We can now estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and then register the signals by rotating signal $i$ by $-\theta_i$.
+R_{i, est} = U_i V_i^T, 
+\end{equation} 
+where $U_i$ and $V_i$ are the left and right singular vectors, respectively, of $\hat{R}_i$.
+%
+We adjust the sign of $\phi_1$ so that $det(R_{i, est}) = +1$, ensuring proper rotations.
+%
+We estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and register the signals by rotating signal $i$ by $-\theta_i$.
 
@@
 
-For diffusion maps, the first step is to construct a matrix $W \in \mathbb{R}^{m \times m}$ that contains information about which pairs of data points are ``close.''
+For diffusion maps, the first step is to construct a matrix $W \in \mathbb{R}^{m \times m}$ that contains information about which pairs of data points are ``close''; more specifically, we want $W_{ij}$ to be large if points $x_i$ and $x_j$ are close.
+%
 %For the above analysis, the key step was having a notion of ``closeness'' between data points so that we can construct the matrix $W$.
 %
-For general data in high-dimensional space, we typically use a Gaussian kernel to construct $W$,
+We typically use a Gaussian kernel, 
 \begin{equation} \label{eq:dmaps_W}
@@
 \end{equation}
-and $\epsilon$ is a characteristic distance between data points.
+where $\epsilon$ is a characteristic distance between data points.
 %
@@
 %
-$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph}); in practice, we choose $\epsilon$ to be the median of the pairwise distances between data points.
-
-We then want to find a coordinate $y$ such that $y(i)$ and $y(j)$ if $W_{ij}$ is large (i.e., $y$ preserves local information, and so points that are close in the original space are also close in the coordinate $y$).
-%
-We therefore want to solve the following optimization problem \cite{Belkin2003}
+$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph}); we take $\epsilon$ to be the median of the pairwise distances between data points.
+
+We want a coordinate $y$ such that $y(i)$ and $y(j)$ if $W_{ij}$ is large (i.e., $y$ preserves local information, and so points that are close in the original space are also close in the coordinate $y$).
+%
+We therefore solve the following optimization problem \cite{Belkin2003}
 \begin{equation} \label{eq:dmaps_opt_problem}
-\argmin_{y} \sum_{ij} W_{ij} (y(i) - y(j))
+\argmin_{y} \sum_{ij} W_{ij} (y(i) - y(j))^2
 \end{equation}
@@
 %
-Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors. 
-%
-Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector.
+%Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors. 
+%
+Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector; this is a trivial solution to \eqref{eq:dmaps_opt_problem}.
 %
@@
 %
-The next eigenvector, $\phi_2$, can be shown to be the solution to \eqref{eq:dmaps_opt_problem}, so that $\phi_2(j)$ gives the coordinate for data point $x_j$.
+The next eigenvector, $\phi_2$, is the (non-trivial) solution to \eqref{eq:dmaps_opt_problem}, so that $\phi_2(j)$ gives the coordinate for data point $x_j$.
 %
@@
 %
-The distance $d(x_i, x_j)$ used to compute $W_{ij}$ is the distance between samples {\em after} pairwise alignment (i.e. it is the minimium obtained in \eqref{eq:opt_pairwise}).
+The distance $d(x_i, x_j)$ used in $W_{ij}$ is the distance between samples {\em after} pairwise alignment (i.e. it is the minimium obtained in \eqref{eq:opt_pairwise}).
 
@@
 %
-If we assume that our data are one-dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, one can show that a parameterization of our data (i.e., the analog of $\phi_2$ from the diffusion maps case) is given by $\psi_{1,d+1}$.
+If we assume that our data are one-dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, the embedding coordinate for our data (i.e., the analog of $\phi_2$ from the diffusion maps case) is given by $\psi_{1,d+1}$.
 %
@@
 %
-However, the typical matrix representation for the group of two-dimensional translations and rotations (using the matrix representation for the affine group \cite{...}) does not satisfy this property.
-%
-Instead, we can (approximately) represent rotations and translations in two-dimensions using three-dimensional rotation matrices.
-%
-We do this by projecting the image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular}. %(see Figure \ref{fig:SO3_picture}).
-%
-Rotations in two-dimensions now correspond to rotations around one principal axis in three-dimensions, and translations in two-dimensions correspond (approximately) to rotations around the other two principal axes in three-dimensions.
-(Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$.
+However, the typical matrix representation for the group of two-dimensional translations and rotations \cite{...} does not satisfy this property. 
+%
+Instead, we (approximately) represent rotations and translations in two-dimensions using three-dimensional rotation matrices, by projecting the image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular}.
+%(see Figure \ref{fig:SO3_picture}).
+%
+Rotation of the image corresponds to rotation around one principal axis in three dimensional, and translation of the image corresponds (approximately) to rotations around the other two principal axes.
+%(Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$.
 However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.)
 %
-These rotation matrices are (by definition) orthogonal, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
-
-The first step in angular synchronization or vector diffusion maps is to compute the translations and rotations to optimally align each pair of data points. 
+%These rotation matrices are (by definition) orthogonal, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
+
+We first compute the translations and rotations to optimally align each pair of data points. 
 %
@@
 \end{matrix}
-} \|g(I_j, \theta, dx, dy) - I_i \|_F^2.
-\end{equation}
-where $\| \cdot \|_F$ denotes the Frobenius norm, $\Delta$ is the maximum number of pixels by which we are allowed to translate the image, and $g(I_j, \theta, dx, dy)$ is a function that first rotates the image $I_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction.
-%
-The image rotation is done with the \texttt{imrotate} function in Matlab, using nearest neighbor interpolation to estimate the pixel intensities, since there is no direct mapping of the pixels of the unrotated image to the pixels of the rotated image.
+} \|g(I_j, \theta, dx, dy) - I_i \|^2.
+\end{equation}
+TODO: fix norm
+where $\| \cdot \|$ denotes the Euclidean norm between the pixel intensities for each color, $\Delta$ is the maximum number of pixels by which we translate the image, and $g(I_j, \theta, dx, dy)$ is a function that first rotates the image $I_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction.
+%
+The image rotation is done with the \texttt{imrotate} function in Matlab, using nearest neighbor interpolation to estimate the pixel intensities after rotation.%, since there is no direct mapping of the pixels of the unrotated image to the pixels of the rotated image.
 %
@@
 %
-The translations are done by shifting the pixels with periodic boundary conditions.
+The translations are implemented by shifting the pixels with periodic boundary conditions.
 %
@@
 %
-Therefore, instead of using an optimization procedure to compute the solution to \eqref{eq:opt_pairwise}, we choose to discretize the search space and exhaustively search to estimate the solution.
-%
-We discritize the search space of rotations into 20 possible rotations ($d\theta  \in \{0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10 \}$), and 11 possible translations in both the $x$ and $y$ directions ($dx, dy \in \{-20, -16, -12, \dots, 12, 16, 20 \}$).
+Therefore, instead of using an optimization procedure, we discretize the search space and exhaustively search to estimate the solution.
+%
+We discritize the search space of rotations into 20 possible rotations %($d\theta  \in \{0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10 \}$), 
+and 11 possible translations in both the $x$ and $y$ directions. %($dx, dy \in \{-20, -16, -12, \dots, 12, 16, 20 \}$). 
 %
@@
 %
-Although this can be somewhat time intensive, it is not prohibitive for the size of data sets which we are considering, and can be trivially parallelized for larger data sets if necessary.
-%
-The solution to this search will not be the exact solution to \eqref{eq:opt_pairwise}, but it will (most likely) be a close approximation.
+Although this can be somewhat time intensive, it is not prohibitive for the data sets we consider, and can be trivially parallelized if necessary.
+%
+The solution will not be the exact solution to \eqref{eq:opt_pairwise}, but it will (most likely) be a close approximation.
 %
@@
 %
-In the space of three-dimensional rotations, the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$ correspond to
+We first compute the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$,
 \begin{equation} \label{eq:angle_relations}
@@
 %
-We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere in $\mathbb{R}^3$.
-%
-From here, we can write rotations around the three principal axes in terms of the three Euler angles
-\small
+We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere.
+%
+We write rotations around the three principal axes, $R^x(\alpha)$, $R^y(\beta)$, and $R^z(\gamma)$, in terms of the three Euler angles
+%\small
 \begin{equation}
@@
     0 & 0 & 1 
-    \end{bmatrix}
+    \end{bmatrix}.
 \end{aligned}
 \end{equation}
-\normalsize
-where $R^x(\alpha)$, $R^y(\beta)$, and $R^z(\gamma)$ correspond to rotations around the $x$, $y$, and $z$ axes, respectively.
-%
-We then write the total rotation $R_{ij} \in SO(3)$ as 
+%\normalsize
+%
+The total rotation, $R_{ij} \in SO(3)$, is 
 \begin{equation} \label{eq:total_rot}
@@
 %
-\eqref{eq:total_rot} corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).
-
-From the three-dimensional rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$,
+%\eqref{eq:total_rot} corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).
+
+From the rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$,
 we build the matrix $H$ in \eqref{eq:H_to_R} and compute the estimates $R_{i, est}$ in \eqref{eq:R_est}.
 %
-From the matrices $R_{1, est}, \dots, R_{m, est}$, we now wish to find the corresponding translations and rotations of the images.
-%
-Because we are permitted to adjust our rotations by a global rotations, we first multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
+From the matrices $R_{1, est}, \dots, R_{m, est}$, we then find the corresponding translations and rotations of the images.
+%
+We first multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
 %
@@
 %
-We then invert \eqref{eq:angle_relations} to compute the optimal translation and rotation for each image from the Euler angles.
+We then invert \eqref{eq:angle_relations} to compute the optimal translation and rotation for each image.
 %
 We round the optimal translations to the nearest integer to remove the need for interpolation when translating the images.
-
-\section{Rank correlation}
-
-We use Spearman's rank correlation coefficient as a metric to evaluate the quality of our orderings.
-%
-Given data $x_1, \dots, x_m$ and $y_1, \dots, y_m$ for which we would like to compute the rank correlation coefficient, we first compute the ranks $X_i$ and $Y_i$ of each data point $x_i$ and $y_i$ in their respective data sets.
-%
-The rank correlation coefficient, $\rho$, is then given by
-\begin{equation}
-\rho = \frac{\sum_{i=1}^m (X_i - \overline{X})(Y_i - \overline{Y})}{\sqrt{\sum_{i=1}^m (X_i - \overline{X})^2 \sum_{i=1}^m (Y_i - \overline{Y})^2}}
-\end{equation}
-where $\overline{X}$ and $\overline{Y}$ are the means of $X_1, \dots, X_m$ and $Y_1, \dots, Y_m$, respectively.
 
@@
 \begin{figure}
-\begin{subfigure}{0.25\textwidth}
-\includegraphics[width=\textwidth]{fish1}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.25\textwidth}
-\includegraphics[width=\textwidth]{fish2}
-\caption{}
-\end{subfigure}
+\includegraphics{figures_0329_mac_1}
 \caption{Schematic illustration of reconstructing dynamics from cross-sectional data.
@@
 \begin{figure}
-\includegraphics[width=0.5\textwidth]{unregistered_unordered_2d}
-\caption{Fluorescent images of {\em Drosophila} embryos. Each embryo has been stained from Dorsal (green) and dpERK (red).}
-\label{fig:fluorescent_images}
+\includegraphics{figures_0329_mac_2}
+\caption{(a) Fluorescent images of {\em Drosophila} embryos.
+Each embryo has been stained from Dorsal (green) and dpERK (red).
+(b) Selected image from data set.
+(c) Profile of Dorsal (green) and dpERK (red) around the circumfrence of the embryo in (b).}
+\label{fig:images_unordered}
 \end{figure}
 
+\begin{figure}
+\includegraphics{figures_0329_mac_3}
+\caption{Illustration of diffusion maps and angular synchronization (a) Data points (in blue) which lie on a one-dimensional nonlinear curve in two dimensions.
+Each pair of data points is connected by an edge, and the edge intensity is related to the distance between the points (close data points are connected by dark edges).
+(b) The data in (a), colored by the first (non-trivial) eigenvector from diffusion maps, $\phi_2$.
+The color parameterizes the one-dimensional curve.
+(c) Set of vectors, each in a different orientation.
+The angles to align one vector to another are indicated.
+(d) The vectors from (c), each rotated so that the set of vectors is aligned.
+Note that the angles are consistent with the pairwise alignments in (c).
+}
+\label{fig:schematics}
+\end{figure}
+
 \begin{figure*}
-\begin{subfigure}{0.17\textwidth}
-\includegraphics[width=\textwidth]{illustrate_toy}
-\caption{}
-\label{subfig:1d_image_example}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{unregistered_unordered_toy}
-\caption{}
-\label{subfig:1d_unaligned_unordered}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{registered_unordered_toy}
-\caption{}
-\label{subfig:1d_aligned_unordered}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{registered_ordered_toy}
-\caption{}
-\label{subfig:1d_aligned_ordered}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{registered_ordered_vdm_toy}
-\caption{}
-\label{subfig:1d_aligned_ordered_vdm}
-\end{subfigure}
+\includegraphics{figures_0329_mac_4}
 \caption{Illustration of methodologies using one-dimensional profiles.
@@
 Rotation of the ring corresponds to shifting the line left and right (with periodic boundary conditions).
-(b) Many concentration profiles, each of the form described in (a).
+(b) One-dimensional profile correspondsing to the profile in (a).
+This profile is similar to the profile in Figure~\ref{fig:images_unordered}(c).
+(c) Many concentration profiles, each of the form described in (a).
 Each row in the array corresponds to one image.
 The overall dynamics are a Gaussian bump growing in time, but this is not readily apparent from the data because the profiles are unregistered and unordered.
-(c) The concentration profiles in (b), now registered using angular synchronization.
-(d) The concentration profiles in (c), now temporally ordered using diffusion maps.
+(d) The concentration profiles in (b), now registered using angular synchronization.
+(e) The concentration profiles in (c), now temporally ordered using diffusion maps.
 The profiles are sorted according to the value of the first non-trivial DMAPS component, $\phi_2$.
-(e) The concentration profiles in (b), now simultaneously registered and temporally ordered using vector diffusion maps.
+(f) The concentration profiles in (b), now simultaneously registered and temporally ordered using vector diffusion maps.
 The profiles are sorted according to the value of the first non-trivial VDM component, $\psi_{1, 3}$.}
@@
 
+\begin{figure*}
+\includegraphics{figures_0329_mac_5}
+\caption{The images in Figure~\ref{fig:images_unordered}, aligned and temporally ordered using vector diffusion maps.
+The images are sorted according to the value of the first non-trivial vector diffusion maps component, $\psi_{1, 4}$.}
+\label{fig:images_ordered}
+\end{figure*}
+
 \begin{figure}
-\begin{subfigure}{0.25\textwidth}
-\includegraphics[width=\textwidth]{dmaps_schematic_edges}
-\caption{}
-\label{subfig:dmaps_edges}
-\end{subfigure}
-\begin{subfigure}{0.25\textwidth}
-\includegraphics[width=\textwidth]{dmaps_schematic_color}
-\caption{}
-\label{subfig:dmaps_color}
-\end{subfigure}
-\caption{Diffusion maps illustration.
-(a) Data points (in blue) which lie on a one-dimensional nonlinear curve in two dimensions.
-Each pair of data points is connected by an edge, and the edge intensity is related to the distance between the points (close data points are connected by dark edges).
-(b) The data in (a), colored by the first (non-trivial) DMAPS eigenvector, $\phi_2$.
-The color parameterizes the one-dimensional curve.
+\includegraphics{figures_0329_mac_6}
+\caption{(a) Images, stained for membranes (gray) and Dorsal (green) at different levels of development.
+(b) The one-dimensional concentration profiles, extracted, registered, and ordered by hand.
+The profiles are extracted using the nuclei positions in the image, registered using the location of the Dorsal peak, and ordered using cellularization.(c) The one-dimensional concentration profiles corresponding to the images in Figure~\ref{fig:images_ordered}.
 }
-\label{fig:dmaps_schematic}
+\label{fig:membrane_compare}
 \end{figure}
-
-\begin{figure*}
-\begin{minipage}{0.6\textwidth}
-\begin{subfigure}{\textwidth}
-\includegraphics[width=\textwidth]{registered_ordered_vdm_2d}
-\caption{}
-\label{subfig:images_vdm}
-\end{subfigure}
-\end{minipage}
-\begin{minipage}{0.3\textwidth}
-\begin{subfigure}{\textwidth}
-\includegraphics[width=\textwidth]{registered_ordered_1dfrom2d}
-\caption{}
-\label{subfig:1dprofiles_vdm}
-\end{subfigure}
-\begin{subfigure}{\textwidth}
-\includegraphics[width=\textwidth]{registered_ordered_hand}
-\caption{}
-\label{subfig:1dprofiles_hand}
-\end{subfigure}
-\end{minipage}
-\caption{(a) The images in Figure~\ref{fig:fluorescent_images}, aligned and temporally ordered using vector diffusion maps.
-The images are sorted according to the value of the first non-trivial VDM component, $\psi_{1, 4}$.
-(b) The one-dimensional concentration profiles corresponding to the images in (a).
-Each row of the array corresponds to one image.
-(c) The one-dimensional concentration profiles, extracted, registered, and ordered by hand.
-The profiles are extracted using the nuclei positions in the image, registered using the location of the Dorsal peak, and ordered using cellularization.
-}
-\end{figure*}
 

### 2014-03-31 image_analysis_paper/make_figures_small_dataset_color.m

@@ -249,7 +249,7 @@ end
 figure;
 imshow(uint8(data_1d))
 
-[~, I] = sort(mem_lengths);
+[~, I] = sort(mem_lengths, 'descend');
 figure;
 set(gcf, 'paperunits', 'centimeters')
 set(gcf, 'papersize', [4 4])
@@ -297,4 +297,38 @@ xlabel('position', 'fontsize', fontsize)
 saveas(gcf, sprintf('%s/fig2c', im_save_dir), 'epsc');
 
 
+%%
+load('Dl_dpERK.mat');
+dpERK = nERK;
+Dl = nDl;
+
+dpERK = flipud(dpERK);
+Dl = flipud(Dl);
+
+[m, n] = size(dpERK);
+
+data_1d = zeros(m, n, 3);
+data_1d(:,:,1) = dpERK;
+data_1d(:,:,2) = Dl;
+
+for i=1:2
+    data_1d(:,:,i) = data_1d(:,:,i) / max(max(data_1d(:,:,i))) * 256;
+end
+
+[~, shift_amt] = max(sum(Dl));
+data_1d = circshift(data_1d, [0 round(n/2)-shift_amt 0]);
+
+figure;
+imshow(uint8(data_1d))
+
+figure;
+set(gcf, 'paperunits', 'centimeters')
+set(gcf, 'papersize', [4 4])
+set(gcf, 'paperposition',[0 0 4 4])
+image(uint8(data_1d))
+set(gca, 'xtick', [])
+set(gca, 'ytick', [])
+xlabel('position', 'fontsize', fontsize)
+ylabel('sample', 'fontsize', fontsize)
+saveas(gcf, sprintf('%s/fig6c', im_save_dir), 'epsc');
 

### 2014-03-31 image_analysis_paper/pnas_draft.tex
updated paper; has figures throughtout text

@@
 \DeclareMathOperator*{\argmin}{arg\,min}
+
+\makeatletter
+\newcommand{\customlabel}[2]{%
+\protected@write \@auxout {}{\string \newlabel {#1}{{#2}{}}}}
+\makeatother
 
@@
 
-\author{Carmeline J. Dsilva\affil{1}{Departmnent of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08544}, 
+\author{Carmeline J. Dsilva\affil{1}{Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08544}, 
 Bomyi Lim\affil{1}{}, Stanislav Y.
-Shvartsman\affil{1}{} \affil{2}{Lewis--Seigler Instittue for Integrative Genomics, Princeton University, Princeton, NJ 08544}, \and Ioannis G.
+Shvartsman\affil{1}{} \affil{2}{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, NJ 08544}, \and Ioannis G.
 Kevrekidis\affil{1}{} \affil{3}{Program in Applied and Computational Mathematics, Princeton University, Princeton, NJ 08544}}
@@
 %
-In this case, temporal ordering can be accomplished by arranging the fish by size, which correlates with the progress of development, and image alignment is based on obvious morphological landmarks, such as positions of head and fins.
+In this case, temporal ordering can be accomplished by arranging the fish by size, which correlates with the progress of development, and image registration is based on obvious morphological landmarks, such as positions of head and fins.
 %
@@
 
+\begin{figure}[t]
+\includegraphics[width=8.4cm]{fig1}
+\caption{Schematic illustration of reconstructing dynamics from cross-sectional data.
+(a) Fish, each in a different orientation and a different stage of development.
+(b) Fish, now registered and temporally ordered.
+For this example, the registration and ordering is easy to do by hand because the data set is small and the developmental changes are easy to recognize.}
+\label{fig:fish}
+\customlabel{subfig:fish_unordered}{\ref{fig:fish}a}
+\customlabel{subfig:fish_ordered}{\ref{fig:fish}b}
+\end{figure}
+
 The fruit fly embryo is shaped like a rice grain, approximately $1/2$~mm long. 
 %
-By end of the $2^{nd}$ hour of development, sequential divisions of a fertilized egg generated a system where $\sim 6,000$ nuclei are uniformly arranged in under the common plasma membrane. 
-%
-During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclosing nuclei into different cells grow in length (Figure~\ref{fig:membrane_compare}a). 
+By end of the $2^{nd}$ hour of development, sequential divisions of a fertilized egg generate a system where $\sim 6,000$ nuclei are uniformly arranged in under the common plasma membrane. 
+%
+During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclose nuclei into different cells grow in length (Figure~\ref{subfig:membranes}). 
 %
@@
 %
-Images in Figure~\ref{fig:images_unordered} show optical sections through the short axis of the embryos. 
-%
-The embryos were fixed and stained with chemicals that visualize nuclei as well as the spatial distributions of two different proteins. 
-%
-Each of the images reveals a uniform arrangement of nuclei and nonuniform distribution of two proteins (green and red) involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.
-
-Each image in Figure~\ref{fig:images_unordered} comes from a different embryo, captured at a different orientation and at a different and unknown point in time. 
+Images in Figure~\ref{subfig:images_unordered} show optical sections through the short axis of the embryos. 
+%
+The embryos were fixed and stained with chemicals that visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red). 
+%
+Each of the images reveals a uniform arrangement of nuclei and nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.
+
+\begin{figure}[t]
+\includegraphics[width=8.4cm]{fig2}
+\caption{(a) Fluorescent images of {\em Drosophila} embryos.
+Each embryo has been stained from Dorsal (green) and dpERK (red).
+(b) Selected image from data set.
+(c) Profile of Dorsal (green) and dpERK (red) around the circumfrence of the embryo in (b).}
+\label{fig:images_unordered}
+\customlabel{subfig:images_unordered}{\ref{fig:images_unordered}a}
+\customlabel{subfig:one_image}{\ref{fig:images_unordered}b}
+\customlabel{subfig:image_profiles}{\ref{fig:images_unordered}c}
+\end{figure}
+
+Each image in Figure~\ref{subfig:images_unordered} comes from a different embryo, captured at a different orientation and at a different and unknown point in time. 
 %
@@
 %
-We will also show how both steps can be combined using methodology called vector diffusion maps \cite{singer2012vector}.
+We will also show how both steps can be combined using vector diffusion maps \cite{singer2012vector}.
 %
@@
 
-Our goal is to temporally order images, such as the images shown in Figure~\ref{fig:images_unordered}.
-%
-Our ordering algorithm will require distances between data points. 
-%
-However, because each image is in a different orientation, we need to first align or {\em register} the images.
-%
-Once all the images have been aligned, we can compute meaningful distances and temporally order the profiles to reconstruct the developmental trajectory.
+We would like to temporally order images, such as the images shown in Figure~\ref{subfig:images_unordered}.
+%
+We will use diffusion maps, a dimensionality reduction algorithm, to automatically order our data.
+%
+However, this algorithm requires distances between data points. 
+%
+Because each image is in a different orientation, we need to first align or {\em register} the images before computing distances.
+%
+Once all the images have been aligned, we can compute distances and use diffusion maps to temporally order the data.
 
@@
 %
-We construct concentration profiles defined on a ring, and we assume that each ring is in an arbitrairy orientation; an example is shown in Figure~\ref{fig:1d_demo}a.
-%
-Each concentration profile is a Gaussian (shown in Figure~\ref{fig:1d_demo}b), and the Gaussians increase in intensity as  a function of time.
+We construct concentration profiles defined on a ring, and we assume that each ring is in an arbitrary orientation; an example is shown in Figure~\ref{subfig:1d_example}.
+%
+Each concentration profile is a Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
 %
@@
 %
-Rotation of the ring in Figure~\ref{fig:1d_demo}a corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown below the ring. 
-%
-Figure \ref{fig:1d_demo}c shows all the concentration profiles in our data set. 
+Rotation of the ring in Figure~\ref{subfig:1d_example} corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown below the ring. 
+%
+Figure \ref{subfig:1d_unaligned_unordered} shows all the concentration profiles in our data set. 
 %
@@
 %
-Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian growing in time) are not apparant.
+Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian growing in time) are not readily apparent.
+
+\begin{figure}[t]
+%\includegraphics{figures_0329_mac_4}
+\includegraphics[width=8.4cm]{fig4}
+\caption{Illustration of methodologies using one-dimensional profiles.
+(a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
+Rotation of the ring corresponds to shifting the line left and right (with periodic boundary conditions).
+(b) One-dimensional profile correspondsing to the profile in (a).
+This profile is similar to the profile in Figure~\ref{fig:images_unordered}(c).
+(c) Many concentration profiles, each of the form described in (a).
+Each row in the array corresponds to one image.
+The overall dynamics are a Gaussian bump growing in time, but this is not readily apparent from the data because the profiles are unregistered and unordered.
+(d) The concentration profiles in (b), now registered using angular synchronization.
+(e) The concentration profiles in (c), now temporally ordered using diffusion maps.
+The profiles are sorted according to the value of the first non-trivial DMAPS component, $\phi_2$.
+(f) The concentration profiles in (b), now simultaneously registered and temporally ordered using vector diffusion maps.
+The profiles are sorted according to the value of the first non-trivial VDM component, $\psi_{1, 3}$.}
+\label{fig:1d_demo}
+\customlabel{subfig:1d_example}{\ref{fig:1d_demo}a}
+\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}b}
+\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}c}
+\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}d}
+\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}e}
+\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}f}
+\end{figure}
 
@@
 
-Our first step is to register the set of images or profiles.
+Our first task is to register the set of images or profiles.
 %
@@
 Angular synchronization first computes the pairwise alignments %(rotations for our one-dimensional profiles in Figure \ref{fig:1d_demo}c, or translations and rotations for the two-dimensional images in Figure \ref{fig:images_unordered})
-for every possible pair of data points.%, effectively using each image as a template for every other image.
-%
-An illustration is shown in Figure~\ref{fig:schematics}c;
+for every possible pair of data points. %, effectively using each image as a template for every other image.
+%
+An illustration is shown in Figure~\ref{subfig:synch1};
 each data point is a vector, and $\theta_{ij}$ denotes the angle of rotation needed to align vectors $i$ and $j$ (these are the pairwise alignments).
 %
-From this large collection of pairwise alignments, angular synchronization then tries to find the alignment for each data point, such that these individual alignments are most consistent (in some metric) with the pairwise alignments.
+From this large collection of pairwise alignments, angular synchronization then rotates each data point, such that the individual alignments are most consistent (in some metric) with the pairwise alignments.
 % (i.e., we would like the difference between the optimal rotations for two different concentration profiles to be as close as possible to the pairwise rotation needed to align the two profiles).
 %
-Figure~\ref{fig:schematics}d shows each vector from Figure~\ref{fig:schematics}c, now rotated so that the entire set is aligned.
+Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned.
 %-- these are the angles we would like to recover from the pairwise information. 
 %
-Note that the differences between the rotation angles are consistent with the pairwise angles needed to rotate one vector to another, i.e. $\theta_i - \theta_j \approx \theta_{ij}$.
-%
-We formulate this task as an optimization problem, and then relax it to an eigenproblem.
-%
-The top eigenvector(s) of a matrix (which contains information about the pairwise alignments) give the alignment for each data point.
+Note that the differences between the rotation angles are consistent with the pairwise angles needed to rotate one vector to another, i.e., $\theta_i - \theta_j \approx \theta_{ij}$.
+%
+To compute these alignments, we formulate the task as an optimization problem, and then relax it to an eigenproblem.
+%
+The top eigenvector(s) of a matrix (which contains information about the pairwise alignments) give the alignments for each data point.
 %
@@
 
-Figure \ref{fig:1d_demo}d shows the concentration profiles from Figure~\ref{fig:1d_demo}c, now aligned using angular synchronization.
+\begin{figure}[t]
+\includegraphics[width=8.4cm]{fig3}
+\caption{Illustration of diffusion maps and angular synchronization (a) Data points (in blue) which lie on a one-dimensional nonlinear curve in two dimensions.
+Each pair of data points is connected by an edge, and the edge intensity is related to the distance between the points (close data points are connected by dark edges).
+(b) The data in (a), colored by the first (non-trivial) eigenvector from diffusion maps, $\phi_2$.
+The color parameterizes the one-dimensional curve.
+(c) Set of vectors, each in a different orientation.
+The angles to align one vector to another are indicated.
+(d) The vectors from (c), each rotated so that the set of vectors is aligned.
+Note that the angles are consistent with the pairwise alignments in (c).
+}
+\label{fig:schematics}
+\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}a}
+\customlabel{subfig:dmaps_color}{\ref{fig:schematics}b}
+\customlabel{subfig:synch1}{\ref{fig:schematics}c}
+\customlabel{subfig:synch2}{\ref{fig:schematics}d}
+\end{figure}
+
+Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization.
 %
@@
 
-After registering the concentration profiles, we want to order them in time to reconstuct the developmental dynamics. 
+After registering the images or concentration profiles, we want to order them in time to reconstuct the developmental dynamics. 
 %
@@
 %
-Figure~\ref{fig:schematics}a shows an example of a two-dimensional data set (the ``high-dimensional data'') which lies on a one-dimensional nonlinear curve (the low-dimensional structure).
-%
-Figure~\ref{fig:schematics}b shows the same data, now colored by the first diffusion maps coordinate.
+Figure~\ref{subfig:dmaps_edges} shows an example of a two-dimensional data set (the ``high-dimensional data'') which lies on a one-dimensional nonlinear curve (the low-dimensional structure).
+%
+Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the first diffusion maps coordinate.
 %
@@
 %
-We assume that our imaging data also falls on a one-dimensional, nonlinear curve, and that this curve is parameterized by time (e.g., in Figure~\ref{fig:schematics}b, we assume that the data points are ordered from youngest to oldest along the curve).
-%
-Therefore, ordering our data along this curve will allow us to order our data in time.
+We assume that our imaging data also falls on a one-dimensional, nonlinear curve, and that this curve is parameterized by time (e.g., in Figure~\ref{subfig:dmaps_color}, we assume that the data points are ordered from youngest to oldest along the curve).
+%
+Therefore, ordering our data along this curve will effectively order our data in time.
 %
@@
 
-To uncover a parameterization of this curve, we assume that data points which are close in the high-dimensional space should also be close on the curve (for example, we would like images which look similar to be close to each other in our developmental trajectory).
+To uncover a parameterization of this curve, we assume that data points which are close in the high-dimensional space should also be close on the curve (we would like images which look similar to be close together in our developmental trajectory).
 %
@@
 %
-As an illustration of how local neighbors are important for parameterizing a curve, the edges between the pairs of data points in Figure~\ref{fig:schematics}a are colored according to the distance between the data points, so that close data points have a dark edge between them.
+As an illustration of how local neighbors are important for parameterizing a curve, the edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored according to the distance between the data points, so that close data points have a dark edge between them.
 %
@@
 %
-Furthermore, the light edges are not informative and do not deptict any meaningful structure. 
-%
-We therefore want points which are close, i.e., connected by dark edges, to be close in our parameterization.
+Furthermore, the light edges are not informative and do not depict any meaningful structure. 
+%
+We want points which are close in the high-dimensional space, i.e., connected by dark edges, to be close in our low-dimensional parameterization.
 %
@@
 
-Given our set of registered signals, shown in Figure \ref{fig:1d_demo}d, we can use diffusion maps to order them in time.
-%
-We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{fig:1d_demo}d) as our pairwise distances between data points.
-%
-Figure \ref{fig:1d_demo}e shows the data from Figure \ref{fig:1d_demo}d, now sorted by the value of the first (non-trivial) eigenvector, $\phi_2$. 
-%
-As expected, the intensity of the dpERK peaks grows as a function of time. 
-%
-Although diffusion maps can order the data, it cannot tell us about the absolute time; we have no information about the speed of this trajectory.
+Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we can use diffusion maps to order them in time.
+%
+We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances between data points.
+%
+Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non-trivial) eigenvector, $\phi_2$. 
+%
+As expected, the intensity of the peak grows as a function of time. 
+%
+Although diffusion maps can order the data, it cannot tell us about the speed of this trajectory.
 %
@@
 %
-Figure \ref{fig:1d_demo}f shows the result of simultaneously registering and ordering the data in Figure \ref{fig:1d_demo}c using vector diffusion maps. 
+Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps. 
 %
@@
 %
-We used vector diffusion maps to align and order the images shown in Figure~\ref{fig:images_unordered}.
-%
-The results are shown in Figure~\ref{fig:images_ordered}.
+We used vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_unordered};
+the results are shown in Figure~\ref{fig:images_ordered}.
 %
@@
 %
-At the same time, temporal ordering can be done based on the monotonic progress of cellularization (Figure~\ref{fig:membrane_compare}a). 
+At the same time, temporal ordering can be done based on the monotonic progress of cellularization (Figure~\ref{subfig:membranes}). 
 %
@@
 %
-We can also visually compare the two orderings by comparing the arrays of one-dimensional profiles (similar to the format in Figure~\ref{fig:1d_demo});
-these results are shown in Figure~\ref{fig:membrane_compare}.
+We can also visually compare the alignments and orderings by comparing the arrays of one-dimensional profiles, similar to the format in Figure~\ref{fig:1d_demo};
+these results are shown in Figures~\ref{subfig:1d_membranes}~and~\ref{subfig:1d_vdm}.
 %
@@
 %
-However, Figure~\ref{fig:membrane_compare}c, which was registered and ordered using our mathematical algorithms, shows a better clustering of the outer peaks in the top corners of the image.
+However, Figure~\ref{subfig:1d_vdm}, which was registered and ordered using our mathematical algorithms, shows a better clustering of the outer peaks in the top corners of the image.
 %
@@
 %
-Furthermore, the membranes stop growing towards the end of this trajectory, making the ordering of the late signals in Figure~\ref{fig:membrane_compare}b less accurate.
+Furthermore, the membranes stop growing towards the end of this trajectory, making the ordering of the late signals in Figure~\ref{subfig:1d_membranes} less accurate.
 
@@
 
+\begin{figure}[t]
+\includegraphics[width=8.4cm]{fig5}
+\caption{The images in Figure~\ref{fig:images_unordered}, aligned and temporally ordered using vector diffusion maps.
+The images are sorted according to the value of the first non-trivial vector diffusion maps component, $\psi_{1, 4}$.}
+\label{fig:images_ordered}
+\end{figure}
+
+\begin{figure}[t]
+%\includegraphics{figures_0329_mac_6}
+\includegraphics[width=8.4cm]{fig6}
+\caption{(a) Images, stained for membranes (gray) and Dorsal (green) at different levels of development.
+(b) Membrane length as a function of time.
+(c) Distribution of embryo ages in the data set.
+(d) The one-dimensional concentration profiles, extracted, registered, and ordered by hand.
+The profiles are extracted using the nuclei positions in the image, registered using the location of the Dorsal peak, and ordered using cellularization.
+(e) The one-dimensional concentration profiles corresponding to the images in Figure~\ref{fig:images_ordered}.
+}
+\label{fig:membrane_compare}
+\customlabel{subfig:membranes}{\ref{fig:membrane_compare}a}
+\customlabel{subfig:membrane_curve}{\ref{fig:membrane_compare}b}
+\customlabel{subfig:membrane_dist}{\ref{fig:membrane_compare}c}
+\customlabel{subfig:1d_membranes}{\ref{fig:membrane_compare}d}
+\customlabel{subfig:1d_vdm}{\ref{fig:membrane_compare}e}
+\end{figure}
+
 \section{Conclusions}
@@
 %
-To our knowledge, temporal ordering of {\em Drosophila} images has not been automated to the extent we present here.
-%
-Recently, machine learning algorithms have been used to sort {\em Drosophila} images into different developmental classes \cite{yuan2014automated};
-however, this has the disadvantages that the user must define the approrpiate classes, and that hand-labeled training data is required.
+To our knowledge, temporal ordering of {\em Drosophila} images has not been automated to this extent.
+%
+Recently, machine learning algorithms have been used to group {\em Drosophila} images into different developmental classes \cite{yuan2014automated}.
+%
+However, these methods have the disadvantages that the user must define the appropriate classes, and that hand-labeled training data is required.
 
@@
 %
-In general, diffusion maps have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{tenenbaum2000global} \cite{balasubramanian2002isomap}; we therefore expect them to perform better for noisy data than the algorithms used in previous work.
-%
-Furthermore, although we use only the first diffusion maps coordinate to order the data (we assume the data are one-dimensional), diffusion mapss are not limited to one-dimensional data and can be used to analyze data containing more complex structure.
-%
-For example, in data which contains several branches or several classes, the first few coordinates will delineate the branches or separate the classes.
+In general, they are more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}; we therefore expect them to perform better than the ordering algorithms used in previous work.
+%
+Furthermore, although we use only the first diffusion maps coordinate to order the data (we assume the data are one-dimensional), diffusion maps are not limited to one-dimensional data and can be used to extract more complex structures.
+%
+For example, in data which contains several branches or classes, the first few coordinates will delineate the branches or separate the classes.
 
@@
 %
-In our more general setting of applications in developmental biology, appropriate features may not be known {\em a priori}.
-%
-The registration algorithm we use, angular synchronization/vector diffusion maps, has been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}.
-%
-However, in this application, each image is a picture of (essentially) the {\em same} molecule; variation in the data comes from differences in viewing angle (rotations of the molecule) and from instrument/measurement noise.
-%
-In contrast, our data contains variation from the different embryo orientations, as well as the temporal variation of the underling developmental dynamics.
+In our developmental biology applications, appropriate features may not be known {\em a priori}, and so we require a registration algorithm that does not rely on predefined features.
+%
+Angular synchronization and vector diffusion maps have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}.
+%
+However, in this application, each image is a picture of (essentially) the {\em same} molecule; variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
+%
+In contrast, our data contains variation from the different embryo orientations, as well as variation from the underling developmental dynamics.
 
@@
 %
-For the methods we present here, we require a sufficient amount of data for the one-dimensional curve to be ``seen'' by the algorithms (i.e., in Figure ~\ref{fig:schematics}a, there are a sufficient number of data points so that our eye can see the curve; if there were only 4 data points, it would most likely be difficult for our eye to detect the one-dimensional structure).
+For the methods we present here, we require a sufficient amount of data for the one-dimensional curve to be ``seen'' by the algorithms (i.e., in Figure~\ref{subfig:dmaps_edges}, there are a sufficient number of data points so that our eye can see the curve; if there were only 4 data points, it would most likely be difficult for our eye to detect the one-dimensional structure).
 %
@@
 %
-For example, in the data presented in Figure~\ref{fig:images_unordered}, the green peaks delineate the position of the ventral axis in the embryo, and the intensity of the red peaks grow over time.
+For example, in the data presented in Figure~\ref{subfig:images_unordered}, the green peaks delineate the position of the ventral axis in the embryo, and the intensity of the red peaks grow over time.
 %
@@
 
-For diffusion maps, the first step is to construct a matrix $W \in \mathbb{R}^{m \times m}$ that contains information about which pairs of data points are ``close''; more specifically, we want $W_{ij}$ to be large if points $x_i$ and $x_j$ are close.
+For diffusion maps, the first step is to construct a matrix $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if points $x_i$ and $x_j$ are ``close.'' 
 %
@@
 \begin{equation} \label{eq:dmaps_W}
-W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right)
+%W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right)
+W_{ij} = e^{ -\frac{d^2(x_i, x_j)}{\epsilon^2}}
 \end{equation}
@@
 %
-Therefore, points closer than $\epsilon$ apart are considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
+Therefore, points less than $\epsilon$ apart are considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
 %
@@
 
-We want a coordinate $y$ such that $y(i)$ and $y(j)$ if $W_{ij}$ is large (i.e., $y$ preserves local information, and so points that are close in the original space are also close in the coordinate $y$).
-%
-We therefore solve the following optimization problem \cite{Belkin2003}
+We want to find a coordinate $y$ such that $y(i)$ and $y(j)$ if $W_{ij}$ is large (i.e., $y$ preserves local information, and so points that are close in the original space are also close in the coordinate $y$).
+%
+We want solve the following optimization problem \cite{Belkin2003}
 \begin{equation} \label{eq:dmaps_opt_problem}
@@
 %Therefore, $\phi_2$ provides a one-dimensional parameterization of the data, and ordering the data points by the value of $\phi_2$ will order them in time.
-
+%
 To solve \eqref{eq:dmaps_opt_problem}, we compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{m} W_{ij}$, and the matrix $A$, where
@@
 %
-In our setup, we assume that this one dimension, parameterized by $\phi_2$, is correlated with time.
-%
-Therefore, ordering our data by $\phi_2(j)$ will allow us to automatically order our data in time.
+In our applicatins, we assume that this one dimension, parameterized by $\phi_2$, is correlated with time.
+%
+Therefore, ordering our data by $\phi_2(j)$ will effectively order our data in time.
 
@@
 
-For vector diffusion maps, we construct the matrix $S \in \mathbb{R}^{md \times md}$, with
+For vector diffusion maps, we first construct the matrix $S \in \mathbb{R}^{md \times md}$, with
 \begin{equation}
@@
 %
-The distance $d(x_i, x_j)$ used in $W_{ij}$ is the distance between samples {\em after} pairwise alignment (i.e. it is the minimium obtained in \eqref{eq:opt_pairwise}).
+The distance $d(x_i, x_j)$ used in $W_{ij}$ is the distance between samples {\em after} pairwise alignment (i.e., it is the minimium obtained in \eqref{eq:opt_pairwise}).
 
@@
 %
-Again, the top (block) eigenvector of $S$ contains approximations of the optimal rotations for each image, and the optimal rotations can be computed as described for angular synchronization.
+As in angular synchronization, the top (block) eigenvector of $S$ contains approximations of the optimal rotations, and the optimal rotations can be computed in the same way from \eqref{eq:R_est}.
 %
@@
 %
-If no such claims can be made, then one often runs a second step of diffusion maps, using $\psi_{k,l}$ as the coordinates, in order to uncover the true manifold modulo symmetries.
+If no such claims can be made, then one often runs a second step of diffusion maps, using $\psi_{k,l}$ as the coordinates, to uncover a parameterization of the manifold modulo symmetries.
 
@@
 
-We first compute the translations and rotations to optimally align each pair of data points. 
+To register the images, we first compute the translations and rotations to optimally align each pair of data points. 
 %
@@
 (\theta_{ij}, dx_{ij}, dy_{ij}) = \argmin_{
-\begin{matrix}
+\scriptsize \begin{matrix}
 \theta \in [0, 2\pi) \\ 
@@
 \end{equation}
-TODO: fix norm
-where $\| \cdot \|$ denotes the Euclidean norm between the pixel intensities for each color, $\Delta$ is the maximum number of pixels by which we translate the image, and $g(I_j, \theta, dx, dy)$ is a function that first rotates the image $I_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction.
+where $\Delta$ is the maximum number of pixels by which we translate the image, and $g(I_j, \theta, dx, dy)$ is a function that first rotates the image $I_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction.
+%
+The norm we use, $\| \cdot \|$, is the Euclidean norm between the pixel intensities for both the red and green channels.
 %
@@
 	R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})	 = R^z(\gamma_{ij})  R^y(\beta_{ij})  R^x(\alpha_{ij})
+%R(\alpha_{ij}, \beta_{ij}, \gamma_{ij}) = 
+%\begin{bmatrix}
+%\cos(\beta) \cos(\gamma) & \cos(\gamma)\sin(\alpha)\sin(\beta)-\cos(\alpha) \sin(\gamma) & \cos(\alpha)\cos(\gamma)\sin(\beta)+ \sin(\alpha)\sin(\gamma) \\
+%\cos(\beta) \sin(\gamma) & \cos(\alpha) \cos(\gamma)+\sin(\alpha) \sin(\beta) \sin(\gamma) & \cos(\alpha) \sin(\beta) \sin(\gamma)-\cos(\gamma) \sin(\alpha) \\
+%-\sin(\beta) & \cos(\beta) \sin(\alpha) & \cos(\alpha) \cos(\beta))
+%\end{bmatrix}
 \end{equation}
@@
 %% \end{table*}
-
-\begin{figure}
-\includegraphics{figures_0329_mac_1}
-\caption{Schematic illustration of reconstructing dynamics from cross-sectional data.
-(a) Fish, each in a different orientation and a different stage of development.
-(b) Fish, now registered and temporally ordered.
-For this example, the registration and ordering is easy to do by hand because the data set is small and the developmental changes are easy to recognize.}
-\label{fig:fish}
-\end{figure}
-
-\begin{figure}
-\includegraphics{figures_0329_mac_2}
-\caption{(a) Fluorescent images of {\em Drosophila} embryos.
-Each embryo has been stained from Dorsal (green) and dpERK (red).
-(b) Selected image from data set.
-(c) Profile of Dorsal (green) and dpERK (red) around the circumfrence of the embryo in (b).}
-\label{fig:images_unordered}
-\end{figure}
-
-\begin{figure}
-\includegraphics{figures_0329_mac_3}
-\caption{Illustration of diffusion maps and angular synchronization (a) Data points (in blue) which lie on a one-dimensional nonlinear curve in two dimensions.
-Each pair of data points is connected by an edge, and the edge intensity is related to the distance between the points (close data points are connected by dark edges).
-(b) The data in (a), colored by the first (non-trivial) eigenvector from diffusion maps, $\phi_2$.
-The color parameterizes the one-dimensional curve.
-(c) Set of vectors, each in a different orientation.
-The angles to align one vector to another are indicated.
-(d) The vectors from (c), each rotated so that the set of vectors is aligned.
-Note that the angles are consistent with the pairwise alignments in (c).
-}
-\label{fig:schematics}
-\end{figure}
-
-\begin{figure*}
-\includegraphics{figures_0329_mac_4}
-\caption{Illustration of methodologies using one-dimensional profiles.
-(a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
-Rotation of the ring corresponds to shifting the line left and right (with periodic boundary conditions).
-(b) One-dimensional profile correspondsing to the profile in (a).
-This profile is similar to the profile in Figure~\ref{fig:images_unordered}(c).
-(c) Many concentration profiles, each of the form described in (a).
-Each row in the array corresponds to one image.
-The overall dynamics are a Gaussian bump growing in time, but this is not readily apparent from the data because the profiles are unregistered and unordered.
-(d) The concentration profiles in (b), now registered using angular synchronization.
-(e) The concentration profiles in (c), now temporally ordered using diffusion maps.
-The profiles are sorted according to the value of the first non-trivial DMAPS component, $\phi_2$.
-(f) The concentration profiles in (b), now simultaneously registered and temporally ordered using vector diffusion maps.
-The profiles are sorted according to the value of the first non-trivial VDM component, $\psi_{1, 3}$.}
-\label{fig:1d_demo}
-\end{figure*}
-
-\begin{figure*}
-\includegraphics{figures_0329_mac_5}
-\caption{The images in Figure~\ref{fig:images_unordered}, aligned and temporally ordered using vector diffusion maps.
-The images are sorted according to the value of the first non-trivial vector diffusion maps component, $\psi_{1, 4}$.}
-\label{fig:images_ordered}
-\end{figure*}
-
-\begin{figure}
-\includegraphics{figures_0329_mac_6}
-\caption{(a) Images, stained for membranes (gray) and Dorsal (green) at different levels of development.
-(b) The one-dimensional concentration profiles, extracted, registered, and ordered by hand.
-The profiles are extracted using the nuclei positions in the image, registered using the location of the Dorsal peak, and ordered using cellularization.(c) The one-dimensional concentration profiles corresponding to the images in Figure~\ref{fig:images_ordered}.
-}
-\label{fig:membrane_compare}
-\end{figure}
 

### 2014-04-01 image_analysis_paper/pnas_draft.tex
Added stas comments to paper

@@
 We present an automatic way to register and temporally order cross--sectional data sets of images. 
-We use relatively new mathematical techniques (angular synchronization for image registration, and diffusion maps for temporal ordering) which are applicable to a wide variety of data sets and require little {\em a priori} knowledge of the image features or the developmental dynamics.
+%
+We use mathematical techniques (angular synchronization for image registration, and diffusion maps for temporal ordering) which are applicable to a wide variety of data sets and require little {\em a priori} knowledge of the image features or the developmental dynamics.
 %
@@
 For this example, the registration and ordering is easy to do by hand because the data set is small and the developmental changes are easy to recognize.}
-\label{fig:fish}
+%\label{fig:fish}
+\customlabel{fig:fish}{1}
 \customlabel{subfig:fish_unordered}{\ref{fig:fish}a}
@@
 %
-Concurrently with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs. 
+Concurrently with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs \cite{lim2013kinetics}.
 %
@@
 %
-The embryos were fixed and stained with chemicals that visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red). 
+The embryos were fixed and stained with chemicals that visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red) \cite{chung2010microfluidic}.
 %
@@
 (c) Profile of Dorsal (green) and dpERK (red) around the circumfrence of the embryo in (b).}
-\label{fig:images_unordered}
+%\label{fig:images_unordered}
+\customlabel{fig:images_unordered}{2}
 \customlabel{subfig:images_unordered}{\ref{fig:images_unordered}a}
@@
 %
-The paper is organized as follows. 
-%
-First, we use a synthetic dataset, where the correct orientation and ordering of data points is known, to formalize the tasks of image registration and temporal ordering. 
-%
-Then, we show that both of these tasks can be effectively combined in a single step. 
-%
-Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align them and order them correctly in time. 
-%
-We conclude by discussing the general requirements that must be satisfied for our approach to work and compare it to other approaches and ideas in cross-sectional studies of biological dynamics.
+%The paper is organized as follows. 
+%%
+%First, we use a synthetic dataset, where the correct orientation and ordering of data points is known, to formalize the tasks of image registration and temporal ordering. 
+%%
+%Then, we show that both of these tasks can be effectively combined in a single step. 
+%%
+%Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align them and order them correctly in time. 
+%%
+%We conclude by discussing the general requirements that must be satisfied for our approach to work and compare it to other approaches and ideas in cross-sectional studies of biological dynamics.
 
@@
 
-We would like to temporally order images, such as the images shown in Figure~\ref{subfig:images_unordered}.
-%
-We will use diffusion maps, a dimensionality reduction algorithm, to automatically order our data.
-%
-However, this algorithm requires distances between data points. 
-%
-Because each image is in a different orientation, we need to first align or {\em register} the images before computing distances.
-%
-Once all the images have been aligned, we can compute distances and use diffusion maps to temporally order the data.
+%We would like to temporally order images, such as the images shown in Figure~\ref{subfig:images_unordered}.
+%%
+%We will use diffusion maps, a dimensionality reduction algorithm, to automatically order our data.
+%%
+%However, this algorithm requires distances between data points. 
+%%
+%Because each image is in a different orientation, we need to first align or {\em register} the images before computing distances.
+%%
+%Once all the images have been aligned, we can compute distances and use diffusion maps to temporally order the data.
 
@@
 %
-Each concentration profile is a Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
+Each concentration profile is a noisy Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
 %
@@
 (a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
-Rotation of the ring corresponds to shifting the line left and right (with periodic boundary conditions).
-(b) One-dimensional profile correspondsing to the profile in (a).
-This profile is similar to the profile in Figure~\ref{fig:images_unordered}(c).
+(b) Intensity corresponding to the profile in (a).
 (c) Many concentration profiles, each of the form described in (a).
 Each row in the array corresponds to one image.
-The overall dynamics are a Gaussian bump growing in time, but this is not readily apparent from the data because the profiles are unregistered and unordered.
 (d) The concentration profiles in (b), now registered using angular synchronization.
 (e) The concentration profiles in (c), now temporally ordered using diffusion maps.
-The profiles are sorted according to the value of the first non-trivial DMAPS component, $\phi_2$.
-(f) The concentration profiles in (b), now simultaneously registered and temporally ordered using vector diffusion maps.
-The profiles are sorted according to the value of the first non-trivial VDM component, $\psi_{1, 3}$.}
-\label{fig:1d_demo}
+(f) The concentration profiles in (b), now simultaneously registered and temporally ordered using vector diffusion maps.}
+%\label{fig:1d_demo}
+\customlabel{fig:1d_demo}{3}
 \customlabel{subfig:1d_example}{\ref{fig:1d_demo}a}
@@
 %
-We will use a relatively new technique called angular synchronization\cite{singer2011angular}, which is robust to noise and requires no {\em a priori} knowledge of image features.
+We will use angular synchronization\cite{singer2011angular}, a technique which is robust to noise and requires no {\em a priori} knowledge of image features.
 %
@@
 }
-\label{fig:schematics}
+%\label{fig:schematics}
+\customlabel{fig:schematics}{4}
 \customlabel{subfig:dmaps_edges}{\ref{fig:schematics}a}
@@
 %
-The peaks are now aligned, even though we input no information about the existence or location of these peaks into our algorithm.
-%
-We would like to note that, because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered up to a global rotation.
+The peaks are now aligned, even though our algorithm use no information about the existence or location of these peaks.
+%
+Because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered up to a global rotation.
 
@@
 %
-We will use diffusion maps (DMAPS) \cite{coifman2005geometric} to temporally order our data.
-%
-In general, diffusion maps is a dimensionality reduction technique, which aims to uncover low-dimensional, nonlinear structure in a high-dimensional data set. 
+We will use diffusion maps (DMAPS) to temporally order our data.
+%
+In general, diffusion maps is a dimensionality reduction technique, which aims to uncover low-dimensional, nonlinear structure in a high-dimensional data set \cite{coifman2005geometric}. 
 %
@@
 %
-Clearly, the color parameterizes the one-dimensional curve that our eye can visually detect.
+The color parameterizes the one-dimensional curve that our eye can visually detect.
 %
@@
 %
-Therefore, ordering our data along this curve will effectively order our data in time.
+Therefore, ordering data along this curve will effectively order data in time.
 %
@@
 %
-As an illustration of how local neighbors are important for parameterizing a curve, the edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored according to the distance between the data points, so that close data points have a dark edge between them.
+To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored according to the distance between the data points, so that close data points have a dark edge between them.
 %
@@
 %
-Although diffusion maps can order the data, it cannot tell us about the speed of this trajectory.
-%
-Furthermore, the direction of the trajectory is also unknown; determining which end of the trajectory corresponds to the beginning of development is a task that needs to be done by hand.
+%Although diffusion maps can order the data, it cannot tell us about the speed of this trajectory.
+%
+%Furthermore, the direction of the trajectory is also unknown; determining which end of the trajectory corresponds to the beginning of development is a task that needs to be done by hand.
 
@@
 
-We are often interested in both registering and ordering our data.
+%We are often interested in both registering and ordering our data.
 %
@@
 
-In these {\em Drosophila} experiments, both registration and temporal ordering can be done by hand using prior knowledge about the system. 
-%
-It is known that the peak in the distribution of Dorsal (visualized by the green channel) specifies the ventralmost position in the embryo, and that the nuclei (shown in gray) delineate the circumfrence of the embryo.
-%
-Based on this, images can be translated to make the centers of the embryos coincide and rotated to place the peak of the green channel at the same angular position. 
-%
-At the same time, temporal ordering can be done based on the monotonic progress of cellularization (Figure~\ref{subfig:membranes}). 
+In these experiments, we can evaluate the quality of our automatic registration and ordering using prior knowledge about the system.
+%
+It is known that the peak in the distribution of Dorsal (visualized by the green channel) specifies the ventralmost position in the embryo, and that the nuclei (shown in gray) delineate the circumference of the embryo.
+%
+%Based on this, images can be translated to make the centers of the embryos coincide and rotated to place the peak of the green channel at the same angular position. 
+%
+%At the same time, 
+Temporal ordering can be done based on the monotonic progress of cellularization (Figure~\ref{subfig:membranes}) \cite{figard2013plasma}. 
 %
@@
 %
-However, Figure~\ref{subfig:1d_vdm}, which was registered and ordered using our mathematical algorithms, shows a better clustering of the outer peaks in the top corners of the image.
+However, Figure~\ref{subfig:1d_vdm}, which was automatically registered and ordered, shows a better clustering of the outer peaks in the top corners of the image.
 %
@@
 %
-Furthermore, the membranes stop growing towards the end of this trajectory, making the ordering of the late signals in Figure~\ref{subfig:1d_membranes} less accurate.
+%Furthermore, the membranes stop growing towards the end of this trajectory, making the ordering of the late signals in Figure~\ref{subfig:1d_membranes} less accurate.
 
@@
 \includegraphics[width=8.4cm]{fig5}
-\caption{The images in Figure~\ref{fig:images_unordered}, aligned and temporally ordered using vector diffusion maps.
+\caption{The images in Figure~\ref{subfig:images_unordered}, aligned and temporally ordered using vector diffusion maps.
 The images are sorted according to the value of the first non-trivial vector diffusion maps component, $\psi_{1, 4}$.}
-\label{fig:images_ordered}
+%\label{fig:images_ordered}
+\customlabel{fig:images_ordered}{5}
 \end{figure}
@@
 (b) Membrane length as a function of time.
-(c) Distribution of embryo ages in the data set.
-(d) The one-dimensional concentration profiles, extracted, registered, and ordered by hand.
+(c) The one-dimensional concentration profiles, extracted, registered, and ordered by hand.
 The profiles are extracted using the nuclei positions in the image, registered using the location of the Dorsal peak, and ordered using cellularization.
-(e) The one-dimensional concentration profiles corresponding to the images in Figure~\ref{fig:images_ordered}.
+(d) The one-dimensional concentration profiles corresponding to the images in Figure~\ref{fig:images_ordered}.
 }
-\label{fig:membrane_compare}
+%\label{fig:membrane_compare}
+\customlabel{fig:membrane_compare}{6}
 \customlabel{subfig:membranes}{\ref{fig:membrane_compare}a}
 \customlabel{subfig:membrane_curve}{\ref{fig:membrane_compare}b}
-\customlabel{subfig:membrane_dist}{\ref{fig:membrane_compare}c}
-\customlabel{subfig:1d_membranes}{\ref{fig:membrane_compare}d}
-\customlabel{subfig:1d_vdm}{\ref{fig:membrane_compare}e}
+\customlabel{subfig:1d_membranes}{\ref{fig:membrane_compare}c}
+\customlabel{subfig:1d_vdm}{\ref{fig:membrane_compare}d}
 \end{figure}
@@
 
-We have presented mathematical techniques for registration and temporal ordering of images arising in studies of developmental biology.
+We have shown how dimensionality reduction techniques can greatly acceleration the registration and temporal ordering of cross-sectional imaging data.
 %
@@
 %
-Furthermore, although we use only the first diffusion maps coordinate to order the data (we assume the data are one-dimensional), diffusion maps are not limited to one-dimensional data and can be used to extract more complex structures.
-%
-For example, in data which contains several branches or classes, the first few coordinates will delineate the branches or separate the classes.
+%Furthermore, although we use only the first diffusion maps coordinate to order the data (we assume the data are one-dimensional), diffusion maps are not limited to one-dimensional data and can be used to extract more complex structures.
+%
+%For example, in data which contains several branches or classes, the first few coordinates will delineate the branches or separate the classes.
 
@@
 %
-In our developmental biology applications, appropriate features may not be known {\em a priori}, and so we require a registration algorithm that does not rely on predefined features.
+%In our developmental biology applications, 
+However, in general, appropriate features may not be known {\em a priori}, and so we require a registration algorithm that does not rely on predefined features.
 %
@@
 %
-We also require the data we collect to contain enough information so that the alignments and dynamics are apparent.
+We also require the data to contain enough information so that the alignments and dynamics are apparent.
 %
@@
 
-Rather than work with the angles $\theta_{ij}$ directly, it will be more convenient to work with the rotation matrices 
+Rather than work with the angles $\theta_{ij}$ directly, it is more convenient to work with the rotation matrices 
 \begin{equation} \label{eq:R_theta}
@@
 %
-Successive rotations now correspond to multiplication of rotation matrices, i.e., 
-$R(\alpha_1 + \alpha_2) = R(\alpha_1) R(\alpha_2)$,
-and, due to the orthogonality of rotation matrices, $R(-\alpha) = R(\alpha)^T$.
+Successive rotations now correspond to multiplication of rotation matrices: $R(\alpha_1 + \alpha_2) = R(\alpha_1) R(\alpha_2)$.
+%
+Due to the orthogonality of rotation matrices, $R(-\alpha) = R(\alpha)^T$.
 
@@
 %
-However, $\hat{R}_i$ is not guaranteed to be a rotation matrix; the closet rotation matrix, $R_{i, est}$, is given by
+However, $\hat{R}_i$ is not guaranteed to be a rotation matrix; the closest rotation matrix, $R_{i, est}$, is given by
 \begin{equation} \label{eq:R_est}
@@
 %
-However, the typical matrix representation for the group of two-dimensional translations and rotations \cite{...} does not satisfy this property. 
+However, the typical matrix representation for the group of two-dimensional translations and rotations does not satisfy this property. 
 %
@@
 %
-Although this can be somewhat time intensive, it is not prohibitive for the data sets we consider, and can be trivially parallelized if necessary.
-%
-The solution will not be the exact solution to \eqref{eq:opt_pairwise}, but it will (most likely) be a close approximation.
-%
-Since our techniques are robust to noise, close approximations will be sufficient to obtain accurate results.
+This direct enumeration approach is not prohibitive for this step.
+%
+%Although this can be somewhat time intensive, it is not prohibitive for the data sets we consider, and can be trivially parallelized if necessary.
+%
+%The solution will not be the exact solution to \eqref{eq:opt_pairwise}, but it will (most likely) be a close approximation.
+%
+%Since our techniques are robust to noise, close approximations will be sufficient to obtain accurate results.
 
@@
 We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere.
+%
+This portion of the sphere is small enough so that curvature effects are not seen in the image.
 %

### 2014-04-04 image_analysis_paper/pnas_draft.tex
added henry's comments

@@
 
-\author{Carmeline J. Dsilva\affil{1}{Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08544}, 
-Bomyi Lim\affil{1}{}, Stanislav Y.
-Shvartsman\affil{1}{} \affil{2}{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, NJ 08544}, \and Ioannis G.
-Kevrekidis\affil{1}{} \affil{3}{Program in Applied and Computational Mathematics, Princeton University, Princeton, NJ 08544}}
+\author{Carmeline~J.~Dsilva\affil{1}{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
+Bomyi~Lim\affil{1}{}, 
+Thomas~J.~Levario\affil{2}{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
+Hang~Lu\affil{2}{},
+Stanislav~Y.~Shvartsman\affil{1}{} \affil{3}{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA}, 
+\and 
+Ioannis~G.~Kevrekidis\affil{1}{} \affil{4}{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA}}
 
@@
 %
-In a cross-sectional study, one embryo contributes only one snapshot of a chemical or morphological process along developmental trajectory which must be reconstructed from multiple snapshots of different embryos.
+In a cross-sectional study, one embryo contributes only one snapshot of a chemical or morphological process along a developmental trajectory, which must be reconstructed from multiple snapshots of different embryos.
 %
@@
 %
-Recent advances in large scale physical manipulation of embryos and their imaging produce rapidly increasing volumes of data from such experiments, in which every embryo is observed at a different orientation and at a different time point.
+Recent advances in large-scale physical manipulation and imaging of embryos have produced rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different orientation and time point.
 %
@@
 %
-Dealing with real data is far more nontrivial. 
+Dealing with real data is far more nontrivial, due to the number of images and similarities between them. 
 %
@@
 %
-This dataset comprises $81$ fluorescence microscopy images of different embryos fixed during the $3^{rd}$ hour of their development.
+This dataset comprises $49$ fluorescence microscopy images of different embryos fixed during the $3^{rd}$ hour of their development.
 
@@
 (a) Fish, each in a different orientation and a different stage of development.
-(b) Fish, now registered and temporally ordered.
+(b) Fish, now registered and temporally ordered by hand.
 For this example, the registration and ordering is easy to do by hand because the data set is small and the developmental changes are easy to recognize.}
@@
 \caption{(a) Fluorescent images of {\em Drosophila} embryos.
-Each embryo has been stained from Dorsal (green) and dpERK (red).
+Each embryo has been stained for Dorsal (green) and dpERK (red).
 (b) Selected image from data set.
@@
 %
-The concentration profiles have been stacked in an array, such that each row corresponds to one data point.
+The concentration profiles have been stacked in an array, such that each row corresponds to one profile.
 %
@@
 \begin{figure}[t]
-%\includegraphics{figures_0329_mac_4}
-\includegraphics[width=8.4cm]{fig4}
+\includegraphics[width=8.4cm]{fig3}
 \caption{Illustration of methodologies using one-dimensional profiles.
@@
 Each row in the array corresponds to one image.
-(d) The concentration profiles in (b), now registered using angular synchronization.
+(d) The concentration profiles in (c), now registered using angular synchronization.
 (e) The concentration profiles in (c), now temporally ordered using diffusion maps.
-(f) The concentration profiles in (b), now simultaneously registered and temporally ordered using vector diffusion maps.}
+(f) The concentration profiles in (c), now simultaneously registered and temporally ordered using vector diffusion maps.}
 %\label{fig:1d_demo}
@@
 \begin{figure}[t]
-\includegraphics[width=8.4cm]{fig3}
-\caption{Illustration of diffusion maps and angular synchronization (a) Data points (in blue) which lie on a one-dimensional nonlinear curve in two dimensions.
+\includegraphics[width=8.4cm]{fig4}
+\caption{(a) Set of vectors, each in a different orientation.
+The angles to align one vector to another are indicated.
+(b) The vectors from (a), each rotated so that the set of vectors is aligned.
+Note that the angles are consistent with the pairwise alignments in (a).
+(c) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
 Each pair of data points is connected by an edge, and the edge intensity is related to the distance between the points (close data points are connected by dark edges).
-(b) The data in (a), colored by the first (non-trivial) eigenvector from diffusion maps, $\phi_2$.
+(d) The data in (c), colored by the first (non-trivial) eigenvector from diffusion maps, $\phi_2$.
 The color parameterizes the one-dimensional curve.
-(c) Set of vectors, each in a different orientation.
-The angles to align one vector to another are indicated.
-(d) The vectors from (c), each rotated so that the set of vectors is aligned.
-Note that the angles are consistent with the pairwise alignments in (c).
 }
@@
 \begin{figure}[t]
-%\includegraphics{figures_0329_mac_6}
 \includegraphics[width=8.4cm]{fig6}
@@
 (b) Membrane length as a function of time.
-(c) The one-dimensional concentration profiles, extracted, registered, and ordered by hand.
-The profiles are extracted using the nuclei positions in the image, registered using the location of the Dorsal peak, and ordered using cellularization.
-(d) The one-dimensional concentration profiles corresponding to the images in Figure~\ref{fig:images_ordered}.
+(c) The one-dimensional concentration profiles, registered by peak Dorsal fluorescence and ordered by membrane length.
+(d) The one-dimensional concentration profiles corresponding to the automatically registered and ordered images in Figure~\ref{fig:images_ordered}.
 }
@@
 
-We note that, under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, 
-\begin{equation} 
-H_{ij} = R(\theta_{ij}) \approx R(\theta_i - \theta_j) = R(\theta_i) R(-\theta_j) = R(\theta_i) R(\theta_j)^T,
-\end{equation}
+We note that, under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, $H_{ij} \approx R(\theta_i) R(\theta_j)^T$,
+%\begin{equation} 
+%H_{ij} = R(\theta_{ij}) \approx R(\theta_i - \theta_j) = R(\theta_i) R(-\theta_j) = R(\theta_i) R(\theta_j)^T,
+%\end{equation}
  and so
@@
 %
-If we assume that our data are one-dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, the embedding coordinate for our data (i.e., the analog of $\phi_2$ from the diffusion maps case) is given by $\psi_{1,d+1}$.
+If we assume that our data are one-dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, the embedding coordinate for our data %(i.e., the analog of $\phi_2$ from the diffusion maps case)
+is given by $\psi_{1,d+1}$.
 %
@@
 \begin{equation} \label{eq:angle_relations}
-\begin{aligned}
-	\alpha_{ij} &= \theta_{ij} \\
-	\beta_{ij} &= \frac{dx_{ij}}{n} \times \eta_{proj} \\
-	\gamma_{ij} &= \frac{dy_{ij}}{n} \times \eta_{proj} \\
-\end{aligned}
+%\begin{aligned}
+%	\alpha_{ij} &= \theta_{ij} \\
+%	\beta_{ij} &= \frac{dx_{ij}}{n} \times \eta_{proj} \\
+%	\gamma_{ij} &= \frac{dy_{ij}}{n} \times \eta_{proj} \\
+%\end{aligned}
+	\alpha_{ij} = \theta_{ij}, \; \; \;
+	\beta_{ij} = \frac{dx_{ij}}{n} \times \eta_{proj}, \; \; \;
+	\gamma_{ij} = \frac{dy_{ij}}{n} \times \eta_{proj} 
 \end{equation}
@@
 \begin{aligned}
-R_{1,1} & = \cos(\beta)\cos(\gamma) \\
+R_{1,1} & = \cos(\beta)\cos(\gamma) &
 R_{2,1} & = \cos(\beta)\sin(\gamma) \\
-R_{3,1} & = -\sin(\beta) \\
+R_{3,1} & = -\sin(\beta) &
 R_{3,2} & = \sin(\alpha)\cos(\beta) \\
-R_{3,3} & = \cos(\alpha)\cos(\beta) 
+R_{3,3} & = \cos(\alpha)\cos(\beta)  &
 \end{aligned}

### 2014-04-08 image_analysis_paper/pnas_draft.tex
added edits to draft

@@
 
-Let $x_1, \dots, x_m \in \mathbb{R}^n$ denote the signals that we wish to align with respect to rotations. 
+%TODO: work on
+%Let $x_1, \dots, x_m \in \mathbb{R}^n$ denote the signals that we wish to align with respect to rotations. 
+Let $x_1, \dots, x_m$ denote the signals that we wish to align with respect to rotations. 
+%
+Each signal is a function defined on the unit circle, discretized at $n$ equally-spaced points;
+rotating the function then corresponds to cyclically shifting the elements of $x_i$.
 %
@@
 \end{equation}
-where the function $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ degrees, and $\xi_i$ is a noise term (often Gaussian noise). 
-%
+where the function $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ radians, and $\xi_i$ is a (typically Gaussian) noise term. 
+%
+For the discretized signals, rotation by $\theta_i$ radians corresponds to cyclically shifting the vectors by $\frac{\theta_i}{2 \pi} n$ elements (rounded to the nearest integer to obtain a valid shift).
+
 Our goal is to obtain estimates of $\theta_1, \dots, \theta_m$.
@@
 We can obtain an {\em estimate} of $\theta_i - \theta_j$ by computing the rotation that optimally aligns $x_j$ to $x_i$, 
-i.e., $\theta_{ij} \approx \theta_i - \theta_j$, where
+i.e., %$\theta_{ij} \approx \theta_i - \theta_j$, where
 %
 \begin{equation} \label{eq:opt_angle}
-\theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2.
-\end{equation}
-%
-For the one-dimensional profiles shown in Figure~\ref{fig:1d_demo}, to compute the optimal angles in \eqref{eq:opt_angle}, we exhaustively search over all 100 possible rotations of the signals.
-
-Rather than work with the angles $\theta_{ij}$ directly, it is more convenient to work with the rotation matrices 
+\theta_i - \theta_j \approx \theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2.
+\end{equation}
+%
+For the one-dimensional profiles shown in Figure~\ref{fig:1d_demo}, we exhaustively search over all $n=100$ possible shifts of the signals to compute the optimal angles in \eqref{eq:opt_angle}.
+
+Rather than work with the angles $\theta_{ij}$ directly, it is more convenient to work with the rotation matrices, 
 \begin{equation} \label{eq:R_theta}
 R(\theta_{ij}) = \begin{bmatrix}
-\cos(\theta_{ij}) & \sin(\theta_{ij}) \\
--\sin(\theta_{ij}) & \cos(\theta_{ij})
-\end{bmatrix}
-\end{equation}
-%
-Successive rotations now correspond to multiplication of rotation matrices: $R(\alpha_1 + \alpha_2) = R(\alpha_1) R(\alpha_2)$.
+\cos(\theta_{ij}) & -\sin(\theta_{ij}) \\
+\sin(\theta_{ij}) & \cos(\theta_{ij})
+\end{bmatrix}.
+\end{equation}
+%
+Successive rotations correspond to multiplication of rotation matrices: $R(\alpha_1 + \alpha_2) = R(\alpha_1) R(\alpha_2)$.
 %
@@
 
-Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$; we will write our procedure for general $d$ because we will later consider three-dimensional rotations).
-%
-We construct the matrix $H \in \mathbb{R}^{md \times md}$, where $H$ is an $m \times m$ matrix of $d \times d$ blocks with
+Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$; we write our procedure for general $d$ because we will later consider three-dimensional rotations).
+%
+We construct the matrix $H \in \mathbb{R}^{md \times md}$, where $H$ is an $m \times m$ matrix of $d \times d$ blocks, with the $i,j^{th}$ block of $H$, $H_{ij}$, defined as
 \begin{equation} \label{eq:H_to_R}
@@
 
-We note that, under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, $H_{ij} \approx R(\theta_i) R(\theta_j)^T$,
+Under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, $H_{ij} \approx R(\theta_i) R(\theta_j)^T$
 %\begin{equation} 
@@
 %\end{equation}
- and so
+ and 
 \begin{equation} \label{eq:H_low_rank}
@@
 %
-From \eqref{eq:H_low_rank}, we can see that the top block eigenvector of $H$, which we will denote $\hat{R}$, contains estimates of $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$.
+From \eqref{eq:H_low_rank}, we can see that the top block eigenvector of $H$, %which we will denote $\hat{R}$, 
+contains estimates of $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$.
 %
@@
 \end{equation}
-and $\hat{R}_i \in \mathbb{R}^{d \times d}$ is the estimate for $R(\theta_i)$. 
+where $\hat{R}_i \in \mathbb{R}^{d \times d}$ is the estimate for $R(\theta_i)$. 
 %
@@
 \end{equation}
-where $\epsilon$ is a characteristic distance between data points.
+where $d(x_i, x_j)$ is the distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic distance between data points.
 %
@@
 %
-$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph}); we take $\epsilon$ to be the median of the pairwise distances between data points.
+%$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph}); 
+We take $\epsilon$ to be the median of the pairwise distances between data points.
 
@@
 %
-In our applicatins, we assume that this one dimension, parameterized by $\phi_2$, is correlated with time.
+In our applications, we assume that this one dimension, parameterized by $\phi_2$, is correlated with time.
 %
@@
 
-For vector diffusion maps, we first construct the matrix $S \in \mathbb{R}^{md \times md}$, with
+For vector diffusion maps, we first construct the matrix $S \in \mathbb{R}^{md \times md}$, with the $i,j^{th}$ block of $S$, $S_{ij}$, defined as
 \begin{equation}
@@
 %
-where $A$ is defined in \eqref{eq:dmaps_A} and $H$ is defined in \eqref{eq:H_to_R}.
-%
-The distance $d(x_i, x_j)$ used in $W_{ij}$ is the distance between samples {\em after} pairwise alignment (i.e., it is the minimium obtained in \eqref{eq:opt_pairwise}).
+where $A_{ij} \in \mathbb{R}$ is defined in \eqref{eq:dmaps_A} and $H_{ij} \in \mathbb{R}^{d \times d}$ is defined in \eqref{eq:H_to_R}.
+%
+The distance $d(x_i, x_j)$ used in \eqref{eq:dmaps_W} to compute $W_{ij}$ is the distance between samples {\em after} pairwise alignment (i.e., it is the minimium obtained in \eqref{eq:opt_pairwise}).
 
@@
 \theta \in [0, 2\pi) \\ 
-dx \in [-\Delta, \Delta]\\ 
-dy \in [-\Delta, \Delta]
+dx, dy \in [-\Delta, \Delta]\\ 
+%dy \in [-\Delta, \Delta]
 \end{matrix}
@@
 %
-The total rotation, $R_{ij} \in SO(3)$, is 
+The total rotation, $R_{ij} \in \mathbb{R}^{3 \times 3}$, is 
 \begin{equation} \label{eq:total_rot}
-	R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})	 = R^z(\gamma_{ij})  R^y(\beta_{ij})  R^x(\alpha_{ij})
+	R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})	 = R^z(\gamma_{ij})  R^y(\beta_{ij})  R^x(\alpha_{ij}).
 %R(\alpha_{ij}, \beta_{ij}, \gamma_{ij}) = 
@@
 %
-We then compute the Euler angles $\alpha$, $\beta$, and $\gamma$ from the rotation matrix $R$ using the following relationships
+$R_{i,est}^T$ then gives the rotation matrix which registers image $i$. 
+%
+From a rotation matrix $R$, we can compute the Euler angles $\alpha$, $\beta$, and $\gamma$ using the following relationships
 \begin{equation}
@@
 %
-We then invert \eqref{eq:angle_relations} to compute the optimal translation and rotation for each image.
+We then invert \eqref{eq:angle_relations} to compute the corresponding translation and rotation.
 %

### 2014-04-09 image_analysis_paper/pnas_draft.tex
updated draft with some of Yannis comments

@@
 Hang~Lu\affil{2}{},
-Stanislav~Y.~Shvartsman\affil{1}{} \affil{3}{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA}, 
+Amit~Singer\affil{3}{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \affil{4}{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA},
+Stanislav~Y.~Shvartsman\affil{1}{} \affil{5}{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA},
 \and
@@
 %
-If each data point is a two-dimensional image, the images must first be aligned or registered before they can be temporally ordered. 
-%
-When such data sets are large, and/or if the developmental changes are subtle, these tasks can be difficult to do by hand. 
-%
-We present an automatic way to register and temporally order cross--sectional data sets of images. 
-%
-We use mathematical techniques (angular synchronization for image registration, and diffusion maps for temporal ordering) which are applicable to a wide variety of data sets and require little {\em a priori} knowledge of the image features or the developmental dynamics.
+If each data point is a two-dimensional image, the images must first be registered (aligned) before they can be temporally ordered.
+%
+When such data sets are large, noisy, and/or if the developmental changes are subtle, these tasks can be difficult to do by hand.
+%
+We present an automatic approach to registering {\em and} temporally order cross--sectional data sets of images.
+%
+The mathematical techniques (angular synchronization for image registration, diffusion maps for temporal ordering and
+vector diffusion maps for simultaneously performing both tasks) are applicable to a wide variety of data sets and 
+require little {\em a priori} knowledge of the image features or the developmental dynamics.
 %
@@
 
-\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional. 
-%
-In longitudinal studies, developmental progress is monitored over time within the same embryo. 
-%
-In a cross-sectional study, one embryo contributes only one snapshot of a chemical or morphological process along a developmental trajectory, which must be reconstructed from multiple snapshots of different embryos.
-%
-Both of these experimental designs have their advantages and limitations and are extensively used by developmental biologists. 
+\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal \cite{...} and cross-sectional \cite{...}.
+%
+In longitudinal studies, developmental progress is monitored over time for the same embryo.
+%
+In a cross-sectional study, one embryo contributes only a single snapshot of a chemical or morphological process along its developmental trajectory; the latter must be reconstructed from multiple snapshots of different embryos.
+%
+Both of these sampling schemes have their advantages and limitations, and both are extensively used by developmental biologists.
 %
@@
 %
-In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that visualize a handful of cellular processes. 
-%
-Fixed embryos are then imaged using any given number of microscopic techniques. 
-%
-Recent advances in large-scale physical manipulation and imaging of embryos have produced rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different orientation and time point.
+In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that help visualize a handful of cellular processes.
+%
+Fixed embryos are then imaged using any given number of microscopy techniques.
+%
+Recent advances in large-scale physical manipulation and imaging of embryos have produced rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different geometric orientation and developmental time point.
 %
@@
 %
-In general, it is known that a collection of embryos belongs to a certain time window.
-%
-In order to mine such datasets for underlying developmental dynamics, snapshots of different embryos must be spatially aligned and ordered in time. 
-%
-We present a general framework that greatly accelerates both of these tasks.
-
-Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are apparent. 
-%
-As an example, Figure~\ref{fig:fish} shows a schematic of fish development, combining the processes of growth and patterning.  
-%
-In this case, temporal ordering can be accomplished by arranging the fish by size, which correlates with the progress of development, and image registration is based on obvious morphological landmarks, such as positions of head and fins.
-%
-Dealing with real data is far more nontrivial, due to the number of images and similarities between them. 
-%
-To illustrate most of the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of pattern formation in the early fruit fly embryo, one of the leading systems for studies of developmental dynamics.
+In general, it is only known that a collection of embryos belongs to a certain time window.
+%
+In order to mine such image datasets for underlying developmental dynamics, snapshots of different embryos must be spatially aligned ("registered")  and ordered in time.
+%
+We present a general computational framework that greatly accelerates -and can combine- both of these tasks.
+
+Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are visually apparent.
+%
+As an example, Figure~\ref{fig:fish} shows a caricature of fish development, combining the processes of growth and color patterning.
+%
+In this case, temporal ordering can be accomplished by arranging the fish by size, which uniquely correlates with the progress of development; image registration is then based on obvious morphological landmarks, such as positions of head and fins.
+%
+Real data will, of course, provide nontrivial challenges, due to the number of images, due to variability and measurement noise, and due to the subtlety of several developmental changes.
+
+%
+To illustrate the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of regulatory protein expression 
+%%%
+%%% check whether "regulatory protein" is the correct term
+%%%
+pattern formation in the early fruit fly embryo, one of the leading model systems for studies of developmental dynamics.
 %
@@
 (a) Fish, each in a different orientation and a different stage of development.
-(b) Fish, now registered and temporally ordered by hand.
-For this example, the registration and ordering is easy to do by hand because the data set is small and the developmental changes are easy to recognize.}
+(b) Fish, now rotationally registered and temporally ordered.
+For this example, the registration and ordering is easy to do ``by hand" because the data set is small and the developmental changes are easy to recognize.}
 %\label{fig:fish}
@@
 %
-By end of the $2^{nd}$ hour of development, sequential divisions of a fertilized egg generate a system where $\sim 6,000$ nuclei are uniformly arranged in under the common plasma membrane. 
-%
-During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclose nuclei into different cells grow in length (Figure~\ref{subfig:membranes}). 
+By end of the $2^{nd}$ hour of development, sequential divisions of a fertilized egg generate a system where $\sim 6,000$ nuclei are uniformly arranged under the common plasma membrane.
+%
+During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclose nuclei into different cells grow in thickness (Figure~\ref{subfig:membranes}).
 %
@@
 %
-Images in Figure~\ref{subfig:images_unordered} show optical sections through the short axis of the embryos. 
-%
-The embryos were fixed and stained with chemicals that visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red) \cite{chung2010microfluidic}.
-%
-Each of the images reveals a uniform arrangement of nuclei and nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.
+Images in Figure~\ref{subfig:images_unordered} show optical sections perpendicular to the long axis of the embryo.
+%%%
+%%% we have to somehow say that it is consistent
+%%%
+%
+%%%
+%%% ask if "visualize" is typical as active verb
+%%%
+The embryos were fixed and stained with chemicals that help visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red) \cite{chung2010microfluidic}.
+%
+Each of the images reveals a uniform arrangement of nuclei and an azimuthally nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.
 
@@
 (b) Selected image from data set.
-(c) Profile of Dorsal (green) and dpERK (red) around the circumfrence of the embryo in (b).}
+(c) Profile of Dorsal (green) and dpERK (red) around the circumference of the embryo in (b).}
 %\label{fig:images_unordered}
@@
 
+%%% 
+%%% orientation does not read good to Y below -
+%%% it is the same orientation, but a different rotation around the axis
+%%% how to say this ?
+%%%
 Each image in Figure~\ref{subfig:images_unordered} comes from a different embryo, captured at a different orientation and at a different and unknown point in time.
 %
-The goal is to align these images in a globally consistent way and order them in a way that is most representative of the underlying developmental trajectory. 
-%
-Below we show that both of these tasks can be fully automated.
-%
-We will use angular synchronization \cite{singer2011angular}, a registration algorithm that is robust to noise, to align the images in a consistent way.
-%
-We will use diffusion maps \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
-%
-We will also show how both steps can be combined using vector diffusion maps \cite{singer2012vector}.
+The goal is to register these images in a globally consistent way and order them in a way that is most representative of the underlying developmental trajectory.
+%
+Both of these tasks can be fully automated, and even combined.
+%
+We will use {\em angular synchronization} \cite{singer2011angular}, a registration algorithm that is robust to noise, to align the images in a consistent way.
+%
+We will use {\em diffusion maps} \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
+%
+We will also show how both steps can be combined using {\em vector diffusion maps} \cite{singer2012vector}.
 %
 %The paper is organized as follows.
-%%
+%
 %First, we use a synthetic dataset, where the correct orientation and ordering of data points is known, to formalize the tasks of image registration and temporal ordering.
-%%
+%
 %Then, we show that both of these tasks can be effectively combined in a single step.
-%%
+%
 %Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align them and order them correctly in time.
-%%
+%
 %We conclude by discussing the general requirements that must be satisfied for our approach to work and compare it to other approaches and ideas in cross-sectional studies of biological dynamics.
@@
 %We would like to temporally order images, such as the images shown in Figure~\ref{subfig:images_unordered}.
-%%
+%
 %We will use diffusion maps, a dimensionality reduction algorithm, to automatically order our data.
-%%
+%
 %However, this algorithm requires distances between data points.
-%%
+%
 %Because each image is in a different orientation, we need to first align or {\em register} the images before computing distances.
-%%
+%
 %Once all the images have been aligned, we can compute distances and use diffusion maps to temporally order the data.
 
+%%%
+%%%  Y to discuss with S
+%%%  I think we should have the real data
+%%%
+%%%  also, these dynamics "are like" our dynamics
+%%%
+%%%   even if this is kept, one should say that this chosen because it LOOKS LIKE what we have
+%%% another reason I want to have the real data is because in the discussion I want to talk about the flips 
+%%%
+
 We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set with relatively simple dynamics.
 %
-We construct concentration profiles defined on a ring, and we assume that each ring is in an arbitrary orientation; an example is shown in Figure~\ref{subfig:1d_example}.
+We construct concentration profiles defined on a ring, and we assume that each ring is arbitrarily rotated around its center; an example is shown in Figure~\ref{subfig:1d_example}.
 %
@@
 %
-We discritize the profiles into 100 points, so our data will be 100-dimensional vectors. 
-%
-Rotation of the ring in Figure~\ref{subfig:1d_example} corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown below the ring. 
-%
-Figure \ref{subfig:1d_unaligned_unordered} shows all the concentration profiles in our data set. 
-%
-The concentration profiles have been stacked in an array, such that each row corresponds to one profile.
-%
-Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian growing in time) are not readily apparent.
+We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors.
+%
+Rotation of the ring in Figure~\ref{subfig:1d_example} corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown below the ring (the continuous symmetry group is the group of all two-dimensional proper rotations, denoted $SO(2)$; the corresponding discrete symmetry group for the discretized profiles is $\mathbb{Z}_n$, the group of integers modulo $n$).
+%
+Figure \ref{subfig:1d_unaligned_unordered} shows all the concentration profiles in our data set; they have been stacked in an array, so that each row corresponds to a single profile.
+%
+Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.
 
@@
 (b) Intensity corresponding to the profile in (a).
-(c) Many concentration profiles, each of the form described in (a).
-Each row in the array corresponds to one image.
+(c) An ensemble of concentration profiles, each of the form described in (a).
+Each row in the array corresponds to a single image.
 (d) The concentration profiles in (c), now registered using angular synchronization.
 (e) The concentration profiles in (c), now temporally ordered using diffusion maps.
-(f) The concentration profiles in (c), now simultaneously registered and temporally ordered using vector diffusion maps.}
+(f) The concentration profiles in (c), registered and temporally ordered in a single step using vector diffusion maps.}
 %\label{fig:1d_demo}
@@
 
-Our first task is to register the set of images or profiles.
+We start by registering the set of images or profiles.
 %
@@
 %
-Angular synchronization first computes the pairwise alignments %(rotations for our one-dimensional profiles in Figure \ref{fig:1d_demo}c, or translations and rotations for the two-dimensional images in Figure \ref{fig:images_unordered})
-for every possible pair of data points. %, effectively using each image as a template for every other image.
+Angular synchronization first aligns every possible pair of data points; that is, it computes the rotation angle
+required to make them have the maximum possible overlap (see {\it Materials and Methods}).
+%%%
+%%% so I do not like the figure with the vectors - it is correct, but people have to take the
+%%% function, discretize and make it a vector in their heads, and may be "one step too much"
+%%%
 %
@@
 %
-From this large collection of pairwise alignments, angular synchronization then rotates each data point, such that the individual alignments are most consistent (in some metric) with the pairwise alignments.
-% (i.e., we would like the difference between the optimal rotations for two different concentration profiles to be as close as possible to the pairwise rotation needed to align the two profiles).
-%
-Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned.
-%-- these are the angles we would like to recover from the pairwise information. 
-%
-Note that the differences between the rotation angles are consistent with the pairwise angles needed to rotate one vector to another, i.e., $\theta_i - \theta_j \approx \theta_{ij}$.
-%
-To compute these alignments, we formulate the task as an optimization problem, and then relax it to an eigenproblem.
-%
-The top eigenvector(s) of a matrix (which contains information about the pairwise alignments) give the alignments for each data point.
+From this large collection of pairwise alignments, angular synchronization then selects a unique (up to a global rotation)
+set of angles $\theta_i$, one for each data point, such that the resulting pairwise alignments $\theta_i-\theta_j$ are {\em most consistent} in some metric.
+%
+Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned
+(the rotation angles we computed are also indicated).
+%
+Note that the differences between the rotation angles assigned to each vector are consistent with the pairwise angles needed to rotate one vector to another, i.e., $\theta_i - \theta_j \approx \theta_{ij}$.
+%
+To compute these alignments, the task is formulated as an optimization problem, and then relaxed to an eigenproblem.
+%
+The top eigenvector(s) of a matrix (which contains information about all pairwise alignments) 
+provides the (optimal) rotation angles assigned to each data point.
 %
@@
 The angles to align one vector to another are indicated.
-(b) The vectors from (a), each rotated so that the set of vectors is aligned.
-Note that the angles are consistent with the pairwise alignments in (a).
+(b) The vectors from (a), each rotated so that the set is globally aligned.
+Note that the chosen rotation angles are consistent with the pairwise alignments in (a).
 (c) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
-Each pair of data points is connected by an edge, and the edge intensity is related to the distance between the points (close data points are connected by dark edges).
-(d) The data in (c), colored by the first (non-trivial) eigenvector from diffusion maps, $\phi_2$.
-The color parameterizes the one-dimensional curve.
+Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \eqref{eq:dmaps_W}), so that close data points are connected by dark (``stronger'') edges.
+(d) The data in (c), colored by the first (non-trivial) eigenvector from the diffusion map computational procedure, $\phi_2$.
+The color intensity is monotonic with the arclength, thus parameterizing the curve.
 }
@@
 %
-The peaks are now aligned, even though our algorithm use no information about the existence or location of these peaks.
-%
-Because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered up to a global rotation.
+It is reassuring that the intensity peaks appear aligned, 
+even though the algorithm uses no specific information about the existence or location of such peaks.
+%
+We note once more that, since angular synchronization only uses pairwise rotation information, 
+the optimal rotations are recovered up to a global rotation.
 
@@
 
-After registering the images or concentration profiles, we want to order them in time to reconstuct the developmental dynamics. 
-%
-We will use diffusion maps (DMAPS) to temporally order our data.
-%
-In general, diffusion maps is a dimensionality reduction technique, which aims to uncover low-dimensional, nonlinear structure in a high-dimensional data set \cite{coifman2005geometric}. 
-%
-Figure~\ref{subfig:dmaps_edges} shows an example of a two-dimensional data set (the ``high-dimensional data'') which lies on a one-dimensional nonlinear curve (the low-dimensional structure).
-%
-Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the first diffusion maps coordinate.
+After geometrically registering the data (images or concentration profiles), we want to order them in time to reconstruct the developmental dynamics.
+%
+Our main assumption is that data variability
+strongly correlates with time (see below).
+%
+We use diffusion maps to temporally order our data.
+%
+In general, diffusion maps is a dimensionality reduction technique, which aims to uncover low-dimensional, nonlinear structure in high-dimensional data sets \cite{coifman2005geometric}.
+%
+Figure~\ref{subfig:dmaps_edges} shows an example of a two-dimensional data set (the ``high-dimensional" data) which lies on a one-dimensional nonlinear curve (the ``low-dimensional" manifold).
+%
+Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the leading diffusion map coordinate.
 %
@@
 %
-We assume that our imaging data also falls on a one-dimensional, nonlinear curve, and that this curve is parameterized by time (e.g., in Figure~\ref{subfig:dmaps_color}, we assume that the data points are ordered from youngest to oldest along the curve).
+Assume that our (one dimensional, or two-dimensional image) data also lies along a one-dimensional, nonlinear curve in a high-dimensional space, and that this curve is parameterized by time (e.g., in Figure~\ref{subfig:dmaps_color}, we assume that the data points are ordered from youngest to oldest along the curve).
 %
@@
 
-To uncover a parameterization of this curve, we assume that data points which are close in the high-dimensional space should also be close on the curve (we would like images which look similar to be close together in our developmental trajectory).
+To uncover a parameterization of this curve, we assume that data points which are close in the high-dimensional space should also be close {\em on the curve} (we would like images which look similar to be close together in our developmental trajectory).
 %
@@
 %
-To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored according to the distance between the data points, so that close data points have a dark edge between them.
-%
-Visually, the dark edges are meaningful and delineate the one-dimensional curve.
-%
-Furthermore, the light edges are not informative and do not depict any meaningful structure. 
-%
-We want points which are close in the high-dimensional space, i.e., connected by dark edges, to be close in our low-dimensional parameterization.
-%
-This problem can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a matrix which contains information about pairwise distances between the data points.
-%
-In general, this procedure is called diffusion maps \cite{coifman2005geometric}.%; the formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).
+To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored according to the distance between the data points, so that close data points have dark (``strong", ``meaningful") edges connecting them.
+%
+Visually, these dark edges delineate the one-dimensional curve; in contrast, the light edges are much less informative 
+about the low-dimensional structure.
+%
+%We want points which are close in the high-dimensional space, i.e., connected by dark edges, to be close in our low-dimensional %parameterization.
+%
+This problem can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a Markov matrix which contains information about pairwise distances between the data points.
+%
+This procedure has been named diffusion maps \cite{coifman2005geometric}.
+%
+%The formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).
 
@@
 %
-We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances between data points.
+We use the Euclidean distance between the (discretized in $\mathbb{R}^{100}$) aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances between data points.
 %
@@
 %
-As expected, the intensity of the peak grows as a function of time. 
+As expected, the intensity of the peak grows as a function of $\phi_2$ (since we constructed these data
+to have a peak growing in time).
 %
@@
 
-\subsection{Vector diffusion maps for registration and ordering}
+\subsection{Vector diffusion maps for simultaneous registration and ordering}
 
@@
 %
-We can combine angular synchronization (for registration) and diffusion maps (for temporal ordering) into one eigencomputation that allows us to {\em simultaneously} recover the optimal alignments and the temporal ordering.
+Angular synchronization (for registration) can be combined with diffusion maps (for temporal ordering) in a single eigencomputation that allows us to {\em simultaneously} recover the optimal alignments {\em and} the right temporal ordering.
 %
@@
 %
-The data has been ordered by $\psi_{1, 3}$, the first (non-trivial) vector diffusion maps coordinate.
+The data has been ordered by $\psi_{1, 3}$, the first (non-trivial) vector diffusion maps coordinate (see {\it Materials and Methods}).
 
@@
 
-The same methodologies (with small modifications) can be applied to two-dimensional images.
-%
-Instead of $100$-dimensional concentration vectors, we now work with $20000$-dimensional images ($100 \times 100$ pixel images with two color channels).
-%
-We register the images with respect to translations and rotations 
-(we convert the two-dimensional translations and rotations to three-dimensional rotations so that we can apply angular synchronization or vector diffusion maps). 
-%
-We used vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_unordered};
+The same methodologies (with small modifications) can be applied directly to our two-dimensional images.
+%
+Instead of one-dimensional profiles (discretized to $100$-dimensional concentration vectors), we now work with 
+two color, two-dimensional images (discretized into $100 \times 100$ pixels, or $20000$-dimensional vectors).
+%
+We first register the images with respect to translations and rotations (the appropriate continuum symmetry group is $ISO(2)$);
+it is expedient to instead approximate the two-dimensional translations and rotations by {\em three-dimensional rotations}, for which the symmetry group $SO(3)$ facilitates the application of the 
+vector diffusion map algorithm (see both {\it Materials and Methods} and {\it SI Appendix})
+%
+%%%
+%%% AGAIN ?
+%%%
+We use vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_unordered};
 the results are shown in Figure~\ref{fig:images_ordered}.
 %
-Both the red and green peaks are now aligned, and the intensity of the red peaks grows as a function of time.
-
-In these experiments, we can evaluate the quality of our automatic registration and ordering using prior knowledge about the system.
-%
-It is known that the peak in the distribution of Dorsal (visualized by the green channel) specifies the ventralmost position in the embryo, and that the nuclei (shown in gray) delineate the circumference of the embryo.
+Both the red and green peaks are now aligned, and the intensity of the red peaks grows as a function of time (or of the leading vector diffusion map coordinate).
+
+In these experiments, we are fortunate enough to be able to validate our computations of automatic registration and ordering using prior physical knowledge about the system.
+%
+For registration, it is known that the peak in the distribution of Dorsal (visualized by the green channel) specifies the ventralmost position in the embryo, and that the nuclei (shown in gray) delineate the circumference of the embryo.
+%%%
+%%% maybe you can also quantify this guy
+%%%
 %
@@
 %
-%At the same time, 
-Temporal ordering can be done based on the monotonic progress of cellularization (Figure~\ref{subfig:membranes}) \cite{figard2013plasma}. 
-%
-We can evaluate the quality our automatic orderings by comparing with the orderings obtained from cellularization.
-%
-The rank correlation coefficient between the two orderings is 0.9149, indicating that our automatic ordering is consistent with the ordering obtained from cellularization.
-%
-We can also visually compare the alignments and orderings by comparing the arrays of one-dimensional profiles, similar to the format in Figure~\ref{fig:1d_demo};
-these results are shown in Figures~\ref{subfig:1d_membranes}~and~\ref{subfig:1d_vdm}.
+Temporal ordering, on the other hand, can be done based on the monotonic progress of cellularization (Figure~\ref{subfig:membranes}) \cite{figard2013plasma}; this helps validate our vector diffusion map-based ordering.
+%
+The rank correlation coefficient between these two temporal orderings is 0.9149, indicating that our automatic ordering is consistent with the ordering obtained from cellularization.
+%
+
+The simultaneous registrations/orderings can be visually compared through the sequences of one-dimensional profiles (similar to the format in Figure~\ref{fig:1d_demo}), shown in Figures~\ref{subfig:1d_membranes}~and~\ref{subfig:1d_vdm}.
 %
 The general trends are consistent:
-the green peaks are aligned, and the red peaks grow as a function of time. 
-%
-However, Figure~\ref{subfig:1d_vdm}, which was automatically registered and ordered, shows a better clustering of the outer peaks in the top corners of the image.
+both green  and red peaks are aligned, and the red peaks grow as a function of time (or of the leading vector diffusion map coordinate).
+%
+However, Figure~\ref{subfig:1d_vdm}, which was automatically registered and ordered, shows an arguably better clustering of the ``outer" peaks in the top corners of the image.
 %
@@
 %
-These peaks are known to arise later in development, and so it is reassuring that they are at the end of the trajectory.
+These peaks are known to arise later in development, and so it is reassuring that they appear at the end of the (recovered) trajectory.
 %
@@
 %Although our methods are fairly robust, they do require a sufficient amount of data so that DMAPS can ``see'' a smooth trajectory.
-%%
+%
 %We evaluated how much data is required by subsampling our data set (with replacement), and for each subsample, computing the temporal ordering using VDM and calculating the rank correlation between this ordering and the ordering from cellularization.
-%%
+%
 %The results are shown in Figure~\ref{subfig:bootstrap} (each data point is averaged over 50 random subsamples).
-%%
+%
 %For small data sets, the average rank correlation is very low, indicating that the orderings from VDM are inaccurate.
-%%
-%However, for moderately sized data sets (approximately 30 data points), the 
-%rank correlation begins to plateau and VDM can accurately order the data.
+%
+%However, for moderately sized data sets (approximately 30 data points), the rank correlation begins to plateau and VDM can accurately order the data.
 
@@
 (b) Membrane length as a function of time.
+The points correspoinding to the images in (a) are indicated with arrows.
 (c) The one-dimensional concentration profiles, registered by peak Dorsal fluorescence and ordered by membrane length.
@@
 
-We have shown how dimensionality reduction techniques can greatly acceleration the registration and temporal ordering of cross-sectional imaging data.
-%
-Although we illustrated our methods using a specific data set from a study of {\em Drosophila} embryogenesis, our framework is very general and requires little knowledge of the image features or the developmental dynamics.
+We have shown how symmetry and dimensionality reduction techniques can greatly accelerate the registration and temporal ordering of cross-sectional developmental imaging data.
+%
+Although we illustrated our methods using a specific data set from a study of {\em Drosophila} embryogenesis, the framework is quite general and requires little knowledge of the image features or the developmental dynamics.
 %
@@
 %
-Recently, machine learning algorithms have been used to group {\em Drosophila} images into different developmental classes \cite{yuan2014automated}.
-%
-However, these methods have the disadvantages that the user must define the appropriate classes, and that hand-labeled training data is required.
+Recently, machine learning algorithms have been used to group {\em Drosophila} images into different developmental classes \cite{yuan2014automated}; yet there, the appropriate classes must be defined by the user, and hand-labeled training data are required.
 
@@
 %
-Cross-sectional RNA sequencing data has been temporally ordered \cite{anavy2014blind}, where each data point is the transcriptome for a single sample.
-%
-Gene expression data from multiple patients has been studied \cite{gupta2008extracting} \cite{qiu2011discovering},
-%
-where each data point is a vector containing the expression levels of the various genes.
-%
-Snapshots of cells throughout the cell cycle can be described by a vector of features (e.g., amount of DNA) which describe the cell's state  \cite{kafri2013dynamics}.
+Cross-sectional RNA sequencing data has been temporally ordered \cite{anavy2014blind, trapnell2014dynamics}, where each data point is the transcriptome for a single sample.
+%
+Gene expression data from multiple patients has been studied in \cite{gupta2008extracting, qiu2011discovering};
+%
+here each data point is a vector containing the expression levels of the various genes.
+%
+Snapshots of cells throughout the cell cycle can be described by a vector of features (e.g., amount of DNA) which quantify the cell's state  \cite{kafri2013dynamics}.
 %
@@
 The majority of these works temporally order the data either by solving a traveling salesman problem on the data, or by constructing a minimum spanning tree of the data,
-in the hopes that these structures (the path of the travelling salesman, or the minimum spanning tree) characterize the majority of the structure within the data and provide an accurate picture of the dynamical progression.
+expecting that these structures (the path of the travelling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate picture of the dynamical progression.
 %
@@
 %
-Diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003} \cite{tenenbaum2000global} \cite{Donoho2003} \cite{Roweis2000}.
-%
-In general, they are more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}; we therefore expect them to perform better than the ordering algorithms used in previous work.
+Diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003, tenenbaum2000global, Donoho2003, Roweis2000}.
+%
+They have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}; we therefore suspect they may perform better than the ordering algorithms used in previous work.
+%%%
+%%%   we need help here 
+%%%
 %
@@
 
-The task of image registration is widely studied, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
+The task of image registration has been widely studied, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
 %
@@
 %
-Temporal ordering of images has been studied, for example, in ordering images taken of a construction site over time \cite{schindler2007inferring}, 
-but these methods are also based on first extracting features from the images. 
-%
-%In our developmental biology applications, 
-However, in general, appropriate features may not be known {\em a priori}, and so we require a registration algorithm that does not rely on predefined features.
-%
-Angular synchronization and vector diffusion maps have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}.
-%
-However, in this application, each image is a picture of (essentially) the {\em same} molecule; variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
-%
-In contrast, our data contains variation from the different embryo orientations, as well as variation from the underling developmental dynamics.
-
-In all of these data-driven methods, the essential question is how much and what type of data is needed to accurately reconstruct the dynamics. 
-%
-For the methods we present here, we require a sufficient amount of data for the one-dimensional curve to be ``seen'' by the algorithms (i.e., in Figure~\ref{subfig:dmaps_edges}, there are a sufficient number of data points so that our eye can see the curve; if there were only 4 data points, it would most likely be difficult for our eye to detect the one-dimensional structure).
-%
-We also require the data to contain enough information so that the alignments and dynamics are apparent.
-%
-For example, in the data presented in Figure~\ref{subfig:images_unordered}, the green peaks delineate the position of the ventral axis in the embryo, and the intensity of the red peaks grow over time.
-%
-Therefore, these images contain information about rotational symmetries of the data as well as the dynamics of the data.
-%
-If we removed the green signal, and only used the red signal in our analysis, the registration would not be as accurate.
-%
-Furthermore, if we only used the green signal, then there would be no dynamic information (the shape or intensity of Dorsal does not change as a function of time in the developmental stage we are considering), and the temporal ordering would be inaccurate.
-%
-However, we are confident that for many applications, these requirements - a sufficient number of data points, and data which contain enough information - are easily met. 
+Temporal ordering of images has been studied, for example, in ordering images taken of a construction site over time \cite{schindler2007inferring}; these methods are also based on first extracting features from the images.
+%
+However, in general, appropriate features may not be known {\em a priori}, and so we require a registration algorithm that does not rely on predefined features, but only on the inherent geometric symmetry of the problem.
+%
+Angular synchronization and vector diffusion maps have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}; there each image is a picture of (essentially) the {\em same} molecule, and variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
+%
+In contrast, our data contains variation from the different embryo orientations, as well as variation from the underlying developmental dynamics (which underpins the development of vector diffusion map algorithms).
+
+In all of these data-driven methods, an essential question is how much and what type of data is needed to accurately reconstruct the dynamics.
+%
+For diffusion maps to temporally order the data, we require the developmental trajectory to be well-sampled, and that the principal direction of variability in the data (the coordinate which diffusion maps will recover) be monotonic in time.
+%
+In addition, the data should contain spatial variability for synchronization to successfully register the images.
+%
+In our example, the data set is sufficiently large so that the dynamics are well-sampled.
+%
+We have included both the red and the green channels so that the principal direction of variability is monotonic in time.
+%
+Visually, one can see that the green channel does not change appreciably in time; if we only used this channel in our analysis, the principal direction of variability would not be correlated with time, and we could not temporally order the data using diffusion maps.
+%
+Both the red and green channels contain spatial variability; the green channel has one peak at a specific angular around the embryo, and the red channel has two peaks which grow in time.
+%
+This variability allows us to register the data points.
+%
+We include the green channel because the location of the red peaks is noisy, and so inclusion of the (less noisy) green channel allows us to more accurately register the images.
+%
+%For the methods we present here, we require a sufficient amount of data for the one-dimensional evolution curve to be ``seen'' by the algorithms (i.e., in Figure~\ref{subfig:dmaps_edges}, there are a sufficient number of data points so that our eye can see the curve; if there were only 4 data points, it would most likely be difficult for our eye to detect the one-dimensional structure).
+%
+%We also require the data to contain enough information so that the alignments and dynamics are apparent.
+%
+%For example, in the data presented in Figure~\ref{subfig:images_unordered}, the green peaks delineate the position of the ventral axis in the embryo, and the intensity of the red peaks grow over time.
+%
+%Therefore, these images contain information about rotational symmetries of the data as well as the dynamics of the data.
+%
+%If we removed the green signal, and only used the red signal in our analysis, the registration would not be as accurate.
+%
+%Furthermore, if we only used the green signal, then there would be no dynamic information (the shape or intensity of Dorsal does not change as a function of time in the developmental stage we are considering), and the temporal ordering would be inaccurate.
+%
+We are confident that for many applications, these requirements - data which adequately samples the developmental trajectory, and data with sufficient spatial and temporal variability - are easily met.
 %
 In these cases, our methods allow for the rapid, automated analysis of imaging data sets to help uncover complex dynamics and structure.
+
+%%%   you should sample well enough in time
+%%% your dmap coordinate should have a nice slope with time
+%% you should have variation in angle ---and we know we are random in the rotations
+%%
+%% and now the green is an issue of noise.....
+%% the green "sees" angle but not time,
+%% the red sees angle and time, but angle badly
+%% the ocmbination sees both well enough 
+%%
+%%
+%%   we need to say things about "other methods" -- Mallat, AmitNow, Procrustes...
+%% and if we change the artificial stuff, but even if not
+%% we need to say something about flips.
+%%
+%%   Y is afraid of the compsci literature being "fairly represented" 
+%%
+%%
 
@@
 
-%TODO: work on
-%Let $x_1, \dots, x_m \in \mathbb{R}^n$ denote the signals that we wish to align with respect to rotations. 
-Let $x_1, \dots, x_m$ denote the signals that we wish to align with respect to rotations. 
-%
-Each signal is a function defined on the unit circle, discretized at $n$ equally-spaced points;
-rotating the function then corresponds to cyclically shifting the elements of $x_i$.
-%
-We assume that each signal $x_i$ is a noisy rotated copy of the underlying signal $x_{true}$ (which we are {\em not} given), such that 
+Let $x_1, \dots, x_m$ denote the signals that we wish to align with respect to rotations;
+%
+Each signal is a function defined on the unit circle (on the plane).
+%
+%Practically, it is discretized in a $n$-long vector (the local intensity at $n$ equidistant points around the circle);
+%rotating the function by an angle $\theta$ then corresponds to cyclically shifting the elements of $x_i$ by $\frac{\theta_i}{2 \pi} n$ (rounded to the nearest integer to obtain a valid shift).
+%
+First assume that each signal $x_i$ is a {\em noisy} rotated copy of the underlying signal $x_{true}$ 
+(which we are {\em not} given), such that
 \begin{equation}
@@
 %
-For the discretized signals, rotation by $\theta_i$ radians corresponds to cyclically shifting the vectors by $\frac{\theta_i}{2 \pi} n$ elements (rounded to the nearest integer to obtain a valid shift).
 
@@
 %
-For the one-dimensional profiles shown in Figure~\ref{fig:1d_demo}, we exhaustively search over all $n=100$ possible shifts of the signals to compute the optimal angles in \eqref{eq:opt_angle}.
-
-Rather than work with the angles $\theta_{ij}$ directly, it is more convenient to work with the rotation matrices, 
+Practically, the signals are discretized in a $n$-long vector (the local intensity at $n$ equidistant points around the circle);
+rotating the function by an angle $\theta$ then corresponds to cyclically shifting the elements of $x_i$
+by $\frac{\theta_i}{2 \pi} n$ (rounded to the nearest integer to obtain a valid shift).
+%
+For the one-dimensional discretized profiles shown in Figure~\ref{fig:1d_demo}, we exhaustively search over all $n=100$ possible shifts of the signals to compute the optimal angles in \eqref{eq:opt_angle}.
+%
+Alternatively, for continuous signals, an optimization algorithm
+can be used to find the solution to \eqref{eq:opt_angle} \cite{ahuja2007template}.
+
+Rather than work with the angles $\theta_{ij}$ directly, it is more convenient to consider work with the rotation matrices,
 \begin{equation} \label{eq:R_theta}
@@
 \end{equation}
-%
-Successive rotations correspond to multiplication of rotation matrices: $R(\alpha_1 + \alpha_2) = R(\alpha_1) R(\alpha_2)$.
+which we can think of as operating on the points of the unit circle, on which our signal is defined.
+%
+Successive rotations correspond to multiplication of the corresponding rotation matrices: $R(\alpha_1 + \alpha_2) = R(\alpha_1) R(\alpha_2)$.
 %
@@
 %
-
+%
 Under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, $H_{ij} \approx R(\theta_i) R(\theta_j)^T$
@@
 %
-From \eqref{eq:H_low_rank}, we can see that the top block eigenvector of $H$, %which we will denote $\hat{R}$, 
-contains estimates of $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$.
+It follows directly from \eqref{eq:H_low_rank}  that the top block eigenvector of $H$, %which we will denote $\hat{R}$,
+contains our best estimates of $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$.
 %
@@
 %
-However, $\hat{R}_i$ is not guaranteed to be a rotation matrix; the closest rotation matrix, $R_{i, est}$, is given by
+However, $\hat{R}_i$ is not guaranteed to be an orthogonal matrix;
+the closest orthogonal matrix, $R_{i, est}$, is given by
 \begin{equation} \label{eq:R_est}
@@
 We estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and register the signals by rotating signal $i$ by $-\theta_i$.
+%
+We note that, in our actual computations, the pairwise rotations $\theta_{ij}$ are computed in a discrete setting, then the overall
+synchronization is performed in the continuum context to obtain $\theta_i$, and the results are rounded to give the closest
+discrete shift.
 
 %Furthermore, this formulation also considers {\em higher-order} consistency information.
-%%
+%
 %For example, given our pairwise estimates $R_{ij}$, we know that relationships of the form
@@
 %should also hold.
-%%
+%
 %Note that
@@
 %therefore, {\em all} infomation of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$.
-%%
+%
 %Because $H$ and $H^2$ have the same eigenvectors, our problem formulation accounts for not only pairwise alignment information, but also these higher-order considerations.
 
-\section{Diffusion Maps \cite{coifman2005geometric}}
-
-%We will uncover a parameterization of this curve by only considering {\em local} neighbors of each data point.
-%%
-%To illustrate how local neighbors can allow us to uncover a global parameterization, assume we have data $x_1, x_2, \dots, x_m \in \mathbb{R}$ such that $x_1 < x_2 < \dots < x_m$.
-%%
-%We consider $x_i$ ``close to'' $x_{i+1}$ and $x_{i-1}$, and $x_i$ ``far away from'' all other data points. 
-%%
-%We then construct the (symmetric) weight matrix, $W \in \mathbb{R}^{m \times m}$,
-%\begin{equation}
-%W = 
-%\begin{bmatrix}
-%	1 & 1/2 & 0 & 0 & \dots & 0 \\
-%	1/2 & 1 & 1/2 & 0 & \dots & 0 \\
-%	\ddots & \ddots & \ddots & \ddots & \ddots & \ddots \\
-%	0 & 0 & 0 & \dots & 1 & 1/2 \\
-%	0 & 0 & 0 & \dots & 1/2 & 1 
-%\end{bmatrix}.
-%\end{equation}
-%%
-%$W_{ij}$ is large if $x_i$ is close to $x_j$,
-%%
-%If we normalize this matrix such that $\sum_j W_{ij} = 1$, then one can show that the first eigenvector of this normalized matrix is the all-ones vector, $\phi_1 = [1 1 \dots 1]^T$, and the second eigenvector is a cosine mode, $\phi_2 = [cos(\pi/m) \cos(2 \pi/ m) \cdots \cos((m-1) \pi / m) \cos(\pi)]^T$.
-%%
-%Therefore, the second eigenvector, $\phi_2$ is {\em one-to-one} with the parameterization of the data, i.e. $\phi_2(1) > \phi_2(2) > \dots > \phi_2(m)$, just as $x_1 < x_2 < \dots < x_m$. 
-%%
-%Therefore, sorting our data by the value of the entries in $\phi_2$ will allow us to order our data.
-
-For diffusion maps, the first step is to construct a matrix $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if points $x_i$ and $x_j$ are ``close.'' 
-%
-%For the above analysis, the key step was having a notion of ``closeness'' between data points so that we can construct the matrix $W$.
+\section{Ordering using Diffusion Maps \cite{coifman2005geometric}}
+
+Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional space), we want to find a coordinate transformation (in our case, we are interested in finding a single coordinate) $y(x)$ that preserves local information: points that are close in the original space should also be close in the coordinate $y$.
+%
+The first step is to construct the matrix $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if points $x_i$ and $x_j$ are ``close.''
 %
@@
 \end{equation}
-where $d(x_i, x_j)$ is the distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic distance between data points.
-%
-Therefore, points less than $\epsilon$ apart are considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
-%
-%$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph}); 
-We take $\epsilon$ to be the median of the pairwise distances between data points.
-
-We want to find a coordinate $y$ such that $y(i)$ and $y(j)$ if $W_{ij}$ is large (i.e., $y$ preserves local information, and so points that are close in the original space are also close in the coordinate $y$).
-%
-We want solve the following optimization problem \cite{Belkin2003}
+where $d(x_i, x_j)$ is the pairwise distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic scale.
+%
+Points less than $\epsilon$ apart are thus considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
+%
+$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph}); here, we take $\epsilon$ to be the median of the pairwise distances between data points.
+
+We want a coordinate $y$ such that $y(x_i)$ and $y(x_j)$ are close if $W_{ij}$ is large.
+%
+This is accomplished solving following optimization problem \cite{Belkin2003}
 \begin{equation} \label{eq:dmaps_opt_problem}
-\argmin_{y} \sum_{ij} W_{ij} (y(i) - y(j))^2
-\end{equation}
-%
-%The solution \ref{eq:dmaps_opt_problem} is $\phi_2$, the second eigenvector of $A$, where $A = D^{-1} W$, and $D$ is a diagonal matrix with $D_{ii} = \sum_j W_{ij}$ \cite{...}.
-%%
-%Therefore, $\phi_2$ provides a one-dimensional parameterization of the data, and ordering the data points by the value of $\phi_2$ will order them in time.
+\argmin_{y} \sum_{ij} W_{ij} (y(x_i) - y(x_j))^2
+\end{equation}
 %
@@
 %
-The next eigenvector, $\phi_2$, is the (non-trivial) solution to \eqref{eq:dmaps_opt_problem}, so that $\phi_2(j)$ gives the coordinate for data point $x_j$.
-%
-In our applications, we assume that this one dimension, parameterized by $\phi_2$, is correlated with time.
-%
-Therefore, ordering our data by $\phi_2(j)$ will effectively order our data in time.
+The next eigenvector, $\phi_2$, is the (non-trivial) solution to \eqref{eq:dmaps_opt_problem}, so that $\phi_2(j)=y(x_j)$ gives the ``new" coordinate for data point $x_j$.
+%
+In our application we have assumed that this single direction of variability, parameterized by $\phi_2$, is one-to-one with time.
+%
+Ordering the data by $\phi_2(j)$ will then, effectively, order them in time.
+%
+The procedure generalizes when the data lie on higher-dimensional manifolds (not just curves) in data space.
 
@@
 
-For vector diffusion maps, we first construct the matrix $S \in \mathbb{R}^{md \times md}$, with the $i,j^{th}$ block of $S$, $S_{ij}$, defined as
+Given data points $x_1, \dots, x_m$, for vector diffusion maps, one first constructs the matrix $S \in \mathbb{R}^{md \times md}$, with the $i,j^{th}$ block of $S$, $S_{ij}$, defined as
 \begin{equation}
@@
 %
-where $A_{ij} \in \mathbb{R}$ is defined in \eqref{eq:dmaps_A} and $H_{ij} \in \mathbb{R}^{d \times d}$ is defined in \eqref{eq:H_to_R}.
-%
-The distance $d(x_i, x_j)$ used in \eqref{eq:dmaps_W} to compute $W_{ij}$ is the distance between samples {\em after} pairwise alignment (i.e., it is the minimium obtained in \eqref{eq:opt_pairwise}).
-
-We then compute the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_{md}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{md}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$.
-%
-As in angular synchronization, the top (block) eigenvector of $S$ contains approximations of the optimal rotations, and the optimal rotations can be computed in the same way from \eqref{eq:R_est}.
-%
-However, the eigenvectors now also contain information about the embedding coordinates for our images.
-%
-In general, the embedding coordinates are given by 
+where $A_{ij} \in \mathbb{R}$ pertains to the diffusion kernel for the pair of data points $x_i, x_j$ and is defined in \eqref{eq:dmaps_A}, and $H_{ij} \in \mathbb{R}^{d \times d}$ pertains to the pairwise alignment between the data points and is defined in \eqref{eq:H_to_R}.
+%
+It is important to note that distance $d(x_i, x_j)$ used in the diffusion kernel in \eqref{eq:dmaps_W} is the distance between data points {\em after} after pairwise alignment; that is, the minimum obtained in \eqref{eq:opt_pairwise} among all possible distances between all possible shifts of the two data points.
+%
+In the language of symmetry groups, this distance is a metric between the orbits of the data points induced by the symmetry group.
+
+One then computes the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_{md}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{md}$ of $S$, and orders them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$.
+%
+These eigenvectors contain information about {\em both} the optimal rotations (the ``synchronization" component) and the 
+variation of the data {\em after} the spatial symmetries have been factored out - in our case, their temporal variation.
+%
+Assuming that the data (after symmetries have been factored out) are relatively closely clustered, it is reasonable
+to expect, as in angular synchronization, that the top (block) eigenvector of $S$ contains approximations of the optimal rotations, 
+which can be computed in the same way from \eqref{eq:R_est}.
+%
+We then expect some subsequent eigenvector to contain information relevant to the ordering of the data,
+i.e., about the main direction(s) of data variability modulo the geometric symmetries.
+%%%
+%%% somewhere we need to talk about compact groups
+%%% discussion might be a good place because SCALING is obviously important and
+%%% it is not compact......
+%%%
+%However, the eigenvectors now also contain information about the embedding coordinates for our images.
+%
+In general, the embedding coordinates will be given by
 \begin{equation}
@@
 %
-If we assume that our data are one-dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, the embedding coordinate for our data %(i.e., the analog of $\phi_2$ from the diffusion maps case)
-is given by $\psi_{1,d+1}$.
-%
-If no such claims can be made, then one often runs a second step of diffusion maps, using $\psi_{k,l}$ as the coordinates, to uncover a parameterization of the manifold modulo symmetries.
+If we assume that our data mainly vary along a one-dimensional manifold, and that the rotations and the dynamics are uncoupled and therefore separable, one expects the embedding coordinate for our data %(i.e., the analog of $\phi_2$ from the diffusion maps case)
+to be  $\psi_{1,d+1}$.
+%
+Our example is, by physical considerations, a separable one; if no separability claims can be made,
+it is conceivable that a second diffusion map step, using $\psi_{k,l}$ as the coordinates, might uncover a 
+useful parameterization of the data manifold modulo symmetries.
+%%%
+%%% \cite....
+%%%
 
@@
 %
-For \eqref{eq:H_low_rank} to be satisfied, we require that the matrices used to represent the symmetry group be orthogonal, i.e. $R(\alpha)^{-1} = R(\alpha)^T$. 
-%
-However, the typical matrix representation for the group of two-dimensional translations and rotations does not satisfy this property. 
-%
-Instead, we (approximately) represent rotations and translations in two-dimensions using three-dimensional rotation matrices, by projecting the image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular}.
-%(see Figure \ref{fig:SO3_picture}).
+For \eqref{eq:H_low_rank} to be satisfied, we require that the matrices used to represent the symmetry group be orthogonal, i.e., $R(\alpha)^{-1} = R(\alpha)^T$.
+%
+However, the typical matrix representation for the group of two-dimensional translations and rotations, $ISO(2)$, does not satisfy this property.
+%
+Instead, we (approximately) represent rotations and translations in two-dimensions using three-dimensional rotation matrices, by projecting the image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular} (see {\it SI Appendix}).
 %
@@
 %
-For each image pair $I_i$ and $I_j$, we solve
+Practically, the images are discretized at pixels (rather than continuous functions on the unit sphere).
+%
+We therefore discretize the optimization problem, and for each image pair $I_i$ and $I_j$, we solve
 \begin{equation}\label{eq:opt_pairwise}
@@
 We only consider shifts that correspond to an integer number of pixels, to remove the need for interpolation when doing the translations.
-
+%
 The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
@@
 %
-We then convert the two-dimensional translation and rotation to a three dimensional rotation. 
+We then convert the two-dimensional translation and rotation to a (continuous) three-dimensional rotation.
 %
@@
 %
-From the matrices $R_{1, est}, \dots, R_{m, est}$, we then find the corresponding translations and rotations of the images.
+From the matrices $R_{1, est}, \dots, R_{m, est}$, we can then recover the corresponding translations and rotations of the images.
 %

### 2014-04-09 image_analysis_paper/pnas_draft.tex
Added more edits

@@
 
-\title{Temporal ordering and registration of imaging data in cross--sectional studies of development}
+\title{Unscrambling arrested development: temporal ordering and registration of cross-sectional imaging data}
 
@@
 
-\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal \cite{...} and cross-sectional \cite{...}.
+\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional \cite{diggle2002analysis}.
 %
@@
 %
-In order to mine such image datasets for underlying developmental dynamics, snapshots of different embryos must be spatially aligned ("registered")  and ordered in time.
+In order to mine such image datasets for underlying developmental dynamics, snapshots of different embryos must be spatially aligned ("registered") to factor out the relevant symmetries (translations and rotations),  and ordered in time.
 %
@@
 This dataset comprises $49$ fluorescence microscopy images of different embryos fixed during the $3^{rd}$ hour of their development.
-
+%
 \begin{figure}[t]
@@
 \end{figure}
-
+%
 The fruit fly embryo is shaped like a rice grain, approximately $1/2$~mm long.
@@
 %
-The main point to note here is that, in most of these applications, registration first requires defining a set of {\em features} or landmark points, and then locating these features within each image.
+The main point to note here is that many registration algorithms require the locations of landmarks or features within each image.
 %
@@
 %
+Template methods are also common for registration, where one optimally aligns each image to a predefined template function.
+%
+However, these methods can suffer when the images are noisy, and the success of such methods can depend strongly on the template function \cite{shatsky2009method}.
+%
 Angular synchronization and vector diffusion maps have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}; there each image is a picture of (essentially) the {\em same} molecule, and variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
@@
 In contrast, our data contains variation from the different embryo orientations, as well as variation from the underlying developmental dynamics (which underpins the development of vector diffusion map algorithms).
+%
+Other synchronization algorithms have been proposed \cite{bandeira2014multireference}.
+%
+These synchronization methods are formulated for general symmetry groups.
+%
+Depending on the specific data, one might consider other symmetries, such as reflections and scaling/dilation.
+%
+For synchronization, our symmetry group must be such that...
+
+Temporal ordering can also be accomplished without first registering the images, by constructing {\em features} which are invariant to the underlying symmetry group. 
+%
+In such methods, one computes functions of the images which are invariant under translations and rotations. 
+%
+These functions are then used as the data in the diffusion maps calculation; since the functions do not change under translations and rotations, the resulting diffusion maps embedding is invariant to symmetries.
+%
+Such functions include the scattering transform \cite{mallat2012group} and the bispectrum \cite{zhao2014rotationally}.
+%
+These have the disadvantage that a we do not obtain a registered set of images, but only a temporal ordering of the images.
+%
+However, such methods can often be faster for large data sets, since sets of features are computed per image, whereas synchronization requires pairwise computations.
 
@@
 %% we need to say something about flips.
+%% we need to say symmetry early at the beginning -- important
+%%  the expression "factoring out symmetry", the reference to Ben....  something about symmetry groups 
+%%    the smoothness when one wants to make movies.
+%%  
 %%
@@
 Given data points $x_1, \dots, x_m$, for vector diffusion maps, one first constructs the matrix $S \in \mathbb{R}^{md \times md}$, with the $i,j^{th}$ block of $S$, $S_{ij}$, defined as
-\begin{equation}
+\begin{equation} \label{eq:vdm_S}
 	S_{ij} = A_{ij} H_{ij}
@@
 %
-For \eqref{eq:H_low_rank} to be satisfied, we require that the matrices used to represent the symmetry group be orthogonal, i.e., $R(\alpha)^{-1} = R(\alpha)^T$.
-%
-However, the typical matrix representation for the group of two-dimensional translations and rotations, $ISO(2)$, does not satisfy this property.
-%
-Instead, we (approximately) represent rotations and translations in two-dimensions using three-dimensional rotation matrices, by projecting the image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular} (see {\it SI Appendix}).
-%
-Rotation of the image corresponds to rotation around one principal axis in three dimensional, and translation of the image corresponds (approximately) to rotations around the other two principal axes.
-%(Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$.
-However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.)
-%
-%These rotation matrices are (by definition) orthogonal, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
-
-To register the images, we first compute the translations and rotations to optimally align each pair of data points.
-%
-Practically, the images are discretized at pixels (rather than continuous functions on the unit sphere).
-%
-We therefore discretize the optimization problem, and for each image pair $I_i$ and $I_j$, we solve
+%We first compute the translations and rotations to optimally align each pair of data points.
+%
+Practically, we have planar square images discretized as pixels (rather than continuous functions on the plane).
+%
+We therefore solve discrete pairwise optimization problems: for each image pair $I_i$ and $I_j$ we find
 \begin{equation}\label{eq:opt_pairwise}
@@
 \end{equation}
-where $\Delta$ is the maximum number of pixels by which we translate the image, and $g(I_j, \theta, dx, dy)$ is a function that first rotates the image $I_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction.
+where $g(I_j, \theta, dx, dy)$ is a function that results from first rotating the image $I_j$ around the center of the square by $\theta$ degrees, then translating the resulting
+image by $dx$ pixel widths in the $x$ direction, and finally translating by $dy$ pixel widths in the $y$ direction.
 %
@@
 %
-The image rotation is done with the \texttt{imrotate} function in Matlab, using nearest neighbor interpolation to estimate the pixel intensities after rotation.%, since there is no direct mapping of the pixels of the unrotated image to the pixels of the rotated image.
-%
-The missing pixels in the rotated image due to corner effects are taken to be black.
-%
-The translations are implemented by shifting the pixels with periodic boundary conditions.
-%
-In \eqref{eq:opt_pairwise}, $\Delta$ is chosen such that only edge pixels with little to no signal are shifted, and the main image is not split or separated when translating (we take $\Delta=20$, which corresponds to a 20\% shift in the image).
+The norm \eqref{eq:opt_pairwise} involve an integral over the square, which is not left invariant by our rotations and
+translations; noting however that the pixels close to the border of the square have zero intensity allows a meaningful
+computation of this norm as long as the ``main feature" (the embryo) does not move out of the original square (no ``corner effects").
+%
+This also explains the $\Delta$ limits in the translation: 
+%
+in \eqref{eq:opt_pairwise}, $\Delta$ is chosen such that only edge pixels with little to no signal are shifted, and the main image is not split or separated when translating (we take $\Delta=20$, which corresponds to a 20\% shift in the image); it does not make sense to shift the images so much that
+the embryo would ``move out" of the original square.
+
+%
+Image rotation is performed with the \texttt{imrotate} function in Matlab, using nearest neighbor interpolation to estimate the pixel intensities after rotation.
+%, since there is no direct mapping of the pixels of the unrotated image to the pixels of the rotated image.
+%
+The missing pixels in the rotated image (the ``corner effects" we mentioned above) are taken to be black (zero intensity).
+%
+The translations are implemented by shifting the pixels (for computational speed we used periodic boundary conditions).
 %
@@
 %
-Therefore, instead of using an optimization procedure, we discretize the search space and exhaustively search to estimate the solution.
-%
-We discritize the search space of rotations into 20 possible rotations %($d\theta  \in \{0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10 \}$),
-and 11 possible translations in both the $x$ and $y$ directions. %($dx, dy \in \{-20, -16, -12, \dots, 12, 16, 20 \}$).
-%
-We then check all possible combinations for the best rotation and translations that align $I_j$ to $I_i$.
-%
-This direct enumeration approach is not prohibitive for this step.
+Therefore, instead of using an optimization procedure, we discretize the search space and exhaustively search to find the solution.
+%
+Our ``quantum translation" is chosen to be $1/10^{th}$ of the domain in each of the $x$ and $y$ directions, while
+our ``quantum rotation" is chosen to be $1/20^{th}$ of $2 \pi$. 
+%
+%We discritize the search space of rotations into 20 possible rotations %($d\theta  \in \{0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10 \}$),
+%and 11 possible translations in both . %($dx, dy \in \{-20, -16, -12, \dots, 12, 16, 20 \}$).
+%
+We then check all possible combinations for the rotation and translations that best align $I_j$ to $I_i$.
+%
+Although computationally demanding, this ``embarrassingly parallelizable" direct enumeration approach is not prohibitive.
 %
@@
 %
-We then convert the two-dimensional translation and rotation to a (continuous) three-dimensional rotation.
+
+The requirements for the synchronization step (e.g.
+for \eqref{eq:H_low_rank} to be satisfied), include the condition that the matrices used to represent the symmetry group be orthogonal, i.e., $R(\alpha)^{-1} = R(\alpha)^T$ \cite{singer2013spectral}.
+%
+However, the typical matrix representation for the group of two-dimensional translations and rotations, $ISO(2)$, does not satisfy this property.
+%
+Instead, we {\em approximately} represent rotations and translations in two-dimensions using {\em three-dimensional rotation matrices}, by projecting the (two-dimensional) image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular} (see {\it SI Appendix}).
+%
+Rotation of the image corresponds to rotation around one principal axis in three dimensions, while translation of the image corresponds (approximately) to rotations around the other two principal axes.
+%(Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$.
+However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.)
+%
+%These rotation matrices are (by definition) orthogonal, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
+
+We thus convert the optimal two-dimensional translation and rotation to corresponding (continuous) three-dimensional rotations.
 %
@@
 %
-This portion of the sphere is small enough so that curvature effects are not seen in the image.
+This portion of the sphere is small enough so that curvature effects are not prominent in the image.
 %
@@
 
-From the rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$,
-we build the matrix $H$ in \eqref{eq:H_to_R} and compute the estimates $R_{i, est}$ in \eqref{eq:R_est}.
-%
-From the matrices $R_{1, est}, \dots, R_{m, est}$, we can then recover the corresponding translations and rotations of the images.
+These rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$ are used in formulating the matrix $H$ in \eqref{eq:H_to_R} for angular synchronization, or the matrix $S$ in \eqref{eq:vdm_S} for vector diffusion maps, to compute the estimates of the $R_{i, est}$ in \eqref{eq:R_est}.
+%
+From the matrices $R_{1, est}, \dots, R_{m, est}$, we can then recover the corresponding (approximately) optimal translations and rotations of the images.
 %
@@
 %
-We then invert \eqref{eq:angle_relations} to compute the corresponding translation and rotation.
-%
-We round the optimal translations to the nearest integer to remove the need for interpolation when translating the images.
+Inverting \eqref{eq:angle_relations} yields the corresponding approximate translations and rotation;
+optimal translations to the nearest integer to remove the need for interpolation when translating the images.
 

### 2014-04-09 image_analysis_paper/pnas_draft.tex
added mroe edits

@@
 %
-If each data point is a two-dimensional image, the images must first be registered (aligned) before they can be temporally ordered.
+If each data point is a two-dimensional image, the images must first be registered before they can be temporally ordered.
 %
@@
 %
-We present an automatic approach to registering {\em and} temporally order cross--sectional data sets of images.
-%
-The mathematical techniques (angular synchronization for image registration, diffusion maps for temporal ordering and
+We present an automatic approach to register {\em and} temporally order cross-sectional data sets of images.
+%
+The mathematical techniques (angular synchronization for image registration, diffusion maps for temporal ordering, and
 vector diffusion maps for simultaneously performing both tasks) are applicable to a wide variety of data sets and
@@
 %
-In a cross-sectional study, one embryo contributes only a single snapshot of a chemical or morphological process along its developmental trajectory; the latter must be reconstructed from multiple snapshots of different embryos.
+In a cross-sectional study, one embryo contributes only a single snapshot of a chemical or morphological process along its developmental trajectory, and the developmental dynamics must be reconstructed from multiple snapshots of different embryos.
 %
@@
 %
-In order to mine such image datasets for underlying developmental dynamics, snapshots of different embryos must be spatially aligned ("registered") to factor out the relevant symmetries (translations and rotations),  and ordered in time.
-%
-We present a general computational framework that greatly accelerates -and can combine- both of these tasks.
+In order to mine such image datasets for the developmental dynamics, snapshots of different embryos must be spatially aligned or registered to factor out the relevant symmetries (i.e., translations and rotations), and ordered in time.
+%
+We present a general computational framework which greatly accelerates -and can combine- both of these tasks.
 
@@
 %
-In this case, temporal ordering can be accomplished by arranging the fish by size, which uniquely correlates with the progress of development; image registration is then based on obvious morphological landmarks, such as positions of head and fins.
-%
-Real data will, of course, provide nontrivial challenges, due to the number of images, due to variability and measurement noise, and due to the subtlety of several developmental changes.
-
-%
-To illustrate the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of regulatory protein expression
-%%%
-%%% check whether "regulatory protein" is the correct term
-%%%
-pattern formation in the early fruit fly embryo, one of the leading model systems for studies of developmental dynamics.
+In this case, temporal ordering can be accomplished by arranging the fish by size, which uniquely correlates with the progress of development.
+%
+Image registration is based on obvious morphological landmarks, such as the position of head and fins.
+%
+On the other hand, real data poses nontrivial challenges, due to the number of images, measurement noise and variability, and the subtlety of the developmental changes.
+
+To illustrate the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of signaling protein expression in the early fruit fly embryo, one of the leading model systems for studies of developmental dynamics.
 %
@@
 %
-Each of the images reveals a uniform arrangement of nuclei and an azimuthally nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.
+Each of the images reveals a uniform arrangement of nuclei and a nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.
 
@@
 %%%
-Each image in Figure~\ref{subfig:images_unordered} comes from a different embryo, captured at a different orientation and at a different and unknown point in time.
-%
-The goal is to register these images in a globally consistent way and order them in a way that is most representative of the underlying developmental trajectory.
+Each image in Figure~\ref{subfig:images_unordered} comes from a different embryo, captured at a different rotation about the long axis and at a different and unknown point in time.
+%
+The goal is to register these images in a globally consistent way and order them to be representative of the underlying developmental trajectory.
 %
@@
 
-We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set with relatively simple dynamics.
+We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set whose features resemble those in our imaging data set in Figure~\ref{subfig:images_unordered}.
+%
+This data set will allow us to illustrate the main points of the different algorithms, and will also have relatively simple dynamics which are easy to visualize.
 %
@@
 %
+Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown in the bottom of Figure~\ref{subfig:1d_example}; the symmetry group is the group of all two-dimensional proper rotations, denoted $SO(2)$.
+%
 Each concentration profile is a noisy Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
 %
-We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors.
-%
-Rotation of the ring in Figure~\ref{subfig:1d_example} corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown below the ring (the continuous symmetry group is the group of all two-dimensional proper rotations, denoted $SO(2)$; the corresponding discrete symmetry group for the discretized profiles is $\mathbb{Z}_n$, the group of integers modulo $n$).
-%
-Figure \ref{subfig:1d_unaligned_unordered} shows all the concentration profiles in our data set; they have been stacked in an array, so that each row corresponds to a single profile.
+We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_n$, the group of integers modulo $n$).
+%
+Figure \ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
 %
@@
 
-We start by registering the set of images or profiles.
-%
-%``Template-based'' registration \cite{ahuja2007template} is a common technique, where one selects a specific image or {\em template}, and then translates and rotates each image in the data set until it is optimally aligned with this template (optimal for us means minimizing the norm between the difference in pixel intensities).
-%
-%However, this method can result in many errors if the images are noisy, and can be problematic if a ``good'' template image is not known {\em a priori}.
-%
-We will use angular synchronization\cite{singer2011angular}, a technique which is robust to noise and requires no {\em a priori} knowledge of image features.
-%
-Angular synchronization first aligns every possible pair of data points; that is, it computes the rotation angle
-required to make them have the maximum possible overlap (see {\it Materials and Methods}).
+We begin by registering the set of profiles using angular synchronization\cite{singer2011angular}, a technique which is robust to noise and requires no {\em a priori} knowledge of image features.
+%
+Angular synchronization first aligns every possible pair of data points, i.e., it computes the rotation angles
+required to make each pair have the maximum possible overlap (see {\it Materials and Methods}).
 %%%
@@
 An illustration is shown in Figure~\ref{subfig:synch1};
-each data point is a vector, and $\theta_{ij}$ denotes the angle of rotation needed to align vectors $i$ and $j$ (these are the pairwise alignments).
-%
-From this large collection of pairwise alignments, angular synchronization then selects a unique (up to a global rotation)
-set of angles $\theta_i$, one for each data point, such that the resulting pairwise alignments $\theta_i-\theta_j$ are {\em most consistent} in some metric.
-%
-Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned
-(the rotation angles we computed are also indicated).
+each data point is a vector, and $\theta_{ij}$ denotes the angle of rotation needed to align vectors $i$ and $j$.
+%
+From this large collection of pairwise alignments, the algorithm then selects a set of rotation angles $\theta_i$, one for each data point, such that differences in rotation angles $\theta_i-\theta_j$ are {\em most consistent} (in some metric) with the pairwise alignments $\theta_{ij}$.
+%
+Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned; the computed rotation angles are also indicated.
 %
@@
 The top eigenvector(s) of a matrix (which contains information about all pairwise alignments)
-provides the (optimal) rotation angles assigned to each data point.
+provides the rotation angles assigned to each data point (see {\it Materials and Methods}).
 %
@@
 \customlabel{fig:schematics}{4}
-\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}a}
-\customlabel{subfig:dmaps_color}{\ref{fig:schematics}b}
-\customlabel{subfig:synch1}{\ref{fig:schematics}c}
-\customlabel{subfig:synch2}{\ref{fig:schematics}d}
+\customlabel{subfig:synch1}{\ref{fig:schematics}a}
+\customlabel{subfig:synch2}{\ref{fig:schematics}b}
+\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}c}
+\customlabel{subfig:dmaps_color}{\ref{fig:schematics}d}
 \end{figure}
@@
 %
-It is reassuring that the intensity peaks appear aligned,
+It is reassuring that the peaks are aligned,
 even though the algorithm uses no specific information about the existence or location of such peaks.
 %
-We note once more that, since angular synchronization only uses pairwise rotation information,
-the optimal rotations are recovered up to a global rotation.
+%We note once more that, since angular synchronization only uses pairwise rotation information,
+%the optimal rotations are recovered up to a global rotation.
 
@@
 
-After geometrically registering the data (images or concentration profiles), we want to order them in time to reconstruct the developmental dynamics.
-%
-Our main assumption is that data variability
-strongly correlates with time (see below).
-%
-We use diffusion maps to temporally order our data.
-%
-In general, diffusion maps is a dimensionality reduction technique, which aims to uncover low-dimensional, nonlinear structure in high-dimensional data sets \cite{coifman2005geometric}.
+After geometrically registering the data (concentration profiles or images), we want to order them in time to reconstruct the developmental dynamics.
+%
+%Our main assumption is that most of the variation in the data correlates with time.
+%
+We use diffusion maps \cite{coifman2005geometric}, a dimensionality reduction technique which aims to uncover low-dimensional, nonlinear structure in high-dimensional data sets, to temporally order our data.
 %
@@
 %
-Assume that our (one dimensional, or two-dimensional image) data also lies along a one-dimensional, nonlinear curve in a high-dimensional space, and that this curve is parameterized by time (e.g., in Figure~\ref{subfig:dmaps_color}, we assume that the data points are ordered from youngest to oldest along the curve).
+Our main assumptions are that our (one-dimensional profile, or two-dimensional image) data also lie on a one-dimensional, nonlinear curve, {\em and} that this curve is parameterized by time (e.g., in Figure~\ref{subfig:dmaps_color}, we assume that the data points are ordered from youngest to oldest along the curve).
 %
@@
 %
-To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored according to the distance between the data points, so that close data points have dark (``strong", ``meaningful") edges connecting them.
-%
-Visually, these dark edges delineate the one-dimensional curve; in contrast, the light edges are much less informative
+To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored according to the distance between the data points, so that close data points have dark/strong
+edges connecting them.
+%
+Visually, these dark edges delineate the one-dimensional curve; in contrast, the light edges are less informative
 about the low-dimensional structure.
@@
 %
-This problem can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a Markov matrix which contains information about pairwise distances between the data points.
+Finding such a parameterization can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a Markov matrix which contains information about pairwise distances between the data points (see {\it Materials and Methods}).
 %
@@
 %
-Angular synchronization (for registration) can be combined with diffusion maps (for temporal ordering) in a single eigencomputation that allows us to {\em simultaneously} recover the optimal alignments {\em and} the right temporal ordering.
-%
-This technique is called vector diffusion maps \cite{singer2012vector}.
+Angular synchronization (for registration) can be combined with diffusion maps (for temporal ordering) in a single eigencomputation that allows us to {\em simultaneously} recover the alignments {\em and} the temporal ordering.
+%
+This technique is called vector diffusion maps \cite{singer2012vector} (see {\it Materials and Methods}).
 %
@@
 %
-The data has been ordered by $\psi_{1, 3}$, the first (non-trivial) vector diffusion maps coordinate (see {\it Materials and Methods}).
+The data has been ordered by $\psi_{1, 3}$, the first (non-trivial) vector diffusion maps coordinate.
 
@@
 Instead of one-dimensional profiles (discretized to $100$-dimensional concentration vectors), we now work with
-two color, two-dimensional images (discretized into $100 \times 100$ pixels, or $20000$-dimensional vectors).
-%
-We first register the images with respect to translations and rotations (the appropriate continuum symmetry group is $ISO(2)$);
-it is expedient to instead approximate the two-dimensional translations and rotations by {\em three-dimensional rotations}, for which the symmetry group $SO(3)$ facilitates the application of the
-vector diffusion map algorithm (see both {\it Materials and Methods} and {\it SI Appendix})
-%
-%%%
-%%% AGAIN ?
-%%%
+two color, two-dimensional images (discretized into $100 \times 100$ pixels/$20000$-dimensional vectors), and we register the images with respect to translations and rotations (the symmetry group is $ISO(2)$).
+%
+To facilitate the application of vector diffusion maps, we approximate the two-dimensional translations and rotations by {\em three-dimensional rotations} (symmetry group $SO(3)$; see {\it Materials and Methods}, {\it SI Appendix}).
+%
 We use vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_unordered};
@@
 %
-Both the red and green peaks are now aligned, and the intensity of the red peaks grows as a function of time (or of the leading vector diffusion map coordinate).
-
-In these experiments, we are fortunate enough to be able to validate our computations of automatic registration and ordering using prior physical knowledge about the system.
-%
-For registration, it is known that the peak in the distribution of Dorsal (visualized by the green channel) specifies the ventralmost position in the embryo, and that the nuclei (shown in gray) delineate the circumference of the embryo.
+Both the red and green peaks are now aligned, and the intensity of the red peaks grows as a function of the leading vector diffusion map coordinate.
+
+In these experiments, we are fortunate enough to be able to validate our  automatic registration and ordering using prior physical knowledge about the system.
+%
+For registration, it is known that the peak in the distribution of Dorsal (shown in green) specifies the ventralmost position in the embryo, and that the nuclei (shown in gray) delineate the circumference of the embryo.
 %%%
@@
 %
-Temporal ordering, on the other hand, can be done based on the monotonic progress of cellularization (Figure~\ref{subfig:membranes}) \cite{figard2013plasma}; this helps validate our vector diffusion map-based ordering.
-%
-The rank correlation coefficient between these two temporal orderings is 0.9149, indicating that our automatic ordering is consistent with the ordering obtained from cellularization.
-%
-
-The simultaneous registrations/orderings can be visually compared through the sequences of one-dimensional profiles (similar to the format in Figure~\ref{fig:1d_demo}), shown in Figures~\ref{subfig:1d_membranes}~and~\ref{subfig:1d_vdm}.
+Temporal ordering, on the other hand, can be done based on the monotonic progress of cellularization (Figure~\ref{subfig:membranes}) \cite{figard2013plasma}.%; this helps validate our vector diffusion map-based ordering.
+%
+The results can be visually compared through the sequences of one-dimensional profiles (similar to the format in Figure~\ref{fig:1d_demo}), shown in Figures~\ref{subfig:1d_membranes}~and~\ref{subfig:1d_vdm}.
 %
@@
 %Furthermore, the membranes stop growing towards the end of this trajectory, making the ordering of the late signals in Figure~\ref{subfig:1d_membranes} less accurate.
+%
+Quantitatively, the rank correlation coefficient between these two temporal orderings is 0.9149, indicating that our automatic ordering is consistent with the ordering obtained from cellularization.
 
@@
 %
-Recently, machine learning algorithms have been used to group {\em Drosophila} images into different developmental classes \cite{yuan2014automated}; yet there, the appropriate classes must be defined by the user, and hand-labeled training data are required.
+Recently, machine learning algorithms have been used to group {\em Drosophila} images into different developmental classes \cite{yuan2014automated}, yet there, the appropriate classes must be defined by the user and hand-labeled training data are required.
 
@@
 Gene expression data from multiple patients has been studied in \cite{gupta2008extracting, qiu2011discovering};
-%
-here each data point is a vector containing the expression levels of the various genes.
+here, each data point is a vector containing the expression levels of the various genes.
 %
@@
 %
-These problems are no different than our task of ordering images, each of which can be thought of as a large vector of pixel intensities.
+These problems are no different than our task of ordering images, each of which is a large vector of pixel intensities.
 %
@@
 %
-They have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}; we therefore suspect they may perform better than the ordering algorithms used in previous work.
+They have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}, and so we suspect they may perform better than the ordering algorithms used in previous work.
 %%%
@@
 %
-For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then compare images by comparing these features.
-%
-Temporal ordering of images has been studied, for example, in ordering images taken of a construction site over time \cite{schindler2007inferring}; these methods are also based on first extracting features from the images.
-%
-However, in general, appropriate features may not be known {\em a priori}, and so we require a registration algorithm that does not rely on predefined features, but only on the inherent geometric symmetry of the problem.
-%
-Template methods are also common for registration, where one optimally aligns each image to a predefined template function.
+For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then register images by aligning these features.
+%
+Images from a construction site over time have been temporally ordered, but these methods are also based on first identifying features within the images \cite{schindler2007inferring}.
+%
+In general, appropriate features may not be known {\em a priori}, and so we require a registration algorithm that does not rely on predefined features, but only on the inherent geometric symmetry of the problem.
+%
+Template-based methods are also common for registration, where one optimally aligns each image to a predefined template function.
 %
 However, these methods can suffer when the images are noisy, and the success of such methods can depend strongly on the template function \cite{shatsky2009method}.
-%
-Angular synchronization and vector diffusion maps have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}; there each image is a picture of (essentially) the {\em same} molecule, and variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
+
+Angular synchronization and vector diffusion maps require no feature points, and no template function.
+%
+These methods have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}, where each image is a picture of (essentially) the {\em same} molecule; variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
 %
@@
 %
-Other synchronization algorithms have been proposed \cite{bandeira2014multireference}.
-%
-These synchronization methods are formulated for general symmetry groups.
-%
-Depending on the specific data, one might consider other symmetries, such as reflections and scaling/dilation.
-%
-For synchronization, our symmetry group must be such that...
-
-Temporal ordering can also be accomplished without first registering the images, by constructing {\em features} which are invariant to the underlying symmetry group. 
-%
-In such methods, one computes functions of the images which are invariant under translations and rotations. 
+%Other synchronization algorithms have been proposed \cite{bandeira2014multireference}.
+%
+These methods are formulated for general symmetry groups, and one might consider other symmetries, such as reflections and scaling/dilation.
+%
+The main requirement is that the symmetry group must be represented by orthogonal or unitary matrices \cite{singer2013spectral}.
+
+Temporal ordering can also be accomplished without first registering the images, by constructing functions of the images which are invariant under translations and rotations. 
 %
@@
 %
-However, such methods can often be faster for large data sets, since sets of features are computed per image, whereas synchronization requires pairwise computations.
+However, such methods can often be faster for large data sets, since the functions are computed per image, whereas synchronization requires the computation of all pairwise alignments.
 
@@
 %
-In addition, the data should contain spatial variability for synchronization to successfully register the images.
+In addition, the data should have spatial variability/structure for synchronization to successfully register the images.
 %
@@
 %
-Visually, one can see that the green channel does not change appreciably in time; if we only used this channel in our analysis, the principal direction of variability would not be correlated with time, and we could not temporally order the data using diffusion maps.
-%
-Both the red and green channels contain spatial variability; the green channel has one peak at a specific angular around the embryo, and the red channel has two peaks which grow in time.
-%
-This variability allows us to register the data points.
+If we only used the green channel in our analysis, we could not temporally order the data using diffusion maps, as this signal does not change appreciably in time.
+%
+Both the red and green channels contain spatial variability: the green channel has one peak at a specific angular around the embryo, and the red channel has two peaks which grow in time.
+%
+This allows us to register the data points.
 %
@@
 \sin(\theta_{ij}) & \cos(\theta_{ij})
-\end{bmatrix}.
-\end{equation}
-which we can think of as operating on the points of the unit circle, on which our signal is defined.
+\end{bmatrix},
+\end{equation}
+which we can think of as operating on the points of the unit circle on which our signal is defined.
 %
@@
 
-Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional space), we want to find a coordinate transformation (in our case, we are interested in finding a single coordinate) $y(x)$ that preserves local information: points that are close in the original space should also be close in the coordinate $y$.
+Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional space), we want to find a coordinate transformation $y(x)$ that preserves local information: points that are close in the original space should also be close in the coordinate $y$.
 %
@@
 %
-We typically use a Gaussian kernel,
+We use a diffusion kernel,
 \begin{equation} \label{eq:dmaps_W}
@@
 
-We want a coordinate $y$ such that $y(x_i)$ and $y(x_j)$ are close if $W_{ij}$ is large.
-%
-This is accomplished solving following optimization problem \cite{Belkin2003}
+We find a coordinate $y$ such that $y(x_i)$ and $y(x_j)$ are close if $W_{ij}$ is large by solving following optimization problem \cite{Belkin2003}
 \begin{equation} \label{eq:dmaps_opt_problem}
@@
 %
-The next eigenvector, $\phi_2$, is the (non-trivial) solution to \eqref{eq:dmaps_opt_problem}, so that $\phi_2(j)=y(x_j)$ gives the ``new" coordinate for data point $x_j$.
-%
-In our application we have assumed that this single direction of variability, parameterized by $\phi_2$, is one-to-one with time.
+The next eigenvector, $\phi_2$, is the (non-trivial) solution to \eqref{eq:dmaps_opt_problem}, so that $\phi_2(j)$, the $j^{th}$ entry of $\phi_2$, gives the ``new" coordinate for data point $x_j$ (i.e., $\phi_2(j) = y(x_j)$).
+%
+In our application, we have assumed that this single direction of variability, parameterized by $\phi_2$, is one-to-one with time.
 %
@@
 %
-where $A_{ij} \in \mathbb{R}$ pertains to the diffusion kernel for the pair of data points $x_i, x_j$ and is defined in \eqref{eq:dmaps_A}, and $H_{ij} \in \mathbb{R}^{d \times d}$ pertains to the pairwise alignment between the data points and is defined in \eqref{eq:H_to_R}.
-%
-It is important to note that distance $d(x_i, x_j)$ used in the diffusion kernel in \eqref{eq:dmaps_W} is the distance between data points {\em after} after pairwise alignment; that is, the minimum obtained in \eqref{eq:opt_pairwise} among all possible distances between all possible shifts of the two data points.
+where $A_{ij} \in \mathbb{R}$ (defined in \eqref{eq:dmaps_A}) pertains to the diffusion kernel between data points, and $H_{ij} \in \mathbb{R}^{d \times d}$ (defined in \eqref{eq:H_to_R}) pertains to the pairwise alignment between data points.
+%
+It is important to note that distance $d(x_i, x_j)$ used in the diffusion kernel in \eqref{eq:dmaps_W} is the distance between data points {\em after} after pairwise alignment, i.e., the minimum distance between all possible shifts of the two data points (which is obtained in \eqref{eq:opt_pairwise}).
 %
@@
 %
-Our example is, by physical considerations, a separable one; if no separability claims can be made,
-it is conceivable that a second diffusion map step, using $\psi_{k,l}$ as the coordinates, might uncover a
+Our example is, by physical considerations, a separable one; if no separability claims can be made, a second diffusion map step, using $\psi_{k,l}$ as the coordinates, may uncover a
 useful parameterization of the data manifold modulo symmetries.
@@
 \end{equation}
-where $g(I_j, \theta, dx, dy)$ is a function that results from first rotating the image $I_j$ around the center of the square by $\theta$ degrees, then translating the resulting
+where $g(I_j, \theta, dx, dy)$ is the image that results from first rotating the image $I_j$ around the center of the square by $\theta$ degrees, then translating the resulting
 image by $dx$ pixel widths in the $x$ direction, and finally translating by $dy$ pixel widths in the $y$ direction.
@@
 %
-The norm \eqref{eq:opt_pairwise} involve an integral over the square, which is not left invariant by our rotations and
-translations; noting however that the pixels close to the border of the square have zero intensity allows a meaningful
-computation of this norm as long as the ``main feature" (the embryo) does not move out of the original square (no ``corner effects").
-%
-This also explains the $\Delta$ limits in the translation: 
-%
-in \eqref{eq:opt_pairwise}, $\Delta$ is chosen such that only edge pixels with little to no signal are shifted, and the main image is not split or separated when translating (we take $\Delta=20$, which corresponds to a 20\% shift in the image); it does not make sense to shift the images so much that
-the embryo would ``move out" of the original square.
-
+The domain of the image (a square) is not invariant to our rotations and translations; however, the pixels close to the border of the square have zero intensity, and so the norm can be computed as long as the main image (the embryo) does not move out of the original square.
+%
+%The norm \eqref{eq:opt_pairwise} involve an integral over the square, which is not left invariant by our rotations and
+%translations; noting however that the pixels close to the border of the square have zero intensity allows a meaningful
+%computation of this norm as long as the ``main feature" (the embryo) does not move out of the original square (no ``corner effects").
+%
+This also justifies the $\Delta$ limits in the translation: 
+in \eqref{eq:opt_pairwise}, $\Delta$ is chosen such that only edge pixels with little to no signal are shifted, and the main image is not split or separated when translating (we take $\Delta=20$, which corresponds to a 20\% shift in the image).%; it does not make sense to shift the images so much that
+%the embryo would ``move out" of the original square.
 %
@@
 %
-The missing pixels in the rotated image (the ``corner effects" we mentioned above) are taken to be black (zero intensity).
-%
-The translations are implemented by shifting the pixels (for computational speed we used periodic boundary conditions).
+The missing pixels in the rotated image are taken to have zero intensity.
+%
+The translations are implemented by shifting the pixels (for computational speed, we used periodic boundary conditions).
 %
@@
 %
-Our ``quantum translation" is chosen to be $1/10^{th}$ of the domain in each of the $x$ and $y$ directions, while
-our ``quantum rotation" is chosen to be $1/20^{th}$ of $2 \pi$. 
+Our discrete translation steps are $1/10^{th}$ of the domain, and our discrete rotations steps are $1/20^{th}$ of $2 \pi$. 
 %
@@
 %
-We then check all possible combinations for the rotation and translations that best align $I_j$ to $I_i$.
+We check all possible combinations for the rotation and translations that best align $I_j$ to $I_i$.
 %
@@
 
-The requirements for the synchronization step (e.g.
-for \eqref{eq:H_low_rank} to be satisfied), include the condition that the matrices used to represent the symmetry group be orthogonal, i.e., $R(\alpha)^{-1} = R(\alpha)^T$ \cite{singer2013spectral}.
-%
-However, the typical matrix representation for the group of two-dimensional translations and rotations, $ISO(2)$, does not satisfy this property.
+The requirements for the synchronization algorithm (i.e., for \eqref{eq:H_low_rank} to be satisfied) include the condition that the matrices used to represent the symmetry group be orthogonal \cite{singer2013spectral}.
+%
+However, the typical matrix representation for $ISO(2)$, the group of two-dimensional translations and rotations, does not satisfy this property.
 %
@@
 
-We thus convert the optimal two-dimensional translation and rotation to corresponding (continuous) three-dimensional rotations.
-%
-We first compute the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$,
+To convert the optimal two-dimensional translation and rotation to corresponding (continuous) three-dimensional rotations,
+we first compute the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$,
 \begin{equation} \label{eq:angle_relations}
@@
 %
-We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere.
-%
-This portion of the sphere is small enough so that curvature effects are not prominent in the image.
+We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere;
+this portion of the sphere is small enough so that curvature effects are not prominent in the image.
 %
@@
 %
-The total rotation, $R_{ij} \in \mathbb{R}^{3 \times 3}$, is
+The total rotation, $R_{ij} \in SO(3) \subset \mathbb{R}^{3 \times 3}$, is
 \begin{equation} \label{eq:total_rot}
@@
 
-These rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$ are used in formulating the matrix $H$ in \eqref{eq:H_to_R} for angular synchronization, or the matrix $S$ in \eqref{eq:vdm_S} for vector diffusion maps, to compute the estimates of the $R_{i, est}$ in \eqref{eq:R_est}.
-%
-From the matrices $R_{1, est}, \dots, R_{m, est}$, we can then recover the corresponding (approximately) optimal translations and rotations of the images.
-%
-We first multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
-%
+These rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$ are used in formulating the matrix $H$ in \eqref{eq:H_to_R} for angular synchronization, or the matrix $S$ in \eqref{eq:vdm_S} for vector diffusion maps.
+%
+We can then compute the (approximately) optimal rotations (and, in the case of vector diffusion maps, temporal orderings) for the images.
+%
+From the three-dimensional rotations $R_{1, est}, \dots, R_{m, est}$, we  require the corresponding translations and rotations of the images.
+%
+We first multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began;
 $R_{i,est}^T$ then gives the rotation matrix which registers image $i$.
 %
-From a rotation matrix $R$, we can compute the Euler angles $\alpha$, $\beta$, and $\gamma$ using the following relationships
+From a rotation matrix $R$, we compute the Euler angles $\alpha$, $\beta$, and $\gamma$ using the following relationships
 \begin{equation}
@@
 %
-Inverting \eqref{eq:angle_relations} yields the corresponding approximate translations and rotation;
-optimal translations to the nearest integer to remove the need for interpolation when translating the images.
+Inverting \eqref{eq:angle_relations} yields the corresponding approximate translations and rotation, and we round optimal translations to the nearest integer to remove the need for interpolation when translating the images.
 

### 2014-04-09 image_analysis_paper/pnas_draft.tex
added more edits

@@
 \includegraphics[width=8.4cm]{fig3}
-\caption{Illustration of methodologies using one-dimensional profiles.
-(a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
+\caption{(a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
 (b) Intensity corresponding to the profile in (a).
 (c) An ensemble of concentration profiles, each of the form described in (a).
-Each row in the array corresponds to a single image.
-(d) The concentration profiles in (c), now registered using angular synchronization.
-(e) The concentration profiles in (c), now temporally ordered using diffusion maps.
-(f) The concentration profiles in (c), registered and temporally ordered in a single step using vector diffusion maps.}
+Each row in the array corresponds to a single profile.
+(d) The profiles in (c), now registered using angular synchronization.
+(e) The profiles in (c), now temporally ordered using diffusion maps.
+(f) The profiles in (c), registered and temporally ordered in a single step using vector diffusion maps.}
 %\label{fig:1d_demo}
@@
 \caption{(a) Set of vectors, each in a different orientation.
-The angles to align one vector to another are indicated.
+The pariwise alignment angles are indicated.
 (b) The vectors from (a), each rotated so that the set is globally aligned.

### 2014-04-10 image_analysis_paper/pnas_draft.tex
more edits to paper

@@
 %
-In order to mine such image datasets for the developmental dynamics, snapshots of different embryos must be spatially aligned or registered to factor out the relevant symmetries (i.e., translations and rotations), and ordered in time.
+In order to recover the developmental dynamics from such image datasets, snapshots of different embryos must be spatially aligned or registered to factor out the relevant symmetries (i.e., translations and rotations), and then ordered in time.
 %
@@
 %
-In this case, temporal ordering can be accomplished by arranging the fish by size, which uniquely correlates with the progress of development.
+In this case, temporal ordering can be accomplished by arranging the fish by size, which is monotonic with the developmental progress.
 %
@@
 \includegraphics[width=8.4cm]{fig1}
-\caption{Schematic illustration of reconstructing dynamics from cross-sectional data.
-(a) Fish, each in a different orientation and a different stage of development.
+\caption{(a) Fish, each in a different orientation and a different stage of development.
 (b) Fish, now rotationally registered and temporally ordered.
-For this example, the registration and ordering is easy to do ``by hand" because the data set is small and the developmental changes are easy to recognize.}
+For this caricature, the registration and ordering is easy to do ``by hand" because the data set is small and the developmental changes are easy to recognize.}
 %\label{fig:fish}
@@
 %
-Concurrently with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs \cite{lim2013kinetics}.
-%
-Images in Figure~\ref{subfig:images_unordered} show optical sections perpendicular to the long axis of the embryo.
+In conjunction with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs \cite{lim2013kinetics}.
+%
+Images in Figure~\ref{subfig:images_unordered} show optical sections at a fixed depth, perpendicular to the long axis of the embryo.
 %%%
@@
 %
-The goal is to register these images in a globally consistent way and order them to be representative of the underlying developmental trajectory.
+The goal is to register these images in a globally consistent way, and order them to be representative of the underlying developmental trajectory.
 %
@@
 
-We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set whose features resemble those in our imaging data set in Figure~\ref{subfig:images_unordered}.
-%
-This data set will allow us to illustrate the main points of the different algorithms, and will also have relatively simple dynamics which are easy to visualize.
+We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set whose features resemble the image features in Figure~\ref{subfig:images_unordered}.
+%
+The relatively simple dynamics of this data set will allow us to easily visualize and illustrate the main points of the different algorithms.
 %
@@
 %
-Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown in the bottom of Figure~\ref{subfig:1d_example}; the symmetry group is the group of all two-dimensional proper rotations, denoted $SO(2)$.
+Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown in the bottom of Figure~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
 %
@@
 (d) The profiles in (c), now registered using angular synchronization.
-(e) The profiles in (c), now temporally ordered using diffusion maps.
+(e) The profiles in (d), now temporally ordered using diffusion maps.
 (f) The profiles in (c), registered and temporally ordered in a single step using vector diffusion maps.}
@@
 Angular synchronization first aligns every possible pair of data points, i.e., it computes the rotation angles
-required to make each pair have the maximum possible overlap (see {\it Materials and Methods}).
+which make each pair have the maximum possible overlap (see {\it Materials and Methods}).
 %%%
@@
 %
-From this large collection of pairwise alignments, the algorithm then selects a set of rotation angles $\theta_i$, one for each data point, such that differences in rotation angles $\theta_i-\theta_j$ are {\em most consistent} (in some metric) with the pairwise alignments $\theta_{ij}$.
+From this large collection of pairwise alignments, the algorithm then selects a set of rotation angles, $\theta_i$ (one for each data point), such that differences in rotation angles, $\theta_i-\theta_j$, are {\em most consistent} (in some metric) with the pairwise alignments $\theta_{ij}$.
 %
@@
 %
-To compute these alignments, the task is formulated as an optimization problem, and then relaxed to an eigenproblem.
+The problem of finding these alignments is formulated as an optimization problem, and then relaxed to an eigenproblem.
 %
@@
 %Because the method considers {\em all} pairwise alignments, angular synchronization is often robust to noise.
-
+%
 \begin{figure}[t]
@@
 \end{figure}
-
+%
 Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization.
@@
 %
-We use diffusion maps \cite{coifman2005geometric}, a dimensionality reduction technique which aims to uncover low-dimensional, nonlinear structure in high-dimensional data sets, to temporally order our data.
-%
-Figure~\ref{subfig:dmaps_edges} shows an example of a two-dimensional data set (the ``high-dimensional" data) which lies on a one-dimensional nonlinear curve (the ``low-dimensional" manifold).
+Our main assumptions are that the high-dimensional data (images have one dimension per color pixel intensity, and concentration profiles have one dimension per discretization point) lie on a one-dimensional, nonlinear curve, {\em and} that this curve is parameterized by time.
+%
+Figure~\ref{subfig:dmaps_edges} shows an schematic of a two-dimensional data set (the ``high-dimensional" data) which lies on a one-dimensional nonlinear curve (the ``low-dimensional" manifold).
+%
+We assume that the data points are ordered from youngest to oldest along the curve.
+%
+Because our data are very high dimensional, we cannot extract this curve by eye.
+%
+We instead use diffusion maps \cite{coifman2005geometric}, a dimensionality reduction technique which aims to uncover low-dimensional, nonlinear structure in high-dimensional data sets, to automatically extract this one-dimensional curve.
 %
@@
 %
-The color parameterizes the one-dimensional curve that our eye can visually detect.
-%
-Our main assumptions are that our (one-dimensional profile, or two-dimensional image) data also lie on a one-dimensional, nonlinear curve, {\em and} that this curve is parameterized by time (e.g., in Figure~\ref{subfig:dmaps_color}, we assume that the data points are ordered from youngest to oldest along the curve).
-%
-Therefore, ordering data along this curve will effectively order data in time.
-%
-Because our data will be very high dimensional, we cannot do this by eye, and instead require an algorithm to automatically extract this one-dimensional structure.
-
-To uncover a parameterization of this curve, we assume that data points which are close in the high-dimensional space should also be close {\em on the curve} (we would like images which look similar to be close together in our developmental trajectory).
+The color parameterizes the one-dimensional curve that our eye can visually detect, and ordering the data by the leading diffusion map coordinate will effectively order data in time.
+
+To uncover the diffusion maps parameterization, we assume that data points which are close in the high-dimensional space should also be close {\em on the curve} (we would like images which look similar to be close together in our developmental trajectory).
 %
@@
 %
-To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored according to the distance between the data points, so that close data points have dark/strong
-edges connecting them.
-%
-Visually, these dark edges delineate the one-dimensional curve; in contrast, the light edges are less informative
-about the low-dimensional structure.
+To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored according to the distance between the data points, so that pairs of points which are close have dark/strong edges connecting them.
+%
+Visually, the dark edges delineate the one-dimensional curve, whereas the light edges are less informative about the low-dimensional structure.
 %
@@
 %
-This procedure has been named diffusion maps \cite{coifman2005geometric}.
+%This procedure has been named diffusion maps \cite{coifman2005geometric}.
 %
@@
 
-Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we can use diffusion maps to order them in time.
+Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we use diffusion maps to order them in time.
 %
@@
 %
-Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non-trivial) eigenvector, $\phi_2$.
+Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the leading diffusion maps coordinate, $\phi_2$.
 %
@@
 %
-The data has been ordered by $\psi_{1, 3}$, the first (non-trivial) vector diffusion maps coordinate.
+The data has been ordered by $\psi_{1, 3}$, the leading vector diffusion maps coordinate.
 
@@
 %
-The results can be visually compared through the sequences of one-dimensional profiles (similar to the format in Figure~\ref{fig:1d_demo}), shown in Figures~\ref{subfig:1d_membranes}~and~\ref{subfig:1d_vdm}.
+The two methods for registration and ordering can be visually compared through the sequences of one-dimensional profiles extracted from the circumference of the embryos (similar to the format in Figure~\ref{fig:1d_demo}), shown in Figures~\ref{subfig:1d_membranes}~and~\ref{subfig:1d_vdm}.
 %
 The general trends are consistent:
-both green  and red peaks are aligned, and the red peaks grow as a function of time (or of the leading vector diffusion map coordinate).
+both green  and red peaks are aligned, and the red peaks grow as a function of time/leading vector diffusion map coordinate.
 %
@@
 %
-Quantitatively, the rank correlation coefficient between these two temporal orderings is 0.9149, indicating that our automatic ordering is consistent with the ordering obtained from cellularization.
+Quantitatively, the rank correlation coefficient between these two temporal orderings is 0.9158, indicating that our automatic ordering is consistent with the ordering obtained from cellularization.
 
@@
 %
-We, instead, use diffusion maps, to construct a one-dimensional embedding of the data.
+We, instead, use diffusion maps to construct a one-dimensional parameterization of the data.
 %
@@
 
-The task of image registration has been widely studied, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
+The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
 %
@@
 %
-For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then register images by aligning these features.
-%
-Images from a construction site over time have been temporally ordered, but these methods are also based on first identifying features within the images \cite{schindler2007inferring}.
+For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then register images by aligning these features \cite{zhao2003face}.
+%
+Temporally ordering images of a construction site over time is also based on first identifying features within the images \cite{schindler2007inferring}.
 %
@@
 
-Angular synchronization and vector diffusion maps require no feature points, and no template function.
+Angular synchronization and vector diffusion maps require no feature points and no template function.
 %
@@
 %
-In contrast, our data contains variation from the different embryo orientations, as well as variation from the underlying developmental dynamics (which underpins the development of vector diffusion map algorithms).
-%
-%Other synchronization algorithms have been proposed \cite{bandeira2014multireference}.
-%
-These methods are formulated for general symmetry groups, and one might consider other symmetries, such as reflections and scaling/dilation.
-%
-The main requirement is that the symmetry group must be represented by orthogonal or unitary matrices \cite{singer2013spectral}.
+In contrast, our data contains variation from the different embryo orientations, as well as variation from the underlying developmental dynamics.
+% (which underpins the development of vector diffusion map algorithms).
+%
+In addition to the eigenproblem relaxation used here, other synchronization algorithms have been proposed \cite{bandeira2014multireference}.
+%
+Synchronization is formulated for general symmetry groups, and depending on the data, one might consider other symmetries, such as reflections and scaling/dilation.
+%
+The main requirement is that the symmetry group be represented by orthogonal or unitary matrices \cite{singer2013spectral}.
 
@@
 %
-Such functions include the scattering transform \cite{mallat2012group} and the bispectrum \cite{zhao2014rotationally}.
-%
-These have the disadvantage that a we do not obtain a registered set of images, but only a temporal ordering of the images.
-%
-However, such methods can often be faster for large data sets, since the functions are computed per image, whereas synchronization requires the computation of all pairwise alignments.
+Such functions include the scale invariant feature transform \cite{lowe1999object, lowe2004distinctive}, the scattering transform \cite{mallat2012group}, and the bispectrum \cite{zhao2014rotationally}.
+%
+These methods have the disadvantage that we do not obtain a registered set of images, but only a temporal ordering.
+%
+However, such methods can often be faster for large data sets, since the functions are computed per image, rather than computing pairwise quantities in synchronization.
 
@@
 %
-
-Our goal is to obtain estimates of $\theta_1, \dots, \theta_m$.
+Our goal is to recover $\theta_1, \dots, \theta_m$.
 %
@@
 Alternatively, for continuous signals, an optimization algorithm
-can be used to find the solution to \eqref{eq:opt_angle} \cite{ahuja2007template}.
-
-Rather than work with the angles $\theta_{ij}$ directly, it is more convenient to consider work with the rotation matrices,
+can be used \cite{ahuja2007template}.
+
+Rather than work with the angles $\theta_{ij}$ directly, it is more convenient to consider the	 rotation matrices,
 \begin{equation} \label{eq:R_theta}
@@
 %
-It follows directly from \eqref{eq:H_low_rank}  that the top block eigenvector of $H$, %which we will denote $\hat{R}$,
-contains our best estimates of $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$.
+It follows directly from \eqref{eq:H_low_rank} that the top block eigenvector of $H$ contains our best estimates of $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$.
 %
@@
 \end{equation}
-where $\hat{R}_i \in \mathbb{R}^{d \times d}$ is the estimate for $R(\theta_i)$.
-%
-However, $\hat{R}_i$ is not guaranteed to be an orthogonal matrix;
-the closest orthogonal matrix, $R_{i, est}$, is given by
+where $\hat{R}_i \in \mathbb{R}^{d \times d}$ is (nearly) the estimate for $R(\theta_i)$.
+%
+To obtain our estimate of $R(\theta_i$), denoted $R_{i, est}$, we project $\hat{R}_i$ onto the closest orthogonal matrix, 
 \begin{equation} \label{eq:R_est}
@@
 
-We find a coordinate $y$ such that $y(x_i)$ and $y(x_j)$ are close if $W_{ij}$ is large by solving following optimization problem \cite{Belkin2003}
+To find this coordinate $y$, we want solve the following optimization problem \cite{Belkin2003}
 \begin{equation} \label{eq:dmaps_opt_problem}
@@
 %
-To solve \eqref{eq:dmaps_opt_problem}, we compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{m} W_{ij}$, and the matrix $A$, where
+We first compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{m} W_{ij}$, and the matrix $A$, where
 \begin{equation} \label{eq:dmaps_A}
@@
 %
-We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_m$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_m$ and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$.
+We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_m$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$.
 %
@@
 
-Given data points $x_1, \dots, x_m$, for vector diffusion maps, one first constructs the matrix $S \in \mathbb{R}^{md \times md}$, with the $i,j^{th}$ block of $S$, $S_{ij}$, defined as
+In vector diffusion maps, given data points $x_1, \dots, x_m$, one first constructs the matrix $S \in \mathbb{R}^{md \times md}$, with the $i,j^{th}$ block of $S$, $S_{ij}$, defined as
 \begin{equation} \label{eq:vdm_S}
@@
 %
-In the language of symmetry groups, this distance is a metric between the orbits of the data points induced by the symmetry group.
-
-One then computes the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_{md}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{md}$ of $S$, and orders them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$.
+In the language of symmetry groups, this distance is a metric between the orbits induced by the symmetry group.
+
+One then computes the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_{md}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{md}$ of $S$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$.
 %
 These eigenvectors contain information about {\em both} the optimal rotations (the ``synchronization" component) and the
-variation of the data {\em after} the spatial symmetries have been factored out - in our case, their temporal variation.
+variation of the data {\em after} the spatial symmetries have been factored out (in our case, their temporal variation).
 %
@@
 %
-We then expect some subsequent eigenvector to contain information relevant to the ordering of the data,
-i.e., about the main direction(s) of data variability modulo the geometric symmetries.
+We then expect subsequent eigenvectors to contain information about the main direction(s) of data variability modulo the geometric symmetries.
 %%%
@@
 %
-Our example is, by physical considerations, a separable one; if no separability claims can be made, a second diffusion map step, using $\psi_{k,l}$ as the coordinates, may uncover a
+Our example is, by physical considerations, a separable one, and so sorting our data by $\psi_{1,d+1}$ will order our data in time.
+%
+If no separability claims can be made, a second diffusion map step, using $\psi_{k,l}$ as the coordinates, may uncover a
 useful parameterization of the data manifold modulo symmetries.
@@
 
-Registering images with respect to rotations {\em and translations} requires some additional care.
+Using angular synchronization or vector diffusion maps to register images with respect to rotations {\em and translations} requires some additional care.
 %
@@
 %
-Practically, we have planar square images discretized as pixels (rather than continuous functions on the plane).
-%
-We therefore solve discrete pairwise optimization problems: for each image pair $I_i$ and $I_j$ we find
+The first step (similar to the case of only rotations) is to compute the optimal alignments between pairs of images.
+%
+Practically, we have square images discretized as pixels (rather than continuous functions on the plane).
+%
+For each image pair $I_i$ and $I_j$ we compute
 \begin{equation}\label{eq:opt_pairwise}
@@
 \end{equation}
-where $g(I_j, \theta, dx, dy)$ is the image that results from first rotating the image $I_j$ around the center of the square by $\theta$ degrees, then translating the resulting
-image by $dx$ pixel widths in the $x$ direction, and finally translating by $dy$ pixel widths in the $y$ direction.
-%
-The norm we use, $\| \cdot \|$, is the Euclidean norm between the pixel intensities for both the red and green channels.
-%
-The domain of the image (a square) is not invariant to our rotations and translations; however, the pixels close to the border of the square have zero intensity, and so the norm can be computed as long as the main image (the embryo) does not move out of the original square.
+where $g(I_j, \theta, dx, dy)$ is image $I_j$ rotated around the center of the square by $\theta$ radians, then translated by $dx$ pixel widths in the $x$ direction, and finally translated by $dy$ pixel widths in the $y$ direction.
+%
+The norm, $\| \cdot \|$, is the Euclidean norm between the pixel intensities for both the red and green channels.
+%
+The domain of the image (a square) is not invariant to our rotations and translations; however, the pixels near the border of the square have zero intensity, and so the norm can be computed as long as the main image does not move out of the original square.
 %
@@
 %
-This also justifies the $\Delta$ limits in the translation: 
-in \eqref{eq:opt_pairwise}, $\Delta$ is chosen such that only edge pixels with little to no signal are shifted, and the main image is not split or separated when translating (we take $\Delta=20$, which corresponds to a 20\% shift in the image).%; it does not make sense to shift the images so much that
-%the embryo would ``move out" of the original square.
+This also justifies the $\Delta$ limits in \eqref{eq:opt_pairwise}: $\Delta$ is chosen such that the main image is not split or separated when translating (we take $\Delta=10$, which corresponds to a 10\% shift in the image).
+%; it does not make sense to shift the images so much that the embryo would ``move out" of the original square.
 %
@@
 %
-The missing pixels in the rotated image are taken to have zero intensity.
-%
-The translations are implemented by shifting the pixels (for computational speed, we used periodic boundary conditions).
-%
-We only consider shifts that correspond to an integer number of pixels, to remove the need for interpolation when doing the translations.
+The missing pixels in the corners of the rotated image are taken to have zero intensity.
+%
+The translations are implemented by shifting the pixels;
+we only consider shifts that correspond to an integer number of pixels to remove the need for interpolation when doing the translations.
 %
@@
 %
-Our discrete translation steps are $1/10^{th}$ of the domain, and our discrete rotations steps are $1/20^{th}$ of $2 \pi$. 
+Our discrete translation steps are $\Delta/5$, and our discrete rotations steps are $\pi/10$. 
 %
@@
 %
-We check all possible combinations for the rotation and translations that best align $I_j$ to $I_i$.
+%We check all possible combinations for the rotation and translations that best align $I_j$ to $I_i$.
 %
@@
 
-The requirements for the synchronization algorithm (i.e., for \eqref{eq:H_low_rank} to be satisfied) include the condition that the matrices used to represent the symmetry group be orthogonal \cite{singer2013spectral}.
+The synchronization formulation requires that the matrices used to represent the symmetry group be orthogonal \cite{singer2013spectral}.  
+%(for \eqref{eq:H_low_rank} to be satisfied).
 %
@@
 %
-Instead, we {\em approximately} represent rotations and translations in two-dimensions using {\em three-dimensional rotation matrices}, by projecting the (two-dimensional) image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular} (see {\it SI Appendix}).
+Instead, we {\em approximately} represent rotations and translations in two dimensions using {\em three-dimensional rotation matrices}, by projecting the (two-dimensional) image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular} (see {\it SI Appendix}).
 %
@@
 
-These rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$ are used in formulating the matrix $H$ in \eqref{eq:H_to_R} for angular synchronization, or the matrix $S$ in \eqref{eq:vdm_S} for vector diffusion maps.
-%
-We can then compute the (approximately) optimal rotations (and, in the case of vector diffusion maps, temporal orderings) for the images.
+These rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$ are used in formulating the matrix $H$ in \eqref{eq:H_to_R} for angular synchronization, or the matrix $S$ in \eqref{eq:vdm_S} for vector diffusion maps to compute the rotations (and, in the case of vector diffusion maps, temporal orderings) for the images.
 %
@@
 %
-We first multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began;
+We first multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
+%
 $R_{i,est}^T$ then gives the rotation matrix which registers image $i$.

### 2014-04-10 image_analysis_paper/pnas_draft.tex
added more edits to paper

@@
 
-We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set whose features resemble the image features in Figure~\ref{subfig:images_unordered}.
+We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set whose features resemble the image features in Figure~\ref{fig:images_unordered}.
 %
@@
 %
-We construct concentration profiles defined on a ring, and we assume that each ring is arbitrarily rotated around its center; an example is shown in Figure~\ref{subfig:1d_example}.
-%
-Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown in the bottom of Figure~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
+We construct concentration profiles defined on a ring (as in Figure~\ref{subfig:one_image}), and we assume that each ring is arbitrarily rotated around its center; an example is shown in Figure~\ref{subfig:1d_example}.
+%
+Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown at the bottom of Figure~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
+%
+{\bf We should discuss/mention that the eggs could be flipped.}
 %
@@
 %
-We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_n$, the group of integers modulo $n$).
+We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_{100}$, the group of integers modulo $100$).
 %
@@
 %
+{\bf It is not clear how we are rotating the arrows (around the center vs. around the tail).}
+%
 From this large collection of pairwise alignments, the algorithm then selects a set of rotation angles, $\theta_i$ (one for each data point), such that differences in rotation angles, $\theta_i-\theta_j$, are {\em most consistent} (in some metric) with the pairwise alignments $\theta_{ij}$.
@@
 The top eigenvector(s) of a matrix (which contains information about all pairwise alignments)
-provides the rotation angles assigned to each data point (see {\it Materials and Methods}).
+provides the rotation angles (up to a constant rotation) assigned to each data point (see {\it Materials and Methods}).
 %
@@
 %
-It is reassuring that the peaks are aligned,
+It is reassuring that the peaks appear aligned,
 even though the algorithm uses no specific information about the existence or location of such peaks.
@@
 %
-Our main assumptions are that the high-dimensional data (images have one dimension per color pixel intensity, and concentration profiles have one dimension per discretization point) lie on a one-dimensional, nonlinear curve, {\em and} that this curve is parameterized by time.
-%
-Figure~\ref{subfig:dmaps_edges} shows an schematic of a two-dimensional data set (the ``high-dimensional" data) which lies on a one-dimensional nonlinear curve (the ``low-dimensional" manifold).
+Our main assumptions are that the high-dimensional data (one dimension per discretization point in the concentration profiles/one dimension per color pixel intensity in the images) lie on a one-dimensional, nonlinear curve, {\em and} that this curve is parameterized by time.
+%
+Figure~\ref{subfig:dmaps_edges} shows a schematic of a two-dimensional data set (the ``high-dimensional" data) which lies on a one-dimensional nonlinear curve (the ``low-dimensional" manifold).
 %
@@
 %
-Because our data are very high dimensional, we cannot extract this curve by eye.
+When our data are very high dimensional, we cannot extract this curve by eye.
 %
@@
 %
-Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the leading diffusion map coordinate.
+Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the leading (non-trivial) diffusion map coordinate.
 %
@@
 
-To uncover the diffusion maps parameterization, we assume that data points which are close in the high-dimensional space should also be close {\em on the curve} (we would like images which look similar to be close together in our developmental trajectory).
+To uncover the diffusion map parameterization, we assume that data points which are close in the high-dimensional space should also be close {\em on the curve} (we would like images which look similar to be close together in our developmental trajectory).
 %
@@
 %
-To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored according to the distance between the data points, so that pairs of points which are close have dark/strong edges connecting them.
+To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored based on the distance between the data points (using the diffusion kernel of the distance), so that pairs of points which are close have dark/strong edges connecting them.
 %
@@
 %
-Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the leading diffusion maps coordinate, $\phi_2$.
+Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the leading diffusion map coordinate, $\phi_2$.
 %
@@
 %
-Angular synchronization (for registration) can be combined with diffusion maps (for temporal ordering) in a single eigencomputation that allows us to {\em simultaneously} recover the alignments {\em and} the temporal ordering.
+Angular synchronization (for registration) can be combined with diffusion maps (for temporal ordering) in a single eigencomputation that allows us to {\em simultaneously} recover the registrations {\em and} the temporal ordering.
 %
@@
 %
-The data has been ordered by $\psi_{1, 3}$, the leading vector diffusion maps coordinate.
+The data has been ordered by $\psi_{1, 3}$, the leading vector diffusion map coordinate.
 
@@
 
-The same methodologies (with small modifications) can be applied directly to our two-dimensional images.
+The same methods (with small modifications) can be applied directly to our two-dimensional images.
 %
@@
 %
-%Furthermore, the membranes stop growing towards the end of this trajectory, making the ordering of the late signals in Figure~\ref{subfig:1d_membranes} less accurate.
+Furthermore, the membranes stop growing towards the end of this trajectory, making the ordering of the late signals in Figure~\ref{subfig:1d_membranes} less accurate.
 %
@@
 \caption{The images in Figure~\ref{subfig:images_unordered}, aligned and temporally ordered using vector diffusion maps.
-The images are sorted according to the value of the first non-trivial vector diffusion maps component, $\psi_{1, 4}$.}
+The images are sorted according to the value of the first non-trivial vector diffusion map component, $\psi_{1, 4}$.}
 %\label{fig:images_ordered}
@@
 
-We have shown how symmetry and dimensionality reduction techniques can greatly accelerate the registration and temporal ordering of cross-sectional developmental imaging data.
+We have shown how symmetry and dimensionality reduction techniques can successfully register and order cross-sectional developmental imaging data.
 %
@@
 %
-Gene expression data from multiple patients has been studied in \cite{gupta2008extracting, qiu2011discovering};
+Gene expression data from multiple patients have been studied in \cite{gupta2008extracting, qiu2011discovering};
 here, each data point is a vector containing the expression levels of the various genes.
@@
 The majority of these works temporally order the data either by solving a traveling salesman problem on the data, or by constructing a minimum spanning tree of the data,
-expecting that these structures (the path of the travelling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate picture of the dynamical progression.
+expecting that these structures (the path of the travelling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate encoding of the dynamical progression.
 %
@@
 They have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}, and so we suspect they may perform better than the ordering algorithms used in previous work.
+%
+{\bf Yannis says ``careful'', he thinks this might be too strong of a statement.}
 %%%
@@
 %
-The main point to note here is that many registration algorithms require the locations of landmarks or features within each image.
+The main point to note here is that many registration algorithms require definition of appropriate landmarks or features for the images and the location of these within each image.
 %
@@
 However, these methods can suffer when the images are noisy, and the success of such methods can depend strongly on the template function \cite{shatsky2009method}.
+%
+{\bf Need to discuss consistency.}
 
@@
 %
-The main requirement is that the symmetry group be represented by orthogonal or unitary matrices \cite{singer2013spectral}.
+Here, the main requirement is that the symmetry group be represented by orthogonal or unitary matrices \cite{singer2013spectral}.
 
@@
 %
-These functions are then used as the data in the diffusion maps calculation; since the functions do not change under translations and rotations, the resulting diffusion maps embedding is invariant to symmetries.
+These functions are then used as the data in the diffusion maps calculation; since these representative functions do not change under translations and rotations, the resulting diffusion map embedding is invariant to symmetries.
 %
@@
 %
-If we only used the green channel in our analysis, we could not temporally order the data using diffusion maps, as this signal does not change appreciably in time.
-%
-Both the red and green channels contain spatial variability: the green channel has one peak at a specific angular around the embryo, and the red channel has two peaks which grow in time.
-%
-This allows us to register the data points.
+If we only used the green channel in our analysis, we could not successfully order the data using diffusion maps, as this signal does not change appreciably in time.
+%
+Both the red and green channels contain spatial variability which allows us to register the data points: the green channel has one peak at a specific angular around the embryo, and the red channel has two peaks which grow in time.
 %
@@
 %
-In these cases, our methods allow for the rapid, automated analysis of imaging data sets to help uncover complex dynamics and structure.
+In these cases, our methods allow for the rapid, automated analysis of imaging data sets to help uncover possibly complex dynamics and structure.
+{\bf Something about smoothness in making movies}
+%
+%{\bf procrustes referece}
+%{\bf allso the guy form Donoho who worried about how much to do syncjhronization}
 
@@
 
-\section{Angular synchronization for rotations\cite{singer2011angular}}
+\section{Angular synchronization \cite{singer2011angular}}
 
 Let $x_1, \dots, x_m$ denote the signals that we wish to align with respect to rotations;
-%
-Each signal is a function defined on the unit circle (on the plane).
+each signal is a function defined on the unit circle (on the plane).
 %
@@
 \end{equation}
-which we can think of as operating on the points of the unit circle on which our signal is defined.
+which we can think of as operating on the points of the unit circle (on the plane) on which our signal is defined.
 %
@@
 
-\section{Ordering using Diffusion Maps \cite{coifman2005geometric}}
+\section{Diffusion maps \cite{coifman2005geometric}}
 
@@
 \begin{equation} \label{eq:dmaps_opt_problem}
-\argmin_{y} \sum_{ij} W_{ij} (y(x_i) - y(x_j))^2
+\argmin_{y} \sum_{ij} W_{ij} (y(x_i) - y(x_j))^2.
 \end{equation}
@@
 
-\section{Vector Diffusion Maps \cite{singer2012vector}}
+\section{Vector diffusion maps\cite{singer2012vector}}
 
@@
 
-\section{Registering Images with Respect to Translations and Rotations} \label{subsec:trans_rot_register}
-
-Using angular synchronization or vector diffusion maps to register images with respect to rotations {\em and translations} requires some additional care.
+\section{Registering images with respect to translations and rotations} \label{subsec:trans_rot_register}
+
+Using angular synchronization or vector diffusion maps to register images with respect to rotations {\em and translations} requires some additional effort.
 %
@@
 %
-The domain of the image (a square) is not invariant to our rotations and translations; however, the pixels near the border of the square have zero intensity, and so the norm can be computed as long as the main image does not move out of the original square.
+The domain of the image (a square) is not invariant to our rotations and translations; however, the pixels near the border of the square have zero intensity, and so the norm can be meaningfully computed as long as the main image does not ``move out of'' the original square.
 %
@@
 %
-Although computationally demanding, this ``embarrassingly parallelizable" direct enumeration approach is not prohibitive.
+Although computationally demanding, this ``embarrassingly parallelizable" direct enumeration approach is not prohibitive here.
 %
@@
 %
-Inverting \eqref{eq:angle_relations} yields the corresponding approximate translations and rotation, and we round optimal translations to the nearest integer to remove the need for interpolation when translating the images.
+Inverting \eqref{eq:angle_relations} yields the desired approximate translations and rotation, and we round optimal translations to the nearest integer to remove the need for interpolation when translating the images.
 

### 2014-04-11 image_analysis_paper/make_figures_large_dataset_1d_color.m
Added more edits to paper

@@ -121,6 +121,18 @@ axis off
 
 print(sprintf('%s/illustrate_1d', im_save_dir), fmt, res);
 
+figure;
+set(gcf, 'paperposition',[0 0 8 8])
+plot(dpERK_unaligned(idx,:, 1), '-r', 'linewidth', 3);
+hold on
+plot(dpERK_unaligned(idx,:, 2), '-g', 'linewidth', 3);
+set(gca, 'xtick', [])
+set(gca, 'ytick', [])
+xlabel('position', 'fontsize', fontsize)
+ylabel('intensity', 'fontsize', fontsize)
+print(sprintf('%s/illustrate_1d_intensity', im_save_dir), fmt, res);
+
+return
 %%
 
 dim = 2;

### 2014-04-14 image_analysis_paper/pnas_draft.tex
Added more edits to paper
Moved things to supplement for paper

@@
 %\def\s{\sigma}
+
+%\DeclareMathSizes{9}{8}{7}{7}
 
@@
 
-\title{Unscrambling arrested development: temporal ordering and registration of cross-sectional imaging data}
+\title{Temporal ordering and registration of cross-sectional imaging data}
 
@@
 %
-We present an automatic approach to register {\em and} temporally order cross-sectional data sets of images.
+We present an automatic approach to register {\it and} temporally order cross-sectional data sets of images.
 %
@@
 vector diffusion maps for simultaneously performing both tasks) are applicable to a wide variety of data sets and
-require little {\em a priori} knowledge of the image features or the developmental dynamics.
-%
-We demonstrate the utility of our methods using a collection of images from a study of {\em Drosophila} embryogenesis.
+require little {\it a priori} knowledge of the image features or the developmental dynamics.
+%
+We demonstrate the utility of our methods using a collection of images from a study of {\it Drosophila} embryogenesis.
 \end{abstract}
@@
 \includegraphics[width=8.4cm]{fig2}
-\caption{(a) Fluorescent images of {\em Drosophila} embryos.
+\caption{(a) Fluorescent images of {\it Drosophila} embryos.
 Each embryo has been stained for Dorsal (green) and dpERK (red).
@@
 %
-We will use {\em angular synchronization} \cite{singer2011angular}, a registration algorithm that is robust to noise, to align the images in a consistent way.
-%
-We will use {\em diffusion maps} \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
-%
-We will also show how both steps can be combined using {\em vector diffusion maps} \cite{singer2012vector}.
+We will use {\it angular synchronization} \cite{singer2011angular}, a registration algorithm that is robust to noise, to align the images in a consistent way.
+%
+We will use {\it diffusion maps} \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
+%
+We will also show how both steps can be combined using {\it vector diffusion maps} \cite{singer2012vector}.
 %
@@
 %
-%Because each image is in a different orientation, we need to first align or {\em register} the images before computing distances.
+%Because each image is in a different orientation, we need to first align or {\it register} the images before computing distances.
 %
@@
 
-We begin by registering the set of profiles using angular synchronization\cite{singer2011angular}, a technique which is robust to noise and requires no {\em a priori} knowledge of image features.
+We begin by registering the set of profiles using angular synchronization\cite{singer2011angular}, a technique which is robust to noise and requires no {\it a priori} knowledge of image features.
 %
@@
 %
-From this large collection of pairwise alignments, the algorithm then selects a set of rotation angles, $\theta_i$ (one for each data point), such that differences in rotation angles, $\theta_i-\theta_j$, are {\em most consistent} (in some metric) with the pairwise alignments $\theta_{ij}$.
+From this large collection of pairwise alignments, the algorithm then selects a set of rotation angles, $\theta_i$ (one for each data point), such that differences in rotation angles, $\theta_i-\theta_j$, are {\it most consistent} (in some metric) with the pairwise alignments $\theta_{ij}$.
 %
@@
 %
-%Because the method considers {\em all} pairwise alignments, angular synchronization is often robust to noise.
+%Because the method considers {\it all} pairwise alignments, angular synchronization is often robust to noise.
 %
@@
 %
-Our main assumptions are that the high-dimensional data (one dimension per discretization point in the concentration profiles/one dimension per color pixel intensity in the images) lie on a one-dimensional, nonlinear curve, {\em and} that this curve is parameterized by time.
+Our main assumptions are that the high-dimensional data (one dimension per discretization point in the concentration profiles/one dimension per color pixel intensity in the images) lie on a one-dimensional, nonlinear curve, {\it and} that this curve is parameterized by time.
 %
@@
 
-To uncover the diffusion map parameterization, we assume that data points which are close in the high-dimensional space should also be close {\em on the curve} (we would like images which look similar to be close together in our developmental trajectory).
-%
-%We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\em as close as possible} in the parameterization.
+To uncover the diffusion map parameterization, we assume that data points which are close in the high-dimensional space should also be close {\it on the curve} (we would like images which look similar to be close together in our developmental trajectory).
+%
+%We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\it as close as possible} in the parameterization.
 %
@@
 %
-Angular synchronization (for registration) can be combined with diffusion maps (for temporal ordering) in a single eigencomputation that allows us to {\em simultaneously} recover the registrations {\em and} the temporal ordering.
+Angular synchronization (for registration) can be combined with diffusion maps (for temporal ordering) in a single eigencomputation that allows us to {\it simultaneously} recover the registrations {\it and} the temporal ordering.
 %
@@
 %
-To facilitate the application of vector diffusion maps, we approximate the two-dimensional translations and rotations by {\em three-dimensional rotations} (symmetry group $SO(3)$; see {\it Materials and Methods}, {\it SI Appendix}).
+To facilitate the application of vector diffusion maps, we approximate the two-dimensional translations and rotations by {\it three-dimensional rotations} (symmetry group $SO(3)$; see {\it SI Appendix}).
 %
@@
 %
-Although we illustrated our methods using a specific data set from a study of {\em Drosophila} embryogenesis, the framework is quite general and requires little knowledge of the image features or the developmental dynamics.
-%
-To our knowledge, temporal ordering of {\em Drosophila} images has not been automated to this extent.
-%
-Recently, machine learning algorithms have been used to group {\em Drosophila} images into different developmental classes \cite{yuan2014automated}, yet there, the appropriate classes must be defined by the user and hand-labeled training data are required.
+Although we illustrated our methods using a specific data set from a study of {\it Drosophila} embryogenesis, the framework is quite general and requires little knowledge of the image features or the developmental dynamics.
+%
+To our knowledge, temporal ordering of {\it Drosophila} images has not been automated to this extent.
+%
+Recently, machine learning algorithms have been used to group {\it Drosophila} images into different developmental classes \cite{yuan2014automated}, yet there, the appropriate classes must be defined by the user and hand-labeled training data are required.
 
@@
 The majority of these works temporally order the data either by solving a traveling salesman problem on the data, or by constructing a minimum spanning tree of the data,
-expecting that these structures (the path of the travelling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate encoding of the dynamical progression.
+expecting that these structures (the path of the traveling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate encoding of the dynamical progression.
 %
@@
 %
-In general, appropriate features may not be known {\em a priori}, and so we require a registration algorithm that does not rely on predefined features, but only on the inherent geometric symmetry of the problem.
+In general, appropriate features may not be known {\it a priori}, and so we require a registration algorithm that does not rely on predefined features, but only on the inherent geometric symmetry of the problem.
 %
@@
 %
-These methods have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}, where each image is a picture of (essentially) the {\em same} molecule; variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
+These methods have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}, where each image is a picture of (essentially) the {\it same} molecule; variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
 %
@@
 %
-First assume that each signal $x_i$ is a {\em noisy} rotated copy of the underlying signal $x_{true}$
-(which we are {\em not} given), such that
+First assume that each signal $x_i$ is a {\it noisy} rotated copy of the underlying signal $x_{true}$
+(which we are {\it not} given), such that
 \begin{equation}
@@
 %
-We can obtain an {\em estimate} of $\theta_i - \theta_j$ by computing the rotation that optimally aligns $x_j$ to $x_i$,
+We can obtain an {\it estimate} of $\theta_i - \theta_j$ by computing the rotation that optimally aligns $x_j$ to $x_i$,
 i.e., %$\theta_{ij} \approx \theta_i - \theta_j$, where
@@
 discrete shift.
-
-%Furthermore, this formulation also considers {\em higher-order} consistency information.
+%
+Details of registering two-dimensional images using synchronization are given in the {\it SI Appendix}.
+
+%Furthermore, this formulation also considers {\it higher-order} consistency information.
 %
@@
 %\end{equation}
-%therefore, {\em all} infomation of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$.
+%therefore, {\it all} infomation of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$.
 %
@@
 \begin{equation} \label{eq:dmaps_W}
-%W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right)
-W_{ij} = e^{ -\frac{d^2(x_i, x_j)}{\epsilon^2}}
+W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right)
+%W_{ij} = e^{ -\frac{d^2(x_i, x_j)}{\epsilon^2}}
 \end{equation}
@@
 %
-It is important to note that distance $d(x_i, x_j)$ used in the diffusion kernel in \eqref{eq:dmaps_W} is the distance between data points {\em after} after pairwise alignment, i.e., the minimum distance between all possible shifts of the two data points (which is obtained in \eqref{eq:opt_pairwise}).
+It is important to note that distance $d(x_i, x_j)$ used in the diffusion kernel in \eqref{eq:dmaps_W} is the distance between data points {\it after} after pairwise alignment, i.e., the minimum distance between all possible shifts of the two data points (which is obtained in \eqref{eq:opt_pairwise}).
 %
@@
 %
-These eigenvectors contain information about {\em both} the optimal rotations (the ``synchronization" component) and the
-variation of the data {\em after} the spatial symmetries have been factored out (in our case, their temporal variation).
+These eigenvectors contain information about {\it both} the optimal rotations (the ``synchronization" component) and the
+variation of the data {\it after} the spatial symmetries have been factored out (in our case, their temporal variation).
 %
@@
 
+\end{materials}
+
+%% Optional Appendix or Appendices
+%% \appendix Appendix text...
+%% or, for appendix with title, use square brackets:
+%% \appendix[Appendix Title]
+
+\begin{acknowledgments}
+- text of acknowledgments here, including grant info -
+\end{acknowledgments}
+
+%% PNAS does not support submission of supporting .tex files such as BibTeX.
+%% Instead all references must be included in the article .tex document.
+%% If you currently use BibTeX, your bibliography is formed because the
+%% command \verb+\bibliography{}+ brings the <filename>.bbl file into your
+%% .tex document. To conform to PNAS requirements, copy the reference listings
+%% from your .bbl file and add them to the article .tex file, using the
+%% bibliography environment described above.
+
+%%  Contact pnas@nas.edu if you need assistance with your
+%%  bibliography.
+
+% Sample bibliography item in PNAS format:
+%% \bibitem{in-text reference} comma-separated author names up to 5,
+%% for more than 5 authors use first author last name et al. (year published)
+%% article title  {\it Journal Name} volume #: start page-end page.
+%% ie,
+% \bibitem{Neuhaus} Neuhaus J-M, Sitcher L, Meins F, Jr, Boller T (1991)
+% A short C-terminal sequence is necessary and sufficient for the
+% targeting of chitinases to the plant vacuole.
+% {\it Proc Natl Acad Sci USA} 88:10362-10366.
+
+%% Enter the largest bibliography number in the facing curly brackets
+%% following \begin{thebibliography}
+
+\bibliographystyle{pnas}
+\bibliography{background_reading/references,../../references/references}
+
+%\begin{thebibliography}{}
+
+%\end{thebibliography}
+
+\end{article}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%% Adding Figure and Table References
+%% Be sure to add figures and tables after \end{article}
+%% and before \end{document}
+
+%% For figures, put the caption below the illustration.
+%%
+%% \begin{figure}
+%% \caption{Almost Sharp Front}\label{afoto}
+%% \end{figure}
+
+%% For Tables, put caption above table
+%%
+%% Table caption should start with a capital letter, continue with lower case
+%% and not have a period at the end
+%% Using @{\vrule height ?? depth ?? width0pt} in the tabular preamble will
+%% keep that much space between every line in the table.
+
+%% \begin{table}
+%% \caption{Repeat length of longer allele by age of onset class}
+%% \begin{tabular}{@{\vrule height 10.5pt depth4pt  width0pt}lrcccc}
+%% table text
+%% \end{tabular}
+%% \end{table}
+
+%% For two column figures and tables, use the following:
+
+%% \begin{figure*}
+%% \caption{Almost Sharp Front}\label{afoto}
+%% \end{figure*}
+
+%% \begin{table*}
+%% \caption{Repeat length of longer allele by age of onset class}
+%% \begin{tabular}{ccc}
+%% table text
+%% \end{tabular}
+%% \end{table*}
+
+%\begin{figure}
+%\begin{minipage}{2.9cm}
+%\centering
+%\includegraphics[width=\textwidth]{illustrate_1d}\\
+%\vspace{-0.25cm}
+%(a)\\
+%\includegraphics[width=\textwidth]{registered_unordered_1d}\\
+%(d)
+%\end{minipage}
+%\begin{minipage}{2.9cm}
+%\centering
+%\includegraphics[width=\textwidth]{illustrate_1d_intensity}\\
+%(b)\\
+%\includegraphics[width=\textwidth]{registered_ordered_1d}\\
+%(e)
+%\end{minipage}
+%\begin{minipage}{2.9cm}
+%\centering
+%\includegraphics[width=\textwidth]{unregistered_unordered_1d}\\
+%(c)\\
+%\includegraphics[width=\textwidth]{registered_ordered_vdm_1d}\\
+%\centering
+%(f)
+%\end{minipage}
+%\caption{Analysis of one-dimensional profiles obtained from experimental data.
+(a) One-dimensional, two-component concentration profile on a ring (top), and the corresponding profile on a line (bottom).
+The concentration profile was extracted from an image of the type shown in Figure~\ref{fig:images_unordered}.
+(b) Intensity corresponding to the profile in (a).
+(c) An ensemble of concentration profiles, each of the form described in (a).
+Each row in the array corresponds to a single profile obtained from an image of the type in Figure~\ref{subfig:images_unordered}.
+(d) The profiles in (c), now registered using angular synchronization.
+(e) The profiles in (d), now temporally ordered using diffusion maps.
+(f) The profiles in (c), registered and temporally ordered in a single step using vector diffusion maps.}
+%\end{figure}
+%
+\begin{figure*}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{sphere_1}
+\includegraphics[width=\textwidth]{sphere2_1}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{sphere_2}
+\includegraphics[width=\textwidth]{sphere2_2}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{sphere_3}
+\includegraphics[width=\textwidth]{sphere2_3}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.2\textwidth}
+\includegraphics[width=\textwidth]{sphere_4}
+\includegraphics[width=\textwidth]{sphere2_4}
+\caption{}
+\end{subfigure}
+\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions.
+(a) The original image.
+(b) Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image.
+(c) Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation.
+(d) Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.
+The top row shows the three-dimensional spheres, and the bottom row shows the (two-dimensional) surface of the sphere in which we are interested.}
+\label{fig:SO3_picture}
+\end{figure*}
+
 \section{Registering images with respect to translations and rotations} \label{subsec:trans_rot_register}
 
-Using angular synchronization or vector diffusion maps to register images with respect to rotations {\em and translations} requires some additional effort.
-%
-%We are not only interested in registering the one-dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two-dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\em and} translations.
+Using angular synchronization or vector diffusion maps to register images with respect to rotations {\it and translations} requires some additional effort.
+%
+%We are not only interested in registering the one-dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two-dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\it and} translations.
 %
@@
 %
-Instead, we {\em approximately} represent rotations and translations in two dimensions using {\em three-dimensional rotation matrices}, by projecting the (two-dimensional) image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular} (see {\it SI Appendix}).
+Instead, we {\it approximately} represent rotations and translations in two dimensions using {\it three-dimensional rotation matrices}, by projecting the (two-dimensional) image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular}.
 %
@@
 
-\end{materials}
-
-%% Optional Appendix or Appendices
-%% \appendix Appendix text...
-%% or, for appendix with title, use square brackets:
-%% \appendix[Appendix Title]
-
-\begin{acknowledgments}
-- text of acknowledgments here, including grant info -
-\end{acknowledgments}
-
-%% PNAS does not support submission of supporting .tex files such as BibTeX.
-%% Instead all references must be included in the article .tex document.
-%% If you currently use BibTeX, your bibliography is formed because the
-%% command \verb+\bibliography{}+ brings the <filename>.bbl file into your
-%% .tex document. To conform to PNAS requirements, copy the reference listings
-%% from your .bbl file and add them to the article .tex file, using the
-%% bibliography environment described above.
-
-%%  Contact pnas@nas.edu if you need assistance with your
-%%  bibliography.
-
-% Sample bibliography item in PNAS format:
-%% \bibitem{in-text reference} comma-separated author names up to 5,
-%% for more than 5 authors use first author last name et al. (year published)
-%% article title  {\it Journal Name} volume #: start page-end page.
-%% ie,
-% \bibitem{Neuhaus} Neuhaus J-M, Sitcher L, Meins F, Jr, Boller T (1991)
-% A short C-terminal sequence is necessary and sufficient for the
-% targeting of chitinases to the plant vacuole.
-% {\it Proc Natl Acad Sci USA} 88:10362-10366.
-
-%% Enter the largest bibliography number in the facing curly brackets
-%% following \begin{thebibliography}
-
-\bibliographystyle{pnas}
-\bibliography{background_reading/references,../../references/references}
-
-%\begin{thebibliography}{}
-
-%\end{thebibliography}
-
-\end{article}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Adding Figure and Table References
-%% Be sure to add figures and tables after \end{article}
-%% and before \end{document}
-
-%% For figures, put the caption below the illustration.
-%%
-%% \begin{figure}
-%% \caption{Almost Sharp Front}\label{afoto}
-%% \end{figure}
-
-%% For Tables, put caption above table
-%%
-%% Table caption should start with a capital letter, continue with lower case
-%% and not have a period at the end
-%% Using @{\vrule height ?? depth ?? width0pt} in the tabular preamble will
-%% keep that much space between every line in the table.
-
-%% \begin{table}
-%% \caption{Repeat length of longer allele by age of onset class}
-%% \begin{tabular}{@{\vrule height 10.5pt depth4pt  width0pt}lrcccc}
-%% table text
-%% \end{tabular}
-%% \end{table}
-
-%% For two column figures and tables, use the following:
-
-%% \begin{figure*}
-%% \caption{Almost Sharp Front}\label{afoto}
-%% \end{figure*}
-
-%% \begin{table*}
-%% \caption{Repeat length of longer allele by age of onset class}
-%% \begin{tabular}{ccc}
-%% table text
-%% \end{tabular}
-%% \end{table*}
-
-%\begin{figure*}
-%\begin{subfigure}{0.2\textwidth}
-%\includegraphics[width=\textwidth]{sphere_1}
-%\includegraphics[width=\textwidth]{sphere2_1}
-%\caption{}
-%\end{subfigure}
-%\begin{subfigure}{0.2\textwidth}
-%\includegraphics[width=\textwidth]{sphere_2}
-%\includegraphics[width=\textwidth]{sphere2_2}
-%\caption{}
-%\end{subfigure}
-%\begin{subfigure}{0.2\textwidth}
-%\includegraphics[width=\textwidth]{sphere_3}
-%\includegraphics[width=\textwidth]{sphere2_3}
-%\caption{}
-%\end{subfigure}
-%\begin{subfigure}{0.2\textwidth}
-%\includegraphics[width=\textwidth]{sphere_4}
-%\includegraphics[width=\textwidth]{sphere2_4}
-%\caption{}
-%\end{subfigure}
-%\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions.
-(a) The original image.
-(b) Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image.
-(c) Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation.
-(d) Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.
-The top row shows the three-dimensional spheres, and the bottom row shows the (two-dimensional) surface of the sphere in which we are interested.}
-%\label{fig:SO3_picture}
-%\end{figure*}
-
 \end{document}

### 2014-04-16 image_analysis_paper/pnas_draft.tex
updated formatting so that fonts are correct

@@
 %\usepackage{PNASoneF}
-%\usepackage{PNAStwoF}
+\usepackage{PNAStwoF}
 
@@
 
-\usepackage{subcaption}
+%\usepackage{subcaption}
 \graphicspath{ {paper_figures/} }
+
+%\usepackage{sansmath}
+
+%\usepackage[font={sf, small}]{caption}
 
@@
 
-%\DeclareMathSizes{9}{8}{7}{7}
+\DeclareMathSizes{9}{8}{7}{7}
 
@@
 \includegraphics[width=8.4cm]{fig1}
-\caption{(a) Fish, each in a different orientation and a different stage of development.
-(b) Fish, now rotationally registered and temporally ordered.
+\caption{{\it (A)} Fish, each in a different orientation and a different stage of development.
+{\it (B)} Fish, now rotationally registered and temporally ordered.
 For this caricature, the registration and ordering is easy to do ``by hand" because the data set is small and the developmental changes are easy to recognize.}
@@
 \customlabel{fig:fish}{1}
-\customlabel{subfig:fish_unordered}{\ref{fig:fish}a}
-\customlabel{subfig:fish_ordered}{\ref{fig:fish}b}
+\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
+\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
 \end{figure}
@@
 \includegraphics[width=8.4cm]{fig2}
-\caption{(a) Fluorescent images of {\it Drosophila} embryos.
+\caption{{\it (A)} Fluorescent images of {\it Drosophila} embryos.
 Each embryo has been stained for Dorsal (green) and dpERK (red).
-(b) Selected image from data set.
-(c) Profile of Dorsal (green) and dpERK (red) around the circumference of the embryo in (b).}
+{\it (B)} Selected image from data set.
+{\it (C)} Profile of Dorsal (green) and dpERK (red) around the circumference of the embryo in {\it B}.
+}
 %\label{fig:images_unordered}
 \customlabel{fig:images_unordered}{2}
-\customlabel{subfig:images_unordered}{\ref{fig:images_unordered}a}
-\customlabel{subfig:one_image}{\ref{fig:images_unordered}b}
-\customlabel{subfig:image_profiles}{\ref{fig:images_unordered}c}
+\customlabel{subfig:images_unordered}{\ref{fig:images_unordered}{\it A}}
+\customlabel{subfig:one_image}{\ref{fig:images_unordered}{\it B}}
+\customlabel{subfig:image_profiles}{\ref{fig:images_unordered}{\it C}}
 \end{figure}
@@
 \includegraphics[width=8.4cm]{fig3}
-\caption{(a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
-(b) Intensity corresponding to the profile in (a).
-(c) An ensemble of concentration profiles, each of the form described in (a).
+\caption{{\it (A)} One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
+{\it (B)} Intensity corresponding to the profile in {\it A}.
+{\it (C)} An ensemble of concentration profiles, each of the form described in {\it A}.
 Each row in the array corresponds to a single profile.
-(d) The profiles in (c), now registered using angular synchronization.
-(e) The profiles in (d), now temporally ordered using diffusion maps.
-(f) The profiles in (c), registered and temporally ordered in a single step using vector diffusion maps.}
+{\it (D)} The profiles in {\it C}, now registered using angular synchronization.
+{\it (E)} The profiles in {\it D}, now temporally ordered using diffusion maps.
+{\it (F)} The profiles in {\it C}, registered and temporally ordered in a single step using vector diffusion maps.}
 %\label{fig:1d_demo}
 \customlabel{fig:1d_demo}{3}
-\customlabel{subfig:1d_example}{\ref{fig:1d_demo}a}
-\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}b}
-\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}c}
-\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}d}
-\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}e}
-\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}f}
+\customlabel{subfig:1d_example}{\ref{fig:1d_demo}{\it A}}
+\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}{\it B}}
+\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}{\it C}}
+\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}{\it D}}
+\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}{\it E}}
+\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}{\it F}}
 \end{figure}
@@
 %
+%
+Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization.
+%
+It is reassuring that the peaks appear aligned,
+even though the algorithm uses no specific information about the existence or location of such peaks.
+%
+%We note once more that, since angular synchronization only uses pairwise rotation information,
+%the optimal rotations are recovered up to a global rotation.
+
 \begin{figure}[t]
 \includegraphics[width=8.4cm]{fig4}
-\caption{(a) Set of vectors, each in a different orientation.
+\caption{ {\it (A)} Set of vectors, each in a different orientation.
 The pariwise alignment angles are indicated.
-(b) The vectors from (a), each rotated so that the set is globally aligned.
-Note that the chosen rotation angles are consistent with the pairwise alignments in (a).
-(c) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
+{\it (B)} The vectors from {\it A}, each rotated so that the set is globally aligned.
+Note that the chosen rotation angles are consistent with the pairwise alignments in {\it A}.
+{\it (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
 Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \eqref{eq:dmaps_W}), so that close data points are connected by dark (``stronger'') edges.
-(d) The data in (c), colored by the first (non-trivial) eigenvector from the diffusion map computational procedure, $\phi_2$.
-The color intensity is monotonic with the arclength, thus parameterizing the curve.
-}
+{\it (D)} The data in {\it C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure, $\phi_2$.
+The color intensity is monotonic with the arclength, thus parameterizing the curve.}
 %\label{fig:schematics}
 \customlabel{fig:schematics}{4}
-\customlabel{subfig:synch1}{\ref{fig:schematics}a}
-\customlabel{subfig:synch2}{\ref{fig:schematics}b}
-\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}c}
-\customlabel{subfig:dmaps_color}{\ref{fig:schematics}d}
+\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
+\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
+\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}{\it C}}
+\customlabel{subfig:dmaps_color}{\ref{fig:schematics}{\it D}}
 \end{figure}
-%
-Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization.
-%
-It is reassuring that the peaks appear aligned,
-even though the algorithm uses no specific information about the existence or location of such peaks.
-%
-%We note once more that, since angular synchronization only uses pairwise rotation information,
-%the optimal rotations are recovered up to a global rotation.
 
@@
 
-Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we use diffusion maps to order them in time.
-%
-We use the Euclidean distance between the (discretized in $\mathbb{R}^{100}$) aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances between data points.
-%
-Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the leading diffusion map coordinate, $\phi_2$.
+Given our set of registered signals, shown in Figure~\ref{subfig:1d_aligned_unordered}, we use diffusion maps to order them in time.
+%
+We use the Euclidean distance between the (discretized in $\mathbb{R}^{100}$) aligned concentration profiles (the rows of Figure~\ref{subfig:1d_aligned_unordered}) as our pairwise distances between data points.
+%
+Figure~\ref{subfig:1d_aligned_ordered} shows the data from Figure~\ref{subfig:1d_aligned_unordered}, now sorted by the value of the leading diffusion map coordinate, $\phi_2$.
 %
@@
 %
-Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps.
+Figure~\ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure~\ref{subfig:1d_unaligned_unordered} using vector diffusion maps.
 %
@@
 \includegraphics[width=8.4cm]{fig6}
-\caption{(a) Images, stained for membranes (gray) and Dorsal (green) at different levels of development.
-(b) Membrane length as a function of time.
-The points correspoinding to the images in (a) are indicated with arrows.
-(c) The one-dimensional concentration profiles, registered by peak Dorsal fluorescence and ordered by membrane length.
-(d) The one-dimensional concentration profiles corresponding to the automatically registered and ordered images in Figure~\ref{fig:images_ordered}.
+\caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different levels of development.
+{\it (B)} Membrane length as a function of time.
+The points corresponding to the images in {\it A} are indicated with arrows.
+{\it (C)} The one-dimensional concentration profiles, registered by peak Dorsal fluorescence and ordered by membrane length.
+{\it (D)} The one-dimensional concentration profiles corresponding to the automatically registered and ordered images in Figure~\ref{fig:images_ordered}.
 }
@@
 \customlabel{fig:membrane_compare}{6}
-\customlabel{subfig:membranes}{\ref{fig:membrane_compare}a}
-\customlabel{subfig:membrane_curve}{\ref{fig:membrane_compare}b}
-\customlabel{subfig:1d_membranes}{\ref{fig:membrane_compare}c}
-\customlabel{subfig:1d_vdm}{\ref{fig:membrane_compare}d}
+\customlabel{subfig:membranes}{\ref{fig:membrane_compare}{\it A}}
+\customlabel{subfig:membrane_curve}{\ref{fig:membrane_compare}{\it B}}
+\customlabel{subfig:1d_membranes}{\ref{fig:membrane_compare}{\it C}}
+\customlabel{subfig:1d_vdm}{\ref{fig:membrane_compare}{\it D}}
 \end{figure}
@@
 
-Let $x_1, \dots, x_m$ denote the signals that we wish to align with respect to rotations;
+%\sansmath
+
+Let $ x_1, \dots, x_m$ denote the signals that we wish to align with respect to rotations;
 each signal is a function defined on the unit circle (on the plane).
@@
 %
+
 \begin{figure*}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{sphere_1}
-\includegraphics[width=\textwidth]{sphere2_1}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{sphere_2}
-\includegraphics[width=\textwidth]{sphere2_2}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{sphere_3}
-\includegraphics[width=\textwidth]{sphere2_3}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{sphere_4}
-\includegraphics[width=\textwidth]{sphere2_4}
-\caption{}
-\end{subfigure}
+\begin{minipage}{0.2\textwidth}
+\centering
+\includegraphics[width=\textwidth]{sphere_1}\\
+\includegraphics[width=\textwidth]{sphere2_1}\\
+A
+\end{minipage}
+%
+\begin{minipage}{0.2\textwidth}
+\centering
+\includegraphics[width=\textwidth]{sphere_2}\\
+\includegraphics[width=\textwidth]{sphere2_2}\\
+B
+\end{minipage}
+\begin{minipage}{0.2\textwidth}
+\centering
+\includegraphics[width=\textwidth]{sphere_3}\\
+\includegraphics[width=\textwidth]{sphere2_3}\\
+C
+\end{minipage}
+\begin{minipage}{0.2\textwidth} 
+\centering
+\includegraphics[width=\textwidth]{sphere_4}\\
+\includegraphics[width=\textwidth]{sphere2_4}\\
+D
+\end{minipage}
 \caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions.
-(a) The original image.
-(b) Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image.
-(c) Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation.
-(d) Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.
+{\it (A)} The original image.
+{\it (B)} Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image.
+{\it (C)} Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation.
+{\it (D)} Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.
 The top row shows the three-dimensional spheres, and the bottom row shows the (two-dimensional) surface of the sphere in which we are interested.}
@@
 (\theta_{ij}, dx_{ij}, dy_{ij}) = \argmin_{
-\scriptsize \begin{matrix}
+\begin{matrix}
 \theta \in [0, 2\pi) \\

### 2014-04-17 image_analysis_paper/test_PCA.m
did PCA on images

@@ -1,0 +1,179 @@
+clear all
+close all
+
+%res = '-r300';
+%fmt = '-djpeg';
+
+fontsize = 10;
+
+%im_save_dir = 'paper_figures';
+
+%% directories where things are stored
+time_data = '../membrane_lengths/oct16.mat';
+image_dir = '../membrane_pictures/membrane2/dpERK_staining';
+
+%% load membrane lengths
+
+load(time_data);
+mem_lengths = L(:,1);
+
+% remove some embryos
+ind = setdiff(1:52, [9, 15, 22]);
+
+m = length(ind);
+
+mem_lengths = mem_lengths(ind);
+
+% compute ranks from membrane lengths
+ranks_from_membranes = compute_ranks(mem_lengths);
+
+%% image parameters
+
+npixels = 100;
+
+%set image plotting parameters
+subplot_dim1 = ceil(sqrt(m));
+subplot_dim2 = ceil(m / subplot_dim1);
+[Y, X] = meshgrid((subplot_dim1:-1:1)/subplot_dim1, (1:subplot_dim2)/subplot_dim2);
+
+nchannels = 3;
+image_set = zeros(npixels, npixels, nchannels, m, 'uint8');
+image_set_raw = zeros(1024, 1024, nchannels, m, 'uint8');
+
+nuclei = zeros(npixels, npixels, m, 'uint8');
+
+%% load images
+
+for i=1:m
+    % read image
+    im1 = imread(sprintf('%s/emb%02d.tif', image_dir, ind(i)));
+    
+    if i==1
+        im1 = circshift(im1, [0 150 0]);
+    end
+    image_set_raw(:, :, :, i) = im1;
+    
+    % resize image
+    im1 = imresize(im1, [npixels npixels]);
+    
+    nuclei(:, :, i) = im1(:, :, 1);
+    
+    im1(:, :, 1) = im1(:, :, 2);
+    im1(:, :, 2) = im1(:, :, 3);
+    im1(:, :, 3) = 0;
+    
+    %store image
+    image_set(:, :, :, i) = im1;
+    
+end
+
+%% raw dpERK images-- synchronization
+
+angle_proj = pi/8;
+shift_max = 10;
+shift_step = 2;
+dim = 3;
+
[…]

### 2014-05-06 image_analysis_paper/pnas_draft2.tex
Added stuff for new version of paper with two data sets

--- 2014-04-16 image_analysis_paper/pnas_draft.tex
@@
 %\usepackage{subcaption}
-\graphicspath{ {paper_figures/} }
+\graphicspath{ {paper_figures2/} }
 
@@
 
-Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are visually apparent.
-%
-As an example, Figure~\ref{fig:fish} shows a caricature of fish development, combining the processes of growth and color patterning.
-%
-In this case, temporal ordering can be accomplished by arranging the fish by size, which is monotonic with the developmental progress.
-%
-Image registration is based on obvious morphological landmarks, such as the position of head and fins.
-%
-On the other hand, real data poses nontrivial challenges, due to the number of images, measurement noise and variability, and the subtlety of the developmental changes.
-
-To illustrate the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of signaling protein expression in the early fruit fly embryo, one of the leading model systems for studies of developmental dynamics.
-%
-This dataset comprises $49$ fluorescence microscopy images of different embryos fixed during the $3^{rd}$ hour of their development.
-%
 \begin{figure}[t]
@@
 %
+Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are visually apparent.
+%
+As an example, Figure~\ref{fig:fish} shows a caricature of fish development, combining the processes of growth and color patterning.
+%
+In this case, temporal ordering can be accomplished by arranging the fish by size, which is monotonic with the developmental progress.
+%
+Image registration is based on obvious morphological landmarks, such as the position of head and fins.
+%
+On the other hand, real data poses nontrivial challenges, due to the number of images, measurement noise and variability, and the subtlety of the developmental changes.
+
+To illustrate the issues associated with temporal ordering and registration of images, we use two data sets from an imaging study of signaling protein expression in the early fruit fly embryo, one of the leading model systems for studies of developmental dynamics.
+%
 The fruit fly embryo is shaped like a rice grain, approximately $1/2$~mm long.
@@
 %
+During the $4^{th}$ hour of development, ...
+%
 Images in Figure~\ref{subfig:images_unordered} show optical sections at a fixed depth, perpendicular to the long axis of the embryo.
-%%%
-%%% we have to somehow say that it is consistent
-%%%
 %
-%%%
-%%% ask if "visualize" is typical as active verb
-%%%
 The embryos were fixed and stained with chemicals that help visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red) \cite{chung2010microfluidic}.
@@
 
-\begin{figure}[t]
-\includegraphics[width=8.4cm]{fig2}
-\caption{{\it (A)} Fluorescent images of {\it Drosophila} embryos.
-Each embryo has been stained for Dorsal (green) and dpERK (red).
-{\it (B)} Selected image from data set.
-{\it (C)} Profile of Dorsal (green) and dpERK (red) around the circumference of the embryo in {\it B}.
-}
-%\label{fig:images_unordered}
-\customlabel{fig:images_unordered}{2}
-\customlabel{subfig:images_unordered}{\ref{fig:images_unordered}{\it A}}
-\customlabel{subfig:one_image}{\ref{fig:images_unordered}{\it B}}
-\customlabel{subfig:image_profiles}{\ref{fig:images_unordered}{\it C}}
-\end{figure}
-
-%%%
-%%% orientation does not read good to Y below -
-%%% it is the same orientation, but a different rotation around the axis
-%%% how to say this ?
-%%%
 Each image in Figure~\ref{subfig:images_unordered} comes from a different embryo, captured at a different rotation about the long axis and at a different and unknown point in time.
@@
 %
-Both of these tasks can be fully automated, and even combined.
+We will demonstrate that both of these tasks can be fully automated.
+
+\begin{figure*}[t]
+\includegraphics[width=6.5cm]{raw_data1}
+\hfill
+\includegraphics[width=6.5cm]{PCA_ordered}
+\hfill
+\begin{minipage}[b]{3.3cm}
+\vspace{0in}
+\includegraphics[width=3.3cm]{PCA_eigenimage1}
+\includegraphics[width=3.3cm]{PCA_corr}
+\end{minipage}
+%\label{fig:fish}
+\customlabel{fig:raw_data}{2}
+\customlabel{subfig:raw_data1}{\ref{fig:fish}{\it A}}
+\customlabel{subfig:raw_data2}{\ref{fig:fish}{\it B}}
+\end{figure*}
 %
-We will use {\it angular synchronization} \cite{singer2011angular}, a registration algorithm that is robust to noise, to align the images in a consistent way.
-%
-We will use {\it diffusion maps} \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
-%
-We will also show how both steps can be combined using {\it vector diffusion maps} \cite{singer2012vector}.
-%
-%The paper is organized as follows.
-%
-%First, we use a synthetic dataset, where the correct orientation and ordering of data points is known, to formalize the tasks of image registration and temporal ordering.
-%
-%Then, we show that both of these tasks can be effectively combined in a single step.
-%
-%Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align them and order them correctly in time.
-%
-%We conclude by discussing the general requirements that must be satisfied for our approach to work and compare it to other approaches and ideas in cross-sectional studies of biological dynamics.
 
@@
 
-%We would like to temporally order images, such as the images shown in Figure~\ref{subfig:images_unordered}.
+\subsection{Image Registration}
+
+Because each embryo is imaged in a different orientation, the first task is to align or {\em register} the images so that they are in a consistent frame of reference.
 %
-%We will use diffusion maps, a dimensionality reduction algorithm, to automatically order our data.
+We use angular synchronization to register our images. 
 %
-%However, this algorithm requires distances between data points.
+This technique is very general, as it requires no template image or feature points. 
 %
-%Because each image is in a different orientation, we need to first align or {\it register} the images before computing distances.
+Furthermore, it has been shown to be robust to noise.
 %
-%Once all the images have been aligned, we can compute distances and use diffusion maps to temporally order the data.
+Implementation details are given in the {\it SI appendix}.
 
-%%%
-%%%  Y to discuss with S
-%%%  I think we should have the real data
-%%%
-%%%  also, these dynamics "are like" our dynamics
-%%%
-%%%   even if this is kept, one should say that this chosen because it LOOKS LIKE what we have
-%%% another reason I want to have the real data is because in the discussion I want to talk about the flips
-%%%
+Figure ... shows the images from Figure ..., registered using angular synchronization.
+%
+The important features that our eye can visually detect (e.g., the green Dorsal peaks) have been aligned, even though the algorithm does not require identification of such features.
 
-We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set whose features resemble the image features in Figure~\ref{fig:images_unordered}.
+\subsection{Temporal Ordering using Principal Component Analysis}
+
+We will use dimensionality reduction techniques to temporally order our data.
 %
-The relatively simple dynamics of this data set will allow us to easily visualize and illustrate the main points of the different algorithms.
+Although our images are inherently high-dimensional (number of pixels $\times$ number of channels), we assume that they lie on a low-dimensional curve, and that this curve is parameterized by time.
 %
-We construct concentration profiles defined on a ring (as in Figure~\ref{subfig:one_image}), and we assume that each ring is arbitrarily rotated around its center; an example is shown in Figure~\ref{subfig:1d_example}.
+
+For the images shown in Figure ...., the dynamics are relatively simple, and we can use principal component analysis (PCA) to order the images.
 %
-Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown at the bottom of Figure~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
+PCA extracts a set of {\it eigenimages} from the data, such that each data point can be (approximately) described as a linear combination of these eigenimages. 
 %
-{\bf We should discuss/mention that the eggs could be flipped.}
+In our dataset, each data point can be effectively described by a multiple of a single eigenimage, i.e., the images lie on a line in a very high-dimensional space.
 %
-Each concentration profile is a noisy Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
+We can therefore sort these images along this line to temporally order them in time.
+
+Figure ... shows the images, now ordered using PCA. 
 %
-We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_{100}$, the group of integers modulo $100$).
+The first eigenimage is also shown.
+
+In this simple example, we can validate our ordering using an independent time marker.
 %
-Figure \ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
+Figure ... shows the rank correlation betrween the two orderings.
 %
-Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.
+Rank correlation is 0.9199.
 
-\begin{figure}[t]
-\includegraphics[width=8.4cm]{fig3}
-\caption{{\it (A)} One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
-{\it (B)} Intensity corresponding to the profile in {\it A}.
-{\it (C)} An ensemble of concentration profiles, each of the form described in {\it A}.
-Each row in the array corresponds to a single profile.
-{\it (D)} The profiles in {\it C}, now registered using angular synchronization.
-{\it (E)} The profiles in {\it D}, now temporally ordered using diffusion maps.
-{\it (F)} The profiles in {\it C}, registered and temporally ordered in a single step using vector diffusion maps.}
-%\label{fig:1d_demo}
-\customlabel{fig:1d_demo}{3}
-\customlabel{subfig:1d_example}{\ref{fig:1d_demo}{\it A}}
-\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}{\it B}}
-\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}{\it C}}
-\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}{\it D}}
-\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}{\it E}}
-\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}{\it F}}
-\end{figure}
+Although PCA can successfully order the data set shown in Figure ..., it can only do so because the data lie on in a one-dimensional, {\em linear} subspace.
+%
+In general, the developmental trajectory is not guaranteed to be linear. 
+%
+This can be seen in the late images of Figure ... that the ordering is not entirely correct.
+%
+We can also see this from the projection onto the first two eigenimages, shown in Figure ...
+%
+The data clearly lie on a one-dimensional curve, but the curve is nonlinear at late times. 
+%
+This motivates us to use nonlinear dimensioanlity reduction techniques to temporally order our data.
 
-\subsection{Angular synchronization for image registration}
+\subsection{Temporal Ordering Using Diffusion Maps}
+We will use diffusion maps, a technique which is robust to noise and broadly applicable to a wide variety of data types.
+%
 
-We begin by registering the set of profiles using angular synchronization\cite{singer2011angular}, a technique which is robust to noise and requires no {\it a priori} knowledge of image features.
+We can use diffusion maps to analyze more complex data sets where the dynamics are clearly nonlinear.
 %
-Angular synchronization first aligns every possible pair of data points, i.e., it computes the rotation angles
-which make each pair have the maximum possible overlap (see {\it Materials and Methods}).
-%%%
-%%% so I do not like the figure with the vectors - it is correct, but people have to take the
-%%% function, discretize and make it a vector in their heads, and may be "one step too much"
-%%%
+Figure ... shows a data set from the $4^{th}$ hour of development.
 %
-An illustration is shown in Figure~\ref{subfig:synch1};
-each data point is a vector, and $\theta_{ij}$ denotes the angle of rotation needed to align vectors $i$ and $j$.
+The dynamics are significantly more complex. 
 %
-{\bf It is not clear how we are rotating the arrows (around the center vs. around the tail).}
+However, we can still use diffusion maps to temporally order the data.
+
+We have no time marker with which to validate our orderings in this time period. 
 %
-From this large collection of pairwise alignments, the algorithm then selects a set of rotation angles, $\theta_i$ (one for each data point), such that differences in rotation angles, $\theta_i-\theta_j$, are {\it most consistent} (in some metric) with the pairwise alignments $\theta_{ij}$.
+However, there are noticable features in the ordering that we can see are consistent with peviously known dynamics. 
 %
-Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned; the computed rotation angles are also indicated.
-%
-Note that the differences between the rotation angles assigned to each vector are consistent with the pairwise angles needed to rotate one vector to another, i.e., $\theta_i - \theta_j \approx \theta_{ij}$.
-%
-The problem of finding these alignments is formulated as an optimization problem, and then relaxed to an eigenproblem.
-%
-The top eigenvector(s) of a matrix (which contains information about all pairwise alignments)
-provides the rotation angles (up to a constant rotation) assigned to each data point (see {\it Materials and Methods}).
-%
-%Because the method considers {\it all} pairwise alignments, angular synchronization is often robust to noise.
-%
-%
-Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization.
-%
-It is reassuring that the peaks appear aligned,
-even though the algorithm uses no specific information about the existence or location of such peaks.
-%
-%We note once more that, since angular synchronization only uses pairwise rotation information,
-%the optimal rotations are recovered up to a global rotation.
+For example, ....
 
-\begin{figure}[t]
-\includegraphics[width=8.4cm]{fig4}
-\caption{ {\it (A)} Set of vectors, each in a different orientation.
-The pariwise alignment angles are indicated.
-{\it (B)} The vectors from {\it A}, each rotated so that the set is globally aligned.
-Note that the chosen rotation angles are consistent with the pairwise alignments in {\it A}.
-{\it (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
-Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \eqref{eq:dmaps_W}), so that close data points are connected by dark (``stronger'') edges.
-{\it (D)} The data in {\it C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure, $\phi_2$.
-The color intensity is monotonic with the arclength, thus parameterizing the curve.}
-%\label{fig:schematics}
-\customlabel{fig:schematics}{4}
-\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
-\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
-\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}{\it C}}
-\customlabel{subfig:dmaps_color}{\ref{fig:schematics}{\it D}}
-\end{figure}
-
-\subsection{Diffusion maps for temporal ordering}
-
-After geometrically registering the data (concentration profiles or images), we want to order them in time to reconstruct the developmental dynamics.
-%
-%Our main assumption is that most of the variation in the data correlates with time.
-%
-Our main assumptions are that the high-dimensional data (one dimension per discretization point in the concentration profiles/one dimension per color pixel intensity in the images) lie on a one-dimensional, nonlinear curve, {\it and} that this curve is parameterized by time.
-%
-Figure~\ref{subfig:dmaps_edges} shows a schematic of a two-dimensional data set (the ``high-dimensional" data) which lies on a one-dimensional nonlinear curve (the ``low-dimensional" manifold).
-%
-We assume that the data points are ordered from youngest to oldest along the curve.
-%
-When our data are very high dimensional, we cannot extract this curve by eye.
-%
-We instead use diffusion maps \cite{coifman2005geometric}, a dimensionality reduction technique which aims to uncover low-dimensional, nonlinear structure in high-dimensional data sets, to automatically extract this one-dimensional curve.
-%
-Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the leading (non-trivial) diffusion map coordinate.
-%
-The color parameterizes the one-dimensional curve that our eye can visually detect, and ordering the data by the leading diffusion map coordinate will effectively order data in time.
-
-To uncover the diffusion map parameterization, we assume that data points which are close in the high-dimensional space should also be close {\it on the curve} (we would like images which look similar to be close together in our developmental trajectory).
-%
-%We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\it as close as possible} in the parameterization.
-%
-To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored based on the distance between the data points (using the diffusion kernel of the distance), so that pairs of points which are close have dark/strong edges connecting them.
-%
-Visually, the dark edges delineate the one-dimensional curve, whereas the light edges are less informative about the low-dimensional structure.
-%
-%We want points which are close in the high-dimensional space, i.e., connected by dark edges, to be close in our low-dimensional %parameterization.
-%
-Finding such a parameterization can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a Markov matrix which contains information about pairwise distances between the data points (see {\it Materials and Methods}).
-%
-%This procedure has been named diffusion maps \cite{coifman2005geometric}.
-%
-%The formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).
-
-Given our set of registered signals, shown in Figure~\ref{subfig:1d_aligned_unordered}, we use diffusion maps to order them in time.
-%
-We use the Euclidean distance between the (discretized in $\mathbb{R}^{100}$) aligned concentration profiles (the rows of Figure~\ref{subfig:1d_aligned_unordered}) as our pairwise distances between data points.
-%
-Figure~\ref{subfig:1d_aligned_ordered} shows the data from Figure~\ref{subfig:1d_aligned_unordered}, now sorted by the value of the leading diffusion map coordinate, $\phi_2$.
-%
-As expected, the intensity of the peak grows as a function of $\phi_2$ (since we constructed these data
-to have a peak growing in time).
-%
-%Although diffusion maps can order the data, it cannot tell us about the speed of this trajectory.
-%
-%Furthermore, the direction of the trajectory is also unknown; determining which end of the trajectory corresponds to the beginning of development is a task that needs to be done by hand.
-
-\subsection{Vector diffusion maps for simultaneous registration and ordering}
-
-%We are often interested in both registering and ordering our data.
-%
-%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
-%
-Angular synchronization (for registration) can be combined with diffusion maps (for temporal ordering) in a single eigencomputation that allows us to {\it simultaneously} recover the registrations {\it and} the temporal ordering.
-%
-This technique is called vector diffusion maps \cite{singer2012vector} (see {\it Materials and Methods}).
-%
-It both reduces the number of required eigencomputations, and is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
-%
-Figure~\ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure~\ref{subfig:1d_unaligned_unordered} using vector diffusion maps.
-%
-The data has been ordered by $\psi_{1, 3}$, the leading vector diffusion map coordinate.
-
-\subsection{Analysis of two-dimensional images}
-
-The same methods (with small modifications) can be applied directly to our two-dimensional images.
-%
-Instead of one-dimensional profiles (discretized to $100$-dimensional concentration vectors), we now work with
-two color, two-dimensional images (discretized into $100 \times 100$ pixels/$20000$-dimensional vectors), and we register the images with respect to translations and rotations (the symmetry group is $ISO(2)$).
-%
-To facilitate the application of vector diffusion maps, we approximate the two-dimensional translations and rotations by {\it three-dimensional rotations} (symmetry group $SO(3)$; see {\it SI Appendix}).
-%
-We use vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_unordered};
-the results are shown in Figure~\ref{fig:images_ordered}.
-%
-Both the red and green peaks are now aligned, and the intensity of the red peaks grows as a function of the leading vector diffusion map coordinate.
-
-In these experiments, we are fortunate enough to be able to validate our  automatic registration and ordering using prior physical knowledge about the system.
-%
-For registration, it is known that the peak in the distribution of Dorsal (shown in green) specifies the ventralmost position in the embryo, and that the nuclei (shown in gray) delineate the circumference of the embryo.
-%%%
-%%% maybe you can also quantify this guy
-%%%
-%
-%Based on this, images can be translated to make the centers of the embryos coincide and rotated to place the peak of the green channel at the same angular position.
-%
-Temporal ordering, on the other hand, can be done based on the monotonic progress of cellularization (Figure~\ref{subfig:membranes}) \cite{figard2013plasma}.%; this helps validate our vector diffusion map-based ordering.
-%
-The two methods for registration and ordering can be visually compared through the sequences of one-dimensional profiles extracted from the circumference of the embryos (similar to the format in Figure~\ref{fig:1d_demo}), shown in Figures~\ref{subfig:1d_membranes}~and~\ref{subfig:1d_vdm}.
-%
-The general trends are consistent:
-both green  and red peaks are aligned, and the red peaks grow as a function of time/leading vector diffusion map coordinate.
-%
-However, Figure~\ref{subfig:1d_vdm}, which was automatically registered and ordered, shows an arguably better clustering of the ``outer" peaks in the top corners of the image.
-%
-%These peaks are more scrambled in Figure~\ref{fig:membrane_compare}b, which was analyzed using cellularization.
-%
-These peaks are known to arise later in development, and so it is reassuring that they appear at the end of the (recovered) trajectory.
-%
-Furthermore, the membranes stop growing towards the end of this trajectory, making the ordering of the late signals in Figure~\ref{subfig:1d_membranes} less accurate.
-%
-Quantitatively, the rank correlation coefficient between these two temporal orderings is 0.9158, indicating that our automatic ordering is consistent with the ordering obtained from cellularization.
-
-%Although our methods are fairly robust, they do require a sufficient amount of data so that DMAPS can ``see'' a smooth trajectory.
-%
-%We evaluated how much data is required by subsampling our data set (with replacement), and for each subsample, computing the temporal ordering using VDM and calculating the rank correlation between this ordering and the ordering from cellularization.
-%
-%The results are shown in Figure~\ref{subfig:bootstrap} (each data point is averaged over 50 random subsamples).
-%
-%For small data sets, the average rank correlation is very low, indicating that the orderings from VDM are inaccurate.
-%
-%However, for moderately sized data sets (approximately 30 data points), the rank correlation begins to plateau and VDM can accurately order the data.
-
-\begin{figure}[t]
-\includegraphics[width=8.4cm]{fig5}
-\caption{The images in Figure~\ref{subfig:images_unordered}, aligned and temporally ordered using vector diffusion maps.
-The images are sorted according to the value of the first non-trivial vector diffusion map component, $\psi_{1, 4}$.}
-%\label{fig:images_ordered}
-\customlabel{fig:images_ordered}{5}
-\end{figure}
-
-\begin{figure}[t]
-\includegraphics[width=8.4cm]{fig6}
-\caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different levels of development.
-{\it (B)} Membrane length as a function of time.
-The points corresponding to the images in {\it A} are indicated with arrows.
-{\it (C)} The one-dimensional concentration profiles, registered by peak Dorsal fluorescence and ordered by membrane length.
-{\it (D)} The one-dimensional concentration profiles corresponding to the automatically registered and ordered images in Figure~\ref{fig:images_ordered}.
-}
-%\label{fig:membrane_compare}
-\customlabel{fig:membrane_compare}{6}
-\customlabel{subfig:membranes}{\ref{fig:membrane_compare}{\it A}}
-\customlabel{subfig:membrane_curve}{\ref{fig:membrane_compare}{\it B}}
-\customlabel{subfig:1d_membranes}{\ref{fig:membrane_compare}{\it C}}
-\customlabel{subfig:1d_vdm}{\ref{fig:membrane_compare}{\it D}}
-\end{figure}
+\begin{figure*}[t]
+\includegraphics[width=8.4cm]{raw_data2}
+\hfill
+\includegraphics[width=8.4cm]{VDM_ordered} \\
+\centering
+\includegraphics[width=16.8cm]{average_trajectory}
+%\label{fig:fish}
+\customlabel{fig:raw_data}{2}
+\customlabel{subfig:raw_data1}{\ref{fig:fish}{\it A}}
+\customlabel{subfig:raw_data2}{\ref{fig:fish}{\it B}}
+\end{figure*}
 
@@
 
-We have shown how symmetry and dimensionality reduction techniques can successfully register and order cross-sectional developmental imaging data.
-%
-Although we illustrated our methods using a specific data set from a study of {\it Drosophila} embryogenesis, the framework is quite general and requires little knowledge of the image features or the developmental dynamics.
-%
-To our knowledge, temporal ordering of {\it Drosophila} images has not been automated to this extent.
-%
-Recently, machine learning algorithms have been used to group {\it Drosophila} images into different developmental classes \cite{yuan2014automated}, yet there, the appropriate classes must be defined by the user and hand-labeled training data are required.
-
-Temporal ordering of cross-sectional biological data has been studied in a variety of other contexts.
-%
-Cross-sectional RNA sequencing data has been temporally ordered \cite{anavy2014blind, trapnell2014dynamics}, where each data point is the transcriptome for a single sample.
-%
-Gene expression data from multiple patients have been studied in \cite{gupta2008extracting, qiu2011discovering};
-here, each data point is a vector containing the expression levels of the various genes.
-%
-Snapshots of cells throughout the cell cycle can be described by a vector of features (e.g., amount of DNA) which quantify the cell's state  \cite{kafri2013dynamics}.
-%
-These problems are no different than our task of ordering images, each of which is a large vector of pixel intensities.
-%
-The majority of these works temporally order the data either by solving a traveling salesman problem on the data, or by constructing a minimum spanning tree of the data,
-expecting that these structures (the path of the traveling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate encoding of the dynamical progression.
-%
-We, instead, use diffusion maps to construct a one-dimensional parameterization of the data.
-%
-Diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003, tenenbaum2000global, Donoho2003, Roweis2000}.
-%
-They have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}, and so we suspect they may perform better than the ordering algorithms used in previous work.
-%
-{\bf Yannis says ``careful'', he thinks this might be too strong of a statement.}
-%%%
-%%%   we need help here
-%%%
-%
-%Furthermore, although we use only the first diffusion maps coordinate to order the data (we assume the data are one-dimensional), diffusion maps are not limited to one-dimensional data and can be used to extract more complex structures.
-%
-%For example, in data which contains several branches or classes, the first few coordinates will delineate the branches or separate the classes.
-
-The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
-%
-The main point to note here is that many registration algorithms require definition of appropriate landmarks or features for the images and the location of these within each image.
-%
-For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then register images by aligning these features \cite{zhao2003face}.
-%
-Temporally ordering images of a construction site over time is also based on first identifying features within the images \cite{schindler2007inferring}.
-%
-In general, appropriate features may not be known {\it a priori}, and so we require a registration algorithm that does not rely on predefined features, but only on the inherent geometric symmetry of the problem.
-%
-Template-based methods are also common for registration, where one optimally aligns each image to a predefined template function.
-%
-However, these methods can suffer when the images are noisy, and the success of such methods can depend strongly on the template function \cite{shatsky2009method}.
-%
-{\bf Need to discuss consistency.}
-
-Angular synchronization and vector diffusion maps require no feature points and no template function.
-%
-These methods have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}, where each image is a picture of (essentially) the {\it same} molecule; variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
-%
-In contrast, our data contains variation from the different embryo orientations, as well as variation from the underlying developmental dynamics.
-% (which underpins the development of vector diffusion map algorithms).
-%
-In addition to the eigenproblem relaxation used here, other synchronization algorithms have been proposed \cite{bandeira2014multireference}.
-%
-Synchronization is formulated for general symmetry groups, and depending on the data, one might consider other symmetries, such as reflections and scaling/dilation.
-%
-Here, the main requirement is that the symmetry group be represented by orthogonal or unitary matrices \cite{singer2013spectral}.
-
-Temporal ordering can also be accomplished without first registering the images, by constructing functions of the images which are invariant under translations and rotations. 
-%
-These functions are then used as the data in the diffusion maps calculation; since these representative functions do not change under translations and rotations, the resulting diffusion map embedding is invariant to symmetries.
-%
-Such functions include the scale invariant feature transform \cite{lowe1999object, lowe2004distinctive}, the scattering transform \cite{mallat2012group}, and the bispectrum \cite{zhao2014rotationally}.
-%
-These methods have the disadvantage that we do not obtain a registered set of images, but only a temporal ordering.
-%
-However, such methods can often be faster for large data sets, since the functions are computed per image, rather than computing pairwise quantities in synchronization.
-
-In all of these data-driven methods, an essential question is how much and what type of data is needed to accurately reconstruct the dynamics.
-%
-For diffusion maps to temporally order the data, we require the developmental trajectory to be well-sampled, and that the principal direction of variability in the data (the coordinate which diffusion maps will recover) be monotonic in time.
-%
-In addition, the data should have spatial variability/structure for synchronization to successfully register the images.
-%
-In our example, the data set is sufficiently large so that the dynamics are well-sampled.
-%
-We have included both the red and the green channels so that the principal direction of variability is monotonic in time.
-%
-If we only used the green channel in our analysis, we could not successfully order the data using diffusion maps, as this signal does not change appreciably in time.
-%
-Both the red and green channels contain spatial variability which allows us to register the data points: the green channel has one peak at a specific angular around the embryo, and the red channel has two peaks which grow in time.
-%
-We include the green channel because the location of the red peaks is noisy, and so inclusion of the (less noisy) green channel allows us to more accurately register the images.
-%
-%For the methods we present here, we require a sufficient amount of data for the one-dimensional evolution curve to be ``seen'' by the algorithms (i.e., in Figure~\ref{subfig:dmaps_edges}, there are a sufficient number of data points so that our eye can see the curve; if there were only 4 data points, it would most likely be difficult for our eye to detect the one-dimensional structure).
-%
-%We also require the data to contain enough information so that the alignments and dynamics are apparent.
-%
-%For example, in the data presented in Figure~\ref{subfig:images_unordered}, the green peaks delineate the position of the ventral axis in the embryo, and the intensity of the red peaks grow over time.
-%
-%Therefore, these images contain information about rotational symmetries of the data as well as the dynamics of the data.
-%
-%If we removed the green signal, and only used the red signal in our analysis, the registration would not be as accurate.
-%
-%Furthermore, if we only used the green signal, then there would be no dynamic information (the shape or intensity of Dorsal does not change as a function of time in the developmental stage we are considering), and the temporal ordering would be inaccurate.
-%
-We are confident that for many applications, these requirements - data which adequately samples the developmental trajectory, and data with sufficient spatial and temporal variability - are easily met.
-%
-In these cases, our methods allow for the rapid, automated analysis of imaging data sets to help uncover possibly complex dynamics and structure.
-{\bf Something about smoothness in making movies}
-%
-%{\bf procrustes referece}
-%{\bf allso the guy form Donoho who worried about how much to do syncjhronization}
-
-%%%   you should sample well enough in time
-%%% your dmap coordinate should have a nice slope with time
-%% you should have variation in angle ---and we know we are random in the rotations
-%%
-%% and now the green is an issue of noise.....
-%% the green "sees" angle but not time,
-%% the red sees angle and time, but angle badly
-%% the ocmbination sees both well enough
-%%
-%%
-%%   we need to say things about "other methods" -- Mallat, AmitNow, Procrustes...
-%% and if we change the artificial stuff, but even if not
-%% we need to say something about flips.
-%% we need to say symmetry early at the beginning -- important
-%%  the expression "factoring out symmetry", the reference to Ben....  something about symmetry groups 
-%%    the smoothness when one wants to make movies.
-%%  
-%%
-%%   Y is afraid of the compsci literature being "fairly represented"
-%%
-%%
-
-%% == end of paper:
-
-%% Optional Materials and Methods Section
-%% The Materials and Methods section header will be added automatically.
-
-%% Enter any subheads and the Materials and Methods text below.
-
-\begin{materials}
-
-\section{Angular synchronization \cite{singer2011angular}}
-
-%\sansmath
-
-Let $ x_1, \dots, x_m$ denote the signals that we wish to align with respect to rotations;
-each signal is a function defined on the unit circle (on the plane).
-%
-%Practically, it is discretized in a $n$-long vector (the local intensity at $n$ equidistant points around the circle);
-%rotating the function by an angle $\theta$ then corresponds to cyclically shifting the elements of $x_i$ by $\frac{\theta_i}{2 \pi} n$ (rounded to the nearest integer to obtain a valid shift).
-%
-First assume that each signal $x_i$ is a {\it noisy} rotated copy of the underlying signal $x_{true}$
-(which we are {\it not} given), such that
-\begin{equation}
-x_i = f(x_{true}, \theta_i) + \xi_i
-\end{equation}
-where the function $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ radians, and $\xi_i$ is a (typically Gaussian) noise term.
-%
-Our goal is to recover $\theta_1, \dots, \theta_m$.
-%
-Up to noise,
-\begin{equation} \label{eq:pairwise_rot}
-x_i \approx f(x_j, \theta_i - \theta_j) ;
-\end{equation}
-note that \eqref{eq:pairwise_rot} does not require knowledge of $x_{true}$.
-%
-We can obtain an {\it estimate} of $\theta_i - \theta_j$ by computing the rotation that optimally aligns $x_j$ to $x_i$,
-i.e., %$\theta_{ij} \approx \theta_i - \theta_j$, where
-%
-\begin{equation} \label{eq:opt_angle}
-\theta_i - \theta_j \approx \theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2.
-\end{equation}
-%
-Practically, the signals are discretized in a $n$-long vector (the local intensity at $n$ equidistant points around the circle);
-rotating the function by an angle $\theta$ then corresponds to cyclically shifting the elements of $x_i$
-by $\frac{\theta_i}{2 \pi} n$ (rounded to the nearest integer to obtain a valid shift).
-%
-For the one-dimensional discretized profiles shown in Figure~\ref{fig:1d_demo}, we exhaustively search over all $n=100$ possible shifts of the signals to compute the optimal angles in \eqref{eq:opt_angle}.
-%
-Alternatively, for continuous signals, an optimization algorithm
-can be used \cite{ahuja2007template}.
-
-Rather than work with the angles $\theta_{ij}$ directly, it is more convenient to consider the	 rotation matrices,
-\begin{equation} \label{eq:R_theta}
-R(\theta_{ij}) = \begin{bmatrix}
-\cos(\theta_{ij}) & -\sin(\theta_{ij}) \\
-\sin(\theta_{ij}) & \cos(\theta_{ij})
-\end{bmatrix},
-\end{equation}
-which we can think of as operating on the points of the unit circle (on the plane) on which our signal is defined.
-%
-Successive rotations correspond to multiplication of the corresponding rotation matrices: $R(\alpha_1 + \alpha_2) = R(\alpha_1) R(\alpha_2)$.
-%
-Due to the orthogonality of rotation matrices, $R(-\alpha) = R(\alpha)^T$.
-
-Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$; we write our procedure for general $d$ because we will later consider three-dimensional rotations).
-%
-We construct the matrix $H \in \mathbb{R}^{md \times md}$, where $H$ is an $m \times m$ matrix of $d \times d$ blocks, with the $i,j^{th}$ block of $H$, $H_{ij}$, defined as
-\begin{equation} \label{eq:H_to_R}
-H_{ij} = R(\theta_{ij}).
-\end{equation}
-%
-%
-Under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, $H_{ij} \approx R(\theta_i) R(\theta_j)^T$
-%\begin{equation}
-%H_{ij} = R(\theta_{ij}) \approx R(\theta_i - \theta_j) = R(\theta_i) R(-\theta_j) = R(\theta_i) R(\theta_j)^T,
-%\end{equation}
- and
-\begin{equation} \label{eq:H_low_rank}
-	H \approx
-	\begin{bmatrix}
-	R(\theta_1) \\
-	R(\theta_2) \\
-	\vdots \\
-	R(\theta_m)
-	\end{bmatrix}
-	\begin{bmatrix}
-	R(\theta_1)^T R(\theta_2)^T \dots R(\theta_m)^T
-	\end{bmatrix}.
-\end{equation}
-%
-It follows directly from \eqref{eq:H_low_rank} that the top block eigenvector of $H$ contains our best estimates of $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$.
-%
-Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$.
-%
-Then,
-\begin{equation}
-\hat{R} =
-\begin{bmatrix}
-\hat{R}_1 \\
-\hat{R}_2 \\
-\vdots \\
-\hat{R}_m
-\end{bmatrix} =
-\begin{bmatrix}
-| & | & & | \\
-\phi_1 & \phi_2 & \dots & \phi_d \\
-| & | & & |
-\end{bmatrix},
-\end{equation}
-where $\hat{R}_i \in \mathbb{R}^{d \times d}$ is (nearly) the estimate for $R(\theta_i)$.
-%
-To obtain our estimate of $R(\theta_i$), denoted $R_{i, est}$, we project $\hat{R}_i$ onto the closest orthogonal matrix, 
-\begin{equation} \label{eq:R_est}
-R_{i, est} = U_i V_i^T,
-\end{equation}
-where $U_i$ and $V_i$ are the left and right singular vectors, respectively, of $\hat{R}_i$.
-%
-We adjust the sign of $\phi_1$ so that $det(R_{i, est}) = +1$, ensuring proper rotations.
-%
-We estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and register the signals by rotating signal $i$ by $-\theta_i$.
-%
-We note that, in our actual computations, the pairwise rotations $\theta_{ij}$ are computed in a discrete setting, then the overall
-synchronization is performed in the continuum context to obtain $\theta_i$, and the results are rounded to give the closest
-discrete shift.
-%
-Details of registering two-dimensional images using synchronization are given in the {\it SI Appendix}.
-
-%Furthermore, this formulation also considers {\it higher-order} consistency information.
-%
-%For example, given our pairwise estimates $R_{ij}$, we know that relationships of the form
-%\begin{equation} \label{eq:triplet_consistency}
-%R(\theta_{ik}) R(\theta_{kj}) \approx R(\theta_i) R(\theta_k)^T R(\theta_k) R(\theta_j)^T = R(\theta_i) R(\theta_j)^T
-%\end{equation}
-%should also hold.
-%
-%Note that
-%\begin{equation}
-%(H^2)_{ij} = \sum_k R(\theta_{ik}) R(\theta_{kj});
-%\end{equation}
-%therefore, {\it all} infomation of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$.
-%
-%Because $H$ and $H^2$ have the same eigenvectors, our problem formulation accounts for not only pairwise alignment information, but also these higher-order considerations.
-
-\section{Diffusion maps \cite{coifman2005geometric}}
-
-Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional space), we want to find a coordinate transformation $y(x)$ that preserves local information: points that are close in the original space should also be close in the coordinate $y$.
-%
-The first step is to construct the matrix $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if points $x_i$ and $x_j$ are ``close.''
-%
-We use a diffusion kernel,
-\begin{equation} \label{eq:dmaps_W}
-W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right)
-%W_{ij} = e^{ -\frac{d^2(x_i, x_j)}{\epsilon^2}}
-\end{equation}
-where $d(x_i, x_j)$ is the pairwise distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic scale.
-%
-Points less than $\epsilon$ apart are thus considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
-%
-$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph}); here, we take $\epsilon$ to be the median of the pairwise distances between data points.
-
-To find this coordinate $y$, we want solve the following optimization problem \cite{Belkin2003}
-\begin{equation} \label{eq:dmaps_opt_problem}
-\argmin_{y} \sum_{ij} W_{ij} (y(x_i) - y(x_j))^2.
-\end{equation}
-%
-We first compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{m} W_{ij}$, and the matrix $A$, where
-\begin{equation} \label{eq:dmaps_A}
-A = D^{-1} W.
-\end{equation}
-%
-We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_m$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$.
-%
-%Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors.
-%
-Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector; this is a trivial solution to \eqref{eq:dmaps_opt_problem}.
-%
-%In general, the next few eigenvectors $\phi_2, \dots, \phi_m$ give ``meaningful'' embedding coordinates for the data, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point.
-%
-The next eigenvector, $\phi_2$, is the (non-trivial) solution to \eqref{eq:dmaps_opt_problem}, so that $\phi_2(j)$, the $j^{th}$ entry of $\phi_2$, gives the ``new" coordinate for data point $x_j$ (i.e., $\phi_2(j) = y(x_j)$).
-%
-In our application, we have assumed that this single direction of variability, parameterized by $\phi_2$, is one-to-one with time.
-%
-Ordering the data by $\phi_2(j)$ will then, effectively, order them in time.
-%
-The procedure generalizes when the data lie on higher-dimensional manifolds (not just curves) in data space.
-
-\section{Vector diffusion maps\cite{singer2012vector}}
-
-In vector diffusion maps, given data points $x_1, \dots, x_m$, one first constructs the matrix $S \in \mathbb{R}^{md \times md}$, with the $i,j^{th}$ block of $S$, $S_{ij}$, defined as
-\begin{equation} \label{eq:vdm_S}
-	S_{ij} = A_{ij} H_{ij}
-\end{equation}
-%
-where $A_{ij} \in \mathbb{R}$ (defined in \eqref{eq:dmaps_A}) pertains to the diffusion kernel between data points, and $H_{ij} \in \mathbb{R}^{d \times d}$ (defined in \eqref{eq:H_to_R}) pertains to the pairwise alignment between data points.
-%
-It is important to note that distance $d(x_i, x_j)$ used in the diffusion kernel in \eqref{eq:dmaps_W} is the distance between data points {\it after} after pairwise alignment, i.e., the minimum distance between all possible shifts of the two data points (which is obtained in \eqref{eq:opt_pairwise}).
-%
-In the language of symmetry groups, this distance is a metric between the orbits induced by the symmetry group.
-
-One then computes the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_{md}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{md}$ of $S$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$.
-%
-These eigenvectors contain information about {\it both} the optimal rotations (the ``synchronization" component) and the
-variation of the data {\it after} the spatial symmetries have been factored out (in our case, their temporal variation).
-%
-Assuming that the data (after symmetries have been factored out) are relatively closely clustered, it is reasonable
-to expect, as in angular synchronization, that the top (block) eigenvector of $S$ contains approximations of the optimal rotations,
-which can be computed in the same way from \eqref{eq:R_est}.
-%
-We then expect subsequent eigenvectors to contain information about the main direction(s) of data variability modulo the geometric symmetries.
-%%%
-%%% somewhere we need to talk about compact groups
-%%% discussion might be a good place because SCALING is obviously important and
-%%% it is not compact......
-%%%
-%However, the eigenvectors now also contain information about the embedding coordinates for our images.
-%
-In general, the embedding coordinates will be given by
-\begin{equation}
-\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle
-\end{equation}
-where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$.
-%
-If we assume that our data mainly vary along a one-dimensional manifold, and that the rotations and the dynamics are uncoupled and therefore separable, one expects the embedding coordinate for our data %(i.e., the analog of $\phi_2$ from the diffusion maps case)
-to be  $\psi_{1,d+1}$.
-%
-Our example is, by physical considerations, a separable one, and so sorting our data by $\psi_{1,d+1}$ will order our data in time.
-%
-If no separability claims can be made, a second diffusion map step, using $\psi_{k,l}$ as the coordinates, may uncover a
-useful parameterization of the data manifold modulo symmetries.
-%%%
-%%% \cite....
-%%%
-
-\end{materials}
-
-%% Optional Appendix or Appendices
-%% \appendix Appendix text...
-%% or, for appendix with title, use square brackets:
-%% \appendix[Appendix Title]
-
-\begin{acknowledgments}
-- text of acknowledgments here, including grant info -
-\end{acknowledgments}
-
-%% PNAS does not support submission of supporting .tex files such as BibTeX.
-%% Instead all references must be included in the article .tex document.
-%% If you currently use BibTeX, your bibliography is formed because the
-%% command \verb+\bibliography{}+ brings the <filename>.bbl file into your
-%% .tex document. To conform to PNAS requirements, copy the reference listings
-%% from your .bbl file and add them to the article .tex file, using the
-%% bibliography environment described above.
-
-%%  Contact pnas@nas.edu if you need assistance with your
-%%  bibliography.
-
-% Sample bibliography item in PNAS format:
-%% \bibitem{in-text reference} comma-separated author names up to 5,
-%% for more than 5 authors use first author last name et al. (year published)
-%% article title  {\it Journal Name} volume #: start page-end page.
-%% ie,
-% \bibitem{Neuhaus} Neuhaus J-M, Sitcher L, Meins F, Jr, Boller T (1991)
-% A short C-terminal sequence is necessary and sufficient for the
-% targeting of chitinases to the plant vacuole.
-% {\it Proc Natl Acad Sci USA} 88:10362-10366.
-
-%% Enter the largest bibliography number in the facing curly brackets
-%% following \begin{thebibliography}
-
-\bibliographystyle{pnas}
-\bibliography{background_reading/references,../../references/references}
-
-%\begin{thebibliography}{}
-
-%\end{thebibliography}
-
 \end{article}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Adding Figure and Table References
-%% Be sure to add figures and tables after \end{article}
-%% and before \end{document}
-
-%% For figures, put the caption below the illustration.
-%%
-%% \begin{figure}
-%% \caption{Almost Sharp Front}\label{afoto}
-%% \end{figure}
-
-%% For Tables, put caption above table
-%%
-%% Table caption should start with a capital letter, continue with lower case
-%% and not have a period at the end
-%% Using @{\vrule height ?? depth ?? width0pt} in the tabular preamble will
-%% keep that much space between every line in the table.
-
-%% \begin{table}
-%% \caption{Repeat length of longer allele by age of onset class}
-%% \begin{tabular}{@{\vrule height 10.5pt depth4pt  width0pt}lrcccc}
-%% table text
-%% \end{tabular}
-%% \end{table}
-
-%% For two column figures and tables, use the following:
-
-%% \begin{figure*}
-%% \caption{Almost Sharp Front}\label{afoto}
-%% \end{figure*}
-
-%% \begin{table*}
-%% \caption{Repeat length of longer allele by age of onset class}
-%% \begin{tabular}{ccc}
-%% table text
-%% \end{tabular}
-%% \end{table*}
-
-%\begin{figure}
-%\begin{minipage}{2.9cm}
-%\centering
-%\includegraphics[width=\textwidth]{illustrate_1d}\\
-%\vspace{-0.25cm}
-%(a)\\
-%\includegraphics[width=\textwidth]{registered_unordered_1d}\\
-%(d)
-%\end{minipage}
-%\begin{minipage}{2.9cm}
-%\centering
-%\includegraphics[width=\textwidth]{illustrate_1d_intensity}\\
-%(b)\\
-%\includegraphics[width=\textwidth]{registered_ordered_1d}\\
-%(e)
-%\end{minipage}
-%\begin{minipage}{2.9cm}
-%\centering
-%\includegraphics[width=\textwidth]{unregistered_unordered_1d}\\
-%(c)\\
-%\includegraphics[width=\textwidth]{registered_ordered_vdm_1d}\\
-%\centering
-%(f)
-%\end{minipage}
-%\caption{Analysis of one-dimensional profiles obtained from experimental data.
-(a) One-dimensional, two-component concentration profile on a ring (top), and the corresponding profile on a line (bottom).
-The concentration profile was extracted from an image of the type shown in Figure~\ref{fig:images_unordered}.
-(b) Intensity corresponding to the profile in (a).
-(c) An ensemble of concentration profiles, each of the form described in (a).
-Each row in the array corresponds to a single profile obtained from an image of the type in Figure~\ref{subfig:images_unordered}.
-(d) The profiles in (c), now registered using angular synchronization.
-(e) The profiles in (d), now temporally ordered using diffusion maps.
-(f) The profiles in (c), registered and temporally ordered in a single step using vector diffusion maps.}
-%\end{figure}
-%
-
-\begin{figure*}
-\begin{minipage}{0.2\textwidth}
-\centering
-\includegraphics[width=\textwidth]{sphere_1}\\
-\includegraphics[width=\textwidth]{sphere2_1}\\
-A
-\end{minipage}
-%
-\begin{minipage}{0.2\textwidth}
-\centering
-\includegraphics[width=\textwidth]{sphere_2}\\
-\includegraphics[width=\textwidth]{sphere2_2}\\
-B
-\end{minipage}
-\begin{minipage}{0.2\textwidth}
-\centering
-\includegraphics[width=\textwidth]{sphere_3}\\
-\includegraphics[width=\textwidth]{sphere2_3}\\
-C
-\end{minipage}
-\begin{minipage}{0.2\textwidth} 
-\centering
-\includegraphics[width=\textwidth]{sphere_4}\\
-\includegraphics[width=\textwidth]{sphere2_4}\\
-D
-\end{minipage}
-\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions.
-{\it (A)} The original image.
-{\it (B)} Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image.
-{\it (C)} Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation.
-{\it (D)} Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.
-The top row shows the three-dimensional spheres, and the bottom row shows the (two-dimensional) surface of the sphere in which we are interested.}
-\label{fig:SO3_picture}
-\end{figure*}
-
-\section{Registering images with respect to translations and rotations} \label{subsec:trans_rot_register}
-
-Using angular synchronization or vector diffusion maps to register images with respect to rotations {\it and translations} requires some additional effort.
-%
-%We are not only interested in registering the one-dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two-dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\it and} translations.
-%
-%We first compute the translations and rotations to optimally align each pair of data points.
-%
-The first step (similar to the case of only rotations) is to compute the optimal alignments between pairs of images.
-%
-Practically, we have square images discretized as pixels (rather than continuous functions on the plane).
-%
-For each image pair $I_i$ and $I_j$ we compute
-\begin{equation}\label{eq:opt_pairwise}
-(\theta_{ij}, dx_{ij}, dy_{ij}) = \argmin_{
-\begin{matrix}
-\theta \in [0, 2\pi) \\
-dx, dy \in [-\Delta, \Delta]\\
-%dy \in [-\Delta, \Delta]
-\end{matrix}
-} \|g(I_j, \theta, dx, dy) - I_i \|^2.
-\end{equation}
-where $g(I_j, \theta, dx, dy)$ is image $I_j$ rotated around the center of the square by $\theta$ radians, then translated by $dx$ pixel widths in the $x$ direction, and finally translated by $dy$ pixel widths in the $y$ direction.
-%
-The norm, $\| \cdot \|$, is the Euclidean norm between the pixel intensities for both the red and green channels.
-%
-The domain of the image (a square) is not invariant to our rotations and translations; however, the pixels near the border of the square have zero intensity, and so the norm can be meaningfully computed as long as the main image does not ``move out of'' the original square.
-%
-%The norm \eqref{eq:opt_pairwise} involve an integral over the square, which is not left invariant by our rotations and
-%translations; noting however that the pixels close to the border of the square have zero intensity allows a meaningful
-%computation of this norm as long as the ``main feature" (the embryo) does not move out of the original square (no ``corner effects").
-%
-This also justifies the $\Delta$ limits in \eqref{eq:opt_pairwise}: $\Delta$ is chosen such that the main image is not split or separated when translating (we take $\Delta=10$, which corresponds to a 10\% shift in the image).
-%; it does not make sense to shift the images so much that the embryo would ``move out" of the original square.
-%
-Image rotation is performed with the \texttt{imrotate} function in Matlab, using nearest neighbor interpolation to estimate the pixel intensities after rotation.
-%, since there is no direct mapping of the pixels of the unrotated image to the pixels of the rotated image.
-%
-The missing pixels in the corners of the rotated image are taken to have zero intensity.
-%
-The translations are implemented by shifting the pixels;
-we only consider shifts that correspond to an integer number of pixels to remove the need for interpolation when doing the translations.
-%
-The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
-%
-Therefore, instead of using an optimization procedure, we discretize the search space and exhaustively search to find the solution.
-%
-Our discrete translation steps are $\Delta/5$, and our discrete rotations steps are $\pi/10$. 
-%
-%We discritize the search space of rotations into 20 possible rotations %($d\theta  \in \{0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10 \}$),
-%and 11 possible translations in both . %($dx, dy \in \{-20, -16, -12, \dots, 12, 16, 20 \}$).
-%
-%We check all possible combinations for the rotation and translations that best align $I_j$ to $I_i$.
-%
-Although computationally demanding, this ``embarrassingly parallelizable" direct enumeration approach is not prohibitive here.
-%
-%Although this can be somewhat time intensive, it is not prohibitive for the data sets we consider, and can be trivially parallelized if necessary.
-%
-%The solution will not be the exact solution to \eqref{eq:opt_pairwise}, but it will (most likely) be a close approximation.
-%
-%Since our techniques are robust to noise, close approximations will be sufficient to obtain accurate results.
-
-%As discussed above, for angular synchronization, the underlying symmetry group needs to have a real and orthogonal matrix representation.
-%
-%We therefore convert $ISO(2)$, the group of two-dimensional translations and rotations, to $SO(3)$, the group of three-dimensional rotations.
-%
-
-The synchronization formulation requires that the matrices used to represent the symmetry group be orthogonal \cite{singer2013spectral}.  
-%(for \eqref{eq:H_low_rank} to be satisfied).
-%
-However, the typical matrix representation for $ISO(2)$, the group of two-dimensional translations and rotations, does not satisfy this property.
-%
-Instead, we {\it approximately} represent rotations and translations in two dimensions using {\it three-dimensional rotation matrices}, by projecting the (two-dimensional) image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular}.
-%
-Rotation of the image corresponds to rotation around one principal axis in three dimensions, while translation of the image corresponds (approximately) to rotations around the other two principal axes.
-%(Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$.
-However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.)
-%
-%These rotation matrices are (by definition) orthogonal, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
-
-To convert the optimal two-dimensional translation and rotation to corresponding (continuous) three-dimensional rotations,
-we first compute the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$,
-\begin{equation} \label{eq:angle_relations}
-%\begin{aligned}
-%	\alpha_{ij} &= \theta_{ij} \\
-%	\beta_{ij} &= \frac{dx_{ij}}{n} \times \eta_{proj} \\
-%	\gamma_{ij} &= \frac{dy_{ij}}{n} \times \eta_{proj} \\
-%\end{aligned}
-	\alpha_{ij} = \theta_{ij}, \; \; \;
-	\beta_{ij} = \frac{dx_{ij}}{n} \times \eta_{proj}, \; \; \;
-	\gamma_{ij} = \frac{dy_{ij}}{n} \times \eta_{proj}
-\end{equation}
-where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project the image.
-%
-We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere;
-this portion of the sphere is small enough so that curvature effects are not prominent in the image.
-%
-We write rotations around the three principal axes, $R^x(\alpha)$, $R^y(\beta)$, and $R^z(\gamma)$, in terms of the three Euler angles
-%\small
-\begin{equation}
-\begin{aligned}
-	R^x(\alpha) &= \begin{bmatrix}
-	1 & 0 & 0 \\
-    0 & \cos(\alpha) & -\sin(\alpha) \\
-    0 & \sin(\alpha) & \cos(\alpha)
-	\end{bmatrix} \\
-	R^y(\beta) &= \begin{bmatrix}
-	\cos(\beta) & 0 & \sin(\beta) \\
-    0 & 1 & 0 \\
-    -\sin(\beta) & 0 & \cos(\beta)
-    \end{bmatrix} \\
-	R^z(\gamma) &= \begin{bmatrix}
-	\cos(\gamma) & -\sin(\gamma) & 0 \\
-    \sin(\gamma) & \cos(\gamma) & 0 \\
-    0 & 0 & 1
-    \end{bmatrix}.
-\end{aligned}
-\end{equation}
-%\normalsize
-%
-The total rotation, $R_{ij} \in SO(3) \subset \mathbb{R}^{3 \times 3}$, is
-\begin{equation} \label{eq:total_rot}
-	R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})	 = R^z(\gamma_{ij})  R^y(\beta_{ij})  R^x(\alpha_{ij}).
-%R(\alpha_{ij}, \beta_{ij}, \gamma_{ij}) =
-%\begin{bmatrix}
-%\cos(\beta) \cos(\gamma) & \cos(\gamma)\sin(\alpha)\sin(\beta)-\cos(\alpha) \sin(\gamma) & \cos(\alpha)\cos(\gamma)\sin(\beta)+ \sin(\alpha)\sin(\gamma) \\
-%\cos(\beta) \sin(\gamma) & \cos(\alpha) \cos(\gamma)+\sin(\alpha) \sin(\beta) \sin(\gamma) & \cos(\alpha) \sin(\beta) \sin(\gamma)-\cos(\gamma) \sin(\alpha) \\
-%-\sin(\beta) & \cos(\beta) \sin(\alpha) & \cos(\alpha) \cos(\beta))
-%\end{bmatrix}
-\end{equation}
-%
-%\eqref{eq:total_rot} corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).
-
-These rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$ are used in formulating the matrix $H$ in \eqref{eq:H_to_R} for angular synchronization, or the matrix $S$ in \eqref{eq:vdm_S} for vector diffusion maps to compute the rotations (and, in the case of vector diffusion maps, temporal orderings) for the images.
-%
-From the three-dimensional rotations $R_{1, est}, \dots, R_{m, est}$, we  require the corresponding translations and rotations of the images.
-%
-We first multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
-%
-$R_{i,est}^T$ then gives the rotation matrix which registers image $i$.
-%
-From a rotation matrix $R$, we compute the Euler angles $\alpha$, $\beta$, and $\gamma$ using the following relationships
-\begin{equation}
-\begin{aligned}
-R_{1,1} & = \cos(\beta)\cos(\gamma) &
-R_{2,1} & = \cos(\beta)\sin(\gamma) \\
-R_{3,1} & = -\sin(\beta) &
-R_{3,2} & = \sin(\alpha)\cos(\beta) \\
-R_{3,3} & = \cos(\alpha)\cos(\beta)  &
-\end{aligned}
-\end{equation}
-%
-Inverting \eqref{eq:angle_relations} yields the desired approximate translations and rotation, and we round optimal translations to the nearest integer to remove the need for interpolation when translating the images.
 

### 2014-05-06 image_analysis_paper/Initial_classification_noReflection.m

@@ -1,0 +1,125 @@
+function [ class, rot, corr, FBsPCA_data, timing ] = Initial_classification_noReflection(data, r_max, n_nbor, isrann )
+%This function does initial classfication on preprocessed data.
+%   Input: 
+%           data:  data should be phase flipped, prewhitened and roughly centered.
+%           r_max: the radius of the region of interest.
+%           n_nbor: number of nearest neighbors used for class averaging.
+%           isrann: using randomized algorithm for finding k nearest
+%           neighbors. isrann=0: brute force nearest neighbor search.
+%           isrann=1: randomized nearest neighbor search
+%   Output:
+%           class: a matrix of size n x n_nbor. This provides a list of
+%           nearest neighbors for each image.
+%           rot: rotational alginment
+%           corr: normalized cross correlation between nearest neighbors
+%           timing: timing of different steps in the algorithm
+%	    FBsPCA_data: FBsPCA denoising data. It includes r_max--radius of 
+%           region of interest, UU--eigenvectors, Freqs--associated angular 
+%           frequence, Coeff--image expansion coefficients on UU, Mean--estimated
+%           rotationally invariant mean of the data, and W--weight for wiener type filtering.
+% Zhizhen Zhao Updated Feb 2014
+
+% Get date dimension:
+L=size(data, 1);
+P=size(data, 3);
+N=floor(L/2);
+trunc = 200; % maximum number of FBsPCA leading componnents chosen. 
+  
+[x, y]=meshgrid(-N:N, -N:N);    %generate square grids
+r=sqrt(x.^2+y.^2);      %compute the radius at each grid point
+test=reshape(data, L^2, P);
+test=test(r>r_max, :);
+noise_variance=var(test(:)); %Estimate the noise variance from ear region.
+
+%Use Fourier Bessel steerable PCA to denoise and compress the original data
+%set.
+tic_FBsPCA=tic;
[…]

### 2014-05-06 image_analysis_paper/VDM_noReflection.m

@@ -1,0 +1,74 @@
+function [ class_VDM, angle, VV_LP ] = VDM_noReflection(class, corr, rot, k, flag, n_nbor) 
+%Function for using VDM to search for nearest neighbors more robust to
+%noise
+% Input:
+%       class: 
+%           Pxl matrix, where l>k. This gives a list of l-nearest
+%           neighbors for P data points.
+%       corr: 
+%           Pxl matrix, normalized cross correlation between each data
+%           point and its l-nearest neighbors.
+%       rot: 
+%           Pxl matrix, in-plane rotational alignment in degrees between
+%           each data point and its l-nearest neighbors.
+%       k: 
+%           number of nearest neighbors to choose to make the graph.
+%       flag:
+%           indicates generating k-nn graph using union rule (flag==0) or
+%           k-nn graph using linear programming.
+%       n_nbor: 
+%           number of nearest neighbors user needs.
+%Output:
+%       class_VDM: 
+%           Pxn_nbor matrix. It gives n_nbor nearest neighbors for each
+%           data point.
+%       angle:
+%           Pxn_nbor matrix. The rotational alignment in degrees between
+%           each point and its nearest neighbors.
+%       VV_LP:
+%           VDM eigenfunctions
+%
+%Zhizhen Zhao Aug 2013
+
+
+P=size(class, 1); 
+
+[ X, Ah, rows, cols ] = sort_list_weights_wrefl(class(:, 1:k), sqrt(2-2*real(corr(:, 1:k))), rot(:, 1:k), ones(P, k) );
+if flag==0
+    W = ones(length(X), 1); %Generating knn graph using union rule
+else
+    W =script_find_graph_weights_v3(X, [rows, cols], 2*P, 5); %Generating k-nn graph using linear programming. It makes sure that the weight in each row is the same.
+end;
+W2=W.*Ah;
+H2=sparse(rows(W>0.001), cols(W>0.001), W2(W>0.001), 2*P, 2*P);
+H2 = H2(1:P, 1:P);
+H2 =H2+H2';
+
+[r_VDM_LP, ~, VV_LP]=VDM_LP(H2, 3);
+
+if P<=10^4
[…]

### 2014-05-06 image_analysis_paper/test_FBsPCA2.m

@@ -1,0 +1,83 @@
+clear all
+close all
+
+%% directories where things are stored
+time_data = '../membrane_lengths/oct16.mat';
+image_dir = '../membrane_pictures/membrane2/dpERK_staining';
+
+%% load membrane lengths
+
+load(time_data);
+mem_lengths = L(:,1);
+
+% remove some embryos
+ind = setdiff(1:52, [9, 15, 22]);
+
+m = length(ind);
+
+mem_lengths = mem_lengths(ind);
+
+% compute ranks from membrane lengths
+ranks_from_membranes = compute_ranks(mem_lengths);
+
+%% image parameters
+
+npixels = 101;
+
+%set image plotting parameters
+subplot_dim1 = ceil(sqrt(m));
+subplot_dim2 = ceil(m / subplot_dim1);
+[Y, X] = meshgrid((subplot_dim1:-1:1)/subplot_dim1, (1:subplot_dim2)/subplot_dim2);
+
+data = zeros(npixels, npixels, m);
+
+%% load images
+
+for i=1:m
+    % read image
+    im1 = imread(sprintf('%s/emb%02d.tif', image_dir, ind(i)));
+    
+    % resize image
+    im1 = imresize(im1, [npixels npixels]);
+    
+    im1 = im1(:, :, 2);
+    
+    %store image
+    data(:, :, i) = im1;
+    
+end
+
+%%
+
+r_max = floor(npixels/2);
+n_nbor = 5;
+isrann = 0;
+
+[ class, rot, corr, FBsPCA_data, timing ] = Initial_classification_noReflection(data, r_max, n_nbor, isrann );
+%[ class, class_refl, rot, corr, FBsPCA_data, timing ] = Initial_classification(data, r_max, n_nbor, isrann );
+
+%%
+
+k = n_nbor;
+flag = 0;
+[ class_VDM, angle, VV_LP ] = VDM_noReflection(class, corr, rot, k, flag, n_nbor) ;
+
+%%
+
+coord = zeros(m, 1);
+theta = zeros(m, 1);
+
+for i=1:m
+    coord(i) = VV_LP(i, 2) * conj(VV_LP(i, 1));
+    theta(i) = atan2(imag(VV_LP(i, 1)), real(VV_LP(i, 1))) * (180/pi);
+end
+
+[~, I] = sort(real(coord));
+
+figure;
+for i=1:m
+    subplot(subplot_dim1, subplot_dim2, i)
[…]

### 2014-05-06 image_analysis_paper/test_PCA.m

@@ -6,7 +6,7 @@ close all
 
 fontsize = 10;
 
-%im_save_dir = 'paper_figures';
+im_save_dir = 'paper_figures2';
 
 %% directories where things are stored
 time_data = '../membrane_lengths/oct16.mat';
@@ -109,15 +109,24 @@ for i=1:m
     imshow(im1);
     
 end
+saveas(gcf,sprintf('%s/raw_data1', im_save_dir), 'pdf')
 
 image_set_aligned = zeros(size(image_set), 'uint8');
+image_set_aligned_withnuclei = zeros(size(image_set), 'uint8');
 for i=1:m
     im_tmp = image_set(:,:,:,i);
-    %     for j=1:3
-    %         im_tmp(:,:,j) = im_tmp(:,:,j) + nuclei(:,:,i);
-    %     end
+    %         for j=1:3
+    %             im_tmp(:,:,j) = im_tmp(:,:,j) + nuclei(:,:,i);
+    %         end
     im1 = rotate_image(R_opt(dim*(i-1)+1:dim*i,:), im_tmp, angle_proj);
     image_set_aligned(:,:,:,i) = im1;
+    
+    im_tmp = image_set(:,:,:,i);
+    for j=1:3
+        im_tmp(:,:,j) = im_tmp(:,:,j) + nuclei(:,:,i);
+    end
+    im1 = rotate_image(R_opt(dim*(i-1)+1:dim*i,:), im_tmp, angle_proj);
+    image_set_aligned_withnuclei(:,:,:,i) = im1;
 end
 
 figure;
@@ -133,47 +142,78 @@ end
 
 %%
 
-npixels2 = 20;
+fontsize = 8;
 
-data_reshaped = zeros(m, npixels2*npixels2*nchannels);
+data_reshaped = zeros(m, npixels*npixels*nchannels);
+
+[eigenimages, D, proj_coeff] = PCA_images(image_set_aligned, 9);
 
 figure;
+for i=1:9
+    subplot(3, 3, i)
+    imshow(eigenimages(:, :, :, i))
+end
+
+figure;
+set(gcf, 'paperunits', 'centimeters')
+set(gcf, 'papersize', [4 4])
+set(gcf, 'paperposition',[0 0 4 4])
+imshow(eigenimages(:, :, :, 1),'border','tight')
[…]

### 2014-05-07 image_analysis_paper/pnas_draft2.tex
more edits
Added updates to paper draft

@@
 %
-During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclose nuclei into different cells grow in thickness (Figure~\ref{subfig:membranes}).
+During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclose nuclei into different cells grow in thickness.
 %
@@
 %
-During the $4^{th}$ hour of development, ...
+Following cellularization, the spatial arrangement of nuclei and the concentration profiles of regulatory molecules change dramatically.
 %
-Images in Figure~\ref{subfig:images_unordered} show optical sections at a fixed depth, perpendicular to the long axis of the embryo.
+Images in Figure~\ref{subfig:raw_data1} and Figure~\ref{subfig:raw_data2} show optical sections at a fixed depth, perpendicular to the long axis of the embryo during and following cellularization, respectively.
 %
@@
 %
-Each of the images reveals a uniform arrangement of nuclei and a nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.
+%Each of the images reveals a uniform arrangement of nuclei and a nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.
 
-Each image in Figure~\ref{subfig:images_unordered} comes from a different embryo, captured at a different rotation about the long axis and at a different and unknown point in time.
+Each image in Figures~\ref{subfig:raw_data1}~and~\ref{subfig:raw_data2} comes from a different embryo, captured at a different rotation about the long axis and at a different and unknown point in time.
 %
@@
 %
-We will demonstrate that both of these tasks can be fully automated.
+We will demonstrate that both of these tasks can be fully automated using a sufficiently general framework which can readily be applied to a variety of biological imaging data sets.
+
+\section{Results and Discussion}
 
 \begin{figure*}[t]
-\includegraphics[width=6.5cm]{raw_data1}
+\includegraphics[width=5.4cm]{raw_data1}
 \hfill
-\includegraphics[width=6.5cm]{PCA_ordered}
+\includegraphics[width=5.4cm]{PCA_ordered}
 \hfill
-\begin{minipage}[b]{3.3cm}
+\begin{minipage}[b]{5.6cm}
 \vspace{0in}
-\includegraphics[width=3.3cm]{PCA_eigenimage1}
-\includegraphics[width=3.3cm]{PCA_corr}
+\includegraphics[width=2.7cm]{PCA_eigenimage1}
+\includegraphics[width=2.7cm]{PCA_corr}
+\includegraphics[width=5.6cm]{PCA_12}
 \end{minipage}
-%\label{fig:fish}
-\customlabel{fig:raw_data}{2}
-\customlabel{subfig:raw_data1}{\ref{fig:fish}{\it A}}
-\customlabel{subfig:raw_data2}{\ref{fig:fish}{\it B}}
+\caption{(A) Images collected during cellularization.
+(B) Images, now registered using angular synchronization and ordering by the first PCA projection coefficient.
+(C) First eigenimage.
+(D) Correlation between first PCA projection coefficient and membrane thickness.
+(E) Projection onto first two principal components.
+}
+\customlabel{fig:data1}{2}
+\customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
+\customlabel{subfig:ordered_data1}{\ref{fig:data1}{\it B}}
+\customlabel{subfig:eigenimage1}{\ref{fig:data1}{\it C}}
+\customlabel{subfig:PCA_corr}{\ref{fig:data1}{\it D}}
+\customlabel{subfig:PCA_12}{\ref{fig:data1}{\it E}}
 \end{figure*}
-%
-
-\section{Results and Discussion}
 
@@
 %
-We use angular synchronization to register our images. 
+We use angular synchronization \cite{singer2011angular} to register our images with respect to translations and rotations. 
 %
-This technique is very general, as it requires no template image or feature points. 
+Angular synchronization is a very general technique which requires no template image or feature points. 
 %
-Furthermore, it has been shown to be robust to noise.
+Although we could have used other image registration algorithms (e.g.
+...) for the data sets we present here, we chose instead to use an algorithm that can be applied to a wide variety of data sets with little {\em a priori} knowledge.
 %
@@
 
-Figure ... shows the images from Figure ..., registered using angular synchronization.
+The images in Figure \ref{subfig:ordered_data1} have been registered using angular synchronization.
 %
@@
 
+After registering our images so that they are optimally aligned, we would like to organize our images to extract meaningful patterns and structure. 
+%
+In our data, the main variation in our images arises due to developmental dynamics, and we would like to order our data in time. 
+%
 We will use dimensionality reduction techniques to temporally order our data.
 %
-Although our images are inherently high-dimensional (number of pixels $\times$ number of channels), we assume that they lie on a low-dimensional curve, and that this curve is parameterized by time.
+Although our images are inherently high-dimensional (number of pixels $\times$ number of channels), we assume that they lie on a low-dimensional curve which is monotonic in time. 
 %
+Furthermore, we assume that the main source of variability in this data is due to time/dynamics. 
+%
+Therefore, uncovering a one-dimensional parameterization of this curve will allow us to order our data in time.
 
-For the images shown in Figure ...., the dynamics are relatively simple, and we can use principal component analysis (PCA) to order the images.
+For the images shown in Figure~\ref{subfig:raw_data1}, the dynamics are relatively simple, and we can use principal component analysis (PCA) to order the images.
 %
@@
 %
-In our dataset, each data point can be effectively described by a multiple of a single eigenimage, i.e., the images lie on a line in a very high-dimensional space.
+In our dataset, the dynamics are such that each data point can be effectively described by a multiple of a single eigenimage, i.e., the images lie on a line in a very high-dimensional space.
 %
-We can therefore sort these images along this line to temporally order them in time.
+We can organize the images by sorting them along this line.
 
-Figure ... shows the images, now ordered using PCA. 
+Figure~\ref{subfig:ordered_data1} shows the images, registered using angular synchronization and ordered using PCA. 
 %
-The first eigenimage is also shown.
+The first eigenimage is shown in Figure~\ref{subfig:eigenimage1}.
 
-In this simple example, we can validate our ordering using an independent time marker.
+For these experiments in this particular time window, we are fortunate enough to be able to validate our  ordering using prior physical knowledge about the system.
 %
-Figure ... shows the rank correlation betrween the two orderings.
+Temporal ordering, on the other hand, can be done based on the monotonic progress of cellularization \cite{figard2013plasma}.%; this helps validate our vector diffusion map-based ordering.
 %
-Rank correlation is 0.9199.
+Figure~\ref{subfig:PCA_corr} shows the rank correlation between the two orderings.
+%
+The rank correlation coefficient is 0.9199, indicating that we do a fairly good job of recovering the correct temporal ordering of the images.
 
-Although PCA can successfully order the data set shown in Figure ..., it can only do so because the data lie on in a one-dimensional, {\em linear} subspace.
+Although PCA can successfully order the data set shown in Figure \ref{subfig:raw_data1}, it can only do so because the data lie on in a one-dimensional, {\it linear} subspace.
 %
@@
 %
-This can be seen in the late images of Figure ... that the ordering is not entirely correct.
+This can be seen in the late images of Figure~\ref{subfig:ordered_data1} that the ordering is not entirely correct.
 %
-We can also see this from the projection onto the first two eigenimages, shown in Figure ...
+We can also see this from the projection onto the first two eigenimages, shown in Figure~\ref{subfig:PCA_12}.
 %
-The data clearly lie on a one-dimensional curve, but the curve is nonlinear at late times. 
+The data clearly lie on a one-dimensional, {\em nonlinear} curve.
 %
-This motivates us to use nonlinear dimensioanlity reduction techniques to temporally order our data.
+This motivates us to use nonlinear dimensionality reduction techniques to temporally order our data.
 
 \subsection{Temporal Ordering Using Diffusion Maps}
-We will use diffusion maps, a technique which is robust to noise and broadly applicable to a wide variety of data types.
+
+We can use diffusion maps \cite{coifman2005geometric} to analyze more complex data sets where the dynamics are clearly nonlinear.
 %
-
-We can use diffusion maps to analyze more complex data sets where the dynamics are clearly nonlinear.
+Diffusion maps is similar to PCA, in that it aims uncover low-dimensional structure in high-dimensional data.
 %
-Figure ... shows a data set from the $4^{th}$ hour of development.
+However, unlike PCA, diffusion maps is a {\it nonlinear} reduction technique. 
 %
-The dynamics are significantly more complex. 
+Figure~\ref{subfig:raw_data2} shows a data set consisting of embryos following cellularization.
+%
+Clearly, the dynamics are significantly more complex than the dynamics in Figure~\ref{subfig:raw_data1}, and the first PCA eigenimage will not capture all of the relevant dynamics.
 %
 However, we can still use diffusion maps to temporally order the data.
+%
+Figure~\ref{subfig:ordered_data2} shows the images in Figure~\ref{subfig:raw_data2}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
 
-We have no time marker with which to validate our orderings in this time period. 
+Unlike the images in Figure~\ref{subfig:raw_data1}, 
+for Figure~\ref{fig:data2}, we have no time marker with which to validate our orderings.
 %
-However, there are noticable features in the ordering that we can see are consistent with peviously known dynamics. 
+However, there are noticeable features in the ordering that we can see are consistent with previously known dynamics. 
 %
-For example, ....
+We can also construct a smoothed trajectory by averaging over neighboring images.
+%
+Figure~\ref{subfig:averaged_trajectory} shows the images from Figure~\ref{subfig:ordered_data2}, divided into 12 groups and averaged. 
+%
+The dynamics are readily apparent; we can see gastrulation, ....
 
+\newpage
 \begin{figure*}[t]
@@
 \includegraphics[width=16.8cm]{average_trajectory}
-%\label{fig:fish}
-\customlabel{fig:raw_data}{2}
-\customlabel{subfig:raw_data1}{\ref{fig:fish}{\it A}}
-\customlabel{subfig:raw_data2}{\ref{fig:fish}{\it B}}
+\caption{(A) Images collected following cellularization.
+Each image is of a different embryo in a different rotational orientation.
+(B) Data from A, registered and ordered using vector diffusion maps.
+(C) Average trajectory of the images shown in B.}
+\customlabel{fig:data2}{3}
+\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
+\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
+\customlabel{subfig:averaged_trajectory}{\ref{fig:data2}{\it C}}
 \end{figure*}
@@
 
+We present algorithms for registration and temporal ordering of images.
+%
+The algorithms are sufficiently general that they can be applied to a wide variety of biological imaging data.
+
+Although the examples here focus on temporally ordering the data, tasks such as classification and comparison of different mutations can also be accomplished using the same techniques. 
+%
+
+\bibliographystyle{pnas}
+\bibliography{background_reading/references,../../references/references}
+
 \end{article}

### 2014-05-09 image_analysis_paper/pnas_draft2.tex
Changed figures for paper after discussing with Stas

--- 2014-05-06 image_analysis_paper/pnas_draft2.tex
@@
 
-\title{Temporal ordering and registration of cross-sectional imaging data}
+\title{Reduction of biological imaging data}
 
@@
 \begin{abstract}
-In studies of development, researchers are often presented with cross-sectional data, where each data point is a sample from a population fixed at a slightly different developmental time.
+
+We demonstrate how data reduction techniques can effectively organize large sets of biological images. 
 %
-The goal is then to temporally order the data to reconstruct the developmental dynamics.
+We use existing algorithms to both register images within a data set so that they are in a consistent frame of reference, as well as extract meaningful structure within the image data set.
 %
-If each data point is a two-dimensional image, the images must first be registered before they can be temporally ordered.
+We illustrate the utility of such algorithms using two data sets from a study of {\it Drosophila} emrbyogenesis.
 %
-When such data sets are large, noisy, and/or if the developmental changes are subtle, these tasks can be difficult to do by hand.
+Here, we show that we can temporally order the images to obtain a representative picture of the embryogenesis dynamics. 	
 %
-We present an automatic approach to register {\it and} temporally order cross-sectional data sets of images.
-%
-The mathematical techniques (angular synchronization for image registration, diffusion maps for temporal ordering, and
-vector diffusion maps for simultaneously performing both tasks) are applicable to a wide variety of data sets and
-require little {\it a priori} knowledge of the image features or the developmental dynamics.
-%
-We demonstrate the utility of our methods using a collection of images from a study of {\it Drosophila} embryogenesis.
+In general, these algorithms allow us to extract meaningful structure from imaging data sets where the patterns are not known {\it a priori}.
+
 \end{abstract}
@@
 
-\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional \cite{diggle2002analysis}.
+Imaging data is becoming increasingly prevalent in biology research.
 %
-In longitudinal studies, developmental progress is monitored over time for the same embryo.
+General motivation about biology data ...
+
+Need methods to organize data, visualize, ...
+
+REDUCTION is important, reduction because of symmetry and reduction because of content
+
+There are a bunch of recent techniques- some established (like dmaps), some very fresh, like synchronization or VDMaps,  that have had their own reasons to be developed but they are as if tailor made for what we would want to do
+
+\section{Results and Discussion}
+
+We apply the methods to two imaging data sets from a study of {\it Drosophila} embryogenesis.
 %
-In a cross-sectional study, one embryo contributes only a single snapshot of a chemical or morphological process along its developmental trajectory, and the developmental dynamics must be reconstructed from multiple snapshots of different embryos.
+The first data set (Figure~\ref{subfig:raw_data1}) is from stage 5, when cellularization occurs. 
 %
-Both of these sampling schemes have their advantages and limitations, and both are extensively used by developmental biologists.
+The second data set (Figure~\ref{subfig:raw_data2}) is from the end of stage 5 to stage 7, which captures formation of the ventral furrow and gastrulation.
 %
-Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms.
+In both of these data sets, each image is taken from a different embryo, fixed at a different developmental time and imaged in a different rotational orientation.
 %
-In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that help visualize a handful of cellular processes.
+We would like to organize these data sets in a meaningful and informative way.
 %
-Fixed embryos are then imaged using any given number of microscopy techniques.
+This requires rotating and translating each image so the set is in a consistent frame of reference ({\it registration}), and then organizing the images. 
 %
-Recent advances in large-scale physical manipulation and imaging of embryos have produced rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different geometric orientation and developmental time point.
+For these data sets, we know (by experimental design) that the main source of variability comes from the different developmental time of each image.
 %
-Importantly, the ``age'' of any given embryo arrested in its development is not known to high accuracy.
+We will show that the algorithms can effectively order the images in time and therefore give us meaningful picture of the underlying developmental dynamics.
+
+\subsection{Image Registration}
+
+Because each embryo is imaged in a different orientation, the first task is to align or {\em register} the images so that they are in a consistent frame of reference.
 %
-In general, it is only known that a collection of embryos belongs to a certain time window.
+We use angular synchronization \cite{singer2011angular} to register our images with respect to translations and rotations. 
 %
-In order to recover the developmental dynamics from such image datasets, snapshots of different embryos must be spatially aligned or registered to factor out the relevant symmetries (i.e., translations and rotations), and then ordered in time.
+The task of image registration has been broadly studied in a variety of contexts (see, for example, ...).
 %
-We present a general computational framework which greatly accelerates -and can combine- both of these tasks.
+We chose to use angular synchronization because it is a very general technique which requires no {\em a priori} knowledge of a good template image or feature points. 
+%
+Therefore, this algorithm can readily be extended to other imaging data sets. 
+%
+Implementation details are given in the {\it SI appendix}.
+
+The images in Figure \ref{subfig:ordered_data1} have been registered using angular synchronization.
+%
+The important features that our eye can visually detect (e.g., the green Dorsal peaks) have been aligned, even though the algorithm does not require explicit identification of such features.
 
@@
 \end{figure}
-%
-Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are visually apparent.
-%
-As an example, Figure~\ref{fig:fish} shows a caricature of fish development, combining the processes of growth and color patterning.
-%
-In this case, temporal ordering can be accomplished by arranging the fish by size, which is monotonic with the developmental progress.
-%
-Image registration is based on obvious morphological landmarks, such as the position of head and fins.
-%
-On the other hand, real data poses nontrivial challenges, due to the number of images, measurement noise and variability, and the subtlety of the developmental changes.
 
-To illustrate the issues associated with temporal ordering and registration of images, we use two data sets from an imaging study of signaling protein expression in the early fruit fly embryo, one of the leading model systems for studies of developmental dynamics.
-%
-The fruit fly embryo is shaped like a rice grain, approximately $1/2$~mm long.
-%
-By end of the $2^{nd}$ hour of development, sequential divisions of a fertilized egg generate a system where $\sim 6,000$ nuclei are uniformly arranged under the common plasma membrane.
-%
-During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclose nuclei into different cells grow in thickness (Figure~\ref{subfig:membranes}).
-%
-In conjunction with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs \cite{lim2013kinetics}.
-%
-During the $4^{th}$ hour of development, ...
-%
-Images in Figure~\ref{subfig:images_unordered} show optical sections at a fixed depth, perpendicular to the long axis of the embryo.
-%
-The embryos were fixed and stained with chemicals that help visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red) \cite{chung2010microfluidic}.
-%
-Each of the images reveals a uniform arrangement of nuclei and a nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.
-
-Each image in Figure~\ref{subfig:images_unordered} comes from a different embryo, captured at a different rotation about the long axis and at a different and unknown point in time.
-%
-The goal is to register these images in a globally consistent way, and order them to be representative of the underlying developmental trajectory.
-%
-We will demonstrate that both of these tasks can be fully automated.
-
-\begin{figure*}[t]
-\includegraphics[width=6.5cm]{raw_data1}
-\hfill
-\includegraphics[width=6.5cm]{PCA_ordered}
-\hfill
-\begin{minipage}[b]{3.3cm}
-\vspace{0in}
-\includegraphics[width=3.3cm]{PCA_eigenimage1}
-\includegraphics[width=3.3cm]{PCA_corr}
-\end{minipage}
-%\label{fig:fish}
-\customlabel{fig:raw_data}{2}
-\customlabel{subfig:raw_data1}{\ref{fig:fish}{\it A}}
-\customlabel{subfig:raw_data2}{\ref{fig:fish}{\it B}}
-\end{figure*}
-%
-
-\section{Results and Discussion}
-
-\subsection{Image Registration}
-
-Because each embryo is imaged in a different orientation, the first task is to align or {\em register} the images so that they are in a consistent frame of reference.
-%
-We use angular synchronization to register our images. 
-%
-This technique is very general, as it requires no template image or feature points. 
-%
-Furthermore, it has been shown to be robust to noise.
-%
-Implementation details are given in the {\it SI appendix}.
-
-Figure ... shows the images from Figure ..., registered using angular synchronization.
-%
-The important features that our eye can visually detect (e.g., the green Dorsal peaks) have been aligned, even though the algorithm does not require identification of such features.
+\begin{figure}[t]
+\includegraphics[width=8.4cm]{fig2} 
+\caption{{\it (A)} Fluorescent images of {\it Drosophila} embryos in stage 5.
+Each embryo has been stained for Dorsal (green) and dpERK (red).
+Each image is of a different embryo in a different rotational orientation.
+{\it (B)} Selected image from data set.
+{\it (C)} Profile of Dorsal (green) and dpERK (red) around the circumference of the embryo in {\it B}.
+}
+\customlabel{fig:data1}{2}
+\customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
+\customlabel{subfig:select_image}{\ref{fig:data1}{\it B}}
+\customlabel{subfig:1d_profile}{\ref{fig:data1}{\it C}}
+\end{figure}
 
@@
 
-We will use dimensionality reduction techniques to temporally order our data.
+After registering our images so that they are in a consistent frame of reference, we would like to organize our images to extract meaningful patterns and structure. 
 %
-Although our images are inherently high-dimensional (number of pixels $\times$ number of channels), we assume that they lie on a low-dimensional curve, and that this curve is parameterized by time.
+Although our images are inherently high-dimensional (number of pixels $\times$ number of channels), we assume that there are only a few modes/directions of variability.
 %
+We will use dimensionality reduction techniques to extract these main directions of variability and efficiently organize our images.
+%
+%In our examples, we assume that the main source of variability in this data is due to time/developmental dynamics. 
+%
+%We will show that dimensionality reduction techniques uncover a one-dimensional parameterization/organization of our images, and that the images are temporally ordered in this parameterization.
 
-For the images shown in Figure ...., the dynamics are relatively simple, and we can use principal component analysis (PCA) to order the images.
+Principal component analysis (PCA) \cite{shlens2005tutorial} is arguably the most common dimensionality reduction technique.
 %
-PCA extracts a set of {\it eigenimages} from the data, such that each data point can be (approximately) described as a linear combination of these eigenimages. 
+PCA extracts the main directions/modes of variability in a data set. 
 %
-In our dataset, each data point can be effectively described by a multiple of a single eigenimage, i.e., the images lie on a line in a very high-dimensional space.
+We applied PCA to the data set shown in Figure~\ref{subfig:raw_data1}.
 %
-We can therefore sort these images along this line to temporally order them in time.
+The first two modes, or eigenimages, are shown in Figure~\ref{subfig:eigenimage1}~and~\ref{subfig:eigenimage2}.
+%
+Clearly, these modes extract/capture meaningful features that we can visually detect in the data set. 
+%
+We can plot the data projected onto the first two eigenimages; this is shown in Figure~\ref{subfig:PCA_12}.
+%
+From this plot, we can see that the data fall effectively on a one-dimensional curve that is parameterized by the first projection coefficient.
+%
+Therefore, an approrpiate way to organize the images would be to order them according to the first projection coefficient.
 
-Figure ... shows the images, now ordered using PCA. 
+Figure~\ref{subfig:ordered_data1} shows the images from Figure~\ref{subfig:raw_data1}, now ordered by the first projection coefficient.
 %
-The first eigenimage is also shown.
+We can now easily see the relevant structure in the data set:
+the red peaks grow in intensity.
+%
+Furthermore, for this data set, we know what the relevant structure is.
+%
+The data consists of embryos fixed at different developmental times. 
+%
+During stage 5, we can measure the developmental time based on the monotonic progress of cellularization (Figure~\ref{subfig:membranes}) \cite{figard2013plasma}.%; this helps validate our vector diffusion map-based ordering.
+%
+The ordering we extract using the first eigenimage is consistent with the temporal ordering of the images using cellularzation.
+%
+Quantitatively, the rank correlation coefficient between these two orderings is 0.9199.
+%
+The two methods for registration and ordering can be visually compared through the sequences of one-dimensional profiles extracted from the circumference of the embryos, shown in Figures~\ref{subfig:1d_membranes}~and~\ref{subfig:1d_vdm}.
+%
+The general trends are consistent:
+both green  and red peaks are aligned, and the red peaks grow as a function of time/first PCA projection coefficient.
 
-In this simple example, we can validate our ordering using an independent time marker.
+Although PCA can successfully order the data set shown in Figure \ref{subfig:raw_data1}, in general, such data is not guaranteed to lie in a low-dimensional, {\it linear} subspace.
 %
-Figure ... shows the rank correlation betrween the two orderings.
+When the data have nonlinear structure, PCA may not provide the most parsimonious organization of the data.
 %
-Rank correlation is 0.9199.
+This can be seen in the late images of Figure~\ref{subfig:ordered_data1};
+images where the ventral furrow begins to form are interspersed with images containing no obvious dent.
+%
+This is because the dynamics at later stages are nonlinear and and not effectively described by a single eigenimage.
+%
+We can see this from the projection onto the first two eigenimages, shown in Figure~\ref{subfig:PCA_12}.
+%
+The data clearly lie on a one-dimensional, {\em nonlinear} curve.
+%
+This motivates us to use nonlinear dimensionality reduction techniques to organize our data.
 
-Although PCA can successfully order the data set shown in Figure ..., it can only do so because the data lie on in a one-dimensional, {\em linear} subspace.
-%
-In general, the developmental trajectory is not guaranteed to be linear. 
-%
-This can be seen in the late images of Figure ... that the ordering is not entirely correct.
-%
-We can also see this from the projection onto the first two eigenimages, shown in Figure ...
-%
-The data clearly lie on a one-dimensional curve, but the curve is nonlinear at late times. 
-%
-This motivates us to use nonlinear dimensioanlity reduction techniques to temporally order our data.
+\begin{figure}[t]
+\raisebox{3.5cm}{{\figtextfont A}}
+\includegraphics[width=4.2cm]{PCA_eigenimage1}
+\raisebox{3.5cm}{{\figtextfont B}}
+\includegraphics[width=4.2cm]{PCA_eigenimage2}
+
+\raisebox{4cm}{{\figtextfont C}}
+\includegraphics[width=8.4cm]{PCA_12}
+
+\raisebox{8cm}{{\figtextfont D}}
+\includegraphics[width=8.4cm]{PCA_ordered}
+\caption{{\it (A)} First eigenimage from the data set shown in Figure~\ref{subfig:raw_data1}.
+{\it (B)} Second eigenimage from the data set shown in Figure~\ref{subfig:raw_data1}.
+{\it (B)} Projection of  data set shown in Figure~\ref{subfig:raw_data1} onto the first two eigenimages.
+{\it (D)} Ordering of  data set shown in Figure~\ref{subfig:raw_data1} by the first projection coefficient.
+}
+\customlabel{fig:data1b}{3}
+\customlabel{subfig:eigenimage1}{\ref{fig:data1b}{\it A}}
+\customlabel{subfig:eigenimage2}{\ref{fig:data1b}{\it B}}
+\customlabel{subfig:PCA_12}{\ref{fig:data1b}{\it C}}
+\customlabel{subfig:ordered_data1}{\ref{fig:data1b}{\it D}}
+\end{figure}
+
+\begin{figure}[t]
+\includegraphics[width=8.4cm]{fig6}
+\caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different levels of development.
+{\it (B)} Membrane length as a function of time.
+The points corresponding to the images in {\it A} are indicated with arrows.
+{\it (C)} The one-dimensional concentration profiles, registered by peak Dorsal fluorescence and ordered by membrane length.
+{\it (D)} The one-dimensional concentration profiles corresponding to the automatically registered and ordered images in Figure~\ref{fig:images_ordered}.
+}
+%\label{fig:membrane_compare}
+\customlabel{fig:membrane_compare}{4}
+\customlabel{subfig:membranes}{\ref{fig:membrane_compare}{\it A}}
+\customlabel{subfig:membrane_curve}{\ref{fig:membrane_compare}{\it B}}
+\customlabel{subfig:1d_membranes}{\ref{fig:membrane_compare}{\it C}}
+\customlabel{subfig:1d_vdm}{\ref{fig:membrane_compare}{\it D}}
+\end{figure}
 
 \subsection{Temporal Ordering Using Diffusion Maps}
-We will use diffusion maps, a technique which is robust to noise and broadly applicable to a wide variety of data types.
+
+We can use diffusion maps \cite{coifman2005geometric} to analyze data sets with nonlinear structure.
 %
+Diffusion maps is similar to PCA, in that it aims uncover low-dimensional structure in high-dimensional data.
+%
+However, unlike PCA, diffusion maps is a {\it nonlinear} reduction technique.
 
-We can use diffusion maps to analyze more complex data sets where the dynamics are clearly nonlinear.
+Figure~\ref{subfig:raw_data2} shows a data set of images from stages 5--7.
 %
-Figure ... shows a data set from the $4^{th}$ hour of development.
-%
-The dynamics are significantly more complex. 
+Clearly, the dynamics are significantly more complex than the dynamics in Figure~\ref{subfig:raw_data1}, and the first eigenimage will not capture all of the relevant dynamics.
 %
 However, we can still use diffusion maps to temporally order the data.
+%
+In addition, we can combine the steps image registration and temporal ordering using vector diffusion maps.
 
-We have no time marker with which to validate our orderings in this time period. 
+Figure~\ref{subfig:ordered_data2} shows the images in Figure~\ref{subfig:raw_data2}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
 %
-However, there are noticable features in the ordering that we can see are consistent with peviously known dynamics. 
+Unlike the images in Figure~\ref{subfig:raw_data1}, 
+for Figure~\ref{fig:data2}, we have no time marker with which to validate our orderings.
 %
-For example, ....
+However, there are noticeable features in the ordering that we can see are consistent with previously known dynamics,
+and we can construct a smoothed trajectory by averaging groups of neighboring images.
+%
+Figure~\ref{subfig:averaged_trajectory} shows the images from Figure~\ref{subfig:ordered_data2}, divided into 12 groups and averaged. 
+%
+We can easily see the formation of the ventral furrow and germ band elongation, which concludes with ventral cells shifted to the dorsal side of the embryo.
 
+\newpage
 \begin{figure*}[t]
+\raisebox{10.8cm}{{\figtextfont A}}
 \includegraphics[width=8.4cm]{raw_data2}
 \hfill
-\includegraphics[width=8.4cm]{VDM_ordered} \\
+\raisebox{10.8cm}{{\figtextfont B}}
+\includegraphics[width=8.4cm]{VDM_ordered}
+
+\vspace{0.2cm}
 \centering
+\raisebox{0.6cm}{{\figtextfont C}}
 \includegraphics[width=16.8cm]{average_trajectory}
-%\label{fig:fish}
-\customlabel{fig:raw_data}{2}
-\customlabel{subfig:raw_data1}{\ref{fig:fish}{\it A}}
-\customlabel{subfig:raw_data2}{\ref{fig:fish}{\it B}}
+
+\caption{(A) Images from stages 5-7 of {\it Drosophila} embryogenesis.
+Each image is of a different embryo in a different rotational orientation.
+(B) Data from A, registered and ordered using vector diffusion maps.
+(C) Average trajectory of the images shown in B.}
+\customlabel{fig:data2}{5}
+\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
+\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
+\customlabel{subfig:averaged_trajectory}{\ref{fig:data2}{\it C}}
 \end{figure*}
@@
 
+We present algorithms for registration and temporal ordering of images.
+%
+The algorithms are sufficiently general that they can be applied to a wide variety of biological imaging data.
+
+Although the examples here focus on temporally ordering the data, tasks such as classification and comparison of different mutations can also be accomplished using the same techniques. 
+%
+
+\begin{itemize}
+\item What are other people doing  (maybe critically)
+\item Movies  smoothness
+\item Discrete symmetries like flips
+\item Local information ? (features of the image) (associate with Fourier-Bessel and bispectra)
+\item Mutant-wildtype stuff  IMPORTANT
+\item And in general, MULTIPLE variablities is important (mutants, guys in another cycle, etc.)
+
+\item size is an issue
+\item  small tilts is an issue
+\item   could one do three-d image processing ?  of the floating thing, without the device
+\item how do we go about merging datasets ?
+\end{itemize}
+
+\bibliographystyle{pnas}
+\bibliography{background_reading/references,../../references/references}
+
 \end{article}

### 2014-05-09 image_analysis_paper/test_PCA.m

@@ -162,6 +162,13 @@ set(gcf, 'paperposition',[0 0 4 4])
 imshow(eigenimages(:, :, :, 1),'border','tight')
 saveas(gcf,sprintf('%s/PCA_eigenimage1', im_save_dir), 'pdf')
 
+figure;
+set(gcf, 'paperunits', 'centimeters')
+set(gcf, 'papersize', [4 4])
+set(gcf, 'paperposition',[0 0 4 4])
+imshow(eigenimages(:, :, :, 2),'border','tight')
+saveas(gcf,sprintf('%s/PCA_eigenimage2', im_save_dir), 'pdf')
+
 [~, I] = sort(proj_coeff(:, 1));
 figure;
 set(gcf, 'paperunits', 'centimeters')
@@ -212,6 +219,9 @@ xlabel('\phi_2')
 ylabel('membrane thickness')
 
 [~, I] = sort(V(:,2));
+if corr(V(:,2), mem_lengths) < 0
+    V(:,2) = -V(:,2);
+end
 figure;
 for i=1:m
     im1 = image_set_aligned(:,:,:,I(i));

### 2014-05-11 image_analysis_paper/pnas_draft2.tex
Added small edits to paper

@@
 
+\begin{figure}[tb]
+\includegraphics[width=8.4cm]{fig4}
+\caption{ {\it (A)} Set of vectors, each in a different orientation.
+The pariwise alignment angles are indicated.
+{\it (B)} The vectors from {\it A}, each rotated so that the set is globally aligned.
+Note that the chosen rotation angles are consistent with the pairwise alignments in {\it A}.
+{\it (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
+Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \eqref{eq:dmaps_W}), so that close data points are connected by dark (``stronger'') edges.
+{\it (D)} The data in {\it C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure, $\phi_2$.
+The color intensity is monotonic with the arclength, thus parameterizing the curve.}
+%\label{fig:schematics}
+\customlabel{fig:schematics}{4}
+\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
+\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
+\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}{\it C}}
+\customlabel{subfig:dmaps_color}{\ref{fig:schematics}{\it D}}
+\end{figure}
+
 \begin{figure}[t]
@@
 \customlabel{subfig:1d_profile}{\ref{fig:data1}{\it C}}
+\end{figure}
+
+\newpage
+\begin{figure}[tb]
+\raisebox{8.1cm}{{\figtextfont A}}
+\includegraphics[width=8.4cm]{PCA_ordered}
+\vspace{0.2cm}
+\raisebox{0.9cm}{{\figtextfont B}}
+\includegraphics[width=8.4cm]{average_trajectory_PCA}
+\caption{{\it (A)}  Ordering of  data set shown in Figure~\ref{subfig:raw_data1} by the first projection coefficient.  {\it (B)} Averaged trajectory of images in A.}
+\customlabel{fig:data1b}{3}
+\customlabel{subfig:eigenimage1}{\ref{fig:data1b}{\it A}}
+\customlabel{subfig:eigenimage2}{\ref{fig:data1b}{\it B}}
+\customlabel{subfig:PCA_12}{\ref{fig:data1b}{\it C}}
+\customlabel{subfig:ordered_data1}{\ref{fig:data1b}{\it D}}
 \end{figure}
@@
 
-\begin{figure}[t]
-\raisebox{3.5cm}{{\figtextfont A}}
-\includegraphics[width=4.2cm]{PCA_eigenimage1}
-\raisebox{3.5cm}{{\figtextfont B}}
-\includegraphics[width=4.2cm]{PCA_eigenimage2}
-
-\raisebox{4cm}{{\figtextfont C}}
-\includegraphics[width=8.4cm]{PCA_12}
-
-\raisebox{8cm}{{\figtextfont D}}
-\includegraphics[width=8.4cm]{PCA_ordered}
-\caption{{\it (A)} First eigenimage from the data set shown in Figure~\ref{subfig:raw_data1}.
-{\it (B)} Second eigenimage from the data set shown in Figure~\ref{subfig:raw_data1}.
-{\it (B)} Projection of  data set shown in Figure~\ref{subfig:raw_data1} onto the first two eigenimages.
-{\it (D)} Ordering of  data set shown in Figure~\ref{subfig:raw_data1} by the first projection coefficient.
-}
-\customlabel{fig:data1b}{3}
-\customlabel{subfig:eigenimage1}{\ref{fig:data1b}{\it A}}
-\customlabel{subfig:eigenimage2}{\ref{fig:data1b}{\it B}}
-\customlabel{subfig:PCA_12}{\ref{fig:data1b}{\it C}}
-\customlabel{subfig:ordered_data1}{\ref{fig:data1b}{\it D}}
-\end{figure}
-
-\begin{figure}[t]
-\includegraphics[width=8.4cm]{fig6}
-\caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different levels of development.
-{\it (B)} Membrane length as a function of time.
-The points corresponding to the images in {\it A} are indicated with arrows.
-{\it (C)} The one-dimensional concentration profiles, registered by peak Dorsal fluorescence and ordered by membrane length.
-{\it (D)} The one-dimensional concentration profiles corresponding to the automatically registered and ordered images in Figure~\ref{fig:images_ordered}.
-}
-%\label{fig:membrane_compare}
-\customlabel{fig:membrane_compare}{4}
-\customlabel{subfig:membranes}{\ref{fig:membrane_compare}{\it A}}
-\customlabel{subfig:membrane_curve}{\ref{fig:membrane_compare}{\it B}}
-\customlabel{subfig:1d_membranes}{\ref{fig:membrane_compare}{\it C}}
-\customlabel{subfig:1d_vdm}{\ref{fig:membrane_compare}{\it D}}
-\end{figure}
-
 \subsection{Temporal Ordering Using Diffusion Maps}
@@
 \newpage
-\begin{figure*}[t]
-\raisebox{10.8cm}{{\figtextfont A}}
-\includegraphics[width=8.4cm]{raw_data2}
-\hfill
-\raisebox{10.8cm}{{\figtextfont B}}
-\includegraphics[width=8.4cm]{VDM_ordered}
+\begin{figure*}[b]
+\raisebox{5.3cm}{{\figtextfont A}}
+\includegraphics[width=16.8cm]{raw_data2}
+%\hfill
+%\raisebox{10.8cm}{{\figtextfont B}}
+%\includegraphics[width=8.4cm]{VDM_ordered}
 
@@
 \centering
-\raisebox{0.6cm}{{\figtextfont C}}
+\raisebox{0.6cm}{{\figtextfont B}}
 \includegraphics[width=16.8cm]{average_trajectory}
 
-\caption{(A) Images from stages 5-7 of {\it Drosophila} embryogenesis.
+\caption{{\it (A)} Images from stages 5-7 of {\it Drosophila} embryogenesis.
 Each image is of a different embryo in a different rotational orientation.
-(B) Data from A, registered and ordered using vector diffusion maps.
-(C) Average trajectory of the images shown in B.}
+{\it (B)} Average trajectory of the images shown in A.}
 \customlabel{fig:data2}{5}

### 2014-05-11 image_analysis_paper/test_PCA.m

@@ -219,9 +219,7 @@ for i=1:nstages
     %subplot('position', [(i-1)/nstages 0 1/nstages-0.005 1])
     make_subplot(nstages, 1, 0.01, i);
     stage_indices = I(max(1, round((i-1)*m/nstages)+1):min(m,round(i*m/nstages)));
-    im1 = uint8(mean(double(image_set_aligned(:,:,:,stage_indices)), 4));
-    im1(:, :, 1) = im1(:, :, 1)  + im1(:, :, 3);
-    im1(:, :, 2) = im1(:, :, 2)  + im1(:, :, 3);
+    im1 = uint8(mean(double(image_set_aligned_withnuclei(:,:,:,stage_indices)), 4));
     
     imshow(im1,'initialmagnification','fit','border','tight')
 end
@@ -264,9 +262,7 @@ for i=1:nstages
     %subplot('position', [(i-1)/nstages 0 1/nstages-0.005 1])
     make_subplot(nstages, 1, 0.01, i);
     stage_indices = I(max(1, round((i-1)*m/nstages)+1):min(m,round(i*m/nstages)));
-    im1 = uint8(mean(double(image_set_aligned(:,:,:,stage_indices)), 4));
-    im1(:, :, 1) = im1(:, :, 1)  + im1(:, :, 3);
-    im1(:, :, 2) = im1(:, :, 2)  + im1(:, :, 3);
+    im1 = uint8(mean(double(image_set_aligned_withnuclei(:,:,:,stage_indices)), 4));
     
     imshow(im1,'initialmagnification','fit','border','tight')
 end

### 2014-05-11 image_analysis_paper/test_late_images.m

@@ -83,10 +83,10 @@ ind = 1:m;
 % use for pictures
 %ind = setdiff(1:m, [1 2 3 5 12 17 19 21 29 35 57 64 71 74 76 92 93 105 54 131 32 102]);
 %%THIS ONE
-%ind = setdiff(1:m, [1 2 3 5 12 17 19 21 29 35 57 64 71 74 76 92 93 105 54 131 32 102 101 100]);
+ind = setdiff(1:m, [1 2 3 5 12 17 19 21 29 35 57 64 71 74 76 92 93 105 54 131 32 102 101 100]);
 
 % used for pictures on 5/5
-ind = setdiff(1:m, [1 2 3 5 12 17 19 21 29 35 57 64 71 74 76 92 93 105]);
+%ind = setdiff(1:m, [1 2 3 5 12 17 19 21 29 35 57 64 71 74 76 92 93 105]);
 
 % used for pictures on 5/7
 %ind = setdiff(1:m, [1 2 3 35 57 64 71 74 76 92 93 105]);
@@ -104,10 +104,10 @@ R = R(R_ind, R_ind);
 
 m = length(ind);
 
-subplot_dim1 = floor(sqrt(m));
-subplot_dim2 = ceil(m / subplot_dim1);
-%subplot_dim1 = 9;
-%subplot_dim2 = 12;
+%subplot_dim1 = floor(sqrt(m));
+%subplot_dim2 = ceil(m / subplot_dim1);
+subplot_dim1 = 18;
+subplot_dim2 = 6;
 
 [Y, X] = meshgrid((subplot_dim2:-1:1)/subplot_dim2, (1:subplot_dim1)/subplot_dim1);
 
@@ -129,7 +129,7 @@ for i=1:m
     make_subplot(subplot_dim1, subplot_dim2, 0.01, i);
     imshow(im1);
 end
-%saveas(gcf,sprintf('%s/raw_data2', im_save_dir), 'pdf')
+saveas(gcf,sprintf('%s/raw_data2', im_save_dir), 'pdf')
 
 
 image_set_aligned = zeros(size(image_set), 'uint8');

### 2014-05-12 image_analysis_paper/pnas_draft2.tex
edited paper figures

--- 2014-05-07 image_analysis_paper/pnas_draft2.tex
@@
 %
-We present a general computational framework which greatly accelerates -and can combine- both of these tasks.
+We show how existing dimensionality reduction algorithms can greatly accelerate -and combine- both of these tasks.
+
+Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are visually apparent.
+%
+As an example, Figure~\ref{fig:fish} shows a caricature of fish development, combining the processes of growth and color patterning.
+%
+In this case, temporal ordering can be accomplished by arranging the fish by size, which is monotonic with the developmental progress.
+%
+Image registration is based on obvious morphological landmarks, such as the position of head and fins.
+%
+On the other hand, real data poses nontrivial challenges, due to the number of images, measurement noise and variability, and the subtlety of the developmental changes.
 
@@
 \end{figure}
+
+We will demonstrate our techniques using two data sets from a study of {\it Dropophila} embryogenesis.
 %
-Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are visually apparent.
+The first data set, shown in Figure~\ref{subfig:raw_data1} is relatively simple and will allow us to both illustrate and validate our methods. 
 %
-As an example, Figure~\ref{fig:fish} shows a caricature of fish development, combining the processes of growth and color patterning.
+In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex, and we will show that we can recover a reasonable trajectory that is qualitatively consistent with current knowledge.
+
+In both of these data sets, each image is taken from a different embryo, fixed at a different developmental time and imaged in a different rotational orientation.
 %
-In this case, temporal ordering can be accomplished by arranging the fish by size, which is monotonic with the developmental progress.
+We would like to organize these data sets in a meaningful and informative way.
 %
-Image registration is based on obvious morphological landmarks, such as the position of head and fins.
+This requires rotating and translating each image so the set is in a consistent frame of reference ({\it registration}), and then organizing the images. 
 %
-On the other hand, real data poses nontrivial challenges, due to the number of images, measurement noise and variability, and the subtlety of the developmental changes.
+We will show that the algorithms can effectively order the images in time and therefore give us meaningful picture of the underlying developmental dynamics.
 
-To illustrate the issues associated with temporal ordering and registration of images, we use two data sets from an imaging study of signaling protein expression in the early fruit fly embryo, one of the leading model systems for studies of developmental dynamics.
+\begin{itemize}
+\item Imaging data is becoming increasingly prevalent in biology research.
 %
-The fruit fly embryo is shaped like a rice grain, approximately $1/2$~mm long.
+\item General motivation about biology data ...
 %
-By end of the $2^{nd}$ hour of development, sequential divisions of a fertilized egg generate a system where $\sim 6,000$ nuclei are uniformly arranged under the common plasma membrane.
+\item Need methods to organize data, visualize, ...
 %
-During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclose nuclei into different cells grow in thickness.
+\item REDUCTION is important, reduction because of symmetry and reduction because of content
 %
-In conjunction with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs \cite{lim2013kinetics}.
-%
-Following cellularization, the spatial arrangement of nuclei and the concentration profiles of regulatory molecules change dramatically.
-%
-Images in Figure~\ref{subfig:raw_data1} and Figure~\ref{subfig:raw_data2} show optical sections at a fixed depth, perpendicular to the long axis of the embryo during and following cellularization, respectively.
-%
-The embryos were fixed and stained with chemicals that help visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red) \cite{chung2010microfluidic}.
-%
-%Each of the images reveals a uniform arrangement of nuclei and a nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.
-
-Each image in Figures~\ref{subfig:raw_data1}~and~\ref{subfig:raw_data2} comes from a different embryo, captured at a different rotation about the long axis and at a different and unknown point in time.
-%
-The goal is to register these images in a globally consistent way, and order them to be representative of the underlying developmental trajectory.
-%
-We will demonstrate that both of these tasks can be fully automated using a sufficiently general framework which can readily be applied to a variety of biological imaging data sets.
+\item There are a bunch of recent techniques- some established (like dmaps), some very fresh, like synchronization or VDMaps,  that have had their own reasons to be developed but they are as if tailor made for what we would want to do
+\end{itemize}
 
@@
 
+%Rank corr for data set 1 using VDM=  0.9127
+
+%Variance captured by PC 1 for data set 2 = 0.2155
+
+\subsection{Vector diffusion maps for registration and temporal ordering}
+
+We use vector diffusion maps \cite{singer2012vector} to automatically register and order our images.
+%
+Vector diffusion maps combines two algorithms, angular synchroniation \cite{singer2011angular} for registeratin and diffusion maps \cite{coifman2005geometric} for extracting low-dimensional structure, into one computation that allows us to simulatenously register and temporally order our images.
+
+Registration via angular synchronizatoin/vector diffusion maps uses pairwise alignment information about the images to globally register the images in a consistent way.
+%
+A schematic illustation is shown in Figure~\ref{subfig:synchron1}, where each image is depicted as a vector, and we would like to align the set of vectors. 
+%
+We first compute the angles needed to align pairs of vectors (or images).%
+This is rather simple to do, and requires no template function, but can be noisy and some images may be misaligned.
+%
+Using the angles between all pairs of vectors (or images), angular synchronization/ vector diffusion maps then finds the best rotation angle for each vector that is most consistent with the pairwise measurements.
+
+Diffusion maps \cite{coifman2005geometric} uncovers a parameterization of data that lies on a low-dimensional (perhaps nonlinear) manifold in high-diemnsional space. 
+%
+This is illustrated in Figure~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-diemsnional nonlinear curve. 
+%
+The color in Figure~\ref{subfig:dmaps2} depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
+%
+In our examples, each data point will be an image with many pixels, and so extracting the low-dimensional structure visually will be difficult.
+%
+Instead, we will rely on diffusion maps to automatically extract this structure from our high-dimensional data.
+
+Both of these steps are accomplished simulataneously using vector diffusion maps.
+%
+Vector diffusion maps finds a good parameterization of the data, while simultaneously registering data points which are close.
+
+\begin{figure}[t]
+\includegraphics[width=8.4cm]{fig4}
+\caption{ {\it (A)} Set of vectors, each in a different orientation.
+The pariwise alignment angles are indicated.
+{\it (B)} The vectors from {\it A}, each rotated so that the set is globally aligned.
+Note that the chosen rotation angles are consistent with the pairwise alignments in {\it A}.
+{\it (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
+Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\it SI Appendix}), so that close data points are connected by dark (``stronger'') edges.
+{\it (D)} The data in {\it C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure, $\phi_2$.
+The color intensity is monotonic with the arclength, thus parameterizing the curve.}
+%\label{fig:schematics}
+\customlabel{fig:schematics}{4}
+\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
+\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
+\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
+\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
+\end{figure}
+
+\subsection{Stage 5 images in {\it Drosophila} embryogenesis}
+
+Our first data set consists of $49$ images from stage 5 of {\em Drosophila} embryogenesis, shown in Figure~\ref{subfig:raw_data1}. 
+%
+Each image is from a different embryo in a different orientation and at a different developmental time. 
+%
+We registered and ordered the images using vector diffusion maps.
+%
+The results are shown in Figure~\ref{subfig:ordered_data1}.
+
+Furthermore, we can gain a better understanding of the developmental trajectory by looking at an averaged trajectroy.
+%
+Figure~\ref{subfig:average_data1} shows the moving average of the trajectory in Figure~\ref{subfig:ordered_data1} at different points. 
+%
+We can easily see the developmental dynamics.
+
+Furthermore, for this data set, we can validate our proposed ordering of the data.
+%
+During stage 5, we can measure the developmental time based on the monotonic progress of cellularization.
+%
+We can compare this temporal ordering with the ordering we extract from vector diffusion maps. 
+%
+Quantitatively, the rank correlation coefficient between these two orderings is 0.9127.
+
 \begin{figure*}[t]
-\includegraphics[width=5.4cm]{raw_data1}
+\raisebox{8.1cm}{{\figtextfont A}}
+\includegraphics[width=8.4cm]{raw_data1}
 \hfill
-\includegraphics[width=5.4cm]{PCA_ordered}
-\hfill
-\begin{minipage}[b]{5.6cm}
-\vspace{0in}
-\includegraphics[width=2.7cm]{PCA_eigenimage1}
-\includegraphics[width=2.7cm]{PCA_corr}
-\includegraphics[width=5.6cm]{PCA_12}
-\end{minipage}
-\caption{(A) Images collected during cellularization.
-(B) Images, now registered using angular synchronization and ordering by the first PCA projection coefficient.
-(C) First eigenimage.
-(D) Correlation between first PCA projection coefficient and membrane thickness.
-(E) Projection onto first two principal components.
-}
-\customlabel{fig:data1}{2}
+\raisebox{8.1cm}{{\figtextfont B}}
+\includegraphics[width=8.4cm]{VDM_data1_ordered}\\
+\vspace{0.2cm}
+\centering
+\raisebox{0.5cm}{{\figtextfont C}}
+\includegraphics[width=8.4cm]{average_trajectory_VDM}
+\caption{{\it (A)} Images from stage 5 of {\it Drosophila} embryogenesis.
+Each image is of a different embryo in a different rotational orientation.
+{\it (B)} Images from {\it A}, now registered and ordered using vector diffusion maps.
+{\it (C)} Average trajectory of the images shown in {\it B}.}
+\customlabel{fig:data1}{3}
 \customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
 \customlabel{subfig:ordered_data1}{\ref{fig:data1}{\it B}}
-\customlabel{subfig:eigenimage1}{\ref{fig:data1}{\it C}}
-\customlabel{subfig:PCA_corr}{\ref{fig:data1}{\it D}}
-\customlabel{subfig:PCA_12}{\ref{fig:data1}{\it E}}
+\customlabel{subfig:averaged_data1}{\ref{fig:data1}{\it C}}
 \end{figure*}
 
-\subsection{Image Registration}
+\subsection{Stage 5--7 Images of {\it Drosophila} embryogenesis}
 
-Because each embryo is imaged in a different orientation, the first task is to align or {\em register} the images so that they are in a consistent frame of reference.
+The images in Figure~\ref{fig:data1} are relatively simple, and the methods presented here are perhaps an overkill for these images.
 %
-We use angular synchronization \cite{singer2011angular} to register our images with respect to translations and rotations. 
+By visual inspection, these data are such that principal component analysis \cite{shlens2005tutorial} could uncover meaningful structure, and projection onto the first principal component would be sufficient to order the data.
 %
-Angular synchronization is a very general technique which requires no template image or feature points. 
+However, often, developmental data are significantly more complex, and nonlinear techniques such as diffusion maps/vector diffusion maps are required to extract meaningful structure in the data. 
 %
-Although we could have used other image registration algorithms (e.g.
-...) for the data sets we present here, we chose instead to use an algorithm that can be applied to a wide variety of data sets with little {\em a priori} knowledge.
+Figure~\ref{subfig:raw_data2} shows a data set of images from stages 5--7.
 %
-Implementation details are given in the {\it SI appendix}.
+Only 22\% of the variability in the data is captured by the first principal component, and so we are forced to use more sophisticated techniques.
 
-The images in Figure \ref{subfig:ordered_data1} have been registered using angular synchronization.
+Figure~\ref{subfig:ordered_data2} shows the images in Figure~\ref{subfig:raw_data2}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
 %
-The important features that our eye can visually detect (e.g., the green Dorsal peaks) have been aligned, even though the algorithm does not require identification of such features.
-
-\subsection{Temporal Ordering using Principal Component Analysis}
-
-After registering our images so that they are optimally aligned, we would like to organize our images to extract meaningful patterns and structure. 
-%
-In our data, the main variation in our images arises due to developmental dynamics, and we would like to order our data in time. 
-%
-We will use dimensionality reduction techniques to temporally order our data.
-%
-Although our images are inherently high-dimensional (number of pixels $\times$ number of channels), we assume that they lie on a low-dimensional curve which is monotonic in time. 
-%
-Furthermore, we assume that the main source of variability in this data is due to time/dynamics. 
-%
-Therefore, uncovering a one-dimensional parameterization of this curve will allow us to order our data in time.
-
-For the images shown in Figure~\ref{subfig:raw_data1}, the dynamics are relatively simple, and we can use principal component analysis (PCA) to order the images.
-%
-PCA extracts a set of {\it eigenimages} from the data, such that each data point can be (approximately) described as a linear combination of these eigenimages. 
-%
-In our dataset, the dynamics are such that each data point can be effectively described by a multiple of a single eigenimage, i.e., the images lie on a line in a very high-dimensional space.
-%
-We can organize the images by sorting them along this line.
-
-Figure~\ref{subfig:ordered_data1} shows the images, registered using angular synchronization and ordered using PCA. 
-%
-The first eigenimage is shown in Figure~\ref{subfig:eigenimage1}.
-
-For these experiments in this particular time window, we are fortunate enough to be able to validate our  ordering using prior physical knowledge about the system.
-%
-Temporal ordering, on the other hand, can be done based on the monotonic progress of cellularization \cite{figard2013plasma}.%; this helps validate our vector diffusion map-based ordering.
-%
-Figure~\ref{subfig:PCA_corr} shows the rank correlation between the two orderings.
-%
-The rank correlation coefficient is 0.9199, indicating that we do a fairly good job of recovering the correct temporal ordering of the images.
-
-Although PCA can successfully order the data set shown in Figure \ref{subfig:raw_data1}, it can only do so because the data lie on in a one-dimensional, {\it linear} subspace.
-%
-In general, the developmental trajectory is not guaranteed to be linear. 
-%
-This can be seen in the late images of Figure~\ref{subfig:ordered_data1} that the ordering is not entirely correct.
-%
-We can also see this from the projection onto the first two eigenimages, shown in Figure~\ref{subfig:PCA_12}.
-%
-The data clearly lie on a one-dimensional, {\em nonlinear} curve.
-%
-This motivates us to use nonlinear dimensionality reduction techniques to temporally order our data.
-
-\subsection{Temporal Ordering Using Diffusion Maps}
-
-We can use diffusion maps \cite{coifman2005geometric} to analyze more complex data sets where the dynamics are clearly nonlinear.
-%
-Diffusion maps is similar to PCA, in that it aims uncover low-dimensional structure in high-dimensional data.
-%
-However, unlike PCA, diffusion maps is a {\it nonlinear} reduction technique. 
-%
-Figure~\ref{subfig:raw_data2} shows a data set consisting of embryos following cellularization.
-%
-Clearly, the dynamics are significantly more complex than the dynamics in Figure~\ref{subfig:raw_data1}, and the first PCA eigenimage will not capture all of the relevant dynamics.
-%
-However, we can still use diffusion maps to temporally order the data.
-%
-Figure~\ref{subfig:ordered_data2} shows the images in Figure~\ref{subfig:raw_data2}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
-
 Unlike the images in Figure~\ref{subfig:raw_data1}, 
@@
 %
-However, there are noticeable features in the ordering that we can see are consistent with previously known dynamics. 
-%
-We can also construct a smoothed trajectory by averaging over neighboring images.
+However, there are noticeable features in the ordering that we can see are consistent with previously known dynamics,
+and we can construct a smoothed trajectory by averaging groups of neighboring images.
 %
@@
 %
-The dynamics are readily apparent; we can see gastrulation, ....
+We can easily see the formation of the ventral furrow and germ band elongation, which concludes with ventral cells shifted to the dorsal side of the embryo.
 
@@
 \begin{figure*}[t]
-\includegraphics[width=8.4cm]{raw_data2}
-\hfill
-\includegraphics[width=8.4cm]{VDM_ordered} \\
-\centering
+\raisebox{5.3cm}{{\figtextfont A}}
+\includegraphics[width=16.8cm]{raw_data2}
+
+\vspace{0.2cm}
+\raisebox{5.3cm}{{\figtextfont B}}
+\includegraphics[width=16.8cm]{VDM_ordered}
+
+\vspace{0.2cm}
+\raisebox{0.6cm}{{\figtextfont C}}
 \includegraphics[width=16.8cm]{average_trajectory}
-\caption{(A) Images collected following cellularization.
+\caption{{\it (A)} Images from stages 5-7 of {\it Drosophila} embryogenesis.
 Each image is of a different embryo in a different rotational orientation.
-(B) Data from A, registered and ordered using vector diffusion maps.
-(C) Average trajectory of the images shown in B.}
-\customlabel{fig:data2}{3}
+{\it(B)} Data from {\it A}, registred and ordered using vector diffusion maps.
+{\it (C)} Average trajectory of the images shown in {\it B}.}
+\customlabel{fig:data2}{5}
 \customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
 \customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
-\customlabel{subfig:averaged_trajectory}{\ref{fig:data2}{\it C}}
+\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
 \end{figure*}
@@
 
+We have shown how symmetry and dimensionality reduction techniques can successfully register and order cross-sectional developmental imaging data.
+%
+Although we illustrated our methods using a specific data set from a study of {\it Drosophila} embryogenesis, the framework is quite general and requires little knowledge of the image features or the developmental dynamics.
+%
+To our knowledge, temporal ordering of {\it Drosophila} images has not been automated to this extent.
+%
+Recently, machine learning algorithms have been used to group {\it Drosophila} images into different developmental classes \cite{yuan2014automated}, yet there, the appropriate classes must be defined by the user and hand-labeled training data are required.
+
+Temporal ordering of cross-sectional biological data has been studied in a variety of other contexts.
+%
+Cross-sectional RNA sequencing data has been temporally ordered \cite{anavy2014blind, trapnell2014dynamics}, where each data point is the transcriptome for a single sample.
+%
+Gene expression data from multiple patients have been studied in \cite{gupta2008extracting, qiu2011discovering};
+here, each data point is a vector containing the expression levels of the various genes.
+%
+Snapshots of cells throughout the cell cycle can be described by a vector of features (e.g., amount of DNA) which quantify the cell's state  \cite{kafri2013dynamics}.
+%
+These problems are no different than our task of ordering images, each of which is a large vector of pixel intensities.
+%
+The majority of these works temporally order the data either by solving a traveling salesman problem on the data, or by constructing a minimum spanning tree of the data,
+expecting that these structures (the path of the traveling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate encoding of the dynamical progression.
+%
+We, instead, use diffusion maps to construct a one-dimensional parameterization of the data.
+%
+Diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003, tenenbaum2000global, Donoho2003, Roweis2000}.
+%
+They have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}, and so we suspect they may perform better than the ordering algorithms used in previous work.
+%
+{\bf Yannis says ``careful'', he thinks this might be too strong of a statement.}
+%%%
+%%%   we need help here
+%%%
+%
+%Furthermore, although we use only the first diffusion maps coordinate to order the data (we assume the data are one-dimensional), diffusion maps are not limited to one-dimensional data and can be used to extract more complex structures.
+%
+%For example, in data which contains several branches or classes, the first few coordinates will delineate the branches or separate the classes.
+
+The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
+%
+The main point to note here is that many registration algorithms require definition of appropriate landmarks or features for the images and the location of these within each image.
+%
+For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then register images by aligning these features \cite{zhao2003face}.
+%
+Temporally ordering images of a construction site over time is also based on first identifying features within the images \cite{schindler2007inferring}.
+%
+In general, appropriate features may not be known {\it a priori}, and so we require a registration algorithm that does not rely on predefined features, but only on the inherent geometric symmetry of the problem.
+%
+Template-based methods are also common for registration, where one optimally aligns each image to a predefined template function.
+%
+However, these methods can suffer when the images are noisy, and the success of such methods can depend strongly on the template function \cite{shatsky2009method}.
+%
+{\bf Need to discuss consistency.}
+
+Angular synchronization and vector diffusion maps require no feature points and no template function.
+%
+These methods have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}, where each image is a picture of (essentially) the {\it same} molecule; variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
+%
+In contrast, our data contains variation from the different embryo orientations, as well as variation from the underlying developmental dynamics.
+% (which underpins the development of vector diffusion map algorithms).
+%
+In addition to the eigenproblem relaxation used here, other synchronization algorithms have been proposed \cite{bandeira2014multireference}.
+%
+Synchronization is formulated for general symmetry groups, and depending on the data, one might consider other symmetries, such as reflections and scaling/dilation.
+%
+Here, the main requirement is that the symmetry group be represented by orthogonal or unitary matrices \cite{singer2013spectral}.
+
+Temporal ordering can also be accomplished without first registering the images, by constructing functions of the images which are invariant under translations and rotations. 
+%
+These functions are then used as the data in the diffusion maps calculation; since these representative functions do not change under translations and rotations, the resulting diffusion map embedding is invariant to symmetries.
+%
+Such functions include the scale invariant feature transform \cite{lowe1999object, lowe2004distinctive}, the scattering transform \cite{mallat2012group}, and the bispectrum \cite{zhao2014rotationally}.
+%
+These methods have the disadvantage that we do not obtain a registered set of images, but only a temporal ordering.
+%
+However, such methods can often be faster for large data sets, since the functions are computed per image, rather than computing pairwise quantities in synchronization.
+
+In all of these data-driven methods, an essential question is how much and what type of data is needed to accurately reconstruct the dynamics.
+%
+For diffusion maps to temporally order the data, we require the developmental trajectory to be well-sampled, and that the principal direction of variability in the data (the coordinate which diffusion maps will recover) be monotonic in time.
+%
+In addition, the data should have spatial variability/structure for synchronization to successfully register the images.
+%
+In our example, the data set is sufficiently large so that the dynamics are well-sampled.
+%
+We have included both the red and the green channels so that the principal direction of variability is monotonic in time.
+%
+If we only used the green channel in our analysis, we could not successfully order the data using diffusion maps, as this signal does not change appreciably in time.
+%
+Both the red and green channels contain spatial variability which allows us to register the data points: the green channel has one peak at a specific angular around the embryo, and the red channel has two peaks which grow in time.
+%
+We include the green channel because the location of the red peaks is noisy, and so inclusion of the (less noisy) green channel allows us to more accurately register the images.
+%
+%For the methods we present here, we require a sufficient amount of data for the one-dimensional evolution curve to be ``seen'' by the algorithms (i.e., in Figure~\ref{subfig:dmaps_edges}, there are a sufficient number of data points so that our eye can see the curve; if there were only 4 data points, it would most likely be difficult for our eye to detect the one-dimensional structure).
+%
+%We also require the data to contain enough information so that the alignments and dynamics are apparent.
+%
+%For example, in the data presented in Figure~\ref{subfig:images_unordered}, the green peaks delineate the position of the ventral axis in the embryo, and the intensity of the red peaks grow over time.
+%
+%Therefore, these images contain information about rotational symmetries of the data as well as the dynamics of the data.
+%
+%If we removed the green signal, and only used the red signal in our analysis, the registration would not be as accurate.
+%
+%Furthermore, if we only used the green signal, then there would be no dynamic information (the shape or intensity of Dorsal does not change as a function of time in the developmental stage we are considering), and the temporal ordering would be inaccurate.
+%
+We are confident that for many applications, these requirements - data which adequately samples the developmental trajectory, and data with sufficient spatial and temporal variability - are easily met.
+%
+In these cases, our methods allow for the rapid, automated analysis of imaging data sets to help uncover possibly complex dynamics and structure.
+
 Although the examples here focus on temporally ordering the data, tasks such as classification and comparison of different mutations can also be accomplished using the same techniques. 
 %
+
+\begin{itemize}
+\item What are other people doing  (maybe critically)
+\item Movies  smoothness
+\item Discrete symmetries like flips
+\item Local information ? (features of the image) (associate with Fourier-Bessel and bispectra)
+\item Mutant-wildtype stuff  IMPORTANT
+\item And in general, MULTIPLE variablities is important (mutants, guys in another cycle, etc.)
+
+\item size is an issue
+\item  small tilts is an issue
+\item   could one do three-d image processing ?  of the floating thing, without the device
+\item how do we go about merging datasets ?
+\end{itemize}
 

### 2014-05-12 image_analysis_paper/test_PCA.m

@@ -114,12 +114,14 @@ saveas(gcf,sprintf('%s/raw_data1', im_save_dir), 'pdf')
 
 image_set_aligned = zeros(size(image_set), 'uint8');
 image_set_aligned_withnuclei = zeros(size(image_set), 'uint8');
+theta_adjust = -58;
 for i=1:m
     im_tmp = image_set(:,:,:,i);
     %         for j=1:3
     %             im_tmp(:,:,j) = im_tmp(:,:,j) + nuclei(:,:,i);
     %         end
     im1 = rotate_image(R_opt(dim*(i-1)+1:dim*i,:), im_tmp, angle_proj);
+    im1 = imrotate(im1, theta_adjust, 'crop');
     image_set_aligned(:,:,:,i) = im1;
     
     im_tmp = image_set(:,:,:,i);
@@ -127,6 +129,7 @@ for i=1:m
         im_tmp(:,:,j) = im_tmp(:,:,j) + nuclei(:,:,i);
     end
     im1 = rotate_image(R_opt(dim*(i-1)+1:dim*i,:), im_tmp, angle_proj);
+    im1 = imrotate(im1, theta_adjust, 'crop');
     image_set_aligned_withnuclei(:,:,:,i) = im1;
 end
 
@@ -147,7 +150,9 @@ fontsize = 8;
 
 data_reshaped = zeros(m, npixels*npixels*nchannels);
 
-[eigenimages, D, proj_coeff] = PCA_images(image_set_aligned, 9);
+[eigenimages, D_PCA, proj_coeff] = PCA_images(image_set_aligned, m);
+
+diag(D_PCA) / sum(diag(D_PCA))
 
 figure;
 for i=1:9
@@ -226,7 +231,7 @@ end
 saveas(gcf,sprintf('%s/average_trajectory_PCA', im_save_dir), 'pdf')
 
 
-%%
+%% DMAPS on PCA
 W2 = squareform(pdist(proj_coeff(:, 1:2))).^2;
 eps2 = median(W2(:));
 
@@ -267,6 +272,76 @@ for i=1:nstages
     imshow(im1,'initialmagnification','fit','border','tight')
 end
 saveas(gcf,sprintf('%s/average_trajectory_DMAPS', im_save_dir), 'pdf')
+
+%% VDM
+
+eps = median(W(:))/10;
+neigs = 10;
+
+[R_opt, embed_coord, embed_idx, D] = vdm(R, W, eps, neigs);
+
[…]

### 2014-05-12 image_analysis_paper/test_late_images.m

@@ -33,14 +33,11 @@ for i=1:m
     
     % resize image
     im1 = imresize(im1, [npixels npixels]);
-    
+
     %     blur image
-    %used on 5/5
-    H = fspecial('gaussian',[5 5], 2.0);
+    H = fspecial('disk',5);
     im1(:, :, 1) = imfilter(im1(:, :, 1),H,'replicate');
-    im1(:, :, 1) = 0.5*imadjust(im1(:, :, 1));
-    %im1(:, :, 1) = 0;
-    
+    im1(:, :, 1) = 0.5*imadjust(im1(:, :, 1));    
     
     im1(:, :, 3) = 1.5 * im1(:, :, 3);
     
@@ -53,7 +50,6 @@ for i=1:m
     image_set(:, :, :, i) = im1;
     
 end
-
 
 %% raw dpERK images-- synchronization
 
@@ -69,7 +65,11 @@ dim = 3;
 %save('pairwise_alignments_late_noblur.mat', 'R', 'W');
 %save('pairwise_alignments_late.mat', 'R', 'W');
 %save('pairwise_alignments_nonuclei.mat', 'R', 'W');
+
+
+
 load('pairwise_alignments_late.mat');
+
 %load('pairwise_alignments_nonuclei.mat');
 
 
@@ -77,6 +77,7 @@ load('pairwise_alignments_late.mat');
 % all
 ind = 1:m;
 
+%ind = setdiff(1:m, [1 2 3 7 17 29]);
 % used for pictures on 5/6
 %ind = setdiff(1:m, [1 2 3 5 12 17 19 21 29 35 57 64 71 74 76 92 93 105 54 131 50 102]);
 
@@ -104,8 +105,8 @@ R = R(R_ind, R_ind);
 
 m = length(ind);
 
-%subplot_dim1 = floor(sqrt(m));
-%subplot_dim2 = ceil(m / subplot_dim1);
+% subplot_dim1 = floor(sqrt(m));
+% subplot_dim2 = ceil(m / subplot_dim1);
 subplot_dim1 = 18;
 subplot_dim2 = 6;
 
@@ -196,7 +197,7 @@ end
 
 %% try PCA
 
-[eigenimages, D, proj_coeff] = PCA_images(image_set_aligned, 9);
+[eigenimages, D_PCA, proj_coeff] = PCA_images(image_set_aligned, m);
 
 [~, I] = sort(proj_coeff(:, 1));
 figure;
[…]

### 2014-05-13 image_analysis_paper/pnas_draft2.tex
edited draft

@@
 %
-The mathematical techniques (angular synchronization for image registration, diffusion maps for temporal ordering, and
-vector diffusion maps for simultaneously performing both tasks) are applicable to a wide variety of data sets and
+The mathematical techniques (vector diffusion maps) are applicable to a wide variety of data sets and
 require little {\it a priori} knowledge of the image features or the developmental dynamics.
@@
 
-We will demonstrate our techniques using two data sets from a study of {\it Dropophila} embryogenesis.
+We will show that we can use existing dimensionality reduction techniques to register and temporally order large sets of images from developmental biology studies. 
 %
-The first data set, shown in Figure~\ref{subfig:raw_data1} is relatively simple and will allow us to both illustrate and validate our methods. 
+We will demonstrate these techniques using two data sets from a study of {\it Dropophila} embryogenesis.
 %
-In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex, and we will show that we can recover a reasonable trajectory that is qualitatively consistent with current knowledge.
+The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate our methods. 
+%
+In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex; here, we will show that we can recover a reasonable trajectory that is qualitatively consistent with current knowledge.
 
@@
 %
-This requires rotating and translating each image so the set is in a consistent frame of reference ({\it registration}), and then organizing the images. 
+This requires rotating and translating each image so the set is in a consistent frame of reference ({\it registration}), and then temporally ordering the images. 
 %
-We will show that the algorithms can effectively order the images in time and therefore give us meaningful picture of the underlying developmental dynamics.
+We will show that dimensionality reduction algorithms can accomplish both of these tasks and therefore give us a meaningful picture of the underlying developmental dynamics.
 
@@
 %
-Vector diffusion maps combines two algorithms, angular synchroniation \cite{singer2011angular} for registeratin and diffusion maps \cite{coifman2005geometric} for extracting low-dimensional structure, into one computation that allows us to simulatenously register and temporally order our images.
+Vector diffusion maps is a general technique, developed for data sets which contain both symmetries (e.g., translations and rotations) which one would like to factor out, and other directions of variability which one would like to uncover.
+%
+It combines two algorithms, angular synchronization \cite{singer2011angular} for registration and diffusion maps \cite{coifman2005geometric} for extracting low-dimensional structure, into one computation that allows us to simultaneously register and temporally order our images.
+%
+Vector diffusion maps has been applied to cryo-EM images \cite{...}, where each image is a (very noisy) projection of a molecule, taken from a different viewing direction and in a different rotation. 
+%
+Here, we will use it to register the images with respect to rotations and translations, as well as uncover the main direction of variability, which is parameterized by time. 
+%
+The noise in our experiments is primarily from interembryo variability, rather than white noise from a measurement instrument (as in the cryo-EM example).
 
-Registration via angular synchronizatoin/vector diffusion maps uses pairwise alignment information about the images to globally register the images in a consistent way.
+Registration via angular synchronization/vector diffusion maps uses pairwise alignment information to globally register the images in a consistent way.
 %
-A schematic illustation is shown in Figure~\ref{subfig:synchron1}, where each image is depicted as a vector, and we would like to align the set of vectors. 
+A schematic illustration is shown in Figure~\ref{subfig:synchron1}; each image is depicted as a vector, and the goal is to align the set of vectors. 
 %
-We first compute the angles needed to align pairs of vectors (or images).%
-This is rather simple to do, and requires no template function, but can be noisy and some images may be misaligned.
+We first compute the angles needed to align pairs of vectors (or images).
 %
-Using the angles between all pairs of vectors (or images), angular synchronization/ vector diffusion maps then finds the best rotation angle for each vector that is most consistent with the pairwise measurements.
+This is rather simple to do, and requires no template function; however, when the images are noisy, some pairwise alignment angles may be inaccurate.
+%
+Using the angles between all pairs of vectors (or images), angular synchronization/vector diffusion maps then finds the set of rotation angles, one angle for each vector, that is most consistent with the pairwise measurements.
 
-Diffusion maps \cite{coifman2005geometric} uncovers a parameterization of data that lies on a low-dimensional (perhaps nonlinear) manifold in high-diemnsional space. 
+In general, diffusion maps \cite{coifman2005geometric} is a nonlinear dimensionality reduction technique;
+it uncovers a parameterization of data that lies on a low-dimensional (perhaps nonlinear) manifold in high-dimensional space. 
 %
-This is illustrated in Figure~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-diemsnional nonlinear curve. 
+This idea is illustrated in Figure~\ref{subfig:dmaps1}: the data are two-dimensional points which lie on a one-dimensional nonlinear curve. 
 %
@@
 %
-In our examples, each data point will be an image with many pixels, and so extracting the low-dimensional structure visually will be difficult.
+In our examples, each data point will be very high dimensional (e.g, an image with many pixels), and so we cannot extract this low-dimensional structure visually.
 %
-Instead, we will rely on diffusion maps to automatically extract this structure from our high-dimensional data.
-
-Both of these steps are accomplished simulataneously using vector diffusion maps.
+Instead, we will rely on diffusion maps to automatically extract a parameterization of our high-dimensional data.
 %
-Vector diffusion maps finds a good parameterization of the data, while simultaneously registering data points which are close.
+We will assume that our data are (approximately) one-dimensional, and that this dimension is parameterized by time.
+%
+Therefore, ordering our data along this main dimension/direction will temporally order our data. 
+%
+Vector diffusion maps combines angular synchronization and diffusion maps to find a parameterization of the data while simultaneously registering data points which are close.
 
@@
 \caption{ {\it (A)} Set of vectors, each in a different orientation.
-The pariwise alignment angles are indicated.
+The pairwise alignment angles are indicated.
 {\it (B)} The vectors from {\it A}, each rotated so that the set is globally aligned.
@@
 
-Our first data set consists of $49$ images from stage 5 of {\em Drosophila} embryogenesis, shown in Figure~\ref{subfig:raw_data1}. 
+Our first data set, shown in Figure~\ref{subfig:raw_data1}, consists of $49$ images from stage 5 of {\em Drosophila} embryogenesis.
 %
-Each image is from a different embryo in a different orientation and at a different developmental time. 
+Each image is from a different embryo imaged in a different orientation and fixed at a different developmental time. 
 %
@@
 The results are shown in Figure~\ref{subfig:ordered_data1}.
+%
+The features that our eyes can visually detect (the green Dorsal peak, and the red dpERK peaks) are now consistently registered, even though the algorithm does not require any information about the existence or location of these peaks.
+%
+In addition, the red dpERK peaks grow in intensity over the developmental trajectory, as expected.
 
-Furthermore, we can gain a better understanding of the developmental trajectory by looking at an averaged trajectroy.
+When dealing with large sets of images with interembryo variability, we can gain a better understanding of the developmental trajectory by looking at an averaged trajectory.
 %
-Figure~\ref{subfig:average_data1} shows the moving average of the trajectory in Figure~\ref{subfig:ordered_data1} at different points. 
+Figure~\ref{subfig:average_data1} shows the moving average of the trajectory in Figure~\ref{subfig:ordered_data1}. 
 %
-We can easily see the developmental dynamics.
+We can now easily see the developmental dynamics: the red dpERK peaks grow in time, and the ventral furrow (the dent in green) begins to form at the end of stage 5.
 
-Furthermore, for this data set, we can validate our proposed ordering of the data.
+For this data set, we can validate our proposed vector diffusion maps ordering of the data.
 %
-During stage 5, we can measure the developmental time based on the monotonic progress of cellularization.
+During stage 5, we can measure the developmental time based on the monotonic progress of cellularization (details are given in the {\it SI Appendix}).
 %
-We can compare this temporal ordering with the ordering we extract from vector diffusion maps. 
+We can compare the temporal ordering obtained from cellularization with the ordering we extract from vector diffusion maps. 
 %
-Quantitatively, the rank correlation coefficient between these two orderings is 0.9127.
+The rank correlation coefficient between these two orderings is 0.9127, indicating that there is very good agreement and vector diffusion maps is uncovering the correct temporal ordering of the images.
 
@@
 
-The images in Figure~\ref{fig:data1} are relatively simple, and the methods presented here are perhaps an overkill for these images.
+The images in Figure~\ref{fig:data1} are relatively simple, and the methods presented here are perhaps unnecessarily complex for these images. 
 %
-By visual inspection, these data are such that principal component analysis \cite{shlens2005tutorial} could uncover meaningful structure, and projection onto the first principal component would be sufficient to order the data.
+These data are effectively a single mode growing in time.
 %
-However, often, developmental data are significantly more complex, and nonlinear techniques such as diffusion maps/vector diffusion maps are required to extract meaningful structure in the data. 
+Therefore, principal component analysis \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data.
 %
-Figure~\ref{subfig:raw_data2} shows a data set of images from stages 5--7.
+However, developmental dynamics are often significantly more complex, and nonlinear techniques such as diffusion maps/vector diffusion maps are required to extract meaningful structure in the data. 
 %
-Only 22\% of the variability in the data is captured by the first principal component, and so we are forced to use more sophisticated techniques.
+Figure~\ref{subfig:raw_data2} shows a data set of images from stages 5--7 of {\it Drosophila} embryogenesis.
+%
+Only 22\% of the variability in the data is captured by the first principal component, and so we are forced to use more sophisticated techniques, such as vector diffusion maps, to temporally order the data.
 
@@
 %
-However, there are noticeable features in the ordering that we can see are consistent with previously known dynamics,
-and we can construct a smoothed trajectory by averaging groups of neighboring images.
+However, there are noticeable features in the ordering that we can see are consistent with previously known dynamics.
 %
-Figure~\ref{subfig:averaged_trajectory} shows the images from Figure~\ref{subfig:ordered_data2}, divided into 12 groups and averaged. 
+We can again construct a smoothed trajectory by looking at the moving average of the images, shown in Figure~\ref{sub fig:average_data2}.
 %
 We can easily see the formation of the ventral furrow and germ band elongation, which concludes with ventral cells shifted to the dorsal side of the embryo.
+%
+This assures us that we can obtained an ordering of the data which is developmentally meaningful.
 
@@
 
-We present algorithms for registration and temporal ordering of images.
+We have demonstrated how existing reduction techniques (vector diffusion maps) can be used to register and temporally order images from developmental biology studies. 
 %
-The algorithms are sufficiently general that they can be applied to a wide variety of biological imaging data.
+We acknowledge that the tasks of both registration and ordering can be accomplished using a variety of techniques.
+%
+We chose the techniques outlined here, not because they are particularly suited to this specific problem, but rather, because they are sufficiently general and we are confident that they can be applied to a myriad of biological imaging applications.
 
-We have shown how symmetry and dimensionality reduction techniques can successfully register and order cross-sectional developmental imaging data.
+The task of image registration has been studied in a variety of contexts \cite{...}.
 %
-Although we illustrated our methods using a specific data set from a study of {\it Drosophila} embryogenesis, the framework is quite general and requires little knowledge of the image features or the developmental dynamics.
+However, most of these applications first the definition and identification of good {\it features} within each image. 
 %
-To our knowledge, temporal ordering of {\it Drosophila} images has not been automated to this extent.
+For our biological applications, good features may not be known {\it a priori}.
 %
-Recently, machine learning algorithms have been used to group {\it Drosophila} images into different developmental classes \cite{yuan2014automated}, yet there, the appropriate classes must be defined by the user and hand-labeled training data are required.
+We therefore use an algorithm which requires no features, but instead, uses the images directly.
 
-Temporal ordering of cross-sectional biological data has been studied in a variety of other contexts.
+The task of temporal ordering has also been studied, for data such as ...
 %
-Cross-sectional RNA sequencing data has been temporally ordered \cite{anavy2014blind, trapnell2014dynamics}, where each data point is the transcriptome for a single sample.
-%
-Gene expression data from multiple patients have been studied in \cite{gupta2008extracting, qiu2011discovering};
-here, each data point is a vector containing the expression levels of the various genes.
-%
-Snapshots of cells throughout the cell cycle can be described by a vector of features (e.g., amount of DNA) which quantify the cell's state  \cite{kafri2013dynamics}.
-%
-These problems are no different than our task of ordering images, each of which is a large vector of pixel intensities.
-%
-The majority of these works temporally order the data either by solving a traveling salesman problem on the data, or by constructing a minimum spanning tree of the data,
+Most of these applications order the data by solving a traveling salesman problem or constructing a minimum spanning tree on the data,
 expecting that these structures (the path of the traveling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate encoding of the dynamical progression.
@@
 % 
-{\bf Yannis says ``careful'', he thinks this might be too strong of a statement.}
-%%%
-%%%   we need help here
-%%%
+Furthermore, techniques such as diffusion maps do not constrain us to one-dimensional parameterizations/organizations of the data.
 % 
-%Furthermore, although we use only the first diffusion maps coordinate to order the data (we assume the data are one-dimensional), diffusion maps are not limited to one-dimensional data and can be used to extract more complex structures.
+In this particular setup, we expect our data to be one-dimensional, as the only (main) source of variability is time.
 %
-%For example, in data which contains several branches or classes, the first few coordinates will delineate the branches or separate the classes.
-
-The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
-%
-The main point to note here is that many registration algorithms require definition of appropriate landmarks or features for the images and the location of these within each image.
-%
-For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then register images by aligning these features \cite{zhao2003face}.
-%
-Temporally ordering images of a construction site over time is also based on first identifying features within the images \cite{schindler2007inferring}.
-%
-In general, appropriate features may not be known {\it a priori}, and so we require a registration algorithm that does not rely on predefined features, but only on the inherent geometric symmetry of the problem.
-%
-Template-based methods are also common for registration, where one optimally aligns each image to a predefined template function.
-%
-However, these methods can suffer when the images are noisy, and the success of such methods can depend strongly on the template function \cite{shatsky2009method}.
-%
-{\bf Need to discuss consistency.}
-
-Angular synchronization and vector diffusion maps require no feature points and no template function.
-%
-These methods have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}, where each image is a picture of (essentially) the {\it same} molecule; variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
-%
-In contrast, our data contains variation from the different embryo orientations, as well as variation from the underlying developmental dynamics.
-% (which underpins the development of vector diffusion map algorithms).
-%
-In addition to the eigenproblem relaxation used here, other synchronization algorithms have been proposed \cite{bandeira2014multireference}.
-%
-Synchronization is formulated for general symmetry groups, and depending on the data, one might consider other symmetries, such as reflections and scaling/dilation.
-%
-Here, the main requirement is that the symmetry group be represented by orthogonal or unitary matrices \cite{singer2013spectral}.
-
-Temporal ordering can also be accomplished without first registering the images, by constructing functions of the images which are invariant under translations and rotations. 
-%
-These functions are then used as the data in the diffusion maps calculation; since these representative functions do not change under translations and rotations, the resulting diffusion map embedding is invariant to symmetries.
-%
-Such functions include the scale invariant feature transform \cite{lowe1999object, lowe2004distinctive}, the scattering transform \cite{mallat2012group}, and the bispectrum \cite{zhao2014rotationally}.
-%
-These methods have the disadvantage that we do not obtain a registered set of images, but only a temporal ordering.
-%
-However, such methods can often be faster for large data sets, since the functions are computed per image, rather than computing pairwise quantities in synchronization.
-
-In all of these data-driven methods, an essential question is how much and what type of data is needed to accurately reconstruct the dynamics.
-%
-For diffusion maps to temporally order the data, we require the developmental trajectory to be well-sampled, and that the principal direction of variability in the data (the coordinate which diffusion maps will recover) be monotonic in time.
-%
-In addition, the data should have spatial variability/structure for synchronization to successfully register the images.
-%
-In our example, the data set is sufficiently large so that the dynamics are well-sampled.
-%
-We have included both the red and the green channels so that the principal direction of variability is monotonic in time.
-%
-If we only used the green channel in our analysis, we could not successfully order the data using diffusion maps, as this signal does not change appreciably in time.
-%
-Both the red and green channels contain spatial variability which allows us to register the data points: the green channel has one peak at a specific angular around the embryo, and the red channel has two peaks which grow in time.
-%
-We include the green channel because the location of the red peaks is noisy, and so inclusion of the (less noisy) green channel allows us to more accurately register the images.
-%
-%For the methods we present here, we require a sufficient amount of data for the one-dimensional evolution curve to be ``seen'' by the algorithms (i.e., in Figure~\ref{subfig:dmaps_edges}, there are a sufficient number of data points so that our eye can see the curve; if there were only 4 data points, it would most likely be difficult for our eye to detect the one-dimensional structure).
-%
-%We also require the data to contain enough information so that the alignments and dynamics are apparent.
-%
-%For example, in the data presented in Figure~\ref{subfig:images_unordered}, the green peaks delineate the position of the ventral axis in the embryo, and the intensity of the red peaks grow over time.
-%
-%Therefore, these images contain information about rotational symmetries of the data as well as the dynamics of the data.
-%
-%If we removed the green signal, and only used the red signal in our analysis, the registration would not be as accurate.
-%
-%Furthermore, if we only used the green signal, then there would be no dynamic information (the shape or intensity of Dorsal does not change as a function of time in the developmental stage we are considering), and the temporal ordering would be inaccurate.
-%
-We are confident that for many applications, these requirements - data which adequately samples the developmental trajectory, and data with sufficient spatial and temporal variability - are easily met.
-%
-In these cases, our methods allow for the rapid, automated analysis of imaging data sets to help uncover possibly complex dynamics and structure.
-
-Although the examples here focus on temporally ordering the data, tasks such as classification and comparison of different mutations can also be accomplished using the same techniques. 
-%
+However, other sources of variability would also be uncovered using diffusion maps, 
+and tasks such as classification and comparison of different mutations can also be accomplished using the same techniques.
 

### 2014-05-13 image_analysis_paper/pnas_draft2.tex
edited draft

@@
 The features that our eyes can visually detect (the green Dorsal peak, and the red dpERK peaks) are now consistently registered, even though the algorithm does not require any information about the existence or location of these peaks.
-%
-In addition, the red dpERK peaks grow in intensity over the developmental trajectory, as expected.
 
-When dealing with large sets of images with interembryo variability, we can gain a better understanding of the developmental trajectory by looking at an averaged trajectory.
+When dealing with large sets of images with interembryo variability, we can gain a better understanding of the developmental dynamics by looking at an averaged trajectory.
 %
@@
 %
-During stage 5, we can measure the developmental time based on the monotonic progress of cellularization (details are given in the {\it SI Appendix}).
+During stage 5, we can measure the developmental time of each embryo based on the monotonic progress of cellularization (details are given in the {\it SI Appendix}).
 %
-We can compare the temporal ordering obtained from cellularization with the ordering we extract from vector diffusion maps. 
+We compared the temporal ordering obtained from cellularization with the ordering we extract from vector diffusion maps. 
 %
@@
 
-The images in Figure~\ref{fig:data1} are relatively simple, and the methods presented here are perhaps unnecessarily complex for these images. 
+The images in Figure~\ref{fig:data1} are relatively simple, and the methods presented here are perhaps unnecessarily complex for this particular data set. 
 %
-These data are effectively a single mode growing in time.
-%
-Therefore, principal component analysis \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data.
+Because these data are effectively a single mode growing in time,
+principal component analysis \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data.
 %
@@
 %
+Again, visually detectable features, such as the green Twist peaks, are consistently aligned. 
+%
 Unlike the images in Figure~\ref{subfig:raw_data1}, 
@@
 %
-We can easily see the formation of the ventral furrow and germ band elongation, which concludes with ventral cells shifted to the dorsal side of the embryo.
+We can see the formation of the ventral furrow and germ band elongation, which concludes with ventral cells shifted to the dorsal side of the embryo, which is consistent with the known picture of {\it Drosophila} embryogeneisis.
 %
@@
 %
-However, most of these applications first the definition and identification of good {\it features} within each image. 
+However, most of these applications rely on the definition and identification of good {\it features} within each image. 
 %
-For our biological applications, good features may not be known {\it a priori}.
-%
-We therefore use an algorithm which requires no features, but instead, uses the images directly.
+For our biological applications, good features may not be known {\it a priori}, 
+and we instead use an algorithm which requires no features, but instead, uses the images directly.
 
@@
 %
-However, other sources of variability would also be uncovered using diffusion maps, 
-and tasks such as classification and comparison of different mutations can also be accomplished using the same techniques.
+However, other sources of variability would also be uncovered using (vector) diffusion maps, 
+and tasks such as classification of embryos by size or viewing angle, and comparison of different mutations can also be accomplished using the same techniques.
 

### 2014-05-13 image_analysis_paper/pnas_draft2.tex
edited draft a little

@@
 %\label{fig:schematics}
-\customlabel{fig:schematics}{4}
+\customlabel{fig:schematics}{2}
 \customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
@@
 {\it (B)} Images from {\it A}, now registered and ordered using vector diffusion maps.
-{\it (C)} Average trajectory of the images shown in {\it B}.}
+{\it (C)} Average trajectory of the images shown in {\it B}.
+Each image is the average of $7$ consecutive images in {\it B}.
+}
 \customlabel{fig:data1}{3}
@@
 %
-Figure~\ref{subfig:raw_data2} shows a data set of images from stages 5--7 of {\it Drosophila} embryogenesis.
+Figure~\ref{subfig:raw_data2} shows a data set of $108$ images from stages 5--7 of {\it Drosophila} embryogenesis.
 %
@@
 %
-We can again construct a smoothed trajectory by looking at the moving average of the images, shown in Figure~\ref{sub fig:average_data2}.
+We can again construct a smoothed trajectory by looking at the moving average of the images, shown in Figure~\ref{subfig:averaged_data2}.
 %
@@
 Each image is of a different embryo in a different rotational orientation.
-{\it(B)} Data from {\it A}, registred and ordered using vector diffusion maps.
-{\it (C)} Average trajectory of the images shown in {\it B}.}
-\customlabel{fig:data2}{5}
+{\it(B)} Data from {\it A}, registered and ordered using vector diffusion maps.
+{\it (C)} Average trajectory of the images shown in {\it B}.
+Each image is the average of $9$ consecutive images in {\it B}.
+}
+\customlabel{fig:data2}{4}
 \customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}

### 2014-05-14 image_analysis_paper/pnas_draft2.tex
edited paper more

@@
 %
-We demonstrate the utility of our methods using a collection of images from a study of {\it Drosophila} embryogenesis.
+We demonstrate the utility of these methods using a collection of images from a study of {\it Drosophila} embryogenesis.
 \end{abstract}
@@
 %
-In a cross-sectional study, one embryo contributes only a single snapshot of a chemical or morphological process along its developmental trajectory, and the developmental dynamics must be reconstructed from multiple snapshots of different embryos.
+In a cross-sectional study, developmental dynamics must be reconstructed from multiple snapshots of different embryos, each of which contributes only a single snapshot of a chemical or morphological process along its developmental trajectory.
 %
-Both of these sampling schemes have their advantages and limitations, and both are extensively used by developmental biologists.
+Both of these sampling schemes have their advantages and limitations, and both are extensively used by biologists.
 %
@@
 %
-In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that help visualize a handful of cellular processes.
+In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that visualize a handful of cellular processes.
 %
@@
 %
-Recent advances in large-scale physical manipulation and imaging of embryos have produced rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different geometric orientation and developmental time point.
+Recent advances in large-scale physical manipulation and imaging of embryos produce rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different geometric orientation and developmental time point.
 %
@@
 %
-In order to recover the developmental dynamics from such image datasets, snapshots of different embryos must be spatially aligned or registered to factor out the relevant symmetries (i.e., translations and rotations), and then ordered in time.
+In order to recover the developmental dynamics from such datasets, snapshots of different embryos must be spatially aligned or registered to factor out the relevant symmetries (i.e., translations and rotations), and then ordered in time.
 %
-We show how existing dimensionality reduction algorithms can greatly accelerate -and combine- both of these tasks.
+We show how existing dimensionality reduction algorithms can greatly accelerate {\it and combine} both of these tasks.
 
@@
 %
-As an example, Figure~\ref{fig:fish} shows a caricature of fish development, combining the processes of growth and color patterning.
+As an example, Figure~\ref{fig:fish} shows a caricature of fish development which illustrates the processes of growth and patterning.
 %
@@
 %
-On the other hand, real data poses nontrivial challenges, due to the number of images, measurement noise and variability, and the subtlety of the developmental changes.
+In contrast to this simple illustration, real data poses nontrivial challenges for both registration and temporal ordering. 
+%
+In general, the landmarks needed for registration, such as the heads and fins in our example, as well as the features that can be used to order the data, such as the body length, are not known {\it a priori}.
+%
+Additional challenges arise from embryo-to-embryo variability, sample size, and measurement noise.
 
@@
 
-We will show that we can use existing dimensionality reduction techniques to register and temporally order large sets of images from developmental biology studies. 
+Here, we show how recently developed dimensionality reduction techniques can successfully address these challenges and reveal developmental trajectories from collections of unregistered and unordered imaging data.
 %
@@
 %
-In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex; here, we will show that we can recover a reasonable trajectory that is qualitatively consistent with current knowledge.
+In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex; here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
 
-In both of these data sets, each image is taken from a different embryo, fixed at a different developmental time and imaged in a different rotational orientation.
+%\begin{itemize}
+%\item Imaging data is becoming increasingly prevalent in biology research.
+%
+%\item General motivation about biology data ...
+%
+%\item Need methods to organize data, visualize, ...
+%
+%\item REDUCTION is important, reduction because of symmetry and reduction because of content
+%
+%\item There are a bunch of recent techniques- some established (like dmaps), some very fresh, like synchronization or VDMaps,  that have had their own reasons to be developed but they are as if tailor made for what we would want to do
+%\end{itemize}
+
+\section{Results and Discussion}
+
+\subsection{Vector diffusion maps for registration and temporal ordering}
+
+A typical dataset from cross-sectional studies of developmental processes is a collection of images of different embryos, each of which is fixed at a different developmental time and imaged in a different rotational orientation.
 %
@@
 
-\begin{itemize}
-\item Imaging data is becoming increasingly prevalent in biology research.
+Our approach is based on vector diffusion maps \cite{singer2012vector}, a general technique developed for data sets which contain two sources of variability:
+symmetries (such as translations and rotations of the images due to the imaging setup) which one would like to factor out, and other directions of variability (such as temporal dynamics) which one would like to uncover.
 %
-\item General motivation about biology data ...
+Vector diffusion maps combine two algorithms, angular synchronization \cite{singer2011angular} for image registration and diffusion maps \cite{coifman2005geometric} for extracting low-dimensional structure, into one computation that allows us to simultaneously register and temporally order our images.
 %
-\item Need methods to organize data, visualize, ...
+Vector diffusion maps has been applied to cryo-EM images \cite{...}, where each image is a projection of a molecule, taken from a different viewing direction and in a different rotation. 
 %
-\item REDUCTION is important, reduction because of symmetry and reduction because of content
+Here, we will use the algorithm to register images of {\it Drosophila} embryos with respect to rotations and translations to remove variability from the imaging procedure, as well as uncover the main direction of variability {\it after} removing translational and rotational symmetries.
 %
-\item There are a bunch of recent techniques- some established (like dmaps), some very fresh, like synchronization or VDMaps,  that have had their own reasons to be developed but they are as if tailor made for what we would want to do
-\end{itemize}
+We assume that, in this set of images, this main direction of variability is parameterized by time. 
+%
+Unlike the noise in the cryo-EM application, which is approximately Gaussian, the noise in our experiments is primarily from interembryo variability and does not have a known simple structure.
 
-\section{Results and Discussion}
-
-%Rank corr for data set 1 using VDM=  0.9127
-
-%Variance captured by PC 1 for data set 2 = 0.2155
-
-\subsection{Vector diffusion maps for registration and temporal ordering}
-
-We use vector diffusion maps \cite{singer2012vector} to automatically register and order our images.
+Registration via angular synchronization/vector diffusion maps uses pairwise alignment information to register the set of images in a globally consistent way.
 %
-Vector diffusion maps is a general technique, developed for data sets which contain both symmetries (e.g., translations and rotations) which one would like to factor out, and other directions of variability which one would like to uncover.
-%
-It combines two algorithms, angular synchronization \cite{singer2011angular} for registration and diffusion maps \cite{coifman2005geometric} for extracting low-dimensional structure, into one computation that allows us to simultaneously register and temporally order our images.
-%
-Vector diffusion maps has been applied to cryo-EM images \cite{...}, where each image is a (very noisy) projection of a molecule, taken from a different viewing direction and in a different rotation. 
-%
-Here, we will use it to register the images with respect to rotations and translations, as well as uncover the main direction of variability, which is parameterized by time. 
-%
-The noise in our experiments is primarily from interembryo variability, rather than white noise from a measurement instrument (as in the cryo-EM example).
-
-Registration via angular synchronization/vector diffusion maps uses pairwise alignment information to globally register the images in a consistent way.
-%
-A schematic illustration is shown in Figure~\ref{subfig:synchron1}; each image is depicted as a vector, and the goal is to align the set of vectors. 
+A schematic illustration is shown in Figure~\ref{subfig:synch1}; each image is depicted as a vector, and the goal is to align the set of vectors. 
 %
@@
 %
-This is rather simple to do, and requires no template function; however, when the images are noisy, some pairwise alignment angles may be inaccurate.
+In general, this requires no template function or image landmarks; however, when the images are noisy, some pairwise alignment angles may be inaccurate.
 %
-Using the angles between all pairs of vectors (or images), angular synchronization/vector diffusion maps then finds the set of rotation angles, one angle for each vector, that is most consistent with the pairwise measurements.
+Using the angles between all pairs of vectors (or images), angular synchronization/vector diffusion maps then finds the set of rotation angles, one angle for each vector (or image), that is most consistent with the pairwise measurements.
+%
+In this schematic, registration via angular synchronization is trivial as the pairwise measurements contain no noise. 
+%
+Angular synchronization/vector diffusion maps can successfully register sets of vectors (or images) even when many of the pairwise measurements are corrupted by noise. 
+%
 
-In general, diffusion maps \cite{coifman2005geometric} is a nonlinear dimensionality reduction technique;
-it uncovers a parameterization of data that lies on a low-dimensional (perhaps nonlinear) manifold in high-dimensional space. 
+%
+After removing the variability due to translations and rotations of the images, the remaining variability arises (mostly) due to developmental dynamics.
+% 
+These dynamics can be revealed by ordering the data along a one-dimensional manifold that parametrizes the data. 
+%
+Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique;
+this algorithm uncovers a parameterization of data that lies on a low-dimensional, perhaps nonlinear, manifold in high-dimensional space. 
 %
 This idea is illustrated in Figure~\ref{subfig:dmaps1}: the data are two-dimensional points which lie on a one-dimensional nonlinear curve. 
+%
+We use {\it local} information about the data to find a ``good'' parameterization: we want points which are close in high-dimensional space (i.e., images which look similar) to be close in our parameterization.
+%
+This idea of locality is illustrated by the edges in Figure~\ref{subfig:dmaps1}; data points which are close are connected by dark edges, and clearly, the dark edges are more informative about the low-dimensional structure of the data.
 %
@@
 %
-We will assume that our data are (approximately) one-dimensional, and that this dimension is parameterized by time.
+We will assume that our data are approximately one-dimensional, and that this dimension is parameterized by time, such that ordering our data along this main dimension/direction will temporally order our data.
+
+Both angular synchronization and diffusion maps use pairwise information to extract a global structure in the data.
 %
-Therefore, ordering our data along this main dimension/direction will temporally order our data. 
+Angular synchronization uses pairwise alignment information to find translations and rotations which are globally consistent, and diffusion maps uses pairwise distances to find parameterizations which are globally informative.
 %
-Vector diffusion maps combines angular synchronization and diffusion maps to find a parameterization of the data while simultaneously registering data points which are close.
+Vector diffusion maps combines both steps into one computation;
+the algorithm uses information about pairwise alignments and pairwise distances to find a global parameterization of the data, while registering images which are ``close''.
+%
+The algorithm is ``tailor-made'' for our task of registering and temporally ordering cross-sectional imaging data; we will use it to uncover a temporal ordering of our images, as well as register images which are developmentally similar, to obtain a global picture of development.
 
@@
 
-\subsection{Stage 5 images in {\it Drosophila} embryogenesis}
+\subsection{Images of {\it Drosophila} embryos during cellularization}
 
-Our first data set, shown in Figure~\ref{subfig:raw_data1}, consists of $49$ images from stage 5 of {\em Drosophila} embryogenesis.
 %
-Each image is from a different embryo imaged in a different orientation and fixed at a different developmental time. 
+As a first illustration of this approach to temporal ordering and registration of images, we used a dataset where the correct registration and temporal order is known. 
 %
-We registered and ordered the images using vector diffusion maps.
+This dataset is obtained by fluorescent imaging of {\it Drosophila} embryos during the third hour after fertilization.
 %
-The results are shown in Figure~\ref{subfig:ordered_data1}.
+By this stage of development, sequential divisions of the fertilized egg have generated a system where most of the nuclei are arranged in a monolayer under the common plasma membrane.
 %
-The features that our eyes can visually detect (the green Dorsal peak, and the red dpERK peaks) are now consistently registered, even though the algorithm does not require any information about the existence or location of these peaks.
+Figure 3a shows optical cross-sections of each embryo, with nuclei (gray) labelled by a DNA stain. 
+%
+The spatially uniform arrangement of nuclei is patterned by the concentration profiles of chemical signals that act as dose-dependent regulators of cell differentiation.
+%
+One of these signals is provided by the nuclear localization of transcription factor Dorsal (Dl, green), which subdivides the embryo into the territories that give rise to the muscle, nerve, and skin tissues.
+%
+Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK), an enzyme that specifies a subset of neuronal cells. 
+%
 
-When dealing with large sets of images with interembryo variability, we can gain a better understanding of the developmental dynamics by looking at an averaged trajectory.
+Previous studies \cite{...} established that throughout the 3rd hour of development, the nuclear localization of Dl is unimodal and peaked at the ventral side of the embryo. 
 %
-Figure~\ref{subfig:average_data1} shows the moving average of the trajectory in Figure~\ref{subfig:ordered_data1}. 
+The spatiotemporal pattern of dpERK is more complex: dpERK is first detected in two lateral stripes, which increase in intensity towards the end of the 3rd hour. 
 %
-We can now easily see the developmental dynamics: the red dpERK peaks grow in time, and the ventral furrow (the dent in green) begins to form at the end of stage 5.
+This bimodal pattern is then augmented by a peak at the dorsal side of the embryo.
+%
+Both of these features become readily apparent when images are registered and ordered using vector diffusion maps: the peak of nuclar Dl is consistent throughout the images and intensity of the dpERK signal in the lateral regions of the embryo monotonically increases across the dataset.
+%
+The quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the temporal ordering provided by an independent approach.
+%
+Specifically, during this developmental timespan, the embryo cellularizes: nuclei located under common plasma membrane are enclosed by lateral membranes which grow inward, separating nuclei into individual cells.
+%
+The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo within 1-3 minutes  \cite{...} (see {\it SI Appendix}).    
+%
+The rank correlation coefficient between the orderings provided by membranes and vector diffusion maps is 0.9127. 
+%
+Thus, we have validated a dimensionality reduction approach to temporal ordering and registration of data  on a dataset where image orientation and ordering can be assessed independently.
 
-For this data set, we can validate our proposed vector diffusion maps ordering of the data.
+%Furthermore, the dynamics uncovered by vector diffusion maps is relatively simple, which suggests that a simpler approach could have been used to order the data.
 %
-During stage 5, we can measure the developmental time of each embryo based on the monotonic progress of cellularization (details are given in the {\it SI Appendix}).
-%
-We compared the temporal ordering obtained from cellularization with the ordering we extract from vector diffusion maps. 
-%
-The rank correlation coefficient between these two orderings is 0.9127, indicating that there is very good agreement and vector diffusion maps is uncovering the correct temporal ordering of the images.
+%Indeed, a principal component analysis (PCA) of registered data reveals that data can be accurately ordered by projecting each image onto the first PCA mode.
 
@@
 \includegraphics[width=8.4cm]{VDM_data1_ordered}\\
-\vspace{0.2cm}
-\centering
-\raisebox{0.5cm}{{\figtextfont C}}
-\includegraphics[width=8.4cm]{average_trajectory_VDM}
-\caption{{\it (A)} Images from stage 5 of {\it Drosophila} embryogenesis.
+%\vspace{0.2cm}
+%\centering
+%\raisebox{0.5cm}{{\figtextfont C}}
+%\includegraphics[width=8.4cm]{average_trajectory_VDM}
+\caption{{\it (A)} Images of {\it Drosophila} embryos during cellularization.
 Each image is of a different embryo in a different rotational orientation.
 {\it (B)} Images from {\it A}, now registered and ordered using vector diffusion maps.
-{\it (C)} Average trajectory of the images shown in {\it B}.
+}%{\it (C)} Average trajectory of the images shown in {\it B}.
 Each image is the average of $7$ consecutive images in {\it B}.
@@
 
-\subsection{Stage 5--7 Images of {\it Drosophila} embryogenesis}
+\subsection{Images of gastrulating {\it Drosophila} embryos}
 
-The images in Figure~\ref{fig:data1} are relatively simple, and the methods presented here are perhaps unnecessarily complex for this particular data set. 
+Because the images in Figure~\ref{fig:data1} are relatively simple, simpler techniques could have been used to temporally order this particular data set. 
 %
-Because these data are effectively a single mode growing in time,
+Because these data are effectively a single mode (the two lateral dpERK peaks) growing in time,
 principal component analysis \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data.
@@
 %
-Figure~\ref{subfig:raw_data2} shows a data set of $108$ images from stages 5--7 of {\it Drosophila} embryogenesis.
+Figure~\ref{subfig:raw_data2} shows a data set of $108$ images of {\it Drosophila} embryos;
+these images cover a thirty minute time interval and span late cellularization through gastrulation, when tissues within the embryo begin to morph and deform. 
 %
-Only 22\% of the variability in the data is captured by the first principal component, and so we are forced to use more sophisticated techniques, such as vector diffusion maps, to temporally order the data.
+Each image shows the optical cross-section of an embryo fixed at a different developmental time.
+%
+The nuclei (gray) are again labeled with a DNA marker.
+%
+However, during this developmental time period, and the nuclei are no longer uniformly arranged around the periphery of the embryo, as the tissue begins to deform and morph.
+%
+Twist (Twi, shown in green) denotes the ventralmost point of the embryo (similar to Dl in the previous data set); during this developmental time span, the ventral furrow is formed, where the ventral side invaginates towards the center of the embryo.
+%
+dpERK (red) again appears as two lateral peaks at the ventral side of the embryo.
+%
+These two peaks merge together during invagination, eventually forming (together with Twi) a characteristic ``omega'' shape.
+%
+During this developmental timespan, germ band elongation causes cells from the ventral side of the embryo to move towards the poles, and then wrap around to the dorsal side of the embryo.
+%
+At the end of this developmental timespan, cells which were originally on the ventral side of the embryo have migrated to the dorsal side of the embryo, causing similar ``omega'' shaped patterns (formed by Twi and dpERK) to arise on the dorsal side.
+%
+Although we know about this general developmental progression, unlike the previous data set, we have no independent way to measure the precise developmental time of each embryo.
+%
+We are forced to use other methods to temporally order these images.
 
+The dynamics in this data set are highly nonlinear, and only 22\% of the variability in the data is captured by the first principal component.
+%
+We require a more sophisticated approach, such as vector diffusion maps, to temporally order the data.
+%
 Figure~\ref{subfig:ordered_data2} shows the images in Figure~\ref{subfig:raw_data2}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
 %
-Again, visually detectable features, such as the green Twist peaks, are consistently aligned. 
+The ordering is consistent with the known developmental dynamics outlined above.
+
+However, because each image comes from a {\it different} embryo, the recovered trajectory is still noisy due to interembryo variability.
 %
-Unlike the images in Figure~\ref{subfig:raw_data1}, 
-for Figure~\ref{fig:data2}, we have no time marker with which to validate our orderings.
+We would like to extract an average developmental trajectory from this set of registered and temporally ordered images to obtain a representative picture of the dynamics.
 %
-However, there are noticeable features in the ordering that we can see are consistent with previously known dynamics.
+A simple method for doing this is dividing the ordered data set into groups or stages, and then computing the average image at each stage. 
 %
-We can again construct a smoothed trajectory by looking at the moving average of the images, shown in Figure~\ref{subfig:averaged_data2}.
-%
-We can see the formation of the ventral furrow and germ band elongation, which concludes with ventral cells shifted to the dorsal side of the embryo, which is consistent with the known picture of {\it Drosophila} embryogeneisis.
-%
-This assures us that we can obtained an ordering of the data which is developmentally meaningful.
+Figure~\ref{sub fig:averaged_data2} shows the average trajectory for our data set, and it is now very straightforward to see the underlying developmental dynamics.
 
@@
 \includegraphics[width=16.8cm]{average_trajectory}
-\caption{{\it (A)} Images from stages 5-7 of {\it Drosophila} embryogenesis.
+\caption{{\it (A)} Images of gastrulating {\it Drosophila} embryos.
 Each image is of a different embryo in a different rotational orientation.
@@
 
-We have demonstrated how existing reduction techniques (vector diffusion maps) can be used to register and temporally order images from developmental biology studies. 
+We have demonstrated how vector diffusion maps, an existing reduction technique, can be used to register and temporally order images from developmental biology studies. 
 %
@@
 %
-We chose the techniques outlined here, not because they are particularly suited to this specific problem, but rather, because they are sufficiently general and we are confident that they can be applied to a myriad of biological imaging applications.
+We chose the techniques outlined here, not because they are specifically tailored to our particular data sets, but rather, because they are sufficiently general and we are confident that they can be applied to a myriad of biological imaging applications.
 
-The task of image registration has been studied in a variety of contexts \cite{...}.
+The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
 %
@@
 %
-For our biological applications, good features may not be known {\it a priori}, 
-and we instead use an algorithm which requires no features, but instead, uses the images directly.
+For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then register images by aligning these features \cite{zhao2003face}.
+%
+%Temporally ordering images of a construction site over time is also based on first identifying features within the images \cite{schindler2007inferring}.
+%
+For our biological applications, good features may not be known {\it a priori}.
+%
+We instead use an algorithm which requires no features, but instead, uses only the raw pixel images and the inherent geometric symmetry of the problem.
+%
+%Template-based methods are also common for registration, where one optimally aligns each image to a predefined template function.
+%
+%However, these methods can suffer when the images are noisy, and the success of such methods can depend strongly on the template function \cite{shatsky2009method}.
+%
 
-The task of temporal ordering has also been studied, for data such as ...
+The task of temporal ordering has also been studied in a variety of biological applications. 
+%
+Data sets such as  RNA sequencing data \cite{anavy2014blind, trapnell2014dynamics} (where each data point is the transcriptome for a single sample), gene expression data from multiple patients \cite{gupta2008extracting, qiu2011discovering} (where each data point is a vector containing the expression levels of the various genes), and snapshots of cells throughout the cell cycle (where each point is a vector of features, e.g., amount of DNA which quantify the cell's state)  \cite{kafri2013dynamics} have all been temporally ordered to extract the trajectory of the underlying dynamic process.
 %
@@
 %
-We, instead, use diffusion maps to construct a one-dimensional parameterization of the data.
+Furthermore, many of these algorithms use PCA (a dimensionality reduction technique) as an initial preprocessing/denoising step. 
 %
-Diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003, tenenbaum2000global, Donoho2003, Roweis2000}.
+We, instead, use vector diffusion maps to construct a one-dimensional parameterization of the data.
+%
+Because vector diffusion maps is a dimensionality reduction technique, it acts as a denoising step (as indicated by the smoothness of our averaged trajectories).
+%
+However, because it is nonlinear, it affords us many of the same geometric flexibilies as the MST or TSP formulations. 
+%
+Vector diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003, tenenbaum2000global, Donoho2003, Roweis2000}.
 %
 They have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}, and so we suspect they may perform better than the ordering algorithms used in previous work.
+
+Furthermore, dimensionality reduction techniques such as vector diffusion maps do not constrain us to one-dimensional parameterizations/organizations of the data.
 %
-Furthermore, techniques such as diffusion maps do not constrain us to one-dimensional parameterizations/organizations of the data.
+For this particular application, we expect our data to be one-dimensional, as the only (main) source of variability is time.
 %
-In this particular setup, we expect our data to be one-dimensional, as the only (main) source of variability is time.
+However, other sources of variability, such as size, viewing angle, and mutation state, could also be uncovered using vector diffusion maps.
 %
-However, other sources of variability would also be uncovered using (vector) diffusion maps, 
-and tasks such as classification of embryos by size or viewing angle, and comparison of different mutations can also be accomplished using the same techniques.
+We could therefore use these same algorithms for tasks such as classification of embryos by size or viewing angle, and comparison of the developmental dynamics in different mutations.
 
-\begin{itemize}
-\item What are other people doing  (maybe critically)
-\item Movies  smoothness
-\item Discrete symmetries like flips
-\item Local information ? (features of the image) (associate with Fourier-Bessel and bispectra)
-\item Mutant-wildtype stuff  IMPORTANT
-\item And in general, MULTIPLE variablities is important (mutants, guys in another cycle, etc.)
+After registering and temporally ordering images, one can perform a variety of post-processing tasks to gain further insight into the developmental processes.
+%
+We have demonstrated one post-processing technique (window averaging) that allows us to extract a more representative developmental trajectory from the set of noisy snapshots. 
+%
+However, tasks such as constructing a smooth movie from these cross-sectional snapshots would also be of biological interest.
+%
+One could look not only at the window averages, but also at the top principal components in each window to see the main dynamical features along the trajectory.
+%
+All of these questions first require that the data be registered and temporally ordered.
+%
+We are confident that, as imaging becomes easier and more robust, the techniques we present here will play an important role in the data analysis pipeline.
 
-\item size is an issue
-\item  small tilts is an issue
-\item   could one do three-d image processing ?  of the floating thing, without the device
-\item how do we go about merging datasets ?
-\end{itemize}
+%\begin{itemize}
+%\item What are other people doing  (maybe critically)
+%\item Movies  smoothness
+%\item Discrete symmetries like flips
+%\item Local information ? (features of the image) (associate with Fourier-Bessel and bispectra)
+%\item Mutant-wildtype stuff  IMPORTANT
+%\item And in general, MULTIPLE variablities is important (mutants, guys in another cycle, etc.)
+
+%\item size is an issue
+%\item  small tilts is an issue
+%\item   could one do three-d image processing ?  of the floating thing, without the device
+%\item how do we go about merging datasets ?
+%\end{itemize}
 

### 2014-05-14 image_analysis_paper/pnas_draft2.tex
edited draft more

@@
 %
-In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex; here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
+In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex.
+%
+Here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
 
@@
 %
-We will show that dimensionality reduction algorithms can accomplish both of these tasks and therefore give us a meaningful picture of the underlying developmental dynamics.
+We will show that dimensionality reduction algorithms can accomplish both of these tasks to give us a meaningful picture of the underlying developmental dynamics.
 
 Our approach is based on vector diffusion maps \cite{singer2012vector}, a general technique developed for data sets which contain two sources of variability:
-symmetries (such as translations and rotations of the images due to the imaging setup) which one would like to factor out, and other directions of variability (such as temporal dynamics) which one would like to uncover.
+symmetries (such as translations and rotations of the images due to the experimental setup) which one would like to factor out, and other directions of variability (such as temporal dynamics) which one would like to uncover.
 %
@@
 %
-Vector diffusion maps has been applied to cryo-EM images \cite{...}, where each image is a projection of a molecule, taken from a different viewing direction and in a different rotation. 
+Vector diffusion maps has been used to classify images from cryo-electron microscopy (cryo-EM) experiments \cite{...}, where each image is a projection of a molecule, taken from a different viewing direction and in a different rotation.
 %
-Here, we will use the algorithm to register images of {\it Drosophila} embryos with respect to rotations and translations to remove variability from the imaging procedure, as well as uncover the main direction of variability {\it after} removing translational and rotational symmetries.
+Here, we will use the algorithm to register images of {\it Drosophila} embryos with respect to rotations and translations, as well as uncover the main direction of variability {\it after} removing translational and rotational symmetries.
 %
-We assume that, in this set of images, this main direction of variability is parameterized by time. 
+We assume that, in this set of images, this main direction of variability is parameterized by time, and so uncovering this direction will allow us to recover the developmental dynamics. 
 %
@@
 
-Registration via angular synchronization/vector diffusion maps uses pairwise alignment information to register the set of images in a globally consistent way.
+Registration via angular synchronization/vector diffusion maps uses pairwise alignment information to register a set of images in a globally consistent way.
 %
-A schematic illustration is shown in Figure~\ref{subfig:synch1}; each image is depicted as a vector, and the goal is to align the set of vectors. 
+A schematic illustration is shown in Figure~\ref{subfig:synch1}, where each image is depicted as a vector, and the goal is to align the set of vectors. 
 %
@@
 %
-In general, this requires no template function or image landmarks; however, when the images are noisy, some pairwise alignment angles may be inaccurate.
+In general, this requires no template function or image landmarks; however, when the images are noisy, some pairwise alignments may be inaccurate.
 %
-Using the angles between all pairs of vectors (or images), angular synchronization/vector diffusion maps then finds the set of rotation angles, one angle for each vector (or image), that is most consistent with the pairwise measurements.
+Using the alignment angles between all pairs of vectors (or images), angular synchronization then finds the set of rotation angles, one angle for each vector (or image), that is most consistent with the pairwise measurements.
 %
@@
 %
-Angular synchronization/vector diffusion maps can successfully register sets of vectors (or images) even when many of the pairwise measurements are corrupted by noise. 
+However, angular synchronization can successfully register sets of vectors (or images) even when many of the pairwise measurements are corrupted by noise. 
 %
@@
 %
-After removing the variability due to translations and rotations of the images, the remaining variability arises (mostly) due to developmental dynamics.
+After removing the variability due to translations and rotations of the images, the remaining variability is (mostly) due to developmental dynamics.
 % 
-These dynamics can be revealed by ordering the data along a one-dimensional manifold that parametrizes the data. 
+These dynamics can be revealed by ordering the data along the one-dimensional manifold that parametrizes the data. 
 %
-Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique;
-this algorithm uncovers a parameterization of data that lies on a low-dimensional, perhaps nonlinear, manifold in high-dimensional space. 
+Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique.
 %
-This idea is illustrated in Figure~\ref{subfig:dmaps1}: the data are two-dimensional points which lie on a one-dimensional nonlinear curve. 
+This algorithm uncovers a parameterization of data that lies on a low-dimensional, perhaps nonlinear, manifold in high-dimensional space. 
 %
-We use {\it local} information about the data to find a ``good'' parameterization: we want points which are close in high-dimensional space (i.e., images which look similar) to be close in our parameterization.
+This idea is illustrated in Figure~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve. 
+%
+We use {\it local} information about the data to find a ``good'' parameterization: we want points which are close in high-dimensional space (e.g., images which look similar) to be close in our parameterization.
 %
@@
 %
-We will assume that our data are approximately one-dimensional, and that this dimension is parameterized by time, such that ordering our data along this main dimension/direction will temporally order our data.
+%We will assume that our data are approximately one-dimensional, and that this dimension is parameterized by time, such that ordering our data along this main dimension/direction will temporally order our data.
 
-Both angular synchronization and diffusion maps use pairwise information to extract a global structure in the data.
+Both angular synchronization and diffusion maps use pairwise information to uncover global structure in the data.
 %
@@
 %
-The ordering is consistent with the known developmental dynamics outlined above.
+Visual inspection reveals that ordering is consistent with the known developmental dynamics outlined above.
+%
+However, the dynamics are not immediately apparent from this set of images for two reasons.
+%
+Firstly, the sheer number of images makes visual assimilation of the entire data set (even after registration and ordering) difficult.
+%
+Secondly, the ordered data is not entirely smooth due to noise from interembryo variability.
+%
+To highlight the developmental dynamics that we uncover using vector diffusion maps, we construct an average developmental trajectory.
+%
+We do this simply by dividing the registered and ordered data into windows or classes (each window contains consecutive images in the ordered data), and then calculating the average image in each window.
+%
+Figure~\ref{subfig:averaged_data2} shows the average trajectory for our data set; it is now very straightforward to see the underlying developmental dynamics.
 
-However, because each image comes from a {\it different} embryo, the recovered trajectory is still noisy due to interembryo variability.
-%
-We would like to extract an average developmental trajectory from this set of registered and temporally ordered images to obtain a representative picture of the dynamics.
-%
-A simple method for doing this is dividing the ordered data set into groups or stages, and then computing the average image at each stage. 
-%
-Figure~\ref{sub fig:averaged_data2} shows the average trajectory for our data set, and it is now very straightforward to see the underlying developmental dynamics.
-
-\newpage
+%\newpage
 \begin{figure*}[t]
@@
 %
-We acknowledge that the tasks of both registration and ordering can be accomplished using a variety of techniques.
+We acknowledge that the tasks of both registration and ordering could have been accomplished using a variety of techniques.
 %
@@
 %
-For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then register images by aligning these features \cite{zhao2003face}.
+For example, in face recognition, the first step is often to locate the eyes, nose, and mouth in each face, and then register images by aligning these features \cite{zhao2003face}.
 %
@@
 %
-We instead use an algorithm which requires no features, but instead, uses only the raw pixel images and the inherent geometric symmetry of the problem.
+We therefore use an algorithm which requires no features, but instead, uses only the raw pixel images and the inherent geometric symmetry of the problem.
 %
@@
 %
-Data sets such as  RNA sequencing data \cite{anavy2014blind, trapnell2014dynamics} (where each data point is the transcriptome for a single sample), gene expression data from multiple patients \cite{gupta2008extracting, qiu2011discovering} (where each data point is a vector containing the expression levels of the various genes), and snapshots of cells throughout the cell cycle (where each point is a vector of features, e.g., amount of DNA which quantify the cell's state)  \cite{kafri2013dynamics} have all been temporally ordered to extract the trajectory of the underlying dynamic process.
+Data sets such as  RNA sequencing data (where each data point is the transcriptome for a single sample) \cite{anavy2014blind, trapnell2014dynamics}, gene expression data from multiple patients (where each data point is a vector containing the expression levels of the various genes) \cite{gupta2008extracting, qiu2011discovering}, and snapshots of cells throughout the cell cycle (where each point is a vector of features, such as the amount of DNA, which quantify the cell's state)  \cite{kafri2013dynamics} have all been temporally ordered to extract a representative trajectory of the underlying dynamic process.
 %
@@
 %
-Furthermore, many of these algorithms use PCA (a dimensionality reduction technique) as an initial preprocessing/denoising step. 
+These algorithms often use dimensionality reduction techniques (typically PCA or variants of) as a preprocessing step; the TSP or MST is solved/constructed not in the original data space, but rather in the lower-dimensional space of the first few principal components.
 %
+In these approaches, PCA serves two purposes: it denoises the data, making the TSP/MST construction more robust, and allows for visualization of the TSP/MST if one only uses the first two or three principal components.
+
 We, instead, use vector diffusion maps to construct a one-dimensional parameterization of the data.
 %
-Because vector diffusion maps is a dimensionality reduction technique, it acts as a denoising step (as indicated by the smoothness of our averaged trajectories).
+Because vector diffusion maps is a dimensionality reduction technique, it serves as a denoising step and yields smooth averaged trajectories even under noisy conditions. 
 %
-However, because it is nonlinear, it affords us many of the same geometric flexibilies as the MST or TSP formulations. 
+However, because the algorithm is nonlinear, it affords us many of the same geometric flexibilies as the MST or TSP formulations. 
 %
@@
 %
-We have demonstrated one post-processing technique (window averaging) that allows us to extract a more representative developmental trajectory from the set of noisy snapshots. 
+We have demonstrated one such technique (window averaging) that allows us to extract a more representative developmental trajectory from the set of noisy snapshots. 
 %
-However, tasks such as constructing a smooth movie from these cross-sectional snapshots would also be of biological interest.
+However, constructing a smooth movie from the registered and ordered cross-sectional snapshots would also be of biological interest.
 %
-One could look not only at the window averages, but also at the top principal components in each window to see the main dynamical features along the trajectory.
+Examining the top principal components in each window would allow one to visualize the main dynamical features along the trajectory.
 %

### 2014-05-14 image_analysis_paper/pnas_draft2.tex
changes to draft

@@
 %
-We assume that, in this set of images, this main direction of variability is parameterized by time, and so uncovering this direction will allow us to recover the developmental dynamics. 
+We assume that, in this set of images, this main direction of variability is parameterized by the developmental time of each embryo, and so uncovering this direction will allow us to recover the developmental dynamics.
 %
@@
 
-Registration via angular synchronization/vector diffusion maps uses pairwise alignment information to register a set of images in a globally consistent way.
+Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
 %
@@
 %
-This idea is illustrated in Figure~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve. 
+The idea is illustrated in Figure~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve. 
 %
-We use {\it local} information about the data to find a ``good'' parameterization: we want points which are close in high-dimensional space (e.g., images which look similar) to be close in our parameterization.
+We use {\it local} information about the data to find a ``good'' parameterization which respects the underlying manifold geometry: we want points which are close in high-dimensional space (e.g., images which look similar) to be close in our parameterization.
 %
@@
 %
-In our examples, each data point will be very high dimensional (e.g, an image with many pixels), and so we cannot extract this low-dimensional structure visually.
+However, in our examples, each data point will be very high dimensional (e.g, an image with many pixels), and so we cannot extract this low-dimensional structure visually.
 %
@@
 %
-The algorithm is ``tailor-made'' for our task of registering and temporally ordering cross-sectional imaging data; we will use it to uncover a temporal ordering of our images, as well as register images which are developmentally similar, to obtain a global picture of development.
+The algorithm is perfectly suited for our task of registering and temporally ordering cross-sectional imaging data; we will use it to uncover a temporal ordering of our images, as well as register images which are developmentally similar, to obtain a global picture of development.
 
@@
 
-Because the images in Figure~\ref{fig:data1} are relatively simple, simpler techniques could have been used to temporally order this particular data set. 
+Because the images in Figure~\ref{fig:data1} are relatively simple, other techniques could have been used to temporally order them. 
 %
-Because these data are effectively a single mode (the two lateral dpERK peaks) growing in time,
-principal component analysis \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data.
+Because these data are effectively a single mode (the two dpERK peaks) growing in time,
+principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data.
 %
@@
 %
-Figure~\ref{subfig:raw_data2} shows a data set of $108$ images of {\it Drosophila} embryos;
-these images cover a thirty minute time interval and span late cellularization through gastrulation, when tissues within the embryo begin to morph and deform. 
+Figure~\ref{subfig:raw_data2} shows a data set of $108$ images of {\it Drosophila} embryos.
+%
+These images cover a thirty minute time interval and span late cellularization (the early images in this data set overlap with the late images in the previous set) through gastrulation, when tissues within the embryo begin to morph and deform.
 %
@@
 %
-However, during this developmental time period, and the nuclei are no longer uniformly arranged around the periphery of the embryo, as the tissue begins to deform and morph.
+However, during this developmental time period, the nuclei are no longer uniformly arranged around the periphery of the embryo as the tissue begins to morph.
 %
-Twist (Twi, shown in green) denotes the ventralmost point of the embryo (similar to Dl in the previous data set); during this developmental time span, the ventral furrow is formed, where the ventral side invaginates towards the center of the embryo.
+Twist (Twi, shown in green) denotes the ventralmost point of the embryo (similar to Dl in the previous data set).
+%
+During this developmental time span, the ventral furrow is formed, where the ventral side invaginates towards the center of the embryo.
 %
@@
 %
-During this developmental timespan, germ band elongation causes cells from the ventral side of the embryo to move towards the poles, and then wrap around to the dorsal side of the embryo.
+During this developmental timespan, germ band elongation causes cells from the ventral side of the embryo to move towards the poles of the embryo, and then wrap around to the dorsal side.
 %
@@
 %
-Although we know about this general developmental progression, unlike the previous data set, we have no independent way to measure the precise developmental time of each embryo.
-%
-We are forced to use other methods to temporally order these images.
+Although we know about this general developmental progression, unlike the previous data set, we have no method to measure the precise developmental time of each embryo, 
+and we are forced to use other methods to temporally order these images.
 
-The dynamics in this data set are highly nonlinear, and only 22\% of the variability in the data is captured by the first principal component.
+The dynamics in this data set are highly nonlinear, as only 22\% of the variability in the data is captured by the first principal component.
 %
@@
 %
-Visual inspection reveals that ordering is consistent with the known developmental dynamics outlined above.
+Visual inspection verifies that this ordering is consistent with the known developmental dynamics outlined above.
 %
-However, the dynamics are not immediately apparent from this set of images for two reasons.
+However, the dynamics are not immediately apparent from this set of registered and ordered images for two reasons.
 %
-Firstly, the sheer number of images makes visual assimilation of the entire data set (even after registration and ordering) difficult.
-%
-Secondly, the ordered data is not entirely smooth due to noise from interembryo variability.
+First, the sheer number of images makes visual assimilation of the entire data set (even after registration and ordering) difficult, and
+second, the ordered data is not entirely smooth due to noise from interembryo variability.
 %
@@
 %
-We do this simply by dividing the registered and ordered data into windows or classes (each window contains consecutive images in the ordered data), and then calculating the average image in each window.
+We divide the registered and ordered data into windows or classes (each window contains a group of consecutive images in the ordered data), and then calculate the average image in each window.
 %
@@
 %
-We acknowledge that the tasks of both registration and ordering could have been accomplished using a variety of techniques.
+We acknowledge that the tasks of both registration and ordering could have been accomplished using a variety of other techniques.
 %
-We chose the techniques outlined here, not because they are specifically tailored to our particular data sets, but rather, because they are sufficiently general and we are confident that they can be applied to a myriad of biological imaging applications.
+We chose the algorithm outlined here, not because it is specifically tailored to our particular data sets, but rather, because it is sufficiently general and we are confident that they can be applied to a myriad of biological imaging applications.
 
@@
 %
-Most of these applications order the data by solving a traveling salesman problem or constructing a minimum spanning tree on the data,
+Most of these applications order the data by solving a traveling salesman problem (TSP) or constructing a minimum spanning tree (MST) on the data,
 expecting that these structures (the path of the traveling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate encoding of the dynamical progression.
 %
-These algorithms often use dimensionality reduction techniques (typically PCA or variants of) as a preprocessing step; the TSP or MST is solved/constructed not in the original data space, but rather in the lower-dimensional space of the first few principal components.
+These algorithms often use dimensionality reduction techniques (typically PCA or variants of) as a preprocessing step.
 %
-In these approaches, PCA serves two purposes: it denoises the data, making the TSP/MST construction more robust, and allows for visualization of the TSP/MST if one only uses the first two or three principal components.
+The TSP or MST is solved/constructed not in the original data space, but rather in the lower-dimensional space of the first few principal components.
+%
+In these approaches, PCA serves two purposes: it denoises the data, making the TSP/MST construction more robust, and allows for easy visualization of the TSP/MST if one only uses the first two or three principal components.
 
@@
 %
-They have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}, and so we suspect they may perform better than the ordering algorithms used in previous work.
+Diffusion maps have been shown to be more robust to noise than isomap \cite{balasubramanian2002isomap}, a path-based algorithm, and so we suspect the methods presented here may perform better than the ordering algorithms used in previous work.
 
@@
 %
-However, constructing a smooth movie from the registered and ordered cross-sectional snapshots would also be of biological interest.
+Constructing a smooth movie from the registered and ordered cross-sectional snapshots would also be of biological interest.
 %
-Examining the top principal components in each window would allow one to visualize the main dynamical features along the trajectory.
+Furthermore, examining the top principal components in each window would allow one to visualize the main dynamical features along the trajectory.
 %
@@
 %
-We are confident that, as imaging becomes easier and more robust, the techniques we present here will play an important role in the data analysis pipeline.
+We are confident that, as biological becomes easier and more robust, the techniques we present here will play an important role in the data analysis pipeline.
 

### 2014-05-14 image_analysis_paper/si_information.tex
started supplemental information for paper

--- 2014-04-16 image_analysis_paper/pnas_draft.tex
@@
-%% PNAStmpl.tex
-%% Template file to use for PNAS articles prepared in LaTeX
-%% Version: Apr 14, 2008
+\documentclass[12pt]{article}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% BASIC CLASS FILE
-%% PNAStwo for two column articles is called by default.
-%% Uncomment PNASone for single column articles. One column class
-%% and style files are available upon request from pnas@nas.edu.
-%% (uncomment means get rid of the '%' in front of the command)
+\usepackage[margin=1in]{geometry}
 
-%\documentclass{pnasone}
-\documentclass{pnastwo}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Changing position of text on physical page:
-%% Since not all printers position
-%% the printed page in the same place on the physical page,
-%% you can change the position yourself here, if you need to:
-
-% \advance\voffset -.5in % Minus dimension will raise the printed page on the
-                         %  physical page; positive dimension will lower it.
-
-%% You may set the dimension to the size that you need.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% OPTIONAL GRAPHICS STYLE FILE
-
-%% Requires graphics style file (graphicx.sty), used for inserting
-%% .eps files into LaTeX articles.
-%% Note that inclusion of .eps files is for your reference only;
-%% when submitting to PNAS please submit figures separately.
-
-%% Type into the square brackets the name of the driver program
-%% that you are using. If you don't know, try dvips, which is the
-%% most common PC driver, or textures for the Mac. These are the options:
-
-% [dvips], [xdvi], [dvipdf], [dvipdfm], [dvipdfmx], [pdftex], [dvipsone],
-% [dviwindo], [emtex], [dviwin], [pctexps], [pctexwin], [pctexhp], [pctex32],
-% [truetex], [tcidvi], [vtex], [oztex], [textures], [xetex]
-
-%\usepackage[dvips]{graphicx}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% OPTIONAL POSTSCRIPT FONT FILES
-
-%% PostScript font files: You may need to edit the PNASoneF.sty
-%% or PNAStwoF.sty file to make the font names match those on your system.
-%% Alternatively, you can leave the font style file commands commented out
-%% and typeset your article using the default Computer Modern
-%% fonts (recommended). If accepted, your article will be typeset
-%% at PNAS using PostScript fonts.
-
-% Choose PNASoneF for one column; PNAStwoF for two column:
-%\usepackage{PNASoneF}
-\usepackage{PNAStwoF}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% ADDITIONAL OPTIONAL STYLE FILES
-
-%% The AMS math files are commonly used to gain access to useful features
-%% like extended math fonts and math commands.
+\usepackage{graphicx}
+\graphicspath{ {paper_figures2/} }
 
 \usepackage{amssymb,amsfonts,amsmath}
-
-%\usepackage{subcaption}
-\graphicspath{ {paper_figures/} }
-
-%\usepackage{sansmath}
-
-%\usepackage[font={sf, small}]{caption}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% OPTIONAL MACRO FILES
-%% Insert self-defined macros here.
-%% \newcommand definitions are recommended; \def definitions are supported
-
-%\newcommand{\mfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2}}
-%\def\s{\sigma}
-
-\DeclareMathSizes{9}{8}{7}{7}
-
 \DeclareMathOperator*{\argmin}{arg\,min}
-
-\makeatletter
-\newcommand{\customlabel}[2]{%
-\protected@write \@auxout {}{\string \newlabel {#1}{{#2}{}}}}
-\makeatother
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Don't type in anything in the following section:
-%%%%%%%%%%%%
-%% For PNAS Only:
-\contributor{Submitted to Proceedings
-of the National Academy of Sciences of the United States of America}
-\url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
-\copyrightyear{2008}
-\issuedate{Issue Date}
-\volume{Volume}
-\issuenumber{Issue Number}
-%%%%%%%%%%%%
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Experimental Methods}
 
-%% For titles, only capitalize the first letter
-%% \title{Almost sharp fronts for the surface quasi-geostrophic equation}
+\subsection{Measuring developmental time during cellularization}
 
-\title{Temporal ordering and registration of cross-sectional imaging data}
-
-%% Enter authors via the \author command.
-%% Use \affil to define affiliations.
-%% (Leave no spaces between author name and \affil command)
-
-%% Note that the \thanks{} command has been disabled in favor of
-%% a generic, reserved space for PNAS publication footnotes.
-
-%% \author{<author name>
-%% \affil{<number>}{<Institution>}} One number for each institution.
-%% The same number should be used for authors that
-%% are affiliated with the same institution, after the first time
-%% only the number is needed, ie, \affil{number}{text}, \affil{number}{}
-%% Then, before last author ...
-%% \and
-%% \author{<author name>
-%% \affil{<number>}{}}
-
-%% For example, assuming Garcia and Sonnery are both affiliated with
-%% Universidad de Murcia:
-%% \author{Roberta Graff\affil{1}{University of Cambridge, Cambridge,
-%% United Kingdom},
-%% Javier de Ruiz Garcia\affil{2}{Universidad de Murcia, Bioquimica y Biologia
-%% Molecular, Murcia, Spain}, \and Franklin Sonnery\affil{2}{}}
-
-\author{Carmeline~J.~Dsilva\affil{1}{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
-Bomyi~Lim\affil{1}{},
-Thomas~J.~Levario\affil{2}{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
-Hang~Lu\affil{2}{},
-Amit~Singer\affil{3}{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \affil{4}{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA},
-Stanislav~Y.~Shvartsman\affil{1}{} \affil{5}{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA},
-\and
-Ioannis~G.~Kevrekidis\affil{1}{} \affil{4}{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA}}
-
-\contributor{Submitted to Proceedings of the National Academy of Sciences
-of the United States of America}
-
-%% The \maketitle command is necessary to build the title page.
-\maketitle
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{article}
-
-\begin{abstract}
-In studies of development, researchers are often presented with cross-sectional data, where each data point is a sample from a population fixed at a slightly different developmental time.
-%
-The goal is then to temporally order the data to reconstruct the developmental dynamics.
-%
-If each data point is a two-dimensional image, the images must first be registered before they can be temporally ordered.
-%
-When such data sets are large, noisy, and/or if the developmental changes are subtle, these tasks can be difficult to do by hand.
-%
-We present an automatic approach to register {\it and} temporally order cross-sectional data sets of images.
-%
-The mathematical techniques (angular synchronization for image registration, diffusion maps for temporal ordering, and
-vector diffusion maps for simultaneously performing both tasks) are applicable to a wide variety of data sets and
-require little {\it a priori} knowledge of the image features or the developmental dynamics.
-%
-We demonstrate the utility of our methods using a collection of images from a study of {\it Drosophila} embryogenesis.
-\end{abstract}
-
-%% When adding keywords, separate each term with a straight line: |
-\keywords{temporal ordering | image registration}
-
-%% Optional for entering abbreviations, separate the abbreviation from
-%% its definition with a comma, separate each pair with a semicolon:
-%% for example:
-%% \abbreviations{SAM, self-assembled monolayer; OTS,
-%% octadecyltrichlorosilane}
-
-% \abbreviations{}
-
-%% The first letter of the article should be drop cap: \dropcap{}
-%\dropcap{I}n this article we study the evolution of ''almost-sharp'' fronts
-
-%% Enter the text of your article beginning here and ending before
-%% \begin{acknowledgements}
-%% Section head commands for your reference:
-%% \section{}
-%% \subsection{}
-%% \subsubsection{}
-
-\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional \cite{diggle2002analysis}.
-%
-In longitudinal studies, developmental progress is monitored over time for the same embryo.
-%
-In a cross-sectional study, one embryo contributes only a single snapshot of a chemical or morphological process along its developmental trajectory, and the developmental dynamics must be reconstructed from multiple snapshots of different embryos.
-%
-Both of these sampling schemes have their advantages and limitations, and both are extensively used by developmental biologists.
-%
-Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms.
-%
-In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that help visualize a handful of cellular processes.
-%
-Fixed embryos are then imaged using any given number of microscopy techniques.
-%
-Recent advances in large-scale physical manipulation and imaging of embryos have produced rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different geometric orientation and developmental time point.
-%
-Importantly, the ``age'' of any given embryo arrested in its development is not known to high accuracy.
-%
-In general, it is only known that a collection of embryos belongs to a certain time window.
-%
-In order to recover the developmental dynamics from such image datasets, snapshots of different embryos must be spatially aligned or registered to factor out the relevant symmetries (i.e., translations and rotations), and then ordered in time.
-%
-We present a general computational framework which greatly accelerates -and can combine- both of these tasks.
-
-Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are visually apparent.
-%
-As an example, Figure~\ref{fig:fish} shows a caricature of fish development, combining the processes of growth and color patterning.
-%
-In this case, temporal ordering can be accomplished by arranging the fish by size, which is monotonic with the developmental progress.
-%
-Image registration is based on obvious morphological landmarks, such as the position of head and fins.
-%
-On the other hand, real data poses nontrivial challenges, due to the number of images, measurement noise and variability, and the subtlety of the developmental changes.
-
-To illustrate the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of signaling protein expression in the early fruit fly embryo, one of the leading model systems for studies of developmental dynamics.
-%
-This dataset comprises $49$ fluorescence microscopy images of different embryos fixed during the $3^{rd}$ hour of their development.
-%
-\begin{figure}[t]
-\includegraphics[width=8.4cm]{fig1}
-\caption{{\it (A)} Fish, each in a different orientation and a different stage of development.
-{\it (B)} Fish, now rotationally registered and temporally ordered.
-For this caricature, the registration and ordering is easy to do ``by hand" because the data set is small and the developmental changes are easy to recognize.}
-%\label{fig:fish}
-\customlabel{fig:fish}{1}
-\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
-\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
-\end{figure}
-%
-The fruit fly embryo is shaped like a rice grain, approximately $1/2$~mm long.
-%
-By end of the $2^{nd}$ hour of development, sequential divisions of a fertilized egg generate a system where $\sim 6,000$ nuclei are uniformly arranged under the common plasma membrane.
-%
-During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclose nuclei into different cells grow in thickness (Figure~\ref{subfig:membranes}).
-%
-In conjunction with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs \cite{lim2013kinetics}.
-%
-Images in Figure~\ref{subfig:images_unordered} show optical sections at a fixed depth, perpendicular to the long axis of the embryo.
-%%%
-%%% we have to somehow say that it is consistent
-%%%
-%
-%%%
-%%% ask if "visualize" is typical as active verb
-%%%
-The embryos were fixed and stained with chemicals that help visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red) \cite{chung2010microfluidic}.
-%
-Each of the images reveals a uniform arrangement of nuclei and a nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.
-
-\begin{figure}[t]
-\includegraphics[width=8.4cm]{fig2}
-\caption{{\it (A)} Fluorescent images of {\it Drosophila} embryos.
-Each embryo has been stained for Dorsal (green) and dpERK (red).
-{\it (B)} Selected image from data set.
-{\it (C)} Profile of Dorsal (green) and dpERK (red) around the circumference of the embryo in {\it B}.
-}
-%\label{fig:images_unordered}
-\customlabel{fig:images_unordered}{2}
-\customlabel{subfig:images_unordered}{\ref{fig:images_unordered}{\it A}}
-\customlabel{subfig:one_image}{\ref{fig:images_unordered}{\it B}}
-\customlabel{subfig:image_profiles}{\ref{fig:images_unordered}{\it C}}
+\begin{figure}[h!]
+\includegraphics[width=8.4cm, trim=0cm 5.6cm 0cm 0cm, clip]{fig6}
+\caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different levels of development.
+{\it (B)} Membrane length as a function of time.
+The points corresponding to the images in {\it A} are indicated with arrows.}
+\label{fig:membrane_compare}
 \end{figure}
 
-%%%
-%%% orientation does not read good to Y below -
-%%% it is the same orientation, but a different rotation around the axis
-%%% how to say this ?
-%%%
-Each image in Figure~\ref{subfig:images_unordered} comes from a different embryo, captured at a different rotation about the long axis and at a different and unknown point in time.
-%
-The goal is to register these images in a globally consistent way, and order them to be representative of the underlying developmental trajectory.
-%
-Both of these tasks can be fully automated, and even combined.
-%
-We will use {\it angular synchronization} \cite{singer2011angular}, a registration algorithm that is robust to noise, to align the images in a consistent way.
-%
-We will use {\it diffusion maps} \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
-%
-We will also show how both steps can be combined using {\it vector diffusion maps} \cite{singer2012vector}.
-%
-%The paper is organized as follows.
-%
-%First, we use a synthetic dataset, where the correct orientation and ordering of data points is known, to formalize the tasks of image registration and temporal ordering.
-%
-%Then, we show that both of these tasks can be effectively combined in a single step.
-%
-%Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align them and order them correctly in time.
-%
-%We conclude by discussing the general requirements that must be satisfied for our approach to work and compare it to other approaches and ideas in cross-sectional studies of biological dynamics.
+\section{PCA for gastrulating images}
 
-\section{Results and Discussion}
+\begin{figure}[h!]
+\raisebox{5.2cm}{A}
+\includegraphics[width=16.8cm]{data2_PCA_ordered}
 
-%We would like to temporally order images, such as the images shown in Figure~\ref{subfig:images_unordered}.
-%
-%We will use diffusion maps, a dimensionality reduction algorithm, to automatically order our data.
-%
-%However, this algorithm requires distances between data points.
-%
-%Because each image is in a different orientation, we need to first align or {\it register} the images before computing distances.
-%
-%Once all the images have been aligned, we can compute distances and use diffusion maps to temporally order the data.
-
-%%%
-%%%  Y to discuss with S
-%%%  I think we should have the real data
-%%%
-%%%  also, these dynamics "are like" our dynamics
-%%%
-%%%   even if this is kept, one should say that this chosen because it LOOKS LIKE what we have
-%%% another reason I want to have the real data is because in the discussion I want to talk about the flips
-%%%
-
-We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set whose features resemble the image features in Figure~\ref{fig:images_unordered}.
-%
-The relatively simple dynamics of this data set will allow us to easily visualize and illustrate the main points of the different algorithms.
-%
-We construct concentration profiles defined on a ring (as in Figure~\ref{subfig:one_image}), and we assume that each ring is arbitrarily rotated around its center; an example is shown in Figure~\ref{subfig:1d_example}.
-%
-Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown at the bottom of Figure~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
-%
-{\bf We should discuss/mention that the eggs could be flipped.}
-%
-Each concentration profile is a noisy Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
-%
-We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_{100}$, the group of integers modulo $100$).
-%
-Figure \ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
-%
-Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.
-
-\begin{figure}[t]
-\includegraphics[width=8.4cm]{fig3}
-\caption{{\it (A)} One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
-{\it (B)} Intensity corresponding to the profile in {\it A}.
-{\it (C)} An ensemble of concentration profiles, each of the form described in {\it A}.
-Each row in the array corresponds to a single profile.
-{\it (D)} The profiles in {\it C}, now registered using angular synchronization.
-{\it (E)} The profiles in {\it D}, now temporally ordered using diffusion maps.
-{\it (F)} The profiles in {\it C}, registered and temporally ordered in a single step using vector diffusion maps.}
-%\label{fig:1d_demo}
-\customlabel{fig:1d_demo}{3}
-\customlabel{subfig:1d_example}{\ref{fig:1d_demo}{\it A}}
-\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}{\it B}}
-\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}{\it C}}
-\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}{\it D}}
-\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}{\it E}}
-\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}{\it F}}
+\raisebox{3.7cm}{B}
+\includegraphics[width=8.4cm]{data2_PCA_variance}
+\caption{{\it A} Images shown in Figure~4 of the paper, now ordered by the first PCA projection coefficient.
+Note that the ordering is not consistent with the developmental dynamics.
+{\it B} The eigenvalue spectrum from PCA.
+Note that subsequent components have a nonnegligible contribution on the data.}
 \end{figure}
 
-\subsection{Angular synchronization for image registration}
+\section{Mathematical Algorithms}
 
-We begin by registering the set of profiles using angular synchronization\cite{singer2011angular}, a technique which is robust to noise and requires no {\it a priori} knowledge of image features.
-%
-Angular synchronization first aligns every possible pair of data points, i.e., it computes the rotation angles
-which make each pair have the maximum possible overlap (see {\it Materials and Methods}).
-%%%
-%%% so I do not like the figure with the vectors - it is correct, but people have to take the
-%%% function, discretize and make it a vector in their heads, and may be "one step too much"
-%%%
-%
-An illustration is shown in Figure~\ref{subfig:synch1};
-each data point is a vector, and $\theta_{ij}$ denotes the angle of rotation needed to align vectors $i$ and $j$.
-%
-{\bf It is not clear how we are rotating the arrows (around the center vs. around the tail).}
-%
-From this large collection of pairwise alignments, the algorithm then selects a set of rotation angles, $\theta_i$ (one for each data point), such that differences in rotation angles, $\theta_i-\theta_j$, are {\it most consistent} (in some metric) with the pairwise alignments $\theta_{ij}$.
-%
-Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned; the computed rotation angles are also indicated.
-%
-Note that the differences between the rotation angles assigned to each vector are consistent with the pairwise angles needed to rotate one vector to another, i.e., $\theta_i - \theta_j \approx \theta_{ij}$.
-%
-The problem of finding these alignments is formulated as an optimization problem, and then relaxed to an eigenproblem.
-%
-The top eigenvector(s) of a matrix (which contains information about all pairwise alignments)
-provides the rotation angles (up to a constant rotation) assigned to each data point (see {\it Materials and Methods}).
-%
-%Because the method considers {\it all} pairwise alignments, angular synchronization is often robust to noise.
-%
-%
-Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization.
-%
-It is reassuring that the peaks appear aligned,
-even though the algorithm uses no specific information about the existence or location of such peaks.
-%
-%We note once more that, since angular synchronization only uses pairwise rotation information,
-%the optimal rotations are recovered up to a global rotation.
-
-\begin{figure}[t]
-\includegraphics[width=8.4cm]{fig4}
-\caption{ {\it (A)} Set of vectors, each in a different orientation.
-The pariwise alignment angles are indicated.
-{\it (B)} The vectors from {\it A}, each rotated so that the set is globally aligned.
-Note that the chosen rotation angles are consistent with the pairwise alignments in {\it A}.
-{\it (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
-Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \eqref{eq:dmaps_W}), so that close data points are connected by dark (``stronger'') edges.
-{\it (D)} The data in {\it C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure, $\phi_2$.
-The color intensity is monotonic with the arclength, thus parameterizing the curve.}
-%\label{fig:schematics}
-\customlabel{fig:schematics}{4}
-\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
-\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
-\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}{\it C}}
-\customlabel{subfig:dmaps_color}{\ref{fig:schematics}{\it D}}
-\end{figure}
-
-\subsection{Diffusion maps for temporal ordering}
-
-After geometrically registering the data (concentration profiles or images), we want to order them in time to reconstruct the developmental dynamics.
-%
-%Our main assumption is that most of the variation in the data correlates with time.
-%
-Our main assumptions are that the high-dimensional data (one dimension per discretization point in the concentration profiles/one dimension per color pixel intensity in the images) lie on a one-dimensional, nonlinear curve, {\it and} that this curve is parameterized by time.
-%
-Figure~\ref{subfig:dmaps_edges} shows a schematic of a two-dimensional data set (the ``high-dimensional" data) which lies on a one-dimensional nonlinear curve (the ``low-dimensional" manifold).
-%
-We assume that the data points are ordered from youngest to oldest along the curve.
-%
-When our data are very high dimensional, we cannot extract this curve by eye.
-%
-We instead use diffusion maps \cite{coifman2005geometric}, a dimensionality reduction technique which aims to uncover low-dimensional, nonlinear structure in high-dimensional data sets, to automatically extract this one-dimensional curve.
-%
-Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the leading (non-trivial) diffusion map coordinate.
-%
-The color parameterizes the one-dimensional curve that our eye can visually detect, and ordering the data by the leading diffusion map coordinate will effectively order data in time.
-
-To uncover the diffusion map parameterization, we assume that data points which are close in the high-dimensional space should also be close {\it on the curve} (we would like images which look similar to be close together in our developmental trajectory).
-%
-%We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\it as close as possible} in the parameterization.
-%
-To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored based on the distance between the data points (using the diffusion kernel of the distance), so that pairs of points which are close have dark/strong edges connecting them.
-%
-Visually, the dark edges delineate the one-dimensional curve, whereas the light edges are less informative about the low-dimensional structure.
-%
-%We want points which are close in the high-dimensional space, i.e., connected by dark edges, to be close in our low-dimensional %parameterization.
-%
-Finding such a parameterization can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a Markov matrix which contains information about pairwise distances between the data points (see {\it Materials and Methods}).
-%
-%This procedure has been named diffusion maps \cite{coifman2005geometric}.
-%
-%The formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).
-
-Given our set of registered signals, shown in Figure~\ref{subfig:1d_aligned_unordered}, we use diffusion maps to order them in time.
-%
-We use the Euclidean distance between the (discretized in $\mathbb{R}^{100}$) aligned concentration profiles (the rows of Figure~\ref{subfig:1d_aligned_unordered}) as our pairwise distances between data points.
-%
-Figure~\ref{subfig:1d_aligned_ordered} shows the data from Figure~\ref{subfig:1d_aligned_unordered}, now sorted by the value of the leading diffusion map coordinate, $\phi_2$.
-%
-As expected, the intensity of the peak grows as a function of $\phi_2$ (since we constructed these data
-to have a peak growing in time).
-%
-%Although diffusion maps can order the data, it cannot tell us about the speed of this trajectory.
-%
-%Furthermore, the direction of the trajectory is also unknown; determining which end of the trajectory corresponds to the beginning of development is a task that needs to be done by hand.
-
-\subsection{Vector diffusion maps for simultaneous registration and ordering}
-
-%We are often interested in both registering and ordering our data.
-%
-%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
-%
-Angular synchronization (for registration) can be combined with diffusion maps (for temporal ordering) in a single eigencomputation that allows us to {\it simultaneously} recover the registrations {\it and} the temporal ordering.
-%
-This technique is called vector diffusion maps \cite{singer2012vector} (see {\it Materials and Methods}).
-%
-It both reduces the number of required eigencomputations, and is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
-%
-Figure~\ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure~\ref{subfig:1d_unaligned_unordered} using vector diffusion maps.
-%
-The data has been ordered by $\psi_{1, 3}$, the leading vector diffusion map coordinate.
-
-\subsection{Analysis of two-dimensional images}
-
-The same methods (with small modifications) can be applied directly to our two-dimensional images.
-%
-Instead of one-dimensional profiles (discretized to $100$-dimensional concentration vectors), we now work with
-two color, two-dimensional images (discretized into $100 \times 100$ pixels/$20000$-dimensional vectors), and we register the images with respect to translations and rotations (the symmetry group is $ISO(2)$).
-%
-To facilitate the application of vector diffusion maps, we approximate the two-dimensional translations and rotations by {\it three-dimensional rotations} (symmetry group $SO(3)$; see {\it SI Appendix}).
-%
-We use vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_unordered};
-the results are shown in Figure~\ref{fig:images_ordered}.
-%
-Both the red and green peaks are now aligned, and the intensity of the red peaks grows as a function of the leading vector diffusion map coordinate.
-
-In these experiments, we are fortunate enough to be able to validate our  automatic registration and ordering using prior physical knowledge about the system.
-%
-For registration, it is known that the peak in the distribution of Dorsal (shown in green) specifies the ventralmost position in the embryo, and that the nuclei (shown in gray) delineate the circumference of the embryo.
-%%%
-%%% maybe you can also quantify this guy
-%%%
-%
-%Based on this, images can be translated to make the centers of the embryos coincide and rotated to place the peak of the green channel at the same angular position.
-%
-Temporal ordering, on the other hand, can be done based on the monotonic progress of cellularization (Figure~\ref{subfig:membranes}) \cite{figard2013plasma}.%; this helps validate our vector diffusion map-based ordering.
-%
-The two methods for registration and ordering can be visually compared through the sequences of one-dimensional profiles extracted from the circumference of the embryos (similar to the format in Figure~\ref{fig:1d_demo}), shown in Figures~\ref{subfig:1d_membranes}~and~\ref{subfig:1d_vdm}.
-%
-The general trends are consistent:
-both green  and red peaks are aligned, and the red peaks grow as a function of time/leading vector diffusion map coordinate.
-%
-However, Figure~\ref{subfig:1d_vdm}, which was automatically registered and ordered, shows an arguably better clustering of the ``outer" peaks in the top corners of the image.
-%
-%These peaks are more scrambled in Figure~\ref{fig:membrane_compare}b, which was analyzed using cellularization.
-%
-These peaks are known to arise later in development, and so it is reassuring that they appear at the end of the (recovered) trajectory.
-%
-Furthermore, the membranes stop growing towards the end of this trajectory, making the ordering of the late signals in Figure~\ref{subfig:1d_membranes} less accurate.
-%
-Quantitatively, the rank correlation coefficient between these two temporal orderings is 0.9158, indicating that our automatic ordering is consistent with the ordering obtained from cellularization.
-
-%Although our methods are fairly robust, they do require a sufficient amount of data so that DMAPS can ``see'' a smooth trajectory.
-%
-%We evaluated how much data is required by subsampling our data set (with replacement), and for each subsample, computing the temporal ordering using VDM and calculating the rank correlation between this ordering and the ordering from cellularization.
-%
-%The results are shown in Figure~\ref{subfig:bootstrap} (each data point is averaged over 50 random subsamples).
-%
-%For small data sets, the average rank correlation is very low, indicating that the orderings from VDM are inaccurate.
-%
-%However, for moderately sized data sets (approximately 30 data points), the rank correlation begins to plateau and VDM can accurately order the data.
-
-\begin{figure}[t]
-\includegraphics[width=8.4cm]{fig5}
-\caption{The images in Figure~\ref{subfig:images_unordered}, aligned and temporally ordered using vector diffusion maps.
-The images are sorted according to the value of the first non-trivial vector diffusion map component, $\psi_{1, 4}$.}
-%\label{fig:images_ordered}
-\customlabel{fig:images_ordered}{5}
-\end{figure}
-
-\begin{figure}[t]
-\includegraphics[width=8.4cm]{fig6}
-\caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different levels of development.
-{\it (B)} Membrane length as a function of time.
-The points corresponding to the images in {\it A} are indicated with arrows.
-{\it (C)} The one-dimensional concentration profiles, registered by peak Dorsal fluorescence and ordered by membrane length.
-{\it (D)} The one-dimensional concentration profiles corresponding to the automatically registered and ordered images in Figure~\ref{fig:images_ordered}.
-}
-%\label{fig:membrane_compare}
-\customlabel{fig:membrane_compare}{6}
-\customlabel{subfig:membranes}{\ref{fig:membrane_compare}{\it A}}
-\customlabel{subfig:membrane_curve}{\ref{fig:membrane_compare}{\it B}}
-\customlabel{subfig:1d_membranes}{\ref{fig:membrane_compare}{\it C}}
-\customlabel{subfig:1d_vdm}{\ref{fig:membrane_compare}{\it D}}
-\end{figure}
-
-\section{Conclusions}
-
-We have shown how symmetry and dimensionality reduction techniques can successfully register and order cross-sectional developmental imaging data.
-%
-Although we illustrated our methods using a specific data set from a study of {\it Drosophila} embryogenesis, the framework is quite general and requires little knowledge of the image features or the developmental dynamics.
-%
-To our knowledge, temporal ordering of {\it Drosophila} images has not been automated to this extent.
-%
-Recently, machine learning algorithms have been used to group {\it Drosophila} images into different developmental classes \cite{yuan2014automated}, yet there, the appropriate classes must be defined by the user and hand-labeled training data are required.
-
-Temporal ordering of cross-sectional biological data has been studied in a variety of other contexts.
-%
-Cross-sectional RNA sequencing data has been temporally ordered \cite{anavy2014blind, trapnell2014dynamics}, where each data point is the transcriptome for a single sample.
-%
-Gene expression data from multiple patients have been studied in \cite{gupta2008extracting, qiu2011discovering};
-here, each data point is a vector containing the expression levels of the various genes.
-%
-Snapshots of cells throughout the cell cycle can be described by a vector of features (e.g., amount of DNA) which quantify the cell's state  \cite{kafri2013dynamics}.
-%
-These problems are no different than our task of ordering images, each of which is a large vector of pixel intensities.
-%
-The majority of these works temporally order the data either by solving a traveling salesman problem on the data, or by constructing a minimum spanning tree of the data,
-expecting that these structures (the path of the traveling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate encoding of the dynamical progression.
-%
-We, instead, use diffusion maps to construct a one-dimensional parameterization of the data.
-%
-Diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003, tenenbaum2000global, Donoho2003, Roweis2000}.
-%
-They have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}, and so we suspect they may perform better than the ordering algorithms used in previous work.
-%
-{\bf Yannis says ``careful'', he thinks this might be too strong of a statement.}
-%%%
-%%%   we need help here
-%%%
-%
-%Furthermore, although we use only the first diffusion maps coordinate to order the data (we assume the data are one-dimensional), diffusion maps are not limited to one-dimensional data and can be used to extract more complex structures.
-%
-%For example, in data which contains several branches or classes, the first few coordinates will delineate the branches or separate the classes.
-
-The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
-%
-The main point to note here is that many registration algorithms require definition of appropriate landmarks or features for the images and the location of these within each image.
-%
-For example, in face recognition, the first step is often to find the eyes, nose, and mouth within the face, and then register images by aligning these features \cite{zhao2003face}.
-%
-Temporally ordering images of a construction site over time is also based on first identifying features within the images \cite{schindler2007inferring}.
-%
-In general, appropriate features may not be known {\it a priori}, and so we require a registration algorithm that does not rely on predefined features, but only on the inherent geometric symmetry of the problem.
-%
-Template-based methods are also common for registration, where one optimally aligns each image to a predefined template function.
-%
-However, these methods can suffer when the images are noisy, and the success of such methods can depend strongly on the template function \cite{shatsky2009method}.
-%
-{\bf Need to discuss consistency.}
-
-Angular synchronization and vector diffusion maps require no feature points and no template function.
-%
-These methods have been used to register two-dimensional cryo-electron microscopy images of three-dimensional molecules \cite{singer2011three}, where each image is a picture of (essentially) the {\it same} molecule; variation in the data comes from differences in the viewing angles (rotations of the molecule) and from instrument/measurement noise.
-%
-In contrast, our data contains variation from the different embryo orientations, as well as variation from the underlying developmental dynamics.
-% (which underpins the development of vector diffusion map algorithms).
-%
-In addition to the eigenproblem relaxation used here, other synchronization algorithms have been proposed \cite{bandeira2014multireference}.
-%
-Synchronization is formulated for general symmetry groups, and depending on the data, one might consider other symmetries, such as reflections and scaling/dilation.
-%
-Here, the main requirement is that the symmetry group be represented by orthogonal or unitary matrices \cite{singer2013spectral}.
-
-Temporal ordering can also be accomplished without first registering the images, by constructing functions of the images which are invariant under translations and rotations. 
-%
-These functions are then used as the data in the diffusion maps calculation; since these representative functions do not change under translations and rotations, the resulting diffusion map embedding is invariant to symmetries.
-%
-Such functions include the scale invariant feature transform \cite{lowe1999object, lowe2004distinctive}, the scattering transform \cite{mallat2012group}, and the bispectrum \cite{zhao2014rotationally}.
-%
-These methods have the disadvantage that we do not obtain a registered set of images, but only a temporal ordering.
-%
-However, such methods can often be faster for large data sets, since the functions are computed per image, rather than computing pairwise quantities in synchronization.
-
-In all of these data-driven methods, an essential question is how much and what type of data is needed to accurately reconstruct the dynamics.
-%
-For diffusion maps to temporally order the data, we require the developmental trajectory to be well-sampled, and that the principal direction of variability in the data (the coordinate which diffusion maps will recover) be monotonic in time.
-%
-In addition, the data should have spatial variability/structure for synchronization to successfully register the images.
-%
-In our example, the data set is sufficiently large so that the dynamics are well-sampled.
-%
-We have included both the red and the green channels so that the principal direction of variability is monotonic in time.
-%
-If we only used the green channel in our analysis, we could not successfully order the data using diffusion maps, as this signal does not change appreciably in time.
-%
-Both the red and green channels contain spatial variability which allows us to register the data points: the green channel has one peak at a specific angular around the embryo, and the red channel has two peaks which grow in time.
-%
-We include the green channel because the location of the red peaks is noisy, and so inclusion of the (less noisy) green channel allows us to more accurately register the images.
-%
-%For the methods we present here, we require a sufficient amount of data for the one-dimensional evolution curve to be ``seen'' by the algorithms (i.e., in Figure~\ref{subfig:dmaps_edges}, there are a sufficient number of data points so that our eye can see the curve; if there were only 4 data points, it would most likely be difficult for our eye to detect the one-dimensional structure).
-%
-%We also require the data to contain enough information so that the alignments and dynamics are apparent.
-%
-%For example, in the data presented in Figure~\ref{subfig:images_unordered}, the green peaks delineate the position of the ventral axis in the embryo, and the intensity of the red peaks grow over time.
-%
-%Therefore, these images contain information about rotational symmetries of the data as well as the dynamics of the data.
-%
-%If we removed the green signal, and only used the red signal in our analysis, the registration would not be as accurate.
-%
-%Furthermore, if we only used the green signal, then there would be no dynamic information (the shape or intensity of Dorsal does not change as a function of time in the developmental stage we are considering), and the temporal ordering would be inaccurate.
-%
-We are confident that for many applications, these requirements - data which adequately samples the developmental trajectory, and data with sufficient spatial and temporal variability - are easily met.
-%
-In these cases, our methods allow for the rapid, automated analysis of imaging data sets to help uncover possibly complex dynamics and structure.
-{\bf Something about smoothness in making movies}
-%
-%{\bf procrustes referece}
-%{\bf allso the guy form Donoho who worried about how much to do syncjhronization}
-
-%%%   you should sample well enough in time
-%%% your dmap coordinate should have a nice slope with time
-%% you should have variation in angle ---and we know we are random in the rotations
-%%
-%% and now the green is an issue of noise.....
-%% the green "sees" angle but not time,
-%% the red sees angle and time, but angle badly
-%% the ocmbination sees both well enough
-%%
-%%
-%%   we need to say things about "other methods" -- Mallat, AmitNow, Procrustes...
-%% and if we change the artificial stuff, but even if not
-%% we need to say something about flips.
-%% we need to say symmetry early at the beginning -- important
-%%  the expression "factoring out symmetry", the reference to Ben....  something about symmetry groups 
-%%    the smoothness when one wants to make movies.
-%%  
-%%
-%%   Y is afraid of the compsci literature being "fairly represented"
-%%
-%%
-
-%% == end of paper:
-
-%% Optional Materials and Methods Section
-%% The Materials and Methods section header will be added automatically.
-
-%% Enter any subheads and the Materials and Methods text below.
-
-\begin{materials}
-
-\section{Angular synchronization \cite{singer2011angular}}
-
-%\sansmath
+\subsection{Angular synchronization \cite{singer2011angular}}
 
@@
 
-\section{Diffusion maps \cite{coifman2005geometric}}
+\subsection{Diffusion maps \cite{coifman2005geometric}}
 
@@
 
-\section{Vector diffusion maps\cite{singer2012vector}}
+\subsection{Vector diffusion maps\cite{singer2012vector}}
 
@@
 useful parameterization of the data manifold modulo symmetries.
-%%%
-%%% \cite....
-%%%
-
-\end{materials}
-
-%% Optional Appendix or Appendices
-%% \appendix Appendix text...
-%% or, for appendix with title, use square brackets:
-%% \appendix[Appendix Title]
-
-\begin{acknowledgments}
-- text of acknowledgments here, including grant info -
-\end{acknowledgments}
-
-%% PNAS does not support submission of supporting .tex files such as BibTeX.
-%% Instead all references must be included in the article .tex document.
-%% If you currently use BibTeX, your bibliography is formed because the
-%% command \verb+\bibliography{}+ brings the <filename>.bbl file into your
-%% .tex document. To conform to PNAS requirements, copy the reference listings
-%% from your .bbl file and add them to the article .tex file, using the
-%% bibliography environment described above.
-
-%%  Contact pnas@nas.edu if you need assistance with your
-%%  bibliography.
-
-% Sample bibliography item in PNAS format:
-%% \bibitem{in-text reference} comma-separated author names up to 5,
-%% for more than 5 authors use first author last name et al. (year published)
-%% article title  {\it Journal Name} volume #: start page-end page.
-%% ie,
-% \bibitem{Neuhaus} Neuhaus J-M, Sitcher L, Meins F, Jr, Boller T (1991)
-% A short C-terminal sequence is necessary and sufficient for the
-% targeting of chitinases to the plant vacuole.
-% {\it Proc Natl Acad Sci USA} 88:10362-10366.
-
-%% Enter the largest bibliography number in the facing curly brackets
-%% following \begin{thebibliography}
-
-\bibliographystyle{pnas}
-\bibliography{background_reading/references,../../references/references}
-
-%\begin{thebibliography}{}
-
-%\end{thebibliography}
-
-\end{article}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Adding Figure and Table References
-%% Be sure to add figures and tables after \end{article}
-%% and before \end{document}
-
-%% For figures, put the caption below the illustration.
-%%
-%% \begin{figure}
-%% \caption{Almost Sharp Front}\label{afoto}
-%% \end{figure}
-
-%% For Tables, put caption above table
-%%
-%% Table caption should start with a capital letter, continue with lower case
-%% and not have a period at the end
-%% Using @{\vrule height ?? depth ?? width0pt} in the tabular preamble will
-%% keep that much space between every line in the table.
-
-%% \begin{table}
-%% \caption{Repeat length of longer allele by age of onset class}
-%% \begin{tabular}{@{\vrule height 10.5pt depth4pt  width0pt}lrcccc}
-%% table text
-%% \end{tabular}
-%% \end{table}
-
-%% For two column figures and tables, use the following:
-
-%% \begin{figure*}
-%% \caption{Almost Sharp Front}\label{afoto}
-%% \end{figure*}
-
-%% \begin{table*}
-%% \caption{Repeat length of longer allele by age of onset class}
-%% \begin{tabular}{ccc}
-%% table text
-%% \end{tabular}
-%% \end{table*}
-
-%\begin{figure}
-%\begin{minipage}{2.9cm}
-%\centering
-%\includegraphics[width=\textwidth]{illustrate_1d}\\
-%\vspace{-0.25cm}
-%(a)\\
-%\includegraphics[width=\textwidth]{registered_unordered_1d}\\
-%(d)
-%\end{minipage}
-%\begin{minipage}{2.9cm}
-%\centering
-%\includegraphics[width=\textwidth]{illustrate_1d_intensity}\\
-%(b)\\
-%\includegraphics[width=\textwidth]{registered_ordered_1d}\\
-%(e)
-%\end{minipage}
-%\begin{minipage}{2.9cm}
-%\centering
-%\includegraphics[width=\textwidth]{unregistered_unordered_1d}\\
-%(c)\\
-%\includegraphics[width=\textwidth]{registered_ordered_vdm_1d}\\
-%\centering
-%(f)
-%\end{minipage}
-%\caption{Analysis of one-dimensional profiles obtained from experimental data.
-(a) One-dimensional, two-component concentration profile on a ring (top), and the corresponding profile on a line (bottom).
-The concentration profile was extracted from an image of the type shown in Figure~\ref{fig:images_unordered}.
-(b) Intensity corresponding to the profile in (a).
-(c) An ensemble of concentration profiles, each of the form described in (a).
-Each row in the array corresponds to a single profile obtained from an image of the type in Figure~\ref{subfig:images_unordered}.
-(d) The profiles in (c), now registered using angular synchronization.
-(e) The profiles in (d), now temporally ordered using diffusion maps.
-(f) The profiles in (c), registered and temporally ordered in a single step using vector diffusion maps.}
-%\end{figure}
-%
-
-\begin{figure*}
-\begin{minipage}{0.2\textwidth}
-\centering
-\includegraphics[width=\textwidth]{sphere_1}\\
-\includegraphics[width=\textwidth]{sphere2_1}\\
-A
-\end{minipage}
-%
-\begin{minipage}{0.2\textwidth}
-\centering
-\includegraphics[width=\textwidth]{sphere_2}\\
-\includegraphics[width=\textwidth]{sphere2_2}\\
-B
-\end{minipage}
-\begin{minipage}{0.2\textwidth}
-\centering
-\includegraphics[width=\textwidth]{sphere_3}\\
-\includegraphics[width=\textwidth]{sphere2_3}\\
-C
-\end{minipage}
-\begin{minipage}{0.2\textwidth} 
-\centering
-\includegraphics[width=\textwidth]{sphere_4}\\
-\includegraphics[width=\textwidth]{sphere2_4}\\
-D
-\end{minipage}
-\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions.
-{\it (A)} The original image.
-{\it (B)} Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image.
-{\it (C)} Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation.
-{\it (D)} Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.
-The top row shows the three-dimensional spheres, and the bottom row shows the (two-dimensional) surface of the sphere in which we are interested.}
-\label{fig:SO3_picture}
-\end{figure*}
 
@@
 
+\bibliographystyle{pnas}
+\bibliography{background_reading/references,../../references/references}
+
 \end{document}

### 2014-05-14 image_analysis_paper/si_information.tex
added more to si information

@@
 
+\section{Image parameters}
+
+For the images shown in Figure~3 of the paper, we only use the red and green channels (dpERK and Dl) to register the images.
+
+For the images shown in Figure~4 of the paper, we use all three channels (dpERK, Twi, and nuclei) to register the images.
+%
+We blur the nuclei using a disc filter with radius $5$~pixels.
+%
+We also adjust the intensity of the nuclear signal such that 1\% of the pixels are saturated.
+
 \section{PCA for gastrulating images}
@@
 \includegraphics[width=8.4cm]{data2_PCA_variance}
-\caption{{\it A} Images shown in Figure~4 of the paper, now ordered by the first PCA projection coefficient.
+\caption{{\it (A)} Images shown in Figure~4 of the paper, now ordered by the first PCA projection coefficient.
 Note that the ordering is not consistent with the developmental dynamics.
-{\it B} The eigenvalue spectrum from PCA.
+{\it (B)} The eigenvalue spectrum from PCA.
 Note that subsequent components have a nonnegligible contribution on the data.}
@@
 \subsection{Angular synchronization \cite{singer2011angular}}
+
+\begin{figure}[t]
+\includegraphics[width=8.4cm]{fig3}
+\caption{{\it (A)} One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
+{\it (B)} Intensity corresponding to the profile in {\it A}.
+{\it (C)} An ensemble of concentration profiles, each of the form described in {\it A}.
+Each row in the array corresponds to a single profile.
+{\it (D)} The profiles in {\it C}, now registered using angular synchronization.
+{\it (E)} The profiles in {\it D}, now temporally ordered using diffusion maps.
+{\it (F)} The profiles in {\it C}, registered and temporally ordered in a single step using vector diffusion maps.}
+\label{fig:1d_demo}
+\end{figure}
 

### 2014-05-17 image_analysis_paper/pnas_draft2.tex
Added draft edits

@@
 %
-In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that visualize a handful of cellular processes.
+In a typical cross-sectional study, a group of embryos is fixed using a procedure that arrests their development and stained with chemicals that visualize a handful of cellular processes.
 %
@@
 \includegraphics[width=8.4cm]{fig1}
-\caption{{\it (A)} Fish, each in a different orientation and a different stage of development.
+\caption{Caricature illustrating the tasks of image registration and ordering.
+{\it (A)} Fish, each in a different orientation and a different stage of development.
 {\it (B)} Fish, now rotationally registered and temporally ordered.
@@
 %
-We will demonstrate these techniques using two data sets from a study of {\it Dropophila} embryogenesis.
+Based on the low-dimensional parameterization of high-dimensional vectors, such as images, these techniques have been used for data mining in a wide range of applications, from cryo-EM analysis of individual molecules to face recognition.
+%
+Here we adapt these techniques for temporal ordering and registration of imaging data from cross-sectional studies of developmental dynamics, with the main objective to reveal stereotypic developmental trajectories.
+%
+We demonstrate these techniques using two data sets from a study of {\it Dropophila} embryogenesis, one of the best experimental models for studies of developmental dynamics.
 %
@@
 Here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
+%
+Add another sentence to say that in addition to registration and ordering, we can discover other…sources of order…such as the presence of subgroups of embryos which follow distinct developmental paths..
 
@@
 %
-We would like to organize these data sets in a meaningful and informative way.
+Our goal is to organize these data sets in a meaningful and informative way.
 %
@@
 %
+
 Vector diffusion maps has been used to classify images from cryo-electron microscopy (cryo-EM) experiments \cite{...}, where each image is a projection of a molecule, taken from a different viewing direction and in a different rotation.
@@
 %
-Using the alignment angles between all pairs of vectors (or images), angular synchronization then finds the set of rotation angles, one angle for each vector (or image), that is most consistent with the pairwise measurements.
+Using the alignment angles between all pairs of vectors (or images), angular synchronization then finds the set of rotation angles, one angle for each vector, that is most consistent with the pairwise measurements.
 %
@@
 %
-After removing the variability due to translations and rotations of the images, the remaining variability is (mostly) due to developmental dynamics.
+After removing the variability due to translations and rotations of the images, the remaining variability may be due to developmental dynamics.
 % 
-These dynamics can be revealed by ordering the data along the one-dimensional manifold that parametrizes the data. 
+These dynamics can be revealed by ordering the data along the one-dimensional manifold that parameterizes the data. 
 %
@@
 %
+
 The idea is illustrated in Figure~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve. 
@@
 \includegraphics[width=8.4cm]{fig4}
-\caption{ {\it (A)} Set of vectors, each in a different orientation.
+\caption{Schematic illustrating angular synchronization and diffusion maps.
+{\it (A)} Set of vectors, each in a different orientation.
 The pairwise alignment angles are indicated.
@@
 \end{figure}
-
-\subsection{Images of {\it Drosophila} embryos during cellularization}
-
-%
-As a first illustration of this approach to temporal ordering and registration of images, we used a dataset where the correct registration and temporal order is known. 
-%
-This dataset is obtained by fluorescent imaging of {\it Drosophila} embryos during the third hour after fertilization.
-%
-By this stage of development, sequential divisions of the fertilized egg have generated a system where most of the nuclei are arranged in a monolayer under the common plasma membrane.
-%
-Figure 3a shows optical cross-sections of each embryo, with nuclei (gray) labelled by a DNA stain. 
-%
-The spatially uniform arrangement of nuclei is patterned by the concentration profiles of chemical signals that act as dose-dependent regulators of cell differentiation.
-%
-One of these signals is provided by the nuclear localization of transcription factor Dorsal (Dl, green), which subdivides the embryo into the territories that give rise to the muscle, nerve, and skin tissues.
-%
-Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK), an enzyme that specifies a subset of neuronal cells. 
-%
-
-Previous studies \cite{...} established that throughout the 3rd hour of development, the nuclear localization of Dl is unimodal and peaked at the ventral side of the embryo. 
-%
-The spatiotemporal pattern of dpERK is more complex: dpERK is first detected in two lateral stripes, which increase in intensity towards the end of the 3rd hour. 
-%
-This bimodal pattern is then augmented by a peak at the dorsal side of the embryo.
-%
-Both of these features become readily apparent when images are registered and ordered using vector diffusion maps: the peak of nuclar Dl is consistent throughout the images and intensity of the dpERK signal in the lateral regions of the embryo monotonically increases across the dataset.
-%
-The quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the temporal ordering provided by an independent approach.
-%
-Specifically, during this developmental timespan, the embryo cellularizes: nuclei located under common plasma membrane are enclosed by lateral membranes which grow inward, separating nuclei into individual cells.
-%
-The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo within 1-3 minutes  \cite{...} (see {\it SI Appendix}).    
-%
-The rank correlation coefficient between the orderings provided by membranes and vector diffusion maps is 0.9127. 
-%
-Thus, we have validated a dimensionality reduction approach to temporal ordering and registration of data  on a dataset where image orientation and ordering can be assessed independently.
-
-%Furthermore, the dynamics uncovered by vector diffusion maps is relatively simple, which suggests that a simpler approach could have been used to order the data.
-%
-%Indeed, a principal component analysis (PCA) of registered data reveals that data can be accurately ordered by projecting each image onto the first PCA mode.
 
@@
 
+\subsection{Images of {\it Drosophila} embryos during cellularization}
+
+%
+As a first illustration of this approach to temporal ordering and registration of images, we used a dataset where the correct registration and temporal order is known {\it a priori}. 
+%
+This dataset is obtained by fluorescent imaging of {\it Drosophila} embryos during the third hour after fertilization.
+%
+By this stage of development, sequential divisions of the fertilized egg have generated a system where most of the nuclei are arranged in a monolayer under the common plasma membrane \cite{foe1983studies}.
+{\bf TODO: We need to have a supplementary figure that shows the lateral view, that makes it even more clear that we work with optical sections.}
+%
+Figure 3a shows optical cross-sections of each embryo, with nuclei (gray) labeled by DAPI, a DNA stain. 
+%
+During the 3rd hour of development, the spatially uniform and two-dimensional arrangement of nuclei is patterned by the concentration profiles of chemical signals that control cell differentiation.
+%
+One of these signals is provided by the nuclear localization of transcription factor Dorsal (Dl, green), which subdivides the early embryo into the territories that give rise to the muscle, nerve, and skin tissues.
+%
+Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK), an enzyme that specifies a subset of neuronal cells. 
+%
+
+Previous studies \cite{rushlow2012temporal} established that throughout the 3rd hour of development, the nuclear localization of Dl is unimodal and peaked at the ventral side of the embryo. 
+%
+The spatiotemporal pattern of dpERK is more complex: dpERK is first detected in two lateral stripes, which increase in intensity towards the end of the 3rd hour \cite{lim2013kinetics}.  
+%
+This bimodal pattern is then augmented by a peak at the dorsal side of the embryo.
+%
+Both of these features become readily apparent when images are registered and ordered using vector diffusion maps: the peak of nuclear Dl is consistent throughout the images and intensity of the dpERK signal in the lateral regions of the embryo monotonically increases across the dataset.
+%
+
+For this dataset, the quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the ordering based on an independent approach.
+%
+Specifically, during the 3rd hour of development, the embryo cellularizes. 
+%
+In this process, nuclei located under common plasma membrane are enclosed by lateral membranes which grow inward, enclosing nuclei into individual cells.
+%
+The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo within 1-3 minutes \cite{figard2013plasma} (see {\it SI Appendix}).    
+%
+The rank correlation coefficient between the orderings estimated based on the progress of cellularization and vector diffusion maps is 0.9127. 
+%
+Thus, we have validated a dimensionality reduction approach to temporal ordering and registration of data on a dataset where image orientation and ordering can be assessed independently.
+
+{\bf TODO: add something about size variability}
+
+%Furthermore, the dynamics uncovered by vector diffusion maps is relatively simple, which suggests that a simpler approach could have been used to order the data.
+%
+%Indeed, a principal component analysis (PCA) of registered data reveals that data can be accurately ordered by projecting each image onto the first PCA mode.
+
 \subsection{Images of gastrulating {\it Drosophila} embryos}
@@
 %
+The next dataset illustrates this point.
+
 Figure~\ref{subfig:raw_data2} shows a data set of $108$ images of {\it Drosophila} embryos.
 %
-These images cover a thirty minute time interval and span late cellularization (the early images in this data set overlap with the late images in the previous set) through gastrulation, when tissues within the embryo begin to morph and deform.
+These images cover a thirty minute time interval and span late cellularization (the early images in this data set overlap with the late images in the previous set) through gastrulation, when the two-dimensional sheet of newly formed cells begins to deform on a way to forming the three-dimensional structures of the future organs.
 %
-Each image shows the optical cross-section of an embryo fixed at a different developmental time.
+As before, each image is an optical cross-section of an embryo fixed at a different developmental time.
 %
@@
 %
-Twist (Twi, shown in green) denotes the ventralmost point of the embryo (similar to Dl in the previous data set).
+Instead of Dl, which is difficult to visualize at this point, we used Twist (Twi, shown in green), a transcription factor that is activated by high levels of Dl, specifying the cells of the future muscle tissue.
 %
-During this developmental time span, the ventral furrow is formed, where the ventral side invaginates towards the center of the embryo.
+
+%
+During this developmental time span, the ventral furrow is formed, where the ventral side of the embryo buckles towards the center of the embryo, bringing in the future muscle cells.
 %
@@
 %
-During this developmental timespan, germ band elongation causes cells from the ventral side of the embryo to move towards the poles of the embryo, and then wrap around to the dorsal side.
+During this time window, germband elongation causes cells from the ventral side of the embryo to move towards the poles of the embryo, and then wrap around to the dorsal side.
 %
-At the end of this developmental timespan, cells which were originally on the ventral side of the embryo have migrated to the dorsal side of the embryo, causing similar ``omega'' shaped patterns (formed by Twi and dpERK) to arise on the dorsal side.
-%
-Although we know about this general developmental progression, unlike the previous data set, we have no method to measure the precise developmental time of each embryo, 
-and we are forced to use other methods to temporally order these images.
+At the end of this process, cells which were originally on the ventral side of the embryo have migrated to the dorsal side of the embryo, causing similar ``omega'' shaped patterns (formed by Twi and dpERK) to arise on the dorsal side.
 
@@
 %
-To highlight the developmental dynamics that we uncover using vector diffusion maps, we construct an average developmental trajectory.
+To highlight the developmental dynamics revealed by vector diffusion maps, we construct an average developmental trajectory.
 %
-We divide the registered and ordered data into windows or classes (each window contains a group of consecutive images in the ordered data), and then calculate the average image in each window.
+This is done by dividing the registered and ordered data into windows or classes (each window contains a group of consecutive images in the ordered data), and then calculate the average image in each window.
 %
 Figure~\ref{subfig:averaged_data2} shows the average trajectory for our data set; it is now very straightforward to see the underlying developmental dynamics.
+%
+We therefore accomplished the same task presented in the caricature in Figure~\ref{fig:fish}, but without knowledge about appropriate landmarks and or the appropriate statistic (e.g., body size) by which to order.
 
@@
 
+\subsection{Identifying distinct developmental trajectories}
+
+Vector diffusion maps not only allows us temporally order data, but also allows us to uncover other sources of variability.
+%
+Until now, we have been assuming that our experiments contain one main direction of variability, and that this direction is parameterized by time.
+%
+However, this is not alw
+
+\begin{figure*}[t]
+\raisebox{8.1cm}{{\figtextfont A}}
+\includegraphics[height=8.4cm]{raw_data3}
+\hspace{0.5cm}
+\raisebox{8.1cm}{{\figtextfont B}}
+\includegraphics[width=8.4cm]{mutants_VDM2}
+\caption{{\it (A)} Images of cellularizing {\it Drosophila} embryos.
+Each image is of a different embryo in a different rotational orientation.
+$19$ of the images are of wild type embryos, and $23$ of the images are of sna null mutant embryos.
+{\it(B)} Data from {\it A}, organized using the first two vector diffusion maps coordinates.}
+\customlabel{fig:data3}{5}
+\customlabel{subfig:raw_data3}{\ref{fig:data3}{\it A}}
+\customlabel{subfig:ordered_data3}{\ref{fig:data3}{\it B}}
+\end{figure*}
+
 \section{Conclusions}
@@
 %
-For our biological applications, good features may not be known {\it a priori}.
+Because of complexity of chemical and morphological changes in developing tissues, good features may not be known {\it a priori}.
 %
@@
 %
-Data sets such as  RNA sequencing data (where each data point is the transcriptome for a single sample) \cite{anavy2014blind, trapnell2014dynamics}, gene expression data from multiple patients (where each data point is a vector containing the expression levels of the various genes) \cite{gupta2008extracting, qiu2011discovering}, and snapshots of cells throughout the cell cycle (where each point is a vector of features, such as the amount of DNA, which quantify the cell's state)  \cite{kafri2013dynamics} have all been temporally ordered to extract a representative trajectory of the underlying dynamic process.
+Data sets such as  transcriptional profiling data (where each data point is the transcriptome for a single sample, such as a cell or tissue sample from a patient) \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering}, and snapshots of cells throughout the cell cycle (where each point is a vector of features, such as the amount of DNA, which quantify the cell's state)  \cite{kafri2013dynamics} have all been temporally ordered to extract a representative trajectory of the underlying dynamic process.
 %
@@
 
+{\bf TODO: add about how we can correlate physical variables with dmaps variables.}
+%
 After registering and temporally ordering images, one can perform a variety of post-processing tasks to gain further insight into the developmental processes.

### 2014-05-17 image_analysis_paper/pnas_draft2.tex
all edits before sending to Yannis

@@
 %
+Furthermore, these techniques can separate multiple trajectories within the same data set.
+%
 We demonstrate the utility of these methods using a collection of images from a study of {\it Drosophila} embryogenesis.
@@
 %% When adding keywords, separate each term with a straight line: |
-\keywords{temporal ordering | image registration}
+\keywords{temporal ordering | image registration | vector diffusion maps}
 
@@
 %
-In general, the landmarks needed for registration, such as the heads and fins in our example, as well as the features that can be used to order the data, such as the body length, are not known {\it a priori}.
+In general, the landmarks needed for registration, such as the heads and fins, as well as the attributes which can be used to order the data, such as the body length, are not known {\it a priori}.
 %
@@
 \includegraphics[width=8.4cm]{fig1}
-\caption{Caricature illustrating the tasks of image registration and ordering.
-{\it (A)} Fish, each in a different orientation and a different stage of development.
-{\it (B)} Fish, now rotationally registered and temporally ordered.
+\caption{Caricature illustrating the tasks of image registration and temporal ordering.
+{\xfigtextfontit (A)} Fish, each in a different orientation and a different stage of development.
+{\xfigtextfontit (B)} Fish, now rotationally registered and temporally ordered.
 For this caricature, the registration and ordering is easy to do ``by hand" because the data set is small and the developmental changes are easy to recognize.}
@@
 %
-Based on the low-dimensional parameterization of high-dimensional vectors, such as images, these techniques have been used for data mining in a wide range of applications, from cryo-EM analysis of individual molecules to face recognition.
+Based on manifold learning algorithms for high-dimensional data (such as images), these techniques have been used for data mining in a wide range of applications, from cryo-EM analysis of individual molecules to face recognition.
 %
-Here we adapt these techniques for temporal ordering and registration of imaging data from cross-sectional studies of developmental dynamics, with the main objective to reveal stereotypic developmental trajectories.
+Here we adapt these techniques for temporal ordering and registration of imaging data from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
 %
-We demonstrate these techniques using two data sets from a study of {\it Dropophila} embryogenesis, one of the best experimental models for studies of developmental dynamics.
+We demonstrate these techniques using three data sets from a study of {\it Dropophila} embryogenesis, one of the best experimental models for studies of developmental dynamics.
 %
@@
 %
-Add another sentence to say that in addition to registration and ordering, we can discover other…sources of order…such as the presence of subgroups of embryos which follow distinct developmental paths..
+The third data set, shown in Figure~\ref{subfig:raw_data3}, consists of data collected from both wild type and mutant embryos.
+%
+We will demonstrate that these algorithms can not only extract stereotypic dynamics, but also separate data sampled from distinct developmental paths.
 
@@
 
-A typical dataset from cross-sectional studies of developmental processes is a collection of images of different embryos, each of which is fixed at a different developmental time and imaged in a different rotational orientation.
+A typical dataset from cross-sectional studies of developmental processes is a collection of images of different embryos, each of which is fixed at a different developmental time and imaged in a different rotational and translational orientation.
 %
@@
 %
-We will show that dimensionality reduction algorithms can accomplish both of these tasks to give us a meaningful picture of the underlying developmental dynamics.
+We will show that dimensionality reduction algorithms can simultaneously accomplish both of these tasks to give us a meaningful picture of the underlying developmental dynamics.
 
@@
 %
-Vector diffusion maps combine two algorithms, angular synchronization \cite{singer2011angular} for image registration and diffusion maps \cite{coifman2005geometric} for extracting low-dimensional structure, into one computation that allows us to simultaneously register and temporally order our images.
+Vector diffusion maps combine two algorithms, angular synchronization \cite{singer2011angular} for image registration and diffusion maps \cite{coifman2005geometric} for extracting low-dimensional structure in data, into one computation that allows us to simultaneously register and temporally order our images.
 %
 
-Vector diffusion maps has been used to classify images from cryo-electron microscopy (cryo-EM) experiments \cite{...}, where each image is a projection of a molecule, taken from a different viewing direction and in a different rotation.
+Vector diffusion maps has been used to classify images from cryo-electron microscopy (cryo-EM) experiments, where each image is a projection of a molecule, taken from a different viewing direction and in a different rotation.
 %
@@
 %
-Unlike the noise in the cryo-EM application, which is approximately Gaussian, the noise in our experiments is primarily from interembryo variability and does not have a known simple structure.
+Unlike the noise in the cryo-EM application, which is approximately Gaussian and arises from the measurement instrument, the noise in our experiments arises primarily from interembryo variability and does not have a known simple structure.
 
@@
 %
-Using the alignment angles between all pairs of vectors (or images), angular synchronization then finds the set of rotation angles, one angle for each vector, that is most consistent with the pairwise measurements.
+Using the alignment angles between all pairs of vectors (or images), angular synchronization then finds the set of rotation angles, one angle for each vector, that is most consistent with the pairwise measurements (see {\it SI Appendix}).
 %
@@
 %
-However, angular synchronization can successfully register sets of vectors (or images) even when many of the pairwise measurements are corrupted by noise. 
+However, the algorithm can successfully register sets of vectors (or images) even when many of the pairwise measurements are corrupted by noise. 
 %
@@
 %
-After removing the variability due to translations and rotations of the images, the remaining variability may be due to developmental dynamics.
+In the data sets we will consider, after removing the variability due to translations and rotations of the images, the remaining variability may be due to developmental dynamics.
 % 
@@
 %
-This idea of locality is illustrated by the edges in Figure~\ref{subfig:dmaps1}; data points which are close are connected by dark edges, and clearly, the dark edges are more informative about the low-dimensional structure of the data.
+This idea of locality is denoted by the edges in Figure~\ref{subfig:dmaps1}.
+%
+Data points which are close are connected by dark edges, and clearly, the dark edges are more informative about the low-dimensional structure of the data. 
 %
@@
 %
-Instead, we will rely on diffusion maps to automatically extract a parameterization of our high-dimensional data.
+Instead, we will rely on diffusion maps to automatically uncover a parameterization of our high-dimensional data (see {\it SI Appendix}).
 %
+Diffusion maps generalizes from one-dimensional nonlinear curves to higher-dimensional manifolds. 
+%
+When our experiments have additional sources of variability (such as mutations) in addition to developmental dynamics, diffusion maps can uncover both of these variables and provide an informative organization of our data.
 %We will assume that our data are approximately one-dimensional, and that this dimension is parameterized by time, such that ordering our data along this main dimension/direction will temporally order our data.
@@
 Vector diffusion maps combines both steps into one computation;
-the algorithm uses information about pairwise alignments and pairwise distances to find a global parameterization of the data, while registering images which are ``close''.
+the algorithm uses information about pairwise alignments and pairwise distances to find a global parameterization of the data, while registering images which are ``close'' (see {\it SI Appendix}).
 %
-The algorithm is perfectly suited for our task of registering and temporally ordering cross-sectional imaging data; we will use it to uncover a temporal ordering of our images, as well as register images which are developmentally similar, to obtain a global picture of development.
+The algorithm is perfectly suited for our task of registering and temporally ordering cross-sectional imaging data.
+%
+We will use it to uncover a temporal ordering of our images, as well as register images which are similar, to obtain a global picture of developmental dynamics.
 
@@
 \caption{Schematic illustrating angular synchronization and diffusion maps.
-{\it (A)} Set of vectors, each in a different orientation.
+{\xfigtextfontit (A)} Set of vectors, each in a different orientation.
 The pairwise alignment angles are indicated.
-{\it (B)} The vectors from {\it A}, each rotated so that the set is globally aligned.
-Note that the chosen rotation angles are consistent with the pairwise alignments in {\it A}.
-{\it (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
-Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\it SI Appendix}), so that close data points are connected by dark (``stronger'') edges.
-{\it (D)} The data in {\it C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure, $\phi_2$.
+{\xfigtextfontit (B)} The vectors from {\xfigtextfontit A}, each rotated about their midpoint so that the set is globally aligned.
+Note that the chosen rotation angles are consistent with the pairwise alignments in {\xfigtextfontit A}: the differences between a pair of angles in {\xfigtextfontit B} is the same as the pairwise angle in {\xfigtextfontit A}.
+{\xfigtextfontit (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
+Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\xfigtextfontit SI Appendix}), so that close data points are connected by dark (``stronger'') edges.
+{\xfigtextfontit (D)} The data in {\xfigtextfontit C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure.
 The color intensity is monotonic with the arclength, thus parameterizing the curve.}
@@
 %\includegraphics[width=8.4cm]{average_trajectory_VDM}
-\caption{{\it (A)} Images of {\it Drosophila} embryos during cellularization.
-Each image is of a different embryo in a different rotational orientation.
-{\it (B)} Images from {\it A}, now registered and ordered using vector diffusion maps.
+\caption{{\xfigtextfontit Drosophila} during cellularization.
+{\xfigtextfontit (A)} Images of {\xfigtextfontit Drosophila} embryos collected during cellularization.
+Each image is of a different embryo in a different rotational and translational orientation.
+{\xfigtextfontit (B)} Images from {\xfigtextfontit A}, now registered and ordered using vector diffusion maps.
 }%{\it (C)} Average trajectory of the images shown in {\it B}.
@@
 
-\subsection{Images of {\it Drosophila} embryos during cellularization}
+\subsection{Images of {\subsectionitfont Drosophila} embryos during cellularization}
 
-%
 As a first illustration of this approach to temporal ordering and registration of images, we used a dataset where the correct registration and temporal order is known {\it a priori}. 
@@
 By this stage of development, sequential divisions of the fertilized egg have generated a system where most of the nuclei are arranged in a monolayer under the common plasma membrane \cite{foe1983studies}.
-{\bf TODO: We need to have a supplementary figure that shows the lateral view, that makes it even more clear that we work with optical sections.}
 %
@@
 %
-During the 3rd hour of development, the spatially uniform and two-dimensional arrangement of nuclei is patterned by the concentration profiles of chemical signals that control cell differentiation.
+During the third hour of development, the spatially uniform two-dimensional arrangement of nuclei is patterned by the concentration profiles of chemical signals that control cell differentiation.
 %
-One of these signals is provided by the nuclear localization of transcription factor Dorsal (Dl, green), which subdivides the early embryo into the territories that give rise to the muscle, nerve, and skin tissues.
+One of these signals is provided by the nuclear localization of the transcription factor Dorsal (Dl, green), which subdivides the early embryo into the territories that give rise to the muscle, nerve, and skin tissues.
 %
@@
 
-Previous studies \cite{rushlow2012temporal} established that throughout the 3rd hour of development, the nuclear localization of Dl is unimodal and peaked at the ventral side of the embryo. 
+Previous studies \cite{rushlow2012temporal} established that throughout the third hour of development, the nuclear localization of Dl is unimodal and peaked at the ventral side of the embryo (see {\it SI Appendix}).
 %
-The spatiotemporal pattern of dpERK is more complex: dpERK is first detected in two lateral stripes, which increase in intensity towards the end of the 3rd hour \cite{lim2013kinetics}.  
+The spatiotemporal pattern of dpERK is more complex: dpERK is first detected in two lateral stripes, which increase in intensity towards the end of the third hour \cite{lim2013kinetics}.  
 %
@@
 %
-Both of these features become readily apparent when images are registered and ordered using vector diffusion maps: the peak of nuclear Dl is consistent throughout the images and intensity of the dpERK signal in the lateral regions of the embryo monotonically increases across the dataset.
+Both of these features become readily apparent when images are registered and ordered using vector diffusion maps (see Figure~\ref{subfig:ordered_data1}): the peak of nuclear Dl is consistent throughout the images and intensity of the dpERK signal in the lateral regions of the embryo monotonically increases across the dataset.
 %
@@
 %
-Specifically, during the 3rd hour of development, the embryo cellularizes. 
-%
-In this process, nuclei located under common plasma membrane are enclosed by lateral membranes which grow inward, enclosing nuclei into individual cells.
+During the third hour of development, the embryo cellularizes, where nuclei located under common plasma membrane are enclosed by lateral membranes which grow inward, separating the nuclei into individual cells.
 %
@@
 %
-The rank correlation coefficient between the orderings estimated based on the progress of cellularization and vector diffusion maps is 0.9127. 
+The rank correlation coefficient between the orderings obtained based on the progress of cellularization and vector diffusion maps is 0.9127. 
 %
@@
 
-{\bf TODO: add something about size variability}
+Although the first vector diffusion maps coordinate accurately captures the developmental dynamics, there are other sources of variability within this data set that could also be extracted using our techniques.
+%
+For example, the images contain size variability. 
+%
+In this example, the size variation is relatively small relative to the developmental changes. 
+%
+However, in other applications where the size variation is significant or the developmental dynamics are more subtle, it may be important to factor out size or scaling symmetries.
 
@@
 
-\subsection{Images of gastrulating {\it Drosophila} embryos}
+\subsection{Images of gastrulating  {\subsectionitfont Drosophila} embryos}
 
@@
 %
-However, developmental dynamics are often significantly more complex, and nonlinear techniques such as diffusion maps/vector diffusion maps are required to extract meaningful structure in the data. 
+However, developmental dynamics are often significantly more complex, and nonlinear techniques such as (vector) diffusion maps are required to extract meaningful structure in the data. 
 %
@@
 %
-These images cover a thirty minute time interval and span late cellularization (the early images in this data set overlap with the late images in the previous set) through gastrulation, when the two-dimensional sheet of newly formed cells begins to deform on a way to forming the three-dimensional structures of the future organs.
+These images cover a thirty minute time interval and span late cellularization (the early images in this data set overlap with the late images in the previous set) through gastrulation, when the two-dimensional sheet of newly formed cells begins to deform into three-dimensional structures of the future organs.
 %
@@
 %
-Instead of Dl, which is difficult to visualize at this point, we used Twist (Twi, shown in green), a transcription factor that is activated by high levels of Dl, specifying the cells of the future muscle tissue.
+Instead of Dl, which is difficult to visualize at this point, we stained for Twist (Twi, shown in green), a transcription factor that is activated by high levels of Dl, specifying the cells of the future muscle tissue.
 %
@@
 %
-At the end of this process, cells which were originally on the ventral side of the embryo have migrated to the dorsal side of the embryo, causing similar ``omega'' shaped patterns (formed by Twi and dpERK) to arise on the dorsal side.
+At the end of this process, cells which were originally on the ventral side have migrated to the dorsal side of the embryo, causing similar ``omega'' shaped patterns (formed by Twi and dpERK) to arise on the dorsal side.
 
-The dynamics in this data set are highly nonlinear, as only 22\% of the variability in the data is captured by the first principal component.
+\begin{figure*}[t]
+\raisebox{5.3cm}{{\figtextfont A}}
+\includegraphics[width=16.8cm]{raw_data2}
+
+\vspace{0.2cm}
+\raisebox{5.3cm}{{\figtextfont B}}
+\includegraphics[width=16.8cm]{VDM_ordered}
+
+\vspace{0.2cm}
+\raisebox{0.6cm}{{\figtextfont C}}
+\includegraphics[width=16.8cm]{average_trajectory}
+\caption{{\xfigtextfontit Drosophila} during gastrulation.
+{\xfigtextfontit (A)} Images of gastrulating {\xfigtextfontit Drosophila} embryos.
+Each image is of a different embryo in a different rotational and translational orientation.
+{\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps.
+{\xfigtextfontit (C)} Average trajectory of the images shown in {\xfigtextfontit B}.
+Each image is the average of $9$ consecutive images in {\xfigtextfontit B}.
+}
+\customlabel{fig:data2}{4}
+\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
+\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
+\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
+\end{figure*}
+
+The dynamics in this data set are highly nonlinear, and only 22\% of the variability in the data is captured by the first principal component.
 %
@@
 %
-This is done by dividing the registered and ordered data into windows or classes (each window contains a group of consecutive images in the ordered data), and then calculate the average image in each window.
+This is done by dividing the registered and ordered data into windows or classes, where each window contains a group of consecutive images in the ordered data, then calculating the average image in each window.
 %
-Figure~\ref{subfig:averaged_data2} shows the average trajectory for our data set; it is now very straightforward to see the underlying developmental dynamics.
+Figure~\ref{subfig:averaged_data2} shows the average trajectory for this data set; it is now very straightforward to see the underlying developmental dynamics.
 %
-We therefore accomplished the same task presented in the caricature in Figure~\ref{fig:fish}, but without knowledge about appropriate landmarks and or the appropriate statistic (e.g., body size) by which to order.
-
-%\newpage
-\begin{figure*}[t]
-\raisebox{5.3cm}{{\figtextfont A}}
-\includegraphics[width=16.8cm]{raw_data2}
-
-\vspace{0.2cm}
-\raisebox{5.3cm}{{\figtextfont B}}
-\includegraphics[width=16.8cm]{VDM_ordered}
-
-\vspace{0.2cm}
-\raisebox{0.6cm}{{\figtextfont C}}
-\includegraphics[width=16.8cm]{average_trajectory}
-\caption{{\it (A)} Images of gastrulating {\it Drosophila} embryos.
-Each image is of a different embryo in a different rotational orientation.
-{\it(B)} Data from {\it A}, registered and ordered using vector diffusion maps.
-{\it (C)} Average trajectory of the images shown in {\it B}.
-Each image is the average of $9$ consecutive images in {\it B}.
-}
-\customlabel{fig:data2}{4}
-\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
-\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
-\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
-\end{figure*}
+In this example, we accomplished the same task presented in the caricature in Figure~\ref{fig:fish}, but without knowledge about the appropriate landmarks (e.g., fins) for registration or the appropriate statistic (e.g., body size) by which to order.
 
@@
 
-Vector diffusion maps not only allows us temporally order data, but also allows us to uncover other sources of variability.
+Vector diffusion maps can not only temporally order data, but can also uncover other sources of variability.
 %
-Until now, we have been assuming that our experiments contain one main direction of variability, and that this direction is parameterized by time.
+Up to this point, we have assumed that our experiments contain one main direction of variability, and that this direction is parameterized by time.
 %
-However, this is not alw
+However, this is not always the case.
+%
+As an example, Figure~\ref{subfig:raw_data3} shows a data set consisting of $42$ embryos undergoing cellularization during the third hour of development. 
+%
+Again, each image is an optical cross-section of an embryo fixed at a different developmental time, and the embryos have been stained for nuclei (gray), Dl (green), and dpERK (red).
+%
+In this data set, $21$ of the embryos are wild type, and $21$ of the embryos are an sna null mutant.
+%
+We therefore expect two sources of variability in this data set: differences due to developmental time, and differences between mutant and wild type dynamics.
+
+Figure~\ref{subfig:projection_data3} shows the data, projected onto the first two vector diffusion maps coordinates;
+each data point corresponds to one image in our data set.
+%
+For the previous examples, where we only uncovered developmental progression, we used the first vector diffusion maps coordinate (i.e., the x-axis of this plot) to order the data. 
+%
+However, our data now contains two main sources of variation, and we can see two distinct trajectories, highlighted in blue and red, emerge in this two-dimensional progression.
+%
+Both curves originate from a common point, and then diverge.
+
+We can compute the average trajectory over each of the two curves to gain insight into the different developmental dynamics.
+%
+The average over the blue curve, shown in Figure~\ref{subfig:average_wt_data3}, shows the emergence of two distinct dpERK peaks, as expected.
+%
+The average over the red curve, shown in Figure~\ref{subfig:average_mut_data3}, reveals one more concentrated dpERK peak.
+%
+This is consistent with previously known dynamics.
+%
+At early times, the wild type and mutant embryos both have little dpERK signal, and are therefore indistinguishable, whereas at later times, the two groups separate as the wild type embryos develop two distinct dpERK peaks and sna null mutants develop only a single peak.
 
 \begin{figure*}[t]
-\raisebox{8.1cm}{{\figtextfont A}}
-\includegraphics[height=8.4cm]{raw_data3}
-\hspace{0.5cm}
-\raisebox{8.1cm}{{\figtextfont B}}
-\includegraphics[width=8.4cm]{mutants_VDM2}
-\caption{{\it (A)} Images of cellularizing {\it Drosophila} embryos.
+\raisebox{8.8cm}{{\figtextfont A}}
+\includegraphics[height=9cm]{raw_data3}
+\begin{minipage}[b]{8.5cm}
+\raisebox{5.3cm}{{\figtextfont B}}
+\includegraphics[width=8.4cm]{mut_wt_vdm_embedding}
+\raisebox{1.3cm}{{\figtextfont C}}
+\includegraphics[width=8.4cm]{wt_trajectory}
+\raisebox{1.3cm}{{\figtextfont D}}
+\includegraphics[width=8.4cm]{mut_trajectory}
+\end{minipage}
+\caption{Mutant and wild type {\xfigtextfontit Drosophila}.
+{\xfigtextfontit (A)} Images of cellularizing {\xfigtextfontit Drosophila} embryos.
 Each image is of a different embryo in a different rotational orientation.
-$19$ of the images are of wild type embryos, and $23$ of the images are of sna null mutant embryos.
-{\it(B)} Data from {\it A}, organized using the first two vector diffusion maps coordinates.}
+Half of the images are of wild type embryos, and half of the images are of sna null mutant embryos.
+{\xfigtextfontit(B)} Data from {\xfigtextfontit A}, projected onto the first two vector diffusion maps coordinates.
+The two distinct trajectories are highlighted in blue and red.
+{\xfigtextfontit (C)} Average trajectory of the dpERK signal over the blue curve in {\xfigtextfontit B}.
+{\xfigtextfontit (D)} Average trajectory of the dpERK signal over the red curve in {\xfigtextfontit B}.}
 \customlabel{fig:data3}{5}
 \customlabel{subfig:raw_data3}{\ref{fig:data3}{\it A}}
-\customlabel{subfig:ordered_data3}{\ref{fig:data3}{\it B}}
+\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it B}}
+\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it C}}
+\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it D}}
 \end{figure*}
@@
 %
-We chose the algorithm outlined here, not because it is specifically tailored to our particular data sets, but rather, because it is sufficiently general and we are confident that they can be applied to a myriad of biological imaging applications.
+We chose the algorithm outlined here, not because it is specifically tailored to our particular data sets, but rather, because it is sufficiently general and we are confident that it can be applied to a myriad of biological imaging applications.
 
@@
 %
-However, most of these applications rely on the definition and identification of good {\it features} within each image. 
+However, most of these applications rely on the definition and identification of good landmarks or features within each image. 
 %
@@
 %
-Because of complexity of chemical and morphological changes in developing tissues, good features may not be known {\it a priori}.
+Because of the complexity of chemical and morphological changes in developing tissues, appropriate features may not be known {\it a priori}.
 %
@@
 %
-Data sets such as  transcriptional profiling data (where each data point is the transcriptome for a single sample, such as a cell or tissue sample from a patient) \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering}, and snapshots of cells throughout the cell cycle (where each point is a vector of features, such as the amount of DNA, which quantify the cell's state)  \cite{kafri2013dynamics} have all been temporally ordered to extract a representative trajectory of the underlying dynamic process.
+Data sets such as transcriptional profiling data (where each data point is the transcriptome for a single sample, such as a cell or tissue sample from a patient) \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering} and snapshots of cells throughout the cell cycle (where each point is a vector of features, such as the amount of DNA, which quantify the cell's state)  \cite{kafri2013dynamics} have been temporally ordered to extract representative trajectories of the underlying dynamic processes.
 %
 Most of these applications order the data by solving a traveling salesman problem (TSP) or constructing a minimum spanning tree (MST) on the data,
-expecting that these structures (the path of the traveling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate encoding of the dynamical progression.
+expecting that these structures (the path of the traveling salesman, or the minimum spanning tree) characterize the majority of the patterns and variability within the data and provide an accurate encoding of the dynamical progression.
 %
@@
 %
-In these approaches, PCA serves two purposes: it denoises the data, making the TSP/MST construction more robust, and allows for easy visualization of the TSP/MST if one only uses the first two or three principal components.
+In these approaches, PCA serves two purposes: it denoises the data, making the TSP/MST construction more robust, and also allows for easy visualization of the TSP/MST if one only uses the first two or three principal components.
 
@@
 
-Furthermore, dimensionality reduction techniques such as vector diffusion maps do not constrain us to one-dimensional parameterizations/organizations of the data.
+Furthermore, dimensionality reduction techniques such as vector diffusion maps do not constrain us to one-dimensional parameterizations/organizations of the data, as illustrated in the mutant data set example (Figure~\ref{fig:data3}).
 %
-For this particular application, we expect our data to be one-dimensional, as the only (main) source of variability is time.
+Other sources of variability, such as size or viewing angle, could also be uncovered using vector diffusion maps.
 %
-However, other sources of variability, such as size, viewing angle, and mutation state, could also be uncovered using vector diffusion maps.
-%
-We could therefore use these same algorithms for tasks such as classification of embryos by size or viewing angle, and comparison of the developmental dynamics in different mutations.
+These techniques would also allow us to compare the developmental dynamics of several different mutations to see how mutations are developmentally related.
 
-{\bf TODO: add about how we can correlate physical variables with dmaps variables.}
-%
 After registering and temporally ordering images, one can perform a variety of post-processing tasks to gain further insight into the developmental processes.
@@
 %
-Constructing a smooth movie from the registered and ordered cross-sectional snapshots would also be of biological interest.
+Constructing a smooth movie from the registered and ordered cross-sectional snapshots would also be of biological interest, and
+examining the top principal components in each window would allow one to visualize the main dynamical features along the trajectory.
 %
-Furthermore, examining the top principal components in each window would allow one to visualize the main dynamical features along the trajectory.
+These questions first require that the data be registered and temporally ordered.
 %
-All of these questions first require that the data be registered and temporally ordered.
+Correlating vector diffusion maps coordinates with physical variables is another post-processing task that can give physical insight into developmental dynamics.
 %
-We are confident that, as biological becomes easier and more robust, the techniques we present here will play an important role in the data analysis pipeline.
+One can test which physical variable, among a set of candidate physical variables (e.g., total intensity of one signal, roundness of the embryo), is most correlated with the uncovered vector diffusion maps coordinate to help understand which physical quantities change across the developmental trajectory.
+%
+We are confident that, as biological becomes easier and more robust, the dimensionality reduction techniques we present here will play an important role in the data analysis pipeline.
 
@@
 
+%% == end of paper:
+
+%% Optional Materials and Methods Section
+%% The Materials and Methods section header will be added automatically.
+
+%% Enter any subheads and the Materials and Methods text below.
+
+\begin{materials}
+
+\end{materials}
+
+%% Optional Appendix or Appendices
+%% \appendix Appendix text...
+%% or, for appendix with title, use square brackets:
+%% \appendix[Appendix Title]
+
+\begin{acknowledgments}
+- text of acknowledgments here, including grant info -
+\end{acknowledgments}
+
+%% PNAS does not support submission of supporting .tex files such as BibTeX.
+%% Instead all references must be included in the article .tex document.
+%% If you currently use BibTeX, your bibliography is formed because the
+%% command \verb+\bibliography{}+ brings the <filename>.bbl file into your
+%% .tex document. To conform to PNAS requirements, copy the reference listings
+%% from your .bbl file and add them to the article .tex file, using the
+%% bibliography environment described above.
+
+%%  Contact pnas@nas.edu if you need assistance with your
+%%  bibliography.
+
+% Sample bibliography item in PNAS format:
+%% \bibitem{in-text reference} comma-separated author names up to 5,
+%% for more than 5 authors use first author last name et al. (year published)
+%% article title  {\it Journal Name} volume #: start page-end page.
+%% ie,
+% \bibitem{Neuhaus} Neuhaus J-M, Sitcher L, Meins F, Jr, Boller T (1991)
+% A short C-terminal sequence is necessary and sufficient for the
+% targeting of chitinases to the plant vacuole.
+% {\it Proc Natl Acad Sci USA} 88:10362-10366.
+
+%% Enter the largest bibliography number in the facing curly brackets
+%% following \begin{thebibliography}
+
 \bibliographystyle{pnas}
 \bibliography{background_reading/references,../../references/references}
+
+%\begin{thebibliography}{}
+
+%\end{thebibliography}
 

### 2014-05-17 image_analysis_paper/si_information.tex
all edits before sending to Yannis

@@
 
+\makeatletter
+\newcommand{\customlabel}[2]{%
+\protected@write \@auxout {}{\string \newlabel {#1}{{#2}{}}}}
+\makeatother
+
 \begin{document}
@@
 \subsection{Measuring developmental time during cellularization}
+
+Todo:
+\begin{itemize}
+\item We need to have a supplementary figure that shows the lateral view, that makes it even more clear that we work with optical sections.
+
+\item Need image labeled with D/V
+
+\end{itemize}
 
@@
 
-For the images shown in Figure~3 of the paper, we only use the red and green channels (dpERK and Dl) to register the images.
-
+Each image is subsampled to $100 \times 100$ pixels.
+%
+For the images shown in Figures~3~and~5 of the paper, we only use the red and green channels (dpERK and Dl) in the vector diffusion maps computations.
+%
+The nuclei are shown only for visual reference.
+%
 For the images shown in Figure~4 of the paper, we use all three channels (dpERK, Twi, and nuclei) to register the images.
 %
-We blur the nuclei using a disc filter with radius $5$~pixels.
+We blur the nuclear signal using a disc filter with a $5$~pixel radius (using the \texttt{imfilter} function in Matlab with a disc filter).
 %
-We also adjust the intensity of the nuclear signal such that 1\% of the pixels are saturated.
+We also adjust the intensity of the nuclear signal in each image such that 1\% of the pixels are saturated (using the \texttt{imadjust} function in Matlab).
 
-\section{PCA for gastrulating images}
+\section{Principal component analysis for gastrulating images}
+
+We performed principal component analysis on the images shown in Figure~4 of the paper.
+%
+We first registered the images using angular synchronization, and then computed the principal components for the data set.
+%
+We ordered the images by the projection onto the first principal component; the results are shown in Figure~\ref{fig:PCA_results}.
+%
+The dynamics that we uncover are not consistent with the expected biological progression.
+%
+Furthermore, the eigenvalue spectrum from PCA indicates that a significant amount of the variance is not captured by the first principal component, therefore motivating us to use nonlinear techniques.
 
@@
 \caption{{\it (A)} Images shown in Figure~4 of the paper, now ordered by the first PCA projection coefficient.
-Note that the ordering is not consistent with the developmental dynamics.
-{\it (B)} The eigenvalue spectrum from PCA.
-Note that subsequent components have a nonnegligible contribution on the data.}
+The recovered ordering is not consistent with the known developmental dynamics.
+{\it (B)} The fraction of the variance captured by each of the first 10 principal components.
+Note that components 2--10 have a nonnegligible contribution to the data.}
+\label{fig:PCA_results}
 \end{figure}
@@
 \section{Mathematical Algorithms}
+
+We will demonstrate our techniques for registration and temporal ordering using a synthetic data set.
+%
+The relatively simple dynamics of this data set will allow us to easily visualize and illustrate the main points of the different algorithms.
+%
+We construct concentration profiles defined on a ring, and we assume that each ring is arbitrarily rotated around its center; an example is shown in Figure~\ref{subfig:1d_example}.
+%
+Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown at the bottom of Figure~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
+%
+Each concentration profile is a noisy Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
+%
+We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_{100}$, the group of integers modulo $100$).
+%
+Figure \ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
+%
+Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.
 
@@
 \label{fig:1d_demo}
+%\customlabel{fig:1d_demo}{3}
+\customlabel{subfig:1d_example}{\ref{fig:1d_demo}{\it A}}
+\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}{\it B}}
+\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}{\it C}}
+\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}{\it D}}
+\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}{\it E}}
+\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}{\it F}}
 \end{figure}
@@
 discrete shift.
+
+Furthermore, this formulation also considers {\it higher-order} consistency information.
 %
-Details of registering two-dimensional images using synchronization are given in the {\it SI Appendix}.
-
-%Furthermore, this formulation also considers {\it higher-order} consistency information.
+For example, given our pairwise estimates $R_{ij}$, we know that relationships of the form
+\begin{equation} \label{eq:triplet_consistency}
+R(\theta_{ik}) R(\theta_{kj}) \approx R(\theta_i) R(\theta_k)^T R(\theta_k) R(\theta_j)^T = R(\theta_i) R(\theta_j)^T
+\end{equation}
+should also hold.
 %
-%For example, given our pairwise estimates $R_{ij}$, we know that relationships of the form
-%\begin{equation} \label{eq:triplet_consistency}
-%R(\theta_{ik}) R(\theta_{kj}) \approx R(\theta_i) R(\theta_k)^T R(\theta_k) R(\theta_j)^T = R(\theta_i) R(\theta_j)^T
-%\end{equation}
-%should also hold.
+Note that
+\begin{equation}
+(H^2)_{ij} = \sum_k R(\theta_{ik}) R(\theta_{kj});
+\end{equation}
+therefore, {\it all} information of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$.
 %
-%Note that
-%\begin{equation}
-%(H^2)_{ij} = \sum_k R(\theta_{ik}) R(\theta_{kj});
-%\end{equation}
-%therefore, {\it all} infomation of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$.
-%
-%Because $H$ and $H^2$ have the same eigenvectors, our problem formulation accounts for not only pairwise alignment information, but also these higher-order considerations.
+Because $H$ and $H^2$ have the same eigenvectors, our problem formulation accounts for not only pairwise alignment information, but also these higher-order considerations.
 
@@
 %
-%Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors.
+Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors.
 %
@@
 %
-The procedure generalizes when the data lie on higher-dimensional manifolds (not just curves) in data space.
+The procedure generalizes when the data lie on higher-dimensional manifolds (not just curves) in data space, where lower eigenvectors give subsequent embedding coordinates for the data.
 
@@
 %
-In general, the embedding coordinates will be given by
+
+If we assume that the rotations and the dynamics are uncoupled and therefore separable, then the eigenvectors of $S$ have the following structure: each block eigenvector contains estimates of the optimal rotations multiplied by the corresponding embedding coordinate.
+%
+As the first diffusion maps coordinate is constant, the first block eigenvector contains only the optimal rotations.
+%
+The second block eigenvector (eigenvectors $d+1$ through $2d$) contains the optimal rotations, each multiplied by their second diffusion maps coordinate. 
+%
+We can therefore recover this diffusion maps coordinate by taking inner products of the columns of the second block eigenvector with columns of the first block eigenvector.
+%
+As rotation matrices are orthonormal, many of these will be zero, 
+and the nonzero inner products give us the embedding coordinates.
+%
+More specifically, the $j^{th}$ embedding coordinate will be given by
 \begin{equation}
-\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle
+\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle, \: 0 < k \le d, (j-1)d < l \le jd
 \end{equation}
-where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$.
+where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$, 
+and we select values of $k$ and $l$ such that the coordinate $\psi_{k, l}$ has the largest variability. 
 %
-If we assume that our data mainly vary along a one-dimensional manifold, and that the rotations and the dynamics are uncoupled and therefore separable, one expects the embedding coordinate for our data %(i.e., the analog of $\phi_2$ from the diffusion maps case)
-to be  $\psi_{1,d+1}$.
+
+\subsection{Registering images with respect to translations and rotations} \label{subsec:trans_rot_register}
+
+\begin{figure*}
+\begin{minipage}{0.2\textwidth}
+\centering
+\includegraphics[width=\textwidth]{sphere_1}\\
+\includegraphics[width=\textwidth]{sphere2_1}\\
+A
+\end{minipage}
 %
-Our example is, by physical considerations, a separable one, and so sorting our data by $\psi_{1,d+1}$ will order our data in time.
-%
-If no separability claims can be made, a second diffusion map step, using $\psi_{k,l}$ as the coordinates, may uncover a
-useful parameterization of the data manifold modulo symmetries.
-
-\section{Registering images with respect to translations and rotations} \label{subsec:trans_rot_register}
+\begin{minipage}{0.2\textwidth}
+\centering
+\includegraphics[width=\textwidth]{sphere_2}\\
+\includegraphics[width=\textwidth]{sphere2_2}\\
+B
+\end{minipage}
+\begin{minipage}{0.2\textwidth}
+\centering
+\includegraphics[width=\textwidth]{sphere_3}\\
+\includegraphics[width=\textwidth]{sphere2_3}\\
+C
+\end{minipage}
+\begin{minipage}{0.2\textwidth} 
+\centering
+\includegraphics[width=\textwidth]{sphere_4}\\
+\includegraphics[width=\textwidth]{sphere2_4}\\
+D
+\end{minipage}
+\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions.
+{\it (A)} The original image.
+{\it (B)} Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image.
+{\it (C)} Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation.
+{\it (D)} Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.
+The top row shows the three-dimensional spheres, and the bottom row shows the (two-dimensional) surface of the sphere in which we are interested.}
+\label{fig:SO3_picture}
+\end{figure*}
 
@@
 %
-The norm, $\| \cdot \|$, is the Euclidean norm between the pixel intensities for both the red and green channels.
+The norm, $\| \cdot \|$, is the Euclidean norm between the pixel intensities of the channels. 
 %
@@
 Rotation of the image corresponds to rotation around one principal axis in three dimensions, while translation of the image corresponds (approximately) to rotations around the other two principal axes.
-%(Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$.
+(Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$.
 However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.)
 %
-%These rotation matrices are (by definition) orthogonal, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
+These rotation matrices are (by definition) orthogonal, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
 

### 2014-05-17 image_analysis_paper/test_mutants_big.m

@@ -16,6 +16,9 @@ mut = [7 11 13 15 16 17 18 20 21 29 32 33 34 36 37 38 40 41 48 49 ...
     52 53 54 55 56 57 60 61 62 63 68 71 72 73 74 81 82 83 84 85 ...
     86 87 90 93 96 97 98]; %47
 
+mutation = true(m, 1);
+mutation(wt) = false;
+
 ind = 1:m;
 
 %% image parameters
@@ -48,7 +51,8 @@ for i=1:m
     nuclei(:, :, i) = im1(:, :, 1);
     
     im1(:, :, 1) = im1(:, :, 2);
-    im1(:, :, 2) = uint8(0.5*double(im1(:, :, 3)));
+    %im1(:, :, 2) = uint8(0.5*double(im1(:, :, 3)));
+    im1(:, :, 2) = 0;
     im1(:, :, 3) = 0;
     
     %store image
@@ -58,7 +62,6 @@ for i=1:m
     imshow(im1);
 end
 
-
 %% raw dpERK images-- synchronization
 
 angle_proj = pi/8;
@@ -66,7 +69,7 @@ shift_max = 10;
 shift_step = 2;
 dim = 3;
 
-matlabpool open 4;
+matlabpool open 2;
 [R, W] = compute_pairwise_alignments_color(image_set, angle_proj, shift_max, shift_step);
 matlabpool close
 
@@ -78,19 +81,7 @@ matlabpool close
 
 ind = setdiff(1:m, [17]);
 
-wt = [1 2 3 4 5 6 8 9 10 12 14 19 22 23 24 25 26 27 28 30 ...
-    31 35 39 42 43 44 45 46 47 50 51 58 59 64 65 66 67 69 70 75 ...
-    76 77 78 79 80 88 89 91 92 94 95]; %51
-
-mut = [7 11 13 15 16 18 20 21 29 32 33 34 36 37 38 40 41 48 49 ...
-    52 53 54 55 56 57 60 61 62 63 68 71 72 73 74 81 82 83 84 85 ...
-    86 87 90 93 96 97 98]; %47
-
-i = find(wt > 17);
-wt(i) = wt(i) - 1;
-i = find(mut > 17);
-mut(i) = mut(i) - 1;
-
+mutation = mutation(ind);
 
 image_set = image_set(:, :, :, ind);
 image_set_raw = image_set_raw(:, :, :, ind);
[…]

### 2014-05-19 image_analysis_paper/pnas_draft2.tex
movie code, edited mutant code, etc.

@@
 %
-We demonstrate these techniques using three data sets from a study of {\it Dropophila} embryogenesis, one of the best experimental models for studies of developmental dynamics.
+We demonstrate these techniques using three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics.
 %
@@
 %
-However, in other applications where the size variation is significant or the developmental dynamics are more subtle, it may be important to factor out size or scaling symmetries.
+However, in other applications where the size variation is significant or the developmental dynamics are more subtle, it may be important to factor out size or scaling symmetries, or consider additional components.
 
@@
 %
-At the end of this process, cells which were originally on the ventral side have migrated to the dorsal side of the embryo, causing similar ``omega'' shaped patterns (formed by Twi and dpERK) to arise on the dorsal side.
+At the end of this process, cells which were originally on the ventral side have migrated to the dorsal side of the embryo, causing similar ``omega'' shaped patterns (formed by Twi and dpERK) to arise on the dorsal side (shown in the late images in Figure~\ref{subfig:averaged_data2}).
 
@@
 %
-In this data set, $21$ of the embryos are wild type, and $21$ of the embryos are an sna null mutant.
+In this data set, $21$ of the embryos are wild type, and $21$ of the embryos are mutant embryos where dpERK is expressed in one broad peak at the ventral side, instead of in two bimodal peaks.
 %
@@
 %
-At early times, the wild type and mutant embryos both have little dpERK signal, and are therefore indistinguishable, whereas at later times, the two groups separate as the wild type embryos develop two distinct dpERK peaks and sna null mutants develop only a single peak.
+At early times, the wild type and mutant embryos both have little dpERK signal, and are therefore indistinguishable, whereas at later times, the two groups separate as the wild type embryos develop two distinct dpERK peaks and the mutant embryos develop only a single peak.
+%
+We have therefore demonstrated that vector diffusion maps can not only recover dynamics, but also distinguish different expression patterns and developmental trajectories.
 
-\begin{figure*}[t]
-\raisebox{8.8cm}{{\figtextfont A}}
-\includegraphics[height=9cm]{raw_data3}
-\begin{minipage}[b]{8.5cm}
-\raisebox{5.3cm}{{\figtextfont B}}
+\begin{figure}[t]
+%\raisebox{8.8cm}{{\figtextfont A}}
+%\includegraphics[height=9cm]{raw_data3}
+%\begin{minipage}[b]{8.5cm}
+\raisebox{6cm}{{\figtextfont B}}
 \includegraphics[width=8.4cm]{mut_wt_vdm_embedding}
-\raisebox{1.3cm}{{\figtextfont C}}
+\raisebox{2cm}{{\figtextfont C}}
 \includegraphics[width=8.4cm]{wt_trajectory}
-\raisebox{1.3cm}{{\figtextfont D}}
+\raisebox{2cm}{{\figtextfont D}}
 \includegraphics[width=8.4cm]{mut_trajectory}
-\end{minipage}
+%\end{minipage}
 \caption{Mutant and wild type {\xfigtextfontit Drosophila}.
@@
 \customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it D}}
-\end{figure*}
+\end{figure}
 
@@
 
-\bibliographystyle{pnas}
-\bibliography{background_reading/references,../../references/references}
+%\bibliographystyle{pnas}
+%\bibliography{background_reading/references,../../references/references}
 

### 2014-05-19 image_analysis_paper/test_late_images.m

@@ -33,11 +33,11 @@ for i=1:m
     
     % resize image
     im1 = imresize(im1, [npixels npixels]);
-
+    
     %     blur image
     H = fspecial('disk',5);
     im1(:, :, 1) = imfilter(im1(:, :, 1),H,'replicate');
-    im1(:, :, 1) = 0.5*imadjust(im1(:, :, 1));    
+    im1(:, :, 1) = 0.5*imadjust(im1(:, :, 1));
     
     im1(:, :, 3) = 1.5 * im1(:, :, 3);
     
@@ -206,7 +206,7 @@ set(gcf, 'papersize', [8 8*subplot_dim2/subplot_dim1])
 set(gcf, 'paperposition',[0 0 8 8*subplot_dim2/subplot_dim1])
 for i=1:m
     im1 = image_set_aligned(:,:,:,I(i));
-        im1(:, :, 1) = im1(:, :, 1)  + im1(:, :, 3);
+    im1(:, :, 1) = im1(:, :, 1)  + im1(:, :, 3);
     im1(:, :, 2) = im1(:, :, 2)  + im1(:, :, 3);
     %subplot(subplot_dim1, subplot_dim2, i);
     %subplot('position', [X(i)-1/subplot_dim1 Y(i)-1/subplot_dim2 1/subplot_dim1-0.01 1/subplot_dim2-0.01])
@@ -307,7 +307,7 @@ for i=1:m
     
     imshow(im1);
 end
-saveas(gcf,sprintf('%s/VDM_ordered', im_save_dir), 'pdf')
+%saveas(gcf,sprintf('%s/VDM_ordered', im_save_dir), 'pdf')
 
 %%
 
@@ -331,3 +331,117 @@ end
 saveas(gcf,sprintf('%s/average_trajectory', im_save_dir), 'pdf')
 
 
+%%
+
+% window_delta = 5;
+% window_eps = 5;
+% window_weights = exp(-(-window_delta:window_delta).^2/window_eps);
+%
+% writerObj = VideoWriter('gastrulation.avi');
+% writerObj.FrameRate = 15;
+% open(writerObj);
+%
+% i = window_delta + 1;
+% stage_indices = I(i-window_delta:i+window_delta);
+% im1 = window_weights(1) * double(image_set_raw_aligned(:,:,:,stage_indices(1)));
+% for j=2:window_delta+1
[…]

### 2014-05-19 image_analysis_paper/test_mutants.m

@@ -296,9 +296,134 @@ figure;
 plot(embed_coord(:,idx1),embed_coord(:,idx2),'.')
 
 
-% [cluster_idx, C, SUMD] = kmeans(embed_coord(:, [idx1 idx2]), 3);
-% figure;
-% scatter(embed_coord(:,idx1),embed_coord(:,idx2),50, cluster_idx, '.')
+
+%%
+figure;
+set(gcf, 'paperunits', 'centimeters')
+set(gcf, 'papersize', [8 6])
+set(gcf, 'paperposition',[0 0 8 6])
+coord2_scale = 0.6;
+
+plot(embed_coord(~mutation,idx1),coord2_scale*embed_coord(~mutation,idx2),'xk', 'markersize', 5)
+hold on
+plot(embed_coord(mutation,idx1),coord2_scale*embed_coord(mutation,idx2),'ok', 'markersize', 5)
+hold on
+mut_draw = [2 37 5 22 27];
+wt_draw = [19 7 10 9 34];
+im_delta = 0.004;
+draw_delta = 0.008;
+hold on
+plot(embed_coord(mut_draw,idx1),coord2_scale*embed_coord(mut_draw,idx2),'ob', 'markersize', 5, 'linewidth', 1)
+
+plot(embed_coord(wt_draw,idx1),coord2_scale*embed_coord(wt_draw,idx2),'xb', 'markersize', 5, 'linewidth', 1)
+for i=mut_draw
+   image('cdata', image_set_aligned(:, :, :, i), 'xdata', [embed_coord(i,idx1)-sign(embed_coord(i,idx1)-0.01)*draw_delta-im_delta embed_coord(i,idx1)-sign(embed_coord(i,idx1)-0.01)*draw_delta+im_delta], 'ydata', [coord2_scale*embed_coord(i,idx2)-draw_delta-im_delta coord2_scale*embed_coord(i,idx2)-draw_delta+im_delta])
+end
+for i=wt_draw
+   image('cdata', image_set_aligned(:, :, :, i), 'xdata', [embed_coord(i,idx1)-draw_delta-im_delta embed_coord(i,idx1)-draw_delta+im_delta], 'ydata', [coord2_scale*embed_coord(i,idx2)-im_delta coord2_scale*embed_coord(i,idx2)+im_delta])
+end
+axis equal
+%axis([-0.025 0.055 -0.07 0.07])
+set(gca, 'xtick', [])
+set(gca, 'ytick', [])
+
[…]

### 2014-05-20 image_analysis_paper/pnas_draft2.tex
edits before meeting with Yannis

@@
 \graphicspath{ {paper_figures2/} }
+
+\usepackage{tikz}
 
@@
 %
-Based on manifold learning algorithms for high-dimensional data (such as images), these techniques have been used for data mining in a wide range of applications, from cryo-EM analysis of individual molecules to face recognition.
+Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, as have dimensionality reduction techniques \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear} and their applications \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering, kafri2013dynamics}.
 %
-Here we adapt these techniques for temporal ordering and registration of imaging data from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
+Here, we use a manifold learning algorithm \cite{singer2012vector} for high-dimensional data (such as images) which both factors out symmetries to align images and extracts low-dimensional structure in the imaging data set; these techniques have been used for data mining in a wide range of applications, from cryo-EM analysis of individual molecules to face recognition.
+%
+We adapt these techniques for temporal ordering and registration of imaging data from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
 %
@@
 %
-The third data set, shown in Figure~\ref{subfig:raw_data3}, consists of data collected from both wild type and mutant embryos.
+The third data set consists of data collected from both wild type and mutant embryos.
 %
@@
 
+We can further use this collection of registered and ordered snapshots to construct a smooth movie of the developmental dynamics. 
+%
+Using a Gaussian kernel, we calculate the average image for each frame of a movie which (approximately) spans the ordered images. 
+%
+This movie is included in the {\it SI Appendix}.
+%
+The dynamics are now apparent.
+%
+There are artifacts in the movie due to size variation in the samples.
+%
+Furthermore, the dynamics at the end of the movie are not as smooth, as we have fewer samples in this window. 
+%
+We would like to note that, although this movie provides a picture of the dynamics, the speed of the movie is not constant with respect to the real time dynamics.
+
 \subsection{Identifying distinct developmental trajectories}
@@
 \begin{figure}[t]
-%\raisebox{8.8cm}{{\figtextfont A}}
-%\includegraphics[height=9cm]{raw_data3}
-%\begin{minipage}[b]{8.5cm}
-\raisebox{6cm}{{\figtextfont B}}
-\includegraphics[width=8.4cm]{mut_wt_vdm_embedding}
-\raisebox{2cm}{{\figtextfont C}}
+\raisebox{5.7cm}{{\figtextfont A}}
+\begin{tikzpicture}
+     \node at (0,0)
+       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding_background}};
+     \node at (0,0)
+       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding}};  
+\end{tikzpicture}
+\raisebox{0.7cm}{{\figtextfont B}}
 \includegraphics[width=8.4cm]{wt_trajectory}
-\raisebox{2cm}{{\figtextfont D}}
+\raisebox{0.7cm}{{\figtextfont C}}
 \includegraphics[width=8.4cm]{mut_trajectory}
-%\end{minipage}
 \caption{Mutant and wild type {\xfigtextfontit Drosophila}.
-{\xfigtextfontit (A)} Images of cellularizing {\xfigtextfontit Drosophila} embryos.
-Each image is of a different embryo in a different rotational orientation.
-Half of the images are of wild type embryos, and half of the images are of sna null mutant embryos.
-{\xfigtextfontit(B)} Data from {\xfigtextfontit A}, projected onto the first two vector diffusion maps coordinates.
+{\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in {\xfigtextfontit SI Appendix}), projected onto the first two vector diffusion maps coordinates.
 The two distinct trajectories are highlighted in blue and red.
-{\xfigtextfontit (C)} Average trajectory of the dpERK signal over the blue curve in {\xfigtextfontit B}.
-{\xfigtextfontit (D)} Average trajectory of the dpERK signal over the red curve in {\xfigtextfontit B}.}
+{\xfigtextfontit (B)} dpERK signal at select wild type points in {\xfigtextfontit A} (denoted by bold symbols).
+{\xfigtextfontit (C)} dpERK signal at select mutant points in {\xfigtextfontit A} (denoted by bold symbols).}
 \customlabel{fig:data3}{5}
-\customlabel{subfig:raw_data3}{\ref{fig:data3}{\it A}}
+\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
+\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it B}}
+\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
+\end{figure}
+
+\begin{figure*}[t]
+\raisebox{8cm}{{\figtextfont A}}
+\includegraphics[height=8.4cm]{wt_trajectory2}
+%\raisebox{8cm}{{\figtextfont B}}
+\begin{tikzpicture}
+     \node at (0,0)
+       {\includegraphics[height=8.4cm]{mut_wt_vdm_embedding_background}};
+     \node at (0,0)
+       {\includegraphics[height=8.4cm]{mut_wt_vdm_embedding}};  
+     \node at (-5cm, 3.8cm)
+       {{\figtextfont B}};
+\end{tikzpicture}
+\raisebox{8cm}{{\figtextfont C}}
+\includegraphics[height=8.4cm]{mut_trajectory2}
+\caption{Mutant and wild type {\xfigtextfontit Drosophila}.
+{\xfigtextfontit (A)} dpERK signal at select wild type points in {\xfigtextfontit B} (denoted by bold symbols).
+{\xfigtextfontit(B)} Data collected from wild type and mutant embryos (original data set is shown in {\xfigtextfontit SI Appendix}), projected onto the first two vector diffusion maps coordinates.
+The two distinct trajectories are highlighted in blue and red.
+{\xfigtextfontit (C)} dpERK signal at select mutant points in {\xfigtextfontit B} (denoted by bold symbols).}
+\customlabel{fig:data3}{5}
+\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it A}}
 \customlabel{subfig:projection_data3}{\ref{fig:data3}{\it B}}
-\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it C}}
-\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it D}}
-\end{figure}
+\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
+\end{figure*}
 
@@
 We chose the algorithm outlined here, not because it is specifically tailored to our particular data sets, but rather, because it is sufficiently general and we are confident that it can be applied to a myriad of biological imaging applications.
+%
+Such algorithms are often tested on databases containing hundreds or thousands of images.
+%
+Here, we showed that we can successfully recover the underlying dynamics with relatively small data sets.
+%
+In general, the number of images required to accurately recover the dynamics will be a function of the noise in the data, as well as the dynamic range and complexity.
 
@@
 
-%\bibliographystyle{pnas}
-%\bibliography{background_reading/references,../../references/references}
+\bibliographystyle{pnas}
+\bibliography{background_reading/references,../../references/references}
 

### 2014-05-20 image_analysis_paper/si_information.tex
edits before meeting with Yannis

@@
 
+\usepackage{tikz}
+
+\usepackage{float}
 \usepackage{amssymb,amsfonts,amsmath}
@@
 \begin{document}
+
+\section{Raw data of mutant and wild type embryos}
+
+\begin{figure}[H]
+\includegraphics[height=9cm]{raw_data3}
+\caption{Images of cellularizing {\it Drosophila} embryos.
+Each image is of a different embryo in a different rotational orientation.
+Half of the images are of wild type embryos, and half of the images are of mutant embryos.}
+\end{figure}
 
@@
 
-\begin{figure}[h!]
+\begin{figure}[H]
 \includegraphics[width=8.4cm, trim=0cm 5.6cm 0cm 0cm, clip]{fig6}
@@
 
-\begin{figure}[h!]
+\begin{figure}[H]
 \raisebox{5.2cm}{A}
@@
 
-\begin{figure}[t]
+\begin{figure}[H]
 \includegraphics[width=8.4cm]{fig3}
@@
 
-\begin{figure*}
+\begin{figure}[H]
 \begin{minipage}{0.2\textwidth}
@@
 \label{fig:SO3_picture}
-\end{figure*}
+\end{figure}
 
@@
 
+\section{Eigenvalue Spectra from VDM}
+
+\begin{figure}[H]
+\includegraphics[width=5cm]{data1_evals}
+\includegraphics[width=5cm]{data2_evals}
+\includegraphics[width=5cm]{data3_evals}
+\caption{Eigenvalue spectra for three examples.}
+\end{figure}
+
+\begin{figure}[H]
+\includegraphics[width=5cm]{data1_coord_var}
+\includegraphics[width=5cm]{data2_coord_var}
+\includegraphics[width=5cm]{data3_coord_var}
+\caption{Variance of $\psi_{k,l}$ for three examples. We use the ones that have the largest variance.}
+\end{figure}
+
+\begin{figure}[H]
+\begin{tikzpicture}
+     \node at (0,0)
+       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding_background}};
+     \node at (0,0)
+       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding}};  
+\end{tikzpicture}
+\begin{tikzpicture}
+     \node at (0,0)
+       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding2_background}};
+     \node at (0,0)
+       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding2}};  
+\end{tikzpicture}
+\caption{Embedding for third example using two different $\epsilon$ values. The right one (smaller $\epsilon$) has a tighter structure, but the mutants are more removed.}
+\end{figure}
+
+\section{FBsPCA}
+
+I computed the FBsPCA basis for the set of images (using only the dpERK signal).
+%
+I then computed the bispectrum coefficients for each image expanded in this basis, and used these coefficients in a DMAPS calculation.
+%
+The results are shown below.
+%
+However, because the bispectrum is not stable to deformations, it does not seem to be able to identify neighboring images.
+
+\begin{figure}[H]
+\includegraphics[width=7cm]{FBsPCA_images}
+\includegraphics[width=10cm]{FBsPCA_dmaps}
+\caption{(Left) Images used for FBsPCA.
+(Right) First 4 dmaps variables, computed from the bispectrum coefficeints, plotted against the membrane thickness.
+Note there is no appreciable correlation: each dmaps variable simply selects one image from the data set as ``abnormal''.}
+\end{figure}
+
 \bibliographystyle{pnas}

### 2014-05-20 image_analysis_paper/test_FBsPCA.m

@@ -5,6 +5,9 @@ close all
 time_data = '../membrane_lengths/oct16.mat';
 image_dir = '../membrane_pictures/membrane2/dpERK_staining';
 
+im_save_dir = 'paper_figures2';
+
+
 %% load membrane lengths
 
 load(time_data);
@@ -38,7 +41,7 @@ for i=1:m
     % read image
     im1 = imread(sprintf('%s/emb%02d.tif', image_dir, ind(i)));
     
-
+    
     % resize image
     im1 = imresize(im1, [npixels npixels]);
     
@@ -59,10 +62,14 @@ K = m;
 %%
 
 figure;
+set(gcf, 'paperunits', 'centimeters')
+set(gcf, 'papersize', [16 16])
+set(gcf, 'paperposition',[0 0 16 16 ]);
 for i=1:m
-    subplot(subplot_dim1, subplot_dim2, i)
+    make_subplot(subplot_dim1, subplot_dim2, 0.01, i);
     imshow(uint8(data(:, :, i)));
 end
+saveas(gcf,sprintf('%s/FBsPCA_images', im_save_dir), 'pdf')
 
 
 figure;
@@ -134,7 +141,7 @@ fprintf('\nThe number of components is %d \n', size(Freqs, 1));
 % Coeff(Freqs==0, :)=Coeff(Freqs==0, :)*sqrt(2);
 
 [ Coeff_b2, Coeff_b_r2, toc_bispec ] = Bispec_2Drot( Coeff2, Freqs );
-    
+
 %%
 % W_dmaps = zeros(m);
 % for i=1:m
@@ -144,14 +151,22 @@ fprintf('\nThe number of components is %d \n', size(Freqs, 1));
 % end
 % W_dmaps = W_dmaps + W_dmaps';
 W_dmaps = squareform(pdist([real(Coeff_b') imag(Coeff_b') real(Coeff_b2') imag(Coeff_b2')])).^2;
-eps_dmaps = median(W_dmaps(:));
+eps_dmaps = median(W_dmaps(:))/2;
 
 [V, D] = dmaps(W_dmaps, eps_dmaps, 10);
 
 figure;
-plot(V(:,2),mem_lengths,'.')
-xlabel('\phi_2');
-ylabel('membrane thickness')
+set(gcf, 'paperunits', 'centimeters')
+set(gcf, 'papersize', [16 16])
[…]

### 2014-05-20 image_analysis_paper/test_mutants.m

@@ -254,7 +254,7 @@ axis equal
 %% VDM
 
 eps = median(W(:));
-neigs = 9;
+neigs = 12;
 
 [R_opt, embed_coord, embed_idx, D] = vdm(R, W, eps, neigs);
 
@@ -280,10 +280,32 @@ for i=1:m
     image_set_aligned_withnuclei(:,:,:,i) = im1;
 end
 
+figure; 
+set(gcf, 'paperunits', 'centimeters')
+set(gcf, 'papersize', [8 8])
+set(gcf, 'paperposition',[0 0 8 8])
+plot(diag(D),'.')
+xlabel('k')
+ylabel('\lambda_k')
+saveas(gcf,sprintf('%s/data3_evals', im_save_dir), 'pdf')
+
+
+figure;  
+set(gcf, 'paperunits', 'centimeters')
+set(gcf, 'papersize', [8 8])
+set(gcf, 'paperposition',[0 0 8 8])
+scatter(embed_idx(1,:), embed_idx(2, :), 500, var(embed_coord),'.')
+colorbar
+xlabel('k')
+ylabel('l')
+saveas(gcf,sprintf('%s/data3_coord_var', im_save_dir), 'pdf')
+
+
 %% find relevant VDM coordinates
 
 idx1 = find(embed_idx(1,:) == 4 & embed_idx(2,:) == 1);
 idx2 = find(embed_idx(1,:) == 7 & embed_idx(2,:) == 3);
+% idx2 = find(embed_idx(1,:) == 7 & embed_idx(2,:) == 1);
 
 if mean(embed_coord(:, idx1)) < 0
     embed_coord(:, idx1) = -embed_coord(:, idx1);
@@ -329,7 +351,7 @@ xlabel('first VDM coordinate')
 ylabel('second VDM coordinate')
 lgnd = legend('wild type','mutant', 'location','northeast');
 set(lgnd,'fontsize',6);
-print(sprintf('%s/mut_wt_vdm_embedding2', im_save_dir), '-dpdf', '-r600')
+%print(sprintf('%s/mut_wt_vdm_embedding2', im_save_dir), '-dpdf', '-r600')
 
 %% clustering
 
@@ -375,6 +397,43 @@ figure;
 plot(embed_coord(wt_ind,idx1),embed_coord(wt_ind,idx2),'.')
 hold on
 plot(embed_coord(mut_ind,idx1),embed_coord(mut_ind,idx2),'o')
+
[…]

### 2014-05-21 image_analysis_paper/pnas_draft2.tex
Added Yannis edits from 5/20

@@
 
+%%%
+%%%YGK   should "development" be in the title ? maybe yes, maybe no.
+%%%
+
 %% Enter authors via the \author command.
@@
 \begin{abstract}
-In studies of development, researchers are often presented with cross-sectional data, where each data point is a sample from a population fixed at a slightly different developmental time.
+In studies of development, researchers are often presented with cross-sectional data, where each data point is a sample from a population fixed at a 
+%slightly 
+different developmental time.
 %
@@
 %
-If each data point is a two-dimensional image, the images must first be registered before they can be temporally ordered.
+Data points in the form of two-dimensional image must first be registered before they can be temporally ordered.
 %
-When such data sets are large, noisy, and/or if the developmental changes are subtle, these tasks can be difficult to do by hand.
+When such data sets are large, noisy, and/or if the developmental changes are subtle, these tasks can be difficult to perform by hand.
 %
-We present an automatic approach to register {\it and} temporally order cross-sectional data sets of images.
+We present an automated approach to registering {\it and} temporally ordering cross-sectional data sets of images.
 %
@@
 Furthermore, these techniques can separate multiple trajectories within the same data set.
+%%%
+%%%YGK  The approach can detect multiple sources of variability in the data 
+%%%     and thus help visualize multiple trajectories within a data set.
 %
@@
 %
-In a cross-sectional study, developmental dynamics must be reconstructed from multiple snapshots of different embryos, each of which contributes only a single snapshot of a chemical or morphological process along its developmental trajectory.
+In a cross-sectional study, developmental dynamics must be reconstructed from multiple embryos, each of which contributes only a single snapshot of a chemical or morphological process along its developmental trajectory.
 %
 Both of these sampling schemes have their advantages and limitations, and both are extensively used by biologists.
+%%%
+%%%YGK sounds condescending
+%%%
 %
 Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms.
+%%%
+%%%YGK  one or two refs ?
+%%%
 %
@@
 %
-Recent advances in large-scale physical manipulation and imaging of embryos produce rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different geometric orientation and developmental time point.
+Recent advances in large-scale 
+%%%
+%%%YGK  large scale ? high throughput
+%%%
+physical manipulation and imaging of embryos produce rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different developmental time point but also at a different geometric orientation.
 %
-Importantly, the ``age'' of any given embryo arrested in its development is not known to high accuracy.
+Importantly, the ``age'' of any given embryo arrested in its development is not quantitatively known; typically what is known is 
+a certain time window to which a collection of embryos belongs.
+%to high accuracy.
 %
-In general, it is only known that a collection of embryos belongs to a certain time window.
+%%%YGK   datasets consistent one or two
 %
-In order to recover the developmental dynamics from such datasets, snapshots of different embryos must be spatially aligned or registered to factor out the relevant symmetries (i.e., translations and rotations), and then ordered in time.
+In order to recover the developmental dynamics from such datasets, snapshots of different embryos must be spatially aligned or {\em registered} to factor out the relevant geometric symmetries (e.g., translations and rotations), and then ordered in time.
 %
-We show how existing dimensionality reduction algorithms can greatly accelerate {\it and combine} both of these tasks.
+We show how existing dimensionality reduction algorithms can automate, greatly accelerate {\it and combine} both of these tasks.
 
-Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are visually apparent.
-%
-As an example, Figure~\ref{fig:fish} shows a caricature of fish development which illustrates the processes of growth and patterning.
+An example where 
+temporal ordering and registration of images is straightforward
+%%%YGK can be done by hand
+because the number of images is small and the differences between them are visually apparent is shown
+in Figure~\ref{fig:fish}: a caricature of fish development which illustrates the processes of growth and patterning.
 %
@@
 %
-In general, the landmarks needed for registration, such as the heads and fins, as well as the attributes which can be used to order the data, such as the body length, are not known {\it a priori}.
+In general, the landmarks needed for registration (here the heads and fins) as well as the attributes which can be used to order the data, such as the body length, are not known {\it a priori}.
 %
@@
 
+%%%YGK put an arrow in the disc
+
 Here, we show how recently developed dimensionality reduction techniques can successfully address these challenges and reveal developmental trajectories from collections of unregistered and unordered imaging data.
 %
-Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, as have dimensionality reduction techniques \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear} and their applications \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering, kafri2013dynamics}.
+Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, and so have dimensionality reduction techniques \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear} and their applications \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering, kafri2013dynamics}, especially in the last decade.
 %
-Here, we use a manifold learning algorithm \cite{singer2012vector} for high-dimensional data (such as images) which both factors out symmetries to align images and extracts low-dimensional structure in the imaging data set; these techniques have been used for data mining in a wide range of applications, from cryo-EM analysis of individual molecules to face recognition.
+Here, we use a manifold learning algorithm \cite{singer2012vector} for high-dimensional data (such as images) which both factors out geometric symmetries to register and extracts low-dimensional structure in the imaging data set.
 %
-We adapt these techniques for temporal ordering and registration of imaging data from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
+Such techniques have been used for data mining in 
+applications ranging from cryo-EM analysis of individual molecules to face recognition; here they are adapted
+for simultaneous registration and temporal ordering of imaging data from cross-sectional studies of developmental dynamics.
 %
-We demonstrate these techniques using three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics.
+The main objective of revealing stereotypic developmental trajectories.
+%
+We analyze three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics.
 %
@@
 %
-The third data set consists of data collected from both wild type and mutant embryos.
-%
-We will demonstrate that these algorithms can not only extract stereotypic dynamics, but also separate data sampled from distinct developmental paths.
+The third set consists of data collected from both wild type and mutant embryos; joint processing of the data 
+helps visualize the different sources of variability along distinct developmental paths.
 
@@
 
+%%%YGK need to repeat ? maybe not ?  English rephrase.
+%%%
 A typical dataset from cross-sectional studies of developmental processes is a collection of images of different embryos, each of which is fixed at a different developmental time and imaged in a different rotational and translational orientation.
+%%%YGK  when I read this I think of Eiffel tower pictures, not of ours - there is a transition... careful where we do it.
 %
 Our goal is to organize these data sets in a meaningful and informative way.
+%
+%%%YGK   maybe here is where this transition should be  -- I again have a feeling of repetition
 %
@@
 
-Our approach is based on vector diffusion maps \cite{singer2012vector}, a general technique developed for data sets which contain two sources of variability:
-symmetries (such as translations and rotations of the images due to the experimental setup) which one would like to factor out, and other directions of variability (such as temporal dynamics) which one would like to uncover.
+Our approach is based on vector diffusion maps \cite{singer2012vector}, a general technique developed for data sets which contain two types of sources of variability:
+geometric symmetries (such as translations and rotations of the images due to the experimental setup) which one would like to factor out, 
+and ``additional" directions of variability (such as temporal dynamics) which one wants to uncover.
 %
-Vector diffusion maps combine two algorithms, angular synchronization \cite{singer2011angular} for image registration and diffusion maps \cite{coifman2005geometric} for extracting low-dimensional structure in data, into one computation that allows us to simultaneously register and temporally order our images.
+Vector diffusion maps combine two algorithms, {\em angular synchronization} \cite{singer2011angular} for image registration and {\em diffusion maps} \cite{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation that allows us to simultaneously register and temporally order our images.
 %
 
-Vector diffusion maps has been used to classify images from cryo-electron microscopy (cryo-EM) experiments, where each image is a projection of a molecule, taken from a different viewing direction and in a different rotation.
+Vector diffusion maps has been used to classify images from cryo-electron microscopy (cryo-EM) experiments, where each image is a projection 
+of a randomly oriented molecule.
 %
+%%%YGK to imply which rotations we need to have discussed it above
+%%
 Here, we will use the algorithm to register images of {\it Drosophila} embryos with respect to rotations and translations, as well as uncover the main direction of variability {\it after} removing translational and rotational symmetries.
 %
-We assume that, in this set of images, this main direction of variability is parameterized by the developmental time of each embryo, and so uncovering this direction will allow us to recover the developmental dynamics.
+We assume that, in this set of images, this main direction of variability is parameterized by the developmental time of each embryo, so taht uncovering this direction will allow us to recover the developmental dynamics.
 %
@@
 
+%%%somewhere I want a sentence that says this is for general symmetry groups of some kind, and HERE it is for..... (maybe in SI)
 Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
@@
 %
-In general, this requires no template function or image landmarks; however, when the images are noisy, some pairwise alignments may be inaccurate.
+In general, this requires no template function or image landmarks; 
+%%%
+%%%YGK  ref template landmark
+%%%
+however, when the images are noisy, some pairwise alignments may be inaccurate.
 %
-Using the alignment angles between all pairs of vectors (or images), angular synchronization then finds the set of rotation angles, one angle for each vector, that is most consistent with the pairwise measurements (see {\it SI Appendix}).
+Using the alignment angles between all pairs of vectors (or images), angular synchronization finds the set of rotation angles (one angle for each vector) most consistent with all pairwise measurements (see {\it SI Appendix}).
 %
@@
 %
-However, the algorithm can successfully register sets of vectors (or images) even when many of the pairwise measurements are corrupted by noise. 
+However, the algorithm has been shown to successfully register sets of vectors (or images) even when many of the pairwise measurements are corrupted by noise \cite{...}.
 %
@@
 These dynamics can be revealed by ordering the data along the one-dimensional manifold that parameterizes the data.
+%%%
+%%%YGK feels like repeats
+%%%
 %
@@
 %
-This algorithm uncovers a parameterization of data that lies on a low-dimensional, perhaps nonlinear, manifold in high-dimensional space. 
+This algorithm uncovers an intrinsic parameterization of data that lies on a low-dimensional, perhaps nonlinear, manifold in high-dimensional space.
 %
@@
 %
-This idea of locality is denoted by the edges in Figure~\ref{subfig:dmaps1}.
+This idea of locality is denoted by the color intensity of the edges (denoting edge weight) in Figure~\ref{subfig:dmaps1}.
 %
-Data points which are close are connected by dark edges, and clearly, the dark edges are more informative about the low-dimensional structure of the data. 
+Data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
 %
@@
 %
-However, in our examples, each data point will be very high dimensional (e.g, an image with many pixels), and so we cannot extract this low-dimensional structure visually.
+In our working examples, each data point will be much higher dimensional (e.g, an image with many pixels), and so we cannot extract this low-dimensional structure visually.
 %
@@
 %
-Diffusion maps generalizes from one-dimensional nonlinear curves to higher-dimensional manifolds. 
+Diffusion maps will also generalize from one-dimensional nonlinear curves to higher-dimensional manifolds.
 %
-When our experiments have additional sources of variability (such as mutations) in addition to developmental dynamics, diffusion maps can uncover both of these variables and provide an informative organization of our data.
+Thus, when our experiments have additional sources of variability (such as mutations) in addition to developmental dynamics, diffusion maps can uncover both of these variability sources, and provide an informative organization of our data.
 %We will assume that our data are approximately one-dimensional, and that this dimension is parameterized by time, such that ordering our data along this main dimension/direction will temporally order our data.
@@
 %
-Vector diffusion maps combines both steps into one computation;
-the algorithm uses information about pairwise alignments and pairwise distances to find a global parameterization of the data, while registering images which are ``close'' (see {\it SI Appendix}).
+Vector diffusion maps combines both steps into a single computation;
+the algorithm uses information about pairwise alignments {\em and} pairwise distances to find a global parameterization of the data, while registering images which are ``close'' (see {\it SI Appendix}).
 %
-The algorithm is perfectly suited for our task of registering and temporally ordering cross-sectional imaging data.
+It is thus perfectly suited for our task of registering and temporally ordering cross-sectional imaging data.
 %
 We will use it to uncover a temporal ordering of our images, as well as register images which are similar, to obtain a global picture of developmental dynamics.
+%
+%%%YGK fees like a repeat
 
@@
 %
-One of these signals is provided by the nuclear localization of the transcription factor Dorsal (Dl, green), which subdivides the early embryo into the territories that give rise to the muscle, nerve, and skin tissues.
+One of these signals is provided by the nuclear localization of the transcription factor Dorsal (Dl, green in the figure), which subdivides the early embryo into the territories that give rise to the muscle, nerve, and skin tissues.
 %
-Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK), an enzyme that specifies a subset of neuronal cells. 
+Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, red in the figure), an enzyme that specifies a subset of neuronal cells.
 %
+
+%%%YGK say a simple thing in the caption about Dorsal-Ventral
 
@@
 
-For this dataset, the quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the ordering based on an independent approach.
+For this dataset, the quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the ordering obtained through an independent approach.
 %
-During the third hour of development, the embryo cellularizes, where nuclei located under common plasma membrane are enclosed by lateral membranes which grow inward, separating the nuclei into individual cells.
+During the third hour of development, the embryo cellularizes: nuclei located under common plasma membrane are enclosed by lateral membranes which grow inward, separating the nuclei into individual cells.
 %
-The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo within 1-3 minutes \cite{figard2013plasma} (see {\it SI Appendix}).    
+The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo to within 1--3 minutes \cite{figard2013plasma} (see {\it SI Appendix}).
 %
 The rank correlation coefficient between the orderings obtained based on the progress of cellularization and vector diffusion maps is 0.9127.
+%%%
+%%%YGK how many digits are significant ?
+%%%
 %
@@
 
-Although the first vector diffusion maps coordinate accurately captures the developmental dynamics, there are other sources of variability within this data set that could also be extracted using our techniques.
+Although the first vector diffusion maps coordinate accurately captures the developmental dynamics, there are other sources of variability within this data set that could also be extracted using these techniques.
 %
-For example, the images contain size variability. 
+An obvious one is variability due to size; here the size variation is relatively small relative to the developmental changes.
 %
-In this example, the size variation is relatively small relative to the developmental changes. 
-%
-However, in other applications where the size variation is significant or the developmental dynamics are more subtle, it may be important to factor out size or scaling symmetries, or consider additional components.
+However, in other applications where the size variation is significant or the developmental dynamics are more subtle, it may be important to factor out size or {\em scaling} symmetries.
+%%%
+%%%YGK   for teh "additional components" - don't have in the same sentence - say it somewhere else.... INCLUDE IT (maybe in the mutants)
 

### 2014-05-21 image_analysis_paper/fig5.tex
edits prior to Yannis on 5/21

\documentclass[12pt]{minimal}

\usepackage{graphicx}
\graphicspath{ {paper_figures2/} }

\usepackage{tikz}

\usepackage[margin=0in, paperwidth=8.7cm, paperheight=9.5cm]{geometry}

\begin{document}

\centering
%\raisebox{5.4cm}{\textsf{A}}
\begin{tikzpicture}
     \node at (0,0)
       {\includegraphics[width=8cm]{mut_wt_vdm_embedding_background}};
     \node at (0,0)
       {\includegraphics[width=8cm]{mut_wt_vdm_embedding}};
      \node at (-4cm, 2.4cm) {\textsf{A}};
\end{tikzpicture}\\
\raisebox{0.6cm}{\textsf{B}}
\includegraphics[width=8cm]{wt_trajectory}\\
\raisebox{0.6cm}{\textsf{C}}
\includegraphics[width=8cm]{mut_trajectory}

\end{document}

### 2014-05-21 image_analysis_paper/pnas_draft2.tex
edits prior to Yannis on 5/21

@@
 
-\title{Temporal ordering and registration of cross-sectional imaging data}
-
-%%%
-%%%YGK   should "development" be in the title ? maybe yes, maybe no.
-%%%
+\title{Temporal ordering and registration of images in studies of developmental dynamics}
 
@@
 %
-Furthermore, these techniques can separate multiple trajectories within the same data set.
-%%%
-%%%YGK  The approach can detect multiple sources of variability in the data 
-%%%     and thus help visualize multiple trajectories within a data set.
+%Furthermore, these techniques can separate multiple trajectories within the same data set.
+%
+Furthermore, the approach can detect multiple sources of variability 
+and thus help visualize multiple trajectories within a data set.
 %
@@
 %
-%%%YGK   datasets consistent one or two
+Furthermore, each embryo appears at an arbitrary translation and rotation within the image frame (see {\it SI Appendix}). 
 %
-In order to recover the developmental dynamics from such datasets, snapshots of different embryos must be spatially aligned or {\em registered} to factor out the relevant geometric symmetries (e.g., translations and rotations), and then ordered in time.
+In order to recover the developmental dynamics from such data sets, snapshots of different embryos must be spatially aligned or {\em registered} to factor out the relevant geometric symmetries (e.g., translations and rotations), and then ordered in time.
 %
@@
 An example where 
-temporal ordering and registration of images is straightforward
-%%%YGK can be done by hand
+temporal ordering and registration of images can be done manually
 because the number of images is small and the differences between them are visually apparent is shown
@@
 %
-In general, the landmarks needed for registration (here the heads and fins) as well as the attributes which can be used to order the data, such as the body length, are not known {\it a priori}.
+In general, the landmarks needed for registration (here the heads and fins) as well as the attributes which can be used to order the data (such as the body length) are not known {\it a priori}.
 %
@@
 \begin{figure}[t]
-\includegraphics[width=8.4cm]{fig1}
+\includegraphics[width=8.7cm]{fig1}
 \caption{Caricature illustrating the tasks of image registration and temporal ordering.
@@
 
-Here, we show how recently developed dimensionality reduction techniques can successfully address these challenges and reveal developmental trajectories from collections of unregistered and unordered imaging data.
+We show how recently developed dimensionality reduction techniques can successfully address these challenges and reveal developmental trajectories from collections of unregistered and unordered imaging data.
 %
@@
 %
-Here, we use a manifold learning algorithm \cite{singer2012vector} for high-dimensional data (such as images) which both factors out geometric symmetries to register and extracts low-dimensional structure in the imaging data set.
+Here, we use a manifold learning algorithm \cite{singer2012vector} for high-dimensional imaging data.% which both factors out geometric symmetries to register the images and extracts low-dimensional structure in the imaging data set.
 %
 Such techniques have been used for data mining in 
-applications ranging from cryo-EM analysis of individual molecules to face recognition; here they are adapted
-for simultaneous registration and temporal ordering of imaging data from cross-sectional studies of developmental dynamics.
-%
-The main objective of revealing stereotypic developmental trajectories.
+applications ranging from analysis of cryo-electron microscopy images of individual molecules to face recognition; here they are adapted
+for simultaneous registration and temporal ordering of imaging data from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
 %
@@
 %
-The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate our methods.
+The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate this approach.
 %
@@
 %
-The third set consists of data collected from both wild type and mutant embryos; joint processing of the data 
-helps visualize the different sources of variability along distinct developmental paths.
+The third set consists of data collected from both wild type and mutant embryos, and joint processing of the data will help to visualize the different sources of variability along distinct developmental paths.
 
@@
 
-%%%YGK need to repeat ? maybe not ?  English rephrase.
-%%%
-A typical dataset from cross-sectional studies of developmental processes is a collection of images of different embryos, each of which is fixed at a different developmental time and imaged in a different rotational and translational orientation.
 %%%YGK  when I read this I think of Eiffel tower pictures, not of ours - there is a transition... careful where we do it.
 %
-Our goal is to organize these data sets in a meaningful and informative way.
+Our goal is to organize data sets from cross-sectional studies of developmental processes in a meaningful and informative way.
 %
-%%%YGK   maybe here is where this transition should be  -- I again have a feeling of repetition
+%We will show that dimensionality reduction algorithms can simultaneously register and temporally order the images to give us a meaningful picture of the underlying developmental dynamics.
 %
-This requires rotating and translating each image so the set is in a consistent frame of reference ({\it registration}), and then temporally ordering the images.
-%
-We will show that dimensionality reduction algorithms can simultaneously accomplish both of these tasks to give us a meaningful picture of the underlying developmental dynamics.
-
 Our approach is based on vector diffusion maps \cite{singer2012vector}, a general technique developed for data sets which contain two types of sources of variability:
@@
 
-Vector diffusion maps has been used to classify images from cryo-electron microscopy (cryo-EM) experiments, where each image is a projection 
-of a randomly oriented molecule.
+Vector diffusion maps has been used to classify images from cryo-electron microscopy (cryo-EM) experiments, where each image is a projection of a randomly oriented molecule.
 %
@@
 %
-In general, this requires no template function or image landmarks; 
-%%%
-%%%YGK  ref template landmark
-%%%
+In general, this requires no template function \cite{ahuja2007template} or image landmarks \cite{ian1998statistical}; 
 however, when the images are noisy, some pairwise alignments may be inaccurate.
@@
 %
-However, the algorithm has been shown to successfully register sets of vectors (or images) even when many of the pairwise measurements are corrupted by noise \cite{...}.
+However, the algorithm has been shown to successfully register sets of vectors (or images) even when many of the pairwise measurements are corrupted by noise \cite{singer2011angular}.
 %
@@
 %
-In the data sets we will consider, after removing the variability due to translations and rotations of the images, the remaining variability may be due to developmental dynamics.
-%
-These dynamics can be revealed by ordering the data along the one-dimensional manifold that parameterizes the data.
+After removing the variability due to translations and rotations of the images, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the variability in the data.
 %%%
@@
 %
-We will use it to uncover a temporal ordering of our images, as well as register images which are similar, to obtain a global picture of developmental dynamics.
-%
-%%%YGK fees like a repeat
 
 \begin{figure}[t]
-\includegraphics[width=8.4cm]{fig4}
+\includegraphics[width=8.7cm]{fig2}
 \caption{Schematic illustrating angular synchronization and diffusion maps.
@@
 \begin{figure*}[t]
-\raisebox{8.1cm}{{\figtextfont A}}
-\includegraphics[width=8.4cm]{raw_data1}
-\hfill
-\raisebox{8.1cm}{{\figtextfont B}}
-\includegraphics[width=8.4cm]{VDM_data1_ordered}\\
-%\vspace{0.2cm}
-%\centering
-%\raisebox{0.5cm}{{\figtextfont C}}
-%\includegraphics[width=8.4cm]{average_trajectory_VDM}
+\includegraphics{fig3}
 \caption{{\xfigtextfontit Drosophila} during cellularization.
@@
 {\xfigtextfontit (B)} Images from {\xfigtextfontit A}, now registered and ordered using vector diffusion maps.
-}%{\it (C)} Average trajectory of the images shown in {\it B}.
-Each image is the average of $7$ consecutive images in {\it B}.
-}
+The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.}
 \customlabel{fig:data1}{3}
@@
 
-As a first illustration of this approach to temporal ordering and registration of images, we used a dataset where the correct registration and temporal order is known {\it a priori}.
+As a first illustration of this approach to temporal ordering and registration of images, we analyze a data set where the correct registration and temporal order is known {\it a priori}.
 %
-This dataset is obtained by fluorescent imaging of {\it Drosophila} embryos during the third hour after fertilization.
+This data set is obtained by fluorescent imaging of {\it Drosophila} embryos during the third hour after fertilization.
 %
@@
 %
-Figure 3a shows optical cross-sections of each embryo, with nuclei (gray) labeled by DAPI, a DNA stain.
+Figure~\ref{subfig:raw_data1} shows optical cross-sections of each embryo, with nuclei (gray in the figure) labeled by DAPI, a DNA stain.
 %
@@
 %
-Both of these features become readily apparent when images are registered and ordered using vector diffusion maps (see Figure~\ref{subfig:ordered_data1}): the peak of nuclear Dl is consistent throughout the images and intensity of the dpERK signal in the lateral regions of the embryo monotonically increases across the dataset.
+Both of these features become readily apparent when images are registered and ordered using vector diffusion maps (see Figure~\ref{subfig:ordered_data1}): the peak of nuclear Dl is consistent throughout the images and intensity of the dpERK signal in the lateral regions of the embryo monotonically increases across the data set.
 %
 
-For this dataset, the quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the ordering obtained through an independent approach.
+For this data set, the quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the ordering obtained through an independent approach.
 %
@@
 %
-The rank correlation coefficient between the orderings obtained based on the progress of cellularization and vector diffusion maps is 0.9127.
-%%%
-%%%YGK how many digits are significant ?
-%%%
+Spearman's rank correlation coefficient between the orderings obtained based on the progress of cellularization and vector diffusion maps is 0.91.
 %
-Thus, we have validated a dimensionality reduction approach to temporal ordering and registration of data on a dataset where image orientation and ordering can be assessed independently.
+Thus, we have validated a dimensionality reduction approach to temporal ordering and registration of data on a data set where image orientation and ordering can be assessed independently.
 
@@
 %
-An obvious one is variability due to size; here the size variation is relatively small relative to the developmental changes.
+An obvious one is variability due to size; here, the size variation is relatively small relative to the developmental changes.
 %
-However, in other applications where the size variation is significant or the developmental dynamics are more subtle, it may be important to factor out size or {\em scaling} symmetries.
+However, in other applications where the size variation is significant or the developmental dynamics are more subtle, it may be important to factor out size or {\it scaling} symmetries.
 %%%
@@
 
-Because the images in Figure~\ref{fig:data1} are relatively simple, other techniques could have been used to temporally order them.
+As the images in Figure~\ref{fig:data1} are relatively simple, other techniques could have been used to temporally order them.
 %
-Because these data are effectively a single mode (the two dpERK peaks) growing in time,
+Because this data set is effectively a single mode (the two dpERK peaks) growing in time,
 principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data.
@@
 %
-The next dataset illustrates this point.
+The next data set illustrates this point.
 
@@
 %
-These images cover a thirty minute time interval and span late cellularization (the early images in this data set overlap with the late images in the previous set) through gastrulation, when the two-dimensional sheet of newly formed cells begins to deform into three-dimensional structures of the future organs.
+These images cover a thirty minute time interval and span late cellularization (the early images in this data set overlap with the late images in the previous set) through gastrulation, when the two-dimensional sheet of newly formed cells begins to deform into three-dimensional structures which will form the future organs.
 %
@@
 %
-However, during this developmental time period, the nuclei are no longer uniformly arranged around the periphery of the embryo as the tissue begins to morph.
+However, during this developmental time period, the nuclei are no longer uniformly arranged around the periphery of the embryo.
 %
@@
 %
-During this time window, germband elongation causes cells from the ventral side of the embryo to move towards the poles of the embryo, and then wrap around to the dorsal side.
+During this time window, germband elongation causes cells from the ventral side to move towards the poles of the embryo, and then wrap around to the dorsal side.
 %
@@
 \begin{figure*}[t]
-\raisebox{5.3cm}{{\figtextfont A}}
-\includegraphics[width=16.8cm]{raw_data2}
-
-\vspace{0.2cm}
-\raisebox{5.3cm}{{\figtextfont B}}
-\includegraphics[width=16.8cm]{VDM_ordered}
-
-\vspace{0.2cm}
-\raisebox{0.6cm}{{\figtextfont C}}
-\includegraphics[width=16.8cm]{average_trajectory}
+\includegraphics{fig4}
 \caption{{\xfigtextfontit Drosophila} during gastrulation.
@@
 
-The dynamics in this data set are highly nonlinear, and only 22\% of the variability in the data is captured by the first principal component.
+Visually, the dynamics in this data set are highly nonlinear, and only 22\% of the variability in the data is captured by the first principal component.
 %
@@
 %
-To highlight the developmental dynamics revealed by vector diffusion maps, we construct an average developmental trajectory.
+To highlight the developmental dynamics revealed by vector diffusion maps, we construct a movie of an averaged developmental trajectory by applying a Gaussian filter to the set of ordered images (more details are given in the {\it SI Appendix}).
 %
-This is done by dividing the registered and ordered data into windows or classes, where each window contains a group of consecutive images in the ordered data, then calculating the average image in each window.
+This movie is included in the {\it SI Appendix},
+and representative snapshots from this movie are shown in Figure~\ref{subfig:averaged_data2}.
 %
-Figure~\ref{subfig:averaged_data2} shows the average trajectory for this data set; it is now very straightforward to see the underlying developmental dynamics.
+The stereotypic dynamics are now readily apparent.
 %
-In this example, we accomplished the same task presented in the caricature in Figure~\ref{fig:fish}, but without knowledge about the appropriate landmarks (e.g., fins) for registration or the appropriate statistic (e.g., body size) by which to order.
-
-We can further use this collection of registered and ordered snapshots to construct a smooth movie of the developmental dynamics.
+However, there are some artifacts in the movie due to size variation, the dynamics at the end of the movie are not as smooth, as we have fewer samples in this time window.
 %
-Using a Gaussian kernel, we calculate the average image for each frame of a movie which (approximately) spans the ordered images.
+We would like to note that, although this movie provides us with a representative picture of the dynamics, the speed of the movie is not constant with time, as we have no true time information about the images.
 %
-This movie is included in the {\it SI Appendix}.
-%
-The dynamics are now apparent.
-%
-There are artifacts in the movie due to size variation in the samples.
-%
-Furthermore, the dynamics at the end of the movie are not as smooth, as we have fewer samples in this window.
-%
-We would like to note that, although this movie provides a picture of the dynamics, the speed of the movie is not constant with respect to the real time dynamics.
+We have thus accomplished the same task presented in the caricature in Figure~\ref{fig:fish}, but without knowledge about the appropriate landmarks (e.g., fins) for registration or the appropriate statistic (e.g., body size) by which to order.
 
@@
 %
-As an example, Figure~\ref{subfig:raw_data3} shows a data set consisting of $42$ embryos undergoing cellularization during the third hour of development.
+As an example, we have collected a data set consisting of $42$ embryos undergoing cellularization during the third hour of development (see {\it SI Appendix}); $21$ of the embryos are wild type, and $21$ of the embryos are mutant embryos where dpERK is expressed in one broad peak at the ventral side.
 %
@@
 %
-In this data set, $21$ of the embryos are wild type, and $21$ of the embryos are mutant embryos where dpERK is expressed in one broad peak at the ventral side, instead of in two bimodal peaks.
-%
-We therefore expect two sources of variability in this data set: differences due to developmental time, and differences between mutant and wild type dynamics.
+We expect two sources of variability in this data set: differences due to developmental time, and differences between mutant and wild type dynamics.
 
@@
 %
-However, our data now contains two main sources of variation, and we can see two distinct trajectories, highlighted in blue and red, emerge in this two-dimensional progression.
+However, our data now contains two main sources of variation, and we can see two distinct trajectories, highlighted in blue and red, emerge in this two-dimensional projection.
 %
 Both curves originate from a common point, and then diverge.
-
-We can compute the average trajectory over each of the two curves to gain insight into the different developmental dynamics.
 %
-The average over the blue curve, shown in Figure~\ref{subfig:average_wt_data3}, shows the emergence of two distinct dpERK peaks, as expected.
+Representative images along the blue and red curves are shown in Figures~\ref{subfig:average_wt_data3}~and~\ref{subfig:average_mut_data3}.
 %
-The average over the red curve, shown in Figure~\ref{subfig:average_mut_data3}, reveals one more concentrated dpERK peak.
+Figure~\ref{subfig:average_wt_data3}, corresponding to the blue curve, shows the emergence of two distinct dpERK peaks, and Figure~\ref{subfig:average_mut_data3} reveals one more concentrated dpERK peak.
 %
-This is consistent with previously known dynamics.
+This is consistent with previously known dynamics:
+at early times, the wild type and mutant embryos both have little dpERK signal, and are therefore indistinguishable, and at later times, the two groups separate as the wild type embryos develop two distinct dpERK peaks and the mutant embryos develop only a single peak.
 %
-At early times, the wild type and mutant embryos both have little dpERK signal, and are therefore indistinguishable, whereas at later times, the two groups separate as the wild type embryos develop two distinct dpERK peaks and the mutant embryos develop only a single peak.
-%
-We have therefore demonstrated that vector diffusion maps can not only recover dynamics, but also distinguish different expression patterns and developmental trajectories.
+We have therefore shown that vector diffusion maps can not only recover dynamics, but also distinguish different expression patterns and developmental trajectories.
 
 \begin{figure}[t]
-\raisebox{5.7cm}{{\figtextfont A}}
-\begin{tikzpicture}
-     \node at (0,0)
-       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding_background}};
-     \node at (0,0)
-       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding}};
-\end{tikzpicture}
-\raisebox{0.7cm}{{\figtextfont B}}
-\includegraphics[width=8.4cm]{wt_trajectory}
-\raisebox{0.7cm}{{\figtextfont C}}
-\includegraphics[width=8.4cm]{mut_trajectory}
+\includegraphics{fig5}
 \caption{Mutant and wild type {\xfigtextfontit Drosophila}.
@@
 
-\begin{figure*}[t]
-\raisebox{8cm}{{\figtextfont A}}
-\includegraphics[height=8.4cm]{wt_trajectory2}
-%\raisebox{8cm}{{\figtextfont B}}
-\begin{tikzpicture}
-     \node at (0,0)
-       {\includegraphics[height=8.4cm]{mut_wt_vdm_embedding_background}};
-     \node at (0,0)
-       {\includegraphics[height=8.4cm]{mut_wt_vdm_embedding}};
-     \node at (-5cm, 3.8cm)
-       {{\figtextfont B}};
-\end{tikzpicture}
-\raisebox{8cm}{{\figtextfont C}}
-\includegraphics[height=8.4cm]{mut_trajectory2}
-\caption{Mutant and wild type {\xfigtextfontit Drosophila}.
+%\begin{figure*}[t]
+%\raisebox{8cm}{{\figtextfont A}}
+%\includegraphics[height=8.4cm]{wt_trajectory2}
+%%\raisebox{8cm}{{\figtextfont B}}
+%\begin{tikzpicture}
+%     \node at (0,0)
+%       {\includegraphics[height=8.4cm]{mut_wt_vdm_embedding_background}};
+%     \node at (0,0)
+%       {\includegraphics[height=8.4cm]{mut_wt_vdm_embedding}};
+%     \node at (-5cm, 3.8cm)
+%       {{\figtextfont B}};
+%\end{tikzpicture}
+%\raisebox{8cm}{{\figtextfont C}}
+%\includegraphics[height=8.4cm]{mut_trajectory2}
+%\caption{Mutant and wild type {\xfigtextfontit Drosophila}.
 {\xfigtextfontit (A)} dpERK signal at select wild type points in {\xfigtextfontit B} (denoted by bold symbols).
@@
 {\xfigtextfontit (C)} dpERK signal at select mutant points in {\xfigtextfontit B} (denoted by bold symbols).}
-\customlabel{fig:data3}{5}
-\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it A}}
-\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it B}}
-\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
-\end{figure*}
+%\customlabel{fig:data3}{5}
+%\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it A}}
+%\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it B}}
+%\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
+%\end{figure*}
 
@@
 %
-Such algorithms are often tested on databases containing hundreds or thousands of images.
+Furthermore, such data mining algorithms are often tested on databases containing hundreds or thousands of images.
 %
-Here, we showed that we can successfully recover the underlying dynamics with relatively small data sets.
+Here, we could successfully recover the underlying dynamics with relatively small data sets.
 %
-In general, the number of images required to accurately recover the dynamics will be a function of the noise in the data, as well as the dynamic range and complexity.
+In general, the number of images required to accurately recover the dynamics will be a function of the noise in the data, as well as the dynamic range and complexity of the underlying process.
 
@@
 %
-Most of these applications order the data by solving a traveling salesman problem (TSP) or constructing a minimum spanning tree (MST) on the data,
+Many of these applications order the data by solving a traveling salesman problem (TSP) or constructing a minimum spanning tree (MST) on the data,
 expecting that these structures (the path of the traveling salesman, or the minimum spanning tree) characterize the majority of the patterns and variability within the data and provide an accurate encoding of the dynamical progression.
@@
 %
+These techniques would also allow us to compare the developmental dynamics of several different mutations to see how mutations are developmentally related.
+%
 Other sources of variability, such as size or viewing angle, could also be uncovered using vector diffusion maps.
-%
-These techniques would also allow us to compare the developmental dynamics of several different mutations to see how mutations are developmentally related.
 
@@
 %
-We have demonstrated one such technique (window averaging) that allows us to extract a more representative developmental trajectory from the set of noisy snapshots.
+We have demonstrated one such technique (Gaussian filtering) that allows us to extract a more representative developmental trajectory from the set of noisy snapshots.
 %
-Constructing a smooth movie from the registered and ordered cross-sectional snapshots would also be of biological interest, and
-examining the top principal components in each window would allow one to visualize the main dynamical features along the trajectory.
+%Constructing a smooth movie from the registered and ordered cross-sectional snapshots would also be of biological interest, and
+Examining the top principal components in different time windows would allow one to visualize the main dynamical features along the trajectory.
 %
-These questions first require that the data be registered and temporally ordered.
+%These questions first require that the data be registered and temporally ordered.
 %
@@
 
-%\begin{itemize}
-%\item What are other people doing  (maybe critically)
-%\item Movies  smoothness
-%\item Discrete symmetries like flips
-%\item Local information ? (features of the image) (associate with Fourier-Bessel and bispectra)
-%\item Mutant-wildtype stuff  IMPORTANT
-%\item And in general, MULTIPLE variablities is important (mutants, guys in another cycle, etc.)
+\begin{itemize}
+\item What are other people doing  (maybe critically)
+\item Movies  smoothness
+\item Discrete symmetries like flips
+\item Local information ? (features of the image) (associate with Fourier-Bessel and bispectra)
+\item Mutant-wildtype stuff  IMPORTANT
+\item And in general, MULTIPLE variablities is important (mutants, guys in another cycle, etc.)
 
 %\item size is an issue
-%\item  small tilts is an issue
-%\item   could one do three-d image processing ?  of the floating thing, without the device
-%\item how do we go about merging datasets ?
-%\end{itemize}
+\item  small tilts is an issue
+\item   could one do three-d image processing ?  of the floating thing, without the device
+\item how do we go about merging data sets ?
+\end{itemize}
 

### 2014-05-21 image_analysis_paper/si_information.tex
edits prior to Yannis on 5/21

@@
 \begin{figure}[H]
-\includegraphics[width=8.4cm, trim=0cm 5.6cm 0cm 0cm, clip]{fig6}
+\includegraphics[width=8.4cm, trim=0cm 5.6cm 0cm 0cm, clip]{SI_fig6}
 \caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different levels of development.
@@
 %
-For the images shown in Figures~3~and~5 of the paper, we only use the red and green channels (dpERK and Dl) in the vector diffusion maps computations.
+For the first and third data sets in the paper (images taken during cellularization), we only use the red and green channels (dpERK and Dl) in the vector diffusion maps computations, as the spatial location of the nuclei does not change appreciably during this time window.
 %
-The nuclei are shown only for visual reference.
+The nuclei in the corresponding figures are shown only for visual reference.
 %
-For the images shown in Figure~4 of the paper, we use all three channels (dpERK, Twi, and nuclei) to register the images.
+For the second data set of images during gastrulation, we use all three channels (dpERK, Twi, and nuclei) in the vector diffusion maps computations.
 %
@@
 
-\section{Principal component analysis for gastrulating images}
+When imaging, the laser settings were calibrated so that images have good dynamic range, while not saturating the detector. 
+% 
+The same laser setting is maintained within each experiment.
+
+{\bf discuss how we choose weights of signals}
+
+\section{Principal Component Analysis for gastrulating images}
 
@@
 %
-We first registered the images using angular synchronization, and then computed the principal components for the data set.
+We first registered the images using angular synchronization \cite{singer2011angular}, and then computed the principal components for the data set.
 %
@@
 %
-The dynamics that we uncover are not consistent with the expected biological progression.
+The dynamics that we uncover are not consistent with the expected biological progression described in the main text.
 %
-Furthermore, the eigenvalue spectrum from PCA indicates that a significant amount of the variance is not captured by the first principal component, therefore motivating us to use nonlinear techniques.
+Furthermore, the eigenvalue spectrum from PCA indicates that a significant amount of the variance is not captured by the first principal component, thereby motivating us to use nonlinear techniques.
 
@@
 \includegraphics[width=8.4cm]{data2_PCA_variance}
+\raisebox{3.7cm}{C}
+\includegraphics[width=8.4cm]{data2_PCA_proj}
+
 \caption{{\it (A)} Images shown in Figure~4 of the paper, now ordered by the first PCA projection coefficient.
@@
 {\it (B)} The fraction of the variance captured by each of the first 10 principal components.
-Note that components 2--10 have a nonnegligible contribution to the data.}
+Note that components 2--10 have a nonnegligible contribution to the data.
+{\it (C)} Projection onto the first two principal components.}
 \label{fig:PCA_results}
@@
 
-We will demonstrate our techniques for registration and temporal ordering using a synthetic data set.
+We will demonstrate the algorithms for registration and temporal ordering using a synthetic data set.
 %
@@
 
-\subsection{Angular synchronization \cite{singer2011angular}}
-
 \begin{figure}[H]
-\includegraphics[width=8.4cm]{fig3}
+\includegraphics[width=8.4cm]{SI_fig3}
 \caption{{\it (A)} One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
@@
 \end{figure}
+
+\subsection{Angular synchronization \cite{singer2011angular}}
 
@@
 Then,
-\begin{equation}
+\begin{equation} \label{eq:R_hat}
 \hat{R} =
@@
 \begin{equation} \label{eq:dmaps_W}
-W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right)
+W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right),
 %W_{ij} = e^{ -\frac{d^2(x_i, x_j)}{\epsilon^2}}
 \end{equation}
-where $d(x_i, x_j)$ is the pairwise distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic scale.
+where $d(x_i, x_j)$ is a pairwise distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic scale.
 %
@@
 %
-$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph}); here, we take $\epsilon$ to be the median of the pairwise distances between data points.
+$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/10^{th}$ of the median of the pairwise distances between data points for the first two data sets, and the median of the pairwise distances between data points for the third data set.
 
@@
 More specifically, the $j^{th}$ embedding coordinate will be given by
-\begin{equation}
+\begin{equation} \label{eq:vdm_coord}
 \psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle, \: 0 < k \le d, (j-1)d < l \le jd
@@
 
-\subsection{Registering images with respect to translations and rotations} \label{subsec:trans_rot_register}
+\subsection{Registering images with respect to rotations and translations} \label{subsec:trans_rot_register}
 
@@
 From a rotation matrix $R$, we compute the Euler angles $\alpha$, $\beta$, and $\gamma$ using the following relationships
-\begin{equation}
+\begin{equation} \label{eq:angle_invert}
 \begin{aligned}
@@
 Inverting \eqref{eq:angle_relations} yields the desired approximate translations and rotation, and we round optimal translations to the nearest integer to remove the need for interpolation when translating the images.
+
+\subsection{Algorithms}
+
+\subsubsection{Angular synchronization}
+
+\begin{enumerate}
+\item Compute the pairwise rotations (described in \eqref{eq:opt_angle}) or rotations + translations (described in \eqref{eq:opt_pairwise}) needed to register every pair of data points.
+\item Construct the corresponding rotation matrices $R_{ij}$ for every pair $i, j$ of images (see \eqref{eq:R_theta} for rotations or \eqref{eq:total_rot} for rotations + translations).
+\item Construct the matrix $H$ as in \eqref{eq:H_to_R}.
+\item Compute the top $2$ (for rotations) or $3$ (for rotations + translations) eigenvectors of $H$.
+\item Recover the estimates $\hat{R}_1, \dots, \hat{R}_m$ from the top 2 or 3 eigenvectors as in \eqref{eq:R_hat}.
+\item Project these estimates onto the closest proper rotation matrix using \eqref{eq:R_est}
+\item Recover the optimal rotation angle $\theta_i$ using \eqref{eq:R_theta} or rotation + translation using \eqref{eq:angle_invert} for each image
+\end{enumerate}
+
+\subsubsection{Diffusion Maps}
+
+\begin{enumerate}
+\item Construct the matrix $W$ as in \eqref{eq:dmaps_W}
+\item Construct the matrix $A$ as in \eqref{eq:dmaps_A}
+\item Compute the top few eigenvectors of $A$, ordered with decreasing eigenvalues
+\item If we only want to order the data, then sort the data by the value of the entries in the second eigenvector
+\item For higher-dimensional embeddings (e.g., the mutants), we use the entries of successive eigenvectors (the second eigenvector, the third eigenvector, etc.) as the embedding coordinates
+\end{enumerate}
+
+\subsubsection{Vector Diffusion Maps}
+
+\begin{enumerate}
+\item Compute the pairwise rotations (described in \eqref{eq:opt_angle}) or rotations + translations (described in \eqref{eq:opt_pairwise}) needed to register every pair of data points.
+%
+\item Construct the corresponding rotation matrices $R_{ij}$ for every pair $i, j$ of images (see \eqref{eq:R_theta} for rotations or \eqref{eq:total_rot} for rotations + translations).
+%
+\item Construct the matrix $W$ as in \eqref{eq:dmaps_W}, using the minimum distance obtained for each pair in step 1 as the pairwise distances in the Gaussian kernel.
+%
+\item Construct the matrix $H$ as in \eqref{eq:H_low_rank}.
+%
+\item Construct the matrix $A$ as in \eqref{eq:dmaps_A}.
+%
+\item Construct the matrix $S$ as in \eqref{eq:vdm_S}.
+%
+\item Compute the top few eigenvectors of $S$.
+%
+\item Recover the estimates $\hat{R}_1, \dots, \hat{R}_m$ from the top 2 or 3 eigenvectors as in \eqref{eq:R_hat}.
+%
+\item Project these estimates onto the closest proper rotation matrix using \eqref{eq:R_est}.
+%
+\item Recover the optimal rotation angle $\theta_i$ using \eqref{eq:R_theta} or rotation + translation using \eqref{eq:angle_invert} for each image
+%
+\item To compute the first (non-trivial) embedding coordinate for the data, compute coordinates as in \eqref{eq:vdm_coord}, taking $k=1, 2$ and $l=3, 4$ for the rotation case, or $k = 1, 2, 3$ and $l = 4, 5, 6$ for the rotation + translation case.
+%
+\item Use the $(k, l)$ pairing which gives the inner products with largest variance as the next embedding coordinate.
+%
+\item For successive coordinates, use successive blocks for $l$ (e.g., for the next coordinate, under translations and rotations, use $k=1, 2, 3$ and $l = 7, 8, 9$, compute the coordinates in \eqref{eq:vdm_coord}, and select the $(k, l)$ pairing with largest variance)
+
+\end{enumerate}
+
+\subsection{Constructing movies from registered and ordered images}
+
+Once we have registered and ordered the images, we smooth the resulting trajectory to obtain a movie of the average developmental dynamics. 
+%
+Let $x_1, \dots, x_m$ denote the set of {\em registered and ordered} images (so $x_i$ is the $i^{th}$ image in the ordered set).
+%
+We define the frame at time $\tau$, denoted $\overline{x}_{\tau}$ (where $1 \le \tau \le m$), as a (Gaussian) weighted average of the images,
+\begin{equation}
+\overline{x}_{\tau} =  \frac{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma} \right) x_i}{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma^2} \right) }
+\end{equation}
+where $\sigma$ is the scale of the Gaussian filter. 
+%
+For the images and movie in Figure~4, we take $\sigma^2 = 20$.
+%
 
@@
 
-\begin{figure}[H]
-\begin{tikzpicture}
-     \node at (0,0)
-       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding_background}};
-     \node at (0,0)
-       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding}};  
-\end{tikzpicture}
-\begin{tikzpicture}
-     \node at (0,0)
-       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding2_background}};
-     \node at (0,0)
-       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding2}};  
-\end{tikzpicture}
-\caption{Embedding for third example using two different $\epsilon$ values. The right one (smaller $\epsilon$) has a tighter structure, but the mutants are more removed.}
-\end{figure}
+%\begin{figure}[H]
+%\begin{tikzpicture}
+%     \node at (0,0)
+%       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding_background}};
+%     \node at (0,0)
+%       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding}};  
+%\end{tikzpicture}
+%\begin{tikzpicture}
+%     \node at (0,0)
+%       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding2_background}};
+%     \node at (0,0)
+%       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding2}};  
+%\end{tikzpicture}
+%\caption{Embedding for third example using two different $\epsilon$ values. The right one (smaller $\epsilon$) has a tighter structure, but the mutants are more removed.}
+%\end{figure}
 

### 2014-05-21 image_analysis_paper/test_mutants.m

@@ -529,7 +529,7 @@ set(gcf, 'paperposition',[0 0 8 8/nstages])
 for i=1:nstages
     
     make_subplot(nstages, 1,  0.01, i);
-    im1 = image_set_aligned(:,:,:,wt_draw(i));
+    im1 = image_set_aligned_withnuclei(:,:,:,wt_draw(i));
     im1(:, :, 2) = im1(:, :, 3);
     
     imshow(im1,'initialmagnification','fit','border','tight')
@@ -544,7 +544,7 @@ set(gcf, 'paperposition',[0 0 8 8/nstages])
 for i=1:nstages
     
     make_subplot( nstages,1,  0.01, i);
-    im1 = image_set_aligned(:,:,:,mut_draw(i));
+    im1 = image_set_aligned_withnuclei(:,:,:,mut_draw(i));
     im1(:, :, 2) = im1(:, :, 3);
     
     imshow(im1,'initialmagnification','fit','border','tight')
@@ -558,7 +558,7 @@ set(gcf, 'paperposition',[0 0 8/nstages 8])
 for i=1:nstages
     
     make_subplot(1, nstages,  0.01, i);
-    im1 = image_set_aligned(:,:,:,wt_draw(i));
+    im1 = image_set_aligned_withnuclei(:,:,:,wt_draw(i));
     im1(:, :, 2) = im1(:, :, 3);
     
     imshow(im1,'initialmagnification','fit','border','tight')
@@ -573,7 +573,7 @@ set(gcf, 'paperposition',[0 0 8/nstages 8])
 for i=1:nstages
     
     make_subplot(1, nstages,  0.01, i);
-    im1 = image_set_aligned(:,:,:,mut_draw(i));
+    im1 = image_set_aligned_withnuclei(:,:,:,mut_draw(i));
     im1(:, :, 2) = im1(:, :, 3);
     
     imshow(im1,'initialmagnification','fit','border','tight')

### 2014-05-26 image_analysis_paper/pnas_draft2.tex
updated mutant figures and some text for paepr

@@
 
-As the images in Figure~\ref{fig:data1} are relatively simple, other techniques could have been used to temporally order them.
+%As the images in Figure~\ref{fig:data1} are relatively simple, other techniques could have been used to temporally order them.
 %
-Because this data set is effectively a single mode (the two dpERK peaks) growing in time,
-principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data.
+Because this data set in Figure~\ref{fig:data1} are relatively simple, effectively consisting of a single mode (the two dpERK peaks) growing in time,
+established techniques such as principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data in time.
 %
@@
 %
-The next data set illustrates this point.
 
-Figure~\ref{subfig:raw_data2} shows a data set of $108$ images of {\it Drosophila} embryos.
+Figure~\ref{subfig:raw_data2} shows a richer data set of $108$ images of {\it Drosophila} embryos.
 %
@@
 %
-dpERK (red) again appears as two lateral peaks at the ventral side of the embryo.
+dpERK (red in the figure) again appears as two lateral peaks at the ventral side of the embryo.
 %
@@
 
-Visually, the dynamics in this data set are highly nonlinear, and only 22\% of the variability in the data is captured by the first principal component.
+Visually, the dynamics in this data set are much more complicated, and only 22\% of the variability in the images is captured by the first principal component; simple time-ordering based on this projection is therefore not possible.
 %
-We require a more sophisticated approach, such as vector diffusion maps, to temporally order the data.
+Even when projecting on several leading principal components (see SI), a more sophisticated, nonlinear approach, such as vector diffusion maps, 
+is required to temporally order the data.
 %
@@
 %
-Visual inspection verifies that this ordering is consistent with the known developmental dynamics outlined above.
+Visual inspection confirms this ordering is consistent with the known developmental dynamics outlined above.
 %
-However, the dynamics are not immediately apparent from this set of registered and ordered images for two reasons.
+However, assimilating the dynamics from this concatenation of registered and ordered images is difficult for two reasons.
 %
-First, the sheer number of images makes visual assimilation of the entire data set (even after registration and ordering) difficult, and
+First, the sheer number of images makes visual processing of the entire data set (even after registration and ordering) nontrivial, and
 second, the ordered data is not entirely smooth due to noise from interembryo variability.
 %
-To highlight the developmental dynamics revealed by vector diffusion maps, we construct a movie of an averaged developmental trajectory by applying a Gaussian filter to the set of ordered images (more details are given in the {\it SI Appendix}).
+To highlight the developmental dynamics revealed by vector diffusion maps, we construct a movie of an averaged developmental trajectory 
+%
+It is worth noting that in the process of making such a movie, one has to decide the (monotonic) relation between
+``movie time" (which is {\em not} physical time) and the parametrization of the data through the first VDM component; and that making a smooth movie
+requires the application of filter (we used a Gaussian one, more details are given in the {\it SI Appendix}) to the set of ordered images.
 %
 This movie is included in the {\it SI Appendix},
-and representative snapshots from this movie are shown in Figure~\ref{subfig:averaged_data2}.
+and representative snapshots from it are shown in Figure~\ref{subfig:averaged_data2}, making the ``stereotypic" dynamics readily apparent;
 %
-The stereotypic dynamics are now readily apparent.
+However, there are some artifacts in the movie due to size variation, and the dynamics towards the end are not as smooth, as we have fewer samples in this time window.
 %
-However, there are some artifacts in the movie due to size variation, the dynamics at the end of the movie are not as smooth, as we have fewer samples in this time window.
+%We would like to note that, although this movie provides us with a representative picture of the dynamics, the speed of the movie is not constant %with time, as we have no true time information about the images.
 %
-We would like to note that, although this movie provides us with a representative picture of the dynamics, the speed of the movie is not constant with time, as we have no true time information about the images.
+We have thus accomplished the same task presented in the caricature in Figure~\ref{fig:fish}, knowing neither the appropriate landmarks (e.g., fins) for registration nor the appropriate statistic (e.g., body size) by which to order.
+
+\subsection{Additional variability: Identifying distinct developmental trajectories}
+
+When time is the main source of variability,
+vector diffusion maps can temporally order data; but it can also uncover multiple sources of variability.
 %
-We have thus accomplished the same task presented in the caricature in Figure~\ref{fig:fish}, but without knowledge about the appropriate landmarks (e.g., fins) for registration or the appropriate statistic (e.g., body size) by which to order.
-
-\subsection{Identifying distinct developmental trajectories}
-
-Vector diffusion maps can not only temporally order data, but can also uncover other sources of variability.
+%Up to this point, we have assumed that our experiments contain one main direction of variability, and that this direction is parameterized by %time.
 %
-Up to this point, we have assumed that our experiments contain one main direction of variability, and that this direction is parameterized by time.
+%However, this is not always the case.
 %
-However, this is not always the case.
-%
-As an example, we have collected a data set consisting of $42$ embryos undergoing cellularization during the third hour of development (see {\it SI Appendix}); $21$ of the embryos are wild type, and $21$ of the embryos are mutant embryos where dpERK is expressed in one broad peak at the ventral side.
+To illustrate this, we have collected a data set consisting of $90$ embryos undergoing cellularization during the third hour of development (see {\it SI Appendix}); $48$ of the embryos are wild type, and $42$ of the embryos are mutant embryos where dpERK is expressed in one broad peak at the ventral side.
 %
@@
 For the previous examples, where we only uncovered developmental progression, we used the first vector diffusion maps coordinate (i.e., the x-axis of this plot) to order the data.
+%%%
+%%%YGK we need something about dmap eignvalues and how many are not harmonics
 %
-However, our data now contains two main sources of variation, and we can see two distinct trajectories, highlighted in blue and red, emerge in this two-dimensional projection.
+Here it is clear even just visually that one dimension is not enough: two dimensions are required to embed the data,
+(and this is also supported by the magnitudes of eigenvalues returned by the VDM process.
 %
-Both curves originate from a common point, and then diverge.
+Furthermore, two distinct developmental trajectories emerge in this two-dimensional embedding space, 
+one for the wiltype highlighted in blue  (to guide the eye) and one for the mutant (in red); 
+they both originate from a common point, and later diverge.
 %
-Representative images along the blue and red curves are shown in Figures~\ref{subfig:average_wt_data3}~and~\ref{subfig:average_mut_data3}.
+Representative images along the blue and red ``curves" are shown in Figures~\ref{subfig:average_wt_data3}~and~\ref{subfig:average_mut_data3}.
 %
-Figure~\ref{subfig:average_wt_data3}, corresponding to the blue curve, shows the emergence of two distinct dpERK peaks, and Figure~\ref{subfig:average_mut_data3} reveals one more concentrated dpERK peak.
+Figure~\ref{subfig:average_wt_data3}, (corresponding to the wildtype/blue), shows the emergence of two distinct dpERK peaks, and Figure~\ref{subfig:average_mut_data3} (mutant/red) reveals one more concentrated dpERK peak.
 %
 This is consistent with previously known dynamics:
-at early times, the wild type and mutant embryos both have little dpERK signal, and are therefore indistinguishable, and at later times, the two groups separate as the wild type embryos develop two distinct dpERK peaks and the mutant embryos develop only a single peak.
+at early times, the wild type and mutant embryos both have little dpERK signal, and are therefore indistinguishable;
+at later times, the two groups separate as the wild type embryos develop two distinct dpERK peaks and the mutant embryos develop only a single peak.
 %
-We have therefore shown that vector diffusion maps can not only recover dynamics, but also distinguish different expression patterns and developmental trajectories.
+We have therefore demonstrated that vector diffusion maps can not only recover dynamics, but also help distinguish different expression patterns and developmental trajectories.
 
@@
 
-We have demonstrated how vector diffusion maps, an existing reduction technique, can be used to register and temporally order images from developmental biology studies.
+We presented a unified approach to temporal ordering and registration of images in cross-sectional studies of developmental dynamics. 
 %
-We acknowledge that the tasks of both registration and ordering could have been accomplished using a variety of other techniques.
+To the best of our knowledge, the algorithmic approaches to these tasks have been pursued relatively independently of each other. 
 %
-We chose the algorithm outlined here, not because it is specifically tailored to our particular data sets, but rather, because it is sufficiently general and we are confident that it can be applied to a myriad of biological imaging applications.
+In particular, temporal ordering of imaging data was done manually for relatively small datasets, with a significant amount of human supervision, and using registered images as a starting point \cite{...}.
 %
-Furthermore, such data mining algorithms are often tested on databases containing hundreds or thousands of images.
+In parallel, ordering large-scale cross-sectional data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different species, such as expression levels of different genes \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering}.
 %
-Here, we could successfully recover the underlying dynamics with relatively small data sets.
+Automated temporal ordering of such datasets was done by first projecting them on a low-dimensional subspace, which can provided by principal or independent components, and then using this projection to solve a traveling salesman problem or construct a minimum spanning tree, providing in this way ordering for multiple developmental snapshots.
 %
-In general, the number of images required to accurately recover the dynamics will be a function of the noise in the data, as well as the dynamic range and complexity of the underlying process.
+The presented approach is different because it handles registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace. 
+%
+Furthermore, in contrast to most of the existing approaches to image registration, which rely on the information about some landmarks in the data, such as the eyes in face recognition applications, registration algorithms based on angular synchronization (such as vector diffusion maps) can readily register images even in the absence of such information.
 
-The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
+The presented approach was demonstrated by applying it to three different experimental datasets, collected specifically for this study. 
 %
-However, most of these applications rely on the definition and identification of good landmarks or features within each image.
+In the first case, correct orientation and temporal order of images was known a priori, providing a test case for the algorithm.  
 %
-For example, in face recognition, the first step is often to locate the eyes, nose, and mouth in each face, and then register images by aligning these features \cite{zhao2003face}.
+Both of these features were recovered by our technique with high accuracy. For this particular dataset, a simpler approach would suffice. 
 %
-%Temporally ordering images of a construction site over time is also based on first identifying features within the images \cite{schindler2007inferring}.
+Because of the relative simplicity of developmental dynamics (small tissue deformation and monotonic increase in the amplitude of a localized signaling pattern), temporal order of images could have been recovered by projection of registered images on the first principal component.
 %
-Because of the complexity of chemical and morphological changes in developing tissues, appropriate features may not be known {\it a priori}.
+This approach was clearly not sufficient for the second dataset, collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.  
 %
-We therefore use an algorithm which requires no features, but instead, uses only the raw pixel images and the inherent geometric symmetry of the problem.
+However, vector diffusion maps could readily extract a highly nontrivial developmental trajectory. 
 %
-%Template-based methods are also common for registration, where one optimally aligns each image to a predefined template function.
+Finally, we applied our algorithm to a heterogeneous dataset, composed of images from embryos that follow distinct developmental paths. 
 %
-%However, these methods can suffer when the images are noisy, and the success of such methods can depend strongly on the template function \cite{shatsky2009method}.
+We found that low-dimensional projection of registered data reveals these distinct paths and can correctly identify embryos of different genotypes.
+
+Angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy imaging data \cite{singer2012vector}. 
 %
+Because of the high levels of noise in these images, thousands of images are needed for successful shape reconstruction. 
+%
+Based on our work, we think that a much smaller dataset may be needed for successful reconstruction of developmental trajectories, providing the first step in solving a large number of systems level tasks in studies of embryogenesis.
+%
+In one of these tasks, the functional data analysis tools can be used for the quantitative estimation of both the average trajectory and its variability.  
+%
+Another important application is related to the joint analysis of datasets provided by different imaging modalities, such as live imaging data of tissue morphogenesis and snapshots from fixed embryos stained with probes for specific molecular components.
+%
+Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we are expect that dimensionality reduction approaches discussed in this work will find numerous applications and motivate future algorithmic advances.
 
-The task of temporal ordering has also been studied in a variety of biological applications.
+%**OLD***
 %
-Data sets such as transcriptional profiling data (where each data point is the transcriptome for a single sample, such as a cell or tissue sample from a patient) \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering} and snapshots of cells throughout the cell cycle (where each point is a vector of features, such as the amount of DNA, which quantify the cell's state)  \cite{kafri2013dynamics} have been temporally ordered to extract representative trajectories of the underlying dynamic processes.
+%We have demonstrated how vector diffusion maps, an existing reduction technique, can be used to register and temporally order images from developmental biology studies.
+%%
+%We acknowledge that the tasks of both registration and ordering could have been accomplished using a variety of other techniques.
+%%
+%We chose the algorithm outlined here, not because it is specifically tailored to our particular data sets, but rather, because it is sufficiently general and we are confident that it can be applied to a myriad of biological imaging applications.
+%%
+%Furthermore, such data mining algorithms are often tested on databases containing hundreds or thousands of images.
+%%
+%Here, we could successfully recover the underlying dynamics with relatively small data sets.
+%%
+%In general, the number of images required to accurately recover the dynamics will be a function of the noise in the data, as well as the dynamic range and complexity of the underlying process.
 %
-Many of these applications order the data by solving a traveling salesman problem (TSP) or constructing a minimum spanning tree (MST) on the data,
-expecting that these structures (the path of the traveling salesman, or the minimum spanning tree) characterize the majority of the patterns and variability within the data and provide an accurate encoding of the dynamical progression.
+%The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
+%%
+%However, most of these applications rely on the definition and identification of good landmarks or features within each image.
+%%
+%For example, in face recognition, the first step is often to locate the eyes, nose, and mouth in each face, and then register images by aligning these features \cite{zhao2003face}.
+%%
+%%Temporally ordering images of a construction site over time is also based on first identifying features within the images \cite{schindler2007inferring}.
+%%
+%Because of the complexity of chemical and morphological changes in developing tissues, appropriate features may not be known {\it a priori}.
+%%
+%We therefore use an algorithm which requires no features, but instead, uses only the raw pixel images and the inherent geometric symmetry of the problem.
+%%
+%%Template-based methods are also common for registration, where one optimally aligns each image to a predefined template function.
+%%
+%%However, these methods can suffer when the images are noisy, and the success of such methods can depend strongly on the template function \cite{shatsky2009method}.
+%%
 %
-These algorithms often use dimensionality reduction techniques (typically PCA or variants of) as a preprocessing step.
+%The task of temporal ordering has also been studied in a variety of biological applications.
+%%
+%Data sets such as transcriptional profiling data (where each data point is the transcriptome for a single sample, such as a cell or tissue sample from a patient) \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering} and snapshots of cells throughout the cell cycle (where each point is a vector of features, such as the amount of DNA, which quantify the cell's state)  \cite{kafri2013dynamics} have been temporally ordered to extract representative trajectories of the underlying dynamic processes.
+%%
+%Many of these applications order the data by solving a traveling salesman problem (TSP) or constructing a minimum spanning tree (MST) on the data,
+%expecting that these structures (the path of the traveling salesman, or the minimum spanning tree) characterize the majority of the patterns and variability within the data and provide an accurate encoding of the dynamical progression.
+%%
+%These algorithms often use dimensionality reduction techniques (typically PCA or variants of) as a preprocessing step.
+%%
+%The TSP or MST is solved/constructed not in the original data space, but rather in the lower-dimensional space of the first few principal components.
+%%
+%In these approaches, PCA serves two purposes: it denoises the data, making the TSP/MST construction more robust, and also allows for easy visualization of the TSP/MST if one only uses the first two or three principal components.
 %
-The TSP or MST is solved/constructed not in the original data space, but rather in the lower-dimensional space of the first few principal components.
+%We, instead, use vector diffusion maps to construct a one-dimensional parameterization of the data.
+%%
+%Because vector diffusion maps is a dimensionality reduction technique, it serves as a denoising step and yields smooth averaged trajectories even under noisy conditions.
+%%
+%However, because the algorithm is nonlinear, it affords us many of the same geometric flexibilies as the MST or TSP formulations.
+%%
+%Vector diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003, tenenbaum2000global, Donoho2003, Roweis2000}.
+%%
+%Diffusion maps have been shown to be more robust to noise than isomap \cite{balasubramanian2002isomap}, a path-based algorithm, and so we suspect the methods presented here may perform better than the ordering algorithms used in previous work.
 %
-In these approaches, PCA serves two purposes: it denoises the data, making the TSP/MST construction more robust, and also allows for easy visualization of the TSP/MST if one only uses the first two or three principal components.
+%Furthermore, dimensionality reduction techniques such as vector diffusion maps do not constrain us to one-dimensional parameterizations/organizations of the data, as illustrated in the mutant data set example (Figure~\ref{fig:data3}).
+%%
+%These techniques would also allow us to compare the developmental dynamics of several different mutations to see how mutations are developmentally related.
+%%
+%Other sources of variability, such as size or viewing angle, could also be uncovered using vector diffusion maps.
+%
+%After registering and temporally ordering images, one can perform a variety of post-processing tasks to gain further insight into the developmental processes.
+%%
+%We have demonstrated one such technique (Gaussian filtering) that allows us to extract a more representative developmental trajectory from the set of noisy snapshots.
+%%
+%%Constructing a smooth movie from the registered and ordered cross-sectional snapshots would also be of biological interest, and
+%Examining the top principal components in different time windows would allow one to visualize the main dynamical features along the trajectory.
+%%
+%%These questions first require that the data be registered and temporally ordered.
+%%
+%Correlating vector diffusion maps coordinates with physical variables is another post-processing task that can give physical insight into developmental dynamics.
+%%
+%One can test which physical variable, among a set of candidate physical variables (e.g., total intensity of one signal, roundness of the embryo), is most correlated with the uncovered vector diffusion maps coordinate to help understand which physical quantities change across the developmental trajectory.
+%%
+%We are confident that, as biological becomes easier and more robust, the dimensionality reduction techniques we present here will play an important role in the data analysis pipeline.
 
-We, instead, use vector diffusion maps to construct a one-dimensional parameterization of the data.
+%\begin{itemize}
+%\item What are other people doing  (maybe critically)
+%\item Movies  smoothness
+%\item Discrete symmetries like flips
+%\item Local information ? (features of the image) (associate with Fourier-Bessel and bispectra)
+%\item And in general, MULTIPLE variablities is important (mutants, guys in another cycle, etc.)
 %
-Because vector diffusion maps is a dimensionality reduction technique, it serves as a denoising step and yields smooth averaged trajectories even under noisy conditions.
-%
-However, because the algorithm is nonlinear, it affords us many of the same geometric flexibilies as the MST or TSP formulations.
-%
-Vector diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003, tenenbaum2000global, Donoho2003, Roweis2000}.
-%
-Diffusion maps have been shown to be more robust to noise than isomap \cite{balasubramanian2002isomap}, a path-based algorithm, and so we suspect the methods presented here may perform better than the ordering algorithms used in previous work.
-
-Furthermore, dimensionality reduction techniques such as vector diffusion maps do not constrain us to one-dimensional parameterizations/organizations of the data, as illustrated in the mutant data set example (Figure~\ref{fig:data3}).
-%
-These techniques would also allow us to compare the developmental dynamics of several different mutations to see how mutations are developmentally related.
-%
-Other sources of variability, such as size or viewing angle, could also be uncovered using vector diffusion maps.
-
-After registering and temporally ordering images, one can perform a variety of post-processing tasks to gain further insight into the developmental processes.
-%
-We have demonstrated one such technique (Gaussian filtering) that allows us to extract a more representative developmental trajectory from the set of noisy snapshots.
-%
-%Constructing a smooth movie from the registered and ordered cross-sectional snapshots would also be of biological interest, and
-Examining the top principal components in different time windows would allow one to visualize the main dynamical features along the trajectory.
-%
-%These questions first require that the data be registered and temporally ordered.
-%
-Correlating vector diffusion maps coordinates with physical variables is another post-processing task that can give physical insight into developmental dynamics.
-%
-One can test which physical variable, among a set of candidate physical variables (e.g., total intensity of one signal, roundness of the embryo), is most correlated with the uncovered vector diffusion maps coordinate to help understand which physical quantities change across the developmental trajectory.
-%
-We are confident that, as biological becomes easier and more robust, the dimensionality reduction techniques we present here will play an important role in the data analysis pipeline.
-
-\begin{itemize}
-\item What are other people doing  (maybe critically)
-\item Movies  smoothness
-\item Discrete symmetries like flips
-\item Local information ? (features of the image) (associate with Fourier-Bessel and bispectra)
-\item Mutant-wildtype stuff  IMPORTANT
-\item And in general, MULTIPLE variablities is important (mutants, guys in another cycle, etc.)
-
-%\item size is an issue
-\item  small tilts is an issue
-\item   could one do three-d image processing ?  of the floating thing, without the device
-\item how do we go about merging data sets ?
-\end{itemize}
+%%\item size is an issue
+%\item  small tilts is an issue
+%\item   could one do three-d image processing ?  of the floating thing, without the device
+%\item how do we go about merging data sets -- microscopes setting / more generally -- different observables?
+%\end{itemize}
 
@@
 \begin{materials}
+
+\textbf{Fly strain and whole-mount immunostaining.}
+%
+Oregon-R (OreR) and spaghetti squash (Sqh)-GFP were used as wild type strains. 
+%
+{\it Sna \textsuperscript{IIG05}/Cyo hb-lacZ} flies were used to study embryos with one peak of dpERK expression.  
+%
+Embryos were collected and fixed at 22$^\circ$~C. 
+%
+Monoclonal rabbit anti-dpERK (1:100, Cell signaling), mouse anti-Dorsal (1:100, Developmental Studies Hybridoma Bank), rat anti-Twist (1:500, a gift from Eric Wieschaus), and mouse anti-beta-galactosidase (1:500, DSHB) were used to stain proteins of interest.
+DAPI (1:10,000, Vector Laboratories) was used to visualize nuclei, and Alexa Fluors (1:500, Invitrogen) were used as secondary antibodies.
+
+\textbf{Microscopy.}
+%
+Nikon A1-RS scanning confocal microscope, and the Nikon 60X Plan-Apo oil objective was used to image embryos.
+Embryos were collected, stained, and imaged together under the same microscope setting.
+End-on imaging was performed by using the microfluidics device described previously.
+Images were collected at the focal plane $\sim$90~$\mu m$ from the posterior pole of an embryo.
+
+\textbf{Development time during cellularization.} 
+%
+The progression of membrane invagination during cellularization was used to assign time point to each embryo in the third hour of development.
+Time-lapse imaging of Sqh-GFP embryo was used to generate correlation curve between the membrane ingression length and the developmental time (n=7) (Sup Fig.
+1B).
+The rate at which membrane invaginates was highly reproducible from embryo to embryo.
+In the experiments with fixed embryos, we measured the membrane length by co-staining embryos with a membrane marker (Sqh-GFP) or by taking phase-contrast images (Sup Fig.
+1A).
+We used the correlation curve to assign each embryo to a time window of 2--3~minutes.
 

### 2014-05-26 image_analysis_paper/make_figures_small_dataset_color.m

@@ -118,17 +118,36 @@ for i=1:m
     image_set_aligned(:,:,:,i) = im1;
 end
 
+image_set_raw_aligned = zeros(size(image_set_raw), 'uint8');
+for i=1:m
+    im_tmp = image_set_raw(:,:,:,i);
+    im1 = rotate_image(R_opt(dim*(i-1)+1:dim*i,:), im_tmp, angle_proj);
+    image_set_raw_aligned(:,:,:,i) = im1;
+end
+
 figure;
+set(gcf, 'paperunits', 'centimeters')
+set(gcf, 'papersize', [8 8])
 set(gcf, 'paperposition',[0 0 8 8])
 for i=1:m
-    im1 = image_set_aligned(:,:,:,i);
+    im1 = image_set_raw_aligned(:, :, :, i);
+    
+    nuclei_tmp = im1(:, :, 1);
+    
+    im1(:, :, 1) = im1(:, :, 2);
+    im1(:, :, 2) = im1(:, :, 3);
+    im1(:, :, 3) = 0;
+    
+    for j=1:3
+        im1(:,:,j) = im1(:,:,j) + nuclei_tmp;
+    end
     
     %subplot(subplot_dim1, subplot_dim2, i);
     subplot('position', [X(i)-1/subplot_dim1 Y(i)-1/subplot_dim2 1/subplot_dim1-0.01 1/subplot_dim2-0.01])
     imshow(im1);
     
 end
-%print(sprintf('%s/registered_unordered_2d', im_save_dir), fmt, res);
+print(sprintf('%s/images_angsynch', im_save_dir), fmt, res);
 
 %% order
 
@@ -154,15 +173,28 @@ end
 
 [~, I] = sort(V(:,2));
 figure;
-set(gcf, 'paperposition', [0 0 8 8])
+set(gcf, 'paperunits', 'centimeters')
+set(gcf, 'papersize', [8 8])
+set(gcf, 'paperposition',[0 0 8 8])
 for i=1:m
-    %subplot(subplot_dim1, subplot_dim2, i);
-    subplot('position', [X(i)-1/subplot_dim1 Y(i)-1/subplot_dim2 1/subplot_dim1 1/subplot_dim2])
+    im1 = image_set_raw_aligned(:, :, :, I(i));
     
-    imshow(image_set_aligned(:,:,:,I(i)));
+    nuclei_tmp = im1(:, :, 1);
+    
+    im1(:, :, 1) = im1(:, :, 2);
[…]

### 2014-05-26 image_analysis_paper/test_FBsPCA.m

@@ -69,7 +69,7 @@ for i=1:m
     make_subplot(subplot_dim1, subplot_dim2, 0.01, i);
     imshow(uint8(data(:, :, i)));
 end
-saveas(gcf,sprintf('%s/FBsPCA_images', im_save_dir), 'pdf')
+% saveas(gcf,sprintf('%s/FBsPCA_images', im_save_dir), 'pdf')
 
 
 figure;
@@ -156,6 +156,9 @@ eps_dmaps = median(W_dmaps(:))/2;
 [V, D] = dmaps(W_dmaps, eps_dmaps, 10);
 
 figure;
+imshow(W_dmaps(ranks_from_membranes,ranks_from_membranes))
+
+figure;
 set(gcf, 'paperunits', 'centimeters')
 set(gcf, 'papersize', [16 16])
 set(gcf, 'paperposition',[0 0 16 16 ]);
@@ -165,7 +168,7 @@ for i=2:5
     xlabel(sprintf('\\phi_%d',i));
     ylabel('membrane thickness')
 end
-saveas(gcf,sprintf('%s/FBsPCA_dmaps', im_save_dir), 'pdf')
+% saveas(gcf,sprintf('%s/FBsPCA_dmaps', im_save_dir), 'pdf')
 
 
 %%

### 2014-05-26 image_analysis_paper/test_mutants_big.m

@@ -52,7 +52,7 @@ for i=1:m
     
     im1(:, :, 1) = im1(:, :, 2);
     %im1(:, :, 2) = uint8(0.5*double(im1(:, :, 3)));
-    im1(:, :, 2) = 0;
+    im1(:, :, 2) = im1(:, :, 3);
     im1(:, :, 3) = 0;
     
     %store image
@@ -69,18 +69,20 @@ shift_max = 10;
 shift_step = 2;
 dim = 3;
 
-matlabpool open 2;
-[R, W] = compute_pairwise_alignments_color(image_set, angle_proj, shift_max, shift_step);
-matlabpool close
+% matlabpool open 2;
+% [R, W] = compute_pairwise_alignments_color(image_set, angle_proj, shift_max, shift_step);
+% matlabpool close
 
 %save('pairwise_alignments_snanull.mat', 'R', 'W');
-% load('pairwise_alignments_snanull.mat');
+load('pairwise_alignments_snanull.mat');
 
 
 %% select which images to use
 
-ind = setdiff(1:m, [17]);
-
+% ind = setdiff(1:m, [17]);
+% ind = setdiff(1:m, [73 37 34 29 15 21 87 62 17 55 96 18]);
+% ind = setdiff(1:m, [37 34 29 15 21 2 4 6]);
+ind = setdiff(1:m, [37 34 29 15 21 2 4 6]);
 mutation = mutation(ind);
 
 image_set = image_set(:, :, :, ind);
@@ -97,112 +99,132 @@ m = length(ind);
 subplot_dim1 = floor(sqrt(m));
 subplot_dim2 = ceil(m / subplot_dim1);
 
+figure;
+set(gcf, 'paperunits', 'centimeters')
+set(gcf, 'papersize', [subplot_dim1 subplot_dim2])
+set(gcf, 'paperposition',[0 0 subplot_dim1 subplot_dim2])
+for i=1:m
+    im1 = image_set(:,:,:,i);
+    
+    for j=1:3
+        im1(:,:,j) = im1(:,:,j) + nuclei(:,:,i);
+    end
+    
+    make_subplot(subplot_dim1, subplot_dim2, 0.01, i);
+    imshow(im1);
+    
+end
+saveas(gcf,sprintf('%s/raw_data3b', im_save_dir), 'pdf')
+
 %% synchronization
 
[…]

### 2014-05-27 image_analysis_paper/pnas_draft2.tex
added little edits to stuff

@@
 %
-To illustrate this, we have collected a data set consisting of $90$ embryos undergoing cellularization during the third hour of development (see {\it SI Appendix}); $48$ of the embryos are wild type, and $42$ of the embryos are mutant embryos where dpERK is expressed in one broad peak at the ventral side.
+To illustrate this, we have collected a data set consisting of $42$ embryos undergoing cellularization during the third hour of development (see {\it SI Appendix}); $21$ of the embryos are wild type, and $21$ of the embryos are mutant embryos where dpERK is expressed in one broad peak at the ventral side.
 %
@@
 %
+In addition, the task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
+%
 The presented approach is different because it handles registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace. 
 %
-Furthermore, in contrast to most of the existing approaches to image registration, which rely on the information about some landmarks in the data, such as the eyes in face recognition applications, registration algorithms based on angular synchronization (such as vector diffusion maps) can readily register images even in the absence of such information.
+Furthermore, in contrast to most of the existing approaches to image registration, which rely on the information about some landmarks in the data, such as the eyes in face recognition applications \cite{zhao2003face}, registration algorithms based on angular synchronization (such as vector diffusion maps) can readily register images even in the absence of such information.
 
@@
 %
-In the first case, correct orientation and temporal order of images was known a priori, providing a test case for the algorithm.  
+In the first case, correct orientation and temporal order of images was known {\it a priori}, which allowed us to validate the results of the algorithm.
 %
-Both of these features were recovered by our technique with high accuracy. For this particular dataset, a simpler approach would suffice. 
+Both the orientation and temporal order were recovered with high accuracy. 
 %
-Because of the relative simplicity of developmental dynamics (small tissue deformation and monotonic increase in the amplitude of a localized signaling pattern), temporal order of images could have been recovered by projection of registered images on the first principal component.
+However, for this particular dataset, a simpler approach would suffice. 
 %
-This approach was clearly not sufficient for the second dataset, collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.  
+Because of the relative simplicity of developmental dynamics (monotonic increase in the amplitude of a localized signaling pattern), temporal order could have been recovered by projection of registered images on the first principal component.
+%
+However, PCA would clearly not suffice for the second dataset, collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.  
 %
@@
 
-Angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy imaging data \cite{singer2012vector}. 
+Angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy imaging data \cite{shkolnisky2012viewing, singer2012vector}. 
 %
-Because of the high levels of noise in these images, thousands of images are needed for successful shape reconstruction. 
+Due to the high levels of noise in these images, thousands of images are needed for successful shape reconstruction. 
 %
@@
 
-%\begin{itemize}
-%\item What are other people doing  (maybe critically)
-%\item Movies  smoothness
-%\item Discrete symmetries like flips
-%\item Local information ? (features of the image) (associate with Fourier-Bessel and bispectra)
-%\item And in general, MULTIPLE variablities is important (mutants, guys in another cycle, etc.)
-%
-%%\item size is an issue
-%\item  small tilts is an issue
-%\item   could one do three-d image processing ?  of the floating thing, without the device
-%\item how do we go about merging data sets -- microscopes setting / more generally -- different observables?
-%\end{itemize}
+\begin{itemize}
+\item What are other people doing  (maybe critically)
+\item Movies  smoothness
+\item Discrete symmetries like flips
+\item Local information ? (features of the image) (associate with Fourier-Bessel and bispectra)
+\item And in general, MULTIPLE variablities is important (mutants, guys in another cycle, etc.)
+
+\item size is an issue
+\item  small tilts is an issue
+\item   could one do three-d image processing ?  of the floating thing, without the device
+\item how do we go about merging data sets -- microscopes setting / more generally -- different observables?
+\end{itemize}
 

### 2014-05-27 image_analysis_paper/si_information.tex
updated mutant figures and some text for paepr
added little edits to stuff

@@
 \begin{figure}[H]
-\includegraphics[height=9cm]{raw_data3}
+\includegraphics[height=9cm]{raw_data3b}
 \caption{Images of cellularizing {\it Drosophila} embryos.
 Each image is of a different embryo in a different rotational orientation.
-Half of the images are of wild type embryos, and half of the images are of mutant embryos.}
+$48$ of the images are of wild type embryos, and $42$ of the images are of mutant embryos.}
 \end{figure}
@@
 %
-$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/10^{th}$ of the median of the pairwise distances between data points for the first two data sets, and the median of the pairwise distances between data points for the third data set.
+$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/10^{th}$ of the median of the pairwise distances between data points for the first two data sets, and half of the median of the pairwise distances between data points for the third data set.
 
@@
 
-If we assume that the rotations and the dynamics are uncoupled and therefore separable, then the eigenvectors of $S$ have the following structure: each block eigenvector contains estimates of the optimal rotations multiplied by the corresponding embedding coordinate.
+In general, the embedding coordinates are given by
+\begin{equation} \label{eq:vdm_coord}
+\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle,
+\end{equation}
+where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$, 
+%
+If we assume that the rotations and the dynamics are uncoupled and therefore separable, then the eigenvectors of $S$ have the following structure: each block eigenvector contains estimates of the optimal rotations (up to a constant rotation) multiplied by the corresponding embedding coordinate.
 %
@@
 %
-As rotation matrices are orthonormal, many of these will be zero, 
-and the nonzero inner products give us the embedding coordinates.
-%
-More specifically, the $j^{th}$ embedding coordinate will be given by
-\begin{equation} \label{eq:vdm_coord}
-\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle, \: 0 < k \le d, (j-1)d < l \le jd
-\end{equation}
-where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$, 
-and we select values of $k$ and $l$ such that the coordinate $\psi_{k, l}$ has the largest variability. 
+The $j^{th}$ embedding coordinate will be given by $\psi_{k,l}$ such that $0 < k \le d, jd +1 < l \le (j+1)d$,
+and we select the $(k,l)$ pair such that the coordinate $\psi_{k, l}$ has the largest variability. 
 %
@@
 \item Construct the corresponding rotation matrices $R_{ij}$ for every pair $i, j$ of images (see \eqref{eq:R_theta} for rotations or \eqref{eq:total_rot} for rotations + translations).
-\item Construct the matrix $H$ as in \eqref{eq:H_to_R}.
+\item Construct the matrix $H$ as in \eqref{eq:H_to_R}, which is a block matrix with entries $R_{ij}$.
 \item Compute the top $2$ (for rotations) or $3$ (for rotations + translations) eigenvectors of $H$.
@@
 \item Compute the top few eigenvectors of $A$, ordered with decreasing eigenvalues
-\item If we only want to order the data, then sort the data by the value of the entries in the second eigenvector
-\item For higher-dimensional embeddings (e.g., the mutants), we use the entries of successive eigenvectors (the second eigenvector, the third eigenvector, etc.) as the embedding coordinates
+\item If we only want to order the data, then sort the data by the entries in the second eigenvector.
+\item For higher-dimensional embeddings (e.g., the mutants), use the entries of successive eigenvectors (the second eigenvector, the third eigenvector, etc.) as the embedding coordinates
 \end{enumerate}
@@
 %
+\item Construct the matrix $H$ as in \eqref{eq:H_low_rank}.
+%
 \item Construct the matrix $W$ as in \eqref{eq:dmaps_W}, using the minimum distance obtained for each pair in step 1 as the pairwise distances in the Gaussian kernel.
-%
-\item Construct the matrix $H$ as in \eqref{eq:H_low_rank}.
 %
@@
 %
-\item To compute the first (non-trivial) embedding coordinate for the data, compute coordinates as in \eqref{eq:vdm_coord}, taking $k=1, 2$ and $l=3, 4$ for the rotation case, or $k = 1, 2, 3$ and $l = 4, 5, 6$ for the rotation + translation case.
+\item To compute the first (non-trivial) embedding coordinate for the data, compute $\psi_{k,l}$ as in \eqref{eq:vdm_coord}, taking $k=1, 2$ and $l=3, 4$ for the rotation case, or $k = 1, 2, 3$ and $l = 4, 5, 6$ for the rotation + translation case.
 %
-\item Use the $(k, l)$ pairing which gives the inner products with largest variance as the next embedding coordinate.
+Select the $(k, l)$ coordinate which has the largest variance as the embedding coordinate.
 %
-\item For successive coordinates, use successive blocks for $l$ (e.g., for the next coordinate, under translations and rotations, use $k=1, 2, 3$ and $l = 7, 8, 9$, compute the coordinates in \eqref{eq:vdm_coord}, and select the $(k, l)$ pairing with largest variance)
+\item To compute subsequent coordinates, use lower blocks for $l$ (e.g., for the next coordinate, under translations and rotations, use $k=1, 2, 3$ and $l = 7, 8, 9$).
+Again, compute $\psi_{k, l}$ as in \eqref{eq:vdm_coord}, and select the $(k, l)$ coordinate with largest variance.
 
@@
 \begin{equation}
-\overline{x}_{\tau} =  \frac{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma} \right) x_i}{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma^2} \right) }
+\overline{x}_{\tau} =  \frac{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma^2} \right) x_i}{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma^2} \right) }
 \end{equation}
@@
 
-\begin{figure}[H]
-\includegraphics[width=5cm]{data1_coord_var}
-\includegraphics[width=5cm]{data2_coord_var}
-\includegraphics[width=5cm]{data3_coord_var}
-\caption{Variance of $\psi_{k,l}$ for three examples. We use the ones that have the largest variance.}
-\end{figure}
+%\begin{figure}[H]
+%\includegraphics[width=5cm]{data1_coord_var}
+%\includegraphics[width=5cm]{data2_coord_var}
+%\includegraphics[width=5cm]{data3_coord_var}
+%\caption{Variance of $\psi_{k,l}$ for three examples. We use the ones that have the largest variance.}
+%\end{figure}
 

### 2014-05-27 image_analysis_paper/test_mutants_big.m

@@ -290,7 +290,7 @@ end
 %%
 
 mut_draw = [27 26 88 79 65];
-wt_draw = [3 20 70 58 28];
+wt_draw = [3 20 70 28 2];
 nstages = 5;
 
 wt_ind = find(embed_coord(:, idx2) > 0);
@@ -302,7 +302,7 @@ hold on
 plot(embed_coord(mut_ind,idx1),embed_coord(mut_ind,idx2),'o')
 
 p_wt = polyfit(embed_coord(wt_ind,idx1),embed_coord(wt_ind,idx2),1);
-p_mut = polyfit(embed_coord(mut_ind,idx1),embed_coord(mut_ind,idx2),1);
+p_mut = polyfit(embed_coord(mut_ind,idx1),embed_coord(mut_ind,idx2),2);
 
 x_wt = linspace(-0.002, 0.02, 100);
 x_mut = linspace(-0.002, 0.055, 100);
@@ -326,11 +326,11 @@ hold on
 plot(embed_coord(mut_draw,idx1),embed_coord(mut_draw,idx2),'ok', 'markersize', 5, 'linewidth', 2)
 % curve_delta_wt = std(embed_coord(wt_ind,idx2)-polyval(p_wt, embed_coord(wt_ind,idx2)));
 % curve_delta_mut = std(embed_coord(mut_ind,idx2)-polyval(p_mut, embed_coord(mut_ind,idx2)));
-curve_delta_wt = 0.01;
-curve_delta_mut = 0.01;
+% curve_delta_wt = 0.01;
+% curve_delta_mut = 0.006;
 % patch([x_wt fliplr(x_wt)], [polyval(p_wt,x_wt)+curve_delta_wt polyval(p_wt,fliplr(x_wt))-curve_delta_wt], 'b', 'facealpha', 0.5, 'edgecolor','none')
 % patch([x_mut fliplr(x_mut)], [polyval(p_mut,x_mut)+curve_delta_mut polyval(p_mut,fliplr(x_mut))-curve_delta_mut], 'r', 'facealpha', 0.5, 'edgecolor','none')
-axis([-0.002 0.055 -0.06 0.02])
+axis([-0.002 0.055 -0.06 0.03])
 set(gca, 'xtick', [])
 set(gca, 'ytick', [])
 xlabel('first VDM coordinate')
@@ -352,11 +352,11 @@ set(gcf, 'paperposition',[0 0 8 6])
[…]

### 2014-05-27 image_analysis_paper/fig5.tex
updated mutant figures and some text for paepr
edited paper; added references, changed figure

@@
 
-\usepackage[margin=0in, paperwidth=8.7cm, paperheight=9.5cm]{geometry}
+%\usepackage[margin=0in, paperwidth=8.7cm, paperheight=9.5cm]{geometry}
+\usepackage[margin=0in, paperwidth=8.7cm, paperheight=16.5cm]{geometry}
 
@@
 \centering
-%\raisebox{5.4cm}{\textsf{A}}
+%%\raisebox{5.4cm}{\textsf{A}}
+%\begin{tikzpicture}
+%     \node at (0,0)
+%       {\includegraphics[width=8cm]{mut_wt_vdm_embedding_background}};
+%     \node at (0,0)
+%       {\includegraphics[width=8cm]{mut_wt_vdm_embedding}};
+%      \node at (-4cm, 2.4cm) {\textsf{A}};
+%\end{tikzpicture}\\
+\raisebox{5.2cm}{\textsf{A}}
+\includegraphics[width=6cm, angle=90]{data3_embed_bw}\\
+\raisebox{5.2cm}{\textsf{B}}
+\includegraphics[width=8cm]{data3_embed_color}\\
 \begin{tikzpicture}
-     \node at (0,0)
-       {\includegraphics[width=8cm]{mut_wt_vdm_embedding_background}};
-     \node at (0,0)
-       {\includegraphics[width=8cm]{mut_wt_vdm_embedding}};
-      \node at (-4cm, 2.4cm) {\textsf{A}};
+	\node at (-4.25cm, 0cm) {\textsf{C}};
+	\node at (0,0) {\includegraphics[width=8cm]{wt_trajectory}};
+	%\draw[help lines,step=.5] (-4,-1) grid (4,1);
+	\draw[->, color=white, thick] (-1.1cm, -0.1cm) -- (-1.3cm, -0.3cm);
+	\draw[->, color=white, thick] (-0.7cm, -0.1cm) -- (-0.5cm, -0.2cm);
+
+	\draw[->, color=white, thick] (0.75cm, -0.1cm) -- (0.6cm, -0.3cm);
+	\draw[->, color=white, thick] (1.25cm, -0.1cm) -- (1.4cm, -0.3cm);
+
+	\draw[->, color=white, thick] (2.8cm, -0.1cm) -- (2.7cm, -0.3cm);
+	\draw[->, color=white, thick] (3.3cm, -0.1cm) -- (3.5cm, -0.3cm);
+
 \end{tikzpicture}\\
-\raisebox{0.6cm}{\textsf{B}}
-\includegraphics[width=8cm]{wt_trajectory}\\
-\raisebox{0.6cm}{\textsf{C}}
-\includegraphics[width=8cm]{mut_trajectory}
+\begin{tikzpicture}
+	\node at (-4.25cm, 0cm) {\textsf{D}};
+	\node at (0,0) {\includegraphics[width=8cm]{mut_trajectory}};
+
+	\draw[->, color=white, thick] (-1cm, -0.2cm) -- (-1cm, -0.5cm);
+
+	\draw[->, color=white, thick] (1cm, -0.2cm) -- (1cm, -0.5cm);
+
+	\draw[->, color=white, thick] (3cm, -0.2cm) -- (3cm, -0.5cm);
+
+\end{tikzpicture}
 

### 2014-05-27 image_analysis_paper/pnas_draft2.tex
edited paper; added references, changed figure

--- 2014-05-26 image_analysis_paper/pnas_draft2.tex
@@
 %
-In longitudinal studies, developmental progress is monitored over time for the same embryo.
+In longitudinal studies, developmental progress is monitored over time for the same embryo \cite{roelens2013live, keller2013imaging}.
 %
-In a cross-sectional study, developmental dynamics must be reconstructed from multiple embryos, each of which contributes only a single snapshot of a chemical or morphological process along its developmental trajectory.
+In a cross-sectional study, developmental dynamics must be reconstructed from multiple embryos, each of which contributes only a single snapshot of a chemical or morphological process along its developmental trajectory \cite{jaeger2004dynamic, fowlkes2008quantitative}.
 %
@@
 Importantly, the ``age'' of any given embryo arrested in its development is not quantitatively known; typically what is known is
-a certain time window to which a collection of embryos belongs.
+a certain time window to which a collection of embryos belongs, e.g., \cite{ng2012large, richardson2014emage, castro2009automatic}.
 %to high accuracy.
@@
 %
-Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, and so have dimensionality reduction techniques \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear} and their applications \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering, kafri2013dynamics}, especially in the last decade.
+Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, as have dimensionality reduction techniques \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear} and their applications \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering, kafri2013dynamics}, especially in the last decade.
 %
-Here, we use a manifold learning algorithm \cite{singer2012vector} for high-dimensional imaging data.% which both factors out geometric symmetries to register the images and extracts low-dimensional structure in the imaging data set.
+Here, we use a manifold learning algorithm \cite{singer2012vector} for high-dimensional imaging data.
+% which both factors out geometric symmetries to register the images and extracts low-dimensional structure in the imaging data set.
 %
@@
 %
-We analyze three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics.
+We analyze three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics \cite{jaeger2012drosophila}.
 %
@@
 
-\subsection{Images of {\subsectionitfont Drosophila} embryos during cellularization}
+\subsection{Images of embryos during cellularization}
 
@@
 
-\subsection{Images of gastrulating  {\subsectionitfont Drosophila} embryos}
+\subsection{Images of gastrulating embryos}
 
@@
 ``movie time" (which is {\em not} physical time) and the parametrization of the data through the first VDM component; and that making a smooth movie
-requires the application of filter (we used a Gaussian one, more details are given in the {\it SI Appendix}) to the set of ordered images.
+requires the application of filter (we used a Gaussian filter, more details are given in the {\it SI Appendix}) to the set of ordered images.
 %
@@
 %
-However, there are some artifacts in the movie due to size variation, and the dynamics towards the end are not as smooth, as we have fewer samples in this time window.
+%However, there are some artifacts in the movie due to size variation, and the dynamics towards the end are not as smooth, as we have fewer samples in this time window.
 %
@@
 
+We computed the vector diffusion maps embedding for our data.
+%
+However, in this example, the eigenvalue spectrum from vector diffusion maps (see {\it SI Appendix}) shows that two coordinates may be required to obtain an adequate description of the data. 
+%
 Figure~\ref{subfig:projection_data3} shows the data, projected onto the first two vector diffusion maps coordinates;
@@
 %
-Here it is clear even just visually that one dimension is not enough: two dimensions are required to embed the data,
-(and this is also supported by the magnitudes of eigenvalues returned by the VDM process.
+In this two-dimensional embedding space, two distinct branches emerge, which originate from a common point.
+
+Figure~\ref{subfig:projection_colored_data3} shows the same two-dimensional projection as Figure~\ref{subfig:projection_data3}, now colored by species.
 %
-Furthermore, two distinct developmental trajectories emerge in this two-dimensional embedding space, 
-one for the wiltype highlighted in blue  (to guide the eye) and one for the mutant (in red); 
-they both originate from a common point, and later diverge.
+Clearly, the two branches of the plot correspond to the mutant and wild type data.
 %
-Representative images along the blue and red ``curves" are shown in Figures~\ref{subfig:average_wt_data3}~and~\ref{subfig:average_mut_data3}.
+Selected images along the wild type and mutant trajectories are shown in Figures~Figures~\ref{subfig:average_wt_data3}~and~\ref{subfig:average_mut_data3}, respectively.
 %
-Figure~\ref{subfig:average_wt_data3}, (corresponding to the wildtype/blue), shows the emergence of two distinct dpERK peaks, and Figure~\ref{subfig:average_mut_data3} (mutant/red) reveals one more concentrated dpERK peak.
+We can see that the lower-right portion of the embedded space, where the two species are mixed, corresponds to early in the developmental trajectory, where there is little signaling in the wild type and thew mutant.
 %
-This is consistent with previously known dynamics:
-at early times, the wild type and mutant embryos both have little dpERK signal, and are therefore indistinguishable;
-at later times, the two groups separate as the wild type embryos develop two distinct dpERK peaks and the mutant embryos develop only a single peak.
+When the two paths diverge, we can see that the wild type data develops two distinct peaks, whereas the mutant data develops a single peak.
 %
@@
 {\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in {\xfigtextfontit SI Appendix}), projected onto the first two vector diffusion maps coordinates.
-The two distinct trajectories are highlighted in blue and red.
-{\xfigtextfontit (B)} dpERK signal at select wild type points in {\xfigtextfontit A} (denoted by bold symbols).
-{\xfigtextfontit (C)} dpERK signal at select mutant points in {\xfigtextfontit A} (denoted by bold symbols).}
+{\xfigtextfontit (B)} The data in {\xfigtextfontit A}, now colored by species.
+The two ``branches'' clearly correspond to mutant and wild type samples.
+{\xfigtextfontit (C)} dpERK signal at select wild type points in {\xfigtextfontit B} (denoted by circles).
+The two dpERK peaks which arise during development are indicated by arrows.
+{\xfigtextfontit (C)} dpERK signal at select mutant points in {\xfigtextfontit B} (denoted by circle).
+The single dpERK peak which arises is indicated by an arrow.
+}
 \customlabel{fig:data3}{5}
 \customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
-\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it B}}
-\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
+\customlabel{subfig:projection_colored_data3}{\ref{fig:data3}{\it B}}
+\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it C}}
+\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it D}}
 \end{figure}
@@
 %
-To the best of our knowledge, the algorithmic approaches to these tasks have been pursued relatively independently of each other. 
+To the best of our knowledge, algorithmic approaches to these two tasks have primarily been explored independently of each other. 
 %
-In particular, temporal ordering of imaging data was done manually for relatively small datasets, with a significant amount of human supervision, and using registered images as a starting point \cite{...}.
+In particular, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point, e.g., \cite{yuan2014automated, surkova2008characterization}.
 %
-In parallel, ordering large-scale cross-sectional data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different species, such as expression levels of different genes \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering}.
+In parallel, ordering of large-scale cross-sectional data was done in the context of molecular profiling studies that lacked spatial resolution, in which data are vectors describing the amounts of different chemical species such as mRNA abundances \cite{trapnell2014dynamics, anavy2014blind, gupta2008extracting}.
 %
-Automated temporal ordering of such datasets was done by first projecting them on a low-dimensional subspace, which can provided by principal or independent components, and then using this projection to solve a traveling salesman problem or construct a minimum spanning tree, providing in this way ordering for multiple developmental snapshots.
+Automated temporal ordering of such data sets was accomplished by first projecting the data on a low-dimensional subspace spanned by the leading principal or independent components, and then solving a traveling salesman problem or constructing a minimum spanning tree on the projected data to order multiple snapshots.
 %
-The presented approach is different because it handles registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace. 
+The approach presented here is different because it addresses image registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace. 
 %
-Furthermore, in contrast to most of the existing approaches to image registration, which rely on the information about some landmarks in the data, such as the eyes in face recognition applications, registration algorithms based on angular synchronization (such as vector diffusion maps) can readily register images even in the absence of such information.
+The task of image registration has also been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
+%
+In contrast to most of the existing approaches to image registration which rely on the knowledge about some landmarks in the data \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization (such as vector diffusion maps) can register images even in the absence of such information.
+%TODO: look at Eisen Cell paper about registration
 
-The presented approach was demonstrated by applying it to three different experimental datasets, collected specifically for this study. 
+The presented approach was demonstrated on three different experimental data sets, collected specifically for this study. 
 %
-In the first case, correct orientation and temporal order of images was known a priori, providing a test case for the algorithm.  
+In the first example, correct orientation and temporal order of images was known {\it a priori}, providing a test case for the algorithm in which both of these features were recovered with high accuracy.
 %
-Both of these features were recovered by our technique with high accuracy. For this particular dataset, a simpler approach would suffice. 
+For this particular data set, a simpler approach would suffice. 
 %
-Because of the relative simplicity of developmental dynamics (small tissue deformation and monotonic increase in the amplitude of a localized signaling pattern), temporal order of images could have been recovered by projection of registered images on the first principal component.
+Indeed, because of the relative simplicity of developmental dynamics (monotonic increase in the amplitude of a localized signaling pattern and relatively small tissue deformation), temporal order could have been inferred by projection of registered images on the first principal component of the data set.
 %
-This approach was clearly not sufficient for the second dataset, collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.  
+This approach was clearly not sufficient for the second data set (see Figure~S3), collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.
 %
-However, vector diffusion maps could readily extract a highly nontrivial developmental trajectory. 
+However, vector diffusion maps could readily extract a highly nontrivial spatiotemporal trajectory. 
 %
-Finally, we applied our algorithm to a heterogeneous dataset, composed of images from embryos that follow distinct developmental paths. 
+Finally, the algorithm was applied to a heterogeneous data set, composed of images from embryos that follow distinct developmental paths. 
 %
-We found that low-dimensional projection of registered data reveals these distinct paths and can correctly identify embryos of different genotypes.
+We found that low-dimensional projection of registered data reveals these distinct paths and can correctly distinguish embryos of different genotypes.
 
-Angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy imaging data \cite{singer2012vector}. 
+Angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy imaging data \cite{singer2012vector, singer2011three}. 
 %
@@
 %
-Based on our work, we think that a much smaller dataset may be needed for successful reconstruction of developmental trajectories, providing the first step in solving a large number of systems level tasks in studies of embryogenesis.
+Based on our work, we expect that a much smaller data set may be sufficient for successful reconstruction of developmental trajectories, providing the first step in systems level studies of embryogenesis.
 %
-In one of these tasks, the functional data analysis tools can be used for the quantitative estimation of both the average trajectory and its variability.  
+In the future, functional data analysis tools can be used for the quantitative estimation of both the average trajectory and its variability \cite{silverman2005functional}.  
 %
-Another important application is related to the joint analysis of datasets provided by different imaging modalities, such as live imaging data of tissue morphogenesis and snapshots from fixed embryos stained with probes for specific molecular components.
+One can also test which physical variable, among a set of candidate physical variables, is most correlated with the uncovered vector diffusion maps coordinate to help understand which physical quantities change across the developmental trajectory, (e.g., total intensity of one signal, roundness of the embryo).
 %
-Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we are expect that dimensionality reduction approaches discussed in this work will find numerous applications and motivate future algorithmic advances.
+Another important application is related to the joint analysis of data sets provided by different imaging approaches, such as live imaging data of tissue morphogenesis and snapshots of cell signaling and gene expression from fixed embryos \cite{krzic2012multiview, ichikawa2014live, rubel2010coupling}.
+%
+Here, we have used vector diffusion maps to register the images with respect to translations and rotations; however, it is conceviable that one could factor out other sources of variability, such as reflections or scaling, in the images.
+%
+Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems and using multiple imaging frameworks, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.
 
@@
 %
-%%\item size is an issue
+%\item size is an issue
 %\item  small tilts is an issue
@@
 %
-{\it Sna \textsuperscript{IIG05}/Cyo hb-lacZ} flies were used to study embryos with one peak of dpERK expression.  
+{\it Sna\textsuperscript{\it IIG05}/Cyo hb-lacZ} flies were used to study embryos with one peak of dpERK expression.  
 %
-Embryos were collected and fixed at 22$^\circ$~C. 
+Embryos were collected and fixed at 22$^\circ$C. 
 %
@@
 End-on imaging was performed by using the microfluidics device described previously.
-Images were collected at the focal plane $\sim$90~$\mu m$ from the posterior pole of an embryo.
+Images were collected at the focal plane $\sim$90~$\mathrm{\mu m}$ from the posterior pole of an embryo.
 
@@
 The progression of membrane invagination during cellularization was used to assign time point to each embryo in the third hour of development.
-Time-lapse imaging of Sqh-GFP embryo was used to generate correlation curve between the membrane ingression length and the developmental time (n=7) (Sup Fig.
-1B).
+Time-lapse imaging of Sqh-GFP embryo was used to generate correlation curve between the membrane ingression length and the developmental time (n=7) (Figure S1B).
 The rate at which membrane invaginates was highly reproducible from embryo to embryo.
-In the experiments with fixed embryos, we measured the membrane length by co-staining embryos with a membrane marker (Sqh-GFP) or by taking phase-contrast images (Sup Fig.
-1A).
+In the experiments with fixed embryos, we measured the membrane length by co-staining embryos with a membrane marker (Sqh-GFP) or by taking phase-contrast images (Figure S1A).
 We used the correlation curve to assign each embryo to a time window of 2--3~minutes.

### 2014-05-27 image_analysis_paper/si_information.tex
edited paper; added references, changed figure

@@
 
+\renewcommand\thefigure{S\arabic{figure}}
+
 \begin{document}
@@
 Each image is of a different embryo in a different rotational orientation.
-$48$ of the images are of wild type embryos, and $42$ of the images are of mutant embryos.}
+Half of the images are of wild type embryos, and half of the images are of mutant embryos.}
 \end{figure}
@@
 \begin{figure}[H]
-\raisebox{5.2cm}{A}
+\raisebox{5.2cm}{\sffamily{A}}
 \includegraphics[width=16.8cm]{data2_PCA_ordered}
 
-\raisebox{3.7cm}{B}
+\raisebox{3.7cm}{\sffamily{B}}
 \includegraphics[width=8.4cm]{data2_PCA_variance}
-\raisebox{3.7cm}{C}
+\raisebox{3.7cm}{\sffamily{C}}
 \includegraphics[width=8.4cm]{data2_PCA_proj}
@@
 %
-The $j^{th}$ embedding coordinate will be given by $\psi_{k,l}$ such that $0 < k \le d, jd +1 < l \le (j+1)d$,
-and we select the $(k,l)$ pair such that the coordinate $\psi_{k, l}$ has the largest variability. 
-%
+The $j^{th}$ embedding coordinate will be given by $\psi_{k,1}$, where $jd +1 < k \le (j+1)d$,
+and we select $k$ such that the coordinate $\psi_{k, 1}$ has the largest variability, i.e., the $j^{th}$ coordinate is $\psi_{k,1}$, where $k$ is
+$$ \max_{jd +1 < k \le (j+1)d} \sum_i \psi_{k,1} (i)^2 $$
 
@@
 \includegraphics[width=\textwidth]{sphere2_1}\\
-A
+\sffamily{A}
 \end{minipage}
@@
 \includegraphics[width=\textwidth]{sphere2_2}\\
-B
+\sffamily{B}
 \end{minipage}
@@
 \includegraphics[width=\textwidth]{sphere2_3}\\
-C
+\sffamily{C}
 \end{minipage}
@@
 \includegraphics[width=\textwidth]{sphere2_4}\\
-D
+\sffamily{D}
 \end{minipage}
@@
 
-\subsection{Algorithms}
-
-\subsubsection{Angular synchronization}
-
-\begin{enumerate}
-\item Compute the pairwise rotations (described in \eqref{eq:opt_angle}) or rotations + translations (described in \eqref{eq:opt_pairwise}) needed to register every pair of data points.
-\item Construct the corresponding rotation matrices $R_{ij}$ for every pair $i, j$ of images (see \eqref{eq:R_theta} for rotations or \eqref{eq:total_rot} for rotations + translations).
-\item Construct the matrix $H$ as in \eqref{eq:H_to_R}, which is a block matrix with entries $R_{ij}$.
-\item Compute the top $2$ (for rotations) or $3$ (for rotations + translations) eigenvectors of $H$.
-\item Recover the estimates $\hat{R}_1, \dots, \hat{R}_m$ from the top 2 or 3 eigenvectors as in \eqref{eq:R_hat}.
-\item Project these estimates onto the closest proper rotation matrix using \eqref{eq:R_est}
-\item Recover the optimal rotation angle $\theta_i$ using \eqref{eq:R_theta} or rotation + translation using \eqref{eq:angle_invert} for each image
-\end{enumerate}
-
-\subsubsection{Diffusion Maps}
-
-\begin{enumerate}
-\item Construct the matrix $W$ as in \eqref{eq:dmaps_W}
-\item Construct the matrix $A$ as in \eqref{eq:dmaps_A}
-\item Compute the top few eigenvectors of $A$, ordered with decreasing eigenvalues
-\item If we only want to order the data, then sort the data by the entries in the second eigenvector.
-\item For higher-dimensional embeddings (e.g., the mutants), use the entries of successive eigenvectors (the second eigenvector, the third eigenvector, etc.) as the embedding coordinates
-\end{enumerate}
-
-\subsubsection{Vector Diffusion Maps}
-
-\begin{enumerate}
-\item Compute the pairwise rotations (described in \eqref{eq:opt_angle}) or rotations + translations (described in \eqref{eq:opt_pairwise}) needed to register every pair of data points.
+%\subsection{Algorithms}
 %
-\item Construct the corresponding rotation matrices $R_{ij}$ for every pair $i, j$ of images (see \eqref{eq:R_theta} for rotations or \eqref{eq:total_rot} for rotations + translations).
+%\subsubsection{Angular synchronization}
 %
-\item Construct the matrix $H$ as in \eqref{eq:H_low_rank}.
+%\begin{enumerate}
+%\item Compute the pairwise rotations (described in \eqref{eq:opt_angle}) or rotations + translations (described in \eqref{eq:opt_pairwise}) needed to register every pair of data points.
+%\item Construct the corresponding rotation matrices $R_{ij}$ for every pair $i, j$ of images (see \eqref{eq:R_theta} for rotations or \eqref{eq:total_rot} for rotations + translations).
+%\item Construct the matrix $H$ as in \eqref{eq:H_to_R}, which is a block matrix with entries $R_{ij}$.
+%\item Compute the top $2$ (for rotations) or $3$ (for rotations + translations) eigenvectors of $H$.
+%\item Recover the estimates $\hat{R}_1, \dots, \hat{R}_m$ from the top 2 or 3 eigenvectors as in \eqref{eq:R_hat}.
+%\item Project these estimates onto the closest proper rotation matrix using \eqref{eq:R_est}
+%\item Recover the optimal rotation angle $\theta_i$ using \eqref{eq:R_theta} or rotation + translation using \eqref{eq:angle_invert} for each image
+%\end{enumerate}
 %
-\item Construct the matrix $W$ as in \eqref{eq:dmaps_W}, using the minimum distance obtained for each pair in step 1 as the pairwise distances in the Gaussian kernel.
+%\subsubsection{Diffusion Maps}
 %
-\item Construct the matrix $A$ as in \eqref{eq:dmaps_A}.
+%\begin{enumerate}
+%\item Construct the matrix $W$ as in \eqref{eq:dmaps_W}
+%\item Construct the matrix $A$ as in \eqref{eq:dmaps_A}
+%\item Compute the top few eigenvectors of $A$, ordered with decreasing eigenvalues
+%\item If we only want to order the data, then sort the data by the entries in the second eigenvector.
+%\item For higher-dimensional embeddings (e.g., the mutants), use the entries of successive eigenvectors (the second eigenvector, the third eigenvector, etc.) as the embedding coordinates
+%\end{enumerate}
 %
-\item Construct the matrix $S$ as in \eqref{eq:vdm_S}.
+%\subsubsection{Vector Diffusion Maps}
 %
-\item Compute the top few eigenvectors of $S$.
+%\begin{enumerate}
+%\item Compute the pairwise rotations (described in \eqref{eq:opt_angle}) or rotations + translations (described in \eqref{eq:opt_pairwise}) needed to register every pair of data points.
+%%
+%\item Construct the corresponding rotation matrices $R_{ij}$ for every pair $i, j$ of images (see \eqref{eq:R_theta} for rotations or \eqref{eq:total_rot} for rotations + translations).
+%%
+%\item Construct the matrix $H$ as in \eqref{eq:H_low_rank}.
+%%
+%\item Construct the matrix $W$ as in \eqref{eq:dmaps_W}, using the minimum distance obtained for each pair in step 1 as the pairwise distances in the Gaussian kernel.
+%%
+%\item Construct the matrix $A$ as in \eqref{eq:dmaps_A}.
+%%
+%\item Construct the matrix $S$ as in \eqref{eq:vdm_S}.
+%%
+%\item Compute the top few eigenvectors of $S$.
+%%
+%\item Recover the estimates $\hat{R}_1, \dots, \hat{R}_m$ from the top 2 or 3 eigenvectors as in \eqref{eq:R_hat}.
+%%
+%\item Project these estimates onto the closest proper rotation matrix using \eqref{eq:R_est}.
+%%
+%\item Recover the optimal rotation angle $\theta_i$ using \eqref{eq:R_theta} or rotation + translation using \eqref{eq:angle_invert} for each image
+%%
+%\item To compute the first (non-trivial) embedding coordinate for the data, compute $\psi_{k,l}$ as in \eqref{eq:vdm_coord}, taking $k=1, 2$ and $l=3, 4$ for the rotation case, or $k = 1, 2, 3$ and $l = 4, 5, 6$ for the rotation + translation case.
+%%
+%Select the $(k, l)$ coordinate which has the largest variance as the embedding coordinate.
+%%
+%\item To compute more coordinates, use the subsequent blocks for $l$ (e.g., for the next coordinate, under translations and rotations, use $k=1, 2, 3$ and $l = 7, 8, 9$).
+Again, compute $\psi_{k, l}$ as in \eqref{eq:vdm_coord}, and select the $(k, l)$ coordinate with largest variance.
 %
-\item Recover the estimates $\hat{R}_1, \dots, \hat{R}_m$ from the top 2 or 3 eigenvectors as in \eqref{eq:R_hat}.
-%
-\item Project these estimates onto the closest proper rotation matrix using \eqref{eq:R_est}.
-%
-\item Recover the optimal rotation angle $\theta_i$ using \eqref{eq:R_theta} or rotation + translation using \eqref{eq:angle_invert} for each image
-%
-\item To compute the first (non-trivial) embedding coordinate for the data, compute $\psi_{k,l}$ as in \eqref{eq:vdm_coord}, taking $k=1, 2$ and $l=3, 4$ for the rotation case, or $k = 1, 2, 3$ and $l = 4, 5, 6$ for the rotation + translation case.
-%
-Select the $(k, l)$ coordinate which has the largest variance as the embedding coordinate.
-%
-\item To compute subsequent coordinates, use lower blocks for $l$ (e.g., for the next coordinate, under translations and rotations, use $k=1, 2, 3$ and $l = 7, 8, 9$).
-Again, compute $\psi_{k, l}$ as in \eqref{eq:vdm_coord}, and select the $(k, l)$ coordinate with largest variance.
-
-\end{enumerate}
+%\end{enumerate}
 
@@
 
-I computed the FBsPCA basis for the set of images (using only the dpERK signal).
+I computed the FBsPCA basis for the set of images (separately for the red and green signals).
 %
-I then computed the bispectrum coefficients for each image expanded in this basis, and used these coefficients in a DMAPS calculation.
+I then computed the bispectrum coefficients for each image expanded in the FBsPCA basis, and used these coefficients in a DMAPS calculation.
 %
@@
 \begin{figure}[H]
-\includegraphics[width=7cm]{FBsPCA_images}
-\includegraphics[width=10cm]{FBsPCA_dmaps}
-\caption{(Left) Images used for FBsPCA.
-(Right) First 4 dmaps variables, computed from the bispectrum coefficeints, plotted against the membrane thickness.
+\raisebox{7.6cm}{\sffamily{A}}
+\includegraphics[width=8cm]{raw_data1}
+\hfill
+\raisebox{7.6cm}{\sffamily{B}}
+\includegraphics[width=7cm]{FBsPCA_evals}
+
+\raisebox{2.2cm}{\sffamily{C}}
+\hspace{-1cm}
+\includegraphics[width=4cm, angle=90]{FBsPCA_dmaps}
+\caption{(A) Images used for FBsPCA.
+(B) Eigenvalue spectrum from dmaps, using bispectrum coefficients.
+(C) First 4 dmaps variables, computed from the bispectrum coefficeints, plotted against the membrane thickness.
 Note there is no appreciable correlation: each dmaps variable simply selects one image from the data set as ``abnormal''.}

### 2014-05-27 image_analysis_paper/test_FBsPCA.m

@@ -151,24 +151,34 @@ fprintf('\nThe number of components is %d \n', size(Freqs, 1));
 % end
 % W_dmaps = W_dmaps + W_dmaps';
 W_dmaps = squareform(pdist([real(Coeff_b') imag(Coeff_b') real(Coeff_b2') imag(Coeff_b2')])).^2;
-eps_dmaps = median(W_dmaps(:))/2;
+eps_dmaps = median(W_dmaps(:))/4;
 
-[V, D] = dmaps(W_dmaps, eps_dmaps, 10);
+[V, D] = dmaps(W_dmaps, eps_dmaps, m);
+
+figure;
+set(gcf, 'paperunits', 'centimeters')
+set(gcf, 'papersize', [8 8])
+set(gcf, 'paperposition',[0 0 8 8]);
+plot(diag(abs(D)),'.')
+xlabel('k')
+ylabel('|\lambda_k|')
+saveas(gcf,sprintf('%s/FBsPCA_evals', im_save_dir), 'pdf')
 
 figure;
 imshow(W_dmaps(ranks_from_membranes,ranks_from_membranes))
 
 figure;
 set(gcf, 'paperunits', 'centimeters')
-set(gcf, 'papersize', [16 16])
-set(gcf, 'paperposition',[0 0 16 16 ]);
+set(gcf, 'papersize', [16 4])
+set(gcf, 'paperposition',[0 0 16 4]);
 for i=2:5
-    subplot(2, 2, i-1)
+    subplot(1,4, i-1)
     plot(V(:,i),mem_lengths,'.')
     xlabel(sprintf('\\phi_%d',i));
-    ylabel('membrane thickness')
+    ylabel('membrane length')
+    axis([-1.1 1.1 -inf inf])
 end
-% saveas(gcf,sprintf('%s/FBsPCA_dmaps', im_save_dir), 'pdf')
+saveas(gcf,sprintf('%s/FBsPCA_dmaps', im_save_dir), 'pdf')
 
 
 %%

### 2014-05-27 image_analysis_paper/test_mutants.m

@@ -260,6 +260,16 @@ neigs = 42;
 
 theta_adjust = -55;
 
+% mut_draw = [ 37 32 1 22 27];
+% wt_draw = [19 7 10 9 34];
+% nstages = 5;
+% mut_draw = [ 37  1 22 27];
+% wt_draw = [19  10 9 34];
+% nstages = 4;
+mut_draw = [29 5 1 27];
+wt_draw = [21  10 9 34];
+nstages = 4;
+
 image_set_aligned = zeros(size(image_set), 'uint8');
 image_set_aligned_withnuclei = zeros(size(image_set), 'uint8');
 for i=1:m
@@ -287,7 +297,7 @@ set(gcf, 'paperposition',[0 0 8 8])
 plot(abs(diag(D)),'.')
 xlabel('k')
 ylabel('|\lambda_k|')
-saveas(gcf,sprintf('%s/data3_evals', im_save_dir), 'pdf')
+% saveas(gcf,sprintf('%s/data3_evals', im_save_dir), 'pdf')
 
 figure;  
 set(gcf, 'paperunits', 'centimeters')
@@ -316,13 +326,26 @@ end
 figure;
 plot(embed_coord(:,idx1),embed_coord(:,idx2),'.')
 
+
 figure;
 set(gcf, 'paperunits', 'centimeters')
 set(gcf, 'papersize', [8 6])
 set(gcf, 'paperposition',[0 0 8 6])
-plot(embed_coord(~mutation,idx1),embed_coord(~mutation,idx2),'xk', 'markersize', 5)
+plot(embed_coord(:,idx1),embed_coord(:,idx2),'.k', 'markersize', 10)
+set(gca, 'xtick', [])
+set(gca, 'ytick', [])
+xlabel('first VDM coordinate')
+ylabel('second VDM coordinate')
+saveas(gcf,sprintf('%s/data3_embed_bw', im_save_dir), 'pdf')
+
+
+figure;
+set(gcf, 'paperunits', 'centimeters')
+set(gcf, 'papersize', [8 6])
+set(gcf, 'paperposition',[0 0 8 6])
+plot(embed_coord(~mutation,idx1),embed_coord(~mutation,idx2),'xb', 'markersize', 5)
 hold on
-plot(embed_coord(mutation,idx1),embed_coord(mutation,idx2),'ok', 'markersize', 5)
+plot(embed_coord(mutation,idx1),embed_coord(mutation,idx2),'or', 'markersize', 5)
 set(gca, 'xtick', [])
[…]

### 2014-05-27 image_analysis_paper/test_mutants_big.m

@@ -81,8 +81,8 @@ load('pairwise_alignments_snanull.mat');
 
 % ind = setdiff(1:m, [17]);
 % ind = setdiff(1:m, [73 37 34 29 15 21 87 62 17 55 96 18]);
+ind = setdiff(1:m, [73 37 34 29 15 21 87 62 17 55]);
 % ind = setdiff(1:m, [37 34 29 15 21 2 4 6]);
-ind = setdiff(1:m, [37 34 29 15 21 2 4 6]);
 mutation = mutation(ind);
 
 image_set = image_set(:, :, :, ind);
@@ -114,7 +114,7 @@ for i=1:m
     imshow(im1);
     
 end
-saveas(gcf,sprintf('%s/raw_data3b', im_save_dir), 'pdf')
+% saveas(gcf,sprintf('%s/raw_data3b', im_save_dir), 'pdf')
 
 %% synchronization
 
@@ -216,7 +216,7 @@ saveas(gcf,sprintf('%s/raw_data3b', im_save_dir), 'pdf')
 
 %% VDM
 
-eps = median(W(:))/5;
+eps = median(W(:))/4;
 neigs = 42;
 
 [R_opt, embed_coord, embed_idx, D] = vdm(R, W, eps, neigs);
@@ -253,7 +253,7 @@ set(gcf, 'paperposition',[0 0 8 8])
 plot(abs(diag(D)),'.')
 xlabel('k')
 ylabel('|\lambda_k|')
-saveas(gcf,sprintf('%s/data3_evalsb', im_save_dir), 'pdf')
+% saveas(gcf,sprintf('%s/data3_evalsb', im_save_dir), 'pdf')
 
 %%
 
@@ -275,6 +275,8 @@ figure;
 plot(embed_coord(~mutation,idx1),embed_coord(~mutation,idx2),'x')
 hold on
 plot(embed_coord(mutation,idx1),embed_coord(mutation,idx2),'o')
+
+return
 
 %%
 [~, I ] =sort(embed_coord(:,idx1));
@@ -337,7 +339,7 @@ xlabel('first VDM coordinate')
 ylabel('second VDM coordinate')
 lgnd = legend('wild type','mutant', 'location','northeast');
 set(lgnd,'fontsize',6);
-saveas(gcf,sprintf('%s/mut_wt_vdm_embeddingb', im_save_dir), 'pdf')
+% saveas(gcf,sprintf('%s/mut_wt_vdm_embeddingb', im_save_dir), 'pdf')
 
 figure;
[…]

### 2014-05-28 image_analysis_paper/pnas_draft2.tex
small changes

@@
 of the United States of America}
+
+\significancetext{blahhhhh}
 
@@
 %
-In particular, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point, e.g., \cite{yuan2014automated, surkova2008characterization}.
+In particular, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \cite{yuan2014automated, surkova2008characterization}.
 %
@@
 
-\textbf{Fly strain and whole-mount immunostaining.}
+\section{Fly strain and whole-mount immunostaining}
 %
@@
 
-\textbf{Microscopy.}
+\section{Microscopy}
 %
@@
 
-\textbf{Development time during cellularization.} 
+\section{Development time during cellularization} 
 %

### 2014-05-28 image_analysis_paper/pnas_draft2.tex
added edits to paper and si

@@
 
-\DeclareMathSizes{9}{8}{7}{7}
+%\DeclareMathSizes{9}{8}{7}{7}
 
-\DeclareMathOperator*{\argmin}{arg\,min}
+%\DeclareMathOperator*{\argmin}{arg\,min}
+
+\newcommand{\SI}[0]{{\it SI Materials and Methods}}
+\newcommand{\fig}[0]{Fig.}
 
@@
 
-\significancetext{blahhhhh}
+\significancetext{Cross-sectional imaging studies are commonly used to study developmental dynamics in multiple organisms.
+In these studies, a representative developmental trajectory is reconstructed from snapshots of different embryos, each of which has been arrested at a different point of its development.
+One of the first steps in this reconstruction involves registration and temporal ordering of images.
+Here we show that both of these tasks can be automated and combined in a single step using recently developed advances in data mining.
+To illustrate this approach, we temporally order and register several datasets collected in cross-sectional studies of cell signaling in the early fruit fly embryo.}
 
@@
 \begin{abstract}
-In studies of development, researchers are often presented with cross-sectional data, where each data point is a sample from a population fixed at a
-%slightly
-different developmental time.
+
+Imaging studies provide unique insights into dynamics of developmental processes. 
 %
-The goal is then to temporally order the data to reconstruct the developmental dynamics.
+A large number of imaging studies are based on a cross-sectional experimental design, where developmental progress is arrested in a population of embryos, each of which is at a different point along its developmental trajectory.
 %
-Data points in the form of two-dimensional image must first be registered before they can be temporally ordered.
+The goal is then to temporally order the data to reconstruct a representative developmental dynamics from snapshots provided by the collection of embryos.
 %
-When such data sets are large, noisy, and/or if the developmental changes are subtle, these tasks can be difficult to perform by hand.
+Images of different biological samples must first be registered before they can be temporally ordered.
 %
-We present an automated approach to registering {\it and} temporally ordering cross-sectional data sets of images.
+When such data sets are large, noisy, and/or if the developmental changes are subtle, these tasks can be difficult to perform manually.
 %
-The mathematical techniques (vector diffusion maps) are applicable to a wide variety of data sets and
-require little {\it a priori} knowledge of the image features or the developmental dynamics.
+We present an automated approach to registering and temporally ordering data from such cross-sectional studies.
+%
+The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require {\it a priori} knowledge of image features or developmental dynamics.
 %
@@
 %
-Furthermore, the approach can detect multiple sources of variability
-and thus help visualize multiple trajectories within a data set.
-%
-We demonstrate the utility of these methods using a collection of images from a study of {\it Drosophila} embryogenesis.
+We demonstrate the utility of these methods using a collection of images from a cross-sectional study of cell signaling during dorsoventral patterning of the {\it Drosophila} embryo.
 \end{abstract}
@@
 %
-In longitudinal studies, developmental progress is monitored over time for the same embryo \cite{roelens2013live, keller2013imaging}.
+In longitudinal studies, developmental progress is monitored over time in the same embryo \cite{roelens2013live, keller2013imaging}.
 %
-In a cross-sectional study, developmental dynamics must be reconstructed from multiple embryos, each of which contributes only a single snapshot of a chemical or morphological process along its developmental trajectory \cite{jaeger2004dynamic, fowlkes2008quantitative}.
-%
-Both of these sampling schemes have their advantages and limitations, and both are extensively used by biologists.
-%%%
-%%%YGK sounds condescending
-%%%
+In a cross-sectional study, developmental dynamics is reconstructed from multiple embryos, each of which contributes only a snapshot of a chemical or morphological process along its developmental trajectory \cite{jaeger2004dynamic, fowlkes2008quantitative}.
 %
 Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms.
-%%%
-%%%YGK  one or two refs ?
-%%%
 %
@@
 %
-Recent advances in large-scale
-%%%
-%%%YGK  large scale ? high throughput
-%%%
-physical manipulation and imaging of embryos produce rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different developmental time point but also at a different geometric orientation.
+Recent advances in physical manipulation and imaging of embryos produce rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different developmental time point and at a different geometric orientation.
 %
 Importantly, the ``age'' of any given embryo arrested in its development is not quantitatively known; typically what is known is
-a certain time window to which a collection of embryos belongs, e.g., \cite{ng2012large, richardson2014emage, castro2009automatic}.
+a certain time window to which a collection of embryos belongs \cite{ng2012large, richardson2014emage, castro2009automatic}.
 %to high accuracy.
-%
-Furthermore, each embryo appears at an arbitrary translation and rotation within the image frame (see {\it SI Appendix}).
 %
@@
 %
-We show how existing dimensionality reduction algorithms can automate, greatly accelerate {\it and combine} both of these tasks.
+We show how existing dimensionality reduction algorithms can automate {\it and combine} both of these tasks.
 
-An example where
-temporal ordering and registration of images can be done manually
-because the number of images is small and the differences between them are visually apparent is shown
-in Figure~\ref{fig:fish}: a caricature of fish development which illustrates the processes of growth and patterning.
+Temporal ordering and registration of images can be done manually
+when the number of images is small and the differences between them are visually apparent. 
+%
+\fig~\ref{fig:fish} shows a caricature of fish development which illustrates the processes of growth and patterning.
 %
@@
 %
-In contrast to this simple illustration, real data poses nontrivial challenges for both registration and temporal ordering.
+In contrast to this example, real data poses nontrivial challenges for both registration and temporal ordering.
 %
-In general, the landmarks needed for registration (here the heads and fins) as well as the attributes which can be used to order the data (such as the body length) are not known {\it a priori}.
+In general, the landmarks needed for registration (here, the heads and fins) as well as the attributes which can be used to order the data (such as the body length) are not known {\it a priori}.
 %
@@
 \caption{Caricature illustrating the tasks of image registration and temporal ordering.
-{\xfigtextfontit (A)} Fish, each in a different orientation and a different stage of development.
-{\xfigtextfontit (B)} Fish, now rotationally registered and temporally ordered.
-For this caricature, the registration and ordering is easy to do ``by hand" because the data set is small and the developmental changes are easy to recognize.}
+{\xfigtextfontit (A)} Images of ``samples'', each in a different orientation and a different stage of development.
+{\xfigtextfontit (B)} Registered and ordered samples.
+For this caricature, the registration and ordering is straightforward because the data set is small and the developmental changes are easy to recognize.}
 %\label{fig:fish}
@@
 %
-Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, as have dimensionality reduction techniques \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear} and their applications \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering, kafri2013dynamics}, especially in the last decade.
+Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, as have dimensionality reduction techniques \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear} and their applications \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering, kafri2013dynamics}.
 %
-Here, we use a manifold learning algorithm \cite{singer2012vector} for high-dimensional imaging data.
-% which both factors out geometric symmetries to register the images and extracts low-dimensional structure in the imaging data set.
+Here, we use a manifold learning algorithm \cite{singer2012vector} that has been recently developed for high-dimensional imaging data. 
 %
-Such techniques have been used for data mining in
-applications ranging from analysis of cryo-electron microscopy images of individual molecules to face recognition; here they are adapted
-for simultaneous registration and temporal ordering of imaging data from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
+This technique has been used for data mining in
+applications ranging from analysis of cryo-electron microscopy images of individual molecules to face recognition, and is sufficiently general to be applicable to a wide variety of imaging studies. 
+%
+Here it is adapted for simultaneous registration and temporal ordering of images from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
 %
@@
 %
-The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate this approach.
+%The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate this approach.
 %
-In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex.
+%In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex.
 %
-Here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
+%Here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
 %
-The third set consists of data collected from both wild type and mutant embryos, and joint processing of the data will help to visualize the different sources of variability along distinct developmental paths.
+%The third set consists of data collected from both wild type and mutant embryos, where joint processing of the data will help to visualize the different sources of variability along distinct developmental paths.
 
@@
 %
-Our approach is based on vector diffusion maps \cite{singer2012vector}, a general technique developed for data sets which contain two types of sources of variability:
+The approach is based on vector diffusion maps \cite{singer2012vector}, a manifold learning
+technique developed for data sets which contain two types of sources of variability:
 geometric symmetries (such as translations and rotations of the images due to the experimental setup) which one would like to factor out,
@@
 %
-We assume that, in this set of images, this main direction of variability is parameterized by the developmental time of each embryo, so taht uncovering this direction will allow us to recover the developmental dynamics.
-%
-Unlike the noise in the cryo-EM application, which is approximately Gaussian and arises from the measurement instrument, the noise in our experiments arises primarily from interembryo variability and does not have a known simple structure.
+We assume that, in these sets of images, the main direction of variability is parameterized by the developmental time of each embryo, so that uncovering this direction will allow us to recover the developmental dynamics.
 
@@
 %
-A schematic illustration is shown in Figure~\ref{subfig:synch1}, where each image is depicted as a vector, and the goal is to align the set of vectors.
+A schematic illustration of angular synchronization is shown in \fig~\ref{subfig:synch1}, where each image is represented as a vector, and the goal is to align the set of vectors.
 %
@@
 %
-Using the alignment angles between all pairs of vectors (or images), angular synchronization finds the set of rotation angles (one angle for each vector) most consistent with all pairwise measurements (see {\it SI Appendix}).
+Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with all pairwise measurements (see {\it SI Materials and Methods}).
 %
@@
 %
-However, the algorithm has been shown to successfully register sets of vectors (or images) even when many of the pairwise measurements are corrupted by noise \cite{singer2011angular}.
+However, the algorithm can successfully register data sets even when many of the pairwise measurements are corrupted by noise \cite{singer2011angular}.
 %
@@
 %
-After removing the variability due to translations and rotations of the images, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the variability in the data.
+After removing variability due to translations and rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
 %%%
@@
 %
-Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique.
+Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parameterization of data that lies on a low-dimensional, perhaps nonlinear, manifold in high-dimensional space.
 %
-This algorithm uncovers an intrinsic parameterization of data that lies on a low-dimensional, perhaps nonlinear, manifold in high-dimensional space.
+The idea is illustrated in \fig~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
 %
-
-The idea is illustrated in Figure~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
+We use {\it local} information about the data to find a parameterization which respects the underlying manifold geometry: we want points which are close in high-dimensional space (e.g., images which look similar) to be close in our parameterization.
 %
-We use {\it local} information about the data to find a ``good'' parameterization which respects the underlying manifold geometry: we want points which are close in high-dimensional space (e.g., images which look similar) to be close in our parameterization.
-%
-This idea of locality is denoted by the color intensity of the edges (denoting edge weight) in Figure~\ref{subfig:dmaps1}.
+This idea of locality is denoted by the edges in \fig~\ref{subfig:dmaps1}.
 %
@@
 %
-The color in Figure~\ref{subfig:dmaps2} depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
+The color in \fig~\ref{subfig:dmaps2} depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
 %
@@
 %
-Instead, we will rely on diffusion maps to automatically uncover a parameterization of our high-dimensional data (see {\it SI Appendix}).
+Instead, we will use diffusion maps to automatically uncover a parameterization of our high-dimensional data (see \SI).
 %
-Diffusion maps will also generalize from one-dimensional nonlinear curves to higher-dimensional manifolds.
+Diffusion maps also generalizes from one-dimensional nonlinear curves to higher-dimensional manifolds.
 %
-Thus, when our experiments have additional sources of variability (such as mutations) in addition to developmental dynamics, diffusion maps can uncover both of these variability sources, and provide an informative organization of our data.
+Thus, when our experiments have additional sources of variability (such as distinct populations of embryos) in addition to developmental dynamics, diffusion maps can provide an informative organization of our data which highlights both of these variability sources.
 %We will assume that our data are approximately one-dimensional, and that this dimension is parameterized by time, such that ordering our data along this main dimension/direction will temporally order our data.
@@
 %
-Vector diffusion maps combines both steps into a single computation;
-the algorithm uses information about pairwise alignments {\em and} pairwise distances to find a global parameterization of the data, while registering images which are ``close'' (see {\it SI Appendix}).
+Vector diffusion maps combines both steps into a single computation.
+%
+The algorithm uses information about pairwise alignments {\em and} pairwise distances to find a global parameterization of the data, while registering images which are ``close'' (see \SI).
 %
@@
 {\xfigtextfontit (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
-Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\xfigtextfontit SI Appendix}), so that close data points are connected by dark (``stronger'') edges.
+Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\xfigtextfontit \SI}), so that close data points are connected by dark (``stronger'') edges.
 {\xfigtextfontit (D)} The data in {\xfigtextfontit C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure.
@@
 
-As a first illustration of this approach to temporal ordering and registration of images, we analyze a data set where the correct registration and temporal order is known {\it a priori}.
+As a first illustration of using vector diffusion maps to temporally order and register images, we analyze a data set where the correct registration and temporal order is known {\it a priori}.
 %
@@
 %
-Figure~\ref{subfig:raw_data1} shows optical cross-sections of each embryo, with nuclei (gray in the figure) labeled by DAPI, a DNA stain.
+\fig~\ref{subfig:raw_data1} shows optical cross-sections of each embryo, with nuclei (gray in the figure) labeled by DAPI, a DNA stain.
 %
@@
 
-Previous studies \cite{rushlow2012temporal} established that throughout the third hour of development, the nuclear localization of Dl is unimodal and peaked at the ventral side of the embryo (see {\it SI Appendix}).
+Previous studies established that throughout the third hour of development, the nuclear localization of Dl is unimodal and peaked at the ventral side of the embryo, reviewed in \cite{rushlow2012temporal}.
 %
-The spatiotemporal pattern of dpERK is more complex: dpERK is first detected in two lateral stripes, which increase in intensity towards the end of the third hour \cite{lim2013kinetics}.
+The spatiotemporal pattern of dpERK is more complex: dpERK is first detected in two lateral stripes, which increase in intensity towards the end of the third hour of development \cite{lim2013kinetics}.
 %
@@
 %
-Both of these features become readily apparent when images are registered and ordered using vector diffusion maps (see Figure~\ref{subfig:ordered_data1}): the peak of nuclear Dl is consistent throughout the images and intensity of the dpERK signal in the lateral regions of the embryo monotonically increases across the data set.
+Both of these features become readily apparent when images are registered and ordered using vector diffusion maps (see \fig~\ref{subfig:ordered_data1}): the peak of nuclear Dl is consistent throughout the images and the intensity of the dpERK signal in the lateral regions of the embryo increases monotonically across the data set.
 %
@@
 %
-During the third hour of development, the embryo cellularizes: nuclei located under common plasma membrane are enclosed by lateral membranes which grow inward, separating the nuclei into individual cells.
+During the third hour of development, the embryo cellularizes: in this process, nuclei located under a common plasma membrane are enclosed by lateral membranes which grow inward, separating the nuclei into individual cells.
 %
-The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo to within 1--3 minutes \cite{figard2013plasma} (see {\it SI Appendix}).
+The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo to within 1--3 minutes \cite{figard2013plasma} (see \fig~S1).
 %
-Spearman's rank correlation coefficient between the orderings obtained based on the progress of cellularization and vector diffusion maps is 0.91.
+The Spearman rank correlation coefficient between the orderings obtained based on the progress of cellularization and vector diffusion maps is 0.91, indicating that vector diffusion maps can accurately recover the temporal order.
 %
@@
 
-Although the first vector diffusion maps coordinate accurately captures the developmental dynamics, there are other sources of variability within this data set that could also be extracted using these techniques.
+%Although the first vector diffusion maps coordinate accurately captures the developmental dynamics, there are other sources of variability within this data set that could also be extracted using these techniques.
 %
-An obvious one is variability due to size; here, the size variation is relatively small relative to the developmental changes.
+%An obvious one is variability due to size; here, the size variation is relatively small relative to the developmental changes and does not significantly obstruct the reconstruction of the temporal order.
 %
-However, in other applications where the size variation is significant or the developmental dynamics are more subtle, it may be important to factor out size or {\it scaling} symmetries.
+%However, in other applications where the size variation is significant %or the developmental dynamics are more subtle, it may be important to %factor out size or {\it scaling} symmetries.
 %%%
@@
 %
-Because this data set in Figure~\ref{fig:data1} are relatively simple, effectively consisting of a single mode (the two dpERK peaks) growing in time,
-established techniques such as principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data in time.
+Because the dynamics of the data set shown in \fig~\ref{fig:data1} are relatively simple, effectively consisting of a single mode (the two dpERK peaks) growing in time, established techniques such as principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data in time (see \fig~S6).
 %
-However, developmental dynamics are often significantly more complex, and nonlinear techniques such as (vector) diffusion maps are required to extract meaningful structure in the data.
+However, developmental dynamics are often significantly more complex, and nonlinear dimensionality reduction techniques are required to extract meaningful structure in the data.
 %
 
-Figure~\ref{subfig:raw_data2} shows a richer data set of $108$ images of {\it Drosophila} embryos.
+\fig~\ref{subfig:raw_data2} shows a richer data set of $108$ images of {\it Drosophila} embryos.
 %
-These images cover a thirty minute time interval and span late cellularization (the early images in this data set overlap with the late images in the previous set) through gastrulation, when the two-dimensional sheet of newly formed cells begins to deform into three-dimensional structures which will form the future organs.
+These images cover a thirty minute time interval and span late cellularization through gastrulation, when the two-dimensional sheet of newly formed cells begins to deform into three-dimensional structures which will form the future organs \cite{leptin2005gastrulation}.
 %
-As before, each image is an optical cross-section of an embryo fixed at a different developmental time.
+As before, each image is an optical cross-section of a vertically oriented embryo fixed at a different developmental time.
 %
@@
 %
-Instead of Dl, which is difficult to visualize at this point, we stained for Twist (Twi, shown in green), a transcription factor that is activated by high levels of Dl, specifying the cells of the future muscle tissue.
+Instead of Dl, which is difficult to visualize at this point, embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor that is activated by high levels of Dl, specifying the cells of the future muscle tissue.
 %
+dpERK is again shown in red.
 
 %
-During this developmental time span, the ventral furrow is formed, where the ventral side of the embryo buckles towards the center of the embryo, bringing in the future muscle cells.
+During this window, the ventral furrow is formed, where the ventral side of the embryo buckles towards the center of the embryo, bringing in the future muscle cells.
 %
-dpERK (red in the figure) again appears as two lateral peaks at the ventral side of the embryo.
+dpERK appears as two lateral peaks at the ventral side of the embryo.
 %
@@
 %
-During this time window, germband elongation causes cells from the ventral side to move towards the poles of the embryo, and then wrap around to the dorsal side.
+During this time window, germband extension causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side.
 %
-At the end of this process, cells which were originally on the ventral side have migrated to the dorsal side of the embryo, causing similar ``omega'' shaped patterns (formed by Twi and dpERK) to arise on the dorsal side (shown in the late images in Figure~\ref{subfig:averaged_data2}).
+At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing similar ``omega'' shaped patterns to arise on the dorsal side; these patterns are most readily seen in the last image of \fig~\ref{subfig:averaged_data2}.
 
@@
 {\xfigtextfontit (C)} Average trajectory of the images shown in {\xfigtextfontit B}.
-Each image is the average of $9$ consecutive images in {\xfigtextfontit B}.
+Each image is a (Gaussian-weighted) average of the images in {\xfigtextfontit B}.
 }
@@
 
-Visually, the dynamics in this data set are much more complicated, and only 22\% of the variability in the images is captured by the first principal component; simple time-ordering based on this projection is therefore not possible.
+Visually, the dynamics in this data set are much more complicated, and only 22\% of the variability in the images is captured by the first principal component.
 %
-Even when projecting on several leading principal components (see SI), a more sophisticated, nonlinear approach, such as vector diffusion maps, 
-is required to temporally order the data.
+Simple time-ordering based on this linear projection is therefore not possible (see \fig~S7).
 %
-Figure~\ref{subfig:ordered_data2} shows the images in Figure~\ref{subfig:raw_data2}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
+%Even when projecting on several leading principal components (see SI), a nonlinear approach, such as vector diffusion maps, is required to temporally order the data.
+%
+\fig~\ref{subfig:ordered_data2} shows the images in \fig~\ref{subfig:raw_data2}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
 %
@@
 %
-However, assimilating the dynamics from this concatenation of registered and ordered images is difficult for two reasons.
+However, assimilating the dynamics even from this concatenation of images is difficult for two reasons, even after registration and ordering.
 %
-First, the sheer number of images makes visual processing of the entire data set (even after registration and ordering) nontrivial, and
-second, the ordered data is not entirely smooth due to noise from interembryo variability.
+First, the sheer number of images makes visual processing of the entire data set nontrivial, and
+second, the ordered data is not entirely smooth due to inter-embryo variability.
 %
-To highlight the developmental dynamics revealed by vector diffusion maps, we construct a movie of an averaged developmental trajectory 
+To highlight the developmental dynamics revealed by vector diffusion maps, we applied a Gaussian filter to the trajectory of registered and ordered images to obtain an averaged developmental path.  
 %
-It is worth noting that in the process of making such a movie, one has to decide the (monotonic) relation between
-``movie time" (which is {\em not} physical time) and the parametrization of the data through the first VDM component; and that making a smooth movie
-requires the application of filter (we used a Gaussian filter, more details are given in the {\it SI Appendix}) to the set of ordered images.
+Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:averaged_data2}, serve as a summary of a stereotypic developmental dynamics reconstructed from cross-sectional data.
 %
-This movie is included in the {\it SI Appendix},
-and representative snapshots from it are shown in Figure~\ref{subfig:averaged_data2}, making the ``stereotypic" dynamics readily apparent;
+Thus, vector diffusion maps can accomplish the image registration and ordering tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental dynamics (e.g., correlation of age with body size).
+
+\subsection{Additional variability: identifying distinct developmental trajectories}
+
+When time is the main source of variability, we have shown vector diffusion maps can recover developmental progression from cross-sectional data.
 %
-%However, there are some artifacts in the movie due to size variation, and the dynamics towards the end are not as smooth, as we have fewer samples in this time window.
+However, vector diffusion maps can also provide an organization of data which contains multiple sources of variability.
 %
-%We would like to note that, although this movie provides us with a representative picture of the dynamics, the speed of the movie is not constant %with time, as we have no true time information about the images.
+To illustrate this, we collected a data set consisting of $42$ embryos undergoing cellularization during the third hour of development (see \fig~S2).
 %
-We have thus accomplished the same task presented in the caricature in Figure~\ref{fig:fish}, knowing neither the appropriate landmarks (e.g., fins) for registration nor the appropriate statistic (e.g., body size) by which to order.
-
-\subsection{Additional variability: Identifying distinct developmental trajectories}
-
-When time is the main source of variability,
-vector diffusion maps can temporally order data; but it can also uncover multiple sources of variability.
-%
-%Up to this point, we have assumed that our experiments contain one main direction of variability, and that this direction is parameterized by %time.
-%
-%However, this is not always the case.
-%
-To illustrate this, we have collected a data set consisting of $42$ embryos undergoing cellularization during the third hour of development (see {\it SI Appendix}); $21$ of the embryos are wild type, and $21$ of the embryos are mutant embryos where dpERK is expressed in one broad peak at the ventral side.
+Half of the embryos are wild type, and half of the embryos are mutants where dpERK is expressed in one broad peak at the ventral side.
 %
@@
 
-We computed the vector diffusion maps embedding for our data.
-%
-However, in this example, the eigenvalue spectrum from vector diffusion maps (see {\it SI Appendix}) shows that two coordinates may be required to obtain an adequate description of the data. 
-%
-Figure~\ref{subfig:projection_data3} shows the data, projected onto the first two vector diffusion maps coordinates;
-each data point corresponds to one image in our data set.
-%
-For the previous examples, where we only uncovered developmental progression, we used the first vector diffusion maps coordinate (i.e., the x-axis of this plot) to order the data.
-%%%
-%%%YGK we need something about dmap eignvalues and how many are not harmonics
-%
-In this two-dimensional embedding space, two distinct branches emerge, which originate from a common point.
-
-Figure~\ref{subfig:projection_colored_data3} shows the same two-dimensional projection as Figure~\ref{subfig:projection_data3}, now colored by species.
-%
-Clearly, the two branches of the plot correspond to the mutant and wild type data.
-%
-Selected images along the wild type and mutant trajectories are shown in Figures~Figures~\ref{subfig:average_wt_data3}~and~\ref{subfig:average_mut_data3}, respectively.
-%
-We can see that the lower-right portion of the embedded space, where the two species are mixed, corresponds to early in the developmental trajectory, where there is little signaling in the wild type and thew mutant.
-%
-When the two paths diverge, we can see that the wild type data develops two distinct peaks, whereas the mutant data develops a single peak.
-%
-We have therefore demonstrated that vector diffusion maps can not only recover dynamics, but also help distinguish different expression patterns and developmental trajectories.
-
-\begin{figure}[t]
+\begin{figure}[h!]
 \includegraphics{fig5}
 \caption{Mutant and wild type {\xfigtextfontit Drosophila}.
-{\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in {\xfigtextfontit SI Appendix}), projected onto the first two vector diffusion maps coordinates.
-{\xfigtextfontit (B)} The data in {\xfigtextfontit A}, now colored by species.
-The two ``branches'' clearly correspond to mutant and wild type samples.
+{\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~S2), projected onto the first two vector diffusion maps coordinates.
+{\xfigtextfontit (B)} The data in {\xfigtextfontit A}, now colored by genotype.
+The two branches separate mutant and wild type samples.
 {\xfigtextfontit (C)} dpERK signal at select wild type points in {\xfigtextfontit B} (denoted by circles).
 The two dpERK peaks which arise during development are indicated by arrows.
-{\xfigtextfontit (C)} dpERK signal at select mutant points in {\xfigtextfontit B} (denoted by circle).
-The single dpERK peak which arises is indicated by an arrow.
+{\xfigtextfontit (C)} dpERK signal at select mutant points in {\xfigtextfontit B} (denoted by circles).
+The single dpERK peak is indicated by an arrow.
 }
@@
 
-%\begin{figure*}[t]
-%\raisebox{8cm}{{\figtextfont A}}
-%\includegraphics[height=8.4cm]{wt_trajectory2}
-%%\raisebox{8cm}{{\figtextfont B}}
-%\begin{tikzpicture}
-%     \node at (0,0)
-%       {\includegraphics[height=8.4cm]{mut_wt_vdm_embedding_background}};
-%     \node at (0,0)
-%       {\includegraphics[height=8.4cm]{mut_wt_vdm_embedding}};
-%     \node at (-5cm, 3.8cm)
-%       {{\figtextfont B}};
-%\end{tikzpicture}
-%\raisebox{8cm}{{\figtextfont C}}
-%\includegraphics[height=8.4cm]{mut_trajectory2}
-%\caption{Mutant and wild type {\xfigtextfontit Drosophila}.
-{\xfigtextfontit (A)} dpERK signal at select wild type points in {\xfigtextfontit B} (denoted by bold symbols).
-{\xfigtextfontit(B)} Data collected from wild type and mutant embryos (original data set is shown in {\xfigtextfontit SI Appendix}), projected onto the first two vector diffusion maps coordinates.
-The two distinct trajectories are highlighted in blue and red.
-{\xfigtextfontit (C)} dpERK signal at select mutant points in {\xfigtextfontit B} (denoted by bold symbols).}
-%\customlabel{fig:data3}{5}
-%\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it A}}
-%\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it B}}
-%\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
-%\end{figure*}
+We computed the vector diffusion maps embedding for our data, and the eigenvalue spectrum (see \fig~S5) indicates that two coordinates may be required for a thorough description of the data. 
+%
+\fig~\ref{subfig:projection_data3} shows the data projected onto the first two vector diffusion maps coordinates, where each point in this projection corresponds to one image in our data set.
+%%%
+%%%YGK we need something about dmap eignvalues and how many are not harmonics
+%
+Two distinct branches appear to emerge in this two-dimensional embedding space.
+
+\fig~\ref{subfig:projection_colored_data3} shows the same two-dimensional projection as \fig~\ref{subfig:projection_data3}, but the data points are colored to indicate whether they come from the wild type or mutant.
+%
+Clearly, the two branches in the embedding separate wild type and mutant embryos.
+%
+Selected images along the wild type and mutant trajectories are shown in \fig~\ref{subfig:average_wt_data3} and \fig~\ref{subfig:average_mut_data3}, respectively.
+%
+The portion of the embedding where the two genotypes are intermixed (lower-right region of the plot) contains embryos who are early in their developmental trajectory, where there is little dpERK signaling and the two groups of embryos are indistinguishable.
+%
+When the two paths diverge, the wild type data develops two distinct peaks, whereas the mutant data develops a single peak.
+%
+Therefore, vector diffusion maps can not only recover dynamics, but also help distinguish different expression patterns and developmental trajectories.
 
@@
 %
-To the best of our knowledge, algorithmic approaches to these two tasks have primarily been explored independently of each other. 
+To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other. 
 %
@@
 %
-In parallel, ordering of large-scale cross-sectional data was done in the context of molecular profiling studies that lacked spatial resolution, in which data are vectors describing the amounts of different chemical species such as mRNA abundances \cite{trapnell2014dynamics, anavy2014blind, gupta2008extracting}.
+In parallel, ordering of large-scale cross-sectional data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different chemical species such as mRNA abundances \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
 %
@@
 %
-In the first example, correct orientation and temporal order of images was known {\it a priori}, providing a test case for the algorithm in which both of these features were recovered with high accuracy.
+In the first example, correct orientation and temporal order of images was known {\it a priori}, providing a test case for the algorithm in which both of these features were accurately recovered. 
 %
-For this particular data set, a simpler approach would suffice. 
+For this particular data set, a simpler approach would have sufficed. 
 %
@@
 %
-This approach was clearly not sufficient for the second data set (see Figure~S3), collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.
+This approach was clearly not sufficient for the second data set (see \fig~S3), collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.
 %
-However, vector diffusion maps could readily extract a highly nontrivial spatiotemporal trajectory. 
+However, vector diffusion maps could readily extract a nontrivial spatiotemporal trajectory. 
 %
-Finally, the algorithm was applied to a heterogeneous data set, composed of images from embryos that follow distinct developmental paths. 
+Finally, the algorithm was applied to a heterogeneous data set, composed of images from embryos of different genotypes. 
 %
-We found that low-dimensional projection of registered data reveals these distinct paths and can correctly distinguish embryos of different genotypes.
+In this case, low-dimensional projection of registered data revealed distinct paths that correlated with the developmental trajectories of the mutant and wild type populations.
 
-Angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy imaging data \cite{singer2012vector, singer2011three}. 
+Previously, angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy imaging data \cite{singer2012vector, singer2011three}. 
 %
-Because of the high levels of noise in these images, thousands of images are needed for successful shape reconstruction. 
+Because of the high levels of instrument noise in these images, thousands of images were needed for successful shape reconstruction. 
 %
-Based on our work, we expect that a much smaller data set may be sufficient for successful reconstruction of developmental trajectories, providing the first step in systems level studies of embryogenesis.
+Based on our work, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories, providing the first step in systems level studies of embryogenesis.
 %
@@
 %
-One can also test which physical variable, among a set of candidate physical variables, is most correlated with the uncovered vector diffusion maps coordinate to help understand which physical quantities change across the developmental trajectory, (e.g., total intensity of one signal, roundness of the embryo).
+One can also test which physical variable, among a set of candidate variables, is most correlated with the uncovered vector diffusion maps coordinate to gain insight into which physical quantities change across the developmental trajectory (e.g., total intensity of one signal, roundness of the embryo).
 %
-Another important application is related to the joint analysis of data sets provided by different imaging approaches, such as live imaging data of tissue morphogenesis and snapshots of cell signaling and gene expression from fixed embryos \cite{krzic2012multiview, ichikawa2014live, rubel2010coupling}.
+Here, we have used vector diffusion maps to register the images with respect to translations and rotations; however, it is conceivable that one could factor out other sources of variability, such as reflections or dilations, in the images.
 %
-Here, we have used vector diffusion maps to register the images with respect to translations and rotations; however, it is conceviable that one could factor out other sources of variability, such as reflections or scaling, in the images.
+Another important application is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \cite{krzic2012multiview, ichikawa2014live, rubel2010coupling}.
 %
-Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems and using multiple imaging frameworks, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.
+Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.
 
@@
 %
-%Furthermore, dimensionality reduction techniques such as vector diffusion maps do not constrain us to one-dimensional parameterizations/organizations of the data, as illustrated in the mutant data set example (Figure~\ref{fig:data3}).
+%Furthermore, dimensionality reduction techniques such as vector diffusion maps do not constrain us to one-dimensional parameterizations/organizations of the data, as illustrated in the mutant data set example (\fig~\ref{fig:data3}).
 %%
@@
 The progression of membrane invagination during cellularization was used to assign time point to each embryo in the third hour of development.
-Time-lapse imaging of Sqh-GFP embryo was used to generate correlation curve between the membrane ingression length and the developmental time (n=7) (Figure S1B).
+Time-lapse imaging of Sqh-GFP embryo was used to generate correlation curve between the membrane ingression length and the developmental time (n=7) (\fig S1B).
 The rate at which membrane invaginates was highly reproducible from embryo to embryo.
-In the experiments with fixed embryos, we measured the membrane length by co-staining embryos with a membrane marker (Sqh-GFP) or by taking phase-contrast images (Figure S1A).
+In the experiments with fixed embryos, we measured the membrane length by co-staining embryos with a membrane marker (Sqh-GFP) or by taking phase-contrast images (\fig S1A).
 We used the correlation curve to assign each embryo to a time window of 2--3~minutes.
@@
 
-\bibliographystyle{pnas}
+\bibliographystyle{pnas2}
 \bibliography{background_reading/references,../../references/references}

### 2014-05-28 image_analysis_paper/si_information.tex
added edits to paper and si

@@
-\documentclass[12pt]{article}
+\documentclass{pnastwo}
+%\documentclass[10pt,twocolumn]{article}
 
-\usepackage[margin=1in]{geometry}
-
-\usepackage{graphicx}
+%\usepackage[margin=1in]{geometry}
+%\usepackage{graphicx}
 \graphicspath{ {paper_figures2/} }
@@
 
+\newcommand{\fig}[0]{Fig.}
+
 \begin{document}
 
-\section{Raw data of mutant and wild type embryos}
+\begin{article}
 
-\begin{figure}[H]
-\includegraphics[height=9cm]{raw_data3b}
-\caption{Images of cellularizing {\it Drosophila} embryos.
-Each image is of a different embryo in a different rotational orientation.
-Half of the images are of wild type embryos, and half of the images are of mutant embryos.}
-\end{figure}
-
-\section{Experimental Methods}
+\section{Materials and Methods}
 
@@
 
-Todo:
 \begin{itemize}
@@
 
-\begin{figure}[H]
-\includegraphics[width=8.4cm, trim=0cm 5.6cm 0cm 0cm, clip]{SI_fig6}
-\caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different levels of development.
-{\it (B)} Membrane length as a function of time.
-The points corresponding to the images in {\it A} are indicated with arrows.}
-\label{fig:membrane_compare}
-\end{figure}
-
-\section{Image parameters}
+\subsection{Image parameters}
 
@@
 
-\section{Principal Component Analysis for gastrulating images}
-
-We performed principal component analysis on the images shown in Figure~4 of the paper.
-%
-We first registered the images using angular synchronization \cite{singer2011angular}, and then computed the principal components for the data set.
-%
-We ordered the images by the projection onto the first principal component; the results are shown in Figure~\ref{fig:PCA_results}.
-%
-The dynamics that we uncover are not consistent with the expected biological progression described in the main text.
-%
-Furthermore, the eigenvalue spectrum from PCA indicates that a significant amount of the variance is not captured by the first principal component, thereby motivating us to use nonlinear techniques.
-
-\begin{figure}[H]
-\raisebox{5.2cm}{\sffamily{A}}
-\includegraphics[width=16.8cm]{data2_PCA_ordered}
-
-\raisebox{3.7cm}{\sffamily{B}}
-\includegraphics[width=8.4cm]{data2_PCA_variance}
-\raisebox{3.7cm}{\sffamily{C}}
-\includegraphics[width=8.4cm]{data2_PCA_proj}
-
-\caption{{\it (A)} Images shown in Figure~4 of the paper, now ordered by the first PCA projection coefficient.
-The recovered ordering is not consistent with the known developmental dynamics.
-{\it (B)} The fraction of the variance captured by each of the first 10 principal components.
-Note that components 2--10 have a nonnegligible contribution to the data.
-{\it (C)} Projection onto the first two principal components.}
-\label{fig:PCA_results}
-\end{figure}
-
-\section{Mathematical Algorithms}
+\subsection{Mathematical Algorithms}
 
@@
 %
-We construct concentration profiles defined on a ring, and we assume that each ring is arbitrarily rotated around its center; an example is shown in Figure~\ref{subfig:1d_example}.
+We construct concentration profiles defined on a ring, and we assume that each ring is arbitrarily rotated around its center; an example is shown in \fig~\ref{subfig:1d_example}.
 %
-Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown at the bottom of Figure~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
+Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown at the bottom of \fig~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
 %
-Each concentration profile is a noisy Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
+Each concentration profile is a noisy Gaussian (shown in \fig~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
 %
@@
 %
-Figure \ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
+\fig~\ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
 %
@@
 
-\begin{figure}[H]
-\includegraphics[width=8.4cm]{SI_fig3}
-\caption{{\it (A)} One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
-{\it (B)} Intensity corresponding to the profile in {\it A}.
-{\it (C)} An ensemble of concentration profiles, each of the form described in {\it A}.
-Each row in the array corresponds to a single profile.
-{\it (D)} The profiles in {\it C}, now registered using angular synchronization.
-{\it (E)} The profiles in {\it D}, now temporally ordered using diffusion maps.
-{\it (F)} The profiles in {\it C}, registered and temporally ordered in a single step using vector diffusion maps.}
-\label{fig:1d_demo}
-%\customlabel{fig:1d_demo}{3}
-\customlabel{subfig:1d_example}{\ref{fig:1d_demo}{\it A}}
-\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}{\it B}}
-\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}{\it C}}
-\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}{\it D}}
-\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}{\it E}}
-\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}{\it F}}
-\end{figure}
-
-\subsection{Angular synchronization \cite{singer2011angular}}
+\subsubsection{Angular synchronization \cite{singer2011angular}}
 
@@
 %
-For the one-dimensional discretized profiles shown in Figure~\ref{fig:1d_demo}, we exhaustively search over all $n=100$ possible shifts of the signals to compute the optimal angles in \eqref{eq:opt_angle}.
+For the one-dimensional discretized profiles shown in \fig~\ref{fig:1d_demo}, we exhaustively search over all $n=100$ possible shifts of the signals to compute the optimal angles in \eqref{eq:opt_angle}.
 %
@@
 
-\subsection{Diffusion maps \cite{coifman2005geometric}}
+\subsubsection{Diffusion maps \cite{coifman2005geometric}}
 
@@
 The procedure generalizes when the data lie on higher-dimensional manifolds (not just curves) in data space, where lower eigenvectors give subsequent embedding coordinates for the data.
+%
+The magnitude of the eigenvalues, $|\lambda_i|$, gives a measure of ``importance'' of the corresponding embedding coordinate $\phi_i$.
+%
+Often, one looks for a spectral gap in the eigenvalues to determine how many dimensions are necessary to adequately describe the data.
 
-\subsection{Vector diffusion maps\cite{singer2012vector}}
+\subsubsection{Vector diffusion maps\cite{singer2012vector}}
 
@@
 The $j^{th}$ embedding coordinate will be given by $\psi_{k,1}$, where $jd +1 < k \le (j+1)d$,
-and we select $k$ such that the coordinate $\psi_{k, 1}$ has the largest variability, i.e., the $j^{th}$ coordinate is $\psi_{k,1}$, where $k$ is
-$$ \max_{jd +1 < k \le (j+1)d} \sum_i \psi_{k,1} (i)^2 $$
+and we select $k$ such that the coordinate $\psi_{k, 1}$ has the largest variability, i.e., the $j^{th}$ coordinate is $\psi_{k,1}$, where $k$ is the solution to
+$$ \max_{jd +1 < k \le (j+1)d} \sum_i \psi_{k,1} (i)^2. $$
+%
+As in diffusion maps, the magnitude of the eigenvalues is related to the importance of the coordinates and gives an indication of the inherent dimensionality of the data. 
+%
+Again, if the rotations and dynamic variation are separable, then we expect the each eigenvalue to appear with multiplicity $d$, as each block eigenvector will yield $d$ eigenvalues of approximately the same value.
+%
+We therefore look for a spectral gap in the ``blocks'' of eigenvalues to determine how many components are necessary to describe the data.
 
 \subsection{Registering images with respect to rotations and translations} \label{subsec:trans_rot_register}
-
-\begin{figure}[H]
-\begin{minipage}{0.2\textwidth}
-\centering
-\includegraphics[width=\textwidth]{sphere_1}\\
-\includegraphics[width=\textwidth]{sphere2_1}\\
-\sffamily{A}
-\end{minipage}
-%
-\begin{minipage}{0.2\textwidth}
-\centering
-\includegraphics[width=\textwidth]{sphere_2}\\
-\includegraphics[width=\textwidth]{sphere2_2}\\
-\sffamily{B}
-\end{minipage}
-\begin{minipage}{0.2\textwidth}
-\centering
-\includegraphics[width=\textwidth]{sphere_3}\\
-\includegraphics[width=\textwidth]{sphere2_3}\\
-\sffamily{C}
-\end{minipage}
-\begin{minipage}{0.2\textwidth} 
-\centering
-\includegraphics[width=\textwidth]{sphere_4}\\
-\includegraphics[width=\textwidth]{sphere2_4}\\
-\sffamily{D}
-\end{minipage}
-\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions.
-{\it (A)} The original image.
-{\it (B)} Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image.
-{\it (C)} Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation.
-{\it (D)} Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.
-The top row shows the three-dimensional spheres, and the bottom row shows the (two-dimensional) surface of the sphere in which we are interested.}
-\label{fig:SO3_picture}
-\end{figure}
 
@@
 
-\subsection{Constructing movies from registered and ordered images}
+\subsection{Constructing smooth trajectories from registered and ordered images}
 
@@
 %
-For the images and movie in Figure~4, we take $\sigma^2 = 20$.
-%
+For the images in \fig~4, we take $\sigma^2 = 20$.
 
-\section{Eigenvalue Spectra from VDM}
+\bibliographystyle{pnas2}
+\bibliography{background_reading/references,../../references/references}
 
-\begin{figure}[H]
-\includegraphics[width=5cm]{data1_evals}
-\includegraphics[width=5cm]{data2_evals}
-\includegraphics[width=5cm]{data3_evals}
-\caption{Eigenvalue spectra for three examples.}
+\end{article}
+
+\begin{figure}
+\includegraphics[width=8.4cm, trim=0cm 5.6cm 0cm 0cm, clip]{SI_fig6}
+\caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different times during development.
+The membrane length is indicated by the red bars.
+{\it (B)} Membrane length as a function of time.
+The points corresponding to the images in {\it A} are indicated with arrows.}
+\label{fig:membrane_compare}
 \end{figure}
 
-%\begin{figure}[H]
-%\includegraphics[width=5cm]{data1_coord_var}
-%\includegraphics[width=5cm]{data2_coord_var}
-%\includegraphics[width=5cm]{data3_coord_var}
-%\caption{Variance of $\psi_{k,l}$ for three examples. We use the ones that have the largest variance.}
-%\end{figure}
+\begin{figure}
+\includegraphics[height=9cm]{raw_data3}
+\caption{Images of mutant and wild type {\it Drosophila} embryos during cellularization, used for the analysis in \fig~5.
+Each image is of a different embryo in a different rotational orientation.
+Half of the images are of wild type embryos, and half of the images are of mutant embryos.}
+\end{figure}
 
-%\begin{figure}[H]
-%\begin{tikzpicture}
-%     \node at (0,0)
-%       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding_background}};
-%     \node at (0,0)
-%       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding}};  
-%\end{tikzpicture}
-%\begin{tikzpicture}
-%     \node at (0,0)
-%       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding2_background}};
-%     \node at (0,0)
-%       {\includegraphics[width=8.4cm]{mut_wt_vdm_embedding2}};  
-%\end{tikzpicture}
-%\caption{Embedding for third example using two different $\epsilon$ values. The right one (smaller $\epsilon$) has a tighter structure, but the mutants are more removed.}
-%\end{figure}
+\begin{figure}
+\includegraphics[width=8.4cm]{SI_fig3}
+\caption{Synthetic data set used to illustrate the mathematical algorithms.
+{\it (A)} One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
+{\it (B)} Intensity corresponding to the profile in {\it A}.
+{\it (C)} An ensemble of concentration profiles, each of the form described in {\it A}.
+Each row in the array corresponds to a single profile.
+{\it (D)} The profiles in {\it C}, now registered using angular synchronization.
+{\it (E)} The profiles in {\it D}, now temporally ordered using diffusion maps.
+{\it (F)} The profiles in {\it C}, registered and temporally ordered in a single step using vector diffusion maps.}
+\label{fig:1d_demo}
+%\customlabel{fig:1d_demo}{3}
+\customlabel{subfig:1d_example}{\ref{fig:1d_demo}{\it A}}
+\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}{\it B}}
+\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}{\it C}}
+\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}{\it D}}
+\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}{\it E}}
+\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}{\it F}}
+\end{figure}
 
-\section{FBsPCA}
+\begin{figure*}
+\begin{minipage}{0.2\textwidth}
+\centering
+\includegraphics[width=\textwidth]{sphere_1}\\
+\includegraphics[width=\textwidth]{sphere2_1}\\
+\sffamily{A}
+\end{minipage}
+%
+\begin{minipage}{0.2\textwidth}
+\centering
+\includegraphics[width=\textwidth]{sphere_2}\\
+\includegraphics[width=\textwidth]{sphere2_2}\\
+\sffamily{B}
+\end{minipage}
+\begin{minipage}{0.2\textwidth}
+\centering
+\includegraphics[width=\textwidth]{sphere_3}\\
+\includegraphics[width=\textwidth]{sphere2_3}\\
+\sffamily{C}
+\end{minipage}
+\begin{minipage}{0.2\textwidth} 
+\centering
+\includegraphics[width=\textwidth]{sphere_4}\\
+\includegraphics[width=\textwidth]{sphere2_4}\\
+\sffamily{D}
+\end{minipage}
+\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions.
+{\it (A)} The original image (shown in red), projected onto a portion of a sphere.
+{\it (B)} Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image.
+{\it (C)} Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation.
+{\it (D)} Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.
+The top row shows the three-dimensional spheres, and the bottom row shows the (two-dimensional) surface of the sphere in which we are interested.}
+\label{fig:SO3_picture}
+\end{figure*}
 
-I computed the FBsPCA basis for the set of images (separately for the red and green signals).
+\begin{figure*}
+\raisebox{4.5cm}{\sffamily{A}}
+\includegraphics[width=5cm]{data1_evals}
+\raisebox{4.5cm}{\sffamily{B}}
+\includegraphics[width=5cm]{data2_evals}
+\raisebox{4.5cm}{\sffamily{C}}
+\includegraphics[width=5cm]{data3_evals}
+
+\caption{The eigenvalue spectra from vector diffusion maps analysis for the three data sets. 
 %
-I then computed the bispectrum coefficients for each image expanded in the FBsPCA basis, and used these coefficients in a DMAPS calculation.
+As expected, the eigenvalues appear in groups of 3, as $d=3$ when using vector diffusion maps to register with respect to translations and rotations. 
 %
-The results are shown below.
+(A) Spectrum from images during cellularization (\fig~3).
+As expected, there is a spectral gap after the first (non-trivial) group of eigenvalues: the first three eigenvalues correspond to the rotational and translational alignment, the second three eigenvalues correspond to the first VDM component which uncovers the developmental dynamics, and the then there is an appreciable gap before the third group of eigenvalues.
 %
-However, because the bispectrum is not stable to deformations, it does not seem to be able to identify neighboring images.
+(B) Spectrum from images during gastrulation (\fig~4). Again, 
+there is a spectral gap after the first (non-trivial) group of eigenvalues.
+%
+(C) Spectrum from images of mutant and wild type embryos during cellularization (\fig~5).
+We see a group of three eigenvalues corresponding to the translational and rotational alignments, one group that is clearly significant, and then a second group that, while less important, is still above the level of noise of the subsequent eigenvalues, indicating that two components are necessary to describe the data.
+}
+\label{fig:eval_spectra}
+\end{figure*}
 
-\begin{figure}[H]
+\begin{figure}
+\includegraphics[width=8.7cm]{PCA_ordered}
+\caption{Images of embryos during cellularization (from \fig~3), registered using angular synchronization \cite{singer2011angular} and ordered by the first PCA projection coefficient.
+The recovered dynamics are qualitatively consistent with those recovered by vector diffusion maps.
+}
+\label{fig:PCA_data1}
+\end{figure}
+
+\begin{figure*}
+\raisebox{5.2cm}{\sffamily{A}}
+\includegraphics[width=16.8cm]{data2_PCA_ordered}
+
+\raisebox{3.7cm}{\sffamily{B}}
+\includegraphics[width=8.4cm]{data2_PCA_variance}
+\raisebox{3.7cm}{\sffamily{C}}
+\includegraphics[width=8.4cm]{data2_PCA_proj}
+
+\caption{{\it (A)} Images shown in \fig~4 of the paper, registered using angular synchronization \cite{singer2011angular} and ordered by the first PCA projection coefficient.
+The recovered ordering is not consistent with the known developmental dynamics.
+{\it (B)} The fraction of the variance captured by each of the first 10 principal components.
+Note that components 2--10 have a nonnegligible contribution to the data.
+{\it (C)} Projection onto the first two principal components.}
+\label{fig:PCA_results}
+\end{figure*}
+
+\begin{figure*}[t]
 \raisebox{7.6cm}{\sffamily{A}}
@@
 \includegraphics[width=4cm, angle=90]{FBsPCA_dmaps}
-\caption{(A) Images used for FBsPCA.
+\caption{I computed the FBsPCA basis for the set of images (separately for the red and green signals).
+%
+I then computed the bispectrum coefficients for each image expanded in the FBsPCA basis, and used these coefficients in a DMAPS calculation.
+%
+The results are shown below.
+%
+However, because the bispectrum is not stable to deformations, it does not seem to be able to identify neighboring images.
+%
+(A) Images used for FBsPCA.
 (B) Eigenvalue spectrum from dmaps, using bispectrum coefficients.
-(C) First 4 dmaps variables, computed from the bispectrum coefficeints, plotted against the membrane thickness.
+There is no spectral gap after the first component, as would be expected for this particular data set.
+(C) First 4 diffusion maps variables, computed from the bispectrum coefficeints, plotted against the membrane thickness.
 Note there is no appreciable correlation: each dmaps variable simply selects one image from the data set as ``abnormal''.}
-\end{figure}
-
-\bibliographystyle{pnas}
-\bibliography{background_reading/references,../../references/references}
+\end{figure*}
 

### 2014-05-28 image_analysis_paper/test_PCA.m

@@ -175,6 +175,9 @@ imshow(eigenimages(:, :, :, 2),'border','tight')
 % saveas(gcf,sprintf('%s/PCA_eigenimage2', im_save_dir), 'pdf')
 
 [~, I] = sort(proj_coeff(:, 1));
+if corr(I, mem_lengths) < 0 
+    I = flipud(I);
+end
 figure;
 set(gcf, 'paperunits', 'centimeters')
 set(gcf, 'papersize', [8 8])
@@ -190,6 +193,7 @@ for i=1:m
 end
 % saveas(gcf,sprintf('%s/PCA_ordered', im_save_dir), 'pdf')
 
+return
 figure;
 set(gcf, 'paperunits', 'centimeters')
 set(gcf, 'papersize', [4 4])

### 2014-05-28 image_analysis_paper/pnas_draft2.tex
Added small edits

@@
 %
-Recent advances in physical manipulation and imaging of embryos produce rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different developmental time point and at a different geometric orientation.
+%Recent advances in physical manipulation and imaging of embryos produce rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different developmental time point and at a different geometric orientation.
 %
@@
 %
-Our goal is to organize data sets from cross-sectional studies of developmental processes in a meaningful and informative way.
+%Our goal is to organize data sets from cross-sectional studies of developmental processes in a meaningful and informative way.
 %
@@
 %
-Vector diffusion maps combines both steps into a single computation.
+Vector diffusion maps combines both steps into a single computation (see \SI).
 %
-The algorithm uses information about pairwise alignments {\em and} pairwise distances to find a global parameterization of the data, while registering images which are ``close'' (see \SI).
+%The algorithm uses information about pairwise alignments {\em and} pairwise distances to find a global parameterization of the data, while registering images which are ``close'' .
 %
-It is thus perfectly suited for our task of registering and temporally ordering cross-sectional imaging data.
+%It is thus perfectly suited for our task of registering and temporally ordering cross-sectional imaging data.
 %

### 2014-05-29 image_analysis_paper/paper_figures2/fig5.tex
edited paper

--- 2014-05-27 image_analysis_paper/fig5.tex
@@
 %\usepackage[margin=0in, paperwidth=8.7cm, paperheight=9.5cm]{geometry}
-\usepackage[margin=0in, paperwidth=8.7cm, paperheight=16.5cm]{geometry}
+\usepackage[margin=0in, paperwidth=8.7cm, paperheight=17cm]{geometry}
 
@@
 \centering
-%%\raisebox{5.4cm}{\textsf{A}}
-%\begin{tikzpicture}
-%     \node at (0,0)
-%       {\includegraphics[width=8cm]{mut_wt_vdm_embedding_background}};
-%     \node at (0,0)
-%       {\includegraphics[width=8cm]{mut_wt_vdm_embedding}};
-%      \node at (-4cm, 2.4cm) {\textsf{A}};
-%\end{tikzpicture}\\
-\raisebox{5.2cm}{\textsf{A}}
-\includegraphics[width=6cm, angle=90]{data3_embed_bw}\\
-\raisebox{5.2cm}{\textsf{B}}
-\includegraphics[width=8cm]{data3_embed_color}\\
+\begin{tikzpicture}
+	\node at (0cm, 0cm) {\includegraphics[width=6cm, angle=90]{data3_embed_bw}};
+	\node at (-4cm, 2.5cm) {\textsf{A}};
+\end{tikzpicture}\\
+\begin{tikzpicture}
+	\node at (0cm, 0cm) {\includegraphics[width=8cm]{data3_embed_color}};
+	\node at (-4cm, 2.5cm) {\textsf{B}};
+	%\draw[help lines,step=.5] (-4,-4) grid (4,4);
+	\draw [draw=black!20!green, ->, ultra thick] (-2, -2) .. controls (-1.1, 0.2) and  (1, 0)  .. (2, -2);
+	\draw [blue, ->, ultra thick] (-2.65, -1.8)  .. controls (-2.3, 0.9) and (-1.7, 1.5) .. (-1.4, 2.1); 
+\end{tikzpicture}\\
 \begin{tikzpicture}

### 2014-05-29 image_analysis_paper/pnas_draft2.tex
edited paper

@@
 
-We show how recently developed dimensionality reduction techniques can successfully address these challenges and reveal developmental trajectories from collections of unregistered and unordered imaging data.
+Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}.
 %
-Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, as have dimensionality reduction techniques \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear} and their applications \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering, kafri2013dynamics}.
+We use a manifold learning algorithm \cite{singer2012vector} which simultaneously addresses the problems of registration and temporal ordering. 
 %
-Here, we use a manifold learning algorithm \cite{singer2012vector} that has been recently developed for high-dimensional imaging data. 
+This algorithm is one of many nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}.
 %
@@
 %
-The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo to within 1--3 minutes \cite{figard2013plasma} (see \fig~S1).
+The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo to within 1--3 minutes \cite{figard2013plasma} (see \fig~S2).
 %
@@
 %
-Because the dynamics of the data set shown in \fig~\ref{fig:data1} are relatively simple, effectively consisting of a single mode (the two dpERK peaks) growing in time, established techniques such as principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data in time (see \fig~S6).
+Because the dynamics of the data set shown in \fig~\ref{fig:data1} are relatively simple, effectively consisting of a single mode (the two dpERK peaks) growing in time, established techniques such as principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data in time (see \fig~S7).
 %
@@
 %
-Simple time-ordering based on this linear projection is therefore not possible (see \fig~S7).
+Simple time-ordering based on this linear projection is therefore not possible (see \fig~S8).
 %
@@
 %
-To illustrate this, we collected a data set consisting of $42$ embryos undergoing cellularization during the third hour of development (see \fig~S2).
+To illustrate this, we collected a data set consisting of $42$ embryos undergoing cellularization during the third hour of development (see \fig~S3).
 %
@@
 
-\begin{figure}[h!]
-\includegraphics{fig5}
-\caption{Mutant and wild type {\xfigtextfontit Drosophila}.
-{\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~S2), projected onto the first two vector diffusion maps coordinates.
-{\xfigtextfontit (B)} The data in {\xfigtextfontit A}, now colored by genotype.
-The two branches separate mutant and wild type samples.
-{\xfigtextfontit (C)} dpERK signal at select wild type points in {\xfigtextfontit B} (denoted by circles).
-The two dpERK peaks which arise during development are indicated by arrows.
-{\xfigtextfontit (C)} dpERK signal at select mutant points in {\xfigtextfontit B} (denoted by circles).
-The single dpERK peak is indicated by an arrow.
-}
-\customlabel{fig:data3}{5}
-\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
-\customlabel{subfig:projection_colored_data3}{\ref{fig:data3}{\it B}}
-\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it C}}
-\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it D}}
-\end{figure}
-
-We computed the vector diffusion maps embedding for our data, and the eigenvalue spectrum (see \fig~S5) indicates that two coordinates may be required for a thorough description of the data. 
+We computed the vector diffusion maps embedding for our data, and the eigenvalue spectrum (see \fig~S6) indicates that two coordinates may be required for a thorough description of the data. 
 %
@@
 Therefore, vector diffusion maps can not only recover dynamics, but also help distinguish different expression patterns and developmental trajectories.
+
+\begin{figure}[h!]
+\includegraphics{fig5}
+\caption{Mutant and wild type {\xfigtextfontit Drosophila}.
+{\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~S3), projected onto the first two vector diffusion maps coordinates.
+{\xfigtextfontit (B)} The data in {\xfigtextfontit A}, now colored by genotype.
+The two branches separate mutant and wild type samples, and the direction of the developmental trajectories are indicated by arrows.
+{\xfigtextfontit (C)} dpERK signal at select wild type points in {\xfigtextfontit B} (denoted by circles).
+The two dpERK peaks which arise during development are indicated by arrows.
+{\xfigtextfontit (C)} dpERK signal at select mutant points in {\xfigtextfontit B} (denoted by circles).
+The single dpERK peak is indicated by an arrow.
+}
+\customlabel{fig:data3}{5}
+\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
+\customlabel{subfig:projection_colored_data3}{\ref{fig:data3}{\it B}}
+\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it C}}
+\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it D}}
+\end{figure}
 
@@
 %
-This approach was clearly not sufficient for the second data set (see \fig~S3), collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.
+This approach was clearly not sufficient for the second data set (see \fig~S4), collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.
 %

### 2014-05-29 image_analysis_paper/si_information.tex
edited paper

@@
 %
-The magnitude of the eigenvalues, $|\lambda_i|$, gives a measure of ``importance'' of the corresponding embedding coordinate $\phi_i$.
-%
-Often, one looks for a spectral gap in the eigenvalues to determine how many dimensions are necessary to adequately describe the data.
 
@@
 $$ \max_{jd +1 < k \le (j+1)d} \sum_i \psi_{k,1} (i)^2. $$
-%
-As in diffusion maps, the magnitude of the eigenvalues is related to the importance of the coordinates and gives an indication of the inherent dimensionality of the data. 
-%
-Again, if the rotations and dynamic variation are separable, then we expect the each eigenvalue to appear with multiplicity $d$, as each block eigenvector will yield $d$ eigenvalues of approximately the same value.
-%
-We therefore look for a spectral gap in the ``blocks'' of eigenvalues to determine how many components are necessary to describe the data.
 
@@
 
+\subsection{Determining dimensionality from eigenvalue spectrum}
+
+We can use the distribution of the eigenvalues from (vector) diffusion maps to help determine the dimensionality of the data.
+%
+For diffusion maps, the largest eigenvalue will always be 1 and correspond to the trivial (constant) eigenvector. 
+%
+The eigenvalues corresponding to noise in the data form a semicircle distribution centered at 0. 
+%
+The eigenvalues corresponding to descriptive/meaningful embedding coordinates will be positive and removed from the semicircle. 
+%
+For vector diffusion maps, the sam structure holds, except each coordinate gives rise to $d$ eigenvalues with (approximately) the same value (as each block eigenvector corresponds to $d$ eigenvalues).
+
 %\subsection{Algorithms}
@@
 
-Once we have registered and ordered the images, we smooth the resulting trajectory to obtain a movie of the average developmental dynamics. 
+Once we have registered and ordered the images, we smooth the resulting trajectory to obtain a picture of the stereotypic developmental dynamics. 
 %
@@
 %
-We define the frame at time $\tau$, denoted $\overline{x}_{\tau}$ (where $1 \le \tau \le m$), as a (Gaussian) weighted average of the images,
+We define the average image at time $\tau$, denoted $\overline{x}_{\tau}$ (where $1 \le \tau \le m$), as a (Gaussian) weighted average of the images,
 \begin{equation}
@@
 \end{article}
+
+\begin{figure}
+\includegraphics[width=8.4cm]{ap_dv}
+\caption{(Left) A lateral view of a {\it Drosophila} embryo stained with DAPI (gray), dpERK (red), and Dorsal (green).
+Embryo is presented so that the anterior (A) side is to the left and the posterior (P) side is to the right.
+The arrow indicates the position where the cross-section of an embryo is imaged.
+(Right) A dorsoventral view of a {\it Drosophila} embryo.
+The dorsal (D) side is up and the ventral (V) side is down.
+Images were collected at the focal plane ~18\% from either the anterior or posterior pole of an embryo (arrow in the left image).
+}
+\end{figure}
 
@@
 \begin{figure*}
-\raisebox{4.5cm}{\sffamily{A}}
-\includegraphics[width=5cm]{data1_evals}
-\raisebox{4.5cm}{\sffamily{B}}
-\includegraphics[width=5cm]{data2_evals}
-\raisebox{4.5cm}{\sffamily{C}}
-\includegraphics[width=5cm]{data3_evals}
-
-\caption{The eigenvalue spectra from vector diffusion maps analysis for the three data sets. 
+%\raisebox{4.5cm}{\sffamily{A}}
+%\includegraphics[width=5cm]{data1_evals}
+\begin{tikzpicture}
+	\node at (0,0) {\includegraphics[width=5cm]{data1_evals_dist}};
+	\node at (-2.5cm, 2cm) {\sffamily{A}};
+	\draw[<->, ultra thick, cyan] (-1.4cm, -1.9cm) -- (1.7cm, -1.9cm);
+	\draw[->, ultra thick, red] (1.97cm, -0.5cm) -- (1.97cm, -1cm);
+\end{tikzpicture}
+%\raisebox{4.5cm}{\sffamily{B}}
+%\includegraphics[width=5cm]{data2_evals}
+\begin{tikzpicture}
+	\node at (0,0) {\includegraphics[width=5cm]{data2_evals_dist}};
+	\node at (-2.5cm, 2cm) {\sffamily{B}};
+	\draw[<->, ultra thick, cyan] (-1.2cm, -1.9cm) -- (1.5cm, -1.9cm);
+	\draw[->, ultra thick, red] (1.97, -0.85cm) -- (1.97cm, -1.34cm);
+\end{tikzpicture}
+%\raisebox{4.5cm}{\sffamily{C}}
+%\includegraphics[width=5cm]{data3_evals}
+\begin{tikzpicture}
+	\node at (0,0) {\includegraphics[width=5cm]{data3_evals_dist}};
+	\node at (-2.5cm, 2cm) {\sffamily{C}};
+	\draw[<->, ultra thick, cyan] (-0.25cm, -1.9cm) -- (0.35cm, -1.9cm);
+	\draw[->, ultra thick, red] (1.97, -1.1cm) -- (1.97cm, -1.6cm);
+	\draw[->, ultra thick, red] (1.52, -1.1cm) -- (1.52cm, -1.6cm);
+	\draw[->, ultra thick, red] (0.64, -1.1cm) -- (0.64cm, -1.6cm);
+\end{tikzpicture}
+\caption{The histograms of the eigenvalues from vector diffusion maps analysis for the three data sets. 
 %
-As expected, the eigenvalues appear in groups of 3, as $d=3$ when using vector diffusion maps to register with respect to translations and rotations. 
+We expect to see an approximately symmetric semicircle distribution centered at 0, which corresponds to noise, and positive eigenvalues separated from the semicircle which correspond to meaningful coordinates.
 %
-(A) Spectrum from images during cellularization (\fig~3).
+Each coordinate yields $d=3$ eigenvalues of (approximately) the same magnitude, and the top $d=3$ eigenvalues ($\lambda=1$) correspond to a trivial embedding coordinate.
+%
+(A) Histogram of eigenvalues from images during cellularization (\fig~3).
+As expected, we see an approximately symmetric distribution that ranges from $-0.8$ to $0.8$ (indicated by the blue arrow), and $6$ eigenvalues that are outside of this distribution (indicated by the red arrow).
+$3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $3$ eigenvalues correspond to the first VDM component which uncovers the developmental dynamics, and the then there is an appreciable gap before the third group of eigenvalues.
+%
+(B) Histogram of eigenvalues from images during gastrulation (\fig~4).
+We see an approximately symmetric distribution that ranges from $-0.7$ to $0.7$ (indicated by the blue arrow), and $6$ eigenvalues that are outside of this distribution (indicated by the red arrow).
+$3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $3$ eigenvalues correspond to the first VDM component which uncovers the developmental dynamics.
+%
+(C) Histogram of eigenvalues from images of mutant and wild type embryos during cellularization (\fig~5).
+We see an approximately symmetric distribution that ranges from $-0.2$ to $0.2$ (indicated by the blue arrow), and $9$ eigenvalues that are outside of this distribution (indicated by the red arrows).
+$3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $2$ eigenvalues correspond to the first two VDM components which uncovers the developmental dynamics of the two genotypes.}
+%there is a spectral gap after the first (non-trivial) group of eigenvalues.
+%\caption{The eigenvalue spectra from vector diffusion maps analysis for the three data sets. 
+%%
+%As expected, the eigenvalues appear in groups of 3, as $d=3$ when using vector diffusion maps to register with respect to translations and rotations. 
+%%
+%(A) Spectrum from images during cellularization (\fig~3).
 As expected, there is a spectral gap after the first (non-trivial) group of eigenvalues: the first three eigenvalues correspond to the rotational and translational alignment, the second three eigenvalues correspond to the first VDM component which uncovers the developmental dynamics, and the then there is an appreciable gap before the third group of eigenvalues.
-%
-(B) Spectrum from images during gastrulation (\fig~4). Again, 
-there is a spectral gap after the first (non-trivial) group of eigenvalues.
-%
-(C) Spectrum from images of mutant and wild type embryos during cellularization (\fig~5).
-We see a group of three eigenvalues corresponding to the translational and rotational alignments, one group that is clearly significant, and then a second group that, while less important, is still above the level of noise of the subsequent eigenvalues, indicating that two components are necessary to describe the data.
+%%
+%(B) Spectrum from images during gastrulation (\fig~4). Again, 
+%there is a spectral gap after the first (non-trivial) group of eigenvalues.
+%%
+%(C) Spectrum from images of mutant and wild type embryos during cellularization (\fig~5).
+%We see a group of three eigenvalues corresponding to the translational and rotational alignments, one group that is clearly significant, and then a second group that, while less important, is still above the level of noise of the subsequent eigenvalues, indicating that two components are necessary to describe the data.
 }

### 2014-05-29 image_analysis_paper/test_late_images.m

@@ -262,7 +262,7 @@ ylabel('PCA projection 2')
 %% VDM
 
 eps = median(W(:))/10;
-neigs = 42;
+neigs = 3*m;
 [R_opt, embed_coord, embed_idx, D] = vdm(R, W, eps, neigs);
 
 figure;
@@ -272,7 +272,27 @@ set(gcf, 'paperposition',[0 0 8 8])
 plot(abs(diag(D)),'.')
 xlabel('k')
 ylabel('|\lambda_k|')
-saveas(gcf,sprintf('%s/data2_evals', im_save_dir), 'pdf')
+% saveas(gcf,sprintf('%s/data2_evals', im_save_dir), 'pdf')
+
+figure;
+set(gcf, 'paperunits', 'centimeters')
+set(gcf, 'papersize', [8 8])
+set(gcf, 'paperposition',[0 0 8 8])
+nbins = 25;
+hist(diag(D), nbins)
+set(gca, 'xlim', [-1 1])
+grid on
+set(gca, 'ytick', 0:3:45)
+set(gca, 'xtick', -1:0.2:1)
+xlabel('\lambda')
+ylabel('Pr(\lambda)')
+% hold on
+% [nelements, centers] = hist(diag(D(7:end, 7:end)), nbins);
+% [alpha, rad] = fit_semicircle(centers, nelements);
+% plot(centers, wigner_semicircle(centers, rad, alpha))
+saveas(gcf,sprintf('%s/data2_evals_dist', im_save_dir), 'pdf')
+
+return
 
 figure;
 set(gcf, 'paperunits', 'centimeters')

### 2014-05-29 image_analysis_paper/wigner_semicircle.m

@@ -1,0 +1,5 @@
+function y = wigner_semicircle(x, R, alpha)
+
+y = alpha * sqrt(R^2-x.^2);
+
+y(abs(x) > R) = 0;

### 2014-05-29 image_analysis_paper/paper_figures2/spheres.tex
Added edits to paper

\documentclass{minimal}

\usepackage{tikz}           %for TikZ graphics
\usepackage{tikz-3dplot} %for 3dplot functionality

%\usepackage[active,tightpage]{preview}      %generates a tightly fitting border around the work
%\PreviewEnvironment{tikzpicture}
%\setlength\PreviewBorder{2mm}

\def\R{1}

\begin{document}
\tdplotsetmaincoords{70}{135}
\tdplotsetrotatedcoords{45}{0}{0}
\begin{tikzpicture}[line join=bevel,tdplot_rotated_coords, fill opacity=.7]

\tdplotsphericalsurfaceplot{36}{18}%
{\R}{black}{red}%
    {\draw[color=black,thick,<->] (-2,0,0) -- (2,0,0) node[anchor=north east]{$y$};}%
    {\draw[color=black,thick,<->] (0,-2,0) -- (0,2,0) node[anchor=north west]{$x$};}%
    {\draw[color=black,thick,<->] (0,0,-2) -- (0,0,2) node[anchor=south]{$z$};}%

    \tdplotsetpolarplotrange{70}{120}{60}{120}
\tdplotsphericalsurfaceplot{36}{18}%
{\R}{black}{blue}%
{}{}{}%

\end{tikzpicture}

\end{document}

### 2014-05-29 image_analysis_paper/pnas_draft2.tex
Added edits to paper

@@
 
-Imaging studies provide unique insights into dynamics of developmental processes. 
+Imaging studies provide unique insights into the dynamics of developmental processes. 
 %
@@
 %
-The goal is then to temporally order the data to reconstruct a representative developmental dynamics from snapshots provided by the collection of embryos.
+The goal is then to temporally order the data to reconstruct representative developmental dynamics from snapshots provided by the collection of embryos.
 %
@@
 
-Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}.
+Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, 
+and most registration algorithms are based on landmarks, such as the head and fins in \fig~\ref{fig:fish}.
 %
-We use a manifold learning algorithm \cite{singer2012vector} which simultaneously addresses the problems of registration and temporal ordering. 
+In contrast, our approach requires no such information. 
+%
+It is based on a manifold learning algorithm \cite{singer2012vector} which simultaneously addresses the problems of registration and temporal ordering. 
 %
@@
 This technique has been used for data mining in
-applications ranging from analysis of cryo-electron microscopy images of individual molecules to face recognition, and is sufficiently general to be applicable to a wide variety of imaging studies. 
+applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules to face recognition, and is sufficiently general to be applicable to a wide variety of biological imaging studies.
 %
-Here it is adapted for simultaneous registration and temporal ordering of images from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
+Here the algorithm is adapted for simultaneous registration and temporal ordering of images from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
 %
-We analyze three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics \cite{jaeger2012drosophila}.
+To illustrate our approach, we analyze three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics \cite{jaeger2012drosophila}.
 %
@@
 %
-The approach is based on vector diffusion maps \cite{singer2012vector}, a manifold learning
+Our approach is based on vector diffusion maps \cite{singer2012vector}, a manifold learning
 technique developed for data sets which contain two types of sources of variability:
@@
 %
-
-Vector diffusion maps has been used to classify images from cryo-electron microscopy (cryo-EM) experiments, where each image is a projection of a randomly oriented molecule.
+%Vector diffusion maps has been used to classify images from cryo-electron microscopy (cryo-EM) experiments, where each image is a projection of a randomly oriented molecule.
 %
@@
 %%
-Here, we will use the algorithm to register images of {\it Drosophila} embryos with respect to rotations and translations, as well as uncover the main direction of variability {\it after} removing translational and rotational symmetries.
+We will use the algorithm to register images of {\it Drosophila} embryos with respect to rotations and translations, as well as uncover the main direction of variability {\it after} removing translational and rotational symmetries.
 %
@@
 %
-Vector diffusion maps combines both steps into a single computation (see \SI).
+Vector diffusion maps combines both steps into a single computation (see \SI), simultaneously addressing the tasks of registration and temporal ordering.
 %
@@
 
-Previous studies established that throughout the third hour of development, the nuclear localization of Dl is unimodal and peaked at the ventral side of the embryo, reviewed in \cite{rushlow2012temporal}.
+Previous studies established that throughout the third hour of development, the nuclear localization of Dl is unimodal and peaked at the ventral side of the embryo (reviewed in \cite{rushlow2012temporal}).
 %
@@
 %
-The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo to within 1--3 minutes \cite{figard2013plasma} (see \fig~S2).
+The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo to within 2--3 minutes \cite{figard2013plasma} (see \fig~S2).
 %
@@
 %
-Thus, we have validated a dimensionality reduction approach to temporal ordering and registration of data on a data set where image orientation and ordering can be assessed independently.
+Thus, we have validated a dimensionality reduction approach to temporal ordering and registration on a data set where image orientation and ordering can be assessed independently.
 
@@
 %
-During this time window, germband extension causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side.
+During this time window, germband extension causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
 %
@@
 %
-To illustrate this, we collected a data set consisting of $42$ embryos undergoing cellularization during the third hour of development (see \fig~S3).
+To illustrate this point, we collected a data set consisting of $42$ embryos undergoing cellularization during the third hour of development (see \fig~S3).
 %
@@
 
-We computed the vector diffusion maps embedding for our data, and the eigenvalue spectrum (see \fig~S6) indicates that two coordinates may be required for a thorough description of the data. 
+We computed the vector diffusion maps embedding for our data, and the eigenvalue distribution (see \fig~S6) indicates that two coordinates may be required for a thorough description of the data. 
 %
@@
 %
-In parallel, ordering of large-scale cross-sectional data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different chemical species such as mRNA abundances \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
+In parallel, ordering of large-scale cross-sectional data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different chemical species, such as mRNA abundances \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
 %
@@
 %
-Indeed, because of the relative simplicity of developmental dynamics (monotonic increase in the amplitude of a localized signaling pattern and relatively small tissue deformation), temporal order could have been inferred by projection of registered images on the first principal component of the data set.
+Indeed, because of the relative simplicity of developmental dynamics (monotonic increase in the amplitude of a localized signaling pattern and relatively small tissue deformation), temporal order could have been inferred by projection of registered images on the first principal component of the data set (see \fig~S7).
 %
-This approach was clearly not sufficient for the second data set (see \fig~S4), collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.
+This approach was clearly not sufficient for the second data set (see \fig~S8), collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.
 %
@@
 The progression of membrane invagination during cellularization was used to assign time point to each embryo in the third hour of development.
-Time-lapse imaging of Sqh-GFP embryo was used to generate correlation curve between the membrane ingression length and the developmental time (n=7) (\fig S1B).
+Time-lapse imaging of Sqh-GFP embryo was used to generate correlation curve between the membrane ingression length and the developmental time (n=7) (\fig S2B).
 The rate at which membrane invaginates was highly reproducible from embryo to embryo.
-In the experiments with fixed embryos, we measured the membrane length by co-staining embryos with a membrane marker (Sqh-GFP) or by taking phase-contrast images (\fig S1A).
+In the experiments with fixed embryos, we measured the membrane length by co-staining embryos with a membrane marker (Sqh-GFP) or by taking phase-contrast images (\fig S2A).
 We used the correlation curve to assign each embryo to a time window of 2--3~minutes.
@@
 
-\bibliographystyle{pnas2}
-\bibliography{background_reading/references,../../references/references}
+%\bibliographystyle{pnas2}
+%\bibliography{background_reading/references,../../references/references}
 
-%\begin{thebibliography}{}
+\begin{thebibliography}{10}
 
-%\end{thebibliography}
+\bibitem{diggle2002analysis}
+Diggle P, Heagerty P, Liang K~Y, Zeger S (2002) \textit{Analysis of
+  longitudinal data} (Oxford University Press).
+
+\bibitem{roelens2013live}
+Roelens B, Chanes J~D~L~H, Dostatni N (2013) Live imaging of bicoid-dependent
+  transcription in \textit{{D}rosophila} embryos. \textit{Curr Biol}
+  23:1--5.
+
+\bibitem{keller2013imaging}
+Keller P~J (2013) Imaging morphogenesis: technological advances and biological
+  insights. \textit{Science} 340(6137).
+
+\bibitem{jaeger2004dynamic}
+Jaeger J, Surkova S, Blagov M, Janssens H, Kosman D, \textit{et~al.} (2004)
+  Dynamic control of positional information in the early \textit{{D}rosophila}
+  embryo. \textit{Nature} 430(6997):368--371.
+
+\bibitem{fowlkes2008quantitative}
+Fowlkes C~C, Hendriks C~L~L, Ker{\"a}nen S~V, Weber G~H, R{\"u}bel O,
+  \textit{et~al.} (2008) A quantitative spatiotemporal atlas of gene expression
+  in the \textit{{D}rosophila} blastoderm. \textit{Cell} 133(2):364--374.
+
+\bibitem{ng2012large}
+Ng L~L, Sunkin S~M, Feng D, Lau C, Dang C, \textit{et~al.} (2012) Large-scale
+  neuroinformatics for in situ hybridization data in the mouse brain.
+  \textit{Int Rev Neurobiol} 104:159--182.
+
+\bibitem{richardson2014emage}
+Richardson L, Stevenson P, Venkataraman S, Yang Y, Burton N, \textit{et~al.}
+  (2014) Emage: Electronic mouse atlas of gene expression, \textit{Mouse
+  Molecular Embryology} (Springer), pp. 61--79.
+
+\bibitem{castro2009automatic}
+Castro C, Luengo-Oroz M, Desnoulez S, Duloquin L, Fernandez-de Manuel L,
+  \textit{et~al.} (2009) An automatic quantification and registration strategy
+  to create a gene expression atlas of zebrafish embryogenesis,
+  \textit{Conf Proc IEEE Eng Med Biol Soc}, pp. 1469--1472.
+
+\bibitem{zitova2003image}
+Zitova B, Flusser J (2003) Image registration methods: a survey. \textit{Image Vis Comput} 21(11):977--1000.
+
+\bibitem{rowley1998rotation}
+Rowley H~A, Baluja S, Kanade T (1998) Rotation invariant neural network-based
+  face detection, \textit{Computer Vision and Pattern Recognition, 1998.
+  Proceedings. 1998 IEEE Computer Society Conference on} (IEEE), pp. 38--44.
+
+\bibitem{hajnal2010medical}
+Hajnal J~V, Hill D~L (2010) \textit{Medical image registration} (CRC press).
+
+\bibitem{greenspan1994rotation}
+Greenspan H, Belongie S, Goodman R, Perona P (1994) Rotation invariant texture
+  recognition using a steerable pyramid, \textit{Pattern Recognition, 1994.
+  Vol. 2-Conference B: Computer Vision \& Image Processing., Proceedings of the
+  12th IAPR International. Conference on} (IEEE), volume~2, pp. 162--167.
+
+\bibitem{zhao2003face}
+Zhao W, Chellappa R, Phillips P~J, Rosenfeld A (2003) Face recognition: A
+  literature survey. \textit{ACM Computing Surveys (CSUR)} 35(4):399--458.
+
+\bibitem{singer2012vector}
+Singer A, Wu H~T (2012) Vector diffusion maps and the connection laplacian.
+  \textit{Commun Pure Appl Math} 65(8):1067--1144.
+
+\bibitem{Belkin2003}
+Belkin M, Niyogi P (2003) Laplacian eigenmaps for dimensionality reduction and
+  data representation. \textit{Neural Comput} 15:1373--1396.
+
+\bibitem{coifman2005geometric}
+Coifman R~R, Lafon S, Lee A~B, Maggioni M, Nadler B, \textit{et~al.} (2005)
+  Geometric diffusions as a tool for harmonic analysis and structure definition
+  of data: Diffusion maps. \textit{Proc Natl Acad Sci USA} 102(21):7426--7431.
+
+\bibitem{coifman2006geometric}
+Coifman R~R, Lafon S (2006) Geometric harmonics: a novel tool for multiscale
+  out-of-sample extension of empirical functions. \textit{Appl Comput Harmon Anal} 21(1):31--52.
+
+\bibitem{tenenbaum2000global}
+Tenenbaum J~B, De~Silva V, Langford J~C (2000) A global geometric framework for
+  nonlinear dimensionality reduction. \textit{Science} 290(5500):2319--2323.
+
+\bibitem{roweis2000nonlinear}
+Roweis S~T, Saul L~K (2000) Nonlinear dimensionality reduction by locally
+  linear embedding. \textit{Science} 290(5500):2323--2326.
+
+\bibitem{jaeger2012drosophila}
+Jaeger J, Manu, Reinitz J (2012) \textit{{D}rosophila} blastoderm patterning.
+  \textit{Curr Opin Genet Dev} 22(6):533 -- 541, genetics of system biology.
+
+\bibitem{singer2011angular}
+Singer A (2011) Angular synchronization by eigenvectors and semidefinite
+  programming. \textit{Appl Comput Harmon Anal}
+  30(1):20--36.
+
+\bibitem{ahuja2007template}
+Ahuja S, Kevrekidis I~G, Rowley C~W (2007) Template-based stabilization of
+  relative equilibria in systems with continuous symmetry. \textit{Journal of
+  Nonlinear Science} 17(2):109--143.
+
+\bibitem{ian1998statistical}
+Dryden I~L, Mardia K (1998) \textit{Statistical shape analysis}, Wiley series
+  in probability and statistics: Probability and statistics (J. Wiley).
+
+\bibitem{foe1983studies}
+Foe V~E, Alberts B~M (1983) Studies of nuclear and cytoplasmic behaviour during
+  the five mitotic cycles that precede gastrulation in \textit{{D}rosophila}
+  embryogenesis. \textit{J Cell Sci} 61(1):31--70.
+
+\bibitem{rushlow2012temporal}
+Rushlow C~A, Shvartsman S~Y (2012) Temporal dynamics, spatial range, and
+  transcriptional interpretation of the dorsal morphogen gradient.
+  \textit{Curr Opin Genet Dev} 22(6):542--546.
+
+\bibitem{lim2013kinetics}
+Lim B, Samper N, Lu H, Rushlow C, Jim{\'e}nez G, \textit{et~al.} (2013)
+  Kinetics of gene derepression by erk signaling. \textit{Proc Natl Acad Sci USA} 110(25):10330--10335.
+
+\bibitem{figard2013plasma}
+Figard L, Xu H, Garcia H~G, Golding I, Sokac A~M (2013) The plasma membrane
+  flattens out to fuel cell-surface growth during \textit{{D}rosophila}
+  cellularization. \textit{Dev Cell} 27(6):648--655.
+
+\bibitem{shlens2005tutorial}
+Shlens J (2005) A tutorial on principal component analysis. \textit{Systems
+  Neurobiology Laboratory, University of California at San Diego} .
+
+\bibitem{leptin2005gastrulation}
+Leptin M (2005) Gastrulation movements: the logic and the nuts and bolts.
+  \textit{Dev Cell} 8(3):305--320.
+
+\bibitem{yuan2014automated}
+Yuan L, Pan C, Ji S, McCutchan M, Zhou Z~H, \textit{et~al.} (2014) Automated
+  annotation of developmental stages of \textit{{D}rosophila} embryos in images
+  containing spatial patterns of expression. \textit{Bioinformatics}
+  30(2):266--273.
+
+\bibitem{surkova2008characterization}
+Surkova S, Kosman D, Kozlov K, Myasnikova E, Samsonova A~A, \textit{et~al.}
+  (2008) Characterization of the \textit{{D}rosophila} segment determination
+  morphome. \textit{Dev Biol} 313(2):844--862.
+
+\bibitem{anavy2014blind}
+Anavy L, Levin M, Khair S, Nakanishi N, Fernandez-Valverde S~L, \textit{et~al.}
+  (2014) Blind ordering of large-scale transcriptomic developmental
+  timecourses. \textit{Development} 141(5):1161--1166.
+
+\bibitem{trapnell2014dynamics}
+Trapnell C, Cacchiarelli D, Grimsby J, Pokharel P, Li S, \textit{et~al.} (2014)
+  The dynamics and regulators of cell fate decisions are revealed by
+  pseudotemporal ordering of single cells. \textit{Nat Biotechnol} 32(4):381--386.
+
+\bibitem{gupta2008extracting}
+Gupta A, Bar-Joseph Z (2008) Extracting dynamics from static cancer expression
+  data. \textit{IEEE/ACM Trans Comput Biol Bioinform} 5(2):172--182.
+
+\bibitem{singer2011three}
+Singer A, Shkolnisky Y (2011) Three-dimensional structure determination from
+  common lines in cryo-em by eigenvectors and semidefinite programming.
+  \textit{SIAM J Imaging Sci} 4(2):543--572.
+
+\bibitem{silverman2005functional}
+Ramsay J~O, Silverman B~W (2005) \textit{Functional Data Analysis}, Springer
+  Series in Statistics (Springer).
+
+\bibitem{krzic2012multiview}
+Krzic U, Gunther S, Saunders T~E, Streichan S~J, Hufnagel L (2012) Multiview
+  light-sheet microscope for rapid in toto imaging. \textit{Nat Methods}
+  9(7):730--733.
+
+\bibitem{ichikawa2014live}
+Ichikawa T, Nakazato K, Keller P~J, Kajiura-Kobayashi H, Stelzer E~H,
+  \textit{et~al.} (2014) Live imaging and quantitative analysis of gastrulation
+  in mouse embryos using light-sheet microscopy and 3d tracking tools.
+  \textit{Nat Protoc} 9(3):575--585.
+
+\bibitem{rubel2010coupling}
+R{\"u}bel O, Ahern S, Bethel E, Biggin M~D, Childs H, \textit{et~al.} (2010)
+  Coupling visualization and data analysis for knowledge discovery from
+  multi-dimensional scientific data. \textit{Procedia Comput Sci}
+  1(1):1757--1764.
+
+\end{thebibliography}
 

### 2014-05-29 image_analysis_paper/si_information.tex
Added edits to paper

@@
 
-\subsection{Measuring developmental time during cellularization}
-
-\begin{itemize}
-\item We need to have a supplementary figure that shows the lateral view, that makes it even more clear that we work with optical sections.
-
-\item Need image labeled with D/V
-
-\end{itemize}
-
 \subsection{Image parameters}
@@
 The same laser setting is maintained within each experiment.
-
-{\bf discuss how we choose weights of signals}
 
@@
 %
-The eigenvalues corresponding to noise in the data form a semicircle distribution centered at 0. 
+The eigenvalues corresponding to noise in the data form a semicircle distribution centered at 0 \cite{wigner1955characteristic, wigner1958distribution}. 
 %
@@
 %
-For vector diffusion maps, the sam structure holds, except each coordinate gives rise to $d$ eigenvalues with (approximately) the same value (as each block eigenvector corresponds to $d$ eigenvalues).
+Therefore, by counting how many eigenvalues are removed from the semicircle (excluding the trivial $\lambda=1$ eigenvalue), we can obtain an estimate of the dimensionality of the data.
+
+For vector diffusion maps, the same structure holds: we expect trivial eigenvalues, a semicircle corresponding to noise, and positive eigenvalues corresponding to meaningful embedding coordinates.
+%
+However, in vector diffusion maps, each coordinate gives rise to $d$ eigenvalues with (approximately) the same value (as each block eigenvector corresponds to $d$ eigenvalues).
+%
+Therefore, by counting how many sets of $d$ eigenvalues are removed from the semicircle (excluding the trivial set of $d$ eigenvalues, all with  $\lambda=1$), we can obtain an estimate of the dimensionality of the data.
 
@@
 \caption{(Left) A lateral view of a {\it Drosophila} embryo stained with DAPI (gray), dpERK (red), and Dorsal (green).
-Embryo is presented so that the anterior (A) side is to the left and the posterior (P) side is to the right.
+The embryo is presented so that the anterior (A) side is to the left and the posterior (P) side is to the right.
 The arrow indicates the position where the cross-section of an embryo is imaged.
@@
 \begin{figure*}
-\begin{minipage}{0.2\textwidth}
+\begin{minipage}{0.22\textwidth}
 \centering
-\includegraphics[width=\textwidth]{sphere_1}\\
+%\includegraphics[width=\textwidth]{sphere_1}\\
 \includegraphics[width=\textwidth]{sphere2_1}\\
@@
 %
-\begin{minipage}{0.2\textwidth}
+\hfill
+%
+\begin{minipage}{0.22\textwidth}
 \centering
-\includegraphics[width=\textwidth]{sphere_2}\\
+%\includegraphics[width=\textwidth]{sphere_2}\\
 \includegraphics[width=\textwidth]{sphere2_2}\\
@@
 \end{minipage}
-\begin{minipage}{0.2\textwidth}
+\begin{minipage}{0.22\textwidth}
 \centering
-\includegraphics[width=\textwidth]{sphere_3}\\
+%\includegraphics[width=\textwidth]{sphere_3}\\
 \includegraphics[width=\textwidth]{sphere2_3}\\
@@
 \end{minipage}
-\begin{minipage}{0.2\textwidth} 
+\begin{minipage}{0.22\textwidth} 
 \centering
-\includegraphics[width=\textwidth]{sphere_4}\\
+%\includegraphics[width=\textwidth]{sphere_4}\\
 \includegraphics[width=\textwidth]{sphere2_4}\\
@@
 {\it (C)} Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation.
-{\it (D)} Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.
-The top row shows the three-dimensional spheres, and the bottom row shows the (two-dimensional) surface of the sphere in which we are interested.}
+{\it (D)} Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.}
 \label{fig:SO3_picture}
@@
 	\node at (-2.5cm, 2cm) {\sffamily{A}};
-	\draw[<->, ultra thick, cyan] (-1.4cm, -1.9cm) -- (1.7cm, -1.9cm);
+%	\draw[<->, ultra thick, cyan] (-1.4cm, -1.9cm) -- (1.7cm, -1.9cm);
+	\draw[<->, ultra thick, cyan] (-1.4cm, -1.85cm) -- (1.7cm, -1.85cm);
 	\draw[->, ultra thick, red] (1.97cm, -0.5cm) -- (1.97cm, -1cm);
@@
 	\node at (-2.5cm, 2cm) {\sffamily{B}};
-	\draw[<->, ultra thick, cyan] (-1.2cm, -1.9cm) -- (1.5cm, -1.9cm);
+	\draw[<->, ultra thick, cyan] (-1.2cm, -1.85cm) -- (1.5cm, -1.85cm);
 	\draw[->, ultra thick, red] (1.97, -0.85cm) -- (1.97cm, -1.34cm);
@@
 	\node at (-2.5cm, 2cm) {\sffamily{C}};
-	\draw[<->, ultra thick, cyan] (-0.25cm, -1.9cm) -- (0.35cm, -1.9cm);
+	\draw[<->, ultra thick, cyan] (-0.25cm, -1.85cm) -- (0.35cm, -1.85cm);
 	\draw[->, ultra thick, red] (1.97, -1.1cm) -- (1.97cm, -1.6cm);

### 2014-05-29 image_analysis_paper/make_sphere.m

@@ -8,7 +8,7 @@ fmt = '-djpeg';
 %set(0,'DefaultLineLineWidth',2)
 %set(0,'DefaultAxesFontSize',20)
 
-im_save_dir = 'paper_figures';
+im_save_dir = 'paper_figures2';
 
 n = 25;
 
@@ -28,18 +28,20 @@ dir = [0 0 0;
     1 0 0];
 theta = 0:0.1:pi;
 r = 0.2;
-r1 = 1.25;
+r1 = 1.4;
 
 for i=1:4
     figure;
-    set(gcf, 'papersize', [0.1 0.1])
+    set(gcf, 'units', 'centimeters')
+    set(gcf, 'papersize', [4 4])
+    set(gcf, 'paperposition', [0 0 4 4])
     h = surf(X, Y, Z, C, 'edgecolor','k');
-    set(gca,'position',[0 0 1 1],'units','normalized')
+     set(gca,'position',[0 0 1 1],'units','normalized')
     camlight(45, -20);
     lighting phong
     %material dull
-    set(h, 'diffusestrength', 1.0);
-    set(h, 'specularstrength', 0.25);
+    set(h, 'diffusestrength', 1);
+    set(h, 'specularstrength', 1);
     %set(gca, 'ambientlightcolor', [1 0 0]);
     set(h, 'AmbientStrength', 0.7);
     %set(h, 'SpecularColorReflectance', 0.25);
@@ -55,31 +57,37 @@ for i=1:4
         rotate(h,dir(i, :),25)
     end
     if i == 2
-        plot3(r*cos(theta), -r1*ones(size(theta)),r*sin(theta), 'linewidth', 2)
-        plot3(r*cos(theta(end)), -r1,r*sin(theta(end)), '>', 'linewidth', 2, 'markerfacecolor','b')
-        text(-r-0.25, -r1, r+0.25,  '\alpha', 'color','b', 'fontsize', 24)
-        
+        plot3(r*cos(theta), -1.75*ones(size(theta)),r*sin(theta), 'linewidth', 2)
+        plot3(r*cos(theta(1)), -1.75,r*sin(theta(1)), '<', 'linewidth', 2, 'markerfacecolor','b')
+        text(r+0.1, -r1,0.1,  '\alpha', 'color','b', 'fontsize', 24)
+            plot3(v0, 1.1*v1, v0, 'k', 'linewidth', 2)
+    axis equal
     end
     
     if i == 3
[…]

### 2014-05-29 image_analysis_paper/test_PCA.m

@@ -300,8 +300,9 @@ set(gcf, 'paperposition',[0 0 8 8])
 nbins = 25;
 hist(diag(D), nbins)
 set(gca, 'xlim', [-1 1])
+set(gca, 'ylim', [0 45])
 grid on
-set(gca, 'ytick', 0:3:30)
+set(gca, 'ytick', 0:3:45)
 set(gca, 'xtick', -1:0.2:1)
 xlabel('\lambda')
 ylabel('Pr(\lambda)')

### 2014-05-29 image_analysis_paper/test_mutants.m

@@ -307,6 +307,7 @@ set(gcf, 'paperposition',[0 0 8 8])
 nbins = 25;
 hist(diag(D), nbins)
 set(gca, 'xlim', [-1 1])
+set(gca, 'ylim', [0 45])
 grid on
 set(gca, 'ytick', 0:3:45)
 set(gca, 'xtick', -1:0.2:1)

### 2014-05-30 image_analysis_paper/pnas_draft2.tex
added yannis edits to paper

@@
 
+\customlabel{fig:DV_view}{S1}
+\customlabel{fig:membrane_curve}{S2}
+\customlabel{fig:raw_data3}{S3}
+\customlabel{fig:1d_example}{S4}
+\customlabel{fig:spheres}{S5}
+\customlabel{fig:eigenvalues}{S6}
+\customlabel{fig:PCA_data1}{S7}
+\customlabel{fig:PCA_data2}{S8}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \significancetext{Cross-sectional imaging studies are commonly used to study developmental dynamics in multiple organisms. 
+%
 In these studies, a representative developmental trajectory is reconstructed from snapshots of different embryos, each of which has been arrested at a different point of its development. 
+%
 One of the first steps in this reconstruction involves registration and temporal ordering of images. 
-Here we show that both of these tasks can be automated and combined in a single step using recently developed advances in data mining.
+%
+Here we show that both of these tasks can be automated and combined in a single step using recent advances in data mining. 
+%
 To illustrate this approach, we temporally order and register several datasets collected in cross-sectional studies of cell signaling in the early fruit fly embryo.}
@@
 %
-We present an automated approach to registering and temporally ordering data from such cross-sectional studies.
+We present an automated approach to simultaneously register and temporally order data from such cross-sectional studies.
 %
-The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require {\it a priori} knowledge of image features or developmental dynamics.
+The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require {\it a priori} knowledge of image features or a parametric model of the developmental dynamics.
 %
@@
 %
-We demonstrate the utility of these methods using a collection of images from a cross-sectional study of cell signaling during dorsoventral patterning of the {\it Drosophila} embryo.
+We use this method to extract developmental dynamics from collections of images from a cross-sectional study of cell signaling during dorsoventral patterning of the {\it Drosophila} embryo.
 \end{abstract}
@@
 %
-In order to recover the developmental dynamics from such data sets, snapshots of different embryos must be spatially aligned or {\em registered} to factor out the relevant geometric symmetries (e.g., translations and rotations), and then ordered in time.
+In order to recover the developmental dynamics from such data sets, snapshots of different embryos must first be spatially aligned or {\em registered} to factor out the relevant geometric symmetries (e.g., translations and rotations), and then ordered in time.
 %
-We show how existing dimensionality reduction algorithms can automate {\it and combine} both of these tasks.
+We show how recently developed dimensionality reduction algorithms can automate {\it and combine} both of these tasks.
 
@@
 {\xfigtextfontit (B)} Registered and ordered samples.
-For this caricature, the registration and ordering is straightforward because the data set is small and the developmental changes are easy to recognize.}
+For this caricature, the registration and ordering is straightforward because the data set is small and the developmental changes are easy to recognize visually.}
 %\label{fig:fish}
@@
 %
-It is based on a manifold learning algorithm \cite{singer2012vector} which simultaneously addresses the problems of registration and temporal ordering. 
+It is based on a manifold learning algorithm (vector diffusion maps, VDM \cite{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
 %
-This algorithm is one of many nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}.
+This algorithm is one of several influential nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}.
 %
-This technique has been used for data mining in
+The technique has been used for data mining in
 applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules to face recognition, and is sufficiently general to be applicable to a wide variety of biological imaging studies.
@@
 
-\section{Results and Discussion}
+\section{Results}
 
@@
 %
-In general, this requires no template function \cite{ahuja2007template} or image landmarks \cite{ian1998statistical};
-however, when the images are noisy, some pairwise alignments may be inaccurate.
+In general, this requires no template function \cite{ahuja2007template} or image landmarks \cite{ian1998statistical}.
 %
-Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with all pairwise measurements (see {\it SI Materials and Methods}).
+Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see {\it SI Materials and Methods}); this is illustrated in \fig~\ref{subfig:synch2}.
 %
-In this schematic, registration via angular synchronization is trivial as the pairwise measurements contain no noise.
+In this schematic, registration via angular synchronization is trivial, as the pairwise measurements contain no noise.
 %
-However, the algorithm can successfully register data sets even when many of the pairwise measurements are corrupted by noise \cite{singer2011angular}.
+However, the algorithm can successfully register data sets even when many of the pairwise measurements are inaccurate \cite{singer2011angular}.
 %
@@
 %
-Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parameterization of data that lies on a low-dimensional, perhaps nonlinear, manifold in high-dimensional space.
+Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parametrization of data that lies on a low-dimensional, perhaps nonlinear, manifold in high-dimensional space.
 %
@@
 %
-We use {\it local} information about the data to find a parameterization which respects the underlying manifold geometry: we want points which are close in high-dimensional space (e.g., images which look similar) to be close in our parameterization.
+We use {\it local} information about the data to find a parametrization  which respects the underlying manifold geometry: we want points which are close in high-dimensional space (e.g., images which look similar) to be close in our parametrization.
 %
@@
 %
-The color in \fig~\ref{subfig:dmaps2} depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
+The color in \fig~\ref{subfig:dmaps2} depicts the one-dimensional parametrization or ordering of the data that we can detect visually.
 %
-In our working examples, each data point will be much higher dimensional (e.g, an image with many pixels), and so we cannot extract this low-dimensional structure visually.
+In our working examples, each data point will be much higher dimensional (e.g., a pixelated image), and so we cannot extract this low-dimensional structure visually.
 %
-Instead, we will use diffusion maps to automatically uncover a parameterization of our high-dimensional data (see \SI).
+Instead, we will use diffusion maps to automatically uncover a parametrization of our high-dimensional data (see \SI).
 %
-Diffusion maps also generalizes from one-dimensional nonlinear curves to higher-dimensional manifolds.
+Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.
 %
@@
 %
-Angular synchronization uses pairwise alignment information to find translations and rotations which are globally consistent, and diffusion maps uses pairwise distances to find parameterizations which are globally informative.
+Angular synchronization uses pairwise alignment information to find translations and rotations which are globally consistent, and diffusion maps uses pairwise distances to find global parametrizations which are (geometrically) informative.
 %
@@
 %
-%The algorithm uses information about pairwise alignments {\em and} pairwise distances to find a global parameterization of the data, while registering images which are ``close'' .
+%The algorithm uses information about pairwise alignments {\em and} pairwise distances to find a global parametrization of the data, while registering images which are ``close'' .
 %
@@
 {\xfigtextfontit (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
-Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\xfigtextfontit \SI}), so that close data points are connected by dark (``stronger'') edges.
+Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\xfigtextfontit \SI}), so that close data points are connected by darker (``stronger'') edges.
 {\xfigtextfontit (D)} The data in {\xfigtextfontit C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure.
-The color intensity is monotonic with the arclength, thus parameterizing the curve.}
+The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
 %\label{fig:schematics}
@@
 \includegraphics{fig3}
-\caption{{\xfigtextfontit Drosophila} during cellularization.
-{\xfigtextfontit (A)} Images of {\xfigtextfontit Drosophila} embryos collected during cellularization.
-Each image is of a different embryo in a different rotational and translational orientation.
+\caption{{\xfigtextfontit Drosophila} embryos during cellularization.
+{\xfigtextfontit (A)} Images of embryos collected during cellularization.
+Each image is of a different embryo arrested at a different developmental time, and in a different rotational and translational orientation.
 {\xfigtextfontit (B)} Images from {\xfigtextfontit A}, now registered and ordered using vector diffusion maps.
+The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right position.
 The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.}
@@
 %
-During the third hour of development, the spatially uniform two-dimensional arrangement of nuclei is patterned by the concentration profiles of chemical signals that control cell differentiation.
+During the third hour of development, this arrangement of nuclei is patterned by the concentration profiles of chemical signals that control cell differentiation.
 %
@@
 %
-The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo to within 2--3 minutes \cite{figard2013plasma} (see \fig~S2).
+The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo to within 2--3 minutes \cite{figard2013plasma} (see \fig~\ref{fig:membrane_curve}).
 %
@@
 %
-Because the dynamics of the data set shown in \fig~\ref{fig:data1} are relatively simple, effectively consisting of a single mode (the two dpERK peaks) growing in time, established techniques such as principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data in time (see \fig~S7).
+Because the dynamics of the data set shown in \fig~\ref{fig:data1} are relatively simple, effectively consisting of a single mode (the two dpERK peaks) growing in time, established techniques such as principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data in time (see \fig~\ref{fig:PCA_data1}).
 %
-However, developmental dynamics are often significantly more complex, and nonlinear dimensionality reduction techniques are required to extract meaningful structure in the data.
+However, developmental dynamics are often significantly more complex, and nonlinear dimensionality reduction techniques are required to extract meaningful structure.
 %
-
 \fig~\ref{subfig:raw_data2} shows a richer data set of $108$ images of {\it Drosophila} embryos.
@@
 %
-During this window, the ventral furrow is formed, where the ventral side of the embryo buckles towards the center of the embryo, bringing in the future muscle cells.
+During this time window the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, bringing in the future muscle cells.
 %
@@
 %
-During this time window, germband extension causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
+Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
 %
@@
 \includegraphics{fig4}
-\caption{{\xfigtextfontit Drosophila} during gastrulation.
-{\xfigtextfontit (A)} Images of gastrulating {\xfigtextfontit Drosophila} embryos.
-Each image is of a different embryo in a different rotational and translational orientation.
+\caption{{\xfigtextfontit Drosophila} embryos during gastrulation.
+{\xfigtextfontit (A)} Images of gastrulating embryos.
+Each image is of a different embryo arrested at a different developmental time, and in a different translation and rotational orientation.
 {\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps.
-{\xfigtextfontit (C)} Average trajectory of the images shown in {\xfigtextfontit B}.
+The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right position.
+{\xfigtextfontit (C)} A representative ``developmental path'' from local averaging of the images.
 Each image is a (Gaussian-weighted) average of the images in {\xfigtextfontit B}.
@@
 %
-Simple time-ordering based on this linear projection is therefore not possible (see \fig~S8).
+Simple time-ordering based on this linear projection is therefore not possible (see \fig~\ref{fig:PCA_data2}).
 %
@@
 %
-\fig~\ref{subfig:ordered_data2} shows the images in \fig~\ref{subfig:raw_data2}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
-%
-Visual inspection confirms this ordering is consistent with the known developmental dynamics outlined above.
+\fig~\ref{subfig:ordered_data2} shows the images registered and ordered using vector diffusion maps \cite{singer2012vector}, 
+and visual inspection confirms this ordering is consistent with the known developmental dynamics outlined above.
 %
@@
 %
-First, the sheer number of images makes visual processing of the entire data set nontrivial, and
-second, the ordered data is not entirely smooth due to inter-embryo variability.
+First, the sheer number of images makes visual processing of the entire data set nontrivial;
+and second, the ordered data is not entirely smooth due to inter-embryo variability.
 %
-To highlight the developmental dynamics revealed by vector diffusion maps, we applied a Gaussian filter to the trajectory of registered and ordered images to obtain an averaged developmental path.  
+To highlight the developmental dynamics revealed by vector diffusion maps, we applied a Gaussian filter to the trajectory of registered and ordered images to obtain an averaged developmental path (see also \cite{kemelmacher2011exploring}).
 %
-Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:averaged_data2}, serve as a summary of a stereotypic developmental dynamics reconstructed from cross-sectional data.
+Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:averaged_data2}, serve as a summary of the stereotypic developmental dynamics reconstructed from cross-sectional data.
 %
@@
 
-When time is the main source of variability, we have shown vector diffusion maps can recover developmental progression from cross-sectional data.
+When time is the main source of variability, we have shown that vector diffusion maps can recover developmental progression from cross-sectional data.
 %
@@
 %
-To illustrate this point, we collected a data set consisting of $42$ embryos undergoing cellularization during the third hour of development (see \fig~S3).
+To illustrate this point, we collected a data set consisting of $42$ embryos undergoing cellularization during the third hour of development (see \fig~\ref{fig:raw_data3}).
 %
@@
 
-We computed the vector diffusion maps embedding for our data, and the eigenvalue distribution (see \fig~S6) indicates that two coordinates may be required for a thorough description of the data. 
+We computed the vector diffusion maps embedding for our data, and the eigenvalue distribution (see \fig~\ref{fig:eigenvalues}) indicates that two coordinates may be required for a meaningful description of the data.
 %
-\fig~\ref{subfig:projection_data3} shows the data projected onto the first two vector diffusion maps coordinates, where each point in this projection corresponds to one image in our data set.
+\fig~\ref{subfig:projection_data3} shows the data projected onto the first two vector diffusion maps coordinates; each point in this projection corresponds to one image in our data set.
 %%%
@@
 
-\fig~\ref{subfig:projection_colored_data3} shows the same two-dimensional projection as \fig~\ref{subfig:projection_data3}, but the data points are colored to indicate whether they come from the wild type or mutant.
+\begin{figure}[h]
+\includegraphics{fig5_2}
+%\caption{Wild type and mutant {\xfigtextfontit Drosophila} embryos.
+{\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~\ref{fig:raw_data3}), projected onto the first two vector diffusion maps coordinates.
+{\xfigtextfontit (B)} The data in {\xfigtextfontit A}, now colored by genotype.
+The two branches separate mutant and wild type samples, and the direction of the developmental trajectories are indicated by arrows.
+{\xfigtextfontit (C)} dpERK signal at select wild type points in {\xfigtextfontit B} (denoted by circles).
+The two dpERK peaks which arise during development are indicated by arrows.
+{\xfigtextfontit (C)} dpERK signal at select mutant points in {\xfigtextfontit B} (denoted by circles).
+The single dpERK peak is indicated by an arrow.
+}
+\caption{Wild type and mutant {\xfigtextfontit Drosophila} embryos.
+{\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~\ref{fig:raw_data3}), projected onto the first two vector diffusion maps coordinates.
+The direction of the developmental trajectories are indicated by arrows.
+(Inset) The projected data, now colored by genotype.
+The two branches separate mutant and wild type samples.
+{\xfigtextfontit (B)} dpERK signal at select points along the wild type trajectory in {\xfigtextfontit A} (denoted by circles in the inset).
+The two dpERK peaks which arise during development are indicated by arrows.
+{\xfigtextfontit (C)} dpERK signal at select points along the mutant trajectory in {\xfigtextfontit A} (denoted by circles in the inset).
+The single dpERK peak is indicated by an arrow.
+}
+\customlabel{fig:data3}{5}
+\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
+%\customlabel{subfig:projection_colored_data3}{\ref{fig:data3}{\it B}}
+\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it B}}
+\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
+\end{figure}
+
+The inset of \fig~\ref{subfig:projection_data3} shows the two-dimensional projection, with the data points colored to indicate whether they come from the wild type or mutant.
 %
-Clearly, the two branches in the embedding separate wild type and mutant embryos.
+Clearly, the two branches in the embedding help distinguish wild type and mutant embryos.
 %
@@
 %
-Therefore, vector diffusion maps can not only recover dynamics, but also help distinguish different expression patterns and developmental trajectories.
+Therefore, vector diffusion maps can also help distinguish different expression patterns and developmental trajectories.
 
-\begin{figure}[h!]
-\includegraphics{fig5}
-\caption{Mutant and wild type {\xfigtextfontit Drosophila}.
-{\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~S3), projected onto the first two vector diffusion maps coordinates.
-{\xfigtextfontit (B)} The data in {\xfigtextfontit A}, now colored by genotype.
-The two branches separate mutant and wild type samples, and the direction of the developmental trajectories are indicated by arrows.
-{\xfigtextfontit (C)} dpERK signal at select wild type points in {\xfigtextfontit B} (denoted by circles).
-The two dpERK peaks which arise during development are indicated by arrows.
-{\xfigtextfontit (C)} dpERK signal at select mutant points in {\xfigtextfontit B} (denoted by circles).
-The single dpERK peak is indicated by an arrow.
-}
-\customlabel{fig:data3}{5}
-\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
-\customlabel{subfig:projection_colored_data3}{\ref{fig:data3}{\it B}}
-\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it C}}
-\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it D}}
-\end{figure}
-
-\section{Conclusions}
+\section{Discussion}
 
@@
 
-The presented approach was demonstrated on three different experimental data sets, collected specifically for this study. 
+We used the technique to analyze three different experimental data sets, collected specifically for this study. 
 %
@@
 %
-Indeed, because of the relative simplicity of developmental dynamics (monotonic increase in the amplitude of a localized signaling pattern and relatively small tissue deformation), temporal order could have been inferred by projection of registered images on the first principal component of the data set (see \fig~S7).
+Indeed, because of the relative simplicity of developmental dynamics (monotonic increase in the amplitude of a localized signaling pattern and relatively small tissue deformation), temporal order could have been inferred by projection of registered images on the first principal component of the data set (see \fig~\ref{fig:PCA_data1}).
 %
-This approach was clearly not sufficient for the second data set (see \fig~S8), collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.
+This approach was clearly not sufficient for the second data set (see \fig~\ref{fig:PCA_data2}), collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.
 %
-However, vector diffusion maps could readily extract a nontrivial spatiotemporal trajectory. 
+Vector diffusion maps could readily extract a nontrivial spatiotemporal trajectory. 
 %
@@
 %
-Here, we have used vector diffusion maps to register the images with respect to translations and rotations; however, it is conceivable that one could factor out other sources of variability, such as reflections or dilations, in the images.
+Here, we have used vector diffusion maps to register the images with respect to translations and rotations; however, it is conceivable that one could factor out other sources of variability, such as reflections \cite{singer2012vector, goemans1995improved} or dilations, in the images.
+%
+One might even envision automated analysis of embryos imaged in arbitrary three-dimensional orientations, in the spirit of cryo-EM image analysis.
+%
+We believe that algorithms based on image features which are insensitive to small deformations, such as the scattering transform \cite{mallat2012group}, are especially promising.
 %
@@
 Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.
-
-%**OLD***
-%
-%We have demonstrated how vector diffusion maps, an existing reduction technique, can be used to register and temporally order images from developmental biology studies.
-%%
-%We acknowledge that the tasks of both registration and ordering could have been accomplished using a variety of other techniques.
-%%
-%We chose the algorithm outlined here, not because it is specifically tailored to our particular data sets, but rather, because it is sufficiently general and we are confident that it can be applied to a myriad of biological imaging applications.
-%%
-%Furthermore, such data mining algorithms are often tested on databases containing hundreds or thousands of images.
-%%
-%Here, we could successfully recover the underlying dynamics with relatively small data sets.
-%%
-%In general, the number of images required to accurately recover the dynamics will be a function of the noise in the data, as well as the dynamic range and complexity of the underlying process.
-%
-%The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
-%%
-%However, most of these applications rely on the definition and identification of good landmarks or features within each image.
-%%
-%For example, in face recognition, the first step is often to locate the eyes, nose, and mouth in each face, and then register images by aligning these features \cite{zhao2003face}.
-%%
-%%Temporally ordering images of a construction site over time is also based on first identifying features within the images \cite{schindler2007inferring}.
-%%
-%Because of the complexity of chemical and morphological changes in developing tissues, appropriate features may not be known {\it a priori}.
-%%
-%We therefore use an algorithm which requires no features, but instead, uses only the raw pixel images and the inherent geometric symmetry of the problem.
-%%
-%%Template-based methods are also common for registration, where one optimally aligns each image to a predefined template function.
-%%
-%%However, these methods can suffer when the images are noisy, and the success of such methods can depend strongly on the template function \cite{shatsky2009method}.
-%%
-%
-%The task of temporal ordering has also been studied in a variety of biological applications.
-%%
-%Data sets such as transcriptional profiling data (where each data point is the transcriptome for a single sample, such as a cell or tissue sample from a patient) \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering} and snapshots of cells throughout the cell cycle (where each point is a vector of features, such as the amount of DNA, which quantify the cell's state)  \cite{kafri2013dynamics} have been temporally ordered to extract representative trajectories of the underlying dynamic processes.
-%%
-%Many of these applications order the data by solving a traveling salesman problem (TSP) or constructing a minimum spanning tree (MST) on the data,
-%expecting that these structures (the path of the traveling salesman, or the minimum spanning tree) characterize the majority of the patterns and variability within the data and provide an accurate encoding of the dynamical progression.
-%%
-%These algorithms often use dimensionality reduction techniques (typically PCA or variants of) as a preprocessing step.
-%%
-%The TSP or MST is solved/constructed not in the original data space, but rather in the lower-dimensional space of the first few principal components.
-%%
-%In these approaches, PCA serves two purposes: it denoises the data, making the TSP/MST construction more robust, and also allows for easy visualization of the TSP/MST if one only uses the first two or three principal components.
-%
-%We, instead, use vector diffusion maps to construct a one-dimensional parameterization of the data.
-%%
-%Because vector diffusion maps is a dimensionality reduction technique, it serves as a denoising step and yields smooth averaged trajectories even under noisy conditions.
-%%
-%However, because the algorithm is nonlinear, it affords us many of the same geometric flexibilies as the MST or TSP formulations.
-%%
-%Vector diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003, tenenbaum2000global, Donoho2003, Roweis2000}.
-%%
-%Diffusion maps have been shown to be more robust to noise than isomap \cite{balasubramanian2002isomap}, a path-based algorithm, and so we suspect the methods presented here may perform better than the ordering algorithms used in previous work.
-%
-%Furthermore, dimensionality reduction techniques such as vector diffusion maps do not constrain us to one-dimensional parameterizations/organizations of the data, as illustrated in the mutant data set example (\fig~\ref{fig:data3}).
-%%
-%These techniques would also allow us to compare the developmental dynamics of several different mutations to see how mutations are developmentally related.
-%%
-%Other sources of variability, such as size or viewing angle, could also be uncovered using vector diffusion maps.
-%
-%After registering and temporally ordering images, one can perform a variety of post-processing tasks to gain further insight into the developmental processes.
-%%
-%We have demonstrated one such technique (Gaussian filtering) that allows us to extract a more representative developmental trajectory from the set of noisy snapshots.
-%%
-%%Constructing a smooth movie from the registered and ordered cross-sectional snapshots would also be of biological interest, and
-%Examining the top principal components in different time windows would allow one to visualize the main dynamical features along the trajectory.
-%%
-%%These questions first require that the data be registered and temporally ordered.
-%%
-%Correlating vector diffusion maps coordinates with physical variables is another post-processing task that can give physical insight into developmental dynamics.
-%%
-%One can test which physical variable, among a set of candidate physical variables (e.g., total intensity of one signal, roundness of the embryo), is most correlated with the uncovered vector diffusion maps coordinate to help understand which physical quantities change across the developmental trajectory.
-%%
-%We are confident that, as biological becomes easier and more robust, the dimensionality reduction techniques we present here will play an important role in the data analysis pipeline.
-
-%\begin{itemize}
-%\item What are other people doing  (maybe critically)
-%\item Movies  smoothness
-%\item Discrete symmetries like flips
-%\item Local information ? (features of the image) (associate with Fourier-Bessel and bispectra)
-%\item And in general, MULTIPLE variablities is important (mutants, guys in another cycle, etc.)
-%
-%\item size is an issue
-%\item  small tilts is an issue
-%\item   could one do three-d image processing ?  of the floating thing, without the device
-%\item how do we go about merging data sets -- microscopes setting / more generally -- different observables?
-%\end{itemize}
 
@@
 End-on imaging was performed by using the microfluidics device described previously.
-Images were collected at the focal plane $\sim$90~$\mathrm{\mu m}$ from the posterior pole of an embryo.
+Images were collected at the focal plane $\sim$90~$\mathrm{\mu m}$ from the posterior pole of an embryo (see \fig~\ref{fig:DV_view}).
 
@@
 The progression of membrane invagination during cellularization was used to assign time point to each embryo in the third hour of development.
-Time-lapse imaging of Sqh-GFP embryo was used to generate correlation curve between the membrane ingression length and the developmental time (n=7) (\fig S2B).
+Time-lapse imaging of Sqh-GFP embryo was used to generate correlation curve between the membrane ingression length and the developmental time (n=7) (\fig~\ref{fig:membrane_curve}B).
 The rate at which membrane invaginates was highly reproducible from embryo to embryo.
-In the experiments with fixed embryos, we measured the membrane length by co-staining embryos with a membrane marker (Sqh-GFP) or by taking phase-contrast images (\fig S2A).
+In the experiments with fixed embryos, we measured the membrane length by co-staining embryos with a membrane marker (Sqh-GFP) or by taking phase-contrast images (\fig~\ref{fig:membrane_curve}A).
 We used the correlation curve to assign each embryo to a time window of 2--3~minutes.
@@
 \begin{acknowledgments}
-- text of acknowledgments here, including grant info -
+The authors thank Adam Finkelstein,  Thomas Funkhouser, and John Storey for helpful discussions. 
+%
+C.J.D. was supported by the Department of Energy Computational Science Graduate Fellowship (CSGF), grant number DE-FG02-97ER25308.
+%
+B.L. and S.Y.S. were supported by the National Institutes of Health Grant R01GM086537. 
+%
+T.J.L. and H.L. were supported by the National Science Foundation Grant Emerging Frontiers in Research and Innovation (EFRI) 1136913.
+%
+I.G.K. was supported by the National Science Foundation (CS\&E program).
 \end{acknowledgments}
@@
 
+\bibitem{kemelmacher2011exploring}
+Kemelmacher-Shlizerman I, Shechtman E, Garg R, Seit, S~M (2011)
+Exploring photobios. \textit{ACM Transactions on Graphics (TOG)} 30(4):61.
+
 \bibitem{yuan2014automated}
@@
 
+\bibitem{goemans1995improved}
+Goemans M~X, Williamson D~P (1995) Improved approximation algorithms for maximum cut and satisfiability problems using semidefinite programming.
+\textit{Journal of the ACM} 42(6):1115--1145.
+
+\bibitem{mallat2012group}
+Mallat, S (2012) Group invariant scattering. 
+\textit{Communications on Pure and Applied Mathematics}
+65(10):1331--1398.
+
 \bibitem{krzic2012multiview}

### 2014-05-30 image_analysis_paper/si_information.tex
added yannis edits to paper

@@
 % 
-The same laser setting is maintained within each experiment.
+The same laser setting was maintained within each experiment.
 
@@
 %
-The eigenvalues corresponding to noise in the data form a semicircle distribution centered at 0 \cite{wigner1955characteristic, wigner1958distribution}. 
+The eigenvalues corresponding to noise in the data form a semicircle distribution centered at the origin \cite{wigner1955characteristic, wigner1958distribution}. 
 %
@@
 Note that components 2--10 have a nonnegligible contribution to the data.
-{\it (C)} Projection onto the first two principal components.}
+{\it (C)} Projection onto the first two principal components.
+Each principal component is normalized to 1.
+}
 \label{fig:PCA_results}
@@
 
-\raisebox{2.2cm}{\sffamily{C}}
-\hspace{-1cm}
-\includegraphics[width=4cm, angle=90]{FBsPCA_dmaps}
+\raisebox{5.7cm}{\sffamily{C}}
+\includegraphics[width=16cm, trim=2cm 0cm 2cm 0cm, clip]{FBsPCA_dmaps}
 \caption{I computed the FBsPCA basis for the set of images (separately for the red and green signals).
@@
 %
-The results are shown below.
+The results are shown above.
 %
@@
 (A) Images used for FBsPCA.
-(B) Eigenvalue spectrum from dmaps, using bispectrum coefficients.
+(B) Eigenvalue spectrum from diffusion maps, computed using bispectrum coefficients.
 There is no spectral gap after the first component, as would be expected for this particular data set.
-(C) First 4 diffusion maps variables, computed from the bispectrum coefficeints, plotted against the membrane thickness.
-Note there is no appreciable correlation: each dmaps variable simply selects one image from the data set as ``abnormal''.}
+(C, top) First 4 diffusion maps variables, computed from the bispectrum coefficients, plotted against the membrane thickness.
+Note there is no appreciable correlation: each diffusion maps variable simply selects one image from the data set as ``abnormal''.
+(C, bottom) The corresponding images that each diffusion maps mode ``selects.'' In general, these seem to be the late images with ventral furrows beginning to form.
+}
 \end{figure*}

### 2014-05-30 image_analysis_paper/make_sphere.m

@@ -91,10 +91,10 @@ for i=1:4
     h = surf(X, Y, Z, C, 'edgecolor','k');
     set(gca,'position',[0 0 1 1],'units','normalized')
     
-    camlight(45, -20);
+    camlight(-10, 30);
     lighting phong
     %material dull
-    set(h, 'diffusestrength', 1.0);
+    set(h, 'diffusestrength', 3.0);
     set(h, 'specularstrength', 1);
     %set(gca, 'ambientlightcolor', [1 0 0]);
     set(h, 'AmbientStrength', 0.7);

### 2014-05-30 image_analysis_paper/test_FBsPCA.m

@@ -34,6 +34,8 @@ subplot_dim2 = ceil(m / subplot_dim1);
 
 data = zeros(npixels, npixels, m);
 data2 = zeros(npixels, npixels, m);
+image_set = zeros(npixels, npixels, 3, m, 'uint8');
+
 
 %% load images
 
@@ -49,6 +51,12 @@ for i=1:m
     %store image
     data(:, :, i) = im1(:, :, 2);
     data2(:, :, i) = im1(:, :, 3);
+    
+    im1 = circshift(im1, [0 0 -1]);
+    im1(:, :, 1) = im1(:, :, 1) + im1(:, :, 3);
+    im1(:, :, 2) = im1(:, :, 2) + im1(:, :, 3);
+    
+    image_set(:, :, :, i) = im1;
     
 end
 
@@ -167,16 +175,22 @@ saveas(gcf,sprintf('%s/FBsPCA_evals', im_save_dir), 'pdf')
 figure;
 imshow(W_dmaps(ranks_from_membranes,ranks_from_membranes))
 
+ncomp = 6;
 figure;
 set(gcf, 'paperunits', 'centimeters')
-set(gcf, 'papersize', [16 4])
-set(gcf, 'paperposition',[0 0 16 4]);
-for i=2:5
-    subplot(1,4, i-1)
+set(gcf, 'papersize', [24 24/ncomp*2])
+set(gcf, 'paperposition',[0 0 24 24/ncomp*2]);
+for i=2:ncomp+1
+    subplot(2, ncomp, i-1)
     plot(V(:,i),mem_lengths,'.')
     xlabel(sprintf('\\phi_%d',i));
-    ylabel('membrane length')
+    ylabel('mem. length')
     axis([-1.1 1.1 -inf inf])
+    
+    subplot(2, ncomp, ncomp+(i-1))
+    [~, idx] = max(abs(V(:,i)));
+    imshow(image_set(:, :, :, idx), 'border', 'tight')
+    
 end
 saveas(gcf,sprintf('%s/FBsPCA_dmaps', im_save_dir), 'pdf')
 

### 2014-06-02 image_analysis_paper/pnas_draft2.tex
Added yannis edits before sending to Amit

@@
 
-%\usepackage{subcaption}
 \graphicspath{ {paper_figures2/} }
@@
 \usepackage{tikz}
-
-%\usepackage{sansmath}
-
-%\usepackage[font={sf, small}]{caption}
 
@@
 %\def\s{\sigma}
-
-%\DeclareMathSizes{9}{8}{7}{7}
-
-%\DeclareMathOperator*{\argmin}{arg\,min}
 
@@
 We use this method to extract developmental dynamics from collections of images from a cross-sectional study of cell signaling during dorsoventral patterning of the {\it Drosophila} embryo.
+%
+We show that the algorithm recovers the correct orientation and temporal order when they are independently known, reconstructs an average developmental trajectory when such information is unavailable, and can detect distinct trajectories in a heterogeneous population of embryos.
+
 \end{abstract}
@@
 %
-In a cross-sectional study, developmental dynamics is reconstructed from multiple embryos, each of which contributes only a snapshot of a chemical or morphological process along its developmental trajectory \cite{jaeger2004dynamic, fowlkes2008quantitative}.
+In a cross-sectional study, developmental dynamics is reconstructed from multiple embryos, each of which contributes only a snapshot of a chemical or morphological process along its developmental trajectory \cite{jaeger2004dynamic, peter2012predictive, peter2011gene, fowlkes2008quantitative}.
 %
@@
 %
-These images cover a thirty minute time interval and span late cellularization through gastrulation, when the two-dimensional sheet of newly formed cells begins to deform into three-dimensional structures which will form the future organs \cite{leptin2005gastrulation}.
+These images cover a thirty minute time interval and span late cellularization through gastrulation, when the arrangement of newly formed cells begins to deform into structures which will form the future organs \cite{leptin2005gastrulation}.
 %
@@
 
-We computed the vector diffusion maps embedding for our data, and the eigenvalue distribution (see \fig~\ref{fig:eigenvalues}) indicates that two coordinates may be required for a meaningful description of the data.
-%
-\fig~\ref{subfig:projection_data3} shows the data projected onto the first two vector diffusion maps coordinates; each point in this projection corresponds to one image in our data set.
-%%%
-%%%YGK we need something about dmap eignvalues and how many are not harmonics
-%
-Two distinct branches appear to emerge in this two-dimensional embedding space.
-
 \begin{figure}[h]
@@
 \end{figure}
+
+We computed the vector diffusion maps embedding for our data, and the eigenvalue distribution (see \fig~\ref{fig:eigenvalues}) indicates that two coordinates may be required for a meaningful description of the data.
+%
+\fig~\ref{subfig:projection_data3} shows the data projected onto the first two vector diffusion maps coordinates; each point in this projection corresponds to one image in our data set.
+%%%
+%%%YGK we need something about dmap eignvalues and how many are not harmonics
+%
+Two distinct branches appear to emerge in this two-dimensional embedding space.
 
@@
 %
-Here, we have used vector diffusion maps to register the images with respect to translations and rotations; however, it is conceivable that one could factor out other sources of variability, such as reflections \cite{singer2012vector, goemans1995improved} or dilations, in the images.
+Here, we have used vector diffusion maps to register the images with respect to translations and rotations; however, it is conceivable that one could factor out other sources of variability, such as reflections \cite{singer2012vector, goemans1995improved, bandeira2013cheeger} or dilations, in the images.
 %
@@
   embryo. \textit{Nature} 430(6997):368--371.
+
+\bibitem{peter2012predictive}
+Peter IS, Faure E, Davidson EH (2012) Predictive computation of genomic logic processing functions in embryonic development. \textit{Proc Natl Acad Sci USA} 109(41):16434--16442.
+
+\bibitem{peter2011gene}
+Peter IS, Davidson EH (2011) A gene regulatory network controlling the embryonic
+specification of endoderm. \textit{Nature} 474(7353):635--639.
 
@@
 
+\bibitem{bandeira2013cheeger}
+Bandeira A~S, Singer A, Spielman D~A (2013) A Cheeger inequality for the graph connection Laplacian. \textit{SIAM Journal on Matrix Analysis and Applications} 34(4):1611--1630.
+
 \bibitem{mallat2012group}

### 2014-06-02 image_analysis_paper/si_information.tex
Added yannis edits before sending to Amit

@@
 
-Each image is subsampled to $100 \times 100$ pixels.
+The original resolution of each image is $1024 \times 1024$ pixels.
 %
-For the first and third data sets in the paper (images taken during cellularization), we only use the red and green channels (dpERK and Dl) in the vector diffusion maps computations, as the spatial location of the nuclei does not change appreciably during this time window.
+The images are subsampled to $100 \times 100$ pixels for vector diffusion maps analysis.
 %
-The nuclei in the corresponding figures are shown only for visual reference.
+A scale bar for a representative image is shown in \fig~\ref{fig:ap_dv}.
 %
-For the second data set of images during gastrulation, we use all three channels (dpERK, Twi, and nuclei) in the vector diffusion maps computations.
+When imaging, the laser settings are calibrated so that images have good dynamic range, while not saturating the detector.
 %
-We blur the nuclear signal using a disc filter with a $5$~pixel radius (using the \texttt{imfilter} function in Matlab with a disc filter).
+The same laser settings are maintained within each experiment.
 %
-We also adjust the intensity of the nuclear signal in each image such that 1\% of the pixels are saturated (using the \texttt{imadjust} function in Matlab).
+The intensity of each channel is integer-valued, and ranges from $0$ (no signal) to $255$ (saturated). 
+%
+We blur the nuclear signal using a disc filter with a $5$~pixel radius (using the \texttt{imfilter} function in MATLAB\textsuperscript{\textregistered}).
+%
+We also adjust the intensity of the nuclear signal in each image such that 1\% of the pixels are saturated 
+(using the \texttt{imadjust} function in MATLAB), and then scale the intensity of the nuclear signal by half.
 
-When imaging, the laser settings were calibrated so that images have good dynamic range, while not saturating the detector. 
+For the first and third data sets in the paper (images taken during cellularization), we only use the red and green channels (dpERK and Dl) in the vector diffusion maps computations, as the spatial location of the nuclei (the gray channel) does not change appreciably during this time window.
 %
-The same laser setting was maintained within each experiment.
+The nuclei in the corresponding figures are thus shown only for visual reference.
+%
+For the second data set of images (taken during gastrulation), we use all three channels (dpERK, Twi, and nuclei) in the vector diffusion maps computations.
+%
+For this particular data set, we multiply the intensity of the Twi signal by 1.5, as this channel was not completely saturated in the original imaging data.
 
-\subsection{Mathematical Algorithms}
+\subsection{Algorithms}
 
@@
 %
-The relatively simple dynamics of this data set will allow us to easily visualize and illustrate the main points of the different algorithms.
+The relatively simple dynamics of this data set will allow us to easily visualize and illustrate the main features of the different algorithms.
 %
-We construct concentration profiles defined on a ring, and we assume that each ring is arbitrarily rotated around its center; an example is shown in \fig~\ref{subfig:1d_example}.
+Motivated by the geometry of our images, we construct a sequence of concentration profiles defined on a ring, and rotate each ring randomly around its center; an example is shown in \fig~\ref{subfig:1d_example}.
 %
@@
 %
-Each concentration profile is a noisy Gaussian (shown in \fig~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
+Each concentration profile is a noisy Gaussian (shown in \fig~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as a function of ``time".
 %
@@
 %
-For the one-dimensional discretized profiles shown in \fig~\ref{fig:1d_demo}, we exhaustively search over all $n=100$ possible shifts of the signals to compute the optimal angles in \eqref{eq:opt_angle}.
+For the one-dimensional discretized profiles shown in \fig~\ref{fig:1d_demo}, we exhaustively search over all $n=100$ possible shifts of the signals to obtain the optimal angles in \eqref{eq:opt_angle}.
 %
@@
 %
-Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$.
+Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$ ordered so that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$.
 %
@@
 %
-We adjust the sign of $\phi_1$ so that $det(R_{i, est}) = +1$, ensuring proper rotations.
+We adjust the sign of $\phi_1$ so that $det(R_{i, est}) = +1$, ensuring proper rotations 
+(note that systematically incorporating improper rotations is also possible \cite{goemans1995improved, bandeira2013cheeger}).
 %
@@
 
-Furthermore, this formulation also considers {\it higher-order} consistency information.
+Importantly, this formulation also considers {\it higher-order} consistency information.
 %
@@
 \end{equation}
-therefore, {\it all} information of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$.
+therefore, {\it all} information of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$ (and higher order
+consistency information in its higher powers).
 %
@@
 
-Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional space), we want to find a coordinate transformation $y(x)$ that preserves local information: points that are close in the original space should also be close in the coordinate $y$.
+Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional vector space), we want to find a coordinate transformation $y(x)$ that preserves local geometry: points that are ``close" in the original space should also be ``close" in the coordinates $y$.
 %
@@
 %
-$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/10^{th}$ of the median of the pairwise distances between data points for the first two data sets, and half of the median of the pairwise distances between data points for the third data set.
+$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/10^{th}$ of the median of the pairwise distances between data points for the first two data sets (and half of the median of the pairwise distances between data points for the third data set).
 
-To find this coordinate $y$, we want solve the following optimization problem \cite{Belkin2003}
+To find the coordinate $y$, we want solve the following optimization problem \cite{Belkin2003}
 \begin{equation} \label{eq:dmaps_opt_problem}
@@
 %
-In our application, we have assumed that this single direction of variability, parameterized by $\phi_2$, is one-to-one with time.
+In our application, we have assumed that this {\em single} direction of variability, parameterized by $\phi_2$, is one-to-one with time.
 %
@@
 %
-The procedure generalizes when the data lie on higher-dimensional manifolds (not just curves) in data space, where lower eigenvectors give subsequent embedding coordinates for the data.
+The procedure generalizes when the data lie on higher-dimensional manifolds (not just curves) in data space, where leading eigenvectors can give subsequent embedding coordinates for the data.
 %
@@
 %
-In the language of symmetry groups, this distance is a metric between the orbits induced by the symmetry group.
+In the language of symmetry groups, this distance is a metric between the orbits induced by the relevant symmetry group.
 
@@
 %
-If we assume that the rotations and the dynamics are uncoupled and therefore separable, then the eigenvectors of $S$ have the following structure: each block eigenvector contains estimates of the optimal rotations (up to a constant rotation) multiplied by the corresponding embedding coordinate.
+If we assume that the rotations and the dynamics are uncoupled and therefore separable, then the eigenvectors of $S$ have the following structure: each block eigenvector contains estimates of the optimal rotations (up to a constant rotation) multiplied by the corresponding embedding coordinate (a scalar)
 %
-As the first diffusion maps coordinate is constant, the first block eigenvector contains only the optimal rotations.
+As the first diffusion maps coordinate is constant over the data, the first block eigenvector contains only the optimal rotations.
 %
@@
 We can therefore recover this diffusion maps coordinate by taking inner products of the columns of the second block eigenvector with columns of the first block eigenvector.
-%
+%%
+%%YGK just as a favor to me, tell me how consistent the three values are -- 
+%%
 The $j^{th}$ embedding coordinate will be given by $\psi_{k,1}$, where $jd +1 < k \le (j+1)d$,
@@
 %
-Practically, we have square images discretized as pixels (rather than continuous functions on the plane).
+Practically, we have square images discretized as pixels (rather than continuous functions on the plane); note that the center of the
+image square is {\em not} necessarily also the center of the embryo.
 %
@@
 %
-The translations are implemented by shifting the pixels;
+The vertical and horizontal translations are implemented by shifting the pixels;
 we only consider shifts that correspond to an integer number of pixels to remove the need for interpolation when doing the translations.
@@
 
-The synchronization formulation requires that the matrices used to represent the symmetry group be orthogonal \cite{singer2013spectral}.  
+The synchronization formulation requires the matrices used to represent the symmetry group to be orthogonal \cite{singer2013spectral}.
 %(for \eqref{eq:H_low_rank} to be satisfied).
@@
 %
-Rotation of the image corresponds to rotation around one principal axis in three dimensions, while translation of the image corresponds (approximately) to rotations around the other two principal axes.
+Rotation of the image corresponds to rotation around one principal axis in three dimensions, while translation of the image corresponds (approximately) to rotations around the other two principal axes (see \fig~\ref{fig:SO3_picture}).
 (Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$.
-However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.)
+However, the images we will be considering are already ``mostly centered", and so we are only interested in {\em small} translations that are well within the $[0, 2\pi)$ range.)
 %
-These rotation matrices are (by definition) orthogonal, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
+These rotation matrices are (by definition) orthogonal, and successive application of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
 
-To convert the optimal two-dimensional translation and rotation to corresponding (continuous) three-dimensional rotations,
+To convert the optimal two-dimensional translations and rotation to corresponding (continuous) three-dimensional rotations,
 we first compute the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$,
@@
 
-These rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$ are used in formulating the matrix $H$ in \eqref{eq:H_to_R} for angular synchronization, or the matrix $S$ in \eqref{eq:vdm_S} for vector diffusion maps to compute the rotations (and, in the case of vector diffusion maps, temporal orderings) for the images.
+These rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$ are used in formulating the matrix $H$ in \eqref{eq:H_to_R} for angular synchronization, or the matrix $S$ in \eqref{eq:vdm_S} for vector diffusion maps to compute the optimal rotations (and, in the case of vector diffusion maps, deduce optimal temporal orderings) for the images.
 %
-From the three-dimensional rotations $R_{1, est}, \dots, R_{m, est}$, we  require the corresponding translations and rotations of the images.
+The three-dimensional optimal rotations $R_{1, est}, \dots, R_{m, est}$ then give us the corresponding translations and rotations of the images.
 %
-We first multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
+We multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
 %
@@
 %
-From a rotation matrix $R$, we compute the Euler angles $\alpha$, $\beta$, and $\gamma$ using the following relationships
+From a rotation matrix $R$, we compute the corresponding Euler angles $\alpha$, $\beta$, and $\gamma$ using the following relationships
 \begin{equation} \label{eq:angle_invert}
@@
 
-\subsection{Determining dimensionality from eigenvalue spectrum}
+\subsection{The eigenvalue spectrum}
 
-We can use the distribution of the eigenvalues from (vector) diffusion maps to help determine the dimensionality of the data.
+We can use the distribution of the eigenvalues from (vector) diffusion maps to help deduce the dimensionality of the data.
 %
@@
 %
-The eigenvalues corresponding to noise in the data form a semicircle distribution centered at the origin \cite{wigner1955characteristic, wigner1958distribution}. 
+The eigenvalues corresponding to noise in the data form a semicircle distribution centered at the origin (when the noise is a symmetric random matrix whose entries are drawn from a Gaussian distribution) \cite{wigner1955characteristic, wigner1958distribution}.
 %
-The eigenvalues corresponding to descriptive/meaningful embedding coordinates will be positive and removed from the semicircle. 
+We expect the eigenvalues corresponding to descriptive/meaningful embedding coordinates to be positive and ``visibly removed" from the semicircle.
 %
-Therefore, by counting how many eigenvalues are removed from the semicircle (excluding the trivial $\lambda=1$ eigenvalue), we can obtain an estimate of the dimensionality of the data.
+Counting such eigenvalues which are removed from the semicircle (excluding the trivial $\lambda=1$ eigenvalue) can help obtain an estimate of the dimensionality of the data (modulo harmonics in the same dimension \cite{ferguson2010systematic}).
 
-For vector diffusion maps, the same structure holds: we expect trivial eigenvalues, a semicircle corresponding to noise, and positive eigenvalues corresponding to meaningful embedding coordinates.
+For vector diffusion maps, the same structure holds: we expect trivial eigenvalues, a semicircle corresponding to noise, and positive eigenvalues removed from the noise that correspond to meaningful embedding coordinates.
 %
@@
 %
-Therefore, by counting how many sets of $d$ eigenvalues are removed from the semicircle (excluding the trivial set of $d$ eigenvalues, all with  $\lambda=1$), we can obtain an estimate of the dimensionality of the data.
+We must then establish how many sets of $d$ eigenvalues are removed from the noise semicircle (excluding the trivial set of $d$ eigenvalues, all with  $\lambda=1$), to explore the dimensionality of the data.
 
-%\subsection{Algorithms}
-%
-%\subsubsection{Angular synchronization}
-%
-%\begin{enumerate}
-%\item Compute the pairwise rotations (described in \eqref{eq:opt_angle}) or rotations + translations (described in \eqref{eq:opt_pairwise}) needed to register every pair of data points.
-%\item Construct the corresponding rotation matrices $R_{ij}$ for every pair $i, j$ of images (see \eqref{eq:R_theta} for rotations or \eqref{eq:total_rot} for rotations + translations).
-%\item Construct the matrix $H$ as in \eqref{eq:H_to_R}, which is a block matrix with entries $R_{ij}$.
-%\item Compute the top $2$ (for rotations) or $3$ (for rotations + translations) eigenvectors of $H$.
-%\item Recover the estimates $\hat{R}_1, \dots, \hat{R}_m$ from the top 2 or 3 eigenvectors as in \eqref{eq:R_hat}.
-%\item Project these estimates onto the closest proper rotation matrix using \eqref{eq:R_est}
-%\item Recover the optimal rotation angle $\theta_i$ using \eqref{eq:R_theta} or rotation + translation using \eqref{eq:angle_invert} for each image
-%\end{enumerate}
-%
-%\subsubsection{Diffusion Maps}
-%
-%\begin{enumerate}
-%\item Construct the matrix $W$ as in \eqref{eq:dmaps_W}
-%\item Construct the matrix $A$ as in \eqref{eq:dmaps_A}
-%\item Compute the top few eigenvectors of $A$, ordered with decreasing eigenvalues
-%\item If we only want to order the data, then sort the data by the entries in the second eigenvector.
-%\item For higher-dimensional embeddings (e.g., the mutants), use the entries of successive eigenvectors (the second eigenvector, the third eigenvector, etc.) as the embedding coordinates
-%\end{enumerate}
-%
-%\subsubsection{Vector Diffusion Maps}
-%
-%\begin{enumerate}
-%\item Compute the pairwise rotations (described in \eqref{eq:opt_angle}) or rotations + translations (described in \eqref{eq:opt_pairwise}) needed to register every pair of data points.
-%%
-%\item Construct the corresponding rotation matrices $R_{ij}$ for every pair $i, j$ of images (see \eqref{eq:R_theta} for rotations or \eqref{eq:total_rot} for rotations + translations).
-%%
-%\item Construct the matrix $H$ as in \eqref{eq:H_low_rank}.
-%%
-%\item Construct the matrix $W$ as in \eqref{eq:dmaps_W}, using the minimum distance obtained for each pair in step 1 as the pairwise distances in the Gaussian kernel.
-%%
-%\item Construct the matrix $A$ as in \eqref{eq:dmaps_A}.
-%%
-%\item Construct the matrix $S$ as in \eqref{eq:vdm_S}.
-%%
-%\item Compute the top few eigenvectors of $S$.
-%%
-%\item Recover the estimates $\hat{R}_1, \dots, \hat{R}_m$ from the top 2 or 3 eigenvectors as in \eqref{eq:R_hat}.
-%%
-%\item Project these estimates onto the closest proper rotation matrix using \eqref{eq:R_est}.
-%%
-%\item Recover the optimal rotation angle $\theta_i$ using \eqref{eq:R_theta} or rotation + translation using \eqref{eq:angle_invert} for each image
-%%
-%\item To compute the first (non-trivial) embedding coordinate for the data, compute $\psi_{k,l}$ as in \eqref{eq:vdm_coord}, taking $k=1, 2$ and $l=3, 4$ for the rotation case, or $k = 1, 2, 3$ and $l = 4, 5, 6$ for the rotation + translation case.
-%%
-%Select the $(k, l)$ coordinate which has the largest variance as the embedding coordinate.
-%%
-%\item To compute more coordinates, use the subsequent blocks for $l$ (e.g., for the next coordinate, under translations and rotations, use $k=1, 2, 3$ and $l = 7, 8, 9$).
-Again, compute $\psi_{k, l}$ as in \eqref{eq:vdm_coord}, and select the $(k, l)$ coordinate with largest variance.
-%
-%\end{enumerate}
+\subsection{Smooth trajectories from registered and ordered images}
 
-\subsection{Constructing smooth trajectories from registered and ordered images}
-
-Once we have registered and ordered the images, we smooth the resulting trajectory to obtain a picture of the stereotypic developmental dynamics. 
+Once we have registered and ordered the images, we can smooth the resulting trajectory to obtain a ``stereotypic" developmental trajectory.
 %
@@
 For the images in \fig~4, we take $\sigma^2 = 20$.
+%
+See \cite{kemelmacher2011exploring} for a more detailed discussion.
 
@@
 The embryo is presented so that the anterior (A) side is to the left and the posterior (P) side is to the right.
-The arrow indicates the position where the cross-section of an embryo is imaged.
-(Right) A dorsoventral view of a {\it Drosophila} embryo.
+The arrow indicates the position(s) where the cross-section of an embryo is imaged.
+(Right) A dorsoventral view of the cross-section of the {\it Drosophila} embryo.
 The dorsal (D) side is up and the ventral (V) side is down.
-Images were collected at the focal plane ~18\% from either the anterior or posterior pole of an embryo (arrow in the left image).
+For the first and third data sets (cellularization), images were collected at the focal plane $\sim 18\%$ from either the anterior or posterior pole of an embryo (arrows in the left image).
+During cellularization, it is known that the developmental dynamics are identical at these two regions \cite{...}.
+In the second data set (gastrulation), images were collected at the focal plane $\sim 18\%$ from the posterior pole of an embryo (right arrow in the left image).
 }
+\label{fig:ap_dv}
 \end{figure}
@@
 Each image is of a different embryo in a different rotational orientation.
-Half of the images are of wild type embryos, and half of the images are of mutant embryos.}
+Half of the images are of wild type embryos, and half are of mutant embryos.}
 \end{figure}
@@
 \includegraphics[width=8.4cm]{SI_fig3}
-\caption{Synthetic data set used to illustrate the mathematical algorithms.
+\caption{Synthetic data set used to illustrate the data processing algorithms.
 {\it (A)} One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
@@
 \end{minipage}
-\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions.
+\caption{Illustration of the correspondence between three-dimensional rotations to translations and rotations in two dimensions.
 {\it (A)} The original image (shown in red), projected onto a portion of a sphere.
@@
 (A) Histogram of eigenvalues from images during cellularization (\fig~3).
-As expected, we see an approximately symmetric distribution that ranges from $-0.8$ to $0.8$ (indicated by the blue arrow), and $6$ eigenvalues that are outside of this distribution (indicated by the red arrow).
-$3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $3$ eigenvalues correspond to the first VDM component which uncovers the developmental dynamics, and the then there is an appreciable gap before the third group of eigenvalues.
+As expected, we see an approximately symmetric distribution that ranges from $-0.8$ to $0.8$ (indicated by the blue arrow), and $6$ eigenvalues that are removed from this distribution (indicated by the red arrow).
+$3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $3$ eigenvalues correspond to the first nontrivial VDM component which ``uncovers" the developmental dynamics; then there is an appreciable gap before the third group of eigenvalues.
 %
 (B) Histogram of eigenvalues from images during gastrulation (\fig~4).
-We see an approximately symmetric distribution that ranges from $-0.7$ to $0.7$ (indicated by the blue arrow), and $6$ eigenvalues that are outside of this distribution (indicated by the red arrow).
-$3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $3$ eigenvalues correspond to the first VDM component which uncovers the developmental dynamics.
+We see an approximately symmetric distribution that ranges from $-0.7$ to $0.7$ (indicated by the blue arrow), and $6$ eigenvalues that are removed from this distribution (indicated by the red arrow).
+$3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $3$ eigenvalues correspond to the first nontrivial VDM component which ``uncovers" the developmental dynamics.
 %
 (C) Histogram of eigenvalues from images of mutant and wild type embryos during cellularization (\fig~5).
-We see an approximately symmetric distribution that ranges from $-0.2$ to $0.2$ (indicated by the blue arrow), and $9$ eigenvalues that are outside of this distribution (indicated by the red arrows).
-$3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $2$ eigenvalues correspond to the first two VDM components which uncovers the developmental dynamics of the two genotypes.}
+We see an approximately symmetric distribution that ranges from $-0.2$ to $0.2$ (indicated by the blue arrow), and $9$ eigenvalues that are removed from this distribution (indicated by the red arrows).
+$3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $6$ eigenvalues correspond to the first two VDM components which are now necessary to parametrize the developmental dynamics of the two genotypes.}
 %there is a spectral gap after the first (non-trivial) group of eigenvalues.
@@
 \raisebox{3.7cm}{\sffamily{B}}
-\includegraphics[width=8.4cm]{data2_PCA_variance}
+\includegraphics[width=8.4cm, angle=90]{data2_PCA_variance}
 \raisebox{3.7cm}{\sffamily{C}}
@@
 Note that components 2--10 have a nonnegligible contribution to the data.
-{\it (C)} Projection onto the first two principal components.
+{\it (C)} Projection of the data onto the first two principal components.
 Each principal component is normalized to 1.
+Both the projection coefficients and the projection coefficients scaled by the square roots of the corresponding eigenvalues are shown.
 }
@@
 
+%
 \begin{figure*}[t]
+{\bf Item C in the figure below is our main observation.}
+
 \raisebox{7.6cm}{\sffamily{A}}
@@
 \raisebox{7.6cm}{\sffamily{B}}
-\includegraphics[width=7cm]{FBsPCA_evals}
+\includegraphics[width=8cm]{FBsPCA_evals}
 
@@
 %
-However, because the bispectrum is not stable to deformations, it does not seem to be able to identify neighboring images.
+However, (mainly, we suspect) because the bispectrum is not stable to deformations, it does not seem to be able to identify neighboring images.
 %
@@
 (B) Eigenvalue spectrum from diffusion maps, computed using bispectrum coefficients.
-There is no spectral gap after the first component, as would be expected for this particular data set.
+There is no visible spectral gap after the first component, as would be expected for this particular data set.
 (C, top) First 4 diffusion maps variables, computed from the bispectrum coefficients, plotted against the membrane thickness.
-Note there is no appreciable correlation: each diffusion maps variable simply selects one image from the data set as ``abnormal''.
-(C, bottom) The corresponding images that each diffusion maps mode ``selects.'' In general, these seem to be the late images with ventral furrows beginning to form.
+Note there is no appreciable correlation: {\bf each diffusion maps variable simply selects one image from the data set as ``abnormal''.} (C, bottom) The corresponding images that each diffusion maps mode ``selects.'' In general, these seem to be the late images with ventral furrows beginning to form.
 }

### 2014-06-02 image_analysis_paper/test_late_images.m

@@ -243,20 +243,39 @@ end
 
 figure;
 set(gcf, 'paperunits', 'centimeters')
-set(gcf, 'papersize', [8 4])
-set(gcf, 'paperposition',[0 0 8 4])
+set(gcf, 'papersize', [12 12])
+set(gcf, 'paperposition',[0 0 12 12])
 bar(diag(D_PCA(1:10, 1:10)) / sum(diag(D_PCA)))
 xlabel('k')
-ylabel('\lambda_k')
+ylabel('fraction of variance captured by k^{th} PC')
 % saveas(gcf,sprintf('%s/data2_PCA_variance', im_save_dir), 'pdf')
 
 figure;
 set(gcf, 'paperunits', 'centimeters')
-set(gcf, 'papersize', [8 4])
-set(gcf, 'paperposition',[0 0 8 4])
-plot(proj_coeff(:,1),proj_coeff(:,2),'.')
-xlabel('PCA projection 1')
-ylabel('PCA projection 2')
+set(gcf, 'papersize', [12 12])
+set(gcf, 'paperposition',[0 0 12 12])
+line(proj_coeff(:,1), proj_coeff(:,2),'marker', '.','linestyle', 'none')
+xlabel('$< x, PC1 >$', 'interpreter','latex')
+ylabel('$< x, PC2 >$', 'interpreter','latex')
+xlim = get(gca, 'xlim');
+ylim = get(gca, 'ylim');
+haxes1 = gca;
+haxes1_pos = get(haxes1,'Position'); % store position of first axes
+haxes2 = axes('Position',haxes1_pos,...
+              'XAxisLocation','top',...
+              'YAxisLocation','right', 'color','none');
+%line(sqrt(D_PCA(1,1))*proj_coeff(:,1), sqrt(D_PCA(2,2))*proj_coeff(:,2),'Parent',haxes2 ,'marker', '.','color','b','linestyle', 'none')
+set(gca, 'xlim', sqrt(D_PCA(1,1))*xlim);
+set(gca, 'ylim', sqrt(D_PCA(2,2))*ylim);
+set(gca,'xticklabel',num2str(get(gca,'xtick')','%g'))
+set(gca,'yticklabel',num2str(get(gca,'ytick')','%g'))
+xlabel('$\sqrt{\lambda_1} < x, PC1 >$', 'interpreter','latex')
[…]

### 2014-06-05 image_analysis_paper/paper_figures2/fig5_2.tex
Added edits to paper
Added comments from Amit

--- 2014-05-29 image_analysis_paper/paper_figures2/fig5.tex
@@
 %\usepackage[margin=0in, paperwidth=8.7cm, paperheight=9.5cm]{geometry}
-\usepackage[margin=0in, paperwidth=8.7cm, paperheight=17cm]{geometry}
+\usepackage[margin=0in, paperwidth=8.7cm, paperheight=11.7cm]{geometry}
 
@@
 \begin{tikzpicture}
-	\node at (0cm, 0cm) {\includegraphics[width=6cm, angle=90]{data3_embed_bw}};
-	\node at (-4cm, 2.5cm) {\textsf{A}};
+	\node at (0cm, 0cm) {\includegraphics[width=6.9cm, angle=90]{data3_embed_bw}};
+	\node at (-4.35cm, 2.75cm) {\textsf{A}};
+	%\draw[help lines,step=.5] (-4,-4) grid (4,4);
+	\draw [draw=white!50!gray, ->, ultra thick] (-2.2, -2) .. controls (-1.1, 0.5) and  (1, 0)  .. (2, -2);
+	\draw [draw=white!50!gray, ->, ultra thick] (-3, -1.8)  .. controls (-2.6, 0.9) and (-1.9, 1.5) .. (-1.7, 2.1); 
+	\node at (2.05cm, 1.6cm) {\includegraphics[width=4.1cm]{data3_embed_color}};
 \end{tikzpicture}\\
 \begin{tikzpicture}
-	\node at (0cm, 0cm) {\includegraphics[width=8cm]{data3_embed_color}};
-	\node at (-4cm, 2.5cm) {\textsf{B}};
-	%\draw[help lines,step=.5] (-4,-4) grid (4,4);
-	\draw [draw=black!20!green, ->, ultra thick] (-2, -2) .. controls (-1.1, 0.2) and  (1, 0)  .. (2, -2);
-	\draw [blue, ->, ultra thick] (-2.65, -1.8)  .. controls (-2.3, 0.9) and (-1.7, 1.5) .. (-1.4, 2.1); 
-\end{tikzpicture}\\
-\begin{tikzpicture}
-	\node at (-4.25cm, 0cm) {\textsf{C}};
+	\node at (-4.25cm, 0cm) {\textsf{B}};
 	\node at (0,0) {\includegraphics[width=8cm]{wt_trajectory}};
@@
 \begin{tikzpicture}
-	\node at (-4.25cm, 0cm) {\textsf{D}};
+	\node at (-4.25cm, 0cm) {\textsf{C}};
 	\node at (0,0) {\includegraphics[width=8cm]{mut_trajectory}};

### 2014-06-05 image_analysis_paper/pnas_draft2.tex
Added comments from Amit

@@
 The technique has been used for data mining in
-applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules to face recognition, and is sufficiently general to be applicable to a wide variety of biological imaging studies.
+applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules to face recognition, and is sufficiently general to be applicable to a wide variety of biological imaging studies {\color{red} does this sentence need references?}.
 %
-Here the algorithm is adapted for simultaneous registration and temporal ordering of images from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
+Here the algorithm is adapted for the analysis of images from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
 %
@@
 %
-Vector diffusion maps combine two algorithms, {\em angular synchronization} \cite{singer2011angular} for image registration and {\em diffusion maps} \cite{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation that allows us to simultaneously register and temporally order our images.
+Vector diffusion maps combine two algorithms, {\em angular synchronization} \cite{singer2011angular} for image registration and {\em diffusion maps} \cite{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
 %
@@
 
-As a first illustration of using vector diffusion maps to temporally order and register images, we analyze a data set where the correct registration and temporal order is known {\it a priori}.
+As a first illustration, we analyze a data set where the correct registration and temporal order is known {\it a priori}.
 %
@@
 %
-Thus, vector diffusion maps can accomplish the image registration and ordering tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental dynamics (e.g., correlation of age with body size).
+Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental dynamics (e.g., correlation of age with body size).
 
@@
 We expect two sources of variability in this data set: differences due to developmental time, and differences between mutant and wild type dynamics.
+
+We computed the vector diffusion maps embedding for our data, and the eigenvalue distribution (see \fig~\ref{fig:eigenvalues}) indicates that two coordinates may be required for a meaningful description of the data.
+%
+\fig~\ref{subfig:projection_data3} shows the data projected onto the first two vector diffusion maps coordinates; each point in this projection corresponds to one image in our data set.
+%%%
+%%%YGK we need something about dmap eignvalues and how many are not harmonics
+%
+Two distinct branches appear to emerge in this two-dimensional embedding space.
 
@@
 \end{figure}
-
-We computed the vector diffusion maps embedding for our data, and the eigenvalue distribution (see \fig~\ref{fig:eigenvalues}) indicates that two coordinates may be required for a meaningful description of the data.
-%
-\fig~\ref{subfig:projection_data3} shows the data projected onto the first two vector diffusion maps coordinates; each point in this projection corresponds to one image in our data set.
-%%%
-%%%YGK we need something about dmap eignvalues and how many are not harmonics
-%
-Two distinct branches appear to emerge in this two-dimensional embedding space.
 
@@
 In contrast to most of the existing approaches to image registration which rely on the knowledge about some landmarks in the data \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization (such as vector diffusion maps) can register images even in the absence of such information.
+
 %TODO: look at Eisen Cell paper about registration
@@
 
-Previously, angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy imaging data \cite{singer2012vector, singer2011three}. 
+{\color{red} Previously, angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy imaging data \cite{singer2012vector, zhao2014rotationally, singer2011viewing}.}
 %
@@
 %
-We believe that algorithms based on image features which are insensitive to small deformations, such as the scattering transform \cite{mallat2012group}, are especially promising.
+{\color{red} In these situations, vector diffusion maps is essential, as global registration is topologically impossible \cite{zhao2014rotationally}.}
+%
+{\color{red} We believe that algorithms based on rotation- and translation-invariant image features which are stable to small deformations, such as the scattering transform \cite{mallat2012group} or steerable PCA and the bispectrum \cite{zhao2014rotationally}, are especially promising; such methods circumvent the need for registration by using these invariant features in the data analysis.}
 %
@@
 T.J.L. and H.L. were supported by the National Science Foundation Grant Emerging Frontiers in Research and Innovation (EFRI) 1136913.
+%
+A.S. was supported by the Air Force Office of Scientific Research Grant
+FA9550-12-1-0317.
 %
@@
 
-\bibitem{singer2011three}
-Singer A, Shkolnisky Y (2011) Three-dimensional structure determination from
-  common lines in cryo-em by eigenvectors and semidefinite programming.
-  \textit{SIAM J Imaging Sci} 4(2):543--572.
+\bibitem{zhao2014rotationally}
+Zhao Z, Singer A (2014) Rotationally invariant image representation for
+viewing direction classification in cryo-EM. \textit{J Struct Biol} 186(1):153--166.
+
+\bibitem{singer2011viewing}
+Singer A, Zhao Z, Shkolnisky Y, Hadani R (2011) Viewing angle
+classification of cryo-electron microscopy images using eigenvectors. \textit{SIAM J Imaging Sci} 4(2):723--759.
+
+%\bibitem{singer2011three}
+%Singer A, Shkolnisky Y (2011) Three-dimensional structure determination from
+%  common lines in cryo-em by eigenvectors and semidefinite programming.
+%  \textit{SIAM J Imaging Sci} 4(2):543--572.
 

### 2014-06-06 image_analysis_paper/pnas_draft2.tex
added more amit edits to paper

@@
 %
-%Furthermore, these techniques can separate multiple trajectories within the same data set.
-%
 We use this method to extract developmental dynamics from collections of images from a cross-sectional study of cell signaling during dorsoventral patterning of the {\it Drosophila} embryo.
@@
 %
-%Recent advances in physical manipulation and imaging of embryos produce rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different developmental time point and at a different geometric orientation.
-%
 Importantly, the ``age'' of any given embryo arrested in its development is not quantitatively known; typically what is known is
 a certain time window to which a collection of embryos belongs \cite{ng2012large, richardson2014emage, castro2009automatic}.
-%to high accuracy.
 %
@@
 
-%%%YGK put an arrow in the disc
-
 Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, 
@@
 %
-The technique has been used for data mining in
-applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules to face recognition, and is sufficiently general to be applicable to a wide variety of biological imaging studies {\color{red} does this sentence need references?}.
+These techniques has been used for data mining in
+applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \cite{zhao2014rotationally, singer2011viewing} to face recognition \cite{lafon2006data}, and are sufficiently general to be applicable to a wide variety of biological imaging studies.
 %
-Here the algorithm is adapted for the analysis of images from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
+Here, the vector diffusion maps algorithm is adapted for the analysis of images from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
 %
 To illustrate our approach, we analyze three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics \cite{jaeger2012drosophila}.
-%
-%The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate this approach.
-%
-%In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex.
-%
-%Here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
-%
-%The third set consists of data collected from both wild type and mutant embryos, where joint processing of the data will help to visualize the different sources of variability along distinct developmental paths.
-
-%\begin{itemize}
-%\item Imaging data is becoming increasingly prevalent in biology research.
-%
-%\item General motivation about biology data ...
-%
-%\item Need methods to organize data, visualize, ...
-%
-%\item REDUCTION is important, reduction because of symmetry and reduction because of content
-%
-%\item There are a bunch of recent techniques- some established (like dmaps), some very fresh, like synchronization or VDMaps,  that have had their own reasons to be developed but they are as if tailor made for what we would want to do
-%\end{itemize}
 
@@
 
-%%%YGK  when I read this I think of Eiffel tower pictures, not of ours - there is a transition... careful where we do it.
-%
-%Our goal is to organize data sets from cross-sectional studies of developmental processes in a meaningful and informative way.
-%
-%We will show that dimensionality reduction algorithms can simultaneously register and temporally order the images to give us a meaningful picture of the underlying developmental dynamics.
-%
 Our approach is based on vector diffusion maps \cite{singer2012vector}, a manifold learning
@@
 %
-%Vector diffusion maps has been used to classify images from cryo-electron microscopy (cryo-EM) experiments, where each image is a projection of a randomly oriented molecule.
-%
-%%%YGK to imply which rotations we need to have discussed it above
-%%
 We will use the algorithm to register images of {\it Drosophila} embryos with respect to rotations and translations, as well as uncover the main direction of variability {\it after} removing translational and rotational symmetries.
@@
 
-%%%somewhere I want a sentence that says this is for general symmetry groups of some kind, and HERE it is for..... (maybe in SI)
 Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
@@
 After removing variability due to translations and rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
-%%%
-%%%YGK feels like repeats
-%%%
 %
@@
 Thus, when our experiments have additional sources of variability (such as distinct populations of embryos) in addition to developmental dynamics, diffusion maps can provide an informative organization of our data which highlights both of these variability sources.
-%We will assume that our data are approximately one-dimensional, and that this dimension is parameterized by time, such that ordering our data along this main dimension/direction will temporally order our data.
 
@@
 Vector diffusion maps combines both steps into a single computation (see \SI), simultaneously addressing the tasks of registration and temporal ordering.
-%
-%The algorithm uses information about pairwise alignments {\em and} pairwise distances to find a global parametrization of the data, while registering images which are ``close'' .
-%
-%It is thus perfectly suited for our task of registering and temporally ordering cross-sectional imaging data.
 %
@@
 Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, red in the figure), an enzyme that specifies a subset of neuronal cells.
-%
-
-%%%YGK say a simple thing in the caption about Dorsal-Ventral
 
@@
 
-%Although the first vector diffusion maps coordinate accurately captures the developmental dynamics, there are other sources of variability within this data set that could also be extracted using these techniques.
-%
-%An obvious one is variability due to size; here, the size variation is relatively small relative to the developmental changes and does not significantly obstruct the reconstruction of the temporal order.
-%
-%However, in other applications where the size variation is significant %or the developmental dynamics are more subtle, it may be important to %factor out size or {\it scaling} symmetries.
-%%%
-%%%YGK   for teh "additional components" - don't have in the same sentence - say it somewhere else.... INCLUDE IT (maybe in the mutants)
-
-%Furthermore, the dynamics uncovered by vector diffusion maps is relatively simple, which suggests that a simpler approach could have been used to order the data.
-%
-%Indeed, a principal component analysis (PCA) of registered data reveals that data can be accurately ordered by projecting each image onto the first PCA mode.
-
 \subsection{Images of gastrulating embryos}
 
-%As the images in Figure~\ref{fig:data1} are relatively simple, other techniques could have been used to temporally order them.
-%
 Because the dynamics of the data set shown in \fig~\ref{fig:data1} are relatively simple, effectively consisting of a single mode (the two dpERK peaks) growing in time, established techniques such as principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data in time (see \fig~\ref{fig:PCA_data1}).
@@
 %
-%Even when projecting on several leading principal components (see SI), a nonlinear approach, such as vector diffusion maps, is required to temporally order the data.
-%
 \fig~\ref{subfig:ordered_data2} shows the images registered and ordered using vector diffusion maps \cite{singer2012vector}, 
@@
 \includegraphics{fig5_2}
-%\caption{Wild type and mutant {\xfigtextfontit Drosophila} embryos.
-{\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~\ref{fig:raw_data3}), projected onto the first two vector diffusion maps coordinates.
-{\xfigtextfontit (B)} The data in {\xfigtextfontit A}, now colored by genotype.
-The two branches separate mutant and wild type samples, and the direction of the developmental trajectories are indicated by arrows.
-{\xfigtextfontit (C)} dpERK signal at select wild type points in {\xfigtextfontit B} (denoted by circles).
-The two dpERK peaks which arise during development are indicated by arrows.
-{\xfigtextfontit (C)} dpERK signal at select mutant points in {\xfigtextfontit B} (denoted by circles).
-The single dpERK peak is indicated by an arrow.
-}
 \caption{Wild type and mutant {\xfigtextfontit Drosophila} embryos.
@@
 \customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
-%\customlabel{subfig:projection_colored_data3}{\ref{fig:data3}{\it B}}
 \customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it B}}
@@
 
-%TODO: look at Eisen Cell paper about registration
-
 We used the technique to analyze three different experimental data sets, collected specifically for this study. 
@@
 %
-{\color{red} We believe that algorithms based on rotation- and translation-invariant image features which are stable to small deformations, such as the scattering transform \cite{mallat2012group} or steerable PCA and the bispectrum \cite{zhao2014rotationally}, are especially promising; such methods circumvent the need for registration by using these invariant features in the data analysis.}
+{\color{red} In addition, we believe that algorithms based on rotation- and translation-invariant image features which are stable to small deformations, such as the scattering transform \cite{mallat2012group} or the bispectrum and steerable PCA \cite{zhao2014rotationally}, are especially promising; such methods circumvent the need for registration by using invariant features of the images in the data analysis.}
 %
@@
 
+\bibitem{zhao2014rotationally}
+Zhao Z, Singer A (2014) Rotationally invariant image representation for
+viewing direction classification in cryo-EM. \textit{J Struct Biol} 186(1):153--166.
+
+\bibitem{singer2011viewing}
+Singer A, Zhao Z, Shkolnisky Y, Hadani R (2011) Viewing angle
+classification of cryo-electron microscopy images using eigenvectors. \textit{SIAM J Imaging Sci} 4(2):723--759.
+
+\bibitem{lafon2006data}
+Lafon S, Keller Y, Coifman R~R (2006) 
+Data fusion and multicue data matching by diffusion maps. \textit{IEEE Trans Pattern Anal Mach Intell}
+28(11):1784--1797.
+
 \bibitem{jaeger2012drosophila}
@@
 
-\bibitem{zhao2014rotationally}
-Zhao Z, Singer A (2014) Rotationally invariant image representation for
-viewing direction classification in cryo-EM. \textit{J Struct Biol} 186(1):153--166.
-
-\bibitem{singer2011viewing}
-Singer A, Zhao Z, Shkolnisky Y, Hadani R (2011) Viewing angle
-classification of cryo-electron microscopy images using eigenvectors. \textit{SIAM J Imaging Sci} 4(2):723--759.
-
 %\bibitem{singer2011three}

### 2014-06-06 image_analysis_paper/pnas_draft2.tex
added more amit edits to paper

@@
 %
-A large number of imaging studies are based on a cross-sectional experimental design, where developmental progress is arrested in a population of embryos, each of which is at a different point along its developmental trajectory.
+Many experiments are based on a cross-sectional design, where developmental progress is arrested in a population of embryos, each of which is at a different point along its developmental trajectory. 
 %
@@
 %
-We use this method to extract developmental dynamics from collections of images from a cross-sectional study of cell signaling during dorsoventral patterning of the {\it Drosophila} embryo.
+We use this method to extract developmental trajectories from a cross-sectional imaging study of cell signaling during dorsoventral patterning of the {\it Drosophila} embryo.
 %
-We show that the algorithm recovers the correct orientation and temporal order when they are independently known, reconstructs an average developmental trajectory when such information is unavailable, and can detect distinct trajectories in a heterogeneous population of embryos.
+We show that the algorithm recovers the correct orientation and temporal order in a data set where they are independently known, reconstructs an average developmental trajectory when such information is unavailable, and can detect distinct trajectories in a heterogeneous population of embryos.
 
@@
 %
-In a typical cross-sectional study, a group of embryos is fixed using a procedure that arrests their development and stained with chemicals that visualize a handful of cellular processes.
+In a typical cross-sectional study, a group of embryos is fixed using a procedure that arrests their development, and stained with chemicals that visualize a handful of cellular processes.
 %
@@
 %
-Image registration is based on obvious morphological landmarks, such as the position of head and fins.
+Image registration is based on obvious morphological landmarks, such as the positions of the head and the fins.
 %
@@
 %
-In general, the landmarks needed for registration (here, the heads and fins) as well as the attributes which can be used to order the data (such as the body length) are not known {\it a priori}.
+In general, the landmarks needed for registration, as well as the attributes which can be used to order the data, are not known {\it a priori}.
 %
@@
 {\xfigtextfontit (B)} Registered and ordered samples.
-For this caricature, the registration and ordering is straightforward because the data set is small and the developmental changes are easy to recognize visually.}
+For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
 %\label{fig:fish}
@@
 Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, 
-and most registration algorithms are based on landmarks, such as the head and fins in \fig~\ref{fig:fish}.
+most of which require the definition and identification of appropriate landmarks.
 %
@@
 %
-These techniques has been used for data mining in
+These techniques has been used in
 applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \cite{zhao2014rotationally, singer2011viewing} to face recognition \cite{lafon2006data}, and are sufficiently general to be applicable to a wide variety of biological imaging studies.
@@
 
-Our approach is based on vector diffusion maps \cite{singer2012vector}, a manifold learning
+Vector diffusion maps \cite{singer2012vector} is a manifold learning
 technique developed for data sets which contain two types of sources of variability:
@@
 %
-Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see {\it SI Materials and Methods}); this is illustrated in \fig~\ref{subfig:synch2}.
+Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see \SI); this is illustrated in \fig~\ref{subfig:synch2}.
 %
@@
 %
-Thus, when our experiments have additional sources of variability (such as distinct populations of embryos) in addition to developmental dynamics, diffusion maps can provide an informative organization of our data which highlights both of these variability sources.
+Thus, when our experiments have additional sources of variability (such as distinct populations of embryos) in addition to developmental dynamics, the algorithm can provide an informative organization of our data which highlights both of these variability sources.
 
@@
 {\xfigtextfontit (B)} Images from {\xfigtextfontit A}, now registered and ordered using vector diffusion maps.
-The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right position.
+The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right.
 The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.}
@@
 %
-Both of these features become readily apparent when images are registered and ordered using vector diffusion maps (see \fig~\ref{subfig:ordered_data1}): the peak of nuclear Dl is consistent throughout the images and the intensity of the dpERK signal in the lateral regions of the embryo increases monotonically across the data set.
+Both of these features become readily apparent when images are registered and ordered using vector diffusion maps (see \fig~\ref{subfig:ordered_data1}).
+%
+The peak of nuclear Dl is consistent throughout the images and the intensity of the dpERK signal in the lateral regions of the embryo increases monotonically across the data set.
 %
 
-For this data set, the quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the ordering obtained through an independent approach.
+For this data, the quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the ordering obtained through an independent approach.
 %
@@
 %
-These images cover a thirty minute time interval and span late cellularization through gastrulation, when the arrangement of newly formed cells begins to deform into structures which will form the future organs \cite{leptin2005gastrulation}.
+These images cover a thirty minute time interval spanning late cellularization through gastrulation, when the arrangement of newly formed cells begins to deform into structures which will form the future organs \cite{leptin2005gastrulation}.
 %
@@
 %
-However, during this developmental time period, the nuclei are no longer uniformly arranged around the periphery of the embryo.
+However, during this developmental time period, the nuclei no longer remain uniformly arranged around the periphery of the embryo.
 %
@@
 {\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps.
-The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right position.
+The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right.
 {\xfigtextfontit (C)} A representative ``developmental path'' from local averaging of the images.
@@
 %
-\fig~\ref{subfig:projection_data3} shows the data projected onto the first two vector diffusion maps coordinates; each point in this projection corresponds to one image in our data set.
+\fig~\ref{subfig:projection_data3} shows the data projected onto the first two vector diffusion maps coordinates, where each point in this projection corresponds to one image in our data set.
 %%%
@@
 {\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~\ref{fig:raw_data3}), projected onto the first two vector diffusion maps coordinates.
-The direction of the developmental trajectories are indicated by arrows.
+The directions of the developmental trajectories are indicated by arrows.
 (Inset) The projected data, now colored by genotype.
@@
 
-The inset of \fig~\ref{subfig:projection_data3} shows the two-dimensional projection, with the data points colored to indicate whether they come from the wild type or mutant.
+The inset of \fig~\ref{subfig:projection_data3} shows the two-dimensional projection, with the data points now colored to indicate whether they correspond to wild type or mutant embryos.
 %
-Clearly, the two branches in the embedding help distinguish wild type and mutant embryos.
+Clearly, the two branches in the embedding help distinguish the two genotypes.
 %
@@
 %
-The portion of the embedding where the two genotypes are intermixed (lower-right region of the plot) contains embryos who are early in their developmental trajectory, where there is little dpERK signaling and the two groups of embryos are indistinguishable.
+The portion of the embedding where the two genotypes are intermixed (lower-left region of the plot) contains embryos which are early in their developmental trajectory, where there is little dpERK signaling and the two groups of embryos are indistinguishable.
 %
@@
 %
-Automated temporal ordering of such data sets was accomplished by first projecting the data on a low-dimensional subspace spanned by the leading principal or independent components, and then solving a traveling salesman problem or constructing a minimum spanning tree on the projected data to order multiple snapshots.
+Automated temporal ordering of such data sets was accomplished by first projecting the data onto a low-dimensional subspace spanned by the leading principal or independent components, and then solving a traveling salesman problem or constructing a minimum spanning tree on the projected data to order multiple snapshots.
 %
@@
 %
-In contrast to most of the existing approaches to image registration which rely on the knowledge about some landmarks in the data \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization (such as vector diffusion maps) can register images even in the absence of such information.
+In contrast to most of the existing approaches to image registration which rely on the knowledge about some landmarks in the data \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information.
 
-We used the technique to analyze three different experimental data sets, collected specifically for this study. 
+We used vector diffusion maps to analyze three different experimental data sets, collected specifically for this study. 
 %
@@
 %
-For this particular data set, a simpler approach would have sufficed. 
-%
-Indeed, because of the relative simplicity of developmental dynamics (monotonic increase in the amplitude of a localized signaling pattern and relatively small tissue deformation), temporal order could have been inferred by projection of registered images on the first principal component of the data set (see \fig~\ref{fig:PCA_data1}).
+For this particular data set, a simpler approach would have sufficed;
+because of the relative simplicity of developmental dynamics (monotonic increase in the amplitude of a localized signaling pattern and relatively small tissue deformation), temporal order could have been inferred by projection of registered images on the first principal component of the data set (see \fig~\ref{fig:PCA_data1}).
 %
@@
 %
-In this case, low-dimensional projection of registered data revealed distinct paths that correlated with the developmental trajectories of the mutant and wild type populations.
+In this case, low-dimensional projection of the data revealed distinct paths which correlated with the developmental trajectories of the mutant and wild type populations.
 
@@
 %
-Here, we have used vector diffusion maps to register the images with respect to translations and rotations; however, it is conceivable that one could factor out other sources of variability, such as reflections \cite{singer2012vector, goemans1995improved, bandeira2013cheeger} or dilations, in the images.
+Here, we have used vector diffusion maps to register the images with respect to translations and rotations; however, it is conceivable that one could factor out other sources of variability, such as reflections \cite{singer2012vector, goemans1995improved, bandeira2013cheeger} or dilations.
 %
@@
 %
-{\color{red} In these situations, vector diffusion maps is essential, as global registration is topologically impossible \cite{zhao2014rotationally}.}
+{\color{red} For these applications, vector diffusion maps is essential, as global registration of the entire imaging data set is topologically impossible \cite{zhao2014rotationally}.}
 %

### 2014-06-06 image_analysis_paper/ASPIRE/extern/SDPLR-1.03-beta/doc/SDPLR-1.03-beta-usrguide.tex
Added everything before working on submission

\documentclass[12pt]{article}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}

%included packages (some not so standard, let me know if you don't want to use them)
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}

\usepackage[dvipdf,colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}
%\usepackage[pdftex,colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}

%\usepackage[authoryear,round]{natbib}

\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{url}
\usepackage{graphics}
\usepackage{verbatim}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{example}[theorem]{Example}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{assumption}[theorem]{Assumption}
\def\proof{\noindent{\bf Proof.} \ignorespaces}
\def\endproof{{\ \hfill\hbox{%
      \vrule width1.0ex height1.0ex
    }\parfillskip 0pt}\par}

\marginparwidth 0pt \oddsidemargin  0pt \evensidemargin  0pt
\marginparsep 0pt

\setlength{\textwidth}{6.5in} \setlength{\textheight}{9in}
\setlength{\topmargin}{-.45in}
\setlength{\evensidemargin}{-0.14in}
\setlength{\oddsidemargin}{-0.14in}
%\renewcommand{\baselinestretch}{1.2}

\newcommand{\vgap}{\vspace*{.1in}}
\newcommand{\bu}{\bullet}
\newcommand{\A}{{\cal A}}
\newcommand{\K}{{\cal K}}
\newcommand{\Sn}{{\cal S}^n}
\newcommand{\Snp}{{\cal S}^n_+}
\newcommand{\Snpp}{{\cal S}^n_{++}}
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\beqa}{\begin{eqnarray}}
\newcommand{\eeqa}{\end{eqnarray}}
\newcommand{\beqas}{\begin{eqnarray*}}
\newcommand{\eeqas}{\end{eqnarray*}}
\newcommand{\rank}{{\rm rank}\,}
\newcommand{\Image}{{\rm Im}\,}
\def\eqnok#1{(\ref{#1})}
\newcommand{\mdot}{\bullet}
\newcommand{\bSn}{{\cal S}^{n_1 \cdots n_\ell}}
\newcommand{\bSnp}{{\cal S}^{n_1 \cdots n_\ell}_+}
\newcommand{\tr}{\operatorname{trace}}
\newcommand{\eqnref}[1]{(\ref{#1})}
\newcommand{\thmref}[1]{Theorem \ref{#1}}
\newcommand{\sectref}[1]{Section \ref{#1}}
\newcommand{\propref}[1]{Proposition \ref{#1}}
\newcommand{\blockRspace}{\Re^{(n_1 \times r_1)\cdots(n_\ell \times
r_\ell)}}
\renewcommand{\L}{{\cal L}}
\newcommand{\blas}{{\sc blas}}
\newcommand{\blass}{{\sc blas }}
\newcommand{\lapack}{{\sc lapack}}
\newcommand{\lapacks}{{\sc lapack }}
\newcommand{\arpack}{{\sc arpack}}
\newcommand{\arpacks}{{\sc arpack }}
\newcommand{\metis}{{\sc metis}}
\newcommand{\metiss}{{\sc metis }}

\begin{document}

\title{SDPLR 1.03-beta User's Guide (short version)}
%\author{
%Samuel Burer%
%\thanks{Department of Management Sciences, University of Iowa,
%Iowa City, IA 52242-1000, USA. (Email: samuel-burer@uiowa.edu).
%This research was supported in part by NSF Grant CCR-0203426.}
%\and Renato D.C.\ Monteiro%
%\thanks{Department of Industrial and Systems Engineering, Georgia Institute
%of Technology, Atlanta, GA 30332, USA. (Email:
%monteiro@isye.gatech.edu).} \and Changhui Choi%
%\thanks{Department of Management Sciences, University of Iowa,
%Iowa City, IA 52242-1000, USA. (Email: changhui-choi@uiowa.edu).
%This research was supported in part by NSF Grant CCR-0203426.}}

\date{\today}

\maketitle

\section{Summary}

SDPLR is a C software package for solving large-scale semidefinite
programming problems. Source code, binaries, and a Matlab
interface for SDPLR can be downloaded from
\begin{center}
\url{http://dollar.biz.uiowa.edu/~sburer/software/SDPLR/}.
\end{center}
SDPLR accepts the standard sparse SDPA format as well as a special
format that handles low-rank data matrices. People involved with
the development of SDPLR are
\begin{itemize}

\item Samuel Burer (samuel-burer@uiowa.edu)

\item Renato D.C.\ Monteiro (monteiro@isye.gatech.edu)

\item Changhui Choi (changhui.choi@ucdenver.edu)

\end{itemize}

% \section{Binaries and Installation from Source}
\section{Installation from Source}

% Binaries for different platforms are available from the SDPLR
% website. Users' contribution of binaries would be especially
% appreciated. In the event that no binary is available for a
% particular platform, the source code is also available for
% installation from scratch.

It is assumed that the user has downloaded and unpacked the SDPLR
source files.

\subsection{Matlab}

Within Matlab, run {\sl mexinstall.m} from the main SDPLR directory.
Then type {\sl help sdplr} for instructions. Add the SDPLR directory to
your Matlab path.

{\bf Note to Windows users.} The top two lines of {\sl mexinstall.m} may
need to be adjusted for the location of the Matlab's \lapack~ and \blas~
libraries on your system.

\subsection{Command line (UNIX, Linux, etc.)}

\subsubsection*{Step 1}

Make sure \blass and \lapacks are installed on your system and
determine how to link them with other programs. The linking
information will be required for the {\tt LIB} and {\tt LIB\_DIRS}
options in {\sl Makefile.inc\/} (see Step 2).

% Many systems have
% optimized \blass and \lapacks libraries.

% On Linux, Cygwin, and
% Mingw systems, optimized libraries for common architectures can be
% downloaded from
% \begin{center}
% %\url{http://www.scipy.org/download/atlasbinaries/}.
% \url{http://old.scipy.org/download/atlasbinaries/}.
% \end{center}

% \subsubsection{Step 2 (optional)}

% If you would like the ability to use some of SDPLR's advanced
% features, including sparse eigenvalue computations and
% preconditioning for the truncated Newton's method, download and
% compile
% \begin{center}
% \arpack \ \ \ \url{http://www.caam.rice.edu/software/ARPACK/}
% \end{center}
% and
% \begin{center}
% \metis 4.0.1 \ \ \
% \url{http://www-users.cs.umn.edu/~karypis/metis/metis/index.html}.
% \end{center}
% Both compilations produce static libraries (.a), which can be
% copied into the {\sl lib\/} directory of the SDPLR package.
% Whether \arpacks and/or \metiss are available will be required for
% the {\tt DEFINES} option in {\sl Makefile.inc\/} (see Step 3).

\subsubsection*{Step 2}

Tailor the file {\sl Makefile.inc\/} to your system; {\sl
Makefile.inc.linux\/} is provided as an example. There are three
sections:

\begin{itemize}

\item Specify the location and switch for the \lapack~library
using {\tt LAPACK\_LIB\_DIR} and {\tt LAPACK\_LIB}.

\item Specify the location and switch for the \blas~library
using {\tt BLAS\_LIB\_DIR} and {\tt BLAS\_LIB}.

\item Specify the compiling options:

\begin{itemize}

\item {\tt CC}: Specify the compiler, e.g., {\sl gcc} or {\sl
cc}.

\item {\tt CFLAGS}: Specify any flags to pass to the compiler and
linker, e.g., optimization flags such as {\tt -O3}.

\item {\tt LIB\_DIRS}: If necessary, using the {\tt -L} switch, specify
any directory locations for additional libraries that need to be
linked in. Leave items involving \lapack and \blas~ unchanged. It is
not necessary to include the SDPLR {\sl lib} directory since it is
automatically specified.

\item {\tt LIBS}: Specify the flags for the libraries that will be
linked in. This should definitely include {\tt -lgsl}, which is included
with and required by SDPLR, and the flags for the \blass and \lapacks
libraries from Step 1. Remember that linking order is important! For
example, on an Ubuntu 9.04 Linux system, the line might read
\begin{center}
{\tt LIBS = -lgsl -llapack -lblas -lgfortran -lm}
\end{center}
Here, {\tt -lgfortran} indicates a library that is required by \lapacks
and \blas, and {\tt -lm} is the standard math library.

\end{itemize}

% \item {\tt DEFINES}: If \arpacks and \metiss are to be linked in,
% then {\tt -D\_\_ARPACK} and {\tt -D\_\_METIS} should be specified.
% (The {\tt \_\_} symbol is two underscore characters.)

% \item {\tt  MEX}: (optional) If  you will  be using SDPLR  under Matlab,
% specify the Mex binary here.

% \item {\tt MEXEXT}: (optional) If you  will be using SDPLR under Matlab,
% specify the  default Matlab  binary extension  for your  platform (e.g.,
% {\tt mexglx} under Linux).

% \item {\tt LIBSMEX}: (optional) If you  will be using SDPLR under Matlab,
% specify those libraries which Mex requires. These may be the same or
% different from those specified in {\tt LIBS}.

\end{itemize}

\subsubsection*{Step 3}

Type {\sl make} to compile the program. Then type {\sl sdplr
vibra1.dat-s} to test the installation. Your output should be similar to
the file {\sl vibra1.dat-s.out} provided in the installation directory.
Typing {\sl make clean} will remove all object files, and typing {\sl
make cleanall} will remove the SDPLR executable and GSL library as well.

% \subsubsection{Step 5 (optional)}

% If you would like to use SDPLR from within Matlab, type {\sl make
% mex}; this assumes that the Matlab MEX compiler is called `mex'.
% The appropriate Matlab binary will be created. Here again, typing
% {\sl make clean} will remove all object files, and typing {\sl
% make cleanall} will remove the SDPLR executable, the GSL library,
% and the Matlab binary.

\subsection{Windows (via MinGW)}

MinGW is a GNU compiling environment for Windows. To compile SDPLR, you will need

\begin{itemize}

\item The MinGW base system (including the {\sl make\/} utility). Add
MinGW's {\sl bin\/} directory to your Windows path.

\item MinGW's {\sl gcc\/} tools. It is suggested to install the latest
{\sl gcc\/} release with additional packages such as {\sl gfortran\/}.

\item Working \lapack~ and \blas~ libraries within MinGW. The following
guide is suggested: {\tt http://gcc.gnu.org/wiki/LAPACK\%20on\%20Windows}.

\end{itemize}

Then tailor {\sl Makefile.inc\/} as described above; {\sl
Makefile.inc.mingw} is provided as an example. Finally, type
{\sl mingw32-make mingw\/}.

\section{Usage}

\subsection{Matlab}

Make sure that the Matlab binary for SDPLR and the file {\sl sdplr.m\/}
are in your Matlab path. Then type {\sl help sdplr} for more
information.

\subsection{Command line}

The syntax for SDPLR is
\begin{quote}
Usage \#1: {\tt sdplr <input\_file> [params\_file] [soln\_in]
[soln\_out]} \\ Usage \#2: {\tt sdplr gen\_params}
\end{quote}

\noindent Regarding usage \#1:

\begin{itemize}

\item If {\tt params\_file} is not specified, then a set of
default parameters is used.

\item Both files {\tt soln\_in} and {\tt soln\_out} refer to a
particular file format generated and used by SDPLR.

\item If {\tt soln\_in} is specified, then {\tt params\_file} must
be specified.

\item If {\tt soln\_out} is specified, then both {\tt
params\_file} and {\tt soln\_in} must be specified. However, a
dummy filename for {\tt soln\_in} may be used, allowing one to
save an out-file without an in-file.

\end{itemize}

\noindent Regarding usage \#2: This can be used to automatically
generate a valid {\tt params\_file} for SDPLR. The user is
prompted to enter each parameter value, and detailed explanations
of the parameters can be gotten by just typing `{\tt i}'.

\section{Input Formats}

By default, SDPLR accepts the sparse SDPA format, which is
explained at
\begin{center}
% \url{http://www.nmt.edu/~sdplib/FORMAT}
\url{http://euler.nmt.edu/~brian/sdplib/FORMAT}
\end{center}
However, some SDPs are not sparse but still have a great deal of
structure. The SDPLR format, which SDPLR is also capable of
handling, is more or less an extension of sparse SDPA format which
also allows low-rank data matrices to be specified easily.

The structure of the SDPLR format is as follows:

\begin{itemize}

\item The first line contains $m$, the number of constraint
matrices.

\item The second line contains $k$, the number of blocks in the SDP.

\item The next $k$ lines contain the sizes $n_1, \ldots, n_k$ of the
blocks, where a negative $n_j$ indicates a diagonal block.

\item The next line contains $b$, the right-hand side vector, all
on one line.

\item The next line is currently ignored, but should contain a
number. (This line is related to a feature of SDPLR which is in
development. Basically, for +/-1 combinatorial optimization SDPs,
this line will contain the trace of the primal matrix (a
constant), which can be used to provide dual bounds during the
execution of SDPLR.)

\item The remaining portion of the file is divided into $(m+1)k$
sections, giving the $k$ blocks of the $m+1$ objective and constraint
data matrices. The matrices should be listed in order, i.e., the
objective matrix first and then the constraint matrices in order, and
within each matrix, the blocks should be listed in order.

\item Two types of blocks can be specified for the data matrices: sparse
and low-rank.

\item If the $j$-th block of $A_i$ is a sparse data matrix, then the
section is as follows:

\begin{itemize}

\item The first line contains the matrix number $i$ ($i = 0$ for
objective), the block number $j$, the character {\tt s} to indicate
`sparse,' and the number nnz, which is the number nonzeros in the upper
triangular part of $A_i$.

\item The next nnz lines contain the entries in ``$i \ \ j \ \ $
entry'' format, where $i \le j$.

\end{itemize}

\item If the $j$-th block of $A_i$ is a low-rank data matrix of rank
$r$, then a factorization $B D B^T$ of this block must be known such
that $B \in \Re^{n_j \times r}$ and $D \in \Re^{r \times r}$ is
diagonal. Then the section specifying the block is as follows:

\begin{itemize}

\item The first line contains the matrix number (0 for objective), the
block number $j$, the character {\tt l} to indicate `low-rank,' and the
number $r$, which is the rank of the $j$-th block of $A_i$.

\item The next $r$ lines contain the diagonal entries of $D$ in
order.

\item The next $n_j r$ lines contain the entries of $B$ in
column-major format. Note that $B$ is specified as dense.

\end{itemize}

\end{itemize}

A brief example of the SDPLR format is the following, which encodes
the Lov\'asz theta SDP for the 5-cycle (more examples
available on the SDPLR website):
% \newpage
\begin{verbatim}
6
1
5
0.0 0.0 0.0 0.0 0.0 1.0
-1.0
0 1 l 1
-1.0
1.0
1.0
1.0
1.0
1.0
1 1 s 1
1 2 1.0
2 1 s 1
2 3 1.0
3 1 s 1
3 4 1.0
4 1 s 1
4 5 1.0
5 1 s 1
1 5 1.0
6 1 s 5
1 1 1.0
2 2 1.0
3 3 1.0
4 4 1.0
5 5 1.0
\end{verbatim}

%\bibliographystyle{plain}
%\bibliography{../bibliography}

\end{document}

### 2014-06-06 image_analysis_paper/pnas_draft2.tex
Added everything before working on submission

@@
 %
-In a cross-sectional study, developmental dynamics is reconstructed from multiple embryos, each of which contributes only a snapshot of a chemical or morphological process along its developmental trajectory \cite{jaeger2004dynamic, peter2012predictive, peter2011gene, fowlkes2008quantitative}.
+In a cross-sectional study, developmental dynamics is reconstructed from multiple embryos, each of which contributes only a snapshot of a chemical or morphological process along its developmental trajectory \cite{jaeger2004dynamic, peter2011gene, fowlkes2008quantitative}.
 %
@@
 %
-We assume that, in these sets of images, the main direction of variability is parameterized by the developmental time of each embryo, so that uncovering this direction will allow us to recover the developmental dynamics.
+We assume that the main direction of variability in these images is parameterized by the developmental time of each embryo, so that uncovering this direction will allow us to recover the developmental dynamics.
 
@@
 %
-In our working examples, each data point will be much higher dimensional (e.g., a pixelated image), and so we cannot extract this low-dimensional structure visually.
+In our working examples, each data point will be of much higher dimension (e.g., a pixelated image), and so we cannot extract this low-dimensional structure visually.
 %
@@
 
-Both angular synchronization and diffusion maps use pairwise information to uncover global structure in the data.
+%Both angular synchronization and diffusion maps use pairwise information to uncover global structure in the data.
 %
-Angular synchronization uses pairwise alignment information to find translations and rotations which are globally consistent, and diffusion maps uses pairwise distances to find global parametrizations which are (geometrically) informative.
+%Angular synchronization uses pairwise alignment information to find translations and rotations which are globally consistent, and diffusion maps uses pairwise distances to find global parametrizations which are (geometrically) informative.
 %
-Vector diffusion maps combines both steps into a single computation (see \SI), simultaneously addressing the tasks of registration and temporal ordering.
+%Vector diffusion maps combines both steps into a single computation (see \SI), simultaneously addressing the tasks of registration and temporal ordering.
 %
@@
 
-Previous studies established that throughout the third hour of development, the nuclear localization of Dl is unimodal and peaked at the ventral side of the embryo (reviewed in \cite{rushlow2012temporal}).
+Throughout the third hour of development, the nuclear localization of Dl is unimodal and peaked at the ventral side of the embryo (reviewed in \cite{rushlow2012temporal}).
 %
@@
 
-For this data, the quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the ordering obtained through an independent approach.
+For this set of images, the quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the ordering obtained through an independent approach.
 %
@@
 %
-\fig~\ref{subfig:raw_data2} shows a richer data set of $108$ images of {\it Drosophila} embryos.
+As an example, \fig~\ref{subfig:raw_data2} shows a set of $108$ images, which cover a thirty minute time interval spanning late cellularization through gastrulation.
 %
-These images cover a thirty minute time interval spanning late cellularization through gastrulation, when the arrangement of newly formed cells begins to deform into structures which will form the future organs \cite{leptin2005gastrulation}.
+During this time, the arrangement of newly formed cells begins to deform into structures which will form the future organs \cite{leptin2005gastrulation}.
 %
@@
 %
-During this time window the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, bringing in the future muscle cells.
+During this time window the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, bringing in the future muscle cells \cite{leptin2005gastrulation}.
 %
@@
 %
-At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing similar ``omega'' shaped patterns to arise on the dorsal side; these patterns are most readily seen in the last image of \fig~\ref{subfig:averaged_data2}.
+At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing similar ``omega'' shaped patterns to appear on the dorsal side; these patterns are most readily seen in the last image of \fig~\ref{subfig:averaged_data2}.
 
@@
 %
-Simple time-ordering based on this linear projection is therefore not possible (see \fig~\ref{fig:PCA_data2}).
+We found that simple time-ordering based on this linear projection is therefore not possible (see \fig~\ref{fig:PCA_data2}).
 %
@@
 %
-To highlight the developmental dynamics revealed by vector diffusion maps, we applied a Gaussian filter to the trajectory of registered and ordered images to obtain an averaged developmental path (see also \cite{kemelmacher2011exploring}).
+To highlight the developmental dynamics revealed by vector diffusion maps, we applied a Gaussian filter to the trajectory of registered and ordered images to obtain an averaged developmental path (see \SI and  \cite{kemelmacher2011exploring}).
 %
@@
 
-\subsection{Additional variability: identifying distinct developmental trajectories}
+\subsection{Identifying distinct developmental trajectories}
 
@@
 %
-Half of the embryos are wild type, and half of the embryos are mutants where dpERK is expressed in one broad peak at the ventral side.
+Half of the embryos are wild type, and half of the embryos are mutants where dpERK is expressed in one broad peak at the ventral side \cite{snail}.
 %
@@
 Two distinct branches appear to emerge in this two-dimensional embedding space.
+
+The inset of \fig~\ref{subfig:projection_data3} shows the two-dimensional projection, with the data points now colored to indicate whether they correspond to wild type or mutant embryos.
+%
+The two branches in the embedding help distinguish the two genotypes.
+%
+Selected images along the wild type and mutant trajectories are shown in \fig~\ref{subfig:average_wt_data3} and \fig~\ref{subfig:average_mut_data3}, respectively.
+%
+The portion of the embedding where the two genotypes are intermixed (lower-left region of the plot) contains embryos which are early in their developmental trajectory, where there is little dpERK signaling and the two groups of embryos are indistinguishable.
+%
+When the two paths diverge, the wild type data develops two distinct peaks, whereas the mutant data develops a single peak.
+%
+Therefore, vector diffusion maps can also help distinguish different expression patterns and developmental trajectories.
 
@@
 
-The inset of \fig~\ref{subfig:projection_data3} shows the two-dimensional projection, with the data points now colored to indicate whether they correspond to wild type or mutant embryos.
-%
-Clearly, the two branches in the embedding help distinguish the two genotypes.
-%
-Selected images along the wild type and mutant trajectories are shown in \fig~\ref{subfig:average_wt_data3} and \fig~\ref{subfig:average_mut_data3}, respectively.
-%
-The portion of the embedding where the two genotypes are intermixed (lower-left region of the plot) contains embryos which are early in their developmental trajectory, where there is little dpERK signaling and the two groups of embryos are indistinguishable.
-%
-When the two paths diverge, the wild type data develops two distinct peaks, whereas the mutant data develops a single peak.
-%
-Therefore, vector diffusion maps can also help distinguish different expression patterns and developmental trajectories.
-
 \section{Discussion}
@@
 %
-In parallel, ordering of large-scale cross-sectional data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different chemical species, such as mRNA abundances \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
+In parallel, ordering of large-scale cross-sectional data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different chemical species \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
 %
@@
 %
-The approach presented here is different because it addresses image registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace. 
+The approach presented here is different because it addresses image registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace; vector diffusion maps can also register images when global registration of the entire data set is topologically impossible \cite{zhao2014rotationally}.
 %
@@
 
-We used vector diffusion maps to analyze three different experimental data sets, collected specifically for this study. 
+We used vector diffusion maps to analyze three data sets, collected specifically for this study. 
 %
@@
 For this particular data set, a simpler approach would have sufficed;
-because of the relative simplicity of developmental dynamics (monotonic increase in the amplitude of a localized signaling pattern and relatively small tissue deformation), temporal order could have been inferred by projection of registered images on the first principal component of the data set (see \fig~\ref{fig:PCA_data1}).
+because of the relative simplicity of developmental dynamics (monotonic increase in the amplitude of a localized signaling pattern and small tissue deformation), temporal order could have been inferred by projection of registered images on the first principal component of the data set (see \fig~\ref{fig:PCA_data1}).
 %
-This approach was clearly not sufficient for the second data set (see \fig~\ref{fig:PCA_data2}), collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.
+This approach was not sufficient for the second data set (see \fig~\ref{fig:PCA_data2}), collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.
 %
@@
 
-{\color{red} Previously, angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy imaging data \cite{singer2012vector, zhao2014rotationally, singer2011viewing}.}
+%%%AMIT
+Previously, angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy images \cite{singer2012vector, zhao2014rotationally, singer2011viewing}.
 %
-Because of the high levels of instrument noise in these images, thousands of images were needed for successful shape reconstruction. 
+Because of high levels of instrument noise in these data, thousands of images were needed for successful shape reconstruction. 
 %
-Based on our work, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories, providing the first step in systems level studies of embryogenesis.
-%
-In the future, functional data analysis tools can be used for the quantitative estimation of both the average trajectory and its variability \cite{silverman2005functional}.  
-%
-One can also test which physical variable, among a set of candidate variables, is most correlated with the uncovered vector diffusion maps coordinate to gain insight into which physical quantities change across the developmental trajectory (e.g., total intensity of one signal, roundness of the embryo).
+Based on the presented results, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories. 
 %
 Here, we have used vector diffusion maps to register the images with respect to translations and rotations; however, it is conceivable that one could factor out other sources of variability, such as reflections \cite{singer2012vector, goemans1995improved, bandeira2013cheeger} or dilations.
+%In the future, functional data analysis tools can be used for the %quantitative estimation of both the average trajectory and its %variability \cite{silverman2005functional}.  
 %
-One might even envision automated analysis of embryos imaged in arbitrary three-dimensional orientations, in the spirit of cryo-EM image analysis.
+%One can also test which physical variable, among a set of candidate variables, is most correlated with the uncovered vector diffusion maps coordinate to gain insight into which physical quantities change across the developmental trajectory (e.g., total intensity of one signal, roundness of the embryo).
 %
-{\color{red} For these applications, vector diffusion maps is essential, as global registration of the entire imaging data set is topologically impossible \cite{zhao2014rotationally}.}
+
+For the data sets analyzed in this work, the Euclidean norm between the pixelated images was informative for identifying similar data points. 
 %
-{\color{red} In addition, we believe that algorithms based on rotation- and translation-invariant image features which are stable to small deformations, such as the scattering transform \cite{mallat2012group} or the bispectrum and steerable PCA \cite{zhao2014rotationally}, are especially promising; such methods circumvent the need for registration by using invariant features of the images in the data analysis.}
+For more complex data sets, for instance when the imaged samples are significantly deformed, working with raw images might be more challenging.   
 %
-Another important application is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \cite{krzic2012multiview, ichikawa2014live, rubel2010coupling}.
+In this case, one might choose to operate not on the images directly, but on those functions of the images that are invariant to symmetries and insensitive to small deformations.  
+%
+This is the basis of the recently developed scattering transform \cite{mallat2012group}, as well as the bispectrum-based steerable PCA algorithms in \cite{zhao2014rotationally}. 
+%
+Such methods circumvent the need for registration by using invariant features of the images in the data analysis.
+%
+Another direction for future work is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \cite{krzic2012multiview, ichikawa2014live, rubel2010coupling}.
 %
 Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.
+
+%Here, we have used vector diffusion maps to register the images with respect to translations and rotations; however, it is conceivable that one could factor out other sources of variability, such as reflections \cite{singer2012vector, goemans1995improved, bandeira2013cheeger} or dilations.
+%
+%One might even envision automated analysis of embryos imaged in arbitrary three-dimensional orientations, in the spirit of cryo-EM image analysis.
+%
+%%%AMIT
+%For these applications, vector diffusion maps is essential, as global registration of the entire imaging data set is topologically impossible \cite{zhao2014rotationally}.
+%
+%%%AMIT
+%In addition, we believe that algorithms based on rotation- and translation-invariant image features which are stable to small deformations, such as the scattering transform \cite{mallat2012group} or the bispectrum and steerable PCA \cite{zhao2014rotationally}, are especially promising; such methods circumvent the need for registration by using invariant features of the images in the data analysis.
+%
 
@@
 %
-{\it Sna\textsuperscript{\it IIG05}/Cyo hb-lacZ} flies were used to study embryos with one peak of dpERK expression.  
+{\it Sna\textsuperscript{\it IIG05}/Cyo hb-lacZ} flies were used to study mutant embryos with one peak of dpERK activation.  
 %
@@
 The progression of membrane invagination during cellularization was used to assign time point to each embryo in the third hour of development.
-Time-lapse imaging of Sqh-GFP embryo was used to generate correlation curve between the membrane ingression length and the developmental time (n=7) (\fig~\ref{fig:membrane_curve}B).
+Time-lapse imaging of Sqh-GFP embryo was used to generate the calibration curve that relates the membrane ingression length and the developmental time.
+Time-lapsing imaging of 7 embryos was used to construct the calibration curve (\fig~\ref{fig:membrane_curve}B).
 The rate at which membrane invaginates was highly reproducible from embryo to embryo.
@@
 
-\bibitem{peter2012predictive}
-Peter IS, Faure E, Davidson EH (2012) Predictive computation of genomic logic processing functions in embryonic development. \textit{Proc Natl Acad Sci USA} 109(41):16434--16442.
+%\bibitem{peter2012predictive}
+%Peter IS, Faure E, Davidson EH (2012) Predictive computation of genomic %logic processing functions in embryonic development. \textit{Proc Natl %Acad Sci USA} 109(41):16434--16442.
 
@@
 Jaeger J, Manu, Reinitz J (2012) \textit{{D}rosophila} blastoderm patterning.
-  \textit{Curr Opin Genet Dev} 22(6):533 -- 541, genetics of system biology.
+  \textit{Curr Opin Genet Dev} 22(6):533 -- 541.
 
@@
 Exploring photobios. \textit{ACM Transactions on Graphics (TOG)} 30(4):61.
+
+\bibitem{snail}
+Ip Y~T, Park R~E, Kosman D, Bier E, Levine M (1992) 
+The dorsal gradient morphogen regulates stripes of rhomboid expression in the presumptive neuroectoderm of the \textit{Drosophila} embryo. \textit{Genes Dev} 6(9):1728-39.
 

### 2014-06-06 image_analysis_paper/si_information.tex
Added everything before working on submission

@@
 %
-$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/10^{th}$ of the median of the pairwise distances between data points for the first two data sets (and half of the median of the pairwise distances between data points for the third data set).
+$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon^2$ to be $1/10^{th}$ of the median of the squared pairwise distances between data points for the first two data sets (and half of the median of the squared pairwise distances between data points for the third data set).
 
@@
 and we select $k$ such that the coordinate $\psi_{k, 1}$ has the largest variability, i.e., the $j^{th}$ coordinate is $\psi_{k,1}$, where $k$ is the solution to
-$$ \max_{jd +1 < k \le (j+1)d} \sum_i \psi_{k,1} (i)^2. $$
+\begin{equation} \label{eq:first_embed_vdm}
+\max_{jd +1 < k \le (j+1)d} \sum_i \psi_{k,1} (i)^2. 
+\end{equation}
+
+\begin{table}[t]
+%\caption{Outline of algorithm used to temporally order images}
+\setlength{\fboxrule}{2pt}
+\framebox[\columnwidth]{
+\begin{minipage}{0.9\columnwidth}
+{\bf Outline of algorithm used to register and order images}
+\begin{enumerate}
+\item Compute the alignments for each pair of images, as described in \eqref{eq:opt_pairwise}.
+Retain both the optimal pairwise alignments (translations and rotations) as well as the minimum distances obtained when pairs are optimally aligned.
+%
+\item Compute the corresponding pairwise three-dimensional rotations from the translations and rotations, using \eqref{eq:angle_relations}, \eqref{eq:three_rot}, and \eqref{eq:total_rot}.
+%
+\item Select $\epsilon$ for use in the diffusion maps kernel. 
+We often begin with using the median of the pairwise distances as $\epsilon$, and then adjust as needed.
+\item Compute the matrix $S$ in \eqref{eq:vdm_S}. 
+$A_{ij}$ is computed from \eqref{eq:dmaps_W} and \eqref{eq:dmaps_A}, using the pairwise distances from step 1, and $H_{ij} = R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$, which are computed in step 2.
+%
+\item Compute the eigenvalues  $\lambda_1, \lambda_2, \dots, \lambda_{3m}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{3m}$ of $S$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{3m}|$.
+%
+\item Estimate the optimal three-dimensional rotation for each image: stack the first three eigenvectors $\phi_1, \phi_2, \phi_3$ into a $3m \times 3$ matrix, and then divide this matrix into $m$ $3 \times 3$ blocks, $\hat{R}_1, \dots, \hat{R}_m$.
+Compute the estimated rotations for each of these $m$ blocks as in \eqref{eq:R_est}.
+If most of the rotations are improper ($det(R) = -1$), flip the sign of one of the eigenvectors.
+\item Multiply each estimated rotation $R_{i, est}$ by $R_{1, est}^T$ (so that $R_{1, est}$ will become the identity matrix).
+\item For each image $i$, compute the optimal translation and rotation to align the image by converting $R_{i, est}^T$ (note the transpose) to the corresponding translation and rotation, described in \eqref{eq:angle_invert}.
+\item Compute the first embedding coordinate $\psi_{k,1}(i)$ as described in \eqref{eq:vdm_coord} and \eqref{eq:first_embed_vdm}, where $4 \le k \le 6$.
+\item To order the images, sort them by the values of this embedding coordinate $\psi_{k,1}(i)$.
+\end{enumerate} 
+\end{minipage}
+}
+\end{table}
 
@@
 %\small
-\begin{equation}
+\begin{equation} \label{eq:three_rot}
 \begin{aligned}
@@
 %
-\begin{figure*}[t]
-{\bf Item C in the figure below is our main observation.}
-
-\raisebox{7.6cm}{\sffamily{A}}
-\includegraphics[width=8cm]{raw_data1}
-\hfill
-\raisebox{7.6cm}{\sffamily{B}}
-\includegraphics[width=8cm]{FBsPCA_evals}
-
-\raisebox{5.7cm}{\sffamily{C}}
-\includegraphics[width=16cm, trim=2cm 0cm 2cm 0cm, clip]{FBsPCA_dmaps}
-\caption{I computed the FBsPCA basis for the set of images (separately for the red and green signals).
+%\begin{figure*}[t]
+%{\bf Item C in the figure below is our main observation.}
 %
-I then computed the bispectrum coefficients for each image expanded in the FBsPCA basis, and used these coefficients in a DMAPS calculation.
+%\raisebox{7.6cm}{\sffamily{A}}
+%\includegraphics[width=8cm]{raw_data1}
+%\hfill
+%\raisebox{7.6cm}{\sffamily{B}}
+%\includegraphics[width=8cm]{FBsPCA_evals}
 %
-The results are shown above.
-%
-However, (mainly, we suspect) because the bispectrum is not stable to deformations, it does not seem to be able to identify neighboring images.
-%
-(A) Images used for FBsPCA.
+%\raisebox{5.7cm}{\sffamily{C}}
+%\includegraphics[width=16cm, trim=2cm 0cm 2cm 0cm, clip]{FBsPCA_dmaps}
+%\caption{I computed the FBsPCA basis for the set of images (separately for the red and green signals).
+%%
+%I then computed the bispectrum coefficients for each image expanded in the FBsPCA basis, and used these coefficients in a DMAPS calculation.
+%%
+%The results are shown above.
+%%
+%However, (mainly, we suspect) because the bispectrum is not stable to deformations, it does not seem to be able to identify neighboring images.
+%%
+%(A) Images used for FBsPCA.
 (B) Eigenvalue spectrum from diffusion maps, computed using bispectrum coefficients.
@@
 }
-\end{figure*}
+%\end{figure*}
 

### 2014-07-10 image_analysis_paper/SI_text.tex
Added live imaging movies

--- 2014-06-06 image_analysis_paper/si_information.tex
@@
 
-\bibliographystyle{pnas2}
-\bibliography{background_reading/references,../../references/references}
+\begin{thebibliography}{10}
+
+\bibitem{singer2011angular}
+Singer A (2011) Angular synchronization by eigenvectors and semidefinite
+  programming. \textit{Appl Comput Harmon Anal}
+  30(1):20--36.
+
+\bibitem{ahuja2007template}
+Ahuja S, Kevrekidis I~G, Rowley C~W (2007) Template-based stabilization of
+  relative equilibria in systems with continuous symmetry. \textit{Journal of
+  Nonlinear Science} 17(2):109--143.
+
+\bibitem{goemans1995improved}
+Goemans M~X, Williamson D~P (1995) Improved approximation algorithms for maximum cut and satisfiability problems using semidefinite programming.
+\textit{Journal of the ACM} 42(6):1115--1145.
+
+\bibitem{bandeira2013cheeger}
+Bandeira A~S, Singer A, Spielman D~A (2013) A Cheeger inequality for the graph connection Laplacian. \textit{SIAM Journal on Matrix Analysis and Applications} 34(4):1611--1630.
+
+\bibitem{coifman2005geometric}
+Coifman R~R, Lafon S, Lee A~B, Maggioni M, Nadler B, \textit{et~al.} (2005)
+  Geometric diffusions as a tool for harmonic analysis and structure definition
+  of data: Diffusion maps. \textit{Proc Natl Acad Sci USA} 102(21):7426--7431.
+
+\bibitem{coifman2008graph}
+Coifman R~R, Shkolnisky Y, Sigworth F~J, Singer A (2008) Graph {L}aplacian
+  tomography from unknown random projections. \textit{IEEE Trans Image Process} 17(10):1891--1899.
+
+\bibitem{rohrdanz2011determination}
+Rohrdanz M~A, Zheng W, Maggioni M, Clementi C (2011) Determination of reaction
+  coordinates via locally scaled diffusion map. \textit{J Chem Phys} 134:124116.
+
+\bibitem{Belkin2003}
+Belkin M, Niyogi P (2003) Laplacian eigenmaps for dimensionality reduction and
+  data representation. \textit{Neural Comput} 15:1373--1396.
+
+\bibitem{singer2012vector}
+Singer A, Wu H~T (2012) Vector diffusion maps and the connection laplacian.
+  \textit{Commun Pure Appl Math} 65(8):1067--1144.
+
+\bibitem{singer2013spectral}
+Singer A, Wu H~t (2013) Spectral convergence of the connection {L}aplacian from
+  random samples. \textit{arXiv preprint arXiv:13061587} .
+
+\bibitem{wigner1955characteristic}
+Wigner E~P (1955) Characteristic vectors of bordered matrices with infinite
+  dimensions. \textit{The Annals of Mathematics} 62(3):548--564.
+
+\bibitem{wigner1958distribution}
+Wigner E~P (1958) On the distribution of the roots of certain symmetric
+  matrices. \textit{The Annals of Mathematics} 67(2):325--327.
+
+\bibitem{ferguson2010systematic}
+Ferguson A~L, Panagiotopoulos A~Z, Debenedetti P~G, Kevrekidis I~G (2010)
+  Systematic determination of order parameters for chain dynamics using
+  diffusion maps. \textit{Proc Natl Acad Sci USA}
+  107(31):13597--13602.
+
+\bibitem{kemelmacher2011exploring}
+Kemelmacher-Shlizerman I, Shechtman E, Garg R, Seit, S~M (2011)
+Exploring photobios. \textit{ACM Transactions on Graphics (TOG)} 30(4):61.
+
+\bibitem{lim2013kinetics}
+Lim B, Samper N, Lu H, Rushlow C, Jim{\'e}nez G, \textit{et~al.} (2013)
+  Kinetics of gene derepression by erk signaling. \textit{Proc Natl Acad Sci USA} 110(25):10330--10335.
+
+\end{thebibliography}
 
@@
 \begin{figure}
-\includegraphics[width=8.4cm]{ap_dv}
+\includegraphics{figS1}
 \caption{(Left) A lateral view of a {\it Drosophila} embryo stained with DAPI (gray), dpERK (red), and Dorsal (green).
@@
 For the first and third data sets (cellularization), images were collected at the focal plane $\sim 18\%$ from either the anterior or posterior pole of an embryo (arrows in the left image).
-During cellularization, it is known that the developmental dynamics are identical at these two regions \cite{...}.
+During cellularization, it is known that the developmental dynamics are identical at these two regions \cite{lim2013kinetics}.
 In the second data set (gastrulation), images were collected at the focal plane $\sim 18\%$ from the posterior pole of an embryo (right arrow in the left image).
@@
 \begin{figure}
-\includegraphics[width=8.4cm, trim=0cm 5.6cm 0cm 0cm, clip]{SI_fig6}
+\includegraphics{figS2}
 \caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different times during development.
@@
 \begin{figure}
-\includegraphics[height=9cm]{raw_data3}
+\includegraphics{figS3}
 \caption{Images of mutant and wild type {\it Drosophila} embryos during cellularization, used for the analysis in \fig~5.
@@
 \begin{figure}
-\includegraphics[width=8.4cm]{SI_fig3}
+\includegraphics{figS4}
 \caption{Synthetic data set used to illustrate the data processing algorithms.
@@
 \begin{figure*}
-\begin{minipage}{0.22\textwidth}
-\centering
-%\includegraphics[width=\textwidth]{sphere_1}\\
-\includegraphics[width=\textwidth]{sphere2_1}\\
-\sffamily{A}
-\end{minipage}
-%
-\hfill
-%
-\begin{minipage}{0.22\textwidth}
-\centering
-%\includegraphics[width=\textwidth]{sphere_2}\\
-\includegraphics[width=\textwidth]{sphere2_2}\\
-\sffamily{B}
-\end{minipage}
-\begin{minipage}{0.22\textwidth}
-\centering
-%\includegraphics[width=\textwidth]{sphere_3}\\
-\includegraphics[width=\textwidth]{sphere2_3}\\
-\sffamily{C}
-\end{minipage}
-\begin{minipage}{0.22\textwidth}
-\centering
-%\includegraphics[width=\textwidth]{sphere_4}\\
-\includegraphics[width=\textwidth]{sphere2_4}\\
-\sffamily{D}
-\end{minipage}
+\includegraphics{figS5}
 \caption{Illustration of the correspondence between three-dimensional rotations to translations and rotations in two dimensions.
@@
 \begin{figure*}
-%\raisebox{4.5cm}{\sffamily{A}}
-%\includegraphics[width=5cm]{data1_evals}
-\begin{tikzpicture}
-	\node at (0,0) {\includegraphics[width=5cm]{data1_evals_dist}};
-	\node at (-2.5cm, 2cm) {\sffamily{A}};
-%	\draw[<->, ultra thick, cyan] (-1.4cm, -1.9cm) -- (1.7cm, -1.9cm);
-	\draw[<->, ultra thick, cyan] (-1.4cm, -1.85cm) -- (1.7cm, -1.85cm);
-	\draw[->, ultra thick, red] (1.97cm, -0.5cm) -- (1.97cm, -1cm);
-\end{tikzpicture}
-%\raisebox{4.5cm}{\sffamily{B}}
-%\includegraphics[width=5cm]{data2_evals}
-\begin{tikzpicture}
-	\node at (0,0) {\includegraphics[width=5cm]{data2_evals_dist}};
-	\node at (-2.5cm, 2cm) {\sffamily{B}};
-	\draw[<->, ultra thick, cyan] (-1.2cm, -1.85cm) -- (1.5cm, -1.85cm);
-	\draw[->, ultra thick, red] (1.97, -0.85cm) -- (1.97cm, -1.34cm);
-\end{tikzpicture}
-%\raisebox{4.5cm}{\sffamily{C}}
-%\includegraphics[width=5cm]{data3_evals}
-\begin{tikzpicture}
-	\node at (0,0) {\includegraphics[width=5cm]{data3_evals_dist}};
-	\node at (-2.5cm, 2cm) {\sffamily{C}};
-	\draw[<->, ultra thick, cyan] (-0.25cm, -1.85cm) -- (0.35cm, -1.85cm);
-	\draw[->, ultra thick, red] (1.97, -1.1cm) -- (1.97cm, -1.6cm);
-	\draw[->, ultra thick, red] (1.52, -1.1cm) -- (1.52cm, -1.6cm);
-	\draw[->, ultra thick, red] (0.64, -1.1cm) -- (0.64cm, -1.6cm);
-\end{tikzpicture}
+\includegraphics{figS6}
 \caption{The histograms of the eigenvalues from vector diffusion maps analysis for the three data sets.
@@
 $3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $6$ eigenvalues correspond to the first two VDM components which are now necessary to parametrize the developmental dynamics of the two genotypes.}
-%there is a spectral gap after the first (non-trivial) group of eigenvalues.
-%\caption{The eigenvalue spectra from vector diffusion maps analysis for the three data sets.
-%%
-%As expected, the eigenvalues appear in groups of 3, as $d=3$ when using vector diffusion maps to register with respect to translations and rotations.
-%%
-%(A) Spectrum from images during cellularization (\fig~3).
-As expected, there is a spectral gap after the first (non-trivial) group of eigenvalues: the first three eigenvalues correspond to the rotational and translational alignment, the second three eigenvalues correspond to the first VDM component which uncovers the developmental dynamics, and the then there is an appreciable gap before the third group of eigenvalues.
-%%
-%(B) Spectrum from images during gastrulation (\fig~4). Again,
-%there is a spectral gap after the first (non-trivial) group of eigenvalues.
-%%
-%(C) Spectrum from images of mutant and wild type embryos during cellularization (\fig~5).
-%We see a group of three eigenvalues corresponding to the translational and rotational alignments, one group that is clearly significant, and then a second group that, while less important, is still above the level of noise of the subsequent eigenvalues, indicating that two components are necessary to describe the data.
-}
 \label{fig:eval_spectra}
@@
 \begin{figure}
-\includegraphics[width=8.7cm]{PCA_ordered}
+\includegraphics{figS7}
 \caption{Images of embryos during cellularization (from \fig~3), registered using angular synchronization \cite{singer2011angular} and ordered by the first PCA projection coefficient.
@@
 \begin{figure*}
-\raisebox{5.2cm}{\sffamily{A}}
-\includegraphics[width=16.8cm]{data2_PCA_ordered}
-
-\raisebox{3.7cm}{\sffamily{B}}
-\includegraphics[width=8.4cm, angle=90]{data2_PCA_variance}
-\raisebox{3.7cm}{\sffamily{C}}
-\includegraphics[width=8.4cm]{data2_PCA_proj}
+\includegraphics{figS8}
 
@@
 
-%
-%\begin{figure*}[t]
-%{\bf Item C in the figure below is our main observation.}
-%
-%\raisebox{7.6cm}{\sffamily{A}}
-%\includegraphics[width=8cm]{raw_data1}
-%\hfill
-%\raisebox{7.6cm}{\sffamily{B}}
-%\includegraphics[width=8cm]{FBsPCA_evals}
-%
-%\raisebox{5.7cm}{\sffamily{C}}
-%\includegraphics[width=16cm, trim=2cm 0cm 2cm 0cm, clip]{FBsPCA_dmaps}
-%\caption{I computed the FBsPCA basis for the set of images (separately for the red and green signals).
-%%
-%I then computed the bispectrum coefficients for each image expanded in the FBsPCA basis, and used these coefficients in a DMAPS calculation.
-%%
-%The results are shown above.
-%%
-%However, (mainly, we suspect) because the bispectrum is not stable to deformations, it does not seem to be able to identify neighboring images.
-%%
-%(A) Images used for FBsPCA.
-(B) Eigenvalue spectrum from diffusion maps, computed using bispectrum coefficients.
-There is no visible spectral gap after the first component, as would be expected for this particular data set.
-(C, top) First 4 diffusion maps variables, computed from the bispectrum coefficients, plotted against the membrane thickness.
-Note there is no appreciable correlation: {\bf each diffusion maps variable simply selects one image from the data set as ``abnormal''.} (C, bottom) The corresponding images that each diffusion maps mode ``selects.'' In general, these seem to be the late images with ventral furrows beginning to form.
-}
-%\end{figure*}
-
 \end{document}

### 2014-07-10 image_analysis_paper/paper_figures2/figS6.tex
Added live imaging movies

--- 2014-06-02 image_analysis_paper/si_information.tex
@@
-\documentclass{pnastwo}
-%\documentclass[10pt,twocolumn]{article}
+\documentclass[12pt]{minimal}
 
-%\usepackage[margin=1in]{geometry}
-%\usepackage{graphicx}
-\graphicspath{ {paper_figures2/} }
+\usepackage{graphicx}
 
@@
 
-\usepackage{float}
-\usepackage{amssymb,amsfonts,amsmath}
-\DeclareMathOperator*{\argmin}{arg\,min}
-
-\makeatletter
-\newcommand{\customlabel}[2]{%
-\protected@write \@auxout {}{\string \newlabel {#1}{{#2}{}}}}
-\makeatother
-
-\renewcommand\thefigure{S\arabic{figure}}
-
-\newcommand{\fig}[0]{Fig.}
+\usepackage[margin=0in, paperwidth=17.8cm, paperheight=5.3cm]{geometry}
 
@@
 
-\begin{article}
-
-\section{Materials and Methods}
-
-\subsection{Image parameters}
-
-The original resolution of each image is $1024 \times 1024$ pixels.
-%
-The images are subsampled to $100 \times 100$ pixels for vector diffusion maps analysis.
-%
-A scale bar for a representative image is shown in \fig~\ref{fig:ap_dv}.
-%
-When imaging, the laser settings are calibrated so that images have good dynamic range, while not saturating the detector.
-%
-The same laser settings are maintained within each experiment.
-%
-The intensity of each channel is integer-valued, and ranges from $0$ (no signal) to $255$ (saturated). 
-%
-We blur the nuclear signal using a disc filter with a $5$~pixel radius (using the \texttt{imfilter} function in MATLAB\textsuperscript{\textregistered}).
-%
-We also adjust the intensity of the nuclear signal in each image such that 1\% of the pixels are saturated 
-(using the \texttt{imadjust} function in MATLAB), and then scale the intensity of the nuclear signal by half.
-
-For the first and third data sets in the paper (images taken during cellularization), we only use the red and green channels (dpERK and Dl) in the vector diffusion maps computations, as the spatial location of the nuclei (the gray channel) does not change appreciably during this time window.
-%
-The nuclei in the corresponding figures are thus shown only for visual reference.
-%
-For the second data set of images (taken during gastrulation), we use all three channels (dpERK, Twi, and nuclei) in the vector diffusion maps computations.
-%
-For this particular data set, we multiply the intensity of the Twi signal by 1.5, as this channel was not completely saturated in the original imaging data.
-
-\subsection{Algorithms}
-
-We will demonstrate the algorithms for registration and temporal ordering using a synthetic data set.
-%
-The relatively simple dynamics of this data set will allow us to easily visualize and illustrate the main features of the different algorithms.
-%
-Motivated by the geometry of our images, we construct a sequence of concentration profiles defined on a ring, and rotate each ring randomly around its center; an example is shown in \fig~\ref{subfig:1d_example}.
-%
-Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown at the bottom of \fig~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
-%
-Each concentration profile is a noisy Gaussian (shown in \fig~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as a function of ``time".
-%
-We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_{100}$, the group of integers modulo $100$).
-%
-\fig~\ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
-%
-Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.
-
-\subsubsection{Angular synchronization \cite{singer2011angular}}
-
-Let $ x_1, \dots, x_m$ denote the signals that we wish to align with respect to rotations;
-each signal is a function defined on the unit circle (on the plane).
-%
-%Practically, it is discretized in a $n$-long vector (the local intensity at $n$ equidistant points around the circle);
-%rotating the function by an angle $\theta$ then corresponds to cyclically shifting the elements of $x_i$ by $\frac{\theta_i}{2 \pi} n$ (rounded to the nearest integer to obtain a valid shift).
-%
-First assume that each signal $x_i$ is a {\it noisy} rotated copy of the underlying signal $x_{true}$
-(which we are {\it not} given), such that
-\begin{equation}
-x_i = f(x_{true}, \theta_i) + \xi_i
-\end{equation}
-where the function $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ radians, and $\xi_i$ is a (typically Gaussian) noise term.
-%
-Our goal is to recover $\theta_1, \dots, \theta_m$.
-%
-Up to noise,
-\begin{equation} \label{eq:pairwise_rot}
-x_i \approx f(x_j, \theta_i - \theta_j) ;
-\end{equation}
-note that \eqref{eq:pairwise_rot} does not require knowledge of $x_{true}$.
-%
-We can obtain an {\it estimate} of $\theta_i - \theta_j$ by computing the rotation that optimally aligns $x_j$ to $x_i$,
-i.e., %$\theta_{ij} \approx \theta_i - \theta_j$, where
-%
-\begin{equation} \label{eq:opt_angle}
-\theta_i - \theta_j \approx \theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2.
-\end{equation}
-%
-Practically, the signals are discretized in a $n$-long vector (the local intensity at $n$ equidistant points around the circle);
-rotating the function by an angle $\theta$ then corresponds to cyclically shifting the elements of $x_i$
-by $\frac{\theta_i}{2 \pi} n$ (rounded to the nearest integer to obtain a valid shift).
-%
-For the one-dimensional discretized profiles shown in \fig~\ref{fig:1d_demo}, we exhaustively search over all $n=100$ possible shifts of the signals to obtain the optimal angles in \eqref{eq:opt_angle}.
-%
-Alternatively, for continuous signals, an optimization algorithm
-can be used \cite{ahuja2007template}.
-
-Rather than work with the angles $\theta_{ij}$ directly, it is more convenient to consider the rotation matrices,
-\begin{equation} \label{eq:R_theta}
-R(\theta_{ij}) = \begin{bmatrix}
-\cos(\theta_{ij}) & -\sin(\theta_{ij}) \\
-\sin(\theta_{ij}) & \cos(\theta_{ij})
-\end{bmatrix},
-\end{equation}
-which we can think of as operating on the points of the unit circle (on the plane) on which our signal is defined.
-%
-Successive rotations correspond to multiplication of the corresponding rotation matrices: $R(\alpha_1 + \alpha_2) = R(\alpha_1) R(\alpha_2)$.
-%
-Due to the orthogonality of rotation matrices, $R(-\alpha) = R(\alpha)^T$.
-
-Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$; we write our procedure for general $d$ because we will later consider three-dimensional rotations).
-%
-We construct the matrix $H \in \mathbb{R}^{md \times md}$, where $H$ is an $m \times m$ matrix of $d \times d$ blocks, with the $i,j^{th}$ block of $H$, $H_{ij}$, defined as
-\begin{equation} \label{eq:H_to_R}
-H_{ij} = R(\theta_{ij}).
-\end{equation}
-%
-%
-Under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, $H_{ij} \approx R(\theta_i) R(\theta_j)^T$
-%\begin{equation}
-%H_{ij} = R(\theta_{ij}) \approx R(\theta_i - \theta_j) = R(\theta_i) R(-\theta_j) = R(\theta_i) R(\theta_j)^T,
-%\end{equation}
- and
-\begin{equation} \label{eq:H_low_rank}
-	H \approx
-	\begin{bmatrix}
-	R(\theta_1) \\
-	R(\theta_2) \\
-	\vdots \\
-	R(\theta_m)
-	\end{bmatrix}
-	\begin{bmatrix}
-	R(\theta_1)^T R(\theta_2)^T \dots R(\theta_m)^T
-	\end{bmatrix}.
-\end{equation}
-%
-It follows directly from \eqref{eq:H_low_rank} that the top block eigenvector of $H$ contains our best estimates of $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$.
-%
-Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$ ordered so that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$.
-%
-Then,
-\begin{equation} \label{eq:R_hat}
-\hat{R} =
-\begin{bmatrix}
-\hat{R}_1 \\
-\hat{R}_2 \\
-\vdots \\
-\hat{R}_m
-\end{bmatrix} =
-\begin{bmatrix}
-| & | & & | \\
-\phi_1 & \phi_2 & \dots & \phi_d \\
-| & | & & |
-\end{bmatrix},
-\end{equation}
-where $\hat{R}_i \in \mathbb{R}^{d \times d}$ is (nearly) the estimate for $R(\theta_i)$.
-%
-To obtain our estimate of $R(\theta_i$), denoted $R_{i, est}$, we project $\hat{R}_i$ onto the closest orthogonal matrix,
-\begin{equation} \label{eq:R_est}
-R_{i, est} = U_i V_i^T,
-\end{equation}
-where $U_i$ and $V_i$ are the left and right singular vectors, respectively, of $\hat{R}_i$.
-%
-We adjust the sign of $\phi_1$ so that $det(R_{i, est}) = +1$, ensuring proper rotations 
-(note that systematically incorporating improper rotations is also possible \cite{goemans1995improved, bandeira2013cheeger}).
-%
-We estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and register the signals by rotating signal $i$ by $-\theta_i$.
-%
-We note that, in our actual computations, the pairwise rotations $\theta_{ij}$ are computed in a discrete setting, then the overall
-synchronization is performed in the continuum context to obtain $\theta_i$, and the results are rounded to give the closest
-discrete shift.
-
-Importantly, this formulation also considers {\it higher-order} consistency information.
-%
-For example, given our pairwise estimates $R_{ij}$, we know that relationships of the form
-\begin{equation} \label{eq:triplet_consistency}
-R(\theta_{ik}) R(\theta_{kj}) \approx R(\theta_i) R(\theta_k)^T R(\theta_k) R(\theta_j)^T = R(\theta_i) R(\theta_j)^T
-\end{equation}
-should also hold.
-%
-Note that
-\begin{equation}
-(H^2)_{ij} = \sum_k R(\theta_{ik}) R(\theta_{kj});
-\end{equation}
-therefore, {\it all} information of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$ (and higher order
-consistency information in its higher powers).
-%
-Because $H$ and $H^2$ have the same eigenvectors, our problem formulation accounts for not only pairwise alignment information, but also these higher-order considerations.
-
-\subsubsection{Diffusion maps \cite{coifman2005geometric}}
-
-Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional vector space), we want to find a coordinate transformation $y(x)$ that preserves local geometry: points that are ``close" in the original space should also be ``close" in the coordinates $y$.
-%
-The first step is to construct the matrix $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if points $x_i$ and $x_j$ are ``close.''
-%
-We use a diffusion kernel,
-\begin{equation} \label{eq:dmaps_W}
-W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right),
-%W_{ij} = e^{ -\frac{d^2(x_i, x_j)}{\epsilon^2}}
-\end{equation}
-where $d(x_i, x_j)$ is a pairwise distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic scale.
-%
-Points less than $\epsilon$ apart are thus considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
-%
-$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/10^{th}$ of the median of the pairwise distances between data points for the first two data sets (and half of the median of the pairwise distances between data points for the third data set).
-
-To find the coordinate $y$, we want solve the following optimization problem \cite{Belkin2003}
-\begin{equation} \label{eq:dmaps_opt_problem}
-\argmin_{y} \sum_{ij} W_{ij} (y(x_i) - y(x_j))^2.
-\end{equation}
-%
-We first compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{m} W_{ij}$, and the matrix $A$, where
-\begin{equation} \label{eq:dmaps_A}
-A = D^{-1} W.
-\end{equation}
-%
-We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_m$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$.
-%
-Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors.
-%
-Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector; this is a trivial solution to \eqref{eq:dmaps_opt_problem}.
-%
-%In general, the next few eigenvectors $\phi_2, \dots, \phi_m$ give ``meaningful'' embedding coordinates for the data, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point.
-%
-The next eigenvector, $\phi_2$, is the (non-trivial) solution to \eqref{eq:dmaps_opt_problem}, so that $\phi_2(j)$, the $j^{th}$ entry of $\phi_2$, gives the ``new" coordinate for data point $x_j$ (i.e., $\phi_2(j) = y(x_j)$).
-%
-In our application, we have assumed that this {\em single} direction of variability, parameterized by $\phi_2$, is one-to-one with time.
-%
-Ordering the data by $\phi_2(j)$ will then, effectively, order them in time.
-%
-The procedure generalizes when the data lie on higher-dimensional manifolds (not just curves) in data space, where leading eigenvectors can give subsequent embedding coordinates for the data.
-%
-
-\subsubsection{Vector diffusion maps\cite{singer2012vector}}
-
-In vector diffusion maps, given data points $x_1, \dots, x_m$, one first constructs the matrix $S \in \mathbb{R}^{md \times md}$, with the $i,j^{th}$ block of $S$, $S_{ij}$, defined as
-\begin{equation} \label{eq:vdm_S}
-	S_{ij} = A_{ij} H_{ij}
-\end{equation}
-%
-where $A_{ij} \in \mathbb{R}$ (defined in \eqref{eq:dmaps_A}) pertains to the diffusion kernel between data points, and $H_{ij} \in \mathbb{R}^{d \times d}$ (defined in \eqref{eq:H_to_R}) pertains to the pairwise alignment between data points.
-%
-It is important to note that distance $d(x_i, x_j)$ used in the diffusion kernel in \eqref{eq:dmaps_W} is the distance between data points {\it after} after pairwise alignment, i.e., the minimum distance between all possible shifts of the two data points (which is obtained in \eqref{eq:opt_pairwise}).
-%
-In the language of symmetry groups, this distance is a metric between the orbits induced by the relevant symmetry group.
-
-One then computes the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_{md}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{md}$ of $S$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$.
-%
-These eigenvectors contain information about {\it both} the optimal rotations (the ``synchronization" component) and the
-variation of the data {\it after} the spatial symmetries have been factored out (in our case, their temporal variation).
-%
-Assuming that the data (after symmetries have been factored out) are relatively closely clustered, it is reasonable
-to expect, as in angular synchronization, that the top (block) eigenvector of $S$ contains approximations of the optimal rotations,
-which can be computed in the same way from \eqref{eq:R_est}.
-%
-We then expect subsequent eigenvectors to contain information about the main direction(s) of data variability modulo the geometric symmetries.
-%%%
-%%% somewhere we need to talk about compact groups
-%%% discussion might be a good place because SCALING is obviously important and
-%%% it is not compact......
-%%%
-%However, the eigenvectors now also contain information about the embedding coordinates for our images.
-%
-
-In general, the embedding coordinates are given by
-\begin{equation} \label{eq:vdm_coord}
-\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle,
-\end{equation}
-where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$,
-%
-If we assume that the rotations and the dynamics are uncoupled and therefore separable, then the eigenvectors of $S$ have the following structure: each block eigenvector contains estimates of the optimal rotations (up to a constant rotation) multiplied by the corresponding embedding coordinate (a scalar)
-%
-As the first diffusion maps coordinate is constant over the data, the first block eigenvector contains only the optimal rotations.
-%
-The second block eigenvector (eigenvectors $d+1$ through $2d$) contains the optimal rotations, each multiplied by their second diffusion maps coordinate.
-%
-We can therefore recover this diffusion maps coordinate by taking inner products of the columns of the second block eigenvector with columns of the first block eigenvector.
-%%
-%%YGK just as a favor to me, tell me how consistent the three values are -- 
-%%
-The $j^{th}$ embedding coordinate will be given by $\psi_{k,1}$, where $jd +1 < k \le (j+1)d$,
-and we select $k$ such that the coordinate $\psi_{k, 1}$ has the largest variability, i.e., the $j^{th}$ coordinate is $\psi_{k,1}$, where $k$ is the solution to
-$$ \max_{jd +1 < k \le (j+1)d} \sum_i \psi_{k,1} (i)^2. $$
-
-\subsection{Registering images with respect to rotations and translations} \label{subsec:trans_rot_register}
-
-Using angular synchronization or vector diffusion maps to register images with respect to rotations {\it and translations} requires some additional effort.
-%
-%We are not only interested in registering the one-dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two-dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\it and} translations.
-%
-%We first compute the translations and rotations to optimally align each pair of data points.
-%
-The first step (similar to the case of only rotations) is to compute the optimal alignments between pairs of images.
-%
-Practically, we have square images discretized as pixels (rather than continuous functions on the plane); note that the center of the
-image square is {\em not} necessarily also the center of the embryo.
-%
-For each image pair $I_i$ and $I_j$ we compute
-\begin{equation}\label{eq:opt_pairwise}
-(\theta_{ij}, dx_{ij}, dy_{ij}) = \argmin_{
-\begin{matrix}
-\theta \in [0, 2\pi) \\
-dx, dy \in [-\Delta, \Delta]\\
-%dy \in [-\Delta, \Delta]
-\end{matrix}
-} \|g(I_j, \theta, dx, dy) - I_i \|^2.
-\end{equation}
-where $g(I_j, \theta, dx, dy)$ is image $I_j$ rotated around the center of the square by $\theta$ radians, then translated by $dx$ pixel widths in the $x$ direction, and finally translated by $dy$ pixel widths in the $y$ direction.
-%
-The norm, $\| \cdot \|$, is the Euclidean norm between the pixel intensities of the channels.
-%
-The domain of the image (a square) is not invariant to our rotations and translations; however, the pixels near the border of the square have zero intensity, and so the norm can be meaningfully computed as long as the main image does not ``move out of'' the original square.
-%
-%The norm \eqref{eq:opt_pairwise} involve an integral over the square, which is not left invariant by our rotations and
-%translations; noting however that the pixels close to the border of the square have zero intensity allows a meaningful
-%computation of this norm as long as the ``main feature" (the embryo) does not move out of the original square (no ``corner effects").
-%
-This also justifies the $\Delta$ limits in \eqref{eq:opt_pairwise}: $\Delta$ is chosen such that the main image is not split or separated when translating (we take $\Delta=10$, which corresponds to a 10\% shift in the image).
-%; it does not make sense to shift the images so much that the embryo would ``move out" of the original square.
-%
-Image rotation is performed with the \texttt{imrotate} function in Matlab, using nearest neighbor interpolation to estimate the pixel intensities after rotation.
-%, since there is no direct mapping of the pixels of the unrotated image to the pixels of the rotated image.
-%
-The missing pixels in the corners of the rotated image are taken to have zero intensity.
-%
-The vertical and horizontal translations are implemented by shifting the pixels;
-we only consider shifts that correspond to an integer number of pixels to remove the need for interpolation when doing the translations.
-%
-The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
-%
-Therefore, instead of using an optimization procedure, we discretize the search space and exhaustively search to find the solution.
-%
-Our discrete translation steps are $\Delta/5$, and our discrete rotations steps are $\pi/10$.
-%
-%We discritize the search space of rotations into 20 possible rotations %($d\theta  \in \{0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10 \}$),
-%and 11 possible translations in both . %($dx, dy \in \{-20, -16, -12, \dots, 12, 16, 20 \}$).
-%
-%We check all possible combinations for the rotation and translations that best align $I_j$ to $I_i$.
-%
-Although computationally demanding, this ``embarrassingly parallelizable" direct enumeration approach is not prohibitive here.
-%
-%Although this can be somewhat time intensive, it is not prohibitive for the data sets we consider, and can be trivially parallelized if necessary.
-%
-%The solution will not be the exact solution to \eqref{eq:opt_pairwise}, but it will (most likely) be a close approximation.
-%
-%Since our techniques are robust to noise, close approximations will be sufficient to obtain accurate results.
-
-%As discussed above, for angular synchronization, the underlying symmetry group needs to have a real and orthogonal matrix representation.
-%
-%We therefore convert $ISO(2)$, the group of two-dimensional translations and rotations, to $SO(3)$, the group of three-dimensional rotations.
-%
-
-The synchronization formulation requires the matrices used to represent the symmetry group to be orthogonal \cite{singer2013spectral}.
-%(for \eqref{eq:H_low_rank} to be satisfied).
-%
-However, the typical matrix representation for $ISO(2)$, the group of two-dimensional translations and rotations, does not satisfy this property.
-%
-Instead, we {\it approximately} represent rotations and translations in two dimensions using {\it three-dimensional rotation matrices}, by projecting the (two-dimensional) image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular}.
-%
-Rotation of the image corresponds to rotation around one principal axis in three dimensions, while translation of the image corresponds (approximately) to rotations around the other two principal axes (see \fig~\ref{fig:SO3_picture}).
-(Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$.
-However, the images we will be considering are already ``mostly centered", and so we are only interested in {\em small} translations that are well within the $[0, 2\pi)$ range.)
-%
-These rotation matrices are (by definition) orthogonal, and successive application of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
-
-To convert the optimal two-dimensional translations and rotation to corresponding (continuous) three-dimensional rotations,
-we first compute the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$,
-\begin{equation} \label{eq:angle_relations}
-%\begin{aligned}
-%	\alpha_{ij} &= \theta_{ij} \\
-%	\beta_{ij} &= \frac{dx_{ij}}{n} \times \eta_{proj} \\
-%	\gamma_{ij} &= \frac{dy_{ij}}{n} \times \eta_{proj} \\
-%\end{aligned}
-	\alpha_{ij} = \theta_{ij}, \; \; \;
-	\beta_{ij} = \frac{dx_{ij}}{n} \times \eta_{proj}, \; \; \;
-	\gamma_{ij} = \frac{dy_{ij}}{n} \times \eta_{proj}
-\end{equation}
-where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project the image.
-%
-We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere;
-this portion of the sphere is small enough so that curvature effects are not prominent in the image.
-%
-We write rotations around the three principal axes, $R^x(\alpha)$, $R^y(\beta)$, and $R^z(\gamma)$, in terms of the three Euler angles
-%\small
-\begin{equation}
-\begin{aligned}
-	R^x(\alpha) &= \begin{bmatrix}
-	1 & 0 & 0 \\
-    0 & \cos(\alpha) & -\sin(\alpha) \\
-    0 & \sin(\alpha) & \cos(\alpha)
-	\end{bmatrix} \\
-	R^y(\beta) &= \begin{bmatrix}
-	\cos(\beta) & 0 & \sin(\beta) \\
-    0 & 1 & 0 \\
-    -\sin(\beta) & 0 & \cos(\beta)
-    \end{bmatrix} \\
-	R^z(\gamma) &= \begin{bmatrix}
-	\cos(\gamma) & -\sin(\gamma) & 0 \\
-    \sin(\gamma) & \cos(\gamma) & 0 \\
-    0 & 0 & 1
-    \end{bmatrix}.
-\end{aligned}
-\end{equation}
-%\normalsize
-%
-The total rotation, $R_{ij} \in SO(3) \subset \mathbb{R}^{3 \times 3}$, is
-\begin{equation} \label{eq:total_rot}
-	R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})	 = R^z(\gamma_{ij})  R^y(\beta_{ij})  R^x(\alpha_{ij}).
-%R(\alpha_{ij}, \beta_{ij}, \gamma_{ij}) =
-%\begin{bmatrix}
-%\cos(\beta) \cos(\gamma) & \cos(\gamma)\sin(\alpha)\sin(\beta)-\cos(\alpha) \sin(\gamma) & \cos(\alpha)\cos(\gamma)\sin(\beta)+ \sin(\alpha)\sin(\gamma) \\
-%\cos(\beta) \sin(\gamma) & \cos(\alpha) \cos(\gamma)+\sin(\alpha) \sin(\beta) \sin(\gamma) & \cos(\alpha) \sin(\beta) \sin(\gamma)-\cos(\gamma) \sin(\alpha) \\
-%-\sin(\beta) & \cos(\beta) \sin(\alpha) & \cos(\alpha) \cos(\beta))
-%\end{bmatrix}
-\end{equation}
-%
-%\eqref{eq:total_rot} corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).
-
-These rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$ are used in formulating the matrix $H$ in \eqref{eq:H_to_R} for angular synchronization, or the matrix $S$ in \eqref{eq:vdm_S} for vector diffusion maps to compute the optimal rotations (and, in the case of vector diffusion maps, deduce optimal temporal orderings) for the images.
-%
-The three-dimensional optimal rotations $R_{1, est}, \dots, R_{m, est}$ then give us the corresponding translations and rotations of the images.
-%
-We multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
-%
-$R_{i,est}^T$ then gives the rotation matrix which registers image $i$.
-%
-From a rotation matrix $R$, we compute the corresponding Euler angles $\alpha$, $\beta$, and $\gamma$ using the following relationships
-\begin{equation} \label{eq:angle_invert}
-\begin{aligned}
-R_{1,1} & = \cos(\beta)\cos(\gamma) &
-R_{2,1} & = \cos(\beta)\sin(\gamma) \\
-R_{3,1} & = -\sin(\beta) &
-R_{3,2} & = \sin(\alpha)\cos(\beta) \\
-R_{3,3} & = \cos(\alpha)\cos(\beta)  &
-\end{aligned}
-\end{equation}
-%
-Inverting \eqref{eq:angle_relations} yields the desired approximate translations and rotation, and we round optimal translations to the nearest integer to remove the need for interpolation when translating the images.
-
-\subsection{The eigenvalue spectrum}
-
-We can use the distribution of the eigenvalues from (vector) diffusion maps to help deduce the dimensionality of the data.
-%
-For diffusion maps, the largest eigenvalue will always be 1 and correspond to the trivial (constant) eigenvector.
-%
-The eigenvalues corresponding to noise in the data form a semicircle distribution centered at the origin (when the noise is a symmetric random matrix whose entries are drawn from a Gaussian distribution) \cite{wigner1955characteristic, wigner1958distribution}.
-%
-We expect the eigenvalues corresponding to descriptive/meaningful embedding coordinates to be positive and ``visibly removed" from the semicircle.
-%
-Counting such eigenvalues which are removed from the semicircle (excluding the trivial $\lambda=1$ eigenvalue) can help obtain an estimate of the dimensionality of the data (modulo harmonics in the same dimension \cite{ferguson2010systematic}).
-
-For vector diffusion maps, the same structure holds: we expect trivial eigenvalues, a semicircle corresponding to noise, and positive eigenvalues removed from the noise that correspond to meaningful embedding coordinates.
-%
-However, in vector diffusion maps, each coordinate gives rise to $d$ eigenvalues with (approximately) the same value (as each block eigenvector corresponds to $d$ eigenvalues).
-%
-We must then establish how many sets of $d$ eigenvalues are removed from the noise semicircle (excluding the trivial set of $d$ eigenvalues, all with  $\lambda=1$), to explore the dimensionality of the data.
-
-\subsection{Smooth trajectories from registered and ordered images}
-
-Once we have registered and ordered the images, we can smooth the resulting trajectory to obtain a ``stereotypic" developmental trajectory.
-%
-Let $x_1, \dots, x_m$ denote the set of {\em registered and ordered} images (so $x_i$ is the $i^{th}$ image in the ordered set).
-%
-We define the average image at time $\tau$, denoted $\overline{x}_{\tau}$ (where $1 \le \tau \le m$), as a (Gaussian) weighted average of the images,
-\begin{equation}
-\overline{x}_{\tau} =  \frac{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma^2} \right) x_i}{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma^2} \right) }
-\end{equation}
-where $\sigma$ is the scale of the Gaussian filter.
-%
-For the images in \fig~4, we take $\sigma^2 = 20$.
-%
-See \cite{kemelmacher2011exploring} for a more detailed discussion.
-
-\bibliographystyle{pnas2}
-\bibliography{background_reading/references,../../references/references}
-
-\end{article}
-
-\begin{figure}
-\includegraphics[width=8.4cm]{ap_dv}
-\caption{(Left) A lateral view of a {\it Drosophila} embryo stained with DAPI (gray), dpERK (red), and Dorsal (green).
-The embryo is presented so that the anterior (A) side is to the left and the posterior (P) side is to the right.
-The arrow indicates the position(s) where the cross-section of an embryo is imaged.
-(Right) A dorsoventral view of the cross-section of the {\it Drosophila} embryo.
-The dorsal (D) side is up and the ventral (V) side is down.
-For the first and third data sets (cellularization), images were collected at the focal plane $\sim 18\%$ from either the anterior or posterior pole of an embryo (arrows in the left image).
-During cellularization, it is known that the developmental dynamics are identical at these two regions \cite{...}.
-In the second data set (gastrulation), images were collected at the focal plane $\sim 18\%$ from the posterior pole of an embryo (right arrow in the left image).
-}
-\label{fig:ap_dv}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=8.4cm, trim=0cm 5.6cm 0cm 0cm, clip]{SI_fig6}
-\caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different times during development.
-The membrane length is indicated by the red bars.
-{\it (B)} Membrane length as a function of time.
-The points corresponding to the images in {\it A} are indicated with arrows.}
-\label{fig:membrane_compare}
-\end{figure}
-
-\begin{figure}
-\includegraphics[height=9cm]{raw_data3}
-\caption{Images of mutant and wild type {\it Drosophila} embryos during cellularization, used for the analysis in \fig~5.
-Each image is of a different embryo in a different rotational orientation.
-Half of the images are of wild type embryos, and half are of mutant embryos.}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=8.4cm]{SI_fig3}
-\caption{Synthetic data set used to illustrate the data processing algorithms.
-{\it (A)} One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
-{\it (B)} Intensity corresponding to the profile in {\it A}.
-{\it (C)} An ensemble of concentration profiles, each of the form described in {\it A}.
-Each row in the array corresponds to a single profile.
-{\it (D)} The profiles in {\it C}, now registered using angular synchronization.
-{\it (E)} The profiles in {\it D}, now temporally ordered using diffusion maps.
-{\it (F)} The profiles in {\it C}, registered and temporally ordered in a single step using vector diffusion maps.}
-\label{fig:1d_demo}
-%\customlabel{fig:1d_demo}{3}
-\customlabel{subfig:1d_example}{\ref{fig:1d_demo}{\it A}}
-\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}{\it B}}
-\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}{\it C}}
-\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}{\it D}}
-\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}{\it E}}
-\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}{\it F}}
-\end{figure}
-
-\begin{figure*}
-\begin{minipage}{0.22\textwidth}
 \centering
-%\includegraphics[width=\textwidth]{sphere_1}\\
-\includegraphics[width=\textwidth]{sphere2_1}\\
-\sffamily{A}
-\end{minipage}
-%
-\hfill
-%
-\begin{minipage}{0.22\textwidth}
-\centering
-%\includegraphics[width=\textwidth]{sphere_2}\\
-\includegraphics[width=\textwidth]{sphere2_2}\\
-\sffamily{B}
-\end{minipage}
-\begin{minipage}{0.22\textwidth}
-\centering
-%\includegraphics[width=\textwidth]{sphere_3}\\
-\includegraphics[width=\textwidth]{sphere2_3}\\
-\sffamily{C}
-\end{minipage}
-\begin{minipage}{0.22\textwidth}
-\centering
-%\includegraphics[width=\textwidth]{sphere_4}\\
-\includegraphics[width=\textwidth]{sphere2_4}\\
-\sffamily{D}
-\end{minipage}
-\caption{Illustration of the correspondence between three-dimensional rotations to translations and rotations in two dimensions.
-{\it (A)} The original image (shown in red), projected onto a portion of a sphere.
-{\it (B)} Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image.
-{\it (C)} Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation.
-{\it (D)} Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.}
-\label{fig:SO3_picture}
-\end{figure*}
-
-\begin{figure*}
 %\raisebox{4.5cm}{\sffamily{A}}
@@
 \end{tikzpicture}
-\caption{The histograms of the eigenvalues from vector diffusion maps analysis for the three data sets.
-%
-We expect to see an approximately symmetric semicircle distribution centered at 0, which corresponds to noise, and positive eigenvalues separated from the semicircle which correspond to meaningful coordinates.
-%
-Each coordinate yields $d=3$ eigenvalues of (approximately) the same magnitude, and the top $d=3$ eigenvalues ($\lambda=1$) correspond to a trivial embedding coordinate.
-%
-(A) Histogram of eigenvalues from images during cellularization (\fig~3).
-As expected, we see an approximately symmetric distribution that ranges from $-0.8$ to $0.8$ (indicated by the blue arrow), and $6$ eigenvalues that are removed from this distribution (indicated by the red arrow).
-$3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $3$ eigenvalues correspond to the first nontrivial VDM component which ``uncovers" the developmental dynamics; then there is an appreciable gap before the third group of eigenvalues.
-%
-(B) Histogram of eigenvalues from images during gastrulation (\fig~4).
-We see an approximately symmetric distribution that ranges from $-0.7$ to $0.7$ (indicated by the blue arrow), and $6$ eigenvalues that are removed from this distribution (indicated by the red arrow).
-$3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $3$ eigenvalues correspond to the first nontrivial VDM component which ``uncovers" the developmental dynamics.
-%
-(C) Histogram of eigenvalues from images of mutant and wild type embryos during cellularization (\fig~5).
-We see an approximately symmetric distribution that ranges from $-0.2$ to $0.2$ (indicated by the blue arrow), and $9$ eigenvalues that are removed from this distribution (indicated by the red arrows).
-$3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $6$ eigenvalues correspond to the first two VDM components which are now necessary to parametrize the developmental dynamics of the two genotypes.}
-%there is a spectral gap after the first (non-trivial) group of eigenvalues.
-%\caption{The eigenvalue spectra from vector diffusion maps analysis for the three data sets.
-%%
-%As expected, the eigenvalues appear in groups of 3, as $d=3$ when using vector diffusion maps to register with respect to translations and rotations.
-%%
-%(A) Spectrum from images during cellularization (\fig~3).
-As expected, there is a spectral gap after the first (non-trivial) group of eigenvalues: the first three eigenvalues correspond to the rotational and translational alignment, the second three eigenvalues correspond to the first VDM component which uncovers the developmental dynamics, and the then there is an appreciable gap before the third group of eigenvalues.
-%%
-%(B) Spectrum from images during gastrulation (\fig~4). Again,
-%there is a spectral gap after the first (non-trivial) group of eigenvalues.
-%%
-%(C) Spectrum from images of mutant and wild type embryos during cellularization (\fig~5).
-%We see a group of three eigenvalues corresponding to the translational and rotational alignments, one group that is clearly significant, and then a second group that, while less important, is still above the level of noise of the subsequent eigenvalues, indicating that two components are necessary to describe the data.
-}
-\label{fig:eval_spectra}
-\end{figure*}
-
-\begin{figure}
-\includegraphics[width=8.7cm]{PCA_ordered}
-\caption{Images of embryos during cellularization (from \fig~3), registered using angular synchronization \cite{singer2011angular} and ordered by the first PCA projection coefficient.
-The recovered dynamics are qualitatively consistent with those recovered by vector diffusion maps.
-}
-\label{fig:PCA_data1}
-\end{figure}
-
-\begin{figure*}
-\raisebox{5.2cm}{\sffamily{A}}
-\includegraphics[width=16.8cm]{data2_PCA_ordered}
-
-\raisebox{3.7cm}{\sffamily{B}}
-\includegraphics[width=8.4cm, angle=90]{data2_PCA_variance}
-\raisebox{3.7cm}{\sffamily{C}}
-\includegraphics[width=8.4cm]{data2_PCA_proj}
-
-\caption{{\it (A)} Images shown in \fig~4 of the paper, registered using angular synchronization \cite{singer2011angular} and ordered by the first PCA projection coefficient.
-The recovered ordering is not consistent with the known developmental dynamics.
-{\it (B)} The fraction of the variance captured by each of the first 10 principal components.
-Note that components 2--10 have a nonnegligible contribution to the data.
-{\it (C)} Projection of the data onto the first two principal components.
-Each principal component is normalized to 1.
-Both the projection coefficients and the projection coefficients scaled by the square roots of the corresponding eigenvalues are shown.
-}
-\label{fig:PCA_results}
-\end{figure*}
-
-%
-\begin{figure*}[t]
-{\bf Item C in the figure below is our main observation.}
-
-\raisebox{7.6cm}{\sffamily{A}}
-\includegraphics[width=8cm]{raw_data1}
-\hfill
-\raisebox{7.6cm}{\sffamily{B}}
-\includegraphics[width=8cm]{FBsPCA_evals}
-
-\raisebox{5.7cm}{\sffamily{C}}
-\includegraphics[width=16cm, trim=2cm 0cm 2cm 0cm, clip]{FBsPCA_dmaps}
-\caption{I computed the FBsPCA basis for the set of images (separately for the red and green signals).
-%
-I then computed the bispectrum coefficients for each image expanded in the FBsPCA basis, and used these coefficients in a DMAPS calculation.
-%
-The results are shown above.
-%
-However, (mainly, we suspect) because the bispectrum is not stable to deformations, it does not seem to be able to identify neighboring images.
-%
-(A) Images used for FBsPCA.
-(B) Eigenvalue spectrum from diffusion maps, computed using bispectrum coefficients.
-There is no visible spectral gap after the first component, as would be expected for this particular data set.
-(C, top) First 4 diffusion maps variables, computed from the bispectrum coefficients, plotted against the membrane thickness.
-Note there is no appreciable correlation: {\bf each diffusion maps variable simply selects one image from the data set as ``abnormal''.} (C, bottom) The corresponding images that each diffusion maps mode ``selects.'' In general, these seem to be the late images with ventral furrows beginning to form.
-}
-\end{figure*}
 

### 2014-07-10 image_analysis_paper/pnas_draft2.tex
Added live imaging movies

@@
 %
-When the two paths diverge, the wild type data develops two distinct peaks, whereas the mutant data develops a single peak.
+When the two paths diverge, the wild type data develops two distinct peaks, 
 %
-Therefore, vector diffusion maps can also help distinguish different expression patterns and developmental trajectories.
-
 \begin{figure}[h]
@@
 \end{figure}
+%
+\noindent whereas the mutant data develops a single peak.
+%
+Therefore, vector diffusion maps can also help distinguish different expression patterns and developmental trajectories.
 
@@
 %
-The task of image registration has also been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
+The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
 %
@@
 
-%%%AMIT
 Previously, angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy images \cite{singer2012vector, zhao2014rotationally, singer2011viewing}.
@@
 Here, we have used vector diffusion maps to register the images with respect to translations and rotations; however, it is conceivable that one could factor out other sources of variability, such as reflections \cite{singer2012vector, goemans1995improved, bandeira2013cheeger} or dilations.
-%In the future, functional data analysis tools can be used for the %quantitative estimation of both the average trajectory and its %variability \cite{silverman2005functional}.  
-%
-%One can also test which physical variable, among a set of candidate variables, is most correlated with the uncovered vector diffusion maps coordinate to gain insight into which physical quantities change across the developmental trajectory (e.g., total intensity of one signal, roundness of the embryo).
-%
 
@@
 Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.
-
-%Here, we have used vector diffusion maps to register the images with respect to translations and rotations; however, it is conceivable that one could factor out other sources of variability, such as reflections \cite{singer2012vector, goemans1995improved, bandeira2013cheeger} or dilations.
-%
-%One might even envision automated analysis of embryos imaged in arbitrary three-dimensional orientations, in the spirit of cryo-EM image analysis.
-%
-%%%AMIT
-%For these applications, vector diffusion maps is essential, as global registration of the entire imaging data set is topologically impossible \cite{zhao2014rotationally}.
-%
-%%%AMIT
-%In addition, we believe that algorithms based on rotation- and translation-invariant image features which are stable to small deformations, such as the scattering transform \cite{mallat2012group} or the bispectrum and steerable PCA \cite{zhao2014rotationally}, are especially promising; such methods circumvent the need for registration by using invariant features of the images in the data analysis.
-%
 
@@
 
-\bibitem{silverman2005functional}
-Ramsay J~O, Silverman B~W (2005) \textit{Functional Data Analysis}, Springer
-  Series in Statistics (Springer).
+%\bibitem{silverman2005functional}
+%Ramsay J~O, Silverman B~W (2005) \textit{Functional Data Analysis}, Springer
+%  Series in Statistics (Springer).
 

### 2014-07-10 image_analysis_paper/pnas_manuscript.tex
Added live imaging movies

--- 2014-07-10 image_analysis_paper/pnas_draft2.tex
@@
 
-\graphicspath{ {paper_figures2/} }
+%\graphicspath{ {paper_figures2/} }
 
-\usepackage{tikz}
+%\usepackage{tikz}
 
@@
 %
-To illustrate this approach, we temporally order and register several datasets collected in cross-sectional studies of cell signaling in the early fruit fly embryo.}
+To illustrate this approach, we temporally order and register several data sets collected in cross-sectional studies of cell signaling in the early fruit fly embryo.}
 
@@
 
-\begin{figure}[t]
-\includegraphics[width=8.7cm]{fig1}
-\caption{Caricature illustrating the tasks of image registration and temporal ordering.
+%\begin{figure}[t]
+%\includegraphics{fig1}
+%\caption{Caricature illustrating the tasks of image registration and temporal ordering.
 {\xfigtextfontit (A)} Images of ``samples'', each in a different orientation and a different stage of development.
@@
 For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
-%\label{fig:fish}
-\customlabel{fig:fish}{1}
-\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
-\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
-\end{figure}
+%%\label{fig:fish}
+%\customlabel{fig:fish}{1}
+%\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
+%\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
+%\end{figure}
 
@@
 
-%Both angular synchronization and diffusion maps use pairwise information to uncover global structure in the data.
-%
-%Angular synchronization uses pairwise alignment information to find translations and rotations which are globally consistent, and diffusion maps uses pairwise distances to find global parametrizations which are (geometrically) informative.
-%
-%Vector diffusion maps combines both steps into a single computation (see \SI), simultaneously addressing the tasks of registration and temporal ordering.
-%
-
-\begin{figure}[t]
-\includegraphics[width=8.7cm]{fig2}
-\caption{Schematic illustrating angular synchronization and diffusion maps.
+%\begin{figure}[t]
+%\includegraphics{fig2}
+%\caption{Schematic illustrating angular synchronization and diffusion maps.
 {\xfigtextfontit (A)} Set of vectors, each in a different orientation.
@@
 The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
-%\label{fig:schematics}
-\customlabel{fig:schematics}{2}
-\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
-\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
-\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
-\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
-\end{figure}
+%%\label{fig:schematics}
+%\customlabel{fig:schematics}{2}
+%\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
+%\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
+%\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
+%\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
+%\end{figure}
 
-\begin{figure*}[t]
-\includegraphics{fig3}
-\caption{{\xfigtextfontit Drosophila} embryos during cellularization.
+%\begin{figure*}[t]
+%\includegraphics{fig3}
+%\caption{{\xfigtextfontit Drosophila} embryos during cellularization.
 {\xfigtextfontit (A)} Images of embryos collected during cellularization.
@@
 The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.}
-\customlabel{fig:data1}{3}
-\customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
-\customlabel{subfig:ordered_data1}{\ref{fig:data1}{\it B}}
-\customlabel{subfig:averaged_data1}{\ref{fig:data1}{\it C}}
-\end{figure*}
+%\customlabel{fig:data1}{3}
+%\customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
+%\customlabel{subfig:ordered_data1}{\ref{fig:data1}{\it B}}
+%\customlabel{subfig:averaged_data1}{\ref{fig:data1}{\it C}}
+%\end{figure*}
 
@@
 
-\begin{figure*}[t]
-\includegraphics{fig4}
-\caption{{\xfigtextfontit Drosophila} embryos during gastrulation.
+%\begin{figure*}[t]
+%\includegraphics{fig4}
+%\caption{{\xfigtextfontit Drosophila} embryos during gastrulation.
 {\xfigtextfontit (A)} Images of gastrulating embryos.
@@
 }
-\customlabel{fig:data2}{4}
-\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
-\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
-\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
-\end{figure*}
+%\customlabel{fig:data2}{4}
+%\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
+%\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
+%\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
+%\end{figure*}
 
@@
 \fig~\ref{subfig:projection_data3} shows the data projected onto the first two vector diffusion maps coordinates, where each point in this projection corresponds to one image in our data set.
-%%%
-%%%YGK we need something about dmap eignvalues and how many are not harmonics
 %
@@
 %
-\begin{figure}[h]
-\includegraphics{fig5_2}
-\caption{Wild type and mutant {\xfigtextfontit Drosophila} embryos.
+%\begin{figure}[h]
+%\includegraphics{fig5}
+%\caption{Wild type and mutant {\xfigtextfontit Drosophila} embryos.
 {\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~\ref{fig:raw_data3}), projected onto the first two vector diffusion maps coordinates.
@@
 }
-\customlabel{fig:data3}{5}
-\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
-\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it B}}
-\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
-\end{figure}
+%\customlabel{fig:data3}{5}
+%\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
+%\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it B}}
+%\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
+%\end{figure}
 %
@@
 Ip Y~T, Park R~E, Kosman D, Bier E, Levine M (1992) 
-The dorsal gradient morphogen regulates stripes of rhomboid expression in the presumptive neuroectoderm of the \textit{Drosophila} embryo. \textit{Genes Dev} 6(9):1728-39.
+The dorsal gradient morphogen regulates stripes of rhomboid expression in the presumptive neuroectoderm of the \textit{Drosophila} embryo. \textit{Genes Dev} 6(9):1728--1739.
 
@@
 
+\begin{figure}[t]
+\includegraphics{fig1}
+\caption{Caricature illustrating the tasks of image registration and temporal ordering.
+{\xfigtextfontit (A)} Images of ``samples'', each in a different orientation and a different stage of development.
+{\xfigtextfontit (B)} Registered and ordered samples.
+For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
+%\label{fig:fish}
+\customlabel{fig:fish}{1}
+\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
+\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
+\end{figure}
+
+\begin{figure}[t]
+\includegraphics{fig2}
+\caption{Schematic illustrating angular synchronization and diffusion maps.
+{\xfigtextfontit (A)} Set of vectors, each in a different orientation.
+The pairwise alignment angles are indicated.
+{\xfigtextfontit (B)} The vectors from {\xfigtextfontit A}, each rotated about their midpoint so that the set is globally aligned.
+Note that the chosen rotation angles are consistent with the pairwise alignments in {\xfigtextfontit A}: the differences between a pair of angles in {\xfigtextfontit B} is the same as the pairwise angle in {\xfigtextfontit A}.
+{\xfigtextfontit (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
+Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\xfigtextfontit \SI}), so that close data points are connected by darker (``stronger'') edges.
+{\xfigtextfontit (D)} The data in {\xfigtextfontit C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure.
+The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
+%\label{fig:schematics}
+\customlabel{fig:schematics}{2}
+\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
+\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
+\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
+\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
+\end{figure}
+
+\begin{figure*}[t]
+\includegraphics{fig3}
+\caption{{\xfigtextfontit Drosophila} embryos during cellularization.
+{\xfigtextfontit (A)} Images of embryos collected during cellularization.
+Each image is of a different embryo arrested at a different developmental time, and in a different rotational and translational orientation.
+{\xfigtextfontit (B)} Images from {\xfigtextfontit A}, now registered and ordered using vector diffusion maps.
+The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right.
+The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.}
+\customlabel{fig:data1}{3}
+\customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
+\customlabel{subfig:ordered_data1}{\ref{fig:data1}{\it B}}
+\customlabel{subfig:averaged_data1}{\ref{fig:data1}{\it C}}
+\end{figure*}
+
+\begin{figure*}[t]
+\includegraphics{fig4}
+\caption{{\xfigtextfontit Drosophila} embryos during gastrulation.
+{\xfigtextfontit (A)} Images of gastrulating embryos.
+Each image is of a different embryo arrested at a different developmental time, and in a different translation and rotational orientation.
+{\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps.
+The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right.
+{\xfigtextfontit (C)} A representative ``developmental path'' from local averaging of the images.
+Each image is a (Gaussian-weighted) average of the images in {\xfigtextfontit B}.
+}
+\customlabel{fig:data2}{4}
+\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
+\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
+\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
+\end{figure*}
+
+\begin{figure}[h]
+\includegraphics{fig5}
+\caption{Wild type and mutant {\xfigtextfontit Drosophila} embryos.
+{\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~\ref{fig:raw_data3}), projected onto the first two vector diffusion maps coordinates.
+The directions of the developmental trajectories are indicated by arrows.
+(Inset) The projected data, now colored by genotype.
+The two branches separate mutant and wild type samples.
+{\xfigtextfontit (B)} dpERK signal at select points along the wild type trajectory in {\xfigtextfontit A} (denoted by circles in the inset).
+The two dpERK peaks which arise during development are indicated by arrows.
+{\xfigtextfontit (C)} dpERK signal at select points along the mutant trajectory in {\xfigtextfontit A} (denoted by circles in the inset).
+The single dpERK peak is indicated by an arrow.
+}
+\customlabel{fig:data3}{5}
+\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
+\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it B}}
+\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
+\end{figure}
+
 \end{document}

### 2014-08-19 image_analysis_paper2/pnas_draft.tex
added more for paper edits

--- 2014-07-10 image_analysis_paper/pnas_manuscript.tex
@@
 
-%\graphicspath{ {paper_figures2/} }
+\graphicspath{ {paper_figures2/} }
 
@@
 
-\significancetext{Cross-sectional imaging studies are commonly used to study developmental dynamics in multiple organisms. 
-%
-In these studies, a representative developmental trajectory is reconstructed from snapshots of different embryos, each of which has been arrested at a different point of its development. 
-%
-One of the first steps in this reconstruction involves registration and temporal ordering of images. 
-%
-Here we show that both of these tasks can be automated and combined in a single step using recent advances in data mining. 
-%
-To illustrate this approach, we temporally order and register several data sets collected in cross-sectional studies of cell signaling in the early fruit fly embryo.}
+%\significancetext{Cross-sectional imaging studies are commonly used to study developmental dynamics in multiple organisms. 
+%%
+%In these studies, a representative developmental trajectory is reconstructed from snapshots of different embryos, each of which has been arrested at a different point of its development. 
+%%
+%One of the first steps in this reconstruction involves registration and temporal ordering of images. 
+%%
+%Here we show that both of these tasks can be automated and combined in a single step using recent advances in data mining. 
+%%
+%To illustrate this approach, we temporally order and register several data sets collected in cross-sectional studies of cell signaling in the early fruit fly embryo.}
 
@@
 \keywords{temporal ordering | image registration | vector diffusion maps}
+
+\section{Significance}
+Cross-sectional imaging studies are commonly used to study developmental dynamics in multiple organisms. 
+%
+In these studies, a representative developmental trajectory is reconstructed from snapshots of different embryos, each of which has been arrested at a different point of its development. 
+%
+One of the first steps in this reconstruction involves registration and temporal ordering of images. 
+%
+Here we show that both of these tasks can be automated and combined in a single step using recent advances in data mining. 
+%
+To illustrate this approach, we temporally order and register several data sets collected in cross-sectional studies of cell signaling in the early fruit fly embryo.
+\vspace{0.5cm}
 
@@
 
-%\begin{figure}[t]
-%\includegraphics{fig1}
-%\caption{Caricature illustrating the tasks of image registration and temporal ordering.
+\begin{figure}[t]
+\includegraphics{fig1}
+\caption{Caricature illustrating the tasks of image registration and temporal ordering.
 {\xfigtextfontit (A)} Images of ``samples'', each in a different orientation and a different stage of development.
@@
 For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
-%%\label{fig:fish}
-%\customlabel{fig:fish}{1}
-%\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
-%\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
-%\end{figure}
+%\label{fig:fish}
+\customlabel{fig:fish}{1}
+\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
+\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
+\end{figure}
 
@@
 
-%\begin{figure}[t]
-%\includegraphics{fig2}
-%\caption{Schematic illustrating angular synchronization and diffusion maps.
+\begin{figure}[t]
+\includegraphics{fig2}
+\caption{Schematic illustrating angular synchronization and diffusion maps.
 {\xfigtextfontit (A)} Set of vectors, each in a different orientation.
@@
 The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
-%%\label{fig:schematics}
-%\customlabel{fig:schematics}{2}
-%\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
-%\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
-%\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
-%\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
-%\end{figure}
+%\label{fig:schematics}
+\customlabel{fig:schematics}{2}
+\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
+\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
+\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
+\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
+\end{figure}
 
-%\begin{figure*}[t]
-%\includegraphics{fig3}
-%\caption{{\xfigtextfontit Drosophila} embryos during cellularization.
+\begin{figure*}[t]
+\includegraphics{fig3}
+\caption{{\xfigtextfontit Drosophila} embryos during cellularization.
 {\xfigtextfontit (A)} Images of embryos collected during cellularization.
@@
 The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.}
-%\customlabel{fig:data1}{3}
-%\customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
-%\customlabel{subfig:ordered_data1}{\ref{fig:data1}{\it B}}
-%\customlabel{subfig:averaged_data1}{\ref{fig:data1}{\it C}}
-%\end{figure*}
+\customlabel{fig:data1}{3}
+\customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
+\customlabel{subfig:ordered_data1}{\ref{fig:data1}{\it B}}
+\customlabel{subfig:averaged_data1}{\ref{fig:data1}{\it C}}
+\end{figure*}
 
-\subsection{Images of embryos during cellularization}
+\subsection{Ordering Images From a Movie}
 
@@
 
-%\begin{figure*}[t]
-%\includegraphics{fig4}
-%\caption{{\xfigtextfontit Drosophila} embryos during gastrulation.
+\begin{figure*}[t]
+\includegraphics{fig4}
+\caption{{\xfigtextfontit Drosophila} embryos during gastrulation.
 {\xfigtextfontit (A)} Images of gastrulating embryos.
@@
 }
-%\customlabel{fig:data2}{4}
-%\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
-%\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
-%\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
-%\end{figure*}
+\customlabel{fig:data2}{4}
+\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
+\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
+\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
+\end{figure*}
 
@@
 
-\subsection{Identifying distinct developmental trajectories}
-
-When time is the main source of variability, we have shown that vector diffusion maps can recover developmental progression from cross-sectional data.
-%
-However, vector diffusion maps can also provide an organization of data which contains multiple sources of variability.
-%
-To illustrate this point, we collected a data set consisting of $42$ embryos undergoing cellularization during the third hour of development (see \fig~\ref{fig:raw_data3}).
-%
-Half of the embryos are wild type, and half of the embryos are mutants where dpERK is expressed in one broad peak at the ventral side \cite{snail}.
-%
-Again, each image is an optical cross-section of an embryo fixed at a different developmental time, and the embryos have been stained for nuclei (gray), Dl (green), and dpERK (red).
-%
-We expect two sources of variability in this data set: differences due to developmental time, and differences between mutant and wild type dynamics.
-
-We computed the vector diffusion maps embedding for our data, and the eigenvalue distribution (see \fig~\ref{fig:eigenvalues}) indicates that two coordinates may be required for a meaningful description of the data.
-%
-\fig~\ref{subfig:projection_data3} shows the data projected onto the first two vector diffusion maps coordinates, where each point in this projection corresponds to one image in our data set.
-%
-Two distinct branches appear to emerge in this two-dimensional embedding space.
-
-The inset of \fig~\ref{subfig:projection_data3} shows the two-dimensional projection, with the data points now colored to indicate whether they correspond to wild type or mutant embryos.
-%
-The two branches in the embedding help distinguish the two genotypes.
-%
-Selected images along the wild type and mutant trajectories are shown in \fig~\ref{subfig:average_wt_data3} and \fig~\ref{subfig:average_mut_data3}, respectively.
-%
-The portion of the embedding where the two genotypes are intermixed (lower-left region of the plot) contains embryos which are early in their developmental trajectory, where there is little dpERK signaling and the two groups of embryos are indistinguishable.
-%
-When the two paths diverge, the wild type data develops two distinct peaks, 
-%
-%\begin{figure}[h]
-%\includegraphics{fig5}
-%\caption{Wild type and mutant {\xfigtextfontit Drosophila} embryos.
-{\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~\ref{fig:raw_data3}), projected onto the first two vector diffusion maps coordinates.
-The directions of the developmental trajectories are indicated by arrows.
-(Inset) The projected data, now colored by genotype.
-The two branches separate mutant and wild type samples.
-{\xfigtextfontit (B)} dpERK signal at select points along the wild type trajectory in {\xfigtextfontit A} (denoted by circles in the inset).
-The two dpERK peaks which arise during development are indicated by arrows.
-{\xfigtextfontit (C)} dpERK signal at select points along the mutant trajectory in {\xfigtextfontit A} (denoted by circles in the inset).
-The single dpERK peak is indicated by an arrow.
-}
-%\customlabel{fig:data3}{5}
-%\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
-%\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it B}}
-%\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
-%\end{figure}
-%
-\noindent whereas the mutant data develops a single peak.
-%
-Therefore, vector diffusion maps can also help distinguish different expression patterns and developmental trajectories.
-
 \section{Discussion}
@@
 %
-The approach presented here is different because it addresses image registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace; vector diffusion maps can also register images when global registration of the entire data set is topologically impossible \cite{zhao2014rotationally}.
+The approach presented here is different because it addresses image registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace; vector diffusion maps can also order images when global registration of the entire data set is topologically impossible \cite{zhao2014rotationally}.
 %
@@
 
-\begin{figure}[t]
-\includegraphics{fig1}
-\caption{Caricature illustrating the tasks of image registration and temporal ordering.
+%\begin{figure}[t]
+%\includegraphics{fig1}
+%\caption{Caricature illustrating the tasks of image registration and temporal ordering.
 {\xfigtextfontit (A)} Images of ``samples'', each in a different orientation and a different stage of development.
@@
 %\label{fig:fish}
-\customlabel{fig:fish}{1}
-\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
-\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
-\end{figure}
-
-\begin{figure}[t]
-\includegraphics{fig2}
-\caption{Schematic illustrating angular synchronization and diffusion maps.
+%\customlabel{fig:fish}{1}
+%\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
+%\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
+%\end{figure}
+%
+%\begin{figure}[t]
+%\includegraphics{fig2}
+%\caption{Schematic illustrating angular synchronization and diffusion maps.
 {\xfigtextfontit (A)} Set of vectors, each in a different orientation.
@@
 %\label{fig:schematics}
-\customlabel{fig:schematics}{2}
-\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
-\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
-\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
-\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
-\end{figure}
-
-\begin{figure*}[t]
-\includegraphics{fig3}
-\caption{{\xfigtextfontit Drosophila} embryos during cellularization.
+%\customlabel{fig:schematics}{2}
+%\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
+%\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
+%\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
+%\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
+%\end{figure}
+%
+%\begin{figure*}[t]
+%\includegraphics{fig3}
+%\caption{{\xfigtextfontit Drosophila} embryos during cellularization.
 {\xfigtextfontit (A)} Images of embryos collected during cellularization.
@@
 The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.}
-\customlabel{fig:data1}{3}
-\customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
-\customlabel{subfig:ordered_data1}{\ref{fig:data1}{\it B}}
-\customlabel{subfig:averaged_data1}{\ref{fig:data1}{\it C}}
-\end{figure*}
-
-\begin{figure*}[t]
-\includegraphics{fig4}
-\caption{{\xfigtextfontit Drosophila} embryos during gastrulation.
+%\customlabel{fig:data1}{3}
+%\customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
+%\customlabel{subfig:ordered_data1}{\ref{fig:data1}{\it B}}
+%\customlabel{subfig:averaged_data1}{\ref{fig:data1}{\it C}}
+%\end{figure*}
+%
+%\begin{figure*}[t]
+%\includegraphics{fig4}
+%\caption{{\xfigtextfontit Drosophila} embryos during gastrulation.
 {\xfigtextfontit (A)} Images of gastrulating embryos.
@@
 }
-\customlabel{fig:data2}{4}
-\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
-\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
-\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
-\end{figure*}
-
-\begin{figure}[h]
-\includegraphics{fig5}
-\caption{Wild type and mutant {\xfigtextfontit Drosophila} embryos.
+%\customlabel{fig:data2}{4}
+%\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
+%\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
+%\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
+%\end{figure*}
+%
+%\begin{figure}[h]
+%\includegraphics{fig5}
+%\caption{Wild type and mutant {\xfigtextfontit Drosophila} embryos.
 {\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~\ref{fig:raw_data3}), projected onto the first two vector diffusion maps coordinates.
@@
 }
-\customlabel{fig:data3}{5}
-\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
-\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it B}}
-\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
-\end{figure}
+%\customlabel{fig:data3}{5}
+%\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
+%\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it B}}
+%\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
+%\end{figure}
 

### 2014-08-19 image_analysis_paper2/compute_pairwise_alignments.m

@@ -30,10 +30,11 @@ RI = imref2d([npixels npixels],[-1 1],[-1 1]);
 
 h_waitbar = waitbar(0,'Computing pairwise alignments...');
 for i=1:nrot
-    waitbar(i/nrot, h_waitbar);
+%     waitbar(i/nrot, h_waitbar);
     theta = theta_vec(i);
     for j=1:nshifts
         for k=1:nshifts
+            waitbar(((i-1)*nshifts^2+(j-1)*nshifts+k)/(nrot*nshifts^2), h_waitbar);
             A = affine2d([cosd(theta) -sind(theta) 0; sind(theta) cosd(theta) 0; shifts(j) shifts(k) 1]);
             images_transformed = imwarp(images, RI, A, 'outputview', RI);
             dist_tmp = pdist2(reshape(double(images), [], m)', reshape(double(images_transformed), [], m)');

### 2014-08-19 image_analysis_paper2/crop_image2.m

@@ -1,0 +1,26 @@
+function image2 = crop_image2(image)
+
+edge_tol = 2;
+detection_type = 'canny'; 
+
+npixels = size(image, 1);
+
+if ndims(image) == 3
+    im_tmp = edge(imlincomb(1, image(:, :, 1), 1, image(:, :, 2), 1, image(:, :, 3)), detection_type);
+else
+    im_tmp = edge(image, detection_type);
+end
+
+row_sum = sum(im_tmp);
+idx1 = find(row_sum > edge_tol, 1, 'first');
+idx2 = find(row_sum > edge_tol, 1, 'last');
+
+col_sum = sum(im_tmp, 2);
+idx3 = find(col_sum > edge_tol, 1, 'first');
+idx4 = find(col_sum > edge_tol, 1, 'last');
+
+if ndims(image) == 3
+    image2 = imresize(image(idx3:idx4, idx1:idx2,:), [npixels npixels]);
+elseif ndims(image) == 2
+    image2 = imresize(image(idx3:idx4, idx1:idx2), [npixels npixels]);    
+end

### 2014-08-19 image_analysis_paper2/make_figures_onemovie.m

@@ -23,6 +23,9 @@ channel = 1;
 
 k_print_fig = 2;
 
+theta_err = zeros(nmovies, 1);
+rank_corr = zeros(nmovies, 1);
+
 %%
 for k=1:nmovies
     load(sprintf('movie%d.mat', k));
@@ -73,7 +76,7 @@ for k=1:nmovies
         theta_opt(i) = atan2d(R_tmp(2,1), R_tmp(1,1));
     end
     fprintf('movie %d, average error = %2.2f \n', k, std(mod(theta - theta_opt, 360)))
-    
+    theta_err(k) =  std(mod(theta - theta_opt, 360));
     
     %
     if corr(time, embed_coord(:,1)) < 0
@@ -81,25 +84,33 @@ for k=1:nmovies
     end
     
     fprintf('movie %d, rank corr = %2.4f \n', k, corr(compute_ranks(time), compute_ranks(embed_coord(:,1))))
+    rank_corr(k) = corr(compute_ranks(time), compute_ranks(embed_coord(:,1)));
     
     if k == k_print_fig
         [~, max_idx] = max(mod(theta, 360));
-        make_fig(8, 6);
-        plot(mod(theta, 360), mod(theta_opt-theta_opt(max_idx), 360), '.')
-        xlabel('true angle (degrees)')
-        ylabel('recovered angle (degrees)')
+        make_fig(4,4);
+        plot(mod(theta, 360), mod(theta_opt-theta_opt(max_idx)-1, 360)+1, '.')
+        xlabel('true angle')
+        ylabel('recovered angle')
+        set(gca, 'xtick', [0 180 360]);
+        set(gca, 'ytick', [0 180 360]);
+        set(gca, 'xticklabel', {'0°'; '180°'; '360°'});
+        set(gca, 'yticklabel', {'0°'; '180°'; '360°'});
         saveas(gcf, 'angle_corr.pdf');
-        
-        make_fig(8, 6);
+         
+        make_fig(4,4);
         plot(compute_ranks(time), compute_ranks(embed_coord(:,1)),'.')
-        xlabel('rank from time')
-        ylabel('rank from vdm')
+        xlabel('true rank')
+        ylabel('recovered rank')
+        set(gca, 'xtick', [0 20 40])
+        set(gca, 'ytick', [0 20 40])
+        axis([0 50 0 50])
         saveas(gcf, 'rank_corr.pdf');
         
[…]

### 2014-08-20 image_analysis_paper2/pnas_draft.tex
pnas paper edits

@@
 The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.}
-\customlabel{fig:data1}{3}
-\customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
-\customlabel{subfig:ordered_data1}{\ref{fig:data1}{\it B}}
-\customlabel{subfig:averaged_data1}{\ref{fig:data1}{\it C}}
+\customlabel{fig:movie}{3}
+\customlabel{subfig:movie_unregistered_unordered}{\ref{fig:movie}{\it A}}
+\customlabel{subfig:movie_registered_ordered}{\ref{fig:movie}{\it B}}
+\customlabel{subfig:movie_angle_corr}{\ref{fig:movie}{\it C}}
+\customlabel{subfig:movie_rank_corr}{\ref{fig:movie}{\it D}}
+\customlabel{subfig:movie_error_table}{\ref{fig:movie}{\it E}}
+\customlabel{subfig:movie_bootstrap}{\ref{fig:movie}{\it F}}
 \end{figure*}
 
-\subsection{Ordering Images From a Movie}
+\subsection{Method validation using a live movie}
 
@@
 %
-This data set is obtained by fluorescent imaging of {\it Drosophila} embryos during the third hour after fertilization.
+This data set is obtained by live imaging of a single {\it Drosophila} embryo during the thirty minute time interval spanning late cellularization through gastrulation.
 %
-By this stage of development, sequential divisions of the fertilized egg have generated a system where most of the nuclei are arranged in a monolayer under the common plasma membrane \cite{foe1983studies}.
+During this time window the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, bringing in the future muscle cells and eventually forming a characteristic ``omega'' shape.
 %
-\fig~\ref{subfig:raw_data1} shows optical cross-sections of each embryo, with nuclei (gray in the figure) labeled by DAPI, a DNA stain.
+Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
 %
-During the third hour of development, this arrangement of nuclei is patterned by the concentration profiles of chemical signals that control cell differentiation.
+At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing similar ``omega'' shaped patterns to appear on the dorsal side.
+
+The movie shows an optical cross-section of a vertically oriented developing embryo, where the nuclei are labeled by DAPI, a DNA stain.
 %
-One of these signals is provided by the nuclear localization of the transcription factor Dorsal (Dl, green in the figure), which subdivides the early embryo into the territories that give rise to the muscle, nerve, and skin tissues.
+The movie contains $42$ frames, taken at $30$~sec time intervals;
+\fig~\ref{subfig:movie_unregistered_unordered} shows selected frames from this movie. 
 %
-Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, red in the figure), an enzyme that specifies a subset of neuronal cells.
+The gray nuclear signal has been blurred to obtain a smoother images.
+%
+Each frame has been arbitrarily rotated, and the order of the frames has been scrambled.
 
-Throughout the third hour of development, the nuclear localization of Dl is unimodal and peaked at the ventral side of the embryo (reviewed in \cite{rushlow2012temporal}).
+We used vector diffusion maps to register and order the images. 
 %
-The spatiotemporal pattern of dpERK is more complex: dpERK is first detected in two lateral stripes, which increase in intensity towards the end of the third hour of development \cite{lim2013kinetics}.
+\fig~\ref{subfig:movie_registered_ordered} shows the same selected frames, now registered and ordered.
 %
-This bimodal pattern is then augmented by a peak at the dorsal side of the embryo.
+The real time for each frame is indicated.
 %
-Both of these features become readily apparent when images are registered and ordered using vector diffusion maps (see \fig~\ref{subfig:ordered_data1}).
+\fig~\ref{subfig:movie_angle_corr} and \fig~\ref{subfig:movie_rank_corr}  shows the correlations between the recovered and true angles and ranks for each of the frames.
 %
-The peak of nuclear Dl is consistent throughout the images and the intensity of the dpERK signal in the lateral regions of the embryo increases monotonically across the data set.
+Both the angles and the ranks are recovered to a high degree of accuracy.
+
+To assess the robustness of our proposed methodology, we repeated this procedure with four additional movies.
 %
+The results are shown in \fig~\ref{subfig:movie_error_table}. 
+%
+The errors in the recovered angles are all very low, and the rank correlation coefficients are consistently close to 1, indicating that our methodology can reproducably order data of this type.
 
-For this set of images, the quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the ordering obtained through an independent approach.
+To assess how the quality of our results depends on the number of data points, we used bootstrap sampling. 
 %
-During the third hour of development, the embryo cellularizes: in this process, nuclei located under a common plasma membrane are enclosed by lateral membranes which grow inward, separating the nuclei into individual cells.
+\fig~\ref{subfig:movie_bootstrap} shows the average rank correlation coefficient as a function of the number of images in the data set.
 %
-The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo to within 2--3 minutes \cite{figard2013plasma} (see \fig~\ref{fig:membrane_curve}).
+The data sets were sampled (with replacement) from the original $40$-frame movie.
 %
-The Spearman rank correlation coefficient between the orderings obtained based on the progress of cellularization and vector diffusion maps is 0.91, indicating that vector diffusion maps can accurately recover the temporal order.
+There is a steady increase in the accuracy of the recovered orderings with an increasing number of images.
 %
-Thus, we have validated a dimensionality reduction approach to temporal ordering and registration on a data set where image orientation and ordering can be assessed independently.
+The recovered angles had errors of less than $10^{\circ}$, and the magnitude of the errors did not vary significantly with the size of the data set.
 
@@
 
-Because the dynamics of the data set shown in \fig~\ref{fig:data1} are relatively simple, effectively consisting of a single mode (the two dpERK peaks) growing in time, established techniques such as principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data in time (see \fig~\ref{fig:PCA_data1}).
+We will now demonstrate the utility of our methodogies on a data set where the correct ordering is not known {\em a priori}.
 %
-However, developmental dynamics are often significantly more complex, and nonlinear dimensionality reduction techniques are required to extract meaningful structure.
+\fig~\ref{subfig:fixed_images_unregistered_unordered} shows a set of $126$ images, which cover a thirty minute time interval spanning late cellularization through gastrulation.
 %
-As an example, \fig~\ref{subfig:raw_data2} shows a set of $108$ images, which cover a thirty minute time interval spanning late cellularization through gastrulation.
-%
-During this time, the arrangement of newly formed cells begins to deform into structures which will form the future organs \cite{leptin2005gastrulation}.
-%
-As before, each image is an optical cross-section of a vertically oriented embryo fixed at a different developmental time.
+Each image is an optical cross-section of a {\em different} vertically oriented embryo fixed at a different (and unknown) developmental time.
 %
@@
 %
-However, during this developmental time period, the nuclei no longer remain uniformly arranged around the periphery of the embryo.
+Embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor which specifies the cells of the future muscle tissue.
 %
-Instead of Dl, which is difficult to visualize at this point, embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor that is activated by high levels of Dl, specifying the cells of the future muscle tissue.
-%
-dpERK is again shown in red.
-
-%
-During this time window the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, bringing in the future muscle cells \cite{leptin2005gastrulation}.
-%
-dpERK appears as two lateral peaks at the ventral side of the embryo.
-%
-These two peaks merge together during invagination, eventually forming (together with Twi) a characteristic ``omega'' shape.
-%
-Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
-%
-At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing similar ``omega'' shaped patterns to appear on the dorsal side; these patterns are most readily seen in the last image of \fig~\ref{subfig:averaged_data2}.
+Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, red in the figure), an enzyme that specifies a subset of neuronal cells.
 
@@
 \customlabel{fig:data2}{4}
-\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
-\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
-\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
+\customlabel{subfig:fixed_images_unregistered_unordered}{\ref{fig:data2}{\it A}}
+\customlabel{subfig:fixed_images_registered_ordered}{\ref{fig:data2}{\it B}}
+\customlabel{subfig:fixed_images_average_trajectory}{\ref{fig:data2}{\it C}}
 \end{figure*}
 
-Visually, the dynamics in this data set are much more complicated, and only 22\% of the variability in the images is captured by the first principal component.
-%
-We found that simple time-ordering based on this linear projection is therefore not possible (see \fig~\ref{fig:PCA_data2}).
-%
+FIX!!
+
 \fig~\ref{subfig:ordered_data2} shows the images registered and ordered using vector diffusion maps \cite{singer2012vector}, 
@@
 
+%
+dpERK appears as two lateral peaks at the ventral side of the embryo.
+%
+These two peaks merge together during invagination, eventually forming (together with Twi) a characteristic ``omega'' shape.
+%
+Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
+%
+At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing similar ``omega'' shaped patterns to appear on the dorsal side; these patterns are most readily seen in the last image of \fig~\ref{subfig:fixed_images_average_trajectory}.
+
 \section{Discussion}
 
+\begin{itemize}
+\item High--throughput imaging is becoming increasingly common
+\item We are/wil reach the point where imaging data is being produced more rapidly than it can be analyzed by hand (similar to microarray data)
+\item We need a way to automate the organization of such data
+\item Our algorithms register, which is important for many imaging applications
+\item In general our algorithms find good coordinates/descriptors for the data
+\item We focus on uncovering a single coordinate that can be used to order data
+\item in general, can uncover many coordinates; use coordinates to do other data processing tasks (clustering, model fitting, etc.)
+
+\end{itemize}
 We presented a unified approach to temporal ordering and registration of images in cross-sectional studies of developmental dynamics. 
@@
 In contrast to most of the existing approaches to image registration which rely on the knowledge about some landmarks in the data \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information.
+
+We are approaching the time when the rate of imaging data collection will surpass the rate of manual image analysis. 
+%
+We therefore require automated methodologies to organize such large data sets; these methodogies should be sufficiently general to address a wide variety of applications. 
+%
+Vector diffusion maps allows us to automatically register images, which is an essential task for many applications.
+%
+Simultaneously the algorithm provides us with a parameterization or coordinates for each image.
+%
+In the examples presented here, we have focued on recovering a single coordinate, and using this coordinate to order the images, effectively ordering them in time.
+%
+In general, we can recover an arbitrariy number of coordinates, and use this parametersization to do any of the typical data analysis tasks, such as outlier detection, clustering, and model fitting.
 
@@
 Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.
+
+  \begin{itemize}
+  \item Discuss parameter robustness/sensitity/tuning
+  \item timing/scaling
+  \item mention code
+  \item SNR? Add Gaussian Noise?
+  \end{itemize}
 
@@
 
-%\bibliographystyle{pnas2}
-%\bibliography{background_reading/references,../../references/references}
-
-\begin{thebibliography}{10}
-
-\bibitem{diggle2002analysis}
-Diggle P, Heagerty P, Liang K~Y, Zeger S (2002) \textit{Analysis of
-  longitudinal data} (Oxford University Press).
-
-\bibitem{roelens2013live}
-Roelens B, Chanes J~D~L~H, Dostatni N (2013) Live imaging of bicoid-dependent
-  transcription in \textit{{D}rosophila} embryos. \textit{Curr Biol}
-  23:1--5.
-
-\bibitem{keller2013imaging}
-Keller P~J (2013) Imaging morphogenesis: technological advances and biological
-  insights. \textit{Science} 340(6137).
-
-\bibitem{jaeger2004dynamic}
-Jaeger J, Surkova S, Blagov M, Janssens H, Kosman D, \textit{et~al.} (2004)
-  Dynamic control of positional information in the early \textit{{D}rosophila}
-  embryo. \textit{Nature} 430(6997):368--371.
-
-%\bibitem{peter2012predictive}
-%Peter IS, Faure E, Davidson EH (2012) Predictive computation of genomic %logic processing functions in embryonic development. \textit{Proc Natl %Acad Sci USA} 109(41):16434--16442.
-
-\bibitem{peter2011gene}
-Peter IS, Davidson EH (2011) A gene regulatory network controlling the embryonic
-specification of endoderm. \textit{Nature} 474(7353):635--639.
-
-\bibitem{fowlkes2008quantitative}
-Fowlkes C~C, Hendriks C~L~L, Ker{\"a}nen S~V, Weber G~H, R{\"u}bel O,
-  \textit{et~al.} (2008) A quantitative spatiotemporal atlas of gene expression
-  in the \textit{{D}rosophila} blastoderm. \textit{Cell} 133(2):364--374.
-
-\bibitem{ng2012large}
-Ng L~L, Sunkin S~M, Feng D, Lau C, Dang C, \textit{et~al.} (2012) Large-scale
-  neuroinformatics for in situ hybridization data in the mouse brain.
-  \textit{Int Rev Neurobiol} 104:159--182.
-
-\bibitem{richardson2014emage}
-Richardson L, Stevenson P, Venkataraman S, Yang Y, Burton N, \textit{et~al.}
-  (2014) Emage: Electronic mouse atlas of gene expression, \textit{Mouse
-  Molecular Embryology} (Springer), pp. 61--79.
-
-\bibitem{castro2009automatic}
-Castro C, Luengo-Oroz M, Desnoulez S, Duloquin L, Fernandez-de Manuel L,
-  \textit{et~al.} (2009) An automatic quantification and registration strategy
-  to create a gene expression atlas of zebrafish embryogenesis,
-  \textit{Conf Proc IEEE Eng Med Biol Soc}, pp. 1469--1472.
-
-\bibitem{zitova2003image}
-Zitova B, Flusser J (2003) Image registration methods: a survey. \textit{Image Vis Comput} 21(11):977--1000.
-
-\bibitem{rowley1998rotation}
-Rowley H~A, Baluja S, Kanade T (1998) Rotation invariant neural network-based
-  face detection, \textit{Computer Vision and Pattern Recognition, 1998.
-  Proceedings. 1998 IEEE Computer Society Conference on} (IEEE), pp. 38--44.
-
-\bibitem{hajnal2010medical}
-Hajnal J~V, Hill D~L (2010) \textit{Medical image registration} (CRC press).
-
-\bibitem{greenspan1994rotation}
-Greenspan H, Belongie S, Goodman R, Perona P (1994) Rotation invariant texture
-  recognition using a steerable pyramid, \textit{Pattern Recognition, 1994.
-  Vol. 2-Conference B: Computer Vision \& Image Processing., Proceedings of the
-  12th IAPR International. Conference on} (IEEE), volume~2, pp. 162--167.
-
-\bibitem{zhao2003face}
-Zhao W, Chellappa R, Phillips P~J, Rosenfeld A (2003) Face recognition: A
-  literature survey. \textit{ACM Computing Surveys (CSUR)} 35(4):399--458.
-
-\bibitem{singer2012vector}
-Singer A, Wu H~T (2012) Vector diffusion maps and the connection laplacian.
-  \textit{Commun Pure Appl Math} 65(8):1067--1144.
-
-\bibitem{Belkin2003}
-Belkin M, Niyogi P (2003) Laplacian eigenmaps for dimensionality reduction and
-  data representation. \textit{Neural Comput} 15:1373--1396.
-
-\bibitem{coifman2005geometric}
-Coifman R~R, Lafon S, Lee A~B, Maggioni M, Nadler B, \textit{et~al.} (2005)
-  Geometric diffusions as a tool for harmonic analysis and structure definition
-  of data: Diffusion maps. \textit{Proc Natl Acad Sci USA} 102(21):7426--7431.
-
-\bibitem{coifman2006geometric}
-Coifman R~R, Lafon S (2006) Geometric harmonics: a novel tool for multiscale
-  out-of-sample extension of empirical functions. \textit{Appl Comput Harmon Anal} 21(1):31--52.
-
-\bibitem{tenenbaum2000global}
-Tenenbaum J~B, De~Silva V, Langford J~C (2000) A global geometric framework for
-  nonlinear dimensionality reduction. \textit{Science} 290(5500):2319--2323.
-
-\bibitem{roweis2000nonlinear}
-Roweis S~T, Saul L~K (2000) Nonlinear dimensionality reduction by locally
-  linear embedding. \textit{Science} 290(5500):2323--2326.
-
-\bibitem{zhao2014rotationally}
-Zhao Z, Singer A (2014) Rotationally invariant image representation for
-viewing direction classification in cryo-EM. \textit{J Struct Biol} 186(1):153--166.
-
-\bibitem{singer2011viewing}
-Singer A, Zhao Z, Shkolnisky Y, Hadani R (2011) Viewing angle
-classification of cryo-electron microscopy images using eigenvectors. \textit{SIAM J Imaging Sci} 4(2):723--759.
-
-\bibitem{lafon2006data}
-Lafon S, Keller Y, Coifman R~R (2006) 
-Data fusion and multicue data matching by diffusion maps. \textit{IEEE Trans Pattern Anal Mach Intell}
-28(11):1784--1797.
-
-\bibitem{jaeger2012drosophila}
-Jaeger J, Manu, Reinitz J (2012) \textit{{D}rosophila} blastoderm patterning.
-  \textit{Curr Opin Genet Dev} 22(6):533 -- 541.
-
-\bibitem{singer2011angular}
-Singer A (2011) Angular synchronization by eigenvectors and semidefinite
-  programming. \textit{Appl Comput Harmon Anal}
-  30(1):20--36.
-
-\bibitem{ahuja2007template}
-Ahuja S, Kevrekidis I~G, Rowley C~W (2007) Template-based stabilization of
-  relative equilibria in systems with continuous symmetry. \textit{Journal of
-  Nonlinear Science} 17(2):109--143.
-
-\bibitem{ian1998statistical}
-Dryden I~L, Mardia K (1998) \textit{Statistical shape analysis}, Wiley series
-  in probability and statistics: Probability and statistics (J. Wiley).
-
-\bibitem{foe1983studies}
-Foe V~E, Alberts B~M (1983) Studies of nuclear and cytoplasmic behaviour during
-  the five mitotic cycles that precede gastrulation in \textit{{D}rosophila}
-  embryogenesis. \textit{J Cell Sci} 61(1):31--70.
-
-\bibitem{rushlow2012temporal}
-Rushlow C~A, Shvartsman S~Y (2012) Temporal dynamics, spatial range, and
-  transcriptional interpretation of the dorsal morphogen gradient.
-  \textit{Curr Opin Genet Dev} 22(6):542--546.
-
-\bibitem{lim2013kinetics}
-Lim B, Samper N, Lu H, Rushlow C, Jim{\'e}nez G, \textit{et~al.} (2013)
-  Kinetics of gene derepression by erk signaling. \textit{Proc Natl Acad Sci USA} 110(25):10330--10335.
-
-\bibitem{figard2013plasma}
-Figard L, Xu H, Garcia H~G, Golding I, Sokac A~M (2013) The plasma membrane
-  flattens out to fuel cell-surface growth during \textit{{D}rosophila}
-  cellularization. \textit{Dev Cell} 27(6):648--655.
-
-\bibitem{shlens2005tutorial}
-Shlens J (2005) A tutorial on principal component analysis. \textit{Systems
-  Neurobiology Laboratory, University of California at San Diego} .
-
-\bibitem{leptin2005gastrulation}
-Leptin M (2005) Gastrulation movements: the logic and the nuts and bolts.
-  \textit{Dev Cell} 8(3):305--320.
-
-\bibitem{kemelmacher2011exploring}
-Kemelmacher-Shlizerman I, Shechtman E, Garg R, Seit, S~M (2011)
-Exploring photobios. \textit{ACM Transactions on Graphics (TOG)} 30(4):61.
-
-\bibitem{snail}
-Ip Y~T, Park R~E, Kosman D, Bier E, Levine M (1992) 
-The dorsal gradient morphogen regulates stripes of rhomboid expression in the presumptive neuroectoderm of the \textit{Drosophila} embryo. \textit{Genes Dev} 6(9):1728--1739.
-
-\bibitem{yuan2014automated}
-Yuan L, Pan C, Ji S, McCutchan M, Zhou Z~H, \textit{et~al.} (2014) Automated
-  annotation of developmental stages of \textit{{D}rosophila} embryos in images
-  containing spatial patterns of expression. \textit{Bioinformatics}
-  30(2):266--273.
-
-\bibitem{surkova2008characterization}
-Surkova S, Kosman D, Kozlov K, Myasnikova E, Samsonova A~A, \textit{et~al.}
-  (2008) Characterization of the \textit{{D}rosophila} segment determination
-  morphome. \textit{Dev Biol} 313(2):844--862.
-
-\bibitem{anavy2014blind}
-Anavy L, Levin M, Khair S, Nakanishi N, Fernandez-Valverde S~L, \textit{et~al.}
-  (2014) Blind ordering of large-scale transcriptomic developmental
-  timecourses. \textit{Development} 141(5):1161--1166.
-
-\bibitem{trapnell2014dynamics}
-Trapnell C, Cacchiarelli D, Grimsby J, Pokharel P, Li S, \textit{et~al.} (2014)
-  The dynamics and regulators of cell fate decisions are revealed by
-  pseudotemporal ordering of single cells. \textit{Nat Biotechnol} 32(4):381--386.
-
-\bibitem{gupta2008extracting}
-Gupta A, Bar-Joseph Z (2008) Extracting dynamics from static cancer expression
-  data. \textit{IEEE/ACM Trans Comput Biol Bioinform} 5(2):172--182.
-
-%\bibitem{singer2011three}
-%Singer A, Shkolnisky Y (2011) Three-dimensional structure determination from
-%  common lines in cryo-em by eigenvectors and semidefinite programming.
-%  \textit{SIAM J Imaging Sci} 4(2):543--572.
-
-%\bibitem{silverman2005functional}
-%Ramsay J~O, Silverman B~W (2005) \textit{Functional Data Analysis}, Springer
-%  Series in Statistics (Springer).
-
-\bibitem{goemans1995improved}
-Goemans M~X, Williamson D~P (1995) Improved approximation algorithms for maximum cut and satisfiability problems using semidefinite programming.
-\textit{Journal of the ACM} 42(6):1115--1145.
-
-\bibitem{bandeira2013cheeger}
-Bandeira A~S, Singer A, Spielman D~A (2013) A Cheeger inequality for the graph connection Laplacian. \textit{SIAM Journal on Matrix Analysis and Applications} 34(4):1611--1630.
-
-\bibitem{mallat2012group}
-Mallat, S (2012) Group invariant scattering. 
-\textit{Communications on Pure and Applied Mathematics}
-65(10):1331--1398.
-
-\bibitem{krzic2012multiview}
-Krzic U, Gunther S, Saunders T~E, Streichan S~J, Hufnagel L (2012) Multiview
-  light-sheet microscope for rapid in toto imaging. \textit{Nat Methods}
-  9(7):730--733.
-
-\bibitem{ichikawa2014live}
-Ichikawa T, Nakazato K, Keller P~J, Kajiura-Kobayashi H, Stelzer E~H,
-  \textit{et~al.} (2014) Live imaging and quantitative analysis of gastrulation
-  in mouse embryos using light-sheet microscopy and 3d tracking tools.
-  \textit{Nat Protoc} 9(3):575--585.
-
-\bibitem{rubel2010coupling}
-R{\"u}bel O, Ahern S, Bethel E, Biggin M~D, Childs H, \textit{et~al.} (2010)
-  Coupling visualization and data analysis for knowledge discovery from
-  multi-dimensional scientific data. \textit{Procedia Comput Sci}
-  1(1):1757--1764.
-
-\end{thebibliography}
+\bibliographystyle{pnas2}
+\bibliography{../image_analysis_paper/background_reading/references,../../references/references}
 
@@
 
-%\begin{figure}[t]
-%\includegraphics{fig1}
-%\caption{Caricature illustrating the tasks of image registration and temporal ordering.
-{\xfigtextfontit (A)} Images of ``samples'', each in a different orientation and a different stage of development.
-{\xfigtextfontit (B)} Registered and ordered samples.
-For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
-%\label{fig:fish}
-%\customlabel{fig:fish}{1}
-%\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
-%\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
-%\end{figure}
-%
-%\begin{figure}[t]
-%\includegraphics{fig2}
-%\caption{Schematic illustrating angular synchronization and diffusion maps.
-{\xfigtextfontit (A)} Set of vectors, each in a different orientation.
-The pairwise alignment angles are indicated.
-{\xfigtextfontit (B)} The vectors from {\xfigtextfontit A}, each rotated about their midpoint so that the set is globally aligned.
-Note that the chosen rotation angles are consistent with the pairwise alignments in {\xfigtextfontit A}: the differences between a pair of angles in {\xfigtextfontit B} is the same as the pairwise angle in {\xfigtextfontit A}.
-{\xfigtextfontit (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
-Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\xfigtextfontit \SI}), so that close data points are connected by darker (``stronger'') edges.
-{\xfigtextfontit (D)} The data in {\xfigtextfontit C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure.
-The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
-%\label{fig:schematics}
-%\customlabel{fig:schematics}{2}
-%\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
-%\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
-%\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
-%\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
-%\end{figure}
-%
-%\begin{figure*}[t]
-%\includegraphics{fig3}
-%\caption{{\xfigtextfontit Drosophila} embryos during cellularization.
-{\xfigtextfontit (A)} Images of embryos collected during cellularization.
-Each image is of a different embryo arrested at a different developmental time, and in a different rotational and translational orientation.
-{\xfigtextfontit (B)} Images from {\xfigtextfontit A}, now registered and ordered using vector diffusion maps.
-The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right.
-The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.}
-%\customlabel{fig:data1}{3}
-%\customlabel{subfig:raw_data1}{\ref{fig:data1}{\it A}}
-%\customlabel{subfig:ordered_data1}{\ref{fig:data1}{\it B}}
-%\customlabel{subfig:averaged_data1}{\ref{fig:data1}{\it C}}
-%\end{figure*}
-%
-%\begin{figure*}[t]
-%\includegraphics{fig4}
-%\caption{{\xfigtextfontit Drosophila} embryos during gastrulation.
-{\xfigtextfontit (A)} Images of gastrulating embryos.
-Each image is of a different embryo arrested at a different developmental time, and in a different translation and rotational orientation.
-{\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps.
-The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right.
-{\xfigtextfontit (C)} A representative ``developmental path'' from local averaging of the images.
-Each image is a (Gaussian-weighted) average of the images in {\xfigtextfontit B}.
-}
-%\customlabel{fig:data2}{4}
-%\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
-%\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
-%\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
-%\end{figure*}
-%
-%\begin{figure}[h]
-%\includegraphics{fig5}
-%\caption{Wild type and mutant {\xfigtextfontit Drosophila} embryos.
-{\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~\ref{fig:raw_data3}), projected onto the first two vector diffusion maps coordinates.
-The directions of the developmental trajectories are indicated by arrows.
-(Inset) The projected data, now colored by genotype.
-The two branches separate mutant and wild type samples.
-{\xfigtextfontit (B)} dpERK signal at select points along the wild type trajectory in {\xfigtextfontit A} (denoted by circles in the inset).
-The two dpERK peaks which arise during development are indicated by arrows.
-{\xfigtextfontit (C)} dpERK signal at select points along the mutant trajectory in {\xfigtextfontit A} (denoted by circles in the inset).
-The single dpERK peak is indicated by an arrow.
-}
-%\customlabel{fig:data3}{5}
-%\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
-%\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it B}}
-%\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it C}}
-%\end{figure}
-
 \end{document}

### 2014-08-20 image_analysis_paper2/image_fn_color.m

@@ -1,11 +1,11 @@
 function image2 = image_fn_color(image, channel, npixels)
 
-filt = fspecial('disk', 5);
+filt = fspecial('disk', round(0.05*npixels));
 
 image2 = imresize(image, [npixels npixels]);
 
 image2 = crop_image(image2, channel);
-image2 = padarray(image2, [10 10 0]);
+image2 = padarray(image2, [round(0.1*npixels) round(0.1*npixels) 0]);
 image2 = imresize(image2, [npixels npixels]);
 
 image2(:, :, channel) = adapthisteq(image2(:, :, channel));

### 2014-08-20 image_analysis_paper2/make_figures_onemovie.m

@@ -88,7 +88,7 @@ for k=1:nmovies
     
     if k == k_print_fig
         [~, max_idx] = max(mod(theta, 360));
-        make_fig(4,4);
+        make_fig(4.3,4);
         plot(mod(theta, 360), mod(theta_opt-theta_opt(max_idx)-1, 360)+1, '.')
         xlabel('true angle')
         ylabel('recovered angle')
@@ -96,8 +96,9 @@ for k=1:nmovies
         set(gca, 'ytick', [0 180 360]);
         set(gca, 'xticklabel', {'0°'; '180°'; '360°'});
         set(gca, 'yticklabel', {'0°'; '180°'; '360°'});
+        axis square
         saveas(gcf, 'angle_corr.pdf');
-         
+        
         make_fig(4,4);
         plot(compute_ranks(time), compute_ranks(embed_coord(:,1)),'.')
         xlabel('true rank')
@@ -105,6 +106,7 @@ for k=1:nmovies
         set(gca, 'xtick', [0 20 40])
         set(gca, 'ytick', [0 20 40])
         axis([0 50 0 50])
+        axis square
         saveas(gcf, 'rank_corr.pdf');
         
         nprint_images = 10;
@@ -137,13 +139,13 @@ end
 % xlabel('movie index')
 % ylabel('average angle error (degrees)')
 % saveas(gcf, 'angle_corr_allmovies.pdf');
-% 
+%
 % make_fig(5.5, 5.5);
 % bar(rank_corr([2 1 3 5 6]));
 % xlabel('movie index')
 % ylabel('rank correlation coefficient')
 % saveas(gcf, 'rank_corr_allmovies.pdf');
-% 
+%
 % make_fig(5.5, 5.5)
 % bar_width = 0.15;
 % [ax,h1,h2] = plotyy((1:nmovies)-bar_width/2,theta_err,(1:nmovies)+bar_width/2,rank_corr,'bar');
@@ -256,6 +258,7 @@ xlabel('number of images')
 ylabel('rank correlation')
 axis([0 50 0.5 1.0])
 set(gca, 'xtick', [0 20 40])
+axis square
 saveas(gcf, 'bootstrap_rankcorr.pdf');
 
 % figure;

### 2014-08-20 image_analysis_paper2/make_gray_nuclei.m

@@ -1,8 +1,12 @@
 function image2 = make_gray_nuclei(image)
 
 image2 = image;
+
+scale = 1.25;
+
+image2(:,:,1) = immultiply(image2(:,:,1), scale);
 for i=2:3
-    image2(:,:,i) = imlincomb(1, image2(:,:,i), 1, image2(:,:,1));
+    image2(:,:,i) = imlincomb(1, image2(:,:,i), scale, image2(:,:,1));
 end
 
 image2 = circshift(image2, [0 0 -1]);

### 2014-08-20 image_analysis_paper2/timing_studies.m

@@ -1,0 +1,112 @@
+clear all
+close all
+
+%%
+
+
+nimages_all = 132;
+image_dir = '../membrane_pictures/14_0501_dpERK_late';
+image_name = 'emb';
+
+shift_max = 0.1;
+
+neigs = 10;
+alpha = 0;
+
+channel = 1;
+nchannels = 3;
+
+image_set_raw_all = zeros(1024, 1024, nchannels, nimages_all, 'uint8');
+
+%%
+
+for i=1:nimages_all
+    im_tmp = imread(sprintf('%s/%s%02d.tif', image_dir, image_name, i));
+    image_set_raw_all(:, :, :, i) = im_tmp;
+end
+
+%%
+
+% npixels_vec = [50 100 200];
+npixels_vec = [50 100 200];
+nimages_vec = [20 50 100 120];
+nrot_vec = [10 20 30 40];
+% nshifts_vec = [0 5 10];
+nshifts_vec = [0];
+
+timing = zeros(length(npixels_vec), length(npixels_vec), length(nrot_vec), length(nshifts_vec));
+
+%%
+
+for i=1:length(npixels_vec)
+    image_set_all = imresize(image_set_raw_all, [npixels_vec(i) npixels_vec(i)]);
+    for j=1:length(nimages_vec)
+        nimages = nimages_vec(j);
+        image_set = image_set_all(:, :, :, 1:nimages);
+        for k=1:length(nrot_vec)
+            nrot = nrot_vec(k);
+            for l=1:length(nshifts_vec)
+                nshifts = nshifts_vec(l);
+                
+                tic
+                
+                %
+                [R, W] = compute_pairwise_alignments(image_set, nrot, nshifts, shift_max);
+                dim = size(R, 1) / nimages;
+                
+                %
+                W2 = W.^2;
+                eps = median(W2(:))/10;
+                [R_opt, embed_coord, D2, D] = vdm(R, W2, eps, neigs, alpha);
+                
+                timing(i,j,k,l) = toc;
+            end
+        end
+    end
+end
+
+%%
+
+figure;
+for i=1:length(npixels_vec)
+    for j=1:length(nimages_vec)
+        % for k=1:length(nrot_vec)
+        for l=1:length(nshifts_vec)
+            loglog(nrot_vec, squeeze(timing(i,j,:,l)), '-o')
+            hold on
+        end
+    end
+end
+loglog([10 50], [10 50], '-r')
[…]

### 2014-08-21 image_analysis_paper2/pnas_draft.tex
added recent draft of paper

@@
 
-%\significancetext{Cross-sectional imaging studies are commonly used to study developmental dynamics in multiple organisms. 
-%%
-%In these studies, a representative developmental trajectory is reconstructed from snapshots of different embryos, each of which has been arrested at a different point of its development. 
-%%
-%One of the first steps in this reconstruction involves registration and temporal ordering of images. 
-%%
-%Here we show that both of these tasks can be automated and combined in a single step using recent advances in data mining. 
-%%
-%To illustrate this approach, we temporally order and register several data sets collected in cross-sectional studies of cell signaling in the early fruit fly embryo.}
+\significancetext{
+Cross-sectional imaging studies are commonly used to study developmental dynamics in multiple organisms. 
+%
+In these studies, a representative developmental trajectory is reconstructed from snapshots of different embryos, each of which has been arrested at a different point of its development. 
+%
+One of the first steps in this reconstruction involves registration and temporal ordering of images. 
+%
+Here we show that both of these tasks can be automated and combined in a single step using recent advances in data mining. 
+%
+To illustrate this approach, we temporally order and register several data sets collected in cross-sectional studies of cell signaling in the early fruit fly embryo.}
 
@@
 %
-We show that the algorithm recovers the correct orientation and temporal order in a data set where they are independently known, reconstructs an average developmental trajectory when such information is unavailable, and can detect distinct trajectories in a heterogeneous population of embryos.
+We show that the algorithm reproducible recovers the correct orientation and temporal order in data sets where they are independently known, and can reconstruct an average developmental trajectory when such information is unavailable.
 
@@
 
-\section{Significance}
-Cross-sectional imaging studies are commonly used to study developmental dynamics in multiple organisms. 
-%
-In these studies, a representative developmental trajectory is reconstructed from snapshots of different embryos, each of which has been arrested at a different point of its development. 
-%
-One of the first steps in this reconstruction involves registration and temporal ordering of images. 
-%
-Here we show that both of these tasks can be automated and combined in a single step using recent advances in data mining. 
-%
-To illustrate this approach, we temporally order and register several data sets collected in cross-sectional studies of cell signaling in the early fruit fly embryo.
-\vspace{0.5cm}
+%\section{Significance}
+
+%\vspace{0.5cm}
 
@@
 %
-In order to recover the developmental dynamics from such data sets, snapshots of different embryos must first be spatially aligned or {\em registered} to factor out the relevant geometric symmetries (e.g., translations and rotations), and then ordered in time.
+In order to recover the developmental dynamics from such data sets, snapshots of different embryos must first be spatially aligned or {\em registered} to factor out the relevant geometric symmetries (e.g., rotations), and then ordered in time.
 %
@@
 %
-To illustrate our approach, we analyze three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics \cite{jaeger2012drosophila}.
+To illustrate our approach, we analyze two data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics \cite{jaeger2012drosophila}.
+%
+Our first data set, where the correct rotational orientation and temporal order are independently known, will be used to validate our approach.
+%
+The correct orientation and order is unknown for the second data set; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.
 
@@
 technique developed for data sets which contain two types of sources of variability:
-geometric symmetries (such as translations and rotations of the images due to the experimental setup) which one would like to factor out,
+geometric symmetries (such as rotations of the images due to the experimental setup) which one would like to factor out,
 and ``additional" directions of variability (such as temporal dynamics) which one wants to uncover.
@@
 %
-We will use the algorithm to register images of {\it Drosophila} embryos with respect to rotations and translations, as well as uncover the main direction of variability {\it after} removing translational and rotational symmetries.
+We will use the algorithm to register images of {\it Drosophila} embryos with respect to rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
 %
@@
 However, the algorithm can successfully register data sets even when many of the pairwise measurements are inaccurate \cite{singer2011angular}.
-%
 
-%
-After removing variability due to translations and rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
+After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
 %
@@
 \caption{{\xfigtextfontit Drosophila} embryos during cellularization.
-{\xfigtextfontit (A)} Images of embryos collected during cellularization.
-Each image is of a different embryo arrested at a different developmental time, and in a different rotational and translational orientation.
-{\xfigtextfontit (B)} Images from {\xfigtextfontit A}, now registered and ordered using vector diffusion maps.
-The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right.
-The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.}
+{\xfigtextfontit (A)} Selected images from a live movie of a {\xfigtextfontit Drosophilla} embryo during gastrulation.
+Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
+{\xfigtextfontit (B)} Images from {\xfigtextfontit A} after registration and ordering using vector diffusion maps.
+The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.
+{\xfigtextfontit (C)} The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle for each frame in the movie.
+{\xfigtextfontit (D)} The correlation between the recovered rank (using vector diffusion maps) and the true rank for each frame in the movie.
+{\xfigtextfontit (E)} The average error in the recovered angle and the rank correlation coefficient for $5$ different movies.
+{\xfigtextfontit (F)} The average rank correlation coefficient as a function of the number of images in the data set.
+Each point is averaged over $200$ samples.
+}
 \customlabel{fig:movie}{3}
@@
 
-As a first illustration, we analyze a data set where the correct registration and temporal order is known {\it a priori}.
+As a first illustration, we analyzed a data set where the correct registration and temporal order is known {\it a priori}.
 %
-This data set is obtained by live imaging of a single {\it Drosophila} embryo during the thirty minute time interval spanning late cellularization through gastrulation.
+This data set is obtained by live imaging of a single {\it Drosophila} embryo during the twenty minute time interval spanning late cellularization through gastrulation.
 %
@@
 
-The movie shows an optical cross-section of a vertically oriented developing embryo, where the nuclei are labeled by DAPI, a DNA stain.
+The images show an optical cross-section of a vertically oriented developing embryo, where the nuclei are labeled by DAPI, a DNA stain.
 %
-The movie contains $42$ frames, taken at $30$~sec time intervals;
-\fig~\ref{subfig:movie_unregistered_unordered} shows selected frames from this movie. 
+The data set contains $42$ frames, taken at $30$~sec time intervals;
+\fig~\ref{subfig:movie_unregistered_unordered} shows selected images from this data set. 
 %
-The gray nuclear signal has been blurred to obtain a smoother images.
+The gray nuclear signal was blurred to obtain a smoother images.
 %
-Each frame has been arbitrarily rotated, and the order of the frames has been scrambled.
+Each frame was been arbitrarily rotated, and the order of the frames was scrambled.
 
@@
 %
-\fig~\ref{subfig:movie_registered_ordered} shows the same selected frames, now registered and ordered.
+\fig~\ref{subfig:movie_registered_ordered} shows the same selected images, now registered and ordered, and the real time for each frame is indicated.
 %
-The real time for each frame is indicated.
+Note that the recovered ordering is consistent with the real time for each frame. 
 %
@@
 %
-The errors in the recovered angles are all very low, and the rank correlation coefficients are consistently close to 1, indicating that our methodology can reproducably order data of this type.
+The errors in the recovered angles are all very low, and the rank correlation coefficients are consistently close to 1, indicating that our methodology can reproducibly order data of this type.
 
-To assess how the quality of our results depends on the number of data points, we used bootstrap sampling. 
+We used bootstrap sampling to determine how the accuracy of the recovered orderings varies as a function of the size of the data set.
 %
-\fig~\ref{subfig:movie_bootstrap} shows the average rank correlation coefficient as a function of the number of images in the data set.
+\fig~\ref{subfig:movie_bootstrap} shows the average rank correlation coefficient as a function of the number of images.
 %
-The data sets were sampled (with replacement) from the original $40$-frame movie.
+The images were sampled (with replacement) from the original data set of $42$ images, and each point is the average over $200$ bootstrap samples.
 %
@@
 %
-The recovered angles had errors of less than $10^{\circ}$, and the magnitude of the errors did not vary significantly with the size of the data set.
+The errors in the recovered angles did not vary significantly with the number of images and were consistently less than $10^{\circ}$.
 
@@
 
-We will now demonstrate the utility of our methodogies on a data set where the correct ordering is not known {\em a priori}.
+We can now demonstrate the utility of our methodologies on a data set where the correct ordering is not known {\em a priori}.
 %
@@
 
+\fig~\ref{subfig:fixed_images_registered_ordered} shows the images registered and ordered using vector diffusion maps \cite{singer2012vector}.
+%
+However, assimilating the dynamics even from this concatenation of images is difficult for two reasons, even after registration and ordering.
+%
+First, the sheer number of images makes visual processing of the entire data set nontrivial;
+and second, the ordered data is not entirely smooth due to inter-embryo variability.
+%
+To highlight the developmental dynamics revealed by vector diffusion maps, we divided the ordered trajectory into $14$ groups, and computed the average image in each group.
+%
+Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:fixed_images_average_trajectory}, serve as a summary of the stereotypic developmental dynamics reconstructed from cross-sectional data.
+%
+We can now easily see the developmental progression, which is consistent with known dynamics. 
+%
+dpERK first appears as two lateral peaks at the ventral side of the embryo; a third dpERK peak then appears at the dorsal side of the embryo.
+%
+During invagination, the two ventral dpERK peaks merge together, eventually forming (together with Twi) a characteristic ``omega'' shape.
+%
+The dorsal dpERK peak then dissapears during germband extension as cells from the ventral side wrap around to the dorsal side.
+%
+At the end of this process, similar ``omega'' shaped patterns formed by Twi and dpERK appear on the dorsal side; these patterns are most readily seen in the last image of \fig~\ref{subfig:fixed_images_average_trajectory}.
+%
+Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental dynamics (e.g., correlation of age with body size).
+
+Empirically, we found required computational time to be polynomial in the number of images, number of pixels, and the number of discretization points used to compute the pairwise rotations. 
+%
+The CPU time required scaled linearly with the number of pixels and the number of rotations, and scaled to the 1.5 power with the number of images in the data set.
+%
+For $100$ images of $100 \times 100$ pixels, using $30$ rotation discritezation points, the computations took $15$~seconds on an Intel Core i7 2.93 GHz processor. 
+%
+Furthermore, the computation of the pairwise rotational alignments, which is the most computationally intensive portion of the calculation, is trivially parallelizable.
+%
+Therefore, this analysis is computationally feasible for much larger imaging data sets which are impractical to evaluate manually.
+
 \begin{figure*}[t]
@@
 {\xfigtextfontit (A)} Images of gastrulating embryos.
-Each image is of a different embryo arrested at a different developmental time, and in a different translation and rotational orientation.
+Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation.
 {\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps.
@@
 {\xfigtextfontit (C)} A representative ``developmental path'' from local averaging of the images.
-Each image is a (Gaussian-weighted) average of the images in {\xfigtextfontit B}.
 }
@@
 
-FIX!!
-
-\fig~\ref{subfig:ordered_data2} shows the images registered and ordered using vector diffusion maps \cite{singer2012vector}, 
-and visual inspection confirms this ordering is consistent with the known developmental dynamics outlined above.
-%
-However, assimilating the dynamics even from this concatenation of images is difficult for two reasons, even after registration and ordering.
-%
-First, the sheer number of images makes visual processing of the entire data set nontrivial;
-and second, the ordered data is not entirely smooth due to inter-embryo variability.
-%
-To highlight the developmental dynamics revealed by vector diffusion maps, we applied a Gaussian filter to the trajectory of registered and ordered images to obtain an averaged developmental path (see \SI and  \cite{kemelmacher2011exploring}).
-%
-Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:averaged_data2}, serve as a summary of the stereotypic developmental dynamics reconstructed from cross-sectional data.
-%
-Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental dynamics (e.g., correlation of age with body size).
-
-%
-dpERK appears as two lateral peaks at the ventral side of the embryo.
-%
-These two peaks merge together during invagination, eventually forming (together with Twi) a characteristic ``omega'' shape.
-%
-Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
-%
-At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing similar ``omega'' shaped patterns to appear on the dorsal side; these patterns are most readily seen in the last image of \fig~\ref{subfig:fixed_images_average_trajectory}.
-
 \section{Discussion}
 
-\begin{itemize}
-\item High--throughput imaging is becoming increasingly common
-\item We are/wil reach the point where imaging data is being produced more rapidly than it can be analyzed by hand (similar to microarray data)
-\item We need a way to automate the organization of such data
-\item Our algorithms register, which is important for many imaging applications
-\item In general our algorithms find good coordinates/descriptors for the data
-\item We focus on uncovering a single coordinate that can be used to order data
-\item in general, can uncover many coordinates; use coordinates to do other data processing tasks (clustering, model fitting, etc.)
-
-\end{itemize}
 We presented a unified approach to temporal ordering and registration of images in cross-sectional studies of developmental dynamics. 
@@
 
-We are approaching the time when the rate of imaging data collection will surpass the rate of manual image analysis. 
-%
-We therefore require automated methodologies to organize such large data sets; these methodogies should be sufficiently general to address a wide variety of applications. 
-%
-Vector diffusion maps allows us to automatically register images, which is an essential task for many applications.
-%
-Simultaneously the algorithm provides us with a parameterization or coordinates for each image.
-%
-In the examples presented here, we have focued on recovering a single coordinate, and using this coordinate to order the images, effectively ordering them in time.
-%
-In general, we can recover an arbitrariy number of coordinates, and use this parametersization to do any of the typical data analysis tasks, such as outlier detection, clustering, and model fitting.
-
-We used vector diffusion maps to analyze three data sets, collected specifically for this study. 
-%
-In the first example, correct orientation and temporal order of images was known {\it a priori}, providing a test case for the algorithm in which both of these features were accurately recovered. 
-%
-For this particular data set, a simpler approach would have sufficed;
-because of the relative simplicity of developmental dynamics (monotonic increase in the amplitude of a localized signaling pattern and small tissue deformation), temporal order could have been inferred by projection of registered images on the first principal component of the data set (see \fig~\ref{fig:PCA_data1}).
-%
-This approach was not sufficient for the second data set (see \fig~\ref{fig:PCA_data2}), collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.
-%
-Vector diffusion maps could readily extract a nontrivial spatiotemporal trajectory. 
-%
-Finally, the algorithm was applied to a heterogeneous data set, composed of images from embryos of different genotypes. 
-%
-In this case, low-dimensional projection of the data revealed distinct paths which correlated with the developmental trajectories of the mutant and wild type populations.
-
 Previously, angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy images \cite{singer2012vector, zhao2014rotationally, singer2011viewing}.
@@
 %
-Based on the presented results, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories. 
+Based on the presented results, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories, and that the size of the data set required is a function of both the instrument noise as well as interembryo variablity.
 %
-Here, we have used vector diffusion maps to register the images with respect to translations and rotations; however, it is conceivable that one could factor out other sources of variability, such as reflections \cite{singer2012vector, goemans1995improved, bandeira2013cheeger} or dilations.
+Here, we have used vector diffusion maps to register the images with respect to rotations; however, one could factor out other sources of variability, such as translations, reflections \cite{singer2012vector, goemans1995improved, bandeira2013cheeger} or dilations.
 
-For the data sets analyzed in this work, the Euclidean norm between the pixelated images was informative for identifying similar data points. 
+As experimental imaging becomes faster and less expensive, the rate of collection of imaging data will surpass the rate of manual image analysis. 
 %
-For more complex data sets, for instance when the imaged samples are significantly deformed, working with raw images might be more challenging.   
+We therefore require automated methodologies which are sufficiently general to address a wide variety of applications to organize such large data sets.
 %
-In this case, one might choose to operate not on the images directly, but on those functions of the images that are invariant to symmetries and insensitive to small deformations.  
+Vector diffusion maps allows us to automatically register images, an essential task for many applications.
 %
-This is the basis of the recently developed scattering transform \cite{mallat2012group}, as well as the bispectrum-based steerable PCA algorithms in \cite{zhao2014rotationally}. 
+Simultaneously, the algorithm provides us with coordinates for each image.
 %
-Such methods circumvent the need for registration by using invariant features of the images in the data analysis.
+In the examples presented here, we have focused on recovering a single coordinate which is used to effectively order the images in time.
+%
+In general, we can recover several coordinates that are good descriptors of our data set.
+%
+This parameterization can then be used to do any typical data analysis tasks, such as outlier detection, clustering, and model fitting. 
+%
+Images taken from different viewing directions can be analyzed in this way, as the vector diffusion maps parameterization will organize the images according to the viewing angle.
 %
@@
   \begin{itemize}
-  \item Discuss parameter robustness/sensitity/tuning
-  \item timing/scaling
-  \item mention code
-  \item SNR? Add Gaussian Noise?
+  \item Will this work for more complex data sets (3-D imaging, different developmental processes)?
+  \item How dense does the sampling have to be (I don't think we can quantitatively say anything, but it has to be dense enough so that similar images have a small Euclidean distance)
+  \item Discuss parameter robustness/sensitity/tuning, how much effort is needed, etc.
+  \item How do the results depend on SNR?
+  \item `` One comment: the use of cross-sectional to describe image data from different specimen is highly confusing here, since the data in question do show a cross-section of Drosophila embryo''
+  \item Compare times and quality to Michael/Henry
   \end{itemize}
@@
 
+\section{Software}
+%
+The vector diffusion maps algorithms was implemented in MATLAB\textsuperscript{\textregistered} (R2013b, The MathWorks, Natick, Massachusetts).
+%
+All code, including all the scripts used to generate the figures in this paper, is available at .....
 \end{materials}

### 2014-08-21 image_analysis_paper2/make_figures_onemovie.m

@@ -12,7 +12,7 @@ nmovies = 6;
 idx_end = [56 60 47 44 46 42];
 idx_start = idx_end - 41;
 
-nrot = 40;
+nrot = 36;
 nshifts = 0;
 shift_max = 0.1;
 
@@ -43,7 +43,7 @@ for k=1:nmovies
     
     %
     image_set = image_set(:, :, idx);
-    time = time(idx);
+    time = time(idx) - time(idx(1));
     nimages = length(time);
     
     %

### 2014-08-21 image_analysis_paper2/timing_studies.m

@@ -28,8 +28,8 @@ end
 %%
 
 % npixels_vec = [50 100 200];
-npixels_vec = [50 100 200];
-nimages_vec = [20 50 100 120];
+npixels_vec = [50 100 150 200];
+nimages_vec = [20 50 75 100 120];
 nrot_vec = [10 20 30 40];
 % nshifts_vec = [0 5 10];
 nshifts_vec = [0];
@@ -91,7 +91,7 @@ for i=1:length(npixels_vec)
         end
     end
 end
-loglog([10 100], ([10 100].^2)/100, '-r')
+loglog([10 100], 0.1*[10 100].^1.5, '-r')
 
 xlabel('number of images')
 ylabel('time (sec)')
@@ -106,7 +106,10 @@ figure;
         end
     end
 end
-loglog([10 100], ([10 100].^2)/100, '-r')
-xlabel('number of images')
+loglog([50 150], ([50 150].^2)/100, '-r')
+xlabel('number of pixels')
 ylabel('time (sec)')
+
+%% 
+save('timing_data.mat','npixels_vec','nimages_vec','nrot_vec','nshifts_vec','timing')
 

### 2014-08-23 image_analysis_paper2/pnas_draft.tex
edit pnas paper

@@
 \significancetext{
-Cross-sectional imaging studies are commonly used to study developmental dynamics in multiple organisms. 
+Imaging studies are commonly used to study developmental dynamics in multiple organisms. 
 %
-In these studies, a representative developmental trajectory is reconstructed from snapshots of different embryos, each of which has been arrested at a different point of its development. 
-%
-One of the first steps in this reconstruction involves registration and temporal ordering of images. 
+When many samples are involved in a single study,
+one of the first steps in data analysis involves registration and temporal ordering of images. 
 %
@@
 %
-To illustrate this approach, we temporally order and register several data sets collected in cross-sectional studies of cell signaling in the early fruit fly embryo.}
+To illustrate this approach, we temporally order and register several data sets collected in  studies of cell signaling in the early fruit fly embryo.}
 
@@
 %
-Many experiments are based on a cross-sectional design, where developmental progress is arrested in a population of embryos, each of which is at a different point along its developmental trajectory. 
+Often data is collected in several experiments across many samples, and the first task is to organize the data for further analysis
 %
-The goal is then to temporally order the data to reconstruct representative developmental dynamics from snapshots provided by the collection of embryos.
-%
-Images of different biological samples must first be registered before they can be temporally ordered.
+This includes spatially registering images of different biological samples, and well as
+temporally ordering the data to reconstruct representative developmental dynamics.
 %
@@
 %
-We present an automated approach to simultaneously register and temporally order data from such cross-sectional studies.
+We present an automated approach to simultaneously register and temporally order imaging data sets.
 %
@@
 %
-We use this method to extract developmental trajectories from a cross-sectional imaging study of cell signaling during dorsoventral patterning of the {\it Drosophila} embryo.
-%
-We show that the algorithm reproducible recovers the correct orientation and temporal order in data sets where they are independently known, and can reconstruct an average developmental trajectory when such information is unavailable.
+We validate this approach using a live imaging data set where both the correct rotational orientation and temporal order are known, as well as demonstrate using these methods to extract a developmental trajectory from a data set of fixed images to study dorsoventral patterning of the {\it Drosophila} embryo.
 
@@
 %
-This algorithm is one of several influential nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}.
+This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}.
 %
 These techniques has been used in
-applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \cite{zhao2014rotationally, singer2011viewing} to face recognition \cite{lafon2006data}, and are sufficiently general to be applicable to a wide variety of biological imaging studies.
+applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \cite{zhao2014rotationally, singer2011viewing} to face recognition \cite{lafon2006data}.
 %
-Here, the vector diffusion maps algorithm is adapted for the analysis of images from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
+Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing embryos from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
 %
@@
 Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.
-%
-Thus, when our experiments have additional sources of variability (such as distinct populations of embryos) in addition to developmental dynamics, the algorithm can provide an informative organization of our data which highlights both of these variability sources.
 
@@
 
-\subsection{Method validation using a live movie}
+\subsection{Method validation using live imaging}
 
-As a first illustration, we analyzed a data set where the correct registration and temporal order is known {\it a priori}.
+To validate the proposed approach, we applied our algorithm to a dataset where the true temporal order and rotational orientation are known {\em a priori}.
 %
-This data set is obtained by live imaging of a single {\it Drosophila} embryo during the twenty minute time interval spanning late cellularization through gastrulation.
+This data set was obtained by live imaging of a single {\it Drosophila} embryo during the twenty minutes spanning late cellularization through gastrulation.
 %
-During this time window the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, bringing in the future muscle cells and eventually forming a characteristic ``omega'' shape.
+During this time window, the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, bringing in the future muscle cells and eventually forming a characteristic ``omega'' shape.
 %
@@
 %
-At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing similar ``omega'' shaped patterns to appear on the dorsal side.
+At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing a similar ``omega'' shaped pattern to appear on the dorsal side.
 
-The images show an optical cross-section of a vertically oriented developing embryo, where the nuclei are labeled by DAPI, a DNA stain.
+\fig~\ref{subfig:movie_unregistered_unordered} shows selected images from this live imaging study. 
 %
-The data set contains $42$ frames, taken at $30$~sec time intervals;
-\fig~\ref{subfig:movie_unregistered_unordered} shows selected images from this data set. 
+The entire data set contains $42$ frames, taken at $30$~second time intervals.
 %
-The gray nuclear signal was blurred to obtain a smoother images.
+Each image shows an optical cross-section at a fixed depth of a vertically oriented developing embryo, with the nuclei labeled by Histone-GFP.
 %
-Each frame was been arbitrarily rotated, and the order of the frames was scrambled.
+Each frame was arbitrarily rotated, and the order of the frames was scrambled.
+%
+The task is now to register these images and order them in time to reconstruct the developmental trajectory.
 
@@
 %
-\fig~\ref{subfig:movie_registered_ordered} shows the same selected images, now registered and ordered, and the real time for each frame is indicated.
+\fig~\ref{subfig:movie_registered_ordered} shows the images from \fig~\ref{subfig:movie_unregistered_unordered}, now registered and ordered, and the real time for each frame is indicated.
 %
-Note that the recovered ordering is consistent with the real time for each frame. 
+With a small number of exceptions, the recovered ordering is consistent with the real time for each frame. 
 %
-\fig~\ref{subfig:movie_angle_corr} and \fig~\ref{subfig:movie_rank_corr}  shows the correlations between the recovered and true angles and ranks for each of the frames.
+\fig~\ref{subfig:movie_angle_corr} and \fig~\ref{subfig:movie_rank_corr}  show the correlations between the recovered and true angles and ranks, respectively, for the entire data set. 
 %
-Both the angles and the ranks are recovered to a high degree of accuracy.
+Both the angles and the ranks are recovered with a high degree of accuracy.
 
-To assess the robustness of our proposed methodology, we repeated this procedure with four additional movies.
+To assess the robustness of our proposed methodology, we repeated this procedure with four additional datasets extracted from independent live imaging studies spanning the same developmental time period.
 %
@@
 %
-The errors in the recovered angles are all very low, and the rank correlation coefficients are consistently close to 1, indicating that our methodology can reproducibly order data of this type.
+The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than $0.98$, indicating that our methodology can reproducibly order data of this type.
 
-We used bootstrap sampling to determine how the accuracy of the recovered orderings varies as a function of the size of the data set.
+Clearly, $42$ images are sufficient to accurately recover the rotational orientation and temporal order  in this specific developmental time frame.
 %
-\fig~\ref{subfig:movie_bootstrap} shows the average rank correlation coefficient as a function of the number of images.
+In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{subfig:dmaps1}) is well-sampled;
+the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \cite{singer2011angular}.
 %
-The images were sampled (with replacement) from the original data set of $42$ images, and each point is the average over $200$ bootstrap samples.
+To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to calculate the accuracy of the recovered ordering as a function of the size of the data set.
 %
-There is a steady increase in the accuracy of the recovered orderings with an increasing number of images.
+\fig~\ref{subfig:movie_bootstrap} shows the average rank correlation coefficient as a function of the number of images, where each point is the average over $200$ bootstrap samples.
 %
-The errors in the recovered angles did not vary significantly with the number of images and were consistently less than $10^{\circ}$.
+As expected, the accuracy in the recovered ordering increases with the number of images. 
+%
+The errors in the recovered angles did not vary significantly with the number of images and were consistently less than $10^{\circ}$;
+because our particular data set contains relatively low noise levels, few images are required to obtain accurate rotational alignments. 
+%
+We have therefore demonstrated that the algorithm accurately registers and orders a model data set where both the rotation angles and real times are independently known. 
+%
+We also showed that these results are consistent across multiple data sets, and presented how the accuracy of our methods varies as a function of the number of images.
 
@@
 
-We can now demonstrate the utility of our methodologies on a data set where the correct ordering is not known {\em a priori}.
+%The way it goes now is that its not only not known a priori, but that we have embryo to embryo variability, both in sizes of embryos and in the levels of signals extracted from images.
+To test our algorithm under these more challenges conditions, we collected a dataset where......
+
+The data set presented in the previous section contained no interembryo variability. 
 %
-\fig~\ref{subfig:fixed_images_unregistered_unordered} shows a set of $126$ images, which cover a thirty minute time interval spanning late cellularization through gastrulation.
+We will now demonstrate that our methods are sufficiently robust to order images from multiple embryos. 
+%
+\fig~\ref{subfig:fixed_images_unregistered_unordered} shows selected images from a set of $126$ images (see \SI) which cover a thirty minute time interval spanning late cellularization through gastrulation.
 %
@@
 %
-The nuclei (gray) are again labeled with a DNA marker.
+The nuclei (gray) are labeled with DAPI, a DNA stain.
 %
@@
 
-\fig~\ref{subfig:fixed_images_registered_ordered} shows the images registered and ordered using vector diffusion maps \cite{singer2012vector}.
+\fig~\ref{subfig:fixed_images_registered_ordered} shows the images from \fig~\ref{subfig:fixed_images_unregistered_unordered}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
 %
-However, assimilating the dynamics even from this concatenation of images is difficult for two reasons, even after registration and ordering.
+However,  the selected images are not entirely representative of the developmental dynamics due to interembryo variability. 
 %
-First, the sheer number of images makes visual processing of the entire data set nontrivial;
-and second, the ordered data is not entirely smooth due to inter-embryo variability.
-%
-To highlight the developmental dynamics revealed by vector diffusion maps, we divided the ordered trajectory into $14$ groups, and computed the average image in each group.
+To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory using a Gaussian filter across the set of registered and ordered images. 
 %
@@
 %
-The dorsal dpERK peak then dissapears during germband extension as cells from the ventral side wrap around to the dorsal side.
+The dorsal dpERK peak then disappears during germband extension as cells from the ventral side wrap around to the dorsal side.
 %
@@
 
+Each embryo is fixed prior to imaging, and the developmental time of each embryo cannot be easily estimated.
+%
+TODO: compare to Michael/Henry ordering
+
 Empirically, we found required computational time to be polynomial in the number of images, number of pixels, and the number of discretization points used to compute the pairwise rotations. 
 %
-The CPU time required scaled linearly with the number of pixels and the number of rotations, and scaled to the 1.5 power with the number of images in the data set.
+The CPU time required was $\mathcal{O}(n)$ in the number of pixels and the number of rotations, and $\mathcal{O}(n^{1.5})$ in the number of images in the data set.
 %
@@
 
-We presented a unified approach to temporal ordering and registration of images in cross-sectional studies of developmental dynamics. 
+We presented a unified approach to temporal ordering and registration of images in studies of developmental dynamics. 
 %
@@
 %
-The approach presented here is different because it addresses image registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace; vector diffusion maps can also order images when global registration of the entire data set is topologically impossible \cite{zhao2014rotationally}.
+The presented approach addresses image registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace.
+%; vector diffusion maps can also order images when global registration of the entire data set is topologically impossible \cite{zhao2014rotationally}.
 %
@@
 %
-In contrast to most of the existing approaches to image registration which rely on the knowledge about some landmarks in the data \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information.
+In contrast to most of the existing approaches to image registration which rely on the knowledge about some landmarks in the data \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them applicable to a wide variety of biological applications.
 
@@
 %
-Here, we have used vector diffusion maps to register the images with respect to rotations; however, one could factor out other sources of variability, such as translations, reflections \cite{singer2012vector, goemans1995improved, bandeira2013cheeger} or dilations.
+%Here, we have used vector diffusion maps to register the images with respect to rotations; however, one could factor out other sources of variability, such as translations, reflections \cite{singer2012vector, goemans1995improved, bandeira2013cheeger} or dilations.
 
-As experimental imaging becomes faster and less expensive, the rate of collection of imaging data will surpass the rate of manual image analysis. 
-%
-We therefore require automated methodologies which are sufficiently general to address a wide variety of applications to organize such large data sets.
+As experimental imaging becomes faster and less expensive, the rate of collection of imaging data will surpass the rate of manual image analysis,
+and we will require automated methodologies which are sufficiently general to address a wide variety of applications to organize such large data sets.
 %
@@
 %
-In the examples presented here, we have focused on recovering a single coordinate which is used to effectively order the images in time.
+In the examples presented here, we have focused on recovering a single coordinate which is used to order the images in time.
 %
@@
 %
-Images taken from different viewing directions can be analyzed in this way, as the vector diffusion maps parameterization will organize the images according to the viewing angle.
+Images taken from different viewing directions can be analyzed, as the vector diffusion maps parameterization will organize the images according to the viewing angle.
 %
@@
 
-\section{Development time during cellularization} 
+\section{Image preprocessing}
 %
-The progression of membrane invagination during cellularization was used to assign time point to each embryo in the third hour of development.
-Time-lapse imaging of Sqh-GFP embryo was used to generate the calibration curve that relates the membrane ingression length and the developmental time.
-Time-lapsing imaging of 7 embryos was used to construct the calibration curve (\fig~\ref{fig:membrane_curve}B).
-The rate at which membrane invaginates was highly reproducible from embryo to embryo.
-In the experiments with fixed embryos, we measured the membrane length by co-staining embryos with a membrane marker (Sqh-GFP) or by taking phase-contrast images (\fig~\ref{fig:membrane_curve}A).
-We used the correlation curve to assign each embryo to a time window of 2--3~minutes.
+Borders were cropped from each image using a Canny edge-detection filter to remove any size effects.
+%
+Adaptive histogram equalization was used to normalize the nuclear signal intensity throughout each image, and a 5-pixel radius disk filter was used to smooth the nuclear signal. 
+%
+%For the three-channel fixed images, the nuclear signal was scaled by half relative to the Twi and dpERK signals.
 
@@
 All code, including all the scripts used to generate the figures in this paper, is available at .....
+
 \end{materials}

### 2014-08-23 image_analysis_paper2/make_figures_fixed_images.m

@@ -59,6 +59,15 @@ for i=1:nimages
 end
 saveas(gcf, 'fixed_images_unregistered_unordered.pdf');
 
+nsubsamples = 15;
+subsample_idx = round(linspace(1, nimages, nsubsamples));
+make_fig(17, 17/nsubsamples);
+for i=1:nsubsamples
+    make_subplot(nsubsamples, 1, 0.01, i);
+    imshow(make_gray_nuclei(image_set(:,:,:,subsample_idx(i))))
+end
+saveas(gcf, 'fixed_images_unregistered_unordered_subsampled.pdf');
+
 %%
 
 [R, W] = compute_pairwise_alignments(image_set, nrot, nshifts, shift_max);
@@ -95,6 +104,15 @@ for i=1:nimages
     imshow(imrotate(im_tmp, rot_angle, 'crop'))
 end
 saveas(gcf, 'fixed_images_registered_ordered.pdf');
+
+[~, I2] = sort(embed_coord(subsample_idx,1));
+make_fig(17, 17/nsubsamples);
+for i=1:nsubsamples
+    make_subplot(nsubsamples, 1, 0.01, i);
+    im_tmp = make_gray_nuclei(image_set_aligned(:,:,:,subsample_idx(I2(i))));
+    imshow(imrotate(im_tmp, rot_angle, 'crop'))
+end
+saveas(gcf, 'fixed_images_registered_ordered_subsampled.pdf');
 
 %%
 nstages = 15;

### 2014-08-23 image_analysis_paper2/make_figures_onemovie.m

@@ -127,10 +127,10 @@ for k=1:nmovies
             im_tmp = imrotate(image_set(:, :, plot_idx(I(j))), -theta_opt(plot_idx(I(j)))+88, 'crop');
             make_subplot(subplot_dim1, subplot_dim2, 0.01, j);
             imshow(im_tmp);
-            text(npixels/2, npixels/2, sprintf('%2.2f min', time(plot_idx(I(j)))),'color',0.95*ones(1,3),'HorizontalAlignment','center','VerticalAlignment','middle', 'fontsize', 6)
+            text(npixels/2, npixels/2, sprintf('%2.1f min', time(plot_idx(I(j)))),'color',0.95*ones(1,3),'HorizontalAlignment','center','VerticalAlignment','middle', 'fontsize', 6)
         end
         saveas(gcf, 'movie_registered_ordered.pdf');
-        
+        return
     end
 end
 

### 2014-08-24 image_analysis_paper2/make_figures_onemovie.m
added more stuff for paper

@@ -130,7 +130,6 @@ for k=1:nmovies
             text(npixels/2, npixels/2, sprintf('%2.1f min', time(plot_idx(I(j)))),'color',0.95*ones(1,3),'HorizontalAlignment','center','VerticalAlignment','middle', 'fontsize', 6)
         end
         saveas(gcf, 'movie_registered_ordered.pdf');
-        return
     end
 end
 
@@ -204,7 +203,7 @@ rank_corr = zeros(length(sample_vec), nbootstrap_samples);
 for j = 1:nbootstrap_points
     nimages = sample_vec(j);
     
-    j
+%     j
     
     for j2 = 1:nbootstrap_samples
         

### 2014-08-27 image_analysis_paper2/pnas_draft.tex
Added edits to paper

@@
 %
-Often data is collected in several experiments across many samples, and the first task is to organize the data for further analysis
+Often, data is collected across many samples in several experiments, and the first task is to organize the data for further analysis.
 %
@@
 %
-We validate this approach using a live imaging data set where both the correct rotational orientation and temporal order are known, as well as demonstrate using these methods to extract a developmental trajectory from a data set of fixed images to study dorsoventral patterning of the {\it Drosophila} embryo.
+We validate this approach using a live imaging data set where both the correct rotational orientation and temporal order are known, as well as use these methods to extract a developmental trajectory from a data set of fixed images collected to study dorsoventral patterning of the {\it Drosophila} embryo.
 
@@
 %
-This data set was obtained by live imaging of a single {\it Drosophila} embryo during the twenty minutes spanning late cellularization through gastrulation.
+This data set was obtained through live imaging of a single {\it Drosophila} embryo during the twenty minutes spanning late cellularization through gastrulation.
 %
@@
 %
-The entire data set contains $42$ frames, taken at $30$~second time intervals.
+The entire data set contains $40$ consecutive frames taken at $30$~second time intervals.
 %
@@
 %
-\fig~\ref{subfig:movie_registered_ordered} shows the images from \fig~\ref{subfig:movie_unregistered_unordered}, now registered and ordered, and the real time for each frame is indicated.
+\fig~\ref{subfig:movie_registered_ordered} shows the images from \fig~\ref{subfig:movie_unregistered_unordered}, now registered and ordered; the real time for each frame is also indicated.
 %
-With a small number of exceptions, the recovered ordering is consistent with the real time for each frame. 
+With a small number of exceptions, the recovered ordering is consistent with the real time dynamics. 
 %
@@
 %
-The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than $0.98$, indicating that our methodology can reproducibly order data of this type.
+The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than $0.95$, indicating that our methodology can reproducibly order data of this type.
+%
+However, arguably the most common source of noise within developmental biology experiments is interembryo variability. 
+%
+To measure the sensitivity of our proposed methodology to interembryo variability, we constructed synthetic time course data sets by selecting a frame from a random movie for each time point.
+% 
+When using vector diffusion maps to order these data sets, the median rank correlation coefficient was found to be 0.81, indicating that our methodology can still accurately recover the temporal order even under noisy conditions.
 
-Clearly, $42$ images are sufficient to accurately recover the rotational orientation and temporal order  in this specific developmental time frame.
+Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order  in this specific developmental time frame.
 %
@@
 %
-To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to calculate the accuracy of the recovered ordering as a function of the size of the data set.
+To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
 %
-\fig~\ref{subfig:movie_bootstrap} shows the average rank correlation coefficient as a function of the number of images, where each point is the average over $200$ bootstrap samples.
+\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images, where each point is the average over $200$ bootstrap samples.
 %
@@
 %
-We also showed that these results are consistent across multiple data sets, and presented how the accuracy of our methods varies as a function of the number of images.
-
-\subsection{Images of gastrulating embryos}
-
-%The way it goes now is that its not only not known a priori, but that we have embryo to embryo variability, both in sizes of embryos and in the levels of signals extracted from images.
-To test our algorithm under these more challenges conditions, we collected a dataset where......
-
-The data set presented in the previous section contained no interembryo variability. 
-%
-We will now demonstrate that our methods are sufficiently robust to order images from multiple embryos. 
-%
-\fig~\ref{subfig:fixed_images_unregistered_unordered} shows selected images from a set of $126$ images (see \SI) which cover a thirty minute time interval spanning late cellularization through gastrulation.
-%
-Each image is an optical cross-section of a {\em different} vertically oriented embryo fixed at a different (and unknown) developmental time.
-%
-The nuclei (gray) are labeled with DAPI, a DNA stain.
-%
-Embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor which specifies the cells of the future muscle tissue.
-%
-Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, red in the figure), an enzyme that specifies a subset of neuronal cells.
-
-\fig~\ref{subfig:fixed_images_registered_ordered} shows the images from \fig~\ref{subfig:fixed_images_unregistered_unordered}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
-%
-However,  the selected images are not entirely representative of the developmental dynamics due to interembryo variability. 
-%
-To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory using a Gaussian filter across the set of registered and ordered images. 
-%
-Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:fixed_images_average_trajectory}, serve as a summary of the stereotypic developmental dynamics reconstructed from cross-sectional data.
-%
-We can now easily see the developmental progression, which is consistent with known dynamics. 
-%
-dpERK first appears as two lateral peaks at the ventral side of the embryo; a third dpERK peak then appears at the dorsal side of the embryo.
-%
-During invagination, the two ventral dpERK peaks merge together, eventually forming (together with Twi) a characteristic ``omega'' shape.
-%
-The dorsal dpERK peak then disappears during germband extension as cells from the ventral side wrap around to the dorsal side.
-%
-At the end of this process, similar ``omega'' shaped patterns formed by Twi and dpERK appear on the dorsal side; these patterns are most readily seen in the last image of \fig~\ref{subfig:fixed_images_average_trajectory}.
-%
-Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental dynamics (e.g., correlation of age with body size).
-
-Each embryo is fixed prior to imaging, and the developmental time of each embryo cannot be easily estimated.
-%
-TODO: compare to Michael/Henry ordering
-
-Empirically, we found required computational time to be polynomial in the number of images, number of pixels, and the number of discretization points used to compute the pairwise rotations. 
-%
-The CPU time required was $\mathcal{O}(n)$ in the number of pixels and the number of rotations, and $\mathcal{O}(n^{1.5})$ in the number of images in the data set.
-%
-For $100$ images of $100 \times 100$ pixels, using $30$ rotation discritezation points, the computations took $15$~seconds on an Intel Core i7 2.93 GHz processor. 
-%
-Furthermore, the computation of the pairwise rotational alignments, which is the most computationally intensive portion of the calculation, is trivially parallelizable.
-%
-Therefore, this analysis is computationally feasible for much larger imaging data sets which are impractical to evaluate manually.
+We also showed that these results are consistent across multiple data sets, demonstrated that the method also performs well for data with interembryo variability, and showed how the accuracy of our methods varies as a function of the number of images.
 
@@
 \customlabel{subfig:fixed_images_average_trajectory}{\ref{fig:data2}{\it C}}
+\customlabel{subfig:fixed_images_rank_corr}{\ref{fig:data2}{\it D}}
 \end{figure*}
+
+\subsection{Images of gastrulating embryos}
+
+Live imaging data, such as the images presented in the previous section, represent only a small fraction of the imaging data collected in developmental biology experiments. 
+%
+Often, data is collected from many multiple embryos, each of which is fixed to arrest its development and contributes only a snapshot of a chemical or morphological process along its developmental trajectory \cite{jaeger2004dynamic, peter2011gene, fowlkes2008quantitative}.
+%
+Importantly, the ``age'' of any given embryo arrested in its development is often not quantitatively known; typically what is known is
+a certain time window to which a collection of embryos belongs \cite{ng2012large, richardson2014emage, castro2009automatic}.
+%
+In order to recover the developmental dynamics from such data sets, snapshots of different embryos must be registered and then ordered in time.
+
+\fig~\ref{subfig:fixed_images_unregistered_unordered} shows selected images from a set of $126$ images (see \SI  for the full data set) which cover a thirty minute time interval spanning late cellularization through gastrulation.
+%
+This data set is more complex than the live imaging data sets in that it contains many more images, each with several channels.
+%
+Each image is an optical cross-section of a {\em different} vertically oriented embryo fixed at a different (and unknown) developmental time.
+%
+The nuclei (gray) were labeled with DAPI, a DNA stain.
+%
+Embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor which specifies the cells of the future muscle tissue.
+%
+Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that specifies a subset of neuronal cells.
+
+\fig~\ref{subfig:fixed_images_registered_ordered} shows the images from \fig~\ref{subfig:fixed_images_unregistered_unordered}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
+%
+However, because of interembryo variability, the selected images are not entirely representative of the developmental dynamics across the entire data set.
+%
+To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI). 
+%
+Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:fixed_images_average_trajectory}, serve as a summary of the stereotypic developmental dynamics.
+%
+We can now easily see the developmental progression, which is consistent with known dynamics. 
+%
+dpERK first appears as two lateral peaks at the ventral side of the embryo; a third dpERK peak then appears at the dorsal side of the embryo.
+%
+During invagination, the two ventral dpERK peaks merge together, eventually forming (together with Twi) the ``omega'' shape.
+%
+The dorsal dpERK peak then disappears during germband extension as cells from the ventral side wrap around to the dorsal side.
+%
+At the end of this process, similar ``omega'' shaped patterns formed by Twi and dpERK appear on the dorsal side; these patterns are most readily seen in the last image of \fig~\ref{subfig:fixed_images_average_trajectory}.
+%
+Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental dynamics (e.g., correlation of age with body size).
+
+Because the developmental time of each embryo cannot be easily estimated, we have few options for validating our ordering. 
+%
+We compared the ordering obtained from vector diffusion maps to a manual ordering provided by a trained embryologist who is aware of the developmental progression and the important image features. 
+%
+The results, shown in  \fig~\ref{subfig:fixed_images_rank_corr}, show that our ordering is consistent with the manual ordering.
+%
+However, manually ordering the images is nontrivial for researchers who are not familiar with the developmental progression, and can be tedious and time-consuming for those who are.
+%
+In contrast, our method requires relatively little computational time, taking $15$~seconds on an Intel Core i7 2.93 GHz processor for $100$ images of $100 \times 100$ pixels, using 10$^{\circ}$ rotational discretizations.
+%
+Empirically, we found the computational time to be polynomial in the number of images, number of pixels, and the number of discretization points used to compute the pairwise rotations. 
+%
+The CPU time required was $\mathcal{O}(n)$ in the number of pixels and the number of rotations, and $\mathcal{O}(n^{1.5})$ in the number of images in the data set,.
+%
+Furthermore, the computation of the pairwise rotational alignments, which is the most computationally intensive portion of the calculation, is trivially parallelizable.
+%
+The requisite human intervention and tuning for our method is relatively small. 
+%
+Images must first be preprocessed so that the Euclidean distance between the pixels in informative; for our examples, this amounts to cropping the images to normalize the sizes of the embryo cross-sections, as well as blurring and normalizing the intensity of the nuclear signal, as the absolute intensity of this signal is not meaningful.
+%
+%The image resolution should be chosen as low as possible while still retaining the important image features, as the computational cost increases with the number of pixels.
+%
+The two algorithmic parameters are the number of discretization points required to compute the pairwise alignments, and the kernel scale.
+%
+In our examples, we use 10$^{\circ}$ angle discretizations for the pairwise alignments, and one-tenth of the median pairwise distance for the kernel scale. 
+%
+We found that the results are not particularly sensitive to either of these parameters %(5$^{\circ}$--15$^{\circ}$ angle discretizations and kernel scales on the order of the median pairwise distances all yielded acceptable results)
+, and that these parameter values were appropriate for all of the data sets presented in this paper. 
+%
+Image preprocessing and selecting these two parameters are relatively simple tasks compared to manual registration and ordering of images, especially when the data sets are large.  
+%
+Therefore, this methodology is very promising for much larger imaging data sets which are impractical to evaluate manually.
 
@@
 In contrast to most of the existing approaches to image registration which rely on the knowledge about some landmarks in the data \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them applicable to a wide variety of biological applications.
+%
+Although we have presented results for a specific type of data (cross-sectional images of developing {\em Drosophila} embryos), because the methodologies only rely on computing distances between images, we are confident that they can be readily applied and extnded to many other imaging data sets.
 
@@
 %
-Based on the presented results, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories, and that the size of the data set required is a function of both the instrument noise as well as interembryo variablity.
+Based on the presented results, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories.
+%
+In general, the size of the data set required is a function of the instrument noise, interembryo variablity, and the length and complexity of the developmental dynamics.
 %
@@
 %
-This parameterization can then be used to do any typical data analysis tasks, such as outlier detection, clustering, and model fitting. 
+This parameterization can then be used for typical data analysis tasks, such as outlier detection, clustering, and model fitting. 
 %
@@
 Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.
-
-  \begin{itemize}
-  \item Will this work for more complex data sets (3-D imaging, different developmental processes)?
-  \item How dense does the sampling have to be (I don't think we can quantitatively say anything, but it has to be dense enough so that similar images have a small Euclidean distance)
-  \item Discuss parameter robustness/sensitity/tuning, how much effort is needed, etc.
-  \item How do the results depend on SNR?
-  \item `` One comment: the use of cross-sectional to describe image data from different specimen is highly confusing here, since the data in question do show a cross-section of Drosophila embryo''
-  \item Compare times and quality to Michael/Henry
-  \end{itemize}
 

### 2014-08-27 image_analysis_paper2/compute_pairwise_alignments.m

@@ -1,4 +1,8 @@
-function [R, W] = compute_pairwise_alignments(images, nrot, nshifts, shift_max)
+function [R, W] = compute_pairwise_alignments(images, nrot, nshifts, shift_max, display_waitbar)
+
+if nargin < 5
+    display_waitbar = true;
+end
 
 m = size(images, ndims(images));
 npixels = size(images, 1);
@@ -28,13 +32,19 @@ W = inf(m);
 % reference image
 RI = imref2d([npixels npixels],[-0.5 0.5],[-0.5 0.5]);
 
-h_waitbar = waitbar(0,'Computing pairwise alignments...');
+if display_waitbar
+    h_waitbar = waitbar(0,'Computing pairwise alignments...');
+end
 for i=1:nrot
 %     waitbar(i/nrot, h_waitbar);
     theta = theta_vec(i);
     for j=1:nshifts
         for k=1:nshifts
-            waitbar(((i-1)*nshifts^2+(j-1)*nshifts+k)/(nrot*nshifts^2), h_waitbar);
+            
+            if display_waitbar
+                waitbar(((i-1)*nshifts^2+(j-1)*nshifts+k)/(nrot*nshifts^2), h_waitbar);
+            end
+            
             A = affine2d([cosd(theta) -sind(theta) 0; sind(theta) cosd(theta) 0; shifts(j) shifts(k) 1]);
             images_transformed = imwarp(images, RI, A, 'outputview', RI);
             dist_tmp = pdist2(reshape(double(images), [], m)', reshape(double(images_transformed), [], m)');
@@ -64,7 +74,9 @@ end
 W = 0.5*(W + W');
 R = 0.5*(R + R');
 
-close(h_waitbar);
+if display_waitbar
+    close(h_waitbar);
+end
 
 function R = calc_rot_matrix_3d(dtheta, dx, dy, angle_proj)
 

### 2014-08-27 image_analysis_paper2/make_figures_onemovie.m

@@ -9,8 +9,9 @@ dt = 0.5;
 
 nmovies = 6;
 
+nimages = 40;
 idx_end = [56 60 47 44 46 42];
-idx_start = idx_end - 41;
+idx_start = idx_end - nimages + 1;
 
 nrot = 36;
 nshifts = 0;
@@ -27,7 +28,11 @@ theta_err = zeros(nmovies, 1);
 rank_corr = zeros(nmovies, 1);
 
 %%
-for k=1:nmovies
+movies_to_use = [2 1 3 5 6];
+
+for k1 = 1:length(movies_to_use)
+    k = movies_to_use(k1);
+    
     load(sprintf('movie%d.mat', k));
     idx = idx_start(k):idx_end(k);
     
@@ -75,21 +80,22 @@ for k=1:nmovies
         R_tmp = R_opt(dim*(i-1)+1:dim*i, :);
         theta_opt(i) = atan2d(R_tmp(2,1), R_tmp(1,1));
     end
-    fprintf('movie %d, average error = %2.2f \n', k, std(mod(theta - theta_opt, 360)))
-    theta_err(k) =  std(mod(theta - theta_opt, 360));
+    %     fprintf('movie %d, average error = %2.2f \n', k, std(mod(theta - theta_opt, 360)))
+    theta_err(k) =  min(std(mod(theta - theta_opt, 360)),std(mod(theta - theta_opt+180, 360)));
     
     %
     if corr(time, embed_coord(:,1)) < 0
         embed_coord(:,1) = -embed_coord(:,1);
     end
     
-    fprintf('movie %d, rank corr = %2.4f \n', k, corr(compute_ranks(time), compute_ranks(embed_coord(:,1))))
+    %     fprintf('movie %d, rank corr = %2.4f \n', k, corr(compute_ranks(time), compute_ranks(embed_coord(:,1))))
     rank_corr(k) = corr(compute_ranks(time), compute_ranks(embed_coord(:,1)));
     
     if k == k_print_fig
+        
         [~, max_idx] = max(mod(theta, 360));
         make_fig(4.3,4);
-        plot(mod(theta, 360), mod(theta_opt-theta_opt(max_idx)-1, 360)+1, '.')
+        plot(mod(theta, 360), mod(theta_opt-theta_opt(max_idx)-10, 360)+10, '.')
         xlabel('true angle')
         ylabel('recovered angle')
         set(gca, 'xtick', [0 180 360]);
[…]

### 2014-08-28 image_analysis_paper2/pnas_draft.tex
added stuff for paper

@@
 %
-We found that the results are not particularly sensitive to either of these parameters %(5$^{\circ}$--15$^{\circ}$ angle discretizations and kernel scales on the order of the median pairwise distances all yielded acceptable results)
+We found that the results are not particularly sensitive to either of these parameters (5$^{\circ}$--30$^{\circ}$ angle discretizations and kernel scales of one-fifth to one-twentieth of the median pairwise distances all yielded acceptable results)
 , and that these parameter values were appropriate for all of the data sets presented in this paper. 

### 2014-08-28 image_analysis_paper2/test_parameter_robustness_fixed_images.m

@@ -1,0 +1,99 @@
+clear all
+close all
+
+%%
+
+npixels = 100;
+
+nimages = 132;
+image_dir = '../membrane_pictures/14_0501_dpERK_late';
+image_name = 'emb';
+
+neigs = 10;
+alpha = 0;
+
+channel = 1;
+nchannels = 3;
+
+image_set = zeros(npixels, npixels, nchannels, nimages, 'uint8');
+
+%%
+
+for i=1:nimages
+    im_tmp = imread(sprintf('%s/%s%02d.tif', image_dir, image_name, i));
+    image_set(:, :, :, i) = image_fn_color(im_tmp, channel, npixels);
+end
+
+%%
+
+figure;
+for i=1:nimages
+    make_subplot(10, 14, 0.01, i);
+    imshow(image_set(:,:,:,i))
+end
+
+
+%%
+% ind = setdiff(1:nimages, [1 2 3 5 12 17 19 21 29 35 57 64 71 74 76 92 93 105 54 131 32 102 101 100]);
+%
+% image_set = image_set(:, :, :, ind);
+% nimages = length(ind);
+
+ind = setdiff(1:nimages, [32 91 28 77 46 116 5 17 1 2 3 29]);
+
+image_set = image_set(:, :, :, ind);
+nimages = length(ind);
+
+dim1 = 8;
+dim2 = 15;
+
+%%
+% make_fig(17, dim1*(17/dim2));
+% for i=1:nimages
+%     make_subplot(dim2, dim1, 0.01, i);
+%     imshow(make_gray_nuclei(image_set(:,:,:,i)))
+% end
+
+%%
+
+nrot = 36;
+nshifts = 0;
+shift_max = 0.1;
+
+[R, W] = compute_pairwise_alignments(image_set, nrot, nshifts, shift_max);
+dim = size(R, 1) / nimages;
+
+%
+W2 = W.^2;
+eps = median(W2(:))/10;
+[R_opt, embed_coord_true, D2, D] = vdm(R, W2, eps, neigs, alpha);
+
+%
+if embed_coord_true(1,1) < 0
+    embed_coord_true(:,1) = -embed_coord_true(:,1);
+end
+
+%%
+
+
+nrot = 36;
+nshifts = 0;
+shift_max = 0.1;
+
+[R, W] = compute_pairwise_alignments(image_set, nrot, nshifts, shift_max);
+dim = size(R, 1) / nimages;
+
+%
[…]

### 2014-08-28 image_analysis_paper2/pnas_draft.tex
added paper edits

--- 2014-08-27 image_analysis_paper2/pnas_draft.tex
@@
 
-\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional \cite{diggle2002analysis}.
+\dropcap{I}n one of the common approaches to studies of developmental dynamics, a group of embryos is fixed and stained with chemicals that visualize a particular biochemical or morphological process within a developing tissue.
 %
-In longitudinal studies, developmental progress is monitored over time in the same embryo \cite{roelens2013live, keller2013imaging}.
+The developmental dynamics must then be reconstructed from multiple embryos, each of which contributes only a snapshot of a chemical or morphological process along its developmental trajectory \cite{jaeger2004dynamic, peter2011gene, fowlkes2008quantitative}.
 %
-In a cross-sectional study, developmental dynamics is reconstructed from multiple embryos, each of which contributes only a snapshot of a chemical or morphological process along its developmental trajectory \cite{jaeger2004dynamic, peter2011gene, fowlkes2008quantitative}.
-%
-Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms.
-%
-In a typical cross-sectional study, a group of embryos is fixed using a procedure that arrests their development, and stained with chemicals that visualize a handful of cellular processes.
-%
-Fixed embryos are then imaged using any given number of microscopy techniques.
-%
-Importantly, the ``age'' of any given embryo arrested in its development is not quantitatively known; typically what is known is
+Importantly, the ``age'' of any given embryo arrested in its development is often not quantitatively known; typically what is known is
 a certain time window to which a collection of embryos belongs \cite{ng2012large, richardson2014emage, castro2009automatic}.
 %
-In order to recover the developmental dynamics from such data sets, snapshots of different embryos must first be spatially aligned or {\em registered} to factor out the relevant geometric symmetries (e.g., rotations), and then ordered in time.
+In order to recover the developmental dynamics from such data sets, snapshots of different embryos must first be spatially aligned or {\em registered}, and then ordered in time.
 %
-We show how recently developed dimensionality reduction algorithms can automate {\it and combine} both of these tasks.
+Here, we present an automated approach to both of these tasks.
+%We show how recently developed dimensionality reduction algorithms can automate {\it and combine} both of these tasks.
 
@@
 %
-Image registration is based on obvious morphological landmarks, such as the positions of the head and the fins.
+Image registration can be based on obvious morphological landmarks, such as the positions of the head and the fins.
 %
@@
 
-Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, 
-most of which require the definition and identification of appropriate landmarks.
+We describe an algorithmic approach to registration and temporal ordering that is robust to these sources of error.
 %
-In contrast, our approach requires no such information. 
+In contrast to a number of previous approaches, it does not rely on the {\em a priori} knowledge of spatial landmarks and markers of developmental progression that can be used to align and order images \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}.
 %
@@
 %
-This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}.
-%
-These techniques has been used in
+This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, with
 applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \cite{zhao2014rotationally, singer2011viewing} to face recognition \cite{lafon2006data}.
@@
 %
-Our first data set, where the correct rotational orientation and temporal order are independently known, will be used to validate our approach.
+Our first data set, from a live imaging study where the correct rotational orientation and temporal order are independently known, will be used to validate our approach.
 %
-The correct orientation and order is unknown for the second data set; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.
+The correct orientation and order is unknown for the second data set comprised of images from fixed embryos; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.
 
@@
 technique developed for data sets which contain two types of sources of variability:
-geometric symmetries (such as rotations of the images due to the experimental setup) which one would like to factor out,
-and ``additional" directions of variability (such as temporal dynamics) which one wants to uncover.
+geometric symmetries (such as rotations of the images) which one would like to factor out,
+and ``additional" directions of variability (such as temporal dynamics) which one would like to uncover.
 %
@@
 %
-We will use the algorithm to register images of {\it Drosophila} embryos with respect to rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
+We will use the algorithm to register images of developing embryos with respect to rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
 %
-We assume that the main direction of variability in these images is parameterized by the developmental time of each embryo, so that uncovering this direction will allow us to recover the developmental dynamics.
+We assume that the main direction of variability in these images is parameterized by the developmental time of each embryo.
+%
+As a consequence, uncovering this direction will allow us to recover the developmental dynamics.
 
@@
 %
-We use {\it local} information about the data to find a parametrization  which respects the underlying manifold geometry: we want points which are close in high-dimensional space (e.g., images which look similar) to be close in our parametrization.
+We use {\it local} information about the data to find a parametrization  which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parametrization.
 %
@@
 \includegraphics{fig3}
-\caption{{\xfigtextfontit Drosophila} embryos during cellularization.
-{\xfigtextfontit (A)} Selected images from a live movie of a {\xfigtextfontit Drosophilla} embryo during gastrulation.
+\caption{Method validation using live imaging of {\xfigtextfontit Drosophila} embryos.{\xfigtextfontit (A)} Selected images from a live imaging study of a {\xfigtextfontit Drosophilla} embryo during gastrulation.
 Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
@@
 {\xfigtextfontit (E)} The average error in the recovered angle and the rank correlation coefficient for $5$ different movies.
-{\xfigtextfontit (F)} The average rank correlation coefficient as a function of the number of images in the data set.
-Each point is averaged over $200$ samples.
+{\xfigtextfontit (F)} The median rank correlation coefficient (over $200$ samples) as a function of the number of images in the data set.
 }
@@
 
-To validate the proposed approach, we applied our algorithm to a dataset where the true temporal order and rotational orientation are known {\em a priori}.
+To validate the proposed approach, we applied our algorithm to a dataset where the true temporal order and rotational orientation of the images are known {\em a priori}.
 %
-This data set was obtained through live imaging of a single {\it Drosophila} embryo during the twenty minutes spanning late cellularization through gastrulation.
+This data set was obtained through live imaging of a single {\it Drosophila} embryo during the twenty minutes spanning late stages of cellularization through early gastrulation.
 %
-During this time window, the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, bringing in the future muscle cells and eventually forming a characteristic ``omega'' shape.
+During this time window, the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, internalizing the future muscle cells, eventually forming a characteristic ``omega'' shape.
 %
@@
 %
-At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing a similar ``omega'' shaped pattern to appear on the dorsal side.
+At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing a similar ``omega'' to appear on the dorsal side.
 
@@
 The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than $0.95$, indicating that our methodology can reproducibly order data of this type.
-%
-However, arguably the most common source of noise within developmental biology experiments is interembryo variability. 
-%
-To measure the sensitivity of our proposed methodology to interembryo variability, we constructed synthetic time course data sets by selecting a frame from a random movie for each time point.
-% 
-When using vector diffusion maps to order these data sets, the median rank correlation coefficient was found to be 0.81, indicating that our methodology can still accurately recover the temporal order even under noisy conditions.
 
-Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order  in this specific developmental time frame.
-%
 In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{subfig:dmaps1}) is well-sampled;
 the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \cite{singer2011angular}.
+%
+Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order  in this specific developmental time frame.
 %
@@
 %
-We also showed that these results are consistent across multiple data sets, demonstrated that the method also performs well for data with interembryo variability, and showed how the accuracy of our methods varies as a function of the number of images.
+We also demonstrated that these results are consistent across multiple data sets, and showed how the accuracy of our methods varies as a function of the number of images.
 
-\begin{figure*}[t]
-\includegraphics{fig4}
-\caption{{\xfigtextfontit Drosophila} embryos during gastrulation.
-{\xfigtextfontit (A)} Images of gastrulating embryos.
-Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation.
-{\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps.
-The images are ordered from left to right and from top to bottom, so that the first image appears in the top left position of the array, and the last image appears in the bottom right.
-{\xfigtextfontit (C)} A representative ``developmental path'' from local averaging of the images.
-}
-\customlabel{fig:data2}{4}
-\customlabel{subfig:fixed_images_unregistered_unordered}{\ref{fig:data2}{\it A}}
-\customlabel{subfig:fixed_images_registered_ordered}{\ref{fig:data2}{\it B}}
-\customlabel{subfig:fixed_images_average_trajectory}{\ref{fig:data2}{\it C}}
-\customlabel{subfig:fixed_images_rank_corr}{\ref{fig:data2}{\it D}}
-\end{figure*}
+\subsection{Datasets with interembryo variability}
 
-\subsection{Images of gastrulating embryos}
+Arguably, the most common source of noise within developmental biology experiments is embryo-to-embryo variability. 
+%
+We would like to demonstrate that our methods are robust to such noise sources. 
+%
+We constructed synthetic time course data sets by selecting a frame from a random live imaging experiment for each time point, and used our methodology to register and order these sets. 
+% 
+The median rank correlation coefficient for these synthetic time courses was found to be 0.81, indicating that our methodology can still accurately recover the temporal order even under noisy conditions.
 
-Live imaging data, such as the images presented in the previous section, represent only a small fraction of the imaging data collected in developmental biology experiments. 
+Because live imaging data represents only a small fraction of the data collected in developmental biology experiments, the utility of our proposed methodology is in the analysis of large data sets of fixed images, where the developmental times and spatial orientations are unknown.
 %
-Often, data is collected from many multiple embryos, each of which is fixed to arrest its development and contributes only a snapshot of a chemical or morphological process along its developmental trajectory \cite{jaeger2004dynamic, peter2011gene, fowlkes2008quantitative}.
-%
-Importantly, the ``age'' of any given embryo arrested in its development is often not quantitatively known; typically what is known is
-a certain time window to which a collection of embryos belongs \cite{ng2012large, richardson2014emage, castro2009automatic}.
-%
-In order to recover the developmental dynamics from such data sets, snapshots of different embryos must be registered and then ordered in time.
-
-\fig~\ref{subfig:fixed_images_unregistered_unordered} shows selected images from a set of $126$ images (see \SI  for the full data set) which cover a thirty minute time interval spanning late cellularization through gastrulation.
+To demonstrate such an application, \fig~\ref{subfig:fixed_images_unregistered_unordered} shows selected images from a set of $120$ images (see \SI  for the full data set) which cover a thirty minute time interval spanning late cellularization through gastrulation.
 %
@@
 %
-Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that specifies a subset of neuronal cells.
+Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that, in this context, specifies a subset of neuronal cells.
+
+\begin{figure*}[t]
+\includegraphics{fig4}
+\caption{Fixed {\xfigtextfontit Drosophila} embryos during gastrulation.
+{\xfigtextfontit (A)} Images of gastrulating embryos, stained for nuclei (gray), Twi (green), and dpERK (red).
+Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation.
+{\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps.
+The expert rank for each image is indicated.
+{\xfigtextfontit (C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI).
+{\xfigtextfontit (D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert.
+The rank correlation coefficient is $0.97$.
+}
+\customlabel{fig:data2}{4}
+\customlabel{subfig:fixed_images_unregistered_unordered}{\ref{fig:data2}{\it A}}
+\customlabel{subfig:fixed_images_registered_ordered}{\ref{fig:data2}{\it B}}
+\customlabel{subfig:fixed_images_average_trajectory}{\ref{fig:data2}{\it C}}
+\customlabel{subfig:fixed_images_rank_corr}{\ref{fig:data2}{\it D}}
+\end{figure*}
 
@@
 
-Because the developmental time of each embryo cannot be easily estimated, we have few options for validating our ordering. 
+To evaluate the quality of our registration and ordering, we can use prior knowledge about the developmental dynamics. 
 %
-We compared the ordering obtained from vector diffusion maps to a manual ordering provided by a trained embryologist who is aware of the developmental progression and the important image features. 
+The Twi signal is known to form a Gaussian peak at the ventralmost point of the embryo. 
 %
-The results, shown in  \fig~\ref{subfig:fixed_images_rank_corr}, show that our ordering is consistent with the manual ordering.
+TODO: add more...
+%
+Because the developmental time of each embryo cannot be easily estimated, we have few options for evaluating the quality of our ordering. 
+%
+We compared the ordering obtained from vector diffusion maps to an expert ordering provided by a trained embryologist who is knowledgeable about the developmental progression and the important image features.
+%
+The results (see \fig~\ref{subfig:fixed_images_rank_corr}) show that our ordering is consistent with the expert ordering. 
 %
@@
 %
-In contrast, our method requires relatively little computational time, taking $15$~seconds on an Intel Core i7 2.93 GHz processor for $100$ images of $100 \times 100$ pixels, using 10$^{\circ}$ rotational discretizations.
+In contrast, our method requires relatively little computational time, taking $15$~seconds on an Intel Core i7 2.93 GHz processor.
+%for $100$ images of $100 \times 100$ pixels, using 10$^{\circ}$ rotational discretizations.
 %
-Empirically, we found the computational time to be polynomial in the number of images, number of pixels, and the number of discretization points used to compute the pairwise rotations. 
-%
-The CPU time required was $\mathcal{O}(n)$ in the number of pixels and the number of rotations, and $\mathcal{O}(n^{1.5})$ in the number of images in the data set,.
+Empirically, we found the CPU time required was $\mathcal{O}(n)$ in the number of pixels and the number of rotations, and $\mathcal{O}(n^{1.5})$ in the number of images in the data set.
 %
@@
 %
-Images must first be preprocessed so that the Euclidean distance between the pixels in informative; for our examples, this amounts to cropping the images to normalize the sizes of the embryo cross-sections, as well as blurring and normalizing the intensity of the nuclear signal, as the absolute intensity of this signal is not meaningful.
+Images must first be minimally preprocessed so that the Euclidean distance between the pixels in informative. 
 %
@@
 %
-We found that the results are not particularly sensitive to either of these parameters %(5$^{\circ}$--15$^{\circ}$ angle discretizations and kernel scales on the order of the median pairwise distances all yielded acceptable results)
+We found that the results are robust to both of these parameters% (5$^{\circ}$--30$^{\circ}$ angle discretizations and kernel scales of one-fifth to one-twentieth of the median pairwise distances all yielded acceptable results)
 , and that these parameter values were appropriate for all of the data sets presented in this paper. 
@@
 %
-The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
+%The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
 %
-In contrast to most of the existing approaches to image registration which rely on the knowledge about some landmarks in the data \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them applicable to a wide variety of biological applications.
+In contrast to most of the existing approaches to image registration which rely on the knowledge about some landmarks in the data \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them applicable to a wide variety of applications.
 %
-Although we have presented results for a specific type of data (cross-sectional images of developing {\em Drosophila} embryos), because the methodologies only rely on computing distances between images, we are confident that they can be readily applied and extnded to many other imaging data sets.
+Although we have presented results for a specific type of data (images along the dorsoventral axis of developing {\em Drosophila} embryos), because the methodology only relies on computing distances between images, we are confident that they can be readily applied and extended to many other imaging data sets.
 
@@
 %
-In general, the size of the data set required is a function of the instrument noise, interembryo variablity, and the length and complexity of the developmental dynamics.
+In general, the size of the data set required is a function of the instrument noise, interembryo variablity, and the complexity of the developmental dynamics.
 %

### 2014-08-28 image_analysis_paper2/make_figures_fixed_images.m

@@ -38,12 +38,9 @@ end
 
 
 %%
-% ind = setdiff(1:nimages, [1 2 3 5 12 17 19 21 29 35 57 64 71 74 76 92 93 105 54 131 32 102 101 100]);
-%
-% image_set = image_set(:, :, :, ind);
-% nimages = length(ind);
 
-ind = setdiff(1:nimages, [32 91 28 77 46 116 5 17 1 2 3 29]);
+images_to_remove = [32 91 28 77 46 116 5 17 1 2 3 29];
+ind = setdiff(1:nimages, images_to_remove);
 
 image_set = image_set(:, :, :, ind);
 nimages = length(ind);
@@ -95,6 +92,50 @@ end
 
 %%
 
+% I_manual = [17 29 92 64 5 76 71 35 12 19 74 57 93 21 105  121 54 108 130 2 ...
+% 1 62 3 44 127 20 36 9 111 49 96 120 41 26 131 112 60 123 30 70 ...
+% 102 63 34 117 124 115 104 82 48 97 83 40 81 69 85 67 7 14 58 50 ...
+% 55 27 65 22 129 75 78 18 52 119 16 94 98 125 47 109 101 86 37 114 ...
+% 33 42 110 79 59 38 66 99 24 4 6 39 13 87 90 88 8 106 103 132 ...
+% 56 95 118 80 23 100 68 45 107 126 31 25 128 51 84 53 11 43 10 73 ...
+% 77 32 116 46 122 113 61 89 91 15 72 28];
+
+I_manual = [17 5 29 64 92 35 12 71 57 76 19 74 93 21 105 54 121 130 108 127 ...
+2 3 9 44 62 1 26 20 49 36 111 131 120 123 112 117 96 60 70 30 ...
+41 124 115 102 34 104 63 48 97 82 81 83 40 69 85 67 58 7 14 50 ...
+55 27 65 129 18 22 75 52 78 16 119 47 98 94 125 101 86 114 33 37 ...
+109 42 100 110 80 99 38 59 66 24 6 4 39 90 56 8 13 106 88 87 ...
+132 95 103 118 68 107 31 25 23 79 45 126 53 11 128 51 84 10 43 73 ...
+77 32 116 46 122 15 113 61 89 91 72 28];
+
+for i=sort(images_to_remove, 'descend');
+    I_manual(I_manual == i) = [];
+    idx = find(I_manual > i);
[…]

### 2014-08-28 image_analysis_paper2/pnas_draft.tex
edited paper

@@
 
-\dropcap{I}n one of the common approaches to studies of developmental dynamics, a group of embryos is fixed and stained with chemicals that visualize a particular biochemical or morphological process within a developing tissue.
+\dropcap{I}n one of the common approaches to studies of development, a group of embryos is fixed and stained to visualize a particular biochemical or morphological process within a developing tissue. 
 %
@@
 %
-Here, we present an automated approach to both of these tasks.
 %We show how recently developed dimensionality reduction algorithms can automate {\it and combine} both of these tasks.
@@
 %
-Image registration can be based on obvious morphological landmarks, such as the positions of the head and the fins.
+Image registration is based on obvious morphological landmarks, such as the positions of the head and the fins.
 %
@@
 
-We describe an algorithmic approach to registration and temporal ordering that is robust to these sources of error.
+We present a robust algorithmic approach to registration and temporal ordering.
 %
-In contrast to a number of previous approaches, it does not rely on the {\em a priori} knowledge of spatial landmarks and markers of developmental progression that can be used to align and order images \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}.
+In contrast to a number of previous methodologies, it does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression for ordering images\cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}.
 %
-It is based on a manifold learning algorithm (vector diffusion maps, VDM \cite{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
+Our approach is based on a manifold learning algorithm (vector diffusion maps \cite{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
 %
-This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, with
+This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
 applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \cite{zhao2014rotationally, singer2011viewing} to face recognition \cite{lafon2006data}.
 %
-Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing embryos from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
+Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing embryos in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
 %
-To illustrate our approach, we analyze two data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics \cite{jaeger2012drosophila}.
+To illustrate our approach, we analyze two data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \cite{jaeger2012drosophila}.
 %
@@
 %
-The correct orientation and order is unknown for the second data set comprised of images from fixed embryos; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.
+Our second data set consists of images from fixed embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.
 
@@
 The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.
-{\xfigtextfontit (C)} The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle for each frame in the movie.
-{\xfigtextfontit (D)} The correlation between the recovered rank (using vector diffusion maps) and the true rank for each frame in the movie.
-{\xfigtextfontit (E)} The average error in the recovered angle and the rank correlation coefficient for $5$ different movies.
+{\xfigtextfontit (C)} The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle {\xfigtextfontit (D)} The correlation between the recovered rank (using vector diffusion maps) and the true rank.
+{\xfigtextfontit (E)} The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies.
 {\xfigtextfontit (F)} The median rank correlation coefficient (over $200$ samples) as a function of the number of images in the data set.
@@
 %
-This data set was obtained through live imaging of a single {\it Drosophila} embryo during the twenty minutes spanning late stages of cellularization through early gastrulation.
+This data set was obtained through live imaging of a single {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
 %
-During this time window, the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, internalizing the future muscle cells, eventually forming a characteristic ``omega'' shape.
+During this time window, the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, internalizing the future muscle cells and forming a characteristic ``omega'' shape.
 %
@@
 %
-The entire data set contains $40$ consecutive frames taken at $30$~second time intervals.
+The entire data set contains $40$ consecutive frames taken at $30$~second time intervals at a fixed position within a single embryo.
 %
-Each image shows an optical cross-section at a fixed depth of a vertically oriented developing embryo, with the nuclei labeled by Histone-GFP.
+Each image shows an optical cross-section of a vertically oriented developing embryo, with the nuclei labeled by Histone-GFP.
 %
@@
 %
-The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than $0.95$, indicating that our methodology can reproducibly order data of this type.
+The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than 95\%, indicating that our methodology can reproducibly order data of this type.
 
@@
 %
-The errors in the recovered angles did not vary significantly with the number of images and were consistently less than $10^{\circ}$;
-because our particular data set contains relatively low noise levels, few images are required to obtain accurate rotational alignments. 
+Because our particular data set contains relatively low noise levels, few images are required to obtain accurate rotational alignments and the errors in the recovered angles were consistently less than $10^{\circ}$.
 %
-We have therefore demonstrated that the algorithm accurately registers and orders a model data set where both the rotation angles and real times are independently known. 
+Therefore, the algorithm accurately registers and orders a model data set where both the rotation angles and real times are independently known. 
 %
-We also demonstrated that these results are consistent across multiple data sets, and showed how the accuracy of our methods varies as a function of the number of images.
+In addition, these results are consistent across multiple data sets, and we showed how the accuracy of our methods varies as a function of the number of images.
 
@@
 %
-We would like to demonstrate that our methods are robust to such noise sources. 
+To demonstrate that our methods are robust to such noise sources,
+we constructed a synthetic time course data set by selecting a frame from a random live imaging experiment for each time point.
 %
-We constructed synthetic time course data sets by selecting a frame from a random live imaging experiment for each time point, and used our methodology to register and order these sets. 
-% 
-The median rank correlation coefficient for these synthetic time courses was found to be 0.81, indicating that our methodology can still accurately recover the temporal order even under noisy conditions.
+The median rank correlation coefficient for ordering such a synthetic time course using our methodology was 0.81, indicating that the algorithm can accurately recover the temporal order even under noisy conditions.
 
@@
 %
-This data set is more complex than the live imaging data sets in that it contains many more images, each with several channels.
+This data set is more complex than the live imaging data sets in that it contains significantly more images, each with several channels.
 %
-Each image is an optical cross-section of a {\em different} vertically oriented embryo fixed at a different (and unknown) developmental time.
+Each image shows an optical cross-section of a {\em different} vertically oriented embryo at a different rotational orientation and fixed at a different (and unknown) developmental time.
 %
@@
 \includegraphics{fig4}
-\caption{Fixed {\xfigtextfontit Drosophila} embryos during gastrulation.
-{\xfigtextfontit (A)} Images of gastrulating embryos, stained for nuclei (gray), Twi (green), and dpERK (red).
+\caption{Analyzing images of fixed {\xfigtextfontit Drosophila} embryos.
+{\xfigtextfontit (A)} Images of {\xfigtextfontit Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red).
 Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation.
@@
 %
-We can now easily see the developmental progression, which is consistent with known dynamics. 
+We can now easily see the developmental progression, which is consistent with the known dynamics. 
 %
@@
 %
-During invagination, the two ventral dpERK peaks merge together, eventually forming (together with Twi) the ``omega'' shape.
+During invagination, the two ventral dpERK peaks merge together, eventually forming, together with Twi, the ``omega'' shape.
 %
@@
 
-To evaluate the quality of our registration and ordering, we can use prior knowledge about the developmental dynamics. 
+To evaluate the quality of our registration and ordering, we can use prior knowledge about the developmental dynamics of this specific system. 
 %
-The Twi signal is known to form a Gaussian peak at the ventralmost point of the embryo. 
+The Twi signal is known to form a single peak at the ventralmost point of the embryo. 
 %
-TODO: add more...
+We found that the standard deviation in the location of the peak of the Twi signal was 8$^\circ$ after registration, indicating that our registration is consistent with the known biology. 
 %
@@
 %
-The results (see \fig~\ref{subfig:fixed_images_rank_corr}) show that our ordering is consistent with the expert ordering. 
+The ranks for the images in \fig~\ref{subfig:fixed_images_registered_ordered} as determined by this expert ordering are indicated, and the rank correlation (see \fig~\ref{subfig:fixed_images_rank_corr}) shows that our ordering is consistent with the expert ordering.
 %
@@
 %
-In contrast, our method requires relatively little computational time, taking $15$~seconds on an Intel Core i7 2.93 GHz processor.
+In contrast, our method requires relatively little computational time; registering and ordering this fixed imaging data set took $15$~seconds on an Intel Core i7 2.93 GHz processor.
 %for $100$ images of $100 \times 100$ pixels, using 10$^{\circ}$ rotational discretizations.
 %
-Empirically, we found the CPU time required was $\mathcal{O}(n)$ in the number of pixels and the number of rotations, and $\mathcal{O}(n^{1.5})$ in the number of images in the data set.
+The computational time is a function of the number of images in the dataset, the number of pixels in each point, and the number of angles used to compute the pairwise rotations (see \SI). 
 %
-Furthermore, the computation of the pairwise rotational alignments, which is the most computationally intensive portion of the calculation, is trivially parallelizable.
+Empirically, the CPU time required was $\mathcal{O}(n^{1.5})$ in the number of images, and $\mathcal{O}(n)$ in the number of pixels and the number of angles.
 %
-The requisite human intervention and tuning for our method is relatively small. 
+Furthermore, the computation of the pairwise rotational alignments, which is the most time-intensive portion of the calculation, is trivially parallelizable.
 %
-Images must first be minimally preprocessed so that the Euclidean distance between the pixels in informative. 
+The parameter tuning required for our method is minimal.
 %
-%The image resolution should be chosen as low as possible while still retaining the important image features, as the computational cost increases with the number of pixels.
+Images must first be preprocessed so that the Euclidean distance between the pixels in informative, and two algorithmic parameters, the number of angles to compute the pairwise alignments and the diffusion maps kernel scale (see \SI), must be defined.
 %
-The two algorithmic parameters are the number of discretization points required to compute the pairwise alignments, and the kernel scale.
+We found that the results are robust to both of these parameters, and that consistent parameter values were appropriate for all of the data sets presented in this paper. 
 %
-In our examples, we use 10$^{\circ}$ angle discretizations for the pairwise alignments, and one-tenth of the median pairwise distance for the kernel scale. 
+%In our examples, we use 10$^{\circ}$ angle discretizations for the pairwise alignments, and one-tenth of the median pairwise distance for the kernel scale. 
+% (5$^{\circ}$--30$^{\circ}$ angle discretizations and kernel scales of one-fifth to one-twentieth of the median pairwise distances all yielded acceptable results)
 %
-We found that the results are robust to both of these parameters% (5$^{\circ}$--30$^{\circ}$ angle discretizations and kernel scales of one-fifth to one-twentieth of the median pairwise distances all yielded acceptable results)
-, and that these parameter values were appropriate for all of the data sets presented in this paper. 
-%
-Image preprocessing and selecting these two parameters are relatively simple tasks compared to manual registration and ordering of images, especially when the data sets are large.  
+The tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, especially when the data sets are large.  
 %
@@
 %
-In contrast to most of the existing approaches to image registration which rely on the knowledge about some landmarks in the data \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them applicable to a wide variety of applications.
+In contrast to most of the existing registration approaches which rely on the knowledge of appropriate landmarks in the images \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them relevant for a wide variety of applications.
 %
-Although we have presented results for a specific type of data (images along the dorsoventral axis of developing {\em Drosophila} embryos), because the methodology only relies on computing distances between images, we are confident that they can be readily applied and extended to many other imaging data sets.
+Although we have presented results for a specific type of data (cross-sectional images taken along the dorsoventral axis of developing {\em Drosophila} embryos), because the methodology only relies on computing distances between images, we are confident that it can be readily applied and extended to many other imaging data sets.
 
@@
 %
-In general, the size of the data set required is a function of the instrument noise, interembryo variablity, and the complexity of the developmental dynamics.
+In general, the size of the data set required for accurate registration and ordering is a function of the instrument noise, interembryo variablity, and the complexity of the developmental dynamics.
 %

### 2014-08-29 image_analysis_paper2/SI_text.tex
added paper edits

--- 2014-07-10 image_analysis_paper/SI_text.tex
@@
 
-The original resolution of each image is $1024 \times 1024$ pixels.
-%
-The images are subsampled to $100 \times 100$ pixels for vector diffusion maps analysis.
+The original image resolution is $1024 \times 1024$ pixels for the fixed images, and $512 \times 512$ pixels for the live images; all images are subsampled to $100 \times 100$ pixels for vector diffusion maps analysis.
 %
@@
 %
-We blur the nuclear signal using a disc filter with a $5$~pixel radius (using the \texttt{imfilter} function in MATLAB\textsuperscript{\textregistered}).
+The Canny method is used to detect the edges of the nuclear signal in each image (using the \texttt{edge} function in MATLAB\textsuperscript{\textregistered}), and images are cropped to these edges to remove any size effects.
 %
-We also adjust the intensity of the nuclear signal in each image such that 1\% of the pixels are saturated 
-(using the \texttt{imadjust} function in MATLAB), and then scale the intensity of the nuclear signal by half.
-
-For the first and third data sets in the paper (images taken during cellularization), we only use the red and green channels (dpERK and Dl) in the vector diffusion maps computations, as the spatial location of the nuclei (the gray channel) does not change appreciably during this time window.
+A constant $10$-pixel border is then added to each image. 
 %
-The nuclei in the corresponding figures are thus shown only for visual reference.
+Contrast-limited adaptive histogram equalization (using the \texttt{adapthisteq} function with an $8 \times 8$ tile grid, and a uniform distribution for the intensities with a clip limit of 0.01) is used to normalize the intensities of the nuclear signal.
 %
-For the second data set of images (taken during gastrulation), we use all three channels (dpERK, Twi, and nuclei) in the vector diffusion maps computations.
+The nuclear signal is blurred using a disc filter with a $5$~pixel radius (using the \texttt{imfilter} function).
 %
-For this particular data set, we multiply the intensity of the Twi signal by 1.5, as this channel was not completely saturated in the original imaging data.
+For the multichannel images, the nuclear signal is scaled by half (after normalization and blurring) relative to the other signals.
 
@@
 \end{equation}
-where the function $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ radians, and $\xi_i$ is a (typically Gaussian) noise term.
+where the function $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ degrees, and $\xi_i$ is a (typically Gaussian) noise term.
 %
@@
 
-Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$; we write our procedure for general $d$ because we will later consider three-dimensional rotations).
+Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$).
 %
@@
 %
-$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon^2$ to be $1/10^{th}$ of the median of the squared pairwise distances between data points for the first two data sets (and half of the median of the squared pairwise distances between data points for the third data set).
+$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon^2$ to be $1/10^{th}$ of the median of the squared pairwise distances.
 
@@
 %%
-The $j^{th}$ embedding coordinate will be given by $\psi_{k,1}$, where $jd +1 < k \le (j+1)d$,
-and we select $k$ such that the coordinate $\psi_{k, 1}$ has the largest variability, i.e., the $j^{th}$ coordinate is $\psi_{k,1}$, where $k$ is the solution to
+The $j^{th}$ embedding coordinate will be given by $\psi_{k,l}$, where $jd +1 < k \le (j+1)d$ and $1 \le l \le d$,
+and we select $k, l$ such that the coordinate $\psi_{k, l}$ has the largest variability, i.e., the $j^{th}$ coordinate is $\psi_{k,l}$, where $k, l$ is the solution to
 \begin{equation} \label{eq:first_embed_vdm}
-\max_{jd +1 < k \le (j+1)d} \sum_i \psi_{k,1} (i)^2. 
+\max_{
+\begin{matrix}
+jd +1 \le k \le (j+1)d \\
+1 \le l \le d
+\end{matrix}}
+ \sum_i \psi_{k,l} (i)^2. 
 \end{equation}
@@
 \item Compute the alignments for each pair of images, as described in \eqref{eq:opt_pairwise}.
-Retain both the optimal pairwise alignments (translations and rotations) as well as the minimum distances obtained when pairs are optimally aligned.
+Retain both the optimal pairwise rotations as well as the minimum distances obtained when pairs are optimally aligned.
 %
-\item Compute the corresponding pairwise three-dimensional rotations from the translations and rotations, using \eqref{eq:angle_relations}, \eqref{eq:three_rot}, and \eqref{eq:total_rot}.
+\item Compute the corresponding rotation matrices $R(\theta_{ij})$ from \eqref{eq:R_theta}.
 %
 \item Select $\epsilon$ for use in the diffusion maps kernel. 
-We often begin with using the median of the pairwise distances as $\epsilon$, and then adjust as needed.
+%
+Taking $\epsilon$ to be on the order of the median of the pairwise distances often yields good results. 
 \item Compute the matrix $S$ in \eqref{eq:vdm_S}. 
-$A_{ij}$ is computed from \eqref{eq:dmaps_W} and \eqref{eq:dmaps_A}, using the pairwise distances from step 1, and $H_{ij} = R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$, which are computed in step 2.
+$A_{ij}$ is computed from \eqref{eq:dmaps_W} and \eqref{eq:dmaps_A}, using the pairwise distances from step 1, and $H_{ij} = R(\theta_{ij})$ are computed in step 2.
 %
@@
 \item Multiply each estimated rotation $R_{i, est}$ by $R_{1, est}^T$ (so that $R_{1, est}$ will become the identity matrix).
-\item For each image $i$, compute the optimal translation and rotation to align the image by converting $R_{i, est}^T$ (note the transpose) to the corresponding translation and rotation, described in \eqref{eq:angle_invert}.
-\item Compute the first embedding coordinate $\psi_{k,1}(i)$ as described in \eqref{eq:vdm_coord} and \eqref{eq:first_embed_vdm}, where $4 \le k \le 6$.
+\item For each image $i$, compute the optimal rotation to align the image by converting $R_{i, est}^T$ (note the transpose) to the corresponding angle using \eqref{eq:R_theta}. 
+\item Compute the first embedding coordinate $\psi_{k,1}(i)$ as described in \eqref{eq:vdm_coord} and \eqref{eq:first_embed_vdm}, where $3 \le k \le 4$.
 \item To order the images, sort them by the values of this embedding coordinate $\psi_{k,1}(i)$.
@@
 
-\subsection{Registering images with respect to rotations and translations} \label{subsec:trans_rot_register}
+\subsection{Registering images} \label{subsec:trans_rot_register}
 
-Using angular synchronization or vector diffusion maps to register images with respect to rotations {\it and translations} requires some additional effort.
+To register sets of images, 
+the first step is to compute the optimal alignments between pairs of images.
 %
-%We are not only interested in registering the one-dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two-dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\it and} translations.
-%
-%We first compute the translations and rotations to optimally align each pair of data points.
-%
-The first step (similar to the case of only rotations) is to compute the optimal alignments between pairs of images.
-%
-Practically, we have square images discretized as pixels (rather than continuous functions on the plane); note that the center of the
-image square is {\em not} necessarily also the center of the embryo.
+Practically, we have square images discretized as pixels (rather than continuous functions on the plane).
 %
@@
 \begin{equation}\label{eq:opt_pairwise}
-(\theta_{ij}, dx_{ij}, dy_{ij}) = \argmin_{
-\begin{matrix}
-\theta \in [0, 2\pi) \\
-dx, dy \in [-\Delta, \Delta]\\
-%dy \in [-\Delta, \Delta]
-\end{matrix}
-} \|g(I_j, \theta, dx, dy) - I_i \|^2.
+\theta_{ij} = \argmin_{
+0^\circ \le \theta < 360^\circ }
+ \|g(I_j, \theta) - I_i \|^2.
 \end{equation}
-where $g(I_j, \theta, dx, dy)$ is image $I_j$ rotated around the center of the square by $\theta$ radians, then translated by $dx$ pixel widths in the $x$ direction, and finally translated by $dy$ pixel widths in the $y$ direction.
+where $g(I_j, \theta)$ is image $I_j$ rotated around the center of the square by $\theta$ degrees.
 %
@@
 %
-The domain of the image (a square) is not invariant to our rotations and translations; however, the pixels near the border of the square have zero intensity, and so the norm can be meaningfully computed as long as the main image does not ``move out of'' the original square.
+The domain of the image (a square) is not invariant to our rotations; however, the pixels near the corners of the square have zero intensity, and so the norm can be meaningfully computed as long as the main image does not ``move out of'' the original square.
 %
-%The norm \eqref{eq:opt_pairwise} involve an integral over the square, which is not left invariant by our rotations and
-%translations; noting however that the pixels close to the border of the square have zero intensity allows a meaningful
-%computation of this norm as long as the ``main feature" (the embryo) does not move out of the original square (no ``corner effects").
-%
-This also justifies the $\Delta$ limits in \eqref{eq:opt_pairwise}: $\Delta$ is chosen such that the main image is not split or separated when translating (we take $\Delta=10$, which corresponds to a 10\% shift in the image).
-%; it does not make sense to shift the images so much that the embryo would ``move out" of the original square.
-%
-Image rotation is performed with the \texttt{imrotate} function in Matlab, using nearest neighbor interpolation to estimate the pixel intensities after rotation.
-%, since there is no direct mapping of the pixels of the unrotated image to the pixels of the rotated image.
+Image rotation is performed with the \texttt{imwarp} function in Matlab, using linear interpolation to estimate the pixel intensities after rotation.
 %
@@
 %
-The vertical and horizontal translations are implemented by shifting the pixels;
-we only consider shifts that correspond to an integer number of pixels to remove the need for interpolation when doing the translations.
-%
 The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
 %
-Therefore, instead of using an optimization procedure, we discretize the search space and exhaustively search to find the solution.
-%
-Our discrete translation steps are $\Delta/5$, and our discrete rotations steps are $\pi/10$.
+Therefore, instead of using an optimization procedure, we discretize the search space and exhaustively search to find the solution (for the results presened, we use $10^\circ$ discretization steps).
 %
@@
 %
-%Although this can be somewhat time intensive, it is not prohibitive for the data sets we consider, and can be trivially parallelized if necessary.
+Once we have computed $\theta_{ij}$ for all image pairs, we can proceed with the vector diffusion maps procedure.
 %
-%The solution will not be the exact solution to \eqref{eq:opt_pairwise}, but it will (most likely) be a close approximation.
-%
-%Since our techniques are robust to noise, close approximations will be sufficient to obtain accurate results.
-
-%As discussed above, for angular synchronization, the underlying symmetry group needs to have a real and orthogonal matrix representation.
-%
-%We therefore convert $ISO(2)$, the group of two-dimensional translations and rotations, to $SO(3)$, the group of three-dimensional rotations.
-%
-
-The synchronization formulation requires the matrices used to represent the symmetry group to be orthogonal \cite{singer2013spectral}.
-%(for \eqref{eq:H_low_rank} to be satisfied).
-%
-However, the typical matrix representation for $ISO(2)$, the group of two-dimensional translations and rotations, does not satisfy this property.
-%
-Instead, we {\it approximately} represent rotations and translations in two dimensions using {\it three-dimensional rotation matrices}, by projecting the (two-dimensional) image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular}.
-%
-Rotation of the image corresponds to rotation around one principal axis in three dimensions, while translation of the image corresponds (approximately) to rotations around the other two principal axes (see \fig~\ref{fig:SO3_picture}).
-(Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$.
-However, the images we will be considering are already ``mostly centered", and so we are only interested in {\em small} translations that are well within the $[0, 2\pi)$ range.)
-%
-These rotation matrices are (by definition) orthogonal, and successive application of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
-
-To convert the optimal two-dimensional translations and rotation to corresponding (continuous) three-dimensional rotations,
-we first compute the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$,
-\begin{equation} \label{eq:angle_relations}
-%\begin{aligned}
-%	\alpha_{ij} &= \theta_{ij} \\
-%	\beta_{ij} &= \frac{dx_{ij}}{n} \times \eta_{proj} \\
-%	\gamma_{ij} &= \frac{dy_{ij}}{n} \times \eta_{proj} \\
-%\end{aligned}
-	\alpha_{ij} = \theta_{ij}, \; \; \;
-	\beta_{ij} = \frac{dx_{ij}}{n} \times \eta_{proj}, \; \; \;
-	\gamma_{ij} = \frac{dy_{ij}}{n} \times \eta_{proj}
-\end{equation}
-where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project the image.
-%
-We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere;
-this portion of the sphere is small enough so that curvature effects are not prominent in the image.
-%
-We write rotations around the three principal axes, $R^x(\alpha)$, $R^y(\beta)$, and $R^z(\gamma)$, in terms of the three Euler angles
-%\small
-\begin{equation} \label{eq:three_rot}
-\begin{aligned}
-	R^x(\alpha) &= \begin{bmatrix}
-	1 & 0 & 0 \\
-    0 & \cos(\alpha) & -\sin(\alpha) \\
-    0 & \sin(\alpha) & \cos(\alpha)
-	\end{bmatrix} \\
-	R^y(\beta) &= \begin{bmatrix}
-	\cos(\beta) & 0 & \sin(\beta) \\
-    0 & 1 & 0 \\
-    -\sin(\beta) & 0 & \cos(\beta)
-    \end{bmatrix} \\
-	R^z(\gamma) &= \begin{bmatrix}
-	\cos(\gamma) & -\sin(\gamma) & 0 \\
-    \sin(\gamma) & \cos(\gamma) & 0 \\
-    0 & 0 & 1
-    \end{bmatrix}.
-\end{aligned}
-\end{equation}
-%\normalsize
-%
-The total rotation, $R_{ij} \in SO(3) \subset \mathbb{R}^{3 \times 3}$, is
-\begin{equation} \label{eq:total_rot}
-	R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})	 = R^z(\gamma_{ij})  R^y(\beta_{ij})  R^x(\alpha_{ij}).
-%R(\alpha_{ij}, \beta_{ij}, \gamma_{ij}) =
-%\begin{bmatrix}
-%\cos(\beta) \cos(\gamma) & \cos(\gamma)\sin(\alpha)\sin(\beta)-\cos(\alpha) \sin(\gamma) & \cos(\alpha)\cos(\gamma)\sin(\beta)+ \sin(\alpha)\sin(\gamma) \\
-%\cos(\beta) \sin(\gamma) & \cos(\alpha) \cos(\gamma)+\sin(\alpha) \sin(\beta) \sin(\gamma) & \cos(\alpha) \sin(\beta) \sin(\gamma)-\cos(\gamma) \sin(\alpha) \\
-%-\sin(\beta) & \cos(\beta) \sin(\alpha) & \cos(\alpha) \cos(\beta))
-%\end{bmatrix}
-\end{equation}
-%
-%\eqref{eq:total_rot} corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).
-
-These rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$ are used in formulating the matrix $H$ in \eqref{eq:H_to_R} for angular synchronization, or the matrix $S$ in \eqref{eq:vdm_S} for vector diffusion maps to compute the optimal rotations (and, in the case of vector diffusion maps, deduce optimal temporal orderings) for the images.
-%
-The three-dimensional optimal rotations $R_{1, est}, \dots, R_{m, est}$ then give us the corresponding translations and rotations of the images.
-%
-We multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
-%
-$R_{i,est}^T$ then gives the rotation matrix which registers image $i$.
-%
-From a rotation matrix $R$, we compute the corresponding Euler angles $\alpha$, $\beta$, and $\gamma$ using the following relationships
-\begin{equation} \label{eq:angle_invert}
-\begin{aligned}
-R_{1,1} & = \cos(\beta)\cos(\gamma) &
-R_{2,1} & = \cos(\beta)\sin(\gamma) \\
-R_{3,1} & = -\sin(\beta) &
-R_{3,2} & = \sin(\alpha)\cos(\beta) \\
-R_{3,3} & = \cos(\alpha)\cos(\beta)  &
-\end{aligned}
-\end{equation}
-%
-Inverting \eqref{eq:angle_relations} yields the desired approximate translations and rotation, and we round optimal translations to the nearest integer to remove the need for interpolation when translating the images.
+The rotation matrices returned from vector diffusion maps can be used to calculate the angle of rotation for each image (by inverting \eqref{eq:R_theta}), and the function $g$ as described previously is then used to rotate the images.
 
@@
 %
-For diffusion maps, the largest eigenvalue will always be 1 and correspond to the trivial (constant) eigenvector.
+In diffusion maps, the largest eigenvalue will always be 1 and correspond to the trivial (constant) eigenvector, and $|\lambda_k|$ gives a measure of the importance of coordinate $\phi_k$. 
 %
-The eigenvalues corresponding to noise in the data form a semicircle distribution centered at the origin (when the noise is a symmetric random matrix whose entries are drawn from a Gaussian distribution) \cite{wigner1955characteristic, wigner1958distribution}.
+Therefore, we expect to see a ``spectral gap'' in the eigenvalues which separates the meaningful coordinates from those corresponding to noise.
 %
-We expect the eigenvalues corresponding to descriptive/meaningful embedding coordinates to be positive and ``visibly removed" from the semicircle.
+However, some embedding coordinates which appear meaningful according to the eigenvalues may be harmonics of previous coordinates \cite{ferguson2010systematic}.
 %
-Counting such eigenvalues which are removed from the semicircle (excluding the trivial $\lambda=1$ eigenvalue) can help obtain an estimate of the dimensionality of the data (modulo harmonics in the same dimension \cite{ferguson2010systematic}).
+Therefore, one must visually check for correlations among potential embedding coordinates before deducing the true dimensionality of the data.
 
-For vector diffusion maps, the same structure holds: we expect trivial eigenvalues, a semicircle corresponding to noise, and positive eigenvalues removed from the noise that correspond to meaningful embedding coordinates.
+In vector diffusion maps, the importance of each coordinate is measured by the product of the corresponding eigenvalues (i.e., the importance of $\psi_{k,l}$ is given by $| \lambda_k \lambda_l |$). 
 %
-However, in vector diffusion maps, each coordinate gives rise to $d$ eigenvalues with (approximately) the same value (as each block eigenvector corresponds to $d$ eigenvalues).
-%
-We must then establish how many sets of $d$ eigenvalues are removed from the noise semicircle (excluding the trivial set of $d$ eigenvalues, all with  $\lambda=1$), to explore the dimensionality of the data.
+We again expect to see a ``spectral gap'' in these eigenvalue products between those corresponding to meaningful coordinates (modulo higher harmonics) and those corresponding to noise.
 
@@
 %
-Let $x_1, \dots, x_m$ denote the set of {\em registered and ordered} images (so $x_i$ is the $i^{th}$ image in the ordered set).
+Let $I_1, \dots, I_m$ denote the set of {\em registered and ordered} images (so $I_j$ is the $j^{th}$ image in the ordered set).
 %
-We define the average image at time $\tau$, denoted $\overline{x}_{\tau}$ (where $1 \le \tau \le m$), as a (Gaussian) weighted average of the images,
+We define the average image at time $\tau$, denoted $\overline{I}_{\tau}$ (where $1 \le \tau \le m$), as a (Gaussian) weighted average of the images,
 \begin{equation}
-\overline{x}_{\tau} =  \frac{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma^2} \right) x_i}{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma^2} \right) }
+\overline{I}_{\tau} =  \frac{\sum_j \exp \left( - \frac{|j - \tau|^2}{\sigma^2} \right) I_i}{\sum_j \exp \left( - \frac{|j - \tau|^2}{\sigma^2} \right) }
 \end{equation}
@@
 %
-For the images in \fig~4, we take $\sigma^2 = 20$.
+For the images in \fig~4, we take $\sigma = 3$.
 %
@@
 
-\begin{thebibliography}{10}
-
-\bibitem{singer2011angular}
-Singer A (2011) Angular synchronization by eigenvectors and semidefinite
-  programming. \textit{Appl Comput Harmon Anal}
-  30(1):20--36.
-
-\bibitem{ahuja2007template}
-Ahuja S, Kevrekidis I~G, Rowley C~W (2007) Template-based stabilization of
-  relative equilibria in systems with continuous symmetry. \textit{Journal of
-  Nonlinear Science} 17(2):109--143.
-
-\bibitem{goemans1995improved}
-Goemans M~X, Williamson D~P (1995) Improved approximation algorithms for maximum cut and satisfiability problems using semidefinite programming.
-\textit{Journal of the ACM} 42(6):1115--1145.
-
-\bibitem{bandeira2013cheeger}
-Bandeira A~S, Singer A, Spielman D~A (2013) A Cheeger inequality for the graph connection Laplacian. \textit{SIAM Journal on Matrix Analysis and Applications} 34(4):1611--1630.
-
-\bibitem{coifman2005geometric}
-Coifman R~R, Lafon S, Lee A~B, Maggioni M, Nadler B, \textit{et~al.} (2005)
-  Geometric diffusions as a tool for harmonic analysis and structure definition
-  of data: Diffusion maps. \textit{Proc Natl Acad Sci USA} 102(21):7426--7431.
-
-\bibitem{coifman2008graph}
-Coifman R~R, Shkolnisky Y, Sigworth F~J, Singer A (2008) Graph {L}aplacian
-  tomography from unknown random projections. \textit{IEEE Trans Image Process} 17(10):1891--1899.
-
-\bibitem{rohrdanz2011determination}
-Rohrdanz M~A, Zheng W, Maggioni M, Clementi C (2011) Determination of reaction
-  coordinates via locally scaled diffusion map. \textit{J Chem Phys} 134:124116.
-
-\bibitem{Belkin2003}
-Belkin M, Niyogi P (2003) Laplacian eigenmaps for dimensionality reduction and
-  data representation. \textit{Neural Comput} 15:1373--1396.
-
-\bibitem{singer2012vector}
-Singer A, Wu H~T (2012) Vector diffusion maps and the connection laplacian.
-  \textit{Commun Pure Appl Math} 65(8):1067--1144.
-
-\bibitem{singer2013spectral}
-Singer A, Wu H~t (2013) Spectral convergence of the connection {L}aplacian from
-  random samples. \textit{arXiv preprint arXiv:13061587} .
-
-\bibitem{wigner1955characteristic}
-Wigner E~P (1955) Characteristic vectors of bordered matrices with infinite
-  dimensions. \textit{The Annals of Mathematics} 62(3):548--564.
-
-\bibitem{wigner1958distribution}
-Wigner E~P (1958) On the distribution of the roots of certain symmetric
-  matrices. \textit{The Annals of Mathematics} 67(2):325--327.
-
-\bibitem{ferguson2010systematic}
-Ferguson A~L, Panagiotopoulos A~Z, Debenedetti P~G, Kevrekidis I~G (2010)
-  Systematic determination of order parameters for chain dynamics using
-  diffusion maps. \textit{Proc Natl Acad Sci USA}
-  107(31):13597--13602.
-
-\bibitem{kemelmacher2011exploring}
-Kemelmacher-Shlizerman I, Shechtman E, Garg R, Seit, S~M (2011)
-Exploring photobios. \textit{ACM Transactions on Graphics (TOG)} 30(4):61.
-
-\bibitem{lim2013kinetics}
-Lim B, Samper N, Lu H, Rushlow C, Jim{\'e}nez G, \textit{et~al.} (2013)
-  Kinetics of gene derepression by erk signaling. \textit{Proc Natl Acad Sci USA} 110(25):10330--10335.
-
-\end{thebibliography}
+\bibliographystyle{pnas2}
+\bibliography{../image_analysis_paper/background_reading/references,../../references/references}
 
@@
 The dorsal (D) side is up and the ventral (V) side is down.
-For the first and third data sets (cellularization), images were collected at the focal plane $\sim 18\%$ from either the anterior or posterior pole of an embryo (arrows in the left image).
-During cellularization, it is known that the developmental dynamics are identical at these two regions \cite{lim2013kinetics}.
-In the second data set (gastrulation), images were collected at the focal plane $\sim 18\%$ from the posterior pole of an embryo (right arrow in the left image).
+Images were collected at the focal plane $\sim 18\%$ from the posterior pole of an embryo (right arrow in the left image).
 }
@@
 \includegraphics{figS2}
-\caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different times during development.
-The membrane length is indicated by the red bars.
-{\it (B)} Membrane length as a function of time.
-The points corresponding to the images in {\it A} are indicated with arrows.}
-\label{fig:membrane_compare}
-\end{figure}
-
-\begin{figure}
-\includegraphics{figS3}
-\caption{Images of mutant and wild type {\it Drosophila} embryos during cellularization, used for the analysis in \fig~5.
-Each image is of a different embryo in a different rotational orientation.
-Half of the images are of wild type embryos, and half are of mutant embryos.}
-\end{figure}
-
-\begin{figure}
-\includegraphics{figS4}
 \caption{Synthetic data set used to illustrate the data processing algorithms.
@@
 \begin{figure*}
+\includegraphics{figS3}
+\caption{Eigenvalue spectra.
+{\it (A)} Eigenvalue spectrum for the live imaging data set presented in \fig~3.
+Note that there is a gap after the third coordinate.
+{\it (B), (C)} Second and third embedding coordinates, respectively, versus the first embedding coordinate for the live imaging data set presented in \fig~3.
+Note that coordinates 2 and 3 are harmonics/functions of coordinate 1, and are therefore not informative about structure in the data set.
+We can conclude that the data set is effectively one-dimensional.
+{\it (D)} Eigenvalue spectrum for the live imaging data set presented in \fig~4.
+Note that there is a gap after the first coordinate; we can conclude that the data set is effectively one-dimensional.
+}
+\end{figure*}
+
+\begin{figure*}
+\includegraphics{figS4}
+\caption{Live imaging data set.
+{\it (A)} Live imaging data set used for the results in \fig~3.
+The images have been preprocessed as described above, each image has been arbitrarily rotated, and the order of the images has been shuffled.
+{\it (B)} Data set from A after registration and ordering.
+The real time for each image is indicated.
+}
+\end{figure*}
+
+\begin{figure*}
 \includegraphics{figS5}
-\caption{Illustration of the correspondence between three-dimensional rotations to translations and rotations in two dimensions.
-{\it (A)} The original image (shown in red), projected onto a portion of a sphere.
-{\it (B)} Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image.
-{\it (C)} Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation.
-{\it (D)} Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.}
-\label{fig:SO3_picture}
+\caption{{\it (A-D)} Four additional live imaging data sets used for the analysis in \fig~3. }
 \end{figure*}
@@
 \includegraphics{figS6}
-\caption{The histograms of the eigenvalues from vector diffusion maps analysis for the three data sets.
-%
-We expect to see an approximately symmetric semicircle distribution centered at 0, which corresponds to noise, and positive eigenvalues separated from the semicircle which correspond to meaningful coordinates.
-%
-Each coordinate yields $d=3$ eigenvalues of (approximately) the same magnitude, and the top $d=3$ eigenvalues ($\lambda=1$) correspond to a trivial embedding coordinate.
-%
-(A) Histogram of eigenvalues from images during cellularization (\fig~3).
-As expected, we see an approximately symmetric distribution that ranges from $-0.8$ to $0.8$ (indicated by the blue arrow), and $6$ eigenvalues that are removed from this distribution (indicated by the red arrow).
-$3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $3$ eigenvalues correspond to the first nontrivial VDM component which ``uncovers" the developmental dynamics; then there is an appreciable gap before the third group of eigenvalues.
-%
-(B) Histogram of eigenvalues from images during gastrulation (\fig~4).
-We see an approximately symmetric distribution that ranges from $-0.7$ to $0.7$ (indicated by the blue arrow), and $6$ eigenvalues that are removed from this distribution (indicated by the red arrow).
-$3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $3$ eigenvalues correspond to the first nontrivial VDM component which ``uncovers" the developmental dynamics.
-%
-(C) Histogram of eigenvalues from images of mutant and wild type embryos during cellularization (\fig~5).
-We see an approximately symmetric distribution that ranges from $-0.2$ to $0.2$ (indicated by the blue arrow), and $9$ eigenvalues that are removed from this distribution (indicated by the red arrows).
-$3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $6$ eigenvalues correspond to the first two VDM components which are now necessary to parametrize the developmental dynamics of the two genotypes.}
-\label{fig:eval_spectra}
-\end{figure*}
-
-\begin{figure}
-\includegraphics{figS7}
-\caption{Images of embryos during cellularization (from \fig~3), registered using angular synchronization \cite{singer2011angular} and ordered by the first PCA projection coefficient.
-The recovered dynamics are qualitatively consistent with those recovered by vector diffusion maps.
+\caption{Fixed embryo imaging data set.
+{\it (A)} Fixed images used for the results in \fig~4.
+The images have been preprocessed as described above.
+{\it (B)} Data set from A after registration and ordering.
+The expert rank for each image is indicated.
 }
-\label{fig:PCA_data1}
-\end{figure}
-
-\begin{figure*}
-\includegraphics{figS8}
-
-\caption{{\it (A)} Images shown in \fig~4 of the paper, registered using angular synchronization \cite{singer2011angular} and ordered by the first PCA projection coefficient.
-The recovered ordering is not consistent with the known developmental dynamics.
-{\it (B)} The fraction of the variance captured by each of the first 10 principal components.
-Note that components 2--10 have a nonnegligible contribution to the data.
-{\it (C)} Projection of the data onto the first two principal components.
-Each principal component is normalized to 1.
-Both the projection coefficients and the projection coefficients scaled by the square roots of the corresponding eigenvalues are shown.
-}
-\label{fig:PCA_results}
 \end{figure*}

### 2014-08-29 image_analysis_paper2/pnas_draft.tex
added paper edits

@@
 \customlabel{fig:DV_view}{S1}
-\customlabel{fig:membrane_curve}{S2}
-\customlabel{fig:raw_data3}{S3}
-\customlabel{fig:1d_example}{S4}
-\customlabel{fig:spheres}{S5}
-\customlabel{fig:eigenvalues}{S6}
-\customlabel{fig:PCA_data1}{S7}
-\customlabel{fig:PCA_data2}{S8}
+\customlabel{fig:1d_example}{S2}
+\customlabel{fig:eigenvalues}{S3}
+\customlabel{fig:full_movie_data}{S4}
+\customlabel{fig:all_movies}{S5}
+\customlabel{fig:full_fixed_data}{S6}
 
@@
 %
-To illustrate this approach, we temporally order and register several data sets collected in  studies of cell signaling in the early fruit fly embryo.}
+To illustrate this approach, we temporally order and register data sets collected in  studies of cell signaling in the early fruit fly embryo.}
 
@@
 
-Imaging studies provide unique insights into the dynamics of developmental processes. 
+Imaging studies provide unique insights into the dynamics of pattern formation and morphogenesis in developing tissues.
 %
-Often, data is collected across many samples in several experiments, and the first task is to organize the data for further analysis.
+In one of the most commonly used experimental formats, developmental trajectories are reconstructed from snapshots of chemical or mechanical processes in fixed embryos.
 %
-This includes spatially registering images of different biological samples, and well as
-temporally ordering the data to reconstruct representative developmental dynamics.
+As a first step in these reconstructions, such snapshots must be spatially registered and ordered in time.
 %
-When such data sets are large, noisy, and/or if the developmental changes are subtle, these tasks can be difficult to perform manually.
+Currently, registration and ordering of images is often done manually, requiring a significant amount of expertise with a specific developmental system.
 %
-We present an automated approach to simultaneously register and temporally order imaging data sets.
+However, as the size of imaging datasets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
+%
+To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
 %
@@
 %
-We validate this approach using a live imaging data set where both the correct rotational orientation and temporal order are known, as well as use these methods to extract a developmental trajectory from a data set of fixed images collected to study dorsoventral patterning of the {\it Drosophila} embryo.
+We validate our approach using a live imaging data set where the correct rotational orientation and temporal order are known.
+%
+Finally, we demonstrate that our approach can successfully extract a representative developmental trajectory from a collection of images of fixed {\it Drosophila} embryos.
 
@@
 
-\dropcap{I}n one of the common approaches to studies of development, a group of embryos is fixed and stained to visualize a particular biochemical or morphological process within a developing tissue. 
+\dropcap{I}n one of the common approaches to studies in development, a group of embryos is fixed and stained to visualize a particular biochemical or morphological process within a developing tissue. 
 %
-The developmental dynamics must then be reconstructed from multiple embryos, each of which contributes only a snapshot of a chemical or morphological process along its developmental trajectory \cite{jaeger2004dynamic, peter2011gene, fowlkes2008quantitative}.
+The developmental dynamics must then be reconstructed from multiple embryos, each of which contributes only a snapshot of the relevant process along its developmental trajectory \cite{jaeger2004dynamic, peter2011gene, fowlkes2008quantitative}.
 %
@@
 a certain time window to which a collection of embryos belongs \cite{ng2012large, richardson2014emage, castro2009automatic}.
+%
+Furthermore, images are often collected in different spatial orientations or from different viewing directions.
 %
@@
 %
-In contrast to a number of previous methodologies, it does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression for ordering images\cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}.
+In contrast to a number of previous methodologies \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, it does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression for ordering images.
 %
@@
 %
-The entire data set contains $40$ consecutive frames taken at $30$~second time intervals at a fixed position within a single embryo.
+The entire data set contains $40$ consecutive frames taken at $30$~second time intervals at a fixed position within a single embryo (see \fig~\ref{fig:full_movie_data}).
 %
-Each image shows an optical cross-section of a vertically oriented developing embryo, with the nuclei labeled by Histone-GFP.
+Each image shows an optical cross-section of a vertically oriented developing embryo, with the nuclei labeled by Histone-RFP.
 %
@@
 
-To assess the robustness of our proposed methodology, we repeated this procedure with four additional datasets extracted from independent live imaging studies spanning the same developmental time period.
+To assess the robustness of our proposed methodology, we repeated this procedure with four additional datasets (see \fig~\ref{fig:all_movies}) extracted from independent live imaging studies spanning the same developmental time period.
 %
@@
 %
-Because our particular data set contains relatively low noise levels, few images are required to obtain accurate rotational alignments and the errors in the recovered angles were consistently less than $10^{\circ}$.
+Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignments and the errors in the recovered angles were consistently less than $10^{\circ}$.
 %
-Therefore, the algorithm accurately registers and orders a model data set where both the rotation angles and real times are independently known. 
-%
-In addition, these results are consistent across multiple data sets, and we showed how the accuracy of our methods varies as a function of the number of images.
+Therefore, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.
 
@@
 %
-The median rank correlation coefficient for ordering such a synthetic time course using our methodology was 0.81, indicating that the algorithm can accurately recover the temporal order even under noisy conditions.
+The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.81, indicating that the algorithm can accurately recover the temporal order even under noisy conditions.
 
@@
 %
-To demonstrate such an application, \fig~\ref{subfig:fixed_images_unregistered_unordered} shows selected images from a set of $120$ images (see \SI for the full data set) which cover a thirty minute time interval spanning late cellularization through gastrulation.
+To demonstrate such an application, \fig~\ref{subfig:fixed_images_unregistered_unordered} shows selected images from a set of $120$ images (see \fig~\ref{fig:full_fixed_data}) which cover a thirty minute time interval spanning late cellularization through gastrulation.
 %
@@
 
-To evaluate the quality of our registration and ordering, we can use prior knowledge about the developmental dynamics of this specific system. 
+We can use prior knowledge about the developmental dynamics of this specific system to evaluate the quality of our registration and ordering. 
 %
@@
 %
-We found that the standard deviation in the location of the peak of the Twi signal was 8$^\circ$ after registration, indicating that our registration is consistent with the known biology. 
+We found that the standard deviation in the location of this peak in the set of registered images was 8$^\circ$,
+indicating that the algorithm successfully aligns the ventralmost points of the images. 
 %
@@
 %
-The ranks for the images in \fig~\ref{subfig:fixed_images_registered_ordered} as determined by this expert ordering are indicated, and the rank correlation (see \fig~\ref{subfig:fixed_images_rank_corr}) shows that our ordering is consistent with the expert ordering.
-%
+The ranks from this expert ordering for the images in \fig~\ref{subfig:fixed_images_registered_ordered} are indicated, and the rank correlation (see \fig~\ref{subfig:fixed_images_rank_corr}) shows that our ordering is consistent with the expert ordering.
+
 However, manually ordering the images is nontrivial for researchers who are not familiar with the developmental progression, and can be tedious and time-consuming for those who are.
@@
 %
-The parameter tuning required for our method is minimal.
+The requisite user intervention and parameter tuning required for our method is relatively minimal.
 %
@@
 %
-We found that the results are robust to both of these parameters, and that consistent parameter values were appropriate for all of the data sets presented in this paper. 
+We found that the results are robust to both of these parameters, and that consistent parameter values were appropriate for all data sets presented here. 
 %
@@
 %
-In general, we can recover several coordinates that are good descriptors of our data set.
+In general, we can recover several coordinates that are good descriptors of the data set.
 %
@@
 
-\section{Fly strain and whole-mount immunostaining}
+\section{Experiments}
 %
-Oregon-R (OreR) and spaghetti squash (Sqh)-GFP were used as wild type strains. 
-%
-{\it Sna\textsuperscript{\it IIG05}/Cyo hb-lacZ} flies were used to study mutant embryos with one peak of dpERK activation.  
+Oregon-R was used as wild type strains. 
 %
@@
 %
-Monoclonal rabbit anti-dpERK (1:100, Cell signaling), mouse anti-Dorsal (1:100, Developmental Studies Hybridoma Bank), rat anti-Twist (1:500, a gift from Eric Wieschaus), and mouse anti-beta-galactosidase (1:500, DSHB) were used to stain proteins of interest.
+Monoclonal rabbit anti-dpERK (1:100, Cell signaling) and rat anti-Twist (1:500, a gift from Eric Wieschaus) were used to stain proteins of interest. 
+%
 DAPI (1:10,000, Vector Laboratories) was used to visualize nuclei, and Alexa Fluors (1:500, Invitrogen) were used as secondary antibodies. 
+%
+Histone-RFP strain was used to obtain time-lapse movie of gastrulating embryos at 22$^\circ$C. 
+%
+Embryos were loaded to the microfluidic device with PBST to keep them oxidized.
 
@@
 %
-Nikon A1-RS scanning confocal microscope, and the Nikon 60X Plan-Apo oil objective was used to image embryos.
+Nikon A1-RS scanning confocal microscope, and the Nikon 60x Plan-Apo oil objective was used to image embryos. 
+%
 Embryos were collected, stained, and imaged together under the same microscope setting. 
-End-on imaging was performed by using the microfluidics device described previously.
+%
+End-on imaging was performed by using the microfluidics device described previously \cite{chung2010microfluidic}.
+%
 Images were collected at the focal plane $\sim$90~$\mathrm{\mu m}$ from the posterior pole of an embryo (see \fig~\ref{fig:DV_view}).
-
-\section{Image preprocessing}
-%
-Borders were cropped from each image using a Canny edge-detection filter to remove any size effects.
-%
-Adaptive histogram equalization was used to normalize the nuclear signal intensity throughout each image, and a 5-pixel radius disk filter was used to smooth the nuclear signal. 
-%
-%For the three-channel fixed images, the nuclear signal was scaled by half relative to the Twi and dpERK signals.
 
@@
 %
-The vector diffusion maps algorithms was implemented in MATLAB\textsuperscript{\textregistered} (R2013b, The MathWorks, Natick, Massachusetts).
+All algorithms and analysis were implemented in MATLAB\textsuperscript{\textregistered} (R2013b, The MathWorks, Natick, Massachusetts).
 %
-All code, including all the scripts used to generate the figures in this paper, is available at .....
+Code is available at .....
 

### 2014-08-30 image_analysis_paper2/SI_text.tex
small edits to paper

@@
 
-Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$).
+Let $d$ denote the dimension of the rotation matrices we are considering (for planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$).
 %
@@
 %
-\item Compute the eigenvalues  $\lambda_1, \lambda_2, \dots, \lambda_{3m}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{3m}$ of $S$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{3m}|$.
+\item Compute the eigenvalues  $\lambda_1, \lambda_2, \dots, \lambda_{2m}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{2m}$ of $S$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{2m}|$.
 %
-\item Estimate the optimal three-dimensional rotation for each image: stack the first three eigenvectors $\phi_1, \phi_2, \phi_3$ into a $3m \times 3$ matrix, and then divide this matrix into $m$ $3 \times 3$ blocks, $\hat{R}_1, \dots, \hat{R}_m$.
+\item Estimate the optimal three-dimensional rotation for each image: stack the first two eigenvectors $\phi_1, \phi_2$ into a $2m \times 2$ matrix, and then divide this matrix into $m$ $2 \times 2$ blocks, $\hat{R}_1, \dots, \hat{R}_m$.
 Compute the estimated rotations for each of these $m$ blocks as in \eqref{eq:R_est}.
@@
 \item Compute the first embedding coordinate $\psi_{k,1}(i)$ as described in \eqref{eq:vdm_coord} and \eqref{eq:first_embed_vdm}, where $3 \le k \le 4$.
-\item To order the images, sort them by the values of this embedding coordinate $\psi_{k,1}(i)$.
+\item To order the images, sort them by the values of this embedding coordinate $\psi_{k,l}(i)$.
 \end{enumerate} 
@@
 
-We can use the distribution of the eigenvalues from (vector) diffusion maps to help deduce the dimensionality of the data.
+We can use the eigenvalues from (vector) diffusion maps to help deduce the dimensionality of the data.
 %
@@
 %
-Therefore, we expect to see a ``spectral gap'' in the eigenvalues which separates the meaningful coordinates from those corresponding to noise.
+We therefore expect to see a ``spectral gap'' in the eigenvalues which separates the meaningful coordinates from those corresponding to noise.
 %
-However, some embedding coordinates which appear meaningful according to the eigenvalues may be harmonics of previous coordinates \cite{ferguson2010systematic}.
-%
-Therefore, one must visually check for correlations among potential embedding coordinates before deducing the true dimensionality of the data.
+However, some embedding coordinates which appear meaningful according to the eigenvalues may be harmonics of previous coordinates \cite{ferguson2010systematic}, 
+and one must visually check for correlations among potential embedding coordinates before deducing the true dimensionality of the data.
 
@@
 \includegraphics{figS1}
-\caption{(Left) A lateral view of a {\it Drosophila} embryo stained with DAPI (gray), dpERK (red), and Dorsal (green).
+\caption{(Left) A lateral view of a {\it Drosophila} embryo stained with DAPI (gray), dpERK (red), and Twi (green).
 The embryo is presented so that the anterior (A) side is to the left and the posterior (P) side is to the right.
-The arrow indicates the position(s) where the cross-section of an embryo is imaged.
+The arrow indicates the position where the cross-section of an embryo is imaged.
 (Right) A dorsoventral view of the cross-section of the {\it Drosophila} embryo.
 The dorsal (D) side is up and the ventral (V) side is down.
-Images were collected at the focal plane $\sim 18\%$ from the posterior pole of an embryo (right arrow in the left image).
+Images were collected at the focal plane $\sim 18\%$ from the posterior pole of an embryo (arrow in the left image).
 }
@@
 {\it (B)} Data set from A after registration and ordering.
-The expert rank for each image is indicated.
+The rank provided by a trained embryologist for each image is indicated.
 }

### 2014-08-30 image_analysis_paper2/pnas_draft.tex
small edits to paper

@@
 \significancetext{
-Imaging studies are commonly used to study developmental dynamics in multiple organisms. 
+Spatial registration and temporal ordering of images of developing embryos are one of the first steps in the reconstruction of developmental dynamics. 
 %
-When many samples are involved in a single study,
-one of the first steps in data analysis involves registration and temporal ordering of images. 
+When the number of images is small and the differences between them are easily recognizable, image registration and ordering can be done manually; these tasks become challenging when the data sets are larger and more complex.
 %
-Here we show that both of these tasks can be automated and combined in a single step using recent advances in data mining. 
+Here we show this can be automated and combined in a single
+step using recent advances in data mining. 
 %
-To illustrate this approach, we temporally order and register data sets collected in  studies of cell signaling in the early fruit fly embryo.}
+To illustrate this approach, we temporally order and register data sets collected
+in studies of cell signaling in the early fruit fly embryo.}
 
@@
 %
-In one of the most commonly used experimental formats, developmental trajectories are reconstructed from snapshots of chemical or mechanical processes in fixed embryos.
+In one of the most commonly used experimental formats, developmental progress is reconstructed from snapshots of chemical or mechanical processes in fixed embryos.
 %
@@
 %
-However, as the size of imaging datasets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
+However, as the size of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
 %
@@
 
-\dropcap{I}n one of the common approaches to studies in development, a group of embryos is fixed and stained to visualize a particular biochemical or morphological process within a developing tissue. 
+\dropcap{I}n one of the common approaches to studies of developmental dynamics, a group of embryos is fixed and stained to visualize a particular biochemical or morphological process within a developing tissue.
 %
@@
 %
-Importantly, the ``age'' of any given embryo arrested in its development is often not quantitatively known; typically what is known is
+Importantly, the ``age'' of any given embryo arrested in its development is often only approximately known; typically what is known is
 a certain time window to which a collection of embryos belongs \cite{ng2012large, richardson2014emage, castro2009automatic}.
 %
-Furthermore, images are often collected in different spatial orientations or from different viewing directions.
+Furthermore, images are often collected in different spatial orientations.
 %
@@
 %
-In contrast to a number of previous methodologies \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, it does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression for ordering images.
+In contrast to a number of previous methodologies \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, our methodology does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression for temporal ordering.
 %
@@
 Vector diffusion maps \cite{singer2012vector} is a manifold learning
-technique developed for data sets which contain two types of sources of variability:
+technique developed for data sets which contain two sources of variability:
 geometric symmetries (such as rotations of the images) which one would like to factor out,
@@
 %
-We assume that the main direction of variability in these images is parameterized by the developmental time of each embryo.
+We assume that the main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
 %
-As a consequence, uncovering this direction will allow us to recover the developmental dynamics.
+As a consequence, uncovering this direction should reveal the underlying developmental dynamics.
 
@@
 
-To validate the proposed approach, we applied our algorithm to a dataset where the true temporal order and rotational orientation of the images are known {\em a priori}.
+To validate the proposed approach, we applied our algorithm to a data set where the true temporal order and rotational orientation of the images are known {\em a priori}.
 %
-This data set was obtained through live imaging of a single {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
+This data set was obtained through live imaging of a {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
 %
@@
 
-\fig~\ref{subfig:movie_unregistered_unordered} shows selected images from this live imaging study. 
-%
-The entire data set contains $40$ consecutive frames taken at $30$~second time intervals at a fixed position within a single embryo (see \fig~\ref{fig:full_movie_data}).
+\fig~\ref{subfig:movie_unregistered_unordered} shows selected images from this live imaging data set, which contains $40$ consecutive frames taken at $30$~second time intervals at a fixed position within a single embryo (see \fig~\ref{fig:full_movie_data}).
 %
@@
 
-To assess the robustness of our proposed methodology, we repeated this procedure with four additional datasets (see \fig~\ref{fig:all_movies}) extracted from independent live imaging studies spanning the same developmental time period.
+To assess the robustness of our proposed methodology, we repeated this procedure with four additional data sets (see \fig~\ref{fig:all_movies}) extracted from independent live imaging studies spanning the same developmental time period.
 %
@@
 
-In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{subfig:dmaps1}) is well-sampled;
-the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \cite{singer2011angular}.
+In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{subfig:dmaps1}) is well-sampled.
+%
+At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \cite{singer2011angular}.
 %
@@
 %
-\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images, where each point is the average over $200$ bootstrap samples.
+Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered. 
+%
+\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
 %
@@
 
-\subsection{Datasets with interembryo variability}
+\subsection{Data sets with interembryo variability}
 
-Arguably, the most common source of noise within developmental biology experiments is embryo-to-embryo variability. 
+We have analyzed how our algorithm performs on a model data set, where all images come from a single embryo. 
+%
+Often in practice, each images comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
 %
 To demonstrate that our methods are robust to such noise sources,
-we constructed a synthetic time course data set by selecting a frame from a random live imaging experiment for each time point.
+we constructed a synthetic time course data set by selecting an image from a random live imaging experiment at each time point.
+%
+The resulting data set is therefore spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability. 
 %
@@
 
-Because live imaging data represents only a small fraction of the data collected in developmental biology experiments, the utility of our proposed methodology is in the analysis of large data sets of fixed images, where the developmental times and spatial orientations are unknown.
+Finally, we applied our approach to a data set where the true rotational orientation and temporal order was not known {\it a priori}.
 %
-To demonstrate such an application, \fig~\ref{subfig:fixed_images_unregistered_unordered} shows selected images from a set of $120$ images (see \fig~\ref{fig:full_fixed_data}) which cover a thirty minute time interval spanning late cellularization through gastrulation.
+\fig~\ref{subfig:fixed_images_unregistered_unordered} shows selected images from a set of $120$ images (see \fig~\ref{fig:full_fixed_data}) which cover a thirty minute time interval spanning late cellularization through gastrulation.
 %
-This data set is more complex than the live imaging data sets in that it contains significantly more images, each with several channels.
+This data set is more complex than the live imaging data sets in that it contains significantly more images, each of which provide information about tissue morphology and the spatial distribution of two regulatory proteins.
 %
@@
 %
-During invagination, the two ventral dpERK peaks merge together, eventually forming, together with Twi, the ``omega'' shape.
+During mesoderm invagination, the two ventral dpERK peaks merge together, eventually forming, together with Twi, the ``omega'' shape.
 %
@@
 %
-At the end of this process, similar ``omega'' shaped patterns formed by Twi and dpERK appear on the dorsal side; these patterns are most readily seen in the last image of \fig~\ref{subfig:fixed_images_average_trajectory}.
+At the end of this process, similar ``omegas'' formed by Twi and dpERK appear on the dorsal side of the embryo; these patterns are most readily seen in the last image of \fig~\ref{subfig:fixed_images_average_trajectory}.
 %
@@
 
-We can use prior knowledge about the developmental dynamics of this specific system to evaluate the quality of our registration and ordering. 
+We can use prior knowledge about the developmental system to evaluate the quality of our registration and ordering. 
 %
@@
 %
-Because the developmental time of each embryo cannot be easily estimated, we have few options for evaluating the quality of our ordering. 
+Because the developmental time of each embryo cannot be easily estimated, we have few options for evaluating the quality of our temporal ordering. 
 %
-We compared the ordering obtained from vector diffusion maps to an expert ordering provided by a trained embryologist who is knowledgeable about the developmental progression and the important image features.
+We compared the ordering obtained from vector diffusion maps to the ordering provided by a trained embryologist who is knowledgeable about the developmental progression and the important image features.
 %
-The ranks from this expert ordering for the images in \fig~\ref{subfig:fixed_images_registered_ordered} are indicated, and the rank correlation (see \fig~\ref{subfig:fixed_images_rank_corr}) shows that our ordering is consistent with the expert ordering.
+The ranks from this expert-based ordering for the images in \fig~\ref{subfig:fixed_images_registered_ordered} are indicated, and the rank correlation (see \fig~\ref{subfig:fixed_images_rank_corr}) shows that our ordering is consistent with the expert ordering.
 
-However, manually ordering the images is nontrivial for researchers who are not familiar with the developmental progression, and can be tedious and time-consuming for those who are.
+Manually ordering the images is nontrivial for researchers who are unfamiliar with the developmental progression, and can be tedious and time-consuming for those who are.
 %
@@
 %
-The computational time is a function of the number of images in the dataset, the number of pixels in each point, and the number of angles used to compute the pairwise rotations (see \SI). 
+The computational time is a function of the number of images in the data set, the number of pixels in each point, and the number of angles used to compute the pairwise rotations (see \SI). 
 %
@@
 %
-Images must first be preprocessed so that the Euclidean distance between the pixels in informative, and two algorithmic parameters, the number of angles to compute the pairwise alignments and the diffusion maps kernel scale (see \SI), must be defined.
+Images must first be preprocessed so that the Euclidean distance between the pixels is informative.
+%
+Two algorithmic parameters, the number of angles to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{subfig:dmaps1} and \SI), must also be defined.
 %
@@
 %
-Therefore, this methodology is very promising for much larger imaging data sets which are impractical to evaluate manually.
+Therefore, this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.
 
@@
 %
-In parallel, ordering of large-scale cross-sectional data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different chemical species \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
+In parallel, temporal ordering of large-scale data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different chemical species.
 %
-Automated temporal ordering of such data sets was accomplished by first projecting the data onto a low-dimensional subspace spanned by the leading principal or independent components, and then solving a traveling salesman problem or constructing a minimum spanning tree on the projected data to order multiple snapshots.
+Automated temporal ordering of such data sets was accomplished by first projecting the data onto a low-dimensional subspace spanned by the leading principal or independent components, and then solving a traveling salesman problem or constructing a minimum spanning tree on the projected data to order multiple snapshots \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
 %
@@
 %
-Although we have presented results for a specific type of data (cross-sectional images taken along the dorsoventral axis of developing {\em Drosophila} embryos), because the methodology only relies on computing distances between images, we are confident that it can be readily applied and extended to many other imaging data sets.
+Although we have presented results for a specific type of data (cross-sectional images taken along the dorsoventral axis of developing {\em Drosophila} embryos), our methodology only relies on computing distances between images and may be readily extended to many other imaging data sets.
 
@@
 %
-Based on the presented results, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories.
+Based on the presented results, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories from snapshots of fixed embryos.
 %
@@
 
-As experimental imaging becomes faster and less expensive, the rate of collection of imaging data will surpass the rate of manual image analysis,
-and we will require automated methodologies which are sufficiently general to address a wide variety of applications to organize such large data sets.
+Already at this point, the rate of collection of imaging data has surpassed the rate of manual image analysis.
+%
+This necessitates the development of automated methodologies which are sufficiently general to address a wide variety of applications to organize such large data sets.
 %
@@
 %
-In the examples presented here, we have focused on recovering a single coordinate which is used to order the images in time.
+In the examples presented here, we have focused on ordering the images in time using the first vector diffusion maps coordinate.
 %
-In general, we can recover several coordinates that are good descriptors of the data set.
+In general, we can recover several coordinates that serve as a good parameterization of the data set.
 %
-This parameterization can then be used for typical data analysis tasks, such as outlier detection, clustering, and model fitting. 
+This parameterization can then be used for typical data analysis tasks, such as outlier detection and model fitting.
+%
+Data which comes from several different genetic backgrounds can be clustered according to their spatial expression patterns, and clustering can also be used to separate images into different developmental stages.
 %
@@
 Images were collected at the focal plane $\sim$90~$\mathrm{\mu m}$ from the posterior pole of an embryo (see \fig~\ref{fig:DV_view}).
+
+\section{Image preprocessing}
+%
+Image resolution was subsampled to $100 \times 100$ pixels for vector diffusion maps analysis.
+%
+Borders were cropped from the images using a Canny edge detector applied to the nuclear signal. 
+%
+Contrast-limited adaptive histogram equalization and a a 5-pixel radius disc filter were used to normalize and blur the nuclear signal.
 

### 2014-09-01 image_analysis_paper2/SI_text.tex
added yannis comments

@@
 %
-The Canny method is used to detect the edges of the nuclear signal in each image (using the \texttt{edge} function in MATLAB\textsuperscript{\textregistered}), and images are cropped to these edges to remove any size effects.
+The Canny method \cite{canny1986computational} is used to detect the edges of the nuclear signal in each image (using the \texttt{edge} function in MATLAB\textsuperscript{\textregistered}), and images are cropped to these edges to remove any size effects.
 %
-A constant $10$-pixel border is then added to each image. 
+A constant $10$-pixel-wide border is then added to each image. 
 %
@@
 %
-For the multichannel images, the nuclear signal is scaled by half (after normalization and blurring) relative to the other signals.
+For the multichannel images, the nuclear signal is scaled by half (after normalization and blurring) relative to the other signals, so that the total nuclear signal across the entire image is of a comparable magnitude to the other signals.
 
@@
 
-We will demonstrate the algorithms for registration and temporal ordering using a synthetic data set.
+We demonstrate the algorithms for registration and temporal ordering using a synthetic data set.
 %
-The relatively simple dynamics of this data set will allow us to easily visualize and illustrate the main features of the different algorithms.
+The relatively simple dynamics of this data set allows us to easily visualize and illustrate the main features of the different algorithms.
 %
@@
 {\it (A)} Eigenvalue spectrum for the live imaging data set presented in \fig~3.
-Note that there is a gap after the third coordinate.
+Note that there is a gap after the third eigenvalue.
 {\it (B), (C)} Second and third embedding coordinates, respectively, versus the first embedding coordinate for the live imaging data set presented in \fig~3.
-Note that coordinates 2 and 3 are harmonics/functions of coordinate 1, and are therefore not informative about structure in the data set.
+Note that coordinates 2 and 3 are higher harmonics (and thus simple functions) of coordinate 1, and are therefore not informative about structure in the data set.
 We can conclude that the data set is effectively one-dimensional.
 {\it (D)} Eigenvalue spectrum for the live imaging data set presented in \fig~4.
-Note that there is a gap after the first coordinate; we can conclude that the data set is effectively one-dimensional.
+Note that there is a gap after the first eigenvalue; we can conclude that the data set is effectively one-dimensional.
 }

### 2014-09-01 image_analysis_paper2/pnas_draft.tex
added yannis comments

@@
 %
-However, as the size of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
+However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
 %
@@
 %
-This idea of locality is denoted by the edges in \fig~\ref{subfig:dmaps1}.
-%
-Data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
+This idea of locality is denoted by the color of the edges in \fig~\ref{subfig:dmaps1}:
+data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
 %
@@
 
-To validate the proposed approach, we applied our algorithm to a data set where the true temporal order and rotational orientation of the images are known {\em a priori}.
+To validate the proposed approach, we applied our algorithm to a data set where the true temporal order and rotational orientation of the images were known {\em a priori}.
 %
@@
 %
-Often in practice, each images comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
+Often in practice, each image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
 %
@@
 
-\fig~\ref{subfig:fixed_images_registered_ordered} shows the images from \fig~\ref{subfig:fixed_images_unregistered_unordered}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
+\fig~\ref{subfig:fixed_images_registered_ordered} shows the selected images in \fig~\ref{subfig:fixed_images_unregistered_unordered}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
 %
-However, because of interembryo variability, the selected images are not entirely representative of the developmental dynamics across the entire data set.
+However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set (see \ref{fig:full_fixed_data}).
 %
@@
 %
-We can now easily see the developmental progression, which is consistent with the known dynamics. 
-%
+We can now easily see the developmental progression, which is consistent with the known dynamics: 
 dpERK first appears as two lateral peaks at the ventral side of the embryo; a third dpERK peak then appears at the dorsal side of the embryo.
@@
 %
-Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental dynamics (e.g., correlation of age with body size).
+Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental features (e.g., correlation of age with body size).
 
@@
 %
-The ranks from this expert-based ordering for the images in \fig~\ref{subfig:fixed_images_registered_ordered} are indicated, and the rank correlation (see \fig~\ref{subfig:fixed_images_rank_corr}) shows that our ordering is consistent with the expert ordering.
+The ranks from the ordering provided by the embryologist, which we will refer to as the ``expert rank'', are indicated for the images in \fig~\ref{subfig:fixed_images_registered_ordered} , and the rank correlation (see \fig~\ref{subfig:fixed_images_rank_corr}) shows that our ordering is consistent with the expert ordering.
 
@@
 %
-The computational time is a function of the number of images in the data set, the number of pixels in each point, and the number of angles used to compute the pairwise rotations (see \SI). 
+The computational time is a function of the number of images in the data set, the number of pixels in each point, and the angular resolution to compute the pairwise rotations (see \SI). 
 %
-Empirically, the CPU time required was $\mathcal{O}(n^{1.5})$ in the number of images, and $\mathcal{O}(n)$ in the number of pixels and the number of angles.
+Empirically, the CPU time required was $\mathcal{O}(n^{1.5})$ in the number of images, and $\mathcal{O}(n)$ in the number of pixels and the angular resolution.
 %
@@
 %
-Two algorithmic parameters, the number of angles to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{subfig:dmaps1} and \SI), must also be defined.
+Two algorithmic parameters, the angular discretization to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{subfig:dmaps1} and \SI), must also be defined.
 %
-We found that the results are robust to both of these parameters, and that consistent parameter values were appropriate for all data sets presented here. 
+We found that the results are robust to both of these parameters, and that consistent parameter values served well for all data sets presented here. 
 %
@@
 %
-The tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, especially when the data sets are large.  
-%
-Therefore, this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.
+The tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, and so this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.
 
@@
 %
-In parallel, temporal ordering of large-scale data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different chemical species.
+Temporal ordering of large-scale data which did not require registration was done in the context of molecular profiling studies, in which data are vectors describing the expression levels of different mRNA.
 %
@@
 %
-This necessitates the development of automated methodologies which are sufficiently general to address a wide variety of applications to organize such large data sets.
+This necessitates the development of automated methodologies to organize such large data sets;
+such methods should be sufficiently general to address a wide variety of applications
 %
-Vector diffusion maps allows us to automatically register images, an essential task for many applications.
+Vector diffusion maps allow us to automatically register images, an essential task for many applications.
 %
-Simultaneously, the algorithm provides us with coordinates for each image.
+Simultaneously, the algorithm provides us with parameters to describe each image.
 %
@@
 %
-In general, we can recover several coordinates that serve as a good parameterization of the data set.
+In general, we can recover several parameters which concisely and comprehensively describe data set.
 %

### 2014-09-01 image_analysis_paper2/pnas_draft.tex
added stas comments

@@
 %
-To illustrate this approach, we temporally order and register data sets collected
-in studies of cell signaling in the early fruit fly embryo.}
+To illustrate this approach, we order and register data from imaging studies of {\it Drosophila}embryogenesis. }
 
@@
 %
-In one of the most commonly used experimental formats, developmental progress is reconstructed from snapshots of chemical or mechanical processes in fixed embryos.
+Developmental progress is commonly reconstructed from snapshots of chemical or mechanical processes in fixed embryos.
 %
-As a first step in these reconstructions, such snapshots must be spatially registered and ordered in time.
+As a first step in these reconstructions, snapshots must be spatially registered and ordered in time.
 %
-Currently, registration and ordering of images is often done manually, requiring a significant amount of expertise with a specific developmental system.
+Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific developmental system.
 %
@@
 %
-We validate our approach using a live imaging data set where the correct rotational orientation and temporal order are known.
-%
-Finally, we demonstrate that our approach can successfully extract a representative developmental trajectory from a collection of images of fixed {\it Drosophila} embryos.
+To illustrate our approach, we register and order data from imaging studies of pattern formation and morphogenesis in  {\it Drosophila} embryos.
 
@@
 This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
-applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \cite{zhao2014rotationally, singer2011viewing} to face recognition \cite{lafon2006data}.
+applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \cite{zhao2014rotationally, singer2011viewing} to face recognition \cite{lafon2006data} and classification of CT scans \cite{fernandez2014diffusion}.
 %
@@
 %
-To illustrate our approach, we analyze two data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \cite{jaeger2012drosophila}.
+To illustrate our approach, we show analyze two data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \cite{jaeger2012drosophila}.
 %
@@
 %
-Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parametrization of data that lies on a low-dimensional, perhaps nonlinear, manifold in high-dimensional space.
+Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parameterization of data that lies on a low-dimensional manifold in high-dimensional space.
 %
@@
 %
-We use {\it local} information about the data to find a parametrization  which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parametrization.
+We use {\it local} information about the data to find a parameterization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parameterization.
 %
@@
 %
-The color in \fig~\ref{subfig:dmaps2} depicts the one-dimensional parametrization or ordering of the data that we can detect visually.
+The color in \fig~\ref{subfig:dmaps2} depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
 %
@@
 %
-Instead, we will use diffusion maps to automatically uncover a parametrization of our high-dimensional data (see \SI).
+Instead, we will use diffusion maps to automatically uncover a parameterization of our high-dimensional data (see \SI).
 %
@@
 %
-This data set was obtained through live imaging of a {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
+This data set was obtained through live imaging of a vertically oriented {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
 %
@@
 %
-\fig~\ref{subfig:movie_angle_corr} and \fig~\ref{subfig:movie_rank_corr}  show the correlations between the recovered and true angles and ranks, respectively, for the entire data set. 
+\fig~\ref{subfig:movie_angle_corr} and \fig~\ref{subfig:movie_rank_corr}  show the correlations between the recovered and true angles and ranks, respectively, for the entire data set: 
 %
@@
 % 
+At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
+%
 Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignments and the errors in the recovered angles were consistently less than $10^{\circ}$.
 %
-Therefore, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.
+In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.
 
@@
 %
-Often in practice, each image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
+In practice, each image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
 %
@@
 %
-The resulting data set is therefore spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability. 
+The resulting data set is spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability. 
 %
@@
 %
-This data set is more complex than the live imaging data sets in that it contains significantly more images, each of which provide information about tissue morphology and the spatial distribution of two regulatory proteins.
+This data set is more complex than the live imaging data sets in that it contains significantly more images, each of which provides information about tissue morphology and the spatial distribution of two regulatory proteins.
 %
@@
 %
-Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that, in this context, specifies a subset of neuronal cells.
+Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that, in this context, specifies a subset of neuronal cells \cite{Lim2013kinetics}.
 
@@
 
-We can use prior knowledge about the developmental system to evaluate the quality of our registration and ordering. 
+To evaluate the quality of our registration and ordering, we can use prior knowledge about the developmental system. 
 %
@@
 %
-We found that the standard deviation in the location of this peak in the set of registered images was 8$^\circ$,
+We found that the standard deviation in the location of this peak in the set of registered images was $\sim$8$^\circ$,
 indicating that the algorithm successfully aligns the ventralmost points of the images. 
@@
 %
-The requisite user intervention and parameter tuning required for our method is relatively minimal.
-%
-Images must first be preprocessed so that the Euclidean distance between the pixels is informative.
+The requisite user intervention and parameter tuning required for our method is relatively minimal: Images must first be preprocessed so that the Euclidean distance between the pixels is informative.
 %
@@
 %
-The tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, and so this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.
+Overall, the tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, and so this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.
 
@@
 
-We presented a unified approach to temporal ordering and registration of images in studies of developmental dynamics. 
+Already at this point, the rate of collection of imaging data has surpassed the rate of manual image analysis.
+%
+This necessitates the development of automated methodologies to organize such large data sets. 
 %
@@
 %
-In particular, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \cite{yuan2014automated, surkova2008characterization}.
+Temporal ordering of large-scale data which did not require registration was done in the context of molecular profiling studies, in which data are vectors describing the expression levels of different mRNA \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
 %
-Temporal ordering of large-scale data which did not require registration was done in the context of molecular profiling studies, in which data are vectors describing the expression levels of different mRNA.
-%
-Automated temporal ordering of such data sets was accomplished by first projecting the data onto a low-dimensional subspace spanned by the leading principal or independent components, and then solving a traveling salesman problem or constructing a minimum spanning tree on the projected data to order multiple snapshots \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
-%
-The presented approach addresses image registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace.
-%; vector diffusion maps can also order images when global registration of the entire data set is topologically impossible \cite{zhao2014rotationally}.
+At the same time, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \cite{yuan2014automated, surkova2008characterization}.
 %
@@
 %
-Although we have presented results for a specific type of data (cross-sectional images taken along the dorsoventral axis of developing {\em Drosophila} embryos), our methodology only relies on computing distances between images and may be readily extended to many other imaging data sets.
+Although we have presented results for a specific type of data (cross-sectional images taken along the dorsoventral axis of {\em Drosophila} embryos), our methodology only relies on computing distances between images and may be readily extended to many other data sets.
 
-Previously, angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy images \cite{singer2012vector, zhao2014rotationally, singer2011viewing}.
+Angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy images \cite{singer2012vector, zhao2014rotationally, singer2011viewing}.
 %
@@
 %
-In general, the size of the data set required for accurate registration and ordering is a function of the instrument noise, interembryo variablity, and the complexity of the developmental dynamics.
+In general, the size of the data set required for accurate registration and ordering is a function of the instrument noise, interembryo variability, and the complexity of the developmental dynamics.
 %
-%Here, we have used vector diffusion maps to register the images with respect to rotations; however, one could factor out other sources of variability, such as translations, reflections \cite{singer2012vector, goemans1995improved, bandeira2013cheeger} or dilations.
 
-Already at this point, the rate of collection of imaging data has surpassed the rate of manual image analysis.
-%
-This necessitates the development of automated methodologies to organize such large data sets;
-such methods should be sufficiently general to address a wide variety of applications
-%
 Vector diffusion maps allow us to automatically register images, an essential task for many applications.
@@
 %
-Images taken from different viewing directions can be analyzed, as the vector diffusion maps parameterization will organize the images according to the viewing angle.
+Furthermore, images taken from different viewing directions can be analyzed, as the vector diffusion maps parameterization will organize the images according to the viewing angle.
 %
@@
 %
-Code is available at .....
+Code is available at \texttt{genomics.princeton.edu/stas/publications.html} under ``Codes and Data''.
 
@@
 \bibliographystyle{pnas2}
-\bibliography{../image_analysis_paper/background_reading/references,../../references/references}
+\bibliography{../../references/references,../image_analysis_paper/background_reading/references}
 

### 2014-09-02 image_analysis_paper2/pnas_draft.tex
added small edits

@@
 %
-To illustrate our approach, we show analyze two data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \cite{jaeger2012drosophila}.
+To illustrate our approach, we analyze two data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \cite{jaeger2012drosophila}.
 %
@@
 %
-This data set was obtained through live imaging of a vertically oriented {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
+This data set was obtained through live imaging near the posterior pole of a vertically oriented {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
 %
@@
 %
-In contrast, our method requires relatively little computational time; registering and ordering this fixed imaging data set took $15$~seconds on an Intel Core i7 2.93 GHz processor.
+In contrast, our method requires relatively little computational time; registering and ordering this fixed imaging data set took $26$~seconds on an Intel Core i7 2.93 GHz processor.
 %for $100$ images of $100 \times 100$ pixels, using 10$^{\circ}$ rotational discretizations.

### 2014-09-03 image_analysis_paper2/SI_text.tex
changed tex to have bib things in tex for PNAs submission

@@
 
-\bibliographystyle{pnas2}
-\bibliography{../image_analysis_paper/background_reading/references,../../references/references}
+%\bibliographystyle{pnas2}
+%\bibliography{../image_analysis_paper/background_reading/references,../../references/references}
+
+\begin{thebibliography}{10}
+
+\bibitem{canny1986computational}
+Canny J (1986) A computational approach to edge detection. \textit{IEEE Trans Pattern Anal Mach Intell} (6):679--698.
+
+\bibitem{singer2011angular}
+Singer A (2011) Angular synchronization by eigenvectors and semidefinite
+  programming. \textit{Appl Comput Harmon Anal}
+  30(1):20--36.
+
+\bibitem{ahuja2007template}
+Ahuja S, Kevrekidis I~G, Rowley C~W (2007) Template-based stabilization of
+  relative equilibria in systems with continuous symmetry. \textit{Journal of
+  Nonlinear Science} 17(2):109--143.
+
+\bibitem{goemans1995improved}
+Goemans M~X, Williamson D~P (1995) Improved approximation algorithms for
+  maximum cut and satisfiability problems using semidefinite programming.
+  \textit{Journal of the ACM (JACM)} 42(6):1115--1145.
+
+\bibitem{bandeira2013cheeger}
+Bandeira A~S, Singer A, Spielman D~A (2013) A {C}heeger inequality for the
+  graph connection {L}aplacian. \textit{SIAM Journal on Matrix Analysis and
+  Applications} 34(4):1611--1630.
+
+\bibitem{coifman2005geometric}
+Coifman R~R, Lafon S, Lee A~B, Maggioni M, Nadler B, \textit{et~al.} (2005)
+  Geometric diffusions as a tool for harmonic analysis and structure definition
+  of data: Diffusion maps. \textit{Proc Natl Acad Sci USA} 102(21):7426--7431.
+
+\bibitem{coifman2008graph}
+Coifman R~R, Shkolnisky Y, Sigworth F~J, Singer A (2008) Graph {L}aplacian
+  tomography from unknown random projections. \textit{IEEE Trans Image Process} 17(10):1891--1899.
+
+\bibitem{rohrdanz2011determination}
+Rohrdanz M~A, Zheng W, Maggioni M, Clementi C (2011) Determination of reaction
+  coordinates via locally scaled diffusion map. \textit{The J Chem Phys} 134:124116.
+
+\bibitem{Belkin2003}
+Belkin M, Niyogi P (2003) Laplacian eigenmaps for dimensionality reduction and
+  data representation. \textit{Neural Comput} 15:1373--1396.
+
+\bibitem{singer2012vector}
+Singer A, Wu H~T (2012) Vector diffusion maps and the connection {L}aplacian.
+  \textit{Commun Pure Appl Math} 65(8):1067--1144.
+
+\bibitem{ferguson2010systematic}
+Ferguson A~L, Panagiotopoulos A~Z, Debenedetti P~G, Kevrekidis I~G (2010)
+  Systematic determination of order parameters for chain dynamics using
+  diffusion maps. \textit{Proc Natl Acad Sci USA}
+  107(31):13597--13602.
+
+\bibitem{kemelmacher2011exploring}
+Kemelmacher-Shlizerman I, Shechtman E, Garg R, Seitz S~M (2011) Exploring
+  photobios, \textit{ACM Trans Graph} (ACM), volume~30,
+  p.~61.
+
+\end{thebibliography}
 

### 2014-09-03 image_analysis_paper2/pnas_draft.tex
everything updated for resubmission
changed tex to have bib things in tex for PNAs submission

@@
 
-\graphicspath{ {paper_figures2/} }
-
 %\usepackage{tikz}
@@
 \newcommand{\fig}[0]{Fig.}
+
+\font\xfigtextfontit=\frutigeroblique at 7pt
 
@@
 %
-To illustrate this approach, we order and register data from imaging studies of {\it Drosophila}embryogenesis. }
+To illustrate this approach, we order and register data from imaging studies of {\xfigtextfontit  Drosophila} embryogenesis. }
 
@@
 
-\begin{figure}[t]
-\includegraphics{fig2}
-\caption{Schematic illustrating angular synchronization and diffusion maps.
+%\begin{figure}[t]
+%\includegraphics{fig2}
+%\caption{Schematic illustrating angular synchronization and diffusion maps.
 {\xfigtextfontit (A)} Set of vectors, each in a different orientation.
@@
 The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
-%\label{fig:schematics}
-\customlabel{fig:schematics}{2}
-\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
-\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
-\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
-\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
-\end{figure}
+%%\label{fig:schematics}
+%\customlabel{fig:schematics}{2}
+%\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
+%\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
+%\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
+%\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
+%\end{figure}
 
-\begin{figure*}[t]
-\includegraphics{fig3}
-\caption{Method validation using live imaging of {\xfigtextfontit Drosophila} embryos.{\xfigtextfontit (A)} Selected images from a live imaging study of a {\xfigtextfontit Drosophilla} embryo during gastrulation.
+%\begin{figure*}[t]
+%\includegraphics{fig3}
+%\caption{Method validation using live imaging of {\xfigtextfontit Drosophila} embryos.{\xfigtextfontit (A)} Selected images from a live imaging study of a {\xfigtextfontit Drosophilla} embryo during gastrulation.
 Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
@@
 }
-\customlabel{fig:movie}{3}
-\customlabel{subfig:movie_unregistered_unordered}{\ref{fig:movie}{\it A}}
-\customlabel{subfig:movie_registered_ordered}{\ref{fig:movie}{\it B}}
-\customlabel{subfig:movie_angle_corr}{\ref{fig:movie}{\it C}}
-\customlabel{subfig:movie_rank_corr}{\ref{fig:movie}{\it D}}
-\customlabel{subfig:movie_error_table}{\ref{fig:movie}{\it E}}
-\customlabel{subfig:movie_bootstrap}{\ref{fig:movie}{\it F}}
-\end{figure*}
+%\customlabel{fig:movie}{3}
+%\customlabel{subfig:movie_unregistered_unordered}{\ref{fig:movie}{\it A}}
+%\customlabel{subfig:movie_registered_ordered}{\ref{fig:movie}{\it B}}
+%\customlabel{subfig:movie_angle_corr}{\ref{fig:movie}{\it C}}
+%\customlabel{subfig:movie_rank_corr}{\ref{fig:movie}{\it D}}
+%\customlabel{subfig:movie_error_table}{\ref{fig:movie}{\it E}}
+%\customlabel{subfig:movie_bootstrap}{\ref{fig:movie}{\it F}}
+%\end{figure*}
 
@@
 
-\begin{figure*}[t]
-\includegraphics{fig4}
-\caption{Analyzing images of fixed {\xfigtextfontit Drosophila} embryos.
+%\begin{figure*}[t]
+%\includegraphics{fig4}
+%\caption{Analyzing images of fixed {\xfigtextfontit Drosophila} embryos.
 {\xfigtextfontit (A)} Images of {\xfigtextfontit Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red).
@@
 }
-\customlabel{fig:data2}{4}
-\customlabel{subfig:fixed_images_unregistered_unordered}{\ref{fig:data2}{\it A}}
-\customlabel{subfig:fixed_images_registered_ordered}{\ref{fig:data2}{\it B}}
-\customlabel{subfig:fixed_images_average_trajectory}{\ref{fig:data2}{\it C}}
-\customlabel{subfig:fixed_images_rank_corr}{\ref{fig:data2}{\it D}}
-\end{figure*}
+%\customlabel{fig:data2}{4}
+%\customlabel{subfig:fixed_images_unregistered_unordered}{\ref{fig:data2}{\it A}}
+%\customlabel{subfig:fixed_images_registered_ordered}{\ref{fig:data2}{\it B}}
+%\customlabel{subfig:fixed_images_average_trajectory}{\ref{fig:data2}{\it C}}
+%\customlabel{subfig:fixed_images_rank_corr}{\ref{fig:data2}{\it D}}
+%\end{figure*}
 
@@
 
-\bibliographystyle{pnas2}
-\bibliography{../../references/references,../image_analysis_paper/background_reading/references}
+%\bibliographystyle{pnas2}
+%\bibliography{../../references/references,../image_analysis_paper/background_reading/references}
+
+\begin{thebibliography}{10}
+
+\bibitem{jaeger2004dynamic}
+Jaeger J, Surkova S, Blagov M, Janssens H, Kosman D, \textit{et~al.} (2004)
+  Dynamic control of positional information in the early \textit{{D}rosophila}
+  embryo. \textit{Nature} 430(6997):368--371.
+
+\bibitem{peter2011gene}
+Peter I~S, Davidson E~H (2011) A gene regulatory network controlling the
+  embryonic specification of endoderm. \textit{Nature} 474(7353):635--639.
+
+\bibitem{fowlkes2008quantitative}
+Fowlkes C~C, Hendriks C~L~L, Ker{\"a}nen S~V, Weber G~H, R{\"u}bel O,
+  \textit{et~al.} (2008) A quantitative spatiotemporal atlas of gene expression
+  in the \textit{{D}rosophila} blastoderm. \textit{Cell} 133(2):364--374.
+
+\bibitem{ng2012large}
+Ng L~L, Sunkin S~M, Feng D, Lau C, Dang C, \textit{et~al.} (2012) Large-scale
+  neuroinformatics for in situ hybridization data in the mouse brain.
+  \textit{Int Rev Neurobiol} 104:159--182.
+
+\bibitem{richardson2014emage}
+Richardson L, Stevenson P, Venkataraman S, Yang Y, Burton N, \textit{et~al.}
+  (2014) Emage: Electronic mouse atlas of gene expression, \textit{Mouse
+  Molecular Embryology} (Springer), pp. 61--79.
+
+\bibitem{castro2009automatic}
+Castro C, Luengo-Oroz M, Desnoulez S, Duloquin L, Fernandez-de Manuel L,
+  \textit{et~al.} (2009) An automatic quantification and registration strategy
+  to create a gene expression atlas of zebrafish embryogenesis,
+  \textit{Conf Proc IEEE Eng Med Biol Soc}, pp. 1469--1472.
+
+\bibitem{zitova2003image}
+Zitova B, Flusser J (2003) Image registration methods: a survey. \textit{Image
+   Vis Comput} 21(11):977--1000.
+
+\bibitem{rowley1998rotation}
+Rowley H~A, Baluja S, Kanade T (1998) Rotation invariant neural network-based
+  face detection, \textit{Conf Proc IEEE Comp Soc on Computer Vision and Pattern Recognition} (IEEE), pp. 38--44.
+
+\bibitem{hajnal2010medical}
+Hajnal J~V, Hill D~L (2010) \textit{Medical image registration} (CRC press).
+
+\bibitem{greenspan1994rotation}
+Greenspan H, Belongie S, Goodman R, Perona P (1994) Rotation invariant texture
+  recognition using a steerable pyramid, \textit{Pattern Recognition, 1994.
+  Vol. 2-Conference B: Computer Vision \& Image Processing., Proceedings of the
+  12th IAPR International. Conference on} (IEEE), volume~2, pp. 162--167.
+
+\bibitem{zhao2003face}
+Zhao W, Chellappa R, Phillips P~J, Rosenfeld A (2003) Face recognition: A
+  literature survey. \textit{ACM Computing Surveys (CSUR)} 35(4):399--458.
+
+\bibitem{singer2012vector}
+Singer A, Wu H~T (2012) Vector diffusion maps and the connection {L}aplacian.
+  \textit{Commun Pure Appl Math} 65(8):1067--1144.
+
+\bibitem{Belkin2003}
+Belkin M, Niyogi P (2003) Laplacian eigenmaps for dimensionality reduction and
+  data representation. \textit{Neural Comput} 15:1373--1396.
+
+\bibitem{coifman2005geometric}
+Coifman R~R, Lafon S, Lee A~B, Maggioni M, Nadler B, \textit{et~al.} (2005)
+  Geometric diffusions as a tool for harmonic analysis and structure definition
+  of data: Diffusion maps. \textit{Proc Natl Acad Sci USA} 102(21):7426--7431.
+
+\bibitem{coifman2006geometric}
+Coifman R~R, Lafon S (2006) Geometric harmonics: a novel tool for multiscale
+  out-of-sample extension of empirical functions. \textit{Appl Comput Harmon Anal} 21(1):31--52.
+
+\bibitem{tenenbaum2000global}
+Tenenbaum J~B, De~Silva V, Langford J~C (2000) A global geometric framework for
+  nonlinear dimensionality reduction. \textit{Science} 290(5500):2319--2323.
+
+\bibitem{roweis2000nonlinear}
+Roweis S~T, Saul L~K (2000) Nonlinear dimensionality reduction by locally
+  linear embedding. \textit{Science} 290(5500):2323--2326.
+
+\bibitem{zhao2014rotationally}
+Zhao Z, Singer A (2014) Rotationally invariant image representation for viewing
+  direction classification in cryo-em. \textit{J Struct Biol}
+  186(1):153 -- 166.
+
+\bibitem{singer2011viewing}
+Singer A, Zhao Z, Shkolnisky Y, Hadani R (2011) Viewing angle classification of
+  cryo-electron microscopy images using eigenvectors. \textit{SIAM J Imaging Sci} 4(2):723--759.
+
+\bibitem{lafon2006data}
+Lafon S, Keller Y, Coifman R~R (2006) Data fusion and multicue data matching by
+  diffusion maps. \textit{IEEE Trans Pattern Anal Mach Intell} 28(11):1784--1797.
+
+\bibitem{fernandez2014diffusion}
+Fern{\'a}ndez {\'A}, Rabin N, Coifman R~R, Eckstein J (2014) Diffusion methods
+  for aligning medical datasets: Location prediction in {CT} scan images.
+  \textit{Med Image Anal} 18(2):425--432.
+
+\bibitem{jaeger2012drosophila}
+Jaeger J, Manu, Reinitz J (2012) \textit{{D}rosophila} blastoderm patterning.
+  \textit{Curr Opin Genet Dev} 22(6):533 -- 541, genetics of system biology.
+
+\bibitem{singer2011angular}
+Singer A (2011) Angular synchronization by eigenvectors and semidefinite
+  programming. \textit{Appl Comput Harmon Anal}
+  30(1):20--36.
+
+\bibitem{ahuja2007template}
+Ahuja S, Kevrekidis I~G, Rowley C~W (2007) Template-based stabilization of
+  relative equilibria in systems with continuous symmetry. \textit{Journal of
+  Nonlinear Science} 17(2):109--143.
+
+\bibitem{ian1998statistical}
+Dryden I~L, Mardia K (1998) \textit{Statistical shape analysis}, Wiley series
+  in probability and statistics: Probability and statistics (J. Wiley).
+
+\bibitem{leptin2005gastrulation}
+Leptin M (2005) Gastrulation movements: the logic and the nuts and bolts.
+  \textit{Dev Cell} 8(3):305--320.
+
+\bibitem{Lim2013kinetics}
+Lim B, Samper N, Lu H, Rushlow C, Jim{\'e}nez G, \textit{et~al.} (2013)
+  Kinetics of gene derepression by {ERK} signaling. \textit{Proc Natl Acad Sci USA} 110(25):10330--10335.
+
+\bibitem{anavy2014blind}
+Anavy L, Levin M, Khair S, Nakanishi N, Fernandez-Valverde S~L, \textit{et~al.}
+  (2014) {BLIND} ordering of large-scale transcriptomic developmental
+  timecourses. \textit{Development} 141(5):1161--1166.
+
+\bibitem{trapnell2014dynamics}
+Trapnell C, Cacchiarelli D, Grimsby J, Pokharel P, Li S, \textit{et~al.} (2014)
+  The dynamics and regulators of cell fate decisions are revealed by
+  pseudotemporal ordering of single cells. \textit{Nat Biotechnol}
+  32(4):381--386.
+
+\bibitem{gupta2008extracting}
+Gupta A, Bar-Joseph Z (2008) Extracting dynamics from static cancer expression
+  data. \textit{IEEE/ACM Trans Comput Biol Bioinform} 5(2):172--182.
+
+\bibitem{yuan2014automated}
+Yuan L, Pan C, Ji S, McCutchan M, Zhou Z~H, \textit{et~al.} (2014) Automated
+  annotation of developmental stages of \textit{{D}rosophila} embryos in images
+  containing spatial patterns of expression. \textit{Bioinformatics}
+  30(2):266--273.
+
+\bibitem{surkova2008characterization}
+Surkova S, Kosman D, Kozlov K, Myasnikova E, Samsonova A~A, \textit{et~al.}
+  (2008) Characterization of the \textit{{D}rosophila} segment determination
+  morphome. \textit{Dev Biol} 313(2):844--862.
+
+\bibitem{krzic2012multiview}
+Krzic U, Gunther S, Saunders T~E, Streichan S~J, Hufnagel L (2012) Multiview
+  light-sheet microscope for rapid in toto imaging. \textit{Nat Methods}
+  9(7):730--733.
+
+\bibitem{ichikawa2014live}
+Ichikawa T, Nakazato K, Keller P~J, Kajiura-Kobayashi H, Stelzer E~H,
+  \textit{et~al.} (2014) Live imaging and quantitative analysis of gastrulation
+  in mouse embryos using light-sheet microscopy and 3d tracking tools.
+  \textit{Nat Protoc} 9(3):575--585.
+
+\bibitem{rubel2010coupling}
+R{\"u}bel O, Ahern S, Bethel E, Biggin M~D, Childs H, \textit{et~al.} (2010)
+  Coupling visualization and data analysis for knowledge discovery from
+  multi-dimensional scientific data. \textit{Procedia Comput Sci}
+  1(1):1757--1764.
+
+\bibitem{chung2010microfluidic}
+Chung K, Kim Y, Kanodia J~S, Gong E, Shvartsman S~Y, \textit{et~al.} (2011) A
+  microfluidic array for large-scale ordering and orientation of embryos.
+  \textit{Nat Methods} 8(2):171--176.
+
+\end{thebibliography}
 
@@
 
+\begin{figure}[t]
+\includegraphics{fig1}
+\caption{Caricature illustrating the tasks of image registration and temporal ordering.
+{\xfigtextfontit (A)} Images of ``samples'', each in a different orientation and a different stage of development.
+{\xfigtextfontit (B)} Registered and ordered samples.
+For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
+%\label{fig:fish}
+\customlabel{fig:fish}{1}
+\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
+\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
+\end{figure}
+
+\begin{figure}[t]
+\includegraphics{fig2}
+\caption{Schematic illustrating angular synchronization and diffusion maps.
+{\xfigtextfontit (A)} Set of vectors, each in a different orientation.
+The pairwise alignment angles are indicated.
+{\xfigtextfontit (B)} The vectors from {\xfigtextfontit A}, each rotated about their midpoint so that the set is globally aligned.
+Note that the chosen rotation angles are consistent with the pairwise alignments in {\xfigtextfontit A}: the differences between a pair of angles in {\xfigtextfontit B} is the same as the pairwise angle in {\xfigtextfontit A}.
+{\xfigtextfontit (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
+Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\xfigtextfontit \SI}), so that close data points are connected by darker (``stronger'') edges.
+{\xfigtextfontit (D)} The data in {\xfigtextfontit C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure.
+The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
+%\label{fig:schematics}
+\customlabel{fig:schematics}{2}
+\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
+\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
+\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
+\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
+\end{figure}
+
+\begin{figure*}[t]
+\includegraphics{fig3}
+\caption{Method validation using live imaging of {\xfigtextfontit Drosophila} embryos.{\xfigtextfontit (A)} Selected images from a live imaging study of a {\xfigtextfontit Drosophilla} embryo during gastrulation.
+Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
+{\xfigtextfontit (B)} Images from {\xfigtextfontit A} after registration and ordering using vector diffusion maps.
+The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.
+{\xfigtextfontit (C)} The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle {\xfigtextfontit (D)} The correlation between the recovered rank (using vector diffusion maps) and the true rank.
+{\xfigtextfontit (E)} The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies.
+{\xfigtextfontit (F)} The median rank correlation coefficient (over $200$ samples) as a function of the number of images in the data set.
+}
+\customlabel{fig:movie}{3}
+\customlabel{subfig:movie_unregistered_unordered}{\ref{fig:movie}{\it A}}
+\customlabel{subfig:movie_registered_ordered}{\ref{fig:movie}{\it B}}
+\customlabel{subfig:movie_angle_corr}{\ref{fig:movie}{\it C}}
+\customlabel{subfig:movie_rank_corr}{\ref{fig:movie}{\it D}}
+\customlabel{subfig:movie_error_table}{\ref{fig:movie}{\it E}}
+\customlabel{subfig:movie_bootstrap}{\ref{fig:movie}{\it F}}
+\end{figure*}
+
+\begin{figure*}[t]
+\includegraphics{fig4}
+\caption{Analyzing images of fixed {\xfigtextfontit Drosophila} embryos.
+{\xfigtextfontit (A)} Images of {\xfigtextfontit Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red).
+Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation.
+{\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps.
+The expert rank for each image is indicated.
+{\xfigtextfontit (C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI).
+{\xfigtextfontit (D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert.
+The rank correlation coefficient is $0.97$.
+}
+\customlabel{fig:data2}{4}
+\customlabel{subfig:fixed_images_unregistered_unordered}{\ref{fig:data2}{\it A}}
+\customlabel{subfig:fixed_images_registered_ordered}{\ref{fig:data2}{\it B}}
+\customlabel{subfig:fixed_images_average_trajectory}{\ref{fig:data2}{\it C}}
+\customlabel{subfig:fixed_images_rank_corr}{\ref{fig:data2}{\it D}}
+\end{figure*}
+
 \end{document}

### 2014-11-03 image_analysis_paper_development/manuscript.tex
added stuff to submit to development

--- 2014-09-02 image_analysis_paper2/pnas_draft.tex
@@
-%% PNAStmpl.tex
-%% Template file to use for PNAS articles prepared in LaTeX
-%% Version: Apr 14, 2008
+\documentclass[twocolumn, 10pt]{article}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% BASIC CLASS FILE
-%% PNAStwo for two column articles is called by default.
-%% Uncomment PNASone for single column articles. One column class
-%% and style files are available upon request from pnas@nas.edu.
-%% (uncomment means get rid of the '%' in front of the command)
+\usepackage[margin=0.75in]{geometry}
+\usepackage{amssymb,amsfonts,amsmath}
+\usepackage{graphicx}
 
-%\documentclass{pnasone}
-\documentclass{pnastwo}
+\usepackage{natbib}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Changing position of text on physical page:
-%% Since not all printers position
-%% the printed page in the same place on the physical page,
-%% you can change the position yourself here, if you need to:
+\renewcommand{\figurename}{Fig.}
+\newcommand{\SI}[0]{\textit{SI Materials and Methods}}
+\newcommand{\fig}[0]{Fig.}
 
-% \advance\voffset -.5in % Minus dimension will raise the printed page on the
-                         %  physical page; positive dimension will lower it.
-
-%% You may set the dimension to the size that you need.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% OPTIONAL GRAPHICS STYLE FILE
-
-%% Requires graphics style file (graphicx.sty), used for inserting
-%% .eps files into LaTeX articles.
-%% Note that inclusion of .eps files is for your reference only;
-%% when submitting to PNAS please submit figures separately.
-
-%% Type into the square brackets the name of the driver program
-%% that you are using. If you don't know, try dvips, which is the
-%% most common PC driver, or textures for the Mac. These are the options:
-
-% [dvips], [xdvi], [dvipdf], [dvipdfm], [dvipdfmx], [pdftex], [dvipsone],
-% [dviwindo], [emtex], [dviwin], [pctexps], [pctexwin], [pctexhp], [pctex32],
-% [truetex], [tcidvi], [vtex], [oztex], [textures], [xetex]
-
-%\usepackage[dvips]{graphicx}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% OPTIONAL POSTSCRIPT FONT FILES
-
-%% PostScript font files: You may need to edit the PNASoneF.sty
-%% or PNAStwoF.sty file to make the font names match those on your system.
-%% Alternatively, you can leave the font style file commands commented out
-%% and typeset your article using the default Computer Modern
-%% fonts (recommended). If accepted, your article will be typeset
-%% at PNAS using PostScript fonts.
-
-% Choose PNASoneF for one column; PNAStwoF for two column:
-%\usepackage{PNASoneF}
-\usepackage{PNAStwoF}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% ADDITIONAL OPTIONAL STYLE FILES
-
-%% The AMS math files are commonly used to gain access to useful features
-%% like extended math fonts and math commands.
-
-\usepackage{amssymb,amsfonts,amsmath}
-
-\graphicspath{ {paper_figures2/} }
-
-%\usepackage{tikz}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% OPTIONAL MACRO FILES
-%% Insert self-defined macros here.
-%% \newcommand definitions are recommended; \def definitions are supported
-
-%\newcommand{\mfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2}}
-%\def\s{\sigma}
-
-\newcommand{\SI}[0]{{\it SI Materials and Methods}}
-\newcommand{\fig}[0]{Fig.}
+\hyphenpenalty=10000
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Don't type in anything in the following section:
-%%%%%%%%%%%%
-%% For PNAS Only:
-\contributor{Submitted to Proceedings
-of the National Academy of Sciences of the United States of America}
-\url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
-\copyrightyear{2008}
-\issuedate{Issue Date}
-\volume{Volume}
-\issuenumber{Issue Number}
-%%%%%%%%%%%%
+\title{Temporal ordering and registration of images in studies of developmental dynamics}
+
+\author{Carmeline~J.~Dsilva \thanks{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
+Bomyi~Lim \footnotemark[1],
+Hang~Lu \thanks{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
+Amit~Singer \thanks{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \thanks{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA},
+Stanislav~Y.~Shvartsman \footnotemark[1] \thanks{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA}, \\
+and Ioannis~G.~Kevrekidis \footnotemark[1] \footnotemark[4]}
+
+\date{}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% For titles, only capitalize the first letter
-%% \title{Almost sharp fronts for the surface quasi-geostrophic equation}
-
-\title{Temporal ordering and registration of images in studies of developmental dynamics}
-
-%% Enter authors via the \author command.
-%% Use \affil to define affiliations.
-%% (Leave no spaces between author name and \affil command)
-
-%% Note that the \thanks{} command has been disabled in favor of
-%% a generic, reserved space for PNAS publication footnotes.
-
-%% \author{<author name>
-%% \affil{<number>}{<Institution>}} One number for each institution.
-%% The same number should be used for authors that
-%% are affiliated with the same institution, after the first time
-%% only the number is needed, ie, \affil{number}{text}, \affil{number}{}
-%% Then, before last author ...
-%% \and
-%% \author{<author name>
-%% \affil{<number>}{}}
-
-%% For example, assuming Garcia and Sonnery are both affiliated with
-%% Universidad de Murcia:
-%% \author{Roberta Graff\affil{1}{University of Cambridge, Cambridge,
-%% United Kingdom},
-%% Javier de Ruiz Garcia\affil{2}{Universidad de Murcia, Bioquimica y Biologia
-%% Molecular, Murcia, Spain}, \and Franklin Sonnery\affil{2}{}}
-
-\author{Carmeline~J.~Dsilva\affil{1}{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
-Bomyi~Lim\affil{1}{},
-Thomas~J.~Levario\affil{2}{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
-Hang~Lu\affil{2}{},
-Amit~Singer\affil{3}{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \affil{4}{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA},
-Stanislav~Y.~Shvartsman\affil{1}{} \affil{5}{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA},
-\and
-Ioannis~G.~Kevrekidis\affil{1}{} \affil{4}{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA}}
-
-\contributor{Submitted to Proceedings of the National Academy of Sciences
-of the United States of America}
-
-\significancetext{
-Spatial registration and temporal ordering of images of developing embryos are one of the first steps in the reconstruction of developmental dynamics. 
-%
-When the number of images is small and the differences between them are easily recognizable, image registration and ordering can be done manually; these tasks become challenging when the data sets are larger and more complex.
-%
-Here we show this can be automated and combined in a single
-step using recent advances in data mining. 
-%
-To illustrate this approach, we order and register data from imaging studies of {\it Drosophila}embryogenesis. }
-
-%% The \maketitle command is necessary to build the title page.
 \maketitle
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{article}
 
@@
 %
-The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require {\it a priori} knowledge of image features or a parametric model of the developmental dynamics.
+The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require \textit{a priori} knowledge of image features or a parametric model of the developmental dynamics.
 %
-To illustrate our approach, we register and order data from imaging studies of pattern formation and morphogenesis in  {\it Drosophila} embryos.
+To illustrate our approach, we register and order data from imaging studies of pattern formation and morphogenesis in  \textit{Drosophila} embryos.
+
+\vspace{1em}
+
+\noindent
+KEY WORDS: temporal ordering, image registration, vector diffusion maps
 
@@
 
-%% When adding keywords, separate each term with a straight line: |
-\keywords{temporal ordering | image registration | vector diffusion maps}
+\section*{Introduction}
 
-%\section{Significance}
-
-%\vspace{0.5cm}
-
-%% Optional for entering abbreviations, separate the abbreviation from
-%% its definition with a comma, separate each pair with a semicolon:
-%% for example:
-%% \abbreviations{SAM, self-assembled monolayer; OTS,
-%% octadecyltrichlorosilane}
-
-% \abbreviations{}
-
-%% The first letter of the article should be drop cap: \dropcap{}
-%\dropcap{I}n this article we study the evolution of ''almost-sharp'' fronts
-
-%% Enter the text of your article beginning here and ending before
-%% \begin{acknowledgements}
-%% Section head commands for your reference:
-%% \section{}
-%% \subsection{}
-%% \subsubsection{}
-
-\dropcap{I}n one of the common approaches to studies of developmental dynamics, a group of embryos is fixed and stained to visualize a particular biochemical or morphological process within a developing tissue.
+In one of the common approaches to studies of developmental dynamics, a group of embryos is fixed and stained to visualize a particular biochemical or morphological process within a developing tissue.
 %
-The developmental dynamics must then be reconstructed from multiple embryos, each of which contributes only a snapshot of the relevant process along its developmental trajectory \cite{jaeger2004dynamic, peter2011gene, fowlkes2008quantitative}.
+The developmental dynamics must then be reconstructed from multiple embryos, each of which contributes only a snapshot of the relevant process along its developmental trajectory \citep{jaeger2004dynamic, peter2011gene, fowlkes2008quantitative}.
 %
 Importantly, the ``age'' of any given embryo arrested in its development is often only approximately known; typically what is known is
-a certain time window to which a collection of embryos belongs \cite{ng2012large, richardson2014emage, castro2009automatic}.
+a certain time window to which a collection of embryos belongs \citep{ng2012large, richardson2014emage, castro2009automatic}.
 %
@@
 \caption{Caricature illustrating the tasks of image registration and temporal ordering.
-{\xfigtextfontit (A)} Images of ``samples'', each in a different orientation and a different stage of development.
-{\xfigtextfontit (B)} Registered and ordered samples.
+(A) Images of ``samples'', each in a different orientation and a different stage of development.
+(B) Registered and ordered samples.
 For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
@@
 \customlabel{fig:fish}{1}
-\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
-\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
+\customlabel{subfig:fish_unordered}{\ref{fig:fish}A}
+\customlabel{subfig:fish_ordered}{\ref{fig:fish}B}
 \end{figure}
@@
 %
-In contrast to a number of previous methodologies \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, our methodology does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression for temporal ordering.
+In contrast to a number of previous methodologies \citep{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, our methodology does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression for temporal ordering.
 %
-Our approach is based on a manifold learning algorithm (vector diffusion maps \cite{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
+Our approach is based on a manifold learning algorithm (vector diffusion maps \citep{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
 %
-This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
-applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \cite{zhao2014rotationally, singer2011viewing} to face recognition \cite{lafon2006data} and classification of CT scans \cite{fernandez2014diffusion}.
+This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \citep{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
+applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \citep{zhao2014rotationally, singer2011viewing} to face recognition \citep{lafon2006data} and classification of CT scans \citep{fernandez2014diffusion}.
 %
@@
 %
-To illustrate our approach, we analyze two data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \cite{jaeger2012drosophila}.
+To illustrate our approach, we analyze two data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
 %
@@
 
-\section{Results}
+\section*{Results}
 
-\subsection{Vector diffusion maps for registration and temporal ordering}
+\subsection*{Vector diffusion maps for registration and temporal ordering}
 
-Vector diffusion maps \cite{singer2012vector} is a manifold learning
+Vector diffusion maps \citep{singer2012vector} is a manifold learning
 technique developed for data sets which contain two sources of variability:
@@
 %
-Vector diffusion maps combine two algorithms, {\em angular synchronization} \cite{singer2011angular} for image registration and {\em diffusion maps} \cite{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
+Vector diffusion maps combine two algorithms, {\em angular synchronization} \citep{singer2011angular} for image registration and {\em diffusion maps} \citep{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
 %
@@
 
+\begin{figure}[t]
+\includegraphics{fig2}
+\caption{Schematic illustrating angular synchronization and diffusion maps.
+(A) Set of vectors, each in a different orientation.
+The pairwise alignment angles are indicated.
+(B) The vectors from A, each rotated about their midpoint so that the set is globally aligned.
+Note that the chosen rotation angles are consistent with the pairwise alignments in A: the differences between a pair of angles in B is the same as the pairwise angle in A.
+(C) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
+Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \SI), so that close data points are connected by darker (``stronger'') edges.
+(D) The data in C, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure.
+The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
+%\label{fig:schematics}
+\customlabel{fig:schematics}{2}
+\customlabel{subfig:synch1}{\ref{fig:schematics}A}
+\customlabel{subfig:synch2}{\ref{fig:schematics}B}
+\customlabel{subfig:dmaps1}{\ref{fig:schematics}C}
+\customlabel{subfig:dmaps2}{\ref{fig:schematics}D}
+\end{figure}
+
 Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
@@
 %
-In general, this requires no template function \cite{ahuja2007template} or image landmarks \cite{ian1998statistical}.
+In general, this requires no template function \citep{ahuja2007template} or image landmarks \citep{ian1998statistical}.
 %
@@
 %
-However, the algorithm can successfully register data sets even when many of the pairwise measurements are inaccurate \cite{singer2011angular}.
+However, the algorithm can successfully register data sets even when many of the pairwise measurements are inaccurate \citep{singer2011angular}.
 
@@
 %
-Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parameterization of data that lies on a low-dimensional manifold in high-dimensional space.
+Such a manifold can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parameterization of data that lies on a low-dimensional manifold in high-dimensional space.
 %
@@
 
-\begin{figure}[t]
-\includegraphics{fig2}
-\caption{Schematic illustrating angular synchronization and diffusion maps.
-{\xfigtextfontit (A)} Set of vectors, each in a different orientation.
-The pairwise alignment angles are indicated.
-{\xfigtextfontit (B)} The vectors from {\xfigtextfontit A}, each rotated about their midpoint so that the set is globally aligned.
-Note that the chosen rotation angles are consistent with the pairwise alignments in {\xfigtextfontit A}: the differences between a pair of angles in {\xfigtextfontit B} is the same as the pairwise angle in {\xfigtextfontit A}.
-{\xfigtextfontit (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
-Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\xfigtextfontit \SI}), so that close data points are connected by darker (``stronger'') edges.
-{\xfigtextfontit (D)} The data in {\xfigtextfontit C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure.
-The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
-%\label{fig:schematics}
-\customlabel{fig:schematics}{2}
-\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
-\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
-\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
-\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
-\end{figure}
-
 \begin{figure*}[t]
 \includegraphics{fig3}
-\caption{Method validation using live imaging of {\xfigtextfontit Drosophila} embryos.{\xfigtextfontit (A)} Selected images from a live imaging study of a {\xfigtextfontit Drosophilla} embryo during gastrulation.
+\caption{Method validation using live imaging of \textit{Drosophila} embryos.
+(A) Selected images from a live imaging study of a \textit{Drosophilla} embryo during gastrulation.
 Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
-{\xfigtextfontit (B)} Images from {\xfigtextfontit A} after registration and ordering using vector diffusion maps.
+(B) Images from A after registration and ordering using vector diffusion maps.
 The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.
-{\xfigtextfontit (C)} The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle {\xfigtextfontit (D)} The correlation between the recovered rank (using vector diffusion maps) and the true rank.
-{\xfigtextfontit (E)} The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies.
-{\xfigtextfontit (F)} The median rank correlation coefficient (over $200$ samples) as a function of the number of images in the data set.
+(C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle.
+(D) The correlation between the recovered rank (using vector diffusion maps) and the true rank.
+(E) The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies.
+(F) The median rank correlation coefficient (over $200$ samples) as a function of the number of images in the data set.
 }
 \customlabel{fig:movie}{3}
-\customlabel{subfig:movie_unregistered_unordered}{\ref{fig:movie}{\it A}}
-\customlabel{subfig:movie_registered_ordered}{\ref{fig:movie}{\it B}}
-\customlabel{subfig:movie_angle_corr}{\ref{fig:movie}{\it C}}
-\customlabel{subfig:movie_rank_corr}{\ref{fig:movie}{\it D}}
-\customlabel{subfig:movie_error_table}{\ref{fig:movie}{\it E}}
-\customlabel{subfig:movie_bootstrap}{\ref{fig:movie}{\it F}}
+\customlabel{subfig:movie_unregistered_unordered}{\ref{fig:movie}A}
+\customlabel{subfig:movie_registered_ordered}{\ref{fig:movie}{B}}
+\customlabel{subfig:movie_angle_corr}{\ref{fig:movie}{C}}
+\customlabel{subfig:movie_rank_corr}{\ref{fig:movie}{D}}
+\customlabel{subfig:movie_error_table}{\ref{fig:movie}{E}}
+\customlabel{subfig:movie_bootstrap}{\ref{fig:movie}{F}}
 \end{figure*}
 
-\subsection{Method validation using live imaging}
+\subsection*{Method validation using live imaging}
 
@@
 %
-Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
+Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \citep{leptin2005gastrulation}.
 %
@@
 %
-At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \cite{singer2011angular}.
+At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \citep{singer2011angular}.
 %
@@
 
-\subsection{Data sets with interembryo variability}
+\subsection*{Data sets with interembryo variability}
 
@@
 %
-Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that, in this context, specifies a subset of neuronal cells \cite{Lim2013kinetics}.
+Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that, in this context, specifies a subset of neuronal cells \citep{Lim2013kinetics}.
 
@@
 \includegraphics{fig4}
-\caption{Analyzing images of fixed {\xfigtextfontit Drosophila} embryos.
-{\xfigtextfontit (A)} Images of {\xfigtextfontit Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red).
+\caption{Analyzing images of fixed \textit{Drosophila} embryos.
+{(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red).
 Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation.
-{\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps.
+{(B)} Data from {\it A}, registered and ordered using vector diffusion maps.
 The expert rank for each image is indicated.
-{\xfigtextfontit (C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI).
-{\xfigtextfontit (D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert.
+{(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI).
+{(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert.
 The rank correlation coefficient is $0.97$.
@@
 \customlabel{fig:data2}{4}
-\customlabel{subfig:fixed_images_unregistered_unordered}{\ref{fig:data2}{\it A}}
-\customlabel{subfig:fixed_images_registered_ordered}{\ref{fig:data2}{\it B}}
-\customlabel{subfig:fixed_images_average_trajectory}{\ref{fig:data2}{\it C}}
-\customlabel{subfig:fixed_images_rank_corr}{\ref{fig:data2}{\it D}}
+\customlabel{subfig:fixed_images_unregistered_unordered}{\ref{fig:data2}{A}}
+\customlabel{subfig:fixed_images_registered_ordered}{\ref{fig:data2}{B}}
+\customlabel{subfig:fixed_images_average_trajectory}{\ref{fig:data2}{C}}
+\customlabel{subfig:fixed_images_rank_corr}{\ref{fig:data2}{D}}
 \end{figure*}
 
-\fig~\ref{subfig:fixed_images_registered_ordered} shows the selected images in \fig~\ref{subfig:fixed_images_unregistered_unordered}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
+\fig~\ref{subfig:fixed_images_registered_ordered} shows the selected images in \fig~\ref{subfig:fixed_images_unregistered_unordered}, now registered and ordered using vector diffusion maps \citep{singer2012vector}.
 %
@@
 
-\section{Discussion}
+\section*{Discussion}
 
@@
 %
-Temporal ordering of large-scale data which did not require registration was done in the context of molecular profiling studies, in which data are vectors describing the expression levels of different mRNA \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
+Temporal ordering of large-scale data which did not require registration was done in the context of molecular profiling studies, in which data are vectors describing the expression levels of different mRNA \citep{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
 %
-At the same time, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \cite{yuan2014automated, surkova2008characterization}.
+At the same time, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \citep{yuan2014automated, surkova2008characterization}.
 %
-%The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
+%The task of image registration has been widely studied \citep{zitova2003image}, for applications such as face recognition \citep{rowley1998rotation}, medical image registration \citep{hajnal2010medical}, and texture classification \citep{greenspan1994rotation}.
 %
-In contrast to most of the existing registration approaches which rely on the knowledge of appropriate landmarks in the images \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them relevant for a wide variety of applications.
+In contrast to most of the existing registration approaches which rely on the knowledge of appropriate landmarks in the images \citep{ian1998statistical} (such as the eyes in face recognition applications \citep{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them relevant for a wide variety of applications.
 %
@@
 
-Angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy images \cite{singer2012vector, zhao2014rotationally, singer2011viewing}.
+Angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy images \citep{singer2012vector, zhao2014rotationally, singer2011viewing}.
 %
@@
 %
-Another direction for future work is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \cite{krzic2012multiview, ichikawa2014live, rubel2010coupling}.
+Another direction for future work is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \citep{krzic2012multiview, ichikawa2014live, rubel2010coupling}.
 %
@@
 
-\begin{materials}
+\section*{Materials and Methods}
 
-\section{Experiments}
+\subsection*{Experiments}
 %
@@
 
-\section{Microscopy}
+\subsection*{Microscopy}
 %
@@
 %
-End-on imaging was performed by using the microfluidics device described previously \cite{chung2010microfluidic}.
+End-on imaging was performed by using the microfluidics device described previously \citep{chung2010microfluidic}.
 %
@@
 
-\section{Image preprocessing}
+\subsection*{Image preprocessing}
 %
@@
 
-\section{Software}
+\subsection*{Software}
 %
@@
 
-\end{materials}
+%\end{materials}
 
-%% Optional Appendix or Appendices
-%% \appendix Appendix text...
-%% or, for appendix with title, use square brackets:
-%% \appendix[Appendix Title]
+\subsubsection*{Acknowledgements}
 
-\begin{acknowledgments}
 The authors thank Adam Finkelstein,  Thomas Funkhouser, and John Storey for helpful discussions.
+
+\subsubsection*{Author contributions}
+Scientific approaches were developed by C.J.D, B.L., H.L., A.S., S.Y.S, and I.G.K.
 %
-C.J.D. was supported by the Department of Energy Computational Science Graduate Fellowship (CSGF), grant number DE-FG02-97ER25308.
+Experiments were performed by B.L.
+%
+Data analysis was performed by C.J.D. and B.L.
+%
+The manuscript was prepared and edited by C.J.D, B.L., H.L., A.S., S.Y.S, and I.G.K.
+
+\subsubsection*{Funding}
+C.J.D.
+was supported by the Department of Energy Computational Science Graduate Fellowship (CSGF), grant number DE-FG02-97ER25308, and the National Science Foundation Graduate Research Fellowship, Grant No.
+DGE 1148900.
 %
@@
 %
-T.J.L. and H.L. were supported by the National Science Foundation Grant Emerging Frontiers in Research and Innovation (EFRI) 1136913.
+H.L. was supported by the National Science Foundation Grant Emerging Frontiers in Research and Innovation (EFRI) 1136913.
 %
@@
 I.G.K. was supported by the National Science Foundation (CS\&E program).
-\end{acknowledgments}
 
-%% PNAS does not support submission of supporting .tex files such as BibTeX.
-%% Instead all references must be included in the article .tex document.
-%% If you currently use BibTeX, your bibliography is formed because the
-%% command \verb+\bibliography{}+ brings the <filename>.bbl file into your
-%% .tex document. To conform to PNAS requirements, copy the reference listings
-%% from your .bbl file and add them to the article .tex file, using the
-%% bibliography environment described above.
-
-%%  Contact pnas@nas.edu if you need assistance with your
-%%  bibliography.
-
-% Sample bibliography item in PNAS format:
-%% \bibitem{in-text reference} comma-separated author names up to 5,
-%% for more than 5 authors use first author last name et al. (year published)
-%% article title  {\it Journal Name} volume #: start page-end page.
-%% ie,
-% \bibitem{Neuhaus} Neuhaus J-M, Sitcher L, Meins F, Jr, Boller T (1991)
-% A short C-terminal sequence is necessary and sufficient for the
-% targeting of chitinases to the plant vacuole.
-% {\it Proc Natl Acad Sci USA} 88:10362-10366.
-
-%% Enter the largest bibliography number in the facing curly brackets
-%% following \begin{thebibliography}
-
-\bibliographystyle{pnas2}
-\bibliography{../../references/references,../image_analysis_paper/background_reading/references}
-
-\end{article}
+\bibliographystyle{development}
+\bibliography{references}
 

### 2015-01-25 image_analysis_paper_development2/SI_text.tex
added development paper text

--- 2014-09-01 image_analysis_paper2/SI_text.tex
@@
-\documentclass{pnastwo}
-%\documentclass[10pt,twocolumn]{article}
+\documentclass[10pt,twocolumn]{article}
 
-%\usepackage[margin=1in]{geometry}
-%\usepackage{graphicx}
-\graphicspath{ {paper_figures2/} }
+\usepackage[margin=0.75in]{geometry}
+\usepackage{graphicx}
 
-\usepackage{tikz}
-
-\usepackage{float}
 \usepackage{amssymb,amsfonts,amsmath}
@@
 
+\renewcommand{\figurename}{Fig.}
 \renewcommand\thefigure{S\arabic{figure}}
@@
 
+\usepackage{natbib}
+
+\title{Supplemental Information}
+\date{}
+
 \begin{document}
 
-\begin{article}
+\maketitle
 
-\section{Materials and Methods}
+\section*{Materials and Methods}
 
-\subsection{Image parameters}
+\subsection*{Image parameters}
 
@@
 %
-The Canny method \cite{canny1986computational} is used to detect the edges of the nuclear signal in each image (using the \texttt{edge} function in MATLAB\textsuperscript{\textregistered}), and images are cropped to these edges to remove any size effects.
+The Canny method \citep{canny1986computational} is used to detect the edges of the nuclear signal in each image (using the \texttt{edge} function in MATLAB\textsuperscript{\textregistered}), and images are cropped to these edges to remove any size effects.
 %
@@
 
-\subsection{Algorithms}
+\subsection*{Algorithms}
 
@@
 
-\subsubsection{Angular synchronization \cite{singer2011angular}}
+\subsubsection*{Angular synchronization \citep{singer2011angular}}
 
@@
 Alternatively, for continuous signals, an optimization algorithm
-can be used \cite{ahuja2007template}.
+can be used \citep{ahuja2007template}.
 
@@
 We adjust the sign of $\phi_1$ so that $det(R_{i, est}) = +1$, ensuring proper rotations 
-(note that systematically incorporating improper rotations is also possible \cite{goemans1995improved, bandeira2013cheeger}).
+(note that systematically incorporating improper rotations is also possible \citep{goemans1995improved, bandeira2013cheeger}).
 %
@@
 
-\subsubsection{Diffusion maps \cite{coifman2005geometric}}
+\subsubsection*{Diffusion maps \citep{coifman2005geometric}}
 
@@
 %
-$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon^2$ to be $1/10^{th}$ of the median of the squared pairwise distances.
+$\epsilon$ can be chosen using several techniques (see, for example \citep{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon^2$ to be $1/10^{th}$ of the median of the squared pairwise distances.
 
-To find the coordinate $y$, we want solve the following optimization problem \cite{Belkin2003}
+To find the coordinate $y$, we want solve the following optimization problem \citep{Belkin2003}
 \begin{equation} \label{eq:dmaps_opt_problem}
@@
 
-\subsubsection{Vector diffusion maps\cite{singer2012vector}}
+\subsubsection*{Vector diffusion maps\citep{singer2012vector}}
 
@@
 
-\subsection{Registering images} \label{subsec:trans_rot_register}
+\subsection*{Registering images} \label{subsec:trans_rot_register}
 
@@
 
-\subsection{The eigenvalue spectrum}
+\subsection*{The eigenvalue spectrum}
 
@@
 %
-However, some embedding coordinates which appear meaningful according to the eigenvalues may be harmonics of previous coordinates \cite{ferguson2010systematic}, 
+However, some embedding coordinates which appear meaningful according to the eigenvalues may be harmonics of previous coordinates \citep{ferguson2010systematic}, 
 and one must visually check for correlations among potential embedding coordinates before deducing the true dimensionality of the data.
@@
 
-\subsection{Smooth trajectories from registered and ordered images}
+\subsection*{Smooth trajectories from registered and ordered images}
 
@@
 %
-See \cite{kemelmacher2011exploring} for a more detailed discussion.
+See \citep{kemelmacher2011exploring} for a more detailed discussion.
 
-\bibliographystyle{pnas2}
-\bibliography{../image_analysis_paper/background_reading/references,../../references/references}
+\bibliographystyle{development}
+\bibliography{references}
 
-\end{article}
+\newpage
 

### 2015-01-25 image_analysis_paper_development2/manuscript.tex
added development paper text

--- 2014-11-03 image_analysis_paper_development/manuscript.tex
@@
 
-\begin{figure}[t]
-\includegraphics{fig1}
-\caption{Caricature illustrating the tasks of image registration and temporal ordering.
+%\begin{figure}[t]
+%\includegraphics{fig1}
+%\caption{Caricature illustrating the tasks of image registration and temporal ordering.
 (A) Images of ``samples'', each in a different orientation and a different stage of development.
@@
 For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
-%\label{fig:fish}
-\customlabel{fig:fish}{1}
-\customlabel{subfig:fish_unordered}{\ref{fig:fish}A}
-\customlabel{subfig:fish_ordered}{\ref{fig:fish}B}
-\end{figure}
+%%\label{fig:fish}
+%\customlabel{fig:fish}{1}
+%\customlabel{subfig:fish_unordered}{\ref{fig:fish}A}
+%\customlabel{subfig:fish_ordered}{\ref{fig:fish}B}
+%\end{figure}
 
@@
 
-\begin{figure}[t]
-\includegraphics{fig2}
-\caption{Schematic illustrating angular synchronization and diffusion maps.
+%\begin{figure}[t]
+%\includegraphics{fig2}
+%\caption{Schematic illustrating angular synchronization and diffusion maps.
 (A) Set of vectors, each in a different orientation.
@@
 The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
-%\label{fig:schematics}
-\customlabel{fig:schematics}{2}
-\customlabel{subfig:synch1}{\ref{fig:schematics}A}
-\customlabel{subfig:synch2}{\ref{fig:schematics}B}
-\customlabel{subfig:dmaps1}{\ref{fig:schematics}C}
-\customlabel{subfig:dmaps2}{\ref{fig:schematics}D}
-\end{figure}
+%%\label{fig:schematics}
+%\customlabel{fig:schematics}{2}
+%\customlabel{subfig:synch1}{\ref{fig:schematics}A}
+%\customlabel{subfig:synch2}{\ref{fig:schematics}B}
+%\customlabel{subfig:dmaps1}{\ref{fig:schematics}C}
+%\customlabel{subfig:dmaps2}{\ref{fig:schematics}D}
+%\end{figure}
 
@@
 
-\begin{figure*}[t]
-\includegraphics{fig3}
-\caption{Method validation using live imaging of \textit{Drosophila} embryos.
+%\begin{figure*}[t]
+%\includegraphics{fig3}
+%\caption{Method validation using live imaging of \textit{Drosophila} embryos.
 (A) Selected images from a live imaging study of a \textit{Drosophilla} embryo during gastrulation.
@@
 }
-\customlabel{fig:movie}{3}
-\customlabel{subfig:movie_unregistered_unordered}{\ref{fig:movie}A}
-\customlabel{subfig:movie_registered_ordered}{\ref{fig:movie}{B}}
-\customlabel{subfig:movie_angle_corr}{\ref{fig:movie}{C}}
-\customlabel{subfig:movie_rank_corr}{\ref{fig:movie}{D}}
-\customlabel{subfig:movie_error_table}{\ref{fig:movie}{E}}
-\customlabel{subfig:movie_bootstrap}{\ref{fig:movie}{F}}
-\end{figure*}
+%\customlabel{fig:movie}{3}
+%\customlabel{subfig:movie_unregistered_unordered}{\ref{fig:movie}A}
+%\customlabel{subfig:movie_registered_ordered}{\ref{fig:movie}{B}}
+%\customlabel{subfig:movie_angle_corr}{\ref{fig:movie}{C}}
+%\customlabel{subfig:movie_rank_corr}{\ref{fig:movie}{D}}
+%\customlabel{subfig:movie_error_table}{\ref{fig:movie}{E}}
+%\customlabel{subfig:movie_bootstrap}{\ref{fig:movie}{F}}
+%\end{figure*}
 
 \subsection*{Method validation using live imaging}
+
+\subsubsection*{{\em Drosophila} gastrulation}
+
+\subsubsection*{Zebrafish gastrulation}
 
@@
 
-\subsection*{Data sets with interembryo variability}
+\subsection*{Data sets with intersample variability}
+
+\subsubsection*{Fixed {\em Drosophila} images}
+
+\subsubsection*{Wing disc data}
 
@@
 
-\begin{figure*}[t]
-\includegraphics{fig4}
-\caption{Analyzing images of fixed \textit{Drosophila} embryos.
+%\begin{figure*}[t]
+%\includegraphics{fig4}
+%\caption{Analyzing images of fixed \textit{Drosophila} embryos.
 {(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red).
@@
 }
-\customlabel{fig:data2}{4}
-\customlabel{subfig:fixed_images_unregistered_unordered}{\ref{fig:data2}{A}}
-\customlabel{subfig:fixed_images_registered_ordered}{\ref{fig:data2}{B}}
-\customlabel{subfig:fixed_images_average_trajectory}{\ref{fig:data2}{C}}
-\customlabel{subfig:fixed_images_rank_corr}{\ref{fig:data2}{D}}
-\end{figure*}
+%\customlabel{fig:data2}{4}
+%\customlabel{subfig:fixed_images_unregistered_unordered}{\ref{fig:data2}{A}}
+%\customlabel{subfig:fixed_images_registered_ordered}{\ref{fig:data2}{B}}
+%\customlabel{subfig:fixed_images_average_trajectory}{\ref{fig:data2}{C}}
+%\customlabel{subfig:fixed_images_rank_corr}{\ref{fig:data2}{D}}
+%\end{figure*}
 

### 2015-01-25 image_analysis_paper_development2/manuscript.tex
updated tex

@@
 \makeatother
+
+\graphicspath{{paper_figures/}}
 
@@
 Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.
+
+\begin{figure*}[t]
+\raisebox{0.75cm}{\textsf{A}}
+\hfill
+\includegraphics[width=17.5cm]{drosophila_live_imaging_scrambled}
+
+\raisebox{0.75cm}{\textsf{B}}
+\hfill
+\includegraphics[width=17.5cm]{drosophila_live_imaging_ordered}
+
+\raisebox{4.25cm}{\textsf{C}}
+\includegraphics[width=5cm]{drosophila_live_imaging_angle_corr}
+%
+\hfill
+%
+\raisebox{4.25cm}{\textsf{D}}
+\includegraphics[width=5cm]{drosophila_live_imaging_rank_corr}
+%
+\hfill
+%
+\raisebox{4.25cm}{\textsf{E}}
+%
+\raisebox{2.75cm}{
+{ \textsf{ 
+\begin{tabular}{ |p{1cm} | p{1.25cm}|  p{1.5cm}| }
+\hline
+Movie Index & Average Angle Error & Rank Correlation Coefficient \\
+\hline
+1 & 8.37$^{\circ}$ & 0.9989 \\ 
+2 & 4.98$^{\circ}$ & 0.9994 \\ 
+3 & 6.47$^{\circ}$ & 0.9795 \\ 
+4 & 7.03$^{\circ}$ & 0.9953 \\ 
+5 & 3.04$^{\circ}$ & 0.9959 \\ 
+ \hline
+\end{tabular}}}}
+
+\end{figure*}
+
+\begin{figure*}[t]
+\begin{minipage}[b]{14cm}
+\raisebox{0.75cm}{\textsf{A}}
+\includegraphics[width=13.75cm]{zebrafish_scrambled}
+
+\raisebox{0.75cm}{\textsf{B}}
+\includegraphics[width=13.75cm]{zebrafish_ordered}
+\end{minipage}
+%
+\hfill
+%
+\raisebox{2.5cm}{\textsf{C}}
+\includegraphics[width=3.25cm]{zebrafish_rank_corr}
+
+\end{figure*}
 

### 2015-01-26 image_analysis_paper_development2/manuscript.tex
updated figures for paper

@@
 \end{tabular}}}}
-
+\caption{Method validation using live imaging of \textit{Drosophila} embryos.
+(A) Selected images from a live imaging study of a \textit{Drosophilla} embryo during gastrulation.
+Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
+(B) Images from A after registration and ordering using vector diffusion maps.
+The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.
+(C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle.
+(D) The correlation between the recovered rank (using vector diffusion maps) and the true rank.
+(E) The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies.
+}
 \end{figure*}
@@
 
+\end{figure*}
+
+\begin{figure*}[t]
+\begin{minipage}[b]{12cm}
+\raisebox{0.5cm}{\textsf{A}}
+\includegraphics[width=12.25cm]{drosophila_fixed_images_scrambled}
+
+\raisebox{0.5cm}{\textsf{B}}
+\includegraphics[width=12.25cm]{drosophila_fixed_images_ordered}
+
+\raisebox{0.5cm}{\textsf{C}}
+\includegraphics[width=12.25cm]{drosophila_fixed_images_average}
+\end{minipage}
+%
+\hfill
+%
+\raisebox{3.25cm}{\textsf{D}}
+\includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}
+
+\end{figure*}
+
+\begin{figure*}
+\includegraphics[width=\textwidth]{wing_disc_ordered}
 \end{figure*}

### 2015-01-26 image_analysis_paper_development2/paper_figures/make_fig_drosophila_fixed_images.m

@@ -57,17 +57,31 @@ end
 
 idx_to_plot = randsample(nimages, nsubimages);
 [~, I] = sort(embed_coord(idx_to_plot));
+
 make_fig(17, 17/nsubimages);
 for j=1:nsubimages
     make_subplot(nsubimages, 1, 0.01, j);
     imshow(make_gray_nuclei(images(:,:,:,idx_to_plot(j))));
 end
+saveas(gcf, 'drosophila_fixed_images_scrambled.pdf');
+
+
 make_fig(17, 17/nsubimages);
 for j=1:nsubimages
     make_subplot(nsubimages, 1, 0.01, j);
     imshow(make_gray_nuclei(imrotate(images_registered(:,:,:,idx_to_plot(I(j))), -5, 'crop')));
     text(npixels/2, npixels/2, sprintf('%d', time(idx_to_plot(I(j)))),'color',0.95*ones(1,3),'HorizontalAlignment','center','VerticalAlignment','middle', 'fontsize', 6)
 end
+saveas(gcf, 'drosophila_fixed_images_ordered.pdf');
+
+make_fig(17, 17/nsubimages);
+for j=1:nsubimages
+    make_subplot(nsubimages, 1, 0.01, j);
+    im_tmp = uint8(mean(images_analyzed(:,:,:,(j-1)*nimages/nsubimages+1:j*nimages/nsubimages), 4));
+    imshow(make_gray_nuclei(imrotate(im_tmp, -5, 'crop')));
+end
+saveas(gcf, 'drosophila_fixed_images_average.pdf');
+
 
 make_fig(4,4);
 plot(tiedrank(time), tiedrank(embed_coord),'.k')
@@ -77,6 +91,8 @@ set(gca, 'xtick', [0 50 100])
 set(gca, 'ytick', [0 50 100])
 axis([0 125 0 125])
 axis square
+saveas(gcf, 'drosophila_fixed_images_rank_corr.pdf');
+
 
 
 

### 2015-01-26 image_analysis_paper_development2/manuscript.tex
updated paper

@@
 
-%\begin{figure}[t]
-%\includegraphics{fig1}
-%\caption{Caricature illustrating the tasks of image registration and temporal ordering.
+\begin{figure}[t]
+\includegraphics{fig1}
+\caption{Caricature illustrating the tasks of image registration and temporal ordering.
 (A) Images of ``samples'', each in a different orientation and a different stage of development.
@@
 For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
-%%\label{fig:fish}
-%\customlabel{fig:fish}{1}
-%\customlabel{subfig:fish_unordered}{\ref{fig:fish}A}
-%\customlabel{subfig:fish_ordered}{\ref{fig:fish}B}
-%\end{figure}
+%\label{fig:fish}
+\customlabel{fig:fish}{1}
+\customlabel{subfig:fish_unordered}{\ref{fig:fish}A}
+\customlabel{subfig:fish_ordered}{\ref{fig:fish}B}
+\end{figure}
 
@@
 
-%\begin{figure}[t]
-%\includegraphics{fig2}
-%\caption{Schematic illustrating angular synchronization and diffusion maps.
+\begin{figure}[t]
+\includegraphics{fig2}
+\caption{Schematic illustrating angular synchronization and diffusion maps.
 (A) Set of vectors, each in a different orientation.
@@
 The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
-%%\label{fig:schematics}
-%\customlabel{fig:schematics}{2}
-%\customlabel{subfig:synch1}{\ref{fig:schematics}A}
-%\customlabel{subfig:synch2}{\ref{fig:schematics}B}
-%\customlabel{subfig:dmaps1}{\ref{fig:schematics}C}
-%\customlabel{subfig:dmaps2}{\ref{fig:schematics}D}
-%\end{figure}
+%\label{fig:schematics}
+\customlabel{fig:schematics}{2}
+\customlabel{subfig:synch1}{\ref{fig:schematics}A}
+\customlabel{subfig:synch2}{\ref{fig:schematics}B}
+\customlabel{subfig:dmaps1}{\ref{fig:schematics}C}
+\customlabel{subfig:dmaps2}{\ref{fig:schematics}D}
+\end{figure}
 
@@
 
+\subsection*{Method validation using live imaging}
+
+\subsubsection*{{\em Drosophila} gastrulation}
+
 \begin{figure*}[t]
@@
 { \textsf{ 
-\begin{tabular}{ |p{1cm} | p{1.25cm}|  p{1.5cm}| }
+\begin{tabular}{ |p{1.25cm} | p{1.5cm}|  p{1.75cm}| }
 \hline
@@
 \end{figure*}
-
-\begin{figure*}[t]
-\begin{minipage}[b]{14cm}
-\raisebox{0.75cm}{\textsf{A}}
-\includegraphics[width=13.75cm]{zebrafish_scrambled}
-
-\raisebox{0.75cm}{\textsf{B}}
-\includegraphics[width=13.75cm]{zebrafish_ordered}
-\end{minipage}
-%
-\hfill
-%
-\raisebox{2.5cm}{\textsf{C}}
-\includegraphics[width=3.25cm]{zebrafish_rank_corr}
-
-\end{figure*}
-
-\begin{figure*}[t]
-\begin{minipage}[b]{12cm}
-\raisebox{0.5cm}{\textsf{A}}
-\includegraphics[width=12.25cm]{drosophila_fixed_images_scrambled}
-
-\raisebox{0.5cm}{\textsf{B}}
-\includegraphics[width=12.25cm]{drosophila_fixed_images_ordered}
-
-\raisebox{0.5cm}{\textsf{C}}
-\includegraphics[width=12.25cm]{drosophila_fixed_images_average}
-\end{minipage}
-%
-\hfill
-%
-\raisebox{3.25cm}{\textsf{D}}
-\includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}
-
-\end{figure*}
-
-\begin{figure*}
-\includegraphics[width=\textwidth]{wing_disc_ordered}
-\end{figure*}
-
-%\begin{figure*}[t]
-%\includegraphics{fig3}
-%\caption{Method validation using live imaging of \textit{Drosophila} embryos.
-(A) Selected images from a live imaging study of a \textit{Drosophilla} embryo during gastrulation.
-Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
-(B) Images from A after registration and ordering using vector diffusion maps.
-The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.
-(C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle.
-(D) The correlation between the recovered rank (using vector diffusion maps) and the true rank.
-(E) The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies.
-(F) The median rank correlation coefficient (over $200$ samples) as a function of the number of images in the data set.
-}
-%\customlabel{fig:movie}{3}
-%\customlabel{subfig:movie_unregistered_unordered}{\ref{fig:movie}A}
-%\customlabel{subfig:movie_registered_ordered}{\ref{fig:movie}{B}}
-%\customlabel{subfig:movie_angle_corr}{\ref{fig:movie}{C}}
-%\customlabel{subfig:movie_rank_corr}{\ref{fig:movie}{D}}
-%\customlabel{subfig:movie_error_table}{\ref{fig:movie}{E}}
-%\customlabel{subfig:movie_bootstrap}{\ref{fig:movie}{F}}
-%\end{figure*}
-
-\subsection*{Method validation using live imaging}
-
-\subsubsection*{{\em Drosophila} gastrulation}
-
-\subsubsection*{Zebrafish gastrulation}
 
@@
 %
-Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
-%
-To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
-%
-Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered. 
-%
-\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
-%
-As expected, the accuracy in the recovered ordering increases with the number of images. 
-% 
-At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
-%
+%Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
+%%
+%To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
+%%
+%Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered. 
+%%
+%\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
+%%
+%As expected, the accuracy in the recovered ordering increases with the number of images. 
+%% 
+%At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
+%%
 Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignments and the errors in the recovered angles were consistently less than $10^{\circ}$.
@@
 In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.
+
+\subsubsection*{Zebrafish gastrulation}
+
+\begin{figure*}[t]
+\begin{minipage}[c]{13.75cm}
+\vspace{0cm}
+\raisebox{0.5cm}{\textsf{A}}
+\includegraphics[width=13.5cm]{zebrafish_scrambled}
+
+\raisebox{0.5cm}{\textsf{B}}
+\includegraphics[width=13.5cm]{zebrafish_ordered}
+\end{minipage}
+%
+\hfill
+%
+\begin{minipage}[c]{3.5cm}
+
+\vspace{0cm}
+\raisebox{2.5cm}{\textsf{C}}
+%
+\includegraphics[width=3.15cm]{zebrafish_rank_corr}
+\end{minipage}
+\caption{Method validation using live imaging of a zebrafish embryo. Registered and ordered zebrafish data.}
+\end{figure*}
+
+To illustrate the broad applicability of our approach, we also applied our algorithm to a time-lapse movie of zebrafish embryogenesis. 
+%
+The original movie is available at \texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov} \cite{karlstrom1996flipbook}. 
+%
+We analyzed 120 frames from this movie which span the epiboly stage.
+
+As in the live imaging {\em Drosophila} example, the frames were randomly rotated and shuffled. 
+%
+We then used vector diffusion maps to register and order the frames. 
+%
+The results are shown in \fig~\ref{fig:zebrafish}. 
+%
+The recovered rotations and order are consistent with the expected developmental dynamics. 
+%
+Quantitatively, the rank correlation coefficient for this data set is $0.9935$, and the average error in the recovered angle is $4.23^\circ$.
 
@@
 
-\subsubsection*{Wing disc data}
+\begin{figure*}[t]
+\begin{minipage}[b]{12cm}
+\raisebox{0.5cm}{\textsf{A}}
+\includegraphics[width=12.25cm]{drosophila_fixed_images_scrambled}
+
+\raisebox{0.5cm}{\textsf{B}}
+\includegraphics[width=12.25cm]{drosophila_fixed_images_ordered}
+
+\raisebox{0.5cm}{\textsf{C}}
+\includegraphics[width=12.25cm]{drosophila_fixed_images_average}
+\end{minipage}
+%
+\hfill
+%
+\raisebox{3.25cm}{\textsf{D}}
+\includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}
+%
+\caption{Analyzing images of fixed \textit{Drosophila} embryos.
+{(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red).
+Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation.
+{(B)} Data from {\it A}, registered and ordered using vector diffusion maps.
+The expert rank for each image is indicated.
+{(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI).
+{(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert.
+The rank correlation coefficient is $0.97$.
+}
+\end{figure*}
 
@@
 %
-The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.81, indicating that the algorithm can accurately recover the temporal order even under noisy conditions.
+The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.77, indicating that the algorithm can accurately recover the temporal order even under noisy conditions.
 
@@
 Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that, in this context, specifies a subset of neuronal cells \citep{Lim2013kinetics}.
-
-%\begin{figure*}[t]
-%\includegraphics{fig4}
-%\caption{Analyzing images of fixed \textit{Drosophila} embryos.
-{(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red).
-Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation.
-{(B)} Data from {\it A}, registered and ordered using vector diffusion maps.
-The expert rank for each image is indicated.
-{(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI).
-{(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert.
-The rank correlation coefficient is $0.97$.
-}
-%\customlabel{fig:data2}{4}
-%\customlabel{subfig:fixed_images_unregistered_unordered}{\ref{fig:data2}{A}}
-%\customlabel{subfig:fixed_images_registered_ordered}{\ref{fig:data2}{B}}
-%\customlabel{subfig:fixed_images_average_trajectory}{\ref{fig:data2}{C}}
-%\customlabel{subfig:fixed_images_rank_corr}{\ref{fig:data2}{D}}
-%\end{figure*}
 
@@
 Overall, the tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, and so this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.
+
+\subsubsection*{Wing disc data}
+
+\begin{figure*}[t]
+\begin{minipage}[c]{9cm}
+\vspace{0cm}
+
+\raisebox{6.75cm}{\textsf{A}}
+\includegraphics[width=8cm]{wing_disc_example}
+\end{minipage}
+%
+\begin{minipage}[c]{8cm}
+\caption{Registered and ordered three-dimensional wing-disc data}
+\end{minipage}
+
+\raisebox{0.7cm}{\textsf{B}}
+\includegraphics[width=\textwidth]{wing_disc_ordered}
+
+\end{figure*}
 

### 2015-01-26 image_analysis_paper_development2/paper_figures/make_fig_drosophila_live_imaging.m

@@ -100,17 +100,6 @@ for i=1:nexpt
         end
         saveas(gcf, 'drosophila_live_imaging_ordered.pdf');
         
-        make_fig(4,4);
-        plot(tiedrank(time), tiedrank(embed_coord),'.k')
-        xlabel('true rank')
-        ylabel('recovered rank')
-        set(gca, 'xtick', [0 20 40])
-        set(gca, 'ytick', [0 20 40])
-        axis([0 50 0 50])
-        axis square
-        saveas(gcf, 'drosophila_live_imaging_rank_corr.pdf');
-        
-        
         make_fig(4, 4);
         plot(angles, angles_opt,'.k')
         xlabel('true angle')
@@ -123,13 +112,94 @@ for i=1:nexpt
         axis square
         saveas(gcf, 'drosophila_live_imaging_angle_corr.pdf');
         
-        return
+        make_fig(4,4);
+        plot(tiedrank(time), tiedrank(embed_coord),'.k')
+        xlabel('true rank')
+        ylabel('recovered rank')
+        set(gca, 'xtick', [0 20 40])
+        set(gca, 'ytick', [0 20 40])
+        set(gca, 'xticklabel', {'0 '; '  20 '; '  40 '});
+        set(gca, 'yticklabel', {'0 '; '  20 '; '  40 '});
+        axis([0 50 0 50])
+        axis square
+        saveas(gcf, 'drosophila_live_imaging_rank_corr.pdf');
+        
+        
+        
+        
+        
         
     end
 end
 
 rank_corr
 theta_err
+
+%%
+
+channel_weight = 1;
+channel_blur = 0.05;
+channel_normalize = 1;
+channel_mean_center = 1;
+resize_image = true;
+
+eps_scale = 0.5;
+
+% images_raw_all = zeros(npixels, npixels, nimages, nexpt);
+images_all = zeros(npixels, npixels, nimages, nexpt);
+
+for i=1:nexpt
+    image_dir = sprintf('../code_for_distribution/drosophila_live_imaging/expt%02d', i);
+    
+    [images_raw_tmp, nchannels] = read_images(image_dir, image_name, image_ext, stack_name, nimages, nstack, npixels, dim);
+    
+    fid = fopen(sprintf('%s/times.txt', image_dir), 'r');
+    time = fscanf(fid, '%f');
+    fclose(fid);
+    
+    [~, idx] = sort(time);
+    images_raw_tmp = images_raw_tmp(:,:,idx);
+    
[…]

### 2015-01-26 image_analysis_paper_development2/paper_figures/make_fig_wing_disc.m

@@ -53,46 +53,52 @@ plot_images(immultiply(permute(images_analyzed, [4 1 3 2 5]), 1.5), dim)
 
 %%
 
+plot_wing_disc_projections(images_registered(:,:,:,:,2), []);
+annotation('arrow', [0.1 0.2], [0.1 0.1], 'color', 0.95*ones(1,3))
+annotation('arrow', [0.1 0.1], [0.1 0.2], 'color', 0.95*ones(1,3))
+annotation('textbox', [0.15 0.05 0.1 0.1], 'string','x','color', 0.95*ones(1,3), 'edgecolor','none', 'verticalalignment','middle','horizontalalignment','left')
+annotation('textbox', [0.05 0.15 0.1 0.1], 'string','y','color', 0.95*ones(1,3), 'edgecolor','none', 'verticalalignment','bottom','horizontalalignment','center')
+
+annotation('arrow', [0.85 0.9], [0.1 0.1], 'color', 0.95*ones(1,3))
+annotation('arrow', [0.85 0.85], [0.1 0.15], 'color', 0.95*ones(1,3))
+annotation('textbox', [0.87 0.05 0.1 0.1], 'string','z','color', 0.95*ones(1,3), 'edgecolor','none', 'verticalalignment','middle','horizontalalignment','left')
+annotation('textbox', [0.82 0.11 0.1 0.1], 'string','y','color', 0.95*ones(1,3), 'edgecolor','none', 'verticalalignment','bottom','horizontalalignment','center')
+
+annotation('arrow',  [0.1 0.1], [0.85 0.89],'color', 0.95*ones(1,3))
+annotation('arrow',  [0.1 0.15],[0.85 0.85], 'color', 0.95*ones(1,3))
+annotation('textbox', [0.05 0.84 0.1 0.1], 'string','z','color', 0.95*ones(1,3), 'edgecolor','none', 'verticalalignment','bottom','horizontalalignment','center')
+annotation('textbox', [0.12 0.8 0.1 0.1], 'string','x','color', 0.95*ones(1,3), 'edgecolor','none', 'verticalalignment','middle','horizontalalignment','left')
+
+saveas(gcf, 'wing_disc_example.pdf');
+
+%%
 fid = fopen(sprintf('%s/times.txt', image_dir), 'r');
 time = fscanf(fid, '%f');
 fclose(fid);
 
-
[…]

### 2015-01-26 image_analysis_paper_development2/paper_figures/timing_studies.m

@@ -1,0 +1,174 @@
+clear all
+close all
+
+%%
+
+% addpath ../code_for_distribution
+
+%%
+image_name = 'image';
+image_ext = 'tif';
+stack_name = '';
+nimages = 120;
+nstack = 0;
+% npixels = 100;
+dim = 2;
+
+channel_weight = 1;
+channel_blur = 0;
+channel_normalize = 0;
+channel_mean_center = 1;
+resize_image = false;
+
+nrot = 36;
+
+eps_scale = 0.25;
+
+image_dir = '../code_for_distribution/zebrafish';
+
+npixels_zebrafish = [50 100 150 200 300 400];
+cpu_times_npixels_zebrafish = zeros(size(npixels_zebrafish));
+
+for i=1:length(npixels_zebrafish)
+    
+    [images_raw, nchannels] = read_images(image_dir, image_name, image_ext, stack_name, nimages, nstack, npixels_zebrafish(i), dim);
+    
+    tic
+    images = apply_image_functions(images_raw, dim, channel_weight, channel_blur, channel_normalize, channel_mean_center, resize_image);
+    
+    [R, W] = compute_pairwise_alignments(images, nrot);
+    
+    % compute optimal rotations + embedding coordinates using vector diffusion maps
+    ncomps = 1;
+    [R_opt, embed_coord, D2] = vdm(R, W, eps_scale, ncomps);
+    
+    % register images using optimal rotations
+    images_registered = register_all_images(images, R_opt);
+    
+    images_analyzed = order_all_images(images_registered, embed_coord);
+    
+    cpu_times_npixels_zebrafish(i) = toc;
+end
+
+
+%%
+
+npixels = 100;
+nimages_zebrafish = [20 40 60 80 100 120];
+cpu_times_nimages_zebrafish = zeros(size(nimages_zebrafish));
+for i=1:length(nimages_zebrafish)
+    
+    [images_raw, nchannels] = read_images(image_dir, image_name, image_ext, stack_name, nimages_zebrafish(i), nstack, npixels, dim);
+    
+    tic
+    images = apply_image_functions(images_raw, dim, channel_weight, channel_blur, channel_normalize, channel_mean_center, resize_image);
+    
+    [R, W] = compute_pairwise_alignments(images, nrot);
+    
+    % compute optimal rotations + embedding coordinates using vector diffusion maps
+    ncomps = 1;
+    [R_opt, embed_coord, D2] = vdm(R, W, eps_scale, ncomps);
+    
[…]

### 2015-01-28 image_analysis_paper_development2/SI_text.tex
updated all files

@@
 \newpage
-
-\begin{figure}
-\includegraphics{figS1}
-\caption{(Left) A lateral view of a {\it Drosophila} embryo stained with DAPI (gray), dpERK (red), and Twi (green).
+%
+%\begin{figure}
+%\includegraphics{figS1}
+%\caption{(Left) A lateral view of a {\it Drosophila} embryo stained with DAPI (gray), dpERK (red), and Twi (green).
 The embryo is presented so that the anterior (A) side is to the left and the posterior (P) side is to the right.
@@
 }
-\label{fig:ap_dv}
-\end{figure}
-
-\begin{figure}
-\includegraphics{figS2}
-\caption{Synthetic data set used to illustrate the data processing algorithms.
+%\label{fig:ap_dv}
+%\end{figure}
+%
+%\begin{figure}
+%\includegraphics{figS2}
+%\caption{Synthetic data set used to illustrate the data processing algorithms.
 {\it (A)} One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom).
@@
 {\it (F)} The profiles in {\it C}, registered and temporally ordered in a single step using vector diffusion maps.}
-\label{fig:1d_demo}
-%\customlabel{fig:1d_demo}{3}
-\customlabel{subfig:1d_example}{\ref{fig:1d_demo}{\it A}}
-\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}{\it B}}
-\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}{\it C}}
-\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}{\it D}}
-\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}{\it E}}
-\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}{\it F}}
-\end{figure}
-
-\begin{figure*}
-\includegraphics{figS3}
-\caption{Eigenvalue spectra.
+%\label{fig:1d_demo}
+%%\customlabel{fig:1d_demo}{3}
+%\customlabel{subfig:1d_example}{\ref{fig:1d_demo}{\it A}}
+%\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}{\it B}}
+%\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}{\it C}}
+%\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}{\it D}}
+%\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}{\it E}}
+%\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}{\it F}}
+%\end{figure}
+%
+%\begin{figure*}
+%\includegraphics{figS3}
+%\caption{Eigenvalue spectra.
 {\it (A)} Eigenvalue spectrum for the live imaging data set presented in \fig~3.
@@
 }
-\end{figure*}
-
-\begin{figure*}
-\includegraphics{figS4}
-\caption{Live imaging data set.
+%\end{figure*}
+%
+%\begin{figure*}
+%\includegraphics{figS4}
+%\caption{Live imaging data set.
 {\it (A)} Live imaging data set used for the results in \fig~3.
@@
 }
-\end{figure*}
-
-\begin{figure*}
-\includegraphics{figS5}
-\caption{{\it (A-D)} Four additional live imaging data sets used for the analysis in \fig~3. }
-\end{figure*}
-
-\begin{figure*}	
-\includegraphics{figS6}
-\caption{Fixed embryo imaging data set.
+%\end{figure*}
+%
+%\begin{figure*}
+%\includegraphics{figS5}
+%\caption{{\it (A-D)} Four additional live imaging data sets used for the analysis in \fig~3. }
+%\end{figure*}
+%
+%\begin{figure*}	
+%\includegraphics{figS6}
+%\caption{Fixed embryo imaging data set.
 {\it (A)} Fixed images used for the results in \fig~4.

### 2015-01-28 image_analysis_paper_development2/manuscript.tex
updated all files

@@
 \usepackage{graphicx}
+
+\usepackage{epstopdf}
 
@@
 %
-To illustrate our approach, we register and order data from imaging studies of pattern formation and morphogenesis in  \textit{Drosophila} embryos.
+To illustrate our approach, we register and order data from imaging studies of pattern formation and morphogenesis in three different model systems in developmental biology.
 
@@
 %
-To illustrate our approach, we analyze two data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
+To illustrate our approach, we analyze four data sets common in developmental biology. 
 %
-Our first data set, from a live imaging study where the correct rotational orientation and temporal order are independently known, will be used to validate our approach.
+Our first two data sets come from live imaging studies of {\em Drosophila} and zebrafish embryogenesis; here, the correct rotational orientation and temporal order are independently known and these data sets will be used to validate our approach.
+%from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
 %
-Our second data set consists of images from fixed embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.
+Our third data set consists of images from fixed {\em Drosophila} embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.
+%
+Our final data set consists of z-stacks of wing discs, which we will use to illustrate how our methods can extend to three-dimensional imaging.
 
@@
 The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
-%\label{fig:schematics}
-\customlabel{fig:schematics}{2}
-\customlabel{subfig:synch1}{\ref{fig:schematics}A}
-\customlabel{subfig:synch2}{\ref{fig:schematics}B}
-\customlabel{subfig:dmaps1}{\ref{fig:schematics}C}
-\customlabel{subfig:dmaps2}{\ref{fig:schematics}D}
+\label{fig:schematics}
+%\customlabel{fig:schematics}{2}
+%\customlabel{subfig:synch1}{\ref{fig:schematics}A}
+%\customlabel{subfig:synch2}{\ref{fig:schematics}B}
+%\customlabel{subfig:dmaps1}{\ref{fig:schematics}C}
+%\customlabel{subfig:dmaps2}{\ref{fig:schematics}D}
 \end{figure}
@@
 %
-A schematic illustration of angular synchronization is shown in \fig~\ref{subfig:synch1}, where each image is represented as a vector, and the goal is to align the set of vectors.
+A schematic illustration of angular synchronization is shown in \fig~\ref{fig:schematics}A, where each image is represented as a vector, and the goal is to align the set of vectors.
 %
@@
 %
-Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see \SI); this is illustrated in \fig~\ref{subfig:synch2}.
+Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see \SI); this is illustrated in \fig~\ref{fig:schematics}B.
 %
@@
 %
-The idea is illustrated in \fig~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
+The idea is illustrated in \fig~\ref{fig:schematics}C, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
 %
@@
 %
-This idea of locality is denoted by the color of the edges in \fig~\ref{subfig:dmaps1}:
+This idea of locality is denoted by the color of the edges in \fig~\ref{fig:schematics}C:
 data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
 %
-The color in \fig~\ref{subfig:dmaps2} depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
+The color in \fig~\ref{fig:schematics}D depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
 %
-In our working examples, each data point will be of much higher dimension (e.g., a pixelated image), and so we cannot extract this low-dimensional structure visually.
+In our working examples, each data point will be of much higher dimension (e.g., a pixelated image or three-dimensional voxel data), and so we cannot extract this low-dimensional structure visually.
 %
@@
 \raisebox{0.75cm}{\textsf{A}}
-\hfill
-\includegraphics[width=17.5cm]{drosophila_live_imaging_scrambled}
+\includegraphics[width=17cm]{drosophila_live_imaging_scrambled}
 
 \raisebox{0.75cm}{\textsf{B}}
-\hfill
-\includegraphics[width=17.5cm]{drosophila_live_imaging_ordered}
+\includegraphics[width=17cm]{drosophila_live_imaging_ordered}
 
@@
 }
+\label{fig:drosophila_live_imaging}
 \end{figure*}
@@
 
-\fig~\ref{subfig:movie_unregistered_unordered} shows selected images from this live imaging data set, which contains $40$ consecutive frames taken at $30$~second time intervals at a fixed position within a single embryo (see \fig~\ref{fig:full_movie_data}).
+\fig~\ref{fig:drosophila_live_imaging}A shows selected images from this live imaging data set, which contains $40$ consecutive frames taken at $30$~second time intervals at a fixed position within a single embryo (see \fig~\ref{fig:full_movie_data}).
 %
@@
 %
-\fig~\ref{subfig:movie_registered_ordered} shows the images from \fig~\ref{subfig:movie_unregistered_unordered}, now registered and ordered; the real time for each frame is also indicated.
+\fig~\ref{fig:drosophila_live_imaging}B shows the images from \fig~\ref{fig:drosophila_live_imaging}A, now registered and ordered; the real time for each frame is also indicated.
 %
@@
 %
-\fig~\ref{subfig:movie_angle_corr} and \fig~\ref{subfig:movie_rank_corr}  show the correlations between the recovered and true angles and ranks, respectively, for the entire data set: 
+\fig~\ref{fig:drosophila_live_imaging}C and \fig~\ref{fig:drosophila_live_imaging}D  show the correlations between the recovered and true angles and ranks, respectively, for the entire data set: 
 %
@@
 %
-The results are shown in \fig~\ref{subfig:movie_error_table}. 
+The results are shown in \fig~\ref{fig:drosophila_live_imaging}E. 
 %
@@
 
-In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{subfig:dmaps1}) is well-sampled.
+In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{fig:schematics}C) is well-sampled.
 %
@@
 \raisebox{0.5cm}{\textsf{A}}
-\includegraphics[width=13.5cm]{zebrafish_scrambled}
+\includegraphics[width=13cm]{zebrafish_scrambled}
 
 \raisebox{0.5cm}{\textsf{B}}
-\includegraphics[width=13.5cm]{zebrafish_ordered}
+\includegraphics[width=13cm]{zebrafish_ordered}
 \end{minipage}
@@
 %
-\begin{minipage}[c]{3.5cm}
+\begin{minipage}[c]{3.75cm}
 
@@
 %
-\includegraphics[width=3.15cm]{zebrafish_rank_corr}
+\includegraphics[width=3.25cm]{zebrafish_rank_corr}
 \end{minipage}
-\caption{Method validation using live imaging of a zebrafish embryo. Registered and ordered zebrafish data.}
+\caption{Method validation using live imaging of a zebrafish embryo.
+(A) Selected images from a time-lapse movie of zebrafish embryogenesis.
+Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
+(B) Images from A after registration and ordering using vector diffusion maps.
+The real time of each frame is also indicated.
+(C) Correlation between the rank recovered using vector diffusion maps and the true rank.
+The rank correlation coefficient is $0.9936$.}
+\label{fig:zebrafish}
 \end{figure*}
@@
 %
-Quantitatively, the rank correlation coefficient for this data set is $0.9935$, and the average error in the recovered angle is $4.23^\circ$.
+Quantitatively, the rank correlation coefficient for this data set is $0.9936$, and the average error in the recovered angle is $4.23^\circ$.
 
@@
 {(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert.
-The rank correlation coefficient is $0.97$.
+The rank correlation coefficient is $0.9716$.
 }
+\label{fig:drosophila_fixed_images}
 \end{figure*}
@@
 %
-\fig~\ref{subfig:fixed_images_unregistered_unordered} shows selected images from a set of $120$ images (see \fig~\ref{fig:full_fixed_data}) which cover a thirty minute time interval spanning late cellularization through gastrulation.
+\fig~\ref{fig:drosophila_fixed_images}A shows selected images from a set of $120$ images (see \fig~\ref{fig:full_fixed_data}) which cover a thirty minute time interval spanning late cellularization through gastrulation.
 %
@@
 
-\fig~\ref{subfig:fixed_images_registered_ordered} shows the selected images in \fig~\ref{subfig:fixed_images_unregistered_unordered}, now registered and ordered using vector diffusion maps \citep{singer2012vector}.
+\fig~\ref{fig:drosophila_fixed_images}B shows the selected images in \fig~\ref{fig:drosophila_fixed_images}A, now registered and ordered using vector diffusion maps \citep{singer2012vector}.
 %
@@
 %
-Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:fixed_images_average_trajectory}, serve as a summary of the stereotypic developmental dynamics.
+Sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.
 %
@@
 %
-At the end of this process, similar ``omegas'' formed by Twi and dpERK appear on the dorsal side of the embryo; these patterns are most readily seen in the last image of \fig~\ref{subfig:fixed_images_average_trajectory}.
+At the end of this process, similar ``omegas'' formed by Twi and dpERK appear on the dorsal side of the embryo; these patterns are most readily seen in the last image of \fig~\ref{fig:drosophila_fixed_images}C.
 %
@@
 %
-The ranks from the ordering provided by the embryologist, which we will refer to as the ``expert rank'', are indicated for the images in \fig~\ref{subfig:fixed_images_registered_ordered} , and the rank correlation (see \fig~\ref{subfig:fixed_images_rank_corr}) shows that our ordering is consistent with the expert ordering.
+The ranks from the ordering provided by the embryologist, which we will refer to as the ``expert rank'', are indicated for the images in \fig~\ref{fig:drosophila_fixed_images}B , and the rank correlation (see \fig~\ref{fig:drosophila_fixed_images}D) shows that our ordering is consistent with the expert ordering.
+
+\subsubsection*{Wing disc data}
+
+\begin{figure*}[t]
+\raisebox{5.5cm}{\textsf{A}}
+\includegraphics[width=6cm]{wing_disc_example}
+%
+\hfill
+%
+\raisebox{5.5cm}{\textsf{B}}
+\includegraphics[width=10cm]{wing_disc_ordered}
+%
+\caption{Three-dimensional wing disc data.
+(A) Maximum projections of an example three-dimensional image.
+Projections along the $x$-, $y$-, and $z$-axes are shown.
+(B) Example three-dimensional images, ordered using diffusion maps.
+The time cohort, as assessed by an expert, is indicated for each image, and the rank correlation coefficient between the diffusion maps ordering and the expert timing is 0.9427.}
+\label{fig:wing_disc}
+\end{figure*}
+
+Up to this point, we have focused on two-dimensional imaging studies. 
+%
+However, it is not uncommon to encounter three-dimensional imaging studies. 
+%
+To illustrate how our methods can be extended to three-dimensional data, we analyzed a wing disc data set.
+
+Although the theory of vector diffusion maps readily extends to three-dimensional data, calculation of the pairwise alignments rapidly becomes computational prohibitive for three dimensional data, as one must now search over three Euler angles, rather than a single rotation angle.
+%
+However,  ordering data via diffusion maps is readily possible for three-dimensional data. 
+%
+Therefore, we can apply our methodology to three-dimensional data by first registering the data using an alternative methodology (such as registering max projections, or aligning a single feature of the data), and then ordering the registered data using diffusion maps.
+
+For the wing disc data, we first registered the three-dimensional data by computing the optimal rotations to align the maximum projections of the data using vector diffusion maps.
+%
+Because the maximum projections are two-dimensional images, this step is no more computationally intensive than the previous examples. 
+%
+We then used the computed alignments to register the full three-dimensional data, and then computed the diffusion maps embedding of the three-dimensional registered data. 
+%
+The results of ordering the data using the first (non-trivial) diffusion maps coordinate are shown in \fig~\ref{fig:wing_disc}.
+
+The data was sorted into $6$ time cohorts by an expert; the time cohort for each image in \fig~\ref{fig:wing_disc}B are indicated. 
+%
+The rank correlation coefficient between the ordering from diffusion maps and the expert cohort sorting is 0.9427, indicating that our ordering is consistent with the expert.
+
+\subsection*{Computational requirements}
+
+\begin{figure*}[t]
+\raisebox{5cm}{\textsf{A}}
+\includegraphics[width=5.5cm]{timing_studies_nimages}
+%
+\raisebox{5cm}{\textsf{B}}
+\includegraphics[width=5.5cm]{timing_studies_npixels}
+%
+\raisebox{5cm}{\textsf{C}}
+\includegraphics[width=5.5cm]{timing_studies_nrot}
+
+\raisebox{1.75cm}{\textsf{D}}
+{\renewcommand{\arraystretch}{1.5}
+{ \textsf{ \small 
+\begin{tabular}{ | p{2cm} | c | c | c | c | c | c | }
+\hline
+Data Set & 
+Data Type & 
+\parbox[c]{2cm}{ \centering Number of Channels} & 
+\parbox[c]{2cm}{ \centering \vspace{.5\baselineskip} Number of Images \vspace{.5\baselineskip}} & 
+\parbox[c]{2cm}{ \centering Number of Pixels} & 
+\parbox[c]{2cm}{ \centering Angular Discretization} & 
+\parbox[c]{2cm}{ \centering CPU time}  \\
+\hline
+{\em Drosophila} live~imaging & 2D & 1 & 40 & 100 $\times$ 100 & 10$^\circ$ & 3.2~sec \\ 
+Zebrafish & 2D & 1 & 120 & 100 $\times$ 100 & 10$^\circ$ & 13~sec\\ 
+{\em Drosophila} fixed images & 2D & 3 & 120 & 100 $\times$ 100 & 10$^\circ$ & 29~sec  \\ 
+Wing disc & 3D & 3 & 46 & 100 $\times$ 100 $\times$ 21 & 10$^\circ$ & 12~sec  \\
+ \hline
+\end{tabular}}}}
+\caption{Computational requirements for the presented methodology.
+(A) CPU time as a function of the number of images in the data set (for $100 \times 100$ pixel images, and $10^\circ$ rotational discretizations).
+(B) CPU time as a function of the number of pixels in the images (for $120$ images, and $10^\circ$ rotational discretizations).
+(C) CPU time as a function of the number of rotations (for $120$ images of $100 \times 100$ pixels).
+(D) The algorithm settings and computational requirements for the data sets analyzed.
+All times are reported for an Intel Core i7 2.93 GHz processor.}
+\label{fig:cpu_time}
+\end{figure*}
 
@@
 %
-In contrast, our method requires relatively little computational time; registering and ordering this fixed imaging data set took $26$~seconds on an Intel Core i7 2.93 GHz processor.
-%for $100$ images of $100 \times 100$ pixels, using 10$^{\circ}$ rotational discretizations.
+In contrast, our method requires relatively little computational time.
+%
+The computational costs for an Intel Core i7 2.93 GHz processor are outlined in \fig~\ref{fig:cpu_time}.
 %
@@
 %
-Empirically, the CPU time required was $\mathcal{O}(n^{1.5})$ in the number of images, and $\mathcal{O}(n)$ in the number of pixels and the angular resolution.
+Furthermore, the computation of the pairwise rotational alignments is trivially parallelizable for larger data sets.
 %
-Furthermore, the computation of the pairwise rotational alignments, which is the most time-intensive portion of the calculation, is trivially parallelizable.
+We subsampled all of our data sets to $100 \times 100$ pixels, as we found that this resolution still captured all of the relevant developmental features.
+
 %
@@
 %
-%In our examples, we use 10$^{\circ}$ angle discretizations for the pairwise alignments, and one-tenth of the median pairwise distance for the kernel scale. 
-% (5$^{\circ}$--30$^{\circ}$ angle discretizations and kernel scales of one-fifth to one-twentieth of the median pairwise distances all yielded acceptable results)
-%
 Overall, the tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, and so this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.
 
-\subsubsection*{Wing disc data}
-
-\begin{figure*}[t]
-\begin{minipage}[c]{9cm}
-\vspace{0cm}
-
-\raisebox{6.75cm}{\textsf{A}}
-\includegraphics[width=8cm]{wing_disc_example}
-\end{minipage}
-%
-\begin{minipage}[c]{8cm}
-\caption{Registered and ordered three-dimensional wing-disc data}
-\end{minipage}
-
-\raisebox{0.7cm}{\textsf{B}}
-\includegraphics[width=\textwidth]{wing_disc_ordered}
-
-\end{figure*}
+$\mathcal{O}(n^{1.83})$ in number of pixels
+$\mathcal{O}(n^{1.33})$ in number of images
+$\mathcal{O}(n^{0.77})$ in number of rotations
 

### 2015-01-28 image_analysis_paper_development2/paper_figures/make_fig_drosophila_fixed_images.m

@@ -55,11 +55,13 @@ if corr(time, embed_coord) < 0
     embed_coord = -embed_coord;
 end
 
+fontsize = 8;
+
 % idx_to_plot = randsample(nimages, nsubimages);
 idx_to_plot = round(linspace(1, nimages, nsubimages));
 [~, I] = sort(embed_coord(idx_to_plot));
 
-make_fig(17, 17/nsubimages);
+make_fig(12, 12/nsubimages);
 for j=1:nsubimages
     make_subplot(nsubimages, 1, 0.01, j);
     imshow(make_gray_nuclei(images(:,:,:,idx_to_plot(j))));
@@ -67,15 +69,15 @@ end
 saveas(gcf, 'drosophila_fixed_images_scrambled.pdf');
 
 
-make_fig(17, 17/nsubimages);
+make_fig(12, 12/nsubimages);
 for j=1:nsubimages
     make_subplot(nsubimages, 1, 0.01, j);
     imshow(make_gray_nuclei(imrotate(images_registered(:,:,:,idx_to_plot(I(j))), -5, 'crop')));
-    text(npixels/2, npixels/2, sprintf('%d', time(idx_to_plot(I(j)))),'color',0.95*ones(1,3),'HorizontalAlignment','center','VerticalAlignment','middle', 'fontsize', 6)
+    text(npixels/2, npixels/2, sprintf('%d', time(idx_to_plot(I(j)))),'color',0.95*ones(1,3),'HorizontalAlignment','center','VerticalAlignment','middle', 'fontsize', 4)
 end
 saveas(gcf, 'drosophila_fixed_images_ordered.pdf');
 
-make_fig(17, 17/nsubimages);
+make_fig(12, 12/nsubimages);
 for j=1:nsubimages
     make_subplot(nsubimages, 1, 0.01, j);
     im_tmp = uint8(mean(images_analyzed(:,:,:,(j-1)*nimages/nsubimages+1:j*nimages/nsubimages), 4));
@@ -86,8 +88,9 @@ saveas(gcf, 'drosophila_fixed_images_average.pdf');
 
 make_fig(4,4);
 plot(tiedrank(time), tiedrank(embed_coord),'.k')
-xlabel('expert rank')
-ylabel('algorithm rank')
+xlabel('expert rank', 'fontsize', fontsize)
+ylabel('algorithm rank', 'fontsize', fontsize)
+set(gca, 'fontsize', fontsize)
[…]

### 2015-01-28 image_analysis_paper_development2/paper_figures/make_fig_wing_disc.m

@@ -52,27 +52,29 @@ plot_images(immultiply(permute(images_analyzed, [2 4 3 1 5]), 1.5), dim)
 plot_images(immultiply(permute(images_analyzed, [4 1 3 2 5]), 1.5), dim)
 
 %%
+
+fontsize = 6;
+
 make_fig(6, 6)
 % set(gcf, 'Renderer', 'Painters')
 
 plot_wing_disc_projections(images_registered(:,:,:,:,2), [], 1, 1)
 annotation('arrow', [0.05 0.15], [0.05 0.05], 'color', 0.95*ones(1,3), 'headlength', 5, 'headwidth', 5)
 annotation('arrow', [0.05 0.05], [0.05 0.15], 'color', 0.95*ones(1,3), 'headlength', 5, 'headwidth', 5)
-annotation('textarrow', [0.15 0.05], [0.05 0.05], 'color', 0.95*ones(1,3), 'string','x', 'headstyle','none', 'fontsize', 8)
-annotation('textarrow', [0.05 0.05], [0.15 0.05], 'color', 0.95*ones(1,3), 'string','y', 'headstyle','none', 'fontsize', 8)
+annotation('textarrow', [0.15 0.05], [0.05 0.05], 'color', 0.95*ones(1,3), 'string','x', 'headstyle','none', 'fontsize', fontsize)
+annotation('textarrow', [0.05 0.05], [0.15 0.05], 'color', 0.95*ones(1,3), 'string','y', 'headstyle','none', 'fontsize', fontsize)
 
 annotation('arrow', [0.82 0.87], [0.05 0.05], 'color', 0.95*ones(1,3), 'headlength', 5, 'headwidth', 5)
 annotation('arrow', [0.82 0.82], [0.05 0.1], 'color', 0.95*ones(1,3), 'headlength', 5, 'headwidth', 5)
-annotation('textarrow', [0.87 0.82], [0.05 0.05], 'color', 0.95*ones(1,3), 'string','z', 'headstyle','none', 'fontsize', 8)
-annotation('textarrow', [0.82 0.82], [0.1 0.05], 'color', 0.95*ones(1,3), 'string','y', 'headstyle','none', 'fontsize', 8)
+annotation('textarrow', [0.87 0.82], [0.05 0.05], 'color', 0.95*ones(1,3), 'string','z', 'headstyle','none', 'fontsize', fontsize)
[…]

### 2015-01-28 image_analysis_paper_development2/paper_figures/timing_studies.m

@@ -209,51 +209,73 @@ fontsize = 8;
 figsize = 5;
 
 make_fig(figsize, figsize);
-loglog(npixels_zebrafish, cpu_times_npixels_zebrafish, '.-b');
+loglog(npixels_zebrafish, cpu_times_npixels_zebrafish, '.b', 'markersize', 10);
 hold on
-loglog(npixels_drosophila, cpu_times_npixels_drosophila, '.-r');
+loglog(npixels_drosophila, cpu_times_npixels_drosophila, '.r', 'markersize', 10);
 xlabel('number of pixels', 'fontsize', fontsize)
-ylabel('CPU time', 'fontsize', fontsize)
-h = legend('grayscale','color','location','southoutside');
-set(h, 'fontsize', fontsize);
+ylabel('CPU time (seconds)', 'fontsize', fontsize)
+h = legend('grayscale','color','location','southeast');
+set(h, 'fontsize', 4);
 set(gca, 'fontsize', fontsize)
-set(gca, 'xtick', [50 100 200 400])
-set(gca, 'ytick', [10 60 360])
-set(gca, 'YTickLabel', {'10 sec'; '1 min'; '6 min'})
+set(gca, 'xtick', [100 1000])
+set(gca, 'xticklabel', {'100x100'; '1000x1000'});
+x_data = linspace(50, 1500, 10);
+P = polyfit(log(npixels_zebrafish(2:end)), log(cpu_times_npixels_zebrafish(2:end)),1);
+P(1)
+loglog(x_data, exp(polyval(P, log(x_data))), '--b');
+
+P = polyfit(log(npixels_drosophila(2:end)), log(cpu_times_npixels_drosophila(2:end)),1);
+P(1)
+loglog(x_data, exp(polyval(P, log(x_data))), '--r');
 axis tight
 axis square
 saveas(gcf, 'timing_studies_npixels.pdf');
 
 make_fig(figsize, figsize);
-loglog(nimages_zebrafish, cpu_times_nimages_zebrafish, '.-b');
+loglog(nimages_zebrafish, cpu_times_nimages_zebrafish, '.b', 'markersize', 10);
 hold on
-loglog(nimages_drosophila, cpu_times_nimages_drosophila, '.-r');
+loglog(nimages_drosophila, cpu_times_nimages_drosophila, '.r', 'markersize', 10);
 xlabel('number of images', 'fontsize', fontsize)
[…]

### 2015-01-29 image_analysis_paper_development2/SI_text.tex
updated code and paper

@@
 \subsection*{Image parameters}
+
+The original image resolution is $512 \times 512$ for the live {\em Drosophila} images, $320 \times 288$ for the zebrafish images, $1024 \times 1024$ for the fixed {\em Drosophila} images, and $1024 \times 1024$ for each of the images in the wing disc z-stacks.
+%
+Each of the original wing disc z-stacks contains 30--40 images.
+
+All images were subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
+%
+Each wing disc z-stack was reduced to a 21-image stack, consisting of the brightest image and the 10 images above and below.
+
+The relevant image operations are 
+\begin{enumerate}
+\item {\bf Intensity normalization:} Contrast-limited adaptive histogram equalization (using the \texttt{adapthisteq} function with an $8 \times 8$ tile grid, and a uniform distribution for the intensities with a clip limit of 0.01) is used to normalize the intensities of the nuclear signal.
+\item {\bf Blur} Using the \texttt{imfilter} function. 
+\item {\bf Intensity scaling} 
+\item {\bf Mean-centering} 
+\item {\bf Size scaling}
+\item {\bf Removing corners}
+\end{enumerate}
 

### 2015-01-29 image_analysis_paper_development2/manuscript.tex
updated code and paper

@@
 Our final data set consists of z-stacks of wing discs, which we will use to illustrate how our methods can extend to three-dimensional imaging. 
+%
+For data sets from multiple samples, we will also show how to compute an average trajectory from a set of registered and ordered images to obtain a smooth description of the underlying developmental dynamics.
 
@@
 %
+Although in general, images may contain variations due to rotations, translations, and scaling, we will remove relevant translations and/or scaling via relatively simple preprocessing, and focus only on factoring out rotations using the vector diffusion maps algorithm.
+%
+In the case that all relevant symmetries can be removed with straightforward preprocessing procedures, our algorithms can also be used simply to extract the main direction of variability within the imaging data set.
+%
 We assume that the main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
@@
 (C) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
-Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \SI), so that close data points are connected by darker (``stronger'') edges.
+Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through a Gaussian kernel (see \SI), so that close data points are connected by darker (``stronger'') edges.
 (D) The data in C, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure.
@@
 \label{fig:schematics}
-%\customlabel{fig:schematics}{2}
-%\customlabel{subfig:synch1}{\ref{fig:schematics}A}
-%\customlabel{subfig:synch2}{\ref{fig:schematics}B}
-%\customlabel{subfig:dmaps1}{\ref{fig:schematics}C}
-%\customlabel{subfig:dmaps2}{\ref{fig:schematics}D}
 \end{figure}
@@
 %
-A schematic illustration of angular synchronization is shown in \fig~\ref{fig:schematics}A, where each image is represented as a vector, and the goal is to align the set of vectors.
+A schematic illustration of angular synchronization is shown in \fig~\ref{fig:schematics}A, where each image is represented as a vector, and the goal is to align the entire set of vectors given pairwise alignment measurements.
 %
-We first compute the angles needed to align pairs of vectors (or images).  
+We first compute the angles needed to align pairs of vectors (or images), which in general requires no template function \citep{ahuja2007template}.
 %
-In general, this requires no template function \citep{ahuja2007template} or image landmarks \citep{ian1998statistical}.
+In this work, we aligned pairs of images with respect to rotations by exhaustively searching over a discretized space of rotation angles. 
+%
+However, in other applications, pairwise alignments can be computed by aligning appropriate image landmarks or features. 
+%
+When the data are noisy, these pairwise measurements may by inaccurate, and so we would like to utilize {\em all} pairwise measurements to align the set of images robustly. 
 %
@@
 %
-\hfill
+%\hfill
 %
@@
 %
-\raisebox{2.75cm}{
+\raisebox{2.5cm}{
 { \textsf{ 
-\begin{tabular}{ |p{1.25cm} | p{1.5cm}|  p{1.75cm}| }
+\begin{tabular}{ |c | c| c| }
 \hline
-Movie Index & Average Angle Error & Rank Correlation Coefficient \\
+\parbox[c]{1.5cm}{ \centering Movie Index} & \parbox[c]{1.5cm}{ \vspace{.5\baselineskip} \centering Average Angle Error \vspace{.5\baselineskip}} & \parbox[c]{1.5cm}{ \centering Rank Correlation Coefficient} \\
 \hline
@@
 %
-Each image shows an optical cross-section of a vertically oriented developing embryo, with the nuclei labeled by Histone-RFP.
+Each image shows an optical cross-section near the posterior pole of a vertically oriented developing embryo, with the nuclei labeled by Histone-RFP.
 %
@@
 (C) Correlation between the rank recovered using vector diffusion maps and the true rank.
-The rank correlation coefficient is $0.9936$.}
+The rank correlation coefficient is $0.9954$.}
 \label{fig:zebrafish}
@@
 %
-The original movie is available at \texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov} \cite{karlstrom1996flipbook}. 
+We used a publicly available flipbook movie of zebrafish embryogenesis, available at \texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov} \cite{karlstrom1996flipbook}.  
 %
-We analyzed 120 frames from this movie which span the epiboly stage.
+Taken with a differential interference contrast (DIC) microscope, the movie records the first $17$~hours of zebrafish development, from a single cell stage to a 16-somite stage. 
+%
+We selected $120$ frames from this movie that captures 5.5~hours of epiboly (3.5--9~hours after fertilization). 
+%
+At the start of the time window, cells have divided 10--11 times and are accumulated in a cell mass above the yolk. 
+%
+The cell mass is then compressed and the animal-vegetal axis of the embryo (top/bottom in the figure) shortens to form a spherical embryo shape by the end of the fourth hour of development. 
+%
+Then, the yolk syncytial layer, which forms the boundary between the yolk and the cell mass, moves upward, forming a dome-shaped structure.  
+%
+During this stage, the cells rearrange to form a cell layer with uniform thickness (about 4-cells thick). 
+%
+With time, this cell layer then spreads over across the yolk and expands toward the vegetal pole. 
+%
+At the end of epiboly, the blastoderm completely engulfs the yolk.
 
-As in the live imaging {\em Drosophila} example, the frames were randomly rotated and shuffled. 
+As in the example of {\em Drosophila} embryo live imaging, the frames were randomly rotated and shuffled (\fig~\ref{fig:zebrafish}A). 
 %
@@
 %
-The results are shown in \fig~\ref{fig:zebrafish}. 
+The results are shown in \fig~\ref{fig:zebrafish}B. 
 %
-The recovered rotations and order are consistent with the expected developmental dynamics. 
+The recovered rotations and order are consistent with the expected developmental dynamics, as shown in the correlations between the recovered and true ranks (\fig~\ref{fig:zebrafish}C). 
 %
-Quantitatively, the rank correlation coefficient for this data set is $0.9936$, and the average error in the recovered angle is $4.23^\circ$.
+Some mismatch in early embryos result from rather static cell movement during early developmental stage where cells divide and pile up above the yolk. 
+%
+During epiboly, cell movement is more dynamic and the recovered ordering is more consistent with the real dynamics.
+%
+Quantitatively, the rank correlation coefficient for this data set is 0.9954, and the average error in the recovered angle is $4.14^\circ$.
+
+In summary, we showed that our algorithm is not limited to a specific set of data. 
+%
+The algorithm works for different tissue types ({\em Drosophila} and zebrafish), and also for images taken with different imaging methods (fluorescent and phase contrast images).
+%
+Provided there exist significant dynamics within the data set, the developmental dynamics can be recovered.
 
@@
 
-We have analyzed how our algorithm performs on a model data set, where all images come from a single embryo. 
+We have analyzed how our algorithm performs on two model data sets, where all images come from a single embryo. 
 %
@@
 To demonstrate that our methods are robust to such noise sources,
-we constructed a synthetic time course data set by selecting an image from a random live imaging experiment at each time point.
+we constructed a synthetic time course data set by selecting an image from a random {\em Drosophila} live imaging experiment (those experiments used in \fig~\ref{fig:drosophila_live_imaging}) at each time point.
 %
@@
 
-Finally, we applied our approach to a data set where the true rotational orientation and temporal order was not known {\it a priori}.
+We then applied our approach to a data set where the true rotational orientation and temporal order was not known {\it a priori}.
 %
-\fig~\ref{fig:drosophila_fixed_images}A shows selected images from a set of $120$ images (see \fig~\ref{fig:full_fixed_data}) which cover a thirty minute time interval spanning late cellularization through gastrulation.
+\fig~\ref{fig:drosophila_fixed_images}A shows selected images from a set of $120$ images of developing {\em Drosophila} embryos (see \fig~\ref{fig:full_fixed_data}) which cover a thirty minute time interval spanning late cellularization through gastrulation.
 %
@@
 %
-Each image shows an optical cross-section of a {\em different} vertically oriented embryo at a different rotational orientation and fixed at a different (and unknown) developmental time.
+Each image shows an optical cross-section of the posterior view of a {\em different} embryo at a different rotational orientation and fixed at a different (and unknown) developmental time.
 %
@@
 %
+Each snapshot in the average trajectory is the (weighted) average of a group of successive images from the registered and ordered data set.
+%
 Sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.
-%
-We can now easily see the developmental progression, which is consistent with the known dynamics: 
+
+From this average trajectory, we can now easily see the developmental progression, which is consistent with the known dynamics: 
 dpERK first appears as two lateral peaks at the ventral side of the embryo; a third dpERK peak then appears at the dorsal side of the embryo.
@@
 Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental features (e.g., correlation of age with body size).
+%
 
@@
 \includegraphics[width=10cm]{wing_disc_ordered}
-%
+
+\vspace{0.25cm}
+\raisebox{2.25cm}{\textsf{C}}
+\includegraphics[width=17.5cm]{wing_disc_average}
+
 \caption{Three-dimensional wing disc data.
@@
 (B) Example three-dimensional images, ordered using diffusion maps.
-The time cohort, as assessed by an expert, is indicated for each image, and the rank correlation coefficient between the diffusion maps ordering and the expert timing is 0.9427.}
+The time cohort, as assessed by an expert, is indicated for each image, and the rank correlation coefficient between the diffusion maps ordering and the expert timing is 0.9427.
+(C) The average developmental trajectory for the registered and ordered images.
+}
 \label{fig:wing_disc}
@@
 %
-However, it is not uncommon to encounter three-dimensional imaging studies. 
+However, it is not uncommon to encounter three-dimensional imaging data in the form of z-stacks. 
 %
@@
 %
-However,  ordering data via diffusion maps is readily possible for three-dimensional data. 
+However, ordering data via diffusion maps is readily possible for (registered) three-dimensional data. 
 %
@@
 The rank correlation coefficient between the ordering from diffusion maps and the expert cohort sorting is 0.9427, indicating that our ordering is consistent with the expert. 
+%
+Furthermore, we can extract a stereotypical developmental trajectory by averaging; this trajectory is shown in \fig~\ref{fig:wing_disc}.
 
@@
 (A) CPU time as a function of the number of images in the data set (for $100 \times 100$ pixel images, and $10^\circ$ rotational discretizations).
+Empirically, the CPU time is $\sim \mathcal{O}(n^{1.33})$ in number of images.
 (B) CPU time as a function of the number of pixels in the images (for $120$ images, and $10^\circ$ rotational discretizations).
+Empirically, the CPU time is $\sim \mathcal{O}(n^{1.83})$ in number of pixels.
 (C) CPU time as a function of the number of rotations (for $120$ images of $100 \times 100$ pixels).
+Empirically, the CPU time is $\sim \mathcal{O}(n^{-0.77})$ in the angular discretization.
 (D) The algorithm settings and computational requirements for the data sets analyzed.
@@
 %
-The computational costs for an Intel Core i7 2.93 GHz processor are outlined in \fig~\ref{fig:cpu_time}.
+The computational costs for our methodology are outlined in \fig~\ref{fig:cpu_time}.
 %
@@
 %
-Furthermore, the computation of the pairwise rotational alignments is trivially parallelizable for larger data sets.
+Furthermore, the computation of the pairwise rotational alignments, which accounts for the majority of the computational time, is trivially parallelizable for larger data sets.
 %
-We subsampled all of our data sets to $100 \times 100$ pixels, as we found that this resolution still captured all of the relevant developmental features.
+Becasuse the computational cost increases with the image resolution, we chose to subsample all of our data sets to $100 \times 100$ pixels.
+%
+This resolution allowed us to rapidly analyze our data sets while still capturing all of the relevant developmental features.
+%
+However, as can be seen from the computational costs in \fig~\ref{fig:cpu_time}, it is feasible to use our algorithms to analyze higher-resolution images.
 
+The requisite user intervention and parameter tuning required for our method is relatively minimal.
 %
-The requisite user intervention and parameter tuning required for our method is relatively minimal: Images must first be preprocessed so that the Euclidean distance between the pixels is informative.
+As a first step, images must be preprocessed so that the Euclidean distance between the pixels is informative.
 %
-Two algorithmic parameters, the angular discretization to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{subfig:dmaps1} and \SI), must also be defined.
+Our software provides several preprocessing options (such as blurring, rescaling, and mean-centering), as well as some guidance for what options to select depending on the system of interest. 
+%
+Two algorithmic parameters, the angular discretization to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{subfig:dmaps1} and \SI), must also be defined; we also provide some guidance on selecting these parameters.
 %
@@
 Overall, the tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, and so this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.
-
-$\mathcal{O}(n^{1.83})$ in number of pixels
-$\mathcal{O}(n^{1.33})$ in number of images
-$\mathcal{O}(n^{0.77})$ in number of rotations
 
@@
 %
-Although we have presented results for a specific type of data (cross-sectional images taken along the dorsoventral axis of {\em Drosophila} embryos), our methodology only relies on computing distances between images and may be readily extended to many other data sets.
+We have applied our methodology to several different proprotypical data sets in developmental biology; because our methodology only relies on computing distances between images and may be readily extended to many other data sets.
+
+We acknowledge that our methods, though general, do have some limitations. 
+%
+The first is that we require enough amount of data to sufficiently sample the developmental trajectory.
+%
+Therefore, for very small and/or very noisy data sets, our algorithms may fail. 
+%
+Second, the developmental features need to be large compared to the noise and the image resolution. 
+%
+In all of our examples, the relevant features, such as expression patterns and morphological structures, all span several pixels and are large compared to both the instrument noise and embryo-to-embryo variability.
+%
+This means that the Euclidean distance is a good measure of similarity between images.
+%
+For other imaging data sets which contain very fine scale structures, our algorithms will fail as small perturbations in these fine-scale structures will result in large changes in the Euclidean distances.
+%
+Furthermore, as we have acknowledged, our methods register full three-dimensional data with respect to rotations due to computational limitations. 
+%
+However, the algorithms presented here could be used for three-dimensional given an alternative method for computing pairwise alignments more efficiently (such as using image features or landmarks).
+%
+Unlike traditional statistical methods, our algorithms do not provide confidence intervals on predicted orderings, but could in principle be extracted from bootstrapping \citep{efron1979bootstrap}.
+%
+Furthermore, although our methods provide a temporal ordering for images, the provide no real-time information, neither do they provide a direction for the proposed ordering.
+%
+Determining which ``end'' of the trajectory is early and which is late is a post-processing task which requires {\em a priori} knowledge of the developmental dynamics. 
+%
+Furthermore, real-time information can only be extracted from either {\em a priori} knowledge of the rates of relevant developmental processes, or by making some assumption about the temporal sampling of the data set (e.g., the data set uniformly samples some given time window).
 

### 2015-01-29 image_analysis_paper_development2/paper_figures/make_fig_drosophila_fixed_images.m

@@ -78,9 +78,11 @@ end
 saveas(gcf, 'drosophila_fixed_images_ordered.pdf');
 
 make_fig(12, 12/nsubimages);
+avg_images = compute_average_trajectory(images_analyzed, nsubimages, 4);
 for j=1:nsubimages
     make_subplot(nsubimages, 1, 0.01, j);
-    im_tmp = uint8(mean(images_analyzed(:,:,:,(j-1)*nimages/nsubimages+1:j*nimages/nsubimages), 4));
+%     im_tmp = uint8(mean(images_analyzed(:,:,:,(j-1)*nimages/nsubimages+1:j*nimages/nsubimages), 4));
+    im_tmp = avg_images(:,:,:,j);
     imshow(make_gray_nuclei(imrotate(im_tmp, -5, 'crop')));
 end
 saveas(gcf, 'drosophila_fixed_images_average.pdf');

### 2015-01-29 image_analysis_paper_development2/paper_figures/make_fig_wing_disc.m

@@ -34,6 +34,14 @@ plot_images(images, dim)
 ncomps = 1;
 [R_opt, embed_coord, D2] = vdm(R, W, eps_scale, ncomps);
 
+fid = fopen(sprintf('%s/times.txt', image_dir), 'r');
+time = fscanf(fid, '%f');
+fclose(fid);
+
+if corr(time, embed_coord) < 0
+    embed_coord = -embed_coord;
+end
+
 % register images using optimal rotations
 images_registered = register_all_images(images, R_opt);
 images_registered = imrotate(images_registered, 90);
@@ -77,13 +85,6 @@ annotation('textarrow',  [0.1 0.05],[0.82 0.82], 'color', 0.95*ones(1,3), 'strin
 print('wing_disc_example.eps', '-depsc', '-r300');
 
 %%
-fid = fopen(sprintf('%s/times.txt', image_dir), 'r');
-time = fscanf(fid, '%f');
-fclose(fid);
-
-if corr(time, embed_coord) < 0
-    embed_coord = -embed_coord;
-end
 
 nsubimages = 15;
 [~, I] = sort(embed_coord(20+(1:nsubimages)));
@@ -94,4 +95,13 @@ plot_wing_disc_projections(images_registered(:,:,:,:,I), time(I)-min(time), 5, 3
 % plot_wing_disc_projections(images_registered(:,:,:,:,I), time(I)-min(time));
 % saveas(gcf, 'wing_disc_ordered.pdf');
 print('wing_disc_ordered.eps', '-depsc', '-r300');
+
+%%
+
+nsubimages = 6;
+
+avg_images = compute_average_trajectory(images_analyzed, nsubimages, 4);
+make_fig(17, 17/nsubimages);
+plot_wing_disc_projections(avg_images, [], nsubimages, 1)
+print('wing_disc_average.eps', '-depsc', '-r300');
 

### 2015-01-29 image_analysis_paper_development2/paper_figures/make_tutorial.m

@@ -1,88 +1,0 @@
-clear all
-close all
-
-%%
-% <latex>
-% First, we create an artifical data set of $n=7$ data points which lie 
-% on a one-dimensional curve in two dimensions
-% </latex>
-
-n = 7;
-
-theta = linspace(0, 3*pi/2, n);
-
-x = [0.9*cos(theta); 0.9*sin(theta)]
-
-figure;
-scatter(x(1,:),x(2,:),50, 'k', '.');
-xlabel('x(1)');
-ylabel('x(2)');
-
-%%
-% <latex>
-% Then we compute the $n \times n$ distance matrix between data points
-% </latex>
-
-dist_matrix = dist(x)
-
-%%
-% <latex>
-% We choose $\epsilon$, the kernel scale, as half the median of the
-% pairwise distances
-% </latex>
-
-eps = median(dist_matrix(:))/2
-
-%%
-% <latex>
-% We then compute the $n \times n$ matrix $W$, with 
-% \[ W_{ij} = \exp \left(-\frac{ \| x_i - x_j \|^2}{\epsilon^2} \right)\]
-% </latex>
-
-W = exp(-dist_matrix.^2/eps.^2)
-
-
-%%
-% <latex>
-% We then compute the $n \times n$ diaganol matrix $D$, with 
-% \[ D_{ii} = \sum_{j=1}^n W_{ij} \]
-% </latex>
-
-D = diag(sum(W))
-
-%%
-% <latex>
-% We then compute the $n \times n$ matrix $A = D^{-1}W$
-% </latex>
-
-A = inv(D)*W
-
-%%
-% <latex>
-% We then compute the top eigenvectors and eigenvalues of $A$, and order
-% them by the magnitude of the eigenvalues
-% </latex>
-
-[evecs, evals] = eig(A);
-[~, I] = sort(abs(diag(evals)), 'descend');
-
-evecs = evecs(:,I)
-evals = evals(I,I)
-
-%%
-% <latex>
-% The first eigenvector is a trivial constant eigenvector with eigenvalue
-% 1. The second eigenvector provides an ordering of the data. If we color
-% our data by the value of this second eigenvector, we see it is one-to-one
-% with the arclength along the curve. Furthermore, if we sorted the data by
-% the corresponding entries of the second eigenvector, we would order the
-% data along this one-dimensional nonlinear curve. 
-% </latex>
-
-figure;
[…]

### 2015-01-30 image_analysis_paper_development2/SI_text.tex
updated paper

--- 2015-01-25 image_analysis_paper_development2/SI_text.tex
@@
 
-\subsection*{Image parameters}
+\subsection*{Image preprocessing}
 
-The original image resolution is $1024 \times 1024$ pixels for the fixed images, and $512 \times 512$ pixels for the live images; all images are subsampled to $100 \times 100$ pixels for vector diffusion maps analysis.
+Before applying vector diffusion maps, images must be preprocessed so that the Euclidean distance between time image pixels is informative.
 %
-A scale bar for a representative image is shown in \fig~\ref{fig:ap_dv}.
+We want images who are developmentally similar to have relatively small Euclidean distances, and images who are at disparate developmental time points to have relatively large distances. 
 %
-When imaging, the laser settings are calibrated so that images have good dynamic range, while not saturating the detector.
+We therefore must adjust images to remove any experimental and imaging artifacts, so that most of the variation is relevant to the developmental dynamics.
+
+The relevant image operations are 
+\begin{enumerate}
+\item {\bf Intensity normalization:} Contrast-limited adaptive histogram equalization (using the \texttt{adapthisteq} function with an $8 \times 8$ tile grid, and a uniform distribution for the intensities with a clip limit of 0.01) is used to normalize the intensities of signals whose absolute intensity is not meaningful/informative.
 %
-The same laser settings are maintained within each experiment.
+\item {\bf Blur:} The \texttt{imfilter} function with a disc filter is used to blur signals whose small-scale structure is not informative. 
 %
-The intensity of each channel is integer-valued, and ranges from $0$ (no signal) to $255$ (saturated). 
+\item {\bf Intensity scaling (multichannel images only):} The \texttt{immultiply} function is used to scale signal intensities.
+This is important for multichannel images, and it determines the (relative) contributions of the various signals.
 %
-The Canny method \citep{canny1986computational} is used to detect the edges of the nuclear signal in each image (using the \texttt{edge} function in MATLAB\textsuperscript{\textregistered}), and images are cropped to these edges to remove any size effects.
+\item {\bf Mean-centering:}  The Canny method \citep{canny1986computational} is used to detect the edges of the object in each image (using the \texttt{edge} function).
+The image is then translated so that the object (as determined by extremeties of the detected edges) is centered.
 %
-A constant $10$-pixel-wide border is then added to each image. 
+\item {\bf Size scaling:} For images whose relative size is unimportant to the developmental dynamics, the images are rescaled/dialated so that the object size (as determined by extremeties of the detected edges) is constant (we set this to be 80\% of the total image).
 %
-Contrast-limited adaptive histogram equalization (using the \texttt{adapthisteq} function with an $8 \times 8$ tile grid, and a uniform distribution for the intensities with a clip limit of 0.01) is used to normalize the intensities of the nuclear signal.
+\item {\bf Removing corners:} The image is cropped/filtered using a disc centered in the middle of the image and whose diameter is equal to the number of pixels; any pixels outside of this disc are set to 0 (no intensity).
+This removes any corner effects under rotations.
+\end{enumerate}
+
+\subsubsection*{{\em Drosophila} live imaging}
+
 %
-The nuclear signal is blurred using a disc filter with a $5$~pixel radius (using the \texttt{imfilter} function).
+The original image resolution is $512 \times 512$ for the live {\em Drosophila} images.
 %
-For the multichannel images, the nuclear signal is scaled by half (after normalization and blurring) relative to the other signals, so that the total nuclear signal across the entire image is of a comparable magnitude to the other signals.
+All images were subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
+%
+Images were normalized, and then blurred with a filter of radius 5\% of the total image (5 pixels). 
+%
+Images were not mean-centered or rescaled, as the entire live imaging data set was approximately centered already.
+
+\subsubsection*{Zebrafish images}
+
+The original image resolution was $320 \times 288$ for the zebrafish images.
+%
+All images were subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
+%
+Images were not normalized or blurred.
+%
+Images were mean-centered so that the embryo in each frame was (approximately) centered. 
+%
+Images were not rescaled for size, as changes in overall were are important.
+
+\subsubsection*{{\em Drosophila} fixed images}
+
+The original image resolution was $1024 \times 1024$ for the fixed {\em Drosophila} images.
+%
+All images were subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
+%
+The nuclear channel was normalized, and all channels blurred with a filter of radius 5\% of the total image (5 pixels) to remove the effects of individual nuclei. 
+%
+The nuclear channel was scaled by half; because this signal occupies a larger fraction of the image relative to the other signals, its overall contribution is large, and so we downscaled it so that each signal would have more comparable weight in the algorithm.
+%
+The images were mean-centered and scaled to have a constant size using the nuclear signal to detect the edges of the embryo within the frame.
+
+\subsubsection*{Wing disc images}
+
+The original image resolution was $1024 \times 1024$ for the wing disc z-stacks.
+%
+Each of the original wing disc z-stacks contains 30--40 images. 
+%
+All images were subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
+%
+Each wing disc z-stack was reduced to a 21-image stack, consisting of the brightest image and the 10 images above and below.
+%
+No channels were normalized or blurred in the images, and the channel intensities were kept at their imaging values. 
+%
+Images were mean-centered using the Wingless/Patched signal to detect the edges of the sample. 
+%
+Images were not rescaled for size, as changes in overall were are important.
+
+%The original image resolution is $1024 \times 1024$ pixels for the fixed images, and $512 \times 512$ pixels for the live images; all images are subsampled to $100 \times 100$ pixels for vector diffusion maps analysis.
+%%
+%A scale bar for a representative image is shown in \fig~\ref{fig:ap_dv}.
+%%
+%When imaging, the laser settings are calibrated so that images have good dynamic range, while not saturating the detector.
+%%
+%The same laser settings are maintained within each experiment.
+%%
+%The intensity of each channel is integer-valued, and ranges from $0$ (no signal) to $255$ (saturated). 
+%%
+%The Canny method \citep{canny1986computational} is used to detect the edges of the nuclear signal in each image (using the \texttt{edge} function in MATLAB\textsuperscript{\textregistered}), and images are cropped to these edges to remove any size effects.
+%%
+%A constant $10$-pixel-wide border is then added to each image. 
+%%
+%Contrast-limited adaptive histogram equalization (using the \texttt{adapthisteq} function with an $8 \times 8$ tile grid, and a uniform distribution for the intensities with a clip limit of 0.01) is used to normalize the intensities of the nuclear signal.
+%%
+%The nuclear signal is blurred using a disc filter with a $5$~pixel radius (using the \texttt{imfilter} function).
+%%
+%For the multichannel images, the nuclear signal is scaled by half (after normalization and blurring) relative to the other signals, so that the total nuclear signal across the entire image is of a comparable magnitude to the other signals.
 
@@
 
-\begin{figure*}
-\includegraphics{figS3}
-\caption{Eigenvalue spectra.
+\begin{figure}
+\includegraphics[width=8cm]{drosophila_live_eval_spectrum}
+
+\includegraphics[width=2.5cm]{drosophila_live_evec_corr1}
+\includegraphics[width=2.5cm]{drosophila_live_evec_corr2}
+\includegraphics[width=2.5cm]{drosophila_live_evec_corr3}
+\caption{Eigenvalue spectra for the {\em Drosophila} live imaging data set presented in \fig~3.
+Note that there is a gap after the fourth eigenvalue product.
+Below are the second, third, and fourth embedding coordinate plotted versus the first embedding coordinate.
+Note that coordinates 2--4 are higher harmonics (and thus simple functions) of coordinate 1, and are therefore not informative about structure in the data set.
+We can conclude that the data set is effectively one-dimensional and can be parameterized/ordered by coordinate 1.}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=8cm]{zebrafish_eval_spectrum}
+
+\includegraphics[width=2.5cm]{zebrafish_evec_corr1}
+\includegraphics[width=2.5cm]{zebrafish_evec_corr2}
+\includegraphics[width=2.5cm]{zebrafish_evec_corr3}
+\caption{Eigenvalue spectra for the zebrafish data set presented in \fig~4.
+Note that there is a gap after the fourth eigenvalue product.
+Below are the second, third, and fourth embedding coordinate plotted versus the first embedding coordinate.
+Note that coordinates 2--4 are higher harmonics (and thus simple functions) of coordinate 1, and are therefore not informative about structure in the data set.
+It is not immediately obvious that coordinate 3 is a harmonic of coordinate 1; the distortion in the plot is due to density effects in the data (the developmental changes are slower towards the beginning of the trajectory, and so there is a higher density of images in this portion of the one-dimensional curve), We can conclude that the data set is effectively one-dimensional and can be parameterized/ordered by coordinate 1.}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=8cm]{drosophila_fixed_eval_spectrum}
+\caption{Eigenvalue spectra for the fixed {\em Drosophila} images presented in \fig~5.
+Note that there is a gap after the first eigenvalue product.
+We can conclude that the data set is effectively one-dimensional and can be parameterized/ordered by coordinate 1.}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=8cm]{wing_disc_eval_spectrum}
+\caption{Eigenvalue spectra for the wing disc data set presented in \fig~6.
+Note that there is a gap after the first eigenvalue product.
+We can conclude that the data set is effectively one-dimensional and can be parameterized/ordered by coordinate 1.}
+\end{figure}
+
+%
+%\begin{figure*}
+%\includegraphics{figS3}
+%\caption{Eigenvalue spectra.
 {\it (A)} Eigenvalue spectrum for the live imaging data set presented in \fig~3.

### 2015-01-30 image_analysis_paper_development2/code_for_distribution/README/README.tex
updated paper

\documentclass[12pt]{article}

\usepackage{amsmath, amsfonts, amssymb}
\usepackage{graphicx}

\usepackage{epstopdf}

\begin{document}

\section{Introduction}

This is code for the registration and ordering algorithms described in \\
``Temporal ordering and registration of images in studies of developmental dynamics,'' Carmeline~J.~Dsilva, Bomyi~Lim, Hang~Lu, Amit~Singer , Stanislav~Y.~Shvartsman, and Ioannis~G.~Kevrekidis

The code reads in a set of two-dimensional images or three-dimensional z-stacks, can perform a number of preprocessing operations, and then registers and/or orders the images using the algorithms described in the paper.

\section{Requirements}

All algorithms and analysis were implemented in MATLAB\textsuperscript{\textregistered} (R2013b, The MathWorks, Natick, Massachusetts).
%
The code requires MATLAB along with the Image Processing Toolbox.

\section{How To Use}

There are two ways to use the code
\begin{enumerate}
\item Interactive GUI.
\item Program a simple script.
\end{enumerate}

The basic steps are as follows:
\begin{enumerate}
   \item Read in images
   \item Preprocess the images (smooth and/or equalize channels, mean-center, etc.)
   \item Run vector diffusion maps to register and order preprocessed images
   \item (Optional) Save registered and ordered images
\end{enumerate}

\subsection{Interactive GUI}

\includegraphics[width=10cm, trim=0cm 2cm 5cm 0cm, clip]{gui_screenshot_initial.jpg}

There are five steps
%
\begin{enumerate}
\item Read Images
\item Preprocess Images
\item Analyze Images
\item Save Images
\item Average Trajectory
\end{enumerate}

\subsubsection{Read Images}

The fields are as follows
\begin{description}
\item[2D/3D] Select ``2D'' if your images are traditional 2D images. Select ``3D'' if your images are z-stacks.
%
\item[Image Directory] Enter the directory where your images or z-stacks are stored. This directory can be an absolute path (e.g. \texttt{C:/Documents/Images}) or a path relative to the directory where this code is stored (e.g. \texttt{../images})
%
\end{description}
\subsection{Script}

\subsection*{Read Images}

\begin{par}
The first step is to read in images. This is done using the \texttt{read\_images} function For two-dimensional images, it requires that all the images are stored in a single directory, all with a consistent prefix, and indexed with two digits starting from \texttt{01}. For three-dimensional z-stack, it assumes each z-stack is stored in a single directory. All of the z-stack directories are required to begin with a consistent prefix, and be indexed with two digits starting from \texttt{01}. Similarly, all of the images are required to begin with a consistent prefix, and be indexed with two digits starting from \texttt{01} for each z-stack.
\end{par} \vspace{1em}
\begin{verbatim}
% directory where images are stored
image_dir = 'drosophila_fixed_images';

% prefix for each image
image_name = 'emb';

% image type/extension
image_ext = 'tif';

% no stack name or number of iamges in a stack are required for 2D images
stack_name = '';
nstack = 0;

% number of images
nimages = 120;

% number of pixels
npixels = 100;

% dimension of images (dim=2 indicates standard 2D images, rather than
% z-stacks)
dim = 2;

% read in images
% images are stored in the variable images_raw
[images_raw, nchannels] = read_images(image_dir, image_name, image_ext, ...
    stack_name, nimages, nstack, npixels, dim);
\end{verbatim}

\subsection*{Show images}

\begin{par}
An optional step is to now show the images using the \texttt{plot\_images} function.
\end{par} \vspace{1em}
\begin{verbatim}
% plot the images
% images_raw are the images to plot
% (returned from the read_images function)
% dim is the dimension of the images
% (dim=2 indicates standard 2D images, rather than z-stacks)
plot_images(images_raw, dim)
\end{verbatim}

\includegraphics [width=4in]{sample_input_file_01.eps}

\subsection*{Preprocess Images}

\begin{par}
Now, we must preprocess the images before registration and ordering, to remove any imaging and/or experimental artifacts. We do this via the \texttt{apply\_image\_functions} function For this particular imaging data set, there are three channels. DAPI, a nuclear stain, is in the first (red) channel. dpERK is in the second (green) channel. Dl is in the third (blue) channel.
\end{par} \vspace{1em}
\begin{verbatim}
% channel weights
% because the nuclear signal is wider spread throughout the image, as well
% as noiser, we choose to scale the first channel by half
channel_weight = [0.5 1 1];

% we blur the image slightly (5%) so that small-scale effects of individual
% nuclei are smoothened
channel_blur = [0.05 0.05 0.05];

% because the absolute intensity of the nuclear signal is not informative
% or important (only the presence or absence of signal is important,
% indicating the presence or absence of nuclei), we normalize the first channel
% We do not normalize the other two channels, as changes in intensity are
% informative to the developmental processes.
channel_normalize = [1 0 0];

% we use the nuclear channel to mean center the images, as this channel
% parameterizes/captures the entire embryo
channel_mean_center = [1 0 0];

% we resize the images to all be (approximately) the same size, to remove
% any variations due to size effects, as size effects are known to not be
% important for this developmental system
resize_image = true;

% we then apply these image functions of normalization, blurring,
% reweighting, and mean-centering
images = apply_image_functions(images_raw, dim, channel_weight, ...
    channel_blur, channel_normalize, channel_mean_center, resize_image);

% plot the images (optional)
plot_images(images, dim)
\end{verbatim}

\includegraphics [width=4in]{sample_input_file_02.eps}

\subsection*{Calculate pairwise alignments}

\begin{par}
We now need to calculate the angles needed to align \textit{pairs} of images
\end{par} \vspace{1em}
\begin{verbatim}
% angular discretization when computing pairwise aligments
% this means we search for pairwise aligmemnts over 10 degree increments
ang_dis = 10;

% compute the pairwise alignments
% images are the preprocessed images
% ang_dis is the angular discretization
% R and W store the pairwise alignments and distances, respectively, for
% vector diffusion maps
[R, W] = compute_pairwise_alignments(images, ang_dis);
\end{verbatim}

\subsection*{Apply vector diffusion maps}

\begin{par}
We can now use vector diffusion maps to register and order the images.
\end{par} \vspace{1em}
\begin{verbatim}
% ncomps is the number of components to compute
% we only compute 1 coordinate because we only need to order the images
% (i.e., sort by the first coordinate)
ncomps = 1;

% epsilon scale for diffusion maps kernel
% eps_scale = 0.25 means that the epsilon in the diffusion maps kernel is
% 1/4 of the median of the pairwise distances between data points
eps_scale = 0.25;

% vector diffusion maps calculates optimal rotations and embedding
% coordinate
[R_opt, embed_coord, D2] = vdm(R, W, eps_scale, ncomps);

% register images
images_registered = register_all_images(images, R_opt);

% order registered images by embedding coordinate
images_analyzed = order_all_images(images_registered, embed_coord);

% plot the images (optional)
plot_images(images_analyzed, dim)
\end{verbatim}

\includegraphics [width=4in]{sample_input_file_03.eps}

\end{document}

### 2015-01-30 image_analysis_paper_development2/code_for_distribution/README/sample_input_file.tex
updated paper

--- 2015-01-30 image_analysis_paper_development2/code_for_distribution/README/README.tex
@@
-\documentclass[12pt]{article}
+% This LaTeX was auto-generated from an M-file by MATLAB.
+% To make changes, update the M-file and republish this document.
 
-\usepackage{amsmath, amsfonts, amssymb}
+\documentclass{article}
 \usepackage{graphicx}
+\usepackage{color}
 
-\usepackage{epstopdf}
+\sloppy
+\definecolor{lightgray}{gray}{0.5}
+\setlength{\parindent}{0pt}
 
@@
 
-\section{Introduction}
+\subsection*{Contents}
 
-This is code for the registration and ordering algorithms described in \\
-``Temporal ordering and registration of images in studies of developmental dynamics,'' Carmeline~J.~Dsilva, Bomyi~Lim, Hang~Lu, Amit~Singer , Stanislav~Y.~Shvartsman, and Ioannis~G.~Kevrekidis
+\begin{itemize}
+\setlength{\itemsep}{-1ex}
+   \item Basic Algorithm
+   \item Read Images
+   \item Show images
+   \item Preprocess Images
+   \item Calculate pairwise alignments
+   \item Apply vector diffusion maps
+\end{itemize}
+\begin{verbatim}
+clear all
+close all
+\end{verbatim}
 
-The code reads in a set of two-dimensional images or three-dimensional z-stacks, can perform a number of preprocessing operations, and then registers and/or orders the images using the algorithms described in the paper.
+\subsection*{Basic Algorithm}
 
-\section{Requirements}
-
-All algorithms and analysis were implemented in MATLAB\textsuperscript{\textregistered} (R2013b, The MathWorks, Natick, Massachusetts).
-%
-The code requires MATLAB along with the Image Processing Toolbox.
-
-\section{How To Use}
-
-There are two ways to use the code
+\begin{par}
+The basic steps are as follows:
+\end{par} \vspace{1em}
 \begin{enumerate}
-\item Interactive GUI.
-\item Program a simple script.
-\end{enumerate}
-
-The basic steps are as follows:
-\begin{enumerate}
+\setlength{\itemsep}{-1ex}
    \item Read in images
@@
    \item Run vector diffusion maps to register and order preprocessed images
-   \item (Optional) Save registered and ordered images
 \end{enumerate}
-
-\subsection{Interactive GUI}
-
-\includegraphics[width=10cm, trim=0cm 2cm 5cm 0cm, clip]{gui_screenshot_initial.jpg}
-
-There are five steps
-%
-\begin{enumerate}
-\item Read Images
-\item Preprocess Images
-\item Analyze Images
-\item Save Images
-\item Average Trajectory
-\end{enumerate}
-
-\subsubsection{Read Images}
-
-The fields are as follows
-\begin{description}
-\item[2D/3D] Select ``2D'' if your images are traditional 2D images. Select ``3D'' if your images are z-stacks.
-%
-\item[Image Directory] Enter the directory where your images or z-stacks are stored.
-This directory can be an absolute path (e.g.
-\texttt{C:/Documents/Images}) or a path relative to the directory where this code is stored (e.g.
-\texttt{../images})
-%
-\end{description}
-\subsection{Script}
 

### 2015-01-30 image_analysis_paper_development2/manuscript.tex
updated paper

@@
 \customlabel{fig:1d_example}{S2}
-\customlabel{fig:eigenvalues}{S3}
-\customlabel{fig:full_movie_data}{S4}
-\customlabel{fig:all_movies}{S5}
-\customlabel{fig:full_fixed_data}{S6}
+\customlabel{fig:eigenvalues_drosophila_live}{S3}
+\customlabel{fig:eigenvalues_zebrafish}{S4}
+\customlabel{fig:eigenvalues_drosophila_fixed}{S5}
+\customlabel{fig:eigenvalues_wing_disc}{S6}
 
@@
 
-\fig~\ref{fig:drosophila_live_imaging}A shows selected images from this live imaging data set, which contains $40$ consecutive frames taken at $30$~second time intervals at a fixed position within a single embryo (see \fig~\ref{fig:full_movie_data}).
+\fig~\ref{fig:drosophila_live_imaging}A shows selected images from this live imaging data set, which contains $40$ consecutive frames taken at $30$~second time intervals at a fixed position within a single embryo.
 %
@@
 
-To assess the robustness of our proposed methodology, we repeated this procedure with four additional data sets (see \fig~\ref{fig:all_movies}) extracted from independent live imaging studies spanning the same developmental time period.
+To assess the robustness of our proposed methodology, we repeated this procedure with four additional data sets extracted from independent live imaging studies spanning the same developmental time period.
 %
@@
 %
-\fig~\ref{fig:drosophila_fixed_images}A shows selected images from a set of $120$ images of developing {\em Drosophila} embryos (see \fig~\ref{fig:full_fixed_data}) which cover a thirty minute time interval spanning late cellularization through gastrulation.
+\fig~\ref{fig:drosophila_fixed_images}A shows selected images from a set of $120$ images of developing {\em Drosophila} embryos which cover a thirty minute time interval spanning late cellularization through gastrulation.
 %
@@
 %
-However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set (see \ref{fig:full_fixed_data}).
+However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set.
 %
@@
 
-Up to this point, we have focused on two-dimensional imaging studies. 
+We have demonstrated that our algorithm can recover the underlying developmental dynamics in several different types of imaging data sets, including ones containing interembryo variability. 
 %
-However, it is not uncommon to encounter three-dimensional imaging data in the form of z-stacks. 
+Thus far, we have only looked at two-dimensional data sets where biological samples are imaged as a single confocal slice or as a phase contrast image. 
 %
-To illustrate how our methods can be extended to three-dimensional data, we analyzed a wing disc data set.
+However, many developmental processes are inherently three-dimensional, and the z-axis often contains interesting dynamics which are independent of the dynamics in the x-y plane. 
+%
+Researchers often take multiple z-slices to visualize the three-dimensional structure of a sample;
+with advances in imaging techniques, such three-dimensional data will be more common, and it is important to validtate that our algorithm can be applied to such data.
 
-Although the theory of vector diffusion maps readily extends to three-dimensional data, calculation of the pairwise alignments rapidly becomes computational prohibitive for three dimensional data, as one must now search over three Euler angles, rather than a single rotation angle.
+We used an existing three-dimensional data set of fixed {\em Drosophila} wing imaginal discs, which was published in \citet{hamaratoglu2011dpp}.  
+%
+Imaginal discs are groups of progenitor cells in fly larva that will transform into a specific organ, such as wings, legs, and eyes, in the adult fly during the pupal transformation. 
+%
+The wing disc is an imaginal disc that turns into a fly wing. 
+%
+The data set is composed of 46 fixed wing discs, whose developmental time range from 72 to 112~hours after fertilization. 
+%
+Each disc contains 21 z-slices taken at 1~$\mu$m intervals. 
+%
+The discs were stained with antibodies that recognize Spalt (red), Dad-GFP (green), and Wingless/Patched (gray) (\fig~\ref{fig:wing_disc}A). 
+%
+The disc grows in time; at long times, the disc becomes too large to fit completely within the frame and only a partial disc is imaged.
+
+Although the theory of vector diffusion maps readily extends to three-dimensional data, calculation of the pairwise alignments rapidly becomes computational prohibitive for three dimensional data, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle).
 %
@@
 %
-Therefore, we can apply our methodology to three-dimensional data by first registering the data using an alternative methodology (such as registering max projections, or aligning a single feature of the data), and then ordering the registered data using diffusion maps.
-
-For the wing disc data, we first registered the three-dimensional data by computing the optimal rotations to align the maximum projections of the data using vector diffusion maps.
+Therefore, we can apply our methodology to three-dimensional data by first registering the data using an alternative methodology (such as registering maximum intensity projection, or aligning a single feature of the data), and then ordering the registered data using diffusion maps.
 %
-Because the maximum projections are two-dimensional images, this step is no more computationally intensive than the previous examples. 
+Instead, we used maximum intensity projection and converted the image into 2D, and then used the angular synchronization on 2D data. 
 %
-We then used the computed alignments to register the full three-dimensional data, and then computed the diffusion maps embedding of the three-dimensional registered data. 
+For the wing disc data, we first registered the three-dimensional data by computing the optimal rotations to align the maximum intensity projections of the data using vector diffusion maps.
 %
-The results of ordering the data using the first (non-trivial) diffusion maps coordinate are shown in \fig~\ref{fig:wing_disc}.
-
-The data was sorted into $6$ time cohorts by an expert; the time cohort for each image in \fig~\ref{fig:wing_disc}B are indicated. 
+Because the maximum intensity projections are two-dimensional images, this step is no more computationally intensive than the previous examples. 
 %
-The rank correlation coefficient between the ordering from diffusion maps and the expert cohort sorting is 0.9427, indicating that our ordering is consistent with the expert. 
+Other options for registering samples include using a specific localized signal (such as Twi in \fig~\ref{fig:drosophila_fixed_images}) as a marker, or using other {\em a priori} knowledge.
+%
+We then used diffusion maps to order the registered three-dimensional images. 
+%
+\fig~\ref{fig:wing_disc}B shows selected images from the data set ordered by diffusion maps. 
+%
+In the original data set, each disc was assigned to 6 time classes (72--73~hr, 76.5--77.5~hr, 79--80~hr, 89--90~hr, 100--101~hr, and 110.5--111.5~hr after fertilization); these times are indicated in \fig~\ref{fig:wing_disc}B.
+%
+In the ordered set, the size of wing disc grows, and the intensity of proteins increases as a function of time. 
+%
+The rank correlation coefficient based on the time class is 0.9427. 
 %
 Furthermore, we can extract a stereotypical developmental trajectory by averaging; this trajectory is shown in \fig~\ref{fig:wing_disc}.
+%
+Thus, this result shows that the algorithm can successfully order a three-dimensional data set.
 
@@
 %
-Two algorithmic parameters, the angular discretization to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{subfig:dmaps1} and \SI), must also be defined; we also provide some guidance on selecting these parameters.
+Two algorithmic parameters, the angular discretization to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{fig:schematics}A and \SI), must also be defined; we also provide some guidance on selecting these parameters.
 %
@@
 
-\subsection*{Experiments}
+\subsection*{{\em Drosophila} Experiments}
 %
-Oregon-R was used as wild type strains. 
+Oregon-R was used as wild type {\em Drosophila} strains. 
 %
@@
 
-\subsection*{Microscopy}
+\subsection*{{\em Drosophila} Microscopy}
 %
@@
 \subsection*{Image preprocessing}
+
+Images were subsampled, normalized, blurred, and centered prior to diffusion maps analysis to remove any variations due to the epxperimental and imaging framework. 
 %
-Image resolution was subsampled to $100 \times 100$ pixels for vector diffusion maps analysis.
-%
-Borders were cropped from the images using a Canny edge detector applied to the nuclear signal. 
-%
-Contrast-limited adaptive histogram equalization and a a 5-pixel radius disc filter were used to normalize and blur the nuclear signal.
+Details about the preprocessing operations applied to each imging data set are given in \SI.
 
@@
 
-The authors thank Adam Finkelstein,  Thomas Funkhouser, and John Storey for helpful discussions.
+The authors thank Fisun Hamaratoglu and Markus Affolter for providing the wing disc data. 
+%
+The authors thank Angela DePace, Granton Jindal, Adam Finkelstein,  Thomas Funkhouser, and John Storey for helpful discussions.
 

### 2015-01-30 image_analysis_paper_development2/paper_figures/make_fig_drosophila_fixed_images.m

@@ -99,6 +99,19 @@ axis([0 125 0 125])
 axis square
 saveas(gcf, 'drosophila_fixed_images_rank_corr.pdf');
 
+%%
+
+ncomps = 20;
+[R_opt, embed_coord, D2] = vdm(R, W, eps_scale, ncomps);
+
+make_fig(8,8)
+plot(abs(D2), '.')
+xlabel('embedding coordinate')
+ylabel('product of eigenvalues')
+axis square
+saveas(gcf, 'drosophila_fixed_eval_spectrum.pdf');
+
+
 
 
 

### 2015-01-30 image_analysis_paper_development2/paper_figures/make_fig_drosophila_live_imaging.m

@@ -97,7 +97,7 @@ for i=1:nexpt
         for j=1:nsubimages
             make_subplot(nsubimages, 1, 0.01, j);
             imshow(imrotate(images_registered(:,:,I(j)), 50, 'crop'));
-%             text(npixels/2, npixels/2, sprintf('%2.1f min', time(I(j))),'color',0.95*ones(1,3),'HorizontalAlignment','center','VerticalAlignment','middle', 'fontsize', 6)
+            %             text(npixels/2, npixels/2, sprintf('%2.1f min', time(I(j))),'color',0.95*ones(1,3),'HorizontalAlignment','center','VerticalAlignment','middle', 'fontsize', 6)
             text(npixels, npixels, sprintf('%2.1f min', time(I(j))),'color',0.95*ones(1,3),'HorizontalAlignment','right','VerticalAlignment','bottom', 'fontsize', 4)
         end
         saveas(gcf, 'drosophila_live_imaging_ordered.pdf');
@@ -128,6 +128,27 @@ for i=1:nexpt
         axis square
         saveas(gcf, 'drosophila_live_imaging_rank_corr.pdf');
         
+        ncomps = 20;
+        [R_opt, embed_coord, D2] = vdm(R, W, eps_scale, ncomps);
+        
+        make_fig(8,8)
+        plot(abs(D2), '.')
+        xlabel('embedding coordinate')
+        ylabel('product of eigenvalues')
+        axis square
+        saveas(gcf, 'drosophila_live_eval_spectrum.pdf');
+        
+        for k = 2:4
+            make_fig(3,3)
+            plot(embed_coord(:,1), embed_coord(:,k), '.')
+            xlabel('embed coord 1')
+            ylabel(sprintf('embed coord %d', k))
+            axis tight
+            axis square
+            set(gca, 'xtick', [])
+            set(gca, 'ytick', [])
+            saveas(gcf, sprintf('drosophila_live_evec_corr%d.pdf', k-1));
+        end
         return
         
         

### 2015-01-30 image_analysis_paper_development2/paper_figures/make_fig_wing_disc.m

@@ -105,3 +105,16 @@ make_fig(17, 17/nsubimages);
 plot_wing_disc_projections(avg_images, [], nsubimages, 1)
 print('wing_disc_average.eps', '-depsc', '-r300');
 
+%%
+
+ncomps = 20;
+[embed_coord, D2] = dm(W, eps_scale, ncomps);
+
+make_fig(8,8)
+plot(abs(D2), '.')
+xlabel('embedding coordinate')
+ylabel('product of eigenvalues')
+axis square
+saveas(gcf, 'wing_disc_eval_spectrum.pdf');
+
+

### 2015-01-30 image_analysis_paper_development2/paper_figures/make_fig_zebrafish.m

@@ -85,8 +85,8 @@ for j=1:nsubimages
     make_subplot(nsubimages, 1, 0.01, j);
     imshow(imrotate(images_registered(:,:,I(j)), 70, 'crop'));
     text(npixels, npixels, sprintf('%2.0f min', time(I(j))-min(time)),'color',0.95*ones(1,3),'HorizontalAlignment','right','VerticalAlignment','bottom', 'fontsize', 4)
-
-%     text(npixels/2, npixels/2, sprintf('%2.0f min', time(I(j))-min(time)),'color',0.95*ones(1,3),'HorizontalAlignment','center','VerticalAlignment','middle', 'fontsize', 6)
+    
+    %     text(npixels/2, npixels/2, sprintf('%2.0f min', time(I(j))-min(time)),'color',0.95*ones(1,3),'HorizontalAlignment','center','VerticalAlignment','middle', 'fontsize', 6)
 end
 saveas(gcf, 'zebrafish_ordered.pdf');
 
@@ -112,6 +112,31 @@ set(gca, 'xticklabel', {'0°'; '180°'; '360°'});
 set(gca, 'yticklabel', {'0°'; '180°'; '360°'});
 axis([0 400 0 400])
 axis square
+
+%%
+
+ncomps = 20;
+[R_opt, embed_coord, D2] = vdm(R, W, eps_scale, ncomps);
+
+make_fig(8,8)
+plot(abs(D2), '.')
+xlabel('embedding coordinate')
+ylabel('product of eigenvalues')
+axis square
+saveas(gcf, 'zebrafish_eval_spectrum.pdf');
+
+
+for k = 2:4
+    make_fig(3,3)
+    plot(embed_coord(:,1), embed_coord(:,k), '.')
+    xlabel('embed coord 1')
+    ylabel(sprintf('embed coord %d', k))
+    axis tight
+    axis square
+    set(gca, 'xtick', [])
+    set(gca, 'ytick', [])
+    saveas(gcf, sprintf('zebrafish_evec_corr%d.pdf', k-1));
+end
 
 %% bootstrap
 % nsamples = 10;

### 2015-01-31 image_analysis_paper_development2/code_for_distribution/README/README.tex
updated code so that npixels is a parameter in the preprocessing

@@
 
-\includegraphics[width=10cm, trim=0cm 2cm 5cm 0cm, clip]{gui_screenshot_initial.jpg}
+\includegraphics[width=\textwidth, trim=0cm 3cm 12cm 0cm, clip]{gui_screenshot_initial.jpg}
 
@@
 \begin{description}
+%
 \item[2D/3D] Select ``2D'' if your images are traditional 2D images. Select ``3D'' if your images are z-stacks.
@@
 %
+\item[Image Prefix] The prefix for all image file names. The images are assumed to named as the image prefix, followed by a number. 
+%
+\item[Stack Prefix] (only for 3D z-stacks) The prefix for all stack folder names. Each stack is assumed to be stored in a single folder, named as the stack prefix, followed by a number. 
+%
+\item[Image Extension] The extension of the images (e.g., tif, jpg)
+%
+\item[Number of Images] The number of images or z-stacks.
+Images are assumed to be stored as the image prefix, followed by a two-digit index, where the indices run from 1 to the number of images (e.g., image\_prefix01.tif, image\_prefix02.tif, etc.).
+For three-dimensional images, the z-stacks are assumed to be stored in individual folders whose names are the stack prefix, followed by a two-digit index, where the indices run from 1 to the number of stacks (e.g., z\_stack\_prefix01, z\_stack\_prefix02, etc.)
+%
+\item[Number of Images in Stack] (only for z-stacks) The number of images per z-stack.
+Images are assumed to be stored in the individual stack directories as the stack prefix, followed by a two-digit index, where the indices run from 1 to the number of images in the stack (e.g., stack\_prefix01.tif, stack\_prefix02.tif, etc.).
+%
 \end{description}
+
+After populating the required fields, select the ``Read Images'' button to read in the images (a wait par will appear to show the progress of image reading).
+%
+After reading the images, select the ``Show Images'' button to display the images (z-stacks will be displayed as maximum intensity projections). This step is optional.
+
+\subsubsection{Preprocess Images}
+
+\begin{description}
+\item[Weight] Each channel in the images is scaled by the weight factor, where all weights at 1 does not adjust the channel intensities. For grayscale images, only the first weight is used. 
+%
+\item[Blur] Each channel in the images is blurred using a disc factor with radius equal to a fraction of the image as indicated by the ``blur'' field.
+For example, setting the blur to 5\% means that the image is blurred using a disc filter with radius $0.5 \times$ the number of pixels.
+%
+\item[Normalize] If the normalize option is selected for a specific channel, then Contrast-Limited Adaptive Histogram Equalization is applied to that channel to normalize the intensities across the image.
+This is important for signals whose absolute intensity is not meaningful or informative.
+%
+\item[Use to Mean-Center] If the mean-center option is selected for any of the channels, then the object is centered in the image by detecting the edges of the object given by the selected channels.
+If no mean-center options are selected, then the object is not mean-centered.
+%
+\item[Resize Images] If the resize images option is selected, then each image is scaled/dialated so that the object occupies 80\% of the total image frame.
+This removes any size effects/variations in the images.
+%
+\item[Number of Pixels] The number of pixels wanted in the images for analysis.
+This does {\em not} have to be the same number of pixels as the true resolution of the images.
+The images are assumed to be square.
+%
+\end{description}
+
+After populating the required fields, select the ``Apply Image'' button to apply the selected preprocessing options (a wait bar will appear to show the progress).
+%
+After preprocessing the images, select the ``Show Images'' button to display the preprocessed images (z-stacks will be displayed as maximum intensity projections). This step is optional.
+
+\subsubsection{Analyze Images}
+
+After preprocessing the images, one can apply various analysis algorithms. 
+%
+There are three parameters
+%
+\begin{description}
+%
+\item[Angular Discretization (degrees)] (required for image registration) This is the angular discretization used to search for the pairwise alignments between images.
+For example, if the angular discretization is 10$^\circ$, then the pairwise alignments are computed by computing the image discrepancy after 10$^\circ$ rotation, after a 20$^\circ$ rotation, ..., up to a 360$^\circ$ rotation, and then selecting the rotation which minimizes the discrepancy.
+Lower values of the angular discretization increase the accuracy of the method, but also increase the computational cost.
+%
+\item[Kernel Scale] This is the scaling of the diffusion maps kernel.
+This parameter should be $\mathcal{O}(1)$, and we have empirically found for our data sets that values of $0.25-0.5$ often yield good results.
+%
+\end{description}
+
+After this, select the ``Apply to Raw Images'' option to apply the computed registration and/or ordering to the original images; select the ``Apply to Preprocessed Images'' option to apply the computed registration and/or ordering to the preprocessed images (after blurring, normalization, etc.)
+
+There are four analysis options
+\begin{description}
+%
+\item[Register] (this option is only for two-dimensional images) This registers the imaging data set using vector diffusion maps to compute the optimal rotations. 
+%
+\item[Order] This orders the imaging data set using diffusion maps.
+{\em Important:} This option assumes that the imaging data set is preregistered!
+This option can be used for imaging data sets that were preregistered using some {\em a priori} knowledge about the system.
+%
+\item[Register + Order] (this option is only for two-dimensional images) This registers and orders the imaging data set using vector diffusion maps to compute the optimal rotations and the temporal ordering.
+%
+\item[Register + Order z-stacks] (this option is only for three-dimensional z-stacks) This registers and orders the imaging data set.
+It first registers the z-stacks by computing the optimal rotations to align the (two-dimensional) maximum intensity projections using vector diffusion maps.
+It then orders the data using diffusion maps on the (now registered) three-dimensional z-stacks.
+%
+\end{description}
+
+After analysis select the ``Show Images'' button to display the registered and/or images (z-stacks will be displayed as maximum intensity projections). This step is optional.
+
+\subsubsection{Save Images}
+
+This is the (optional) step of saving the analyzed images.
+%
+The ``Output Image Directory'' field is where the analyzed images will be stored; this directory will be created if it does not exist. 
+%
+The images will be stored in the output directory; the name format will be the same as in the input directory.
+
+\subsubsection{Average Trajectory}
+
+An (optional) step is to compute the average trajectory of the registered and ordered images.
+
+There are two parameters
+\begin{description}
+\item[Number of Images in Average Trajectory] This is the desired number of images in the average trajectory. 
+%
+\item[Width of Averaging Window] This is the (approximate) width of the averaging window.
+Each snapshot in the average trajectory will be the (weighted) average of approximately the width number of images.
+Increasing this value will result in a smoother average trajectory; decreasing this value will reveal more finer-scale detail.
+% 
+\end{description}
+
+The ``Show Average Trajectory'' option will then display the average trajectory images. 
+%
+The ``Save Average Trajectory'' option will save the average trajectory images. 
+%
+The images will be saved in the ''Output Image Directory'' folder. 
+%
+The name format will be the same as the input file format, but with the prefix ``avg\_''.
+
 \subsection{Script}

### 2015-01-31 image_analysis_paper_development2/paper_figures/make_fig_drosophila_fixed_images.m

@@ -26,10 +26,10 @@ eps_scale = 0.25;
 
 image_dir = '../code_for_distribution/drosophila_fixed_images';
 
-[images_raw, nchannels] = read_images(image_dir, image_name, image_ext, stack_name, nimages, nstack, npixels, dim);
+[images_raw, nchannels] = read_images(image_dir, image_name, image_ext, stack_name, nimages, nstack, dim);
 plot_images(images_raw, dim)
 
-images = apply_image_functions(images_raw, dim, channel_weight, channel_blur, channel_normalize, channel_mean_center, resize_image);
+images = apply_image_functions(images_raw, npixels, dim, channel_weight, channel_blur, channel_normalize, channel_mean_center, resize_image);
 plot_images(images, dim)
 
 [R, W] = compute_pairwise_alignments(images, ang_dis);

### 2015-01-31 image_analysis_paper_development2/paper_figures/make_fig_zebrafish.m

@@ -26,10 +26,10 @@ eps_scale = 0.25;
 
 image_dir = '../code_for_distribution/zebrafish';
 
-[images_raw, nchannels] = read_images(image_dir, image_name, image_ext, stack_name, nimages, nstack, npixels, dim);
+[images_raw, nchannels] = read_images(image_dir, image_name, image_ext, stack_name, nimages, nstack, dim);
 plot_images(images_raw, dim)
 
-images = apply_image_functions(images_raw, dim, channel_weight, channel_blur, channel_normalize, channel_mean_center, resize_image);
+images = apply_image_functions(images_raw, npixels, dim, channel_weight, channel_blur, channel_normalize, channel_mean_center, resize_image);
 plot_images(images, dim)
 
 [R, W] = compute_pairwise_alignments(images, ang_dis);

### 2015-01-31 image_analysis_paper_development2/manuscript.tex
updated paper

@@
 %
-Our first two data sets come from live imaging studies of {\em Drosophila} and zebrafish embryogenesis; here, the correct rotational orientation and temporal order are independently known and these data sets will be used to validate our approach.
+Our first two data sets come from live imaging studies of {\em Drosophila} and zebrafish embryogenesis.
+%
+In both of these examples, the correct rotational orientation and temporal order are independently known, and these data sets will be used to validate our approach.
 %from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
@@
 %
-Our final data set consists of z-stacks of wing discs, which we will use to illustrate how our methods can extend to three-dimensional imaging. 
+Our final data set consists of z-stacks of wing discs, which we will use to illustrate how our methods extend to three-dimensional imaging. 
 %
-For data sets from multiple samples, we will also show how to compute an average trajectory from a set of registered and ordered images to obtain a smooth description of the underlying developmental dynamics.
+We will also show how to compute an average trajectory from a set of registered and ordered fixed images to remove noise due to intersample variability and obtain a smooth description of the underlying developmental dynamics.
 

### 2015-01-31 image_analysis_paper_development2/manuscript.tex
updated paper

@@
 (C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle.
+The average absolute error in the recovered angles is $8.37 ^\circ$.
 (D) The correlation between the recovered rank (using vector diffusion maps) and the true rank.
+The rank correlation coefficient is $0.9989$.
 (E) The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies.
@@
 %
-\fig~\ref{fig:drosophila_live_imaging}C and \fig~\ref{fig:drosophila_live_imaging}D  show the correlations between the recovered and true angles and ranks, respectively, for the entire data set: 
+\fig~\ref{fig:drosophila_live_imaging}C and \fig~\ref{fig:drosophila_live_imaging}D  show the correlations between the recovered and true angles and ranks, respectively, for the entire data set. 
 %
@@
 %%
-Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignments and the errors in the recovered angles were consistently less than $10^{\circ}$.
+Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignment, and the high temporal resolution of our live imaging study allows us to order the images accurately.
 %
@@
 \caption{Method validation using live imaging of a zebrafish embryo.
-(A) Selected images from a time-lapse movie of zebrafish embryogenesis.
+(A) Selected images from a flip book movie of zebrafish embryogenesis.
 Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
@@
 (C) Correlation between the rank recovered using vector diffusion maps and the true rank.
-The rank correlation coefficient is $0.9954$.}
+The rank correlation coefficient is $0.9954$.
+The larger errors in the recovered ranks towards the beginning of the trajectory are due to the rather static cell movement within that time window.
+}
 \label{fig:zebrafish}
@@
 %
-We used a publicly available flipbook movie of zebrafish embryogenesis, available at \texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov} \cite{karlstrom1996flipbook}.  
+We used a publicly available flipbook movie of zebrafish embryogenesis, available at \texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov} \citep{karlstrom1996flipbook}.  
 %
@@
 %
-We selected $120$ frames from this movie that captures 5.5~hours of epiboly (3.5--9~hours after fertilization). 
+We selected $120$ consecutive frames from this movie which capture 5.5~hours of epiboly (3.5--9~hours after fertilization). 
 %
@@
 %
-The cell mass is then compressed and the animal-vegetal axis of the embryo (top/bottom in the figure) shortens to form a spherical embryo shape by the end of the fourth hour of development. 
+The cell mass is then compressed and the animal-vegetal axis of the embryo (vertical axis in the \fig~\ref{fig:zebrafish}) shortens to form a spherical embryo shape by the end of the fourth hour of development.
 %
@@
 %
-During this stage, the cells rearrange to form a cell layer with uniform thickness (about 4-cells thick). 
+During this stage, the cells rearrange to form a cell layer with uniform thickness (about four cells thick). 
 %
@@
 %
-Some mismatch in early embryos result from rather static cell movement during early developmental stage where cells divide and pile up above the yolk. 
+Quantitatively, the rank correlation coefficient for this data set is 0.9954, and the average error in the recovered angle is $4.14^\circ$. 
+%
+Some errors in ordering images of the early embryo result from rather static cell movement during the early developmental stage where cells divide and accumulate above the yolk. 
 %
 During epiboly, cell movement is more dynamic and the recovered ordering is more consistent with the real dynamics.
-%
-Quantitatively, the rank correlation coefficient for this data set is 0.9954, and the average error in the recovered angle is $4.14^\circ$.
 
@@
 %
-Provided there exist significant dynamics within the data set, the developmental dynamics can be recovered.
+Provided there exist significant dynamics within the data set and that the developmental trajectory is well-sampled, the developmental dynamics can be recovered.
 
@@
 
-We have analyzed how our algorithm performs on two model data sets, where all images come from a single embryo. 
+We have analyzed how our algorithm performs on two model data sets where all images come from a single embryo. 
 %
-In practice, each image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
+In practice, we are interested in cases where image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
 %
 To demonstrate that our methods are robust to such noise sources,
-we constructed a synthetic time course data set by selecting an image from a random {\em Drosophila} live imaging experiment (those experiments used in \fig~\ref{fig:drosophila_live_imaging}) at each time point.
+we constructed a synthetic time course data set by selecting a random image from one of five {\em Drosophila} live imaging data sets (those data sets used in \fig~\ref{fig:drosophila_live_imaging}) at each time point.
 %
@@
 %
-Sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.
+Averaging successive images removes (some) of the interembryo variability, so that sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.
 
@@
 %
-Researchers often take multiple z-slices to visualize the three-dimensional structure of a sample;
-with advances in imaging techniques, such three-dimensional data will be more common, and it is important to validtate that our algorithm can be applied to such data.
+Researchers often take multiple z-slices to visualize the three-dimensional structure of a sample.
+%
+With advances in imaging techniques, such three-dimensional data will be more common, and it is important to validate that our algorithm can be applied to such data.
 
@@
 %
-The data set is composed of 46 fixed wing discs, whose developmental time range from 72 to 112~hours after fertilization. 
+The data set is composed of 46 fixed wing discs whose developmental times range from 72 to 112~hours after fertilization. 
 %
@@
 %
-However, ordering data via diffusion maps is readily possible for (registered) three-dimensional data. 
+However, ordering data via diffusion maps is readily possible for registered three-dimensional data. 
 %
-Therefore, we can apply our methodology to three-dimensional data by first registering the data using an alternative methodology (such as registering maximum intensity projection, or aligning a single feature of the data), and then ordering the registered data using diffusion maps.
+Therefore, we can apply our methodology to three-dimensional data by first registering the data using an alternative approach, and then ordering the registered data using diffusion maps. 
 %
-Instead, we used maximum intensity projection and converted the image into 2D, and then used the angular synchronization on 2D data. 
-%
-For the wing disc data, we first registered the three-dimensional data by computing the optimal rotations to align the maximum intensity projections of the data using vector diffusion maps.
+For the wing disc data, we registered the three-dimensional data by computing the optimal rotations to align the maximum intensity projections of the data using vector diffusion maps.
 %
@@
 %
-Other options for registering samples include using a specific localized signal (such as Twi in \fig~\ref{fig:drosophila_fixed_images}) as a marker, or using other {\em a priori} knowledge.
+Other options for registering samples include using a specific localized signal (such as Twi in \fig~\ref{fig:drosophila_fixed_images}) as a marker for alignment, or using other {\em a priori} knowledge.
 %
-We then used diffusion maps to order the registered three-dimensional images. 
+We then used diffusion maps to order the registered three-dimensional data. 
 %
@@
 %
-Furthermore, we can extract a stereotypical developmental trajectory by averaging; this trajectory is shown in \fig~\ref{fig:wing_disc}.
+There are some registration errors in the early images; these are primarily due to some wing discs having extra tissue attached to them. 
+%
+Even with such obstructions, we can order the images with high accuracy. 
+%
+Furthermore, we can extract a stereotypical developmental trajectory, shown in \fig~\ref{fig:wing_disc}C, by averaging (see \SI).
 %
@@
 \caption{Computational requirements for the presented methodology.
-(A) CPU time as a function of the number of images in the data set (for $100 \times 100$ pixel images, and $10^\circ$ rotational discretizations).
+(A) CPU time as a function of the number of images in the data set (for $100 \times 100$ pixel images, and $10^\circ$ angular discretization).
 Empirically, the CPU time is $\sim \mathcal{O}(n^{1.33})$ in number of images.
-(B) CPU time as a function of the number of pixels in the images (for $120$ images, and $10^\circ$ rotational discretizations).
+(B) CPU time as a function of the number of pixels in the images (for $120$ images, and $10^\circ$ angular discretization).
 Empirically, the CPU time is $\sim \mathcal{O}(n^{1.83})$ in number of pixels.
@@
 
-Manually ordering the images is nontrivial for researchers who are unfamiliar with the developmental progression, and can be tedious and time-consuming for those who are.
+Manually ordering the images is nontrivial for researchers who are unfamiliar with the relevant developmental progression, and can be tedious and time-consuming for those who are.
 %
@@
 %
-We found that the results are robust to both of these parameters, and that consistent parameter values served well for all data sets presented here. 
+We found that the results are robust to both of these parameters.
 %
@@
 %
-We have applied our methodology to several different proprotypical data sets in developmental biology; because our methodology only relies on computing distances between images and may be readily extended to many other data sets.
+We have applied our methodology to several different proprotypical data sets in developmental biology; because our methodology only relies on computing distances between images,, we are confident it may be readily extended to many other data sets.
 
@@
 %
-Second, the developmental features need to be large compared to the noise and the image resolution. 
+Second, the relevant developmental patterns and structures need to be large compared to the noise and the image resolution. 
 %
@@
 %
-For other imaging data sets which contain very fine scale structures, our algorithms will fail as small perturbations in these fine-scale structures will result in large changes in the Euclidean distances.
+Our algorithms will fail for other imaging data sets which contain very fine scale structures,  as small perturbations in these fine-scale structures will result in large changes in the Euclidean distances.
 %
-Furthermore, as we have acknowledged, our methods register full three-dimensional data with respect to rotations due to computational limitations. 
+In such cases, one requires a more sophisticated distance metric, such as the Earth Mover's Distance \citep{...}, or image description, such as scattering transform coefficients \citep{...}, which are insensitive to such perturbations.
+%
+Furthermore, as we have acknowledged, our methods as implemented cannot register full three-dimensional data with respect to rotations due to computational limitations. 
 %
 However, the algorithms presented here could be used for three-dimensional given an alternative method for computing pairwise alignments more efficiently (such as using image features or landmarks).
-%
+
 Unlike traditional statistical methods, our algorithms do not provide confidence intervals on predicted orderings, but could in principle be extracted from bootstrapping \citep{efron1979bootstrap}.

### 2015-02-01 image_analysis_paper_development2/SI_text.tex
updated paper

@@
 
-Before applying vector diffusion maps, images must be preprocessed so that the Euclidean distance between time image pixels is informative.
+Before applying vector diffusion maps, images must be preprocessed so that the Euclidean distance between the image pixels is informative: we want images who are developmentally similar to have relatively small Euclidean distances, and images who are at disparate developmental time points to have relatively large distances.
 %
-We want images who are developmentally similar to have relatively small Euclidean distances, and images who are at disparate developmental time points to have relatively large distances. 
-%
-We therefore must adjust images to remove any experimental and imaging artifacts, so that most of the variation is relevant to the developmental dynamics.
+We must preprocess images to remove any experimental and imaging artifacts, so that most of the variation is relevant to the developmental dynamics.
 
-The relevant image operations are 
+For our general purposes, the relevant image operations are 
 \begin{enumerate}
@@
 %
-\item {\bf Blur:} The \texttt{imfilter} function with a disc filter is used to blur signals whose small-scale structure is not informative. 
+\item {\bf Blur:} The \texttt{imfilter} function with a disc filter is used to blur signals whose small-scale structure is not informative. For example, 
 %
@@
 %
-$\epsilon$ can be chosen using several techniques (see, for example \citep{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon^2$ to be $1/10^{th}$ of the median of the squared pairwise distances.
+$\epsilon$ can be chosen using several techniques (see, for example \citep{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/4-1/2$ of the median of the pairwise distances.
 
@@
 %
-The domain of the image (a square) is not invariant to our rotations; however, the pixels near the corners of the square have zero intensity, and so the norm can be meaningfully computed as long as the main image does not ``move out of'' the original square.
+The domain of the image (a square) is not invariant to our rotations; however, the pixels near the corners of the square are preprocessed to have zero intensity, and so the norm can be meaningfully computed as long as the main image does not ``move out of'' the original square.
 %
@@
 %
-For the images in \fig~4, we take $\sigma = 3$.
+For the images in \fig~5~and~6, we take $\sigma = 2$.
 %

### 2015-02-01 image_analysis_paper_development2/code_for_distribution/README/README.tex
updated paper

@@
 \usepackage{epstopdf}
+
+\newcommand{\SI}[0]{\textit{SI Materials and Methods}}
 
@@
    \item (Optional) Save registered and ordered images
+   \item (Optional) Compute an average trajectory
 \end{enumerate}
@@
 %
-\item[2D/3D] Select ``2D'' if your images are traditional 2D images. Select ``3D'' if your images are z-stacks.
+\item[2D/3D] Select ``2D'' if the images are traditional 2D images. Select ``3D'' if the images are z-stacks.
 %
-\item[Image Directory] Enter the directory where your images or z-stacks are stored.
+\item[Image Directory] Enter the directory where the images or z-stacks are stored.
 This directory can be an absolute path (e.g.
 \texttt{C:/Documents/Images}) or a path relative to the directory where this code is stored (e.g.
-\texttt{../images})
+\texttt{../Images})
 %
@@
 %
-\item[Stack Prefix] (only for 3D z-stacks) The prefix for all stack folder names. Each stack is assumed to be stored in a single folder, named as the stack prefix, followed by a number. 
+\item[Stack Prefix] (only for 3D z-stacks) The prefix for all stack folder names.
+Each stack is assumed to be stored in a single folder, with the folder's name as the stack prefix, followed by a number.
 %
@@
 Images are assumed to be stored as the image prefix, followed by a two-digit index, where the indices run from 1 to the number of images (e.g., image\_prefix01.tif, image\_prefix02.tif, etc.).
-For three-dimensional images, the z-stacks are assumed to be stored in individual folders whose names are the stack prefix, followed by a two-digit index, where the indices run from 1 to the number of stacks (e.g., z\_stack\_prefix01, z\_stack\_prefix02, etc.)
+For three-dimensional images, the z-stacks are assumed to be stored in individual folders whose names are the stack prefix, followed by a two-digit index, where the indices run from 1 to the number of stacks (e.g., z\_stack\_prefix01/, z\_stack\_prefix02/, etc.)
 %
@@
 
-After populating the required fields, select the ``Read Images'' button to read in the images (a wait par will appear to show the progress of image reading).
+After populating the required fields, select the ``Read Images'' button to read in the images.
 %
-After reading the images, select the ``Show Images'' button to display the images (z-stacks will be displayed as maximum intensity projections). This step is optional.
+After reading the images, (optional) select the ``Show Images'' button to display the images (z-stacks will be displayed as maximum intensity projections).
 
@@
 \item[Blur] Each channel in the images is blurred using a disc factor with radius equal to a fraction of the image as indicated by the ``blur'' field.
-For example, setting the blur to 5\% means that the image is blurred using a disc filter with radius $0.5 \times$ the number of pixels.
+For example, setting the blur to 5\% means that the image is blurred using a disc filter with radius $0.05 \times$ the number of pixels.
 %
@@
 
-After populating the required fields, select the ``Apply Image'' button to apply the selected preprocessing options (a wait bar will appear to show the progress).
+After populating the required fields, select the ``Apply Image'' button to apply the selected preprocessing options.
 %
-After preprocessing the images, select the ``Show Images'' button to display the preprocessed images (z-stacks will be displayed as maximum intensity projections). This step is optional.
+After preprocessing the images, (optional) select the ``Show Images'' button to display the preprocessed images (z-stacks will be displayed as maximum intensity projections).
 
@@
 %
-\item[Kernel Scale] This is the scaling of the diffusion maps kernel.
-This parameter should be $\mathcal{O}(1)$, and we have empirically found for our data sets that values of $0.25-0.5$ often yield good results.
+\item[Kernel Scale (*median)] This is the scaling of the diffusion maps kernel, in multiples of the median of the pairwise distances (so that $\epsilon$ in the diffusion maps algorithm, as outlined in \SI, is this value times the median of the pairwise distances).
+This parameter should be $\mathcal{O}(1)$; empirically, we have found that values of $0.25-0.5$ often yield good results for our data sets.
+%
+\item[Apply to...]  This determines which images are shown and/or saved as the analyzed images.
+Select the ``Apply to Raw Images'' option to apply the computed registration and/or ordering to the original images; select the ``Apply to Preprocessed Images'' option to apply the computed registration and/or ordering to the preprocessed images (after blurring, normalization, etc.)
 %
 \end{description}
-
-After this, select the ``Apply to Raw Images'' option to apply the computed registration and/or ordering to the original images; select the ``Apply to Preprocessed Images'' option to apply the computed registration and/or ordering to the preprocessed images (after blurring, normalization, etc.)
 
@@
 {\em Important:} This option assumes that the imaging data set is preregistered!
-This option can be used for imaging data sets that were preregistered using some {\em a priori} knowledge about the system.
+This option is intended for imaging data sets that were preregistered using some {\em a priori} knowledge about the system.
 %
@@
 
-After analysis select the ``Show Images'' button to display the registered and/or images (z-stacks will be displayed as maximum intensity projections). This step is optional.
+(Optional) After analysis select the ``Show Images'' button to display the registered and/or images (z-stacks will be displayed as maximum intensity projections).
 
@@
 %
-The ``Output Image Directory'' field is where the analyzed images will be stored; this directory will be created if it does not exist. 
+The ``Output Image Directory'' is where the analyzed images will be stored; this directory will be created if it does not exist. 
 %
-The images will be stored in the output directory; the name format will be the same as in the input directory.
+The images will be stored in the output directory, and the name format will be the same as in the input directory.
 
@@
 \item[Width of Averaging Window] This is the (approximate) width of the averaging window.
-Each snapshot in the average trajectory will be the (weighted) average of approximately the width number of images.
-Increasing this value will result in a smoother average trajectory; decreasing this value will reveal more finer-scale detail.
+Each snapshot in the average trajectory will be the (weighted) average of approximately the width number of images (see \SI).
+Increasing this value will result in a smoother average trajectory, while decreasing this value will retain more finer-scale detail.
 % 

### 2015-02-01 image_analysis_paper_development2/manuscript.tex
updated paper

@@
 %
-Instead, we will use diffusion maps to automatically uncover a parameterization of our high-dimensional data (see \SI).
+Instead, we will use diffusion maps to automatically uncover a parameterization of our high-dimensional data from the eigenvectors of the appropriate matrix (see \SI).
+%
+Furthermore, the corresponding eigenvalues will allow us to confirm our assumption that our data lie on a one-dimensional manifold. 
 %
@@
 %
-Determining which ``end'' of the trajectory is early and which is late is a post-processing task which requires {\em a priori} knowledge of the developmental dynamics. 
+Determining the trajectory rate and direction are post-processing tasks which requires {\em a priori} knowledge of the developmental dynamics. 
 %

### 2015-02-02 image_analysis_paper_development2/SI_text.tex
updated report

@@
 For our general purposes, the relevant image operations are 
-\begin{enumerate}
-\item {\bf Intensity normalization:} Contrast-limited adaptive histogram equalization (using the \texttt{adapthisteq} function with an $8 \times 8$ tile grid, and a uniform distribution for the intensities with a clip limit of 0.01) is used to normalize the intensities of signals whose absolute intensity is not meaningful/informative.
+\begin{description}
+
+\item[Intensity normalization] Contrast-limited adaptive histogram equalization (using the \texttt{adapthisteq} function with an $8 \times 8$ tile grid, and a uniform distribution for the intensities with a clip limit of 0.01) is used to normalize the intensities of signals whose absolute intensity is not meaningful/informative.
 %
-\item {\bf Blur:} The \texttt{imfilter} function with a disc filter is used to blur signals whose small-scale structure is not informative. For example, 
+\item[Blur] The \texttt{imfilter} function with a disc filter is used to blur signals whose small-scale structure is not informative. For example, 
 %
-\item {\bf Intensity scaling (multichannel images only):} The \texttt{immultiply} function is used to scale signal intensities.
+\item[Intensity scaling (multichannel images only)] The \texttt{immultiply} function is used to scale signal intensities.
 This is important for multichannel images, and it determines the (relative) contributions of the various signals.
 %
-\item {\bf Mean-centering:}  The Canny method \citep{canny1986computational} is used to detect the edges of the object in each image (using the \texttt{edge} function).
+\item[Mean-centering]  The Canny method \citep{canny1986computational} is used to detect the edges of the object in each image (using the \texttt{edge} function).
 The image is then translated so that the object (as determined by extremeties of the detected edges) is centered.
 %
-\item {\bf Size scaling:} For images whose relative size is unimportant to the developmental dynamics, the images are rescaled/dialated so that the object size (as determined by extremeties of the detected edges) is constant (we set this to be 80\% of the total image).
+\item[Size scaling] For images whose relative size is unimportant to the developmental dynamics, the images are rescaled/dialated so that the object size (as determined by extremeties of the detected edges) is constant (we set this to be 80\% of the total image).
 %
-\item {\bf Removing corners:} The image is cropped/filtered using a disc centered in the middle of the image and whose diameter is equal to the number of pixels; any pixels outside of this disc are set to 0 (no intensity).
+\item[Removing corners] The image is cropped/filtered using a disc centered in the middle of the image and whose diameter is equal to the number of pixels; any pixels outside of this disc are set to 0 (no intensity).
 This removes any corner effects under rotations.
-\end{enumerate}
+\end{description}
 
@@
 %
-Taking $\epsilon$ to be on the order of the median of the pairwise distances often yields good results. 
+Taking $\epsilon$ to be $25-50\%$ of the median of the pairwise distances often yields good results. 
 \item Compute the matrix $S$ in \eqref{eq:vdm_S}. 

### 2015-02-02 image_analysis_paper_development2/code_for_distribution/README/README.tex
updated report

@@
 
-\usepackage{epstopdf}
+%\usepackage{epstopdf}
 
@@
 
-All algorithms and analysis were implemented in MATLAB\textsuperscript{\textregistered} (R2013b, The MathWorks, Natick, Massachusetts).
+All algorithms and analysis are implemented in MATLAB\textsuperscript{\textregistered} (R2013b, The MathWorks, Natick, Massachusetts).
 %
@@
 
-There are two ways to use the code
-\begin{enumerate}
-\item Interactive GUI.
-\item Program a simple script.
-\end{enumerate}
-
+The code can be used either via an interactive GUI
+or by writing a simple script.
+%
 The basic steps are as follows:
@@
 
+Open ``register\_order\_gui.fig'', which is available in the downloaded code directory.
+%
+A window should appear (see below)
+
 \includegraphics[width=\textwidth, trim=0cm 3cm 12cm 0cm, clip]{gui_screenshot_initial.jpg}
-
-There are five steps
-%
-\begin{enumerate}
-\item Read Images
-\item Preprocess Images
-\item Analyze Images
-\item Save Images
-\item Average Trajectory
-\end{enumerate}
 
@@
 
+First populate the relevant parameters for where the images are stored.
+%
 The fields are as follows
@@
 
+Define the relevant preprocessing operations. 
+%
+The options are as follows
+%
 \begin{description}
@@
 
-\subsection*{Read Images}
+First, make sure that the directory which stores the downloaded code is on your MATLAB path, or that you are in that directory in MATLAB.
+
+\subsubsection{Read Images}
 
 \begin{par}
-The first step is to read in images.
-This is done using the \texttt{read\_images} function For two-dimensional images, it requires that all the images are stored in a single directory, all with a consistent prefix, and indexed with two digits starting from \texttt{01}.
+The first step is to read in images, using the \texttt{read\_images} function.
+For two-dimensional images, it requires that all the images are stored in a single directory, all with a consistent prefix, and indexed with two digits starting from \texttt{01}.
 For three-dimensional z-stack, it assumes each z-stack is stored in a single directory.
@@
 
-% number of pixels
-npixels = 100;
-
 % dimension of images (dim=2 indicates standard 2D images, rather than
@@
 % images are stored in the variable images_raw
+% nchannels stores the number of channels per image
 [images_raw, nchannels] = read_images(image_dir, image_name, image_ext, ...
-    stack_name, nimages, nstack, npixels, dim);
+    stack_name, nimages, nstack, dim);
 \end{verbatim}
 
-\subsection*{Show images}
+\subsubsection{Show images}
 
@@
 
-\includegraphics [width=4in]{sample_input_file_01.eps}
+\includegraphics [width=4in]{sample_input_file_01.jpg}
 
-\subsection*{Preprocess Images}
+\subsubsection{Preprocess Images}
 
 \begin{par}
-Now, we must preprocess the images before registration and ordering, to remove any imaging and/or experimental artifacts.
-We do this via the \texttt{apply\_image\_functions} function For this particular imaging data set, there are three channels.
-DAPI, a nuclear stain, is in the first (red) channel.
-dpERK is in the second (green) channel.
-Dl is in the third (blue) channel.
+Now, we must preprocess the images using the \texttt{apply\_image\_functions} function before registration and ordering, to remove any imaging and/or experimental artifacts.
 \end{par} \vspace{1em}
 \begin{verbatim}
+% number of pixels
+% images will be reduced to npixels x npixels resolution
+npixels = 100;
+
 % channel weights
-% because the nuclear signal is wider spread throughout the image, as well
-% as noiser, we choose to scale the first channel by half
+% we scale the first (red) channel by half, and keep the second (green) and
+% third (blue) channels at their input values
 channel_weight = [0.5 1 1];
 
-% we blur the image slightly (5%) so that small-scale effects of individual
-% nuclei are smoothened
+% channel blur
+% we blur each of the channels by 5%
 channel_blur = [0.05 0.05 0.05];
 
-% because the absolute intensity of the nuclear signal is not informative
-% or important (only the presence or absence of signal is important,
-% indicating the presence or absence of nuclei), we normalize the first channel
-% We do not normalize the other two channels, as changes in intensity are
-% informative to the developmental processes.
+% channel normalization
+% we normalize the first (red) channel using histogram equalization
+% we do not normalize the second (green) or third (blue) channels
 channel_normalize = [1 0 0];
 
-% we use the nuclear channel to mean center the images, as this channel
-% parameterizes/captures the entire embryo
+% mean-center
+% we use the first (red) channel to detect the edges of the object in order
+% to mean center the object
 channel_mean_center = [1 0 0];
 
-% we resize the images to all be (approximately) the same size, to remove
-% any variations due to size effects, as size effects are known to not be
-% important for this developmental system
+% resize
+% we choose to resize the images so all objects are (approximately)
+% the same size, to remove any variations due to size effects
 resize_image = true;
@@
 % reweighting, and mean-centering
-images = apply_image_functions(images_raw, dim, channel_weight, ...
+images = apply_image_functions(images_raw, npixels, dim, channel_weight, ...
     channel_blur, channel_normalize, channel_mean_center, resize_image);
@@
 
-\includegraphics [width=4in]{sample_input_file_02.eps}
+\includegraphics [width=4in]{sample_input_file_02.jpg}
 
-\subsection*{Calculate pairwise alignments}
+\subsubsection{Calculate pairwise alignments}
 
@@
 
-\subsection*{Apply vector diffusion maps}
+\subsubsection{Apply vector diffusion maps}
 
@@
 
-\includegraphics [width=4in]{sample_input_file_03.eps}
+\includegraphics [width=4in]{sample_input_file_03.jpg}
+
+\subsubsection{Calculate average trajectory}
+
+\begin{par}
+We can calculate an average trajectory from our set of (registered and ordered) images
+\end{par} \vspace{1em}
+\begin{verbatim}
+% nsubimages is the desired number of images in the average trajectory
+nsubimages = 10;
+
+% avg_width is the (approximate) width of the averaging window used to
+% compute each of the images in the average trajectory
+avg_width = 4;
+
+% compute the average trajectory
+avg_images = compute_average_trajectory(images_analyzed, nsubimages, avg_width);
+
+% plot the images (optional)
+plot_images(avg_images, dim)
+\end{verbatim}
+
+\includegraphics [width=4in]{sample_input_file_04.jpg}
+
+\section{Parameters for Data Sets}
+
+The four data sets used in the paper are available with the code in the relevant subdirectories. The parameters used (as defined in the script in the previous section) are defined below.
+
+\subsection{{\em Drosophila} live imaging}
+
+\begin{verbatim}
+
+image_dir = 'drosophila_live_imaging/expt01';
+image_name = 'image';
+image_ext = 'tif';
+stack_name = '';
+nimages = 40;
+nstack = 0;
+dim = 2;
+
+npixels = 100;
+channel_weight = 1;
+channel_blur = 0.05;
+channel_normalize = 1;
+channel_mean_center = 0;
+resize_image = false;
+
+ang_dis = 10;
+
+eps_scale = 0.25;
+
+\end{verbatim}
+
+\subsection{Zebrafish}
+
+\begin{verbatim}
+
+image_dir = 'zebrafish';
+image_name = 'image';
+image_ext = 'tif';
+stack_name = '';
+nimages = 120;
+nstack = 0;
+dim = 2;
+
+npixels = 100;
+channel_weight = 1;
+channel_blur = 0;
+channel_normalize = 0;
+channel_mean_center = 1;
+resize_image = false;
+
+ang_dis = 10;
+
+eps_scale = 0.25;
+
+\end{verbatim}
+
+\subsection{{\em Drosophila} fixed images}
+
+\begin{verbatim}
+
+image_dir = 'drosophila_fixed_images';
+image_name = 'emb';
+image_ext = 'tif';
+stack_name = '';
+nimages = 120;
+nstack = 0;
+dim = 2;
+
+npixels = 100;
+channel_weight = [0.5 1 1];
+channel_blur = [0.05 0.05 0.05];
+channel_normalize = [1 0 0];
+channel_mean_center = [1 0 0];
+resize_image = true;
+
+ang_dis = 10;
+
+eps_scale = 0.25;
+
+\end{verbatim}
+
+\subsection{Wing disc}
+
+\begin{verbatim}
+
+image_dir = 'wing_disc_stacks';
+image_name = 'disc';
+image_ext = 'tif';
+stack_name = 'z';
+nimages = 46;
+nstack = 21;
+dim = 3;
+
+npixels = 100;
+channel_weight = [1 1 1];
+channel_blur = [0 0 0];
+channel_normalize = [0 0 0];
+channel_mean_center = [0 0 1];
+resize_image = false;
+
+ang_dis = 10;
+
+eps_scale = 0.5;
+\end{verbatim}
+
+\section{Parameter Values and Guidance}
+
+\begin{description}
+
+\item[Weight] The relative weights of the channels in multichannel images in important for computing meaningful distances for diffusion maps.
+The total intensity for each channel should be of the same order if all channels are to be considered as ``equally important.'' If the channels have dramatically different intensities in the raw images, they should be adjusted via the weight functions.
+%
+\item[Blur] The image should be blurred to remove fine-scale noise and variations.
+Typically, blur of 5\%-10\% are appropriate for removing such variability.
+In some cases, blurring is not necessary to recover an accurate registration and ordering.
+%
+\item[Normalize] Channels should be normalized if the absolute intensity of a specific signal is not meaningful.
+For example, DAPI, a DNA marker, identifies the nuclei in the {\em Drosophila} embryo, but the absolute intensity of the signal is not important.
+%
+\item[Use to Mean-Center] Objects should be (approximately) mean-centered in the frame prior to registration.
+The channel(s) that best identify/delineate the periphery of the object should be selected for use in the mean-centering procedure.
+%
+\item[Resize Images] Object should be resized if size differences are not meaningful/informative about the underlying developmental dynamics. 
+%
+\item[Number of Pixels] The image resolution should be chosen as low as possible such that the relevant developmental structures and patterns are still discernible/evident.
+A higher image resolution will result in a higher computational cost.
+%
+\item[Kernel Scale] The kernel scale defines how similar two images have to be to be considered developmentally close.
+Small values of the kernel will preserve finer scale patternds and structure in the ordering algorihtm, but will also be more sensitive to noise.
+We have found values of 0.25-0.5 to often be suitable for our studies.
+Noisy data with a lower temporal resolution (such as the wing disc data) often requires a larger value of the kernel scale than clean data with a high temporal resolution (such as the zebrafish live imaging data).
+%
+\item[Angular Discretization] The angular discretization sets the resolution for computing the pairwise alignments.
+For example, if the angular discretization is 10$^\circ$, then the pairwise alignments are computed by computing the image discrepancy after 10$^\circ$ rotation, after a 20$^\circ$ rotation, ..., up to a 360$^\circ$ rotation, and then selecting the rotation which minimizes the discrepancy.
+Lower values of the angular discretization increase the accuracy of the method, but also increase the computational cost.
+We have found that $10^\circ$ is often a good value of the angular discretization, which balances computational cost and accuracy.
+
+\end{description}
 

### 2015-02-02 image_analysis_paper_development2/code_for_distribution/README/sample_input_file.tex
updated report

@@
    \item Apply vector diffusion maps
+   \item Calculate average trajectory
 \end{itemize}
@@
 % images are stored in the variable images_raw
+% nchannels stores the number of channels per image
 [images_raw, nchannels] = read_images(image_dir, image_name, image_ext, ...
-    stack_name, nimages, nstack, npixels, dim);
+    stack_name, nimages, nstack, dim);
 \end{verbatim}
@@
 
-\includegraphics [width=4in]{sample_input_file_01.eps}
+\includegraphics [width=4in]{sample_input_file_01.jpg}
 
@@
 \begin{par}
-Now, we must preprocess the images before registration and ordering, to remove any imaging and/or experimental artifacts.
-We do this via the \texttt{apply\_image\_functions} function For this particular imaging data set, there are three channels.
-DAPI, a nuclear stain, is in the first (red) channel.
-dpERK is in the second (green) channel.
-Dl is in the third (blue) channel.
+Now, we must preprocess the images using the \texttt{apply\_image\_functions} function before registration and ordering, to remove any imaging and/or experimental artifacts.
 \end{par} \vspace{1em}
 \begin{verbatim}
+% number of pixels
+% images will be reduced to npixels x npixels resolution
+npixels = 100;
+
 % channel weights
-% because the nuclear signal is wider spread throughout the image, as well
-% as noiser, we choose to scale the first channel by half
+% we scale the first (red) channel by half, and keep the second (green) and
+% third (blue) channels at their input values
 channel_weight = [0.5 1 1];
 
-% we blur the image slightly (5%) so that small-scale effects of individual
-% nuclei are smoothened
+% channel blur
+% we blur each of the channels by 5%
 channel_blur = [0.05 0.05 0.05];
 
-% because the absolute intensity of the nuclear signal is not informative
-% or important (only the presence or absence of signal is important,
-% indicating the presence or absence of nuclei), we normalize the first channel
-% We do not normalize the other two channels, as changes in intensity are
-% informative to the developmental processes.
+% channel normalization
+% we normalize the first (red) channel using histogram equalization
+% we do not normalize the second (green) or third (blue) channels
 channel_normalize = [1 0 0];
 
-% we use the nuclear channel to mean center the images, as this channel
-% parameterizes/captures the entire embryo
+% mean-center
+% we use the first (red) channel to detect the edges of the object in order
+% to mean center the object
 channel_mean_center = [1 0 0];
 
-% we resize the images to all be (approximately) the same size, to remove
-% any variations due to size effects, as size effects are known to not be
-% important for this developmental system
+% resize
+% we choose to resize the images so all objects are (approximately)
+% the same size, to remove any variations due to size effects
 resize_image = true;
@@
 
-\includegraphics [width=4in]{sample_input_file_02.eps}
+\includegraphics [width=4in]{sample_input_file_02.jpg}
 
@@
 
-\includegraphics [width=4in]{sample_input_file_03.eps}
+\includegraphics [width=4in]{sample_input_file_03.jpg}
+
+\subsection*{Calculate average trajectory}
+
+\begin{par}
+We can calculate an average trajectory from our set of (registered and ordered) images
+\end{par} \vspace{1em}
+\begin{verbatim}
+% nsubimages is the desired number of images in the average trajectory
+nsubimages = 10;
+
+% avg_width is the (approximate) width of the averaging window used to
+% compute each of the images in the average trajectory
+avg_width = 4;
+
+% compute the average trajectory
+avg_images = compute_average_trajectory(images_analyzed, nsubimages, avg_width);
+
+% plot the images (optional)
+plot_images(avg_images, dim)
+\end{verbatim}
+
+\includegraphics [width=4in]{sample_input_file_04.jpg}
 

### 2015-02-02 image_analysis_paper_development2/manuscript.tex
updated report

@@
 
-\subsubsection*{Fixed {\em Drosophila} images}
+\subsubsection*{Fixed images of {\em Drosophila} gastrulation}
 
@@
 
-\subsubsection*{Wing disc data}
+\subsubsection*{Fixed z-stacks of {\em Drosophila} wing discs}
 
@@
 %
-At the same time, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \citep{yuan2014automated, surkova2008characterization}.
+At the same time, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \citep{yuan2014automated, surkova2008characterization}, or using some {\em a priori} knowledge of the relevant developmental processes \citep{dubuis2013accurate}.
 %
@@
 %
-In such cases, one requires a more sophisticated distance metric, such as the Earth Mover's Distance \citep{...}, or image description, such as scattering transform coefficients \citep{...}, which are insensitive to such perturbations.
+In such cases, one requires a more sophisticated distance metric, such as the Earth Mover's Distance \citep{rubner2000earth}, or image description, such as scattering transform coefficients \citep{mallat2012group}, which are insensitive to such perturbations.
 %
@@
 
-\subsection*{{\em Drosophila} Experiments}
+\subsection*{{\em Drosophila} experiments}
 %
@@
 
-\subsection*{{\em Drosophila} Microscopy}
+\subsection*{{\em Drosophila} microscopy}
 %
@@
 
-\subsection*{Software}
+\subsection*{Software and imaging data}
 %
@@
 %
-Code is available at \texttt{genomics.princeton.edu/stas/publications.html} under ``Codes and Data''.
+Software, including documentation and tutorials, along with the full imaging data sets used in this paper are available at \texttt{genomics.princeton.edu/stas/publications.html} under ``Codes and Data''.
 

### 2015-02-02 image_analysis_paper_development2/paper_figures/make_fig_drosophila_live_imaging.m

@@ -49,6 +49,13 @@ for i=1:nexpt
     ncomps = 1;
     [R_opt, embed_coord, D2] = vdm(R, W, eps_scale, ncomps);
     
+    fid = fopen(sprintf('%s/times.txt', image_dir), 'r');
+    time = fscanf(fid, '%f');
+    fclose(fid);
+    if corr(time, embed_coord) < 0
+            embed_coord = -embed_coord;
+    end
+        
     % register images using optimal rotations
     images_registered = register_all_images(images, R_opt);
     
@@ -56,9 +63,7 @@ for i=1:nexpt
     
     %         plot_images(images_analyzed, dim)
     
-    fid = fopen(sprintf('%s/times.txt', image_dir), 'r');
-    time = fscanf(fid, '%f');
-    fclose(fid);
+    
     
     %     figure;
     %     plot(tiedrank(time), tiedrank(embed_coord),'.')
@@ -82,9 +87,7 @@ for i=1:nexpt
     if i == 1
         nsubimages = 12;
         fontsize = 8;
-        if corr(time, embed_coord) < 0
-            embed_coord = -embed_coord;
-        end
+        
         [~, I] = sort(embed_coord(1:nsubimages));
         make_fig(17, 17/nsubimages);
         for j=1:nsubimages

### 2015-02-02 image_analysis_paper_development2/paper_figures/make_fig_wing_disc.m

@@ -34,14 +34,6 @@ plot_images(images, dim)
 ncomps = 1;
 [R_opt, embed_coord, D2] = vdm(R, W, eps_scale, ncomps);
 
-fid = fopen(sprintf('%s/times.txt', image_dir), 'r');
-time = fscanf(fid, '%f');
-fclose(fid);
-
-if corr(time, embed_coord) < 0
-    embed_coord = -embed_coord;
-end
-
 % register images using optimal rotations
 images_registered = register_all_images(images, R_opt);
 images_registered = imrotate(images_registered, 90);
@@ -50,6 +42,14 @@ plot_images(images_registered, dim)
 
 W = squareform(pdist(reshape(double(images_registered), [], nimages)')).^2;
 [embed_coord, D2] = dm(W, eps_scale, ncomps);
+
+fid = fopen(sprintf('%s/times.txt', image_dir), 'r');
+time = fscanf(fid, '%f');
+fclose(fid);
+
+if corr(time, embed_coord) < 0
+    embed_coord = -embed_coord;
+end
 
 images_analyzed = order_all_images(images_registered, embed_coord);
 

### 2015-02-02 image_analysis_paper_development2/paper_figures/make_fig_zebrafish.m

@@ -38,6 +38,15 @@ plot_images(images, dim)
 ncomps = 1;
 [R_opt, embed_coord, D2] = vdm(R, W, eps_scale, ncomps);
 
+fid = fopen(sprintf('%s/times.txt', image_dir), 'r');
+time = fscanf(fid, '%f');
+fclose(fid);
+if corr(time, embed_coord) < 0
+    embed_coord = -embed_coord;
+end
+
+corr(time, embed_coord,'type','spearman')
+
 % register images using optimal rotations
 images_registered = register_all_images(images, R_opt);
 
@@ -45,14 +54,11 @@ images_analyzed = order_all_images(images_registered, embed_coord);
 
 plot_images(images_analyzed, dim)
 
-fid = fopen(sprintf('%s/times.txt', image_dir), 'r');
-time = fscanf(fid, '%f');
-fclose(fid);
+
 
 figure;
 plot(tiedrank(time), tiedrank(embed_coord),'.')
 
-corr(time, embed_coord,'type','spearman')
 
 fid = fopen(sprintf('%s/angles.txt', image_dir), 'r');
 angles = fscanf(fid, '%f');
@@ -64,9 +70,7 @@ angles_opt = R_to_theta(R_opt);
 %%
 
 nsubimages = 10;
-if corr(time, embed_coord) < 0
-    embed_coord = -embed_coord;
-end
+
 % [~, I] = sort(embed_coord((1:nsubimages)+10));
 % I = I+10;
 [~, I] = sort(embed_coord);

### 2015-02-04 image_analysis_paper_development2/SI_text.tex
after yannis comments; before stas comments

@@
 
-Before applying vector diffusion maps, images must be preprocessed so that the Euclidean distance between the image pixels is informative: we want images who are developmentally similar to have relatively small Euclidean distances, and images who are at disparate developmental time points to have relatively large distances.
+Before applying vector diffusion maps, images must be preprocessed so that the Euclidean distance between the image pixels is informative: we need images who are developmentally similar to have relatively small Euclidean distances, and images who are at disparate developmental time points to have relatively large distances.
 %
@@
 
-For our general purposes, the relevant image operations are 
+The relevant image operations are listed below for our general purposes, along with the relevant MATLAB functions. 
 \begin{description}
@@
 %
-\item[Blur] The \texttt{imfilter} function with a disc filter is used to blur signals whose small-scale structure is not informative. For example, 
+\item[Blur] The \texttt{imfilter} function with a disc filter is used to blur signals whose small-scale structure is not informative. 
 %
 \item[Intensity scaling (multichannel images only)] The \texttt{immultiply} function is used to scale signal intensities.
-This is important for multichannel images, and it determines the (relative) contributions of the various signals.
+This is important for multichannel images, as it determines the (relative) contributions of the various signals.
 %
@@
 
-\subsubsection*{{\em Drosophila} live imaging}
+\subsubsection*{{\em Drosophila} gastrulation (live)}
 
 %
-The original image resolution is $512 \times 512$ for the live {\em Drosophila} images.
+The original image resolution is $512 \times 512$ for the live {\em Drosophila} embryo images.
 %
@@
 
-\subsubsection*{Zebrafish images}
+\subsubsection*{Zebrafish epiboly}
 
@@
 %
-All images were subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set. 
+A $16$-pixel border was removed from the left and right sides to make the images square, and all images were then subsampled to $100 \times 100$ pixels for analysis, as this was found to be a sufficient resolution to retain all of the major developmental features within the data set.
 %
@@
 %
-Images were not rescaled for size, as changes in overall were are important.
+Images were not rescaled for size, as changes in overall size are important.
 
-\subsubsection*{{\em Drosophila} fixed images}
+\subsubsection*{{\em Drosophila} gastrulation (fixed)}
 
@@
 %
-The nuclear channel was normalized, and all channels blurred with a filter of radius 5\% of the total image (5 pixels) to remove the effects of individual nuclei. 
+The nuclei channel was normalized, and all channels were blurred with a filter of radius 5\% of the total image (5 pixels) to remove the effects of individual nuclei. 
 %
-The nuclear channel was scaled by half; because this signal occupies a larger fraction of the image relative to the other signals, its overall contribution is large, and so we downscaled it so that each signal would have more comparable weight in the algorithm.
+The nuclei channel was scaled by half; because this signal occupies a larger fraction of the image relative to the other signals, its overall contribution is large, and so we downscaled it so that each signal would have more comparable weight in the algorithm.
 %
-The images were mean-centered and scaled to have a constant size using the nuclear signal to detect the edges of the embryo within the frame.
+The images were mean-centered and scaled to have a constant size using the nuclei signal to detect the edges of the embryo within the frame.
 
-\subsubsection*{Wing disc images}
+\subsubsection*{{\em Drosophila} wing discs}
 
@@
 %
-Images were not rescaled for size, as changes in overall were are important.
+Images were not rescaled for size, as changes in overall size are important.
 
@@
 %
-Motivated by the geometry of our images, we construct a sequence of concentration profiles defined on a ring, and rotate each ring randomly around its center; an example is shown in \fig~\ref{subfig:1d_example}.
+Motivated by the geometry of our {\em Drosophila} embryo images, we construct a sequence of concentration profiles defined on a ring, and rotate each ring randomly around its center; an example is shown in \fig~\ref{subfig:1d_example}.
 %
@@
 %
-$\epsilon$ can be chosen using several techniques (see, for example \citep{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/4-1/2$ of the median of the pairwise distances.
+$\epsilon$ can be chosen using several techniques (see, for example \citep{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/4$ of the median of the pairwise distances for the two-dimensional images, and $1/2$ of the median of the pairwise distances for the three-dimensional z-stacks.
 
@@
 %
-Image rotation is performed with the \texttt{imwarp} function in Matlab, using linear interpolation to estimate the pixel intensities after rotation.
+Image rotation is performed with the \texttt{imwarp} function in MATLAB, using linear interpolation to estimate the pixel intensities after rotation.
 %
@@
 
-%
-%\begin{figure*}
-%\includegraphics{figS3}
-%\caption{Eigenvalue spectra.
-{\it (A)} Eigenvalue spectrum for the live imaging data set presented in \fig~3.
-Note that there is a gap after the third eigenvalue.
-{\it (B), (C)} Second and third embedding coordinates, respectively, versus the first embedding coordinate for the live imaging data set presented in \fig~3.
-Note that coordinates 2 and 3 are higher harmonics (and thus simple functions) of coordinate 1, and are therefore not informative about structure in the data set.
-We can conclude that the data set is effectively one-dimensional.
-{\it (D)} Eigenvalue spectrum for the live imaging data set presented in \fig~4.
-Note that there is a gap after the first eigenvalue; we can conclude that the data set is effectively one-dimensional.
+\begin{figure}
+\textsf{
+\begin{tabular}{| r | c | c |}
+\hline
+Data set & \parbox[c]{4cm}{ \centering VDM Rank Correlation} & \parbox[c]{4cm}{ \centering PCA Rank Correlation} \\
+\hline
+{\em Drosophila} gastrulation (live) & 0.9989 & 0.8137 \\
+Zebrafish epiboly & 0.9955 & 0.6351 \\
+{\em Drosophila} gastrulation (fixed) & 0.9716 & 0.8658 \\
+{\em Drosophila} wing discs & 0.9436 & 0.9381 \\
+\hline
+\end{tabular}
 }
-%\end{figure*}
-%
-%\begin{figure*}
-%\includegraphics{figS4}
-%\caption{Live imaging data set.
-{\it (A)} Live imaging data set used for the results in \fig~3.
-The images have been preprocessed as described above, each image has been arbitrarily rotated, and the order of the images has been shuffled.
-{\it (B)} Data set from A after registration and ordering.
-The real time for each image is indicated.
+\caption{Comparison between rank correlation coefficients when ordering using the first vector diffusion maps (VDM) embedding coordinate, and rank correlation coefficients when ordering by the first principal component analysis (PCA) \citep{shlens2005tutorial} projection coefficient.
+For ordering using PCA, we computed the first principal component of the registered images, and then ordered the images by the projection coefficients onto this first mode.
+The PCA ordering is always less accurate than the vector diffusion maps ordering.
+In the zebrafish data set, the ordering is much less accurate using PCA, as the dynamics of the morphing and spreading of the cell mass are highly nonlinear.
+The PCA and VDM orderings are comparable for the wing disc data; this is to be expected, as the tissue simply grows in time.
 }
-%\end{figure*}
-%
-%\begin{figure*}
-%\includegraphics{figS5}
-%\caption{{\it (A-D)} Four additional live imaging data sets used for the analysis in \fig~3. }
-%\end{figure*}
-%
-%\begin{figure*}	
-%\includegraphics{figS6}
-%\caption{Fixed embryo imaging data set.
-{\it (A)} Fixed images used for the results in \fig~4.
-The images have been preprocessed as described above.
-{\it (B)} Data set from A after registration and ordering.
-The rank provided by a trained embryologist for each image is indicated.
-}
-%\end{figure*}
+\end{figure}
 

### 2015-02-04 image_analysis_paper_development2/code_for_distribution/README/README.tex
after yannis comments; before stas comments

@@
 The code reads in a set of two-dimensional images or three-dimensional z-stacks, can perform a number of preprocessing operations, and then registers and/or orders the images using the algorithms described in the paper.
+
+\section{License}
+
+This code is covered under the BSD License.\\
+Copyright (c) 2015, Carmeline Dsilva\\
+All rights reserved.
+
+Redistribution and use in source and binary forms, with or without
+modification, are permitted provided that the following conditions are
+met:
+
+\begin{enumerate}
+
+    \item Redistributions of source code must retain the above copyright
+      notice, this list of conditions and the following disclaimer.
+    \item Redistributions in binary form must reproduce the above copyright
+      notice, this list of conditions and the following disclaimer in
+      the documentation and/or other materials provided with the distribution
+    \item Neither the name of the Princeton University nor the names
+      of its contributors may be used to endorse or promote products derived
+      from this software without specific prior written permission.
+\end{enumerate}
+THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS ``AS IS''
+AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
+IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE
+ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT OWNER OR CONTRIBUTORS BE
+LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR
+CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF
+SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS
+INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN
+CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE)
+ARISING IN ANY WAY OUT OF THE USE OF THIS SOFTWARE, EVEN IF ADVISED OF THE
+POSSIBILITY OF SUCH DAMAGE.
 
@@
 \item[Register + Order z-stacks] (this option is only for three-dimensional z-stacks) This registers and orders the imaging data set.
-It first registers the z-stacks by computing the optimal rotations to align the (two-dimensional) maximum intensity projections using vector diffusion maps.
+It first registers the z-stacks by computing the optimal rotations to align the (two-dimensional) maximum intensity projections along the z-axis using angular synchronization.
 It then orders the data using diffusion maps on the (now registered) three-dimensional z-stacks.
@@
 
-\subsection{{\em Drosophila} live imaging}
+\subsection{{\em Drosophila} gastrulation (live)}
 
@@
 
-\subsection{Zebrafish}
+\subsection{Zebrafish epiboly}
 
@@
 
-\subsection{{\em Drosophila} fixed images}
+\subsection{{\em Drosophila} gastrulation (fixed)}
 
@@
 
-\subsection{Wing disc}
+\subsection{{\em Drosophila} wing discs}
 

### 2015-02-04 image_analysis_paper_development2/manuscript.tex
after yannis comments; before stas comments

@@
 \customlabel{fig:eigenvalues_wing_disc}{S6}
+\customlabel{fig:PCA_compare}{S7}
 
@@
 Bomyi~Lim \footnotemark[1],
+Ioannis~G.~Kevrekidis \footnotemark[1] \thanks{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA} ,
 Hang~Lu \thanks{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
-Amit~Singer \thanks{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \thanks{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA},
-Stanislav~Y.~Shvartsman \footnotemark[1] \thanks{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA}, \\
-and Ioannis~G.~Kevrekidis \footnotemark[1] \footnotemark[4]}
+Amit~Singer \thanks{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \footnotemark[2], \\ and
+Stanislav~Y.~Shvartsman \footnotemark[1] \thanks{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA}}
 
@@
 
-We present a robust algorithmic approach to registration and temporal ordering.
+We present a robust algorithmic approach to simultaneous registration and temporal ordering.
 %
-In contrast to a number of previous methodologies \citep{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, our methodology does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression for temporal ordering.
+In contrast to a number of previous methodologies \citep{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face, dubuis2013accurate}, our methodology does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression.
 %
-Our approach is based on a manifold learning algorithm (vector diffusion maps \citep{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
+The approach is based on a manifold learning algorithm (vector diffusion maps \citep{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
 %
@@
 %
-Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing embryos in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
+Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing tissues in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
 %
@@
 %
-Our final data set consists of z-stacks of wing discs, which we will use to illustrate how our methods extend to three-dimensional imaging. 
+Our final data set consists of z-stacks of wing discs, which we will use to illustrate how our methods can be used to analyze specific types of three-dimensional imaging data. 
 %
@@
 %
-We will use the algorithm to register images of developing embryos with respect to rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
+We will use the algorithm to register images of developing tissues with respect to rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
 %
-Although in general, images may contain variations due to rotations, translations, and scaling, we will remove relevant translations and/or scaling via relatively simple preprocessing, and focus only on factoring out rotations using the vector diffusion maps algorithm.
+Although in general, images may contain variations due to rotations, translations, and scaling, we will remove the relevant translational and/or scaling symmetries via relatively simple image preprocessing, and focus here only on factoring out rotations using the vector diffusion maps algorithm.
 %
-In the case that all relevant symmetries can be removed with straightforward preprocessing procedures, our algorithms can also be used simply to extract the main direction of variability within the imaging data set.
+In the case that {\em all} relevant symmetries can be removed with straightforward preprocessing procedures, our algorithms can also be used simply to extract the main direction of variability within the imaging data set.
 %
-We assume that the main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
+We assume that this main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
 %
@@
 (B) The vectors from A, each rotated about their midpoint so that the set is globally aligned.
-Note that the chosen rotation angles are consistent with the pairwise alignments in A: the differences between a pair of angles in B is the same as the pairwise angle in A.
+Note that the chosen rotation angles are consistent with the pairwise alignments in A: the difference between a pair of angles in B is the same as the pairwise angle in A.
 (C) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
-Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through a Gaussian kernel (see \SI), so that close data points are connected by darker (``stronger'') edges.
+Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through a Gaussian kernel (see \SI), so that pairs of data point which are close are connected by darker (``stronger'') edges.
 (D) The data in C, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure.
@@
 %
-We first compute the angles needed to align pairs of vectors (or images), which in general requires no template function \citep{ahuja2007template}.
+We first compute the angles needed to align pairs of vectors (or images), which in general requires no notion of a template function \citep{ahuja2007template}.
 %
@@
 %
-However, in other applications, pairwise alignments can be computed by aligning appropriate image landmarks or features. 
+However, pairwise alignments can be computed by aligning appropriate image landmarks or features \citep{ian1998statistical}. 
 %
-When the data are noisy, these pairwise measurements may by inaccurate, and so we would like to utilize {\em all} pairwise measurements to align the set of images robustly. 
+When the data are noisy, these pairwise measurements may by inaccurate, and so we utilize {\em all} pairwise measurements to align the set of images robustly. 
 %
@@
 %
-However, the algorithm can successfully register data sets even when many of the pairwise measurements are inaccurate \citep{singer2011angular}.
+However, the algorithm can register data sets even when many of the pairwise measurements are inaccurate \citep{singer2011angular}.
 
-After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
+After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional curve that parameterizes most of the remaining variability in the data.
 %
-Such a manifold can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parameterization of data that lies on a low-dimensional manifold in high-dimensional space.
+Such a curve can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that reveals a parameterization of data that lies on a low-dimensional manifold in high-dimensional space.
 %
-The idea is illustrated in \fig~\ref{fig:schematics}C, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
+The idea is illustrated in \fig~\ref{fig:schematics}C, where the data are two-dimensional points which lie on a one-dimensional (nonlinear) curve.
 %
@@
 %
+A detailed example of using vector diffusion maps to register and order synthetic data is given in \fig~\ref{fig:1d_example}, and a step-by-step tutorial of the diffusion maps implementation is included in the available software.
+%
 In our working examples, each data point will be of much higher dimension (e.g., a pixelated image or three-dimensional voxel data), and so we cannot extract this low-dimensional structure visually.
 %
-Instead, we will use diffusion maps to automatically uncover a parameterization of our high-dimensional data from the eigenvectors of the appropriate matrix (see \SI).
+Instead, we will use diffusion maps which automatically uncovers a parameterization of our high-dimensional data from the eigenvectors of the appropriate matrix (see \SI).
 %
-Furthermore, the corresponding eigenvalues will allow us to confirm our assumption that our data lie on a one-dimensional manifold. 
+Furthermore, the corresponding eigenvalues will allow us to confirm our assumption that our data lie on a one-dimensional manifold (see \fig~\ref{fig:eigenvalues_drosophila_live}--\ref{fig:eigenvalues_wing_disc}).
 %
@@
 \caption{Method validation using live imaging of \textit{Drosophila} embryos.
-(A) Selected images from a live imaging study of a \textit{Drosophilla} embryo during gastrulation.
+(A) Selected images from a live imaging study of a \textit{Drosophila} embryo during gastrulation.
 Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
@@
 The rank correlation coefficient is $0.9989$.
-(E) The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies.
+(E) The average error in the recovered angle and the rank correlation coefficient for $5$ independent live imaging studies.
 }
@@
 
-To validate the proposed approach, we applied our algorithm to a data set where the true temporal order and rotational orientation of the images were known {\em a priori}.
+To validate the proposed approach, we first applied our algorithm to a data set where the true temporal order and rotational orientation of the images were known {\em a priori}.
 %
@@
 %
-\fig~\ref{fig:drosophila_live_imaging}C and \fig~\ref{fig:drosophila_live_imaging}D  show the correlations between the recovered and true angles and ranks, respectively, for the entire data set. 
+\fig~\ref{fig:drosophila_live_imaging}C and \fig~\ref{fig:drosophila_live_imaging}D  show the correlations between the recovered and true angles and rank orders, respectively, for the entire data set.
 %
@@
 
-\subsubsection*{Zebrafish gastrulation}
+\subsubsection*{Zebrafish epiboly}
 
@@
 \caption{Method validation using live imaging of a zebrafish embryo.
-(A) Selected images from a flip book movie of zebrafish embryogenesis.
+(A) Selected images from a flip book movie of zebrafish epiboly.
 Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
@@
 The rank correlation coefficient is $0.9954$.
-The larger errors in the recovered ranks towards the beginning of the trajectory are due to the rather static cell movement within that time window.
+The larger errors in the recovered ranks towards the beginning of the trajectory are due to the rather slow cell movement within that time window.
 }
@@
 %
-The cell mass is then compressed and the animal-vegetal axis of the embryo (vertical axis in the \fig~\ref{fig:zebrafish}) shortens to form a spherical embryo shape by the end of the fourth hour of development.
+The cell mass is then compressed and the animal-vegetal axis of the embryo (vertical axis in \fig~\ref{fig:zebrafish}) shortens to form a spherical embryo shape by the end of the fourth hour of development.
 %
@@
 %
-Some errors in ordering images of the early embryo result from rather static cell movement during the early developmental stage where cells divide and accumulate above the yolk. 
+Some errors in ordering images of the early embryo result from rather slow cell movement during the early developmental stage where cells divide and accumulate above the yolk. 
 %
@@
 
-In summary, we showed that our algorithm is not limited to a specific set of data. 
+In summary, we have showed that our algorithm is not limited to a specific set of data. 
 %
@@
 %
-\caption{Analyzing images of fixed \textit{Drosophila} embryos.
+\caption{Analysis of images of fixed \textit{Drosophila} embryos.
 {(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red).
@@
 %
-The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.77, indicating that the algorithm can accurately recover the temporal order even under noisy conditions.
+The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.77, indicating that the algorithm can recover the temporal order even under noisy conditions.
 
@@
 %
-Averaging successive images removes (some) of the interembryo variability, so that sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.
+Averaging successive images removes some of the interembryo variability, so that sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.
 
-From this average trajectory, we can now easily see the developmental progression, which is consistent with the known dynamics: 
-dpERK first appears as two lateral peaks at the ventral side of the embryo; a third dpERK peak then appears at the dorsal side of the embryo.
+From this average trajectory, we can now easily see the developmental progression, which is consistent with the known dynamics.
 %
-During mesoderm invagination, the two ventral dpERK peaks merge together, eventually forming, together with Twi, the ``omega'' shape.
+dpERK first appears as two lateral peaks at the ventrolateral side of the embryo; a third dpERK peak then appears at the dorsal side of the embryo.
+%
+During mesoderm invagination, the two ventrolateral dpERK peaks merge together, eventually forming, together with Twi, the ``omega'' shape.
 %
@@
 
-\caption{Three-dimensional wing disc data.
-(A) Maximum projections of an example three-dimensional image.
+\caption{Analysis of three-dimensional wing disc data.
+(A) Maximum projections of an example three-dimensional z-stack.
 Projections along the $x$-, $y$-, and $z$-axes are shown.
@@
 %
-Thus far, we have only looked at two-dimensional data sets where biological samples are imaged as a single confocal slice or as a phase contrast image. 
+Thus far, we have only looked at two-dimensional data sets where biological samples are imaged as a single confocal slice or as a phase contrast image at a fixed position within the sample. 
 %
-However, many developmental processes are inherently three-dimensional, and the z-axis often contains interesting dynamics which are independent of the dynamics in the x-y plane. 
+However, many developmental processes contain interesting dynamics in all three dimensions.
 %
-Researchers often take multiple z-slices to visualize the three-dimensional structure of a sample.
+With advances in imaging techniques, taking multiple z-slices to visualize the three-dimensional structure of a sample will become more common, and it is important to validate that our algorithm can be applied to such data.
+
+Although the theory of vector diffusion maps readily extends to three-dimensional data with full three-dimensional rotations, calculation of the pairwise alignments rapidly becomes computational prohibitive for such applications, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle).
 %
-With advances in imaging techniques, such three-dimensional data will be more common, and it is important to validate that our algorithm can be applied to such data.
+However, ordering data via diffusion maps is readily possible for registered three-dimensional data. 
+%
+Therefore, we can apply our methodology to three-dimensional data by first registering the data using an alternative approach, and then ordering the registered data using diffusion maps.
 
-We used an existing three-dimensional data set of fixed {\em Drosophila} wing imaginal discs, which was published in \citet{hamaratoglu2011dpp}.  
+To demonstrate this approach, we used an existing three-dimensional data set of fixed {\em Drosophila} wing imaginal discs, which was published in \citet{hamaratoglu2011dpp}.  
 %
@@
 
-Although the theory of vector diffusion maps readily extends to three-dimensional data, calculation of the pairwise alignments rapidly becomes computational prohibitive for three dimensional data, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle).
+For this particular system, the face of the imagingal disc is significantly larger than the edge of the disc, so that images will always be taken with the face towards the microscope. 
 %
-However, ordering data via diffusion maps is readily possible for registered three-dimensional data. 
+Therefore, we need not consider registration in all three dimensions, as the z-axis will always be aligned with the short axis of the disc. 
 %
-Therefore, we can apply our methodology to three-dimensional data by first registering the data using an alternative approach, and then ordering the registered data using diffusion maps. 
+We can instead focus on registering the discs with respect to rotations only in the x-y plane. 
 %
-For the wing disc data, we registered the three-dimensional data by computing the optimal rotations to align the maximum intensity projections of the data using vector diffusion maps.
+We chose to align the maximum intensity projections using angular synchronization, and then used these rotations to register the full three-dimensional in the x-y plane.
 %
@@
 %
+This approach is possible whenever there are distinct major and minor axes within a three-dimensional sample, which reduces the rotational degrees of freedom.
+%
 Other options for registering samples include using a specific localized signal (such as Twi in \fig~\ref{fig:drosophila_fixed_images}) as a marker for alignment, or using other {\em a priori} knowledge.
-%
+
 We then used diffusion maps to order the registered three-dimensional data. 
@@
 %
-The rank correlation coefficient based on the time class is 0.9427. 
+The rank correlation coefficient based on the time class is 0.9436. 
 %
-There are some registration errors in the early images; these are primarily due to some wing discs having extra tissue attached to them. 
+There are some registration errors in the early images, which are primarily due to some wing discs having extra tissue attached to them (such as the image in \fig~\ref{fig:wing_disc}A and the fourth image in \fig~\ref{fig:wing_disc}B).
 %
@@
 %
+Thus, this result shows that the algorithm can successfully order a three-dimensional data set.
+%
 Furthermore, we can extract a stereotypical developmental trajectory, shown in \fig~\ref{fig:wing_disc}C, by averaging (see \SI).
 %
-Thus, this result shows that the algorithm can successfully order a three-dimensional data set.
+We can now easily see the growth of the wing disc; however, averaging several images blurs the finer scale structures within the images
+
+We would like to note that the vector diffusion maps algorithm has been previously used for a different type of three-dimensional data: to reconstruct a three-dimensional object (in this case, a molecule) from two-dimensional (cryo-electron microscopy) pictures \citep{singer2011viewing}.
+%
+In these applications, many two-dimensional pictures of (effectively) the same object are taken from different viewing directions. 
+%
+The vector diffusion maps algorithm then, instead of revealing the temporal order,  reveals the viewing direction of each image, which can then be used to reconstruct the three-dimensional object. 
+%
+This is another approach to three-dimensional imaging which we hope to extend in the future to examples which also contain variability from the developmental progression.
 
@@
 
-\raisebox{1.75cm}{\textsf{D}}
+\raisebox{2.5cm}{\textsf{D}}
 {\renewcommand{\arraystretch}{1.5}
@@
 \hline
-{\em Drosophila} live~imaging & 2D & 1 & 40 & 100 $\times$ 100 & 10$^\circ$ & 3.2~sec \\ 
-Zebrafish & 2D & 1 & 120 & 100 $\times$ 100 & 10$^\circ$ & 13~sec\\ 
-{\em Drosophila} fixed images & 2D & 3 & 120 & 100 $\times$ 100 & 10$^\circ$ & 29~sec  \\ 
-Wing disc & 3D & 3 & 46 & 100 $\times$ 100 $\times$ 21 & 10$^\circ$ & 12~sec  \\
+{\em Drosophila} gastrulation (live) & 2D & 1 & 40 & 100 $\times$ 100 & 10$^\circ$ & 3.2~sec \\ 
+Zebrafish epiboly & 2D & 1 & 120 & 100 $\times$ 100 & 10$^\circ$ & 13~sec\\ 
+{\em Drosophila} gastrulation (fixed) & 2D & 3 & 120 & 100 $\times$ 100 & 10$^\circ$ & 29~sec  \\ 
+{\em Drosophila} wing discs & 3D & 3 & 46 & 100 $\times$ 100 $\times$ 21 & 10$^\circ$ & 12~sec  \\
  \hline
@@
 (B) CPU time as a function of the number of pixels in the images (for $120$ images, and $10^\circ$ angular discretization).
-Empirically, the CPU time is $\sim \mathcal{O}(n^{1.83})$ in number of pixels.
+Empirically, the CPU time is $\sim \mathcal{O}(n^{1.83})$ in the number of pixels.
 (C) CPU time as a function of the number of rotations (for $120$ images of $100 \times 100$ pixels).
@@
 %
-Furthermore, the computation of the pairwise rotational alignments, which accounts for the majority of the computational time, is trivially parallelizable for larger data sets.
+Furthermore, the computation of the pairwise rotational alignments, which accounts for the majority of the computational time, is trivially parallelizable for larger data sets, and only a subsample of the pairwise alignments need to be computed for accurate recovery of the underlying rotations \citep{singer2011angular}.
 %
-Becasuse the computational cost increases with the image resolution, we chose to subsample all of our data sets to $100 \times 100$ pixels.
+Because the computational cost increases with the image resolution, we chose to subsample all of our data sets to $100 \times 100$ pixels.
 %
-This resolution allowed us to rapidly analyze our data sets while still capturing all of the relevant developmental features.
+This resolution allowed us to rapidly analyze our data sets while still retaining all of the relevant developmental features.
 %
@@
 %
-Two algorithmic parameters, the angular discretization to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{fig:schematics}A and \SI), must also be defined; we also provide some guidance on selecting these parameters.
+Two algorithmic parameters, the angular discretization to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{fig:schematics} and \SI), must also be defined.
 %
-We found that the results are robust to both of these parameters.
+We also provide some guidance on selecting these parameters, and found that the results are robust to both of these parameters.
 %
@@
 %
-We have applied our methodology to several different proprotypical data sets in developmental biology; because our methodology only relies on computing distances between images,, we are confident it may be readily extended to many other data sets.
+We have applied our methodology to several different proprotypical data sets in developmental biology; because our methodology only relies on computing distances between images, we are confident it may be readily extended to many other data sets.
+%
+The benefits of our algorithm are twofold. 
+%
+First, the algorithm can accomplish the tasks of registration and ordering in a single step. 
+%
+Furthermore, because our methodology is nonlinear, it can successfully order data sets which contain complex dynamics (see \fig~\ref{fig:PCA_compare} for a comparison of ordering using linear principal component analysis \citep{shlens2005tutorial} versus vector diffusion maps for the data sets presented in this paper); we expect nonlinear techniques to be necessary for larger data sets which span a wider dynamic range.
+%
+We believe that the main utility of our proposed methodology lies in the analysis of data sets containing hundreds of images from systems which have not been well-studied. 
+%
+For such data sets, manually uncovering the relevant features to describe the dynamics and order the images can be nontrivial, and our algorithms can clearly accelerate such analysis.
 
-We acknowledge that our methods, though general, do have some limitations. 
+We acknowledge that our methods, though general, do have limitations. 
 %
-The first is that we require enough amount of data to sufficiently sample the developmental trajectory.
+The first is that we require enough data to sufficiently sample the developmental trajectory.
 %
@@
 %
-Second, the relevant developmental patterns and structures need to be large compared to the noise and the image resolution. 
+Second, the pertinent image features need to be large compared to the noise and the image resolution. 
 %
-In all of our examples, the relevant features, such as expression patterns and morphological structures, all span several pixels and are large compared to both the instrument noise and embryo-to-embryo variability.
+In all of our examples, the relevant expression patterns and morphological structures span several pixels and are large compared to both the instrument noise and embryo-to-embryo variability. 
 %
-This means that the Euclidean distance is a good measure of similarity between images.
+This means that the Euclidean distance between pixels is a good measure of similarity between images.
 %
-Our algorithms will fail for other imaging data sets which contain very fine scale structures,  as small perturbations in these fine-scale structures will result in large changes in the Euclidean distances.
+Our algorithms will fail for other imaging data sets which contain fine-scale structures,  as small perturbations in these structures will result in large changes in the Euclidean distances. 
 %
-In such cases, one requires a more sophisticated distance metric, such as the Earth Mover's Distance \citep{rubner2000earth}, or image description, such as scattering transform coefficients \citep{mallat2012group}, which are insensitive to such perturbations.
+In such cases, one requires a more sophisticated distance metric (such as the Earth Mover's Distance \citep{rubner2000earth}) and/or image description (such as scattering transform coefficients \citep{mallat2012group}) which is insensitive to such perturbations.
+%
+The methodology we have presented here specifically addresses images taken in the same location and/or from the same viewing direction within the organism. 
+%
+However, for organisms with distinct major and minor axes (such as {\em Drosophila} embryos and wing discs), sample preparation will often fix one of the principal directions. 
 %
@@
 %
-However, the algorithms presented here could be used for three-dimensional given an alternative method for computing pairwise alignments more efficiently (such as using image features or landmarks).
+However, the algorithms presented here could be used for three-dimensional voxel data given an alternative method for computing pairwise alignments more efficiently (such as using image features or landmarks).
 
-Unlike traditional statistical methods, our algorithms do not provide confidence intervals on predicted orderings, but could in principle be extracted from bootstrapping \citep{efron1979bootstrap}.
+Unlike traditional statistical methods, our algorithms do not provide confidence intervals on predicted orderings, but these could in principle be extracted from bootstrapping \citep{efron1979bootstrap}.
 %
-Furthermore, although our methods provide a temporal ordering for images, the provide no real-time information, neither do they provide a direction for the proposed ordering.
+Furthermore, although our methods provide a temporal ordering for images, they provide no information about the actual time, neither do they provide a direction for the proposed ordering.
 %
-Determining the trajectory rate and direction are post-processing tasks which requires {\em a priori} knowledge of the developmental dynamics. 
-%
-Furthermore, real-time information can only be extracted from either {\em a priori} knowledge of the rates of relevant developmental processes, or by making some assumption about the temporal sampling of the data set (e.g., the data set uniformly samples some given time window).
+Determining the trajectory rate and direction are post-processing tasks which require {\em a priori} knowledge of the developmental processes or some assumptions about the temporal sampling of the data set (e.g., the data set uniformly samples some given time window).
 
@@
 %
-Based on the presented results, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories from snapshots of fixed embryos.
+Based on the presented results, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories from snapshots of fixed tissues.
 %
@@
 %
-In general, we can recover several parameters which concisely and comprehensively describe data set.
+In general, we can recover several parameters which concisely and comprehensively describe the data set.
 %
@@
 %
-Furthermore, images taken from different viewing directions can be analyzed, as the vector diffusion maps parameterization will organize the images according to the viewing angle.
+Furthermore, images taken from different viewing directions can be analyzed, as the vector diffusion maps parameterization will organize the images according to the viewing angle \cite{singer2011viewing}.
 %
@@
 %
-Embryos were loaded to the microfluidic device with PBST to keep them oxidized.
+Live embryos were loaded to the microfluidic device with PBST to keep them oxidized, and fixed embryos were loaded with 90\% glycerol.
 
@@
 %
-Nikon A1-RS scanning confocal microscope, and the Nikon 60x Plan-Apo oil objective was used to image embryos. 
+Nikon A1-RS scanning confocal microscope, and the Nikon 60x Plan-Apo oil objective was used to image {\em Drosophila} embryos. 
 %

### 2015-02-04 image_analysis_paper_development2/code_for_distribution/register_order_gui.m

@@ -928,8 +928,8 @@ try
     end
     
     % compute optimal rotations + embedding coordinates using vector diffusion maps
-    ncomps = 1;
-    [R_opt, embed_coord, D2] = vdm(handles.R, handles.W, handles.eps_scale, ncomps);
+    max_proj_dim = 2;
+    R_opt = ang_synch(handles.R, max_proj_dim);
     
     if handles.use_raw
         images_registered = register_all_images(handles.images_raw, R_opt);
@@ -940,6 +940,7 @@ try
     h = waitbar(0, 'Registering and ordering images....') ;
     
     W = squareform(pdist(reshape(double(images_registered), [], handles.nimages)')).^2;
+    ncomps = 1;
     [embed_coord, D2] = dm(W, handles.eps_scale, ncomps);
     
     waitbar(0.5, h);

### 2015-02-04 image_analysis_paper_development2/paper_figures/make_fig_wing_disc.m

@@ -32,7 +32,8 @@ plot_images(images, dim)
 
 % compute optimal rotations + embedding coordinates using vector diffusion maps
 ncomps = 1;
-[R_opt, embed_coord, D2] = vdm(R, W, eps_scale, ncomps);
+% [R_opt, embed_coord, D2] = vdm(R, ones(nimages), eps_scale, ncomps);
+R_opt = ang_synch(R, 2);
 
 % register images using optimal rotations
 images_registered = register_all_images(images, R_opt);
@@ -53,6 +54,7 @@ end
 
 images_analyzed = order_all_images(images_registered, embed_coord);
 
+return
 %%
 
 plot_images(immultiply(images_analyzed, 1.5), dim)
@@ -91,7 +93,7 @@ nsubimages = 15;
 I = I + 20;
 
 make_fig(10, 6);
-plot_wing_disc_projections(images_registered(:,:,:,:,I), time(I)-min(time), 5, 3)
+plot_wing_disc_projections(images_registered(:,:,:,:,I), time(I), 5, 3)
 % plot_wing_disc_projections(images_registered(:,:,:,:,I), time(I)-min(time));
 % saveas(gcf, 'wing_disc_ordered.pdf');
 print('wing_disc_ordered.eps', '-depsc', '-r300');

### 2015-02-05 image_analysis_paper_development2/code_for_distribution/diffusion_maps_tutorial.m
updated dmaps tutorial (spelling errors)

@@ -2,7 +2,7 @@ clear all
 close all
 
 %% Create Data Set
-% First, we create an artifical data set |x| of $n=7$ data points which lie 
+% First, we create an artificial data set |x| of $n=7$ data points which lie 
 % on a one-dimensional curve in two dimensions.
 % This data set can be parameterized by t he angular direction $\theta$.
 
@@ -41,7 +41,7 @@ W = exp(-dist_matrix.^2/eps.^2)
 
 
 %% Compute Diffusion Matrix
-% We then compute the $n \times n$ diaganol matrix $D$, with 
+% We then compute the $n \times n$ diagonal matrix $D$, with 
 % $$ D_{ii} = \sum_{j=1}^n W_{ij} $$ and
 % the $n \times n$ matrix $$A = D^{-1}W. $$
 

### 2015-02-05 image_analysis_paper_development2/manuscript.tex
updated manuscript

@@
 
+\usepackage{color}
+
 \usepackage{epstopdf}
@@
 \renewcommand{\figurename}{Fig.}
-\newcommand{\SI}[0]{\textit{SI Materials and Methods}}
+%\newcommand{\SI}[0]{\textit{SI Materials and Methods}}
+\newcommand{\SI}[0]{supplementary material}
 \newcommand{\fig}[0]{Fig.}
@@
 Bomyi~Lim \footnotemark[1],
-Ioannis~G.~Kevrekidis \footnotemark[1] \thanks{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA} ,
 Hang~Lu \thanks{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
-Amit~Singer \thanks{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \footnotemark[2], \\ and
+Amit~Singer \thanks{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \thanks{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA}, Ioannis~G.~Kevrekidis \footnotemark[1] \footnotemark[4] ,\\ and
 Stanislav~Y.~Shvartsman \footnotemark[1] \thanks{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA}}
@@
 
-Imaging studies provide unique insights into the dynamics of pattern formation and morphogenesis in developing tissues.
-%
-Developmental progress is commonly reconstructed from snapshots of chemical or mechanical processes in fixed embryos.
+Dynamics of developmental progress is commonly reconstructed from imaging snapshots of chemical or mechanical processes in fixed embryos.
 %
@@
 %
-Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific developmental system.
+Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific system.
 %
@@
 %
-To illustrate our approach, we register and order data from imaging studies of pattern formation and morphogenesis in three different model systems in developmental biology.
-
+We illustrate this approach by registering and ordering data from imaging studies of pattern formation and morphogenesis in three different model systems.
+%
+We also provide a software implementation of our methodology to aid in the application of our techniques to other experimental data sets. 
 \vspace{1em}
@@
 \section*{Introduction}
+
+\begin{figure}[t]
+\includegraphics{fig1}
+\caption{Caricature illustrating the tasks of image registration and temporal ordering.
+(A) Images of ``samples'', each in a different orientation and a different stage of development.
+(B) Registered and ordered samples.
+For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
+%\label{fig:fish}
+\customlabel{fig:fish}{1}
+\customlabel{subfig:fish_unordered}{\ref{fig:fish}A}
+\customlabel{subfig:fish_ordered}{\ref{fig:fish}B}
+\end{figure}
 
@@
 
-\begin{figure}[t]
-\includegraphics{fig1}
-\caption{Caricature illustrating the tasks of image registration and temporal ordering.
-(A) Images of ``samples'', each in a different orientation and a different stage of development.
-(B) Registered and ordered samples.
-For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
-%\label{fig:fish}
-\customlabel{fig:fish}{1}
-\customlabel{subfig:fish_unordered}{\ref{fig:fish}A}
-\customlabel{subfig:fish_ordered}{\ref{fig:fish}B}
-\end{figure}
-
 We present a robust algorithmic approach to simultaneous registration and temporal ordering.
 %
-In contrast to a number of previous methodologies \citep{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face, dubuis2013accurate}, our methodology does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression.
+In contrast to a number of previous methodologies \citep{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face, dubuis2013accurate}, our methodology does not rely on the {\em a priori} knowledge of landmarks for registration or markers of developmental progression.
 %
-The approach is based on a manifold learning algorithm (vector diffusion maps \citep{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
+The approach is based on vector diffusion maps \citep{singer2012vector}, a manifold learning algorithm which simultaneously addresses the problems of registration and temporal ordering.
 %
@@
 applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \citep{zhao2014rotationally, singer2011viewing} to face recognition \citep{lafon2006data} and classification of CT scans \citep{fernandez2014diffusion}.
-%
+
 Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing tissues in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
@@
 %
-Our final data set consists of z-stacks of wing discs, which we will use to illustrate how our methods can be used to analyze specific types of three-dimensional imaging data. 
+Our final data set consists of z-stacks of {\em Drosophila} wing discs, which we will use to illustrate how our methods can be used to analyze specific types of three-dimensional imaging data.
 %
@@
 %
-Although in general, images may contain variations due to rotations, translations, and scaling, we will remove the relevant translational and/or scaling symmetries via relatively simple image preprocessing, and focus here only on factoring out rotations using the vector diffusion maps algorithm.
+Although in general, images may contain variations due to rotations, translations, and scaling, we will remove the relevant translational and/or scaling symmetries via relatively simple image preprocessing, and focus only on factoring out rotations using the vector diffusion maps algorithm.
 %
-In the case that {\em all} relevant symmetries can be removed with straightforward preprocessing procedures, our algorithms can also be used simply to extract the main direction of variability within the imaging data set.
+In the case that {\em all} relevant symmetries can be removed with straightforward preprocessing, our algorithms can be used simply to extract the main direction of variability within the imaging data set.
 %
@@
 %
-Such a curve can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that reveals a parameterization of data that lies on a low-dimensional manifold in high-dimensional space.
+Such a curve can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that reveals a parametrization of data that lies on a low-dimensional manifold in high-dimensional space.
 %
@@
 %
-We use {\it local} information about the data to find a parameterization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parameterization.
+We use {\it local} information about the data to find a parametrization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parameterization.
 %
@@
 %
-The color in \fig~\ref{fig:schematics}D depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
+The color in \fig~\ref{fig:schematics}D depicts the one-dimensional parametrization or ordering of the data that we can detect visually.
 %
-A detailed example of using vector diffusion maps to register and order synthetic data is given in \fig~\ref{fig:1d_example}, and a step-by-step tutorial of the diffusion maps implementation is included in the available software.
+A detailed example of using vector diffusion maps to register and order synthetic data is given in \fig~\ref{fig:1d_example}, and a step-by-step tutorial of the diffusion maps implementation is included in the \SI.
 %
@@
 %
-Instead, we will use diffusion maps which automatically uncovers a parameterization of our high-dimensional data from the eigenvectors of the appropriate matrix (see \SI).
+Instead, we will use diffusion maps which automatically uncovers a parametrization of our high-dimensional data from the eigenvectors of the appropriate matrix (see \SI).
 %
@@
 %
-Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.
+%Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.
 
@@
 (A) Selected images from a live imaging study of a \textit{Drosophila} embryo during gastrulation.
+Scale bar indicates 50$\mu$m.
 Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
-(B) Images from A after registration and ordering using vector diffusion maps.
+(B) Images from A registered and ordered by vector diffusion maps.
 The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.
@@
 
-To assess the robustness of our proposed methodology, we repeated this procedure with four additional data sets extracted from independent live imaging studies spanning the same developmental time period.
+To assess the robustness of the proposed methodology, we repeated this procedure with four additional data sets extracted from independent live imaging studies spanning the same developmental time period.
 %
@@
 The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than 95\%, indicating that our methodology can reproducibly order data of this type.
-
+%
 In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{fig:schematics}C) is well-sampled.
@@
 %%
-Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignment, and the high temporal resolution of our live imaging study allows us to order the images accurately.
+%Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignment, and the high temporal resolution of our live imaging study allows us to order the images accurately.
 %
-In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.
+%In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.
 
@@
 \caption{Method validation using live imaging of a zebrafish embryo.
-(A) Selected images from a flip book movie of zebrafish epiboly.
+(A) Selected images from a movie of zebrafish epiboly.
 Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
@@
 The rank correlation coefficient is $0.9954$.
-The larger errors in the recovered ranks towards the beginning of the trajectory are due to the rather slow cell movement within that time window.
+The larger errors in the recovered ranks towards the beginning of the trajectory are due to the slow cell movement within that time window.
 }
@@
 
-To illustrate the broad applicability of our approach, we also applied our algorithm to a time-lapse movie  of zebrafish embryogenesis. 
+As another validation for the proposed methodology, we applied our algorithm to a time-lapse movie  of zebrafish embryogenesis.
 %
-We used a publicly available flipbook movie of zebrafish embryogenesis, available at \texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov} \citep{karlstrom1996flipbook}.  
+We used a publicly available live imaging data of zebrafish embryogenesis (\texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov}, \citet{karlstrom1996flipbook}).
 %
@@
 %
-During this stage, the cells rearrange to form a cell layer with uniform thickness (about four cells thick). 
+During this stage, the cells rearrange to form a uniform layer about four cells thick.
 %
@@
 %
-Some errors in ordering images of the early embryo result from rather slow cell movement during the early developmental stage where cells divide and accumulate above the yolk. 
+Some errors in ordering images of the early embryo result from slow cell movement during the early developmental stage where cells divide and accumulate above the yolk.
 %
@@
 
-In summary, we have showed that our algorithm is not limited to a specific set of data. 
+In summary, we have shown that our approach to temporal ordering performs very well on imaging data of two different developmental processes ({\em Drosophila} gastrulation and zebrafish epiboly), taken with two different imaging methods (fluorescent microscopy and DIC) where the true temporal order is known {\em a priori}.
+%image registration and ordering can correctly recover image orientation and  showed that our algorithm is not limited to a specific set of data.
 %
-The algorithm works for different tissue types ({\em Drosophila} and zebrafish), and also for images taken with different imaging methods (fluorescent and phase contrast images).
+%The algorithm works for different tissue types ({\em Drosophila} and zebrafish), and also for images taken with different imaging methods (fluorescent and phase contrast images).
 %
@@
 
-\begin{figure*}[t]
-\begin{minipage}[b]{12cm}
-\raisebox{0.5cm}{\textsf{A}}
-\includegraphics[width=12.25cm]{drosophila_fixed_images_scrambled}
-
-\raisebox{0.5cm}{\textsf{B}}
-\includegraphics[width=12.25cm]{drosophila_fixed_images_ordered}
-
-\raisebox{0.5cm}{\textsf{C}}
-\includegraphics[width=12.25cm]{drosophila_fixed_images_average}
-\end{minipage}
-%
-\hfill
-%
-\raisebox{3.25cm}{\textsf{D}}
-\includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}
-%
-\caption{Analysis of images of fixed \textit{Drosophila} embryos.
-{(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red).
-Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation.
-{(B)} Data from {\it A}, registered and ordered using vector diffusion maps.
-The expert rank for each image is indicated.
-{(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI).
-{(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert.
-The rank correlation coefficient is $0.9716$.
-}
-\label{fig:drosophila_fixed_images}
-\end{figure*}
-
 We have analyzed how our algorithm performs on two model data sets where all images come from a single embryo.
@@
 %
-To demonstrate that our methods are robust to such noise sources,
+To demonstrate that our methods are robust to such variations,
 we constructed a synthetic time course data set by selecting a random image from one of five {\em Drosophila} live imaging data sets (those data sets used in \fig~\ref{fig:drosophila_live_imaging}) at each time point.
@@
 Averaging successive images removes some of the interembryo variability, so that sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.
+
+\begin{figure*}[t!]
+\begin{minipage}[b]{12cm}
+\raisebox{0.5cm}{\textsf{A}}
+\includegraphics[width=12.25cm]{drosophila_fixed_images_scrambled}
+
+\raisebox{0.5cm}{\textsf{B}}
+\includegraphics[width=12.25cm]{drosophila_fixed_images_ordered}
+
+\raisebox{0.5cm}{\textsf{C}}
+\includegraphics[width=12.25cm]{drosophila_fixed_images_average}
+\end{minipage}
+%
+\hfill
+%
+\raisebox{3.25cm}{\textsf{D}}
+\includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}
+%
+\caption{Analysis of images of fixed \textit{Drosophila} embryos.
+{(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red).
+Scale bar indicates approximately 50$\mu$m (images have been rescaled to remove slight interembryo size variations).
+Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation.
+{(B)} Data from {\it A}, registered and ordered using vector diffusion maps.
+The expert rank for each image is indicated.
+{(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI).
+{(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert.
+The rank correlation coefficient is $0.9716$.
+}
+\label{fig:drosophila_fixed_images}
+\end{figure*}
 
@@
 
-\caption{Analysis of three-dimensional wing disc data.
-(A) Maximum projections of an example three-dimensional z-stack.
+\caption{Analysis of three-dimensional {\em Drosophila} wing disc z-stacks.
+(A) Maximum projections of an example three-dimensional {\em Drosophila} wing disc z-stack.
+Discs express the Dad-GFP reporter construct (green), and are stained for Spalt (red), Wingless (gray), and Patched (gray).
 Projections along the $x$-, $y$-, and $z$-axes are shown.
@@
 
-We have demonstrated that our algorithm can recover the underlying developmental dynamics in several different types of imaging data sets, including ones containing interembryo variability. 
+%We have demonstrated that our algorithm can recover the underlying developmental dynamics in several different types of imaging data sets, including ones containing interembryo variability.
 %
-Thus far, we have only looked at two-dimensional data sets where biological samples are imaged as a single confocal slice or as a phase contrast image at a fixed position within the sample. 
+%Thus far, we have only looked at two-dimensional data sets where biological samples are imaged as a single confocal slice or as a phase contrast image at a fixed position within the sample.
 %
-However, many developmental processes contain interesting dynamics in all three dimensions.
-%
-With advances in imaging techniques, taking multiple z-slices to visualize the three-dimensional structure of a sample will become more common, and it is important to validate that our algorithm can be applied to such data.
+%However, with advances in imaging techniques, taking multiple z-slices to visualize the three-dimensional structure of a sample will become more common, and it is important to validate that our algorithm can be applied to such data.
 
@@
 %
-Therefore, we can apply our methodology to three-dimensional data by first registering the data using an alternative approach, and then ordering the registered data using diffusion maps.
+Therefore, we can apply our methodology to three-dimensional data, such as a stack of confocal images taken at different depths within the sample, by first registering the data using an alternative method, and then ordering the registered data using diffusion maps.
 
@@
 %
-Imaginal discs are groups of progenitor cells in fly larva that will transform into a specific organ, such as wings, legs, and eyes, in the adult fly during the pupal transformation. 
+Imaginal discs are groups of progenitor cells in fly larva that will transform into specific organs (such as wings, legs, and eyes) in the adult fly during the pupal metamorphosis.
 %
-The wing disc is an imaginal disc that turns into a fly wing. 
+The wing disc is an imaginal disc that turns into a wing.
 %
@@
 %
-The discs were stained with antibodies that recognize Spalt (red), Dad-GFP (green), and Wingless/Patched (gray) (\fig~\ref{fig:wing_disc}A). 
+The discs were dissected from larvae expressing the Dad-GFP reporter construct (green) and stained with antibodies that recognize Spalt (red), Wingless (gray), and Patched (gray), the molecules that play important roles in disc patterning and growth (\fig~\ref{fig:wing_disc}A).
 %
-The disc grows in time; at long times, the disc becomes too large to fit completely within the frame and only a partial disc is imaged.
+%The disc grows in time; at long times, the disc becomes too large to fit completely within the frame and only a partial disc is imaged.
 
-For this particular system, the face of the imagingal disc is significantly larger than the edge of the disc, so that images will always be taken with the face towards the microscope. 
+For this particular system, the face of the imaginal disc is significantly larger than the edge of the disc, so that images will always be taken with the face towards the microscope.
 %
@@
 %
-We chose to align the maximum intensity projections using angular synchronization, and then used these rotations to register the full three-dimensional in the x-y plane.
+We aligned the maximum intensity projections using angular synchronization, and then used these rotations to register the full three-dimensional in the x-y plane.
 %
@@
 %
-This approach is possible whenever there are distinct major and minor axes within a three-dimensional sample, which reduces the rotational degrees of freedom.
+Such an approach is possible when there are distinct major and minor axes within a three-dimensional sample, which reduces the rotational degrees of freedom.
 %
-Other options for registering samples include using a specific localized signal (such as Twi in \fig~\ref{fig:drosophila_fixed_images}) as a marker for alignment, or using other {\em a priori} knowledge.
+%Other options for registering samples include using a specific localized signal (such as Twi in \fig~\ref{fig:drosophila_fixed_images}) as a marker for alignment, or using other {\em a priori} knowledge.
 
@@
 %
-In the ordered set, the size of wing disc grows, and the intensity of proteins increases as a function of time. 
+In the ordered set, the size of wing disc grows, and the intensity of the Dad-GFP signal increases as a function of time.
 %
@@
 %
-Even with such obstructions, we can order the images with high accuracy. 
+Even with such obstructions, we can order the images with high accuracy and extract a stereotypical developmental trajectory, shown in \fig~\ref{fig:wing_disc}C, by averaging (see \SI).
+%
+We can now easily see the growth of the wing disc; however, averaging several images blurs the finer scale structures within the images
 %
 Thus, this result shows that the algorithm can successfully order a three-dimensional data set.
+
+%The vector diffusion maps algorithm has been previously used for a different type of three-dimensional data: to reconstruct a three-dimensional object (in this case, a molecule) from two-dimensional (cryo-electron microscopy) pictures \citep{singer2011viewing}.
 %
-Furthermore, we can extract a stereotypical developmental trajectory, shown in \fig~\ref{fig:wing_disc}C, by averaging (see \SI).
+%In these applications, many two-dimensional pictures of (effectively) the same object are taken from different viewing directions.
 %
-We can now easily see the growth of the wing disc; however, averaging several images blurs the finer scale structures within the images
-
-We would like to note that the vector diffusion maps algorithm has been previously used for a different type of three-dimensional data: to reconstruct a three-dimensional object (in this case, a molecule) from two-dimensional (cryo-electron microscopy) pictures \citep{singer2011viewing}.
+%The vector diffusion maps algorithm then, instead of revealing the temporal order, reveals the viewing direction of each image, which can then be used to reconstruct the three-dimensional object.
 %
-In these applications, many two-dimensional pictures of (effectively) the same object are taken from different viewing directions. 
-%
-The vector diffusion maps algorithm then, instead of revealing the temporal order,  reveals the viewing direction of each image, which can then be used to reconstruct the three-dimensional object. 
-%
-This is another approach to three-dimensional imaging which we hope to extend in the future to examples which also contain variability from the developmental progression.
+%This is another approach to three-dimensional imaging which we hope to extend in the future to examples which also contain variability from the developmental progression.
 
@@
 
-Manually ordering the images is nontrivial for researchers who are unfamiliar with the relevant developmental progression, and can be tedious and time-consuming for those who are.
+%Manually ordering the images is nontrivial for researchers who are unfamiliar with the relevant developmental progression, and can be tedious and time-consuming for those who are.
 %
-In contrast, our method requires relatively little computational time.
+%In contrast, our method requires relatively little computational time.
 %
@@
 %
-Furthermore, the computation of the pairwise rotational alignments, which accounts for the majority of the computational time, is trivially parallelizable for larger data sets, and only a subsample of the pairwise alignments need to be computed for accurate recovery of the underlying rotations \citep{singer2011angular}.
+Furthermore, the computation of the pairwise rotational alignments, which accounts for the majority of the computational time, is trivially parallelizable, and only a subsample of the pairwise alignments need to be computed for larger data sets for accurate recovery of the underlying rotations \citep{singer2011angular}.
 %
@@
 
-The requisite user intervention and parameter tuning required for our method is relatively minimal.
+The requisite user intervention and parameter tuning required for our method is relatively minor.
 %
@@
 
-Already at this point, the rate of collection of imaging data has surpassed the rate of manual image analysis.
+%Already at this point, the rate of collection of imaging data has surpassed the rate of manual image analysis.
 %
-This necessitates the development of automated methodologies to organize such large data sets. 
+%This necessitates the development of automated methodologies to organize such large data sets.
 %
-To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other. 
+%To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other.
 %
-Temporal ordering of large-scale data which did not require registration was done in the context of molecular profiling studies, in which data are vectors describing the expression levels of different mRNA \citep{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
+Temporal ordering of large-scale data was done in the context of molecular profiling studies, in which data are vectors describing the expression levels of different mRNA \citep{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
 %
-At the same time, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \citep{yuan2014automated, surkova2008characterization}, or using some {\em a priori} knowledge of the relevant developmental processes \citep{dubuis2013accurate}.
+At the same time, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \citep{yuan2014automated, surkova2008characterization, fowlkes2008quantitative}, or using some {\em a priori} knowledge of the relevant developmental processes \citep{dubuis2013accurate}.
 %
@@
 %
-We have applied our methodology to several different proprotypical data sets in developmental biology; because our methodology only relies on computing distances between images, we are confident it may be readily extended to many other data sets.
+
+Angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy images \citep{singer2012vector, zhao2014rotationally, singer2011viewing}.
 %
-The benefits of our algorithm are twofold. 
+Because of high levels of instrument noise in these data, thousands of images were needed for successful shape reconstruction.
+%
+Based on the presented results, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories from snapshots of fixed tissues.
+%
+In general, the size of the data set required for accurate registration and ordering is a function of the instrument noise, interembryo variability, and the complexity of the developmental dynamics.
+
+%
+The benefits of our approach to image data mining are twofold.
 %
@@
 %
-Furthermore, because our methodology is nonlinear, it can successfully order data sets which contain complex dynamics (see \fig~\ref{fig:PCA_compare} for a comparison of ordering using linear principal component analysis \citep{shlens2005tutorial} versus vector diffusion maps for the data sets presented in this paper); we expect nonlinear techniques to be necessary for larger data sets which span a wider dynamic range.
+Furthermore, because our methodology is nonlinear, it can successfully order data sets which contain complex dynamics (see \fig~\ref{fig:PCA_compare} for a comparison of ordering using linear principal component analysis \citep{shlens2005tutorial} versus vector diffusion maps for the data sets presented in this paper).
 %
-We believe that the main utility of our proposed methodology lies in the analysis of data sets containing hundreds of images from systems which have not been well-studied. 
+We expect nonlinear techniques to be necessary for larger data sets which span a wider dynamic range.
 %
-For such data sets, manually uncovering the relevant features to describe the dynamics and order the images can be nontrivial, and our algorithms can clearly accelerate such analysis.
+The main utility of our proposed methodology lies in the analysis of data sets containing hundreds of images from systems which have not been well-studied.
+%
+For such data sets, manual ordering of the images can be nontrivial, and our algorithms can clearly accelerate uncovering the underlying developmental dynamics.
 
@@
 %
-In all of our examples, the relevant expression patterns and morphological structures span several pixels and are large compared to both the instrument noise and embryo-to-embryo variability. 
+In all of our examples, the relevant expression patterns and morphological structures span several pixels and are large compared to both the instrument noise and embryo-to-embryo variability, making the Euclidean distance between pixels a good measure of images similarity.
 %
-This means that the Euclidean distance between pixels is a good measure of similarity between images.
+%The methodology we have presented here only addresses images taken in the same location and/or from the same viewing direction within the organism.
 %
-Our algorithms will fail for other imaging data sets which contain fine-scale structures,  as small perturbations in these structures will result in large changes in the Euclidean distances. 
+%However, for organisms with distinct major and minor axes (such as {\em Drosophila} embryos and wing discs), sample preparation will often fix one of the principal directions.
 %
-In such cases, one requires a more sophisticated distance metric (such as the Earth Mover's Distance \citep{rubner2000earth}) and/or image description (such as scattering transform coefficients \citep{mallat2012group}) which is insensitive to such perturbations.
+%Furthermore, as we have acknowledged, our methods as implemented cannot register three-dimensional data with respect to three-dimensional rotations due to computational limitations.
 %
-The methodology we have presented here specifically addresses images taken in the same location and/or from the same viewing direction within the organism. 
-%
-However, for organisms with distinct major and minor axes (such as {\em Drosophila} embryos and wing discs), sample preparation will often fix one of the principal directions. 
-%
-Furthermore, as we have acknowledged, our methods as implemented cannot register full three-dimensional data with respect to rotations due to computational limitations. 
-%
-However, the algorithms presented here could be used for three-dimensional voxel data given an alternative method for computing pairwise alignments more efficiently (such as using image features or landmarks).
-
-Unlike traditional statistical methods, our algorithms do not provide confidence intervals on predicted orderings, but these could in principle be extracted from bootstrapping \citep{efron1979bootstrap}.
-%
-Furthermore, although our methods provide a temporal ordering for images, they provide no information about the actual time, neither do they provide a direction for the proposed ordering.
-%
-Determining the trajectory rate and direction are post-processing tasks which require {\em a priori} knowledge of the developmental processes or some assumptions about the temporal sampling of the data set (e.g., the data set uniformly samples some given time window).
-
-Angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy images \citep{singer2012vector, zhao2014rotationally, singer2011viewing}.
-%
-Because of high levels of instrument noise in these data, thousands of images were needed for successful shape reconstruction. 
-%
-Based on the presented results, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories from snapshots of fixed tissues.
-%
-In general, the size of the data set required for accurate registration and ordering is a function of the instrument noise, interembryo variability, and the complexity of the developmental dynamics.
+%In the future, the algorithms presented here could be used for three-dimensional voxel data given an alternative method for computing pairwise alignments more efficiently (such as using image features or landmarks).
 %
 
+%Our algorithms will fail for other imaging data sets which contain fine-scale structures,  as small perturbations in these structures will result in large changes in the Euclidean distances.
+%
+%In such cases, one requires a more sophisticated distance metric (such as the Earth Mover's Distance \citep{rubner2000earth}) and/or image description (such as scattering transform coefficients \citep{mallat2012group}) which is insensitive to such perturbations.
+%
+%Unlike traditional statistical methods, our algorithms do not provide confidence intervals on predicted orderings, but these could in principle be extracted from bootstrapping \citep{efron1979bootstrap}.
+%
+%Furthermore, although our methods provide a temporal ordering for images, they provide no information about the actual time, neither do they provide a direction for the proposed ordering.
+%
+%Determining the trajectory rate and direction are post-processing tasks which require {\em a priori} knowledge of the developmental processes or some assumptions about the temporal sampling of the data set (e.g., the data set uniformly samples some given time window).
+
+%
 Vector diffusion maps allow us to automatically register images, an essential task for many applications.
@@
 %
-In general, we can recover several parameters which concisely and comprehensively describe the data set.
+In general, we can recover several coordinates which concisely and comprehensively describe the data set.
 %
-This parameterization can then be used for typical data analysis tasks, such as outlier detection and model fitting.
+This parametrization can then be used for typical data analysis tasks, such as outlier detection and model fitting.
 %
@@
 %
-Furthermore, images taken from different viewing directions can be analyzed, as the vector diffusion maps parameterization will organize the images according to the viewing angle \cite{singer2011viewing}.
+Furthermore, images taken from different viewing directions can be analyzed, as the vector diffusion maps parametrization will organize the images according to the viewing angle \citep{singer2011viewing}.
 %
@@
 
-\subsection*{{\em Drosophila} experiments}
+\subsection*{{\em Drosophila} embryo experiments}
 %
@@
 
-\subsection*{{\em Drosophila} microscopy}
+\subsection*{{\em Drosophila} embryo microscopy}
 %

### 2015-02-10 image_analysis_paper_development2/manuscript.tex
updated wing disc

@@
 
-Dynamics of developmental progress is commonly reconstructed from imaging snapshots of chemical or mechanical processes in fixed embryos.
+Dynamics of developmental progress are commonly reconstructed from imaging snapshots of chemical or mechanical processes in fixed embryos.
 %
@@
 %
-We will use the algorithm to register images of developing tissues with respect to rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
+We will use the algorithm to register images of developing tissues with respect to planar rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
 %
@@
 (D) The data in C, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure.
-The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
+The color intensity is monotonic with the perceived curve arclength, thus parametrizing the curve.}
 \label{fig:schematics}
@@
 %
-We use {\it local} information about the data to find a parametrization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parameterization.
+We use {\it local} information about the data to find a parametrization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parametrization.
 %
@@
 %
-Furthermore, the corresponding eigenvalues will allow us to confirm our assumption that our data lie on a one-dimensional manifold (see \fig~\ref{fig:eigenvalues_drosophila_live}--\ref{fig:eigenvalues_wing_disc}).
+Furthermore, the corresponding eigenvalues will allow us to confirm our assumption that our data approximately lie on a one-dimensional manifold (see \fig~\ref{fig:eigenvalues_drosophila_live}--\ref{fig:eigenvalues_wing_disc}).
 %
@@
 (A) Selected images from a movie of zebrafish epiboly.
+Scale bar indicates 200$\mu$m.
 Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
@@
 %
+In this experiment, embryos were immobilized for imaging so that the position and orientation remained fixed \cite{kane1996zebrafish}.
+%
 At the start of the time window, cells have divided 10--11 times and are accumulated in a cell mass above the yolk.
@@
 
-As in the example of {\em Drosophila} embryo live imaging, the frames were randomly rotated and shuffled (\fig~\ref{fig:zebrafish}A).
+As in the example of {\em Drosophila} embryo live imaging, the two-dimensional frames were randomly rotated and shuffled (\fig~\ref{fig:zebrafish}A).
 %
@@
 %
-In practice, we are interested in cases where image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
+In practice, we are interested in cases where each image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
 %
@@
 (A) Maximum projections of an example three-dimensional {\em Drosophila} wing disc z-stack.
+The anterior-posterior (A-P) and dorsal-ventral (D-V) axes are indicated.
 Discs express the Dad-GFP reporter construct (green), and are stained for Spalt (red), Wingless (gray), and Patched (gray).
@@
 
-Although the theory of vector diffusion maps readily extends to three-dimensional data with full three-dimensional rotations, calculation of the pairwise alignments rapidly becomes computational prohibitive for such applications, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle).
+Although the theory of vector diffusion maps readily extends to three-dimensional data with full three-dimensional rotations, calculation of the pairwise alignments rapidly becomes computationally prohibitive for such applications, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle).
 %
-However, ordering data via diffusion maps is readily possible for registered three-dimensional data.
+However, ordering data via diffusion maps is readily possible for pre-registered three-dimensional data.
 %
@@
 
-For this particular system, the face of the imaginal disc is significantly larger than the edge of the disc, so that images will always be taken with the face towards the microscope.
+In the wing disc, the anterior-posterior and dorsal-ventral axes are significantly larger than the third principal axis (see \fig~\ref{fig:wing_disc}A). 
 %
-Therefore, we need not consider registration in all three dimensions, as the z-axis will always be aligned with the short axis of the disc.
+Therefore, we need not consider registration in all three dimensions, as the z-axis for imaging (the axis perpendicular to the imaging plane) will always be aligned with the short axis of the wing disc.
 %
-We can instead focus on registering the discs with respect to rotations only in the x-y plane.
+We can instead focus on registering the wing discs with respect to rotations only in the x-y plane.
 %
-We aligned the maximum intensity projections using angular synchronization, and then used these rotations to register the full three-dimensional in the x-y plane.
+To register the data, we first aligned the maximum intensity projections using angular synchronization.
+%
+We then used these rotations to register the full three-dimensional data in the x-y plane.
 %

### 2015-02-10 image_analysis_paper_development2/paper_figures/make_fig_wing_disc.m

@@ -38,6 +38,7 @@ R_opt = ang_synch(R, 2);
 % register images using optimal rotations
 images_registered = register_all_images(images, R_opt);
 images_registered = imrotate(images_registered, 90);
+images_registered = images_registered(:,end:-1:1,:,:,:);
 
 plot_images(images_registered, dim)
 
@@ -54,7 +55,6 @@ end
 
 images_analyzed = order_all_images(images_registered, embed_coord);
 
-return
 %%
 
 plot_images(immultiply(images_analyzed, 1.5), dim)
@@ -85,16 +85,16 @@ annotation('textarrow',  [0.05 0.05], [0.87 0.82],'color', 0.95*ones(1,3), 'stri
 annotation('textarrow',  [0.1 0.05],[0.82 0.82], 'color', 0.95*ones(1,3), 'string','x', 'headstyle','none', 'fontsize', fontsize)
 
 annotation('doublearrow',  [0 0.75], [0.76 0.76],'color', 'k', 'headlength', 5, 'headwidth', 5)
-annotation('textbox', [0.75/2-0.25 0.77 0.5 0.2], 'string', '380 \mum', 'fontsize', 5, 'VerticalAlignment','bottom','HorizontalAlignment','center', 'Margin',0, 'EdgeColor','none')
+annotation('textbox', [0.75/2-0.25 0.77 0.5 0.2], 'string', 'A-P axis; 380 \mum', 'fontsize', 5, 'VerticalAlignment','bottom','HorizontalAlignment','center', 'Margin',0, 'EdgeColor','none')
 
 annotation('doublearrow',  [0.8 0.9], [0.76 0.76],'color', 'k', 'headlength', 5, 'headwidth', 5)
 annotation('textbox', [0.85-0.25 0.77 0.5 0.2], 'string', '21 \mum', 'fontsize', 5, 'VerticalAlignment','bottom','HorizontalAlignment','center', 'Margin',0, 'EdgeColor','none')
 
-annotation('doublearrow', [0.79 0.79], [0.8 0.9],'color', 'k', 'headlength', 5, 'headwidth', 5)
-annotation('textarrow', [0.81 0.81], [0.85 0.85], 'string', '21 \mum', 'fontsize', 5, 'VerticalAlignment','bottom','HorizontalAlignment','center', 'textrotation',90, 'linestyle','none','headstyle','none','color', 0.99*ones(1,3), 'textcolor','k')
[…]

### 2015-02-11 image_analysis_paper_development2/manuscript.tex
added yannis/amit comments

@@
 %
-We first compute the angles needed to align pairs of vectors (or images), which in general requires no notion of a template function \citep{ahuja2007template}.
+We first compute the angles needed to align pairs of vectors (or images), which in general requires no notion of a template function \citep{ahuja2007template, sonday2013noisy}.
 %
-In this work, we aligned pairs of images with respect to rotations by exhaustively searching over a discretized space of rotation angles.
+In this work, we aligned pairs of images with respect to rotations by exhaustively searching over a discretized space of rotation angles to minimize the Euclidean distance between the pixels.
 %
@@
 %
-At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \citep{singer2011angular}.
+At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments of the images, as well as the size of the data set \citep{singer2011angular}.
 %
@@
 %
-In this experiment, embryos were immobilized for imaging so that the position and orientation remained fixed \cite{kane1996zebrafish}.
+In this experiment, embryos were immobilized for imaging so that the position and orientation remained fixed \citep{kane1996zebrafish}.
 %
@@
 
-\fig~\ref{fig:drosophila_fixed_images}B shows the selected images in \fig~\ref{fig:drosophila_fixed_images}A, now registered and ordered using vector diffusion maps \citep{singer2012vector}.
+\fig~\ref{fig:drosophila_fixed_images}B shows the selected images in \fig~\ref{fig:drosophila_fixed_images}A, now registered and ordered using vector diffusion maps.
 %
@@
 
-Although the theory of vector diffusion maps readily extends to three-dimensional data with full three-dimensional rotations, calculation of the pairwise alignments rapidly becomes computationally prohibitive for such applications, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle).
+In this section we show that the approach can readily be applied to three-dimensional data. 
 %
-However, ordering data via diffusion maps is readily possible for pre-registered three-dimensional data.
+We restrict ourselves to the case where an obvious fixed axis exists, so that only rotations of the three-dimensional data around this axis need be taken into account.  
 %
-Therefore, we can apply our methodology to three-dimensional data, such as a stack of confocal images taken at different depths within the sample, by first registering the data using an alternative method, and then ordering the registered data using diffusion maps.
+This does not constitute an inherent limitation for vector diffusion maps. 
+%
+While for simplicity here we will not discuss the general  case, incorporating general 3D symmetries is possible \citep{arie2012global, wang2013exact, cucuringu2012eigenvector}.
+
+%Although the theory of vector diffusion maps readily extends to three-dimensional data with full three-dimensional rotations, calculation of the pairwise alignments rapidly becomes computationally prohibitive for such applications, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle).
+%
+%However, ordering data via diffusion maps is readily possible for pre-registered three-dimensional data.
+%
+%Therefore, we can apply our methodology to three-dimensional data, such as a stack of confocal images taken at different depths within the sample, by first registering the data using an alternative method, and then ordering the registered data using diffusion maps.
 
@@
 %
-Even with such obstructions, we can order the images with high accuracy and extract a stereotypical developmental trajectory, shown in \fig~\ref{fig:wing_disc}C, by averaging (see \SI).
+Even with such obstructions, we can accurately order the images and extract a stereotypical developmental trajectory, shown in \fig~\ref{fig:wing_disc}C, by averaging (see \SI).
 %
-We can now easily see the growth of the wing disc; however, averaging several images blurs the finer scale structures within the images
+We can now easily see the growth of the wing disc, even though averaging somewhat blurs some finer scale structures.
 %
-Thus, this result shows that the algorithm can successfully order a three-dimensional data set.
+%Thus, this result shows that the algorithm can successfully order a three-dimensional data set.
 
@@
 %
-Temporal ordering of large-scale data was done in the context of molecular profiling studies, in which data are vectors describing the expression levels of different mRNA \citep{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
+Temporal ordering of large-scale data was done in the context of molecular profiling studies, in which data points are vectors describing the expression levels of different mRNA \citep{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
 %
@@
 %
-Another direction for future work is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \citep{krzic2012multiview, ichikawa2014live, rubel2010coupling}.
+Another direction for future work is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \citep{krzic2012multiview, ichikawa2014live, rubel2010coupling, dsilva2013nonlinear}.
 %
@@
 
-Images were subsampled, normalized, blurred, and centered prior to diffusion maps analysis to remove any variations due to the epxperimental and imaging framework.
+Images were subsampled, normalized, blurred, and centered prior to diffusion maps analysis to remove any variations due to the experimental and imaging framework.
 %

### 2015-02-11 image_analysis_paper_development2/manuscript.tex
small edits to paper

@@
 %
-We also provide a software implementation of our methodology to aid in the application of our techniques to other experimental data sets.
+We also provide software to aid in the application of our methodology to other experimental data sets.
 \vspace{1em}
@@
 %
-Although in general, images may contain variations due to rotations, translations, and scaling, we will remove the relevant translational and/or scaling symmetries via relatively simple image preprocessing, and focus only on factoring out rotations using the vector diffusion maps algorithm.
+Although in general, images may contain variations due to rotations, translations, and scaling, we will remove the relevant translations and/or scaling via relatively simple image preprocessing, and focus only on factoring out rotations using the vector diffusion maps algorithm.
 %
@@
 %
-However, pairwise alignments can be computed by aligning appropriate image landmarks or features \citep{ian1998statistical}.
+However, pairwise alignments can also be computed by aligning appropriate image landmarks or features \citep{ian1998statistical}.
 %
@@
 %
-In the original data set, each disc was assigned to 6 time classes (72--73~hr, 76.5--77.5~hr, 79--80~hr, 89--90~hr, 100--101~hr, and 110.5--111.5~hr after fertilization); these times are indicated in \fig~\ref{fig:wing_disc}B.
+In the original data set, each disc was assigned to one of six time classes (72--73~hr, 76.5--77.5~hr, 79--80~hr, 89--90~hr, 100--101~hr, and 110.5--111.5~hr after fertilization) by an expert; these times are indicated in \fig~\ref{fig:wing_disc}B.
 %
@@
 %
-There are some registration errors in the early images, which are primarily due to some wing discs having extra tissue attached to them (such as the image in \fig~\ref{fig:wing_disc}A and the fourth image in \fig~\ref{fig:wing_disc}B).
+The registration errors are primarily due to some wing discs having extra tissue attached to them (such as the image in \fig~\ref{fig:wing_disc}A and the fourth image in \fig~\ref{fig:wing_disc}B).
 %
@@
 %
-Furthermore, because our methodology is nonlinear, it can successfully order data sets which contain complex dynamics (see \fig~\ref{fig:PCA_compare} for a comparison of ordering using linear principal component analysis \citep{shlens2005tutorial} versus vector diffusion maps for the data sets presented in this paper).
+Furthermore, because our methodology is nonlinear, it can successfully order data sets which contain complex dynamics (see \fig~\ref{fig:PCA_compare} for a comparison of ordering using linear principal component analysis versus vector diffusion maps for the data sets presented in this paper).
 %

### 2015-02-11 image_analysis_paper_development2/manuscript.tex
added stas edits

@@
 
-Dynamics of developmental progress are commonly reconstructed from imaging snapshots of chemical or mechanical processes in fixed embryos.
+Dynamics of developmental progress is commonly reconstructed from imaging snapshots of chemical or mechanical processes in fixed embryos.
 %
@@
 %
-The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require \textit{a priori} knowledge of image features or a parametric model of the developmental dynamics.
+The approach is based on vector diffusion maps, a manifold learning technique that does not require \textit{a priori} knowledge of image features or a parametric model of the developmental dynamics.
 %
@@
 %
-Importantly, the ``age'' of any given embryo arrested in its development is often only approximately known; typically what is known is
+Importantly, the ``age'' of any given embryo arrested in its development is often only approximately known.
+%
+Typically, what is known is
 a certain time window to which a collection of embryos belongs \citep{ng2012large, richardson2014emage, castro2009automatic}.
@@
 %
-In contrast to this example, real data poses nontrivial challenges for both registration and temporal ordering.
+In contrast to this example, real data pose nontrivial challenges for both registration and temporal ordering.
 %
@@
 %
-To illustrate our approach, we analyze four data sets common in developmental biology.
+To illustrate our approach, we analyze four experimental data sets.
 %
@@
 %
-We will also show how to compute an average trajectory from a set of registered and ordered fixed images to remove noise due to intersample variability and obtain a smooth description of the underlying developmental dynamics.
+We also show how to compute an average trajectory from a set of registered and ordered fixed images to remove noise due to intersample variability and obtain a smooth description of the underlying developmental dynamics.
 
@@
 technique developed for data sets which contain two sources of variability:
-geometric symmetries (such as rotations of the images) which one would like to factor out,
-and ``additional" directions of variability (such as temporal dynamics) which one would like to uncover.
+geometric symmetries, such as rotations of the images, which one would like to factor out,
+and ``additional" directions of variability, such as temporal dynamics, which one would like to uncover.
 %
@@
 %
-In the case that {\em all} relevant symmetries can be removed with straightforward preprocessing, our algorithms can be used simply to extract the main direction of variability within the imaging data set.
+In the case that {\em all} relevant symmetries can be removed with straightforward preprocessing, our algorithms can extract the main direction of variability within the imaging data set.
 %
@@
 %
-As a consequence, uncovering this direction should reveal the underlying developmental dynamics.
+As a consequence, uncovering this direction should reveal the underlying dynamics.
 
@@
 %
-Furthermore, the corresponding eigenvalues will allow us to confirm our assumption that our data approximately lie on a one-dimensional manifold (see \fig~\ref{fig:eigenvalues_drosophila_live}--\ref{fig:eigenvalues_wing_disc}).
+Furthermore, the corresponding eigenvalues will allow us to test our assumption that our data approximately lie on a one-dimensional manifold (see \fig~\ref{fig:eigenvalues_drosophila_live}--\ref{fig:eigenvalues_wing_disc}).
 %
@@
 %
-In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{fig:schematics}C) is well-sampled.
+%In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{fig:schematics}C) is well-sampled.
 %
-At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments of the images, as well as the size of the data set \citep{singer2011angular}.
+%At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments of the images, as well as the size of the data set \citep{singer2011angular}.
 %
@@
 %
-However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set.
+Registered and ordered images of individual embryos can then be used to construct a representative average trajectory. 
 %
-To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI).
+%However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set.
 %
-Each snapshot in the average trajectory is the (weighted) average of a group of successive images from the registered and ordered data set.
+%To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI).
+%
+Each snapshot in the average trajectory is the (weighted) average of a group of successive images from the registered and ordered data set (see \SI).
 %
@@
 
-From this average trajectory, we can now easily see the developmental progression, which is consistent with the known dynamics.
+From this average trajectory, we can now easily see the developmental progression consistent with the known dynamics:
 %
-dpERK first appears as two lateral peaks at the ventrolateral side of the embryo; a third dpERK peak then appears at the dorsal side of the embryo.
+dpERK first appears as two lateral peaks at the ventrolateral side of the embryo, and a third dpERK peak then appears at the dorsal side of the embryo.
 %
@@
 %
-Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental features (e.g., correlation of age with body size).
+Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks and without {\it a priori} knowledge of developmental features.
 %
@@
 
-To demonstrate this approach, we used an existing three-dimensional data set of fixed {\em Drosophila} wing imaginal discs, which was published in \citet{hamaratoglu2011dpp}.
+To demonstrate this approach, we used an existing three-dimensional data set of fixed {\em Drosophila} wing imaginal discs \citep{hamaratoglu2011dpp}.
 %
-Imaginal discs are groups of progenitor cells in fly larva that will transform into specific organs (such as wings, legs, and eyes) in the adult fly during the pupal metamorphosis.
+Imaginal discs are groups of progenitor cells in fly larva that will transform into specific organs during metamorphosis.
 %
@@
 %
-The discs were dissected from larvae expressing the Dad-GFP reporter construct (green) and stained with antibodies that recognize Spalt (red), Wingless (gray), and Patched (gray), the molecules that play important roles in disc patterning and growth (\fig~\ref{fig:wing_disc}A).
+The discs were dissected from larvae expressing the Dad-GFP reporter construct (green) and stained with antibodies that recognize Spalt (red), Wingless (gray), and Patched (gray), the factors that play important roles in disc patterning and growth (\fig~\ref{fig:wing_disc}A).
 %
@@
 %
-Therefore, we need not consider registration in all three dimensions, as the z-axis for imaging (the axis perpendicular to the imaging plane) will always be aligned with the short axis of the wing disc.
-%
-We can instead focus on registering the wing discs with respect to rotations only in the x-y plane.
+Therefore, we need not consider registration in all three dimensions, and can instead focus on registering the wing discs with respect to rotations only in the x-y plane.
 %
@@
 %
-We can now easily see the growth of the wing disc, even though averaging somewhat blurs some finer scale structures.
+We can now clearly see the growth of the wing disc, even though averaging somewhat blurs some finer scale structures.
 %
@@
 %
-Data which comes from several different genetic backgrounds can be clustered according to their spatial expression patterns, and clustering can also be used to separate images into different developmental stages.
+%Data which comes from several different genetic backgrounds can be clustered according to their spatial expression patterns, and clustering can also be used to separate images into different developmental stages.
 %
@@
 Another direction for future work is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \citep{krzic2012multiview, ichikawa2014live, rubel2010coupling, dsilva2013nonlinear}.
+%
+In the future, it would also be interesting to explore the connections between our proposed approach and recently developed methods for ordering and classification of face images \citep{kemelmacher2011exploring, kemelmacher2014illumination}
 %

### 2015-02-11 image_analysis_paper_development2/manuscript.tex
added stas edits

@@
 \author{Carmeline~J.~Dsilva \thanks{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
-Bomyi~Lim \footnotemark[1],
+Bomyi~Lim \footnotemark[1], \\
 Hang~Lu \thanks{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
-Amit~Singer \thanks{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \thanks{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA}, Ioannis~G.~Kevrekidis \footnotemark[1] \footnotemark[4] ,\\ and
+Amit~Singer \thanks{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \thanks{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA}, Ioannis~G.~Kevrekidis \footnotemark[1] \footnotemark[4] ,and
 Stanislav~Y.~Shvartsman \footnotemark[1] \thanks{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA}}
@@
 Both the angles and the ranks are recovered with a high degree of accuracy.
+%
+We note that determining which end of the trajectory corresponds to early in the developmental progression is a post-processing task that requires some {\em a priori} information.
 
@@
 %
-In the future, it would also be interesting to explore the connections between our proposed approach and recently developed methods for ordering and classification of face images \citep{kemelmacher2011exploring, kemelmacher2014illumination}
+It would also be interesting to explore the connections between our proposed approach and recently developed methods for ordering and classification of face images \citep{kemelmacher2011exploring, kemelmacher2014illumination}
 %
@@
 %
-Details about the preprocessing operations applied to each imging data set are given in \SI.
+Details about the specific preprocessing operations applied to each imging data set are given in \SI.
 

### 2015-02-11 image_analysis_paper_development2/paper_figures/fig3.tex
added compiled figures

--- 2015-02-11 image_analysis_paper_development2/manuscript.tex
@@
-\documentclass[twocolumn, 10pt]{article}
+\documentclass[10pt]{article}
 
-\usepackage[margin=0.75in]{geometry}
-\usepackage{amssymb,amsfonts,amsmath}
+\usepackage[paperwidth=18cm, paperheight=8cm, margin=0cm]{geometry}
 \usepackage{graphicx}
-
-\usepackage{color}
-
-\usepackage{epstopdf}
-
-\usepackage{natbib}
-
-\renewcommand{\figurename}{Fig.}
-%\newcommand{\SI}[0]{\textit{SI Materials and Methods}}
-\newcommand{\SI}[0]{supplementary material}
-\newcommand{\fig}[0]{Fig.}
-
 \hyphenpenalty=10000
-
-\makeatletter
-\newcommand{\customlabel}[2]{%
-\protected@write \@auxout {}{\string \newlabel {#1}{{#2}{}}}}
-\makeatother
-
-\graphicspath{{paper_figures/}}
-
-\customlabel{fig:DV_view}{S1}
-\customlabel{fig:1d_example}{S2}
-\customlabel{fig:eigenvalues_drosophila_live}{S3}
-\customlabel{fig:eigenvalues_zebrafish}{S4}
-\customlabel{fig:eigenvalues_drosophila_fixed}{S5}
-\customlabel{fig:eigenvalues_wing_disc}{S6}
-\customlabel{fig:PCA_compare}{S7}
-
-\title{Temporal ordering and registration of images in studies of developmental dynamics}
-
-\author{Carmeline~J.~Dsilva \thanks{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
-Bomyi~Lim \footnotemark[1],
-Hang~Lu \thanks{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
-Amit~Singer \thanks{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \thanks{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA}, Ioannis~G.~Kevrekidis \footnotemark[1] \footnotemark[4] ,\\ and
-Stanislav~Y.~Shvartsman \footnotemark[1] \thanks{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA}}
-
-\date{}
 
@@
 
-\maketitle
-
-\begin{abstract}
-
-Dynamics of developmental progress are commonly reconstructed from imaging snapshots of chemical or mechanical processes in fixed embryos.
-%
-As a first step in these reconstructions, snapshots must be spatially registered and ordered in time.
-%
-Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific system.
-%
-However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
-%
-To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
-%
-The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require \textit{a priori} knowledge of image features or a parametric model of the developmental dynamics.
-%
-We illustrate this approach by registering and ordering data from imaging studies of pattern formation and morphogenesis in three different model systems.
-%
-We also provide software to aid in the application of our methodology to other experimental data sets.
-\vspace{1em}
-
-\noindent
-KEY WORDS: temporal ordering, image registration, vector diffusion maps
-
-\end{abstract}
-
-\section*{Introduction}
-
-\begin{figure}[t]
-\includegraphics{fig1}
-\caption{Caricature illustrating the tasks of image registration and temporal ordering.
-(A) Images of ``samples'', each in a different orientation and a different stage of development.
-(B) Registered and ordered samples.
-For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
-%\label{fig:fish}
-\customlabel{fig:fish}{1}
-\customlabel{subfig:fish_unordered}{\ref{fig:fish}A}
-\customlabel{subfig:fish_ordered}{\ref{fig:fish}B}
-\end{figure}
-
-In one of the common approaches to studies of developmental dynamics, a group of embryos is fixed and stained to visualize a particular biochemical or morphological process within a developing tissue.
-%
-The developmental dynamics must then be reconstructed from multiple embryos, each of which contributes only a snapshot of the relevant process along its developmental trajectory \citep{jaeger2004dynamic, peter2011gene, fowlkes2008quantitative}.
-%
-Importantly, the ``age'' of any given embryo arrested in its development is often only approximately known; typically what is known is
-a certain time window to which a collection of embryos belongs \citep{ng2012large, richardson2014emage, castro2009automatic}.
-%
-Furthermore, images are often collected in different spatial orientations.
-%
-In order to recover the developmental dynamics from such data sets, snapshots of different embryos must first be spatially aligned or {\em registered}, and then ordered in time.
-%
-%We show how recently developed dimensionality reduction algorithms can automate {\it and combine} both of these tasks.
-
-Temporal ordering and registration of images can be done manually
-when the number of images is small and the differences between them are visually apparent.
-%
-\fig~\ref{fig:fish} shows a caricature of fish development which illustrates the processes of growth and patterning.
-%
-In this case, temporal ordering can be accomplished by arranging the fish by size, which is monotonic with the developmental progress.
-%
-Image registration is based on obvious morphological landmarks, such as the positions of the head and the fins.
-%
-In contrast to this example, real data poses nontrivial challenges for both registration and temporal ordering.
-%
-In general, the landmarks needed for registration, as well as the attributes which can be used to order the data, are not known {\it a priori}.
-%
-Additional challenges arise from embryo-to-embryo variability, sample size, and measurement noise.
-
-We present a robust algorithmic approach to simultaneous registration and temporal ordering.
-%
-In contrast to a number of previous methodologies \citep{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face, dubuis2013accurate}, our methodology does not rely on the {\em a priori} knowledge of landmarks for registration or markers of developmental progression.
-%
-The approach is based on vector diffusion maps \citep{singer2012vector}, a manifold learning algorithm which simultaneously addresses the problems of registration and temporal ordering.
-%
-This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \citep{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
-applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \citep{zhao2014rotationally, singer2011viewing} to face recognition \citep{lafon2006data} and classification of CT scans \citep{fernandez2014diffusion}.
-
-Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing tissues in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
-%
-To illustrate our approach, we analyze four data sets common in developmental biology.
-%
-Our first two data sets come from live imaging studies of {\em Drosophila} and zebrafish embryogenesis.
-%
-In both of these examples, the correct rotational orientation and temporal order are independently known, and these data sets will be used to validate our approach.
-%from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
-%
-Our third data set consists of images from fixed {\em Drosophila} embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.
-%
-Our final data set consists of z-stacks of {\em Drosophila} wing discs, which we will use to illustrate how our methods can be used to analyze specific types of three-dimensional imaging data.
-%
-We will also show how to compute an average trajectory from a set of registered and ordered fixed images to remove noise due to intersample variability and obtain a smooth description of the underlying developmental dynamics.
-
-\section*{Results}
-
-\subsection*{Vector diffusion maps for registration and temporal ordering}
-
-Vector diffusion maps \citep{singer2012vector} is a manifold learning
-technique developed for data sets which contain two sources of variability:
-geometric symmetries (such as rotations of the images) which one would like to factor out,
-and ``additional" directions of variability (such as temporal dynamics) which one would like to uncover.
-%
-Vector diffusion maps combine two algorithms, {\em angular synchronization} \citep{singer2011angular} for image registration and {\em diffusion maps} \citep{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
-%
-We will use the algorithm to register images of developing tissues with respect to planar rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
-%
-Although in general, images may contain variations due to rotations, translations, and scaling, we will remove the relevant translations and/or scaling via relatively simple image preprocessing, and focus only on factoring out rotations using the vector diffusion maps algorithm.
-%
-In the case that {\em all} relevant symmetries can be removed with straightforward preprocessing, our algorithms can be used simply to extract the main direction of variability within the imaging data set.
-%
-We assume that this main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
-%
-As a consequence, uncovering this direction should reveal the underlying developmental dynamics.
-
-\begin{figure}[t]
-\includegraphics{fig2}
-\caption{Schematic illustrating angular synchronization and diffusion maps.
-(A) Set of vectors, each in a different orientation.
-The pairwise alignment angles are indicated.
-(B) The vectors from A, each rotated about their midpoint so that the set is globally aligned.
-Note that the chosen rotation angles are consistent with the pairwise alignments in A: the difference between a pair of angles in B is the same as the pairwise angle in A.
-(C) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
-Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through a Gaussian kernel (see \SI), so that pairs of data point which are close are connected by darker (``stronger'') edges.
-(D) The data in C, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure.
-The color intensity is monotonic with the perceived curve arclength, thus parametrizing the curve.}
-\label{fig:schematics}
-\end{figure}
-
-Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
-%
-A schematic illustration of angular synchronization is shown in \fig~\ref{fig:schematics}A, where each image is represented as a vector, and the goal is to align the entire set of vectors given pairwise alignment measurements.
-%
-We first compute the angles needed to align pairs of vectors (or images), which in general requires no notion of a template function \citep{ahuja2007template, sonday2013noisy}.
-%
-In this work, we aligned pairs of images with respect to rotations by exhaustively searching over a discretized space of rotation angles to minimize the Euclidean distance between the pixels.
-%
-However, pairwise alignments can also be computed by aligning appropriate image landmarks or features \citep{ian1998statistical}.
-%
-When the data are noisy, these pairwise measurements may by inaccurate, and so we utilize {\em all} pairwise measurements to align the set of images robustly.
-%
-Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see \SI); this is illustrated in \fig~\ref{fig:schematics}B.
-%
-In this schematic, registration via angular synchronization is trivial, as the pairwise measurements contain no noise.
-%
-However, the algorithm can register data sets even when many of the pairwise measurements are inaccurate \citep{singer2011angular}.
-
-After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional curve that parameterizes most of the remaining variability in the data.
-%
-Such a curve can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that reveals a parametrization of data that lies on a low-dimensional manifold in high-dimensional space.
-%
-The idea is illustrated in \fig~\ref{fig:schematics}C, where the data are two-dimensional points which lie on a one-dimensional (nonlinear) curve.
-%
-We use {\it local} information about the data to find a parametrization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parametrization.
-%
-This idea of locality is denoted by the color of the edges in \fig~\ref{fig:schematics}C:
-data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
-%
-The color in \fig~\ref{fig:schematics}D depicts the one-dimensional parametrization or ordering of the data that we can detect visually.
-%
-A detailed example of using vector diffusion maps to register and order synthetic data is given in \fig~\ref{fig:1d_example}, and a step-by-step tutorial of the diffusion maps implementation is included in the \SI.
-%
-In our working examples, each data point will be of much higher dimension (e.g., a pixelated image or three-dimensional voxel data), and so we cannot extract this low-dimensional structure visually.
-%
-Instead, we will use diffusion maps which automatically uncovers a parametrization of our high-dimensional data from the eigenvectors of the appropriate matrix (see \SI).
-%
-Furthermore, the corresponding eigenvalues will allow us to confirm our assumption that our data approximately lie on a one-dimensional manifold (see \fig~\ref{fig:eigenvalues_drosophila_live}--\ref{fig:eigenvalues_wing_disc}).
-%
-%Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.
-
-\subsection*{Method validation using live imaging}
-
-\subsubsection*{{\em Drosophila} gastrulation}
-
-\begin{figure*}[t]
 \raisebox{0.75cm}{\textsf{A}}
@@
 \end{tabular}}}}
-\caption{Method validation using live imaging of \textit{Drosophila} embryos.
-(A) Selected images from a live imaging study of a \textit{Drosophila} embryo during gastrulation.
-Scale bar indicates 50$\mu$m.
-Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
-(B) Images from A registered and ordered by vector diffusion maps.
-The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.
-(C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle.
-The average absolute error in the recovered angles is $8.37 ^\circ$.
-(D) The correlation between the recovered rank (using vector diffusion maps) and the true rank.
-The rank correlation coefficient is $0.9989$.
-(E) The average error in the recovered angle and the rank correlation coefficient for $5$ independent live imaging studies.
-}
-\label{fig:drosophila_live_imaging}
-\end{figure*}
-
-To validate the proposed approach, we first applied our algorithm to a data set where the true temporal order and rotational orientation of the images were known {\em a priori}.
-%
-This data set was obtained through live imaging near the posterior pole of a vertically oriented {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
-%
-During this time window, the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, internalizing the future muscle cells and forming a characteristic ``omega'' shape.
-%
-Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \citep{leptin2005gastrulation}.
-%
-At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing a similar ``omega'' to appear on the dorsal side.
-
-\fig~\ref{fig:drosophila_live_imaging}A shows selected images from this live imaging data set, which contains $40$ consecutive frames taken at $30$~second time intervals at a fixed position within a single embryo.
-%
-Each image shows an optical cross-section near the posterior pole of a vertically oriented developing embryo, with the nuclei labeled by Histone-RFP.
-%
-Each frame was arbitrarily rotated, and the order of the frames was scrambled.
-%
-The task is now to register these images and order them in time to reconstruct the developmental trajectory.
-
-We used vector diffusion maps to register and order the images.
-%
-\fig~\ref{fig:drosophila_live_imaging}B shows the images from \fig~\ref{fig:drosophila_live_imaging}A, now registered and ordered; the real time for each frame is also indicated.
-%
-With a small number of exceptions, the recovered ordering is consistent with the real time dynamics.
-%
-\fig~\ref{fig:drosophila_live_imaging}C and \fig~\ref{fig:drosophila_live_imaging}D  show the correlations between the recovered and true angles and rank orders, respectively, for the entire data set.
-%
-Both the angles and the ranks are recovered with a high degree of accuracy.
-
-To assess the robustness of the proposed methodology, we repeated this procedure with four additional data sets extracted from independent live imaging studies spanning the same developmental time period.
-%
-The results are shown in \fig~\ref{fig:drosophila_live_imaging}E.
-%
-The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than 95\%, indicating that our methodology can reproducibly order data of this type.
-%
-In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{fig:schematics}C) is well-sampled.
-%
-At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments of the images, as well as the size of the data set \citep{singer2011angular}.
-%
-%Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
-%%
-%To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
-%%
-%Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered.
-%%
-%\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
-%%
-%As expected, the accuracy in the recovered ordering increases with the number of images.
-%%
-%At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
-%%
-%Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignment, and the high temporal resolution of our live imaging study allows us to order the images accurately.
-%
-%In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.
-
-\subsubsection*{Zebrafish epiboly}
-
-\begin{figure*}[t]
-\begin{minipage}[c]{13.75cm}
-\vspace{0cm}
-\raisebox{0.5cm}{\textsf{A}}
-\includegraphics[width=13cm]{zebrafish_scrambled}
-
-\raisebox{0.5cm}{\textsf{B}}
-\includegraphics[width=13cm]{zebrafish_ordered}
-\end{minipage}
-%
-\hfill
-%
-\begin{minipage}[c]{3.75cm}
-
-\vspace{0cm}
-\raisebox{2.5cm}{\textsf{C}}
-%
-\includegraphics[width=3.25cm]{zebrafish_rank_corr}
-\end{minipage}
-\caption{Method validation using live imaging of a zebrafish embryo.
-(A) Selected images from a movie of zebrafish epiboly.
-Scale bar indicates 200$\mu$m.
-Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
-(B) Images from A after registration and ordering using vector diffusion maps.
-The real time of each frame is also indicated.
-(C) Correlation between the rank recovered using vector diffusion maps and the true rank.
-The rank correlation coefficient is $0.9954$.
-The larger errors in the recovered ranks towards the beginning of the trajectory are due to the slow cell movement within that time window.
-}
-\label{fig:zebrafish}
-\end{figure*}
-
-As another validation for the proposed methodology, we applied our algorithm to a time-lapse movie  of zebrafish embryogenesis.
-%
-We used a publicly available live imaging data of zebrafish embryogenesis (\texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov}, \citet{karlstrom1996flipbook}).
-%
-Taken with a differential interference contrast (DIC) microscope, the movie records the first $17$~hours of zebrafish development, from a single cell stage to a 16-somite stage.
-%
-We selected $120$ consecutive frames from this movie which capture 5.5~hours of epiboly (3.5--9~hours after fertilization).
-%
-In this experiment, embryos were immobilized for imaging so that the position and orientation remained fixed \citep{kane1996zebrafish}.
-%
-At the start of the time window, cells have divided 10--11 times and are accumulated in a cell mass above the yolk.
-%
-The cell mass is then compressed and the animal-vegetal axis of the embryo (vertical axis in \fig~\ref{fig:zebrafish}) shortens to form a spherical embryo shape by the end of the fourth hour of development.
-%
-Then, the yolk syncytial layer, which forms the boundary between the yolk and the cell mass, moves upward, forming a dome-shaped structure.
-%
-During this stage, the cells rearrange to form a uniform layer about four cells thick.
-%
-With time, this cell layer then spreads over across the yolk and expands toward the vegetal pole.
-%
-At the end of epiboly, the blastoderm completely engulfs the yolk.
-
-As in the example of {\em Drosophila} embryo live imaging, the two-dimensional frames were randomly rotated and shuffled (\fig~\ref{fig:zebrafish}A).
-%
-We then used vector diffusion maps to register and order the frames.
-%
-The results are shown in \fig~\ref{fig:zebrafish}B.
-%
-The recovered rotations and order are consistent with the expected developmental dynamics, as shown in the correlations between the recovered and true ranks (\fig~\ref{fig:zebrafish}C).
-%
-Quantitatively, the rank correlation coefficient for this data set is 0.9954, and the average error in the recovered angle is $4.14^\circ$.
-%
-Some errors in ordering images of the early embryo result from slow cell movement during the early developmental stage where cells divide and accumulate above the yolk.
-%
-During epiboly, cell movement is more dynamic and the recovered ordering is more consistent with the real dynamics.
-
-In summary, we have shown that our approach to temporal ordering performs very well on imaging data of two different developmental processes ({\em Drosophila} gastrulation and zebrafish epiboly), taken with two different imaging methods (fluorescent microscopy and DIC) where the true temporal order is known {\em a priori}.
-%image registration and ordering can correctly recover image orientation and  showed that our algorithm is not limited to a specific set of data.
-%
-%The algorithm works for different tissue types ({\em Drosophila} and zebrafish), and also for images taken with different imaging methods (fluorescent and phase contrast images).
-%
-Provided there exist significant dynamics within the data set and that the developmental trajectory is well-sampled, the developmental dynamics can be recovered.
-
-\subsection*{Data sets with intersample variability}
-
-\subsubsection*{Fixed images of {\em Drosophila} gastrulation}
-
-We have analyzed how our algorithm performs on two model data sets where all images come from a single embryo.
-%
-In practice, we are interested in cases where each image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
-%
-To demonstrate that our methods are robust to such variations,
-we constructed a synthetic time course data set by selecting a random image from one of five {\em Drosophila} live imaging data sets (those data sets used in \fig~\ref{fig:drosophila_live_imaging}) at each time point.
-%
-The resulting data set is spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability.
-%
-The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.77, indicating that the algorithm can recover the temporal order even under noisy conditions.
-
-We then applied our approach to a data set where the true rotational orientation and temporal order was not known {\it a priori}.
-%
-\fig~\ref{fig:drosophila_fixed_images}A shows selected images from a set of $120$ images of developing {\em Drosophila} embryos which cover a thirty minute time interval spanning late cellularization through gastrulation.
-%
-This data set is more complex than the live imaging data sets in that it contains significantly more images, each of which provides information about tissue morphology and the spatial distribution of two regulatory proteins.
-%
-Each image shows an optical cross-section of the posterior view of a {\em different} embryo at a different rotational orientation and fixed at a different (and unknown) developmental time.
-%
-The nuclei (gray) were labeled with DAPI, a DNA stain.
-%
-Embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor which specifies the cells of the future muscle tissue.
-%
-Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that, in this context, specifies a subset of neuronal cells \citep{Lim2013kinetics}.
-
-\fig~\ref{fig:drosophila_fixed_images}B shows the selected images in \fig~\ref{fig:drosophila_fixed_images}A, now registered and ordered using vector diffusion maps.
-%
-However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set.
-%
-To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI).
-%
-Each snapshot in the average trajectory is the (weighted) average of a group of successive images from the registered and ordered data set.
-%
-Averaging successive images removes some of the interembryo variability, so that sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.
-
-\begin{figure*}[t!]
-\begin{minipage}[b]{12cm}
-\raisebox{0.5cm}{\textsf{A}}
-\includegraphics[width=12.25cm]{drosophila_fixed_images_scrambled}
-
-\raisebox{0.5cm}{\textsf{B}}
-\includegraphics[width=12.25cm]{drosophila_fixed_images_ordered}
-
-\raisebox{0.5cm}{\textsf{C}}
-\includegraphics[width=12.25cm]{drosophila_fixed_images_average}
-\end{minipage}
-%
-\hfill
-%
-\raisebox{3.25cm}{\textsf{D}}
-\includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}
-%
-\caption{Analysis of images of fixed \textit{Drosophila} embryos.
-{(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red).
-Scale bar indicates approximately 50$\mu$m (images have been rescaled to remove slight interembryo size variations).
-Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation.
-{(B)} Data from {\it A}, registered and ordered using vector diffusion maps.
-The expert rank for each image is indicated.
-{(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI).
-{(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert.
-The rank correlation coefficient is $0.9716$.
-}
-\label{fig:drosophila_fixed_images}
-\end{figure*}
-
-From this average trajectory, we can now easily see the developmental progression, which is consistent with the known dynamics.
-%
-dpERK first appears as two lateral peaks at the ventrolateral side of the embryo; a third dpERK peak then appears at the dorsal side of the embryo.
-%
-During mesoderm invagination, the two ventrolateral dpERK peaks merge together, eventually forming, together with Twi, the ``omega'' shape.
-%
-The dorsal dpERK peak then disappears during germband extension as cells from the ventral side wrap around to the dorsal side.
-%
-At the end of this process, similar ``omegas'' formed by Twi and dpERK appear on the dorsal side of the embryo; these patterns are most readily seen in the last image of \fig~\ref{fig:drosophila_fixed_images}C.
-%
-Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental features (e.g., correlation of age with body size).
-%
-
-To evaluate the quality of our registration and ordering, we can use prior knowledge about the developmental system.
-%
-The Twi signal is known to form a single peak at the ventralmost point of the embryo.
-%
-We found that the standard deviation in the location of this peak in the set of registered images was $\sim$8$^\circ$,
-indicating that the algorithm successfully aligns the ventralmost points of the images.
-%
-Because the developmental time of each embryo cannot be easily estimated, we have few options for evaluating the quality of our temporal ordering.
-%
-We compared the ordering obtained from vector diffusion maps to the ordering provided by a trained embryologist who is knowledgeable about the developmental progression and the important image features.
-%
-The ranks from the ordering provided by the embryologist, which we will refer to as the ``expert rank'', are indicated for the images in \fig~\ref{fig:drosophila_fixed_images}B , and the rank correlation (see \fig~\ref{fig:drosophila_fixed_images}D) shows that our ordering is consistent with the expert ordering.
-
-\subsubsection*{Fixed z-stacks of {\em Drosophila} wing discs}
-
-\begin{figure*}[t]
-\raisebox{5.5cm}{\textsf{A}}
-\includegraphics[width=6cm]{wing_disc_example}
-%
-\hfill
-%
-\raisebox{5.5cm}{\textsf{B}}
-\includegraphics[width=10cm]{wing_disc_ordered}
-
-\vspace{0.25cm}
-\raisebox{2.25cm}{\textsf{C}}
-\includegraphics[width=17.5cm]{wing_disc_average}
-
-\caption{Analysis of three-dimensional {\em Drosophila} wing disc z-stacks.
-(A) Maximum projections of an example three-dimensional {\em Drosophila} wing disc z-stack.
-The anterior-posterior (A-P) and dorsal-ventral (D-V) axes are indicated.
-Discs express the Dad-GFP reporter construct (green), and are stained for Spalt (red), Wingless (gray), and Patched (gray).
-Projections along the $x$-, $y$-, and $z$-axes are shown.
-(B) Example three-dimensional images, ordered using diffusion maps.
-The time cohort, as assessed by an expert, is indicated for each image, and the rank correlation coefficient between the diffusion maps ordering and the expert timing is 0.9427.
-(C) The average developmental trajectory for the registered and ordered images.
-}
-\label{fig:wing_disc}
-\end{figure*}
-
-%We have demonstrated that our algorithm can recover the underlying developmental dynamics in several different types of imaging data sets, including ones containing interembryo variability.
-%
-%Thus far, we have only looked at two-dimensional data sets where biological samples are imaged as a single confocal slice or as a phase contrast image at a fixed position within the sample.
-%
-%However, with advances in imaging techniques, taking multiple z-slices to visualize the three-dimensional structure of a sample will become more common, and it is important to validate that our algorithm can be applied to such data.
-
-In this section we show that the approach can readily be applied to three-dimensional data. 
-%
-We restrict ourselves to the case where an obvious fixed axis exists, so that only rotations of the three-dimensional data around this axis need be taken into account.  
-%
-This does not constitute an inherent limitation for vector diffusion maps. 
-%
-While for simplicity here we will not discuss the general  case, incorporating general 3D symmetries is possible \citep{arie2012global, wang2013exact, cucuringu2012eigenvector}.
-
-%Although the theory of vector diffusion maps readily extends to three-dimensional data with full three-dimensional rotations, calculation of the pairwise alignments rapidly becomes computationally prohibitive for such applications, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle).
-%
-%However, ordering data via diffusion maps is readily possible for pre-registered three-dimensional data.
-%
-%Therefore, we can apply our methodology to three-dimensional data, such as a stack of confocal images taken at different depths within the sample, by first registering the data using an alternative method, and then ordering the registered data using diffusion maps.
-
-To demonstrate this approach, we used an existing three-dimensional data set of fixed {\em Drosophila} wing imaginal discs, which was published in \citet{hamaratoglu2011dpp}.
-%
-Imaginal discs are groups of progenitor cells in fly larva that will transform into specific organs (such as wings, legs, and eyes) in the adult fly during the pupal metamorphosis.
-%
-The wing disc is an imaginal disc that turns into a wing.
-%
-The data set is composed of 46 fixed wing discs whose developmental times range from 72 to 112~hours after fertilization.
-%
-Each disc contains 21 z-slices taken at 1~$\mu$m intervals.
-%
-The discs were dissected from larvae expressing the Dad-GFP reporter construct (green) and stained with antibodies that recognize Spalt (red), Wingless (gray), and Patched (gray), the molecules that play important roles in disc patterning and growth (\fig~\ref{fig:wing_disc}A).
-%
-%The disc grows in time; at long times, the disc becomes too large to fit completely within the frame and only a partial disc is imaged.
-
-In the wing disc, the anterior-posterior and dorsal-ventral axes are significantly larger than the third principal axis (see \fig~\ref{fig:wing_disc}A). 
-%
-Therefore, we need not consider registration in all three dimensions, as the z-axis for imaging (the axis perpendicular to the imaging plane) will always be aligned with the short axis of the wing disc.
-%
-We can instead focus on registering the wing discs with respect to rotations only in the x-y plane.
-%
-To register the data, we first aligned the maximum intensity projections using angular synchronization.
-%
-We then used these rotations to register the full three-dimensional data in the x-y plane.
-%
-Because the maximum intensity projections are two-dimensional images, this step is no more computationally intensive than the previous examples.
-%
-Such an approach is possible when there are distinct major and minor axes within a three-dimensional sample, which reduces the rotational degrees of freedom.
-%
-%Other options for registering samples include using a specific localized signal (such as Twi in \fig~\ref{fig:drosophila_fixed_images}) as a marker for alignment, or using other {\em a priori} knowledge.
-
-We then used diffusion maps to order the registered three-dimensional data.
-%
-\fig~\ref{fig:wing_disc}B shows selected images from the data set ordered by diffusion maps.
-%
-In the original data set, each disc was assigned to one of six time classes (72--73~hr, 76.5--77.5~hr, 79--80~hr, 89--90~hr, 100--101~hr, and 110.5--111.5~hr after fertilization) by an expert; these times are indicated in \fig~\ref{fig:wing_disc}B.
-%
-In the ordered set, the size of wing disc grows, and the intensity of the Dad-GFP signal increases as a function of time.
-%
-The rank correlation coefficient based on the time class is 0.9436.
-%
-The registration errors are primarily due to some wing discs having extra tissue attached to them (such as the image in \fig~\ref{fig:wing_disc}A and the fourth image in \fig~\ref{fig:wing_disc}B).
-%
-Even with such obstructions, we can accurately order the images and extract a stereotypical developmental trajectory, shown in \fig~\ref{fig:wing_disc}C, by averaging (see \SI).
-%
-We can now easily see the growth of the wing disc, even though averaging somewhat blurs some finer scale structures.
-%
-%Thus, this result shows that the algorithm can successfully order a three-dimensional data set.
-
-%The vector diffusion maps algorithm has been previously used for a different type of three-dimensional data: to reconstruct a three-dimensional object (in this case, a molecule) from two-dimensional (cryo-electron microscopy) pictures \citep{singer2011viewing}.
-%
-%In these applications, many two-dimensional pictures of (effectively) the same object are taken from different viewing directions.
-%
-%The vector diffusion maps algorithm then, instead of revealing the temporal order, reveals the viewing direction of each image, which can then be used to reconstruct the three-dimensional object.
-%
-%This is another approach to three-dimensional imaging which we hope to extend in the future to examples which also contain variability from the developmental progression.
-
-\subsection*{Computational requirements}
-
-\begin{figure*}[t]
-\raisebox{5cm}{\textsf{A}}
-\includegraphics[width=5.5cm]{timing_studies_nimages}
-%
-\raisebox{5cm}{\textsf{B}}
-\includegraphics[width=5.5cm]{timing_studies_npixels}
-%
-\raisebox{5cm}{\textsf{C}}
-\includegraphics[width=5.5cm]{timing_studies_nrot}
-
-\raisebox{2.5cm}{\textsf{D}}
-{\renewcommand{\arraystretch}{1.5}
-{ \textsf{ \small
-\begin{tabular}{ | p{2cm} | c | c | c | c | c | c | }
-\hline
-Data Set &
-Data Type &
-\parbox[c]{2cm}{ \centering Number of Channels} &
-\parbox[c]{2cm}{ \centering \vspace{.5\baselineskip} Number of Images \vspace{.5\baselineskip}} &
-\parbox[c]{2cm}{ \centering Number of Pixels} &
-\parbox[c]{2cm}{ \centering Angular Discretization} &
-\parbox[c]{2cm}{ \centering CPU time}  \\
-\hline
-{\em Drosophila} gastrulation (live) & 2D & 1 & 40 & 100 $\times$ 100 & 10$^\circ$ & 3.2~sec \\
-Zebrafish epiboly & 2D & 1 & 120 & 100 $\times$ 100 & 10$^\circ$ & 13~sec\\
-{\em Drosophila} gastrulation (fixed) & 2D & 3 & 120 & 100 $\times$ 100 & 10$^\circ$ & 29~sec  \\
-{\em Drosophila} wing discs & 3D & 3 & 46 & 100 $\times$ 100 $\times$ 21 & 10$^\circ$ & 12~sec  \\
- \hline
-\end{tabular}}}}
-\caption{Computational requirements for the presented methodology.
-(A) CPU time as a function of the number of images in the data set (for $100 \times 100$ pixel images, and $10^\circ$ angular discretization).
-Empirically, the CPU time is $\sim \mathcal{O}(n^{1.33})$ in number of images.
-(B) CPU time as a function of the number of pixels in the images (for $120$ images, and $10^\circ$ angular discretization).
-Empirically, the CPU time is $\sim \mathcal{O}(n^{1.83})$ in the number of pixels.
-(C) CPU time as a function of the number of rotations (for $120$ images of $100 \times 100$ pixels).
-Empirically, the CPU time is $\sim \mathcal{O}(n^{-0.77})$ in the angular discretization.
-(D) The algorithm settings and computational requirements for the data sets analyzed.
-All times are reported for an Intel Core i7 2.93 GHz processor.}
-\label{fig:cpu_time}
-\end{figure*}
-
-%Manually ordering the images is nontrivial for researchers who are unfamiliar with the relevant developmental progression, and can be tedious and time-consuming for those who are.
-%
-%In contrast, our method requires relatively little computational time.
-%
-The computational costs for our methodology are outlined in \fig~\ref{fig:cpu_time}.
-%
-The computational time is a function of the number of images in the data set, the number of pixels in each point, and the angular resolution to compute the pairwise rotations (see \SI).
-%
-Furthermore, the computation of the pairwise rotational alignments, which accounts for the majority of the computational time, is trivially parallelizable, and only a subsample of the pairwise alignments need to be computed for larger data sets for accurate recovery of the underlying rotations \citep{singer2011angular}.
-%
-Because the computational cost increases with the image resolution, we chose to subsample all of our data sets to $100 \times 100$ pixels.
-%
-This resolution allowed us to rapidly analyze our data sets while still retaining all of the relevant developmental features.
-%
-However, as can be seen from the computational costs in \fig~\ref{fig:cpu_time}, it is feasible to use our algorithms to analyze higher-resolution images.
-
-The requisite user intervention and parameter tuning required for our method is relatively minor.
-%
-As a first step, images must be preprocessed so that the Euclidean distance between the pixels is informative.
-%
-Our software provides several preprocessing options (such as blurring, rescaling, and mean-centering), as well as some guidance for what options to select depending on the system of interest.
-%
-Two algorithmic parameters, the angular discretization to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{fig:schematics} and \SI), must also be defined.
-%
-We also provide some guidance on selecting these parameters, and found that the results are robust to both of these parameters.
-%
-Overall, the tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, and so this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.
-
-\section*{Discussion}
-
-%Already at this point, the rate of collection of imaging data has surpassed the rate of manual image analysis.
-%
-%This necessitates the development of automated methodologies to organize such large data sets.
-%
-%To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other.
-%
-Temporal ordering of large-scale data was done in the context of molecular profiling studies, in which data points are vectors describing the expression levels of different mRNA \citep{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
-%
-At the same time, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \citep{yuan2014automated, surkova2008characterization, fowlkes2008quantitative}, or using some {\em a priori} knowledge of the relevant developmental processes \citep{dubuis2013accurate}.
-%
-%The task of image registration has been widely studied \citep{zitova2003image}, for applications such as face recognition \citep{rowley1998rotation}, medical image registration \citep{hajnal2010medical}, and texture classification \citep{greenspan1994rotation}.
-%
-In contrast to most of the existing registration approaches which rely on the knowledge of appropriate landmarks in the images \citep{ian1998statistical} (such as the eyes in face recognition applications \citep{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them relevant for a wide variety of applications.
-%
-
-Angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy images \citep{singer2012vector, zhao2014rotationally, singer2011viewing}.
-%
-Because of high levels of instrument noise in these data, thousands of images were needed for successful shape reconstruction.
-%
-Based on the presented results, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories from snapshots of fixed tissues.
-%
-In general, the size of the data set required for accurate registration and ordering is a function of the instrument noise, interembryo variability, and the complexity of the developmental dynamics.
-
-%
-The benefits of our approach to image data mining are twofold.
-%
-First, the algorithm can accomplish the tasks of registration and ordering in a single step.
-%
-Furthermore, because our methodology is nonlinear, it can successfully order data sets which contain complex dynamics (see \fig~\ref{fig:PCA_compare} for a comparison of ordering using linear principal component analysis versus vector diffusion maps for the data sets presented in this paper).
-%
-We expect nonlinear techniques to be necessary for larger data sets which span a wider dynamic range.
-%
-The main utility of our proposed methodology lies in the analysis of data sets containing hundreds of images from systems which have not been well-studied.
-%
-For such data sets, manual ordering of the images can be nontrivial, and our algorithms can clearly accelerate uncovering the underlying developmental dynamics.
-
-We acknowledge that our methods, though general, do have limitations.
-%
-The first is that we require enough data to sufficiently sample the developmental trajectory.
-%
-Therefore, for very small and/or very noisy data sets, our algorithms may fail.
-%
-Second, the pertinent image features need to be large compared to the noise and the image resolution.
-%
-In all of our examples, the relevant expression patterns and morphological structures span several pixels and are large compared to both the instrument noise and embryo-to-embryo variability, making the Euclidean distance between pixels a good measure of images similarity.
-%
-%The methodology we have presented here only addresses images taken in the same location and/or from the same viewing direction within the organism.
-%
-%However, for organisms with distinct major and minor axes (such as {\em Drosophila} embryos and wing discs), sample preparation will often fix one of the principal directions.
-%
-%Furthermore, as we have acknowledged, our methods as implemented cannot register three-dimensional data with respect to three-dimensional rotations due to computational limitations.
-%
-%In the future, the algorithms presented here could be used for three-dimensional voxel data given an alternative method for computing pairwise alignments more efficiently (such as using image features or landmarks).
-%
-
-%Our algorithms will fail for other imaging data sets which contain fine-scale structures,  as small perturbations in these structures will result in large changes in the Euclidean distances.
-%
-%In such cases, one requires a more sophisticated distance metric (such as the Earth Mover's Distance \citep{rubner2000earth}) and/or image description (such as scattering transform coefficients \citep{mallat2012group}) which is insensitive to such perturbations.
-%
-%Unlike traditional statistical methods, our algorithms do not provide confidence intervals on predicted orderings, but these could in principle be extracted from bootstrapping \citep{efron1979bootstrap}.
-%
-%Furthermore, although our methods provide a temporal ordering for images, they provide no information about the actual time, neither do they provide a direction for the proposed ordering.
-%
-%Determining the trajectory rate and direction are post-processing tasks which require {\em a priori} knowledge of the developmental processes or some assumptions about the temporal sampling of the data set (e.g., the data set uniformly samples some given time window).
-
-%
-Vector diffusion maps allow us to automatically register images, an essential task for many applications.
-%
-Simultaneously, the algorithm provides us with parameters to describe each image.
-%
-In the examples presented here, we have focused on ordering the images in time using the first vector diffusion maps coordinate.
-%
-In general, we can recover several coordinates which concisely and comprehensively describe the data set.
-%
-This parametrization can then be used for typical data analysis tasks, such as outlier detection and model fitting.
-%
-Data which comes from several different genetic backgrounds can be clustered according to their spatial expression patterns, and clustering can also be used to separate images into different developmental stages.
-%
-Furthermore, images taken from different viewing directions can be analyzed, as the vector diffusion maps parametrization will organize the images according to the viewing angle \citep{singer2011viewing}.
-%
-Another direction for future work is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \citep{krzic2012multiview, ichikawa2014live, rubel2010coupling, dsilva2013nonlinear}.
-%
-Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.
-
-%% == end of paper:
-
-%% Optional Materials and Methods Section
-%% The Materials and Methods section header will be added automatically.
-
-%% Enter any subheads and the Materials and Methods text below.
-
-\section*{Materials and Methods}
-
-\subsection*{{\em Drosophila} embryo experiments}
-%
-Oregon-R was used as wild type {\em Drosophila} strains.
-%
-Embryos were collected and fixed at 22$^\circ$C.
-%
-Monoclonal rabbit anti-dpERK (1:100, Cell signaling) and rat anti-Twist (1:500, a gift from Eric Wieschaus) were used to stain proteins of interest.
-%
-DAPI (1:10,000, Vector Laboratories) was used to visualize nuclei, and Alexa Fluors (1:500, Invitrogen) were used as secondary antibodies.
-%
-Histone-RFP strain was used to obtain time-lapse movie of gastrulating embryos at 22$^\circ$C.
-%
-Live embryos were loaded to the microfluidic device with PBST to keep them oxidized, and fixed embryos were loaded with 90\% glycerol.
-
-\subsection*{{\em Drosophila} embryo microscopy}
-%
-Nikon A1-RS scanning confocal microscope, and the Nikon 60x Plan-Apo oil objective was used to image {\em Drosophila} embryos.
-%
-Embryos were collected, stained, and imaged together under the same microscope setting.
-%
-End-on imaging was performed by using the microfluidics device described previously \citep{chung2010microfluidic}.
-%
-Images were collected at the focal plane $\sim$90~$\mathrm{\mu m}$ from the posterior pole of an embryo (see \fig~\ref{fig:DV_view}).
-
-\subsection*{Image preprocessing}
-
-Images were subsampled, normalized, blurred, and centered prior to diffusion maps analysis to remove any variations due to the experimental and imaging framework.
-%
-Details about the preprocessing operations applied to each imging data set are given in \SI.
-
-\subsection*{Software and imaging data}
-%
-All algorithms and analysis were implemented in MATLAB\textsuperscript{\textregistered} (R2013b, The MathWorks, Natick, Massachusetts).
-%
-Software, including documentation and tutorials, along with the full imaging data sets used in this paper are available at \texttt{genomics.princeton.edu/stas/publications.html} under ``Codes and Data''.
-
-%\end{materials}
-
-\subsubsection*{Acknowledgements}
-
-The authors thank Fisun Hamaratoglu and Markus Affolter for providing the wing disc data.
-%
-The authors thank Angela DePace, Granton Jindal, Adam Finkelstein,  Thomas Funkhouser, and John Storey for helpful discussions.
-
-\subsubsection*{Author contributions}
-Scientific approaches were developed by C.J.D, B.L., H.L., A.S., S.Y.S, and I.G.K.
-%
-Experiments were performed by B.L.
-%
-Data analysis was performed by C.J.D. and B.L.
-%
-The manuscript was prepared and edited by C.J.D, B.L., H.L., A.S., S.Y.S, and I.G.K.
-
-\subsubsection*{Funding}
-C.J.D.
-was supported by the Department of Energy Computational Science Graduate Fellowship (CSGF), grant number DE-FG02-97ER25308, and the National Science Foundation Graduate Research Fellowship, Grant No.
-DGE 1148900.
-%
-B.L. and S.Y.S. were supported by the National Institutes of Health Grant R01GM086537.
-%
-H.L. was supported by the National Science Foundation Grant Emerging Frontiers in Research and Innovation (EFRI) 1136913.
-%
-A.S. was supported by the Air Force Office of Scientific Research Grant
-FA9550-12-1-0317.
-%
-I.G.K. was supported by the National Science Foundation (CS\&E program).
-
-\bibliographystyle{development}
-\bibliography{references}
 

### 2015-02-11 image_analysis_paper_development2/paper_figures/fig5.tex
added compiled figures

--- 2015-02-11 image_analysis_paper_development2/manuscript.tex
@@
-\documentclass[twocolumn, 10pt]{article}
+\documentclass[10pt]{article}
 
-\usepackage[margin=0.75in]{geometry}
-\usepackage{amssymb,amsfonts,amsmath}
+\usepackage[paperwidth=18cm, paperheight=4.5cm, margin=0cm]{geometry}
 \usepackage{graphicx}
-
-\usepackage{color}
-
-\usepackage{epstopdf}
-
-\usepackage{natbib}
-
-\renewcommand{\figurename}{Fig.}
-%\newcommand{\SI}[0]{\textit{SI Materials and Methods}}
-\newcommand{\SI}[0]{supplementary material}
-\newcommand{\fig}[0]{Fig.}
-
 \hyphenpenalty=10000
-
-\makeatletter
-\newcommand{\customlabel}[2]{%
-\protected@write \@auxout {}{\string \newlabel {#1}{{#2}{}}}}
-\makeatother
-
-\graphicspath{{paper_figures/}}
-
-\customlabel{fig:DV_view}{S1}
-\customlabel{fig:1d_example}{S2}
-\customlabel{fig:eigenvalues_drosophila_live}{S3}
-\customlabel{fig:eigenvalues_zebrafish}{S4}
-\customlabel{fig:eigenvalues_drosophila_fixed}{S5}
-\customlabel{fig:eigenvalues_wing_disc}{S6}
-\customlabel{fig:PCA_compare}{S7}
-
-\title{Temporal ordering and registration of images in studies of developmental dynamics}
-
-\author{Carmeline~J.~Dsilva \thanks{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
-Bomyi~Lim \footnotemark[1],
-Hang~Lu \thanks{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
-Amit~Singer \thanks{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \thanks{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA}, Ioannis~G.~Kevrekidis \footnotemark[1] \footnotemark[4] ,\\ and
-Stanislav~Y.~Shvartsman \footnotemark[1] \thanks{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA}}
-
-\date{}
 
@@
 
-\maketitle
-
-\begin{abstract}
-
-Dynamics of developmental progress are commonly reconstructed from imaging snapshots of chemical or mechanical processes in fixed embryos.
-%
-As a first step in these reconstructions, snapshots must be spatially registered and ordered in time.
-%
-Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific system.
-%
-However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
-%
-To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
-%
-The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require \textit{a priori} knowledge of image features or a parametric model of the developmental dynamics.
-%
-We illustrate this approach by registering and ordering data from imaging studies of pattern formation and morphogenesis in three different model systems.
-%
-We also provide software to aid in the application of our methodology to other experimental data sets.
-\vspace{1em}
-
-\noindent
-KEY WORDS: temporal ordering, image registration, vector diffusion maps
-
-\end{abstract}
-
-\section*{Introduction}
-
-\begin{figure}[t]
-\includegraphics{fig1}
-\caption{Caricature illustrating the tasks of image registration and temporal ordering.
-(A) Images of ``samples'', each in a different orientation and a different stage of development.
-(B) Registered and ordered samples.
-For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
-%\label{fig:fish}
-\customlabel{fig:fish}{1}
-\customlabel{subfig:fish_unordered}{\ref{fig:fish}A}
-\customlabel{subfig:fish_ordered}{\ref{fig:fish}B}
-\end{figure}
-
-In one of the common approaches to studies of developmental dynamics, a group of embryos is fixed and stained to visualize a particular biochemical or morphological process within a developing tissue.
-%
-The developmental dynamics must then be reconstructed from multiple embryos, each of which contributes only a snapshot of the relevant process along its developmental trajectory \citep{jaeger2004dynamic, peter2011gene, fowlkes2008quantitative}.
-%
-Importantly, the ``age'' of any given embryo arrested in its development is often only approximately known; typically what is known is
-a certain time window to which a collection of embryos belongs \citep{ng2012large, richardson2014emage, castro2009automatic}.
-%
-Furthermore, images are often collected in different spatial orientations.
-%
-In order to recover the developmental dynamics from such data sets, snapshots of different embryos must first be spatially aligned or {\em registered}, and then ordered in time.
-%
-%We show how recently developed dimensionality reduction algorithms can automate {\it and combine} both of these tasks.
-
-Temporal ordering and registration of images can be done manually
-when the number of images is small and the differences between them are visually apparent.
-%
-\fig~\ref{fig:fish} shows a caricature of fish development which illustrates the processes of growth and patterning.
-%
-In this case, temporal ordering can be accomplished by arranging the fish by size, which is monotonic with the developmental progress.
-%
-Image registration is based on obvious morphological landmarks, such as the positions of the head and the fins.
-%
-In contrast to this example, real data poses nontrivial challenges for both registration and temporal ordering.
-%
-In general, the landmarks needed for registration, as well as the attributes which can be used to order the data, are not known {\it a priori}.
-%
-Additional challenges arise from embryo-to-embryo variability, sample size, and measurement noise.
-
-We present a robust algorithmic approach to simultaneous registration and temporal ordering.
-%
-In contrast to a number of previous methodologies \citep{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face, dubuis2013accurate}, our methodology does not rely on the {\em a priori} knowledge of landmarks for registration or markers of developmental progression.
-%
-The approach is based on vector diffusion maps \citep{singer2012vector}, a manifold learning algorithm which simultaneously addresses the problems of registration and temporal ordering.
-%
-This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \citep{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
-applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \citep{zhao2014rotationally, singer2011viewing} to face recognition \citep{lafon2006data} and classification of CT scans \citep{fernandez2014diffusion}.
-
-Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing tissues in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
-%
-To illustrate our approach, we analyze four data sets common in developmental biology.
-%
-Our first two data sets come from live imaging studies of {\em Drosophila} and zebrafish embryogenesis.
-%
-In both of these examples, the correct rotational orientation and temporal order are independently known, and these data sets will be used to validate our approach.
-%from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
-%
-Our third data set consists of images from fixed {\em Drosophila} embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.
-%
-Our final data set consists of z-stacks of {\em Drosophila} wing discs, which we will use to illustrate how our methods can be used to analyze specific types of three-dimensional imaging data.
-%
-We will also show how to compute an average trajectory from a set of registered and ordered fixed images to remove noise due to intersample variability and obtain a smooth description of the underlying developmental dynamics.
-
-\section*{Results}
-
-\subsection*{Vector diffusion maps for registration and temporal ordering}
-
-Vector diffusion maps \citep{singer2012vector} is a manifold learning
-technique developed for data sets which contain two sources of variability:
-geometric symmetries (such as rotations of the images) which one would like to factor out,
-and ``additional" directions of variability (such as temporal dynamics) which one would like to uncover.
-%
-Vector diffusion maps combine two algorithms, {\em angular synchronization} \citep{singer2011angular} for image registration and {\em diffusion maps} \citep{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
-%
-We will use the algorithm to register images of developing tissues with respect to planar rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
-%
-Although in general, images may contain variations due to rotations, translations, and scaling, we will remove the relevant translations and/or scaling via relatively simple image preprocessing, and focus only on factoring out rotations using the vector diffusion maps algorithm.
-%
-In the case that {\em all} relevant symmetries can be removed with straightforward preprocessing, our algorithms can be used simply to extract the main direction of variability within the imaging data set.
-%
-We assume that this main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
-%
-As a consequence, uncovering this direction should reveal the underlying developmental dynamics.
-
-\begin{figure}[t]
-\includegraphics{fig2}
-\caption{Schematic illustrating angular synchronization and diffusion maps.
-(A) Set of vectors, each in a different orientation.
-The pairwise alignment angles are indicated.
-(B) The vectors from A, each rotated about their midpoint so that the set is globally aligned.
-Note that the chosen rotation angles are consistent with the pairwise alignments in A: the difference between a pair of angles in B is the same as the pairwise angle in A.
-(C) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions.
-Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through a Gaussian kernel (see \SI), so that pairs of data point which are close are connected by darker (``stronger'') edges.
-(D) The data in C, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure.
-The color intensity is monotonic with the perceived curve arclength, thus parametrizing the curve.}
-\label{fig:schematics}
-\end{figure}
-
-Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
-%
-A schematic illustration of angular synchronization is shown in \fig~\ref{fig:schematics}A, where each image is represented as a vector, and the goal is to align the entire set of vectors given pairwise alignment measurements.
-%
-We first compute the angles needed to align pairs of vectors (or images), which in general requires no notion of a template function \citep{ahuja2007template, sonday2013noisy}.
-%
-In this work, we aligned pairs of images with respect to rotations by exhaustively searching over a discretized space of rotation angles to minimize the Euclidean distance between the pixels.
-%
-However, pairwise alignments can also be computed by aligning appropriate image landmarks or features \citep{ian1998statistical}.
-%
-When the data are noisy, these pairwise measurements may by inaccurate, and so we utilize {\em all} pairwise measurements to align the set of images robustly.
-%
-Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see \SI); this is illustrated in \fig~\ref{fig:schematics}B.
-%
-In this schematic, registration via angular synchronization is trivial, as the pairwise measurements contain no noise.
-%
-However, the algorithm can register data sets even when many of the pairwise measurements are inaccurate \citep{singer2011angular}.
-
-After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional curve that parameterizes most of the remaining variability in the data.
-%
-Such a curve can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that reveals a parametrization of data that lies on a low-dimensional manifold in high-dimensional space.
-%
-The idea is illustrated in \fig~\ref{fig:schematics}C, where the data are two-dimensional points which lie on a one-dimensional (nonlinear) curve.
-%
-We use {\it local} information about the data to find a parametrization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parametrization.
-%
-This idea of locality is denoted by the color of the edges in \fig~\ref{fig:schematics}C:
-data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
-%
-The color in \fig~\ref{fig:schematics}D depicts the one-dimensional parametrization or ordering of the data that we can detect visually.
-%
-A detailed example of using vector diffusion maps to register and order synthetic data is given in \fig~\ref{fig:1d_example}, and a step-by-step tutorial of the diffusion maps implementation is included in the \SI.
-%
-In our working examples, each data point will be of much higher dimension (e.g., a pixelated image or three-dimensional voxel data), and so we cannot extract this low-dimensional structure visually.
-%
-Instead, we will use diffusion maps which automatically uncovers a parametrization of our high-dimensional data from the eigenvectors of the appropriate matrix (see \SI).
-%
-Furthermore, the corresponding eigenvalues will allow us to confirm our assumption that our data approximately lie on a one-dimensional manifold (see \fig~\ref{fig:eigenvalues_drosophila_live}--\ref{fig:eigenvalues_wing_disc}).
-%
-%Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.
-
-\subsection*{Method validation using live imaging}
-
-\subsubsection*{{\em Drosophila} gastrulation}
-
-\begin{figure*}[t]
-\raisebox{0.75cm}{\textsf{A}}
-\includegraphics[width=17cm]{drosophila_live_imaging_scrambled}
-
-\raisebox{0.75cm}{\textsf{B}}
-\includegraphics[width=17cm]{drosophila_live_imaging_ordered}
-
-\raisebox{4.25cm}{\textsf{C}}
-\includegraphics[width=5cm]{drosophila_live_imaging_angle_corr}
-%
-%\hfill
-%
-\raisebox{4.25cm}{\textsf{D}}
-\includegraphics[width=5cm]{drosophila_live_imaging_rank_corr}
-%
-%\hfill
-%
-\raisebox{4.25cm}{\textsf{E}}
-%
-\raisebox{2.5cm}{
-{ \textsf{
-\begin{tabular}{ |c | c| c| }
-\hline
-\parbox[c]{1.5cm}{ \centering Movie Index} & \parbox[c]{1.5cm}{ \vspace{.5\baselineskip} \centering Average Angle Error \vspace{.5\baselineskip}} & \parbox[c]{1.5cm}{ \centering Rank Correlation Coefficient} \\
-\hline
-1 & 8.37$^{\circ}$ & 0.9989 \\
-2 & 4.98$^{\circ}$ & 0.9994 \\
-3 & 6.47$^{\circ}$ & 0.9795 \\
-4 & 7.03$^{\circ}$ & 0.9953 \\
-5 & 3.04$^{\circ}$ & 0.9959 \\
- \hline
-\end{tabular}}}}
-\caption{Method validation using live imaging of \textit{Drosophila} embryos.
-(A) Selected images from a live imaging study of a \textit{Drosophila} embryo during gastrulation.
-Scale bar indicates 50$\mu$m.
-Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
-(B) Images from A registered and ordered by vector diffusion maps.
-The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom.
-(C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle.
-The average absolute error in the recovered angles is $8.37 ^\circ$.
-(D) The correlation between the recovered rank (using vector diffusion maps) and the true rank.
-The rank correlation coefficient is $0.9989$.
-(E) The average error in the recovered angle and the rank correlation coefficient for $5$ independent live imaging studies.
-}
-\label{fig:drosophila_live_imaging}
-\end{figure*}
-
-To validate the proposed approach, we first applied our algorithm to a data set where the true temporal order and rotational orientation of the images were known {\em a priori}.
-%
-This data set was obtained through live imaging near the posterior pole of a vertically oriented {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
-%
-During this time window, the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, internalizing the future muscle cells and forming a characteristic ``omega'' shape.
-%
-Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \citep{leptin2005gastrulation}.
-%
-At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing a similar ``omega'' to appear on the dorsal side.
-
-\fig~\ref{fig:drosophila_live_imaging}A shows selected images from this live imaging data set, which contains $40$ consecutive frames taken at $30$~second time intervals at a fixed position within a single embryo.
-%
-Each image shows an optical cross-section near the posterior pole of a vertically oriented developing embryo, with the nuclei labeled by Histone-RFP.
-%
-Each frame was arbitrarily rotated, and the order of the frames was scrambled.
-%
-The task is now to register these images and order them in time to reconstruct the developmental trajectory.
-
-We used vector diffusion maps to register and order the images.
-%
-\fig~\ref{fig:drosophila_live_imaging}B shows the images from \fig~\ref{fig:drosophila_live_imaging}A, now registered and ordered; the real time for each frame is also indicated.
-%
-With a small number of exceptions, the recovered ordering is consistent with the real time dynamics.
-%
-\fig~\ref{fig:drosophila_live_imaging}C and \fig~\ref{fig:drosophila_live_imaging}D  show the correlations between the recovered and true angles and rank orders, respectively, for the entire data set.
-%
-Both the angles and the ranks are recovered with a high degree of accuracy.
-
-To assess the robustness of the proposed methodology, we repeated this procedure with four additional data sets extracted from independent live imaging studies spanning the same developmental time period.
-%
-The results are shown in \fig~\ref{fig:drosophila_live_imaging}E.
-%
-The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than 95\%, indicating that our methodology can reproducibly order data of this type.
-%
-In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{fig:schematics}C) is well-sampled.
-%
-At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments of the images, as well as the size of the data set \citep{singer2011angular}.
-%
-%Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
-%%
-%To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
-%%
-%Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered.
-%%
-%\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
-%%
-%As expected, the accuracy in the recovered ordering increases with the number of images.
-%%
-%At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
-%%
-%Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignment, and the high temporal resolution of our live imaging study allows us to order the images accurately.
-%
-%In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.
-
-\subsubsection*{Zebrafish epiboly}
-
-\begin{figure*}[t]
-\begin{minipage}[c]{13.75cm}
-\vspace{0cm}
-\raisebox{0.5cm}{\textsf{A}}
-\includegraphics[width=13cm]{zebrafish_scrambled}
-
-\raisebox{0.5cm}{\textsf{B}}
-\includegraphics[width=13cm]{zebrafish_ordered}
-\end{minipage}
-%
-\hfill
-%
-\begin{minipage}[c]{3.75cm}
-
-\vspace{0cm}
-\raisebox{2.5cm}{\textsf{C}}
-%
-\includegraphics[width=3.25cm]{zebrafish_rank_corr}
-\end{minipage}
-\caption{Method validation using live imaging of a zebrafish embryo.
-(A) Selected images from a movie of zebrafish epiboly.
-Scale bar indicates 200$\mu$m.
-Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled.
-(B) Images from A after registration and ordering using vector diffusion maps.
-The real time of each frame is also indicated.
-(C) Correlation between the rank recovered using vector diffusion maps and the true rank.
-The rank correlation coefficient is $0.9954$.
-The larger errors in the recovered ranks towards the beginning of the trajectory are due to the slow cell movement within that time window.
-}
-\label{fig:zebrafish}
-\end{figure*}
-
-As another validation for the proposed methodology, we applied our algorithm to a time-lapse movie  of zebrafish embryogenesis.
-%
-We used a publicly available live imaging data of zebrafish embryogenesis (\texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov}, \citet{karlstrom1996flipbook}).
-%
-Taken with a differential interference contrast (DIC) microscope, the movie records the first $17$~hours of zebrafish development, from a single cell stage to a 16-somite stage.
-%
-We selected $120$ consecutive frames from this movie which capture 5.5~hours of epiboly (3.5--9~hours after fertilization).
-%
-In this experiment, embryos were immobilized for imaging so that the position and orientation remained fixed \citep{kane1996zebrafish}.
-%
-At the start of the time window, cells have divided 10--11 times and are accumulated in a cell mass above the yolk.
-%
-The cell mass is then compressed and the animal-vegetal axis of the embryo (vertical axis in \fig~\ref{fig:zebrafish}) shortens to form a spherical embryo shape by the end of the fourth hour of development.
-%
-Then, the yolk syncytial layer, which forms the boundary between the yolk and the cell mass, moves upward, forming a dome-shaped structure.
-%
-During this stage, the cells rearrange to form a uniform layer about four cells thick.
-%
-With time, this cell layer then spreads over across the yolk and expands toward the vegetal pole.
-%
-At the end of epiboly, the blastoderm completely engulfs the yolk.
-
-As in the example of {\em Drosophila} embryo live imaging, the two-dimensional frames were randomly rotated and shuffled (\fig~\ref{fig:zebrafish}A).
-%
-We then used vector diffusion maps to register and order the frames.
-%
-The results are shown in \fig~\ref{fig:zebrafish}B.
-%
-The recovered rotations and order are consistent with the expected developmental dynamics, as shown in the correlations between the recovered and true ranks (\fig~\ref{fig:zebrafish}C).
-%
-Quantitatively, the rank correlation coefficient for this data set is 0.9954, and the average error in the recovered angle is $4.14^\circ$.
-%
-Some errors in ordering images of the early embryo result from slow cell movement during the early developmental stage where cells divide and accumulate above the yolk.
-%
-During epiboly, cell movement is more dynamic and the recovered ordering is more consistent with the real dynamics.
-
-In summary, we have shown that our approach to temporal ordering performs very well on imaging data of two different developmental processes ({\em Drosophila} gastrulation and zebrafish epiboly), taken with two different imaging methods (fluorescent microscopy and DIC) where the true temporal order is known {\em a priori}.
-%image registration and ordering can correctly recover image orientation and  showed that our algorithm is not limited to a specific set of data.
-%
-%The algorithm works for different tissue types ({\em Drosophila} and zebrafish), and also for images taken with different imaging methods (fluorescent and phase contrast images).
-%
-Provided there exist significant dynamics within the data set and that the developmental trajectory is well-sampled, the developmental dynamics can be recovered.
-
-\subsection*{Data sets with intersample variability}
-
-\subsubsection*{Fixed images of {\em Drosophila} gastrulation}
-
-We have analyzed how our algorithm performs on two model data sets where all images come from a single embryo.
-%
-In practice, we are interested in cases where each image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
-%
-To demonstrate that our methods are robust to such variations,
-we constructed a synthetic time course data set by selecting a random image from one of five {\em Drosophila} live imaging data sets (those data sets used in \fig~\ref{fig:drosophila_live_imaging}) at each time point.
-%
-The resulting data set is spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability.
-%
-The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.77, indicating that the algorithm can recover the temporal order even under noisy conditions.
-
-We then applied our approach to a data set where the true rotational orientation and temporal order was not known {\it a priori}.
-%
-\fig~\ref{fig:drosophila_fixed_images}A shows selected images from a set of $120$ images of developing {\em Drosophila} embryos which cover a thirty minute time interval spanning late cellularization through gastrulation.
-%
-This data set is more complex than the live imaging data sets in that it contains significantly more images, each of which provides information about tissue morphology and the spatial distribution of two regulatory proteins.
-%
-Each image shows an optical cross-section of the posterior view of a {\em different} embryo at a different rotational orientation and fixed at a different (and unknown) developmental time.
-%
-The nuclei (gray) were labeled with DAPI, a DNA stain.
-%
-Embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor which specifies the cells of the future muscle tissue.
-%
-Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that, in this context, specifies a subset of neuronal cells \citep{Lim2013kinetics}.
-
-\fig~\ref{fig:drosophila_fixed_images}B shows the selected images in \fig~\ref{fig:drosophila_fixed_images}A, now registered and ordered using vector diffusion maps.
-%
-However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set.
-%
-To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI).
-%
-Each snapshot in the average trajectory is the (weighted) average of a group of successive images from the registered and ordered data set.
-%
-Averaging successive images removes some of the interembryo variability, so that sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.
-
-\begin{figure*}[t!]
 \begin{minipage}[b]{12cm}
@@
 \includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}
-%
-\caption{Analysis of images of fixed \textit{Drosophila} embryos.
-{(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red).
-Scale bar indicates approximately 50$\mu$m (images have been rescaled to remove slight interembryo size variations).
-Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation.
-{(B)} Data from {\it A}, registered and ordered using vector diffusion maps.
-The expert rank for each image is indicated.
-{(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI).
-{(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert.
-The rank correlation coefficient is $0.9716$.
-}
-\label{fig:drosophila_fixed_images}
-\end{figure*}
-
-From this average trajectory, we can now easily see the developmental progression, which is consistent with the known dynamics.
-%
-dpERK first appears as two lateral peaks at the ventrolateral side of the embryo; a third dpERK peak then appears at the dorsal side of the embryo.
-%
-During mesoderm invagination, the two ventrolateral dpERK peaks merge together, eventually forming, together with Twi, the ``omega'' shape.
-%
-The dorsal dpERK peak then disappears during germband extension as cells from the ventral side wrap around to the dorsal side.
-%
-At the end of this process, similar ``omegas'' formed by Twi and dpERK appear on the dorsal side of the embryo; these patterns are most readily seen in the last image of \fig~\ref{fig:drosophila_fixed_images}C.
-%
-Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental features (e.g., correlation of age with body size).
-%
-
-To evaluate the quality of our registration and ordering, we can use prior knowledge about the developmental system.
-%
-The Twi signal is known to form a single peak at the ventralmost point of the embryo.
-%
-We found that the standard deviation in the location of this peak in the set of registered images was $\sim$8$^\circ$,
-indicating that the algorithm successfully aligns the ventralmost points of the images.
-%
-Because the developmental time of each embryo cannot be easily estimated, we have few options for evaluating the quality of our temporal ordering.
-%
-We compared the ordering obtained from vector diffusion maps to the ordering provided by a trained embryologist who is knowledgeable about the developmental progression and the important image features.
-%
-The ranks from the ordering provided by the embryologist, which we will refer to as the ``expert rank'', are indicated for the images in \fig~\ref{fig:drosophila_fixed_images}B , and the rank correlation (see \fig~\ref{fig:drosophila_fixed_images}D) shows that our ordering is consistent with the expert ordering.
-
-\subsubsection*{Fixed z-stacks of {\em Drosophila} wing discs}
-
-\begin{figure*}[t]
-\raisebox{5.5cm}{\textsf{A}}
-\includegraphics[width=6cm]{wing_disc_example}
-%
-\hfill
-%
-\raisebox{5.5cm}{\textsf{B}}
-\includegraphics[width=10cm]{wing_disc_ordered}
-
-\vspace{0.25cm}
-\raisebox{2.25cm}{\textsf{C}}
-\includegraphics[width=17.5cm]{wing_disc_average}
-
-\caption{Analysis of three-dimensional {\em Drosophila} wing disc z-stacks.
-(A) Maximum projections of an example three-dimensional {\em Drosophila} wing disc z-stack.
-The anterior-posterior (A-P) and dorsal-ventral (D-V) axes are indicated.
-Discs express the Dad-GFP reporter construct (green), and are stained for Spalt (red), Wingless (gray), and Patched (gray).
-Projections along the $x$-, $y$-, and $z$-axes are shown.
-(B) Example three-dimensional images, ordered using diffusion maps.
-The time cohort, as assessed by an expert, is indicated for each image, and the rank correlation coefficient between the diffusion maps ordering and the expert timing is 0.9427.
-(C) The average developmental trajectory for the registered and ordered images.
-}
-\label{fig:wing_disc}
-\end{figure*}
-
-%We have demonstrated that our algorithm can recover the underlying developmental dynamics in several different types of imaging data sets, including ones containing interembryo variability.
-%
-%Thus far, we have only looked at two-dimensional data sets where biological samples are imaged as a single confocal slice or as a phase contrast image at a fixed position within the sample.
-%
-%However, with advances in imaging techniques, taking multiple z-slices to visualize the three-dimensional structure of a sample will become more common, and it is important to validate that our algorithm can be applied to such data.
-
-In this section we show that the approach can readily be applied to three-dimensional data. 
-%
-We restrict ourselves to the case where an obvious fixed axis exists, so that only rotations of the three-dimensional data around this axis need be taken into account.  
-%
-This does not constitute an inherent limitation for vector diffusion maps. 
-%
-While for simplicity here we will not discuss the general  case, incorporating general 3D symmetries is possible \citep{arie2012global, wang2013exact, cucuringu2012eigenvector}.
-
-%Although the theory of vector diffusion maps readily extends to three-dimensional data with full three-dimensional rotations, calculation of the pairwise alignments rapidly becomes computationally prohibitive for such applications, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle).
-%
-%However, ordering data via diffusion maps is readily possible for pre-registered three-dimensional data.
-%
-%Therefore, we can apply our methodology to three-dimensional data, such as a stack of confocal images taken at different depths within the sample, by first registering the data using an alternative method, and then ordering the registered data using diffusion maps.
-
-To demonstrate this approach, we used an existing three-dimensional data set of fixed {\em Drosophila} wing imaginal discs, which was published in \citet{hamaratoglu2011dpp}.
-%
-Imaginal discs are groups of progenitor cells in fly larva that will transform into specific organs (such as wings, legs, and eyes) in the adult fly during the pupal metamorphosis.
-%
-The wing disc is an imaginal disc that turns into a wing.
-%
-The data set is composed of 46 fixed wing discs whose developmental times range from 72 to 112~hours after fertilization.
-%
-Each disc contains 21 z-slices taken at 1~$\mu$m intervals.
-%
-The discs were dissected from larvae expressing the Dad-GFP reporter construct (green) and stained with antibodies that recognize Spalt (red), Wingless (gray), and Patched (gray), the molecules that play important roles in disc patterning and growth (\fig~\ref{fig:wing_disc}A).
-%
-%The disc grows in time; at long times, the disc becomes too large to fit completely within the frame and only a partial disc is imaged.
-
-In the wing disc, the anterior-posterior and dorsal-ventral axes are significantly larger than the third principal axis (see \fig~\ref{fig:wing_disc}A). 
-%
-Therefore, we need not consider registration in all three dimensions, as the z-axis for imaging (the axis perpendicular to the imaging plane) will always be aligned with the short axis of the wing disc.
-%
-We can instead focus on registering the wing discs with respect to rotations only in the x-y plane.
-%
-To register the data, we first aligned the maximum intensity projections using angular synchronization.
-%
-We then used these rotations to register the full three-dimensional data in the x-y plane.
-%
-Because the maximum intensity projections are two-dimensional images, this step is no more computationally intensive than the previous examples.
-%
-Such an approach is possible when there are distinct major and minor axes within a three-dimensional sample, which reduces the rotational degrees of freedom.
-%
-%Other options for registering samples include using a specific localized signal (such as Twi in \fig~\ref{fig:drosophila_fixed_images}) as a marker for alignment, or using other {\em a priori} knowledge.
-
-We then used diffusion maps to order the registered three-dimensional data.
-%
-\fig~\ref{fig:wing_disc}B shows selected images from the data set ordered by diffusion maps.
-%
-In the original data set, each disc was assigned to one of six time classes (72--73~hr, 76.5--77.5~hr, 79--80~hr, 89--90~hr, 100--101~hr, and 110.5--111.5~hr after fertilization) by an expert; these times are indicated in \fig~\ref{fig:wing_disc}B.
-%
-In the ordered set, the size of wing disc grows, and the intensity of the Dad-GFP signal increases as a function of time.
-%
-The rank correlation coefficient based on the time class is 0.9436.
-%
-The registration errors are primarily due to some wing discs having extra tissue attached to them (such as the image in \fig~\ref{fig:wing_disc}A and the fourth image in \fig~\ref{fig:wing_disc}B).
-%
-Even with such obstructions, we can accurately order the images and extract a stereotypical developmental trajectory, shown in \fig~\ref{fig:wing_disc}C, by averaging (see \SI).
-%
-We can now easily see the growth of the wing disc, even though averaging somewhat blurs some finer scale structures.
-%
-%Thus, this result shows that the algorithm can successfully order a three-dimensional data set.
-
-%The vector diffusion maps algorithm has been previously used for a different type of three-dimensional data: to reconstruct a three-dimensional object (in this case, a molecule) from two-dimensional (cryo-electron microscopy) pictures \citep{singer2011viewing}.
-%
-%In these applications, many two-dimensional pictures of (effectively) the same object are taken from different viewing directions.
-%
-%The vector diffusion maps algorithm then, instead of revealing the temporal order, reveals the viewing direction of each image, which can then be used to reconstruct the three-dimensional object.
-%
-%This is another approach to three-dimensional imaging which we hope to extend in the future to examples which also contain variability from the developmental progression.
-
-\subsection*{Computational requirements}
-
-\begin{figure*}[t]
-\raisebox{5cm}{\textsf{A}}
-\includegraphics[width=5.5cm]{timing_studies_nimages}
-%
-\raisebox{5cm}{\textsf{B}}
-\includegraphics[width=5.5cm]{timing_studies_npixels}
-%
-\raisebox{5cm}{\textsf{C}}
-\includegraphics[width=5.5cm]{timing_studies_nrot}
-
-\raisebox{2.5cm}{\textsf{D}}
-{\renewcommand{\arraystretch}{1.5}
-{ \textsf{ \small
-\begin{tabular}{ | p{2cm} | c | c | c | c | c | c | }
-\hline
-Data Set &
-Data Type &
-\parbox[c]{2cm}{ \centering Number of Channels} &
-\parbox[c]{2cm}{ \centering \vspace{.5\baselineskip} Number of Images \vspace{.5\baselineskip}} &
-\parbox[c]{2cm}{ \centering Number of Pixels} &
-\parbox[c]{2cm}{ \centering Angular Discretization} &
-\parbox[c]{2cm}{ \centering CPU time}  \\
-\hline
-{\em Drosophila} gastrulation (live) & 2D & 1 & 40 & 100 $\times$ 100 & 10$^\circ$ & 3.2~sec \\
-Zebrafish epiboly & 2D & 1 & 120 & 100 $\times$ 100 & 10$^\circ$ & 13~sec\\
-{\em Drosophila} gastrulation (fixed) & 2D & 3 & 120 & 100 $\times$ 100 & 10$^\circ$ & 29~sec  \\
-{\em Drosophila} wing discs & 3D & 3 & 46 & 100 $\times$ 100 $\times$ 21 & 10$^\circ$ & 12~sec  \\
- \hline
-\end{tabular}}}}
-\caption{Computational requirements for the presented methodology.
-(A) CPU time as a function of the number of images in the data set (for $100 \times 100$ pixel images, and $10^\circ$ angular discretization).
-Empirically, the CPU time is $\sim \mathcal{O}(n^{1.33})$ in number of images.
-(B) CPU time as a function of the number of pixels in the images (for $120$ images, and $10^\circ$ angular discretization).
-Empirically, the CPU time is $\sim \mathcal{O}(n^{1.83})$ in the number of pixels.
-(C) CPU time as a function of the number of rotations (for $120$ images of $100 \times 100$ pixels).
-Empirically, the CPU time is $\sim \mathcal{O}(n^{-0.77})$ in the angular discretization.
-(D) The algorithm settings and computational requirements for the data sets analyzed.
-All times are reported for an Intel Core i7 2.93 GHz processor.}
-\label{fig:cpu_time}
-\end{figure*}
-
-%Manually ordering the images is nontrivial for researchers who are unfamiliar with the relevant developmental progression, and can be tedious and time-consuming for those who are.
-%
-%In contrast, our method requires relatively little computational time.
-%
-The computational costs for our methodology are outlined in \fig~\ref{fig:cpu_time}.
-%
-The computational time is a function of the number of images in the data set, the number of pixels in each point, and the angular resolution to compute the pairwise rotations (see \SI).
-%
-Furthermore, the computation of the pairwise rotational alignments, which accounts for the majority of the computational time, is trivially parallelizable, and only a subsample of the pairwise alignments need to be computed for larger data sets for accurate recovery of the underlying rotations \citep{singer2011angular}.
-%
-Because the computational cost increases with the image resolution, we chose to subsample all of our data sets to $100 \times 100$ pixels.
-%
-This resolution allowed us to rapidly analyze our data sets while still retaining all of the relevant developmental features.
-%
-However, as can be seen from the computational costs in \fig~\ref{fig:cpu_time}, it is feasible to use our algorithms to analyze higher-resolution images.
-
-The requisite user intervention and parameter tuning required for our method is relatively minor.
-%
-As a first step, images must be preprocessed so that the Euclidean distance between the pixels is informative.
-%
-Our software provides several preprocessing options (such as blurring, rescaling, and mean-centering), as well as some guidance for what options to select depending on the system of interest.
-%
-Two algorithmic parameters, the angular discretization to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{fig:schematics} and \SI), must also be defined.
-%
-We also provide some guidance on selecting these parameters, and found that the results are robust to both of these parameters.
-%
-Overall, the tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, and so this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.
-
-\section*{Discussion}
-
-%Already at this point, the rate of collection of imaging data has surpassed the rate of manual image analysis.
-%
-%This necessitates the development of automated methodologies to organize such large data sets.
-%
-%To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other.
-%
-Temporal ordering of large-scale data was done in the context of molecular profiling studies, in which data points are vectors describing the expression levels of different mRNA \citep{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
-%
-At the same time, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \citep{yuan2014automated, surkova2008characterization, fowlkes2008quantitative}, or using some {\em a priori} knowledge of the relevant developmental processes \citep{dubuis2013accurate}.
-%
-%The task of image registration has been widely studied \citep{zitova2003image}, for applications such as face recognition \citep{rowley1998rotation}, medical image registration \citep{hajnal2010medical}, and texture classification \citep{greenspan1994rotation}.
-%
-In contrast to most of the existing registration approaches which rely on the knowledge of appropriate landmarks in the images \citep{ian1998statistical} (such as the eyes in face recognition applications \citep{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them relevant for a wide variety of applications.
-%
-
-Angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy images \citep{singer2012vector, zhao2014rotationally, singer2011viewing}.
-%
-Because of high levels of instrument noise in these data, thousands of images were needed for successful shape reconstruction.
-%
-Based on the presented results, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories from snapshots of fixed tissues.
-%
-In general, the size of the data set required for accurate registration and ordering is a function of the instrument noise, interembryo variability, and the complexity of the developmental dynamics.
-
-%
-The benefits of our approach to image data mining are twofold.
-%
-First, the algorithm can accomplish the tasks of registration and ordering in a single step.
-%
-Furthermore, because our methodology is nonlinear, it can successfully order data sets which contain complex dynamics (see \fig~\ref{fig:PCA_compare} for a comparison of ordering using linear principal component analysis versus vector diffusion maps for the data sets presented in this paper).
-%
-We expect nonlinear techniques to be necessary for larger data sets which span a wider dynamic range.
-%
-The main utility of our proposed methodology lies in the analysis of data sets containing hundreds of images from systems which have not been well-studied.
-%
-For such data sets, manual ordering of the images can be nontrivial, and our algorithms can clearly accelerate uncovering the underlying developmental dynamics.
-
-We acknowledge that our methods, though general, do have limitations.
-%
-The first is that we require enough data to sufficiently sample the developmental trajectory.
-%
-Therefore, for very small and/or very noisy data sets, our algorithms may fail.
-%
-Second, the pertinent image features need to be large compared to the noise and the image resolution.
-%
-In all of our examples, the relevant expression patterns and morphological structures span several pixels and are large compared to both the instrument noise and embryo-to-embryo variability, making the Euclidean distance between pixels a good measure of images similarity.
-%
-%The methodology we have presented here only addresses images taken in the same location and/or from the same viewing direction within the organism.
-%
-%However, for organisms with distinct major and minor axes (such as {\em Drosophila} embryos and wing discs), sample preparation will often fix one of the principal directions.
-%
-%Furthermore, as we have acknowledged, our methods as implemented cannot register three-dimensional data with respect to three-dimensional rotations due to computational limitations.
-%
-%In the future, the algorithms presented here could be used for three-dimensional voxel data given an alternative method for computing pairwise alignments more efficiently (such as using image features or landmarks).
-%
-
-%Our algorithms will fail for other imaging data sets which contain fine-scale structures,  as small perturbations in these structures will result in large changes in the Euclidean distances.
-%
-%In such cases, one requires a more sophisticated distance metric (such as the Earth Mover's Distance \citep{rubner2000earth}) and/or image description (such as scattering transform coefficients \citep{mallat2012group}) which is insensitive to such perturbations.
-%
-%Unlike traditional statistical methods, our algorithms do not provide confidence intervals on predicted orderings, but these could in principle be extracted from bootstrapping \citep{efron1979bootstrap}.
-%
-%Furthermore, although our methods provide a temporal ordering for images, they provide no information about the actual time, neither do they provide a direction for the proposed ordering.
-%
-%Determining the trajectory rate and direction are post-processing tasks which require {\em a priori} knowledge of the developmental processes or some assumptions about the temporal sampling of the data set (e.g., the data set uniformly samples some given time window).
-
-%
-Vector diffusion maps allow us to automatically register images, an essential task for many applications.
-%
-Simultaneously, the algorithm provides us with parameters to describe each image.
-%
-In the examples presented here, we have focused on ordering the images in time using the first vector diffusion maps coordinate.
-%
-In general, we can recover several coordinates which concisely and comprehensively describe the data set.
-%
-This parametrization can then be used for typical data analysis tasks, such as outlier detection and model fitting.
-%
-Data which comes from several different genetic backgrounds can be clustered according to their spatial expression patterns, and clustering can also be used to separate images into different developmental stages.
-%
-Furthermore, images taken from different viewing directions can be analyzed, as the vector diffusion maps parametrization will organize the images according to the viewing angle \citep{singer2011viewing}.
-%
-Another direction for future work is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \citep{krzic2012multiview, ichikawa2014live, rubel2010coupling, dsilva2013nonlinear}.
-%
-Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.
-
-%% == end of paper:
-
-%% Optional Materials and Methods Section
-%% The Materials and Methods section header will be added automatically.
-
-%% Enter any subheads and the Materials and Methods text below.
-
-\section*{Materials and Methods}
-
-\subsection*{{\em Drosophila} embryo experiments}
-%
-Oregon-R was used as wild type {\em Drosophila} strains.
-%
-Embryos were collected and fixed at 22$^\circ$C.
-%
-Monoclonal rabbit anti-dpERK (1:100, Cell signaling) and rat anti-Twist (1:500, a gift from Eric Wieschaus) were used to stain proteins of interest.
-%
-DAPI (1:10,000, Vector Laboratories) was used to visualize nuclei, and Alexa Fluors (1:500, Invitrogen) were used as secondary antibodies.
-%
-Histone-RFP strain was used to obtain time-lapse movie of gastrulating embryos at 22$^\circ$C.
-%
-Live embryos were loaded to the microfluidic device with PBST to keep them oxidized, and fixed embryos were loaded with 90\% glycerol.
-
-\subsection*{{\em Drosophila} embryo microscopy}
-%
-Nikon A1-RS scanning confocal microscope, and the Nikon 60x Plan-Apo oil objective was used to image {\em Drosophila} embryos.
-%
-Embryos were collected, stained, and imaged together under the same microscope setting.
-%
-End-on imaging was performed by using the microfluidics device described previously \citep{chung2010microfluidic}.
-%
-Images were collected at the focal plane $\sim$90~$\mathrm{\mu m}$ from the posterior pole of an embryo (see \fig~\ref{fig:DV_view}).
-
-\subsection*{Image preprocessing}
-
-Images were subsampled, normalized, blurred, and centered prior to diffusion maps analysis to remove any variations due to the experimental and imaging framework.
-%
-Details about the preprocessing operations applied to each imging data set are given in \SI.
-
-\subsection*{Software and imaging data}
-%
-All algorithms and analysis were implemented in MATLAB\textsuperscript{\textregistered} (R2013b, The MathWorks, Natick, Massachusetts).
-%
-Software, including documentation and tutorials, along with the full imaging data sets used in this paper are available at \texttt{genomics.princeton.edu/stas/publications.html} under ``Codes and Data''.
-
-%\end{materials}
-
-\subsubsection*{Acknowledgements}
-
-The authors thank Fisun Hamaratoglu and Markus Affolter for providing the wing disc data.
-%
-The authors thank Angela DePace, Granton Jindal, Adam Finkelstein,  Thomas Funkhouser, and John Storey for helpful discussions.
-
-\subsubsection*{Author contributions}
-Scientific approaches were developed by C.J.D, B.L., H.L., A.S., S.Y.S, and I.G.K.
-%
-Experiments were performed by B.L.
-%
-Data analysis was performed by C.J.D. and B.L.
-%
-The manuscript was prepared and edited by C.J.D, B.L., H.L., A.S., S.Y.S, and I.G.K.
-
-\subsubsection*{Funding}
-C.J.D.
-was supported by the Department of Energy Computational Science Graduate Fellowship (CSGF), grant number DE-FG02-97ER25308, and the National Science Foundation Graduate Research Fellowship, Grant No.
-DGE 1148900.
-%
-B.L. and S.Y.S. were supported by the National Institutes of Health Grant R01GM086537.
-%
-H.L. was supported by the National Science Foundation Grant Emerging Frontiers in Research and Innovation (EFRI) 1136913.
-%
-A.S. was supported by the Air Force Office of Scientific Research Grant
-FA9550-12-1-0317.
-%
-I.G.K. was supported by the National Science Foundation (CS\&E program).
-
-\bibliographystyle{development}
-\bibliography{references}
 

### 2015-02-11 image_analysis_paper_development2/SI_text.tex
updated figures to make compiled panels

@@
 \begin{figure}
-\includegraphics[width=8cm]{drosophila_live_eval_spectrum}
-
-\includegraphics[width=2.5cm]{drosophila_live_evec_corr1}
-\includegraphics[width=2.5cm]{drosophila_live_evec_corr2}
-\includegraphics[width=2.5cm]{drosophila_live_evec_corr3}
+\includegraphics{figS3}
 \caption{Eigenvalue spectra for the {\em Drosophila} live imaging data set presented in \fig~3.
@@
 \begin{figure}
-\includegraphics[width=8cm]{zebrafish_eval_spectrum}
-
-\includegraphics[width=2.5cm]{zebrafish_evec_corr1}
-\includegraphics[width=2.5cm]{zebrafish_evec_corr2}
-\includegraphics[width=2.5cm]{zebrafish_evec_corr3}
+\includegraphics{figS4}
 \caption{Eigenvalue spectra for the zebrafish data set presented in \fig~4.
@@
 \begin{figure}
-\includegraphics[width=8cm]{drosophila_fixed_eval_spectrum}
+\includegraphics{figS5}
 \caption{Eigenvalue spectra for the fixed {\em Drosophila} images presented in \fig~5.
@@
 \begin{figure}
-\includegraphics[width=8cm]{wing_disc_eval_spectrum}
+\includegraphics{figS6}
 \caption{Eigenvalue spectra for the wing disc data set presented in \fig~6.
@@
 
-\begin{figure}
+\begin{table*}
 \textsf{
@@
 }
-\end{figure}
+\end{table*}
 

### 2015-02-11 image_analysis_paper_development2/manuscript.tex
updated figures to make compiled panels

@@
 \begin{figure*}[t]
-\raisebox{0.75cm}{\textsf{A}}
-\includegraphics[width=17cm]{drosophila_live_imaging_scrambled}
-
-\raisebox{0.75cm}{\textsf{B}}
-\includegraphics[width=17cm]{drosophila_live_imaging_ordered}
-
-\raisebox{4.25cm}{\textsf{C}}
-\includegraphics[width=5cm]{drosophila_live_imaging_angle_corr}
-%
-%\hfill
-%
-\raisebox{4.25cm}{\textsf{D}}
-\includegraphics[width=5cm]{drosophila_live_imaging_rank_corr}
-%
-%\hfill
-%
-\raisebox{4.25cm}{\textsf{E}}
-%
-\raisebox{2.5cm}{
-{ \textsf{
-\begin{tabular}{ |c | c| c| }
-\hline
-\parbox[c]{1.5cm}{ \centering Movie Index} & \parbox[c]{1.5cm}{ \vspace{.5\baselineskip} \centering Average Angle Error \vspace{.5\baselineskip}} & \parbox[c]{1.5cm}{ \centering Rank Correlation Coefficient} \\
-\hline
-1 & 8.37$^{\circ}$ & 0.9989 \\
-2 & 4.98$^{\circ}$ & 0.9994 \\
-3 & 6.47$^{\circ}$ & 0.9795 \\
-4 & 7.03$^{\circ}$ & 0.9953 \\
-5 & 3.04$^{\circ}$ & 0.9959 \\
- \hline
-\end{tabular}}}}
+\includegraphics{fig3}
 \caption{Method validation using live imaging of \textit{Drosophila} embryos.
@@
 \begin{figure*}[t]
-\begin{minipage}[c]{13.75cm}
-\vspace{0cm}
-\raisebox{0.5cm}{\textsf{A}}
-\includegraphics[width=13cm]{zebrafish_scrambled}
-
-\raisebox{0.5cm}{\textsf{B}}
-\includegraphics[width=13cm]{zebrafish_ordered}
-\end{minipage}
-%
-\hfill
-%
-\begin{minipage}[c]{3.75cm}
-
-\vspace{0cm}
-\raisebox{2.5cm}{\textsf{C}}
-%
-\includegraphics[width=3.25cm]{zebrafish_rank_corr}
-\end{minipage}
+\includegraphics{fig4}
 \caption{Method validation using live imaging of a zebrafish embryo.
@@
 \begin{figure*}[t!]
-\begin{minipage}[b]{12cm}
-\raisebox{0.5cm}{\textsf{A}}
-\includegraphics[width=12.25cm]{drosophila_fixed_images_scrambled}
-
-\raisebox{0.5cm}{\textsf{B}}
-\includegraphics[width=12.25cm]{drosophila_fixed_images_ordered}
-
-\raisebox{0.5cm}{\textsf{C}}
-\includegraphics[width=12.25cm]{drosophila_fixed_images_average}
-\end{minipage}
-%
-\hfill
-%
-\raisebox{3.25cm}{\textsf{D}}
-\includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}
+\includegraphics{fig5}
 %
@@
 \begin{figure*}[t]
-\raisebox{5.5cm}{\textsf{A}}
-\includegraphics[width=6cm]{wing_disc_example}
-%
-\hfill
-%
-\raisebox{5.5cm}{\textsf{B}}
-\includegraphics[width=10cm]{wing_disc_ordered}
-
-\vspace{0.25cm}
-\raisebox{2.25cm}{\textsf{C}}
-\includegraphics[width=17.5cm]{wing_disc_average}
-
+\includegraphics{fig6}
 \caption{Analysis of three-dimensional {\em Drosophila} wing disc z-stacks.
@@
 \begin{figure*}[t]
-\raisebox{5cm}{\textsf{A}}
-\includegraphics[width=5.5cm]{timing_studies_nimages}
-%
-\raisebox{5cm}{\textsf{B}}
-\includegraphics[width=5.5cm]{timing_studies_npixels}
-%
-\raisebox{5cm}{\textsf{C}}
-\includegraphics[width=5.5cm]{timing_studies_nrot}
-
-\raisebox{2.5cm}{\textsf{D}}
-{\renewcommand{\arraystretch}{1.5}
-{ \textsf{ \small
-\begin{tabular}{ | p{2cm} | c | c | c | c | c | c | }
-\hline
-Data Set &
-Data Type &
-\parbox[c]{2cm}{ \centering Number of Channels} &
-\parbox[c]{2cm}{ \centering \vspace{.5\baselineskip} Number of Images \vspace{.5\baselineskip}} &
-\parbox[c]{2cm}{ \centering Number of Pixels} &
-\parbox[c]{2cm}{ \centering Angular Discretization} &
-\parbox[c]{2cm}{ \centering CPU time}  \\
-\hline
-{\em Drosophila} gastrulation (live) & 2D & 1 & 40 & 100 $\times$ 100 & 10$^\circ$ & 3.2~sec \\
-Zebrafish epiboly & 2D & 1 & 120 & 100 $\times$ 100 & 10$^\circ$ & 13~sec\\
-{\em Drosophila} gastrulation (fixed) & 2D & 3 & 120 & 100 $\times$ 100 & 10$^\circ$ & 29~sec  \\
-{\em Drosophila} wing discs & 3D & 3 & 46 & 100 $\times$ 100 $\times$ 21 & 10$^\circ$ & 12~sec  \\
- \hline
-\end{tabular}}}}
+\includegraphics{fig7}
 \caption{Computational requirements for the presented methodology.

### 2015-02-13 image_analysis_paper_development2/manuscript.tex
final updates before resubmission
added fisun edits

@@
 \renewcommand{\figurename}{Fig.}
-%\newcommand{\SI}[0]{\textit{SI Materials and Methods}}
 \newcommand{\SI}[0]{supplementary material}
@@
 
-Dynamics of developmental progress is commonly reconstructed from imaging snapshots of chemical or mechanical processes in fixed embryos.
+Dynamics of developmental progress is commonly reconstructed from imaging snapshots of chemical or mechanical processes in fixed tissues.
 %
@@
 For this caricature, the registration and ordering is straightforward because the data set is small, the landmarks are visually apparent, and the developmental changes are easy to recognize.}
-%\label{fig:fish}
-\customlabel{fig:fish}{1}
-\customlabel{subfig:fish_unordered}{\ref{fig:fish}A}
-\customlabel{subfig:fish_ordered}{\ref{fig:fish}B}
+\label{fig:fish}
 \end{figure}
@@
 In order to recover the developmental dynamics from such data sets, snapshots of different embryos must first be spatially aligned or {\em registered}, and then ordered in time.
-%
-%We show how recently developed dimensionality reduction algorithms can automate {\it and combine} both of these tasks.
 
@@
 In both of these examples, the correct rotational orientation and temporal order are independently known, and these data sets will be used to validate our approach.
-%from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
 %
@@
 %
-When the data are noisy, these pairwise measurements may by inaccurate, and so we utilize {\em all} pairwise measurements to align the set of images robustly.
+When the data are noisy, these pairwise measurements may be inaccurate, and so we utilize {\em all} pairwise measurements to align the set of images robustly.
 %
@@
 Furthermore, the corresponding eigenvalues will allow us to test our assumption that our data approximately lie on a one-dimensional manifold (see \fig~\ref{fig:eigenvalues_drosophila_live}--\ref{fig:eigenvalues_wing_disc}).
-%
-%Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.
 
@@
 The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than 95\%, indicating that our methodology can reproducibly order data of this type.
-%
-%In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{fig:schematics}C) is well-sampled.
-%
-%At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments of the images, as well as the size of the data set \citep{singer2011angular}.
-%
-%Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
-%%
-%To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
-%%
-%Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered.
-%%
-%\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
-%%
-%As expected, the accuracy in the recovered ordering increases with the number of images.
-%%
-%At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
-%%
-%Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignment, and the high temporal resolution of our live imaging study allows us to order the images accurately.
-%
-%In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.
 
@@
 In summary, we have shown that our approach to temporal ordering performs very well on imaging data of two different developmental processes ({\em Drosophila} gastrulation and zebrafish epiboly), taken with two different imaging methods (fluorescent microscopy and DIC) where the true temporal order is known {\em a priori}.
-%image registration and ordering can correctly recover image orientation and  showed that our algorithm is not limited to a specific set of data.
-%
-%The algorithm works for different tissue types ({\em Drosophila} and zebrafish), and also for images taken with different imaging methods (fluorescent and phase contrast images).
 %
@@
 %
-%However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set.
-%
-%To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI).
-%
 Each snapshot in the average trajectory is the (weighted) average of a group of successive images from the registered and ordered data set (see \SI).
@@
 \includegraphics{fig5}
-%
 \caption{Analysis of images of fixed \textit{Drosophila} embryos.
@@
 Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks and without {\it a priori} knowledge of developmental features.
-%
 
@@
 
-%We have demonstrated that our algorithm can recover the underlying developmental dynamics in several different types of imaging data sets, including ones containing interembryo variability.
-%
-%Thus far, we have only looked at two-dimensional data sets where biological samples are imaged as a single confocal slice or as a phase contrast image at a fixed position within the sample.
-%
-%However, with advances in imaging techniques, taking multiple z-slices to visualize the three-dimensional structure of a sample will become more common, and it is important to validate that our algorithm can be applied to such data.
-
 In this section we show that the approach can readily be applied to three-dimensional data. 
@@
 
-%Although the theory of vector diffusion maps readily extends to three-dimensional data with full three-dimensional rotations, calculation of the pairwise alignments rapidly becomes computationally prohibitive for such applications, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle).
-%
-%However, ordering data via diffusion maps is readily possible for pre-registered three-dimensional data.
-%
-%Therefore, we can apply our methodology to three-dimensional data, such as a stack of confocal images taken at different depths within the sample, by first registering the data using an alternative method, and then ordering the registered data using diffusion maps.
-
 To demonstrate this approach, we used an existing three-dimensional data set of fixed {\em Drosophila} wing imaginal discs \citep{hamaratoglu2011dpp}.
@@
 %
-The wing disc is an imaginal disc that turns into a wing.
+The wing disc is an imaginal disc that gives rise to the wing, thorax, and hinge.
 %
@@
 The discs were dissected from larvae expressing the Dad-GFP reporter construct (green) and stained with antibodies that recognize Spalt (red), Wingless (gray), and Patched (gray), the factors that play important roles in disc patterning and growth (\fig~\ref{fig:wing_disc}A).
-%
-%The disc grows in time; at long times, the disc becomes too large to fit completely within the frame and only a partial disc is imaged.
 
@@
 Such an approach is possible when there are distinct major and minor axes within a three-dimensional sample, which reduces the rotational degrees of freedom.
-%
-%Other options for registering samples include using a specific localized signal (such as Twi in \fig~\ref{fig:drosophila_fixed_images}) as a marker for alignment, or using other {\em a priori} knowledge.
 
@@
 We can now clearly see the growth of the wing disc, even though averaging somewhat blurs some finer scale structures.
-%
-%Thus, this result shows that the algorithm can successfully order a three-dimensional data set.
-
-%The vector diffusion maps algorithm has been previously used for a different type of three-dimensional data: to reconstruct a three-dimensional object (in this case, a molecule) from two-dimensional (cryo-electron microscopy) pictures \citep{singer2011viewing}.
-%
-%In these applications, many two-dimensional pictures of (effectively) the same object are taken from different viewing directions.
-%
-%The vector diffusion maps algorithm then, instead of revealing the temporal order, reveals the viewing direction of each image, which can then be used to reconstruct the three-dimensional object.
-%
-%This is another approach to three-dimensional imaging which we hope to extend in the future to examples which also contain variability from the developmental progression.
 
@@
 
-%Manually ordering the images is nontrivial for researchers who are unfamiliar with the relevant developmental progression, and can be tedious and time-consuming for those who are.
-%
-%In contrast, our method requires relatively little computational time.
-%
 The computational costs for our methodology are outlined in \fig~\ref{fig:cpu_time}.
@@
 
-%Already at this point, the rate of collection of imaging data has surpassed the rate of manual image analysis.
-%
-%This necessitates the development of automated methodologies to organize such large data sets.
-%
-%To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other.
-%
 Temporal ordering of large-scale data was done in the context of molecular profiling studies, in which data points are vectors describing the expression levels of different mRNA \citep{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
@@
 %
-%The task of image registration has been widely studied \citep{zitova2003image}, for applications such as face recognition \citep{rowley1998rotation}, medical image registration \citep{hajnal2010medical}, and texture classification \citep{greenspan1994rotation}.
-%
 In contrast to most of the existing registration approaches which rely on the knowledge of appropriate landmarks in the images \citep{ian1998statistical} (such as the eyes in face recognition applications \citep{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them relevant for a wide variety of applications.
-%
 
@@
 
-%
 The benefits of our approach to image data mining are twofold.
@@
 In all of our examples, the relevant expression patterns and morphological structures span several pixels and are large compared to both the instrument noise and embryo-to-embryo variability, making the Euclidean distance between pixels a good measure of images similarity.
-%
-%The methodology we have presented here only addresses images taken in the same location and/or from the same viewing direction within the organism.
-%
-%However, for organisms with distinct major and minor axes (such as {\em Drosophila} embryos and wing discs), sample preparation will often fix one of the principal directions.
-%
-%Furthermore, as we have acknowledged, our methods as implemented cannot register three-dimensional data with respect to three-dimensional rotations due to computational limitations.
-%
-%In the future, the algorithms presented here could be used for three-dimensional voxel data given an alternative method for computing pairwise alignments more efficiently (such as using image features or landmarks).
-%
 
-%Our algorithms will fail for other imaging data sets which contain fine-scale structures,  as small perturbations in these structures will result in large changes in the Euclidean distances.
-%
-%In such cases, one requires a more sophisticated distance metric (such as the Earth Mover's Distance \citep{rubner2000earth}) and/or image description (such as scattering transform coefficients \citep{mallat2012group}) which is insensitive to such perturbations.
-%
-%Unlike traditional statistical methods, our algorithms do not provide confidence intervals on predicted orderings, but these could in principle be extracted from bootstrapping \citep{efron1979bootstrap}.
-%
-%Furthermore, although our methods provide a temporal ordering for images, they provide no information about the actual time, neither do they provide a direction for the proposed ordering.
-%
-%Determining the trajectory rate and direction are post-processing tasks which require {\em a priori} knowledge of the developmental processes or some assumptions about the temporal sampling of the data set (e.g., the data set uniformly samples some given time window).
-
-%
 Vector diffusion maps allow us to automatically register images, an essential task for many applications.
@@
 %
-%Data which comes from several different genetic backgrounds can be clustered according to their spatial expression patterns, and clustering can also be used to separate images into different developmental stages.
-%
 Furthermore, images taken from different viewing directions can be analyzed, as the vector diffusion maps parametrization will organize the images according to the viewing angle \citep{singer2011viewing}.
@@
 Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.
-
-%% == end of paper:
-
-%% Optional Materials and Methods Section
-%% The Materials and Methods section header will be added automatically.
-
-%% Enter any subheads and the Materials and Methods text below.
 
